### 2013-01-22 tex/AtmosphereProblems.tex
added math, tex and slides folders

\documentclass[11pt]{article}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}
\usepackage{enumitem,subeqnarray}

\newcommand{\delad}{{\nabla_{\rm ad}}}
\newcommand{\Rg}{\mathcal{R}}

\newcommand{\vcs}[1]{\mbox{\boldmath{$\scriptstyle{#1}$}}}
\newcommand{\vc}[1]{\mbox{\boldmath{$#1$}}}
\newcommand{\nab}{\vc{\nabla}}
\DeclareMathSymbol{\varOmega}{\mathord}{letters}{"0A}
\DeclareMathSymbol{\varSigma}{\mathord}{letters}{"06}
\DeclareMathSymbol{\varPsi}{\mathord}{letters}{"09}

\newcommand{\Eq}[1]{Equation\,(\ref{#1})}
\newcommand{\Eqs}[2]{Equations (\ref{#1}) and~(\ref{#2})}
\newcommand{\Eqss}[2]{Equations (\ref{#1})--(\ref{#2})}
\newcommand{\App}[1]{Appendix~\ref{#1}}
\newcommand{\Fig}[1]{Fig.~\ref{#1}}
\newcommand{\Figs}[2]{Figs.~\ref{#1} and \ref{#2}}
\newcommand{\Tab}[1]{Table \ref{#1}}

\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\title{Atmosphere Calculations}
%\author{Andrew}
\date{\today}                                           % Activate to display a given date or no date

\begin{document}
\maketitle
%\section{}
%\subsection{}
\begin{enumerate}
\item {\bf Adiabats} The thermodynamic state of a gas is in general determined by two thermodynamic variables.  For instance any two of pressure $P$, density $\rho$, and temperature $T$ is a complete description since the third follows from an equation of state (EOS).  The most common is the ideal gas law
\begin{equation}
P = \rho \mathcal{R} T ~,
\end{equation} 
where the specific gas constant $\mathcal{R} = n R = k_B/\mu$ can also be expressed in terms of the number of moles per unit mass $n$, the universal gas constant $R$, Boltzmann's constant $k_B$ and the mean molecular weight, $\mu$.

If another constraint exists then you only need one thermodynamic variable to determine the state of the gas.  In this case one has a so-called barotropic equation of state $P(\rho)$ where $\rho$ determines $P$ (and vice versa).  The more general EOS (i.e.\ ideal gas) then determines $T$.  Concrete examples of barotropic relations include an isothermal ($T = $ constant) and an adiabatic (constant entropy $S$) atmosphere.  The simplest forms of an adiabat are
\begin{subeqnarray}\label{eq:adiabats}
P &=& K \rho^\gamma \slabel{ada}\\
T &=& K_{TP} ~P^\delad \slabel{adb}
\end{subeqnarray} 
where $K$ and $K_T$ are constants related to the entropy.  The exponents are the adiabatic index, $\gamma$, and the adiabatic gradient $\delad$ (or just ``del-ad").  Note that the real adiabats of non-ideal gases don't follow simple powerlaws like this, as we'll get to below.
\begin{enumerate}
\item \Eqs{ada}{adb} are redundant, i.e.\ equivalent.  Using the ideal gas law, relate $\gamma$ and $\delad$.  (You could also relate $K$ and $K_T$ but that's less significant, so I wouldn't bother.)
\item Show also that
\begin{equation}
T = K_{T\rho} ~\rho^{\gamma - 1}
\end{equation} 
\end{enumerate}
The first law of thermodynamics expresses energy conservation in terms of internal energy, heat and work as
\begin{equation}
dU = dQ - P dV = dQ - P d (1/\rho)
\end{equation}
For an ideal gas the internal energy changes as $dU = C_V dT$ for the specific heat at constant (specific) volume (i.e.\ density), $C_V$.  Combining the first law and the ideal gas EOS you can show (see 10.4 in Carroll and Ostlie) that the specific heat at constant pressure is
\begin{equation}
C_P = C_V + \mathcal{R}
\end{equation} 
\begin{enumerate}[resume]
\item Use the first law with $dQ = T dS$ to show
\begin{subeqnarray} \label{eq:dS}
dS &=& C_V{dT \over T} - \mathcal{R} {d \rho \over \rho} \\
&=& C_P {dT \over T} - \mathcal{R} {d P \over P} \slabel{b}\\
&=& C_V {d P \over P} - C_P {d \rho \over \rho} \slabel{c}
\end{subeqnarray} 
\item For an adiabatic process $dS = 0$.   Derive the adiabatic relations in \Eq{eq:adiabats} by integrating \Eqs{b}{c} with $dS = 0$.  This will determine $\gamma$ and $\delad$ in terms of the specific heats and $\mathcal{R}$, e.g.\
\begin{equation}
\delad = {\Rg \over C_P} = {\gamma - 1 \over \gamma}
\end{equation} 
\item Now integrate the \Eqs{b}{c} to find the relative entropy, e.g.\ 
\begin{equation}
\int dS = \Delta S = C_P \ln T - \mathcal{R} \ln P - const 
\end{equation} 
and the constant on the RHS can be ignored/absorbed into $\Delta S$ (by changing the reference entropy that's unknown anyway).  Arrange the above result to get $\Delta S$ in terms of $K$. Does increasing $K$ give a higher or lower entropy?  Starting with \Eq{c} determine this for $K_T$ also.
\item For an ideal gas $C_V = 3 \mathcal{R} /2$ for a monatomic gas and  $C_V = 5 \mathcal{R} /2$ for a diatomic gas, i.e.\ 3 vs.\ 5 degrees of freedom.  Calculate $\gamma$ and $\delad$ for both cases.
\end{enumerate}
Note that the differential forms of an adiabat, given by \Eq{eq:dS}, are more fundamental than \Eq{eq:adiabats} because they do not require the specific heats and adiabatic indices to be constant.  Since these quantities are not constant for real gases, the distinction is important.

\item {\bf Structure of Plane Parallel Atmospheres} A static atmosphere is in hydrostatic balance
\begin{equation}\label{eq:HB}
{dP \over dr} = - \rho g(r)
\end{equation} 
In general the atmsopheric structure also requires an energy equation to determine $T$.  However for a barotropic equation of state $P(\rho)$ such as an isotherm or an adiabat no energy equation is required.  There are several levels of approximation for computing $g$.  The simplest is the plane-parallel approximation, where $g$ is treated as constant.  This is valid near the surface of planet or in any thin layer.  In this case its common to replace $r$ with $z$ (i.e. $dP/dz = - \rho g$) and take $z = 0$ as the surface. 
\begin{enumerate}
\item  Show that the column density of a plane-parallel atmosphere
\begin{equation}
\varSigma = \int_0^\infty \rho dz = P_o/g
\end{equation} 
where $P_o$ is the pressure at the base ($z = 0$).  To do this change variables from $z$ to $P$ and use $P = 0$ at the top of the atmosphere ($z = \infty$).  This important result is completely independent of the atmosphere's structure (adiabatic, isothermal, or anthing else) or EOS (it need not be an ideal gas!).  However it only applies for plane-parallel atmospheres.
\item Define the scale height as $H_o =  \Rg T_o/g$, where $T_o$ is the temperature at the base of the atmosphere.  For an ideal gas show that $\varSigma = \rho_o H_o$, where $\rho_o$ is the density at the base.  This result is still independent of structure.
\item Calculate the density profile of an {\bf isothermal, plane-parallel}, ideal gas atmosphere at temperature $T$.  Assume that the density is $\rho_o$ at the base of the atmosphere, so the pressure is $P_o = \rho_o \Rg T$.  This calculation will show that $H = \Rg T/g$ (a constant for the assumptions made) actually is the scale height.
\item You could also calculate the structure of an adiabatic atmosphere, maybe it's best to move on to the more relevant (for us) spherical case first...

\end{enumerate}

\item {\bf Structure of Spherical Atmospheres}
For spherical atmospheres $g = G M/r^2$ is no longer constant.  However if the mass of the atmosphere is small, then can be held fixed at the core mass $M = M_c$ so that $g \propto 1/r^2$ exactly.  This is a non-self-gravitating atmosphere.  To include the mass of the atmosphere, hydrostatic balance must be supplemented with the mass equation
\begin{equation}
{d M \over dr} = 4 \pi \rho r^2\, .
\end{equation}

\begin{enumerate}
\item  Calculate the density profile of a non-self-gravitating, isothermal atmosphere, with density $\rho_b$ at the bottom $r = r_b$.  Express in the form
\begin{equation}\label{eq:rhoiso}
\rho(r) = \rho_b \exp\left[{r_b \over H_b}\left({r_b \over r} - 1 \right)\right]\, .
\end{equation} 
What's $H_b$?  Show that \Eq{eq:rhoiso} is consistent with the plane parallel result for $r - r_c \ll 1$.

\item The Bondi radius
\begin{equation}
R_B = {G M \over \Rg T}
\end{equation} 
defines the typical scale at which the atmosphere matches onto the surrounding disk.  To see this express \Eq{eq:rhoiso} as
\begin{equation}\label{eq:rhoisoB}
\rho(r) = \rho_b \exp \left[ R_B \left( {1 \over r} - {1 \over r_b} \right) \right]\, .
\end{equation}

Then associate the disk density with $\rho(r \rightarrow \infty) = \rho_d$ to get
\begin{equation}\label{eq:rhoisod}
\rho(r) = \rho_d \exp[R_B/r]
\end{equation} 
which makes it clear that $R_B$ is indeed the right length scale for matching onto the disk.

\item The mass of the isothermal atmosphere includes all the mass from the bottom ($r_b$) to the Bondi radius $R_B$.   The analytic integral is messy, but for the case $r_b \ll R_B$, a series expansion gives
\begin{equation}
M_{iso} \approx 4 \pi \rho_d {r_b^4 \over R_B} \exp[R_B/r_b]
\end{equation} 
Mathematica is the easiest way to get this.  Instead show that this result is consistent with the estimate based on plane-parallel atmospheres $M_{iso} = \varSigma 4 \pi r_b^2$.  Recall that we're calculating the mass of the atmosphere even though we neglected that mass in computing hydrostatic balance.

\item Now consider adiabatic atmospheres.  Show that hydrostatic balance along an adiabat gives
\begin{equation}
{dT \over dr} = - \delad{G M \over  \Rg r^2}\, .
\end{equation} 
This is valid even for non-constant $\delad$.  
\item Now assume $\delad$ (and again $M$) are constant and integrate from the core, $T(r_c) = T_c$ to get
\begin{equation}\label{eq:Tad}
T = T_c - \delad{G M \over  \Rg} \left({1 \over r_c} - {1 \over r}\right) \, .
\end{equation} 
This gives the radius $r_b = r(T_d)$ at which the adiabat intersects the isotherm.  Show that $r_b$ ranges from $r_c$ (no convective zone) to $R_B/\delad$ (convective beyond the Bondi radius) as $T_c$ ranges from $T_d$ to
\begin{equation}
T_{\rm max} = \delad {G M \over  \Rg r_c}
\end{equation} 
Note that we won't know $T_{\rm max}$ exactly when $\delad$ varies throughout the atmosphere.  In any even $T_{\rm max}$ is only the maximum temperature we are interested in.  Note that the core could be even hotter,  with convection extending well beyond $R_B$ (e.g. Rafikov 2006) but these low density atmospheres aren't relevant to giant planet formation.

\end{enumerate}
\item {\bf Matching} Consider an atmosphere with a core temperature $T(r_c) = T_c$.  We want to find the core density $\rho_c$ (and thus the pressure and adiabat.  To do this match the density at the boundary of the convective and isothermal (at disk temperature $T_c$) layers.  The density atop the convective layer is
\begin{equation}
\rho_b^- = \rho_c\left(T_d \over T_c\right)^{1 \over \gamma - 1}
\end{equation} 
and the density at the base of the isothermal layer is given by \Eq{eq:rhoisod} at the radius determined by the solution of \Eq{eq:Tad}.  Check whether this gives
\begin{equation}
\rho_b^+ = \rho_d\exp(R_B/r_c) \exp\left[ - {1 \over \delad} \left( {T_c \over T_d} - 1\right)\right] \, ,
\end{equation} 
The first exponential factor ($\gg 1$) would give the core density if the atmosphere was completely isothermal.  Provided $T_c > T_d$ the second factor $\ll 1$, as it must.  Note that $\rho_c$ trivially follows from $\rho_b^- = \rho_b^+$. 
\end{enumerate}
\begin{enumerate}[resume]
\item Write a numerical algorithm that iteratively solves for $\rho_c$ for a given $T_c$, $T_d$, $\rho_d$, core mass, etc.  Before trying to code anything, just write it as a flowchart and we'll go over it.  Think of ways that each analytic step can be replaced by the result of a numerical (shooting) calculation.  My thought is that we will want to take the unusual approach of using temperature as the effective vertical coordinate along the adiabat (it's more common to do this with pressure or mass).

\item {\bf Absolute Entropy} The absolute entropy (per mass) of an ideal gas is given by the Sacker-Tetrode formula as
\begin{equation}
S = \Rg \left[- \ln\left( n \Lambda^3\right) + {5 \over 2} \right] = \Rg \left[{5 \over 2} \ln T - \ln P +  \ln \left\{ (2 \pi)^{3/2} \Rg^{5/2} \mu^4 \over h^3 \right\} + {5 \over 2}\right] 
\end{equation} 
Check that the final result follows from the first for the number density $n$ and the deBroglie wavelength $\Lambda = h/\sqrt{2 \pi \mu k_B T}$.  Check that this result reproduces the entropy values at low $T$ and $P$ in the S95 Helium tables, using $\mu_{\rm He} = 6.646442\times 10^{-24}~{\rm g}$.  (Agreement should be to a part in $10^4$.)
\end{enumerate}

\end{document}

### 2013-01-22 tex/CoreAccretion.tex
added math, tex and slides folders

% !TEX TS-program = pdflatexmk

\documentclass[12pt, preprint,numberedappendix]{emulateapj}
%\documentclass[12pt, preprint]{aastex}

\newcommand\submitms{n}		% set to y to follow AAS ``ms'' names, etc.
\newcommand\bibinc{n}		% set to y if bib pasted in .tex, set to n to use bibtex

\usepackage{pdfsync}
\usepackage{subeqnarray}
\usepackage{natbib}

\bibliographystyle{apj}

\newcommand{\ie}{i.e.\ }
\newcommand{\eg}{e.g.\ }
\newcommand{\p}{\partial}
\newcommand{\xv}{\vc{x}}
\newcommand{\kv}{\vc{k}}
\newcommand{\brak}[1]{\langle #1\rangle}

\newcommand{\gcc}{\;\mathrm{g\; cm^{-3}}}
\newcommand{\gsc}{\;\mathrm{g\; cm^{-2}}}
\newcommand{\cm}{\; {\rm cm}}
\newcommand{\mm}{\; {\rm mm}}
%\newcommand{\ps}{\; {\rm s^{-1}}}
\newcommand{\km}{\; {\rm km}}
%\newcommand{\au}{\; \varpi_{\rm AU}}

\newcommand{\AU}{\; {\rm AU}}
\newcommand{\yr}{\; {\rm yr}}
\def\K{\; {\rm K}}

\newcommand{\vcs}[1]{\mbox{\boldmath{$\scriptstyle{#1}$}}}
\newcommand{\vc}[1]{\mbox{\boldmath{$#1$}}}
\newcommand{\nab}{\vc{\nabla}}
\DeclareMathSymbol{\varOmega}{\mathord}{letters}{"0A}
\DeclareMathSymbol{\varSigma}{\mathord}{letters}{"06}
\DeclareMathSymbol{\varPsi}{\mathord}{letters}{"09}

\newcommand{\Eq}[1]{Equation\,(\ref{#1})}
\newcommand{\Eqs}[2]{Equations (\ref{#1}) and~(\ref{#2})}
\newcommand{\Eqss}[2]{Equations (\ref{#1})--(\ref{#2})}
\newcommand{\Eqsss}[3]{Equations (\ref{#1}), (\ref{#2}) and~(\ref{#3})}
\newcommand{\App}[1]{Appendix~\ref{#1}}
\newcommand{\Sec}[1]{Sect.~\ref{#1}}
\newcommand{\Chap}[1]{Chapter~\ref{#1}}
\newcommand{\Fig}[1]{Fig.~\ref{#1}}
\newcommand{\Figs}[2]{Figs.~\ref{#1} and \ref{#2}}
\newcommand{\Figss}[2]{Figs.~\ref{#1}--\ref{#2}} 
\newcommand{\Tab}[1]{Table \ref{#1}}

\newenvironment{packed_item}{
\begin{itemize}
  \setlength{\itemsep}{1pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}
}{\end{itemize}}

\newcommand{\delad}{\nabla_{\rm ad}}
\newcommand{\delrad}{\nabla_{\rm rad}}
\newcommand{\Rg}{\mathcal{R}}
\newcommand{\RB}{R_{\rm B}}
\newcommand{\co}{_{\rm c}}
\newcommand{\di}{_{\rm o}}
\newcommand{\cb}{_{\rm CB}}
\newcommand{\surf}{_M}
\newcommand{\mc}{m_{\rm c \oplus}}
\newcommand{\mcn}[1] { m_{ \rm c #1 \oplus} }
\newcommand{\MC}{M_{\rm crit}}
\newcommand{\au}{a_\oplus}
\newcommand{\aun}[1]{ a_{#1\oplus} }

\begin{document}

\slugcomment{Draft Modified \today}

\shorttitle{Core Accretion}
\shortauthors{Youdin }

\title{Core Accretion of Self-Gravitating Atmospheres}
\author{Andrew N.\ Youdin}
\affil{Harvard Smithsonian Center for Astrophysics}

%\begin{abstract}
%\end{abstract}
\section{Introduction}
The core accretion model \citep{PerCam74, MizNak78} proposes that giant planets form by the accretion of gas onto a solid protoplanetary core.    Early versions of the theory found a ``critical core mass", $\MC$, above which static solutions do not exist and unstable atmospheric collapse would occur \citep{Har78, Miz80}.  These works used hydrostatic models that include heating by planetesimal accretion, but neglect the heat generated by the gravitational, or Kelvin-Helmholtz (KH), contraction of the atmosphere.

In evolutionary models that include KH contraction \citep{BodPol86}, the atmosphere does not undergo a hydrodynamic collapse.  Instead the quasi-static contraction of the envelope accelerates continuously.  Contraction becomes particularly rapid after the the  ``crossover" mass, when core and atmospheric masses are equal \citep{pollack96}.  The concepts of critical and crossover masses are similar, and even agree numerically when the planetesimal accretion luminosity exceeds the KH luminosity.  However if the planetesimal accretion rate is small and KH contraction dominates, the critical core mass will be smaller than the more relevant  crossover mass \citep{IkoNak00}.

This paper will consider 1D atmospheres  where KH contraction dominates over planetesimal accretion.  Such solutions give realistic lower limits on the core mass required to accrete a massive gaseous atmosphere.  Indeed even models that include planetesimal accretion obtain the fastest formation times by stopping planetesimal accretion after the core reaches a certain mass \citep{HubBod05}.

The early phases of atmospheric growth are usually treated with 1D models where the outer boundary of the atmosphere matches onto the conditions in the disk midplane.   As the atmosphere continues to grow, 1D solutions are no longer adequate.  The atmosphere both notices that the disk is not spherically symmetric and at large enough masses will start to open a gap in the disk.  In this case the accretion of gas is limited by the ability of the disk to supply gas to the growing protoplanet \citep{DAnLub08}.  This later stage evolution is crucial for determining the final mass of the planet, but  not considered in our calculations of the initial collapse.

%Analytic models to explain the critical core mass have their shortcomings. \cite{Ste82} assumed a fully radiative atmosphere, which is inconsistent with realistic luminosities and opacities.  \cite{Wuc93} assumed a fully convective atmosphere which is is inconsistent with the ability of atmospheres to cool.  \citet[hereafter R06]{Raf06} considerably advanced the analytic understanding of core accretion models.  (Explain what's great.)  However R06 neglected the luminosity due to gas accretion.

%Most analyses of the critical core mass considers planetesimal accretion as the dominant luminosity source.  One notable exception is \cite{PapNel05}.  There are several reasons to consider the atmospheres on protoplanets that are not accreting solids.  First the planetesimal accretion rate may be a negligible source of luminosity.  (In this sense our results complement those of R06).  Second the growth of an atmosphere around a preexisting core represents lower limit to the timescale to form a giant planet.  Since atmospheric mass grows sharply with core mass, there may be little loss in assuming that atmospheric accretion begins once core formation stops.  Finally while it is desirable to self-consistently solve for core growth and atmospheric accretion, it is impossible to account for all possible core growth processes, which are quite complicated (refs).  
%For instance planet migration or scattering could bring a core that formed rapidly in one region of the disk to a region of the disk where cores could not form rapidly.

Section \ref{sec:structure} describes the structure of planet atmospheres embedded in a gas disk.   We then derive the critical core mass to trigger runaway collapse of the atmosphere in \S\ref{sec:critmass}.

\section{Disk Model}
As a fiducial disk model, we adopt the version of the minimum mass, passively irradiated model of \citet{cy10}, which gives the surface density and midplane temperature as
\begin{subeqnarray}
\label{eq_sigmag}
\varSigma  &=& 70 \,F_\varSigma \aun{10}^{-3/2} ~{\rm g~cm}^{-2} \\
T &=& 45  \,F_T\, \aun{10}^{-3/7} ~{\rm K}
\end{subeqnarray} 
where $\aun{10} \equiv a/{10~\rm AU}$ scales the disk radius and the normalization factors $F_\varSigma$ and $F_T$ adjust the disk mass and temperature relative to the canonical values.   The resulting disk midplane pressure 
\begin{equation}
P\di = 6.9 \times 10^{-9} F_\varSigma \sqrt{F_T m_\ast} \, \aun{10}^{-45/14}~{\rm bar}
\end{equation} 
for a mean molecular weight of 2.35.  The stellar mass in solar units is $m_\ast \equiv M_\ast/M_\odot$, which only includes the gravitational effects, though stellar mass also influences temperature and probably correlates with average disk mass.  Away from the midplane the pressure drops with height $z$ as a Gaussian with scale-height
\begin{equation}
H = {\sqrt{\Rg T\di} \over \varOmega} = 0.42 \sqrt{F_T \over m_\ast}  \, \aun{10}^{9/7} \AU\, .
\end{equation}

\section{Atmospheric Structure}\label{sec:structure}
In this section we review the structure of atmospheres surrounding a solid protoplanetary core embedded in a gas disk.  The core has mass $M\co$ and radius $R\co = [3 M\co/(4 \pi \rho\co)]^{1/3}$ with internal density $\rho\co \approx 2~{\rm g~cm^{-3}}$.  For protoplanetary disk parameters we use the minimum mass, passively irradiated disk model presented in \citet[hereafter  CY10]{cy10}.  The disk has a midplane gas density $\rho\di$ and temperature $T\di$.

The core can attract a dense atmosphere inside its Bondi radius
\begin{equation} 
R_B = {G M\co \over \mathcal{R} T\di} \approx 0.17 {\mcn{10}  \, \aun{10}^{3/7} \over F_T} ~\AU
\end{equation} 
where $\mcn{10} \equiv M\co/(10~M_\oplus)$ and $\mathcal{R}$ is the specific gas constant.

Low mass planets satisfy a hierarchy of length scales,  $R\co \ll R_B < R_{\rm H} < H$, where %$h\di \approx 3.3 \times 10^6 \au^{9/7}~{\rm km}$ is the gas scale-height and
\begin{equation} 
R_{\rm H} = \left(M\co \over 3 M_\ast \right)^{1/3}a \approx 0.22 {\mcn{10}^{1/3} \, \aun{10} \over m_\ast^{1/3}}~\AU
\end{equation} 
is the Hill radius, inside which the planet's gravity competes with Keplerian shear.

As protoplanets grow in mass the length scales $R_{\rm H} \sim R_{\rm B} \sim H$ are roughly equal at a characteristic ``transitional" mass
\begin{equation}
M_{\rm tr} > {(\Rg T\di)^{3/2} \over G \varOmega} \approx 25 \, {F_T^{3/2} \over \sqrt{m_\ast} } \, \aun{10}^{6/7}~ M_\oplus \, .
\end{equation} 
Technically the length scales re-order in three steps of increasing mass.  First for $M > M_{\rm tr}/\sqrt{3}$ the ordering becomes $R_{\rm H} < R_{\rm B} < H$ as tidal truncation becomes more important than thermal constriants.  While the shear flow past the atmosphere is clearly not spherically symmetric, it is still common take the atmosphere as spherically symmetric with an outer boundary at the Hill radius.  When the mass increases further to $M > M_{\rm tr}$ then  $R_{\rm H} < H < R_{\rm B}$, which does not change the fact that tidal truncation is the stronger effect at the outer boundary.

Finally when $M > 3 M_{\rm tr}$ the high mass scaling $H < R_{\rm H} < R_{\rm B}$ is attained.  The assumption of spherical symmetry fails for two reasons.  First the planet now notices the vertical density stratification of the disk.  Second, the planet's torques on the disk can begin to open gaps, creating further density inhomogeneity.  Since runaway accretion typically begins for masses well below $\sim 75 M_\oplus$, the assumption of spherical symmetry is a good, if not perfect, approximation for understanding whether cores can attrach massive atmospheres.  Moreover the approximation becomes even better in the disk's outer regions.

Another relevant for atmospheric structure is the mass of disk gas that fills the Bondi radius, at the ambient density.  The ratio of this mass to the core mass is 
\begin{equation}
\theta\co \equiv { 4 \pi \rho\di R_B^3 \over 3 M\co} \approx 5 \times 10^{-3} {\mcn{10}^2 \over \aun{10}^{3/2}}{F_\varSigma \sqrt{m_\ast} \over F_T^{7/2}}\, .
\end{equation} 
The smaller this parameter the more gravitational compression is required for an atmosphere to become self-gravitating.

There are two limiting cases of atmospheric structure, isothermal and adiabatic.  Atmospheres that are isothermal at $T\di$ have cooled completely and give the most massive atmosphere.   Convectively unstable atmospheres transport heat very efficiently and are nearly adiabatic.  Adiabatic atmospheres are the opposite extreme of the isothermal case, giving the steepest increase of temperature with atmospheric dept.  The least dense atmospheres are fully adiabatic with the entropy of the disk.

The structure of an embedded protoplanet consists of an inner adiabatic region that matches onto an outer radiative zone, that in some cases, can be approximated as nearly isothermal.  This basic structure was demonstrated by R06 for an atmosphere supported by planetesimal accretion luminosity.  PN05 showed that atmospheres powered by gas accretion alone have a similar structure.  Due to their intense irradiation, ``hot Jupiters" similarly have an isothermal layer that matches onto a deeper adiabatic interior \citep{ab06, ym10}.

\subsection{Isothermal Atmosphere}
We now consider the structure of a low mass, i.e.\ non-self gravitating, isothermal atmosphere.  We assume the atmosphere matches onto a constant background density, $\rho\di$, at a distance $r_{\rm fit} = n_{\rm fit} \RB$.  The resulting density profile is
\begin{equation} \label{eq:rhoiso}
\rho = \rho\di \exp \left({R_B \over r} - {1 \over n_B} \right) \approx   \rho\di \exp \left(R_B \over r  \right)
\end{equation} 
where the approximate inequality holds deep inside the atmosphere ($r \ll \RB$) for any $n_{\rm fit} \gtrsim 1$.  However the choice of boundary condition does have an order unity effect on the density near the Bondi radius.

The mass of the atmosphere typically determined by integrating the density profile from the core to the Bondi radius.  Planets can attract massive atmospheres if $\theta\co \equiv R_B/R\co \gg 1$.  In this case
\begin{equation} \label{eq:MatmISO}
M_{\rm iso} \approx 4 \pi \rho\di {R\co^4 \over R_B} e^{R_B/R\co} = 4 \pi \rho\di {R\co^3 \over \theta\co} e^{\theta\co}\, .
\end{equation} 
This result is the leading order term in a series expansion.  Furthermore, because the atmospheric scale-height at $R\co$ is $H_\rho = |dr /d\ln\rho| = R\co^2/R_B$, the result is intuitively the correct order of magnitude.

From \Eq{eq:MatmISO} the ratio of isothermal atmosphere to core mass is 
\begin{equation} 
{M_{\rm iso} \over M\co} = 3 {\rho\di \over \rho\co}{R\co \over R_B} e^{R_B/R\co} %= {\theta_o \over \theta\co^2} e^{\theta\co}
\end{equation} 
While  $\rho\di \ll \rho\co$, the exponential dependence allows the atmosphere to become massive.  
%A value of $\theta\co = R_B / R\co$ corresponds to a planet mass of
%\begin{equation} 
%M\co = \sqrt{3 \over 4 \pi \rho\co} \left(\theta\co \mathcal{R} T \over G \right)^{3/2} 
%\end{equation} 
As a concrete example, consider a disk with $T = 60$ K and $\rho\di = 3 \times 10^{-11}$ (in cgs) and a core density $\rho\co = 2$.  The atmosphere self-gravitating for $\theta\co = 27.1$ or $M\co > 0.046 M_\oplus$.  The fact that very low mass cores can have very massive isothermal atmospheres is well known \citep{Sas89}.  However such solutions are unrealistic because there is inadequate time to radiate away all the atmosphere's thermal energy.

We not consider the mass exterior to the Bondi radius.  For a meaningful evaluation we only include the mass coming from the overdensity relative to the background density.  The resulting external mass for an isothermal atmosphere is
\begin{subeqnarray}
M_{\rm ext} &=& 4 \pi \int_{\RB}^{r_{\rm fit}} (\rho - \rho\di) r^2 dr \\
&=& M\co \theta\co \int_1^{n_{\rm fit}} 3 \left[ \exp \left({1 \over x} - {1 \over n_{\rm fit}}\right) - 1 \right] x^2 dx  \nonumber \\
&\equiv& M\co \theta\co I(n_{\rm fit})
\end{subeqnarray} 
where the dimensionless integral $I(n_{\rm fit})$ obeys the limits $I(1) = 0$ and $I \rightarrow n_{\rm fit}^2/2$ as $n_{\rm fit} \rightarrow \infty$.  Since this external mass does not converge the choice of an outer boundary does matter in principle.  In practice however the assumption that   $r_{\rm fit} = R_{\rm H}$ limits $n_{\rm fit}$ to modest values
\begin{equation}
n_{\rm fit} = {R_{\rm H} \over \RB} \approx 1.3 {\aun{10}^{4/7} \over \mcn{10}^{2/3}}{F_T \over  m_\ast^{1/3}}
\end{equation} 
since for instance $I(2) = 1.1$ these modest $n_{\rm fit}$ values will only produce a small external mass.

However the effect is still worth including for the effect on the interior mass.  
\subsection{Temperature Contrast at Convective Boundary}
We now show that the temperature contrast between the convective boundary, $T\cb$, and the ambient disk $T\di$, is modest.  We express the radiative lapse rate
\begin{equation}\label{eq:delrad}
\delrad \equiv {d \ln T \over d \ln P} = {3 \kappa P \over 64 \pi  G M \sigma T^4} L = \nabla\di {(P/P\di)^{1+\alpha} \over (T/T\di)^{4-\beta}}
\end{equation}
where $M$ is the sum of the core mass and the atmosphere mass below the pressure level $P$.  If the mass in the radiative zone is small, then we can hold $M = M\co + M_{\rm conv}$ fixed at the sum of core and convective zone masses.  With this assumption and powerlaw opacity, we get a constant value for $\nabla\di$.  The temperature profile then integrates to
\begin{equation}\label{eq:radTP}
\left(T \over T\di\right)^{4-\beta} - 1 = {\nabla\di \over \nabla_\infty} \left[\left({P \over P\di}\right)^{1-\alpha} - 1 \right] \, ,
\end{equation} 
where $\nabla_\infty = (1+\alpha)/(4-\beta)$ is $\delrad$ for $T ,P \rightarrow \infty$.\footnote{Our definitions of $\nabla\di$ and $\nabla_\infty$ are precisely opposite to R06, but consistent with other works and the general convention of labeling a quantity $f$ in the disk as $f\di$.}
We  apply \Eqs{eq:delrad}{eq:radTP} at the convective boundary $\delrad = \delad$ under the assumption that the pressure there is $P\cb \gg P\di$. The resulting temperature contrast at the convective boundary is
\begin{equation}\label{eq:Tcb}
\chi \equiv {T\cb\over T\di} \simeq \left(1 - {\nabla_{\rm ad} \over \nabla_\infty}\right)^{-\frac{1}{4-\beta}} \, .
\end{equation} 
For  the low temperatures in protoplanetary disks, opacity is dominated by dust with $\alpha = 0$ and $\beta \approx 2$ or $1$, giving $\chi \approx 1.5$ or 1.9, respectively, assuming  $\delad = 2/7$.

\subsection{Location of Convective Boundary}
The pressure at the convective boundary  follows from \Eqs{eq:delrad}{eq:Tcb} as
\begin{equation}
{P\cb\over P\di} \simeq \left({\delad/\nabla\di \over 1 - \delad/\nabla_\infty}\right)^{1 \over 1 + \alpha}
\end{equation} 
This pressure contrast can be quite large due to the smallness of $\nabla\di$ in low luminosity atmospheres.

 It is also useful to obtain a relation between $T$ and $P$ that eliminates $\nabla\di$ in favor of $P\cb$:
 \begin{equation}\label{eq:TP}
{T \over T\di} = \left\{1 + {1 \over {\nabla_\infty \over \delad} - 1} \left[ \left({P \over P\cb}\right)^{1-\alpha} -  \left({P\di \over P\cb}\right)^{1-\alpha}  \right] \right\}^{1 \over 4-\beta}\, .
\end{equation}

We can determine the radius of the convective boundary $r\cb$ from the hydrostatic balance equation as 
\begin{equation}\label{eq:RCBint}
{R_B \over r\cb} = \int_{P\di}^{P\cb} {T \over T\di} {dP \over P}\, .
\end{equation} 
An isothermal atmosphere gives a simple logarithmic dependence on $P\cb$.  However using \Eq{eq:TP} in the integral gives
\begin{equation}\label{eq:Rcb}
{R_B \over r\cb} = \ln \left(P\cb \over P\di \right) - \ln \theta \, ,
\end{equation} 
with an extra correction term, $\theta < 1$.  In the $P\cb \gg P\di$ limit the correction term is a constant that depends on $\alpha$, $\beta$ and $\delad$.  The form of $\theta$ was chosen to allow us to express
\begin{equation}\label{eq:PcbRcb}
P\cb = \theta P\di e^{R_B/r\cb} \, .
\end{equation}   
A simple analytic expression for $\theta$ is not possible.

\begin{deluxetable}{cccccc}  % <--- column justification (center/left/right)
\gdef \numcols {6}
\tablecolumns{\numcols}
\tablecaption{Parameters Describing Structure of Radiative Zone.}
\tablehead{   \multicolumn{\numcols}{c} {$\gamma = 7/5$ ($\delad = 2/7$), $\alpha = 0$} }  
\startdata
 $\beta$   		 &1/2  	& 3/4 &1   		& 3/2  		& 2   \\
 $\nabla_\infty$ & 2/7 \tablenotemark{a}  	&  4/13	& 1/3 	& 2/5 	 	& 1/2 \\
 $\chi$ 		 & \nodata &  2.25245 &1.91293 	& 1.65054 	& 1.52753 \\
 $\theta $  		 &\nodata   & 0.145032	&0.285824   &0.456333   & 0.556069   \\
\enddata
\tablenotetext{a}{Since $\delad = \nabla_\infty$ there is no convective transition at depth for this case.}
\end{deluxetable}

The T-P profile can also be combined with the equation of hydrostatic balance to relate the radius and pressure of the convective boundary.  If the convective boundary is well inside the Bondi radius, i.e.\ $r\cb\ll R_B$ and $P\cb\gg P\di$ then
\begin{equation}\label{eq:PcbRcb}
P\cb = \theta P\di e^{R_B/r\cb} \, .
\end{equation}    
The order unity constant $\theta < 1$ accounts for the fact that the radiative layer is not perfectly isothermal, as $\theta$ would be exactly one in that case.  The value of theta depends on the equation of state and the pressure and temperature dependence of the opacity, i.e. the $\alpha (=0)$ and $\beta$ values.  We find $\theta(\beta = 2) \approx 0.556$ and $\theta(\beta = 1) = 0.286$.  Closed form expressions for these integrals are too cumbersome to be useful.

\section{Virial Theorem}
The virial theorem for self-gravitating atmospheres can be applied to protoplanetary atmospheres.    The virial theorem is derived as usual by integrating the equation of hydrostatic balance to give
\begin{eqnarray}\label{eq:virial}
E_G= -3 \int_{M\co}^{M} {P \over \rho}dm + 4 \pi \left( R^3 P\surf - r\co^3 P\co\right) 
\end{eqnarray} 
where the gravitational energy
\begin{equation}\label{eq:EG}
E_G = -\int_{M\co}^{M} {G m \over r} dm \, .
\end{equation} 
At the outer surface $m = M$, $r = R$ and other values are given a subscripted $M$.  This outer surface can be evaluated anywhere (e.g.\ Bondi radius, Hill radius or RCB) that hydrostatic balance holds.  The boundary terms include the effects of finite surface pressure and finite radius of the solid core.

 The integral on the RHS \Eq{eq:virial} is related to the internal energy.  For an ideal gas, $P/\rho = \Rg T = (\gamma -1) u$, where $u$ is the internal energy per mass, and $\gamma = C_P/C_V$.  For a polytrope with constant $\gamma$ we can express the virial theorem in terms of internal energy $E_i = \int u dm$ as
\begin{equation}
E_G = - \zeta E_i + 4 \pi \left(R^3 P\surf - r\co^3 P\co\right) 
\end{equation} 
where $\zeta \equiv 3 (\gamma -1)$.  In this case the total energy $W = E_G + E_i$ becomes
\begin{subeqnarray}
W &=& (1-\zeta)E_i + 4 \pi \left( R^3 P\surf  - r\co^3 P\co \right) \\
 &=& {\zeta - 1 \over \zeta} E_G + {4 \pi \over \zeta} \left( R^3 P\surf  - r\co^3 P\co \right)\, .
\end{subeqnarray} 
The surface pressure acts to unbind the atmosphere by increasing $W$.  While a realistic EOS is neither ideal nor polytropic, \Eq{eq:virial} is general.

%\section{Adiabatic Atmosphere}
%An adiabatic atmosphere with the entropy of the disk gas is only slightly more massive than the disk gas itself (inside the Bondi radius).

\section{Cooling Models}
An isolated sphere satisfies a simple global energy equation
\begin{equation}\label{eq:simplecool}
L = \Gamma - \dot{E}
\end{equation} 
where the luminosity is balanced by the rate of heat generation, $\Gamma$, and the rate at which total (gravitational plus thermal, at least) energy is lost.

For a protoplanetary atmosphere embedded in a gas disk, the cooling equation is more complicated: 
\begin{equation}\label{eq:EB}
L = L\co + \Gamma- \dot{E} + e_{\rm acc} \dot{M}  -  P\surf {\p V\surf \over \p t} \, .
\end{equation}
We explain these extra terms and then derive the result.   \Eq{eq:EB} assumes a well (but arbitrarily) defined surface at radius $R$ and mass $M$, which can evolve in time.  Examples include the Bondi radius or the radiative-convective boundary.  The luminosity $L$ represents the luminosity from that surface.  The luminosity from the solid core is given by $L\co$, and includes planetesimal accretion energy  and radioactive decay.  The core is assumed to have a fixed mass, $M\co$, and radius, $R\co$, to simplify the atmospheric calculation.  The $\Gamma$ and $\dot{E}$ integrals are identical to \Eq{eq:simplecool}, and include the atmosphere from the core to the top boundary.   Sources of direct heating that  contribute to $\Gamma$ include  atmospheric drag on sedimenting planetesimals and the dissipation of any atmospheric turbulence.

Mass accretion at the rate $\dot{M}$ brings specific energy $e_{\rm acc} = u - GM/R$, the sum of internal and gravitational energies.  The energy of accreted matter, $e_{\rm acc}$, is  zero at a modified Bondi radius, $R_{\rm out} = G M/u \approx  (\Rg/C_V) R_B$.  For $R < R_{\rm out}$ ($R > R_{\rm out}$, respectively) the accreted mass has negative (positive) energy and is (un)bound.
 The final term in \Eq{eq:EB} gives the pressure work done on a surface mass element, so the partial time derivative of  volume holds the mass fixed.  Our quasistatic models we neglect bulk kinetic energy.

\subsection{Derivation of Global Energy Equation}
We now derive the global energy equation (\ref{eq:EB}).  We follow the simpler example in \S4.3 of \citet{KipWei94}, adding the effects of finite core radius, surface pressure and mass accretion.  We assume that hydrostatic balance holds.  Integrating the local energy equation from core to surface gives:
\begin{subeqnarray}
L - L\co &=& \int_{M\co}^M {\p L \over \p m} dm \\
&=& \int_{M\co}^M \left(\epsilon - T {\p S \over \p t} \right)dm \\
&=& \Gamma  - \int_{M\co}^M{\p u \over \p t} dm +  \int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm\slabel{eq:DLc}\, .
\end{subeqnarray} 
with $\Gamma = \int \epsilon dm$ the integral of the direct heating rate.

In what follows, we must carefully distinguish between partial time derivatives, $\p / \p t$, (performed at fixed mass) and total time derivatives, denoted with overdots (which include the effect of mass accreted through the outer boundary). % or $d/dt$.  
For instance the evolution of surface radius, $R$, evolves as  
\begin{equation}\label{eq:Rdot}
 \dot{R} = {\p R \over \p t} + {\dot{M} \over 4 \pi R^2 \rho\surf}
\end{equation} 
where $\p R/\p t$ gives the Lagrangian contraction of surface mass elements, and $\dot{M}$ denotes mass accretion rate through the upper boundary.  The subscript $M$  denotes quantities at the upper boundary of total mass $M$ (though it is omitted from $M$ and $R$).  Similarly the volume, $V = (4 \pi/3)r^3$ and pressure at the surface evolve as
\begin{subeqnarray}\label{eq:dot}
\dot{V}_M &=&  {\p V_{\rm M} \over \p t} + {\dot{M} \over \rho_{\rm M}}  \\
 \dot{P}_M &=& {\p P_{\rm M} \over \p t} + {\p P_M \over \p m}\dot{M} \\
 &=&  {\p P_{\rm M} \over \p t} - {G M  \over 4 \pi R^4} \dot{M}\, .
\end{subeqnarray} 
For the purpose of this derivation we will hold the core mass fixed $\dot{M}\co = 0$ which further gives $\dot{P}\co = \p P\co / \p t$.

To derive the global energy equation we must move the (partial) time derivatives in \Eq{eq:DLc} outside their integrals.  The internal energy integral follows simply from  Leibniz's rule as
\begin{equation}\label{eq:udot}
\int_{M\co}^{M(t)}{\p u \over \p t} dm = \dot{E}_i  -  \dot{M}u\surf\, .
\end{equation}

To evaluate the work integral, we derive a pair of expressions for the rate of change of gravitational energy.
The time derivative of \Eq{eq:virial}  gives
\begin{eqnarray}\label{eq:EGdot}
\dot{E}_G &=& 3  \int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm -3 \int_{M\co}^M {\p P\over \p t}{dm \over \rho} \\
&& -  3{P\surf \over \rho\surf} \dot{M}+ 3 \dot{P}\surf V\surf -3 \dot{P}\co V\co  + 3  P\surf {\dot{ V}\surf} \, . \nonumber
%&&+ 4 \pi \left. \left(r^3 {\p P \over \p t}\right)\right|_{M\co}^M  + 3  P\di {\p V \over \p t}  - {G M \over R} \dot{M} \nonumber
\end{eqnarray} 
%where the volumes, $V\surf = 4 \pi R^3/3$ and $V\co = 4 \pi R\co^3/3$. 
The first integral in \Eq{eq:EGdot} is the one we want, but the next one must be eliminated.  The time derivative of \Eq{eq:EG} (times four) gives
\begin{subeqnarray}
 4 \dot{E}_G &=&  -4 {G M \dot{M} \over R} + 4 \int_{M\co}^M {G m \over r^2}{\p r \over \p t} dm\\ 
&=&   -4 {G M \dot{M} \over R} + 4 \pi \int_{M\co}^M r^3{\p \over \p m}{\p P \over \p t} dm \slabel{eq:4EGb} \\
&=&  -4 {G M \dot{M} \over R} -3  \int_{M\co}^M {\p P\over \p t}{dm \over \rho} \slabel{eq:4EGc} \\
&&+ 3 V\surf {\p P\surf \over \p t} -3 V\co {\p P\co \over \p t} \nonumber 
\end{subeqnarray} 
where \Eqs{eq:4EGb}{eq:4EGc} use hydrostatic balance  and integration by parts.

%To eliminate the time derivates of pressure, we take the time derivative of the hydrostatic balance equation for $\p^2 P / \p m\p t$ and integrate over $4\pi r^3 dm$ (as in the virial equation derivation) to get
%\begin{equation}\label{eq:dHBdt}
%3 \dot{P}\surf V\surf -3 \dot{P}\co V\co -3 \int_{M\co}^M {\p P\over \p t}{dm \over \rho}  = 4 \dot{E}_G + 4{G M \over R} \dot{M}  \, .
%\end{equation} 
%Combining \Eqs{eq:EGdot}{eq:dHBdt} gives 
%\begin{eqnarray}\label{eq:rhodot}
%\int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm  &=& - \dot{E}_G - {4 \over 3}{G M\over R} \dot{M} + {P\surf \over \rho\surf} \dot{M} -  P\surf \dot{V}\surf  \, , \nonumber \\
%&=&- \dot{E}_G - {4 \over 3}{G M\over R} \dot{M}  -  P\surf {\p V\surf \over \p t}  \, ,
%\end{eqnarray} 
%where the final step uses \Eq{eq:Rdot}.

Subtracting \Eqs{eq:udot}{eq:4EGc} and rearranging for the desired integral gives
\begin{eqnarray}\label{eq:PdVint}
\int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm  &=&  - \dot{E}_G - {G M \dot{M} \over R} - P\surf {\p \dot{V}\surf \over \p t} \,  
\end{eqnarray} 
with \Eq{eq:dot} used to combine terms.  Combining \Eqsss{eq:DLc}{eq:udot}{eq:PdVint}, we reproduce \Eq{eq:EB} with the accreted energy $e_{\rm acc} \equiv u\surf - GM/R$.

\section{Simplified cooling model}
We consider an analytic cooling model that assumes an ideal gas polytrope for the convective region and neglects self-gravity.

\subsection{Luminosity,  Energy \& Mass}
The luminosity that emerges at the radiative convective boundary is
\begin{equation} \label{eq:Lcb}
L\cb = {64 \pi G M\cb \sigma T\cb^4 \over 3 \kappa P\cb } \nabla_{\rm ad} \approx L\di {P\di \over P\cb} % \left({P\di \over P\cb}\right)%^{1+\alpha}
\end{equation} 
for a dust opacity with $\kappa \propto T^\beta$ and
\begin{equation} 
L\di \equiv {64 \pi G M\cb \sigma T\di^4 \over 3 \kappa(T\di) P\di} \nabla_{\rm ad}\chi^{4-\beta}\, .
\end{equation} 
normalized to disk conditions, with $T = \chi T\di$ at the RCB.

We adopt an opacity law
\begin{equation}
\kappa = 2 F_\kappa \left(T \over 100~{\rm K}\right)^\beta ~{\rm cm^2/g}
\end{equation} 
independent of pressure as appropriate for dust opacities.  For $\beta = 2$ and $F_\kappa = 1$, this gives the \citet{bl94} opacity for icy grains.  By varying $F_\kappa$, dust depletion or enhancement can be considered.  Grain properties affect both $F_\kappa$ and $\beta$ which generally satisfies  $1/2 \lesssim \beta \lesssim 2$ (aside from discontinuities across sublimation regions), see \citet{SemHen03}.

To make further analytic progress, we ignore the self-gravity in the convective zone, holding the mass fixed at $M\co$.  The density profile of an adiabatic atmosphere follows from hydrostatic balance as 
\begin{equation}\label{eq:rhoconv} 
\rho = \rho\cb \left[ 1 + {R_B' \over r} - {R_B' \over r\cb}  \right]^{1/(\gamma -1)}\, .
\end{equation} 
where we define an effective Bondi radius,
\begin{equation}
R_B' \equiv {G M\co \over C_P T\cb} = {\delad \over \chi} R_B
\end{equation} 
to simplify expressions.

  Deep in the atmosphere, where $r \ll r\cb \lesssim R_b'$ the density profile is $\rho \propto r^{-1/(\gamma -1)}$.  Since energy scale as $\rho r^2 \propto r^{(2\gamma -3)/(\gamma - 1)}$ only polytropes with $\gamma < 3/2$ (i.e. $\gamma = 7/5$, but not $\gamma = 5/3$) have the bulk of energy at the bottom of the atmospheres.  We will thus focus on the $\gamma = 7/5$ case, even though dissociation occurs deep in real protoplanetary atmospheres.

The total (thermal and gravitational) energy in an adiabatic atmosphere could be evaluated from the virial theorem.  More simply, we use the result for temperature profiles in deep (but non-self-gravitating) convective regions:
\begin{equation}
T \approx {G M\co \over C_P r} = T\cb {R_B' \over r}\, .
\end{equation} 
The internal energy per unit mass of an ideal gas is thus $u = C_V T = (1 - \delad) G M\co/r$ and the specific energy  deep in the atmosphere is
\begin{equation}
e = e_g + u = -\delad {GM\co \over r} \, .
\end{equation} 
Thus the total energy for $\gamma < 3/2$ is thus
\begin{eqnarray} 
E &=& - 4 \pi \nabla_{\rm ad} G M\co \int_{R\co}^{r\cb} \rho r dr \\
&\approx& - 4 \pi P\cb {R_B'}^{1 \over \nabla_{\rm ad}} \left(\gamma-1 \over 3 - 2 \gamma\right)  R\co^{2\gamma-3\over \gamma-1}  \\
&\approx& - 8 \pi P\cb {R_B'^{7/2} \over \sqrt{R\co}}
%&\approx& -0.31 P\cb {(R_B / \chi)^{7/2} \over \sqrt{R\co}}
%&\approx& -0.31 P\cb {R_B'^{7/2} \over \sqrt{R\co}}
\end{eqnarray} 
where the final expression takes $\gamma = 7/5$. % and then revert to the standard Bondi radius $R_B = G M\co/(\Rg T\di)$ in terms of the disk temperature.

The mass of the adiabatic atmosphere is given by
\begin{eqnarray} 
M_{\rm atm} &=& 4 \pi \int_{R\co}^{r\cb} \rho r^2 dr \\
&=& {5 \pi^2 \over 4} \rho\cb {R_B'}^{5/2} \sqrt{r\cb} % \approx 0.54 \rho\cb {R_B'}^{5/2} \sqrt{r\cb} \\
%&=& 0.54{R_B'^{3} \sqrt{\chi} \over  \sqrt{ \ln[P\cb/(\theta P\di)]}} \, .
\end{eqnarray}
in the limit $R\co \ll r\cb \ll R_B'$. The mass in the isothermal region is $M_{\rm iso} \sim 4 \pi \rho\cb r\cb^4/R_B \ll M_{\rm atm}$, and can be neglected.

We can eliminate $r\cb$ with \Eq{eq:PcbRcb}.  The ratio of atmosphere to core mass is then  
\begin{equation} \label{eq:crit}
{M_{\rm atm} \over M\co} \approx {P\cb /P_M \over  \sqrt{\ln[ P\cb/(\theta P\di)]}}
\end{equation} 
where we introduce a characteristic pressure
\begin{equation} 
P_M \equiv {4 \delad^{3/2} \over 5 \pi^2 \sqrt{\chi} } {G M\co^2 \over {R_B'}^4}\, .
%P_M \approx {1.85 \over \sqrt{\chi}} {G M\co^2 \over {R_B'}^4}\, .
\end{equation}

For the atmosphere to become self-gravitating, with $M_{\rm atm} = M\co$, we thus require
\begin{equation} \label{eq:Pcbc}
P\cb = \xi P_M
\end{equation} 
where  the logarithmic factor
\begin{equation}\label{eq:xi}
\xi \equiv \sqrt{\ln[ P\cb/(\theta P\di)]} = \sqrt{\ln[ \xi P_M / (\theta P\di)]}
\end{equation} 
is found by numerically solving the above transcendental equation. The physical solution has $\xi >1$, but typically order unity. Numerical solutions exit for $\xi \ll 1$, but are unphysical as they imply $P\cb < P\di$.
%  The physical origin of this logarithm is the exponential increase of pressure with depth in the radiative zone.

\subsection{Cooling Time}
Neglecting surface terms, the total time to cool the atmosphere from an initially fully adiabatic state is
\begin{eqnarray} 
t_{\rm  cool} &=& -\int {d E \over L} = - \int_{P\di}^{P\cb } {d E/d P\cb \over L} dP\cb \\
&\approx& 4 \pi {P\cb^{2} \over P\di} {R_B'^{7/2} \over L\di \sqrt{R\co}} \label{eq:tcool}\, ,
%&\approx& {0.16 }{P\cb^{2} \over P\di} {R_B'^{7/2} \over L\di \sqrt{R\co}}
\end{eqnarray} 
where we neglect self-gravity and take $M\cb = M\co$ in $L\di$.

The cooling timescale  for an atmosphere to become self-gravitating is found from \Eqs{eq:tcool}{eq:Pcbc} as
\begin{eqnarray} 
t_{\rm cool} &\approx & 2 \times 10^8 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2  \over \left(\mc \over 10 \right)^{5/3} \left(\au \over 10\right)^{15 \over 14}} \yr
\end{eqnarray} 
for $\beta = 2$.  
Clearly this timescale is too long, and this is likely due to missing physics of self-gravity and EOS.  Nevertheless the trends are informative.   A reduction in opacity naturally gives faster cooling, as long as the optically thick assumption holds.  Lower disk temperatures also give faster cooling.  Even though higher temperatures give higher luminosities, they also give higher dust opacities and lower the Bondi radius and gas density.  In balance higher temperatures suppress cooling in this model.    For different $\beta$ values, the temperature dependence is $t_{\rm cool} \propto F_T^{1/2 + \beta}$.  The cooling timescale only weakly depends on disk mass or pressure, via the  logarithmic factor $\xi$.  This weak dependence is a consequence of cooling coming from the convective boundary.  Note that the scaling laws do not reflect the changes to $\xi$ which remains order unity and is here given the appropriate value for the nominal parameters.

\subsection{Critical Core Mass}\label{sec:critmass}
%We use the results of the previous section to derive the minimum core mass that can form a giant planet by gas accretion.  Since our structure calculations do not include the self-gravity of the atmosphere, we simply assume that runaway growth occurs once the atmosphere's mass (as calculated in the absence of self-gravity) reaches the core mass, $M_{\rm atm} \ge M_{\rm c}$.  This approximation was used by R06 and is consistent with detailed evolution models (refs).

We can define a critical more mass as that which gives a  self-gravitating atmospheres within a typical gas disk lifetime:
\begin{equation}
t_{\rm cool} = 3 \times 10^6 \tau_{\rm cool} ~{\rm yr} \, ,
\end{equation} 
with $\tau_{\rm cool}$ a scaling factor.  The resulting critical core mass is
\begin{equation}\label{eq:Mcrit}
M_{\rm crit} \approx 100 {F_T^{3/2} F_\kappa^{3/5}   \left(\xi \over 2.6 \right)^{6/5} \over \left(\au \over 10\right)^{9 \over 14}} \; M_\oplus
\end{equation} 
as with the cooling time, the critical mass is too large due to missing physics.  The scaling with disk properties is similar, in fact all quantities are simply raised to the $3/5$ power.  The $\xi$ value is changed to match the value for the nominal solution.  The lower value reflects the fact that an atmosphere around such a massive core does not need to contract as much to be self-gravitating.

The numerical values for a different opacity law do not change the numerical answers above very much.  That is mostly because we have chosen a disk location where $T_o$ is not far from 100 K, where our opacity law is normalized.  More generally a higher $\beta$ value is more favorable for core accretion at large distances due to the sharper drop in opacity.

%We use \Eq{eq:tcool} to find $P\cb$, the pressure depth of the convective boundary after a cooling time.   Relative to the disk pressure we find
%\begin{equation}
%{P\cb\over P\di} =  2.0 \times 10^4 \left(a \over 5~{\rm AU}\right)^{2.0} \mc^{-7/6} {\sqrt{\tau_{\rm cool}} \over F  \sqrt{\kappa_\odot}}
%\end{equation}

%This seems totall wrong:
%\begin{equation} 
%{M\co \over M_\oplus} \approx 14 \left(a \over 5~{\rm AU}\right)^{0.19} \left[{\kappa_s (\xi/2)^3 f_T^4 \over F \sqrt{M_\ast}}\right]^{3/20}
%\end{equation}

\subsection{The Opacity Effect}
A  lower opacity  could lower the core mass.  Reducing the opacity by a factor of one hundred cuts the core mass by more than a factor of 10, specifically to 9 $M_\oplus$ for the parameters in \Eq{eq:Mcrit}.  The reduction is not  as strong as the nominal scaling would imply, $0.01^{3/5} \approx 0.06$, because $\xi$ increases.

Even with significantly lower opacities, radiative diffusion remains a good approximation at the convective boundary.  We estimate the optical depth as (for $\beta = 2$):
\begin{equation}
\tau\cb \sim {\kappa P\cb \over g} \sim 7 \times 10^4 {F_T^4 F_\kappa \over \left(\mc \over 10 \right) \left(\au \over 10\right)^{12 \over 7}} 
\end{equation} 
where $P\cb \sim P_M$ for a self-gravitating atmosphere and $g \sim G M\co/R_B^2$, with both approximation good to within the the order unity factor $\xi$.  Clearly $\tau\cb \gg 1$ even for $F_\kappa \lesssim 0.01$ out to very wide separations.

A hotter disk would increase core masses.  Instead of our passive disk model, adopting the standard Hayashi temperature profile would increase core masses by $\sim 50\%$.  A hotter accretion phase would further increase core masses, but such phases are presumably short lived.

\subsection{Surface Terms}
We now check the relevance of the neglected surface terms in \Eq{eq:EB}.  We already showed that accretion energy can be exactly eliminated by choosing an outer boundary near the Bondi radius.   We now show that accretion energy is also a small correction at the RCB.   A rough comparison, (ignoring terms of order $\xi$) of  accretion luminosity vs.\ $\dot{E}$ gives
\begin{equation}
{G M \dot{M} \over R \dot{E}} = {G M  \over R}{dM \over dE} \sim {G M\co^2 \over R_B E}{P\cb \over P_M} \sim \sqrt{R\co \over R_B} \ll 1
\end{equation} \, ,
where we assume $P\cb \sim P_M$  for a massive atmosphere.  Accretion energy at the protoplanetary surface is thus very weak for marginally self-gravitating atmospheres, and even weaker for lower mass atmospheres.  A similar scaling analysis shows that the work term, $P\surf \p V\surf/\p t$ is similarly weak.  Nevertheless numerical calculations should include, or check, the importance of these surface terms in more realistic self-gravitating atmospheres.

\section{Hydrogen Dissociation}
The dissociation of molecular Hydrogen deep in the atmospheres of accreting protoplanets plays a significant role in the energetics of core accretion.  In the high density regions $r  \ll r\cb$ of a convective atmosphere, the thermal plus gravitational energy scales as
\begin{equation}
dE = -4 \pi \nabla_{\rm ad}^{1/\nabla_{\rm ad}} \rho\cb R_B'^{1/(\gamma-1)} r^{\frac{2\gamma - 3}{\gamma - 1}} {dr \over r}
\end{equation} 
If $\gamma < 3/2$ then the main contribution to the energy is at the bottom of the atmosphere, i.e.\ the core.  This is the case for a diatomic ideal gas ($\gamma = 1.4$) or a solar mixture of hydrogen and helium ($\gamma \approx 1.43$).  However a monatomic gas has $\gamma = 5/3 > 3/2$.  In this case, the atmosphere's energy is no longer concentrated near the bottom, but will be concentrated near the top of the convective zone.

A likely structure is an atmosphere that is dissociated near the base, but becomes molecular near the top of the convective zone.  In this case the atmosphere's energy budget would be concentrated near the atomic to molecular transition.

The energy required to dissociate a hydrogen molecule, $I = 4.467$ eV can be significant to the overall energy budget.
Since
\begin{equation}
{I \over \nabla_{\rm ad} G M\co (2 m_{\rm H})/r} \approx 3 \left(M\co \over 10 M_\oplus \right)^{-2/3} {r \over R_{\rm c}}
\end{equation} 
we see that this energy is always relevant.

We can use the Saha equation to determine where dissociation is significant,
\begin{equation}
{n_{\rm H}^2 \over n_{\rm H_2}} = \left(\pi m_{\rm H} k T \over h \right)^{3/2} e^{-I/(kT)}
\end{equation} 
We introduce a reaction coordinate $\delta$ so that $n_{\rm H} = 2 \delta n_o$ and $n_{\rm H_2} = (1-\delta) n_o$ with $n_o = \rho X/(2 m_{\rm H})$ the number density when all hydrogen is molecular.  We express equilibrium as
\begin{equation}
{\delta^2 \over 1-\delta} = f_\mu {P_{\rm diss}(T) \over P}
\end{equation} 
with the characteristic pressure below which dissociation occurs is
\begin{equation}
P_{\rm diss} = {\left(kT\right)^{5/2} \over 4} \left( \pi m_{\rm H} \over h^2\right)^{3/2}  e^{-I/(kT)}
\end{equation} 
and the order unity factor
\begin{equation}
f_\mu = 2\delta + (1-\delta) + Y/2 + Z/\mu_Z
\end{equation} 
accounts for variations in the mean molecular weight with dissociation.  (Take $\mu_Z = 31/2$, but not too significant.)

Thus dissociation occurs where $P \lesssim P_{\rm diss}(T)$.  At disk temperatures (say 150 K) the dissociation pressure is negligibly small ($\sim 10^{-141}$ dyne cm$^{-2}$) and no dissociation occurs.  However at core temperatures the dissociation pressure is quite large especially for massive cores.  Dissociation is guaranteed.

%\section{What is the Core Instability}
%It is often argued that the critical core mass arises from the inability of a massive atmosphere to maintain hydrostatic equilibrium.    The analytic version of this argument in S82 makes several assumptions.  In particular self-gravity is neglected in some places, but included in others.   Presumably the assumptions are reasonable, because numerical  structure models of Mizuno and Papaloizou \& Terquem give similar results.
% The physical nature of the core instability remains unclear (to me at least).
%
%One possibly illuminating test problem would be the integration of the S82 problem (raditive diffusion at constant opacity) with a consistent inclusion of self-gravity.  The integration solves the structure equations for $dT/dr$, $dP/dr$ (or $d \rho /dr$) and $dM/dr$.  With self-gravity it's hard to avoid the numerical nature of the problem with the boundary conditions for $M$ applying at the core, but for $T$ and $P$ in the disk.
%
%One question is how self-gravity changes the temperature profile (and this is relevant for adiabatic profiles as well).  Without self gravity, the optically thick result is $T = G M\co/(4 R_g R\co)$.  S82 assumes the core mass is replaced by the total mass.  This may be reasonable in an average sense?

\section{Structure Models}
%We propse to numerically solve for the evolution of the atmosphere of a protoplanet that cools in the absense of planetesimal accretion.  Thus all heat loss comes form the atmosphere.  Since much of the atmosphere is convective most models require a mixing length model to calculate the luminosity.  Motivated by analytic models, the goal is to avoid the need for a mixing length model by simply calculating the radiative luminsotiy $L$ at a consistently determined RCB.  The goal then is to construct solutions along a sequence of entropy that gradually decreases from the disk entropy.  From a self-consistently determined luminosity, the cooling timescales would then be given consistently.  \footnote{I thought about fixing $L$ letting $S$ adjust, but this seems problematic if (for instance) $L$ is double-valued.} 
%
%A Henyey method would probably work best.  Practical concerns include a good initial guess and ``root finding" to determine $L$ at a given $S$.  Radial windows that cause multiple convective zones are also a concern.

The atmospheric structure equations  in pressure units are
\begin{subeqnarray}\label{eq:struct}
{\p m \over \p P} &=& - {4 \pi r^4 \over G m} \\
{\p r \over \p P} &=& - { r^2 \over G m \rho} \\
{\p T \over \p P} &=&  \nabla {T \over P} \\
{\p L  \over \p P} &=& {\p m \over \p P} \epsilon\, .\slabel{eq:dL}
\end{subeqnarray} 
When the atmospheric luminosity is carried by radiative diffusion, the lapse rate $\nabla$ is
\begin{equation}
\delrad ={ 3 \kappa P \over 64 \pi G m \sigma T^4}L\, .
\end{equation} 
We assume efficient convection when $\delrad > \delad$ so that $\nabla = \min(\delad, \delrad)$.
The local the local energy generation, $\epsilon$, includes heating from  gravitational contraction
\begin{equation}\label{eq:epsg}
\epsilon_g = -T {\p S \over \p t}\, .
\end{equation} 
Other sources of local heating, such as drag on infalling planetesimals could be specified but are not considered here.  To solve these equations we must supply opacity law for $\kappa$ and an equation of state (EOS) to give $\rho$ and $\delad$ as functions of $T$ and $P$ and the assumed composition.  Note that the luminosity and opacity do not affect the structure of convection zones, but do affect their boundaries.  If we can assume that the atmospheric luminosity is primarily generated in the convective zone, then the time dependence in the structure equations can be neglected.

Boundary conditions for the problem exist both at the core and at the outer accretion radius.  At the core, $m = M\co$, $r = R\co$, $L = L\co$.  At the top boundary $T = T\di$ and $P = P\di$ set by conditions in the disk midplane.  In addition to these four boundary conditions, an initial solution is required for the time dependence that arises when  \Eq{eq:epsg} is included and allowed to affect the radiative zone structure.

\subsection{Shooting Method (1a)}\label{sec:shoot1a}
This is the first (and currently only) method we have numerically developed.  It is not described yet!  It is similar to method 1b below, but holds entropy values, $S_i$ fixed instead of mass (or luminosity).  In brief, we obtain solutions by varying the free thermodynamic variable at the core until the radius at the top of the convective boundary matches the value in \Eq{eq:Rcb}.

\subsection{Shooting Method (1b)}\label{sec:shoot1b}
We here describe an algorithm to obtain static solutions by integrating downward from the top of the convective zone.    As in \S\ref{sec:shoot1a} this method uses simplified solutions for the  radiative zone that neglect self gravity and luminosity generation.  With this method we can obtain an evolutionary sequence at fixed values of either mass or luminosity, with the other quantity optimized to get a converged solution.   Fixing the mass intervals has the advantage that total mass increases monotonically, while the luminosity is known to decrease to a minimum value and then increase (as with entropy in \S\ref{sec:shoot1a}).

 The algorithm proceeds as follows:
\begin{enumerate}
\item Guess a value for either the luminosity, $L_{\rm guess}$, or mass, $M_{\rm guess}$, at the top of the convective zone (whichever is not being held fixed).  From this guess get estimates of $P\cb$ from \Eq{eq:Lcb}, $T\cb$ from \Eq{eq:Tcb} and $r\cb$ from \Eq{eq:Rcb}.
\item Integrate the structure equations from the convective boundary to the core, thereby obtaining a model value of the core mass. 
\item Repeat,  varying either $L_{\rm guess}$ or  $M_{\rm guess}$ in step 1 until the core mass in the atmospheric solution matches the assumed value.
\end{enumerate}

Solutions for a range of masses (and luminosities) can be connected in an evolutionary sequence using the global energy equation that we derive from the virial theorem.

For simplified radiative zone solutions, shooting from the top has the advantage that a monotonic sequence of mass can easily be contructed.  On the other hand, integrating from the bottom can be more numerically efficient because the tabulated values of thermodynamic variables along a single adiabat can be used for each solution.  When shooting from the top, the variation of the luminosity (or mass) requires integration along different adiabats, with greater interpolation costs.  (But computational costs aren't really a big limitation.?)

\subsection{Shooting Method (2)}\label{sec:shoot2}
%At this writing, we are using a shooting model (not described here) which uses the radiative convective boundary (RCB) as the fitting point.    The advantage of this approach is that a 1D (Newton Raphson) optimization of a thermodynamic variable at the core (such as temperature at fixed entropy) can be used to find a model.  A limitation of this approach is that the structure of the radiative zone must be computed analytically, making it difficult to include detailed physics, such as variable opacities and self-gravity (because the total mass of the atmosphere is unknown \emph{a priori}).  To overcome these obstacles we consider here a shooting method that extends from the core to the disk and the fitting is done at the outer boundary.  The price of this more detailed model is that a 2D optimization must be performed which includes finding the self-consistent luminosity.

%In traditional stellar models it is difficult to shoot to the outer boundary because of the numerical instability of very low pressures, i.e.\ small numerical errors in the high density region propagate like a whip (Hansen \& Kawaler; Kippenhahn \& Weigert).   These numerical instabilities should (hopefully) be less severe in the protoplanetary problem due to the finite core radius (and lack of central divergences) and the finite bounding pressure of the disk.

Here we describe a method that allows us to add the self-gravity of the radiative zone.  While a time dependent model could be developed (see \S\ref{sec:timedep}),  we first describe the method for constant radiative zone luminosity.  As in \S\ref{sec:shoot1b} we integrate inwards, this time starting from the disk. The outer boundary conditions are $T = T\di$,  $P = P\di$,  $M = M_i$ where $i$ is simply a label for each evolutionary stage, 
\begin{equation}
r = r_{\rm fit} = n_{\rm fit}{ G M_i \over \Rg T\di}\, ,
\end{equation} 
which places the outer radius at a finite number, $n_{\rm fit} > 1$, of Bondi radii.  Analytic solutions like \Eq{eq:rhoiso} show that the precise choice of fit radius has negligible effect on the atmospheric solution (R06).

To find a solution we must vary $L = L_{\rm guess}$,  though as in \S\ref{sec:shoot1b} we could fix $L$ and vary  $M$.   An inward integration for a trial $L_{\rm guess}$  gives the value of the core radius $R_{\rm c, guess}$ implied by the trial solution.  We adjust $L_{\rm guess}$ until $R_{\rm c, guess}$ converges to the actual $R\co$.

With atmospheric self-gravity, shooting outwards is more difficult because two parameters must be varied.  It is impossible to fix the total mass shooting from the bottom, but we can try to construct models for choices of $L_i$.  In this case,  both $P\co$ and $T\co$ must be varied to match $T = T\di$ and $r = r_{\rm fit}$ (using the total mass $M$ from the trial integration itself) at the top, which is defined by $P = P\di$.  Alternatively one could try to find solutions with fixed values of entropy $S = S_i$ for the interior adiabat as in \S\ref{sec:shoot1a}.  This gives only one thermodynamic variable to specify at the core, since the trial choice of $T\co$ gives $P\co$ at fixed entropy.  However trial values of $L\cb$ must also be chosen, and there remain two quantities to vary ($L\cb$ and the core thermodynamic variable) to match the two conditions at the upper boundary.  With only a single variable to optimize, downward integration is more efficient.

\subsection{Time dependent shooting}\label{sec:timedep}
We now consider how to add time dependence to the method of \S\ref{sec:shoot2}.  \emph{For now this is just a sketchy description.}  First a static solution must be obtained for some small mass.  Then the next, slightly larger mass should be chosen.    A solution now requires variation of two quantities, the luminosity and the assumed time difference between the two states.  The time difference is used to calculate the heating in the convective zone using \Eq{eq:epsg}.  Convergence occurs when the atmosphere solution produces both the correct total mass and the timestep matched the global energy change between the two states.

%\begin{eqnarray} 
%y(\xi\co, L)  &=&  T(P\di; \xi\co, L) - T\di\\
%z(\xi\co, L) &=& r(P\di; \xi\co, L) - n_{\rm fit}{ G m_{\rm B}(\xi\co, L) \over \Rg T\di}
%\end{eqnarray} 
%then the boundary conditions are met when $y = z = 0$.  The 2D Newton-Raphson method can be used to search for the optimal values of $\xi\co$ and $L$.  The mass enclosed by the Bondi radius, $m_{\rm B}$, is where the mass first exceeds $r \Rg T\di/G$.  In practice $m_{\rm B}$ will probably differ only slightly from $m(P\di)$ because of the low density of disk material, but this can be checked numerically.

%\begin{figure}[tb!] %  figure placement: here, top, bottom, or page
%\if\submitms y
% 	\includegraphics[width=6in]{f.eps}
% \else
%	\hspace{-1cm}
%   	\includegraphics[width=3.9in]{../figs/.pdf} 
%      \includegraphics[width=3.9in]{../figs/.eps} 
%\fi
%   	\caption{}
%   	\label{fig:}
%\end{figure}

\acknowledgements
%Portions of this project were supported by the {\it NASA} {\it Astrophysics Theory Program} and  {\it Origins of Solar Systems Program}  through grant NNX10AF35G.\\
\appendix
\section{Self-Gravitating Isothermal Atmosphere}

For an isothermal atmosphere, hydrostatic equilibrium gives:
\begin{equation} \label{eq:HBiso}
r^2 {d \ln \rho \over d r} = -{G \over \mathcal{R} T} \left(M\co + 4 \pi \int_{R\co}^r \rho(r') r'^2 dr'\right) 
\end{equation} 
This integro-differential equation is subject to the boundary condition that the density is $\rho\di$ at large distances.  This boundary condition can be applied at $n_B R_B$, where the Bondi radius $R_B = GM\co/\mathcal{R} T$.  The number $n_B > 1$ does not affect the solution since the gravitational perturbation from the planet is weak outside the Bondi radius.

Radial differentiation of \Eq{eq:HBiso} gives:
\begin{equation} 
x^2 {d^2 S \over dx^2} + 2x {d S \over dx} = - \theta_o e^S x^2
\end{equation} 
where $S = \ln(\rho/\rho_o)$, $x = r/R\co$ and $\theta_o = 4 \pi G \rho\di R\co^2 /(\mathcal{R} T)$.  Since $\theta_o$ is the squared ratio of the core radius to the disk's Jeans length, $\theta_o \lll 1$. The outer boundary condition now reads $S(n_B \theta\co) = 0$ where $\theta\co = R_B/R\co > 1$.  An inner boundary condition
\begin{equation} 
{d S \over dx}(x = 1) = - \theta\co
\end{equation} 
must also be satisfied.

\if\bibinc n
\bibliography{refs}
\fi

\if\bibinc y
\begin{thebibliography}
\end{thebibliography}
\fi

\end{document}

### 2013-01-22 tex/CoreAccretionPaperOutline.tex
added math, tex and slides folders

%\documentclass[apj]{emulateapj}
%\documentclass[manuscript]{aastex} 
\documentclass[aps,prd,preprint]{revtex4}

\usepackage{graphicx}                        
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}

%%%%%%%%%%%%%%%0

\begin{document}

\title{Core Accretion Outline}
\author{Ana-Maria Piso}

\begin{enumerate}
\item Introduction \& Motivation
\item Atmosphere model
\begin{enumerate}
\item Two-layer model - inner convective + outer radiative
\begin{enumerate}
\item describe analytical solution assuming ideal gas + no self-gravity?
\end{enumerate}
\item Basic equations
\item Boundary conditions + disk model
\item Real EOS (tables + table extension, most content of 'EOSnotes')
\item Virial theorem, energy budget, how to estimate cooling time
\end{enumerate}
\item Numerical method
\begin{enumerate}
\item Assumptions
\item Shooting method description
\end{enumerate}
\item Results \& Analysis
\begin{enumerate}
\item Plots: P-T, M-T, M-R etc.; Luminosity plots
\item Critical core mass
\begin{enumerate}
\item Mcrit vs. r plots, explain, comparison with polytropes
\item Mcrit vs. r for different disk assumptions?
\end{enumerate}
\end{enumerate}
\item Summary \& Conclusions
\begin{enumerate}
\item Caveats (neglecting self-gravity of radiative region, assumption of constant luminosity)
\item Future prospects: 2D extension
\end{enumerate}
\end{enumerate}

\end{document}

### 2013-01-22 tex/EOSnotes.tex
added math, tex and slides folders

\documentclass[apj]{emulateapj}
%\documentclass[manuscript]{aastex}

\usepackage{graphicx}                        
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}

%%%%%%%%%%%%%%%0

\begin{document}

\title{EOS notes}
\author{Ana-Maria Piso}

\section{Adiabatic atmospheres}
\label{adiabatic}

%\begin{verbatim}
%Most of the derivations are already in the notes \\
% you wrote,  so it didn't make sense to rewrite \\
%  them. I just wrote the structure equation \\
%   we are trying to solve.
%\end{verbatim}

In the case of an ideal gas, the relation between temperature, density and pressure is given by the ideal gas law:

\begin{equation}
\label{eq:idealgas}
P=\rho \mathcal{R} T,
\end{equation}

\noindent where $\mathcal{R}$ is the specific gas constant. Hydrostatic balance along an adiabat then gives:

\begin{equation}
\label{eq:drdtid}
\frac{dr}{dT}=-\frac{\mathcal{R} r^2}{\nabla_{ad} G M}
\end{equation}

In the case of a non-ideal gas, however, equation (\ref{eq:idealgas}) no longer applies. Further on we derive the structure equations for a non-ideal gas, using $\log P$ (with $P$ the pressure) as the integration variable. We start from hydrostatic equilibrium:

\begin{equation}
\label{eq:hydro}
\frac{dP}{dr}=-\rho g=-\rho \frac{G M}{r^2},
\end{equation}

\noindent which gives

\begin{equation}
\label{eq:drdp}
\frac{dr}{d\log P} = -\frac{P}{\rho} \frac{r^2}{G m}
\end{equation}

Mass conservation gives:

\begin{equation}
\label{eq:mcons}
\frac{dm}{dr} = 4 \pi r^2 \rho
\end{equation}

From this and equation (\ref{eq:drdp}) we obtain:

\begin{equation}
\label{eq:dmdp}
\frac{dm}{d \log P} = -\frac{4 \pi P}{G m} r^4
\end{equation}

Furthermore, the gravitation and internal energy can be expressed as

\begin{equation}
\label{eq:degdp}
\frac{d E_g}{d \log P} = 4 \pi P r^3
\end{equation}

\noindent and

\begin{equation}
\label{eq:deidp}
\frac{d E_i}{d \log P} = - \frac{4 \pi u p}{G m} r^4,
\end{equation}

\noindent where $u$ is the internal energy of the gas per unit mass: $u = U/m$.

%we start from the definition of the adiabatic gradient $\nabla_{ad}$, which is valid even in the non-i%deal cases:

%\begin{equation}
%\label{eq:delad1}
%\nabla_{ad}=\frac{\partial \log T}{\partial \log P}=\frac{P}{T}\frac{dT}{dP}
%\end{equation}
%
%It follows that 
%
%\begin{equation}
%\label{eq:dpdt}
%\frac{dP}{dT}=\frac{1}{\nabla_{ad}}\frac{P}{T}
%\end{equation}
%
%Hydrostatic balance gives
%
%\begin{equation}
%\label{eq:hydro}
%\frac{dP}{dr}=-\rho g=-\rho \frac{G M}{r^2}
%\end{equation}
%
%From equations (\ref{eq:dpdt}) and (\ref{eq:hydro}) we thus find:
%
%\begin{equation}
%\label{eq:drdt}
%\frac{dr}{dT}=-\frac{1}{G M} \frac{P}{\rho \nabla_{ad}} \frac{r^2}{T},
%\end{equation}

In the above equations, $\rho$ and $u$ are functions of pressure. In order to solve them we therefore need to know $\rho(P)$ and $u(P)$ for a non-ideal gas. This will be discussed in detail in the next section.

\section{EOS Tables}

We start from the equation of state tables of \cite{saumon95}. These equations of state take into account non ideal interactions, and include physical treatments of dissociation and ionization, caused by both pressure and temperature effects. However, the \cite{saumon95} EOS tables only cover a relatively high range of temperatures and pressures: $2.10 < \log_{10} T(K)<7.06$ and $4<\log_{10}P$(dyn cm$^{-2})<19$. While these ranges may be suitable deep inside the protoplanetary atmosphere, they are no longer valid as we approach the boundary between the convective and radiative regions of the atmosphere. Since the radiative region is approximated as almost isothermal, the temperature at the bottom of this region is close to the disk temperature, which can be as low as a few tens of Kelvin for large semi-major axes. Additionally, the pressure in the disk can be as low as $10^{-1}$ dyn cm$^{-2}$ (\cite{pn05}). As such, for our purposes it is necessary to extend the \cite{saumon95} EOS tables to lower temperature and pressure values.

We choose $\log_{10} T (K)=1$ and$ \log_{10}P$(dyn cm$^{-2})=-4.4$ as our lower boundaries for temperature and pressure, respectively. Our temperature and pressure grid hence becomes: $1 < \log_{10} T(K)<7.06$ and $-4.4<\log_{10}P$(dyn cm$^{-2})<19$. The other quantities in the tables are calculated as follows.

\subsection{Hydrogen}

\label{hydrogen}

For a system of particles, the partition function can be written as the product of all partition functions associated with each type of energy that the system can have:

\begin{equation}
\label{eq:z}
Z=Z_t Z_r Z_v Z_e Z_n,
\end{equation}

\noindent where $Z_t$, $Z_r$, $Z_v$, $Z_e$ and $Z_n$ are the partition functions associated with translation, rotation, vibration, electronic excitation and nuclear excitation, respectively. For hydrogen, electronic and nuclear excitation are only significant at temperatures higher than our region of interest ($\theta_e \approx 12000$ K and $\theta_n >> \theta_e$, where $\theta_e$ and $\theta_n$ are the characteristic temperatures for electronic and nuclear excitation, respectively). As such, we will only take into account the translation, rotation and vibration of the hydrogen molecule:

\begin{equation}
\label{eq:zagain}
Z=Z_t Z_r Z_v
\end{equation}

The partition function associated with the motion of the center of mass of the molecule is given by (in the classical limit):

\begin{equation}
\label{eq:Zt}
Z_t=(m/2 \beta \pi \hbar^2)^{3/2} V,
\end{equation}

\noindent where $\beta=1/(k T)$ and $V$ is the volume. The rotational partition function is generally written as:

\begin{equation}
\label{eq:Zr}
Z_r=\sum_0^\infty (2 j+1) \exp{\Big[\frac{-j (j+1)\theta_r}{T}\Big]},
\end{equation}

\noindent where $\theta_r$ is the characteristic temperature for rotational motion. In the case of hydrogen, $\theta_r \approx 85$ K. However, molecular hydrogen occurs in two isomeric forms: orthohydrogen, with the proton spins aligned parallel to each other, and parahydrogen, with the proton spins aligned antiparallel. Each isomer has a different rotational partition function; when the spin isomers are in equilibrium, then the partition function can be written as:

\begin{equation}
\label{eq:Zrspin}
Z_r=\sum_0^\infty (2-(-1)^j) (2j+1) \exp{\Big[\frac{-j (j+1) \theta_r}{T}\Big]}
\end{equation}

In our range of temperatures of interest, we found that $Z_r$ converges after about 25 terms in the series.

Finally, the partition function for vibrational motion is given by:

\begin{equation}
\label{eq:Zv}
Z_v=[1-\exp{(\theta_v/T)}]^{-1},
\end{equation}

\noindent where $\theta_v$ is the characteristic temperature for vibrational motion, $\theta_v \approx 6140$ K for hydrogen.

If the partition function of a system of $N$ particles is known in terms of $(V, T, N)$, the internal energy and entropy of the system can be determined as follows:

\begin{equation}
\label{eq:U}
U_N=k T^2 \Big(\frac{\partial \log{Z}}{\partial T}\Big)_{V, N}
\end{equation}

\begin{equation}
\label{eq:S}
S_N=k \log{Z} + \frac{U_N}{T}
\end{equation}

The energy and entropy per mass will subsequently be:

\begin{equation}
\label{eq:u}
U=\mathcal{R} T^2 \Big(\frac{\partial \log{Z}}{\partial T}\Big)_{V, N}
\end{equation}

\begin{equation}
\label{eq:s}
S=\mathcal{R} \log{Z} + \frac{U}{T}
\end{equation}

Since $Z=Z_t Z_r Z_v$, it is easy to notice that $U=U_t+U_r+U_v$ and $S=S_t+S_r+S_v$, where $U_t$, $U_r$, $U_v$, $S_t$, $S_r$, $S_v$ are the quantities corresponding to the individual translation, rotation and partition functions, respectively.

It can be shown that the entropy per mass due to translational motion can be expressed as:

\begin{equation}
\label{eq:st}
S_t=\mathcal{R} \Big[ \frac{5}{2} \ln{T} - \ln{P} + \ln \Big( \frac{(2 \pi)^{3/2} \mathcal{R}^{5/2} \mu^4}{h^3}\Big) +\frac{5}{2} \Big]
\end{equation}

\noindent with $\mu$ the mean molecular weight. Equation (\ref{eq:st}) is known as the Sackur-Tetrode formula, and it is only applicable to an ideal gas. It can also be easily shown that the internal energy per mass due to translational motion is given by:

\begin{equation}
\label{eq:ut}
U_t=\frac{3}{2} \mathcal{R} T
\end{equation}

Putting all of the above together, we can now evaluate the thermodynamic quantities needed to extend the \cite{saumon95} EOS tables to low temperatures and pressures.

\begin{enumerate}

\item{\textbf{Density.}} In the low temperature, low pressure regime, hydrogen is molecular and behaves like an ideal gas. As such, the density in this region follows the ideal gas law $P=\rho \mathcal{R} T$.
\item{\textbf{Internal energy per mass.}} $U=U_t+U_r+U_v$, where $U_t$ is given by equation (\ref{eq:ut}), and $U_r$, $U_v$ are determined using equations (\ref{eq:u}), (\ref{eq:Zrspin}) and (\ref{eq:Zv}) above.
\item{\textbf{Entropy per unit mass}}. Similarly, $S=S_t+S_r+S_v$, where $S_t$ is given by equation (\ref{eq:st}), and $S_r$, $S_v$ can be determined from equation (\ref{eq:s}) and the calculated expressions for $U_r$ and $U_v$, respectively.
\item{\textbf{Entropy logarithmic derivatives}}. The logarithmic derivatives $S_T$ and $S_P$ are given by:

\begin{equation}
\label{eq:sT}
S_T=\frac{\partial \log{S}}{\partial \log{T}} \Big |_P
\end{equation}

\noindent and

\begin{equation}
\label{eq:sP}
S_T=\frac{\partial \log{S}}{\partial \log{P}} \Big |_T
\end{equation}

We calculate $S_T$ and $S_P$ using the table values for $S$, $T$ and $P$, and a linear central difference formula.

\item{\textbf{Adiabatic gradient $\nabla_{ad}$}}. The adiabatic gradient is defined as:

\begin{equation}
\label{eq:delad}
\nabla_{ad}=\frac{\partial \log{T}}{\partial \log{P}} \Big |_S = -\frac{S_P}{S_T}
\end{equation}

We evaluate it from the tabulated values for $S_T$ and $S_P$ determined above. Figure 1 shows a contour plot of the adiabatic index for the extended EOS table, while the black lines represent constant entropy curves. The upper right part of the plot ($\log T>2.1$ and $\log P>4$) is based on the \cite{saumon95} EOS table, while the rest of the plot is our extension. We see that the two tables join smoothly for entropy curves between $8.80<\log{S}$(K g$^{-1})<9.07$.

\end{enumerate}

\begin{figure}[h!]
\centering
\includegraphics[scale=.5]{../figs/EOS/delad_H_ext}
\caption{$\nabla_{ad}$ contour plot for the hydrogen extended table. The black curves represent constant entropy curves.}
\end{figure}

\subsection{Helium}

We extend the helium EOS tables based on a similar procedure. Since helium is primarily neutral and atomic at low temperatures and pressures, we treat it as an ideal monoatomic gas, and subsequently only take into account the translational components of the necessary thermodynamic quantities (see subsection \ref{hydrogen} above for details). The analogous $\nabla_{ad}$ contour plot for helium can be seen in Figure 2. We notice that, in the case of helium, the original and extended table join smoothly for entropy curves between $8.29<\log{S}$(K g$^{-1})<8.77$.

\begin{figure}[h!]
\centering
\includegraphics[scale=.5]{../figs/EOS/delad_He_ext}
\caption{$\nabla_{ad}$ contour plot for the helium extended table. The black curves represent constant entropy curves.}
\end{figure}

\subsection{Hydrogen and helium mixtures}

In order to obtain similar EOS tables for hydrogen and helium mixtures, we follow the approach of \cite{saumon95} to calculate the corresponding thermodynamic quantities for the mixture. We denote the helium mass fraction by $Y$. The density of the hydrogen and helium mixture is then given by:

\begin{equation}
\label{eq:rhomixt}
\frac{1}{\rho(P, T)}=\frac{1-Y}{\rho^H(P,T)}+\frac{Y}{\rho^{He}(P,T)}
\end{equation}

Similarly, the internal energy per mass of the mixture is given by:

\begin{equation}
\label{eq:umixt}
U(P, T)=(1-Y)U^H(P,T)+Y U^{He}(P,T)
\end{equation}

The entropy per mass can be calculated as:

\begin{equation}
\label{eq:smixt}
S(P,T)=(1-Y)S^H(P,T)+Y S^{He}(P,T)+S_{mix}(P,T),
\end{equation}

\noindent where $S_{mix}$ is given by:

\begin{eqnarray}
\label{eq:smix}
S_{mix}&=&k_B \frac{1-Y}{m_H}\frac{2}{1+X_H+3X_{H_2}} {\ln(1+\beta \gamma)} \nonumber \\
&-&X_e^H \ln(1+\delta)+\beta \gamma [\ln(1+1/(\beta \gamma)) \nonumber \\
&-&X_e^{He} \ln(1+1/\delta)]
\end{eqnarray}

\noindent with

\begin{equation}
\label{eq:xeH}
X_e^H=\frac{1}{2}(1-X_{H_2}-X_H),
\end{equation}

\begin{equation}
\label{eq:xe}
X_e^{He}=\frac{1}{3}(2-2X_{He}-X_{He^+}),
\end{equation}

\begin{equation}
\label{eq:beta}
\beta=\frac{m_H}{m_{He}} \frac{Y}{1-Y},
\end{equation}

\begin{equation}
\label{eq:gamma}
\gamma=\frac{3}{2}\frac{1+X_H+3+X_{H_2}}{1+2X_{He}+X_{He^+}},
\end{equation}

\begin{equation}
\label{eq:delta}
\delta=\frac{2}{3} \frac{2-2X_{He}-X_{He^+}}{1-X_{H_2}-X_H} \beta \gamma
\end{equation}

The logarithmic derivatives of the entropy (equations(\ref{eq:sT}) and (\ref{eq:sP})) can be evaluated as follows:

\begin{equation}
\label{eq:stmix}
S_T=(1-Y)\frac{S^H}{S}S_T^H+Y \frac{S^{He}}{S}S_T^{He}+\frac{S_{mix}}{S}\frac{\partial \log{S_{mix}}}{\partial \log{T}}\Big |_P,
\end{equation}

\begin{equation}
\label{eq:spmix}
S_T=(1-Y)\frac{S^H}{S}S_P^H+Y \frac{S^{He}}{S}S_P^{He}+\frac{S_{mix}}{S}\frac{\partial \log{S_{mix}}}{\partial \log{P}}\Big |_T,
\end{equation}

The logarithmic derivatives of $S_{mix}$ are calculated from the tabulated values for $S_{mix}$, density and pressure, using a linear central difference formula.

Finally, the adiabatic gradient of the mixture is given by:

\begin{equation}
\label{eq:deladmix}
\nabla_{ad}=-\frac{S_P}{S_T}
\end{equation}

The $\nabla_{ad}$ contour plot analogous to figures 1 and 2 for a hydrogen-helium mixture with $Y=0.3$ can be seen in Figure 3. In the case of the mixture, the original and extended table join smoothly for entropy curves between $8.71<\log{S}$(K g$^{-1})<9.00$.

\begin{figure}[h!]
\centering
\includegraphics[scale=.5]{../figs/EOS/delad_mixt_ext}
\caption{$\nabla_{ad}$ contour plot for the hydrogen-helium mixture ($Y=0.3$) extended table. The black curves represent constant entropy curves.}
\end{figure}

\section{Numerical atmosphere model}

We are developing a two-layer atmosphere model, with an inner convective region and an outer radiative region that matches onto the protoplanetary disk. One of our main goals is determining the critical core mass for a giant planet to be able to form within the lifetime of the disk (typically of the order of $\sim 10^7$ years), for a variety of disk conditions, and at various distances in the disk. In what follows we approximate the structure of the protoplanet as being spherically symmetric. Figure 4 depicts schematically the structure of our model atmosphere.

\begin{figure}[h!]
\centering
\includegraphics[scale=.33]{../figs/two_layer}
\caption{Schematic depiction of the two-layer model atmosphere: the convective region (of radius $r_{CB}$ and mass $M_{ad}$) and the radiative region (which extends out to the Bondi radius $RB$ and has mass $M_{rad}$) are separated by the radiative-convective boundary.}
\end{figure}

\subsection{Inner Convective Region}

We approximate the inner region as being adiabatic. From section \ref{adiabatic}, the equations that describe the structure of an adiabatic atmosphere are:

\begin{equation}
\label{eq:drdp2}
\frac{dr}{d\log P} = -\frac{P}{\rho} \frac{r^2}{G m}
\end{equation}

\begin{equation}
\label{eq:dmdp2}
\frac{dm}{d \log P} = -\frac{4 \pi P}{G m} r^4
\end{equation}

This is a system of two equations and three unknowns, $r(P)$, $m(P)$ and $\rho(P)$. In order to be able to solve the system we therefore need another relationship between thermodynamic quantities. If the entropy of the system $S$ is assumed to be known, then the density $\rho$ can be uniquely determined as a function of pressure and entropy: $\rho=\rho(P, S)$.

\subsubsection{Shooting method}

As we will see in the next section, we can determine the temperature and pressure at the radiative-convective boundary from integration of hydrostatic equilibrium in the radiative zone.

%We denote the radius of the radiative-convective boundary as $r_{CB}$.

%In what follows we approximate the structure of the protoplanet as being spherically symmetric. Furthermore, we assume hydrostatic equilibrium. As such,
%
%\begin{equation}
%\label{eq:hydroeq}
%\frac{dP}{dr}=-\rho g,
%\end{equation}
%
%\noindent where $g=GM(<r)/r^2$, with $M(<r)$ being the total mass interior to radius $r$.  The mass $M(<r)$ satisfies the equation
%
%\begin{equation}
%\label{eq:m}
%\frac{dM}{dr}=4 \pi r^2 \rho
%\end{equation}
%
%From this and equation (\ref{eq:drdt}) it follows that
%
%\begin{equation}
%\label{eq:dmdT}
%\frac{dM}{dT}=-\frac{4 \pi \rho}{G M} \frac{P}{\rho \nabla_{ad}} \frac{r^4}{T}
%\end{equation}

\bibliographystyle{apj}
\bibliography{refs}

\end{document}

### 2013-01-22 tex/draft/draft.tex
added math, tex and slides folders

--- 2013-01-22 tex/CoreAccretion.tex
@@
-% !TEX TS-program = pdflatexmk
-
 \documentclass[12pt, preprint,numberedappendix]{emulateapj}
@@
 \usepackage{natbib}
+\usepackage{color}
 
@@
 \newcommand{\p}{\partial}
-\newcommand{\xv}{\vc{x}}
-\newcommand{\kv}{\vc{k}}
 \newcommand{\brak}[1]{\langle #1\rangle}
@@
 \newcommand{\km}{\; {\rm km}}
-%\newcommand{\au}{\; \varpi_{\rm AU}}
-
+\newcommand{\au}{\; \varpi_{\rm AU}}
 \newcommand{\AU}{\; {\rm AU}}
-\newcommand{\yr}{\; {\rm yr}}
 \def\K{\; {\rm K}}
@@
 \newcommand{\Eqss}[2]{Equations (\ref{#1})--(\ref{#2})}
-\newcommand{\Eqsss}[3]{Equations (\ref{#1}), (\ref{#2}) and~(\ref{#3})}
 \newcommand{\App}[1]{Appendix~\ref{#1}}
@@
 
+\definecolor{gray}{gray}{0.5}
+\newcommand{\emgr}[1]{\emph{ \color{gray} #1}}
+
 \newenvironment{packed_item}{
@@
 
-\newcommand{\delad}{\nabla_{\rm ad}}
-\newcommand{\delrad}{\nabla_{\rm rad}}
-\newcommand{\Rg}{\mathcal{R}}
-\newcommand{\RB}{R_{\rm B}}
-\newcommand{\co}{_{\rm c}}
-\newcommand{\di}{_{\rm o}}
-\newcommand{\cb}{_{\rm CB}}
-\newcommand{\surf}{_M}
-\newcommand{\mc}{m_{\rm c \oplus}}
-\newcommand{\mcn}[1] { m_{ \rm c #1 \oplus} }
-\newcommand{\MC}{M_{\rm crit}}
-\newcommand{\au}{a_\oplus}
-\newcommand{\aun}[1]{ a_{#1\oplus} }
-
 \begin{document}
@@
 
-\shorttitle{Core Accretion}
-\shortauthors{Youdin }
+\shorttitle{Critical Core Mass at Wide Separations}
+\shortauthors{Piso \& Youdin}
 
-\title{Core Accretion of Self-Gravitating Atmospheres}
-\author{Andrew N.\ Youdin}
+\title{Lower Limits on the Core Mass for Giant Planets}
+%\title{The Critical Core Mass of Wide-Separation Giant Planets:  Lower Limits from Kelvin-Helmholtz Cooling}
+\author{Ana-Maria Piso}
 \affil{Harvard Smithsonian Center for Astrophysics}
-
+\author{ Andrew N.\ Youdin}
+\affil{JILA, University of Colorado}
 %\begin{abstract}
 %\end{abstract}
+
 \section{Introduction}
-The core accretion model \citep{PerCam74, MizNak78} proposes that giant planets form by the accretion of gas onto a solid protoplanetary core.
-Early versions of the theory found a ``critical core mass", $\MC$, above which static solutions do not exist and unstable atmospheric collapse would occur \citep{Har78, Miz80}.
-These works used hydrostatic models that include heating by planetesimal accretion, but neglect the heat generated by the gravitational, or Kelvin-Helmholtz (KH), contraction of the atmosphere.
+\emgr{Background topics to mention include: core accretion history, giant planets at wide separations and theoretical challenges posed (GI probably to big, core accretion maybe to slow).}
 
-In evolutionary models that include KH contraction \citep{BodPol86}, the atmosphere does not undergo a hydrodynamic collapse.
-Instead the quasi-static contraction of the envelope accelerates continuously.
-Contraction becomes particularly rapid after the the  ``crossover" mass, when core and atmospheric masses are equal \citep{pollack96}.
-The concepts of critical and crossover masses are similar, and even agree numerically when the planetesimal accretion luminosity exceeds the KH luminosity.
-However if the planetesimal accretion rate is small and KH contraction dominates, the critical core mass will be smaller than the more relevant  crossover mass \citep{IkoNak00}.
+\emgr{Describe goals and approach of this paper.}
 
-This paper will consider 1D atmospheres  where KH contraction dominates over planetesimal accretion.
-Such solutions give realistic lower limits on the core mass required to accrete a massive gaseous atmosphere.
-Indeed even models that include planetesimal accretion obtain the fastest formation times by stopping planetesimal accretion after the core reaches a certain mass \citep{HubBod05}.
+This paper is organized as follows: \emgr{(section summaries)}.
 
-The early phases of atmospheric growth are usually treated with 1D models where the outer boundary of the atmosphere matches onto the conditions in the disk midplane.
-As the atmosphere continues to grow, 1D solutions are no longer adequate.
-The atmosphere both notices that the disk is not spherically symmetric and at large enough masses will start to open a gap in the disk.
-In this case the accretion of gas is limited by the ability of the disk to supply gas to the growing protoplanet \citep{DAnLub08}.
-This later stage evolution is crucial for determining the final mass of the planet, but  not considered in our calculations of the initial collapse.
+\section{Quasistatic Two Layer Model}\label{sec:model}
 
-%Analytic models to explain the critical core mass have their shortcomings.
-\cite{Ste82} assumed a fully radiative atmosphere, which is inconsistent with realistic luminosities and opacities.
-\cite{Wuc93} assumed a fully convective atmosphere which is is inconsistent with the ability of atmospheres to cool.
-\citet[hereafter R06]{Raf06} considerably advanced the analytic understanding of core accretion models.
-(Explain what's great.)  However R06 neglected the luminosity due to gas accretion.
+\subsection{Model Assumptions}
+\emgr{Describe basic assumptions, perhaps in list format.
+Will need to describe basic scales and idea that atmosphere is 1D and matches smoothly onto disk.
+However we may want to separate these standard assumptions from our more unique ones that are part of the two layer model.}
 
-%Most analyses of the critical core mass considers planetesimal accretion as the dominant luminosity source.
-One notable exception is \cite{PapNel05}.
-There are several reasons to consider the atmospheres on protoplanets that are not accreting solids.
-First the planetesimal accretion rate may be a negligible source of luminosity.
-(In this sense our results complement those of R06).
-Second the growth of an atmosphere around a preexisting core represents lower limit to the timescale to form a giant planet.
-Since atmospheric mass grows sharply with core mass, there may be little loss in assuming that atmospheric accretion begins once core formation stops.
-Finally while it is desirable to self-consistently solve for core growth and atmospheric accretion, it is impossible to account for all possible core growth processes, which are quite complicated (refs).
-%For instance planet migration or scattering could bring a core that formed rapidly in one region of the disk to a region of the disk where cores could not form rapidly.
+\subsection{Structure Equations}
+\emgr{Give in pressure coordinates.  Put EOS (polytrope and detailed) and opacity laws (noting that we only consider cool radiative zones with dust opacities).}
 
-Section \ref{sec:structure} describes the structure of planet atmospheres embedded in a gas disk.
-We then derive the critical core mass to trigger runaway collapse of the atmosphere in \S\ref{sec:critmass}.
+\subsection{Boundary Conditions: Core and Disk}
 
-\section{Disk Model}
-As a fiducial disk model, we adopt the version of the minimum mass, passively irradiated model of \citet{cy10}, which gives the surface density and midplane temperature as
-\begin{subeqnarray}
-\label{eq_sigmag}
-\varSigma  &=& 70 \,F_\varSigma \aun{10}^{-3/2} ~{\rm g~cm}^{-2} \\
-T &=& 45  \,F_T\, \aun{10}^{-3/7} ~{\rm K}
-\end{subeqnarray} 
-where $\aun{10} \equiv a/{10~\rm AU}$ scales the disk radius and the normalization factors $F_\varSigma$ and $F_T$ adjust the disk mass and temperature relative to the canonical values.
-The resulting disk midplane pressure
-\begin{equation}
-P\di = 6.9 \times 10^{-9} F_\varSigma \sqrt{F_T m_\ast} \, \aun{10}^{-45/14}~{\rm bar}
-\end{equation} 
-for a mean molecular weight of 2.35.
-The stellar mass in solar units is $m_\ast \equiv M_\ast/M_\odot$, which only includes the gravitational effects, though stellar mass also influences temperature and probably correlates with average disk mass.
-Away from the midplane the pressure drops with height $z$ as a Gaussian with scale-height
-\begin{equation}
-H = {\sqrt{\Rg T\di} \over \varOmega} = 0.42 \sqrt{F_T \over m_\ast}  \, \aun{10}^{9/7} \AU\, .
-\end{equation}
+%\subsection{Convective Zone}
+%\emgr{might not need a (sub)section}
 
-\section{Atmospheric Structure}\label{sec:structure}
-In this section we review the structure of atmospheres surrounding a solid protoplanetary core embedded in a gas disk.
-The core has mass $M\co$ and radius $R\co = [3 M\co/(4 \pi \rho\co)]^{1/3}$ with internal density $\rho\co \approx 2~{\rm g~cm^{-3}}$.
-For protoplanetary disk parameters we use the minimum mass, passively irradiated disk model presented in \citet[hereafter  CY10]{cy10}.
-The disk has a midplane gas density $\rho\di$ and temperature $T\di$.
+\subsection{Radiative Zone Solutions}
+\emgr{Give analytic solution valid for power law opacity and lack of self-gravity.}
 
-The core can attract a dense atmosphere inside its Bondi radius
-\begin{equation} 
-R_B = {G M\co \over \mathcal{R} T\di} \approx 0.17 {\mcn{10}  \, \aun{10}^{3/7} \over F_T} ~\AU
-\end{equation} 
-where $\mcn{10} \equiv M\co/(10~M_\oplus)$ and $\mathcal{R}$ is the specific gas constant.
+\subsection{Matching at the Radiative-Convective Boundary (RCB)}
+\emgr{Include figure of an example atmospheric profile.}
 
-Low mass planets satisfy a hierarchy of length scales,  $R\co \ll R_B < R_{\rm H} < H$, where %$h\di \approx 3.3 \times 10^6 \au^{9/7}~{\rm km}$ is the gas scale-height and
-\begin{equation} 
-R_{\rm H} = \left(M\co \over 3 M_\ast \right)^{1/3}a \approx 0.22 {\mcn{10}^{1/3} \, \aun{10} \over m_\ast^{1/3}}~\AU
-\end{equation} 
-is the Hill radius, inside which the planet's gravity competes with Keplerian shear.
+\subsection{The Gravothermal Catastrophe}
+\emgr{Show and explain double valued solutions, including caveat that upper branch breaks several model assumptions.}
 
-As protoplanets grow in mass the length scales $R_{\rm H} \sim R_{\rm B} \sim H$ are roughly equal at a characteristic ``transitional" mass
-\begin{equation}
-M_{\rm tr} > {(\Rg T\di)^{3/2} \over G \varOmega} \approx 25 \, {F_T^{3/2} \over \sqrt{m_\ast} } \, \aun{10}^{6/7}~ M_\oplus \, .
-\end{equation} 
-Technically the length scales re-order in three steps of increasing mass.
-First for $M > M_{\rm tr}/\sqrt{3}$ the ordering becomes $R_{\rm H} < R_{\rm B} < H$ as tidal truncation becomes more important than thermal constriants.
-While the shear flow past the atmosphere is clearly not spherically symmetric, it is still common take the atmosphere as spherically symmetric with an outer boundary at the Hill radius.
-When the mass increases further to $M > M_{\rm tr}$ then  $R_{\rm H} < H < R_{\rm B}$, which does not change the fact that tidal truncation is the stronger effect at the outer boundary.
+\section{Cooling History}
+\emgr{start with basic idea of quasistatic contraction and $L = -\dot{E}$}
+\subsection{Role of Finite Pressure}
+ \emgr{Give more more general result, but defer derivation to \App{sec:virial}.}
+\subsection{Estimating a Critical Core Mass}
+\emgr{Describe how and why we use the time to gravothermal catastrophe as estimate of minimum core mass}
 
-Finally when $M > 3 M_{\rm tr}$ the high mass scaling $H < R_{\rm H} < R_{\rm B}$ is attained.
-The assumption of spherical symmetry fails for two reasons.
-First the planet now notices the vertical density stratification of the disk.
-Second, the planet's torques on the disk can begin to open gaps, creating further density inhomogeneity.
-Since runaway accretion typically begins for masses well below $\sim 75 M_\oplus$, the assumption of spherical symmetry is a good, if not perfect, approximation for understanding whether cores can attrach massive atmospheres.
-Moreover the approximation becomes even better in the disk's outer regions.
+\section{Critical Core Mass}
+\emgr{These are the take home results!}
 
-Another relevant for atmospheric structure is the mass of disk gas that fills the Bondi radius, at the ambient density.  The ratio of this mass to the core mass is 
-\begin{equation}
-\theta\co \equiv { 4 \pi \rho\di R_B^3 \over 3 M\co} \approx 5 \times 10^{-3} {\mcn{10}^2 \over \aun{10}^{3/2}}{F_\varSigma \sqrt{m_\ast} \over F_T^{7/2}}\, .
-\end{equation} 
-The smaller this parameter the more gravitational compression is required for an atmosphere to become self-gravitating.
+\subsection{Role of Disk Temperature and Pressure}
 
-There are two limiting cases of atmospheric structure, isothermal and adiabatic.
-Atmospheres that are isothermal at $T\di$ have cooled completely and give the most massive atmosphere.
-Convectively unstable atmospheres transport heat very efficiently and are nearly adiabatic.
-Adiabatic atmospheres are the opposite extreme of the isothermal case, giving the steepest increase of temperature with atmospheric dept.
-The least dense atmospheres are fully adiabatic with the entropy of the disk.
+\subsection{Core Mass vs. Disk Radius}
 
-The structure of an embedded protoplanet consists of an inner adiabatic region that matches onto an outer radiative zone, that in some cases, can be approximated as nearly isothermal.
-This basic structure was demonstrated by R06 for an atmosphere supported by planetesimal accretion luminosity.
-PN05 showed that atmospheres powered by gas accretion alone have a similar structure.
-Due to their intense irradiation, ``hot Jupiters" similarly have an isothermal layer that matches onto a deeper adiabatic interior \citep{ab06, ym10}.
+\subsection{Opacity Dependence}
 
-\subsection{Isothermal Atmosphere}
-We now consider the structure of a low mass, i.e.\ non-self gravitating, isothermal atmosphere.
-We assume the atmosphere matches onto a constant background density, $\rho\di$, at a distance $r_{\rm fit} = n_{\rm fit} \RB$.
-The resulting density profile is
-\begin{equation} \label{eq:rhoiso}
-\rho = \rho\di \exp \left({R_B \over r} - {1 \over n_B} \right) \approx   \rho\di \exp \left(R_B \over r  \right)
-\end{equation} 
-where the approximate inequality holds deep inside the atmosphere ($r \ll \RB$) for any $n_{\rm fit} \gtrsim 1$.
-However the choice of boundary condition does have an order unity effect on the density near the Bondi radius.
+\subsection{EOS Tables}
+\emgr{compare polytrope vs. real EOS, this could go first or last.}
 
-The mass of the atmosphere typically determined by integrating the density profile from the core to the Bondi radius.
-Planets can attract massive atmospheres if $\theta\co \equiv R_B/R\co \gg 1$.
-In this case
-\begin{equation} \label{eq:MatmISO}
-M_{\rm iso} \approx 4 \pi \rho\di {R\co^4 \over R_B} e^{R_B/R\co} = 4 \pi \rho\di {R\co^3 \over \theta\co} e^{\theta\co}\, .
-\end{equation} 
-This result is the leading order term in a series expansion.
-Furthermore, because the atmospheric scale-height at $R\co$ is $H_\rho = |dr /d\ln\rho| = R\co^2/R_B$, the result is intuitively the correct order of magnitude.
-
-From \Eq{eq:MatmISO} the ratio of isothermal atmosphere to core mass is 
-\begin{equation} 
-{M_{\rm iso} \over M\co} = 3 {\rho\di \over \rho\co}{R\co \over R_B} e^{R_B/R\co} %= {\theta_o \over \theta\co^2} e^{\theta\co}
-\end{equation} 
-While  $\rho\di \ll \rho\co$, the exponential dependence allows the atmosphere to become massive.  
-%A value of $\theta\co = R_B / R\co$ corresponds to a planet mass of
-%\begin{equation} 
-%M\co = \sqrt{3 \over 4 \pi \rho\co} \left(\theta\co \mathcal{R} T \over G \right)^{3/2} 
-%\end{equation} 
-As a concrete example, consider a disk with $T = 60$ K and $\rho\di = 3 \times 10^{-11}$ (in cgs) and a core density $\rho\co = 2$.
-The atmosphere self-gravitating for $\theta\co = 27.1$ or $M\co > 0.046 M_\oplus$.
-The fact that very low mass cores can have very massive isothermal atmospheres is well known \citep{Sas89}.
-However such solutions are unrealistic because there is inadequate time to radiate away all the atmosphere's thermal energy.
-
-We not consider the mass exterior to the Bondi radius.
-For a meaningful evaluation we only include the mass coming from the overdensity relative to the background density.
-The resulting external mass for an isothermal atmosphere is
-\begin{subeqnarray}
-M_{\rm ext} &=& 4 \pi \int_{\RB}^{r_{\rm fit}} (\rho - \rho\di) r^2 dr \\
-&=& M\co \theta\co \int_1^{n_{\rm fit}} 3 \left[ \exp \left({1 \over x} - {1 \over n_{\rm fit}}\right) - 1 \right] x^2 dx  \nonumber \\
-&\equiv& M\co \theta\co I(n_{\rm fit})
-\end{subeqnarray} 
-where the dimensionless integral $I(n_{\rm fit})$ obeys the limits $I(1) = 0$ and $I \rightarrow n_{\rm fit}^2/2$ as $n_{\rm fit} \rightarrow \infty$.
-Since this external mass does not converge the choice of an outer boundary does matter in principle.
-In practice however the assumption that   $r_{\rm fit} = R_{\rm H}$ limits $n_{\rm fit}$ to modest values
-\begin{equation}
-n_{\rm fit} = {R_{\rm H} \over \RB} \approx 1.3 {\aun{10}^{4/7} \over \mcn{10}^{2/3}}{F_T \over  m_\ast^{1/3}}
-\end{equation} 
-since for instance $I(2) = 1.1$ these modest $n_{\rm fit}$ values will only produce a small external mass.
-
-However the effect is still worth including for the effect on the interior mass.  
-\subsection{Temperature Contrast at Convective Boundary}
-We now show that the temperature contrast between the convective boundary, $T\cb$, and the ambient disk $T\di$, is modest.  We express the radiative lapse rate
-\begin{equation}\label{eq:delrad}
-\delrad \equiv {d \ln T \over d \ln P} = {3 \kappa P \over 64 \pi  G M \sigma T^4} L = \nabla\di {(P/P\di)^{1+\alpha} \over (T/T\di)^{4-\beta}}
-\end{equation}
-where $M$ is the sum of the core mass and the atmosphere mass below the pressure level $P$.
-If the mass in the radiative zone is small, then we can hold $M = M\co + M_{\rm conv}$ fixed at the sum of core and convective zone masses.
-With this assumption and powerlaw opacity, we get a constant value for $\nabla\di$.
-The temperature profile then integrates to
-\begin{equation}\label{eq:radTP}
-\left(T \over T\di\right)^{4-\beta} - 1 = {\nabla\di \over \nabla_\infty} \left[\left({P \over P\di}\right)^{1-\alpha} - 1 \right] \, ,
-\end{equation} 
-where $\nabla_\infty = (1+\alpha)/(4-\beta)$ is $\delrad$ for $T ,P \rightarrow \infty$.\footnote{Our definitions of $\nabla\di$ and $\nabla_\infty$ are precisely opposite to R06, but consistent with other works and the general convention of labeling a quantity $f$ in the disk as $f\di$.}
-We  apply \Eqs{eq:delrad}{eq:radTP} at the convective boundary $\delrad = \delad$ under the assumption that the pressure there is $P\cb \gg P\di$.
-The resulting temperature contrast at the convective boundary is
-\begin{equation}\label{eq:Tcb}
-\chi \equiv {T\cb\over T\di} \simeq \left(1 - {\nabla_{\rm ad} \over \nabla_\infty}\right)^{-\frac{1}{4-\beta}} \, .
-\end{equation} 
-For  the low temperatures in protoplanetary disks, opacity is dominated by dust with $\alpha = 0$ and $\beta \approx 2$ or $1$, giving $\chi \approx 1.5$ or 1.9, respectively, assuming  $\delad = 2/7$.
-
-\subsection{Location of Convective Boundary}
-The pressure at the convective boundary  follows from \Eqs{eq:delrad}{eq:Tcb} as
-\begin{equation}
-{P\cb\over P\di} \simeq \left({\delad/\nabla\di \over 1 - \delad/\nabla_\infty}\right)^{1 \over 1 + \alpha}
-\end{equation} 
-This pressure contrast can be quite large due to the smallness of $\nabla\di$ in low luminosity atmospheres.
-
- It is also useful to obtain a relation between $T$ and $P$ that eliminates $\nabla\di$ in favor of $P\cb$:
- \begin{equation}\label{eq:TP}
-{T \over T\di} = \left\{1 + {1 \over {\nabla_\infty \over \delad} - 1} \left[ \left({P \over P\cb}\right)^{1-\alpha} -  \left({P\di \over P\cb}\right)^{1-\alpha}  \right] \right\}^{1 \over 4-\beta}\, .
-\end{equation}
-
-We can determine the radius of the convective boundary $r\cb$ from the hydrostatic balance equation as 
-\begin{equation}\label{eq:RCBint}
-{R_B \over r\cb} = \int_{P\di}^{P\cb} {T \over T\di} {dP \over P}\, .
-\end{equation} 
-An isothermal atmosphere gives a simple logarithmic dependence on $P\cb$.  However using \Eq{eq:TP} in the integral gives
-\begin{equation}\label{eq:Rcb}
-{R_B \over r\cb} = \ln \left(P\cb \over P\di \right) - \ln \theta \, ,
-\end{equation} 
-with an extra correction term, $\theta < 1$.
-In the $P\cb \gg P\di$ limit the correction term is a constant that depends on $\alpha$, $\beta$ and $\delad$.
-The form of $\theta$ was chosen to allow us to express
-\begin{equation}\label{eq:PcbRcb}
-P\cb = \theta P\di e^{R_B/r\cb} \, .
-\end{equation}   
-A simple analytic expression for $\theta$ is not possible.
-
-\begin{deluxetable}{cccccc}  % <--- column justification (center/left/right)
-\gdef \numcols {6}
-\tablecolumns{\numcols}
-\tablecaption{Parameters Describing Structure of Radiative Zone.}
-\tablehead{   \multicolumn{\numcols}{c} {$\gamma = 7/5$ ($\delad = 2/7$), $\alpha = 0$} }  
-\startdata
- $\beta$   		 &1/2  	& 3/4 &1   		& 3/2  		& 2   \\
- $\nabla_\infty$ & 2/7 \tablenotemark{a}  	&  4/13	& 1/3 	& 2/5 	 	& 1/2 \\
- $\chi$ 		 & \nodata &  2.25245 &1.91293 	& 1.65054 	& 1.52753 \\
- $\theta $  		 &\nodata   & 0.145032	&0.285824   &0.456333   & 0.556069   \\
-\enddata
-\tablenotetext{a}{Since $\delad = \nabla_\infty$ there is no convective transition at depth for this case.}
-\end{deluxetable}
-
-The T-P profile can also be combined with the equation of hydrostatic balance to relate the radius and pressure of the convective boundary.
-If the convective boundary is well inside the Bondi radius, i.e.\ $r\cb\ll R_B$ and $P\cb\gg P\di$ then
-\begin{equation}\label{eq:PcbRcb}
-P\cb = \theta P\di e^{R_B/r\cb} \, .
-\end{equation}    
-The order unity constant $\theta < 1$ accounts for the fact that the radiative layer is not perfectly isothermal, as $\theta$ would be exactly one in that case.
-The value of theta depends on the equation of state and the pressure and temperature dependence of the opacity, i.e.
-the $\alpha (=0)$ and $\beta$ values.
-We find $\theta(\beta = 2) \approx 0.556$ and $\theta(\beta = 1) = 0.286$.
-Closed form expressions for these integrals are too cumbersome to be useful.
-
-\section{Virial Theorem}
-The virial theorem for self-gravitating atmospheres can be applied to protoplanetary atmospheres.    The virial theorem is derived as usual by integrating the equation of hydrostatic balance to give
-\begin{eqnarray}\label{eq:virial}
-E_G= -3 \int_{M\co}^{M} {P \over \rho}dm + 4 \pi \left( R^3 P\surf - r\co^3 P\co\right) 
-\end{eqnarray} 
-where the gravitational energy
-\begin{equation}\label{eq:EG}
-E_G = -\int_{M\co}^{M} {G m \over r} dm \, .
-\end{equation} 
-At the outer surface $m = M$, $r = R$ and other values are given a subscripted $M$.
-This outer surface can be evaluated anywhere (e.g.\ Bondi radius, Hill radius or RCB) that hydrostatic balance holds.
-The boundary terms include the effects of finite surface pressure and finite radius of the solid core.
-
- The integral on the RHS \Eq{eq:virial} is related to the internal energy.
-For an ideal gas, $P/\rho = \Rg T = (\gamma -1) u$, where $u$ is the internal energy per mass, and $\gamma = C_P/C_V$.
-For a polytrope with constant $\gamma$ we can express the virial theorem in terms of internal energy $E_i = \int u dm$ as
-\begin{equation}
-E_G = - \zeta E_i + 4 \pi \left(R^3 P\surf - r\co^3 P\co\right) 
-\end{equation} 
-where $\zeta \equiv 3 (\gamma -1)$.  In this case the total energy $W = E_G + E_i$ becomes
-\begin{subeqnarray}
-W &=& (1-\zeta)E_i + 4 \pi \left( R^3 P\surf  - r\co^3 P\co \right) \\
- &=& {\zeta - 1 \over \zeta} E_G + {4 \pi \over \zeta} \left( R^3 P\surf  - r\co^3 P\co \right)\, .
-\end{subeqnarray} 
-The surface pressure acts to unbind the atmosphere by increasing $W$.  While a realistic EOS is neither ideal nor polytropic, \Eq{eq:virial} is general.
-
-%\section{Adiabatic Atmosphere}
-%An adiabatic atmosphere with the entropy of the disk gas is only slightly more massive than the disk gas itself (inside the Bondi radius).
-
-\section{Cooling Models}
-An isolated sphere satisfies a simple global energy equation
-\begin{equation}\label{eq:simplecool}
-L = \Gamma - \dot{E}
-\end{equation} 
-where the luminosity is balanced by the rate of heat generation, $\Gamma$, and the rate at which total (gravitational plus thermal, at least) energy is lost.
-
-For a protoplanetary atmosphere embedded in a gas disk, the cooling equation is more complicated: 
-\begin{equation}\label{eq:EB}
-L = L\co + \Gamma- \dot{E} + e_{\rm acc} \dot{M}  -  P\surf {\p V\surf \over \p t} \, .
-\end{equation}
-We explain these extra terms and then derive the result.
-\Eq{eq:EB} assumes a well (but arbitrarily) defined surface at radius $R$ and mass $M$, which can evolve in time.
-Examples include the Bondi radius or the radiative-convective boundary.
-The luminosity $L$ represents the luminosity from that surface.
-The luminosity from the solid core is given by $L\co$, and includes planetesimal accretion energy  and radioactive decay.
-The core is assumed to have a fixed mass, $M\co$, and radius, $R\co$, to simplify the atmospheric calculation.
-The $\Gamma$ and $\dot{E}$ integrals are identical to \Eq{eq:simplecool}, and include the atmosphere from the core to the top boundary.
-Sources of direct heating that  contribute to $\Gamma$ include  atmospheric drag on sedimenting planetesimals and the dissipation of any atmospheric turbulence.
-
-Mass accretion at the rate $\dot{M}$ brings specific energy $e_{\rm acc} = u - GM/R$, the sum of internal and gravitational energies.
-The energy of accreted matter, $e_{\rm acc}$, is  zero at a modified Bondi radius, $R_{\rm out} = G M/u \approx  (\Rg/C_V) R_B$.
-For $R < R_{\rm out}$ ($R > R_{\rm out}$, respectively) the accreted mass has negative (positive) energy and is (un)bound.
- The final term in \Eq{eq:EB} gives the pressure work done on a surface mass element, so the partial time derivative of  volume holds the mass fixed.
-Our quasistatic models we neglect bulk kinetic energy.
-
-\subsection{Derivation of Global Energy Equation}
-We now derive the global energy equation (\ref{eq:EB}).
-We follow the simpler example in \S4.3 of \citet{KipWei94}, adding the effects of finite core radius, surface pressure and mass accretion.
-We assume that hydrostatic balance holds.
-Integrating the local energy equation from core to surface gives:
-\begin{subeqnarray}
-L - L\co &=& \int_{M\co}^M {\p L \over \p m} dm \\
-&=& \int_{M\co}^M \left(\epsilon - T {\p S \over \p t} \right)dm \\
-&=& \Gamma  - \int_{M\co}^M{\p u \over \p t} dm +  \int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm\slabel{eq:DLc}\, .
-\end{subeqnarray} 
-with $\Gamma = \int \epsilon dm$ the integral of the direct heating rate.
-
-In what follows, we must carefully distinguish between partial time derivatives, $\p / \p t$, (performed at fixed mass) and total time derivatives, denoted with overdots (which include the effect of mass accreted through the outer boundary).
-% or $d/dt$.
-For instance the evolution of surface radius, $R$, evolves as  
-\begin{equation}\label{eq:Rdot}
- \dot{R} = {\p R \over \p t} + {\dot{M} \over 4 \pi R^2 \rho\surf}
-\end{equation} 
-where $\p R/\p t$ gives the Lagrangian contraction of surface mass elements, and $\dot{M}$ denotes mass accretion rate through the upper boundary.
-The subscript $M$  denotes quantities at the upper boundary of total mass $M$ (though it is omitted from $M$ and $R$).
-Similarly the volume, $V = (4 \pi/3)r^3$ and pressure at the surface evolve as
-\begin{subeqnarray}\label{eq:dot}
-\dot{V}_M &=&  {\p V_{\rm M} \over \p t} + {\dot{M} \over \rho_{\rm M}}  \\
- \dot{P}_M &=& {\p P_{\rm M} \over \p t} + {\p P_M \over \p m}\dot{M} \\
- &=&  {\p P_{\rm M} \over \p t} - {G M  \over 4 \pi R^4} \dot{M}\, .
-\end{subeqnarray} 
-For the purpose of this derivation we will hold the core mass fixed $\dot{M}\co = 0$ which further gives $\dot{P}\co = \p P\co / \p t$.
-
-To derive the global energy equation we must move the (partial) time derivatives in \Eq{eq:DLc} outside their integrals.  The internal energy integral follows simply from  Leibniz's rule as
-\begin{equation}\label{eq:udot}
-\int_{M\co}^{M(t)}{\p u \over \p t} dm = \dot{E}_i  -  \dot{M}u\surf\, .
-\end{equation}
-
-To evaluate the work integral, we derive a pair of expressions for the rate of change of gravitational energy.
-The time derivative of \Eq{eq:virial}  gives
-\begin{eqnarray}\label{eq:EGdot}
-\dot{E}_G &=& 3  \int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm -3 \int_{M\co}^M {\p P\over \p t}{dm \over \rho} \\
-&& -  3{P\surf \over \rho\surf} \dot{M}+ 3 \dot{P}\surf V\surf -3 \dot{P}\co V\co  + 3  P\surf {\dot{ V}\surf} \, . \nonumber
-%&&+ 4 \pi \left. \left(r^3 {\p P \over \p t}\right)\right|_{M\co}^M  + 3  P\di {\p V \over \p t}  - {G M \over R} \dot{M} \nonumber
-\end{eqnarray} 
-%where the volumes, $V\surf = 4 \pi R^3/3$ and $V\co = 4 \pi R\co^3/3$. 
-The first integral in \Eq{eq:EGdot} is the one we want, but the next one must be eliminated.  The time derivative of \Eq{eq:EG} (times four) gives
-\begin{subeqnarray}
- 4 \dot{E}_G &=&  -4 {G M \dot{M} \over R} + 4 \int_{M\co}^M {G m \over r^2}{\p r \over \p t} dm\\ 
-&=&   -4 {G M \dot{M} \over R} + 4 \pi \int_{M\co}^M r^3{\p \over \p m}{\p P \over \p t} dm \slabel{eq:4EGb} \\
-&=&  -4 {G M \dot{M} \over R} -3  \int_{M\co}^M {\p P\over \p t}{dm \over \rho} \slabel{eq:4EGc} \\
-&&+ 3 V\surf {\p P\surf \over \p t} -3 V\co {\p P\co \over \p t} \nonumber 
-\end{subeqnarray} 
-where \Eqs{eq:4EGb}{eq:4EGc} use hydrostatic balance  and integration by parts.
-
-%To eliminate the time derivates of pressure, we take the time derivative of the hydrostatic balance equation for $\p^2 P / \p m\p t$ and integrate over $4\pi r^3 dm$ (as in the virial equation derivation) to get
-%\begin{equation}\label{eq:dHBdt}
-%3 \dot{P}\surf V\surf -3 \dot{P}\co V\co -3 \int_{M\co}^M {\p P\over \p t}{dm \over \rho}  = 4 \dot{E}_G + 4{G M \over R} \dot{M}  \, .
-%\end{equation} 
-%Combining \Eqs{eq:EGdot}{eq:dHBdt} gives 
-%\begin{eqnarray}\label{eq:rhodot}
-%\int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm  &=& - \dot{E}_G - {4 \over 3}{G M\over R} \dot{M} + {P\surf \over \rho\surf} \dot{M} -  P\surf \dot{V}\surf  \, , \nonumber \\
-%&=&- \dot{E}_G - {4 \over 3}{G M\over R} \dot{M}  -  P\surf {\p V\surf \over \p t}  \, ,
-%\end{eqnarray} 
-%where the final step uses \Eq{eq:Rdot}.
-
-Subtracting \Eqs{eq:udot}{eq:4EGc} and rearranging for the desired integral gives
-\begin{eqnarray}\label{eq:PdVint}
-\int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm  &=&  - \dot{E}_G - {G M \dot{M} \over R} - P\surf {\p \dot{V}\surf \over \p t} \,  
-\end{eqnarray} 
-with \Eq{eq:dot} used to combine terms.  Combining \Eqsss{eq:DLc}{eq:udot}{eq:PdVint}, we reproduce \Eq{eq:EB} with the accreted energy $e_{\rm acc} \equiv u\surf - GM/R$.
-
-\section{Simplified cooling model}
-We consider an analytic cooling model that assumes an ideal gas polytrope for the convective region and neglects self-gravity.
-
-\subsection{Luminosity,  Energy \& Mass}
-The luminosity that emerges at the radiative convective boundary is
-\begin{equation} \label{eq:Lcb}
-L\cb = {64 \pi G M\cb \sigma T\cb^4 \over 3 \kappa P\cb } \nabla_{\rm ad} \approx L\di {P\di \over P\cb} % \left({P\di \over P\cb}\right)%^{1+\alpha}
-\end{equation} 
-for a dust opacity with $\kappa \propto T^\beta$ and
-\begin{equation} 
-L\di \equiv {64 \pi G M\cb \sigma T\di^4 \over 3 \kappa(T\di) P\di} \nabla_{\rm ad}\chi^{4-\beta}\, .
-\end{equation} 
-normalized to disk conditions, with $T = \chi T\di$ at the RCB.
-
-We adopt an opacity law
-\begin{equation}
-\kappa = 2 F_\kappa \left(T \over 100~{\rm K}\right)^\beta ~{\rm cm^2/g}
-\end{equation} 
-independent of pressure as appropriate for dust opacities.
-For $\beta = 2$ and $F_\kappa = 1$, this gives the \citet{bl94} opacity for icy grains.
-By varying $F_\kappa$, dust depletion or enhancement can be considered.
-Grain properties affect both $F_\kappa$ and $\beta$ which generally satisfies  $1/2 \lesssim \beta \lesssim 2$ (aside from discontinuities across sublimation regions), see \citet{SemHen03}.
-
-To make further analytic progress, we ignore the self-gravity in the convective zone, holding the mass fixed at $M\co$.
-The density profile of an adiabatic atmosphere follows from hydrostatic balance as
-\begin{equation}\label{eq:rhoconv} 
-\rho = \rho\cb \left[ 1 + {R_B' \over r} - {R_B' \over r\cb}  \right]^{1/(\gamma -1)}\, .
-\end{equation} 
-where we define an effective Bondi radius,
-\begin{equation}
-R_B' \equiv {G M\co \over C_P T\cb} = {\delad \over \chi} R_B
-\end{equation} 
-to simplify expressions.
-
-  Deep in the atmosphere, where $r \ll r\cb \lesssim R_b'$ the density profile is $\rho \propto r^{-1/(\gamma -1)}$.
-Since energy scale as $\rho r^2 \propto r^{(2\gamma -3)/(\gamma - 1)}$ only polytropes with $\gamma < 3/2$ (i.e.
-$\gamma = 7/5$, but not $\gamma = 5/3$) have the bulk of energy at the bottom of the atmospheres.
-We will thus focus on the $\gamma = 7/5$ case, even though dissociation occurs deep in real protoplanetary atmospheres.
-
-The total (thermal and gravitational) energy in an adiabatic atmosphere could be evaluated from the virial theorem.
-More simply, we use the result for temperature profiles in deep (but non-self-gravitating) convective regions:
-\begin{equation}
-T \approx {G M\co \over C_P r} = T\cb {R_B' \over r}\, .
-\end{equation} 
-The internal energy per unit mass of an ideal gas is thus $u = C_V T = (1 - \delad) G M\co/r$ and the specific energy  deep in the atmosphere is
-\begin{equation}
-e = e_g + u = -\delad {GM\co \over r} \, .
-\end{equation} 
-Thus the total energy for $\gamma < 3/2$ is thus
-\begin{eqnarray} 
-E &=& - 4 \pi \nabla_{\rm ad} G M\co \int_{R\co}^{r\cb} \rho r dr \\
-&\approx& - 4 \pi P\cb {R_B'}^{1 \over \nabla_{\rm ad}} \left(\gamma-1 \over 3 - 2 \gamma\right)  R\co^{2\gamma-3\over \gamma-1}  \\
-&\approx& - 8 \pi P\cb {R_B'^{7/2} \over \sqrt{R\co}}
-%&\approx& -0.31 P\cb {(R_B / \chi)^{7/2} \over \sqrt{R\co}}
-%&\approx& -0.31 P\cb {R_B'^{7/2} \over \sqrt{R\co}}
-\end{eqnarray} 
-where the final expression takes $\gamma = 7/5$. % and then revert to the standard Bondi radius $R_B = G M\co/(\Rg T\di)$ in terms of the disk temperature.
-
-The mass of the adiabatic atmosphere is given by
-\begin{eqnarray} 
-M_{\rm atm} &=& 4 \pi \int_{R\co}^{r\cb} \rho r^2 dr \\
-&=& {5 \pi^2 \over 4} \rho\cb {R_B'}^{5/2} \sqrt{r\cb} % \approx 0.54 \rho\cb {R_B'}^{5/2} \sqrt{r\cb} \\
-%&=& 0.54{R_B'^{3} \sqrt{\chi} \over  \sqrt{ \ln[P\cb/(\theta P\di)]}} \, .
-\end{eqnarray}
-in the limit $R\co \ll r\cb \ll R_B'$. The mass in the isothermal region is $M_{\rm iso} \sim 4 \pi \rho\cb r\cb^4/R_B \ll M_{\rm atm}$, and can be neglected.
-
-We can eliminate $r\cb$ with \Eq{eq:PcbRcb}.  The ratio of atmosphere to core mass is then  
-\begin{equation} \label{eq:crit}
-{M_{\rm atm} \over M\co} \approx {P\cb /P_M \over  \sqrt{\ln[ P\cb/(\theta P\di)]}}
-\end{equation} 
-where we introduce a characteristic pressure
-\begin{equation} 
-P_M \equiv {4 \delad^{3/2} \over 5 \pi^2 \sqrt{\chi} } {G M\co^2 \over {R_B'}^4}\, .
-%P_M \approx {1.85 \over \sqrt{\chi}} {G M\co^2 \over {R_B'}^4}\, .
-\end{equation}
-
-For the atmosphere to become self-gravitating, with $M_{\rm atm} = M\co$, we thus require
-\begin{equation} \label{eq:Pcbc}
-P\cb = \xi P_M
-\end{equation} 
-where  the logarithmic factor
-\begin{equation}\label{eq:xi}
-\xi \equiv \sqrt{\ln[ P\cb/(\theta P\di)]} = \sqrt{\ln[ \xi P_M / (\theta P\di)]}
-\end{equation} 
-is found by numerically solving the above transcendental equation.
-The physical solution has $\xi >1$, but typically order unity.
-Numerical solutions exit for $\xi \ll 1$, but are unphysical as they imply $P\cb < P\di$.
-%  The physical origin of this logarithm is the exponential increase of pressure with depth in the radiative zone.
-
-\subsection{Cooling Time}
-Neglecting surface terms, the total time to cool the atmosphere from an initially fully adiabatic state is
-\begin{eqnarray} 
-t_{\rm  cool} &=& -\int {d E \over L} = - \int_{P\di}^{P\cb } {d E/d P\cb \over L} dP\cb \\
-&\approx& 4 \pi {P\cb^{2} \over P\di} {R_B'^{7/2} \over L\di \sqrt{R\co}} \label{eq:tcool}\, ,
-%&\approx& {0.16 }{P\cb^{2} \over P\di} {R_B'^{7/2} \over L\di \sqrt{R\co}}
-\end{eqnarray} 
-where we neglect self-gravity and take $M\cb = M\co$ in $L\di$.
-
-The cooling timescale  for an atmosphere to become self-gravitating is found from \Eqs{eq:tcool}{eq:Pcbc} as
-\begin{eqnarray} 
-t_{\rm cool} &\approx & 2 \times 10^8 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2  \over \left(\mc \over 10 \right)^{5/3} \left(\au \over 10\right)^{15 \over 14}} \yr
-\end{eqnarray} 
-for $\beta = 2$.  
-Clearly this timescale is too long, and this is likely due to missing physics of self-gravity and EOS.
-Nevertheless the trends are informative.
-A reduction in opacity naturally gives faster cooling, as long as the optically thick assumption holds.
-Lower disk temperatures also give faster cooling.
-Even though higher temperatures give higher luminosities, they also give higher dust opacities and lower the Bondi radius and gas density.
-In balance higher temperatures suppress cooling in this model.
-For different $\beta$ values, the temperature dependence is $t_{\rm cool} \propto F_T^{1/2 + \beta}$.
-The cooling timescale only weakly depends on disk mass or pressure, via the  logarithmic factor $\xi$.
-This weak dependence is a consequence of cooling coming from the convective boundary.
-Note that the scaling laws do not reflect the changes to $\xi$ which remains order unity and is here given the appropriate value for the nominal parameters.
-
-\subsection{Critical Core Mass}\label{sec:critmass}
-%We use the results of the previous section to derive the minimum core mass that can form a giant planet by gas accretion.
-Since our structure calculations do not include the self-gravity of the atmosphere, we simply assume that runaway growth occurs once the atmosphere's mass (as calculated in the absence of self-gravity) reaches the core mass, $M_{\rm atm} \ge M_{\rm c}$.
-This approximation was used by R06 and is consistent with detailed evolution models (refs).
-
-We can define a critical more mass as that which gives a  self-gravitating atmospheres within a typical gas disk lifetime:
-\begin{equation}
-t_{\rm cool} = 3 \times 10^6 \tau_{\rm cool} ~{\rm yr} \, ,
-\end{equation} 
-with $\tau_{\rm cool}$ a scaling factor.  The resulting critical core mass is
-\begin{equation}\label{eq:Mcrit}
-M_{\rm crit} \approx 100 {F_T^{3/2} F_\kappa^{3/5}   \left(\xi \over 2.6 \right)^{6/5} \over \left(\au \over 10\right)^{9 \over 14}} \; M_\oplus
-\end{equation} 
-as with the cooling time, the critical mass is too large due to missing physics.
-The scaling with disk properties is similar, in fact all quantities are simply raised to the $3/5$ power.
-The $\xi$ value is changed to match the value for the nominal solution.
-The lower value reflects the fact that an atmosphere around such a massive core does not need to contract as much to be self-gravitating.
-
-The numerical values for a different opacity law do not change the numerical answers above very much.
-That is mostly because we have chosen a disk location where $T_o$ is not far from 100 K, where our opacity law is normalized.
-More generally a higher $\beta$ value is more favorable for core accretion at large distances due to the sharper drop in opacity.
-
-%We use \Eq{eq:tcool} to find $P\cb$, the pressure depth of the convective boundary after a cooling time.   Relative to the disk pressure we find
-%\begin{equation}
-%{P\cb\over P\di} =  2.0 \times 10^4 \left(a \over 5~{\rm AU}\right)^{2.0} \mc^{-7/6} {\sqrt{\tau_{\rm cool}} \over F  \sqrt{\kappa_\odot}}
-%\end{equation}
-
-%This seems totall wrong:
-%\begin{equation} 
-%{M\co \over M_\oplus} \approx 14 \left(a \over 5~{\rm AU}\right)^{0.19} \left[{\kappa_s (\xi/2)^3 f_T^4 \over F \sqrt{M_\ast}}\right]^{3/20}
-%\end{equation}
-
-\subsection{The Opacity Effect}
-A  lower opacity  could lower the core mass.
-Reducing the opacity by a factor of one hundred cuts the core mass by more than a factor of 10, specifically to 9 $M_\oplus$ for the parameters in \Eq{eq:Mcrit}.
-The reduction is not  as strong as the nominal scaling would imply, $0.01^{3/5} \approx 0.06$, because $\xi$ increases.
-
-Even with significantly lower opacities, radiative diffusion remains a good approximation at the convective boundary.  We estimate the optical depth as (for $\beta = 2$):
-\begin{equation}
-\tau\cb \sim {\kappa P\cb \over g} \sim 7 \times 10^4 {F_T^4 F_\kappa \over \left(\mc \over 10 \right) \left(\au \over 10\right)^{12 \over 7}} 
-\end{equation} 
-where $P\cb \sim P_M$ for a self-gravitating atmosphere and $g \sim G M\co/R_B^2$, with both approximation good to within the the order unity factor $\xi$.
-Clearly $\tau\cb \gg 1$ even for $F_\kappa \lesssim 0.01$ out to very wide separations.
-
-A hotter disk would increase core masses.
-Instead of our passive disk model, adopting the standard Hayashi temperature profile would increase core masses by $\sim 50\%$.
-A hotter accretion phase would further increase core masses, but such phases are presumably short lived.
-
-\subsection{Surface Terms}
-We now check the relevance of the neglected surface terms in \Eq{eq:EB}.
-We already showed that accretion energy can be exactly eliminated by choosing an outer boundary near the Bondi radius.
-We now show that accretion energy is also a small correction at the RCB.
-A rough comparison, (ignoring terms of order $\xi$) of  accretion luminosity vs.\ $\dot{E}$ gives
-\begin{equation}
-{G M \dot{M} \over R \dot{E}} = {G M  \over R}{dM \over dE} \sim {G M\co^2 \over R_B E}{P\cb \over P_M} \sim \sqrt{R\co \over R_B} \ll 1
-\end{equation} \, ,
-where we assume $P\cb \sim P_M$  for a massive atmosphere.
-Accretion energy at the protoplanetary surface is thus very weak for marginally self-gravitating atmospheres, and even weaker for lower mass atmospheres.
-A similar scaling analysis shows that the work term, $P\surf \p V\surf/\p t$ is similarly weak.
-Nevertheless numerical calculations should include, or check, the importance of these surface terms in more realistic self-gravitating atmospheres.
-
-\section{Hydrogen Dissociation}
-The dissociation of molecular Hydrogen deep in the atmospheres of accreting protoplanets plays a significant role in the energetics of core accretion.
-In the high density regions $r  \ll r\cb$ of a convective atmosphere, the thermal plus gravitational energy scales as
-\begin{equation}
-dE = -4 \pi \nabla_{\rm ad}^{1/\nabla_{\rm ad}} \rho\cb R_B'^{1/(\gamma-1)} r^{\frac{2\gamma - 3}{\gamma - 1}} {dr \over r}
-\end{equation} 
-If $\gamma < 3/2$ then the main contribution to the energy is at the bottom of the atmosphere, i.e.\ the core.
-This is the case for a diatomic ideal gas ($\gamma = 1.4$) or a solar mixture of hydrogen and helium ($\gamma \approx 1.43$).
-However a monatomic gas has $\gamma = 5/3 > 3/2$.
-In this case, the atmosphere's energy is no longer concentrated near the bottom, but will be concentrated near the top of the convective zone.
-
-A likely structure is an atmosphere that is dissociated near the base, but becomes molecular near the top of the convective zone.
-In this case the atmosphere's energy budget would be concentrated near the atomic to molecular transition.
-
-The energy required to dissociate a hydrogen molecule, $I = 4.467$ eV can be significant to the overall energy budget.
-Since
-\begin{equation}
-{I \over \nabla_{\rm ad} G M\co (2 m_{\rm H})/r} \approx 3 \left(M\co \over 10 M_\oplus \right)^{-2/3} {r \over R_{\rm c}}
-\end{equation} 
-we see that this energy is always relevant.
-
-We can use the Saha equation to determine where dissociation is significant,
-\begin{equation}
-{n_{\rm H}^2 \over n_{\rm H_2}} = \left(\pi m_{\rm H} k T \over h \right)^{3/2} e^{-I/(kT)}
-\end{equation} 
-We introduce a reaction coordinate $\delta$ so that $n_{\rm H} = 2 \delta n_o$ and $n_{\rm H_2} = (1-\delta) n_o$ with $n_o = \rho X/(2 m_{\rm H})$ the number density when all hydrogen is molecular.
-We express equilibrium as
-\begin{equation}
-{\delta^2 \over 1-\delta} = f_\mu {P_{\rm diss}(T) \over P}
-\end{equation} 
-with the characteristic pressure below which dissociation occurs is
-\begin{equation}
-P_{\rm diss} = {\left(kT\right)^{5/2} \over 4} \left( \pi m_{\rm H} \over h^2\right)^{3/2}  e^{-I/(kT)}
-\end{equation} 
-and the order unity factor
-\begin{equation}
-f_\mu = 2\delta + (1-\delta) + Y/2 + Z/\mu_Z
-\end{equation} 
-accounts for variations in the mean molecular weight with dissociation.  (Take $\mu_Z = 31/2$, but not too significant.)
-
-Thus dissociation occurs where $P \lesssim P_{\rm diss}(T)$.
-At disk temperatures (say 150 K) the dissociation pressure is negligibly small ($\sim 10^{-141}$ dyne cm$^{-2}$) and no dissociation occurs.
-However at core temperatures the dissociation pressure is quite large especially for massive cores.
-Dissociation is guaranteed.
-
-%\section{What is the Core Instability}
-%It is often argued that the critical core mass arises from the inability of a massive atmosphere to maintain hydrostatic equilibrium.
-The analytic version of this argument in S82 makes several assumptions.
-In particular self-gravity is neglected in some places, but included in others.
-Presumably the assumptions are reasonable, because numerical  structure models of Mizuno and Papaloizou \& Terquem give similar results.
-% The physical nature of the core instability remains unclear (to me at least).
-%
-%One possibly illuminating test problem would be the integration of the S82 problem (raditive diffusion at constant opacity) with a consistent inclusion of self-gravity.
-The integration solves the structure equations for $dT/dr$, $dP/dr$ (or $d \rho /dr$) and $dM/dr$.
-With self-gravity it's hard to avoid the numerical nature of the problem with the boundary conditions for $M$ applying at the core, but for $T$ and $P$ in the disk.
-%
-%One question is how self-gravity changes the temperature profile (and this is relevant for adiabatic profiles as well).
-Without self gravity, the optically thick result is $T = G M\co/(4 R_g R\co)$.
-S82 assumes the core mass is replaced by the total mass.
-This may be reasonable in an average sense?
-
-\section{Structure Models}
-%We propse to numerically solve for the evolution of the atmosphere of a protoplanet that cools in the absense of planetesimal accretion.
-Thus all heat loss comes form the atmosphere.
-Since much of the atmosphere is convective most models require a mixing length model to calculate the luminosity.
-Motivated by analytic models, the goal is to avoid the need for a mixing length model by simply calculating the radiative luminsotiy $L$ at a consistently determined RCB.
-The goal then is to construct solutions along a sequence of entropy that gradually decreases from the disk entropy.
-From a self-consistently determined luminosity, the cooling timescales would then be given consistently.
-\footnote{I thought about fixing $L$ letting $S$ adjust, but this seems problematic if (for instance) $L$ is double-valued.}
-%
-%A Henyey method would probably work best.
-Practical concerns include a good initial guess and ``root finding" to determine $L$ at a given $S$.
-Radial windows that cause multiple convective zones are also a concern.
-
-The atmospheric structure equations  in pressure units are
-\begin{subeqnarray}\label{eq:struct}
-{\p m \over \p P} &=& - {4 \pi r^4 \over G m} \\
-{\p r \over \p P} &=& - { r^2 \over G m \rho} \\
-{\p T \over \p P} &=&  \nabla {T \over P} \\
-{\p L  \over \p P} &=& {\p m \over \p P} \epsilon\, .\slabel{eq:dL}
-\end{subeqnarray} 
-When the atmospheric luminosity is carried by radiative diffusion, the lapse rate $\nabla$ is
-\begin{equation}
-\delrad ={ 3 \kappa P \over 64 \pi G m \sigma T^4}L\, .
-\end{equation} 
-We assume efficient convection when $\delrad > \delad$ so that $\nabla = \min(\delad, \delrad)$.
-The local the local energy generation, $\epsilon$, includes heating from  gravitational contraction
-\begin{equation}\label{eq:epsg}
-\epsilon_g = -T {\p S \over \p t}\, .
-\end{equation} 
-Other sources of local heating, such as drag on infalling planetesimals could be specified but are not considered here.
-To solve these equations we must supply opacity law for $\kappa$ and an equation of state (EOS) to give $\rho$ and $\delad$ as functions of $T$ and $P$ and the assumed composition.
-Note that the luminosity and opacity do not affect the structure of convection zones, but do affect their boundaries.
-If we can assume that the atmospheric luminosity is primarily generated in the convective zone, then the time dependence in the structure equations can be neglected.
-
-Boundary conditions for the problem exist both at the core and at the outer accretion radius.
-At the core, $m = M\co$, $r = R\co$, $L = L\co$.
-At the top boundary $T = T\di$ and $P = P\di$ set by conditions in the disk midplane.
-In addition to these four boundary conditions, an initial solution is required for the time dependence that arises when  \Eq{eq:epsg} is included and allowed to affect the radiative zone structure.
-
-\subsection{Shooting Method (1a)}\label{sec:shoot1a}
-This is the first (and currently only) method we have numerically developed.
-It is not described yet!
-It is similar to method 1b below, but holds entropy values, $S_i$ fixed instead of mass (or luminosity).
-In brief, we obtain solutions by varying the free thermodynamic variable at the core until the radius at the top of the convective boundary matches the value in \Eq{eq:Rcb}.
-
-\subsection{Shooting Method (1b)}\label{sec:shoot1b}
-We here describe an algorithm to obtain static solutions by integrating downward from the top of the convective zone.
-As in \S\ref{sec:shoot1a} this method uses simplified solutions for the  radiative zone that neglect self gravity and luminosity generation.
-With this method we can obtain an evolutionary sequence at fixed values of either mass or luminosity, with the other quantity optimized to get a converged solution.
-Fixing the mass intervals has the advantage that total mass increases monotonically, while the luminosity is known to decrease to a minimum value and then increase (as with entropy in \S\ref{sec:shoot1a}).
-
- The algorithm proceeds as follows:
-\begin{enumerate}
-\item Guess a value for either the luminosity, $L_{\rm guess}$, or mass, $M_{\rm guess}$, at the top of the convective zone (whichever is not being held fixed).
-From this guess get estimates of $P\cb$ from \Eq{eq:Lcb}, $T\cb$ from \Eq{eq:Tcb} and $r\cb$ from \Eq{eq:Rcb}.
-\item Integrate the structure equations from the convective boundary to the core, thereby obtaining a model value of the core mass. 
-\item Repeat,  varying either $L_{\rm guess}$ or  $M_{\rm guess}$ in step 1 until the core mass in the atmospheric solution matches the assumed value.
-\end{enumerate}
-
-Solutions for a range of masses (and luminosities) can be connected in an evolutionary sequence using the global energy equation that we derive from the virial theorem.
-
-For simplified radiative zone solutions, shooting from the top has the advantage that a monotonic sequence of mass can easily be contructed.
-On the other hand, integrating from the bottom can be more numerically efficient because the tabulated values of thermodynamic variables along a single adiabat can be used for each solution.
-When shooting from the top, the variation of the luminosity (or mass) requires integration along different adiabats, with greater interpolation costs.
-(But computational costs aren't really a big limitation.?)
-
-\subsection{Shooting Method (2)}\label{sec:shoot2}
-%At this writing, we are using a shooting model (not described here) which uses the radiative convective boundary (RCB) as the fitting point.
-The advantage of this approach is that a 1D (Newton Raphson) optimization of a thermodynamic variable at the core (such as temperature at fixed entropy) can be used to find a model.
-A limitation of this approach is that the structure of the radiative zone must be computed analytically, making it difficult to include detailed physics, such as variable opacities and self-gravity (because the total mass of the atmosphere is unknown \emph{a priori}).
-To overcome these obstacles we consider here a shooting method that extends from the core to the disk and the fitting is done at the outer boundary.
-The price of this more detailed model is that a 2D optimization must be performed which includes finding the self-consistent luminosity.
-
-%In traditional stellar models it is difficult to shoot to the outer boundary because of the numerical instability of very low pressures, i.e.\ small numerical errors in the high density region propagate like a whip (Hansen \& Kawaler; Kippenhahn \& Weigert).
-These numerical instabilities should (hopefully) be less severe in the protoplanetary problem due to the finite core radius (and lack of central divergences) and the finite bounding pressure of the disk.
-
-Here we describe a method that allows us to add the self-gravity of the radiative zone.
-While a time dependent model could be developed (see \S\ref{sec:timedep}),  we first describe the method for constant radiative zone luminosity.
-As in \S\ref{sec:shoot1b} we integrate inwards, this time starting from the disk.
-The outer boundary conditions are $T = T\di$,  $P = P\di$,  $M = M_i$ where $i$ is simply a label for each evolutionary stage,
-\begin{equation}
-r = r_{\rm fit} = n_{\rm fit}{ G M_i \over \Rg T\di}\, ,
-\end{equation} 
-which places the outer radius at a finite number, $n_{\rm fit} > 1$, of Bondi radii.
-Analytic solutions like \Eq{eq:rhoiso} show that the precise choice of fit radius has negligible effect on the atmospheric solution (R06).
-
-To find a solution we must vary $L = L_{\rm guess}$,  though as in \S\ref{sec:shoot1b} we could fix $L$ and vary  $M$.
-An inward integration for a trial $L_{\rm guess}$  gives the value of the core radius $R_{\rm c, guess}$ implied by the trial solution.
-We adjust $L_{\rm guess}$ until $R_{\rm c, guess}$ converges to the actual $R\co$.
-
-With atmospheric self-gravity, shooting outwards is more difficult because two parameters must be varied.
-It is impossible to fix the total mass shooting from the bottom, but we can try to construct models for choices of $L_i$.
-In this case,  both $P\co$ and $T\co$ must be varied to match $T = T\di$ and $r = r_{\rm fit}$ (using the total mass $M$ from the trial integration itself) at the top, which is defined by $P = P\di$.
-Alternatively one could try to find solutions with fixed values of entropy $S = S_i$ for the interior adiabat as in \S\ref{sec:shoot1a}.
-This gives only one thermodynamic variable to specify at the core, since the trial choice of $T\co$ gives $P\co$ at fixed entropy.
-However trial values of $L\cb$ must also be chosen, and there remain two quantities to vary ($L\cb$ and the core thermodynamic variable) to match the two conditions at the upper boundary.
-With only a single variable to optimize, downward integration is more efficient.
-
-\subsection{Time dependent shooting}\label{sec:timedep}
-We now consider how to add time dependence to the method of \S\ref{sec:shoot2}.
-\emph{For now this is just a sketchy description.}  First a static solution must be obtained for some small mass.
-Then the next, slightly larger mass should be chosen.
-A solution now requires variation of two quantities, the luminosity and the assumed time difference between the two states.
-The time difference is used to calculate the heating in the convective zone using \Eq{eq:epsg}.
-Convergence occurs when the atmosphere solution produces both the correct total mass and the timestep matched the global energy change between the two states.
-
-%\begin{eqnarray} 
-%y(\xi\co, L)  &=&  T(P\di; \xi\co, L) - T\di\\
-%z(\xi\co, L) &=& r(P\di; \xi\co, L) - n_{\rm fit}{ G m_{\rm B}(\xi\co, L) \over \Rg T\di}
-%\end{eqnarray} 
-%then the boundary conditions are met when $y = z = 0$.
-The 2D Newton-Raphson method can be used to search for the optimal values of $\xi\co$ and $L$.
-The mass enclosed by the Bondi radius, $m_{\rm B}$, is where the mass first exceeds $r \Rg T\di/G$.
-In practice $m_{\rm B}$ will probably differ only slightly from $m(P\di)$ because of the low density of disk material, but this can be checked numerically.
+\section{Conclusions}
 
@@
 \acknowledgements
-%Portions of this project were supported by the {\it NASA} {\it Astrophysics Theory Program} and  {\it Origins of Solar Systems Program}  through grant NNX10AF35G.\\
+\emgr{Acknowledge Ruth, Scott, Jonathan Fortney \& Didier Saumon.}  Portions of this project were supported by the {\it NASA} {\it Astrophysics Theory Program} and  {\it Origins of Solar Systems Program}  through grant NNX10AF35G.
+\emgr{JILA, Ruth's grants?}\\
+
 \appendix
-\section{Self-Gravitating Isothermal Atmosphere}
-
-For an isothermal atmosphere, hydrostatic equilibrium gives:
-\begin{equation} \label{eq:HBiso}
-r^2 {d \ln \rho \over d r} = -{G \over \mathcal{R} T} \left(M\co + 4 \pi \int_{R\co}^r \rho(r') r'^2 dr'\right) 
-\end{equation} 
-This integro-differential equation is subject to the boundary condition that the density is $\rho\di$ at large distances.
-This boundary condition can be applied at $n_B R_B$, where the Bondi radius $R_B = GM\co/\mathcal{R} T$.
-The number $n_B > 1$ does not affect the solution since the gravitational perturbation from the planet is weak outside the Bondi radius.
-
-Radial differentiation of \Eq{eq:HBiso} gives:
-\begin{equation} 
-x^2 {d^2 S \over dx^2} + 2x {d S \over dx} = - \theta_o e^S x^2
-\end{equation} 
-where $S = \ln(\rho/\rho_o)$, $x = r/R\co$ and $\theta_o = 4 \pi G \rho\di R\co^2 /(\mathcal{R} T)$.
-Since $\theta_o$ is the squared ratio of the core radius to the disk's Jeans length, $\theta_o \lll 1$.
-The outer boundary condition now reads $S(n_B \theta\co) = 0$ where $\theta\co = R_B/R\co > 1$.
-An inner boundary condition
-\begin{equation} 
-{d S \over dx}(x = 1) = - \theta\co
-\end{equation} 
-must also be satisfied.
-
+\section{Extending the EOS tables}\label{sec:EOSext}
+\section{Virial Equilibrium and Cooling}\label{sec:virial}
+\section{Analytic Cooling Model}
 \if\bibinc n

### 2013-01-22 python/Atmscripts/cooling_fixed_disklife.py
added plot + scripts: Mcrit vs distance for disk life 3Myrs

@@ -1,19 +1,34 @@
 
 
-a = numpy.linspace(10, 100, 10)
 
 from RadSGPoly.atmseries_poly import mcrit_disktime as mcd
 
 a = 10
+exec_file = 0
+save = 1
 
-Mcrit = 0 * numpy.ndarray(shape = (10), dtype = float)
+if exec_file == 1:
 
-for k in range(10):
-    execfile('coolingtime_plots_sg.py')
-    Mcrit[k] = mcd(Mc, t)
-    a = a + 10
+    Mcrit00 = 0 * numpy.ndarray(shape = (10), dtype = float)
+    Mcrit03 = 0 * numpy.ndarray(shape = (10), dtype = float)
 
+    for k in range(10):
+        Y = 0.0
+        execfile('coolingtime_plots_sg.py')
+        Mcrit00[k] = mcd(Mc, t)
+        
+        Y = 0.3
+        execfile('coolingtime_plots_sg.py')
+        Mcrit03[k] = mcd(Mc, t)
+
+        a = a + 10
     
 au = numpy.linspace(10, 100, 10)
-plt.semilogx(au, Mcrit, '-o')
+plt.semilogx(au, Mcrit00, '-o', label = r'$\nabla_{\mathrm{ad}}=2/7$, $Y=0.0$')
+plt.semilogx(au, Mcrit03, '-rs', label = r'$\nabla_{\mathrm{ad}}=2/7$, $Y=0.3$')
+plt.legend()
+plt.title(r'Critical core mass vs. distance for disk lifetime of 3 Myrs')
+if save == 1:
+    plt.savefig(userpath + \
+                '/figs/ModelAtmospheres/RadSelfGravPoly/Mcrit_vs_a_3Myrs.pdf')
 plt.show()

### 2013-01-22 python/Atmscripts/coolingtime_plots_sg.py

@@ -14,13 +14,16 @@ from RadSGPoly.cooling_poly import cooling_global as cg, cooling_local as cl, cr
 os.chdir(userpath + '/python/Atmscripts')
 
 delad = 2./7
-Y = 0.3
+
 
 load = 1
 cool = 1
 savefig = 0
+nofixeddisk = 1
 
-#a = 10.0
+if nofixeddisk == 0:
+    Y = 0.0
+    a = 100.0
 
 prms = params(Me, Re, a, delad, Y, gamma = gammafn(delad), R = Rfn(Y), \
               Cv = Cvfn(Y, delad), Pd = Pdisk(a, mstar, FSigma, FT), \
@@ -28,19 +31,35 @@ prms = params(Me, Re, a, delad, Y, gamma = gammafn(delad), R = Rfn(Y), \
 
 if load != 0:
 
-    atm5 = atmload('Mc1.0.npz', prms = prms)
-    atm6 = atmload('Mc2.0.npz', prms = prms)
-    atm7 = atmload('Mc3.0.npz', prms = prms)
-    atm8 = atmload('Mc4.0.npz', prms = prms)
-    atm9 = atmload('Mc5.0.npz', prms = prms)
-    atm10 = atmload('Mc6.0.npz', prms = prms)
-    atm11 = atmload('Mc7.0.npz', prms = prms)
-    atm12 = atmload('Mc8.0.npz', prms = prms)
-    atm13 = atmload('Mc9.0.npz', prms = prms)
-    atm14 = atmload('Mc10.0.npz', prms = prms)
-    #atm15 = atmload('Mc13.0.npz', prms = prms)
+    if Y == 0.3:
 
-Mc = numpy.linspace(1, 10, 10)
+        atm5 = atmload('Mc1.0.npz', prms = prms)
+        atm6 = atmload('Mc2.0.npz', prms = prms)
+        atm7 = atmload('Mc3.0.npz', prms = prms)
+        atm8 = atmload('Mc4.0.npz', prms = prms)
+        atm9 = atmload('Mc5.0.npz', prms = prms)
+        atm10 = atmload('Mc6.0.npz', prms = prms)
+        atm11 = atmload('Mc7.0.npz', prms = prms)
+        atm12 = atmload('Mc8.0.npz', prms = prms)
+        atm13 = atmload('Mc9.0.npz', prms = prms)
+        atm14 = atmload('Mc10.0.npz', prms = prms)
+
[…]

### 2013-01-24 python/RadSGPoly/shoot2.py
starting a nb for evolution runs
Still not happy with performance and should go back and work on that.  Had a problem (broke) when trying to go from 5.7 to 5.8 Me.  10 steps works (if slow) but errors vary quite a bit.

@@ -677,7 +677,7 @@ def evolve_dM(M, dM=0, xg=[], prevroot=0, prms=prms, prevgrid=1000,
         if abs(err0[1]) > abs(errg[1]):
             x0, err0 = xg, errg
             if verbose:
-                print "using supplied guess for M = " , M
+                print "using supplied guess for M = {:.3f} M_E".format(M/Me)
     if verbose:
         print "initial guess error ", err0
 #find time dependent solution based on guessed values

### 2013-02-05 tex/draft/draft.tex
updated paper draft

@@
-\documentclass[12pt, preprint,numberedappendix]{emulateapj}
+%\documentclass[12pt, preprint,numberedappendix]{emulateapj}
 %\documentclass[12pt, preprint]{aastex}
+\documentclass[apj]{emulateapj}
 
@@
 
-\newenvironment{packed_item}{
-\begin{itemize}
-  \setlength{\itemsep}{1pt}
-  \setlength{\parskip}{0pt}
-  \setlength{\parsep}{0pt}
-}{\end{itemize}}
+%\newenvironment{packed_item}{
+%\begin{itemize}
+%  \setlength{\itemsep}{1pt}
+%  \setlength{\parskip}{0pt}
+%  \setlength{\parsep}{0pt}
+%}{\end{itemize}}
 
@@
 \subsection{Model Assumptions}
-\emgr{Describe basic assumptions, perhaps in list format.
-Will need to describe basic scales and idea that atmosphere is 1D and matches smoothly onto disk.
-However we may want to separate these standard assumptions from our more unique ones that are part of the two layer model.}
+\emgr{ Define and describe basic length scales; Describe disk conditions: MMSN etc.
+Describe basic assumptions: 1D atmosphere, smooth matching onto the disk, atmosphere composition etc., perhaps in list format; Describe assumptions specific to our model: quasistatic, $L_{KH} \gg L_{acc}$, two-layer, constant $L$.
+Explain how our model is intermediary between purely static models with $L=L_{acc}$ and full time-dependent evolutionary models.
+Could be split into sub-subsections.}
 
 \subsection{Structure Equations}
-\emgr{Give in pressure coordinates.  Put EOS (polytrope and detailed) and opacity laws (noting that we only consider cool radiative zones with dust opacities).}
+\emgr{Give in radius coordinates.
+EOS (polytrope, various assumption for $\nabla_{\mathrm{ad}}$ and $\mu$) and opacity laws (noting that we only consider cool radiative zones with dust opacities).
+Give boundary conditions (probably not worth a whole subsection by itself)}
 
-\subsection{Boundary Conditions: Core and Disk}
+%\subsection{Boundary Conditions: Core and Disk}
 
@@
 
-\subsection{Radiative Zone Solutions}
-\emgr{Give analytic solution valid for power law opacity and lack of self-gravity.}
-
-\subsection{Matching at the Radiative-Convective Boundary (RCB)}
-\emgr{Include figure of an example atmospheric profile.}
-
-\subsection{The Gravothermal Catastrophe}
-\emgr{Show and explain double valued solutions, including caveat that upper branch breaks several model assumptions.}
+%\subsection{Radiative Zone Solutions}
+%\emgr{Give analytic solution valid for power law opacity and lack of self-gravity.}
+%
+%\subsection{Matching at the Radiative-Convective Boundary (RCB)}
+%\emgr{Include figure of an example atmospheric profile.}
+%
+%\subsection{The Gravothermal Catastrophe}
+%\emgr{Show and explain double valued solutions, including caveat that upper branch breaks several model assumptions.}
 
 \section{Cooling History}
-\emgr{start with basic idea of quasistatic contraction and $L = -\dot{E}$}
+\emgr{Start with basic idea of quasistatic contraction and $L = -\dot{E}$, then write out cooling equation and explain terms. Explain where it comes from (local cooling + virial equilibrium)}
 \subsection{Role of Finite Pressure}
- \emgr{Give more more general result, but defer derivation to \App{sec:virial}.}
-\subsection{Estimating a Critical Core Mass}
-\emgr{Describe how and why we use the time to gravothermal catastrophe as estimate of minimum core mass}
+ \emgr{Give more more general result, but defer derivation to Appendix \ref{sec:virial}. Perhaps not worth a subsection by itself.}
+%\subsection{Estimating a Critical Core Mass}
+%\emgr{Describe how and why we use the time to gravothermal catastrophe as estimate of minimum core mass}
+
+\section{Numerical method}
+\emgr{Explain the shooting technique, then how to obtain an evolutionary series from cooling model between subsequent static atmospheres. Show P-T, M-T, M-R etc. plots. Also L-t, L-M plots.}
+\subsection{The Gravothermal Catastrophe \emgr{or} {Entropy Minimum} \emgr{or} {Negative Luminosity} \emgr{or} Something else?}
+\emgr{Describe that time / luminosity becomes negative when $M_{atm} \sim M_c$.}
 
 \section{Critical Core Mass}
-\emgr{These are the take home results!}
+\emgr{These are the take home results! Define critical core mass (= minimum between mass doubling and entropy minimum). Explain the assumptions, show that growth is slowest then (M-t plot?)}
 
 \subsection{Role of Disk Temperature and Pressure}
+\emgr{Show how critical core mass changes as you vary one parameter while keeping the other constant.}
+
+\subsection{Core Mass vs. Cooling Time at Fixed Radius}
 
@@
 \subsection{Opacity Dependence}
+\emgr{Not sure what would go in this section, since we only use one opacity model.}
 
+%\subsection{EOS Tables}
+%\emgr{compare polytrope vs. real EOS, this could go first or last.}
+
+\section{Discussion}
+\subsection{Kelvin-Helmholtz Contraction vs. Planetesimal Accretion}
+\emgr{Compare and contrast our results with Rafikov's, show accretion rates plot; discuss why our fixed core mass assumptions are valid in the regime we are considering.}
+\subsection{Analytic Cooling Model}
+\emgr{Explain assumptions of analytic model but defer equations and derivations to Appendix \ref{sec:analytic}. Show numerical vs. analytic comparison plots and explain role of self gravity. }
 \subsection{EOS Tables}
-\emgr{compare polytrope vs. real EOS, this could go first or last.}
+\emgr{Mention and briefly describe Saumon et al. EOS tables as future work. }
 
@@
 \appendix
-\section{Extending the EOS tables}\label{sec:EOSext}
+%\section{Extending the EOS tables}\label{sec:EOSext}
 \section{Virial Equilibrium and Cooling}\label{sec:virial}
-\section{Analytic Cooling Model}
+\section{Analytic Cooling Model}\label{sec:analytic}
 \if\bibinc n

### 2013-02-06 tex/draft/draft.tex
reshuffling of sections in outline
All methods in section 2, which should gradually build from the general to the specific.  Ana, if you're going to start writing here, just keep it as basic and factual as possible and not too wordy.  I.e. just get the basic equations and a few connecting words to the page.  The flowery language if anywhere belongs elsewhere and later modifications are possible.

@@
 \shorttitle{Critical Core Mass at Wide Separations}
-\shortauthors{Piso \& Youdin}
+\shortauthors{Youdin \& Piso \emph{TBD}}
 
-\title{Lower Limits on the Core Mass for Giant Planets}
+\title{On the Minimum Core Mass for Giant Planet Formation}
+%\title{Lower Limits on the Core Mass for Giant Planets}
 %\title{The Critical Core Mass of Wide-Separation Giant Planets:  Lower Limits from Kelvin-Helmholtz Cooling}
@@
 \affil{Harvard Smithsonian Center for Astrophysics}
-\author{ Andrew N.\ Youdin}
+\author{Andrew N.\ Youdin}
 \affil{JILA, University of Colorado}
@@
 \section{Introduction}
-\emgr{Background topics to mention include: core accretion history, giant planets at wide separations and theoretical challenges posed (GI probably to big, core accretion maybe to slow).}
+\emgr{Background topics to mention include: core accretion history, giant planets at wide separations and theoretical challenges posed (GI probably too big, core accretion maybe to slow).}
 
@@
 
-\section{Quasistatic Two Layer Model}\label{sec:model}
+\section{Atmosphere Models}\label{sec:model}
 
-\subsection{Model Assumptions}
-\emgr{ Define and describe basic length scales; Describe disk conditions: MMSN etc.
-Describe basic assumptions: 1D atmosphere, smooth matching onto the disk, atmosphere composition etc., perhaps in list format; Describe assumptions specific to our model: quasistatic, $L_{KH} \gg L_{acc}$, two-layer, constant $L$.
+\subsection{Basic Assumptions}
+\emgr{ Define and describe basic length scales; Describe disk conditions: MMSN etc. Describe basic assumptions: 1D atmosphere, smooth matching onto the disk, atmosphere composition etc.  \\
+The idea is to start gentle with generalities and then get specific with the 2-layer quasistatic model after the basic eqns. have been given.}
+
+%Describe assumptions specific to our model: quasistatic, $L_{KH} \gg L_{acc}$, two-layer, constant $L$.
 Explain how our model is intermediary between purely static models with $L=L_{acc}$ and full time-dependent evolutionary models.
-Could be split into sub-subsections.}
 
@@
 EOS (polytrope, various assumption for $\nabla_{\mathrm{ad}}$ and $\mu$) and opacity laws (noting that we only consider cool radiative zones with dust opacities).
-Give boundary conditions (probably not worth a whole subsection by itself)}
+Give boundary conditions}
 
-%\subsection{Boundary Conditions: Core and Disk}
+\subsection{Standard Methods of Solution} 
+\emgr{shooting vs.
+Henyey methods to satisfy above BC.
+Static models powered by accretion and full evolutionary calculations.
+Note that there's nothing really to explain about the ``shooting method" .
+Fine to state the BC's and that we use the shooting method.
+That's a difference between a paper and research notes.
+We probably want the radius version of structure ODEs.}
 
-%\subsection{Convective Zone}
-%\emgr{might not need a (sub)section}
+\subsection{Virial Equilibrium and Global Cooling}
+\emgr{Start with basic idea of quasistatic contraction and $L = -\dot{E}$, then write out cooling equation and explain terms.
+Explain where it comes from (local cooling + virial equilibrium).
+Defer some derivation to Appendix \ref{sec:virial}.
+In principle this is still general stuff, i.e.
+the global cooling eqn applies to all hydrostatic models, not just our quasi-static one.}
 
-%\subsection{Radiative Zone Solutions}
-%\emgr{Give analytic solution valid for power law opacity and lack of self-gravity.}
-%
-%\subsection{Matching at the Radiative-Convective Boundary (RCB)}
-%\emgr{Include figure of an example atmospheric profile.}
-%
-%\subsection{The Gravothermal Catastrophe}
-%\emgr{Show and explain double valued solutions, including caveat that upper branch breaks several model assumptions.}
+Before describing our quasistatic method we first describe the general ways that planetary Kelvin-Helmholz contraction differs from the stellar case.
 
-\section{Cooling History}
-\emgr{Start with basic idea of quasistatic contraction and $L = -\dot{E}$, then write out cooling equation and explain terms. Explain where it comes from (local cooling + virial equilibrium)}
-\subsection{Role of Finite Pressure}
- \emgr{Give more more general result, but defer derivation to Appendix \ref{sec:virial}. Perhaps not worth a subsection by itself.}
-%\subsection{Estimating a Critical Core Mass}
-%\emgr{Describe how and why we use the time to gravothermal catastrophe as estimate of minimum core mass}
+\subsection{Quasi-static Two-layer Model}
+\emgr{Two layer model and simplified treatment of radiative zone. How to obtain an evolutionary series from cooling model between subsequent static atmospheres. }
 
-\section{Numerical method}
-\emgr{Explain the shooting technique, then how to obtain an evolutionary series from cooling model between subsequent static atmospheres. Show P-T, M-T, M-R etc. plots. Also L-t, L-M plots.}
-\subsection{The Gravothermal Catastrophe \emgr{or} {Entropy Minimum} \emgr{or} {Negative Luminosity} \emgr{or} Something else?}
+\section{Analytic Cooling Model}
+\emgr{derive model (perhaps deferring some results to an appendix), give and explain expected scalings.  Could go after numerical results, but if first can use to explain some numerical results.}
+
+\section{Quasistatic Kelvin-Helmholz Contraction}
+\emgr{Show instantaneous profiles. Also L-t, L-M evolution.}
+
+\subsection{The End of Quasistatic Evolution}
 \emgr{Describe that time / luminosity becomes negative when $M_{atm} \sim M_c$.}
@@
 
-\section{Discussion}
-\subsection{Kelvin-Helmholtz Contraction vs. Planetesimal Accretion}
+\section{Neglected effects}
+\emgr{EOS, time evolution, planetesimal accretion, inefficient convection (mixing length models).  Subsections as needed.}
+\subsection{Planetesimal Accretion}
 \emgr{Compare and contrast our results with Rafikov's, show accretion rates plot; discuss why our fixed core mass assumptions are valid in the regime we are considering.}
-\subsection{Analytic Cooling Model}
-\emgr{Explain assumptions of analytic model but defer equations and derivations to Appendix \ref{sec:analytic}. Show numerical vs. analytic comparison plots and explain role of self gravity. }
-\subsection{EOS Tables}
-\emgr{Mention and briefly describe Saumon et al. EOS tables as future work. }
 
@@
 \acknowledgements
-\emgr{Acknowledge Ruth, Scott, Jonathan Fortney \& Didier Saumon.}  Portions of this project were supported by the {\it NASA} {\it Astrophysics Theory Program} and  {\it Origins of Solar Systems Program}  through grant NNX10AF35G.
+\emgr{Acknowledge Ruth, Phil Armitage, Scott Kenyon, Jonathan Fortney \& Didier Saumon.}  Portions of this project were supported by the {\it NASA} {\it Astrophysics Theory Program} and  {\it Origins of Solar Systems Program}  through grant NNX10AF35G.
 \emgr{JILA, Ruth's grants?}\\
@@
 %\section{Extending the EOS tables}\label{sec:EOSext}
-\section{Virial Equilibrium and Cooling}\label{sec:virial}
-\section{Analytic Cooling Model}\label{sec:analytic}
+\section{Derivation of Cooling Model}\label{sec:virial}
+\section{Analytic Cooling Model Details}\label{sec:analytic}
+
 \if\bibinc n

### 2013-02-06 python/utils/mynumsci.py
added simple jacobian rootfinder to mynumsci
'dxjac' provides a single guess for the best step and 'ezroot2D' wraps these steps making some changes to jacobian if errors start getting worse.

@@ -8,6 +8,7 @@ import numpy as np
 from utils.zbrac import zbrac
 from utils.myspace import mylogspace, loglinspace, linlogspace
 from utils.constants import Me, Re, Myr
+import sys
 
 def nextguess(xnew, ys, xs, maxdy = 0):
     """
@@ -38,7 +39,7 @@ class nf(float):
         else:
             return '%.1f' % self.__float__()
 
-def jac(func, x, eps=1e-6, direct=1, args=(), kwargs={}):
+def jac(func, x, F0=0, eps=1e-6, direct=1, args=(), kwargs={}):
     """
     Parameters
     ----------
@@ -52,9 +53,12 @@ def jac(func, x, eps=1e-6, direct=1, args=(), kwargs={}):
         0 for symmetric (two half steps, so more expensive)
     args, kwargs: tuple, dict (optional)
         args and kwargs to be passed to `func`
+
+    TODO: optionally supply value func(x) as keyword
     """
-    
-    F0 = func(x, *args, **kwargs)
+
+    if F0 == 0 and direct != 0:#no need to compute if centered deriv 
+        F0 = func(x, *args, **kwargs)
     ni, nj = len(F0), len(x)
     jac = np.empty((ni, nj))
     if np.shape(np.atleast_1d(eps)) == (1,):
@@ -80,3 +84,58 @@ def jac(func, x, eps=1e-6, direct=1, args=(), kwargs={}):
         jac[:, j] = (np.array(Fnew) - np.array(F0)) / h
         
     return jac
+
+def dxjac(func, x0, F0=0, eps=1e-5, direct=1, args=(), kwargs={}):
+    """Uses Jacobian of `func` at `x0` to estimate location of root"""
+
+    if F0 == 0:
+        F0 = func(x0, *args, **kwargs)
+    jaco = jac(func, x0, F0, eps, direct, args=args, kwargs=kwargs)
+    invJac = np.linalg.inv(jaco)
+    dx = -np.dot(invJac, F0)
+    return x0 + dx
+
+def ezroot2D(func, x0, tol=1e-7, nmax = 20, jaceps=1e-5, jacdir=1,
+             verbose=0, args=(), kwargs={}):
+    """find non-linear roots with simple Jacobian steps
+
+    TODO: allow array of tol values
+    """
+
+    err = func(x0, *args, **kwargs)
+    xbest, errbest = x0, err
+    fails = 0
+    
+    for i in range(nmax):#don't need i...
[…]

### 2013-02-06 python/utils/mynumsci.py
python learning notebook added
There were just notes/junk in a previous folder, but am splitting it off as a semi-organized reference.  Also renamed ezroot2D function just ezroot since nothinng specifies 2D (though that is only case tested).

@@ -95,7 +95,7 @@ def dxjac(func, x0, F0=0, eps=1e-5, direct=1, args=(), kwargs={}):
     dx = -np.dot(invJac, F0)
     return x0 + dx
 
-def ezroot2D(func, x0, tol=1e-7, nmax = 20, jaceps=1e-5, jacdir=1,
+def ezroot(func, x0, tol=1e-7, nmax = 20, jaceps=1e-5, jacdir=1,
              verbose=0, args=(), kwargs={}):
     """find non-linear roots with simple Jacobian steps
 

### 2013-02-07 python/RadSGPoly/shoot2.py
s2.root2D modified to add ezroot option
Also modified ezroot to return a namedtuple that matches the form of the output from the scipy rootfinders, e.g. sol.x for solution.  Not as much info though.  Also some improvements to keyword passing with ** unpacking.  That could still be done more to simplify code with shorter argument lists (and also reduce probability of typos).

@@ -28,7 +28,7 @@ import numpy as np
 from utils.constants import G, Me, Re, Msun, sigma, Pdisk, Tdisk, RHill, \
                             kdust, AU, Myr
 from utils.parameters import R, FT, FSigma, mstar, rhoc, a
-from utils.mynumsci import nextguess, zbrac, mylogspace
+from utils.mynumsci import nextguess, zbrac, mylogspace, ezroot
 from scipy.integrate import odeint
 from scipy.optimize import brentq, newton, root, fsolve
 from scipy.interpolate import interp1d
@@ -45,8 +45,8 @@ values (and create new instance).
 """
 Params = namedtuple('Params',
                     'mco, rco, a, R, delad, Po, To,  Mstar, kappa, match')
-prms = Params(mcore, rcore, a, R, 2./7, Mstar = mstar*Msun, match = 'Hill',
-              Po = Pdisk(a,mstar, FSigma, FT) , To = Tdisk(a,FT), kappa = kdust )
+prms = Params(mcore, rcore, a, R, 2./7, Mstar=mstar*Msun, match='Hill',
+              Po=Pdisk(a,mstar, FSigma, FT), To=Tdisk(a,FT), kappa=kdust)
 
 
 def evostruct_r(y, r, dt, prev, prms=prms, verbose=0):
@@ -575,7 +575,8 @@ def errormap(Mtot, lL1, lL2, dt1, dt2, prev, prms=prms, nL=10, ndt=10,
     return lL, dt, errmap
 
 
-def root2D(x0, Mtot, prev, prms=prms, Lnorm=0, method='hybr', full_out=0):
+def root2D(x0, Mtot, prev, prms=prms, Lnorm=0, method='hybr', full_out=0
+           , **rkwargs):
     """
     find root x = (lL, log10(dt)) that gives matched atmosphere solution
 
@@ -594,6 +595,8 @@ def root2D(x0, Mtot, prev, prms=prms, Lnorm=0, method='hybr', full_out=0):
     method : string, optional
         root finding method to pass to scipy.optimize.root, currently only
         `hybr` known to work.
+    **rkwargs : extra keyword arguments
+        passed to root finder
 
     Example:
         >>> Mold, Mnew = 5.8*s2.Me, 5.808*s2.Me
@@ -604,11 +607,15 @@ def root2D(x0, Mtot, prev, prms=prms, Lnorm=0, method='hybr', full_out=0):
     """
 
[…]

### 2013-02-07 python/utils/mynumsci.py

@@ -8,6 +8,7 @@ import numpy as np
 from utils.zbrac import zbrac
 from utils.myspace import mylogspace, loglinspace, linlogspace
 from utils.constants import Me, Re, Myr
+from collections import namedtuple
 import sys
 
 def nextguess(xnew, ys, xs, maxdy = 0):
@@ -99,22 +100,32 @@ def ezroot(func, x0, tol=1e-7, nmax = 20, jaceps=1e-5, jacdir=1,
              verbose=0, args=(), kwargs={}):
     """find non-linear roots with simple Jacobian steps
 
+    Returns
+    -------
+    sol : named tuple
+        Containts solution `x`, func(x) `fun`, and func evals `nfev`
+
     TODO: allow array of tol values
+    TODO: nfev is approximate and assumed 2D problem
     """
 
+    Sol = namedtuple('Sol', 'x, fun, nfev')
     err = func(x0, *args, **kwargs)
     xbest, errbest = x0, err
-    fails = 0
+    fails, nfev = 0, 1
     
     for i in range(nmax):#don't need i...
         xn = dxjac(func, xbest, errbest, jaceps, jacdir, args=args,
                    kwargs=kwargs)
         errn = func(xn, *args, **kwargs)
+        nfev += 3
+        if fails != 0:
+            nfev =+ 2 #for centered jacobian
         if verbose:
             print i, errn
         
         if np.all(np.abs(errn) < tol):
-            return xn, errn
+            return Sol(xn, errn, 2*i + 2)
         elif np.any(np.abs(errn) < np.abs(errbest)):
             xbest, errbest = xn, errn
         elif np.all(np.abs(errn) > np.abs(errbest)):
@@ -134,8 +145,11 @@ def ezroot(func, x0, tol=1e-7, nmax = 20, jaceps=1e-5, jacdir=1,
                         print "lower resolution jacobian"
                 elif fails > 2:
                     print "all errors got worse, settling for", errbest
-                    return xbest, errbest
+                    return Sol(xbest, errbest, 2*i + 7)
                 else:
                     sys.exit("can't count fails dummy")
 
-    return xbest, errbest
+    
+    nfev = 2*i + 7 #estimated, counts centered jacobians for final 3 guesses
+ 
+    return Sol(xbest, errbest, nfev)

### 2013-02-07 python/RadSGPoly/shoot2.py
Evolution working with ezroot!
Only 20 steps so far, but that's a new record and enough.  Need to add restart ability and filesave before starting a real evolution run.  Other possible improvements are a higher order guess, i.e. using 3 previous values instead of just 2, since verbose output shows a systematic (not random) luminosity error of about -1% (in the early evolution at least).

@@ -534,13 +534,12 @@ def error2D(x, Mtot, prev, prms=prms, Lnorm=0, full_out=0, dt_log=1,
         y = odeint(f, bc, rrange, args=stargs)
     Mcint , Lcint = y[-1,0] , y[-1,3] #core mass, Luminosity
 
-    #set scale luminosity errors
-    if Lnorm == 0:
+    #set scale luminosity errors. A numerical Lnorm > 1e6 will pass
+    if Lnorm == 0 or Lnorm == 1:
         Lnorm = 10**lL
-    elif Lnorm < 1e6: #multiply previous error by factor Lnorm
+    elif Lnorm < 1e6: #multiply error by factor Lnorm
         Lnorm = 10**lL/Lnorm
-    else: #let supplied Lnorm pass
-        Lnorm = 1*Lnorm
+    
     err = (Mcint/prms.mco - 1, Lcint/Lnorm)
     
     #track guesses if lists have been initialized elsewhere
@@ -575,7 +574,7 @@ def errormap(Mtot, lL1, lL2, dt1, dt2, prev, prms=prms, nL=10, ndt=10,
     return lL, dt, errmap
 
 
-def root2D(x0, Mtot, prev, prms=prms, Lnorm=0, method='hybr', full_out=0
+def root2D(x0, Mtot, prev, prms=prms, Lnorm=1, method='ezroot', full_out=0
            , **rkwargs):
     """
     find root x = (lL, log10(dt)) that gives matched atmosphere solution
@@ -601,8 +600,7 @@ def root2D(x0, Mtot, prev, prms=prms, Lnorm=0, method='hybr', full_out=0
     Example:
         >>> Mold, Mnew = 5.8*s2.Me, 5.808*s2.Me
         >>> prev = s2.makeprev(Mold)
-        >>> xg = s2.staticguess(Mnew, prev)*array([1.002])
-        >>> #slightly higher luminosity guesses avoid neg L
+        >>> xg = s2.staticguess(Mnew, prev)
         >>> s2.root2D(xg, Mnew, prev)
     """
 
@@ -666,8 +664,8 @@ def staticguess(M, prev, dt_guess=0.01*Myr, prms=prms, dt_log=1, statgrid=50,
 
 
 def evolve_dM(M, dM=0, xg=[], prevroot=0, prms=prms, prevgrid=1000,
-              statgrid=50, retry=0, full_out=0, Lnorm=0, verbose=1
-              , solver='hybr', **rkwargs):
[…]

### 2013-02-07 python/utils/userpath.py
data folder locations added to userpath

@@ -14,15 +14,20 @@ import os
 user = os.environ['USER']
 
 if user == 'youd':
-    userpath = '/Users/youd/Documents/Work/Repositories/aacore'
+    userpath = os.environ['REPOS']+'/aacore'
 elif user == 'ana-mariapiso':
     userpath ='/Users/ana-mariapiso/Documents/core_accretion'
 elif user == 'apiso':
     userpath = '/home/apiso/repos/core_accretion'
-    
+
+#For all users submodule names and locations fixed.
+aydat = userpath + '/dat_andrew'
+apdat = userpath + '/dat_ana'  
+gendat = userpath + '/dat' 
 
 def cduser(s = '/python'):
-    """cd to the users python directory, or other directory in repository as specified by string"""
+    """cd to the users python directory, or other directory in repository 
+    as specified by string"""
     os.chdir(userpath+s)
 
     

### 2013-02-07 python/RadSGPoly/TPeffects.py
added plots showing disk temp. and pressure effects

@@ -1,0 +1,49 @@
+"""
+
+Generates series of atmosphere profiles for a set of gas, disk and core
+conditions and writes them in .npz files. Also contains functions to load
+profiles.
+
+"""
+
+
+
+from utils.constants import G, kb, mp, Rb, Me, Re, Msun, RH, RHe, sigma, \
+     cmperau, RHill, gammafn, mufn, Rfn, Cvfn, kdust, Tdisk, Pdisk, params
+from utils.parameters import mstar, FSigma, FT
+from utils.userpath import userpath
+import numpy
+import scipy
+import math
+import matplotlib.pyplot as plt
+import matplotlib.cm as cm
+from types import FunctionType as function
+from collections import namedtuple
+from scipy import integrate
+from utils import cb_corrections as cb
+from utils.interpolation_functions_logp import interpolations
+from scipy.integrate import odeint
+from scipy.interpolate import interp1d
+from scipy.optimize import fminbound, brentq
+#from shooting_poly import Ltop, shoot, prms
+from profiles_poly import profiles_write as pp
+
+def Tdfixed(a, delad, Y, rhoc, Mc, fact):
+
+    rc = (3 * Mc / (4 * numpy.pi * rhoc))**(1./3)
+    prms = params(Mc, rc, a, delad, Y, gamma = gammafn(delad), R = Rfn(Y), \
+                  Cv = Cvfn(Y, delad), Pd = Pdisk(a, mstar, FSigma, FT) * fact, \
+                  Td = Tdisk(a, FT), kappa = kdust)
+
+    pp(500, 300, 10**21, 10**30, 1.2*Mc, 2.5*Mc, 'Mc' + \
+                     str(Mc/Me)[:4], prms = prms, disk = 0)
+
+def Pdfixed(a, delad, Y, rhoc, Mc, fact):
+
+    rc = (3 * Mc / (4 * numpy.pi * rhoc))**(1./3)
+    prms = params(Mc, rc, a, delad, Y, gamma = gammafn(delad), R = Rfn(Y), \
+                  Cv = Cvfn(Y, delad), Pd = Pdisk(a, mstar, FSigma, FT), \
+                  Td = Tdisk(a, FT) * fact, kappa = kdust)
+
+    pp(500, 300, 10**21, 10**30, 1.2*Mc, 2.5*Mc, 'Mc' + \
+                     str(Mc/Me)[:4], prms = prms, disk = 0)

### 2013-02-07 python/RadSGPoly/profiles_poly.py

@@ -28,7 +28,7 @@ from shooting_poly import Ltop, shoot, prms
 
 
 def profiles_write(n, nMpoints, L1, L2, Mmin, Mmax, filename, prms = prms, \
-                   tol = 10**(-25), log = 1, savefile = 1):
+                   tol = 10**(-25), log = 1, savefile = 1, disk = 1):
 
     """
     Uses the output of shoot (described in shooting_poly) to generate recarrays
@@ -146,19 +146,32 @@ def profiles_write(n, nMpoints, L1, L2, Mmin, Mmax, filename, prms = prms, \
 
 
     if savefile == 1:
-        paramfilename = userpath + '/dat_ana/MODELS/RadSGPoly/' + 'delad' + \
-                        str(prms.delad)[:4] + '/Y' + str(prms.Y) + '/' + \
-                        str(int(prms.a)) + 'AU/' + filename
-        numpy.savez_compressed(paramfilename, model = model, param = param, \
+        if disk == 1:
+            paramfilename = userpath + '/dat_ana/MODELS/RadSGPoly/' + 'delad' + \
+                            str(prms.delad)[:4] + '/Y' + str(prms.Y) + '/' + \
+                            str(int(prms.a)) + 'AU/' + filename
+            numpy.savez_compressed(paramfilename, model = model, param = param, \
+                               prof = prof)
+        else:
+            paramfilename = userpath + '/dat_ana/MODELS/RadSGPoly/' + 'Td' + \
+                            str(prms.Td)[:6] + '_Pd' + str(prms.Pd)[:7] + \
+                            'Mc' + str(prms.Mco/Me)
+            numpy.savez_compressed(paramfilename, model = model, param = param, \
                                prof = prof)
     
     return model, param, prof
 
 
-def atmload(filename, prms = prms):
-    npzdat = numpy.load(userpath + '/dat_ana/MODELS/RadSGPoly/' + 'delad' + \
-                        str(prms.delad)[:4] + '/Y' + str(prms.Y) + '/' + \
-                        str(int(prms.a)) + 'AU/' + filename)
+def atmload(filename, prms = prms, disk = 1):
+    if disk == 1:
+        npzdat = numpy.load(userpath + '/dat_ana/MODELS/RadSGPoly/' + 'delad' + \
+                            str(prms.delad)[:4] + '/Y' + str(prms.Y) + '/' + \
[…]

### 2013-02-07 python/RadSGPoly/shooting_poly.py
plots showing opacity effects

@@ -19,7 +19,7 @@ atmospheres
 
 
 from utils.constants import G, kb, mp, Rb, Me, Re, Msun, RH, RHe, sigma, \
-     cmperau, RHill, gammafn, mufn, Rfn, Cvfn, kdust, Tdisk, Pdisk, params
+     cmperau, RHill, gammafn, mufn, Rfn, Cvfn, kdust, kdust10, Tdisk, Pdisk, params
 from utils.parameters import FT, FSigma, mstar, Y, delad, rhoc, Mc, rc, \
      gamma, Y, a
 import numpy
@@ -41,11 +41,11 @@ prms = params(Mc, rc, a, delad, Y, gamma = gammafn(delad), R = Rfn(Y), \
               Td = Tdisk(a, FT), kappa = kdust) #gas, disk and core parameters
                         #for specific values imported from parameters.py 
 
-def delradfn(p, m, T, L): #radiative temperature gradient
-    return 3 * kdust(T) * p * L / (64 * pi * G * m * sigma * T**4)
+def delradfn(p, m, T, L, prms = prms): #radiative temperature gradient
+    return 3 * prms.kappa(T) * p * L / (64 * pi * G * m * sigma * T**4)
 
-def Del(p, m, T, L): #del = min(delad, delrad)
-    return min(prms.delad, delradfn(p, m, T, L))
+def Del(p, m, T, L, prms = prms): #del = min(delad, delrad)
+    return min(prms.delad, delradfn(p, m, T, L, prms))
 
 def Ltop(Mi, L1, L2, n, tol, prms = prms):
 
@@ -96,7 +96,7 @@ def Ltop(Mi, L1, L2, n, tol, prms = prms):
         """
         
         return numpy.array([ - G * x[2] * x[0] / (r**2 * prms.R * x[1]), \
-                             - Del(x[0], x[2], x[1], x[3]) * G * x[2] / \
+                             - Del(x[0], x[2], x[1], x[3], prms) * G * x[2] / \
                              (prms.R * r**2),
                              4 * pi * r**2 * x[0] / (prms.R * x[1]), \
                              0])              
@@ -205,7 +205,7 @@ def shoot(Mi, L1, L2, n, tol, prms = prms):
         """
         
         return numpy.array([ - G * x[2] * x[0] / (r**2 * prms.R * x[1]), \
-                             - Del(x[0], x[2], x[1], x[3]) * G * x[2] / \
[…]

### 2013-02-07 python/utils/constants.py

@@ -61,6 +61,14 @@ def kdust(T, P = None, b = 2, fsolar = 1.):
     """Opacity (generally applicable for beta = 2)"""
     return 2 * fsolar * (T/100.)**b
 
+def kdust10(T, P = None, b = 2, fsolar = 1.):
+    """Opacity reduced by a factor of 10"""
+    return 2 * fsolar * (T/100.)**b / 10
+
+def kdust100(T, P = None, b = 2, fsolar = 1.):
+    """Opacity reduced by a factor of 100"""
+    return 2 * fsolar * (T/100.)**b / 100
+
 def kfixed(T = None, P = None, const = 0.01):
     """a silly function but useful if kappa functions expects a temperature."""
     return const

### 2013-02-08 python/utils/userpath.py
learning has more on **kwargs
while this helps to keep parameter lists down (and makes changing what those parameters are easier) any keywords in the function itself need to be explicitly written out.  Also a slash added after definition of dat folder (should have done this for userpath, but it's too late now)...

@@ -21,9 +21,9 @@ elif user == 'apiso':
     userpath = '/home/apiso/repos/core_accretion'
 
 #For all users submodule names and locations fixed.
-aydat = userpath + '/dat_andrew'
-apdat = userpath + '/dat_ana'  
-gendat = userpath + '/dat' 
+aydat = userpath + '/dat_andrew/'
+apdat = userpath + '/dat_ana/'  
+gendat = userpath + '/dat/' 
 
 def cduser(s = '/python'):
     """cd to the users python directory, or other directory in repository 

### 2013-02-08 python/RadSGPoly/shoot2.py
work on saving output and restart info for 2D shooting
Currently stuck on problem that the `prev` object for previous solutions is too complicated to save as npz or other python methods.   Previous trick of turning named tuple into a dictionary isn't working, this is a more insidious problem...

@@ -33,8 +33,9 @@ from scipy.integrate import odeint
 from scipy.optimize import brentq, newton, root, fsolve
 from scipy.interpolate import interp1d
 from collections import namedtuple
-from utils.userpath import userpath
-datfolder = userpath + "/dat/MODELS/RadSGPoly2/"
+from utils.userpath import userpath, aydat
+import pickle
+#datfolder = userpath + "/dat/MODELS/RadSGPoly2/" 
 
 p4 = 4*np.pi
 mcore  = 5*Me 
@@ -195,9 +196,9 @@ def polyradstruct_r_simp(y, r, dt = 0, prms = prms, prev = 0):
     """
     simple version of structure eqns. dy/dr for "y" = [m, P , T, L].
 
-    Not currently used, but the simplest form of the full time-dependent structure
-    eqns. for polytrope. 
-    Not intended for static solutions and will break if dt and prev not specified.
+    Not currently used, but the simplest form of the full time-dependent 
+    structure eqns. for polytrope.  Not intended for static solutions and 
+    will break if dt and prev not specified.
     """
     m, P, T, L = y
     rho = P/prms.R/T
@@ -407,11 +408,11 @@ def sol2prof(prev, prms=prms):
 
 def rBondi(m, r, prms):
     """
-    find the Bondi radius for a solution. Used for evaluation, not in obtaining solutions.
+    find the Bondi radius for a solution. Used for evaluation, not in 
+    obtaining solutions.
 
-    When Bondi radius is outside Hill (fit) radius Bondi radius is based on mass within
-    the fit radius.  In any even Bondi radius has little meaning in this case (in 1D
-    models like ours).
+    When Bondi radius is outside Hill (fit) radius Bondi radius is based on 
+    mass within the fit radius.  Bondi radius has little meaning in this case.
     """
     
     if len(m) != len(r):
@@ -730,8 +731,10 @@ def evolve_dM(M, dM=0, xg=[], prevroot=0, prms=prms, prevgrid=1000,
 
     return sol
     
-def evolve_DM(M, DM, ep=0.01, prms=prms, verbose=1):
+def evolve_DM(M0, DM, ep=0.01, savedir='test/', savefile=0, prms=prms, 
+              **dmargs):
     """take a bunch of steps...
[…]

### 2013-02-08 python/utils/mynumsci.py

@@ -144,7 +144,8 @@ def ezroot(func, x0, tol=1e-7, nmax = 20, jaceps=1e-5, jacdir=1,
                     if verbose:
                         print "lower resolution jacobian"
                 elif fails > 2:
-                    print "all errors got worse, settling for", errbest
+                    if verbose:
+                        print "best error", errbest
                     return Sol(xbest, errbest, 2*i + 7)
                 else:
                     sys.exit("can't count fails dummy")

### 2013-02-18 python/Atmscripts/cooling_realgas_sg.py
added scripts for real EOS

@@ -1,0 +1,108 @@
+os.chdir(userpath + '/python')
+
+import numpy
+from utils.constants import Me, Re
+import matplotlib.pyplot as plt
+from utils.userpath import userpath
+from RadSGRealGas.shooting import prms
+from RadSGRealGas.profiles import atmload
+from RadSGRealGas.cooling import cooling_global as cg, cooling_local as cl
+
+os.chdir(userpath + '/python/Atmscripts')
+
+
+load = 1
+savefig = 0
+savefigt = 0
+out = 'rcb'
+
+if load != 0:
+
+    atm = atmload('Mc5.0.npz', prms = prms)
+    model = atm[0]
+    param = atm[1]
+    prof = atm[2]
+
+    Mass = param.MB
+    #M = M[1:]
+
+    cool = cg(param, prof, prms = prms, out = out)
+    t = cool[0]
+    deltae = cool[2]
+    eaccdm = cool[3]
+    PdV = cool[4]
+    Lglobdt = -deltae + eaccdm - PdV
+
+    Llocdt = cl(param, prof, prms = prms, out = out)
+    Llocdtnegl = cl(param, prof, prms = prms, out = 'RHill', onlyrad = 1)
+
+L = param.L
+dM = (Mass[1:] - Mass[:-1]) * Me
+M = Mass[:-1] - model.Mco / Me #only atmosphere mass for plot
+
+plt.loglog(M, - deltae / dM, 'r', label = r'$-dE/dM$')
+plt.loglog(M, deltae / dM, '--r')
+plt.loglog(M, eaccdm / dM, 'g', label = r'$-e_{\mathrm{acc}}$')
+plt.loglog(M, - eaccdm / dM, '--g')
+plt.loglog(M, - PdV / dM, 'y', label = r'$-PdV/dM$')
+plt.loglog(M, PdV / dM, '--y')
+plt.loglog(M, Lglobdt / dM, 'black', label = r'$L_{\mathrm{glob}}dt/dM$')
+plt.loglog(M, - Lglobdt / dM, 'black', linestyle = '--')
+plt.loglog(M, Llocdt / dM, 'b', label = r'$L_{\mathrm{loc}}dt/dM$')
+plt.loglog(M, - Llocdt / dM, 'b', linestyle = '--')
[…]

### 2013-02-18 python/RadSGRealGas/atmseries.py

@@ -1,0 +1,130 @@
+
+"""
+The script generates series of atmospheres for varying core masses at a fixed distance in the disk.
+"""
+
+import numpy
+from scipy import integrate
+from scipy.interpolate import interp1d
+from utils.constants import G, kb, mp, Rb, Me, Re, Msun, RH, RHe, sigma, \
+     cmperau, RHill, gammafn, mufn, Rfn, Cvfn, kdust, Tdisk, Pdisk
+from utils.parameters import mstar, FSigma, FT
+from utils.constants import params
+from RadSGRealGas.profiles import profiles_write
+from RadSGRealGas.cooling import critical
+
+def atmseries(a, rhoc, Y, Mcomin, Mcomax, nMco, n = 500, \
+              nMpoints = 300, L1 = 10**22, L2 = 10**28, minMfrac = 1.05, \
+              maxMfrac = 2.1):
+    
+    """
+
+    Generates atmosphere profiles for a range of core masses at fixed distance
+    in the disk.
+
+    Input
+    -----
+    a:
+        distance in the disk in AU
+    rhoc:
+        core density in g/cm^3; default value used rhoc = 3.2
+    Y:
+        Helium core mass fraction
+    Mcomin:
+        minimum core mass for which we want to generate an atmosphere
+    Mcomax:
+        maximum core mass for which we want to generate an atmosphere
+    nMco:
+        number of core masses for which we want to generate an atmosphere
+    n:
+        number of points for the profile of one atmosphere of fixed mass.
+        The default is 500.
+    nMpoints:
+        number of atmosphere profiles of fixed mass for a given core mass.
+        The default is 300.
+    L1, L2:
+        luminosities that bracket the matched solution in erg/s.
+        Default L1 = 10**21, L2 = 10**30
+    minMfrac:
+        fractional mass (expressed in core masses) of the lowest mass
+        atmosphere (including core mass) for a given core. Default 1.1
+    maxMfrac:
+        fractional mass (expressed in core masses) of the highest mass
+        atmosphere (including core mass) for a given core. Default 1.1
+
+
+    Returns
+    -------
+    .npz files containing atmosphere profiles
+    
+    """
+
+    #setting up disk parameters
+    Pd = Pdisk(a, mstar, FSigma, FT)
+    Td = Tdisk(a, FT)
+
+    Mcore = numpy.linspace(Mcomin, Mcomax, nMco) #probably linear spacing is
[…]

### 2013-02-18 python/RadSGRealGas/cooling.py

@@ -1,0 +1,262 @@
+"""
+
+Computes global and local energy balance for a given series of atmospheres.
+Also used to find the critical core mass.
+
+"""
+
+
+from utils.constants import G, kb, mp, Rb, Me, Re, Msun, RH, RHe, sigma, \
+     cmperau, RHill, gammafn, mufn, Rfn, Cvfn, kdust, Tdisk, Pdisk, params
+from utils.interpolation_functions import interplog10u, interplog10S
+import numpy
+import scipy
+import matplotlib.pyplot as plt
+import matplotlib.cm as cm
+from collections import namedtuple
+from scipy import integrate
+from scipy.integrate import odeint
+from scipy.interpolate import interp1d
+from shooting import prms
+
+#---------------------------------------------------------------------------
+
+def cooling_global(atmset, atmprofile, prms = prms, out = 'rcb'):
+    
+    """
+    Determines the time evolution of a given atmosphere. Cooling time is
+    calculated based on the following formula:
+    dt = - (dE - <ecb> dm + <Pcb> dV  dm) / L,
+    where dV is the volume change at constant mass
+
+    Input
+    -----
+    atmset:
+        the 'param' recarray (see profiles_poly.py)
+    atmprofile:
+        the 'prof' recarray (see profiles_poly.py)
+    prms:
+        gas, disk and core parameters; default prms imported from shoot
+    out:
+        specifies the surfaces where cooling is computed; default RCB
+        
+
+    Output
+    ------
+    t:
+        array of delta t's between subsequent atmospheres in seconds
+    t_cumulative:
+        cumulative cooling time in units of 3 Myrs
+    deltae, eaccav * deltamout, Pout * deltav:
+        the cooling terms in the energy equation
+        
+    """
+    
+    n = len(atmset)
+    L = atmset.L
+    deltav = 0 * numpy.ndarray(shape = (len(L) - 1), dtype = float)
+
+    if out == 'rcb':
+        Tout = atmset.Tcb
+        Pout = atmset.Pcb 
+        rout = atmset.rcb * Re
+        Mout = atmset.Mcb * Me 
+        Eout = atmset.Etotcb
+
+    elif out == 'RHill':
+        Tout = numpy.array([prms.Td] * n)
+        Pout = numpy.array([prms.Pd] * n) 
+        rout = atmset.RHill * Re 
[…]

### 2013-02-18 python/RadSGRealGas/mass_entropy.py

@@ -1,0 +1,81 @@
+import numpy
+from utils.constants import G, kb, mp, Rb, Me, Re, Msun, RH, RHe, sigma, \
+     cmperau, RHill, gammafn, mufn, Rfn, Cvfn, kdust, kdust10, Tdisk, Pdisk, \
+     paramsEOS, params
+from utils.parameters import FT, FSigma, mstar, Y, delad, rhoc
+from RadSGRealGas.profiles import atmload as atmloadreal
+from RadSGPoly.profiles_poly import atmload as atmloadpoly
+from utils.interpolation_functions import interplog10S
+from scipy.interpolate import interp1d
+
+def massent_real(Y, a, log10entropy, atmfile):
+
+    prms = paramsEOS(Me, Re, Y, a, Pd = Pdisk(a, mstar, FSigma, FT), \
+                 Td = Tdisk(a, FT), kappa = kdust)
+    #log10Sref = interplog10S(numpy.log10(Tcb), numpy.log10(Pcb))
+    
+    atm = atmloadreal(atmfile, prms = prms)
+    model = atm[0]
+    param = atm[1]
+    prof = atm[2]
+
+    Tc = param.Tc
+    Pc = param.Pc
+    MB = param.MB
+
+    log10S = interplog10S(numpy.log10(Tc), numpy.log10(Pc))
+
+    for i in range(len(log10S) - 1):
+        if log10S[i] < log10S[i + 1]:
+            break
+    log10S = log10S[:i]
+    MB = MB[:i]
+    
+    fm = interp1d(log10S[::-1], MB[::-1])
+    Mtot = float(fm(log10entropy))
+
+    return Mtot
+
+def massent_poly(Y, delad, a, mass, atmfile):
+
+    prms = params(Me, Re, a, delad, Y, gamma = gammafn(delad), R = Rfn(Y), \
+              Cv = Cvfn(Y, delad), Pd = Pdisk(a, mstar, FSigma, FT), \
+              Td = Tdisk(a, FT), kappa = kdust)
+    
+    atm = atmloadpoly(atmfile, prms = prms)
+    model = atm[0]
+    param = atm[1]
+    prof = atm[2]
+
+    Tcb = param.Tcb
+    Pcb = param.Pcb
+    MB = param.MB
+
+    fT = interp1d(MB, Tcb)
+    fP = interp1d(MB, Pcb)
+
+    Tm = float(fT(mass))
+    Pm = float(fP(mass))
+
[…]

### 2013-02-18 python/RadSGRealGas/profiles.py

@@ -1,0 +1,185 @@
+"""
+
+Generates series of atmosphere profiles for a set of gas, disk and core
+conditions and writes them in .npz files. Also contains functions to load
+profiles.
+
+"""
+
+
+
+from utils.constants import G, kb, mp, Rb, Me, Re, Msun, RH, RHe, sigma, \
+     cmperau, RHill, gammafn, mufn, Rfn, Cvfn, kdust, Tdisk, Pdisk, paramsEOS
+from utils.userpath import userpath
+import numpy
+import scipy
+import math
+import matplotlib.pyplot as plt
+import matplotlib.cm as cm
+from types import FunctionType as function
+from collections import namedtuple
+from scipy import integrate
+from utils import cb_corrections as cb
+from utils.interpolation_functions_logp import interpolations
+from scipy.integrate import odeint
+from scipy.interpolate import interp1d
+from scipy.optimize import fminbound, brentq
+from shooting import Ltop, shoot, prms
+
+
+def profiles_write(n, nMpoints, L1, L2, Mmin, Mmax, filename, prms = prms, \
+                   tol = 10**(-25), log = 1, savefile = 1, disk = 1):
+
+    """
+    Uses the output of shoot (described in shooting_poly) to generate recarrays
+    and write .npz atmosphere files.
+
+    Input
+    -----
+    n:
+        number of grid points for each atmosphere of fixed mass
+    nMpoints:
+        number of atmosphere profiles of fixed mass for the given core mass
+    L1, L2:
+        trial luminosity values. L1 and L2 must bracket the solution.
+    Mmin, Mmax:
+        minimum and maximum atmosphere + core masses in g
+    filename:
+        name of the file in which the data is saved; typically 'Mcx',
+        where x is the core mass in Earth masses
+    prms:
+        gas, disk and core parameters; default prms imported from shoot
+    tol:
+        relative luminosity matching error; default 10**(-25)
+    log:
+        flag used to specify whether the mass spacing should be logarithmic
+        or linear; default 1 (spacing is logarithmic)
+    savefile:
+        flag used to specify whether the recarray should be saved to a file
+        or not; default 1, file is saved
+
+
+    Output
+    ------
+    model:
+        array of gas, disk and core parameters
+    param:
+        recarray of
+            Mtot:
+                total atmosphere mass in Earth masses
+            Mcb:
[…]

### 2013-02-18 python/RadSGRealGas/shooting.py

@@ -1,0 +1,336 @@
+"""
+
+Shooting code that takes into account the self-gravity of the radiative zone,
+assuming constant luminosity in the radiative zone.
+
+Call Ltop to find the luminosity corresponding to a converged atmosphere for a
+given mass; generally L1 = 10**20, L2 = 10**30, as a physical luminosity
+values should lie between the two:
+    e.g. Ltop(10*Me, 10**20, 10**30, 500, 10**(-25))
+
+Call shoot to find the P, T, M profile of a converged atmosphere with a given
+mass.
+    e.g. shoot(10*Me, 10**20, 10**30, 500, 10**(-25))
+
+'shoot' is used in profiles_poly to generate recarrays and .npz files of model
+atmospheres
+
+"""
+
+
+from utils.constants import G, kb, mp, Rb, Me, Re, Msun, RH, RHe, sigma, \
+     cmperau, RHill, gammafn, mufn, Rfn, Cvfn, kdust, kdust10, Tdisk, Pdisk, \
+     paramsEOS
+from utils.parameters import FT, FSigma, mstar, Y, delad, rhoc, Mc, rc, \
+     gamma, Y, a #, R_EOS
+from utils.interpolation_functions import interplog10rho, interpdelad, \
+     interplog10u
+import numpy
+from numpy import pi
+import scipy
+import math
+import matplotlib.pyplot as plt
+import matplotlib.cm as cm
+from collections import namedtuple
+from scipy import integrate, interpolate, optimize
+from scipy.integrate import odeint
+from scipy.interpolate import interp1d
+from scipy.optimize import brentq, brenth, bisect, newton
+
+#-----------------------------------------------------------------------------
+
+prms = paramsEOS(Mc, rc, Y, a, Pd = Pdisk(a, mstar, FSigma, FT), \
+                 Td = Tdisk(a, FT), kappa = kdust) #gas, disk and core parameters
+                        #for specific values imported from parameters.py 
+
+def delradfn(p, m, T, L, prms = prms): #radiative temperature gradient
+    return 3 * prms.kappa(T) * p * L / (64 * pi * G * m * sigma * T**4)
+
+def Del(p, m, T, L, delad, prms = prms): #del = min(delad, delrad)
+    return min(delad, delradfn(p, m, T, L, prms))
+
+def Ltop(Mi, L1, L2, n, tol, prms = prms):
+
+    """
[…]

### 2013-02-18 python/utils/constants.py

@@ -91,9 +91,13 @@ def Cvfn(Y, delad):
     """
     return Rfn(Y) / (gammafn(delad) - 1)
 
-#tuple of gas, disk and core parameters
+#tuple of gas, disk and core parameters for polytrope
 params = namedtuple( \
     'params', 'Mco, rco, a, delad, Y, gamma, R, Cv, Pd, Td, kappa')
+
+#tuple of gas, disk and core parameters for real EOS
+paramsEOS = namedtuple( \
+    'paramsEOS', 'Mco, rco, Y, a, Pd, Td, kappa')
 
 
 

### 2013-02-18 python/utils/interpolation_functions.py

@@ -1,0 +1,43 @@
+"""
+
+This module defines the 2-D interpolation functions which will then be applied to pairs of
+thermodynamic variables; used in the shooting code, shooting_odeint
+
+"""
+
+import numpy
+import scipy
+from scipy import interpolate
+from EOS import EOSread
+from table_size import Nt, Np, Ncol
+from userpath import userpath
+
+
+
+EOS=EOSread.EOSread(Nt, Np, Ncol, userpath + \
+                    '/dat/EOStables/H_HE_TAB_I_30_EXT.txt')[0]
+log10T=EOSread.EOSread(Nt, Np, Ncol, userpath + \
+                  '/dat/EOStables/H_HE_TAB_I_30_EXT.txt')[1]
+
+log10p = EOS[-1, :, 0] #log pressure
+
+log10rho = EOS[:, :, 3] #log density
+log10delad = EOS[:, :, 10]
+log10u = EOS[:, :, 5]
+log10S = EOS[:, :, 4]
+
+#density interpolation from temperature and pressure
+yy, xx = numpy.meshgrid(log10p, log10T)
+pts = numpy.array((xx.ravel(), yy.ravel())).T
+
+srrho = log10rho.ravel()
+srd = log10delad.ravel()
+sru = log10u.ravel()
+srs = log10S.ravel()
+
+interplog10rho = scipy.interpolate.LinearNDInterpolator(pts,srrho)
+interpdelad = scipy.interpolate.LinearNDInterpolator(pts,srd)
+interplog10u = scipy.interpolate.LinearNDInterpolator(pts,sru)
+interplog10S = scipy.interpolate.LinearNDInterpolator(pts,srs) \
+                   #args: log(t), log(p)
+

### 2013-02-18 python/utils/parameters.py

@@ -12,7 +12,7 @@ atmosphere:
 """
 
 import numpy as np
-from constants import Me, Rb
+from constants import Me, Rb, Rfn
 
 
 #disk parameters (relative to MMSN model)
@@ -26,11 +26,12 @@ gamma = 7./5
 delad = 1 - 1/gamma
 
 #parameters for H-He mixture
-Y = 0.0 #0.3
+Y = 0.3
 muH = 2.0
 mu = 4./(2-Y)
 R = Rb/mu
 Cv = R / (gamma - 1)
+R_EOS = Rfn(Y)
 
 
 #core parameters

### 2013-02-18 python/utils/userpath.py

@@ -16,7 +16,7 @@ user = os.environ['USER']
 if user == 'youd':
     userpath = os.environ['REPOS']+'/aacore'
 elif user == 'ana-mariapiso':
-    userpath ='/Users/ana-mariapiso/Documents/core_accretion'
+    userpath ='/Users/ana-mariapiso/core_accretion'
 elif user == 'apiso':
     userpath = '/home/apiso/repos/core_accretion'
 

### 2013-02-19 python/RadSGPoly/atmseries_poly.py
modified scripts to account for opacity effects

@@ -15,7 +15,7 @@ from RadSGPoly.cooling_poly import critical
 
 def atmseries(a, rhoc, delad, Y, Mcomin, Mcomax, nMco, n = 500, \
               nMpoints = 300, L1 = 10**21, L2 = 10**30, minMfrac = 1.1, \
-              maxMfrac = 2.8):
+              maxMfrac = 2.8, opacity = kdust):
     
     """
 
@@ -75,7 +75,7 @@ def atmseries(a, rhoc, delad, Y, Mcomin, Mcomax, nMco, n = 500, \
     for i in range(nMco):
         
         prms = params(Mcore[i], rcore[i], a, delad, Y, gamma, R, Cv, Pd, Td,\
-                      kappa = kdust)
+                      kappa = opacity)
         Mmin = minMfrac * Mcore[i]
         Mmax = maxMfrac * Mcore[i]
 

### 2013-02-19 python/RadSGPoly/profiles_poly.py

@@ -9,7 +9,8 @@ profiles.
 
 
 from utils.constants import G, kb, mp, Rb, Me, Re, Msun, RH, RHe, sigma, \
-     cmperau, RHill, gammafn, mufn, Rfn, Cvfn, kdust, Tdisk, Pdisk, params
+     cmperau, RHill, gammafn, mufn, Rfn, Cvfn, kdust, Tdisk, Pdisk, params, \
+     kdust, kdust10, kdust100
 from utils.userpath import userpath
 import numpy
 import scipy
@@ -147,11 +148,24 @@ def profiles_write(n, nMpoints, L1, L2, Mmin, Mmax, filename, prms = prms, \
 
     if savefile == 1:
         if disk == 1:
-            paramfilename = userpath + '/dat_ana/MODELS/RadSGPoly/' + 'delad' + \
-                            str(prms.delad)[:4] + '/Y' + str(prms.Y) + '/' + \
-                            str(prms.a) + 'AU/' + filename
-            numpy.savez_compressed(paramfilename, model = model, param = param, \
-                               prof = prof)
+            if prms.kappa == kdust:
+                paramfilename = userpath + '/dat_ana/MODELS/RadSGPoly/' + 'delad' + \
+                                str(prms.delad)[:4] + '/Y' + str(prms.Y) + '/' + \
+                                str(prms.a) + 'AU/' + filename
+                numpy.savez_compressed(paramfilename, model = model, param = param, \
+                                   prof = prof)
+            elif prms.kappa == kdust10:
+                paramfilename = userpath + '/dat_ana/MODELS/RadSGPoly/' + 'delad' + \
+                                str(prms.delad)[:4] + '/Y' + str(prms.Y) + '/kdust10/' + \
+                                str(prms.a) + 'AU/' + filename
+                numpy.savez_compressed(paramfilename, model = model, param = param, \
+                                   prof = prof)
+            elif prms.kappa == kdust100:
+                paramfilename = userpath + '/dat_ana/MODELS/RadSGPoly/' + 'delad' + \
+                                str(prms.delad)[:4] + '/Y' + str(prms.Y) + '/kdust100/' + \
+                                str(prms.a) + 'AU/' + filename
+                numpy.savez_compressed(paramfilename, model = model, param = param, \
+                                   prof = prof)
         else:
[…]

### 2013-02-19 tex/draft/draft2.tex
added correct draft

%\documentstyle[aas2pp4,epsf]{article}
%%%\documentstyle[aaspp4,epsf]{article}
%\documentstyle[12pt,aasms]{article}    % this is for a preprint
%(single-spaced)
%\documentstyle[aaspp4,epsf]{article} % this is for small print
%\documentstyle[12pt, aaspp4]{article}

%\documentstyle[11pt,aaspp]{article}
%\documentclass[12pt, preprint]{aastex}

%\documentclass[manuscript]{aastex}
\documentclass[apj]{emulateapj}

%\documentclass[12pt, preprint,numberedappendix]{emulateapj}
%\documentstyle[12pt,aasms]{article}    % this is for submittal
                                       % (double-spaced)

%\documentstyle[12pt,aasms]{article}   \usepackage{emulateapj5}

\usepackage{graphicx} 
\usepackage{graphics}                       
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{subeqnarray}
%\bibliographystyle{apj}

\newcommand{\delad}{\nabla_{\rm ad}}
\newcommand{\delrad}{\nabla_{\rm rad}}
\newcommand{\emgr}[1]{\emph{ \color{gray} #1}}

\newcommand{\ie}{i.e.\ }
\newcommand{\eg}{e.g.\ }
\newcommand{\p}{\partial}
\newcommand{\xv}{\vc{x}}
\newcommand{\kv}{\vc{k}}
\newcommand{\brak}[1]{\langle #1\rangle}

\newcommand{\gcc}{\;\mathrm{g\; cm^{-3}}}
\newcommand{\gsc}{\;\mathrm{g\; cm^{-2}}}
\newcommand{\cm}{\; {\rm cm}}
\newcommand{\mm}{\; {\rm mm}}
%\newcommand{\ps}{\; {\rm s^{-1}}}
\newcommand{\km}{\; {\rm km}}
%\newcommand{\au}{\; \varpi_{\rm AU}}

\newcommand{\AU}{\; {\rm AU}}
\newcommand{\yr}{\; {\rm yr}}
\def\K{\; {\rm K}}

\newcommand{\vcs}[1]{\mbox{\boldmath{$\scriptstyle{#1}$}}}
\newcommand{\vc}[1]{\mbox{\boldmath{$#1$}}}
\newcommand{\nab}{\vc{\nabla}}
\DeclareMathSymbol{\varOmega}{\mathord}{letters}{"0A}
\DeclareMathSymbol{\varSigma}{\mathord}{letters}{"06}
\DeclareMathSymbol{\varPsi}{\mathord}{letters}{"09}

\newcommand{\Eq}[1]{Equation\,(\ref{#1})}
\newcommand{\Eqs}[2]{Equations (\ref{#1}) and~(\ref{#2})}
\newcommand{\Eqss}[2]{Equations (\ref{#1})--(\ref{#2})}
\newcommand{\Eqsss}[3]{Equations (\ref{#1}), (\ref{#2}) and~(\ref{#3})}
\newcommand{\App}[1]{Appendix~\ref{#1}}
\newcommand{\Sec}[1]{Sect.~\ref{#1}}
\newcommand{\Chap}[1]{Chapter~\ref{#1}}
\newcommand{\Fig}[1]{Fig.~\ref{#1}}
\newcommand{\Figs}[2]{Figs.~\ref{#1} and \ref{#2}}
\newcommand{\Figss}[2]{Figs.~\ref{#1}--\ref{#2}} 
\newcommand{\Tab}[1]{Table \ref{#1}}

\newenvironment{packed_item}{
\begin{itemize}
  \setlength{\itemsep}{1pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}
}{\end{itemize}}

%\newcommand{\delad}{\nabla_{\rm ad}}
%\newcommand{\delrad}{\nabla_{\rm rad}}
\newcommand{\Rg}{\mathcal{R}}
\newcommand{\RB}{R_{\rm B}}
\newcommand{\co}{_{\rm c}}
\newcommand{\di}{_{\rm o}}
\newcommand{\cb}{_{\rm RCB}}
\newcommand{\surf}{_M}
\newcommand{\mc}{m_{\rm c \oplus}}
\newcommand{\mcn}[1] { m_{ \rm c #1 \oplus} }
\newcommand{\MC}{M_{\rm crit}}
\newcommand{\au}{a_\oplus}
\newcommand{\aun}[1]{ a_{#1\oplus} }

\begin{document}
\bibliographystyle{apj}

\shortauthors{Piso \& Youdin}

\title{On the Minimum Core Mass for Giant Planet Formation through Kelvin-Helmholtz Contraction}
\author{Ana-Maria A. Piso}
\affil{Harvard-Smithsonian Center for Astrophysics}
\author{Andrew N. Youdin}
\affil{JILA, University of Colorado at Boulder}

\begin{abstract}
The core accretion model proposes that giant planets form by the accretion of gas onto a solid protoplanetary core. Previous studies have found that there exists a ``critical core mass'' past which hydrostatic solutions can no longer be found and unstable atmosphere collapse occurs. In standard calculations of the critical core mass, planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse. In this study we consider the extreme case in which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the planet. We develop a two-layer atmosphere model with an inner convective region and an outer radiative zone that matches onto the protoplanetary disk, and we determine the minimum core mass for a giant planet to form within the typical disk life timescale for a variety of disk conditions, which we denote as  ``critical core mass''.  We find that the absolute minimum core mass required to nucleate atmosphere collapse within the disk lifetime is smaller for planets forming further away from their host stars. Moreover, the critical core mass is strongly dependent on disk temperature, opacity and mean molecular weight of the gas. Our results yield lower mass cores than corresponding studies for large planetesimal accretion rates. We therefore show that it is easier to form a planet by growing the core first, then accreting a massive gaseous envelope, rather than forming the core and atmosphere simultaneously.

\end{abstract}

\section{Introduction \textbf{(Paper II)}}
\label{intro}

Current theories of giant planet formation postulate that these planets form either through core accretion (refs), in which solid planetesimals collide and grow into a massive solid core, which then accretes a gaseous envelope, or due to a gravitational instability in the protoplanetary disk that leads to fragmentation of the disk into self-gravitating clumps (refs, inc. \citealt{dangelo11}).  %quote murray-clay, kratter 10, rafikov 05, etc.

Standard core accretion models (refs) assume that the core and the atmosphere grow at the same time, and that planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse, while the heat generated by the gravitational (Kelvin-Helmholtz) contraction of the atmosphere is neglected. These studies consider that the planet atmosphere is in steady state, in which all the luminosity due to planetesimal accretion is radiated away by the envelope, and  find that there exists a minimum (``critical'') core mass past which hydrostatic solutions can no longer be found and unstable atmosphere collapse occurs.

Forming giant planets at wide separations in the disk poses theoretical challenges. On the one hand, gravitational instability generates objects that are too massive to explain the current observed properties of exoplanets (refs, inc. \citealt{rafikov05}). On the other hand, planetesimal accretion is slow at large distances in the disk, and therefore large cores may not be able to form before the dissipation of the disk (refs). I would therefore be easier if giant planets could form from smaller cores which would need less time to grow.

In this study, we show that giant planets can grow faster from small protoplanetary cores that are fully formed before significant gas accretion occurs. In this scenario, the planetesimal accretion rate is significantly slowed down during the gas contraction phase of the atmosphere. This reduction can arise due to dynamical clearing, or due to the core having formed in the inner parts of the disk and migrated outwards, etc. In this situation, the atmosphere evolution is dominated by the Kelvin-Helmholtz contraction of the envelope. The atmosphere is no longer in a steady state, but rather it accretes gas as it loses energy through radiation.

In our model we therefore assume that the luminosity evolution of the atmosphere is dominated by gas contraction, while the planetesimal accretion rate is negligible. As a result, the protoplanetary core has a fixed mass. We consider that the atmosphere evolves in time through stages of quasi-static equilibrium. Once the mass of the gaseous envelope becomes comparable to the mass of the solid core, the self-gravity of the atmosphere can no longer balance the pressure gradient and unstable hydrodynamic collapse commences. The time required for the atmosphere to grow to this stage is the characteristic growth time of the atmosphere. For a set of fixed gas and disk conditions, there exists a minimum core mass for which the atmosphere can grow on the time scale described above within the life time of the protoplanetary disk, which we define as the ``critical core mass''.

We develop a two-layer atmosphere model, with a convective inner region and a radiative outer region that matches smoothly on to the protoplanetary disk, and develop a cooling model that evolves the atmosphere in time. We aim to find the critical core mass for a giant planet to form before the dissipation of the disk.

This paper is organized as follows. In section \ref{sec2} we describe the assumptions of our atmosphere model, and derive the basic equations that govern the structure and evolution of the atmosphere. In section \ref{analytic}, we present a simplified analytic model that predicts the qualitative behavior of the numerical model. We describe our results in section \ref{KH}, and determined the critical core mass for planet formation during the life time of the protoplanetary disk in section \ref{critical}.  We discuss our results in section \ref{discussion} and summarize our findings in section \ref{conclusions}.

   %We assume that the luminosity is primarily generated in the convective zone of the atmosphere,

%Giant planets play a fundamental role in shaping the orbital structure of planetary systems, and in affecting the delivery of volatiles to terrestrial planets in the

\section{Atmosphere Models \textbf{(Paper I)}}
\label{sec2}

In this section we derive the structure and evolution of a planetary atmosphere embedded in a protoplanetary disk in the absence of planetesimal accretion. In section \ref{model} we describe the assumptions of our model. In section \ref{scales} we describe the properties of our assumed protoplanetary disk and review relevant length scales. Section \ref{struct} presents the equations describing the structure of a static planetary atmosphere and the boundary conditions that we use. In section \ref{cooling} we develop a model to estimate the rate at which the atmosphere cools. Finally, we combine these calculations into a quasi-static model for atmospheric evolution in section \ref{twolayer}.

\subsection{Basic Model Assumptions}
\label{model}

We develop a two-layer atmosphere model with the following assumptions:

\begin{enumerate}
\item The planet consists of a solid core of fixed mass and a two-layer atmosphere composed of an inner convective zone and an outer radiative zone that matches smoothly on to the disk, as mentioned in section. The two regions are separated by the Schwarzschild criterion for convective instability (see section \ref{struct}). We denote the surface between the two regions as the radiative-convective boundary (RCB), which is henceforth defined by a radius $r=R_{RCB}$.
\item The luminosity evolution of the atmosphere is dominated by Kelvin-Helmholtz contraction rather than planetesimal accretion. 
\item The luminosity is assumed to be constant throughout the outer radiative region. Since the structure of the convective zone is independent of its luminosity, we do not need to make additional assumptions about the luminosity in this region.
\item The envelope evolves through stages of quasi-static equilibrium.
\end{enumerate}

We use assumptions for the atmosphere geometry and composition similar to those of previous studies (e.g., \citealt{ikoma00}, \citealt{pn05}), as follows:

\begin{enumerate}
\item The atmosphere is self-gravitating, spherically symmetric and in hydrostatic balance.
%\item The outer boundary of the envelope matches smoothly onto the surrounding nebula.
\item The atmosphere consists of a hydrogen-helium mixture, with hydrogen and helium mass fractions of 0.7 and 0.3, respectively.
\end{enumerate}

The time dependence of the atmosphere structure equations may be neglected or explicitly taken into account. Some previous studies of atmosphere accretion (e.g., \citealt{stevenson82}, \citealt{wuchterl93}, \citealt{rafikov06}) consider static envelopes, in which the luminosity is solely supplied by planetesimal accretion and fully radiated away by the atmosphere. In other studies, the time evolution is explicitly taken into account and full time dependent models are developed (e.g., \citealt{ikoma00}). We follow an intermediate approach and consider quasi-static evolution. Our model for the atmosphere growth time is described in section \ref{cooling}.

\subsection{Disk Model and Length Scales}
\label{scales}

As a disk model, we use the minimum mass, passively irradiated model of  \citet{chiang10}. In this model, the surface density and mid-plane temperature are given by

\begin{subeqnarray}
\label{eq:diskparam}
\Sigma_d&=&2200 F_{\Sigma} a^{-3/2}\,\, \text{g cm}^{-2} \slabel{eq:diska}\\
T_d &=& 120 F_T a^{-3/7} \,\,K, \slabel{eq:diskb}
\end{subeqnarray}

\noindent with $a$ the semi-major axis in AU, and $F_{\Sigma}$ and $F_T$ normalization factors that adjust the disk mass and temperature relative to the minimum mass solar nebula (MMSN). In this study, we assume $F_{\Sigma}=F_T=1$. The resulting mid-plane pressure is given by

\begin{equation}
\label{eq:Pd}
P_d=1.1 \times 10^{-4} F_{\Sigma} \sqrt{F_T m_*} a^{-45/14} \,\, \text{dyne cm}^{-2}
\end{equation}

\noindent for a molecular weight $\mu=2.35$. Here $m_* \equiv M_*/M_{\odot}$, where $M_*$ is the mass of the central star and $M_{\odot}$ is the mass of the Sun. We take $m_*=1$.

We further define several characteristic length scales that are important for this problem. \\

\textbf{The Core Radius} represents the physical radius of the protoplanetary core:

\begin{equation}
\label{eq:rc}
R_c \equiv \Big(\frac{3 M_c}{4 \pi \rho_c}\Big)^{1/3},
\end{equation}

\noindent with $M_c$ and $\rho_c$ the core mass and density, respectively. Since we assume that the core mass does not change with time, the core radius is also fixed for a given model. \\

\textbf{The Bondi Radius} represents the distance from the planet at which the thermal energy of the nebular gas is of the order of the gravitational energy binding the gas to the planet. It is defined as

\begin{equation}
\label{eq:RB}
R_B \equiv \frac{G M_p}{c_s^2}=\frac{G M_p}{\mathcal{R} T_d},
\end{equation}

\noindent where $G$ is the gravitational constant, $M_p$ is the total mass of the planet, $c_s$ is the sound speed, and $\mathcal{R}$ is the reduced gas constant: $\mathcal{R}=k_b/(\mu m_p)$, with $k_b$ the Boltzmann constant and $m_p$ the proton mass. \\

\textbf{The Hill Radius} represents the distance at which the gravitational attraction of the planet and of the host star are comparable. It is given by

\begin{equation}
\label{eq:RHill}
R_H \equiv a \Big(\frac{M_p}{M_{\odot}}\Big)^{1/3}
\end{equation}

\noindent Outside the Hill radius, the gravity of the core is too weak to affect the nebular gas. Consequently, the planet will be able to accrete only gas that lies within its Hill sphere. \\

\textbf{The Disk Scale Height} is defined by

\begin{equation}
\label{eq:H}
H \equiv \frac{c_s^2}{\Omega},
\end{equation}

\noindent where $\Omega$ is the Keplerian angular velocity. The disk scale height is a measure of the thickness of the protoplanetary disk. Using equation (\ref{eq:diska}), we find that the scale height for our assumed disk model can be expressed as

\begin{equation}
\label{eq:H2}
H=0.022 \sqrt{F_T} a^{9/7} \,\, \mathrm{AU}
\end{equation}

For small planet masses, $R_B<R_H<H$. As the atmosphere becomes more massive, tidal truncation becomes more important than thermal constraints, and $R_H<R_B<H$. Finally, for high mass atmospheres the disk scale height becomes smaller than the Hill radius of the planet, and spherical symmetry breaks, as the planet is now affected by the vertical profile of the disk.

%Some previous studies of atmosphere accretion (e.g., \citealt{stevenson82}, \citealt{wuchterl93}, \citealt{rafikov06}) consider static envelopes, in which the luminosity is solely supplied by planetesimal accretion and fully radiated away by the atmosphere. In other studies, the time evolution is explicitly taken into account and full time dependent models are developed (e.g., \citealt{ikoma00}). We follow an intermediate approach and consider quasi static evolution. Our model for the atmosphere growth time is described in section \ref{...}.

%For the atmosphere structure, we develop a two-layer atmosphere model with the following assumptions:
%
%\begin{enumerate}
%\item The planet consists of a solid core of fixed mass and a two-layer atmosphere composed of an inner convective zone and an outer radiative zone that matches smoothly on to the disk, as mentioned above. The two regions are separated by the Schwarzschild criterion for convective instability (e.g., \citealt{thompson06}) 
%\item The luminosity evolution of the atmosphere is dominated by Kelvin-Helmholtz contraction rather than planetesimal accretion. 
%\item The luminosity is assumed to be constant throughout the outer radiative region.
%\item The envelope evolves through stages of quasi-static equilibrium.
%\end{enumerate}

\subsection{Structure Equations and Boundary Conditions}
\label{struct}

An atmosphere in hydrostatic equilibrium is fully described by the following structure equations:

\begin{subeqnarray}
\label{eq:struct}
\frac{dP}{dr}&=&-\frac{G m}{r^2}\rho \slabel{eq:structa} \\
\frac{dm}{dr}&=&4 \pi r^2 \rho\slabel{eq:structb} \\
\frac{dT}{dr}&=&\nabla \frac{T}{P}\frac{dP}{dr}\slabel{eq:structc} \\
\frac{dL}{dr}&=&4 \pi r^2 \rho \epsilon_g\slabel{eq:structd}, 
\end{subeqnarray}

\noindent where $r$ is the radial coordinate, $P$, $T$ and $\rho$ are the gas pressure, temperature and density, respectively, $m$ is the mass enclosed by the radius $r$, $L$ is the luminosity from the surface of radius $r$, and $\epsilon_g \equiv -T \frac{ds}{dt}$ represents the heating per unit mass due to gravitational contraction, with $s$ the specific gas entropy. The temperature gradient $\nabla \equiv \frac{d \ln T}{d \ln P}$ has different expressions depending on the primary means of energy transport throughout the atmosphere. We assume that energy can be transported either through radiation or convection. When the luminosity is carried by radiative diffusion, the temperature gradient is given by

\begin{equation}
\label{eq:delrad}
\nabla = \delrad \equiv \frac{3 \kappa P}{64 \pi G m \sigma T^4} L,
\end{equation}

\noindent where $\sigma$ is the Stefan-Boltzmann constant and $\kappa$ is the opacity. Alternatively, when the energy is transported outwards through convective motions, the temperature gradient becomes

\begin{equation}
\label{eq:delad}
\nabla = \delad \equiv \Big(\frac{d \ln T}{d \ln P}\Big)_{\mathrm{ad}},
\end{equation}
\\

\noindent where $\delad$ is the adiabatic temperature gradient. For an ideal gas of adiabatic index $\gamma$, $\delad=\frac{\gamma-1}{\gamma}$. The process that dominates energy transport throughout the atmosphere is determined by the Schwarzschild criterion (e.g., \citealt{thompson06}): the atmosphere is stable against convection when

\begin{equation}
\label{eq:schwarz}
\nabla < \delad
\end{equation}

\noindent and convectively unstable when the reverse is true. In order for convection to be effective, $\nabla \approx \delad$. Therefore, we find that the temperature gradient is given by $\nabla=\mathrm{min}(\delad, \delrad)$.

In order for equation set (\ref{eq:struct}) to be solvable, it has to be supplemented by an equation of state (EOS) that relates pressure, temperature and density $P=P(\rho, T)$, and an opacity law for $\kappa$. In our study, we assume an ideal gas polytropic EOS

\begin{equation}
\label{eq:polyEOS}
P=K \rho^{\gamma}=K \rho^{1/(1-\delad)},
\end{equation}

\noindent with $K$ the adiabatic constant. The adiabatic gradient is $\delad=2/5$ for a monatomic gas and $\delad=2/7$ for a diatomic gas. Additionally, the ideal gas law relates the pressure, temperature and density through

\begin{equation}
\label{eq:idealgas}
P=\rho \mathcal{R} T,
\end{equation}

\noindent with $\mathcal{R}$ defined in section \ref{model}. The specific gas constant $\mathcal{R}$ depends on the mean molecular weight of the gas $\mu$. $\mu$ is determined by the relative abundance of hydrogen and helium in the atmosphere. In this study we assume that the helium fraction only affects the molecular weight of the gas, and not the adiabatic index. We show in section \ref{critcore} that this assumption does not affect the qualitative behavior of our solutions.

We assume an opacity power law of the form

\begin{equation}
\label{eq:opacitylaw}
\kappa=\kappa_0 \Big(\frac{P}{P_d}\Big)^{\alpha} \Big(\frac{T}{T_d}\Big)^{\beta},
\end{equation}

\noindent with $\alpha$ and $\beta$ constants, and $\kappa_0=2 F_{\kappa}$. To estimate $\alpha$ and $\beta$ we use the \citet{bell94} opacity laws for ice grains: $\alpha =0 $, $\beta=2$ and $F_{\kappa}=1$ (i.e., $\kappa_0=2$). We note that these values are valid only for low disk temperatures: $T_d< \sim 100 K$. As such, we only consider cool atmospheres that form in the outer part of the protoplanetary disk ($a \geq 10$ AU).

\subsubsection{Boundary Conditions}
\label{BCs}

We assume a solid core of fixed mass $M_c$ with a radius given by equation (\ref{eq:rc}). We choose $\rho_c=3.2$ g cm$^{-3}$ (e.g., \citealt{pap99}). Furthermore, we assume the atmosphere extends out to the boundary of the Hill sphere, defined by equation (\ref{eq:RHill}). If the Bondi radius (defined in equation (\ref{eq:RB})) is within the Hill radius, perturbations of the nebular gas still occur past the Bondi radius due to the gravitational influence of the core. If, on the other hand, $R_B>R_H$, then the Hill radius is the more relevant truncation radius since material cannot be gravitationally bound to the protoplanet outside the Hill sphere. As such, it is always safe to choose the Hill radius as the outer boundary.

As stated in section \ref{model}, the atmosphere matches smoothly onto the disk at the outer boundary. The temperature and pressure at the Hill radius have to therefore be given by the nebular temperature and pressure: $T(R_H)=T_d$ and $P(R_H)=P_d$.

%\subsection{Standards Methods of Solution}

%Analogously to the stellar evolution case, simply integrating the structure equations (\ref{eq:struct}) from one boundary to the other is not possible, since the boundary conditions are given both at the center and at the surface. In this case, the standard procedures for numerical integration are the shooting method or the Henyey method (\citealt{kippenhahn90}). The shooting method solves the boundary value problem by reducing it to an initial value problem: trial values are chosen for the parameters at one of the boundaries, then the equations are integrated and the resulting values at the other boundary are compared to the actual boundary conditions. The procedure is repeated until convergence is achieved. Alternatively, inward and outward integrations are carried to an intermediate fitting point, where they are fitted smoothly to each other. In the Henyey method, a trial solution for the whole interval is initially guessed, then gradually adjusted through subsequent iterations until the desired level of accuracy is achieved. In this study we use the shooting method --- we integrate inwards from the disk, and match at the core. The detailed numerical procedure is described in section \ref{twolayer}.

\subsection{Virial Equilibrium and Global Cooling Model}
\label{cooling}

Section \ref{struct} provides the structure model for a static atmosphere. We now derive the time evolution of the atmosphere due to Kelvin-Helmholtz contraction between subsequent static models.

We first describe the general ways in which planetary Klevin-Helmholtz contraction differs from the stellar case.

An isolated sphere, such as a star, satisfies a simple global energy equation:

\begin{equation}
\label{eq:coolingstar}
L=\Gamma - \dot{E},
\end{equation}

\noindent where the total luminosity $L$ is balanced by the rate of heat generation $\Gamma$ (due to nuclear fusion in the case of star) and the rate at which total energy is lost $\dot{E}$. The total energy of the sphere is given by the sum of the gravitational and internal energies.

A protoplanetary atmosphere that is not isolated, but is embedded in a gas disk, satisfies a more complex cooling equation:

\begin{equation}
\label{eq:coolingglobal}
L=L_c+\Gamma-\dot{E}+e_{\mathrm{acc}}\dot{M}-P_M \frac{\partial V_M}{\partial t}
\end{equation}

This cooling model applies at any radius $R$ where the mass enclosed is $M$, for example the Bondi radius or the Hill radius. $L$, $\Gamma$ and $\dot{E}$ are the same as in equation (\ref{eq:coolingstar}), and include the atmosphere from the core to the top boundary. The gravitational energy $E_G$ of an atmosphere with total mass $M$ is given by

\begin{equation}
\label{eq:Eg}
E_G=-\int_{M_c}^M \frac{G m}{r} dm,
\end{equation}

\noindent while the internal energy is

\begin{equation}
\label{eq:U}
U=\int_{M_c}^M u dm,
\end{equation}

\noindent where $u$ is the internal energy per mass. We now explain the additional terms in the cooling equation (\ref{eq:coolingglobal}). $L_c$ is the luminosity from the solid core, and may include planetesimal accretion and radioactive decay. $e_{\mathrm{acc}}$ is the specific total energy brought in by mass accreting at the rate $\dot{M}$: $e_{\mathrm{acc}}=u-G M/R$. The last term represents the work done on a surface mass element.

The global cooling equation can be derived starting from the local cooling balance. Integrating the local energy equation and (\ref{eq:structd}) from the core to the surface and using equation (\ref{eq:structb}) yields

\begin{subeqnarray}
\label{eq:coolinglocal}
L-L_c&=&\int_{M_c}^M \frac{\partial L}{\partial m} dm \\
&=&\int_{M_c}^M\Big(\epsilon - T \frac{\partial s}{\partial t}\Big) dm \slabel{eq:clb} \\
&=&\Gamma-\int_{M_c}^M\frac{\partial u}{\partial t}dm + \int_{M_c}^M \frac{P}{\rho^2}\frac{\partial{\rho}}{\partial t} dm \slabel{eq:clc},
\end{subeqnarray}

\noindent where $\Gamma=\int_{M_c}^M \epsilon dm$ and equation (\ref{eq:clc}) follows from equation (\ref{eq:clb}) through the first law of thermodynamics.

Moreover, a self-gravitating protoplanetary atmosphere has to satisfy virial equilibrium. The virial theorem is derived by integrating the equation of hydrostatic balance (\ref{eq:structa}) and has the form

\begin{equation}
\label{eq:virial}
E_G=-3 \int_{M_c}^M \frac{P}{\rho} dm + 4 \pi (R^3 P_M-R_c^3 Pc)
\end{equation}

Similarly to the cooling equation, it applies at any radius $R$ where the total mass enclosed is $M$. For an ideal gas of adiabatic index $\gamma$, equation (\ref{eq:virial}) leads to the following expression for the total energy:

\begin{subeqnarray}
E&=&(1-\xi)U+4 \pi (R^3 P_M-R_c^3 Pc) \slabel{eq:vira} \\
&=&\frac{\xi-1}{\xi}E_G+\frac{4 \pi}{\xi} (R^3 P_M-R_c^3 Pc) \slabel{eq:virb}
\end{subeqnarray}

The global cooling equation (\ref{eq:coolingglobal}) can therefore be derived using equations(\ref{eq:coolinglocal}) and (\ref{eq:virial}). A detailed derivation is deferred to  \App{sec:virial}.

\subsection{Quasi-Static Two-Layer Model}
\label{twolayer}

%The model and equations presented in this section so far can be applied to any protoplanetary atmosphere that satisfies the assumptions listed in section \ref{model}.

As a consequence of the equations derived in section \ref{cooling}, both the atmosphere structure and the gas accretion rate are uniquely determined by the current atmosphere mass. As this mass accretion rate is slow compared to the time it takes to relax to this solution, we can make a quasi-static model of the atmosphere growth. In this section we therefore describe the procedure to obtain an evolutionary series from the cooling model introduced in \ref{cooling} between consecutive two-layer static atmospheres (cf. \ref{model}).

We use the boundary conditions and opacity laws described in section \ref{struct}. The numerical integration is performed through the shooting method, which solves the boundary value problem by reducing it to an initial value problem: trial values are chosen for the parameters at one of the boundaries, then the equations are integrated and the resulting values at the other boundary are com- pared to the actual boundary conditions. The procedure is repeated until convergence is achieved. We start with a total atmosphere mass $M_i$, where the index $i$ labels each evolutionary stage. For this mass we guess a trial luminosity $L=L_{\mathrm{guess}}$. The assumption of constant luminosity throughout the radiative zone sets the right hand side of equation (\ref{eq:structb}) to zero, i.e. the time dependence is neglected. We integrate the structure equations (\ref{eq:struct}) inwards from $R_{\mathrm{out}}=R_H$, with the boundary conditions $T(R_{\mathrm{out}})=T_d$ and $P(R_{\mathrm{out}})=P_d$. The numerical integration gives a value for the core mass implied by the trial solution $M_c=M_{c, \mathrm{guess}}$. We adjust $L_{\mathrm{guess}}$ until $M_{c, \mathrm{guess}}$ converges to the actual core mass $M_c$.

The time evolution of the atmosphere is obtained using the global cooling model described in section \ref{cooling}. We neglect the luminosity due to planetesimal accretion and direct heat generation; as such, the first two terms in equation (\ref{eq:coolingglobal}) are set to zero. Equation (\ref{eq:coolingglobal}) is then evaluated at the RCB, since we assume that luminosity is primarily generated in the convective zone. In what follows, $X_{RCB}$ denotes quantity $X$ evaluated at the RCB. The time step $\Delta t_i$ between two consecutive models of masses $M_i$ and $M_{i+1}$ is calculated as follows:

\begin{eqnarray}\nonumber
\label{eq:dti}
<L>_i \Delta t_i&=&-\Delta E_{RCB,i} + <e_{\mathrm{acc}, RCB}>_i \Delta M_{RCB, i} \\
&&  - <P_{RCB}>_i \Delta V_{RCB, M, i} 
\end{eqnarray}

\noindent where we denote

\begin{subeqnarray}
\Delta X_i&=&X_{i+1}-X_i \\
<X>_i&=&\frac{X_i+X_{i+1}}{2}
\end{subeqnarray}
\\
\noindent The subscript $M$ in the volume expression above signifies evaluating the change in volume at constant mass.

By connecting sets of subsequent static atmospheres through the procedure described above we therefore obtain an evolutionary atmosphere series.

\section{Analytic Cooling Model \textbf{(Paper I)}}
\label{analytic}

In this section we derive a simplified analytic model for the atmosphere structure and evolution, based on the assumptions and equations described in section \ref{sec2}. We use this model to predict the qualitative behavior of our more detailed numerical model. %, and to explain expected scalings (\textbf{Maybe take out this sentence.})

We consider a two-layer atmosphere structure model as presented in section \ref{model}. We approximate the outer radiative zone of the atmosphere as nearly isothermal. Moreover, we do not take into account the self-gravity of the atmosphere in either region. In the next subsections we derive the structure and cooling behavior of such an atmosphere, deferring some of the calculations to \App{sec:analytic}.

\subsection{Isothermal Atmosphere}
\label{iso}

We now consider the structure of a low mass (non self-gravitating) isothermal atmosphere.  We assume the atmosphere matches onto a constant background density, $\rho\di$, at a distance $r_{\rm fit} = n_{\rm fit} \RB$.  The resulting density profile is
\begin{equation} \label{eq:rhoiso}
\rho = \rho_{\rm d} \exp \left({R_B \over r} - {1 \over n_B} \right) \approx   \rho_{\rm{d}} \exp \left(R_B \over r  \right),
\end{equation} 
where the approximate inequality holds deep inside the atmosphere ($r \ll \RB$) for any $n_{\rm fit} \gtrsim 1$.  However the choice of boundary condition does have an order unity effect on the density near the Bondi radius. In our study, $n_{\rm{fit}}=R_{\rm B}/R_{\rm H}$, when the Bondi radius is larger than the Hill radius, and $n_{\rm fit}=1$ when the opposite is true.

The mass of the atmosphere is typically determined by integrating the density profile from the core to the Bondi radius.  Planets can attract massive atmospheres if $\theta\co \equiv R_B/R\co \gg 1$.  In this case
\begin{equation} \label{eq:MatmISO}
M_{\rm iso} \approx 4 \pi \rho\di {R\co^4 \over R_B} e^{R_B/R\co} = 4 \pi \rho\di {R\co^3 \over \theta\co} e^{\theta\co}\, .
\end{equation} 
This result is the leading order term in a series expansion.  Furthermore, because the atmospheric scale-height at $R\co$ is $H_\rho = |dr /d\ln\rho| = R\co^2/R_B$, the result is intuitively the correct order of magnitude.

\subsection{Temperature and Pressure Corrections at the Radiative-Convective Boundary}
\label{RCBcorr}

%\subsection{Temperature Contrast at Convective Boundary}
We now show that the temperature contrast between the radiative-convective boundary $T_{\rm RCB}$ and the ambient disk $T_{\rm d}$, is modest.  From equation (\ref{eq:delrad}), we express the radiative lapse rate
\begin{equation}\label{eq:delradan}
\delrad = {3 \kappa P \over 64 \pi  G M \sigma T^4} L = \nabla\di {(P/P_{\rm d})^{1+\alpha} \over (T/T_{\rm d})^{4-\beta}}
\end{equation}

\noindent where the second equality follows from the opacity law (\ref{eq:opacitylaw}), and $\nabla \di$ is the radiative temperature gradient at the disk:

\begin{equation}
\label{eq:delo}
\nabla \di \equiv \frac{3 \kappa_0 P_{\rm d}}{64 \pi G M \sigma T_{\rm d}^4}.
\end{equation}

\noindent $M$ is the sum of the core mass and the atmosphere mass below the pressure level $P$.  If the mass in the radiative zone is small, then we can hold $M$ fixed at the sum of the core and convective zone masses: $M = M\co + M_{\rm conv}$.   With this assumption and power-law opacity, we get a constant value for $\nabla\di$.  The temperature profile then integrates to
\begin{equation}\label{eq:radTP}
\left(T \over T_{\rm d}\right)^{4-\beta} - 1 = {\nabla\di \over \nabla_\infty} \left[\left({P \over P_{\rm d}}\right)^{1-\alpha} - 1 \right] \, ,
\end{equation} 
where $\nabla_\infty = (1+\alpha)/(4-\beta)$ is $\delrad$ for $T ,P \rightarrow \infty$.\footnote{Our definitions of $\nabla\di$ and $\nabla_\infty$ are precisely opposite to \citet{rafikov06}, but consistent with other works and the general convention of labeling a quantity $f$ in the disk as $f\di$.}
We  apply \Eqs{eq:delradan}{eq:radTP} at the radiative-convective boundary (where $\delrad = \delad$) under the assumption that the pressure there is $P\cb \gg P_{\rm d}$, motivated by the fact that the exponential density and hence pressure profile is isothermal (see equation (\ref{eq:rhoiso})). The resulting temperature contrast at the radiative-convective boundary is
\begin{equation}\label{eq:Tcb}
\chi \equiv {T\cb\over T_{\rm d}} \simeq \left(1 - {\nabla_{\rm ad} \over \nabla_\infty}\right)^{-\frac{1}{4-\beta}} \, .
\end{equation} 
For  $\alpha=0$ and $\beta =2$,  and assuming $\delad = 2/7$, we find $\chi \approx 1.5$. Values of $\chi$ and $\nabla_\infty$ for other choices of $\beta$ are summarized in \App{sec:analytic}.

The pressure at the convective boundary  follows from \Eqs{eq:delrad}{eq:Tcb} as
\begin{equation}
{P\cb\over P_{\rm d}} \simeq \left({\delad/\nabla\di \over 1 - \delad/\nabla_\infty}\right)^{1 \over 1 + \alpha}
\end{equation} 
This pressure contrast can be quite large due to the smallness of $\nabla\di$ in low luminosity atmospheres.

 It is also useful to obtain a relation between $T$ and $P$ that eliminates $\nabla\di$ in favor of $P\cb$:
 \begin{equation}\label{eq:TP}
{T \over T_{\rm d}} = \left\{1 + {1 \over {\nabla_\infty \over \delad} - 1} \left[ \left({P \over P\cb}\right)^{1-\alpha} -  \left({P_{\rm d} \over P\cb}\right)^{1-\alpha}  \right] \right\}^{1 \over 4-\beta}\, .
\end{equation}

We can determine the radius of the convective boundary $R\cb$ from the hydrostatic balance equation as 
\begin{equation}\label{eq:RCBint}
{R_B \over R\cb} = \int_{P\di}^{P\cb} {T \over T_{\rm d}} {dP \over P}\, .
\end{equation} 
An isothermal atmosphere gives a simple logarithmic dependence on $P\cb$.  However using \Eq{eq:TP} in the integral gives
\begin{equation}\label{eq:Rcb}
{R_B \over r\cb} = \ln \left(P\cb \over P\di \right) - \ln \theta \, ,
\end{equation} 
with an extra correction term, $\theta < 1$.  The form we chose for the correction term allows us to relate the disk and radiative-convective boundary pressures as :
 \begin{equation}\label{eq:PcbRcb}
P\cb = \theta P_{\rm d} e^{R_B/R\cb} \, .
\end{equation}   
In the $P\cb \gg P_{\rm d}$ limit the correction term is an order unity constant that depends on $\alpha$, $\beta$ and $\delad$. Similarly to the temperature correction factor $\chi$,  $\theta$ accounts for the fact that the radiative region is not perfectly isothermal. For $\delad=2/7$, we find $\theta(\alpha=0, \beta=20) \approx 0.556$. Values of $\theta$ for other choices of $\beta$ are summarized in \App{sec:analytic}.  A simple analytic expression for $\theta$ is not possible.

\subsection{Simplified cooling model}
We now use a simplified version of the cooling model introduced in section \ref{cooling} in order to obtain an analytic cooling model. As before, we assume a polytropic equation of state and neglect self-gravity.

\subsubsection{Luminosity,  Energy \& Mass}
From equation (\ref{eq:delradan}), the luminosity that emerges at the radiative-convective boundary is
\begin{equation} \label{eq:Lcb}
L\cb = {64 \pi G M\cb \sigma T\cb^4 \over 3 \kappa P\cb } \nabla_{\rm ad} \approx L\di {P_{\rm d} \over P\cb}, % \left({P\di \over P\cb}\right)%^{1+\alpha}
\end{equation} 
where we drop the pressure dependence from the opacity law (\ref{eq:opacitylaw}) and
\begin{equation} 
L\di \equiv {64 \pi G M\cb \sigma T_{\rm d}^4 \over 3 \kappa(T_{\rm d}) P_{\rm d}} \nabla_{\rm ad}\chi^{4-\beta}\, 
\end{equation} 
normalized to disk conditions, with $T = \chi T_{\rm d}$ at the RCB.

 For $\beta = 2$ and $F_\kappa = 1$ in the opacity law (\ref{eq:opacitylaw}), this gives the \citet{bell94} opacity for icy grains.  By varying $F_\kappa$, dust depletion or enhancement can be considered.  Grain properties affect both $F_\kappa$ and $\beta$ which generally satisfies  $1/2 \lesssim \beta \lesssim 2$ (aside from discontinuities across sublimation regions, see \citet{semenov03}).

To make further analytic progress, we ignore the self-gravity in the convective zone, holding the mass fixed at $M\co$.  The density profile of an adiabatic atmosphere follows from hydrostatic balance as 
\begin{equation}\label{eq:rhoconv} 
\rho = \rho\cb \left[ 1 + {R_B' \over r} - {R_B' \over R\cb}  \right]^{1/(\gamma -1)}\, .
\end{equation} 
where we define an effective Bondi radius,
\begin{equation}
R_B' \equiv {G M\co \over C_P T\cb} = {\delad \over \chi} R_B
\end{equation} 
to simplify expressions.

  Deep in the atmosphere, where $r \ll R\cb \lesssim R_B'$ the density profile is $\rho \propto r^{-1/(\gamma -1)}$.  Since the energy scales as $\rho r^2 \propto r^{(2\gamma -3)/(\gamma - 1)}$ only polytropes with $\gamma < 3/2$ (i.e. $\gamma = 7/5$, but not $\gamma = 5/3$) have the bulk of energy at the bottom of the atmospheres.  We will thus focus on the $\gamma = 7/5$ case, even though dissociation occurs deep in real protoplanetary atmospheres.

The total (thermal and gravitational) energy in an adiabatic atmosphere could be evaluated from the virial theorem (see section \ref{cooling}).  More simply, we use the result for temperature profiles in deep (but non-self-gravitating) convective regions:
\begin{equation}
T \approx {G M\co \over C_P r} = T\cb {R_B' \over r}\, .
\end{equation} 
The internal energy per unit mass of an ideal gas is thus $u = C_V T = (1 - \delad) G M\co/r$ and the specific energy  deep in the atmosphere is
\begin{equation}
e = e_g + u = -\delad {GM\co \over r} \, .
\end{equation} 
Thus the total energy for $\gamma < 3/2$ is thus
\begin{eqnarray} 
E &=& - 4 \pi \nabla_{\rm ad} G M\co \int_{R\co}^{r\cb} \rho r dr \\
&\approx& - 4 \pi P\cb {R_B'}^{1 \over \nabla_{\rm ad}} \left(\gamma-1 \over 3 - 2 \gamma\right)  R\co^{2\gamma-3\over \gamma-1}  \\
&\approx& - 8 \pi P\cb {R_B'^{7/2} \over \sqrt{R\co}}
%&\approx& -0.31 P\cb {(R_B / \chi)^{7/2} \over \sqrt{R\co}}
%&\approx& -0.31 P\cb {R_B'^{7/2} \over \sqrt{R\co}}
\end{eqnarray} 
where the final expression takes $\gamma = 7/5$. % and then revert to the standard Bondi radius $R_B = G M\co/(\Rg T\di)$ in terms of the disk temperature.

The mass of the adiabatic atmosphere is given by
\begin{eqnarray} 
M_{\rm atm} &=& 4 \pi \int_{R\co}^{R\cb} \rho r^2 dr \\
&=& {5 \pi^2 \over 4} \rho\cb {R_B'}^{5/2} \sqrt{R\cb} % \approx 0.54 \rho\cb {R_B'}^{5/2} \sqrt{r\cb} \\
%&=& 0.54{R_B'^{3} \sqrt{\chi} \over  \sqrt{ \ln[P\cb/(\theta P\di)]}} \, .
\end{eqnarray}
in the limit $R\co \ll R\cb \ll R_B'$. The mass in the isothermal region is $M_{\rm iso} \sim 4 \pi \rho\cb r\cb^4/R_B \ll M_{\rm atm}$, and can be neglected.

We can eliminate $R\cb$ using \Eq{eq:PcbRcb}.  The ratio of atmosphere to core mass is then  
\begin{equation} \label{eq:crit}
{M_{\rm atm} \over M\co} \approx {P\cb /P_M \over  \sqrt{\ln[ P\cb/(\theta P_{\rm d})]}}
\end{equation} 
where we introduce a characteristic pressure
\begin{equation} 
P_M \equiv {4 \delad^{3/2} \over 5 \pi^2 \sqrt{\chi} } {G M\co^2 \over {R_B'}^4}\, .
%P_M \approx {1.85 \over \sqrt{\chi}} {G M\co^2 \over {R_B'}^4}\, .
\end{equation}

For the atmosphere to initiate unstable gas accretion, $M_{\rm atm} = M\co$. We thus require
\begin{equation} \label{eq:Pcbc}
P\cb = \xi P_M
\end{equation} 
where  the logarithmic factor
\begin{equation}\label{eq:xi}
\xi \equiv \sqrt{\ln[ P\cb/(\theta P_{\rm d})]} = \sqrt{\ln[ \xi P_M / (\theta P_{\rm d})]}
\end{equation} 
is found by numerically solving the above transcendental equation. The physical solution has $\xi >1$, but typically order unity. Numerical solutions exist for $\xi \ll 1$, but are unphysical as they imply $P\cb < P_{\rm d}$.
%  The physical origin of this logarithm is the exponential increase of pressure with depth in the radiative zone.

\subsubsection{Cooling Time}
Neglecting the surface terms from equation (\ref{eq:coolingglobal}), the total time to cool the atmosphere from an initially fully adiabatic state is \footnote{The relevance of the neglected surface terms is discussed in \App{sec:analytic}}
\begin{eqnarray} 
t_{\rm  cool} &=& -\int {d E \over L} = - \int_{P\di}^{P\cb } {d E/d P\cb \over L} dP\cb \\
&\approx& 4 \pi {P\cb^{2} \over P\di} {R_B'^{7/2} \over L\di \sqrt{R\co}} \label{eq:tcool}\, ,
%&\approx& {0.16 }{P\cb^{2} \over P\di} {R_B'^{7/2} \over L\di \sqrt{R\co}}
\end{eqnarray} 
where we neglect self-gravity and take $M\cb = M\co$ in $L\di$.

The cooling timescale  for an atmosphere to become self-gravitating is found from \Eqs{eq:tcool}{eq:Pcbc} as
\begin{eqnarray} 
t_{\rm cool} &\approx & 2 \times 10^8 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2  \over \left(\mc \over 10 \right)^{5/3} \left(a\over 10\right)^{15 \over 14}} \yr
\end{eqnarray} 
for $\beta = 2$ and $\mc \equiv M_{\rm c} / M_{\oplus}$.  
Clearly this timescale is too long, and this is likely due to missing physics of self-gravity and EOS.  Nevertheless the trends are informative.   A reduction in opacity naturally gives faster cooling, as long as the optically thick assumption holds.  Lower disk temperatures also give faster cooling.  Even though higher temperatures give higher luminosities, they also give higher dust opacities and lower the Bondi radius and gas density.  On the other hand,  higher temperatures suppress cooling in this model.    For different $\beta$ values, the temperature dependence is $t_{\rm cool} \propto F_T^{ \beta + 1/2}$.  The cooling timescale only weakly depends on disk mass or pressure, via the  logarithmic factor $\xi$.  This weak dependence is a consequence of cooling coming from the radiative-convective boundary.  Note that the scaling laws do not reflect the changes to $\xi$ which remains order unity. % and is here given the appropriate value for the nominal parameters.

\subsubsection{Critical Core Mass}\label{sec:critmass}
%We use the results of the previous section to derive the minimum core mass that can form a giant planet by gas accretion.  Since our structure calculations do not include the self-gravity of the atmosphere, we simply assume that runaway growth occurs once the atmosphere's mass (as calculated in the absence of self-gravity) reaches the core mass, $M_{\rm atm} \ge M_{\rm c}$.  This approximation was used by R06 and is consistent with detailed evolution models (refs).

We can define a critical more mass as that which gives a self-gravitating atmosphere within a typical gas disk lifetime:
\begin{equation}
t_{\rm cool} = 3 \times 10^6 \tau_{\rm cool} ~{\rm yr} \, ,
\end{equation} 
with $\tau_{\rm cool}$ a scaling factor.  The resulting critical core mass is
\begin{equation}\label{eq:Mcrit}
M_{\rm crit} \approx 100 {F_T^{3/2} F_\kappa^{3/5}   \left(\xi \over 2.6 \right)^{6/5} \over \left(\au \over 10\right)^{9 \over 14}} \; M_\oplus.
\end{equation} 
As with the cooling time, the critical mass is too large due to missing physics.  The scaling with disk properties is similar, in fact all quantities are simply raised to the $3/5$ power.  The $\xi$ value is changed to match the value for the nominal solution.  The lower value reflects the fact that an atmosphere around such a massive core does not need to contract as much to be self-gravitating.

The numerical values for a different opacity law do not change the numerical answers above very much.  That is mostly because we have chosen a disk location where $T_{\rm d}$ is not far from 100 K, where our opacity law is normalized.  More generally a higher $\beta$ value is more favorable for core accretion at large distances due to the sharper drop in opacity. The effects of opacity on the cooling time and critical core mass are discussed in appendix \App{sec:analytic}

%
%The T-P profile can also be combined with the equation of hydrostatic balance to relate the radius and pressure of the convective boundary.  If the convective boundary is well inside the Bondi radius, i.e.\ $r\cb\ll R_B$ and $P\cb\gg P\di$ then
%\begin{equation}\label{eq:PcbRcb}
%P\cb = \theta P\di e^{R_B/r\cb} \, .
%\end{equation}    
%The order unity constant $\theta < 1$ accounts for the fact that the radiative layer is not perfectly isothermal, as $\theta$ would be exactly one in that case.  The value of theta depends on the equation of state and the pressure and temperature dependence of the opacity, i.e. the $\alpha (=0)$ and $\beta$ values.  We find $\theta(\beta = 2) \approx 0.556$ and $\theta(\beta = 1) = 0.286$.  Closed form expressions for these integrals are too cumbersome to be useful.

\section{Quasi-Static Kelvin-Helmholtz Contraction \textbf{(Paper I)}}
\label{KH}

In this section we describe our choices for the gas, disk and core parameters, and present some typical atmosphere numerical models. As mentioned in section \ref{model}, we assume that the central star is a Sun-like star with $M_*=M_{\odot}$. We are primarily interested in the outer regions of the protoplanetary disk, so we generate atmosphere models between 10 and 100 AU. As a fiducial case, we assume that the atmosphere is composed of a hydrogen-helium mixture with the helium mass fraction $Y=0.3$., resulting in a mean molecular weight $\mu=2.35$. We do not take into account the dependence of the adiabatic index on the helium fraction, and therefore assume the adiabatic index for a diatomic gas $\delad=2/7$. We further explore how the atmosphere evolution varies with adiabatic index and molecular weight. We generate model atmospheres for monatomic adiabatic index $\delad=2/5$ and for a purely molecular hydrogen composition ($\mu=2$). The comparison between different gas compositions is deferred to section \ref{critcore}.

Figure 1 shows instantaneous atmosphere radial profiles at $a=10$ AU accreting around a fixed core of mass $M_{\rm{c}}=5 M_{\oplus}$ for several total planet masses. In what follows,  if the $R_B<R_H$ hierarchy holds, we define the planet mass as the total mass enclosed within the Bondi radius of the atmosphere, and regard the gas between the Bondi and Hill radii as perturbations due to the gravitational influence of the planet. For illustrative purposes we mark the position of the Bondi radius and of the radiative convective boundary. We notice that the pressure and temperature in the convective regions of the atmospheres scale as $1/r$, which is the expected behavior of an ideal gas adiabat with a polytopic equation of state (e.g., \citealt{rafikov06}). In the outer radiative region of the atmosphere, the pressure has a nearly exponential profile, which is in agreement with the predictions of the analytic model (see section \ref{analytic} and \App{sec:analytic}). In addition, the atmosphere temperature at the radiative-convective boundary $T_{\rm{RCB}}$ only differs from the disk temperature $T_{\rm{d}}$ by an order unity factor, and so the atmosphere is nearly isothermal throughout the radiative zone. This is also in agreement with the analytic predictions. %, and justifies

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/PTm_profiles_v2.pdf}
%\vspace{-0.5in}
\caption{Example pressure, temperature and mass profiles as a function of radius.}
\end{figure}

The location of the Bondi radius moves further out as the mass of the atmosphere increases, as the Bondi radius scales with mass. The location of the radiative-convective boundary has a more interesting behavior. As stated in section \ref{intro}, the nebular gas initially settles into equilibrium around the protoplanetary core on a short timescale relative to the thermal timescale, and is isentropic throughout with an entropy equal to that of the disk. In the context of our model, this atmosphere represents the minimum atmosphere mass for which a hydrostatic solution exists. This atmosphere is purely convective. As gas accumulates and the atmosphere starts cooling, an outer radiative region forms. This layer is thin for low mass atmospheres, as can be seen in Fig. 1: for the $5.21 M_{\oplus}$ atmosphere, the ratio between the thickness of the radiative and convective regions is low. An increase in atmosphere mass causes the radiative zone to expand and the convective region to contract. As more and more gas is accumulated, the convective zone eventually starts expanding again; however, the ratio between the radii of the radiative and convective regions continually increases with atmosphere mass, i.e. the radiative regions of these types of atmospheres are getting deeper and deeper.  \textbf{[This needs to be rephrased, somewhat vague and repetitive.]}

We further use the global cooling model developed in section \ref{cooling} to obtain an evolutionary series from static atmosphere profiles. Figure 2 shows the luminosity and cooling time evolution with atmosphere mass. Gas accretion is initially rapid, but slows down significantly after a relatively small mass increase. For comparison, we also plot the analytic results derived in section \ref{analytic}. We see that there is only a very brief agreement between the analytic and full numerical results. At early times, the radiative zone is too shallow and so the approximation that $P_{\rm RCB} \gg P_{\rm d}$ used in section \ref{analytic} breaks down. At late times, the self-gravity of the atmosphere becomes important. %We see that self-gravity starts to matter

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Lt_profiles_v2.pdf}
%\vspace{-0.5in}
\caption{Luminosity and time evolution with mass. The analytic result is plotted for comparison.}
\end{figure}

\subsection{The End of Quasi-Static Evolution}
\label{endoftime}

The right-hand side of equation (\ref{eq:dti}) decreases with increasing atmosphere mass, and eventually becomes negative when the mass of the envelope becomes comparable to the core mass $M_{\rm{atm}} \sim M_{\rm{c}}$. This implies a negative luminosity from the RCB. Physically, the luminosity $L$ represents the total heat flux through the surface defined by the RCB. The heat flux due to convective motions is always directed outwards, and hence it is positive. Moreover, the temperature gradient $dT/dr$ is negative, as the temperature decreases towards the outside. As a result (see equation (\ref{eq:structd})), the heat flux due to radiative diffusion is also positive. This implies that, in the absence of planetesimal accretion and internal heat sources, there is no physical mechanism that can justify a negative luminosity. The quasi-static, constant luminosity approximation breaks down at this point and a more detailed, time-dependent model is necessary.

%As can be inferred from Figure 2, the time steps between subsequent evolutionary models become increasingly smaller as the atmosphere grows in mass. Around the time when the atmosphere mass is comparable to the core mass $M_{\mathrm{atm}} \sim M_{\mathrm{c}}$, we find that the time steps become negative.

\section{Critical Core Mass \textbf{(Paper I, except for $\delad=2/5$ case)}}
\label{critical}

Standard static calculations of atmosphere accretion that consider planetesimal accretion as the main source of luminosity (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{rafikov06}) find that an atmosphere can be stable only if the atmosphere mass is lower than the core mass. As the mass of the envelope becomes comparable to the solid core mass, the pressure gradient can no longer balance the self-gravity of the atmosphere, and hydrostatic equilibrium breaks down. As a result, a phase of rapid gas accumulation is initiated, i.e. runaway accretion. Typically, this instability settles in when $M_{\rm{atm}} \sim M_{\mathrm{c}}$. Static calculations do not take into account the time evolution of the atmosphere explicitly, but rather assume a steady state at which the luminosity due to planetesimal accretion is fully radiated away by the atmosphere. Therefore, for a set of disk and gas conditions, a single value for the critical core mass $M_{\rm{crit}}$ is found. This is the maximum mass the core can have before unstable gas accretion commences.

In our evolutionary model the core mass is assumed to be fixed, so the above definition for the critical core mass is not applicable. Nevertheless, we adopt a similar approach in defining the atmosphere mass past which hydrostatic balance no longer holds. We denote this mass as the minimum between the mass at which $M_{\rm{atm}}=M_{\rm{c}}$ and the mass at which the right-hand side of equation (\ref{eq:dti}) becomes negative (see section \ref{endoftime}). We therefore define the evolutionary time (or growth time, or cooling time) of the atmosphere as the time elapsed until runaway accretion commences according to the criteria above. The evolution of atmosphere mass with time is shown in Figure 3 for $a=10$ AU and $M_{\mathrm{c}}=5 M_{\oplus}$. Here, $\Delta t$ is the instantaneous time step between subsequent static atmosphere models. We see that growth is slowest around $M_{\rm{tot}} \sim 6-7 M_{\oplus}$, after which it is continuously accelerated. The estimate of the atmosphere characteristic growth time is therefore insensitive to the exact choice of atmosphere mass to initiate runaway gas accretion, to order of magnitude, as long as it is past the time at which accretion is slowest. %\textbf{Needs a bit of a cleanup as well.}

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mt_profile.pdf}
%\vspace{-0.5in}
\caption{Instantaneous cooling time evolution with total mass. Accretion is slowest at $M_{\mathrm{tot}} \sim 6-7 M_{\oplus}$.}
\end{figure}

\subsection{Role of Disk Temperature and Pressure}
\label{TPeffects}

In what follows we study the separate effect of disk temperature and pressure on the atmosphere growth time. At a fixed distance in the disk $a=10$ AU and for a core mass $M_{\mathrm{c}}=5 M_{\oplus}$, we fix the disk temperature $T_{\rm{d}}$ to the value given by equation (\ref{eq:diskb}) and vary the disk pressure $P_{\rm{d}}$. The resulting time evolution is plotted in Figure 4. We find that the atmosphere evolution time decreases as the disk pressure increases. This effect is likely due to the fact that a higher disk pressure, and hence density, leads to the atmosphere becoming self-gravitating and therefore initiating runaway accretion on a faster timescale. However, increasing the disk pressure by a factor of $\sim 100$ only reduces the atmosphere growth time by a factor of $\sim 2$. As such, the atmosphere cooling time is only weakly dependent on the nebular pressure.

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Pd_effect.pdf}
%\vspace{-0.5in}
\caption{Disk pressure effect on growth time. Gas accretion is faster for higher pressures.}
\end{figure}

We perform a similar analysis on the effect of disk temperature $T_{\rm{d}}$ on the evolutionary time. For $a=10$ AU and $M_{\rm{c}}=5 M_{\oplus}$, we fix the disk pressure at the value given by equation (\ref{eq:Pd}) and vary $T_{\rm{d}}$. The resulting dependence is shown in Figure 5. In contrast with the pressure effect, an increase in disk temperature causes the growth time to increase as well. This is due to opacity effects: a higher disk temperature increases the opacity in the radiative zone of the atmosphere, causing a decrease in luminosity and therefore a longer Kelvin-Helmholtz accretion time. Moreover, a factor of $\sim 3$ reduction in the disk temperature results in an decrease of the growth time by a factor of $\sim 30$. The atmosphere evolution is therefore strongly dependent on the temperature in the protoplanetary disk.

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Td_effect.pdf}
%\vspace{-0.5in}
\caption{Disk temperature effect on growth time. Gas accretion is faster for lower temperatures.}
\end{figure}

\subsection{Core Mass versus Evolution Time at Fixed Distance}
\label{Mct}

In this section we explore how the core mass affects the cooling time of the atmosphere for a fixed semi-major axis. For consistency, we use the same distance as in the previous subsections $a=10$ AU. Figure 6 shows the growth time behavior for a range of cores with masses between 1 and 10 $M_{\oplus}$. As expected, growth happens on a faster timescale if the core is more massive, as it can bind a gaseous atmosphere quicker due to its higher gravitational attraction. We find that the atmosphere growth time is very sensitive to the core mass: the core mass increase needed to reduce the cooling time by an order of magnitude is only half an Earth mass.

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cumul_coolingtime_vs_Matm_10au_mu235.pdf}
%\vspace{-0.5in}
\caption{Time evolution as a function of atmosphere mass for a range of core masses. The circles mark the time when runaway accretion is initiated.}
\end{figure}

We further analyze how the thermodynamic properties of the gas affect the atmosphere growth time. Specifically, we compare our standard model given by $\delad=2/7$ and $\mu=2.35$ with models with a different adiabatic index and mean molecular weight. This is shown in Figure 7. We discuss the effects of adiabatic index and molecular weight separately.

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/coolingtime_vs_Mc_10au_new.pdf}
%\vspace{-0.5in}
\caption{Time evolution for atmospheres with different adiabatic indices and mean molecular weight. A typical protoplanetary disk life time is plotted for comparison.}
\end{figure}

A decrease in the mean molecular weight of the gas results in a larger cooling time for the same core mass, since an atmosphere composed of heavier molecules is easier to bind. Moreover, a gas with a higher adiabatic gradient $\delad$ is more dense, and therefore more likely to be bound by the gravity of the planet.

\subsection{Critical Core Mass}
\label{critcore}

In the previous subsections we explored and discussed the effect of various parameters on the growth time of an atmosphere accreting around a protoplanetary core of fixed mass. The main question we address right now is under what conditions can such an atmosphere grow before the gas in the protoplanetary disk is dissipated. Specifically, we determine the minimum core mass that is necessary for a protoplanet to initiate runaway gas accretion within the disk lifetime. We define this mass as the \textit{critical core mass} for runaway growth.

Observations of protoplanetary disks show that they dissipate over a time scale of a few million years (e.g., \citealt{lagrange00}, \citealt{haisch01}, \citealt{goldreich04}). We adopt a disk life timescale of $\tau=3 $ Myrs and determine the critical core mass as a function of gas and disk conditions. The results are displayed in Figure 8.

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_new2.pdf}
%\vspace{-0.5in}
\caption{The minimum core mass for an atmosphere to become ÔcriticalÕ within the lifetime of the disk as a function of semi-major axis, for different adiabatic indices and molecular weight of the nebular gas.}
\end{figure}

The critical core mass decreases as we move further away in the disk. This is due to several reasons:

\begin{itemize}
\item The disk temperature $T_d$ is lower in the outer regions of the disk. As we showed in section \ref{TPeffects}, the atmosphere grows faster for lower temperatures, which offsets the need for a larger core mass. 
\item The gravitational attraction of the central star is  weaker at larger distances,  and hence the gravity of the protoplanet can act on a wider range. %resulting in a larger effective radius for the planet (i.e.m the )As a result, the Hill radius of the planet (see equation \ref{eq:RH}) is larger, and hence the gravity of the planet can 
\end{itemize}

Moreover, the critical core mass increases for a lower mean molecular weight and a smaller adiabatic gradient. This is consistent with the results derived in section \ref{Mct}. The critical core mass is strongly dependent on the mean molecular weight $\mu$, but only weakly depended on the adiabatic index $\gamma$ (or $\delad$).

The comparison between our results and standard values for the critical core mass due to planetesimal accretion is deferred to section \ref{acc}. % \textbf{Need to add qualitative explanations for different slopes and curve intersection}

\subsection{Opacity Dependence}
\label{opacity}

 In this section we investigate the effect of opacity reduction on the atmosphere evolution, since a lower opacity in the outer atmosphere regions accelerates gas accretion, resulting in a lower critical core mass. We explore opacity reductions by factors of 10 and 100, while keeping the power law (\ref{eq:opacitylaw}) coefficient constant $\beta=2$. The result is depicted in Figure 9.

 We find that the atmosphere luminosity and growth time are increased and decreased, respectively, by factors of the same order as the opacity reductions.

 \begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/opacity_effect.pdf}
%\vspace{-0.5in}
\caption{Comparison between standard dust opacity and opacity reduced by factor of 10 and 100, respectively. Lower opacity accelerates atmosphere growth.}
\end{figure}

\section{Discussion}
\label{discussion}

In this section we discuss the parameter space of validity of our model. We calculate the conditions under which planetesimal accretion can be ignored, and review the effects of some of the approximations that come into our model.

\subsection{Comparison with Planetesimal Accretion \textbf{(Paper II)}}
\label{acc}

In this study we have considered atmospheres for which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the atmosphere. This is different from standard calculations, in which the atmosphere is heated by planetesimal accretion. In this section we investigate the core accretion rates that are necessary for our regime to be valid. We also discuss the conditions under which runaway gas accretion can be initiated due to the Kelvin-Helmholtz contraction of the atmosphere before it becomes critical due to planetesimal accretion.

We estimate the planetesimal accretion rate consistent with our assumptions that $L_{\mathrm{acc}} \ll L_{\rm{KH}}$. Here $L_{\rm{acc}}$ is the accretion luminosity given by

\begin{equation}
\label{eq:Lacc}
L_{acc}=G \frac{M_{\rm{c}} \dot{M_{\rm{c}}}}{R_{\rm{c}}},
\end{equation}

\noindent where $\dot{M_{\rm{c}}}$ is the core accretion rate, and $L_{\rm{KH}}$ is the luminosity of the atmosphere due to gas contraction as calculated in section \ref{KH}. At the limit, $L_{\rm{acc}}=L_{\rm{KH}}$. For a given atmosphere model we can therefore estimate the maximum planetesimal accretion rate during the gas contraction phase in order for the atmosphere to be dominated by the Kelvin-Helmholtz luminosity. We choose as a fiducial case an atmosphere forming at 10 AU and with a core mass of $10 M_{\oplus}$. For this choice or parameters, the atmosphere cooling time is $\sim$ 1.2 Myrs (see Figure 7), which is within the typical life time of a protoplanetary disk. The results are presented in Figure 10.

 \begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/acc_rates_poster.pdf}
%\vspace{-0.5in}
\caption{Various relevant accretion rates. $\dot{M}_{\rm{atm}}$ is the growth rate of the atmosphere as estimated by our model, and $\dot{M}_{\rm{c,KH}}$ is the maximum planetesimal accretion rate during the gas contraction phase in order for our regime to be valid. For comparison, we plot the core accretion rate $\dot{M}_{\rm{c,acc}}$ necessary to grow the core on the same time scale as the atmosphere $\tau \sim 1.2$ Myrs, and a typical planetesimal accretion rate where the random velocity of the planetesimals is given by the Hill velocity due to the core.}
\end{figure}

We label the resulting minimum core accretion rate as $\dot{M}_{\rm{c,KH}}$. The atmosphere growth rate is also plotted for comparison. We therefore see that the core accretion rate has to be $\sim2-3$ orders of magnitude lower than the atmosphere accretion rate for our assumptions to be valid.  If the core had accreted planetesimals at this constant rate since it started forming, then the formation of a core massive enough to attract an atmosphere would not have been possible within typical disk life timescales, which indicates a slowing down of the planetesimal accretion regime. Possible explanations for that include the core having formed in the inner part of the disk and later migrated outwards, or the core having been depleted of planetesimals due to a giant neighbor. We also estimate the core accretion rate

\begin{equation}
\label{eq:Mcdot}
\dot{M}_{\rm{c,acc}}(M_{\rm{c}}) \equiv \frac{M_{\rm{c}}}{\tau}
\end{equation}

\noindent needed for the core to form on the same timescale as our model atmosphere, $\tau=1.2$ Myrs, as well as a typical planetesimal accretion rate, for which the random velocities of the planetesimals are of the order of the Hill velocity around the protoplanetary core \citep{goldreich04}. We denote this latter rate as $\dot{M}_{\rm{c,typical}}$. This rate, which is at the boundary of the dispersion dominated and shear dominated regimes ,is often quoted as a maximum (refs). However, it is easy to see that  $\dot{M}_{\rm{c,typical}}$ is more than one order of magnitude lower than the gas accretion rate of our model atmosphere $\dot{M}_{\rm{atm}}$, and lower than the core accretion rate $\dot{M}_{\rm{c,acc}}$ needed to grow the core and the atmosphere at the same time within the disk life time. As such, the formation of a giant planet by growing the core first, then letting the atmosphere cool is faster than growing the core and the atmosphere at the same time at a steady planetesimal accretion rate.

Next, we are interested in whether hydrodynamic gas accumulation due to planetesimal accretion can already commence before the atmosphere becomes unstable due to Kelvin-Helmholtz contraction, as our regime is no longer the relevant one under such conditions. A core that forms on the same timescale as our model atmosphere accretes planetesimals at a rate given by equation (\ref{eq:Mcdot}). This accretion rate is dependent on the core mass, which is steadily increasing. We therefore compare the critical core mass due to planetesimal accretion at this rate $M_{\rm{crit,acc}}$ to a core mass assumed fixed $M_{\rm{c,fixed}}$. If $M_{\rm{crit,acc}}<M{\rm_{c, fixed}}$, then the atmosphere has already initiated unstable gas accretion by the time Kelvin-Helmholtz contraction starts dominating.

%Specifically, we compare the critical core mass due to planetesimal accretion at the rate

%we compare the critical core mass $M_{\rm{c,acc}}$ due to planetesimal accretion, for an accretion rate that satisfies $L_{\rm{acc}}<L_{\rm{KH}}$ to the actual core mass $M_{\rm{c}}$ assumed fixed in our model. If $M_{\rm{c,acc}}<M{\rm{c}}$, then the atmosphere has already initiated unstable gas accretion by the time Kelvin-Helmholtz contraction starts dominating.

In order to estimate the critical core mass due to planetesimal accretion $M_{\rm{crit,acc}}$, we use the results of \citet{rafikov06} for low luminosity atmospheres forming in the outer disk ($>2-5$ AU), consistent with our region of interest. By relating his expression for the critical core mass to a given core mass dependent planetesimal accretion rate $\dot{M}(M_{\rm{c}})$, we find the following expression for the critical core mass when accretion luminosity dominates the evolution of the atmosphere:

\begin{equation}
\label{eq:critraf}
M_{\rm{crit, acc}} \sim \Big[\frac{\dot{M}(M_{\rm{c}})}{64 \pi^2} \frac{\kappa_0}{\sigma G^3} \frac{1}{R_c M_c^{1/3}} \Big(\frac{k}{\mu}\Big)^4\Big]^{3/5},
\end{equation}

\noindent with all the constants as defined in previous sections. We therefore calculate $M_{\rm{crit,acc}}$ for a range of core masses. The result is displayed in Figure 11.

 \begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/critcore.pdf}
%\vspace{-0.5in}
\caption{Comparison between the critical core mass $M_{\rm{crit, acc}}$ due to planetesimal accretion and the assumed fixed core mass when gas contraction dominates, for a growth time of $\tau=1.2$ Myrs. Unstable atmosphere collapse occurs due to core accretion before Kelvin-Helmholtz contraction starts dominating for core masses larger than $\sim 40 M_{\oplus}$.}
\end{figure}

We see that the critical core mass due to planetesimal accretion is smaller than the mass of the fixed core for $M_{\rm{c}} \sim 40 M_{\oplus}$. This means that for large core masses, planetesimal accretion will lead to runaway gas accretion before gas contraction starts dominating, and hence our model is not applicable in that regime. However, we have found in section \ref{critcore} that unstable atmospheres collapse occurs within the life time of the disk for protoplanetary cores smaller than this. We therefore confirm that planetesimal accretion can be safely ignored in our regime of interest. Moreover, as displayed in Fig. 10, it is faster to form a planet by growing the core first, then accreting a massive envelope, rather than by growing the core and atmosphere in parallel. \textbf{This works for this particular choice of core, disk etc. conditions, and I am expecting it to work for most of our parameter space, but it would probably be good to explore this numerically, at least to order of magnitude, with the estimates we have.}

As a final check, we investigate whether planetesimal accretion during the gas contraction phase at the rate $\dot{M}_{\rm{KH}}$ imposed by the condition that $L_{\rm{acc}}<L_{\rm{KH}}$ can alter the core mass enough to affect the time evolution of the atmosphere. We can quantitatively estimate the increase in core mass as

\begin{equation}
\label{eq:cminc}
\Delta M_{\rm{c}} = \int_0^{t_{\rm{crit}}} \dot{M_{\rm{c}}} dt \approx \sum_i \dot{M_{\rm{c}}}_i \Delta t_i,
\end{equation}

 \noindent where $t_{\rm{crit}}$ is the time elapsed until runaway gas accretion commences, and the accretion rate $ \dot{M_{\rm{c}}}_i $ is given by

 \begin{equation}
 \label{eq:Mdotexp}
 \dot{M_{\rm{c}}}_i =\frac{L_i R_c}{G M_{\rm{c}}} 
 \end{equation}

 \noindent from equation (\ref{eq:Lacc}), with $L_i$ the luminosity of the atmosphere at time $t_i$ in our model. For $M_{\rm{c}}=10 M_{\oplus}$, we find $\Delta M_{\rm{c}} \approx 0.05 M_{\oplus}$. This mass increase is therefore negligible in comparison with the initial core mass. It follows that a significant increase in core mass that could potentially alter the time evolution of the atmosphere would occur on a longer time scale than the mass-doubling time for the unperturbed atmosphere. Therefore, the time evolution of the atmosphere is insensitive to core mass changes at a rate imposed by the assumption that $L_{\rm{acc}}<L_{\rm{KH}}$.

 \subsection{Neglected Effects}
 \label{neglected}

 In this section we summarize some of the simplifications we have made, and discuss their relative effect on our model.

 \subsubsection{Equation of State \textbf{(Paper II)}}
 \label{EOS}

 In our model we assume an ideal gas law and a polytropic equations of state (EOS), given by equations (\ref{eq:idealgas}) and (\ref{eq:polyEOS}), respectively. However, non-ideal effects such as partial dissociation and ionization have to be taken into account. One solution is using tabulated equation of state tables for hydrogen and helium mixtures. As future work, we plan to use the \citet{saumon95} EOS tables and extend them to lower pressures and temperatures as required by our disk assumptions.

 \subsubsection{Time Evolution \textbf{(Paper I)}}

 In this study we have assumed that the luminosity stays constant throughout the radiative region of the atmosphere. As a result, the time dependence can be neglected in equation (\ref{eq:structd}) and static solution can be obtained and then connected through a global cooling equations, as described in section \ref{cooling}. However, the assumption of constant luminosity breaks down for high atmosphere masses, as described in section \ref{endoftime}; in this case, a time-dependent model is required to describe the atmosphere evolution. We are currently developing a full evolutionary time-dependent atmosphere model. We do not expect qualitatively different results. We showed in section \ref{endoftime} that the quasi-static model breaks down around the time when the atmosphere mass becomes equal to the mass of the core. Nevertheless, we can see from Figures 2 and 3 that the time of slowest accretion is safely before the region in which our model breaks down, and that the time after which the atmosphere mass increases by 25 \% is only a factor of 2 lower than the mass doubling time. As a result, we are confident of our results to order of magnitude.

 \subsubsection{Inefficient Convection\textbf{(Paper I)}}

\textbf{Discussion about mixing length theory...}

\subsubsection{Opacity \textbf{(Paper II)}}

\textbf{In progress, still need to run models for smaller AU and see at what point ice grain opacity is no longer valid. Currently strange behavior at 1 AU for $\delad=2/5$, the critical core mass goes up significantly instead of having a linear trend. $\delad=2/7$ atmospheres do not show this effect. Models for the inner disk would also show / confirm that close-in giant planet formation is hard (need larger core mass).}

 \section{Conclusions}
 \label{conclusions}

 In this paper we have studied the formation of giant planet atmospheres under the assumption that Kelvin-Helmoholtz gas contraction dominates the luminosity evolution of the atmosphere over planetesimal accretion. We built quasi-static two-layer atmosphere models with an inner convective region and an outer radiative region that matches smoothly onto the protoplanetary disk. We derived a cooling model to connect series of quasti-static atmospheres, and thus obtained an evolutionary history of the envelope. We defined the time at which unstable atmosphere collapse commences as $M_{\rm atm}(t)\sim M_{\rm c}$. From this we defined as ``critical core mass'' the minimum core mass for a protoplanet to initiate runaway gas accretion during the lifetime of the protoplanetary disk. We studied this minimum mass for a variety of disk conditions, nebular gas compositions and opacities. We found that the critical core mass decreases as we move further out in the disk, and is smaller for lower disk temperatures and opacities and for higher mean molecular weight of the gas.

 We find that the critical core mass to form a giant planet within the life time of the disk is smaller than the results yielded by studies that assume that the atmosphere evolution is dominated by the luminosity due to planetesimal accretion. We have showed that the planetesimal accretion rate needed to grow the core on a typical disk time scale is larger than the expected planetesimal accretion rates at large separations. As such, it is faster to form a planet by growing the core first in a fast planetesimal accretion regime (e.g., the core forms in the inner disk, then migrates outwards), then significantly reduce planetesimal accretion and allow a massive atmosphere to accumulate.

%Our study assumes that the protoplanetary core forms first, then it r

%---------------------------------------------------------------------------
\bibliographystyle{apj}
\bibliography{refs}

\appendix
\section{Derivation of Global Energy Equation}\label{sec:virial}

We derive here the global energy equation (\ref{eq:coolingglobal}).  We follow the simpler example in \S4.3 of \citet{kippenhahn90}, adding the effects of finite core radius, surface pressure and mass accretion.  We assume that hydrostatic balance holds.  Integrating the local energy equation (\ref{eq:coolinglocal}) from core to surface gives:
\begin{subeqnarray}
L - L\co &=& \int_{M\co}^M {\p L \over \p m} dm \\
&=& \int_{M\co}^M \left(\epsilon - T {\p S \over \p t} \right)dm \\
&=& \Gamma  - \int_{M\co}^M{\p u \over \p t} dm +  \int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm\slabel{eq:DLc}\, .
\end{subeqnarray} 
with $\Gamma = \int \epsilon dm$ the integral of the direct heating rate.

In what follows, we must carefully distinguish between partial time derivatives, $\p / \p t$, (performed at fixed mass) and total time derivatives, denoted with overdots (which include the effect of mass accreted through the outer boundary). % or $d/dt$.  
For instance the surface radius, $R$, evolves as  
\begin{equation}\label{eq:Rdot}
 \dot{R} = {\p R \over \p t} + {\dot{M} \over 4 \pi R^2 \rho\surf}
\end{equation} 
where $\p R/\p t$ gives the Lagrangian contraction of surface mass elements, and $\dot{M}$ denotes mass accretion rate through the upper boundary.  The subscript $M$  denotes quantities at the upper boundary of total mass $M$ (though it is omitted from $M$ and $R$).  Similarly the volume, $V = (4 \pi/3)r^3$ and pressure at the surface evolve as
\begin{subeqnarray}\label{eq:dot}
\dot{V}_M &=&  {\p V_{\rm M} \over \p t} + {\dot{M} \over \rho_{\rm M}}  \\
 \dot{P}_M &=& {\p P_{\rm M} \over \p t} + {\p P_M \over \p m}\dot{M} \\
 &=&  {\p P_{\rm M} \over \p t} - {G M  \over 4 \pi R^4} \dot{M}\, .
\end{subeqnarray} 
For the purpose of this derivation we will hold the core mass fixed $\dot{M}\co = 0$ which further gives $\dot{P}\co = \p P\co / \p t$.

To derive the global energy equation we must move the (partial) time derivatives in \Eq{eq:DLc} outside their integrals.  The internal energy integral follows simply from  Leibniz's rule as
\begin{equation}\label{eq:udot}
\int_{M\co}^{M(t)}{\p u \over \p t} dm = \dot{E}_i  -  \dot{M}u\surf\, .
\end{equation}

To evaluate the work integral, we derive a pair of expressions for the rate of change of gravitational energy.
The time derivative of \Eq{eq:virial}  gives
\begin{eqnarray}\label{eq:EGdot}
\dot{E}_G &=& 3  \int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm -3 \int_{M\co}^M {\p P\over \p t}{dm \over \rho} \\
&& -  3{P\surf \over \rho\surf} \dot{M}+ 3 \dot{P}\surf V\surf -3 \dot{P}\co V\co  + 3  P\surf {\dot{ V}\surf} \, . \nonumber
%&&+ 4 \pi \left. \left(r^3 {\p P \over \p t}\right)\right|_{M\co}^M  + 3  P\di {\p V \over \p t}  - {G M \over R} \dot{M} \nonumber
\end{eqnarray} 
%where the volumes, $V\surf = 4 \pi R^3/3$ and $V\co = 4 \pi R\co^3/3$. 
The first integral in \Eq{eq:EGdot} is the one we want, but the next one must be eliminated.  The time derivative of \Eq{eq:Eg} (times four) gives
\begin{subeqnarray}
 4 \dot{E}_G &=&  -4 {G M \dot{M} \over R} + 4 \int_{M\co}^M {G m \over r^2}{\p r \over \p t} dm\\ 
&=&   -4 {G M \dot{M} \over R} + 4 \pi \int_{M\co}^M r^3{\p \over \p m}{\p P \over \p t} dm \slabel{eq:4EGb} \\
&=&  -4 {G M \dot{M} \over R} -3  \int_{M\co}^M {\p P\over \p t}{dm \over \rho} \slabel{eq:4EGc} \\
&&+ 3 V\surf {\p P\surf \over \p t} -3 V\co {\p P\co \over \p t} \nonumber 
\end{subeqnarray} 
where \Eqs{eq:4EGb}{eq:4EGc} use hydrostatic balance  and integration by parts.

%To eliminate the time derivates of pressure, we take the time derivative of the hydrostatic balance equation for $\p^2 P / \p m\p t$ and integrate over $4\pi r^3 dm$ (as in the virial equation derivation) to get
%\begin{equation}\label{eq:dHBdt}
%3 \dot{P}\surf V\surf -3 \dot{P}\co V\co -3 \int_{M\co}^M {\p P\over \p t}{dm \over \rho}  = 4 \dot{E}_G + 4{G M \over R} \dot{M}  \, .
%\end{equation} 
%Combining \Eqs{eq:EGdot}{eq:dHBdt} gives 
%\begin{eqnarray}\label{eq:rhodot}
%\int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm  &=& - \dot{E}_G - {4 \over 3}{G M\over R} \dot{M} + {P\surf \over \rho\surf} \dot{M} -  P\surf \dot{V}\surf  \, , \nonumber \\
%&=&- \dot{E}_G - {4 \over 3}{G M\over R} \dot{M}  -  P\surf {\p V\surf \over \p t}  \, ,
%\end{eqnarray} 
%where the final step uses \Eq{eq:Rdot}.

Subtracting \Eqs{eq:udot}{eq:4EGc} and rearranging for the desired integral gives
\begin{eqnarray}\label{eq:PdVint}
\int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm  &=&  - \dot{E}_G - {G M \dot{M} \over R} - P\surf {\p \dot{V}\surf \over \p t} \,  
\end{eqnarray} 
with \Eq{eq:dot} used to combine terms.  Combining \Eqsss{eq:DLc}{eq:udot}{eq:PdVint}, we reproduce \Eq{eq:coolingglobal} with the accreted energy $e_{\rm acc} \equiv u\surf - GM/R$.

\section{Analytic Cooling Model Details}\label{sec:analytic}

\subsection{Estimate of Atmosphere Mass Outside the Bondi Radius}

Here we consider the mass exterior to the Bondi radius.  For a meaningful evaluation we only include the mass coming from the overdensity relative to the background density.  The resulting external mass for an isothermal atmosphere is
\begin{subeqnarray}
M_{\rm ext} &=& 4 \pi \int_{\RB}^{r_{\rm fit}} (\rho - \rho\di) r^2 dr \\
&=& M\co \theta\co \int_1^{n_{\rm fit}} 3 \left[ \exp \left({1 \over x} - {1 \over n_{\rm fit}}\right) - 1 \right] x^2 dx  \nonumber \\
&\equiv& M\co \theta\co I(n_{\rm fit})
\end{subeqnarray} 
where $\theta_c=R_B/R_{\rm c}$ and the dimensionless integral $I(n_{\rm fit})$ obeys the limits $I(1) = 0$ and $I \rightarrow n_{\rm fit}^2/2$ as $n_{\rm fit} \rightarrow \infty$.  Since this external mass does not converge, the choice of an outer boundary does matter in principle.  In practice, however, the assumption that   $r_{\rm fit} = R_{\rm H}$ limits $n_{\rm fit}$ to modest values
\begin{equation}
n_{\rm fit} = {R_{\rm H} \over \RB} \approx 1.3 {\aun{10}^{4/7} \over \mcn{10}^{2/3}}{F_T \over  m_\ast^{1/3}}.
\end{equation} 
Since for instance $I(2) = 1.1$, these modest $n_{\rm fit}$ values will only produce a small external mass.

\subsection{The Opacity Effect}
A  lower opacity  could lower the core mass.  Reducing the opacity by a factor of one hundred cuts the core mass by more than a factor of 10, specifically to 9 $M_\oplus$ for the parameters in \Eq{eq:Mcrit}.  The reduction is not  as strong as the nominal scaling would imply, $0.01^{3/5} \approx 0.06$, because $\xi$ increases.

Even with significantly lower opacities, radiative diffusion remains a good approximation at the radiative-convective boundary.  We estimate the optical depth as (for $\beta = 2$):
\begin{equation}
\tau\cb \sim {\kappa P\cb \over g} \sim 7 \times 10^4 {F_T^4 F_\kappa \over \left(\mc \over 10 \right) \left(\au \over 10\right)^{12 \over 7}} 
\end{equation} 
where $P\cb \sim P_M$ for a self-gravitating atmosphere and $g \sim G M\co/R_B^2$, with both approximation good to within the the order unity factor $\xi$.  Clearly $\tau\cb \gg 1$ even for $F_\kappa \lesssim 0.01$ out to very wide separations.

A hotter disk would increase core masses.  Instead of our passive disk model, adopting the standard Hayashi temperature profile would increase core masses by $\sim 50\%$.  A hotter accretion phase would further increase core masses, but such phases are presumably short lived.

\subsection{Surface Terms}
We now check the relevance of the neglected surface terms in \Eq{eq:coolingglobal}.  We already showed that accretion energy can be exactly eliminated by choosing an outer boundary near the Bondi radius.   We now show that accretion energy is also a small correction at the RCB.   A rough comparison, (ignoring terms of order $\xi$) of  accretion luminosity vs.\ $\dot{E}$ gives
\begin{equation}
{G M \dot{M} \over R \dot{E}} = {G M  \over R}{dM \over dE} \sim {G M\co^2 \over R_B E}{P\cb \over P_M} \sim \sqrt{R\co \over R_B} \ll 1,
\end{equation} 
where we assume $P\cb \sim P_M$  for a massive atmosphere.  Accretion energy at the protoplanetary surface is thus very weak for marginally self-gravitating atmospheres, and even weaker for lower mass atmospheres.  A similar scaling analysis shows that the work term, $P\surf \p V\surf/\p t$ is similarly weak.  Nevertheless our numerical calculations include these surface terms in a more realistic and complete model of self-gravitating atmospheres.

\subsection{Hydrogen Dissociation}
The dissociation of molecular hydrogen deep in the atmospheres of accreting protoplanets plays a significant role in the energetics of core accretion.  In the high density regions $r  \ll R\cb$ of a convective atmosphere, the thermal plus gravitational energy scales as
\begin{equation}
dE = -4 \pi \nabla_{\rm ad}^{1/\nabla_{\rm ad}} \rho\cb R_B'^{1/(\gamma-1)} r^{\frac{2\gamma - 3}{\gamma - 1}} {dr \over r}
\end{equation} 
If $\gamma < 3/2$ then the main contribution to the energy is at the bottom of the atmosphere, i.e.\ the core.  This is the case for a diatomic ideal gas ($\gamma = 1.4$) or a solar mixture of hydrogen and helium ($\gamma \approx 1.43$).  However a monatomic gas has $\gamma = 5/3 > 3/2$.  In this case, the atmosphere's energy is no longer concentrated near the bottom, but will be concentrated near the top of the convective zone.

A likely structure is an atmosphere that is dissociated near the base, but becomes molecular near the top of the convective zone.  In this case the atmosphere's energy budget would be concentrated near the atomic to molecular transition.

The energy required to dissociate a hydrogen molecule, $I = 4.467$ eV can be significant to the overall energy budget.
Since
\begin{equation}
{I \over \nabla_{\rm ad} G M\co (2 m_{\rm H})/r} \approx 3 \left(M\co \over 10 M_\oplus \right)^{-2/3} {r \over R_{\rm c}}
\end{equation} 
we see that this energy is always relevant.

We can use the Saha equation to determine where dissociation is significant,
\begin{equation}
{n_{\rm H}^2 \over n_{\rm H_2}} = \left(\pi m_{\rm H} k T \over h \right)^{3/2} e^{-I/(kT)}
\end{equation} 
We introduce a reaction coordinate $\delta$ so that $n_{\rm H} = 2 \delta n_o$ and $n_{\rm H_2} = (1-\delta) n_o$ with $n_o = \rho X/(2 m_{\rm H})$ the number density when all hydrogen is molecular.  We express equilibrium as
\begin{equation}
{\delta^2 \over 1-\delta} = f_\mu {P_{\rm diss}(T) \over P}
\end{equation} 
with the characteristic pressure below which dissociation occurs is
\begin{equation}
P_{\rm diss} = {\left(kT\right)^{5/2} \over 4} \left( \pi m_{\rm H} \over h^2\right)^{3/2}  e^{-I/(kT)}
\end{equation} 
and the order unity factor
\begin{equation}
f_\mu = 2\delta + (1-\delta) + Y/2 + Z/\mu_Z
\end{equation} 
accounts for variations in the mean molecular weight with dissociation.  (Take $\mu_Z = 31/2$, but not too significant.)

Thus dissociation occurs where $P \lesssim P_{\rm diss}(T)$.  At disk temperatures (say 150 K) the dissociation pressure is negligibly small ($\sim 10^{-141}$ dyne cm$^{-2}$) and no dissociation occurs.  However at core temperatures the dissociation pressure is quite large especially for massive cores.  Dissociation is guaranteed.

\begin{deluxetable}{cccccc}  % <--- column justification (center/left/right)
\gdef \numcols {6}
\tablecolumns{\numcols}
\tablecaption{Parameters Describing Structure of Radiative Zone.}
\tablehead{   \multicolumn{\numcols}{c} {$\gamma = 7/5$ ($\delad = 2/7$), $\alpha = 0$} }  
\startdata
 $\beta$   		 &1/2  	& 3/4 &1   		& 3/2  		& 2   \\
 $\nabla_\infty$ & 2/7 \tablenotemark{a}  	&  4/13	& 1/3 	& 2/5 	 	& 1/2 \\
 $\chi$ 		 & \nodata &  2.25245 &1.91293 	& 1.65054 	& 1.52753 \\
 $\theta $  		 &\nodata   & 0.145032	&0.285824   &0.456333   & 0.556069   \\
\enddata
\tablenotetext{a}{Since $\delad = \nabla_\infty$ there is no convective transition at depth for this case.}
\end{deluxetable}

\end{document}

% We then apply the model to one particular super Earth, GJ 1214 b, and generalize it for all super Earths. We summarize our results and findings in the next paragraphs.

%First, we modeled the planetary magnetic field as a dynamo (section \ref{dynamo}). We then presented a model of the magnetic interaction between a super Earth and its host star (section \ref{reconnection}). Based on existing models, we estimated the radio flux emitted by a super Earth and the frequency at which this emission is maximum as a function of the strength of the magnetic field at the surface of the planet (sections \ref{radioemission} and \ref{peakfrequency}). %Since this magnetic field depends on the internal composition of the planet (more precisely, on the radius, density and electrical conductivity of the planetary core), we were able to find a direct relation between the radio flux, peak frequency and some of the internal physical parameters of the planet. 
%Further on, we presented three different models for the internal structure of super Earths (section \ref{composition}):

### 2013-02-19 python/RadSGPoly/plot_helper.py
modified figures for the paper
- bigger labels (i.e. smaller figsizes)
- fig. 3 now shows M / Mdot on the y-axis
- fig. 6 has core masses at endpoints
- fig. 7 only shows mu effect + comparison over the same range of masses

@@ -6,7 +6,7 @@ scripts & ipython notebooks in python/Atmscrips.
 
 import numpy
 from utils.constants import Me, Re, gammafn, Rfn, Cvfn, Pdisk, Tdisk, kdust, \
-     params
+     params, kdust10, kdust100
 from utils.parameters import FSigma, FT, mstar
 import matplotlib.pyplot as plt
 
@@ -15,7 +15,7 @@ from RadSGPoly.cooling_poly import cooling_global as cg, cooling_local as cl,\
      critical
 from RadSGPoly.atmseries_poly import mcrit_disktime as mcd
 
-def t_vs_Mc_fixed_a(delad, Y, a, returnt = 0):
+def t_vs_Mc_fixed_a(delad, Y, a, returnt = 0, opacity = kdust):
 
     """
     Uses already generated atmosphere profiles to calculate the atmosphere
@@ -55,62 +55,71 @@ def t_vs_Mc_fixed_a(delad, Y, a, returnt = 0):
                   Cv = Cvfn(Y, delad), Pd = Pdisk(a, mstar, FSigma, FT), \
                   Td = Tdisk(a, FT), kappa = kdust)
 
+    #if prms.kappa == kdust:
+
     if a == 1.0 or a == 5.0 or a  == 3.0 or a == 2.0 or a == 1.5 or a == 2.5 or a == 1.75:
-    	t = 0 * numpy.ndarray(shape = (5), dtype = float)
-    	dtarr = 0 * numpy.ndarray(shape = (5, 299), dtype = float)
-    	tarr = 0 * numpy.ndarray(shape = (5, 299), dtype = float)
-    	Mtotcrit = 0 * numpy.ndarray(shape = (299), dtype = float)
+        t = 0 * numpy.ndarray(shape = (5), dtype = float)
+        dtarr = 0 * numpy.ndarray(shape = (5, 299), dtype = float)
+        tarr = 0 * numpy.ndarray(shape = (5, 299), dtype = float)
+        Mtotcrit = 0 * numpy.ndarray(shape = (299), dtype = float)
     
     else:
-	t = 0 * numpy.ndarray(shape = (10), dtype = float)
-    	dtarr = 0 * numpy.ndarray(shape = (10, 299), dtype = float)
-    	tarr = 0 * numpy.ndarray(shape = (10, 299), dtype = float)
-    	Mtotcrit = 0 * numpy.ndarray(shape = (299), dtype = float)
[…]

### 2013-02-19 python/Atmscripts/cooling_fixed_disklife.py
updated Mc vs. a plot + opacity effects

@@ -4,7 +4,7 @@ os.chdir('/home/apiso/repos/core_accretion/python')
 
 import numpy
 from utils.constants import Me, Re, gammafn, Rfn, Cvfn, Pdisk, Tdisk, kdust, \
-     params, mufn
+     params, mufn, kdust10, kdust100
 from utils.userpath import userpath
 from utils.parameters import FSigma, FT, mstar
 import matplotlib.pyplot as plt
@@ -12,7 +12,7 @@ from RadSGPoly import plot_helper as ph
 from RadSGPoly.profiles_poly import atmload
 import matplotlib.cm as cm
 
-a = 5.0
+a = 10.0
 delad27 = 2./7
 delad25 = 2./5
 Y00 = 0.0
@@ -36,6 +36,7 @@ prms03m = params(Me, Re, a, delad25, Y03, gamma = gammafn(delad25), R = Rfn(Y03)
 Mct_a10_00 = ph.t_vs_Mc_fixed_a(2. / 7, 0.0, a, returnt = 1)
 Mct_a10_03 = ph.t_vs_Mc_fixed_a(2. / 7, 0.3, a, returnt = 1)
 Mct_a10_03m = ph.t_vs_Mc_fixed_a(2. / 5, 0.3, a, returnt = 1)
+#Mct_a10_03_k10 = ph.t_vs_Mc_fixed_a(2. / 5, 0.3, a, returnt = 1, opacity = kdust10)
 
 Mc00 = Mct_a10_00[0]
 ttot00 = Mct_a10_00[1]
@@ -75,39 +76,43 @@ plt.show()
 au = [1.0, 5.0, 10.0, 20.0, 30.0, 40.0, 50.0, 60.0, 70.0, 80.0, 90.0, 100.0]
 #au = numpy.linspace(10, 100, 10)
 
-crit00 = ph.Mc_vs_a_fixed_t(2. / 7, 0.0, 3 * 10**6)
-crit03 = ph.Mc_vs_a_fixed_t(2. / 7, 0.3, 3 * 10**6)
-crit03m = ph.Mc_vs_a_fixed_t(2. / 5, 0.3, 3 * 10**6)
+#crit00 = ph.Mc_vs_a_fixed_t(2. / 7, 0.0, 3 * 10**6)
+#crit03 = ph.Mc_vs_a_fixed_t(2. / 7, 0.3, 3 * 10**6)
[…]

### 2013-02-19 python/Atmscripts/cooling_realgas_sg.py

@@ -18,7 +18,7 @@ out = 'rcb'
 
 if load != 0:
 
-    atm = atmload('Mc5.0.npz', prms = prms)
+    atm = atmload('Mc6.0.npz', prms = prms)
     model = atm[0]
     param = atm[1]
     prof = atm[2]

### 2013-02-19 python/RadSGPoly/plot_helper.py

@@ -107,7 +107,7 @@ def t_vs_Mc_fixed_a(delad, Y, a, returnt = 0, opacity = kdust):
             elif delad == 2./5 and Y == 0.3:
                 Mc = numpy.linspace(6, 10, 5)
 
-        elif a == 10.0 and delad == 2./7 and Y == 0.3:
+        elif a == 10.0 and delad == 2./7 and Y == 0.3 and opacity == kdust:
                 Mc = numpy.linspace(5, 14, 10)
 
         else:

### 2013-02-19 python/RadSGRealGas/atmseries.py

@@ -9,13 +9,13 @@ from scipy.interpolate import interp1d
 from utils.constants import G, kb, mp, Rb, Me, Re, Msun, RH, RHe, sigma, \
      cmperau, RHill, gammafn, mufn, Rfn, Cvfn, kdust, Tdisk, Pdisk
 from utils.parameters import mstar, FSigma, FT
-from utils.constants import params
+from utils.constants import paramsEOS
 from RadSGRealGas.profiles import profiles_write
 from RadSGRealGas.cooling import critical
 
 def atmseries(a, rhoc, Y, Mcomin, Mcomax, nMco, n = 500, \
-              nMpoints = 300, L1 = 10**22, L2 = 10**28, minMfrac = 1.05, \
-              maxMfrac = 2.1):
+              nMpoints = 200, L1 = 10**21, L2 = 10**28, minMfrac = 1.05, \
+              maxMfrac = 2.25):
     
     """
 
@@ -69,7 +69,7 @@ def atmseries(a, rhoc, Y, Mcomin, Mcomax, nMco, n = 500, \
 
     for i in range(nMco):
         
-        prms = params(Mcore[i], rcore[i], a, Y, Pd, Td, kappa = kdust)
+        prms = paramsEOS(Mcore[i], rcore[i], Y, a, Pd, Td, kappa = kdust)
         Mmin = minMfrac * Mcore[i]
         Mmax = maxMfrac * Mcore[i]
 

### 2013-02-19 python/RadSGRealGas/cooling.py

@@ -188,17 +188,17 @@ def cooling_local(param, prof, prms = prms, out = 'rcb', onlyrad = 0):
 
             success = 0
         
-            #while success != 1:
-            #    try:
-            P1int = fP1(mmid)
-            P2int = fP2(mmid)
-            T1int = fT1(mmid)
-            T2int = fT2(mmid)
+            while success != 1:
+                try:
+            	    P1int = fP1(mmid)
+            	    P2int = fP2(mmid)
+            	    T1int = fT1(mmid)
+            	    T2int = fT2(mmid)
 
-                    #success = 1
+                    success = 1
 
-                #except ValueError:
-                #    mmid = mmid[1:]
+                except ValueError:
+                    mmid = mmid[1:]
 
             Tav = (T1int + T2int) / 2
 

### 2013-02-19 python/RadSGRealGas/profiles.py

@@ -126,10 +126,14 @@ def profiles_write(n, nMpoints, L1, L2, Mmin, Mmax, filename, prms = prms, \
         mass = numpy.linspace(Mmin, Mmax, nMpoints)
 
     for i in range(nMpoints):
-        try:
-            sol = shoot(mass[i], L1, L2, n, tol, prms)
-        except ValueError:
-            pass
+	 
+        #try:
+        sol = shoot(mass[i], L1, L2, n, tol, prms)
+
+        #except ValueError:
+	#    pass 
+            	#L1 = 0.99 * L1
+	    	#L2 = 1.01 * L2
         
         param.Mtot[i], param.Mcb[i], param.MB[i], param.rcb[i], param.RB[i], \
                        param.RHill[i], param.Pc[i], param.Pcb[i], param.PB[i],\
@@ -147,8 +151,10 @@ def profiles_write(n, nMpoints, L1, L2, Mmin, Mmax, filename, prms = prms, \
 
             #step = (Mmax - Mmin) / nMpoints / param.Mtot[i]
             
-            L1 = param.L[i] * 0.5
-            L2 = param.L[i] * 1.5
+        L1 = param.L[i] * 0.5
+        L2 = param.L[i] * 1.5
+
+	print i
 
 
     if savefile == 1:

### 2013-02-19 python/RadSGRealGas/shooting.py

@@ -128,7 +128,11 @@ def Ltop(Mi, L1, L2, n, tol, prms = prms):
         #deltaL = 4 / numpy.pi * numpy.arctan(Mc1 / prms.Mco) - 1
         #relative error; use of the arctan ensures deltaL stays between -1 and 1
         if math.isnan(deltaL): #used to get rid of possible divergences
-                deltaL = 1. *(-1)
+	    #try:
+	    if lum < 10**25:
+            	deltaL = 1. *(-1)
+	    else:
+		deltaL = 1.
         return deltaL
 
     #Lmatch = newton(delta, L1, tol = tol, maxiter = 200)

### 2013-02-19 tex/draft/draft2.tex
draft updated with new figs

@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cumul_coolingtime_vs_Matm_10au_mu235.pdf}
+\includegraphics[width=0.52\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cumul_coolingtime_vs_Matm_10au_mu235.pdf}
 %\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/coolingtime_vs_Mc_10au_new.pdf}
+\includegraphics[width=0.48\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/coolingtime_vs_Mc_10au.pdf}
 %\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_new2.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_new.pdf}
 %\vspace{-0.5in}
@@
 \caption{Comparison between standard dust opacity and opacity reduced by factor of 10 and 100, respectively. Lower opacity accelerates atmosphere growth.}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_opacity.pdf}
+%\vspace{-0.5in}
+\caption{Opacity effect on critical core mass.}
 \end{figure}

### 2013-02-20 python/Analytical/analytic_comp.py
analytic model and plots added
for comparison to  temp and pressure effect and Mcrit dependence including kappa and mu.  As python class.

@@ -3,11 +3,17 @@ abalytic model does not take into account self-gravity in either the convective
 or radiative region.
 Useful for comparison with the full numerical model."""
 
+import sys
 import numpy as np
 from utils.constants import G, kb, mp, Rb, Me, Re, Msun, RH, RHe, sigma, \
-     cmperau, RHill, gammafn, mufn, Rfn, Cvfn, kdust, Tdisk, Pdisk, params
-from utils.parameters import mstar, FSigma, FT
+     cmperau, RHill, gammafn, mufn, Rfn, Cvfn, kdust, Tdisk, Pdisk, params, \
+     Myr
+from utils.parameters import mstar, FSigma, FT, rhoc
+from utils.mynumsci import zbrac
 from scipy.integrate import odeint
+from scipy.optimize import brentq
+
+p4 = 4 * np.pi
 
 def analytic_sols(Pcb, delad, Y, rhoc, a, Mc, beta = 2, theta = 0.556):
 
@@ -49,7 +55,6 @@ def analytic_sols(Pcb, delad, Y, rhoc, a, Mc, beta = 2, theta = 0.556):
     
     """
 
-    p4 = 4 * np.pi
     delinf = 1. / (4 - beta)
 
     rc = (3 * Mc / (4 * np.pi * rhoc))**(1./3)
@@ -128,7 +133,104 @@ def analytic_sols(Pcb, delad, Y, rhoc, a, Mc, beta = 2, theta = 0.556):
     #if desired could switch Mcb to total Matm above.
 
 
+class AnSols:
+    p4 = 4 * np.pi
+    beta = 2
+    delad = 2./7
 
+    delinf = 1. / (4 - beta)
+    chi = (1 - delad / delinf)**(-1. / (4 - beta))
+    theta = 0.556
+
+    #Mcb = Mc * (Pcb / PM) / (np.sqrt(np.log(Pcb / (theta * Pd))))
+
+    """While the true L at convective zone depends on Mcb, the analytic
+    theory only considers Mc to make integrals easier"""
+    #ML = Mc #+ Mcb neglected
+    #Lo = 64 * np.pi * G * ML * sigma * Td**4 /\
+    #           (3 * kdust(Td) * Pd) * delad * chi**(4 - beta)
+
+    #Lcb = Lo * Pd / Pcb
+
+    def __init__(self, mu=2.35, a=10, FT=FT, **kwargs):
+        #disk
[…]

### 2013-02-20 python/RadSGRealGas/atmseries.py
more real EOS data files

@@ -90,7 +90,7 @@ def atmseries(a, rhoc, Y, Mcomin, Mcomax, nMco, n = 500, \
                 success = 1
                 
             except ValueError:
-                Mmin = Mmin * 1.05
+                Mmin = Mmin * 1.01
 
         print "SUCCESS FOR MC = %s !" % str(Mcore[i]/Me)
 

### 2013-02-20 python/Analytical/analytic_comp.py
add fixed xi (ksi) versions of analytic theory
for plots I think we want the full version of solving for ksi.

@@ -228,9 +228,17 @@ class AnSols:
             sys.exit('mcrit bracket not found')
         return brentq(f, m0, m1, args = (tdisk, fM))
         
+    def tcool_fixksi(self, ksi=2, mc=5, fM=0.15, **kwargs):
+        """returns cooling time in Myr"""
 
+        self.setcore(mc, **kwargs)
+        PM, Pd = self.PM, self.Pd
+        #fac = fM * PM / (self.theta * Pd)
+        
+        Pcb = ksi * fM * PM
 
-    
+        tcool = p4 * Pcb**2 / Pd * self.RBprime**(7./2) / (
+            self.Lo * np.sqrt(self.rc))
 
-
+        return tcool/Myr
 

### 2013-02-21 tex/draft/draft2.tex
added new paper figs
- taller figs, removed delad info
- added analytic curves
updated plots

@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Pd_effect.pdf}
-%\vspace{-0.5in}
-\caption{Disk pressure effect on growth time. Gas accretion is faster for higher pressures.}
-\end{figure}
+%\begin{figure}[h]
+%\centering
+%\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Pd_effect.pdf}
+%%\vspace{-0.5in}
+%\caption{Disk pressure effect on growth time. Gas accretion is faster for higher pressures.}
+%\end{figure}
 
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Td_effect.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/TdPd_effect.pdf}
 %\vspace{-0.5in}
@@
 
-Moreover, the critical core mass increases for a lower mean molecular weight and a smaller adiabatic gradient.
+Moreover, the critical core mass increases for a lower mean molecular weight and a smaller adiabatic index.
 This is consistent with the results derived in section \ref{Mct}.

### 2013-02-22 tex/draft/draft2.tex
updated plots

@@
 
-\begin{figure}[h]
+\begin{figure*}[tb]
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/TdPd_effect.pdf}
+\includegraphics[width=1.\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/TdPd_effect.pdf}
 %\vspace{-0.5in}
 \caption{Disk temperature effect on growth time. Gas accretion is faster for lower temperatures.}
-\end{figure}
+\end{figure*}
 
@@
 
+\begin{figure}[tb]
+\centering
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cooling_a10_Mc5_rcb.pdf}
+%\vspace{-0.5in}
+\caption{}
+\end{figure}
+
 \section{Analytic Cooling Model Details}\label{sec:analytic}

### 2013-05-06 tex/draft/draft2.tex
actual commit, hopefully

@@
 \begin{eqnarray} 
+\label{eq:Matman}
 M_{\rm atm} &=& 4 \pi \int_{R\co}^{R\cb} \rho r^2 dr \\
@@
 \begin{eqnarray} 
+\label{eq:tcoolan}
 t_{\rm cool} &\approx & 2 \times 10^8 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2  \over \left(\mc \over 10 \right)^{5/3} \left(a\over 10\right)^{15 \over 14}} \yr
@@
 \caption{Luminosity and time evolution with mass. The analytic result is plotted for comparison.}
+\label{fig:Ltplot}
 \end{figure}
@@
 
-Standard static calculations of atmosphere accretion that consider planetesimal accretion as the main source of luminosity (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{rafikov06}) find that an atmosphere can be stable only if the atmosphere mass is lower than the core mass.
+%Standard static calculations of atmosphere accretion that consider planetesimal accretion as the main source of luminosity (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{rafikov06}) find that an atmosphere can be stable only if the atmosphere mass is lower than the core mass.
 As the mass of the envelope becomes comparable to the solid core mass, the pressure gradient can no longer balance the self-gravity of the atmosphere, and hydrostatic equilibrium breaks down.
@@
 Typically, this instability settles in when $M_{\rm{atm}} \sim M_{\mathrm{c}}$.
-Static calculations do not take into account the time evolution of the atmosphere explicitly, but rather assume a steady state at which the luminosity due to planetesimal accretion is fully radiated away by the atmosphere.
-Therefore, for a set of disk and gas conditions, a single value for the critical core mass $M_{\rm{crit}}$ is found.
+
+Standard static calculations of giant planet formation that consider planetesimal accretion as the main source of luminosity (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{rafikov06}) assume that the atmosphere is at all times in a steady state at which the luminosity due to planetesimal accretion is fully radiated away by the atmosphere.
+The core and the atmosphere grow simultaneously, and the mass of the envelope is a function of the core mass.
+When the atmosphere mass becomes comparable to the mass of the solid core, hydrostatic balance no longer holds and runaway gas accretion commences.
+Therefore, for a set of disk and gas conditions, there is a maximum mass the core can have before the onset of unstable accretion: this is defined as the critical core mass.
+%a single value for the critical core mass $M_{\rm{crit}}$ is found, when the mass of the gaseous envelope is equal to the mass of the core.
 This is the maximum mass the core can have before unstable gas accretion commences.
 
-In our evolutionary model the core mass is assumed to be fixed, so the above definition for the critical core mass is not applicable.
+In our model the atmosphere is no longer in steady state, because energy balance requires gas accretion, and the atmosphere contracts on a Kelvin-Helmholtz time scale.
+Figure \ref{fig:growthtime} shows the evolution of the atmosphere growth time with mass at several locations in our fiducial disk.
+In the early stages when $M_{\rm{atm}}$ is small, the envelope has a thin radiative zone with a high luminosity (see Figure \ref{fig:Ltplot}).
+As a result, the envelope initially grows quickly, i.e.
+on a short time scale, as shown in Figure \ref{fig:growthtime}.
+As the radiative zone becomes thicker, the luminosity decreases and growth is slowed down as a result.
+For our choice of parameters, we see that accretion is slowest for an atmosphere mass of $1-2 M_{\oplus}$.
+This is when the self-gravity of the atmosphere starts to matter; the envelope becomes massive enough for its energy to be large compared to the internal energy of the nebular gas.
+As a result, growth is continuously accelerated, and eventually runaway gas accretion settles in when the mass of the atmosphere becomes comparable to the core mass.
+Since growth is slowest long before $M_{\rm{atm}} \sim M_{\rm c}$, the estimate of the time to initiate runaway gas accretion is insensitive to the exact choice of crossover atmosphere mass to order of magnitude.
+We therefore choose the effective crossover mass, past which hydrostatic balance no longer holds, to be the minimum between $M_{\rm{c}}$ and the $M_{\rm{atm}}<M_{\rm c}$ at which our model becomes unphysical according to equation (\ref{eq:dti}), noting that both of these occur after the slow accretion phase for all our parameter choices.
+We define the \textit{crossover time} $t_{\rm{co}}$ as the time elapsed until runaway accretion commences according to the criteria above.
+
+In this section we explore the dependence of the crossover time on core, disk and gas parameters, and determine the minimum core mass needed to initiate runaway gas accretion within the lifetime of the protoplanetary disk.
+In section \ref{Mct} we show the dependence of the crossover time on the core mass at a fixed distance in the disk, justifying the need for a ``minimum'' core mass.
+We further explore how the crossover time is affected by the position in the disk in section \ref{TPeffects} and by the mean molecular weight and opacity of the nebular gas in section \ref{muopacity}.
+Finally, we determine the minimum core mass necessary for a giant planet to form before the nebular gas dissipates; we define this minimum mass as the \textit{critical core mass} for runaway accretion to commence.
+
+%In our evolutionary model the core mass is assumed to be fixed, so the above definition for the critical core mass is not applicable.
 Nevertheless, we adopt a similar approach in defining the atmosphere mass past which hydrostatic balance no longer holds.
@@
 We therefore define the evolutionary time (or growth time, or cooling time) of the atmosphere as the time elapsed until runaway accretion commences according to the criteria above.
-The evolution of atmosphere mass with time is shown in Figure 3 for $a=10$ AU and $M_{\mathrm{c}}=5 M_{\oplus}$.
-Here, $\Delta t$ is the instantaneous time step between subsequent static atmosphere models.
+The evolution of atmosphere mass with time is shown in Figure \ref{fig:growthtime} at several locations in the disk for a core mass $M_{\mathrm{c}}=5 M_{\oplus}$.
+The atmosphere instantaneous growth time, defined as $\tau=M_{\rm{atm}}/\dot{M}$, is plotted as a function of the envelope mass.
 We see that growth is slowest around $M_{\rm{tot}} \sim 6-7 M_{\oplus}$, after which it is continuously accelerated.
-The estimate of the atmosphere characteristic growth time is therefore insensitive to the exact choice of atmosphere mass to initiate runaway gas accretion, to order of magnitude, as long as it is past the time at which accretion is slowest.
+The estimate of the time to initiate runaway gas accretion is therefore insensitive to the exact choice of atmosphere mass, to order of magnitude, as long as it is past the time at which accretion is slowest.
 %\textbf{Needs a bit of a cleanup as well.}
@@
 %\vspace{-0.5in}
-\caption{Instantaneous cooling time evolution with total mass. Accretion is slowest at $M_{\mathrm{tot}} \sim 6-7 M_{\oplus}$.}
+\caption{Instantaneous cooling time evolution with total mass at 10, 30 and 60 AU. Accretion is slowest at $M_{\mathrm{atm}} \sim 1-2 M_{\oplus}$.}
+\label{fig:growthtime}
+\end{figure}
+
+\subsection{Influence of Core Mass on Crossover Time}
+\label{Mct}
+
+In this section we explore the effect of core mass on the atmosphere evolution time, for a fixed semi-major axis.
+We choose $a=10$ AU.
+Figure \ref{fig:tvsM} shows the crossover time behavior for a range of cores with masses between 1 and 10 $M_{\oplus}$.
+The envelope grows faster if the core is more massive, as it can bind a gaseous atmosphere quicker due to larger gravity.
+As such, for a fixed crossover time, e.g.
+the life timescale of the protoplanetary disk, there is a minimum core mass needed to form a planet within this time frame, which we defined above as the critical core mass.
+We defer the numerical estimates of the critical core mass to section \ref{critcore}.
+
+We also find that the atmosphere growth time is very sensitive to the core mass: for the choice of parameters in Figure \ref{fig:tvsM}, the mass increase needed to reduce the crossover time by a factor of two is only half an Earth mass.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.52\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cumul_coolingtime_vs_Matm_10au_mu235.pdf}
+%\vspace{-0.5in}
+\caption{Time evolution as a function of atmosphere mass for a range of core masses. The circles mark the crossover time. A larger core mass results in a shorter crossover time.}
+\label{fig:tvsM}
 \end{figure}
@@
 
-In what follows we study the separate effect of disk temperature and pressure on the atmosphere growth time.
-At a fixed distance in the disk $a=10$ AU and for a core mass $M_{\mathrm{c}}=5 M_{\oplus}$, we fix the disk temperature $T_{\rm{d}}$ to the value given by equation (\ref{eq:diskb}) and vary the disk pressure $P_{\rm{d}}$.
-The resulting time evolution is plotted in Figure 4.
-We find that the atmosphere evolution time decreases as the disk pressure increases.
-This effect is likely due to the fact that a higher disk pressure, and hence density, leads to the atmosphere becoming self-gravitating and therefore initiating runaway accretion on a faster timescale.
+In what follows we explore how the crossover time is affected by the planet's location in the disk.
+The location only matters because it dictates the disk temperature and pressure (see equations (\ref{eq:diskb}) and (\ref{eq:Pd})).
+We therefore study the separate effects of disk temperature and pressure on the time evolution of the atmosphere.
+At a fixed distance in the disk $a=10$ AU and for a core mass $M_{\mathrm{c}}=5 M_{\oplus}$, we fix the disk temperature $T_{\rm{d}}$ at the value given by equation (\ref{eq:diskb}) and vary the disk pressure $P_{\rm{d}}$.
+We perform a similar analysis on the effect of disk temperature $T_{\rm{d}}$ on the crossover time by fixing the disk pressure at the value given by equation (\ref{eq:Pd}) and varying $T_{\rm{d}}$.
+The resulting time evolution is plotted in Figure \ref{fig:TPeffects}.
+For comparison, we also plot the results predicted by the analytic, non self-gravitating model described in section \ref{analytic}.
+However, we saw in equation (\ref{eq:tcoolan}) that the crossover time as predicted by the analytic results is too long; since we are investigating whether the analytic model predicts the correct scalings as opposed to the actual absolute results, when compared to the numerical model, we estimate the analytic crossover time as the time needed for the atmosphere mass to become a small fraction of the core mass, rather than equal to it: $M_{\rm{atm}}=f M_{\rm c}$.
+We use $f = 0.13$.
+
+\begin{figure*}[tb]
+\centering
+\includegraphics[width=1.\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/TdPd_effect.pdf}
+%\vspace{-0.5in}
+\caption{Disk temperature and pressure effect on the crossover time $t_{\rm{co}}$.
+Gas accretion is faster for lower temperatures and higher pressures.
+The analytic scalings are plotted for comparison, for an effective crossover mass $M_{\rm{co}}$=f M_{\rm c} with $f=0.13$.}
+\label{fig:TPeffects}
+\end{figure*}
+
+We find that the crossover time is shorter for lower disk temperatures and higher pressures. We discuss the two effects separately.
+
+An increase in disk temperature results in a longer crossover time, as a higher temperature causes an increase in opacity and hence a lower luminosity.
+The analytic model predicts $t_{\rm{co}} \propto T^{\beta+ 1/2}$ (see equation (\ref{eq:tcoolan})), where $\beta$ is the constant in the opacity law \ref{eq:opacitylaw}.
+For our choice of $\beta=2$, the crossover time is therefore strongly dependent on the opacity.
+Nevertheless, different opacity assumptions will result in a different scaling with temperature of $t_{\rm{co}}$.
+We notice a generally good agreement between our numerical results and the scaled analytic approximation in the left panel of Figure \ref{fig:TPeffects}.
+A factor of $\sim 3$ reduction in the disk temperature results in an decrease of the growth time by a factor of $\sim 30$.
+The atmosphere evolution is therefore strongly dependent on the temperature in the protoplanetary disk.
+
+In contrast with the temperature effect, the crossover time decreases with disk pressure.
+This effect can also be explained by the analytic model.
+From equations (\ref{eq:tcoolan}) and (\ref{eq:xi}), the crossover time is inversely proportional to the disk pressure.
+A higher disk pressure will therefore result in a shorter crossover time.
+Increasing the disk pressure by a factor of $\sim 100$ only reduces the crossover time by a factor of $\sim 2$.
+As such, the crossover time is only weakly dependent on the nebular pressure.
+
+%From equation (\ref{eq:PcbRcb}), a higher disk pressure will yield the same pressure at the radiative-convective boundary $P_{\cb}$ at a larger radius of the convective zone $R_{\cb}$, and thus at a larger atmosphere mass from equation (\ref{eq:Matman}).
+A higher mass implies the atmosphere will become critical on a shorter timescale.
+
+ %This effect is likely due to the fact that a higher disk pressure, and hence density, leads to the atmosphere becoming self-gravitating and therefore initiating runaway accretion on a faster timescale.
 However, increasing the disk pressure by a factor of $\sim 100$ only reduces the atmosphere growth time by a factor of $\sim 2$.
@@
 
-We perform a similar analysis on the effect of disk temperature $T_{\rm{d}}$ on the evolutionary time.
-For $a=10$ AU and $M_{\rm{c}}=5 M_{\oplus}$, we fix the disk pressure at the value given by equation (\ref{eq:Pd}) and vary $T_{\rm{d}}$.
-The resulting dependence is shown in Figure 5.
-In contrast with the pressure effect, an increase in disk temperature causes the growth time to increase as well.
+%In contrast with the pressure effect, an increase in disk temperature causes the growth time to increase as well.
 This is due to opacity effects: a higher disk temperature increases the opacity in the radiative zone of the atmosphere, causing a decrease in luminosity and therefore a longer Kelvin-Helmholtz accretion time.
@@
 The atmosphere evolution is therefore strongly dependent on the temperature in the protoplanetary disk.
-
-\begin{figure*}[tb]
-\centering
-\includegraphics[width=1.\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/TdPd_effect.pdf}
-%\vspace{-0.5in}
-\caption{Disk temperature effect on growth time. Gas accretion is faster for lower temperatures.}
-\end{figure*}
-
-\subsection{Core Mass versus Evolution Time at Fixed Distance}
-\label{Mct}
-
-In this section we explore how the core mass affects the cooling time of the atmosphere for a fixed semi-major axis.
-For consistency, we use the same distance as in the previous subsections $a=10$ AU.
-Figure 6 shows the growth time behavior for a range of cores with masses between 1 and 10 $M_{\oplus}$.
-As expected, growth happens on a faster timescale if the core is more massive, as it can bind a gaseous atmosphere quicker due to its higher gravitational attraction.
-We find that the atmosphere growth time is very sensitive to the core mass: the core mass increase needed to reduce the cooling time by an order of magnitude is only half an Earth mass.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.52\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cumul_coolingtime_vs_Matm_10au_mu235.pdf}
-%\vspace{-0.5in}
-\caption{Time evolution as a function of atmosphere mass for a range of core masses. The circles mark the time when runaway accretion is initiated.}
-\end{figure}
-
-We further analyze how the thermodynamic properties of the gas affect the atmosphere growth time.
-Specifically, we compare our standard model given by $\delad=2/7$ and $\mu=2.35$ with models with a different adiabatic index and mean molecular weight.
-This is shown in Figure 7.
-We discuss the effects of adiabatic index and molecular weight separately.
+%We also notice that the analytic scaling is
+
+\subsection{Role of Mean Molecular Weight and Opacity}
+\label{muopacity}
+%
+%In this section we explore the effect of the core mass on the atmosphere evolution time, for a fixed semi-major axis.
+We explore again the case of $a=10$ AU.
+Figure \ref{fig:tvsM} shows the growth time behavior for a range of cores with masses between 1 and 10 $M_{\oplus}$.
+As expected, growth happens on a faster timescale if the core is more massive, as it can bind a gaseous atmosphere quicker due to larger gravity.
+We find that the atmosphere growth time is very sensitive to the core mass: the mass increase needed to reduce the cooling time by an order of magnitude is only half an Earth mass.
+%
+%\begin{figure}[h]
+%\centering
+%\includegraphics[width=0.52\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cumul_coolingtime_vs_Matm_10au_mu235.pdf}
+%%\vspace{-0.5in}
+%\caption{Time evolution as a function of atmosphere mass for a range of core masses.
+The circles mark the time when runaway accretion is initiated.
+A larger core mass results in a lower cooling time.}
+%\label{fig:tvsM}
+%\end{figure}
+
+In this work we build evolutionary atmosphere profiles assuming an ideal gas equation of state with constant mean molecular weight and adiabatic index.
+A realistic gas mixture, however, has variable adiabatic gradient and molecular weight, due to non-ideal effects and processes like dissociation or ionization.
+Future work will include real gas equation of state tables, and we expect quantitatively different results.
+In this study we analyze only the effect of the mean molecular weight of the nebular gas on the crossover time.
+Specifically, we compare our standard model given by $\delad=2/7$ and $\mu=2.35$ with a model with the same adiabatic index but a different molecular mass, $\mu=2$ (i.e., purely molecular hydrogen).
+The result is shown in Figure \ref{fig:tvsMcomp}.
+We also plot the analytic results for comparison, using the same fractional mass scaling as in subsection \ref{TPeffects}, $f=0.13$, and notice that they are in good agreement with the numerical models.
+A lower mean molecular weight translates into a longer crossover time for the same core mass, since an atmosphere composed of heavier molecules is easier to bind.
 
@@
 %\vspace{-0.5in}
-\caption{Time evolution for atmospheres with different adiabatic indices and mean molecular weight. A typical protoplanetary disk life time is plotted for comparison.}
+\caption{The crossover time $t_{\rm{co}}$ for atmospheres with different mean molecular weight.
+A typical protoplanetary disk life time is plotted for comparison.
+The crossover time is larger for a lower $\mu$.}
+\label{fig:tvsMcomp}
 \end{figure}
 
-A decrease in the mean molecular weight of the gas results in a larger cooling time for the same core mass, since an atmosphere composed of heavier molecules is easier to bind.
-Moreover, a gas with a higher adiabatic gradient $\delad$ is more dense, and therefore more likely to be bound by the gravity of the planet.
-
-\subsection{Critical Core Mass}
-\label{critcore}
-
-In the previous subsections we explored and discussed the effect of various parameters on the growth time of an atmosphere accreting around a protoplanetary core of fixed mass.
-The main question we address right now is under what conditions can such an atmosphere grow before the gas in the protoplanetary disk is dissipated.
-Specifically, we determine the minimum core mass that is necessary for a protoplanet to initiate runaway gas accretion within the disk lifetime.
-We define this mass as the \textit{critical core mass} for runaway growth.
-
-Observations of protoplanetary disks show that they dissipate over a time scale of a few million years (e.g., \citealt{lagrange00}, \citealt{haisch01}, \citealt{goldreich04}).
-We adopt a disk life timescale of $\tau=3 $ Myrs and determine the critical core mass as a function of gas and disk conditions.
-The results are displayed in Figure 8.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_new.pdf}
-%\vspace{-0.5in}
-\caption{The minimum core mass for an atmosphere to become ÔcriticalÕ within the lifetime of the disk as a function of semi-major axis, for different adiabatic indices and molecular weight of the nebular gas.}
-\end{figure}
-
-The critical core mass decreases as we move further away in the disk. This is due to several reasons:
-
-\begin{itemize}
-\item The disk temperature $T_d$ is lower in the outer regions of the disk.
-As we showed in section \ref{TPeffects}, the atmosphere grows faster for lower temperatures, which offsets the need for a larger core mass.
-\item The gravitational attraction of the central star is  weaker at larger distances,  and hence the gravity of the protoplanet can act on a wider range.
-%resulting in a larger effective radius for the planet (i.e.m the )As a result, the Hill radius of the planet (see equation \ref{eq:RH}) is larger, and hence the gravity of the planet can
-\end{itemize}
-
-Moreover, the critical core mass increases for a lower mean molecular weight and a smaller adiabatic index.
-This is consistent with the results derived in section \ref{Mct}.
-The critical core mass is strongly dependent on the mean molecular weight $\mu$, but only weakly depended on the adiabatic index $\gamma$ (or $\delad$).
-
-The comparison between our results and standard values for the critical core mass due to planetesimal accretion is deferred to section \ref{acc}.
-% \textbf{Need to add qualitative explanations for different slopes and curve intersection}
-
-\subsection{Opacity Dependence}
-\label{opacity}
-
- In this section we investigate the effect of opacity reduction on the atmosphere evolution, since a lower opacity in the outer atmosphere regions accelerates gas accretion, resulting in a lower critical core mass.
-We explore opacity reductions by factors of 10 and 100, while keeping the power law (\ref{eq:opacitylaw}) coefficient constant $\beta=2$.
-The result is depicted in Figure 9.
-
- We find that the atmosphere luminosity and growth time are increased and decreased, respectively, by factors of the same order as the opacity reductions.
+We further investigate the effect of opacity reduction on the atmosphere evolution.
+We explore opacity laws in which the coefficient $\kappa_0$ from equation (\ref{eq:opacitylaw})  is reduced by factors of 10 and 100.
+The result is depicted in Figure \ref{fig:LtvsMopacity}.
+All things equal, a lower opacity increases he luminosity and hence speeds atmosphere growth by the same factor as the opacity reduction (see equations (\ref{eq:structd}) and (\ref{eq:delrad})).
 
@@
 \caption{Comparison between standard dust opacity and opacity reduced by factor of 10 and 100, respectively. Lower opacity accelerates atmosphere growth.}
+\label{fig:LtvsMopacity}
 \end{figure}
+
+\subsection{Critical Core Mass}
+\label{critcore}
+
+In the previous subsections we explored the effect of disk and gas properties on the growth time of an atmosphere accreting around a protoplanetary core of fixed mass.
+The main question we address here is under what conditions the atmosphere can grow before the gas in the protoplanetary disk is dissipated.
+Specifically, we determine the minimum core mass that is necessary for a protoplanet to initiate runaway gas accretion within the disk lifetime, i.e.
+to have a crossover time equal to the disk life timescale.
+We  denoted this mass the \textit{critical core mass} in section \ref{Mct}.
+
+Observations of protoplanetary disks show that they dissipate over a time scale of a few million years (e.g., \citealt{lagrange00}, \citealt{haisch01}, \citealt{goldreich04}).
+We adopt a disk life timescale of $\tau=3 $ Myrs and determine the critical core mass as a function of gas and disk conditions.
+The results are displayed in Figure \ref{fig:Mcvsa}.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_new.pdf}
+%\vspace{-0.5in}
+\caption{The minimum core mass for an atmosphere to become critical within the lifetime of the disk $t \sim 3 Myrs$ as a function of semi-major axis, for different mean molecular weight of the nebular gas.
+The analytic results are plotted for comparison.
+The critical core mass is lower further out in the disk.}
+\label{fig:Mcvsa}
+\end{figure}
+
+The critical core mass decreases as we move further away in the disk, primarily due to the fact the disk temperature is lower in the outer regions of the disk.
+As we showed in section \ref{TPeffects}, the atmosphere grows faster for lower temperatures, which offsets the need for a larger core mass.
+The critical core mass is larger for a lower mean molecular weight, consistent with the results derived in section \ref{Mct}.
+\textbf{Add explanation for analytic curve behavior --- need to chat about this with Andrew}.
+We also explore the opacity effect on the critical core mass.
+The results are shown in Figure \ref{fig:Mcritopacity} For an opacity, and therefore crossover time, reduced by an order of magnitude, the critical core mass is reduced by roughly a factor of 2.
+
+%The comparison between our results and standard values for the critical core mass due to planetesimal accretion is deferred to section \ref{acc}.
+% \textbf{Need to add qualitative explanations for different slopes and curve intersection}
+%
+%\subsection{Opacity Dependence}
+%\label{opacity}
+% 
+% In this section we investigate the effect of opacity reduction on the atmosphere evolution.
+We explore opacity laws in which in which the constant $\kappa_0$ from equation (\ref{eq:opacitylaw})  is reduced by factors of 10 and 100.
+The result is depicted in Figure \ref{fig:LtvsMopacity}.
+All things equal, a lower opacity increases (decreases) the luminosity (cooling time) by the same factor as the opacity reduction (see equations (\ref{eq:structd}) and (\ref{eq:delrad})).
+% 
+%
+% \begin{figure}[h]
+%\centering
+%\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/opacity_effect.pdf}
+%%\vspace{-0.5in}
+%\caption{Comparison between standard dust opacity and opacity reduced by factor of 10 and 100, respectively. Lower opacity accelerates atmosphere growth.}
+%\label{fig:LtvsMopacity}
+%\end{figure}
 
@@
 %\vspace{-0.5in}
-\caption{Opacity effect on critical core mass.}
+\caption{Opacity effect on critical core mass. A factor of ten opacity reduction results in a critical core mass twice as small.}
+\label{fig:Mcritopacity}
 \end{figure}

### 2013-05-09 tex/draft/draft2.tex
slightly updated section 5

@@
 The core and the atmosphere grow simultaneously, and the mass of the envelope is a function of the core mass.
-When the atmosphere mass becomes comparable to the mass of the solid core, hydrostatic balance no longer holds and runaway gas accretion commences.
+As the atmosphere mass becomes comparable to the mass of the solid core, hydrostatic balance no longer holds and runaway gas accretion commences.
 Therefore, for a set of disk and gas conditions, there is a maximum mass the core can have before the onset of unstable accretion: this is defined as the critical core mass.
@@
 In our model the atmosphere is no longer in steady state, because energy balance requires gas accretion, and the atmosphere contracts on a Kelvin-Helmholtz time scale.
-Figure \ref{fig:growthtime} shows the evolution of the atmosphere growth time with mass at several locations in our fiducial disk.
-In the early stages when $M_{\rm{atm}}$ is small, the envelope has a thin radiative zone with a high luminosity (see Figure \ref{fig:Ltplot}).
-As a result, the envelope initially grows quickly, i.e.
-on a short time scale, as shown in Figure \ref{fig:growthtime}.
-As the radiative zone becomes thicker, the luminosity decreases and growth is slowed down as a result.
+Figure \ref{fig:growthtime} shows the evolution of the atmosphere growth time with mass at several locations in our fiducial disk for a fixed core mass.
+In the early stages when $M_{\rm{atm}}$ is small, the envelope has a thin radiative zone and hence a high luminosity (see Figure \ref{fig:Ltplot}).
+As a result, the envelope initially grows quickly, as shown in Figure \ref{fig:growthtime}.
+As the radiative zone becomes thicker, the luminosity decreases and growth slows down.
 For our choice of parameters, we see that accretion is slowest for an atmosphere mass of $1-2 M_{\oplus}$.
@@
 As a result, growth is continuously accelerated, and eventually runaway gas accretion settles in when the mass of the atmosphere becomes comparable to the core mass.
-Since growth is slowest long before $M_{\rm{atm}} \sim M_{\rm c}$, the estimate of the time to initiate runaway gas accretion is insensitive to the exact choice of crossover atmosphere mass to order of magnitude.
+Since growth is slowest before $M_{\rm{atm}} \sim M_{\rm c}$ for the cases that we tried, the estimate of the time to initiate runaway gas accretion is insensitive to the exact choice of crossover atmosphere mass to order of magnitude.
 We therefore choose the effective crossover mass, past which hydrostatic balance no longer holds, to be the minimum between $M_{\rm{c}}$ and the $M_{\rm{atm}}<M_{\rm c}$ at which our model becomes unphysical according to equation (\ref{eq:dti}), noting that both of these occur after the slow accretion phase for all our parameter choices.
-We define the \textit{crossover time} $t_{\rm{co}}$ as the time elapsed until runaway accretion commences according to the criteria above.
+We define the \textit{crossover time} $t_{\rm{co}}$ as the time elapsed until the atmosphere mass becomes equal to the effective crossover mass as defined above, and runaway accretion commences.
 
@@
 %\vspace{-0.5in}
-\caption{Instantaneous cooling time evolution with total mass at 10, 30 and 60 AU. Accretion is slowest at $M_{\mathrm{atm}} \sim 1-2 M_{\oplus}$.}
+\caption{Instantaneous growth time for the atmosphere of a planet with fixed core mass $M_{\rm c}=5 M_{\oplus}$ as a function of atmospheric mass, for a planet located at 10, 30 and 60 AU.
+Accretion is slowest before the mass of the atmosphere becomes comparable to the mass of the core.
+In this example, this happens at $M_{\mathrm{atm}} \sim 1-2 M_{\oplus}$.}
 \label{fig:growthtime}
@@
 
-In this section we explore the effect of core mass on the atmosphere evolution time, for a fixed semi-major axis.
-We choose $a=10$ AU.
-Figure \ref{fig:tvsM} shows the crossover time behavior for a range of cores with masses between 1 and 10 $M_{\oplus}$.
-The envelope grows faster if the core is more massive, as it can bind a gaseous atmosphere quicker due to larger gravity.
+We begin by fixing the disk properties and generating atmosphere profiles for a set of fixed core masses.
+Figure \ref{fig:tvsM} shows the crossover time behavior for a range of cores with masses between 1 and 10 $M_{\oplus}$, at $a=10$ AU in our fiducial disk.
+The envelope grows faster if the core is more massive, as it can bind a gaseous atmosphere more quickly due to larger gravity.
+This can also be seen from the analytical results in equation (\ref{eq:tcoolan}).
 As such, for a fixed crossover time, e.g.
-the life timescale of the protoplanetary disk, there is a minimum core mass needed to form a planet within this time frame, which we defined above as the critical core mass.
-We defer the numerical estimates of the critical core mass to section \ref{critcore}.
+the typical lifetime of the protoplanetary disk, there is a minimum core mass needed to form a planet within this time frame, which we defined above as the critical core mass.
+We defer numerical estimates of the critical core mass to section \ref{critcore}.
 
@@
 %\vspace{-0.5in}
-\caption{Time evolution as a function of atmosphere mass for a range of core masses. The circles mark the crossover time. A larger core mass results in a shorter crossover time.}
+\caption{We generate atmosphere profiles at $a=10$ AU for cores with fixed masses between $5 M_{\oplus}$ and $14 M_{\oplus}$.
+We plot the cumulative cooling time as a function of atmosphere mass.
+The circles mark the crossover time.
+A larger core mass results in a shorter crossover time.}
 \label{fig:tvsM}
@@
 In what follows we explore how the crossover time is affected by the planet's location in the disk.
-The location only matters because it dictates the disk temperature and pressure (see equations (\ref{eq:diskb}) and (\ref{eq:Pd})).
+The location only matters because it dictates the disk temperature (equation (\ref{eq:diskb})) and pressure (equation (\ref{eq:Pd})).
 We therefore study the separate effects of disk temperature and pressure on the time evolution of the atmosphere.
@@
 For comparison, we also plot the results predicted by the analytic, non self-gravitating model described in section \ref{analytic}.
-However, we saw in equation (\ref{eq:tcoolan}) that the crossover time as predicted by the analytic results is too long; since we are investigating whether the analytic model predicts the correct scalings as opposed to the actual absolute results, when compared to the numerical model, we estimate the analytic crossover time as the time needed for the atmosphere mass to become a small fraction of the core mass, rather than equal to it: $M_{\rm{atm}}=f M_{\rm c}$.
-We use $f = 0.13$.
+However, we saw in equation (\ref{eq:tcoolan}) that the crossover time as predicted by the analytic results is too long.
+Here we are investigating whether the analytic model predicts the correct scalings as opposed to the actual absolute results, when compared to the numerical model.
+We therefore estimate the analytic crossover time as the time needed for the atmosphere to reach an effective crossover mass $M_{\rm{atm}}=f M_{\rm c}$, where $f = 0.13$.
 
@@
 %\vspace{-0.5in}
-\caption{Disk temperature and pressure effect on the crossover time $t_{\rm{co}}$.
-Gas accretion is faster for lower temperatures and higher pressures.
-The analytic scalings are plotted for comparison, for an effective crossover mass $M_{\rm{co}}$=f M_{\rm c} with $f=0.13$.}
+\caption{Left panel: crossover time as a function of disk temperature for fixed disk pressure and a core of fixed mass $M_{\rm c}=5 M_{\oplus}$.
+Right panel: crossover time as a function of disk pressure for fixed disk temperature and a core of fixed mass  $M_{\rm c}=5 M_{\oplus}$.
+The analytic scalings are plotted for comparison, for an effective crossover mass $M_{\rm{co}}=f M_{\rm c}$ with $f=0.13$.
+Gas accretion is faster for lower temperatures and higher pressures.}
 \label{fig:TPeffects}
@@
 An increase in disk temperature results in a longer crossover time, as a higher temperature causes an increase in opacity and hence a lower luminosity.
-The analytic model predicts $t_{\rm{co}} \propto T^{\beta+ 1/2}$ (see equation (\ref{eq:tcoolan})), where $\beta$ is the constant in the opacity law \ref{eq:opacitylaw}.
+The analytic model predicts $t_{\rm{co}} \propto T^{\beta+ 1/2}$ (see equation (\ref{eq:tcoolan})), where $\beta$ is the constant in the opacity law (\ref{eq:opacitylaw}).
 For our choice of $\beta=2$, the crossover time is therefore strongly dependent on the opacity.
-Nevertheless, different opacity assumptions will result in a different scaling with temperature of $t_{\rm{co}}$.
+Different opacity assumptions will result in a different scaling with temperature of $t_{\rm{co}}$.
 We notice a generally good agreement between our numerical results and the scaled analytic approximation in the left panel of Figure \ref{fig:TPeffects}.
@@
 
-In contrast with the temperature effect, the crossover time decreases with disk pressure.
+An increase in disk pressure will decrease the crossover time.
 This effect can also be explained by the analytic model.
@@
 In this study we analyze only the effect of the mean molecular weight of the nebular gas on the crossover time.
-Specifically, we compare our standard model given by $\delad=2/7$ and $\mu=2.35$ with a model with the same adiabatic index but a different molecular mass, $\mu=2$ (i.e., purely molecular hydrogen).
-The result is shown in Figure \ref{fig:tvsMcomp}.
+In Figure \ref{fig:tvsMcomp}, we compare our standard model given by $\delad=2/7$ and $\mu=2.35$ with a model with the same adiabatic index but a different molecular mass, $\mu=2$ (i.e., purely molecular hydrogen).
 We also plot the analytic results for comparison, using the same fractional mass scaling as in subsection \ref{TPeffects}, $f=0.13$, and notice that they are in good agreement with the numerical models.
@@
 %\vspace{-0.5in}
-\caption{The crossover time $t_{\rm{co}}$ for atmospheres with different mean molecular weight.
-A typical protoplanetary disk life time is plotted for comparison.
-The crossover time is larger for a lower $\mu$.}
+\caption{The crossover time $t_{\rm{co}}$ as a function of core mass at $a=10$ AU, for atmospheres with different mean molecular weight.
+A typical protoplanetary disk life time $t=3$ Myrs is plotted for comparison.
+The crossover time is larger for a lower mean molecular weight.}
 \label{fig:tvsMcomp}
@@
 
-We further investigate the effect of opacity reduction on the atmosphere evolution.
-We explore opacity laws in which the coefficient $\kappa_0$ from equation (\ref{eq:opacitylaw})  is reduced by factors of 10 and 100.
-The result is depicted in Figure \ref{fig:LtvsMopacity}.
-All things equal, a lower opacity increases he luminosity and hence speeds atmosphere growth by the same factor as the opacity reduction (see equations (\ref{eq:structd}) and (\ref{eq:delrad})).
+A decrease in opacity reduces the number of steps that photons need to diffuse (since it increases their mean free path), resulting in a higher luminosity.
+In Figure \ref{fig:LtvsMopacity}, we explore opacity laws in which the coefficient $\kappa_0$ from equation (\ref{eq:opacitylaw}) is reduced by factors of 10 and 100.
+All things equal, a lower opacity increases the luminosity and hence speeds atmosphere growth by the same factor as the opacity reduction (see equations (\ref{eq:structd}) and (\ref{eq:delrad})).
 
@@
 %\vspace{-0.5in}
-\caption{Comparison between standard dust opacity and opacity reduced by factor of 10 and 100, respectively. Lower opacity accelerates atmosphere growth.}
+\caption{Luminosity and cooling time evolution with total atmosphere mass for planets of fixed core mass $M_{\rm c}=5 M_{\oplus}$ and located at $a=10$ AU, for different opacities.
+The standard dust opacity is reduced by factors of 10 and 100, respectively.
+A lower opacity accelerates atmosphere growth.}
 \label{fig:LtvsMopacity}
@@
 
-In the previous subsections we explored the effect of disk and gas properties on the growth time of an atmosphere accreting around a protoplanetary core of fixed mass.
+We can now combine the results obtained in sections \ref{Mct}, \ref{TPeffects} and \ref{muopacity} to determine the minimum core mass necessary for a protoplanet to initiate runaways gas accretion within the disk lifetime, i.e.
+to have a crossover time equal to the disk life timescale.
+We  denote this mass the \textit{critical core mass}.
+
+%In the previous subsections we explored the effect of disk and gas properties on the growth time of an atmosphere accreting around a protoplanetary core of fixed mass.
 The main question we address here is under what conditions the atmosphere can grow before the gas in the protoplanetary disk is dissipated.
@@
 We adopt a disk life timescale of $\tau=3 $ Myrs and determine the critical core mass as a function of gas and disk conditions.
-The results are displayed in Figure \ref{fig:Mcvsa}.
 
@@
 %\vspace{-0.5in}
-\caption{The minimum core mass for an atmosphere to become critical within the lifetime of the disk $t \sim 3 Myrs$ as a function of semi-major axis, for different mean molecular weight of the nebular gas.
+\caption{The minimum core mass for an atmosphere to initiate runaway gas accretion within the lifetime of the disk $t \sim 3 Myrs$ as a function of semi-major axis, for different mean molecular weight of the nebular gas.
 The analytic results are plotted for comparison.
@@
 
-The critical core mass decreases as we move further away in the disk, primarily due to the fact the disk temperature is lower in the outer regions of the disk.
+ In Figure \ref{fig:Mcvsa}, we see that the critical core mass decreases as we move further away in the disk, primarily due to the fact the disk temperature is lower in the outer regions of the disk.
 As we showed in section \ref{TPeffects}, the atmosphere grows faster for lower temperatures, which offsets the need for a larger core mass.
@@
 \textbf{Add explanation for analytic curve behavior --- need to chat about this with Andrew}.
-We also explore the opacity effect on the critical core mass.
-The results are shown in Figure \ref{fig:Mcritopacity} For an opacity, and therefore crossover time, reduced by an order of magnitude, the critical core mass is reduced by roughly a factor of 2.
+We also explore the opacity effect on the critical core mass in Figure \ref{fig:Mcritopacity} For an opacity, and therefore crossover time, reduced by an order of magnitude, the critical core mass is reduced by roughly a factor of 2.
 
@@
 %\vspace{-0.5in}
-\caption{Opacity effect on critical core mass. A factor of ten opacity reduction results in a critical core mass twice as small.}
+\caption{Critical core mass as a function of semi-major axis for different opacities. A factor of ten opacity reduction results in a critical core mass twice as small.}
 \label{fig:Mcritopacity}

### 2013-05-27 tex/draft/draft2.tex
updated draft

@@
 
-\section{Introduction \textbf{(Paper II)}}
+\section{Introduction}
 \label{intro}
@@
 
-\section{Atmosphere Models \textbf{(Paper I)}}
+\section{Atmosphere Models}
 \label{sec2}
@@
 In section \ref{model} we describe the assumptions of our model.
-In section \ref{scales} we describe the properties of our assumed protoplanetary disk and review relevant length scales.
-Section \ref{struct} presents the equations describing the structure of a static planetary atmosphere and the boundary conditions that we use.
+Section \ref{struct} presents the equations describing the structure of a static planetary atmosphere.
+In section \ref{BCs} we describe discuss our boundary conditions and the properties of our assumed protoplanetary disk, and review relevant length scales.
+and the boundary conditions that we use.
 In section \ref{cooling} we develop a model to estimate the rate at which the atmosphere cools.
@@
 
-\subsection{Disk Model and Length Scales}
-\label{scales}
-
-As a disk model, we use the minimum mass, passively irradiated model of  \citet{chiang10}. In this model, the surface density and mid-plane temperature are given by
-
-\begin{subeqnarray}
-\label{eq:diskparam}
-\Sigma_d&=&2200 F_{\Sigma} a^{-3/2}\,\, \text{g cm}^{-2} \slabel{eq:diska}\\
-T_d &=& 120 F_T a^{-3/7} \,\,K, \slabel{eq:diskb}
-\end{subeqnarray}
-
-\noindent with $a$ the semi-major axis in AU, and $F_{\Sigma}$ and $F_T$ normalization factors that adjust the disk mass and temperature relative to the minimum mass solar nebula (MMSN).
-In this study, we assume $F_{\Sigma}=F_T=1$.
-The resulting mid-plane pressure is given by
-
-\begin{equation}
-\label{eq:Pd}
-P_d=1.1 \times 10^{-4} F_{\Sigma} \sqrt{F_T m_*} a^{-45/14} \,\, \text{dyne cm}^{-2}
-\end{equation}
-
-\noindent for a molecular weight $\mu=2.35$. Here $m_* \equiv M_*/M_{\odot}$, where $M_*$ is the mass of the central star and $M_{\odot}$ is the mass of the Sun. We take $m_*=1$.
-
-We further define several characteristic length scales that are important for this problem. \\
-
-\textbf{The Core Radius} represents the physical radius of the protoplanetary core:
-
-\begin{equation}
-\label{eq:rc}
-R_c \equiv \Big(\frac{3 M_c}{4 \pi \rho_c}\Big)^{1/3},
-\end{equation}
-
-\noindent with $M_c$ and $\rho_c$ the core mass and density, respectively. Since we assume that the core mass does not change with time, the core radius is also fixed for a given model. \\
-
-\textbf{The Bondi Radius} represents the distance from the planet at which the thermal energy of the nebular gas is of the order of the gravitational energy binding the gas to the planet.
-It is defined as
-
-\begin{equation}
-\label{eq:RB}
-R_B \equiv \frac{G M_p}{c_s^2}=\frac{G M_p}{\mathcal{R} T_d},
-\end{equation}
-
-\noindent where $G$ is the gravitational constant, $M_p$ is the total mass of the planet, $c_s$ is the sound speed, and $\mathcal{R}$ is the reduced gas constant: $\mathcal{R}=k_b/(\mu m_p)$, with $k_b$ the Boltzmann constant and $m_p$ the proton mass.
-\\
-
-\textbf{The Hill Radius} represents the distance at which the gravitational attraction of the planet and of the host star are comparable. It is given by
-
-\begin{equation}
-\label{eq:RHill}
-R_H \equiv a \Big(\frac{M_p}{M_{\odot}}\Big)^{1/3}
-\end{equation}
-
-\noindent Outside the Hill radius, the gravity of the core is too weak to affect the nebular gas. Consequently, the planet will be able to accrete only gas that lies within its Hill sphere. \\
-
-\textbf{The Disk Scale Height} is defined by
-
-\begin{equation}
-\label{eq:H}
-H \equiv \frac{c_s^2}{\Omega},
-\end{equation}
-
-\noindent where $\Omega$ is the Keplerian angular velocity.
-The disk scale height is a measure of the thickness of the protoplanetary disk.
-Using equation (\ref{eq:diska}), we find that the scale height for our assumed disk model can be expressed as
-
-\begin{equation}
-\label{eq:H2}
-H=0.022 \sqrt{F_T} a^{9/7} \,\, \mathrm{AU}
-\end{equation}
-
-For small planet masses, $R_B<R_H<H$.
-As the atmosphere becomes more massive, tidal truncation becomes more important than thermal constraints, and $R_H<R_B<H$.
-Finally, for high mass atmospheres the disk scale height becomes smaller than the Hill radius of the planet, and spherical symmetry breaks, as the planet is now affected by the vertical profile of the disk.
+%\subsection{Length Scales}
+%\label{scales}
 
@@
 
-\subsubsection{Boundary Conditions}
+\subsection{Boundary Conditions}
 \label{BCs}
+
+We first define several characteristic length scales that are important for this problem. \\
+
+\textbf{The Core Radius} represents the physical radius of the protoplanetary core:
+
+\begin{equation}
+\label{eq:rc}
+R_c \equiv \Big(\frac{3 M_c}{4 \pi \rho_c}\Big)^{1/3},
+\end{equation}
+
+\noindent with $M_c$ and $\rho_c$ the core mass and density, respectively. Since we assume that the core mass does not change with time, the core radius is also fixed for a given model. \\
+
+\textbf{The Bondi Radius} represents the distance from the planet at which the thermal energy of the nebular gas is of the order of the gravitational energy binding the gas to the planet.
+It is defined as
+
+\begin{equation}
+\label{eq:RB}
+R_B \equiv \frac{G M_p}{c_s^2}=\frac{G M_p}{\mathcal{R} T_d},
+\end{equation}
+
+\noindent where $G$ is the gravitational constant, $M_p$ is the total mass of the planet, $c_s$ is the sound speed, and $\mathcal{R}$ is the reduced gas constant: $\mathcal{R}=k_b/(\mu m_p)$, with $k_b$ the Boltzmann constant and $m_p$ the proton mass.
+\\
+
+\textbf{The Hill Radius} represents the distance at which the gravitational attraction of the planet and of the host star are comparable. It is given by
+
+\begin{equation}
+\label{eq:RHill}
+R_H \equiv a \Big(\frac{M_p}{M_{\odot}}\Big)^{1/3}
+\end{equation}
+
+\noindent Outside the Hill radius, the gravity of the core is too weak to affect the nebular gas. Consequently, the planet will be able to accrete only gas that lies within its Hill sphere. \\
+
+\textbf{The Disk Scale Height} is defined by
+
+\begin{equation}
+\label{eq:H}
+H \equiv \frac{c_s}{\Omega},
+\end{equation}
+
+\noindent where $\Omega$ is the Keplerian angular velocity.
+The disk scale height is a measure of the thickness of the protoplanetary disk.
+Using equation (\ref{eq:diska}), we find that the scale height for our assumed disk model can be expressed as
+
+\begin{equation}
+\label{eq:H2}
+H=0.022 \sqrt{F_T} a^{9/7} \,\, \mathrm{AU}
+\end{equation}
+
+For small planet masses, $R_B<R_H<H$.
+As the atmosphere becomes more massive, tidal truncation becomes more important than thermal constraints, and $R_H<R_B<H$.
+Finally, for high mass atmospheres the disk scale height becomes smaller than the Hill radius of the planet, and spherical symmetry breaks, as the planet is now affected by the vertical profile of the disk.
 
@@
 The temperature and pressure at the Hill radius have to therefore be given by the nebular temperature and pressure: $T(R_H)=T_d$ and $P(R_H)=P_d$.
+As a disk model, we use the minimum mass, passively irradiated model of  \citet{chiang10}.
+In this model, the surface density and mid-plane temperature are given by
+
+\begin{subeqnarray}
+\label{eq:diskparam}
+\Sigma_d&=&2200 F_{\Sigma} a^{-3/2}\,\, \text{g cm}^{-2} \slabel{eq:diska}\\
+T_d &=& 120 F_T a^{-3/7} \,\,K, \slabel{eq:diskb}
+\end{subeqnarray}
+
+\noindent with $a$ the semi-major axis in AU, and $F_{\Sigma}$ and $F_T$ normalization factors that adjust the disk mass and temperature relative to the minimum mass solar nebula (MMSN).
+In this study, we assume $F_{\Sigma}=F_T=1$.
+The resulting mid-plane pressure is given by
+
+\begin{equation}
+\label{eq:Pd}
+P_d=1.1 \times 10^{-4} F_{\Sigma} \sqrt{F_T m_*} a^{-45/14} \,\, \text{dyne cm}^{-2}
+\end{equation}
+
+\noindent for a molecular weight $\mu=2.35$. Here $m_* \equiv M_*/M_{\odot}$, where $M_*$ is the mass of the central star and $M_{\odot}$ is the mass of the Sun. We take $m_*=1$.
 
@@
 
-\subsection{Quasi-Static Two-Layer Model}
-\label{twolayer}
-
-%The model and equations presented in this section so far can be applied to any protoplanetary atmosphere that satisfies the assumptions listed in section \ref{model}.
-
-As a consequence of the equations derived in section \ref{cooling}, both the atmosphere structure and the gas accretion rate are uniquely determined by the current atmosphere mass.
-As this mass accretion rate is slow compared to the time it takes to relax to this solution, we can make a quasi-static model of the atmosphere growth.
-In this section we therefore describe the procedure to obtain an evolutionary series from the cooling model introduced in \ref{cooling} between consecutive two-layer static atmospheres (cf.
-\ref{model}).
-
-We use the boundary conditions and opacity laws described in section \ref{struct}.
-The numerical integration is performed through the shooting method, which solves the boundary value problem by reducing it to an initial value problem: trial values are chosen for the parameters at one of the boundaries, then the equations are integrated and the resulting values at the other boundary are com- pared to the actual boundary conditions.
-The procedure is repeated until convergence is achieved.
-We start with a total atmosphere mass $M_i$, where the index $i$ labels each evolutionary stage.
-For this mass we guess a trial luminosity $L=L_{\mathrm{guess}}$.
-The assumption of constant luminosity throughout the radiative zone sets the right hand side of equation (\ref{eq:structb}) to zero, i.e.
-the time dependence is neglected.
-We integrate the structure equations (\ref{eq:struct}) inwards from $R_{\mathrm{out}}=R_H$, with the boundary conditions $T(R_{\mathrm{out}})=T_d$ and $P(R_{\mathrm{out}})=P_d$.
-The numerical integration gives a value for the core mass implied by the trial solution $M_c=M_{c, \mathrm{guess}}$.
-We adjust $L_{\mathrm{guess}}$ until $M_{c, \mathrm{guess}}$ converges to the actual core mass $M_c$.
-
-The time evolution of the atmosphere is obtained using the global cooling model described in section \ref{cooling}.
-We neglect the luminosity due to planetesimal accretion and direct heat generation; as such, the first two terms in equation (\ref{eq:coolingglobal}) are set to zero.
-Equation (\ref{eq:coolingglobal}) is then evaluated at the RCB, since we assume that luminosity is primarily generated in the convective zone.
-In what follows, $X_{RCB}$ denotes quantity $X$ evaluated at the RCB.
-The time step $\Delta t_i$ between two consecutive models of masses $M_i$ and $M_{i+1}$ is calculated as follows:
-
-\begin{eqnarray}\nonumber
-\label{eq:dti}
-<L>_i \Delta t_i&=&-\Delta E_{RCB,i} + <e_{\mathrm{acc}, RCB}>_i \Delta M_{RCB, i} \\
-&&  - <P_{RCB}>_i \Delta V_{RCB, M, i} 
-\end{eqnarray}
-
-\noindent where we denote
-
-\begin{subeqnarray}
-\Delta X_i&=&X_{i+1}-X_i \\
-<X>_i&=&\frac{X_i+X_{i+1}}{2}
-\end{subeqnarray}
-\\
-\noindent The subscript $M$ in the volume expression above signifies evaluating the change in volume at constant mass.
-
-By connecting sets of subsequent static atmospheres through the procedure described above we therefore obtain an evolutionary atmosphere series.
-
-\section{Analytic Cooling Model \textbf{(Paper I)}}
+\section{Analytic Cooling Model}
 \label{analytic}
@@
 \label{eq:delo}
-\nabla \di \equiv \frac{3 \kappa_0 P_{\rm d}}{64 \pi G M \sigma T_{\rm d}^4}.
+\nabla \di \equiv \frac{3 \kappa_0 P_{\rm d}}{64 \pi G M \sigma T_{\rm d}^4} L.
 \end{equation}
@@
 
-\section{Quasi-Static Kelvin-Helmholtz Contraction \textbf{(Paper I)}}
+\section{Quasi-Static Kelvin-Helmholtz Contraction}
 \label{KH}
 
-In this section we describe our choices for the gas, disk and core parameters, and present some typical atmosphere numerical models.
-As mentioned in section \ref{model}, we assume that the central star is a Sun-like star with $M_*=M_{\odot}$.
-We are primarily interested in the outer regions of the protoplanetary disk, so we generate atmosphere models between 10 and 100 AU.
-As a fiducial case, we assume that the atmosphere is composed of a hydrogen-helium mixture with the helium mass fraction $Y=0.3$., resulting in a mean molecular weight $\mu=2.35$.
-We do not take into account the dependence of the adiabatic index on the helium fraction, and therefore assume the adiabatic index for a diatomic gas $\delad=2/7$.
-We further explore how the atmosphere evolution varies with adiabatic index and molecular weight.
+In this section we generate quasi-static two-layer atmosphere profiles by integrating the equations of hydrostatic balance (\ref{eq:struct}) with the boundary conditions described in section \ref{BCs}.
+We obtain an evolutionary sequence from a series of instantaneous atmosphere profiles using the cooling model outlined in section \ref{cooling}.
+We describe our numerical method in section \ref{twolayer}.
+In section \ref{profiles} we show example plots of instantaneous and evolutionary atmosphere profiles for various choices of disk and core parameters.
+We compare our results with the predictions of the analytic model described in section \ref{analytic} and we explore the effect of changes in opacity on the atmosphere evolution.
+Finally, we discuss the breakdown of the quasi-static model in section \ref{endoftime}.
+
+\subsection{Numerical Method}
+\label{twolayer}
+
+%The model and equations presented in this section so far can be applied to any protoplanetary atmosphere that satisfies the assumptions listed in section \ref{model}.
+
+As a consequence of the equations derived in section \ref{cooling}, both the atmosphere structure and the gas accretion rate are uniquely determined by the current atmosphere mass.
+As this mass accretion rate is slow compared to the time it takes to relax to this solution, we can make a quasi-static model of the atmosphere growth.
+In this section we describe the procedure to obtain an evolutionary series from the cooling model introduced in \ref{cooling} between consecutive two-layer static atmospheres (cf.
+\ref{model}).
+
+We use the boundary conditions and opacity laws described in section \ref{struct}.
+The numerical integration is performed through the shooting method, which solves the boundary value problem by reducing it to an initial value problem: trial values are chosen for the parameters at one of the boundaries, then the equations are integrated and the resulting values at the other boundary are compared to the actual boundary conditions.
+The procedure is repeated until convergence is achieved.
+We start with a total atmosphere mass $M_i$, where the index $i$ labels each evolutionary stage.
+For this mass we guess a trial luminosity $L=L_{\mathrm{guess}}$.
+The assumption of constant luminosity throughout the radiative zone sets the right hand side of equation (\ref{eq:structb}) to zero, i.e.
+the time dependence is neglected.
+We integrate the structure equations (\ref{eq:struct}) inwards from $R_{\mathrm{out}}=R_{\rm H}$, with the boundary conditions $T(R_{\mathrm{out}})=T_{\rm d}$ and $P(R_{\mathrm{out}})=P_{\rm d}$.
+The numerical integration gives a value for the core mass implied by the trial solution $M_{\rm c}=M_{\rm{c, guess}}$.
+We adjust $L_{\mathrm{guess}}$ until $M_{\rm{c,guess}}$ converges to the actual core mass $M_{\rm c}$.
+
+The time evolution of the atmosphere is obtained using the global cooling model described in section \ref{cooling}.
+We neglect the luminosity due to planetesimal accretion and direct heat generation; as such, the first two terms in equation (\ref{eq:coolingglobal}) are set to zero.
+Equation (\ref{eq:coolingglobal}) is then evaluated at the RCB, since we assume that luminosity is primarily generated in the convective zone.
+In what follows, $X_{RCB}$ denotes quantity $X$ evaluated at the RCB.
+The time step $\Delta t_i$ between two consecutive models of masses $M_i$ and $M_{i+1}$ is calculated as follows:
+
+\begin{eqnarray}\nonumber
+\label{eq:dti}
+<L>_i \Delta t_i&=&-\Delta E_{RCB,i} + <e_{\mathrm{acc}, RCB}>_i \Delta M_{RCB, i} \\
+&&  - <P_{RCB}>_i \Delta V_{RCB, M, i} 
+\end{eqnarray}
+
+\noindent where we denote
+
+\begin{subeqnarray}
+\Delta X_i&=&X_{i+1}-X_i \\
+<X>_i&=&\frac{X_i+X_{i+1}}{2}
+\end{subeqnarray}
+\\
+\noindent The subscript $M$ in the volume expression above signifies evaluating the change in volume at constant mass.
+
+By connecting sets of subsequent static atmospheres through the procedure described above we therefore obtain an evolutionary atmosphere series.
+
+\subsection{Atmosphere Profiles}
+\label{profiles}
+
+We consider planets embedded in a protoplanetary disk around a Sun-like star and with a temperature and pressure profile described by equations (\ref{eq:diskb}) and (\ref{eq:Pd}), respectively.
+We are primarily interested in the outer regions of the disk, so we generate atmosphere models at distances between 10 and 100 AU in our fiducial disk.
+We emphasize that the location in the disk only affects the temperature and pressure of the nebular gas.
+As our model applies only when planetesimal accretion occurs too slowly to substantially change the mass of the core, we consider the evolution of an atmosphere around a core of fixed mass.
+
+We assume that the atmosphere consists of an ideal gas described by a polytropic equation of state, and is composed of a hydrogen-helium mixture with the helium mass fraction $Y=0.3$, resulting in a mean molecular weight $\mu=2.35$.
+We do not take into account the dependence of the adiabatic index on the helium fraction, and therefore assume the adiabatic gradient of a diatomic gas $\delad=2/7$.
+%We further explore how the atmosphere evolution varies with adiabatic index and molecular weight.
 We generate model atmospheres for monatomic adiabatic index $\delad=2/5$ and for a purely molecular hydrogen composition ($\mu=2$).
@@
 
-Figure 1 shows instantaneous atmosphere radial profiles at $a=10$ AU accreting around a fixed core of mass $M_{\rm{c}}=5 M_{\oplus}$ for several total planet masses.
-In what follows,  if the $R_B<R_H$ hierarchy holds, we define the planet mass as the total mass enclosed within the Bondi radius of the atmosphere, and regard the gas between the Bondi and Hill radii as perturbations due to the gravitational influence of the planet.
-For illustrative purposes we mark the position of the Bondi radius and of the radiative convective boundary.
-We notice that the pressure and temperature in the convective regions of the atmospheres scale as $1/r$, which is the expected behavior of an ideal gas adiabat with a polytopic equation of state (e.g., \citealt{rafikov06}).
-In the outer radiative region of the atmosphere, the pressure has a nearly exponential profile, which is in agreement with the predictions of the analytic model (see section \ref{analytic} and \App{sec:analytic}).
-In addition, the atmosphere temperature at the radiative-convective boundary $T_{\rm{RCB}}$ only differs from the disk temperature $T_{\rm{d}}$ by an order unity factor, and so the atmosphere is nearly isothermal throughout the radiative zone.
-This is also in agreement with the analytic predictions.
+Figure \ref{fig:profiles} shows instantaneous radial profiles at different stages in the evolution of an atmosphere accreting around a core of fixed mass $M_{\rm{c}}=5 M_{\oplus}$ located at $a=60$ AU.
+We define the atmosphere mass as the mass of the gas enclosed within the Bondi radius.
+The locations of the radiative-convective boundary and of the Bondi radius are displayed.
+Pressure and temperature in the convective region of the atmosphere scale as $1/r$, which is the expected behavior of an ideal gas adiabat (e.g., \citealt{rafikov06}).
+In the outer radiative region of the atmosphere, the pressure has a nearly exponential profile, which is in agreement with the predictions of our analytic model (see section \ref{analytic} and \App{sec:analytic}).
+In addition, the temperature at the radiative-convective boundary $T_{\rm{RCB}}$ only differs from the disk temperature $T_{\rm{d}}$ by an order unity factor, hence the atmosphere is nearly isothermal throughout the radiative zone.
+This is also in agreement with our analytic predictions (see equation \ref{eq:Tcb}).
 %, and justifies
@@
 %\vspace{-0.5in}
-\caption{Example pressure, temperature and mass profiles as a function of radius.}
+\caption{Example pressure, temperature and total mass (core plus atmosphere) profiles as a function of radius at $a=60$ AU and for a fixed core mass $M_{\rm c}=5 M_{\oplus}$.
+The circles and triangles mark the locations of the Bondi radius and of the radiative-convective boundary, respectively.
+The solid, dotted and dashed lines correspond to profiles with total mass (core and atmosphere) of $5.12 M_{\oplus}$, $5.85 M_{\oplus}$ and $9.22 M_{\oplus}$, respectively.}
+\label{fig:profiles}
 \end{figure}
 
-The location of the Bondi radius moves further out as the mass of the atmosphere increases, as the Bondi radius scales with mass.
-The location of the radiative-convective boundary has a more interesting behavior.
-As stated in section \ref{intro}, the nebular gas initially settles into equilibrium around the protoplanetary core on a short timescale relative to the thermal timescale, and is isentropic throughout with an entropy equal to that of the disk.
-In the context of our model, this atmosphere represents the minimum atmosphere mass for which a hydrostatic solution exists.
-This atmosphere is purely convective.
-As gas accumulates and the atmosphere starts cooling, an outer radiative region forms.
-This layer is thin for low mass atmospheres, as can be seen in Fig.
-1: for the $5.21 M_{\oplus}$ atmosphere, the ratio between the thickness of the radiative and convective regions is low.
-An increase in atmosphere mass causes the radiative zone to expand and the convective region to contract.
+In the initial stages of its evolution, the atmosphere is purely convective and isentropic, with an entropy equal to that of the disk.
+In the context of our numerical model, this is the minimum atmosphere mass for which a hydrostatic solution exists.
+Further gas accumulation brings in energy that has to be radiated away, and an initially thin outer radiative region forms, which becomes thicker as the atmosphere becomes more massive.
+
+%This layer is thin for low mass atmospheres, An increase in atmosphere mass causes the radiative zone to expand and the convective region to contract.
 As more and more gas is accumulated, the convective zone eventually starts expanding again; however, the ratio between the radii of the radiative and convective regions continually increases with atmosphere mass, i.e.
@@
 
-We further use the global cooling model developed in section \ref{cooling} to obtain an evolutionary series from static atmosphere profiles.
-Figure 2 shows the luminosity and cooling time evolution with atmosphere mass.
-Gas accretion is initially rapid, but slows down significantly after a relatively small mass increase.
+%as can be seen in Fig. 1: for the $5.21 M_{\oplus}$ atmosphere, the ratio between the thickness of the radiative and convective regions is low.
+
+In what follows we discuss the time evolution of the envelope.
+We use the global cooling model developed in section \ref{cooling} to obtain an evolutionary history from series of static atmosphere profiles, using the numerical procedure described in section \ref{twolayer}.
+Figure \ref{fig:Ltplot} shows the luminosity evolution and the elapsed time as a function of atmosphere mass for the same choice of parameters as in Figure \ref{fig:profiles} ($M_{\rm c}=5 M_{\oplus}$).
+The radiative region of the atmosphere is initially thin, which results in a relatively small number of steps that photons need to diffuse and hence a high luminosity (see equation \ref{eq:Lcb}).
+The luminosity decreases as the atmosphere mass grows and the radiative region becomes deeper.
 For comparison, we also plot the analytic results derived in section \ref{analytic}.
-We see that there is only a very brief agreement between the analytic and full numerical results.
-At early times, the radiative zone is too shallow and so the approximation that $P_{\rm RCB} \gg P_{\rm d}$ used in section \ref{analytic} breaks down.
-At late times, the self-gravity of the atmosphere becomes important.
+We find only a brief agreement between the analytic and numerical results.
+In the early stages, the radiative zone is too shallow and the assumption that $P_{\rm RCB} \gg P_{\rm d}$ used in the analytic model breaks down.
+As the mass of the atmosphere grows, the neglect of self-gravity in the analytic model results in a lower luminosity (equation \ref{eq:Lcb}) and therefore a longer cooling time when compared to the numerical solution.
+While our analytic model does not provide an accurate quantitative description of the atmosphere evolution, we will show in section \ref{critical} that it predicts the correct scalings for the atmosphere crossover time and critical core mass.
 %We see that self-gravity starts to matter
@@
 %\vspace{-0.5in}
-\caption{Luminosity and time evolution with mass. The analytic result is plotted for comparison.}
+\caption{Luminosity and time evolution of the atmosphere of a planet with fixed core mass $M_{\rm c}=5 M_{\oplus}$ and located at $a=60$ AU.
+The analytic result is plotted for comparison.
+The luminosity is initially high, then decreases as the atmosphere grows in mass and the radiative zone becomes deeper.
+Growth is accelerated as a result.}
 \label{fig:Ltplot}
@@
 
+We explain the time evolution from the bottom panel of Figure \ref{fig:Ltplot} by showing how the atmosphere growth time evolves with the atmosphere mass for a fixed core.
+Since the luminosity is high in the early stages, the envelope initially grows quickly; this is shown in Figure \ref{fig:growthtime} for a planet forming around a core of mass $M_{\rm c}=5 M_{\oplus}$ at several locations in our fiducial disk.
+Even though the atmosphere growth time, i.e.
+$M/\dot{M}$, is small in the early evolution stages, the increase in atmosphere mass, i.e.
+$\dot{M}$ is also small.
+As a result, the envelope spends a relatively long time at a low mass, which explains the behavior in the bottom panel of Figure \ref{fig:Ltplot}.
+As the radiative zone becomes thicker, the luminosity decreases and growth slows down.
+For $M_{\rm c}=5 M_{\oplus}$, we see that accretion is slowest for an atmosphere mass of $1-2 M_{\oplus}$.
+This is when the self-gravity of the atmosphere starts to matter; the envelope becomes massive enough for its energy to be large compared to the internal energy of the nebular gas.
+As a result, growth is continuously accelerated, and eventually runaway gas accretion settles in when the mass of the atmosphere becomes comparable to the core mass.
+This explains the flat behavior of the elapsed time in the later stages of the evolution, since this is when the atmosphere growth time continuously decreases, as seen in Figure \ref{fig:growthtime}.
+We note that in all cases presented here, growth is slowest before $M_{\rm{atm}} \sim M_{\rm c}$.
+%and the estimate of the time to initiate runaway gas accretion is insensitive to the exact choice of crossover atmosphere mass to order of magnitude.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mt_profile.pdf}
+%\vspace{-0.5in}
+\caption{Instantaneous growth time for the atmosphere of a planet with fixed core mass $M_{\rm c}=5 M_{\oplus}$ as a function of atmospheric mass, for a planet located at 10, 30 and 60 AU.
+Accretion is slowest before the mass of the atmosphere becomes comparable to the mass of the core.
+In this example, this happens at $M_{\mathrm{atm}} \sim 1-2 M_{\oplus}$.}
+\label{fig:growthtime}
+\end{figure}
+
+The grain opacity is one source of uncertainty in our model, as it may be be lower or higher than the standard dust opacity.
+Dust settling or grain growth result in a lower opacity, while a high metallicity of the host star yields an enhanced opacity.
+Lower opacity is possible to be achieved in our regime, since we consider low planetesimal accretion rates and therefore an atmosphere that is mostly depleted of planetesimals.
+A decrease in opacity reduces the number of steps that photons need to diffuse (since it increases their mean free path), resulting in a higher luminosity.
+In Figure \ref{fig:LtvsMopacity}, we explore opacity laws in which the coefficient $\kappa_0$ from equation (\ref{eq:opacitylaw}) is reduced by factors of 10 and 100.
+All things equal, a lower opacity increases the luminosity and hence speeds atmosphere growth by the same factor as the opacity reduction (see equations \ref{eq:structd} and \ref{eq:delrad}).
+
+ \begin{figure}[h]
+\centering
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/opacity_effect.pdf}
+%\vspace{-0.5in}
+\caption{Luminosity and time evolution as a function of atmosphere mass for planets of fixed core mass $M_{\rm c}=5 M_{\oplus}$ and located at $a=10$ AU, for different opacities.
+The standard dust opacity is reduced by factors of 10 and 100, respectively.
+A lower opacity accelerates atmosphere growth.}
+\label{fig:LtvsMopacity}
+\end{figure}
+
+%\subsection{Growth Time \textbf{temporary title}}
+%\label{growthtime}
+
 \subsection{The End of Quasi-Static Evolution}
@@
 
-The right-hand side of equation (\ref{eq:dti}) decreases with increasing atmosphere mass, and eventually becomes negative when the mass of the envelope becomes comparable to the core mass $M_{\rm{atm}} \sim M_{\rm{c}}$.
-This implies a negative luminosity from the RCB.
-Physically, the luminosity $L$ represents the total heat flux through the surface defined by the RCB.
-The heat flux due to convective motions is always directed outwards, and hence it is positive.
-Moreover, the temperature gradient $dT/dr$ is negative, as the temperature decreases towards the outside.
-As a result (see equation (\ref{eq:structd})), the heat flux due to radiative diffusion is also positive.
+We have shown in section \ref{profiles} that growth is continuously accelerated once the self-gravity of the atmosphere becomes important.
+This is due to both the decrease in luminosity and to the fact that less and less binding energy per unit of added mass is released as the envelope grows in mass, i.e.
+$-dE/dM$ becomes small.
+The right-hand side of equation (\ref{eq:coolingglobal}) therefore decreases with increasing atmosphere mass; in our calculations, we have found that it eventually becomes negative around the time when $M_{\rm{atm}} \sim M_{\rm{c}}$.
+This translates into a negative luminosity from the radiative-convective boundary.
+Physically, the luminosity $L$ represents the total heat flux through the surface defined by the radiative-convective boundary.
+The heat flux due to convective motions is always directed outwards and hence positive.
+The temperature gradient $dT/dr$ is negative, as the temperature decreases towards the outside, and the heat flux due to radiative diffusion is also positive as a result (see equation \ref{eq:structd}).
 This implies that, in the absence of planetesimal accretion and internal heat sources, there is no physical mechanism that can justify a negative luminosity.
 The quasi-static, constant luminosity approximation breaks down at this point and a more detailed, time-dependent model is necessary.
+However, we will show in section \ref{critical} that the existence of this unphysical result does not affect our estimates of the atmosphere evolution time or critical core mass, as the unphysical regime occurs after the slowest growth stage.
+%However, we have shown in section \ref{growthtime} that in all cases presented here, growth is slowest before $M_{\rm{atm}} \sim M_{\rm c}$, and the estimate of the time to initiate runaway gas accretion is therefore insensitive to the exact choice of crossover atmosphere mass to order of magnitude.
 
@@
 
-\section{Critical Core Mass \textbf{(Paper I, except for $\delad=2/5$ case)}}
+\section{Critical Core Mass}
 \label{critical}
@@
 As the atmosphere mass becomes comparable to the mass of the solid core, hydrostatic balance no longer holds and runaway gas accretion commences.
-Therefore, for a set of disk and gas conditions, there is a maximum mass the core can have before the onset of unstable accretion: this is defined as the critical core mass.
+Therefore, for a given set of disk conditions, there is a maximum mass the core can have before the onset of unstable accretion: this is defined as the critical core mass.
 %a single value for the critical core mass $M_{\rm{crit}}$ is found, when the mass of the gaseous envelope is equal to the mass of the core.
@@
 
-In our model the atmosphere is no longer in steady state, because energy balance requires gas accretion, and the atmosphere contracts on a Kelvin-Helmholtz time scale.
-Figure \ref{fig:growthtime} shows the evolution of the atmosphere growth time with mass at several locations in our fiducial disk for a fixed core mass.
-In the early stages when $M_{\rm{atm}}$ is small, the envelope has a thin radiative zone and hence a high luminosity (see Figure \ref{fig:Ltplot}).
-As a result, the envelope initially grows quickly, as shown in Figure \ref{fig:growthtime}.
-As the radiative zone becomes thicker, the luminosity decreases and growth slows down.
-For our choice of parameters, we see that accretion is slowest for an atmosphere mass of $1-2 M_{\oplus}$.
-This is when the self-gravity of the atmosphere starts to matter; the envelope becomes massive enough for its energy to be large compared to the internal energy of the nebular gas.
-As a result, growth is continuously accelerated, and eventually runaway gas accretion settles in when the mass of the atmosphere becomes comparable to the core mass.
-Since growth is slowest before $M_{\rm{atm}} \sim M_{\rm c}$ for the cases that we tried, the estimate of the time to initiate runaway gas accretion is insensitive to the exact choice of crossover atmosphere mass to order of magnitude.
+In our model the atmosphere is no longer in steady state because energy balance requires gas accretion, and the atmosphere contracts on a Kelvin-Helmholtz time scale.
+We have shown in section \ref{profiles} that atmosphere growth is slowest before $M_{\rm{atm}} \sim M_{\rm c}$, and so the estimate of the time to initiate runaway gas accretion is insensitive to the exact choice of crossover atmosphere mass to order of magnitude.
 We therefore choose the effective crossover mass, past which hydrostatic balance no longer holds, to be the minimum between $M_{\rm{c}}$ and the $M_{\rm{atm}}<M_{\rm c}$ at which our model becomes unphysical according to equation (\ref{eq:dti}), noting that both of these occur after the slow accretion phase for all our parameter choices.
-We define the \textit{crossover time} $t_{\rm{co}}$ as the time elapsed until the atmosphere mass becomes equal to the effective crossover mass as defined above, and runaway accretion commences.
+We define the \textit{crossover time} $t_{\rm{co}}$ as the time elapsed until the atmosphere mass becomes equal to the effective crossover mass.
+At $t_{\rm{co}}$, runaway accretion commences.
 
 In this section we explore the dependence of the crossover time on core, disk and gas parameters, and determine the minimum core mass needed to initiate runaway gas accretion within the lifetime of the protoplanetary disk.
-In section \ref{Mct} we show the dependence of the crossover time on the core mass at a fixed distance in the disk, justifying the need for a ``minimum'' core mass.
-We further explore how the crossover time is affected by the position in the disk in section \ref{TPeffects} and by the mean molecular weight and opacity of the nebular gas in section \ref{muopacity}.
-Finally, we determine the minimum core mass necessary for a giant planet to form before the nebular gas dissipates; we define this minimum mass as the \textit{critical core mass} for runaway accretion to commence.
+In section \ref{Mct} we show how the crossover time varies with core mass for fixed disk parameters, demonstrating that higher mass cores achieve runaway accretion more quickly.
+We explore how the crossover time is affected by by the mean molecular weight of the nebular gas in section \ref{muopacity} and by the disk temperature and pressure in section \ref{TPeffects}.
+Finally, we determine the minimum core mass necessary for a giant planet to form before the nebular gas dissipates in section \ref{critcore}; we define this minimum mass as the \textit{critical core mass} for runaway accretion to commence.
 
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mt_profile.pdf}
-%\vspace{-0.5in}
-\caption{Instantaneous growth time for the atmosphere of a planet with fixed core mass $M_{\rm c}=5 M_{\oplus}$ as a function of atmospheric mass, for a planet located at 10, 30 and 60 AU.
-Accretion is slowest before the mass of the atmosphere becomes comparable to the mass of the core.
-In this example, this happens at $M_{\mathrm{atm}} \sim 1-2 M_{\oplus}$.}
-\label{fig:growthtime}
-\end{figure}
-
-\subsection{Influence of Core Mass on Crossover Time}
+\subsection{Crossover Time as a Function of Core Mass}
 \label{Mct}
 
-We begin by fixing the disk properties and generating atmosphere profiles for a set of fixed core masses.
-Figure \ref{fig:tvsM} shows the crossover time behavior for a range of cores with masses between 1 and 10 $M_{\oplus}$, at $a=10$ AU in our fiducial disk.
+We begin by generating atmosphere evolution profiles for cores of different masses under a fixed set of disk conditions.
+Figure \ref{fig:tvsM} displays $t_{\rm{co}}$ for $M_{\rm c}$ between 5 and 14 $M_{\oplus}$ at $a=10$ AU in our fiducial disk.
 The envelope grows faster if the core is more massive, as it can bind a gaseous atmosphere more quickly due to larger gravity.
 This can also be seen from the analytical results in equation (\ref{eq:tcoolan}).
-As such, for a fixed crossover time, e.g.
-the typical lifetime of the protoplanetary disk, there is a minimum core mass needed to form a planet within this time frame, which we defined above as the critical core mass.
+As such, there is a minimum core mass required to form a planet within a fixed crossover time, e.g.
+the typical lifetime of the protoplanetary disk, which we defined above as the critical core mass.
 We defer numerical estimates of the critical core mass to section \ref{critcore}.
@@
 %\vspace{-0.5in}
-\caption{We generate atmosphere profiles at $a=10$ AU for cores with fixed masses between $5 M_{\oplus}$ and $14 M_{\oplus}$.
-We plot the cumulative cooling time as a function of atmosphere mass.
-The circles mark the crossover time.
+\caption{Time to grow an atmosphere of mass $M_{\rm{atm}}$ for cores with fixed masses between $5 M_{\oplus}$ and $14 M_{\oplus}$ at $a=10$ AU in our fiducial disk.
+The circles mark the crossover time where $M_{\rm{atm}} \sim M_{\rm c}$.
+The numbers are labeling the core mass in Earth masses.
 A larger core mass results in a shorter crossover time.}
@@
 
-\subsection{Role of Disk Temperature and Pressure}
-\label{TPeffects}
-
-In what follows we explore how the crossover time is affected by the planet's location in the disk.
-The location only matters because it dictates the disk temperature (equation (\ref{eq:diskb})) and pressure (equation (\ref{eq:Pd})).
-We therefore study the separate effects of disk temperature and pressure on the time evolution of the atmosphere.
-At a fixed distance in the disk $a=10$ AU and for a core mass $M_{\mathrm{c}}=5 M_{\oplus}$, we fix the disk temperature $T_{\rm{d}}$ at the value given by equation (\ref{eq:diskb}) and vary the disk pressure $P_{\rm{d}}$.
-We perform a similar analysis on the effect of disk temperature $T_{\rm{d}}$ on the crossover time by fixing the disk pressure at the value given by equation (\ref{eq:Pd}) and varying $T_{\rm{d}}$.
-The resulting time evolution is plotted in Figure \ref{fig:TPeffects}.
+\subsubsection{Role of Mean Molecular Weight}
+\label{muopacity}
+
+%Figure \ref{fig:tvsMcomp} reproduces the crossover time as a function of core mass at $a=10$ AU from Figure \ref{fig:tvsM}.
+
+ In this work we build evolutionary atmosphere profiles assuming an ideal gas equation of state with constant mean molecular weight and adiabatic index.
+A realistic gas mixture, however, has variable adiabatic gradient and molecular weight, due to non-ideal effects and processes like dissociation or ionization.
+Future work will include realistic equation of state tables, and we expect quantitatively different results.
+In this study we analyze only the effect of the mean molecular weight of the nebular gas on the crossover time.
+In Figure \ref{fig:tvsM} we compare our standard model given by $\delad=2/7$ and $\mu=2.35$ from Figure \ref{fig:tvsM} with a model with the same adiabatic index but a different molecular mass, $\mu=2$ (i.e., purely molecular hydrogen).
 For comparison, we also plot the results predicted by the analytic, non self-gravitating model described in section \ref{analytic}.
@@
 We therefore estimate the analytic crossover time as the time needed for the atmosphere to reach an effective crossover mass $M_{\rm{atm}}=f M_{\rm c}$, where $f = 0.13$.
-
-\begin{figure*}[tb]
-\centering
-\includegraphics[width=1.\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/TdPd_effect.pdf}
-%\vspace{-0.5in}
-\caption{Left panel: crossover time as a function of disk temperature for fixed disk pressure and a core of fixed mass $M_{\rm c}=5 M_{\oplus}$.
-Right panel: crossover time as a function of disk pressure for fixed disk temperature and a core of fixed mass  $M_{\rm c}=5 M_{\oplus}$.
-The analytic scalings are plotted for comparison, for an effective crossover mass $M_{\rm{co}}=f M_{\rm c}$ with $f=0.13$.
-Gas accretion is faster for lower temperatures and higher pressures.}
-\label{fig:TPeffects}
-\end{figure*}
-
-We find that the crossover time is shorter for lower disk temperatures and higher pressures. We discuss the two effects separately.
-
-An increase in disk temperature results in a longer crossover time, as a higher temperature causes an increase in opacity and hence a lower luminosity.
-The analytic model predicts $t_{\rm{co}} \propto T^{\beta+ 1/2}$ (see equation (\ref{eq:tcoolan})), where $\beta$ is the constant in the opacity law (\ref{eq:opacitylaw}).
-For our choice of $\beta=2$, the crossover time is therefore strongly dependent on the opacity.
-Different opacity assumptions will result in a different scaling with temperature of $t_{\rm{co}}$.
-We notice a generally good agreement between our numerical results and the scaled analytic approximation in the left panel of Figure \ref{fig:TPeffects}.
-A factor of $\sim 3$ reduction in the disk temperature results in an decrease of the growth time by a factor of $\sim 30$.
-The atmosphere evolution is therefore strongly dependent on the temperature in the protoplanetary disk.
-
-An increase in disk pressure will decrease the crossover time.
-This effect can also be explained by the analytic model.
-From equations (\ref{eq:tcoolan}) and (\ref{eq:xi}), the crossover time is inversely proportional to the disk pressure.
-A higher disk pressure will therefore result in a shorter crossover time.
-Increasing the disk pressure by a factor of $\sim 100$ only reduces the crossover time by a factor of $\sim 2$.
-As such, the crossover time is only weakly dependent on the nebular pressure.
-
-%From equation (\ref{eq:PcbRcb}), a higher disk pressure will yield the same pressure at the radiative-convective boundary $P_{\cb}$ at a larger radius of the convective zone $R_{\cb}$, and thus at a larger atmosphere mass from equation (\ref{eq:Matman}).
-A higher mass implies the atmosphere will become critical on a shorter timescale.
-
- %This effect is likely due to the fact that a higher disk pressure, and hence density, leads to the atmosphere becoming self-gravitating and therefore initiating runaway accretion on a faster timescale.
-However, increasing the disk pressure by a factor of $\sim 100$ only reduces the atmosphere growth time by a factor of $\sim 2$.
-As such, the atmosphere cooling time is only weakly dependent on the nebular pressure.
-
-%\begin{figure}[h]
-%\centering
-%\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Pd_effect.pdf}
-%%\vspace{-0.5in}
-%\caption{Disk pressure effect on growth time. Gas accretion is faster for higher pressures.}
-%\end{figure}
-
-%In contrast with the pressure effect, an increase in disk temperature causes the growth time to increase as well.
-This is due to opacity effects: a higher disk temperature increases the opacity in the radiative zone of the atmosphere, causing a decrease in luminosity and therefore a longer Kelvin-Helmholtz accretion time.
-Moreover, a factor of $\sim 3$ reduction in the disk temperature results in an decrease of the growth time by a factor of $\sim 30$.
-The atmosphere evolution is therefore strongly dependent on the temperature in the protoplanetary disk.
-%We also notice that the analytic scaling is
-
-\subsection{Role of Mean Molecular Weight and Opacity}
-\label{muopacity}
-%
-%In this section we explore the effect of the core mass on the atmosphere evolution time, for a fixed semi-major axis.
-We explore again the case of $a=10$ AU.
-Figure \ref{fig:tvsM} shows the growth time behavior for a range of cores with masses between 1 and 10 $M_{\oplus}$.
-As expected, growth happens on a faster timescale if the core is more massive, as it can bind a gaseous atmosphere quicker due to larger gravity.
-We find that the atmosphere growth time is very sensitive to the core mass: the mass increase needed to reduce the cooling time by an order of magnitude is only half an Earth mass.
-%
-%\begin{figure}[h]
-%\centering
-%\includegraphics[width=0.52\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cumul_coolingtime_vs_Matm_10au_mu235.pdf}
-%%\vspace{-0.5in}
-%\caption{Time evolution as a function of atmosphere mass for a range of core masses.
-The circles mark the time when runaway accretion is initiated.
-A larger core mass results in a lower cooling time.}
-%\label{fig:tvsM}
-%\end{figure}
-
-In this work we build evolutionary atmosphere profiles assuming an ideal gas equation of state with constant mean molecular weight and adiabatic index.
-A realistic gas mixture, however, has variable adiabatic gradient and molecular weight, due to non-ideal effects and processes like dissociation or ionization.
-Future work will include real gas equation of state tables, and we expect quantitatively different results.
-In this study we analyze only the effect of the mean molecular weight of the nebular gas on the crossover time.
-In Figure \ref{fig:tvsMcomp}, we compare our standard model given by $\delad=2/7$ and $\mu=2.35$ with a model with the same adiabatic index but a different molecular mass, $\mu=2$ (i.e., purely molecular hydrogen).
-We also plot the analytic results for comparison, using the same fractional mass scaling as in subsection \ref{TPeffects}, $f=0.13$, and notice that they are in good agreement with the numerical models.
-A lower mean molecular weight translates into a longer crossover time for the same core mass, since an atmosphere composed of heavier molecules is easier to bind.
+The scaled analytic results are in good agreement with the numerical models.
+A lower mean molecular weight translates into a longer crossover time for the same core mass, since an atmosphere composed of heavier molecules has a smaller scale height, and therefore a weaker pressure gradient to oppose self-gravity.
 
@@
 
-A decrease in opacity reduces the number of steps that photons need to diffuse (since it increases their mean free path), resulting in a higher luminosity.
-In Figure \ref{fig:LtvsMopacity}, we explore opacity laws in which the coefficient $\kappa_0$ from equation (\ref{eq:opacitylaw}) is reduced by factors of 10 and 100.
-All things equal, a lower opacity increases the luminosity and hence speeds atmosphere growth by the same factor as the opacity reduction (see equations (\ref{eq:structd}) and (\ref{eq:delrad})).
-
- \begin{figure}[h]
+\subsubsection{Disk Temperature and Pressure}
+\label{TPeffects}
+
+A planet's location in the protoplanetary disk affects its crossover time only because the disk temperature and pressure vary with stellocentric distance.
+Here, we study the separate effects of disk temperature and pressure on the time evolution of the atmosphere.
+At a fixed distance in the disk $a=10$ AU and for a core mass $M_{\mathrm{c}}=5 M_{\oplus}$, we first fix the disk pressure $P_{\rm{d}}$ at the value given by equation (\ref{eq:Pd}) and vary the disk temperature $T_{\rm{d}}$.
+We then fix the disk temperature at the value given by equation (\ref{eq:diskb}) and vary $P_{\rm{d}}$.
+The resulting variation in crossover time is plotted in Figure \ref{fig:TPeffects}.
+For comparison, we also plot the analytic results, using the same fractional mass scaling as in subsection \ref{muopacity}, $f=0.13$,
+
+\begin{figure*}[tb]
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/opacity_effect.pdf}
+\includegraphics[width=1.\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/TdPd_effect.pdf}
 %\vspace{-0.5in}
-\caption{Luminosity and cooling time evolution with total atmosphere mass for planets of fixed core mass $M_{\rm c}=5 M_{\oplus}$ and located at $a=10$ AU, for different opacities.
-The standard dust opacity is reduced by factors of 10 and 100, respectively.
-A lower opacity accelerates atmosphere growth.}
-\label{fig:LtvsMopacity}
-\end{figure}
+\caption{Left panel: crossover time as a function of disk temperature for fixed disk pressure $P_{\rm d}=0.067$ dyn cm$^{-2}$ given by equation (\ref{eq:Pd}) at 10 AU and a core of fixed mass $M_{\rm c}=5 M_{\oplus}$.
+Right panel: crossover time as a function of disk pressure for fixed disk temperature $T_{\rm d}=44.73$ K given by equation (\ref{eq:diskb}) at 10 AU and a core of fixed mass  $M_{\rm c}=5 M_{\oplus}$.
+The analytic scalings are plotted for comparison, for an effective crossover mass $M_{\rm{co}}=f M_{\rm c}$ with $f=0.13$.
+Gas accretion is faster for lower temperatures and higher pressures.}
+\label{fig:TPeffects}
+\end{figure*}
+
+We find that the crossover time is shorter for lower disk temperatures and higher pressures. We discuss the two effects separately.
+
+An increase in disk temperature results in a longer crossover time, as a higher temperature causes an increase in opacity.
+As a result, photons diffuse through the radiative layer more slowly, producing a lower planetary luminosity.
+Our analytic model predicts $t_{\rm{co}} \propto T^{\beta+ 1/2}$ (see equation \ref{eq:tcoolan}), where $\beta$ is the constant in the opacity law (\ref{eq:opacitylaw}).
+For our choice of $\beta=2$, the crossover time is therefore strongly dependent on the opacity.
+Different opacity assumptions will result in a different scaling with temperature of $t_{\rm{co}}$.
+We notice a generally good agreement between our numerical results and the scaled analytic approximation in the left panel of Figure \ref{fig:TPeffects}.
+A factor of $\sim 3$ reduction in the disk temperature results in an decrease of the growth time by a factor of $\sim 30$.
+The atmosphere evolution is therefore strongly dependent on the temperature in the protoplanetary disk.
+
+From equations (\ref{eq:tcoolan}) and (\ref{eq:xi}), the crossover time is inversely proportional to disk pressure.
+A higher disk pressure therefore results in a shorter crossover time.
+Increasing the disk pressure by a factor of $\sim 100$ only reduces the crossover time by a factor of $\sim 2$.
+As such, the crossover time is only weakly dependent on the nebular pressure.
+
+%From equation (\ref{eq:PcbRcb}), a higher disk pressure will yield the same pressure at the radiative-convective boundary $P_{\cb}$ at a larger radius of the convective zone $R_{\cb}$, and thus at a larger atmosphere mass from equation (\ref{eq:Matman}).
+A higher mass implies the atmosphere will become critical on a shorter timescale.
+
+ %This effect is likely due to the fact that a higher disk pressure, and hence density, leads to the atmosphere becoming self-gravitating and therefore initiating runaway accretion on a faster timescale.
+However, increasing the disk pressure by a factor of $\sim 100$ only reduces the atmosphere growth time by a factor of $\sim 2$.
+As such, the atmosphere cooling time is only weakly dependent on the nebular pressure.
+
+%\begin{figure}[h]
+%\centering
+%\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Pd_effect.pdf}
+%%\vspace{-0.5in}
+%\caption{Disk pressure effect on growth time. Gas accretion is faster for higher pressures.}
+%\end{figure}
+
+%In contrast with the pressure effect, an increase in disk temperature causes the growth time to increase as well.
+This is due to opacity effects: a higher disk temperature increases the opacity in the radiative zone of the atmosphere, causing a decrease in luminosity and therefore a longer Kelvin-Helmholtz accretion time.
+Moreover, a factor of $\sim 3$ reduction in the disk temperature results in an decrease of the growth time by a factor of $\sim 30$.
+The atmosphere evolution is therefore strongly dependent on the temperature in the protoplanetary disk.
+%We also notice that the analytic scaling is
 
@@
 
-We can now combine the results obtained in sections \ref{Mct}, \ref{TPeffects} and \ref{muopacity} to determine the minimum core mass necessary for a protoplanet to initiate runaways gas accretion within the disk lifetime, i.e.
-to have a crossover time equal to the disk life timescale.
+We can now combine the results obtained in section \ref{Mct} to determine the minimum core mass necessary for a protoplanet to initiate runaway gas accretion within the disk lifetime, i.e.
+to have a crossover time less than the disk lifetime.
 We  denote this mass the \textit{critical core mass}.
@@
 Observations of protoplanetary disks show that they dissipate over a time scale of a few million years (e.g., \citealt{lagrange00}, \citealt{haisch01}, \citealt{goldreich04}).
-We adopt a disk life timescale of $\tau=3 $ Myrs and determine the critical core mass as a function of gas and disk conditions.
+We adopt a disk life timescale of $\tau=3 $ Myrs and determine the critical core mass as a function of disk temperature, pressure and mean molecular weight of the nebular gas.
 
@@
 %\vspace{-0.5in}
-\caption{The minimum core mass for an atmosphere to initiate runaway gas accretion within the lifetime of the disk $t \sim 3 Myrs$ as a function of semi-major axis, for different mean molecular weight of the nebular gas.
+\caption{The minimum core mass for an atmosphere to initiate runaway gas accretion within the lifetime of a typical protoplanetary disk $t \sim 3$ Myrs as a function of semi-major axis, for different mean molecular weight of the nebular gas.
 The analytic results are plotted for comparison.
-The critical core mass is lower further out in the disk.}
+The critical core mass is lower at larger stellocentric distances.}
 \label{fig:Mcvsa}
@@
 
- In Figure \ref{fig:Mcvsa}, we see that the critical core mass decreases as we move further away in the disk, primarily due to the fact the disk temperature is lower in the outer regions of the disk.
-As we showed in section \ref{TPeffects}, the atmosphere grows faster for lower temperatures, which offsets the need for a larger core mass.
+ In Figure \ref{fig:Mcvsa}, we see that the critical core mass decreases for larger separations, primarily due to the fact the disk temperature is lower in the outer regions of the disk.
+As we showed in section \ref{Mct}, the atmosphere grows faster for lower temperatures, which offsets the need for a larger core mass.
 The critical core mass is larger for a lower mean molecular weight, consistent with the results derived in section \ref{Mct}.
-\textbf{Add explanation for analytic curve behavior --- need to chat about this with Andrew}.
-We also explore the opacity effect on the critical core mass in Figure \ref{fig:Mcritopacity} For an opacity, and therefore crossover time, reduced by an order of magnitude, the critical core mass is reduced by roughly a factor of 2.
+We also explore the opacity effect on the critical core mass in Figure \ref{fig:Mcritopacity}.
+For an opacity, and therefore crossover time, that is an order of magnitude larger, the critical core mass is increased by roughly a factor of 2.
 
@@
 
-\section{Discussion}
-\label{discussion}
-
-In this section we discuss the parameter space of validity of our model.
+%\section{Discussion}
+%\label{discussion}
+%
+%In this section we discuss the parameter space of validity of our model.
 We calculate the conditions under which planetesimal accretion can be ignored, and review the effects of some of the approximations that come into our model.
 
-\subsection{Comparison with Planetesimal Accretion \textbf{(Paper II)}}
-\label{acc}
-
-In this study we have considered atmospheres for which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the atmosphere.
-This is different from standard calculations, in which the atmosphere is heated by planetesimal accretion.
-In this section we investigate the core accretion rates that are necessary for our regime to be valid.
-We also discuss the conditions under which runaway gas accretion can be initiated due to the Kelvin-Helmholtz contraction of the atmosphere before it becomes critical due to planetesimal accretion.
-
-We estimate the planetesimal accretion rate consistent with our assumptions that $L_{\mathrm{acc}} \ll L_{\rm{KH}}$. Here $L_{\rm{acc}}$ is the accretion luminosity given by
-
-\begin{equation}
-\label{eq:Lacc}
-L_{acc}=G \frac{M_{\rm{c}} \dot{M_{\rm{c}}}}{R_{\rm{c}}},
-\end{equation}
-
-\noindent where $\dot{M_{\rm{c}}}$ is the core accretion rate, and $L_{\rm{KH}}$ is the luminosity of the atmosphere due to gas contraction as calculated in section \ref{KH}.
-At the limit, $L_{\rm{acc}}=L_{\rm{KH}}$.
-For a given atmosphere model we can therefore estimate the maximum planetesimal accretion rate during the gas contraction phase in order for the atmosphere to be dominated by the Kelvin-Helmholtz luminosity.
-We choose as a fiducial case an atmosphere forming at 10 AU and with a core mass of $10 M_{\oplus}$.
-For this choice or parameters, the atmosphere cooling time is $\sim$ 1.2 Myrs (see Figure 7), which is within the typical life time of a protoplanetary disk.
-The results are presented in Figure 10.
-
- \begin{figure}[h]
-\centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/acc_rates_poster.pdf}
-%\vspace{-0.5in}
-\caption{Various relevant accretion rates.
-$\dot{M}_{\rm{atm}}$ is the growth rate of the atmosphere as estimated by our model, and $\dot{M}_{\rm{c,KH}}$ is the maximum planetesimal accretion rate during the gas contraction phase in order for our regime to be valid.
-For comparison, we plot the core accretion rate $\dot{M}_{\rm{c,acc}}$ necessary to grow the core on the same time scale as the atmosphere $\tau \sim 1.2$ Myrs, and a typical planetesimal accretion rate where the random velocity of the planetesimals is given by the Hill velocity due to the core.}
-\end{figure}
-
-We label the resulting minimum core accretion rate as $\dot{M}_{\rm{c,KH}}$.
-The atmosphere growth rate is also plotted for comparison.
-We therefore see that the core accretion rate has to be $\sim2-3$ orders of magnitude lower than the atmosphere accretion rate for our assumptions to be valid.
-If the core had accreted planetesimals at this constant rate since it started forming, then the formation of a core massive enough to attract an atmosphere would not have been possible within typical disk life timescales, which indicates a slowing down of the planetesimal accretion regime.
-Possible explanations for that include the core having formed in the inner part of the disk and later migrated outwards, or the core having been depleted of planetesimals due to a giant neighbor.
-We also estimate the core accretion rate
-
-\begin{equation}
-\label{eq:Mcdot}
-\dot{M}_{\rm{c,acc}}(M_{\rm{c}}) \equiv \frac{M_{\rm{c}}}{\tau}
-\end{equation}
-
-\noindent needed for the core to form on the same timescale as our model atmosphere, $\tau=1.2$ Myrs, as well as a typical planetesimal accretion rate, for which the random velocities of the planetesimals are of the order of the Hill velocity around the protoplanetary core \citep{goldreich04}.
-We denote this latter rate as $\dot{M}_{\rm{c,typical}}$.
-This rate, which is at the boundary of the dispersion dominated and shear dominated regimes ,is often quoted as a maximum (refs).
-However, it is easy to see that  $\dot{M}_{\rm{c,typical}}$ is more than one order of magnitude lower than the gas accretion rate of our model atmosphere $\dot{M}_{\rm{atm}}$, and lower than the core accretion rate $\dot{M}_{\rm{c,acc}}$ needed to grow the core and the atmosphere at the same time within the disk life time.
-As such, the formation of a giant planet by growing the core first, then letting the atmosphere cool is faster than growing the core and the atmosphere at the same time at a steady planetesimal accretion rate.
-
-Next, we are interested in whether hydrodynamic gas accumulation due to planetesimal accretion can already commence before the atmosphere becomes unstable due to Kelvin-Helmholtz contraction, as our regime is no longer the relevant one under such conditions.
-A core that forms on the same timescale as our model atmosphere accretes planetesimals at a rate given by equation (\ref{eq:Mcdot}).
-This accretion rate is dependent on the core mass, which is steadily increasing.
-We therefore compare the critical core mass due to planetesimal accretion at this rate $M_{\rm{crit,acc}}$ to a core mass assumed fixed $M_{\rm{c,fixed}}$.
-If $M_{\rm{crit,acc}}<M{\rm_{c, fixed}}$, then the atmosphere has already initiated unstable gas accretion by the time Kelvin-Helmholtz contraction starts dominating.
-
-%Specifically, we compare the critical core mass due to planetesimal accretion at the rate
-
-%we compare the critical core mass $M_{\rm{c,acc}}$ due to planetesimal accretion, for an accretion rate that satisfies $L_{\rm{acc}}<L_{\rm{KH}}$ to the actual core mass $M_{\rm{c}}$ assumed fixed in our model.
-If $M_{\rm{c,acc}}<M{\rm{c}}$, then the atmosphere has already initiated unstable gas accretion by the time Kelvin-Helmholtz contraction starts dominating.
-
-In order to estimate the critical core mass due to planetesimal accretion $M_{\rm{crit,acc}}$, we use the results of \citet{rafikov06} for low luminosity atmospheres forming in the outer disk ($>2-5$ AU), consistent with our region of interest.
-By relating his expression for the critical core mass to a given core mass dependent planetesimal accretion rate $\dot{M}(M_{\rm{c}})$, we find the following expression for the critical core mass when accretion luminosity dominates the evolution of the atmosphere:
-
-\begin{equation}
-\label{eq:critraf}
-M_{\rm{crit, acc}} \sim \Big[\frac{\dot{M}(M_{\rm{c}})}{64 \pi^2} \frac{\kappa_0}{\sigma G^3} \frac{1}{R_c M_c^{1/3}} \Big(\frac{k}{\mu}\Big)^4\Big]^{3/5},
-\end{equation}
-
-\noindent with all the constants as defined in previous sections. We therefore calculate $M_{\rm{crit,acc}}$ for a range of core masses. The result is displayed in Figure 11.
-
- \begin{figure}[h]
-\centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/critcore.pdf}
-%\vspace{-0.5in}
-\caption{Comparison between the critical core mass $M_{\rm{crit, acc}}$ due to planetesimal accretion and the assumed fixed core mass when gas contraction dominates, for a growth time of $\tau=1.2$ Myrs.
-Unstable atmosphere collapse occurs due to core accretion before Kelvin-Helmholtz contraction starts dominating for core masses larger than $\sim 40 M_{\oplus}$.}
-\end{figure}
-
-We see that the critical core mass due to planetesimal accretion is smaller than the mass of the fixed core for $M_{\rm{c}} \sim 40 M_{\oplus}$.
-This means that for large core masses, planetesimal accretion will lead to runaway gas accretion before gas contraction starts dominating, and hence our model is not applicable in that regime.
-However, we have found in section \ref{critcore} that unstable atmospheres collapse occurs within the life time of the disk for protoplanetary cores smaller than this.
-We therefore confirm that planetesimal accretion can be safely ignored in our regime of interest.
-Moreover, as displayed in Fig.
-10, it is faster to form a planet by growing the core first, then accreting a massive envelope, rather than by growing the core and atmosphere in parallel.
-\textbf{This works for this particular choice of core, disk etc.
-conditions, and I am expecting it to work for most of our parameter space, but it would probably be good to explore this numerically, at least to order of magnitude, with the estimates we have.}
-
-As a final check, we investigate whether planetesimal accretion during the gas contraction phase at the rate $\dot{M}_{\rm{KH}}$ imposed by the condition that $L_{\rm{acc}}<L_{\rm{KH}}$ can alter the core mass enough to affect the time evolution of the atmosphere.
-We can quantitatively estimate the increase in core mass as
-
-\begin{equation}
-\label{eq:cminc}
-\Delta M_{\rm{c}} = \int_0^{t_{\rm{crit}}} \dot{M_{\rm{c}}} dt \approx \sum_i \dot{M_{\rm{c}}}_i \Delta t_i,
-\end{equation}
-
- \noindent where $t_{\rm{crit}}$ is the time elapsed until runaway gas accretion commences, and the accretion rate $ \dot{M_{\rm{c}}}_i $ is given by
-
- \begin{equation}
- \label{eq:Mdotexp}
- \dot{M_{\rm{c}}}_i =\frac{L_i R_c}{G M_{\rm{c}}} 
- \end{equation}
-
- \noindent from equation (\ref{eq:Lacc}), with $L_i$ the luminosity of the atmosphere at time $t_i$ in our model.
-For $M_{\rm{c}}=10 M_{\oplus}$, we find $\Delta M_{\rm{c}} \approx 0.05 M_{\oplus}$.
-This mass increase is therefore negligible in comparison with the initial core mass.
-It follows that a significant increase in core mass that could potentially alter the time evolution of the atmosphere would occur on a longer time scale than the mass-doubling time for the unperturbed atmosphere.
-Therefore, the time evolution of the atmosphere is insensitive to core mass changes at a rate imposed by the assumption that $L_{\rm{acc}}<L_{\rm{KH}}$.
-
- \subsection{Neglected Effects}
+ \section{Neglected Effects}
  \label{neglected}
@@
 
- \subsubsection{Equation of State \textbf{(Paper II)}}
+ \subsection{Planetesimal Accretion}
+
+ In this study we explore the formation of giant planets through core accretion under the assumption that the protoplanetary core is already fully formed, and therefore the planetesimal accretion rate is low.
+We defer quantitative estimates of the parameter space under which this is a valid assumption to Piso et al.
+(2013, in prep).
+
+ \subsection{Equation of State}
  \label{EOS}
@@
 
- \subsubsection{Time Evolution \textbf{(Paper I)}}
+ \subsection{Time Evolution}
 
@@
 
- \subsubsection{Inefficient Convection\textbf{(Paper I)}}
+ \subsection{Inefficient Convection}
 
@@
 
-\subsubsection{Opacity \textbf{(Paper II)}}
-
-\textbf{In progress, still need to run models for smaller AU and see at what point ice grain opacity is no longer valid.
+%\subsubsection{Opacity \textbf{(Paper II)}}
+
+%\textbf{In progress, still need to run models for smaller AU and see at what point ice grain opacity is no longer valid.
 Currently strange behavior at 1 AU for $\delad=2/5$, the critical core mass goes up significantly instead of having a linear trend.

### 2013-06-14 tex/draft/draft2.tex
current paper versions, plus figures for paper II

@@
 \newcommand{\co}{_{\rm c}}
-\newcommand{\di}{_{\rm o}}
+\newcommand{\di}{_{\rm d}}
 \newcommand{\cb}{_{\rm RCB}}
@@
 
-\title{On the Minimum Core Mass for Giant Planet Formation through Kelvin-Helmholtz Contraction}
+\title{On the Minimum Core Mass for Giant Planet Formation}
 \author{Ana-Maria A. Piso}
@@
 Previous studies have found that there exists a ``critical core mass'' past which hydrostatic solutions can no longer be found and unstable atmosphere collapse occurs.
+This core mass is typically quoted to be around $10 M_{\oplus}$.
 In standard calculations of the critical core mass, planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse.
 In this study we consider the extreme case in which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the planet.
-We develop a two-layer atmosphere model with an inner convective region and an outer radiative zone that matches onto the protoplanetary disk, and we determine the minimum core mass for a giant planet to form within the typical disk life timescale for a variety of disk conditions, which we denote as  ``critical core mass''.
+We develop a two-layer atmosphere model with an inner convective region and an outer radiative zone that matches onto the protoplanetary disk, and we determine the minimum core mass for a giant planet to form within the typical disk life timescale for a variety of disk conditions, which we denote as  \textit{critical core mass}.
 We find that the absolute minimum core mass required to nucleate atmosphere collapse within the disk lifetime is smaller for planets forming further away from their host stars.
@@
 Our results yield lower mass cores than corresponding studies for large planetesimal accretion rates.
-We therefore show that it is easier to form a planet by growing the core first, then accreting a massive gaseous envelope, rather than forming the core and atmosphere simultaneously.
+%We therefore show that it is easier to form a planet by growing the core first, then accreting a massive gaseous envelope, rather than forming the core and atmosphere simultaneously.
 
@@
 
-Current theories of giant planet formation postulate that these planets form either through core accretion (refs), in which solid planetesimals collide and grow into a massive solid core, which then accretes a gaseous envelope, or due to a gravitational instability in the protoplanetary disk that leads to fragmentation of the disk into self-gravitating clumps (refs, inc.
-\citealt{dangelo11}).
+Current theories of giant planet formation postulate that these planets form either through core accretion (e.g., \citealt{harris78}, \citealt{mizuno78}, \citealt{stevenson82}, \citealt{boden86}, \citealt{wuchterl93}, \citealt{pollack96}, \citealt{dangelo11}), where solid planetesimals collide and grow into a massive solid core which then accretes a gaseous envelope, or due to a gravitational instability in the protoplanetary disk that leads to fragmentation of the disk into self-gravitating clumps (e.g., \citealt{harris78}, \citealt{boss97}, \citealt{rafikov05}, \citealt{kratter10}, \citealt{dangelo11}).
 %quote murray-clay, kratter 10, rafikov 05, etc.
 
-Standard core accretion models (refs) assume that the core and the atmosphere grow at the same time, and that planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse, while the heat generated by the gravitational (Kelvin-Helmholtz) contraction of the atmosphere is neglected.
-These studies consider that the planet atmosphere is in steady state, in which all the luminosity due to planetesimal accretion is radiated away by the envelope, and  find that there exists a minimum (``critical'') core mass past which hydrostatic solutions can no longer be found and unstable atmosphere collapse occurs.
-
-Forming giant planets at wide separations in the disk poses theoretical challenges.
-On the one hand, gravitational instability generates objects that are too massive to explain the current observed properties of exoplanets (refs, inc.
-\citealt{rafikov05}).
-On the other hand, planetesimal accretion is slow at large distances in the disk, and therefore large cores may not be able to form before the dissipation of the disk (refs).
-I would therefore be easier if giant planets could form from smaller cores which would need less time to grow.
+Standard core accretion models (e.g., \citealt{stevenson82}, \citealt{rafikov06}) assume that the core and the atmosphere grow simultaneously, and that planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse.
+These studies consider that the planet atmosphere is in a steady state at all times, where all the incoming energy from planetesimal accretion is radiated away by the envelope, and  find that there exists a minimum (``critical'') core mass past which hydrostatic solutions can no longer be found and unstable atmosphere collapse occurs.
+Standard calculations find the critical core mass to be about $10 M_{\oplus}$ (e.g., \citealt{stevenson82}, \citealt{rafikov06}).
+
+Forming giant planets at wide separations in the protoplanetary disk poses theoretical challenges.
+On the one hand, gravitational instability generates objects that are too massive to explain the current observed properties of exoplanets (e.g., \citealt{rafikov05}).
+On the other hand, planetesimal accretion is slow at large distances in the disk, and therefore large cores may not be able to form before the dissipation of the gas in the disk (e.g., \citealt{rafikov06}).
+It would therefore be easier if giant planets could form from smaller cores which need less time to grow.
 
@@
 In this scenario, the planetesimal accretion rate is significantly slowed down during the gas contraction phase of the atmosphere.
-This reduction can arise due to dynamical clearing, or due to the core having formed in the inner parts of the disk and migrated outwards, etc.
-In this situation, the atmosphere evolution is dominated by the Kelvin-Helmholtz contraction of the envelope.
+This reduction can arise, for example, due to dynamical clearing, or due to the core having formed in the inner parts of the disk and migrated outwards.
+In this situation, the atmosphere evolution is dominated by the contraction of the envelope rather than planetesimal accretion.
 The atmosphere is no longer in a steady state, but rather it accretes gas as it loses energy through radiation.
@@
 The time required for the atmosphere to grow to this stage is the characteristic growth time of the atmosphere.
-For a set of fixed gas and disk conditions, there exists a minimum core mass for which the atmosphere can grow on the time scale described above within the life time of the protoplanetary disk, which we define as the ``critical core mass''.
-
-We develop a two-layer atmosphere model, with a convective inner region and a radiative outer region that matches smoothly on to the protoplanetary disk, and develop a cooling model that evolves the atmosphere in time.
-We aim to find the critical core mass for a giant planet to form before the dissipation of the disk.
+For a set of fixed gas and disk conditions, there exists a minimum core mass for which the atmosphere can grow on the time scale described above within the life time of the protoplanetary disk, which we define as the \textit{critical core mass}.
+
+We build a two-layer atmosphere model, with a convective inner region and a radiative outer region that matches smoothly on to the protoplanetary disk, and develop a cooling model that evolves the atmosphere in time.
+We aim to find the critical core mass for a giant planet to form before disk dissipation.
 
@@
 In section \ref{analytic}, we present a simplified analytic model that predicts the qualitative behavior of the numerical model.
-We describe our results in section \ref{KH}, and determined the critical core mass for planet formation during the life time of the protoplanetary disk in section \ref{critical}.
-We discuss our results in section \ref{discussion} and summarize our findings in section \ref{conclusions}.
+We describe our numerical results in section \ref{KH}, and determine the critical core mass for planet formation during the life time of the protoplanetary disk in section \ref{critical}.
+We discuss some of the approximations of our model in section \ref{neglected} and summarize our findings in section \ref{conclusions}.
 
@@
 Section \ref{struct} presents the equations describing the structure of a static planetary atmosphere.
-In section \ref{BCs} we describe discuss our boundary conditions and the properties of our assumed protoplanetary disk, and review relevant length scales.
-and the boundary conditions that we use.
+In section \ref{BCs} we discuss our boundary conditions and the properties of our assumed protoplanetary disk, and review relevant length scales.
 In section \ref{cooling} we develop a model to estimate the rate at which the atmosphere cools.
-Finally, we combine these calculations into a quasi-static model for atmospheric evolution in section \ref{twolayer}.
+%Finally, we combine these calculations into a quasi-static model for atmospheric evolution in section \ref{twolayer}.
 
@@
 \begin{enumerate}
-\item The planet consists of a solid core of fixed mass and a two-layer atmosphere composed of an inner convective zone and an outer radiative zone that matches smoothly on to the disk, as mentioned in section.
+\item The planet consists of a solid core of fixed mass and a two-layer atmosphere composed of an inner convective zone and an outer radiative region that matches smoothly on to the protoplanetary disk.
 The two regions are separated by the Schwarzschild criterion for convective instability (see section \ref{struct}).
-We denote the surface between the two regions as the radiative-convective boundary (RCB), which is henceforth defined by a radius $r=R_{RCB}$.
-\item The luminosity evolution of the atmosphere is dominated by Kelvin-Helmholtz contraction rather than planetesimal accretion. 
+We denote the surface between the two regions as the radiative-convective boundary (RCB), which is henceforth defined by a radius $r=R\cb$.
+\item The luminosity evolution of the atmosphere is dominated by Kelvin-Helmholtz contraction, while planetesimal accretion is negligible. As such, the core no longer grows and has a fixed mass.
 \item The luminosity is assumed to be constant throughout the outer radiative region.
@@
 
-In order for equation set (\ref{eq:struct}) to be solvable, it has to be supplemented by an equation of state (EOS) that relates pressure, temperature and density $P=P(\rho, T)$, and an opacity law for $\kappa$.
+In order for the equation set (\ref{eq:struct}) to be solvable, it has to be supplemented by an equation of state (EOS) that relates pressure, temperature and density $P=P(\rho, T)$, and an opacity law for $\kappa$.
 In our study, we assume an ideal gas polytropic EOS
@@
 
-\noindent with $K$ the adiabatic constant.
+\noindent with $K$ the adiabatic constant, determined by $\delad$.
 The adiabatic gradient is $\delad=2/5$ for a monatomic gas and $\delad=2/7$ for a diatomic gas.
@@
 
-\noindent with $\mathcal{R}$ defined in section \ref{model}.
-The specific gas constant $\mathcal{R}$ depends on the mean molecular weight of the gas $\mu$.
-$\mu$ is determined by the relative abundance of hydrogen and helium in the atmosphere.
+\noindent where  $\mathcal{R}$ is the reduced gas constant: $\mathcal{R}=k_b/(\mu m_p)$, with $k_b$ the Boltzmann constant and $m_p$ the proton mass.
+The gas constant $\mathcal{R}$ depends on the mean molecular weight of the gas $\mu$, which is determined by the relative abundance of hydrogen and helium in the atmosphere.
 In this study we assume that the helium fraction only affects the molecular weight of the gas, and not the adiabatic index.
@@
 \label{eq:opacitylaw}
-\kappa=\kappa_0 \Big(\frac{P}{P_d}\Big)^{\alpha} \Big(\frac{T}{T_d}\Big)^{\beta},
-\end{equation}
-
-\noindent with $\alpha$ and $\beta$ constants, and $\kappa_0=2 F_{\kappa}$.
-To estimate $\alpha$ and $\beta$ we use the \citet{bell94} opacity laws for ice grains: $\alpha =0 $, $\beta=2$ and $F_{\kappa}=1$ (i.e., $\kappa_0=2$).
-We note that these values are valid only for low disk temperatures: $T_d< \sim 100 K$.
-As such, we only consider cool atmospheres that form in the outer part of the protoplanetary disk ($a \geq 10$ AU).
+\kappa=\kappa_0 \Big(\frac{P}{P_{\rm{ref}}}\Big)^{\alpha} \Big(\frac{T}{T_{\rm{ref}}}\Big)^{\beta},
+\end{equation}
+
+\noindent with $\alpha$ and $\beta$ constants, $\kappa_0=2 F_{\kappa}$, and $T_{\rm{ref}}$ and $P_{\rm{ref}}$ reference temperature and pressure, respectively.
+To estimate $\alpha$ and $\beta$ we use the \citet{bell94} opacity laws for ice grains: $\alpha =0 $, $\beta=2$, $T_{\rm{ref}}=100$ K, and $F_{\kappa}=1$ (i.e., $\kappa_0=2$).
+We note that these values are valid only for low disk temperatures: $T_d \lesssim 100 K$.
+As such, we only consider cool atmospheres that form in the outer part of the protoplanetary disk ($a \geq 5$ AU).
 
@@
 
-We first define several characteristic length scales that are important for this problem. \\
+In what follows we first define several characteristic length scales that are important for this problem. \\
 
@@
 \label{eq:rc}
-R_c \equiv \Big(\frac{3 M_c}{4 \pi \rho_c}\Big)^{1/3},
-\end{equation}
-
-\noindent with $M_c$ and $\rho_c$ the core mass and density, respectively. Since we assume that the core mass does not change with time, the core radius is also fixed for a given model. \\
+R\co \equiv \Big(\frac{3 M\co}{4 \pi \rho\co}\Big)^{1/3},
+\end{equation}
+
+\noindent with $M\co$ and $\rho\co$ the core mass and density, respectively. Since we assume that the core mass does not change with time, the core radius is also fixed for a given model. \\
 
@@
 \label{eq:RB}
-R_B \equiv \frac{G M_p}{c_s^2}=\frac{G M_p}{\mathcal{R} T_d},
-\end{equation}
-
-\noindent where $G$ is the gravitational constant, $M_p$ is the total mass of the planet, $c_s$ is the sound speed, and $\mathcal{R}$ is the reduced gas constant: $\mathcal{R}=k_b/(\mu m_p)$, with $k_b$ the Boltzmann constant and $m_p$ the proton mass.
-\\
-
-\textbf{The Hill Radius} represents the distance at which the gravitational attraction of the planet and of the host star are comparable. It is given by
+R_{\rm B} \equiv \frac{G M_{\rm p}}{c_s^2}=\frac{G M_{\rm p}}{\mathcal{R} T\di},
+\end{equation}
+
+\noindent where $G$ is the gravitational constant, $M_{\rm p}$ is the total mass of the planet, $c_{\rm s}$ is the isothermal sound speed, defined as $c_{\rm s}=\Big(\frac{k_b T\di}{\mu m_p}\Big)^{1/2}$, and $\mathcal{R}$ is the reduced gas constant defined in section \ref{struct}.\\
+
+\textbf{The Hill Radius} represents the distance at which the gravitational attraction of the planet and the tidal acceleration from the host star are the same. It is given by
 
@@
 \label{eq:RHill}
-R_H \equiv a \Big(\frac{M_p}{M_{\odot}}\Big)^{1/3}
-\end{equation}
-
-\noindent Outside the Hill radius, the gravity of the core is too weak to affect the nebular gas. Consequently, the planet will be able to accrete only gas that lies within its Hill sphere. \\
+R_{\rm H} \equiv a \Big(\frac{M_{\rm p}}{3 M_{\odot}}\Big)^{1/3}
+\end{equation}
+
+\noindent Outside the Hill radius, the gravity of the planet is too weak to affect the nebular gas. Consequently, the planet will be able to accrete only gas that lies within its Hill sphere. \\
 
@@
 \label{eq:H}
-H \equiv \frac{c_s}{\Omega},
+H \equiv \frac{c_{\rm s}}{\Omega},
 \end{equation}
@@
 The disk scale height is a measure of the thickness of the protoplanetary disk.
-Using equation (\ref{eq:diska}), we find that the scale height for our assumed disk model can be expressed as
-
-\begin{equation}
-\label{eq:H2}
-H=0.022 \sqrt{F_T} a^{9/7} \,\, \mathrm{AU}
-\end{equation}
-
-For small planet masses, $R_B<R_H<H$.
-As the atmosphere becomes more massive, tidal truncation becomes more important than thermal constraints, and $R_H<R_B<H$.
+%Using equation (\ref{eq:diska}), we find that the scale height for our assumed disk model can be expressed as
+
+%\begin{equation}
+%\label{eq:H2}
+%H=0.022 \sqrt{F_T} a^{9/7} \,\, \mathrm{AU}
+%\end{equation}
+
+For small planet masses, $R_{\rm B}<R_{\rm H}<H$.
+As the atmosphere becomes more massive, tidal truncation becomes more important than thermal constraints, and $R_{\rm H}<R_{\rm B}<H$.
 Finally, for high mass atmospheres the disk scale height becomes smaller than the Hill radius of the planet, and spherical symmetry breaks, as the planet is now affected by the vertical profile of the disk.
-
-We assume a solid core of fixed mass $M_c$ with a radius given by equation (\ref{eq:rc}).
-We choose $\rho_c=3.2$ g cm$^{-3}$ (e.g., \citealt{pap99}).
+In our study, we only accept atmosphere solutions that are spherically symmetric.
+
+We assume a solid core of fixed mass $M\co$ with a radius given by equation (\ref{eq:rc}).
+We choose $\rho\co=3.2$ g cm$^{-3}$ (e.g., \citealt{pap99}).
 Furthermore, we assume the atmosphere extends out to the boundary of the Hill sphere, defined by equation (\ref{eq:RHill}).
-If the Bondi radius (defined in equation (\ref{eq:RB})) is within the Hill radius, perturbations of the nebular gas still occur past the Bondi radius due to the gravitational influence of the core.
-If, on the other hand, $R_B>R_H$, then the Hill radius is the more relevant truncation radius since material cannot be gravitationally bound to the protoplanet outside the Hill sphere.
+If the Bondi radius (defined in equation \ref{eq:RB}) is within the Hill radius, perturbations of the nebular gas still occur past the Bondi radius due to the gravitational influence of the planet.
+If, on the other hand, $R_{\rm B}>R_{\rm H}$, then the Hill radius is the more relevant truncation radius since material cannot be gravitationally bound to the protoplanet outside the Hill sphere.
 As such, it is always safe to choose the Hill radius as the outer boundary.
-
-As stated in section \ref{model}, the atmosphere matches smoothly onto the disk at the outer boundary.
-The temperature and pressure at the Hill radius have to therefore be given by the nebular temperature and pressure: $T(R_H)=T_d$ and $P(R_H)=P_d$.
-As a disk model, we use the minimum mass, passively irradiated model of  \citet{chiang10}.
+We note, however, that we define the planet mass as only the mass enclosed within the Bondi radius if $R_{\rm B}<R_{\rm H}$.%, i.e.
+$M_{\rm{tot}}=M\co+M_{\rm{atm}}=M(<\text{min}(R_{\rm B}, R_{\rm H}))$.
+
+The atmosphere matches smoothly onto the disk at the outer boundary.
+The temperature and pressure at the Hill radius have to therefore be given by the nebular temperature and pressure: $T(R_{\rm H})=T\di$ and $P(R_{\rm H})=P\di$.
+As a fiducial disk model, we use the minimum mass, passively irradiated model of  \citet{chiang10}.
 In this model, the surface density and mid-plane temperature are given by
@@
 \label{eq:diskparam}
-\Sigma_d&=&2200 F_{\Sigma} a^{-3/2}\,\, \text{g cm}^{-2} \slabel{eq:diska}\\
-T_d &=& 120 F_T a^{-3/7} \,\,K, \slabel{eq:diskb}
+\Sigma\di&=&2200 F_{\Sigma} a^{-3/2}\,\, \text{g cm}^{-2} \slabel{eq:diska}\\
+T\di &=& 120 F_T a^{-3/7} \,\,K, \slabel{eq:diskb}
 \end{subeqnarray}
@@
 In this study, we assume $F_{\Sigma}=F_T=1$.
-The resulting mid-plane pressure is given by
+The resulting mid-plane pressure and disk scale height are given by
 
@@
 \label{eq:Pd}
-P_d=1.1 \times 10^{-4} F_{\Sigma} \sqrt{F_T m_*} a^{-45/14} \,\, \text{dyne cm}^{-2}
-\end{equation}
-
-\noindent for a molecular weight $\mu=2.35$. Here $m_* \equiv M_*/M_{\odot}$, where $M_*$ is the mass of the central star and $M_{\odot}$ is the mass of the Sun. We take $m_*=1$.
+P\di=110  F_{\Sigma} \sqrt{F_T m_*} a^{-45/14} \,\, \text{dyn cm}^{-2}
+\end{equation}
+and
+\begin{equation}
+H=0.022 \sqrt{F_T} a^{9/7}\,\, \text{AU},
+\end{equation}
+
+\noindent respectively, for a molecular weight $\mu=2.35$. Here $m_* \equiv M_*/M_{\odot}$, where $M_*$ is the mass of the central star and $M_{\odot}$ is the mass of the Sun. We take $m_*=1$.
 
@@
 
-We first describe the general ways in which planetary Klevin-Helmholtz contraction differs from the stellar case.
-
-An isolated sphere, such as a star, satisfies a simple global energy equation:
+We first describe the general ways in which planetary Klevin-Helmholtz contraction differs from the stellar case. An isolated sphere, such as a star, satisfies a simple global energy equation:
 
@@
 
-A protoplanetary atmosphere that is not isolated, but is embedded in a gas disk, satisfies a more complex cooling equation:
+A protoplanetary atmosphere that is not isolated, but rather embedded in a gas disk, satisfies a more complex cooling equation:
 
@@
 \label{eq:coolingglobal}
-L=L_c+\Gamma-\dot{E}+e_{\mathrm{acc}}\dot{M}-P_M \frac{\partial V_M}{\partial t}
+L=L\co+\Gamma-\dot{E}+e_{\mathrm{acc}}\dot{M}-P_M \frac{\partial V_M}{\partial t}
 \end{equation}
@@
 \label{eq:Eg}
-E_G=-\int_{M_c}^M \frac{G m}{r} dm,
+E_G=-\int_{M\co}^M \frac{G m}{r} dm,
 \end{equation}
@@
 \label{eq:U}
-U=\int_{M_c}^M u dm,
+U=\int_{M\co}^M u dm,
 \end{equation}
@@
 We now explain the additional terms in the cooling equation (\ref{eq:coolingglobal}).
-$L_c$ is the luminosity from the solid core, and may include planetesimal accretion and radioactive decay.
+$L\co$ is the luminosity from the solid core, and may include planetesimal accretion and radioactive decay.
 $e_{\mathrm{acc}}$ is the specific total energy brought in by mass accreting at the rate $\dot{M}$: $e_{\mathrm{acc}}=u-G M/R$.
-The last term represents the work done on a surface mass element.
-
-The global cooling equation can be derived starting from the local cooling balance.
+The last term represents the work done on a mass element as the atmosphere contracts.
+
+The global cooling equation (\ref{eq:coolingglobal}) can be derived starting from the local cooling balance.
 Integrating the local energy equation and (\ref{eq:structd}) from the core to the surface and using equation (\ref{eq:structb}) yields
@@
 Moreover, we do not take into account the self-gravity of the atmosphere in either region.
+We derive the structure of an isothermal non self-gravitating atmosphere is section \ref{iso}.
+
 In the next subsections we derive the structure and cooling behavior of such an atmosphere, deferring some of the calculations to \App{sec:analytic}.
+We estimate the temperature and pressure corrections at the radiative-convective boundary due to the fact that the outer radiative region of the atmosphere is not purely isothermal in section \ref{RCBcorr}.
+Finally, we derive the time evolution of the atmosphere and determine the minimum core mass for runaway gas accretion in section \ref{coolingan}.
 
@@
 We now consider the structure of a low mass (non self-gravitating) isothermal atmosphere.
-We assume the atmosphere matches onto a constant background density, $\rho\di$, at a distance $r_{\rm fit} = n_{\rm fit} \RB$.
-The resulting density profile is
+We assume the atmosphere matches onto a constant background density, $\rho_{\rm d}$, at a distance $r_{\rm fit} = n_{\rm fit} \RB$, where $R_{\rm B}$ is the Bondi radius defined in equation (\ref{eq:RB}).
+From equation (\ref{eq:structa}) and without taking into account the self-gravity of the atmosphere, the resulting density profile is
 \begin{equation} \label{eq:rhoiso}
-\rho = \rho_{\rm d} \exp \left({R_B \over r} - {1 \over n_B} \right) \approx   \rho_{\rm{d}} \exp \left(R_B \over r  \right),
+\rho = \rho_{\rm d} \exp \left({R_B \over r} - {1 \over n_{\rm{fit}}} \right) \approx   \rho_{\rm{d}} \exp \left(R_B \over r  \right),
 \end{equation} 
 where the approximate inequality holds deep inside the atmosphere ($r \ll \RB$) for any $n_{\rm fit} \gtrsim 1$.
-However the choice of boundary condition does have an order unity effect on the density near the Bondi radius.
-In our study, $n_{\rm{fit}}=R_{\rm B}/R_{\rm H}$, when the Bondi radius is larger than the Hill radius, and $n_{\rm fit}=1$ when the opposite is true.
-
-The mass of the atmosphere is typically determined by integrating the density profile from the core to the Bondi radius.
+However, the choice of boundary condition does have an order unity effect on the density near the Bondi radius.
+In our study, $n_{\rm{fit}}=R_{\rm H}/R_{\rm B}$.
+
+The mass of the atmosphere is determined by integrating equation (\ref{eq:structb}) from the core to the Bondi radius using the density profile (\ref{eq:rhoiso}).
 Planets can attract massive atmospheres if $\theta\co \equiv R_B/R\co \gg 1$.
@@
 %\subsection{Temperature Contrast at Convective Boundary}
-We now show that the temperature contrast between the radiative-convective boundary $T_{\rm RCB}$ and the ambient disk $T_{\rm d}$, is modest.
+In this section we estimate the temperature and pressure corrections at the radiative-convective boundary due to the fact that the radiative region is not purely isothermal.
+We show that the temperature contrast between the temperature at the radiative-convective boundary $T_{\rm RCB}$ and the ambient disk $T_{\rm d}$, is modest.
 From equation (\ref{eq:delrad}), we express the radiative lapse rate
 \begin{equation}\label{eq:delradan}
-\delrad = {3 \kappa P \over 64 \pi  G M \sigma T^4} L = \nabla\di {(P/P_{\rm d})^{1+\alpha} \over (T/T_{\rm d})^{4-\beta}}
-\end{equation}
-
-\noindent where the second equality follows from the opacity law (\ref{eq:opacitylaw}), and $\nabla \di$ is the radiative temperature gradient at the disk:
+\delrad = {3 \kappa P \over 64 \pi  G M \sigma T^4} L = \nabla\di {(P/P_{\rm d})^{1+\alpha} \over (T/T_{\rm d})^{4-\beta}},
+\end{equation}
+
+\noindent where the second equality follows from the opacity law (\ref{eq:opacitylaw}), and $\nabla_{\rm d}$ is the radiative temperature gradient at the disk:
 
@@
 \label{eq:delo}
-\nabla \di \equiv \frac{3 \kappa_0 P_{\rm d}}{64 \pi G M \sigma T_{\rm d}^4} L.
+\nabla \di \equiv \frac{3 \kappa(P{\di}, T{\di}) P{\di}}{64 \pi G M \sigma T_{\rm d}^4} L.
 \end{equation}
@@
 \noindent $M$ is the sum of the core mass and the atmosphere mass below the pressure level $P$.
-If the mass in the radiative zone is small, then we can hold $M$ fixed at the sum of the core and convective zone masses: $M = M\co + M_{\rm conv}$.
-With this assumption and power-law opacity, we get a constant value for $\nabla\di$.
-The temperature profile then integrates to
+If the mass in the radiative zone is small, then we can hold $M$ fixed at the sum of the core and convective zone masses: $M = M\co + M_{\cb}$.
+With this assumption, we get a constant value for $\nabla_{\rm d}$.
+As $\delrad=d \ln T/d \ln P$, the temperature profile integrates to
 \begin{equation}\label{eq:radTP}
-\left(T \over T_{\rm d}\right)^{4-\beta} - 1 = {\nabla\di \over \nabla_\infty} \left[\left({P \over P_{\rm d}}\right)^{1-\alpha} - 1 \right] \, ,
-\end{equation} 
-where $\nabla_\infty = (1+\alpha)/(4-\beta)$ is $\delrad$ for $T ,P \rightarrow \infty$.\footnote{Our definitions of $\nabla\di$ and $\nabla_\infty$ are precisely opposite to \citet{rafikov06}, but consistent with other works and the general convention of labeling a quantity $f$ in the disk as $f\di$.}
-We  apply \Eqs{eq:delradan}{eq:radTP} at the radiative-convective boundary (where $\delrad = \delad$) under the assumption that the pressure there is $P\cb \gg P_{\rm d}$, motivated by the fact that the exponential density and hence pressure profile is isothermal (see equation (\ref{eq:rhoiso})).
+\left(T \over T_{\rm d}\right)^{4-\beta} - 1 = {\nabla\di \over \nabla_\infty} \left[\left({P \over P_{\rm d}}\right)^{1+\alpha} - 1 \right] \, ,
+\end{equation} 
+where $\nabla_\infty = (1+\alpha)/(4-\beta)$ is the radiative temperature gradient for $T ,P \rightarrow \infty$.
+We  apply \Eqs{eq:delradan}{eq:radTP} at the radiative-convective boundary (where $\delrad = \delad$) under the assumption that the pressure there $P\cb$ satisfies the relation $P\cb \gg P_{\rm d}$.
+This is motivated by the fact that the density, and hence pressure profile, falls exponentially (see equation \ref{eq:rhoiso}).
 The resulting temperature contrast at the radiative-convective boundary is
@@
 
-The pressure at the convective boundary  follows from \Eqs{eq:delrad}{eq:Tcb} as
-\begin{equation}
+%\footnote{Our definitions of $\nabla\di$ and $\nabla_\infty$ are precisely opposite to \citet{rafikov06}, but consistent with other works and the general convention of labeling a quantity $f$ in the disk as $f\di$.}
+
+The pressure at the convective boundary  follows from \Eqs{eq:radTP}{eq:Tcb} as
+\begin{equation}
+\label{eq:Pcbapprox}
 {P\cb\over P_{\rm d}} \simeq \left({\delad/\nabla\di \over 1 - \delad/\nabla_\infty}\right)^{1 \over 1 + \alpha}
 \end{equation} 
-This pressure contrast can be quite large due to the smallness of $\nabla\di$ in low luminosity atmospheres.
-
- It is also useful to obtain a relation between $T$ and $P$ that eliminates $\nabla\di$ in favor of $P\cb$:
+This pressure contrast can be quite large since $\nabla_{\rm d} \ll 1$.
+
+ We now eliminate $\nabla\di$ from equation (\ref{eq:Pcbapprox}) to obtain a relation between temperature and pressure as a function of the pressure at the radiative-convective boundary $P\cb$.
+From equation (\ref{eq:radTP}), it follows that
  \begin{equation}\label{eq:TP}
-{T \over T_{\rm d}} = \left\{1 + {1 \over {\nabla_\infty \over \delad} - 1} \left[ \left({P \over P\cb}\right)^{1-\alpha} -  \left({P_{\rm d} \over P\cb}\right)^{1-\alpha}  \right] \right\}^{1 \over 4-\beta}\, .
-\end{equation}
-
-We can determine the radius of the convective boundary $R\cb$ from the hydrostatic balance equation as 
+{T \over T_{\rm d}} = \left\{1 + {1 \over {\nabla_\infty \over \delad} - 1} \left[ \left({P \over P\cb}\right)^{1+\alpha} -  \left({P_{\rm d} \over P\cb}\right)^{1+\alpha}  \right] \right\}^{1 \over 4-\beta}\, .
+\end{equation} 
+ We can then determine the radius of the convective boundary $R\cb$ from the hydrostatic balance equation (\ref{eq:struct a}) as 
 \begin{equation}\label{eq:RCBint}
@@
 \end{equation}   
-In the $P\cb \gg P_{\rm d}$ limit the correction term is an order unity constant that depends on $\alpha$, $\beta$ and $\delad$.
+In the $P\cb \gg P_{\rm d}$ limit, the correction term is an order unity constant that depends on $\alpha$, $\beta$ and $\delad$.
 Similarly to the temperature correction factor $\chi$,  $\theta$ accounts for the fact that the radiative region is not perfectly isothermal.
-For $\delad=2/7$, we find $\theta(\alpha=0, \beta=20) \approx 0.556$.
+For $\delad=2/7$, $\alpha=0$ and $\beta=2$, we find $\theta \approx 0.556$.
 Values of $\theta$ for other choices of $\beta$ are summarized in \App{sec:analytic}.
@@
 \subsection{Simplified cooling model}
+\label{coolingan}
 We now use a simplified version of the cooling model introduced in section \ref{cooling} in order to obtain an analytic cooling model.
@@
 \end{equation} 
-where we drop the pressure dependence from the opacity law (\ref{eq:opacitylaw}) and
-\begin{equation} 
-L\di \equiv {64 \pi G M\cb \sigma T_{\rm d}^4 \over 3 \kappa(T_{\rm d}) P_{\rm d}} \nabla_{\rm ad}\chi^{4-\beta}\, 
-\end{equation} 
-normalized to disk conditions, with $T = \chi T_{\rm d}$ at the RCB.
+where we drop the pressure dependence from the opacity law (\ref{eq:opacitylaw}) and define 
+\begin{equation} 
+L\di \equiv {64 \pi G M\cb \sigma T_{\rm d}^4 \over 3 \kappa(T_{\rm d}) P_{\rm d}} \nabla_{\rm ad}\chi^{4-\beta}, 
+\end{equation} 
+with $T = \chi T_{\rm d}$ at the radiative-convective boundary.
 
@@
 By varying $F_\kappa$, dust depletion or enhancement can be considered.
-Grain properties affect both $F_\kappa$ and $\beta$ which generally satisfies  $1/2 \lesssim \beta \lesssim 2$ (aside from discontinuities across sublimation regions, see \citet{semenov03}).
-
-To make further analytic progress, we ignore the self-gravity in the convective zone, holding the mass fixed at $M\co$.
+Grain properties affect both $F_\kappa$ and $\beta$, which generally satisfies  $1/2 \lesssim \beta \lesssim 2$ (aside from discontinuities across sublimation regions, see \citealt{semenov03}).
+
+To make further analytic progress, we ignore the self-gravity in the convective zone, holding the mass fixed at the core mass $M\co$.
 The density profile of an adiabatic atmosphere follows from hydrostatic balance as
 \begin{equation}\label{eq:rhoconv} 
-\rho = \rho\cb \left[ 1 + {R_B' \over r} - {R_B' \over R\cb}  \right]^{1/(\gamma -1)}\, .
+\rho = \rho\cb \left[ 1 + {R_B' \over r} - {R_B' \over R\cb}  \right]^{1/(\gamma -1)},
 \end{equation} 
@@
 \begin{equation}
-R_B' \equiv {G M\co \over C_P T\cb} = {\delad \over \chi} R_B
-\end{equation} 
-to simplify expressions.
+R_B' \equiv {G M\co \over C_{\rm P} T\cb} = {\delad \over \chi} R_B
+\end{equation} 
+to simplify expressions, with $C_{\rm P}$ the specific heat at constant pressure.
 
   Deep in the atmosphere, where $r \ll R\cb \lesssim R_B'$ the density profile is $\rho \propto r^{-1/(\gamma -1)}$.
-Since the energy scales as $\rho r^2 \propto r^{(2\gamma -3)/(\gamma - 1)}$ only polytropes with $\gamma < 3/2$ (i.e.
+Since the energy scales as $\rho r^2 \propto r^{(2\gamma -3)/(\gamma - 1)}$, only polytropes with $\gamma < 3/2$ (i.e.
 $\gamma = 7/5$, but not $\gamma = 5/3$) have the bulk of energy at the bottom of the atmospheres.
@@
 M_{\rm atm} &=& 4 \pi \int_{R\co}^{R\cb} \rho r^2 dr \\
-&=& {5 \pi^2 \over 4} \rho\cb {R_B'}^{5/2} \sqrt{R\cb} % \approx 0.54 \rho\cb {R_B'}^{5/2} \sqrt{r\cb} \\
+&=& {5 \pi^2 \over 4} \rho\cb {R_B'}^{5/2} \sqrt{R\cb}, % \approx 0.54 \rho\cb {R_B'}^{5/2} \sqrt{r\cb} \\
 %&=& 0.54{R_B'^{3} \sqrt{\chi} \over  \sqrt{ \ln[P\cb/(\theta P\di)]}} \, .
 \end{eqnarray}
-in the limit $R\co \ll R\cb \ll R_B'$. The mass in the isothermal region is $M_{\rm iso} \sim 4 \pi \rho\cb r\cb^4/R_B \ll M_{\rm atm}$, and can be neglected.
+in the limit $R\co \ll R\cb \ll R_B'$.
+From equation (\ref{eq:MatmISO}) and replacing $R_{\rm c}$ with $R\cb$, the mass in the isothermal region is given by $M_{\rm iso} \sim 4 \pi \rho\cb r\cb^4/R_B \ll M_{\rm atm}$, and can be neglected.
 
@@
 \begin{equation} \label{eq:crit}
-{M_{\rm atm} \over M\co} \approx {P\cb /P_M \over  \sqrt{\ln[ P\cb/(\theta P_{\rm d})]}}
+{M_{\rm atm} \over M\co} \approx {P\cb /P_M \over  \sqrt{\ln[ P\cb/(\theta P_{\rm d})]}},
 \end{equation} 
@@
 \end{equation} 
-
 For the atmosphere to initiate unstable gas accretion, $M_{\rm atm} = M\co$. We thus require
@@
 is found by numerically solving the above transcendental equation.
-The physical solution has $\xi >1$, but typically order unity.
+The physical solution has $\xi >1$, but is typically of order unity.
 Numerical solutions exist for $\xi \ll 1$, but are unphysical as they imply $P\cb < P_{\rm d}$.
@@
 
-The cooling timescale  for an atmosphere to become self-gravitating is found from \Eqs{eq:tcool}{eq:Pcbc} as
+We define the timescale for runaway gas accretion to initiate, when $M_{\rm{atm}}=M_{\rm c}$, as the \textit{crossover time} $t_{\rm{co}}$. This time is found from \Eqs{eq:tcool}{eq:Pcbc} to be
 \begin{eqnarray} 
 \label{eq:tcoolan}
-t_{\rm cool} &\approx & 2 \times 10^8 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2  \over \left(\mc \over 10 \right)^{5/3} \left(a\over 10\right)^{15 \over 14}} \yr
+t_{\rm co} &\approx & 2 \times 10^8 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2  \over \left(\mc \over 10 \right)^{5/3} \left(a\over 10\right)^{15 \over 14}} \yr
 \end{eqnarray} 
 for $\beta = 2$ and $\mc \equiv M_{\rm c} / M_{\oplus}$.  
-Clearly this timescale is too long, and this is likely due to missing physics of self-gravity and EOS.
-Nevertheless the trends are informative.
-A reduction in opacity naturally gives faster cooling, as long as the optically thick assumption holds.
-Lower disk temperatures also give faster cooling.
-Even though higher temperatures give higher luminosities, they also give higher dust opacities and lower the Bondi radius and gas density.
-On the other hand,  higher temperatures suppress cooling in this model.
-For different $\beta$ values, the temperature dependence is $t_{\rm cool} \propto F_T^{ \beta + 1/2}$.
+This timescale is too long compared to the average lifetime of a protoplanetary disk --- this is likely due to our neglect of self-gravity.
+Nevertheless, the trends are informative.
+A reduction in opacity gives faster cooling, as long as the optically thick assumption holds.
+Lower disk temperatures also decrease the crossover time;  even though higher temperatures result in higher luminosities, they also give higher dust opacities.
+For different $\beta$ values, the temperature dependence is $t_{\rm co} \propto F_T^{ \beta + 1/2}$.
 The cooling timescale only weakly depends on disk mass or pressure, via the  logarithmic factor $\xi$.
@@
 
-We can define a critical more mass as that which gives a self-gravitating atmosphere within a typical gas disk lifetime:
-\begin{equation}
-t_{\rm cool} = 3 \times 10^6 \tau_{\rm cool} ~{\rm yr} \, ,
-\end{equation} 
-with $\tau_{\rm cool}$ a scaling factor.  The resulting critical core mass is
+We can define a \textit{critical more mass} as the core mass for which $M_{\rm{atm}}=M_{\rm c}$ within a typical gas disk lifetime:
+\begin{equation}
+t_{\rm co} = 3 \times 10^6 ~{\rm yr} \,.
+\end{equation} 
+From equation (\ref{eq:tcoolan}), the resulting critical core mass is
 \begin{equation}\label{eq:Mcrit}
@@
 \end{equation} 
-As with the cooling time, the critical mass is too large due to missing physics.
-The scaling with disk properties is similar, in fact all quantities are simply raised to the $3/5$ power.
-The $\xi$ value is changed to match the value for the nominal solution.
+Similarly to the crossover time, the critical mass is too large due to the neglect of self-gravity.  The scaling with disk properties is similar: all quantities are raised to the $3/5$ power.
+
+%The $\xi$ value is changed to match the value for the nominal solution.
 The lower value reflects the fact that an atmosphere around such a massive core does not need to contract as much to be self-gravitating.
 
-The numerical values for a different opacity law do not change the numerical answers above very much.
-That is mostly because we have chosen a disk location where $T_{\rm d}$ is not far from 100 K, where our opacity law is normalized.
-More generally a higher $\beta$ value is more favorable for core accretion at large distances due to the sharper drop in opacity.
-The effects of opacity on the cooling time and critical core mass are discussed in appendix \App{sec:analytic}
+Different values of the power law coefficient $\beta$ in the opacity law do not change the answers above by much.
+This is mostly because the temperature in our fiducial disk (see equation \ref{eq:diskb}) is relatively close to 100 K, at which our opacity law is normalized.
+A higher $\beta$ value is more favorable for core accretion at large distances due to the sharper drop in opacity.
+The effects of opacity on the cooling time and critical core mass are discussed in \App{sec:analytic}.
 
@@
 
-We also find that the atmosphere growth time is very sensitive to the core mass: for the choice of parameters in Figure \ref{fig:tvsM}, the mass increase needed to reduce the crossover time by a factor of two is only half an Earth mass.
+%We also find that the atmosphere growth time is very sensitive to the core mass: for the choice of parameters in Figure \ref{fig:tvsM}, the mass increase needed to reduce the crossover time by a factor of two is only half an Earth mass.
 
@@
 
- In this section we summarize some of the simplifications we have made, and discuss their relative effect on our model.
+ In this section we summarize some of the simplifications we have made, and discuss future work.
 
@@
 
- In this study we explore the formation of giant planets through core accretion under the assumption that the protoplanetary core is already fully formed, and therefore the planetesimal accretion rate is low.
-We defer quantitative estimates of the parameter space under which this is a valid assumption to Piso et al.
+ %In this study we explore the formation of giant planets through core accretion under the assumption that the protoplanetary core is already fully formed and the planetesimal accretion rate is low.
+
+ Standard calculations in which the atmosphere evolution is dominated by planetesimal accretion find that the critical core mass to form a giant planet is of the order of $10 M_{\oplus}$.
+We have showed in Figure \ref{fig:Mcvsa} that our calculations yield lower critical core masses even in the more inner parts of the disk for a gas with standard composition.
+It is therefore easier, to first order, to first form the core, then lower planetesimal accretion and accumulate a massive atmosphere.
+We defer concrete quantitative comparisons between our values and standard results to Piso et al.
 (2013, in prep).
@@
  In this study we have assumed that the luminosity stays constant throughout the radiative region of the atmosphere.
-As a result, the time dependence can be neglected in equation (\ref{eq:structd}) and static solution can be obtained and then connected through a global cooling equations, as described in section \ref{cooling}.
+As a result, the time dependence can be neglected in equation (\ref{eq:structd}) and static solution can be obtained and then connected through a global cooling equation, as described in section \ref{cooling}.
 However, the assumption of constant luminosity breaks down for high atmosphere masses, as described in section \ref{endoftime}; in this case, a time-dependent model is required to describe the atmosphere evolution.
@@
 We showed in section \ref{endoftime} that the quasi-static model breaks down around the time when the atmosphere mass becomes equal to the mass of the core.
-Nevertheless, we can see from Figures 2 and 3 that the time of slowest accretion is safely before the region in which our model breaks down, and that the time after which the atmosphere mass increases by 25 \% is only a factor of 2 lower than the mass doubling time.
+Nevertheless, we can see from Figure \ref{fig:growthtime} that accretion is slowest before the region in which our model breaks down, and hence the time to initiate runaway gas accretion is insensitive to the exact choice of atmosphere mass at which this process starts.
 As a result, we are confident of our results to order of magnitude.
 
- \subsection{Inefficient Convection}
-
-\textbf{Discussion about mixing length theory...}
+% \subsection{Inefficient Convection}
+
+%\textbf{Discussion about mixing length theory...}
 
@@
 We defined the time at which unstable atmosphere collapse commences as $M_{\rm atm}(t)\sim M_{\rm c}$.
-From this we defined as ``critical core mass'' the minimum core mass for a protoplanet to initiate runaway gas accretion during the lifetime of the protoplanetary disk.
+From this we defined as \textit{critical core mass} the minimum core mass for a protoplanet to initiate runaway gas accretion during the lifetime of the protoplanetary disk.
 We studied this minimum mass for a variety of disk conditions, nebular gas compositions and opacities.
-We found that the critical core mass decreases as we move further out in the disk, and is smaller for lower disk temperatures and opacities and for higher mean molecular weight of the gas.
-
- We find that the critical core mass to form a giant planet within the life time of the disk is smaller than the results yielded by studies that assume that the atmosphere evolution is dominated by the luminosity due to planetesimal accretion.
-We have showed that the planetesimal accretion rate needed to grow the core on a typical disk time scale is larger than the expected planetesimal accretion rates at large separations.
+We found that the critical core mass decreases for larger stellocentric distances, and is smaller for lower disk temperatures and opacities and for higher mean molecular weight of the gas.
+
+ We also find that the critical core mass to form a giant planet within the life time of the disk is smaller than the typical 10 $M_{\oplus}$ value yielded by studies that assume that the atmosphere evolution is dominated by the luminosity due to planetesimal accretion.
 As such, it is faster to form a planet by growing the core first in a fast planetesimal accretion regime (e.g., the core forms in the inner disk, then migrates outwards), then significantly reduce planetesimal accretion and allow a massive atmosphere to accumulate.
@@
 
-\begin{figure}[tb]
-\centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cooling_a10_Mc5_rcb.pdf}
-%\vspace{-0.5in}
-\caption{}
-\end{figure}
+%\begin{figure}[tb]
+%\centering
+%\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cooling_a10_Mc5_rcb.pdf}
+%%\vspace{-0.5in}
+%\caption{}
+%\end{figure}
 

### 2013-06-14 tex/draft_paper2/draft_paper2.tex
current paper versions, plus figures for paper II

%\documentstyle[aas2pp4,epsf]{article}
%%%\documentstyle[aaspp4,epsf]{article}
%\documentstyle[12pt,aasms]{article}    % this is for a preprint
%(single-spaced)
%\documentstyle[aaspp4,epsf]{article} % this is for small print
%\documentstyle[12pt, aaspp4]{article}

%\documentstyle[11pt,aaspp]{article}
%\documentclass[12pt, preprint]{aastex}

%\documentclass[manuscript]{aastex}
\documentclass[apj]{emulateapj}

%\documentclass[12pt, preprint,numberedappendix]{emulateapj}
%\documentstyle[12pt,aasms]{article}    % this is for submittal
                                       % (double-spaced)

%\documentstyle[12pt,aasms]{article}   \usepackage{emulateapj5}

\usepackage{graphicx} 
\usepackage{graphics}                       
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{subeqnarray}
%\bibliographystyle{apj}

\newcommand{\delad}{\nabla_{\rm ad}}
\newcommand{\delrad}{\nabla_{\rm rad}}
\newcommand{\emgr}[1]{\emph{ \color{gray} #1}}

\newcommand{\ie}{i.e.\ }
\newcommand{\eg}{e.g.\ }
\newcommand{\p}{\partial}
\newcommand{\xv}{\vc{x}}
\newcommand{\kv}{\vc{k}}
\newcommand{\brak}[1]{\langle #1\rangle}

\newcommand{\gcc}{\;\mathrm{g\; cm^{-3}}}
\newcommand{\gsc}{\;\mathrm{g\; cm^{-2}}}
\newcommand{\cm}{\; {\rm cm}}
\newcommand{\mm}{\; {\rm mm}}
%\newcommand{\ps}{\; {\rm s^{-1}}}
\newcommand{\km}{\; {\rm km}}
%\newcommand{\au}{\; \varpi_{\rm AU}}

\newcommand{\AU}{\; {\rm AU}}
\newcommand{\yr}{\; {\rm yr}}
\def\K{\; {\rm K}}

\newcommand{\vcs}[1]{\mbox{\boldmath{$\scriptstyle{#1}$}}}
\newcommand{\vc}[1]{\mbox{\boldmath{$#1$}}}
\newcommand{\nab}{\vc{\nabla}}
\DeclareMathSymbol{\varOmega}{\mathord}{letters}{"0A}
\DeclareMathSymbol{\varSigma}{\mathord}{letters}{"06}
\DeclareMathSymbol{\varPsi}{\mathord}{letters}{"09}

\newcommand{\Eq}[1]{Equation\,(\ref{#1})}
\newcommand{\Eqs}[2]{Equations (\ref{#1}) and~(\ref{#2})}
\newcommand{\Eqss}[2]{Equations (\ref{#1})--(\ref{#2})}
\newcommand{\Eqsss}[3]{Equations (\ref{#1}), (\ref{#2}) and~(\ref{#3})}
\newcommand{\App}[1]{Appendix~\ref{#1}}
\newcommand{\Sec}[1]{Sect.~\ref{#1}}
\newcommand{\Chap}[1]{Chapter~\ref{#1}}
\newcommand{\Fig}[1]{Fig.~\ref{#1}}
\newcommand{\Figs}[2]{Figs.~\ref{#1} and \ref{#2}}
\newcommand{\Figss}[2]{Figs.~\ref{#1}--\ref{#2}} 
\newcommand{\Tab}[1]{Table \ref{#1}}

\newenvironment{packed_item}{
\begin{itemize}
  \setlength{\itemsep}{1pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}
}{\end{itemize}}

%\newcommand{\delad}{\nabla_{\rm ad}}
%\newcommand{\delrad}{\nabla_{\rm rad}}
\newcommand{\Rg}{\mathcal{R}}
\newcommand{\RB}{R_{\rm B}}
\newcommand{\co}{_{\rm c}}
\newcommand{\di}{_{\rm d}}
\newcommand{\cb}{_{\rm RCB}}
\newcommand{\surf}{_M}
\newcommand{\mc}{m_{\rm c \oplus}}
\newcommand{\mcn}[1] { m_{ \rm c #1 \oplus} }
\newcommand{\MC}{M_{\rm crit}}
\newcommand{\au}{a_\oplus}
\newcommand{\aun}[1]{ a_{#1\oplus} }

\begin{document}
\bibliographystyle{apj}

\shortauthors{Piso, Youdin \& Murray-Clay}

\title{Equation of State Effects on the Minimum Core Mass for Giant Planet Formation}
\author{Ana-Maria A. Piso}
\affil{Harvard-Smithsonian Center for Astrophysics}
\author{Andrew N. Youdin}
\affil{JILA, University of Colorado at Boulder}
\author{Ruth A. Murray-Clay}
\affil{Harvard-Smithsonian Center for Astrophysics}

\begin{abstract}

The core accretion model assumes that giant planets form through gas accretion on to a solid core. The core and the atmosphere initially grow simultaneously through stages of quasistatic equilibrium; once the core becomes massive enough, the atmosphere is no longer in hydrostatic balance and a rapid phase of runaway gas accretion commences. The minimum core mass for which unstable atmosphere collapse occurs is typically called the ``critical core mass''. In standard calculations of the critical core mass, planetesimal accretion dominates the atmosphere evolution, and the energy deposited by incoming planetesimal on to the core is radiated away by the atmosphere. In this study we consider a low planetesimal accretion regime in which the luminosity evolution of the atmosphere is dominated by Kelvin-Helmholtz contraction. We use the atmosphere structure and cooling model developed in Piso \& Youdin (in prep.) to derive the profile and evolution of atmospheres composed of a gas described by a realistic equation of state. We find that the minimum core mass (which we denote as \textit{critical core mass}) to form a giant planet before the dissipation of the protoplanetary disk is substantially increased compared to an ideal gas polytrope when non-ideal effects such as hydrogen dissociation and ionization are taken into account. Moreover, our results yield lower mass cores than corresponding studies for large planetesimal accretion rates. We therefore show that it is easier to form a planet by growing the core first, then accreting a massive gaseous envelope, rather than forming the core and atmosphere simultaneously.

 %The core accretion model proposes that giant planets form by the accretion of gas onto a solid protoplanetary core. Previous studies have found that there exists a ``critical core mass'' past which hydrostatic solutions can no longer be found and unstable atmosphere collapse occurs. In standard calculations of the critical core mass, planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse. In this study we consider the extreme case in which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the planet. We develop a two-layer atmosphere model with an inner convective region and an outer radiative zone that matches onto the protoplanetary disk, and we determine the minimum core mass for a giant planet to form within the typical disk life timescale for a variety of disk conditions, which we denote as  ``critical core mass''.  We find that the absolute minimum core mass required to nucleate atmosphere collapse within the disk lifetime is smaller for planets forming further away from their host stars. Moreover, the critical core mass is strongly dependent on disk temperature, opacity and mean molecular weight of the gas
\end{abstract}

\section{Introduction}
\label{intro}

%\textbf{Explain the context of the physical problem. Briefly describe the results of paper I, and say how that gave us a qualitative understanding of the numbers (critical core mass, cooling time etc.). Say that, however, in reality gas gas is non-ideal, non-polytropic etc., and that these effects need to be taken into account. Then describe what we do in the paper with the EOS. (Maybe also mention what we actually find in the intro? Although maybe too early to say it in the intro...)}

%Current theories of giant planet formation postulate that these planets form either through core accretion (refs), in which solid planetesimals collide and grow into a massive solid core, which then accretes a gaseous envelope, or due to a gravitational instability in the protoplanetary disk that leads to fragmentation of the disk into self-gravitating clumps (refs, inc. \citealt{dangelo11}).  %quote murray-clay, kratter 10, rafikov 05, etc.
%
%Standard core accretion models (refs) assume that the core and the atmosphere grow at the same time, and that planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse, while the heat generated by the gravitational (Kelvin-Helmholtz) contraction of the atmosphere is neglected. These studies consider that the planet atmosphere is in steady state, in which all the luminosity due to planetesimal accretion is radiated away by the envelope, and  find that there exists a minimum (``critical'') core mass past which hydrostatic solutions can no longer be found and unstable atmosphere collapse occurs. 
%
%Forming giant planets at wide separations in the disk poses theoretical challenges. On the one hand, gravitational instability generates objects that are too massive to explain the current observed properties of exoplanets (refs, inc. \citealt{rafikov05}). On the other hand, planetesimal accretion is slow at large distances in the disk, and therefore large cores may not be able to form before the dissipation of the disk (refs). It would therefore be easier if giant planets could form from smaller cores which would need less time to grow.

One of the prevalent theories of giant planet formation is the core accretion model  (\citealt{mizuno78}, \citealt{stevenson82}, \citealt{boden86}, \citealt{wuchterl93}, \citealt{dangelo11}). In this model, a solid core is first formed; the core grows, and once it becomes large enough it can accumulate a massive atmosphere. In the standard core accretion models, the atmosphere is heated due to accretion of planetesimals, and as a result it radiates away energy. The envelope is therefore in a steady state at all times, and the atmosphere mass is a function of the core mass. It is thus found that there exists a minimum core mass past which hydrostatic equilibrium breaks down and rapid, unstable gas accretion commences: this is called the ``critical core mass''.

However, the planetesimal accretion rate is not necessarily constant at a given location in the protoplanetary disk throughout the disk lifetime (e.g., \citealt{ikoma00}). If the planetesimal accretion rate is very low, the atmosphere can no longer gain energy due to accretion of solids, and is instead dominated by gas accretion, contracting on a Kelvin-Helmholtz timescale. In Piso \& Youdin (in prep.) we studied the formation of giant planet atmospheres under the assumption that Kelvin-Helmoholtz gas contraction dominates the luminosity evolution of the atmosphere over planetesimal accretion. We built quasi-static two-layer atmosphere models with an inner convective region and an outer radiative region that matches smoothly onto the protoplanetary disk. We derived a cooling model to connect series of quasti-static atmospheres, and thus obtained an evolutionary history of the envelope. We defined the time at which unstable atmosphere collapse commences as the \texit{crossover time} $t_{\rm{co}}$, at which  $M_{\rm atm}(t_{\rm{co}})\sim M_{\rm c}$. From this we defined as \textit{critical core mass} the minimum core mass for a protoplanet to initiate runaway gas accretion during the lifetime of the protoplanetary disk. We studied this minimum mass for a variety of disk conditions, nebular gas compositions and opacities. We found that the critical core mass decreases for larger stellocentric distances, and is smaller for lower disk temperatures and opacities and for a higher mean molecular weight of the gas.

 %We found that the critical core mass to form a giant planet within the life time of the disk is smaller than the results yielded by studies that assume that the atmosphere evolution is dominated by the luminosity due to planetesimal accretion. We have showed that the planetesimal accretion rate needed to grow the core on a typical disk time scale is larger than the expected planetesimal accretion rates at large separations. As such, it is faster to form a planet by growing the core first in a fast planetesimal accretion regime (e.g., the core forms in the inner disk, then migrates outwards), then significantly reduce planetesimal accretion and allow a massive atmosphere to accumulate.

 Piso \& Youdin (in prep.) assume that the nebular gas can be described by a polytropic equation of state (EOS) corresponding to an ideal diatomic gas: $\delad=2/7$. In reality, however, non-ideal effects, such as gas dissociation and ionization, have to be taken into account. A realistic hydrogen-helium mixture can be described using tabulated equation of state tables. In this study, we use the \cite{saumon95} EOS tables to describe the nebular gas. We generate atmosphere profiles and estimate atmospheric cooling timescales for a variety of disk conditions. From this, we determine the minimum core mass required for runaway gas accretion to commence within the typical life timescale of the protoplanetary disk.  We find that the realistic equation of state yields larger critical core masses compared to the ideal gas polytrope. %We show, however, that our results are still smaller than corresponding studies for large planetesimal accretion rates.

%Youdin \& Piso (2013) showed that giant planets can grow faster from small protoplanetary cores that are fully formed before significant gas accretion occurs. In this scenario, the planetesimal accretion rate is significantly slowed down during the gas contraction phase of the atmosphere. This reduction can arise due to dynamical clearing, or due to the core having formed in the inner parts of the disk and migrated outwards, etc. In this situation, the atmosphere evolution is dominated by the Kelvin-Helmholtz contraction of the envelope. The atmosphere is no longer in a steady state, but rather it accretes gas as it loses energy through radiation.

%In our model we therefore assume that the luminosity evolution of the atmosphere is dominated by gas contraction, while the planetesimal accretion rate is negligible. As a result, the protoplanetary core has a fixed mass. We consider that the atmosphere evolves in time through stages of quasi-static equilibrium. Once the mass of the gaseous envelope becomes comparable to the mass of the solid core, the self-gravity of the atmosphere can no longer balance the pressure gradient and unstable hydrodynamic collapse commences. The time required for the atmosphere to grow to this stage is the characteristic growth time of the atmosphere. For a set of fixed gas and disk conditions, there exists a minimum core mass for which the atmosphere can grow on the time scale described above within the life time of the protoplanetary disk, which we define as the ``critical core mass''.

%We develop a two-layer atmosphere model, with a convective inner region and a radiative outer region that matches smoothly on to the protoplanetary disk, and develop a cooling model that evolves the atmosphere in time. We aim to find the critical core mass for a giant planet to form before the dissipation of the disk.

This paper is organized as follows. In section \ref{sec2} we review the quasi-static and evolution models derived in Piso \& Youdin (in prep.). We discuss the variations in the adiabatic gradient, and hence in the atmosphere structure, caused by a non-ideal equation of state in section \ref{deladtable}, and discuss the implications of this variability on the atmosphere evolution time in section \ref{EOSeffects}. We determine the minimum core mass to form a giant planet during the disk lifetime when the nebular gas is described by a realistic equation of state in section \ref{critical}. In section \ref{acc} we compare our results to similar results obtained in studies that consider planetesimal accretion as the dominant source of energy. Finally, we summarize our findings in section \ref{conclusions}.

% In section \ref{sec2} we describe the assumptions of our atmosphere model, and derive the basic equations that govern the structure and evolution of the atmosphere. In section \ref{analytic}, we present a simplified analytic model that predicts the qualitative behavior of the numerical model. We describe our results in section \ref{KH}, and determined the critical core mass for planet formation during the life time of the protoplanetary disk in section \ref{critical}.  We discuss our results in section \ref{discussion} and summarize our findings in section \ref{conclusions}.

   %We assume that the luminosity is primarily generated in the convective zone of the atmosphere,

%Giant planets play a fundamental role in shaping the orbital structure of planetary systems, and in affecting the delivery of volatiles to terrestrial planets in the

\section{Model Review}
\label{sec2}

%\textbf{Describe the assumptions of the model: disk, BCs, structure equations, cooling model etc., but with less detail than in paper I (and obviously refer to paper I for more details); for the BCs, emphasize that for a given core, the atmosphere profile and evolution are determined by the outer boundary conditions, i.e. Pout, Tout, Rout --- this will be relevant for section 3.2, i.e. outer boundary effects.}

In this section we review the model developed in Piso \& Youdin (in prep.) for the structure and evolution of a planetary atmosphere embedded in a protoplanetary disk. We describe the assumptions of the model and the properties of our assumed protoplanetary disk in section \ref{model}, and we summarize the equations describing the structure and time evolution of a static atmosphere in section \ref{struct}.

\subsection{Assumptions and Disk Model}
\label{model}

We assume that the planet consists of a solid core of fixed mass and a two-layer atmosphere composed of an inner convective zone and an outer radiative zone that matches smoothly on to the disk. The two regions are separated by the Schwarzschild criterion for convective instability. We denote the surface between the two regions as the radiative-convective boundary (RCB), which is henceforth defined by a radius $r=R_{\cb}$. The evolution of the atmosphere is dominated by Kelvin-Helmholtz contraction rather than planetesimal accretion, and the luminosity is assumed to be constant throughout the outer radiative region. The atmosphere is considered to be spherically symmetric and self-gravitating. It consists of a hydrogen-helium mixture, with hydrogen and helium mass fractions of 0.7 and 0.3, respectively. We further assume that the envelope evolves through stages of quasi-static equilibrium.

%The time dependence of the atmosphere structure equations may be neglected or explicitly taken into account. Some previous studies of atmosphere accretion (e.g., \citealt{stevenson82}, \citealt{wuchterl93}, \citealt{rafikov06}) consider static envelopes, in which the luminosity is solely supplied by planetesimal accretion and fully radiated away by the atmosphere. In other studies, the time evolution is explicitly taken into account and full time dependent models are developed (e.g., \citealt{ikoma00}). We follow an intermediate approach and consider quasi-static evolution. Our model for the atmosphere growth time is described in section \ref{cooling}.

As a fiducial disk model, we use the minimum mass, passively irradiated model of  \citet{chiang10}. In this model, the surface density and mid-plane temperature are given by

\begin{subeqnarray}
\label{eq:diskparam}
\Sigma\di&=&2200 F_{\Sigma} a^{-3/2}\,\, \text{g cm}^{-2} \slabel{eq:diska}\\
T\di &=& 120 F_T a^{-3/7} \,\,K, \slabel{eq:diskb}
\end{subeqnarray}

\noindent with $a$ the semi-major axis in AU, and $F_{\Sigma}$ and $F_T$ normalization factors that adjust the disk mass and temperature relative to the minimum mass solar nebula (MMSN). In this study, we assume $F_{\Sigma}=F_T=1$. The resulting mid-plane pressure is given by

\begin{equation}
\label{eq:Pd}
P\di=110 F_{\Sigma} \sqrt{F_T m_*} a^{-45/14} \,\, \text{dyn cm}^{-2}
\end{equation}

\noindent for a mean molecular weight $\mu=2.35$. Here $m_* \equiv M_*/M_{\odot}$, where $M_*$ is the mass of the central star and $M_{\odot}$ is the mass of the Sun. We choose $m_*=1$.

%We further define several characteristic length scales that are important for this problem. \\
%
%\textbf{The Core Radius} represents the physical radius of the protoplanetary core:
%
%\begin{equation}
%\label{eq:rc}
%R_c \equiv \Big(\frac{3 M_c}{4 \pi \rho_c}\Big)^{1/3},
%\end{equation}
%
%\noindent with $M_c$ and $\rho_c$ the core mass and density, respectively. Since we assume that the core mass does not change with time, the core radius is also fixed for a given model. \\
%
%\textbf{The Bondi Radius} represents the distance from the planet at which the thermal energy of the nebular gas is of the order of the gravitational energy binding the gas to the planet. It is defined as
%
%\begin{equation}
%\label{eq:RB}
%R_B \equiv \frac{G M_p}{c_s^2}=\frac{G M_p}{\mathcal{R} T_d},
%\end{equation}
%
%\noindent where $G$ is the gravitational constant, $M_p$ is the total mass of the planet, $c_s$ is the sound speed, and $\mathcal{R}$ is the reduced gas constant: $\mathcal{R}=k_b/(\mu m_p)$, with $k_b$ the Boltzmann constant and $m_p$ the proton mass. \\
%
%
%\textbf{The Hill Radius} represents the distance at which the gravitational attraction of the planet and of the host star are comparable. It is given by
%
%\begin{equation}
%\label{eq:RHill}
%R_H \equiv a \Big(\frac{M_p}{M_{\odot}}\Big)^{1/3}
%\end{equation}
%
%\noindent Outside the Hill radius, the gravity of the core is too weak to affect the nebular gas. Consequently, the planet will be able to accrete only gas that lies within its Hill sphere. \\
%
%\textbf{The Disk Scale Height} is defined by
%
%\begin{equation}
%\label{eq:H}
%H \equiv \frac{c_s^2}{\Omega},
%\end{equation}
%
%\noindent where $\Omega$ is the Keplerian angular velocity. The disk scale height is a measure of the thickness of the protoplanetary disk. Using equation (\ref{eq:diska}), we find that the scale height for our assumed disk model can be expressed as
%
%\begin{equation}
%\label{eq:H2}
%H=0.022 \sqrt{F_T} a^{9/7} \,\, \mathrm{AU}
%\end{equation}
%
%For small planet masses, $R_B<R_H<H$. As the atmosphere becomes more massive, tidal truncation becomes more important than thermal constraints, and $R_H<R_B<H$. Finally, for high mass atmospheres the disk scale height becomes smaller than the Hill radius of the planet, and spherical symmetry breaks, as the planet is now affected by the vertical profile of the disk.
%
%%Some previous studies of atmosphere accretion (e.g., \citealt{stevenson82}, \citealt{wuchterl93}, \citealt{rafikov06}) consider static envelopes, in which the luminosity is solely supplied by planetesimal accretion and fully radiated away by the atmosphere. In other studies, the time evolution is explicitly taken into account and full time dependent models are developed (e.g., \citealt{ikoma00}). We follow an intermediate approach and consider quasi static evolution. Our model for the atmosphere growth time is described in section \ref{...}.

%For the atmosphere structure, we develop a two-layer atmosphere model with the following assumptions:
%
%\begin{enumerate}
%\item The planet consists of a solid core of fixed mass and a two-layer atmosphere composed of an inner convective zone and an outer radiative zone that matches smoothly on to the disk, as mentioned above. The two regions are separated by the Schwarzschild criterion for convective instability (e.g., \citealt{thompson06}) 
%\item The luminosity evolution of the atmosphere is dominated by Kelvin-Helmholtz contraction rather than planetesimal accretion. 
%\item The luminosity is assumed to be constant throughout the outer radiative region.
%\item The envelope evolves through stages of quasi-static equilibrium.
%\end{enumerate}

\subsection{Structure Equations and Cooling Model}
\label{struct}

An atmosphere in hydrostatic equilibrium is described by the following structure equations:

\begin{subeqnarray}
\label{eq:struct}
\frac{dP}{dr}&=&-\frac{G m}{r^2}\rho \slabel{eq:structa} \\
\frac{dm}{dr}&=&4 \pi r^2 \rho\slabel{eq:structb} \\
\frac{dT}{dr}&=&\nabla \frac{T}{P}\frac{dP}{dr}\slabel{eq:structc} \\
\frac{dL}{dr}&=&4 \pi r^2 \rho \epsilon_g\slabel{eq:structd}, 
\end{subeqnarray}

\noindent where $r$ is the radial coordinate, $P$, $T$ and $\rho$ are the gas pressure, temperature and density, respectively, $m$ is the mass enclosed by the radius $r$, $L$ is the luminosity from the surface of radius $r$, and $\epsilon_g \equiv -T \frac{ds}{dt}$ represents the heating per unit mass due to gravitational contraction, with $s$ the specific gas entropy. The temperature gradient $\nabla \equiv \frac{d \ln T}{d \ln P}$ has different expressions depending on the primary means of energy transport throughout the atmosphere. We assume that energy can be transported either through radiation or convection. When the luminosity is carried by radiative diffusion, the temperature gradient is given by

\begin{equation}
\label{eq:delrad}
\nabla = \delrad \equiv \frac{3 \kappa P}{64 \pi G m \sigma T^4} L,
\end{equation}

\noindent where $\sigma$ is the Stefan-Boltzmann constant and $\kappa$ is the opacity. Alternatively, when the energy is transported outwards through convective motions, the temperature gradient becomes

\begin{equation}
\label{eq:delad}
\nabla = \delad \equiv \Big(\frac{d \ln T}{d \ln P}\Big)_{\mathrm{ad}},
\end{equation}
\\

\noindent where $\delad$ is the adiabatic temperature gradient. The process that dominates energy transport throughout the atmosphere is determined by the Schwarzschild criterion (e.g., \citealt{thompson06}): the atmosphere is stable against convection when

\begin{equation}
\label{eq:schwarz}
\nabla < \delad
\end{equation}

\noindent and convectively unstable when the reverse is true. In order for convection to be effective, $\nabla \approx \delad$. Therefore, we find that the temperature gradient is given by $\nabla=\mathrm{min}(\delad, \delrad)$.

In order for equation set (\ref{eq:struct}) to be solvable, it has to be supplemented by an equation of state (EOS) that relates pressure, temperature and density $P=P(\rho, T)$, and an opacity law for $\kappa$. In this study we use the interpolated EOS tables of \citet{saumon95} for a helium fraction $Y=0.3$. More details on the EOS tables and the methodology of combining the separate tables for hydrogen and helium are presented in section \ref{deladtable} and \App{EOStables}.

We assume an opacity power law of the form

\begin{equation}
\label{eq:opacitylaw}
\kappa=\kappa_0 \Big(\frac{P}{P_{\rm{ref}}}\Big)^{\alpha} \Big(\frac{T}{T_{\rm{ref}}}\Big)^{\beta},
\end{equation}

\noindent with $\alpha$, $\beta$, $\kappa_0$ constants, and $T_{\rm{ref}}$ and $P_{\rm{ref}}$ a normalizing temperature and pressure, respectively. To estimate $\alpha$, $\beta$ and $\kappa_0$ we use the \citet{bell94} opacity laws for ice grains: $\alpha =0 $, $\beta=2$ and $\kappa_0=2$. We note that these values are valid only for low disk temperatures: $T_d \lesssim 100 K$. As such, we only consider cool atmospheres that form in the outer part of the protoplanetary disk ($a \geq 5$ AU).

We now discuss our choice of core parameters and boundary conditions. We assume a solid core of fixed mass $M_c$ with a radius $R_{\rm c}=(3 M_{\rm c}/4 \pi \rho_{\rm c})^{1/3}$, where $\rho_{\rm c}$ is the core density. We choose $\rho_{\rm c}=3.2$ g cm$^{-3}$ (e.g., \citealt{pap99}). Furthermore, we assume the atmosphere matches on to the protoplanetary at a distance equal to its Hill radius, the distance at which the gravitational attraction of the planet and of the host star are comparable. Its radius is given by

\begin{equation}
\label{eq:RHill}
R_{\rm H} \equiv a \Big(\frac{M_{\rm p}}{3 M_{\odot}}\Big)^{1/3}
\end{equation}

\noindent Outside the Hill radius, the gravity of the planet is overcome by the tidal gravity from its host star, and hence only gas that lies within the Hill sphere can be gravitationally bound to the planet. We choose the effective outer boundary of the atmosphere to be the surface defined by the Bondi radius. The Bondi radius represents the distance from the planet at which the thermal energy of the nebular gas is of the order of the gravitational energy binding the gas to the planet. It is defined as

\begin{equation}
\label{eq:RB}
R_{\rm B} \equiv \frac{G M_{\rm p}}{c_s^2}=\frac{G M_{\rm p}}{\mathcal{R} T\di},
\end{equation}

\noindent where $G$ is the gravitational constant, $M_{\rm p}$ is the total mass of the planet, $c_s$ is the sound speed, and $\mathcal{R}$ is the reduced gas constant: $\mathcal{R}=k_b/(\mu m_p)$, with $k_b$ the Boltzmann constant and $m_p$ the proton mass. Outside the Bondi sphere, the core gravity is too weak to significantly affect the nebular gas, justifying the choice of Bondi radius as the relevant scale for the atmosphere. We note that the Hill radius is still the correct scale for matching on to the disk, as perturbations of the nebular gas still occur past the Bondi radius due to the gravitational influence of the core. The above choice for the atmosphere boundary is only applicable when the Bondi radius is smaller than the Hill radius. If, on the other hand, $R_{\rm B}>R_{\rm H}$, the atmosphere only extends out to the Hill radius, since material cannot be gravitationally bound to the protoplanet outside the Hill sphere. At the Hill radius, the temperature and pressure are given by the nebular temperature and pressure: $T(R_{\rm H})=T_{\rm d}$ and $P(R_{\rm H})=P_{\rm d}$. For a given core mass, the atmosphere profile and evolution are therefore uniquely determined by the outer boundary conditions.

%\subsection{Standards Methods of Solution}

%Analogously to the stellar evolution case, simply integrating the structure equations (\ref{eq:struct}) from one boundary to the other is not possible, since the boundary conditions are given both at the center and at the surface. In this case, the standard procedures for numerical integration are the shooting method or the Henyey method (\citealt{kippenhahn90}). The shooting method solves the boundary value problem by reducing it to an initial value problem: trial values are chosen for the parameters at one of the boundaries, then the equations are integrated and the resulting values at the other boundary are compared to the actual boundary conditions. The procedure is repeated until convergence is achieved. Alternatively, inward and outward integrations are carried to an intermediate fitting point, where they are fitted smoothly to each other. In the Henyey method, a trial solution for the whole interval is initially guessed, then gradually adjusted through subsequent iterations until the desired level of accuracy is achieved. In this study we use the shooting method --- we integrate inwards from the disk, and match at the core. The detailed numerical procedure is described in section \ref{twolayer}.

Lastly, we review the cooling model developed in Piso \& Youdin (in prep.) used to determine the time evolution of the atmosphere between subsequent static models. A protoplanetary atmosphere embedded in a gas disk satisfies the following cooling equation:

\begin{equation}
\label{eq:coolingglobal}
L=L_c+\Gamma-\dot{E}+e_{\mathrm{acc}}\dot{M}-P_M \frac{\partial V_M}{\partial t}
\end{equation}

This cooling model applies at any radius $R$ where the mass enclosed is $M$, for example the Bondi radius, the Hill radius, or the radius of the radiative-convective boundary. $L$ is the total luminosity, $L_{\rm c}$ is the luminosity from the solid core, and may include planetesimal accretion and radioactive decay. $\Gamma$ is the heat generation rate, $\dot{E}$ is the rate at which total energy (internal and gravitational) is lost, and  $e_{\mathrm{acc}}$ is the specific total energy brought in by mass accreting at the rate $\dot{M}$: $e_{\mathrm{acc}}=u-G M/R$. The last term represents the work done on a surface mass element.

As a consequence of the equations above, both the atmosphere structure and the gas accretion rate are uniquely determined by the current atmosphere mass. As this mass accretion rate is slow compared to the time it takes to relax to this solution, we can make a quasi-static model of the atmosphere growth. By connecting sets of subsequent static atmospheres through the cooling equation (\ref{eq:coolingglobal}) we can obtain an evolutionary atmosphere series.

%\subsection{Quasi-Static Two-Layer Model}
%\label{twolayer}
%
%%The model and equations presented in this section so far can be applied to any protoplanetary atmosphere that satisfies the assumptions listed in section \ref{model}. 
%
%As a consequence of the equations derived in section \ref{cooling}, both the atmosphere structure and the gas accretion rate are uniquely determined by the current atmosphere mass. As this mass accretion rate is slow compared to the time it takes to relax to this solution, we can make a quasi-static model of the atmosphere growth. In this section we therefore describe the procedure to obtain an evolutionary series from the cooling model introduced in \ref{cooling} between consecutive two-layer static atmospheres (cf. \ref{model}).
%
%We use the boundary conditions and opacity laws described in section \ref{struct}. The numerical integration is performed through the shooting method, which solves the boundary value problem by reducing it to an initial value problem: trial values are chosen for the parameters at one of the boundaries, then the equations are integrated and the resulting values at the other boundary are com- pared to the actual boundary conditions. The procedure is repeated until convergence is achieved. We start with a total atmosphere mass $M_i$, where the index $i$ labels each evolutionary stage. For this mass we guess a trial luminosity $L=L_{\mathrm{guess}}$. The assumption of constant luminosity throughout the radiative zone sets the right hand side of equation (\ref{eq:structb}) to zero, i.e. the time dependence is neglected. We integrate the structure equations (\ref{eq:struct}) inwards from $R_{\mathrm{out}}=R_H$, with the boundary conditions $T(R_{\mathrm{out}})=T_d$ and $P(R_{\mathrm{out}})=P_d$. The numerical integration gives a value for the core mass implied by the trial solution $M_c=M_{c, \mathrm{guess}}$. We adjust $L_{\mathrm{guess}}$ until $M_{c, \mathrm{guess}}$ converges to the actual core mass $M_c$. 
%
%The time evolution of the atmosphere is obtained using the global cooling model described in section \ref{cooling}. We neglect the luminosity due to planetesimal accretion and direct heat generation; as such, the first two terms in equation (\ref{eq:coolingglobal}) are set to zero. Equation (\ref{eq:coolingglobal}) is then evaluated at the RCB, since we assume that luminosity is primarily generated in the convective zone. In what follows, $X_{RCB}$ denotes quantity $X$ evaluated at the RCB. The time step $\Delta t_i$ between two consecutive models of masses $M_i$ and $M_{i+1}$ is calculated as follows:
%
%\begin{eqnarray}\nonumber
%\label{eq:dti}
%<L>_i \Delta t_i&=&-\Delta E_{RCB,i} + <e_{\mathrm{acc}, RCB}>_i \Delta M_{RCB, i} \\
%&&  - <P_{RCB}>_i \Delta V_{RCB, M, i} 
%\end{eqnarray}  
%
%\noindent where we denote 
%
%\begin{subeqnarray}
%\Delta X_i&=&X_{i+1}-X_i \\
%<X>_i&=&\frac{X_i+X_{i+1}}{2}
%\end{subeqnarray}
%\\
%\noindent The subscript $M$ in the volume expression above signifies evaluating the change in volume at constant mass. 
%
%By connecting sets of subsequent static atmospheres through the procedure described above we therefore obtain an evolutionary atmosphere series.

%

%\section{Quasi-Static Kelvin-Helmholtz Contraction}
%\label{KH}
%
%\textbf{Plots of radial / P-T profiles, also M-L, M-t profiles; compare the evolutionary profiles with the polytropic case. The text below is copy-pasted from Paper I - things will be similar, but needs a lot of rephrasing. }
%
%In this section we describe our choices for the gas, disk and core parameters, and present some typical atmosphere numerical models. As mentioned in section \ref{model}, we assume that the central star is a Sun-like star with $M_*=M_{\odot}$. We are primarily interested in the outer regions of the protoplanetary disk, so we generate atmosphere models between 10 and 100 AU. As a fiducial case, we assume that the atmosphere is composed of a hydrogen-helium mixture with the helium mass fraction $Y=0.3$., resulting in a mean molecular weight $\mu=2.35$. We do not take into account the dependence of the adiabatic index on the helium fraction, and therefore assume the adiabatic index for a diatomic gas $\delad=2/7$. We further explore how the atmosphere evolution varies with adiabatic index and molecular weight. We generate model atmospheres for monatomic adiabatic index $\delad=2/5$ and for a purely molecular hydrogen composition ($\mu=2$). The comparison between different gas compositions is deferred to section \ref{critcore}.
%
%Figure 1 shows instantaneous atmosphere radial profiles at $a=10$ AU accreting around a fixed core of mass $M_{\rm{c}}=5 M_{\oplus}$ for several total planet masses. In what follows,  if the $R_B<R_H$ hierarchy holds, we define the planet mass as the total mass enclosed within the Bondi radius of the atmosphere, and regard the gas between the Bondi and Hill radii as perturbations due to the gravitational influence of the planet. For illustrative purposes we mark the position of the Bondi radius and of the radiative convective boundary. We notice that the pressure and temperature in the convective regions of the atmospheres scale as $1/r$, which is the expected behavior of an ideal gas adiabat with a polytopic equation of state (e.g., \citealt{rafikov06}). In the outer radiative region of the atmosphere, the pressure has a nearly exponential profile, which is in agreement with the predictions of the analytic model (see section \ref{analytic} and \App{sec:analytic}). In addition, the atmosphere temperature at the radiative-convective boundary $T_{\rm{RCB}}$ only differs from the disk temperature $T_{\rm{d}}$ by an order unity factor, and so the atmosphere is nearly isothermal throughout the radiative zone. This is also in agreement with the analytic predictions. %, and justifies  
%%
%%\begin{figure}[h]
%%\centering
%%\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/PTm_profiles_v2.pdf}
%%%\vspace{-0.5in}
%%\caption{Example pressure, temperature and mass profiles as a function of radius.}
%%\end{figure}
%
%The location of the Bondi radius moves further out as the mass of the atmosphere increases, as the Bondi radius scales with mass. The location of the radiative-convective boundary has a more interesting behavior. As stated in section \ref{intro}, the nebular gas initially settles into equilibrium around the protoplanetary core on a short timescale relative to the thermal timescale, and is isentropic throughout with an entropy equal to that of the disk. In the context of our model, this atmosphere represents the minimum atmosphere mass for which a hydrostatic solution exists. This atmosphere is purely convective. As gas accumulates and the atmosphere starts cooling, an outer radiative region forms. This layer is thin for low mass atmospheres, as can be seen in Fig. 1: for the $5.21 M_{\oplus}$ atmosphere, the ratio between the thickness of the radiative and convective regions is low. An increase in atmosphere mass causes the radiative zone to expand and the convective region to contract. As more and more gas is accumulated, the convective zone eventually starts expanding again; however, the ratio between the radii of the radiative and convective regions continually increases with atmosphere mass, i.e. the radiative regions of these types of atmospheres are getting deeper and deeper.  \textbf{[This needs to be rephrased, somewhat vague and repetitive.]}
%
%We further use the global cooling model developed in section \ref{cooling} to obtain an evolutionary series from static atmosphere profiles. Figure 2 shows the luminosity and cooling time evolution with atmosphere mass. Gas accretion is initially rapid, but slows down significantly after a relatively small mass increase. For comparison, we also plot the analytic results derived in section \ref{analytic}. We see that there is only a very brief agreement between the analytic and full numerical results. At early times, the radiative zone is too shallow and so the approximation that $P_{\rm RCB} \gg P_{\rm d}$ used in section \ref{analytic} breaks down. At late times, the self-gravity of the atmosphere becomes important. %We see that self-gravity starts to matter 
%
%%\begin{figure}[h]
%%\centering
%%\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Lt_profiles_v2.pdf}
%%%\vspace{-0.5in}
%%\caption{Luminosity and time evolution with mass. The analytic result is plotted for comparison.}
%%\end{figure}
%
%%\subsection{The End of Quasi-Static Evolution}
%%\label{endoftime}
%%
%%
%%The right-hand side of equation (\ref{eq:dti}) decreases with increasing atmosphere mass, and eventually becomes negative when the mass of the envelope becomes comparable to the core mass $M_{\rm{atm}} \sim M_{\rm{c}}$. This implies a negative luminosity from the RCB. Physically, the luminosity $L$ represents the total heat flux through the surface defined by the RCB. The heat flux due to convective motions is always directed outwards, and hence it is positive. Moreover, the temperature gradient $dT/dr$ is negative, as the temperature decreases towards the outside. As a result (see equation (\ref{eq:structd})), the heat flux due to radiative diffusion is also positive. This implies that, in the absence of planetesimal accretion and internal heat sources, there is no physical mechanism that can justify a negative luminosity. The quasi-static, constant luminosity approximation breaks down at this point and a more detailed, time-dependent model is necessary.
%
%
%%As can be inferred from Figure 2, the time steps between subsequent evolutionary models become increasingly smaller as the atmosphere grows in mass. Around the time when the atmosphere mass is comparable to the core mass $M_{\mathrm{atm}} \sim M_{\mathrm{c}}$, we find that the time steps become negative.

\section{Adiabatic Gradient for the Tabulated Equation of State}
\label{deladtable}

%\textbf{Explain the effects separately: dissociation vs. spin effects; show plots that explore these effects separately.}

%\subsection{Interpretation of Adiabatic Gradient Table}
%\label{deladinterp}

In this section we explain the differences in the behavior of thermodynamic variables between an ideal gas polytrope and a realistic equation of state by studying the dependence of the adiabatic gradient (defined in equation \ref{eq:delad}) on temperature and pressure. Figure \ref{fig:deladmap} shows a contour plot of the adiabatic gradient as a function of gas temperature and pressure, which was obtained by interpolating and extending the \citet{saumon95} equation of state tables as referenced in section \ref{sec2} and described in \App{EOStables}.

We distinguish three separate temperature regimes:

%In this section we aim to explain how the variable adiabatic index of the hydrogen-helium mixture described by a real equation of state affects the atmosphere evolution when compared to an ideal gas with constant $\delad$. A contour plot of the adiabatic gradient as extrapolated from the \citet{saumon95} EOS tables is shown in Figure \ref{fig:deladmap}. We distinguish three separate regimes:

\begin{enumerate}
\item Intermediate temperature regime (300 K $\lesssim T \lesssim 3000$), where the hydrogen-helium mixture behaves as an ideal gas characterized by a polytropic equation of state.
\item High temperature regime ($T \gtrsim$ 3000 K), where dissociation of molecular hydrogen occurs, followed by ionization of atomic hydrogen.
\item Low temperature regime ($T \lesssim 300$ K), where the temperature becomes low enough for rotational motion to reduce.
\end{enumerate}

We note that helium behaves like an ideal monatomic gas in our regime of interest ($\delad=2/5$). As such, its presence in the atmosphere only causes a small, constant upper shift in the adiabatic gradient of the mixture.

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/EOS/delad_S_mixt.pdf}
%%\vspace{-0.5in}
\caption{Contour plot of the adiabatic gradient $\delad$ for a hydrogen-helium mixture as a function of gas temperature and pressure. The upper right rectangle encloses the region described by the original \citet{saumon95} EOS tables, while the rest of the plot is our extension to lower temperatures and pressures. The black curves represent constant entropy adiabats, with the labels the natural logarithm of the absolute entropy per unit mass. At high temperatures, hydrogen dissociates and ionizes, while at low temperatures the rotational states of the hydrogen molecule are only partially excited and it therefore no longer behaves like an ideal diatomic gas. The discontinuity between the original EOS tables and our extension is due to the fact that we did not take into account dissociation and ionization. However, in our regime of interest (encompassed by the constant entropy curves), the extension matches smoothly with the original tables.}
\label{fig:deladmap}
\end{figure}

In what follows we explain the behavior of the adiabatic gradient in the three temperature regimes separately.

\vspace{0.2in}

\textbf{1. Intermediate T: Ideal Gas}

For temperatures less than $\sim 2000$ K but larger than $\sim 300$ K, the hydrogen molecule is not energetic enough to dissociate and hydrogen behaves as an ideal diatomic gas. We see this in Fig. \ref{fig:deladmap} for 300 K $\lesssim T \lesssim 3000$ K, where the adiabatic gradient is approximately constant. The helium component of the gas causes a slight increase in the adiabatic index: $\delad \approx 0.3$ in this temperature range rather than 2/7 as is the case for a diatomic gas.

\vspace{0.2in}

\textbf{2. High T: Dissociation and Ionization of Hydrogen.}

At low temperatures, hydrogen exists in molecular form, which has a stable configuration. As the temperature becomes higher than $T \sim 2000-3000$ K, the internal energy becomes large enough to break the covalent bond between the atoms, and hydrogen starts dissociating. At temperatures of the order of $10^4$ K, the internal energy becomes large enough to remove electrons from the atoms, and hydrogen ionizes. In stellar and giant planet interiors there is little overlap between the two processes: hydrogen is almost entirely dissociated into atoms by the time ionization becomes important.

For a mixture of molecular and atomic hydrogen, we expect the adiabatic gradient to have an intermediate value between monatomic and diatomic gas, while for a mixture of protons and electrons the adiabatic index is just $2/5$ as for a monatomic ideal gas. However, we notice in Fig. \ref{fig:deladmap} that the adiabatic gradient decreases significantly in the regions where hydrogen is either partially dissociated or partially ionized. We further explain this behavior.

We first discuss ionization. For a mixture of ideal gases, the total internal energy is given by the sum of the internal energies of the individual gases. When a gas is ionized, however, the energy used to ionize the atoms has to also be taken into account. This energy depends on the ionization fraction and can be determined from the Saha equation (see e.g., \citealt{kippenhahn90}). The ionization fraction only depends on gas temperature and density (see \App{deladioniz}), and hence only the equation of state. An expression of the adiabatic gradient as a function of the ionization fraction can be derived. We present this expression in \App{deladioniz}. The adiabatic gradient is $\delad=2/5$ when there is no ionization (i.e., only atomic hydrogen) or when the plasma is fully ionized, but decreases significantly during partial ionization, reaching a minimum when half of the gas is ionized.

This is consistent with the behavior we see in Figure \ref{fig:deladmap}. At constant entropy, we expect an increase in pressure to increase the internal energy of the system, thus causing the temperature to also rise significantly. In the case of partial ionization, however, part of the internal energy is used to remove the electrons from atoms, and therefore there is less energy available to increase the temperature of the system. This behavior of the constant entropy curves is seen in Figure \ref{fig:deladmap}.

The dissociation of molecular hydrogen is dictated by an equation similar to the Saha equation, with the ionization energy replaced by dissociation energy (equal to 4.27 eV for molecular hydrogen, \citealt{mandl89}). The adiabatic gradient therefore has an analogous behavior, consistent with Fig. \ref{fig:deladmap}.

%We now investigate the effect of the low adiabatic index caused by hydrogen dissociation on the luminosity and cooling time evolution of the atmosphere. The left panel of Figure (x) shows the

%The resulting luminosity and time evolution are shown in Figure \ref{fig:Ltplotall}. We use again instantaneous atmosphere profiles to explain the differences. Figure \ref{fig:ETrprofall} shows the instantaneous temperature and energy profiles, as well as the location of the Bondi radius for a total fixed mass $M_{\rm{tot}}=11.8 M_{\oplus}$. We see that the spin effect at the outer boundary dictates the location of the radiative zone, and therefore the luminosity behavior, while dissociation deep in the atmosphere dictates the energy behavior. As compared to the polytrope, the real EOS therefore generates a deeper radiative zone with a lower luminosity, due to the lower adiabatic index in the outer regions, as well as an atmosphere with the bulk of its energy concentrated at the bottom, due to the low $\delad$ caused by dissociation. As shown above for the ideal gas polytropes, and remembering that $\Delta t \sim -\Delta E/L$, we see that both effects result in a longer time for the atmosphere to evolve.

\vspace{0.2in}

\textbf{3. Low T: Hydrogen Rotation and Spin Isomers}

As a diatomic molecule, hydrogen has five degrees of freedom, three associated with translational motion and two associated with rotation. At room temperature, the rotational states are fully excited and the full rotational effects are seen. The excitation temperature for rotation is $\Theta_r \approx 85$ K for the hydrogen molecule (e.g., \citealt{kittel}); as the gas temperature becomes comparable to $\Theta_r$, fewer rotational states are excited and rotation entirely ceases as $T \rightarrow 0$. %Here we explore the quantum mechanical effects associated with rotation. %We present expressions for the partition function associated with rotation and derive relevant thermodynamic variables in \App{EOStables}. For $T \ll \Theta_r$, the partition function is just equal to one, and no rotational states are excited for very low temperatures and hydrogen behaves like a monatomic gas. On the other hand, if $T \gg \Theta_r$, the molecule is energetic enough for all the rotational states to be fully excited.

In this section we discuss the quantum effects of the hydrogen isomeric forms and the way they affect the rotational energy and heat capacity of the hydrogen molecule at low temperatures, and hence the adiabatic gradient.

 Molecular hydrogen occurs in two isomeric forms: orthohydrogen, with parallel proton spins, and parahydrogen, with antiparallel proton spins. The Pauli exclusion principle requires the total wavefunction of two fermions, such as protons, to be antisymmetric. As such, a symmetric spin wavefunction requires an antisymmetric rotational wavefunction, and vice-versa (\citealt{farkas35}). Parahydrogen has an antisymmetric spin wavefunction, which means that it can only occupy symmetric rotational states and hence the angular quantum number $j$ has to be even. By analogy, orthohydrogen must have an antisymmetric rotational wavefunction and can only occupy states with odd $j$. The partition functions for ortho- and parahydrogen are described in \App{EOStables}.
 At equilibrium, the relative abundance of the ortho and para states is given by the ratio of their partition functions. For very low temperatures there is only parahydrogen, as molecules are in the ground state with $j=0$, which corresponds to the para state. As the temperature is increased, parahydrogen starts converting into orthohydrogen, resulting in an ortho-para equilibrium ratio of 3:1 at room temperature.

%\begin{figure}[h]
%\centering
%\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/EOSeffects/ortho_para_ratio.pdf}
%%%\vspace{-0.5in}
%\caption{Relative abundance of ortho- and parahydrogen as a function of temperature.}
%\label{fig:Zrel}
%\end{figure}

The internal energy and specific heat per unit mass associated with rotation for the individual isomers and for the equilibrium mixture can be derived from equations (\ref{eq:Zpara}), (\ref{eq:Zortho}), (\ref{eq:Zrspin}), (\ref{eq:u}) and (\ref{eq:cv}) and plotted in Figure \ref{fig:ucvr}. The para state has no rotational energy for low temperatures, since all the molecules occupy the rotational level with $j=0$. Orthohydrogen, on the other hand, is in the $j=1$ state, and so has an energy given by the energy of its first rotational level. Since all the hydrogen mixtures behave like monatomic gases at low temperatures, their rotational heat capacity is zero in this region. This is consistent with $\delad=2/5$ at low temperatures as seen in Fig. \ref{fig:deladmap}. There are two significant maxima in the heat capacities of parahydrogen and of the mixture. At very low temperatures, the heat capacity of parahydrogen is zero because only the lowest accessible energy level $j=0$ is occupied and a temperature increase does not provide enough energy to populate the next higher level. When the temperature becomes sufficiently high to populate the second lowest level $j=2$, the heat capacity rapidly increases, passes through a maximum and starts to decrease when the second lowest level becomes saturated. The maxima in the ortho-para mixture appears around the time when parahydrogen starts converting into orthohydrogen. The heat capacity of the equilibrium mixture is not a weighted average of the heat capacities of the individual components because it takes into account both the rotational energy uptake of para- and ortho-hydrogen, and also the shift in their equilibrium concentrations with temperature. At $T=0$, only para-hydrogen is present in the equilibrium mixture; as the temperature is increased, the energetically higher-lying ($j=1$) ortho-hydrogen is formed, and the concomitant energy increase is seen as a peak in the heat capacity. As the adiabatic gradient is inversely proportional to the heat capacity, it means that  the former has to first decrease from 2/5 as the temperature increases, reach a minimum around 50 K ($\delad \approx 0.25$ from Fig. \ref{fig:deladmap}), then gradually increase to 2/7 as for a diatomic gas. This behavior is illustrated in Fig. \ref{fig:deladmap}.

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/EOSeffects/ortho_para_energy.pdf}
%%\vspace{-0.5in}
\caption{Internal energy and heat capacity per unit mass for the hydrogen spins isomers and the equilibrium mixture as a function of temperature.}
\label{fig:ucvr}
\end{figure}

%\vspace{0.1in}

\section{Equation of State Effects on Atmosphere Evolution}
\label{EOSeffects}

Variations in the adiabatic gradient have two competing effects on the atmosphere evolution: they affect the luminosity of the envelope and the amount of energy needed to accrete more mass, i.e. $dE/dM$. Together, they result in changes in the growth time of the atmosphere, and therefore in the crossover time and critical core mass. In this section we discuss how the variable adiabatic gradient discussed in section \ref{deladtable} affects the atmosphere evolution when compared to an ideal gas of constant $\delad$. We explore the effect of partial dissociation and hydrogen spin isomers (see section \ref{deladtable}) on the atmosphere luminosity and cooling time evolution. As the adiabatic gradient is variable throughout the atmosphere profiles described in section (\textbf{some previous section}), we first investigate the differences in atmosphere profiles and evolution for ideal gas polytropes with different adiabatic gradients in section \ref{deladpoly}. We then show how the variable adiabatic gradient affects the time evolution of the atmosphere in section \ref{deladeffect}.

\subsection{Ideal Gas Polytropes with Different Adiabatic Gradient}
\label{deladpoly}

In this section we investigate the differences in luminosity and $dE/dM$, and the resulting time evolution, between ideal gas polytropes with different adiabatic gradients: $\delad=2/7$ (diatomic gas) and $\delad=2/5$ (monatomic gas). We assume both gases have the same mean molecular weight. %We use these results to explain the separate effects of dissociation and spin on the time evolution of the atmosphere.

We generate atmosphere profiles for the two different adiabatic indices at $a=10$ AU and for a core mass $M_{\rm c}=10 M_{\oplus}$, and estimate the luminosity and cooling time evolution as described in section \ref{sec2}. The results are shown in Figure \ref{fig:Ltplotpoly}. We find that the polytrope with the lower adiabatic gradient has both a higher luminosity and a longer cooling time. We use instantaneous atmosphere profiles to explain these effects.

We first discuss the effect of the variable adiabatic gradient on luminosity. Figure \ref{fig:ETrplotpoly}, top panel, shows the temperature profile as well as the location of the radiative-convective boundary for the two polytropes at a fixed total mass $M_{\rm{tot}}=11.8 M_{\oplus}$. The polytrope with a larger adiabatic gradient has a more shallow convective zone, and hence a deeper radiative region, since a larger temperature gradient delays the onset of convection. As the luminosity in the radiative region is inversely proportional to the depth of the radiative zone (see equation (\ref{eq:structd})), a deeper radiative region results in a lower luminosity, which explains the results in the top panel of Figure \ref{fig:Ltplotpoly}.

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/EOSeffects/Ltplot_poly.pdf}
%%\vspace{-0.5in}
\caption{Luminosity and time evolution as a function of total mass (core + atmosphere) for polytropes with different adiabatic indices, for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=10 M_{\oplus}$. A larger adiabatic index results in both a lower luminosity and a shorter cooling time.}
\label{fig:Ltplotpoly}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/EOSeffects/TErplot_poly.pdf}
%%\vspace{-0.5in}
\caption{Instantaneous temperature and total energy profiles as a function of radius for polytropes with different adiabatic indices, for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=10 M_{\oplus}$. The total mass (core + atmosphere) is $11.8 M_{\oplus}$. The location of the radiative-convective boundary is marked. A lower adiabatic gradient results in a more shallow radiative region (upper panel), and in the total energy being concentrated at the bottom of the atmosphere (lower panel).}
\label{fig:ETrplotpoly}
\end{figure}

We study the energy behavior for the two polytropes in order to explain the $dE/dM$ effect. We have shown in Piso \& Youdin (in prep.) that polytropes with $\delad=2/7$ have most of the energy concentrated at the bottom of the atmosphere, while polytropes with $\delad=2/5$ have the bulk of the energy towards the outer boundary. The bottom panel of Figure \ref{fig:ETrplotpoly} shows an instantaneous energy profile for the same total mass $M_{\rm{tot}} = 11.8 M_{\oplus}$, confirming this behavior. It takes more energy to bring in gas deep in the atmosphere for an envelope that has the bulk of its energy concentrated towards the bottom,  resulting in a larger $dE/dM$. The energy effect prevails over the luminosity effect, resulting in a longer cooling time for the envelope, as shown in the bottom panel of Figure \ref{fig:Ltplotpoly}.

%From the adiabatic gradient table shown in Figure \ref{fig:deladmap} we find that $\delad$ has the flattest behavior around $T=500$ K. In this region, the gas behaves like an ideal gas with constant polytropic index $\delad \approx 0.3$, with the shift from diatomic gas caused by the helium in the mixture. We generate three sets of atmosphere profiles. The first one corresponds to an ideal gas of constant adiabatic index $\delad=0.3$. The second one is described by the real EOS for temperatures larger than 500 K and by an ideal gas with $\delad=0.3$ for $T<500$ K. Finally, the third profile consists of an ideal gas polytrope with $\delad=0.3$ for $T>500$ K and a real gas in the low temperature regime. We compare the first two profiles to show the dissociation effects, and the first and third profile to show the effects of ortho- and parahydrogen. The resulting time evolution is shown in Figure \ref{fig:Ltplotall}. We see that both dissociation and spin isomers have a comparable effect on the atmosphere growth, and result in slower cooling, and therefore a longer crossover time, when compared to the polytropic ideal gas equation of state. In what follows we explore the two effects separately.

%\subsection{Effect of Adiabatic Gradient on Atmosphere Cooling Time}
%\label{deladeffect}

%In this section we discuss the effect of the variable adiabatic index described in subsection \ref{deladinterp} on the atmosphere cooling time.

%We now explore the separate effects of hydrogen dissociation at high temperatures and spin isomers at low temperatures. From the adiabatic gradient table shown in Figure \ref{fig:deladmap} we find that $\delad$ has the flattest behavior around $T=500$ K. In this region, the gas behaves like an ideal gas with constant polytropic index $\delad \approx 0.3$, with the shift from diatomic gas caused by the helium in the mixture. We generate three sets of atmosphere profiles. The first one corresponds to an ideal gas of constant adiabatic index $\delad=0.3$. The second one is described by the real EOS for temperatures larger than 500 K and by an ideal gas with $\delad=0.3$ for $T<500$ K. Finally, the third profile consists of an ideal gas polytrope with $\delad=0.3$ for $T>500$ K and a real gas in the low temperature regime. We compare the first two profiles to show the dissociation effects, and the first and third profile to show the effects of ortho- and parahydrogen.

%We now explore the effects of

\subsection{Dissociation and Spin Isomers Effects}
\label{deladeffect}

We now discuss the differences in luminosity and $dE/dM$ between an ideal gas with constant adiabatic gradient and atmospheres with variations in $\delad$ as prescribed by the equation of state discussed in section \ref{EOSeffects}. In what follows we describe our choices of equation of state combinations. From the adiabatic gradient table shown in Figure \ref{fig:deladmap} we find that $\delad$ has the flattest behavior around $T=500$ K. In this region, the gas behaves like an ideal gas with constant polytropic index $\delad \approx 0.3$, with the shift from diatomic gas caused by the helium in the mixture. We generate three sets of atmosphere profiles. The first one corresponds to an ideal gas of constant adiabatic index $\delad=0.3$. The second one is described by the real EOS for temperatures larger than 500 K and by an ideal gas with $\delad=0.3$ for $T<500$ K. Finally, the third profile consists of an ideal gas polytrope with $\delad=0.3$ for $T>500$ K and a real gas in the low temperature regime. We compare the first two profiles to show the dissociation effects, and the first and third profile to show the effects of ortho- and parahydrogen. The resulting time evolution is shown in Figure \ref{fig:tplotall}. We see that both dissociation and spin isomers have a comparable effect on the atmosphere growth, and result in slower cooling, and therefore a longer crossover time, when compared to the polytropic ideal gas equation of state. The cooling time is dependent on both the total energy released due to the contraction of the envelope and the luminosity of the atmosphere. In what follows we explore the relative influence of these two factors separately.

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/EOSeffects/tplot.pdf}
%%\vspace{-0.5in}
\caption{Cooling time evolution as a function of total mass (core + atmosphere) for a variety of EOS combinations, for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=10 M_{\oplus}$. The cooling time is larger both due to hydrogen dissociation and spin effects when compared to an ideal gas polytrope.}
\label{fig:tplotall}
\end{figure}

%We now investigate the effect of the low adiabatic index caused by hydrogen dissociation, on the one hand, and spin isomers on the other hand, on the luminosity and cooling time evolution of the atmosphere, in light of the discussion in section \ref{deladpoly}. The left panel of Figure (x) shows the luminosity evolution with mass for the four combinations of equations of state described in section

The right panel of Figure \ref{fig:TLrplot} shows the luminosity evolution with mass for the three combinations of equations of state described above, as well as for the complete real gas EOS. We use again instantaneous atmosphere profiles to explain the differences. The left panel of Figure \ref{fig:TLrplot} shows the instantaneous temperature profile and the location of the radiative-convective boundary for a total fixed mass $M_{\rm{tot}}=11.8 M_{\oplus}$. The real equation of state for low temperatures is characterized by a lower adiabatic index in the outer regions, due to the spin effects, and is therefore dominant in the radiative zone. As a result, it generates a deeper radiative zone with a lower luminosity, which explains the results in the left panel of Figure \ref{fig:TLrplot}. Moreover, since the cooling time is inversely proportional to the luminosity, the spin effect will result in a longer cooling (and crossover) time.

The energy behavior is shown in Figure \ref{fig:Erplot}. The real EOS for high temperatures has a low adiabatic index deep in the atmosphere, due to hydrogen dissociation, and thus the bulk of its energy concentrated at the bottom of the atmosphere, for the reasons described in section \ref{deladpoly}. It takes more energy to add mass more mass deep in the atmosphere, and $dE/dM$ is larger as a result. 
%

We have seen that the spin effect at the outer boundary dictates the location of the radiative zone, and therefore the luminosity behavior, while dissociation deep in the atmosphere dictates the energy behavior. Overall, both effects result in a longer time for the atmosphere to evolve.

%As compared to the polytrope, the real EOS therefore generates a deeper radiative zone with a lower luminosity, due to the lower adiabatic index in the outer regions, as well as an atmosphere with the bulk of its energy concentrated at the bottom, due to the low $\delad$ caused by dissociation. As shown above for the ideal gas polytropes, and remembering that $\Delta t \sim -\Delta E/L$, we see that both effects result in a longer time for the atmosphere to evolve.

\begin{figure*}[tb]
\centering
\includegraphics[width=\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/EOSeffects/TLr_plot.pdf}
%%\vspace{-0.5in}
\caption{Left panel: Instantaneous emperature profile as a function of radius for a variety of EOS combinations, for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=10 M_{\oplus}$. The total mass (core + atmosphere) is $11.8 M_{\oplus}$. The location of the radiative-convective boundary is marked. The effect of hydrogen spin isomers in the outer region of the atmosphere sets the location of the radiative-convective boundary.  Right panel: Luminosity evolution as a function of total mass (core + atmosphere) for a variety of EOS combinations, for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=5 M_{\oplus}$. Dissociation deep in the atmosphere increases the luminosity when compared to an ideal gas polytrope, while the existence of the hydrogen spin isomers in the outer part of the atmosphere results in a lower luminosity. The two effects combined yield a lower luminosity for the realistic equation of state when compared to the polytrope. }
\label{fig:TLrplot}
\end{figure*}

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/EOSeffects/Er_plot.pdf}
%%\vspace{-0.5in}
\caption{Instantaneous energy profiles as a function on radius for a variety of EOS combinations, for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=10 M_{\oplus}$. The total mass (core + atmosphere) is $11.8 M_{\oplus}$. Hydrogen dissociation deep in the atmospheres causes the bulk of the energy to be concentrated at the bottom of the atmosphere.}
\label{fig:Erplot}
\end{figure}

%\subsection{Outer Boundary Effects}

%\textbf{Reemphasize the fact that the atmosphere structure is determined by your outer boundary conditions: T_{out}, P_{out}, R_{out} $\rightarrow$ explore the separate effects of pressure, temperature and a (since the Hill radius is determined by a); show how temperature is the strongest effect. }

\section{Critical Core Mass}
\label{critical}

%\textbf{Define what it is (refer again to paper I also); show the Mcrit vs a for fixed disk life plot, for both real EOS, gamma 7/5 and gamma 5/3; justify the differences in terms of the EOS effects from 4.2; show that using a real EOS makes a significant difference to the results; however, the core masses we get are still doable. Again, a lot of text below will be used but needs reorganizing/rephrasing.}

%\textbf{Define what it is (also refer to paper I) and emphasize how it's different from $M_{crit}$ in standard calculations. Define the crossover mass and crossover time.}

In this section we put together the results obtained in section \ref{EOSeffects} and determine the minimum core mass to initiate runaway gas accretion during the lifetime of the protoplanetary disk assuming that the nebular gas is described by a realistic equation of state as prescribed by the \citet{saumon95} EOS tables (see section \ref{EOSeffects}). As in Piso \& Youdin (in prep.), we define this minimum core mass as the \textit{critical core mass}. Moreover, we define the time elapsed until runaway gas accretion is initiated when $M_{\rm{atm}} \sim M\co$ as the \textit{crossover time}. In this section we first explore the dependence of the crossover time on the core mass for a fixed semi-major axis. We then determine the critical core mass to form a giant planet before the dissipation of the gas in the protoplanetary disk, assuming that the nebular gas is described by a realistic hydrogen-helium mixture, and we compare this with the results from Piso \& Youdin (in prep.) for an ideal diatomic gas.

%\subsection{Crossover Time as a Function of Core Mass}
%\label{tvsM}

%\textbf{t vs. M at fixed distance, similar to the plot from paper I. Compare scalings.}

Figure \ref{fig:tvsMplot} displays the time evolution and the crossover time for core masses between 7 and 16 $M_{\oplus}$ at $a=10$ AU in our fiducial disk. The crossover time is shorter for lower mass cores, consistent with the results of Piso \& Youdin (in prep.).

\begin{figure}[h!]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/t_vs_M_10au.pdf}
%%\vspace{-0.5in}
\caption{Time to grow an atmosphere of mass $M_{\rm{atm}}$ for cores with fixed masses between $7 M_{\oplus}$ and $16 M_{\oplus}$ at $a=10$ AU in our fiducial disk. The circles mark the crossover time where $M_{\rm{atm}} \sim M_{\rm c}$. The numbers are labeling the core mass in Earth masses. A larger core mass results in a shorter crossover time.}
\label{fig:tvsMplot}
\end{figure}

%\subsection{Critical Core Mass}
%\label{Mcrit}

%\textbf{Mcrit vs. a plot, realistic EOS and polytrope. Discuss the larger critical core mass for the real EOS in light of the effects from section 4.}

Figure \ref{fig:Mvsaplot} shows the critical core mass for a massive atmosphere to form during a typical lifetime of a protoplanetary disk $t=3$ Myrs, for a gas described by a realistic equation of state. For comparison, we also plot the results of Piso \& Youdin (in prep.) for an ideal diatomic gas. The use of a realistic equation of state increases the critical core mass by more than a factor of 2. As such, non-ideal effects substantially affect the core mass needed to form a giant planet  before the dissipation of the protoplanetary disk.

\begin{figure}[h!]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/Mc_vs_a_poly_real_paper.pdf}
%%\vspace{-0.5in}
\caption{The minimum core mass for an atmosphere to initiate runaway gas accretion within the lifetime of a typical protoplanetary disk $t \sim 3$ Myrs as a function of semi-major axis, for a realistic hydrogen-helium mixture. The results of Piso \& Youdin (in prep.) for an ideal diatomic gas are plotted for comparison. The realistic equation of state yields core masses larger by more than a factor of 2 when compared to the polytrope.}
\label{fig:Mvsaplot}
\end{figure}

\section{Model Relevance in Planet Formation Theory}
\label{acc}

In this study we have considered atmospheres for which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the atmosphere. This is different from standard calculations, in which the atmosphere is heated by planetesimal accretion. In this section we compare our results for the critical core mass to analogous results from the standard calculations. We discuss the core accretion rates that are necessary for our regime to be valid in section \ref{raf1}. We then compare our results with planetesimal accretion results under similar assumptions in section \ref{raf2}.

 %In this section we investigate the core accretion rates that are necessary for our regime to be valid. We also discuss the conditions under which runaway gas accretion can be initiated due to the Kelvin-Helmholtz contraction of the atmosphere before it becomes critical due to planetesimal accretion.

\subsection{Planetesimal Accretion Rates}
\label{raf1}

We estimate the planetesimal accretion rate consistent with our assumptions that $L_{\mathrm{acc}} \ll L_{\rm{KH}}$. Here $L_{\rm{acc}}$ is the accretion luminosity given by

\begin{equation}
\label{eq:Lacc}
L_{acc}=G \frac{M_{\rm{c}} \dot{M_{\rm{c}}}}{R_{\rm{c}}},
\end{equation}

\noindent where $\dot{M_{\rm{c}}}$ is the planetesimal accretion rate, and $L_{\rm{KH}}$ is the luminosity of the atmosphere due to gas contraction obtained from our static model described in section \ref{sec2}. At the limit, $L_{\rm{acc}}=L_{\rm{KH}}$. For a given atmosphere model we can therefore estimate the maximum planetesimal accretion rate during the gas contraction phase in order for the atmosphere to be dominated by gas contraction. We choose as a fiducial case an atmosphere forming at 40 AU and with a core mass of $10 M_{\oplus}$, described by a realistic equation of state. For this choice of parameters, the atmosphere crossover time is $t \sim$ 2.7 Myrs, which is within the typical life time of a protoplanetary disk. The results are presented in Figure \ref{fig:accrates}.

 \begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/acc_rates_paper.pdf}
%\vspace{-0.5in}
\caption{Various relevant accretion rates. $\dot{M}_{\rm{atm}}$ is the growth rate of the atmosphere as estimated by our model, and $\dot{M}_{\rm{c,KH}}$ is the maximum planetesimal accretion rate during the gas contraction phase in order for our regime to be valid. For comparison, we plot the core accretion rate $\dot{M}_{\rm{c,acc}}$ necessary to grow the core on the same time scale as the atmosphere $\tau \sim 2.7$ Myrs, and a typical planetesimal accretion rate where the random velocity of the planetesimals is given by the Hill velocity due to the core.}
\label{fig:accrates}
\end{figure}

We label the resulting minimum core accretion rate as $\dot{M}_{\rm{c,KH}}$. The atmosphere growth rate $\dot{M}_{\rm{atm}}$ is also plotted for comparison. We therefore see that the core accretion rate has to be $\sim2-3$ orders of magnitude lower than the atmosphere accretion rate for our assumptions to be valid.  If the core had accreted planetesimals at this constant rate since it started forming, then the formation of a core massive enough to attract an atmosphere would not have been possible within typical disk life timescales, which indicates a slowing down of the planetesimal accretion regime. Possible explanations for that include the core having formed in the inner part of the disk and later migrated outwards, or the core having been depleted of planetesimals due to a giant neighbor. We also estimate the core accretion rate

\begin{equation}
\label{eq:Mcdot}
\dot{M}_{\rm{c,acc}}(M_{\rm{c}}) \equiv \frac{M_{\rm{c}}}{\tau}
\end{equation}

\noindent needed for the core to form on the same timescale as our model atmosphere, $\tau=2.7$ Myrs, as well as a typical planetesimal accretion rate, for which the random velocities of the planetesimals are of the order of the Hill velocity around the protoplanetary core \citep{goldreich04}. We denote this latter rate as $\dot{M}_{\rm{c,typical}}$. This is the accretion rate at the boundary between the dispersion dominated and shear dominated regimes. It is easy to see that  $\dot{M}_{\rm{c,typical}}$ is more than one order of magnitude lower than the gas accretion rate of our model atmosphere $\dot{M}_{\rm{atm}}$, and lower than the core accretion rate $\dot{M}_{\rm{c,acc}}$ needed to grow the core and the atmosphere at the same time within the disk life time. As such, the formation of a giant planet by growing the core first, then letting the atmosphere cool is faster than growing the core and the atmosphere at the same time at a steady planetesimal accretion rate.

\subsection{Comparison with Standard Results}
\label{raf2}

Next, we are interested in whether hydrodynamic gas accumulation due to planetesimal accretion can already commence before the atmosphere becomes unstable due to Kelvin-Helmholtz contraction, as our regime is no longer the relevant one under such conditions. A core that forms on the same timescale as our model atmosphere accretes planetesimals at a rate given by equation (\ref{eq:Mcdot}). This accretion rate is dependent on the core mass, which is steadily increasing. We therefore compare the critical core mass due to planetesimal accretion at this rate $M_{\rm{crit,acc}}$ to the critical core mass as defined in our estimates $M_{\rm{c,crit}}$ (see section \ref{critical}). If $M_{\rm{crit,acc}}<M{\rm_{c, crit}}$, then the atmosphere has already initiated unstable gas accretion by the time Kelvin-Helmholtz contraction starts dominating.

%Specifically, we compare the critical core mass due to planetesimal accretion at the rate

%we compare the critical core mass $M_{\rm{c,acc}}$ due to planetesimal accretion, for an accretion rate that satisfies $L_{\rm{acc}}<L_{\rm{KH}}$ to the actual core mass $M_{\rm{c}}$ assumed fixed in our model. If $M_{\rm{c,acc}}<M{\rm{c}}$, then the atmosphere has already initiated unstable gas accretion by the time Kelvin-Helmholtz contraction starts dominating.

In order to estimate the critical core mass due to planetesimal accretion $M_{\rm{crit,acc}}$, we use the results of \citet{rafikov06} for low luminosity atmospheres forming in the outer disk ($>2-5$ AU), consistent with our region of interest. \citet{rafikov06} assumes an ideal gas polytropic equation of state and a lower opacity than the standard ISM opacity that we use in our calculations (see equation \ref{eq:opacitylaw}), due to grain growth. In Piso \& Youdin (in prep.) we showed that a reduction in opacity results in a shorter crossover time and therefore a lower critical core mass. In this section, we calculate the critical core mass for an ideal gas polytrope with the standard ISM opacity reduced by a factor of 100, which is comparable to the opacity law used by \citet{rafikov06}.

By relating his expression for the critical core mass to a given core mass dependent planetesimal accretion rate $\dot{M}(M_{\rm{c}})$, we find the following expression for the critical core mass when accretion luminosity dominates the evolution of the atmosphere:

\begin{equation}
\label{eq:critraf}
M_{\rm{crit, acc}} \sim \Big[\frac{\dot{M}(M_{\rm{c}})}{64 \pi^2 C} \frac{\kappa_0}{\sigma G^3} \frac{1}{R\co M\co^{1/3}} \Big(\frac{k}{\mu}\Big)^4\Big]^{3/5},
\end{equation}

\noindent with all the constants as defined in previous sections, and $C$ a constant depending on the adiabatic gradient and disk properties (see \citealt{rafikov06}, equation B3). We calculate $M_{\rm{crit,acc}}$ for a range of core masses. The result is displayed in Figure \ref{fig:raf2}.

 \begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/Mc_vs_a_poly_raf2_paper.pdf}
%\vspace{-0.5in}
\caption{Comparison between the critical core mass $M_{\rm{crit, acc}}$ due to planetesimal accretion and the assumed fixed core mass when gas contraction dominates, for a growth time of $\tau=2.6$ Myrs. Our results yield lower core masses than in the standard case.}
\label{fig:raf2}
\end{figure}

We see that the critical core mass due to planetesimal accretion is smaller than in the case in which planetesimal accretion dominates the evolution of the atmosphere. This brings us to two conclusions. First, we confirm that planetesimal accretion can be safely ignored in our regime of interest. Secondly, this comparison tells us that it is easier and more efficient to form a planet by growing the core first, then accreting a massive envelope, rather than by growing the core and atmosphere in parallel. Moreover, our result represents a true, absolute minimum on the core mass that is needed to form a giant planet during the lifetime of the protoplanetary disk, as our core no longer grows.

% \textbf{This works for this particular choice of core, disk etc. conditions, and I am expecting it to work for most of our parameter space, but it would probably be good to explore this numerically, at least to order of magnitude, with the estimates we have.}

%This means that for large core masses, planetesimal accretion will lead to runaway gas accretion before gas contraction starts dominating, and hence our model is not applicable in that regime. However, we have found in section \ref{critcore} that unstable atmospheres collapse occurs within the life time of the disk for protoplanetary cores smaller than this.

As a final check, we investigate whether planetesimal accretion during the gas contraction phase at the rate $\dot{M}_{\rm{KH}}$ imposed by the condition that $L_{\rm{acc}}<L_{\rm{KH}}$ can alter the core mass enough to affect the time evolution of the atmosphere. We can quantitatively estimate the increase in core mass as

\begin{equation}
\label{eq:cminc}
\Delta M_{\rm{c}} = \int_0^{t_{\rm{crit}}} \dot{M_{\rm{c}}} dt \approx \sum_i \dot{M_{\rm{c}}}_i \Delta t_i,
\end{equation}

 \noindent where $t_{\rm{crit}}$ is the time elapsed until runaway gas accretion commences, and the accretion rate $ \dot{M_{\rm{c}}}_i $ is given by

 \begin{equation}
 \label{eq:Mdotexp}
 \dot{M_{\rm{c}}}_i =\frac{L_i R\co}{G M_{\rm{c}}} 
 \end{equation}

 \noindent from equation (\ref{eq:Lacc}), with $L_i$ the luminosity of the atmosphere at time $t_i$ in our model. For $M_{\rm{c}}=10 M_{\oplus}$, we find $\Delta M_{\rm{c}} \approx 0.2 M_{\oplus}$. This mass increase is therefore negligible in comparison with the initial core mass. It follows that a significant increase in core mass that could potentially alter the time evolution of the atmosphere would occur on a longer time scale than the mass-doubling time for the unperturbed atmosphere. Therefore, the time evolution of the atmosphere is insensitive to core mass changes at a rate imposed by the assumption that $L_{\rm{acc}}<L_{\rm{KH}}$.

%\section{Model Validity??? \textbf{(for now, in lack of a better title)}}
%
%\textbf{Discuss geometric effects (spherical symmetry in the inner disk), opacity effects (also in the 	inner disk) and planetesimal accretion; from this build the parameter space where the model is 	valid and confirm from the plots in the previous section that this is where we're looking. Part of the text regarding the planetesimal accretion discussion is below, as initially written for Paper I}
%
%\subsection{Geometric Effects}
%\subsection{Opacity Effects}
%
%
%\section{Discussion}
%\label{discussion}

%In this section we discuss the parameter space of validity of our model. We calculate the conditions under which planetesimal accretion can be ignored, and review the effects of some of the approximations that come into our model.
%
%  
 \section{Conclusions}
 \label{conclusions}

 In this paper we have studied giant planet formation under the assumption that the planetesimal accretion rate is negligible and the atmosphere evolution is dominated by gas contraction. We have used the model developed in Piso \& Youdin (in prep.) to build atmosphere profiles assuming that the nebular gas obeys a realistic equation of state that takes into account non-ideal effects. We found that the variations in the adiabatic index due to the realistic equation of state result in a significantly larger crossover time, and therefore critical core mass, when compared to an ideal gas polytrope. While for an ideal diatomic gas the minimum core mass to form a giant planet under the assumptions of our model is lower than the typically quoted value of $10 M_{\oplus}$ (see Piso \& Youdin in prep.), the inclusion of non-ideal effects brings this values back to around $10 M_{\oplus}$,

 We also compared our results to standard studies that assume that the evolution of the gaseous envelope is dominated by planetesimal accretion. We found that that our model yields lower core masses than the standard results. It is therefore easier to form a giant planet by growing the core first, then reducing the planetesimal accretion rate and let the atmosphere evolve on a Kelvin-Helmholtz time scale. Moreover, our results represent a true minimum on the core mass needed to form a giant planet during the typical lifetime of a protoplanetary disk.

% In this paper we have studied the formation of giant planet atmospheres under the assumption that Kelvin-Helmoholtz gas contraction dominates the luminosity evolution of the atmosphere over planetesimal accretion. We built quasi-static two-layer atmosphere models with an inner convective region and an outer radiative region that matches smoothly onto the protoplanetary disk. We derived a cooling model to connect series of quasti-static atmospheres, and thus obtained an evolutionary history of the envelope. We defined the time at which unstable atmosphere collapse commences as $M_{\rm atm}(t)\sim M_{\rm c}$. From this we defined as ``critical core mass'' the minimum core mass for a protoplanet to initiate runaway gas accretion during the lifetime of the protoplanetary disk. We studied this minimum mass for a variety of disk conditions, nebular gas compositions and opacities. We found that the critical core mass decreases as we move further out in the disk, and is smaller for lower disk temperatures and opacities and for higher mean molecular weight of the gas. 
% 
% We find that the critical core mass to form a giant planet within the life time of the disk is smaller than the results yielded by studies that assume that the atmosphere evolution is dominated by the luminosity due to planetesimal accretion. We have showed that the planetesimal accretion rate needed to grow the core on a typical disk time scale is larger than the expected planetesimal accretion rates at large separations. As such, it is faster to form a planet by growing the core first in a fast planetesimal accretion regime (e.g., the core forms in the inner disk, then migrates outwards), then significantly reduce planetesimal accretion and allow a massive atmosphere to accumulate.

%Our study assumes that the protoplanetary core forms first, then it r

%---------------------------------------------------------------------------
\bibliographystyle{apj}
\bibliography{refs}

\appendix
\section{Equation of State Tables}\label{EOStables}

In this section we explain the procedure for extending and interpolating the \cite{saumon95} equation of state tables. The equation of state takes into account non ideal interactions, and includes physical treatments of dissociation and ionization. However, the \cite{saumon95} EOS tables only cover a relatively high range of temperatures and pressures: $2.10 < \log_{10} T(\rm{K})<7.06$ and $4<\log_{10}P$(dyn cm$^{-2})<19$. We consider cold disks, where the temperature and pressure drop to $\sim 20$ K and $\sim 10^{-4}$ dyn cm$^{-2}$, respectively (see equations (\ref{eq:diskb}) and (\ref{eq:Pd})). As such, it is necessary to extend the \cite{saumon95} EOS tables to lower temperature and pressure values.

We choose $\log_{10} T (\rm{K})=1$ and $ \log_{10}P$(dyn cm$^{-2})=-4.4$ as our lower boundaries for temperature and pressure, respectively. Our temperature and pressure grid becomes: $1 < \log_{10} T(\rm{K})<7.06$ and $-4.4<\log_{10}P$(dyn cm$^{-2})<19$. The other thermodynamic variables in the tables are calculated as follows.

\subsection{Hydrogen}

\label{hydrogen}

For a system of particles, the partition function can be written as the product of all partition functions associated with each type of energy that the system can have:

\begin{equation}
\label{eq:z}
Z=Z_t Z_r Z_v Z_e Z_n,
\end{equation}

\noindent where $Z_t$, $Z_r$, $Z_v$, $Z_e$ and $Z_n$ are the partition functions associated with translation, rotation, vibration, electronic excitation and nuclear excitation, respectively. For hydrogen, electronic and nuclear excitation are only significant at temperatures higher than our region of interest ($\theta_e \approx 12000$ K and $\theta_n >> \theta_e$, where $\theta_e$ and $\theta_n$ are the characteristic temperatures for electronic and nuclear excitation, respectively). As such, we will only take into account the translation, rotation and vibration of the hydrogen molecule:

\begin{equation}
\label{eq:zagain}
Z=Z_t Z_r Z_v
\end{equation}

The partition function associated with the motion of the center of mass of the molecule is given by (in the classical limit):

\begin{equation}
\label{eq:Zt}
Z_t=(m/2 \beta \pi \hbar^2)^{3/2} V,
\end{equation}

\noindent where $\beta=1/(k T)$ and $V$ is the volume. The rotational partition function is generally written as:

\begin{equation}
\label{eq:Zr}
Z_r=\sum_0^\infty (2 j+1) \exp{\Big[\frac{-j (j+1)\Theta_r}{T}\Big]},
\end{equation}

\noindent where $\Theta_r$ is the characteristic temperature for rotational motion. In the case of hydrogen, $\Theta_r \approx 85$ K. However, molecular hydrogen occurs in two isomeric forms: orthohydrogen, with the proton spins aligned parallel to each other, and parahydrogen, with the proton spins aligned antiparallel. Parahydrogen can only a have symmetric (even) wave function associated with rotation, while orthohydrogen can only have an antisymmetric (odd) wave function associated with rotation (see section \ref{deladtable} for an explanation why). The rotational partition functions for ortho- and parahydrogen can thus be written as:

\begin{equation}
\label{eq:Zpara}
Z_{\rm{r,para}}=\frac{1}{2}\sum_0^\infty (1+(-1)^j) (2 j +1) \exp\Big[-\frac{j(j+1)\Theta_r}{T}\Big]
\end{equation}
and
\begin{equation}
\label{eq:Zortho}
Z_{\rm{r,ortho}}=\frac{3}{2}\sum_0^\infty (1-(-1)^j) (2 j +1) \exp\Big[-\frac{j(j+1)\Theta_r}{T}\Big]
\end{equation}

The factor of 3 above accounts for the three-fold degeneracy of the ortho state.

 When the two isomers are in equilibrium, the combined partition function is given by the sum of the individual partition functions, $Z_{\rm r}=Z_{\rm{r, ortho}}+Z_{\rm{r,para}}$ and can be written as:

\begin{equation}
\label{eq:Zrspin}
Z_r=\sum_0^\infty (2-(-1)^j) (2j+1) \exp{\Big[\frac{-j (j+1) \Theta_r}{T}\Big]}
\end{equation}

In our range of temperatures of interest, we found that $Z_r$ converges after about 25 terms in the series.

Finally, the partition function for vibrational motion is given by:

\begin{equation}
\label{eq:Zv}
Z_v=[1-\exp{(\theta_v/T)}]^{-1},
\end{equation}

\noindent where $\theta_v$ is the characteristic temperature for vibrational motion, $\theta_v \approx 6140$ K for hydrogen.

If the partition function of a system of $N$ particles is known in terms of $(V, T, N)$, the internal energy and entropy of the system can be determined as follows:

\begin{equation}
\label{eq:U}
U_N=k T^2 \Big(\frac{\partial \log{Z}}{\partial T}\Big)_{V, N}
\end{equation}

\begin{equation}
\label{eq:S}
S_N=k \log{Z} + \frac{U_N}{T}
\end{equation}

The energy, and entropy per mass and specific heat capacity will subsequently be:

\begin{equation}
\label{eq:u}
U=\mathcal{R} T^2 \Big(\frac{\partial \log{Z}}{\partial T}\Big)_{V, N}
\end{equation}

\begin{equation}
\label{eq:s}
S=\mathcal{R} \log{Z} + \frac{U}{T}
\end{equation}

\begin{equation}
\label{eq:cv}
C_v=\Big(\frac{\partial U}{\partial T}\Big)_{V, N}
\end{equation}

Since $Z=Z_t Z_r Z_v$, it is easy to notice that $U=U_t+U_r+U_v$ and $S=S_t+S_r+S_v$, where $U_t$, $U_r$, $U_v$, $S_t$, $S_r$, $S_v$ are the quantities corresponding to the individual translation, rotation and partition functions, respectively.

It can be shown that the entropy per mass due to translational motion can be expressed as:

\begin{equation}
\label{eq:st}
S_t=\mathcal{R} \Big[ \frac{5}{2} \ln{T} - \ln{P} + \ln \Big( \frac{(2 \pi)^{3/2} \mathcal{R}^{5/2} \mu^4}{h^3}\Big) +\frac{5}{2} \Big]
\end{equation}

\noindent with $\mu$ the mean molecular weight. Equation (\ref{eq:st}) is known as the Sackur-Tetrode formula, and it is only applicable to an ideal gas. It can also be easily shown that the internal energy per mass due to translational motion is given by:

\begin{equation}
\label{eq:ut}
U_t=\frac{3}{2} \mathcal{R} T
\end{equation}

Putting all of the above together, we can now evaluate the thermodynamic quantities needed to extend the \cite{saumon95} EOS tables to low temperatures and pressures.

\begin{enumerate}

\item{\textbf{Density.}} In the low temperature, low pressure regime, hydrogen is molecular and behaves like an ideal gas. As such, the density in this region follows the ideal gas law $P=\rho \mathcal{R} T$.
\item{\textbf{Internal energy per mass.}} $U=U_t+U_r+U_v$, where $U_t$ is given by equation (\ref{eq:ut}), and $U_r$, $U_v$ are determined using equations (\ref{eq:u}), (\ref{eq:Zrspin}) and (\ref{eq:Zv}) above.
\item{\textbf{Entropy per unit mass}}. Similarly, $S=S_t+S_r+S_v$, where $S_t$ is given by equation (\ref{eq:st}), and $S_r$, $S_v$ can be determined from equation (\ref{eq:s}) and the calculated expressions for $U_r$ and $U_v$, respectively.
\item{\textbf{Entropy logarithmic derivatives}}. The logarithmic derivatives $S_T$ and $S_P$ are given by:

\begin{equation}
\label{eq:sT}
S_T=\frac{\partial \log{S}}{\partial \log{T}} \Big |_P
\end{equation}

\noindent and

\begin{equation}
\label{eq:sP}
S_T=\frac{\partial \log{S}}{\partial \log{P}} \Big |_T
\end{equation}

We calculate $S_T$ and $S_P$ using the table values for $S$, $T$ and $P$, and a linear central difference formula.

\item{\textbf{Adiabatic gradient $\nabla_{ad}$}}. The adiabatic gradient is defined as:

\begin{equation}
\label{eq:deladSP}
\nabla_{ad}=\frac{\partial \log{T}}{\partial \log{P}} \Big |_S = -\frac{S_P}{S_T}
\end{equation}

We evaluate it from the tabulated values for $S_T$ and $S_P$ determined above. Figure 1 shows a contour plot of the adiabatic index for the extended EOS table, while the black lines represent constant entropy curves. The upper right part of the plot ($\log T>2.1$ and $\log P>4$) is based on the \cite{saumon95} EOS table, while the rest of the plot is our extension. We see that the two tables join smoothly for entropy curves between $8.80<\log{S}$(K g$^{-1})<9.07$.

\end{enumerate}

\begin{figure}[h!]
\centering
\includegraphics[scale=.8]{../../figs/EOS/delad_S_H.pdf}
\caption{Contour plot of the adiabatic gradient $\delad$ for the hydrogen extended table. The black curves represent constant entropy curves.}
\end{figure}

\subsection{Helium}

We extend the helium EOS tables based on a similar procedure. Since helium is primarily neutral and atomic at low temperatures and pressures, we treat it as an ideal monoatomic gas, and subsequently only take into account the translational components of the necessary thermodynamic quantities (see subsection \ref{hydrogen} above for details). The analogous $\nabla_{ad}$ contour plot for helium can be seen in Figure 2. We notice that, in the case of helium, the original and extended table join smoothly for entropy curves between $8.29<\log{S}$(K g$^{-1})<8.77$.

\begin{figure}[h!]
\centering
\includegraphics[scale=.8]{../../figs/EOS/delad_S_He.pdf}
\caption{Contour plot for the adiabatic gradient $\delad$ for the helium extended table. The black curves represent constant entropy curves.}
\end{figure}

\vspace{0.2in}

Lastly, we obtain the equation of state tables for the hydrogen-helium mixture thorough the procedure described in \citet{saumon95}, for a helium mass fraction $Y=0.3$.

%Using equation (\ref{eq:upartition}) we therefore recover the standard result $U_{\rm r}=\mathcal{R} T$ (refs). Furthermore, we know that the internal energy and entropy per unit mass associated with translation are given by $U_{\rm t}=\frac{3}{2} \mathcal{R}$ and $C_{\rm{v,t}}=\frac{3}{2}\mathcal{R}$, respectively, and so we are able to calculate the total internal energy and specific heat of a diatomic molecule as a function of temperature. An example of the variation of heat capacity with temperature is shown in \citet{kittel}, chapter 3.

%The partition function associated with rotation is generally written as:
%
%\begin{equation}
%\label{eq:Zr}
%Z_{\rm r}=\sum_0^\infty (2 j +1) \exp\Big[-\frac{j(j+1)\Theta_r}{T}\Big],
%\end{equation}
%with $j$ the angular momentum quantum number \citep{kittel}. Various thermodynamic quantities can be derived from the partition function. For example, the internal energy per unit mass can be written as:
%
%\begin{equation}
%\label{eq:upartition}
%u_{\rm r}=\mathcal{R}T^2 \frac{\partial \log Z}{\partial T}
%\end{equation}
%
%The specific heat at constant volume then easily follows as
%
%\begin{equation}
%\label{eq:cvpartition}
%c_{\rm{v,r}}=\Big(\frac{\partial u}{\partial T}\Big)_V
%\end{equation}

\section{Adiabatic Gradient during Partial Ionization}\label{deladioniz}

For a partially ionized gas, the total internal energy includes contributions from the individual internal energies of neutral atoms, ions and electrons, as well as from the ionization energy. Specifically, if we denote the internal energies of neutral hydrogen, protons and electrons as $U_{H}$, $U_+$ and $U_e$, respectively, then the total internal energy of the gas is given by:

\begin{equation}
U=U_H+U_+ + U_e + x \chi,
\end{equation}

\noindent where $x$ is the ionization fraction and $\chi$ is the ionization energy (equal to -13.6 eV for hydrogen). The ionization fraction can be determined from the Saha equation (see e.g., \citealt{kippenhahn90}).

\begin{equation}
\label{eq:saha}
\frac{x^2}{1-x} \frac{\rho}{m_H}=\frac{(2 \pi m_e k_B T)^{3/2}}{h^3} e^{-\chi/k_B T},
\end{equation}

\noindent where $m_e$ is the mass of the electron and $h$ is Planck's constant. It can be seen from the Saha equation that the ionization fraction depends only on the gas temperature and density: $x=x(T, \rho)$. As such, all the thermodynamic quantities also depend only on the gas temperature and density, and hence on the equation of state. The adiabatic gradient is given by (see \citealt{kippenhahn90}, chapter 14 for a derivation):

\begin{equation}
\delad=\frac{2+x (1-x) \Phi_H}{5+x (1-x) \Phi_H^2},
\end{equation} 
with $\Phi_H=\frac{5}{2}+\frac{\chi}{k T}$. Figure \ref{fig:deladion} shows the behavior of $\delad$ for partially ionized hydrogen. We recover $\delad=2/5$ for $x=0$ (pure atomic hydrogen) and $x=1$ (fully ionized plasma). The adiabatic gradient decreases significantly for intermediate values of $x$, becoming smaller than 0.1 at its minimum (for $x=0.5$).

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/EOSeffects/delad_ionization.pdf}
%%\vspace{-0.5in}
\caption{Adiabatic gradient as a function of the hydrogen ionization fraction $x$. The adiabatic gradient is $\delad=2/5$ for pure atomic hydrogen ($x=0$) and fully ionized hydrogen ($x=1$), and drops to low values during partial ionization.}
\label{fig:deladion}
\end{figure}

\end{document}

### 2013-06-17 tex/draft/draft2.tex
current paper versions

@@
 
-\newcommand{\Eq}[1]{Equation\,(\ref{#1})}
-\newcommand{\Eqs}[2]{Equations (\ref{#1}) and~(\ref{#2})}
-\newcommand{\Eqss}[2]{Equations (\ref{#1})--(\ref{#2})}
-\newcommand{\Eqsss}[3]{Equations (\ref{#1}), (\ref{#2}) and~(\ref{#3})}
+\newcommand{\Eq}[1]{equation\,(\ref{#1})}
+\newcommand{\Eqs}[2]{equations (\ref{#1}) and~(\ref{#2})}
+\newcommand{\Eqss}[2]{equations (\ref{#1})--(\ref{#2})}
+\newcommand{\Eqsss}[3]{equations (\ref{#1}), (\ref{#2}) and~(\ref{#3})}
 \newcommand{\App}[1]{Appendix~\ref{#1}}
@@
 Standard core accretion models (e.g., \citealt{stevenson82}, \citealt{rafikov06}) assume that the core and the atmosphere grow simultaneously, and that planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse.
-These studies consider that the planet atmosphere is in a steady state at all times, where all the incoming energy from planetesimal accretion is radiated away by the envelope, and  find that there exists a minimum (``critical'') core mass past which hydrostatic solutions can no longer be found and unstable atmosphere collapse occurs.
+These studies consider that the planet atmosphere is in a steady state at all times, where all the incoming energy from planetesimal accretion is radiated away by the envelope.
+The core and the atmosphere grow simultaneously, and the mass of the envelope is a function of the core mass.
+As the atmosphere mass becomes comparable to the mass of the solid core, hydrostatic balance no longer holds and runaway gas accretion commences.
+For a set of disk conditions it is therefore found that there is one well-defined core mass past which hydrostatic solutions no longer exist and unstable atmosphere collapse occurs: this is defined as the ``critical core mass''.
 Standard calculations find the critical core mass to be about $10 M_{\oplus}$ (e.g., \citealt{stevenson82}, \citealt{rafikov06}).
+
+%Standard static calculations of giant planet formation that consider planetesimal accretion as the main source of luminosity (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{rafikov06}) assume that the atmosphere is at all times in a steady state at which the luminosity due to planetesimal accretion is fully radiated away by the atmosphere.
+Therefore, for a given set of disk conditions, there is a maximum mass the core can have before the onset of unstable accretion: this is defined as the critical core mass.
 
@@
 
-The time dependence of the atmosphere structure equations may be neglected or explicitly taken into account.
+The time dependence of the atmosphere structure equations may be taken into account either explicitly or implicitly.
 Some previous studies of atmosphere accretion (e.g., \citealt{stevenson82}, \citealt{wuchterl93}, \citealt{rafikov06}) consider static envelopes, in which the luminosity is solely supplied by planetesimal accretion and fully radiated away by the atmosphere.
@@
 \label{eq:struct}
-\frac{dP}{dr}&=&-\frac{G m}{r^2}\rho \slabel{eq:structa} \\
-\frac{dm}{dr}&=&4 \pi r^2 \rho\slabel{eq:structb} \\
-\frac{dT}{dr}&=&\nabla \frac{T}{P}\frac{dP}{dr}\slabel{eq:structc} \\
-\frac{dL}{dr}&=&4 \pi r^2 \rho \epsilon_g\slabel{eq:structd}, 
+\frac{d P}{d r}&=&-\frac{G m}{r^2}\rho \slabel{eq:structa} \\
+\frac{d m}{d r}&=&4 \pi r^2 \rho\slabel{eq:structb} \\
+\frac{d T}{d r}&=&\nabla \frac{T}{P}\frac{d P}{d r}\slabel{eq:structc} \\
+\frac{d L}{d r}&=&4 \pi r^2 \rho (\epsilon + \epsilon_g)\slabel{eq:structd}, 
 \end{subeqnarray}
 
-\noindent where $r$ is the radial coordinate, $P$, $T$ and $\rho$ are the gas pressure, temperature and density, respectively, $m$ is the mass enclosed by the radius $r$, $L$ is the luminosity from the surface of radius $r$, and $\epsilon_g \equiv -T \frac{ds}{dt}$ represents the heating per unit mass due to gravitational contraction, with $s$ the specific gas entropy.
+\noindent where $r$ is the radial coordinate, $P$, $T$ and $\rho$ are the gas pressure, temperature and density, respectively, $m$ is the mass enclosed by the radius $r$, $L$ is the luminosity from the surface of radius $r$, $\epsilon$ is the rate at which heat is generated due to internal sources, and $\epsilon_g \equiv -T \frac{ds}{dt}$ represents the heating per unit mass due to gravitational contraction, with $s$ the specific gas entropy.
+In our study we assume that there are no internal heat sources, and therefore $\epsilon=0$.
 The temperature gradient $\nabla \equiv \frac{d \ln T}{d \ln P}$ has different expressions depending on the primary means of energy transport throughout the atmosphere.
@@
 \noindent where $\sigma$ is the Stefan-Boltzmann constant and $\kappa$ is the opacity.
-Alternatively, when the energy is transported outwards through convective motions, the temperature gradient becomes
+Equation (\ref{eq:delrad}) is valid when the atmosphere is optically thick, which is true throughout the entire atmosphere for our choice of opacity.
+Alternatively, when energy is transported outwards through convective motions, the temperature gradient becomes
 
@@
 
-\noindent where  $\mathcal{R}$ is the reduced gas constant: $\mathcal{R}=k_b/(\mu m_p)$, with $k_b$ the Boltzmann constant and $m_p$ the proton mass.
+\noindent with  $\mathcal{R}$ the reduced gas constant: $\mathcal{R}=k_b/(\mu m_p)$, $k_b$ the Boltzmann constant and $m_p$ the proton mass.
 The gas constant $\mathcal{R}$ depends on the mean molecular weight of the gas $\mu$, which is determined by the relative abundance of hydrogen and helium in the atmosphere.
-In this study we assume that the helium fraction only affects the molecular weight of the gas, and not the adiabatic index.
-We show in section \ref{critcore} that this assumption does not affect the qualitative behavior of our solutions.
+In this study we assume that the helium fraction only affects the molecular weight of the gas, and not the adiabatic index.% We show in section \ref{critcore} that this assumption does not affect the qualitative behavior of our solutions.
 
@@
 \label{eq:RHill}
-R_{\rm H} \equiv a \Big(\frac{M_{\rm p}}{3 M_{\odot}}\Big)^{1/3}
-\end{equation}
-
-\noindent Outside the Hill radius, the gravity of the planet is too weak to affect the nebular gas. Consequently, the planet will be able to accrete only gas that lies within its Hill sphere. \\
+R_{\rm H} \equiv a \Big(\frac{M_{\rm p}}{3 M_{\odot}}\Big)^{1/3}.
+\end{equation}
+
+\noindent Outside the Hill radius, the gravity of the planet is too weak to affect the nebular gas. Consequently, the planet can only accrete gas that lies within its Hill sphere. \\
 
@@
 \label{eq:diskparam}
-\Sigma\di&=&2200 F_{\Sigma} a^{-3/2}\,\, \text{g cm}^{-2} \slabel{eq:diska}\\
-T\di &=& 120 F_T a^{-3/7} \,\,K, \slabel{eq:diskb}
+\Sigma\di&=&2200 F_{\Sigma} (a/\text{AU})^{-3/2}\,\, \text{g cm}^{-2} \slabel{eq:diska}\\
+T\di &=& 120 F_T (a/\text{AU})^{-3/7} \,\,K, \slabel{eq:diskb}
 \end{subeqnarray}
 
-\noindent with $a$ the semi-major axis in AU, and $F_{\Sigma}$ and $F_T$ normalization factors that adjust the disk mass and temperature relative to the minimum mass solar nebula (MMSN).
+\noindent with $a$ the semi-major axis, and $F_{\Sigma}$ and $F_T$ normalization factors that adjust the disk mass and temperature relative to the minimum mass solar nebula (MMSN).
 In this study, we assume $F_{\Sigma}=F_T=1$.
@@
 \label{eq:Pd}
-P\di=110  F_{\Sigma} \sqrt{F_T m_*} a^{-45/14} \,\, \text{dyn cm}^{-2}
+P\di=110  F_{\Sigma} \sqrt{F_T m_*} (a/\text{AU})^{-45/14} \,\, \text{dyn cm}^{-2}
 \end{equation}
@@
 \begin{equation}
-H=0.022 \sqrt{F_T} a^{9/7}\,\, \text{AU},
+H=0.022 \sqrt{F_T} (a/\text{AU})^{9/7}\,\, \text{AU},
 \end{equation}
@@
 
-\noindent where the total luminosity $L$ is balanced by the rate of heat generation $\Gamma$ (due to nuclear fusion in the case of star) and the rate at which total energy is lost $\dot{E}$.
-The total energy of the sphere is given by the sum of the gravitational and internal energies.
+\noindent where the total luminosity $L$ is balanced by the rate of heat generation $\Gamma$ (due to nuclear fusion in the case of a star) and the rate of change of the total energy of the sphere $\dot{E}$.
+The total energy of the sphere $E$ is given by the sum of the gravitational and internal energies: $E=E_{\rm G}+U$, with
+
+\begin{equation}
+\label{eq:Eg}
+E_{\rm G}=-\int_{M\co}^M \frac{G m}{r} dm,
+\end{equation}
+and
+\begin{equation}
+\label{eq:U}
+U=\int_{M\co}^M u\, dm,
+\end{equation}
+where $u$ is the internal energy per mass.
 
@@
 This cooling model applies at any radius $R$ where the mass enclosed is $M$, for example the Bondi radius or the Hill radius.
-$L$, $\Gamma$ and $\dot{E}$ are the same as in equation (\ref{eq:coolingstar}), and include the atmosphere from the core to the top boundary.
-The gravitational energy $E_G$ of an atmosphere with total mass $M$ is given by
-
-\begin{equation}
-\label{eq:Eg}
-E_G=-\int_{M\co}^M \frac{G m}{r} dm,
-\end{equation}
-
-\noindent while the internal energy is
-
-\begin{equation}
-\label{eq:U}
-U=\int_{M\co}^M u dm,
-\end{equation}
-
-\noindent where $u$ is the internal energy per mass.
-We now explain the additional terms in the cooling equation (\ref{eq:coolingglobal}).
-$L\co$ is the luminosity from the solid core, and may include planetesimal accretion and radioactive decay.
-$e_{\mathrm{acc}}$ is the specific total energy brought in by mass accreting at the rate $\dot{M}$: $e_{\mathrm{acc}}=u-G M/R$.
+The $L$, $\Gamma$ and $\dot{E}$ terms are the same as in equation (\ref{eq:coolingstar}), and include the atmosphere from the core to the top boundary.
+In what follows we explain the additional terms in the cooling equation (\ref{eq:coolingglobal}).
+The first term $L\co$ is the luminosity from the solid core, and may include planetesimal accretion and radioactive decay.
+Since planetesimal accretion is negligible in our model, we set $L\co=0$.
+The fourth term on the right-hand side of equation (\ref{eq:coolingglobal}) accounts for the fact that gas is being brought in from a finite radius $R$ with a specific accretion energy $e_{\mathrm{acc}}$ given by $e_{\mathrm{acc}}=u-G M/R$.
 The last term represents the work done on a mass element as the atmosphere contracts.
-
-The global cooling equation (\ref{eq:coolingglobal}) can be derived starting from the local cooling balance.
-Integrating the local energy equation and (\ref{eq:structd}) from the core to the surface and using equation (\ref{eq:structb}) yields
-
-\begin{subeqnarray}
-\label{eq:coolinglocal}
-L-L_c&=&\int_{M_c}^M \frac{\partial L}{\partial m} dm \\
-&=&\int_{M_c}^M\Big(\epsilon - T \frac{\partial s}{\partial t}\Big) dm \slabel{eq:clb} \\
-&=&\Gamma-\int_{M_c}^M\frac{\partial u}{\partial t}dm + \int_{M_c}^M \frac{P}{\rho^2}\frac{\partial{\rho}}{\partial t} dm \slabel{eq:clc},
-\end{subeqnarray}
-
-\noindent where $\Gamma=\int_{M_c}^M \epsilon dm$ and equation (\ref{eq:clc}) follows from equation (\ref{eq:clb}) through the first law of thermodynamics.
-
-Moreover, a self-gravitating protoplanetary atmosphere has to satisfy virial equilibrium.
+In the partial derivative the enclosed mass variable $m$ is kept fixed, and the derivative is evaluated at the surface with total enclosed mass $M$.
+We can think of this as rewriting the atmosphere structure equations (\ref{eq:struct}) in terms of $m$ rather than $r$, and taking into account that the thermodynamic quantities now also depend on the time $t$ (whereas time dependence is neglected in our static profiles).
+As such, $\partial / \partial t$ is the partial time derivative taken at constant mass.
+
+ %As we now take into account the time-dependence of the structure equations (\ref{eq:struct}),  In the
+
+ %In the cooling model given by equation (\ref{eq:coolingglobal}), the enclosed mass variable $m$ is kept fixed.
+and take into account the time-dependence of the thermodynamic variables, which we neglected when obtaining static atmosphere models from equation (\ref{eq:struct}).
+This is equivalent to rewriting equation set (\ref{eq:struct}) with $m$ as the variable of integration; the partial
+
+%In the partial derivative we are keeping fixed the enclosed mass variable $m$ and we evaluate this term at the surface with total enclosed mass $M$.
+This is equivalent to rewriting the structure equations (\ref{eq:struct}) with $m$ as the variable of integration, and taking into account the time dependence of the thermodynamic variables explicitly.
+The partial derivative $\partial / \partial t$ is therefore taken at fixed mass.
+
+%The partial derivative accounts for the fact that we are calculating this term at a fixed atmosphere mass, rather than total envelope mass that includes the mass accreted through the upper boundary.
+
+The global cooling equation (\ref{eq:coolingglobal}) can be derived from the local cooling equation (\ref{eq:structd}) and the virial theorem, which relates the internal and gravitational energy of the planet as follows:
+\begin{equation}
+\label{eq:virial}
+E_G=-3 \int_{M_c}^M \frac{P}{\rho} dm + 4 \pi (R^3 P_M-R\co^3 P\co),
+\end{equation}
+where $P_M$ is the pressure at radius $R$ where the mass enclosed is $M$ and $P\co$ is the pressure at the core.
+The virial theorem is derived from the structure equations (\ref{eq:struct}).
+We defer the explicit derivation of equation (\ref{eq:coolingglobal}) to \App{sec:virial}.
+
+%The global cooling equation (\ref{eq:coolingglobal}) can be derived starting from the local cooling balance.
+
+%Moreover, a self-gravitating protoplanetary atmosphere has to satisfy virial equilibrium.
 The virial theorem is derived by integrating the equation of hydrostatic balance (\ref{eq:structa}) and has the form
 
-\begin{equation}
-\label{eq:virial}
-E_G=-3 \int_{M_c}^M \frac{P}{\rho} dm + 4 \pi (R^3 P_M-R_c^3 Pc)
-\end{equation}
-
-Similarly to the cooling equation, it applies at any radius $R$ where the total mass enclosed is $M$.
+%\begin{equation}
+%\label{eq:virial}
+%E_G=-3 \int_{M_c}^M \frac{P}{\rho} dm + 4 \pi (R^3 P_M-R_c^3 Pc)
+%\end{equation}
+%
+%Similarly to the cooling equation, it applies at any radius $R$ where the total mass enclosed is $M$.
 For an ideal gas of adiabatic index $\gamma$, equation (\ref{eq:virial}) leads to the following expression for the total energy:
-
-\begin{subeqnarray}
-E&=&(1-\xi)U+4 \pi (R^3 P_M-R_c^3 Pc) \slabel{eq:vira} \\
-&=&\frac{\xi-1}{\xi}E_G+\frac{4 \pi}{\xi} (R^3 P_M-R_c^3 Pc) \slabel{eq:virb}
-\end{subeqnarray}
-
-The global cooling equation (\ref{eq:coolingglobal}) can therefore be derived using equations(\ref{eq:coolinglocal}) and (\ref{eq:virial}). A detailed derivation is deferred to  \App{sec:virial}.
+%
+%\begin{subeqnarray}
+%E&=&(1-\xi)U+4 \pi (R^3 P_M-R_c^3 Pc) \slabel{eq:vira} \\
+%&=&\frac{\xi-1}{\xi}E_G+\frac{4 \pi}{\xi} (R^3 P_M-R_c^3 Pc) \slabel{eq:virb}
+%\end{subeqnarray}
+%
+%The global cooling equation (\ref{eq:coolingglobal}) can therefore be derived using equations(\ref{eq:coolinglocal}) and (\ref{eq:virial}). A detailed derivation is deferred to  \App{sec:virial}.
 
@@
 
-In the next subsections we derive the structure and cooling behavior of such an atmosphere, deferring some of the calculations to \App{sec:analytic}.
+In the next subsections we derive the structure and cooling behavior of a two-layer non self-gravitating atmosphere, deferring some of the calculations to \App{sec:analytic}.
 We estimate the temperature and pressure corrections at the radiative-convective boundary due to the fact that the outer radiative region of the atmosphere is not purely isothermal in section \ref{RCBcorr}.
@@
 We use the boundary conditions and opacity laws described in section \ref{struct}.
-The numerical integration is performed through the shooting method, which solves the boundary value problem by reducing it to an initial value problem: trial values are chosen for the parameters at one of the boundaries, then the equations are integrated and the resulting values at the other boundary are compared to the actual boundary conditions.
+The numerical integration is performed through the shooting method (e.g., \citealt{kippenhahn90}, \citealt{press92}), which solves the boundary value problem by reducing it to an initial value problem: trial values are chosen for the parameters at one of the boundaries, then the equations are integrated and the resulting values at the other boundary are compared to the actual boundary conditions.
 The procedure is repeated until convergence is achieved.
@@
 For this mass we guess a trial luminosity $L=L_{\mathrm{guess}}$.
-The assumption of constant luminosity throughout the radiative zone sets the right hand side of equation (\ref{eq:structb}) to zero, i.e.
+The assumption of constant luminosity throughout the radiative zone sets the right hand side of equation (\ref{eq:structd}) to zero, i.e.
 the time dependence is neglected.
@@
 Equation (\ref{eq:coolingglobal}) is then evaluated at the RCB, since we assume that luminosity is primarily generated in the convective zone.
-In what follows, $X_{RCB}$ denotes quantity $X$ evaluated at the RCB.
+In what follows, $X\cb$ denotes quantity $X$ evaluated at the RCB.
 The time step $\Delta t_i$ between two consecutive models of masses $M_i$ and $M_{i+1}$ is calculated as follows:
@@
 \label{eq:dti}
-<L>_i \Delta t_i&=&-\Delta E_{RCB,i} + <e_{\mathrm{acc}, RCB}>_i \Delta M_{RCB, i} \\
-&&  - <P_{RCB}>_i \Delta V_{RCB, M, i} 
+<L>_i \Delta t_i&=&-\Delta E\cb_{,i} + <e_{\mathrm{acc, RCB}}>_i \Delta M\cb_{, i} \\
+&&  - <P\cb>_i \Delta V\cb_{, M, i} 
 \end{eqnarray}
@@
 We consider planets embedded in a protoplanetary disk around a Sun-like star and with a temperature and pressure profile described by equations (\ref{eq:diskb}) and (\ref{eq:Pd}), respectively.
-We are primarily interested in the outer regions of the disk, so we generate atmosphere models at distances between 10 and 100 AU in our fiducial disk.
+We are primarily interested in the outer regions of the disk, so we generate atmosphere models at distances between 5 and 100 AU in our fiducial disk.
 We emphasize that the location in the disk only affects the temperature and pressure of the nebular gas.
@@
 The circles and triangles mark the locations of the Bondi radius and of the radiative-convective boundary, respectively.
-The solid, dotted and dashed lines correspond to profiles with total mass (core and atmosphere) of $5.12 M_{\oplus}$, $5.85 M_{\oplus}$ and $9.22 M_{\oplus}$, respectively.}
+The solid, dotted and dashed lines correspond to profiles with total mass (core and atmosphere) of $5.12 M_{\oplus}$, $5.85 M_{\oplus}$ and $9.22 M_{\oplus}$, respectively.
+The 5.12 $M_{\oplus}$ and 9.22 $M_{\oplus}$ values represent the minimum and maximum total mass, respectively, that our model allows for this particular choice of disk parameters and core mass.}
 \label{fig:profiles}
@@
 We use the global cooling model developed in section \ref{cooling} to obtain an evolutionary history from series of static atmosphere profiles, using the numerical procedure described in section \ref{twolayer}.
-Figure \ref{fig:Ltplot} shows the luminosity evolution and the elapsed time as a function of atmosphere mass for the same choice of parameters as in Figure \ref{fig:profiles} ($M_{\rm c}=5 M_{\oplus}$).
+Figure \ref{fig:Ltplot} shows the luminosity evolution and the elapsed time as a function of atmosphere mass for the same choice of parameters as in Figure \ref{fig:profiles}, $a=60$ AU and $M_{\rm c}=5 M_{\oplus}$.
 The radiative region of the atmosphere is initially thin, which results in a relatively small number of steps that photons need to diffuse and hence a high luminosity (see equation \ref{eq:Lcb}).
@@
 The analytic result is plotted for comparison.
-The luminosity is initially high, then decreases as the atmosphere grows in mass and the radiative zone becomes deeper.
-Growth is accelerated as a result.}
+The luminosity is initially high, then decreases as the atmosphere grows in mass and the radiative zone becomes deeper.}
 \label{fig:Ltplot}
@@
 We have shown in section \ref{profiles} that growth is continuously accelerated once the self-gravity of the atmosphere becomes important.
-This is due to both the decrease in luminosity and to the fact that less and less binding energy per unit of added mass is released as the envelope grows in mass, i.e.
+This is due to the fact that less and less binding energy per unit of added mass is released as the envelope grows in mass, i.e.
 $-dE/dM$ becomes small.
-The right-hand side of equation (\ref{eq:coolingglobal}) therefore decreases with increasing atmosphere mass; in our calculations, we have found that it eventually becomes negative around the time when $M_{\rm{atm}} \sim M_{\rm{c}}$.
+The right-hand side of the cooling equation (\ref{eq:coolingglobal}) is dominated by the energy loss rate $-\dot{E}$ (as we show in \App{sec:analytic}), and as a result the right-hand side of equation (\ref{eq:coolingglobal}) decreases with increasing atmosphere mass; in our calculations, we have found that it eventually becomes negative around the time when $M_{\rm{atm}} \sim M_{\rm{c}}$.
 This translates into a negative luminosity from the radiative-convective boundary.
@@
 
-Standard static calculations of giant planet formation that consider planetesimal accretion as the main source of luminosity (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{rafikov06}) assume that the atmosphere is at all times in a steady state at which the luminosity due to planetesimal accretion is fully radiated away by the atmosphere.
-The core and the atmosphere grow simultaneously, and the mass of the envelope is a function of the core mass.
-As the atmosphere mass becomes comparable to the mass of the solid core, hydrostatic balance no longer holds and runaway gas accretion commences.
-Therefore, for a given set of disk conditions, there is a maximum mass the core can have before the onset of unstable accretion: this is defined as the critical core mass.
  %a single value for the critical core mass $M_{\rm{crit}}$ is found, when the mass of the gaseous envelope is equal to the mass of the core.
@@
 
+Standard calculations that consider planetesimal accretion as the main source of luminosity (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{rafikov06}) assume a steady state at all times and find that, for a given set of disk conditions, there is a maximum mass the core can have before the onset of unstable accretion: this is defined as the critical core mass (see section \ref{intro}).
 In our model the atmosphere is no longer in steady state because energy balance requires gas accretion, and the atmosphere contracts on a Kelvin-Helmholtz time scale.
@@
 We therefore choose the effective crossover mass, past which hydrostatic balance no longer holds, to be the minimum between $M_{\rm{c}}$ and the $M_{\rm{atm}}<M_{\rm c}$ at which our model becomes unphysical according to equation (\ref{eq:dti}), noting that both of these occur after the slow accretion phase for all our parameter choices.
-We define the \textit{crossover time} $t_{\rm{co}}$ as the time elapsed until the atmosphere mass becomes equal to the effective crossover mass.
+As in section \ref{analytic}, we define the \textit{crossover time} $t_{\rm{co}}$ as the time elapsed until the atmosphere mass becomes equal to the effective crossover mass.
 At $t_{\rm{co}}$, runaway accretion commences.
@@
 Different opacity assumptions will result in a different scaling with temperature of $t_{\rm{co}}$.
-We notice a generally good agreement between our numerical results and the scaled analytic approximation in the left panel of Figure \ref{fig:TPeffects}.
-A factor of $\sim 3$ reduction in the disk temperature results in an decrease of the growth time by a factor of $\sim 30$.
+We note, however, that the power law dependence of the crossover time on disk temperature found in our analytic model breaks down for the numerical model for low $T\di$.
+Nevertheless, the trends predicted by the scaled analytic approximation in the left panel of Figure \ref{fig:TPeffects} are still informative.
+For the full numerical solution, a factor of $\sim 3$ reduction in the disk temperature results in an decrease of the growth time by a factor of $\sim 30$.
 The atmosphere evolution is therefore strongly dependent on the temperature in the protoplanetary disk.
@@
 \caption{The minimum core mass for an atmosphere to initiate runaway gas accretion within the lifetime of a typical protoplanetary disk $t \sim 3$ Myrs as a function of semi-major axis, for different mean molecular weight of the nebular gas.
-The analytic results are plotted for comparison.
+The analytic scaled results are plotted for comparison.
 The critical core mass is lower at larger stellocentric distances.}
@@
  In this study we have assumed that the luminosity stays constant throughout the radiative region of the atmosphere.
-As a result, the time dependence can be neglected in equation (\ref{eq:structd}) and static solution can be obtained and then connected through a global cooling equation, as described in section \ref{cooling}.
+As a result, the time dependence can be neglected in equation (\ref{eq:structd}) and static solutions can be obtained and then connected through a global cooling equation, as described in section \ref{cooling}.
 However, the assumption of constant luminosity breaks down for high atmosphere masses, as described in section \ref{endoftime}; in this case, a time-dependent model is required to describe the atmosphere evolution.
@@
 
- \section{Conclusions}
+ \section{Summary}
  \label{conclusions}
 
- In this paper we have studied the formation of giant planet atmospheres under the assumption that Kelvin-Helmoholtz gas contraction dominates the luminosity evolution of the atmosphere over planetesimal accretion.
-We built quasi-static two-layer atmosphere models with an inner convective region and an outer radiative region that matches smoothly onto the protoplanetary disk.
-We derived a cooling model to connect series of quasti-static atmospheres, and thus obtained an evolutionary history of the envelope.
-We defined the time at which unstable atmosphere collapse commences as $M_{\rm atm}(t)\sim M_{\rm c}$.
+ In this paper we have studied the formation of giant planet atmospheres under the assumption of a low planetesimal regime in which Kelvin-Helmoholtz gas contraction dominates the luminosity evolution of the atmosphere.
+We built quasi-static two-layer atmosphere models with an inner convective region and an outer radiative region that matches smoothly onto the protoplanetary disk at the Hill radius.
+In our static models, we used an ideal gas polytropic equation of state, and assumed a standard ISM power-law opacity.
+We then derived a cooling model to connect series of quasti-static atmospheres, and thus obtained an evolutionary history of the envelope.
+We defined the time $t$ at which unstable atmosphere collapse commences as the \textit{crossover time} $t_{\rm{co}}$, for which $M_{\rm atm}(t_{\rm{co}})\sim M_{\rm c}$.
 From this we defined as \textit{critical core mass} the minimum core mass for a protoplanet to initiate runaway gas accretion during the lifetime of the protoplanetary disk.
 We studied this minimum mass for a variety of disk conditions, nebular gas compositions and opacities.
-We found that the critical core mass decreases for larger stellocentric distances, and is smaller for lower disk temperatures and opacities and for higher mean molecular weight of the gas.
-
- We also find that the critical core mass to form a giant planet within the life time of the disk is smaller than the typical 10 $M_{\oplus}$ value yielded by studies that assume that the atmosphere evolution is dominated by the luminosity due to planetesimal accretion.
+We found that the critical core mass decreases for lower disk temperatures, and hence for larger stellocentric distances.
+Moreover, we found that an opacity reduction also lowers the critical core mass significantly.
+On the other hand, a lower mean molecular weight of the nebular gas results in a larger critical core mass.
+
+ We also find that the critical core mass to form a giant planet within the life time of the disk (of about 3 Myrs) is smaller than the typical 10 $M_{\oplus}$ value yielded by studies that assume that the atmosphere evolution is dominated by the luminosity due to planetesimal accretion.
 As such, it is faster to form a planet by growing the core first in a fast planetesimal accretion regime (e.g., the core forms in the inner disk, then migrates outwards), then significantly reduce planetesimal accretion and allow a massive atmosphere to accumulate.
@@
 We assume that hydrostatic balance holds.
-Integrating the local energy equation (\ref{eq:coolinglocal}) from core to surface gives:
+Integrating the local energy equation (\ref{eq:structd}) from core to surface gives:
 \begin{subeqnarray}
+\label{eq:coolinglocal}
 L - L\co &=& \int_{M\co}^M {\p L \over \p m} dm \\
@@
 \end{subeqnarray} 
-with $\Gamma = \int \epsilon dm$ the integral of the direct heating rate.
+with $\Gamma \equiv \int \epsilon dm$ the rate of intern heat generation.
 
@@
 % or $d/dt$.
-For instance the surface radius, $R$, evolves as  
+For instance, the surface radius $R$ evolves as  
 \begin{equation}\label{eq:Rdot}
@@
 
-To derive the global energy equation we must move the (partial) time derivatives in \Eq{eq:DLc} outside their integrals.  The internal energy integral follows simply from  Leibniz's rule as
+To derive the global energy equation (\ref{eq:coolingglobal}) we must move the (partial) time derivatives in \Eq{eq:DLc} outside their integrals.
+The internal energy integral follows simply from  Leibniz's rule as
 \begin{equation}\label{eq:udot}

### 2013-06-17 tex/draft_paper2/draft_paper2.tex
current paper versions

@@
 
-\title{Equation of State Effects on the Minimum Core Mass for Giant Planet Formation}
+\title{Quantitative Estimates of Minimum Core Masses for Giant Planet Formation}
 \author{Ana-Maria A. Piso}
@@
 The core accretion model assumes that giant planets form through gas accretion on to a solid core.
-The core and the atmosphere initially grow simultaneously through stages of quasistatic equilibrium; once the core becomes massive enough, the atmosphere is no longer in hydrostatic balance and a rapid phase of runaway gas accretion commences.
+The core and the atmosphere initially grow simultaneously through stages of quasistatic equilibrium.
+Once the core becomes massive enough, the atmosphere is no longer in hydrostatic balance and a rapid phase of runaway gas accretion commences.
 The minimum core mass for which unstable atmosphere collapse occurs is typically called the ``critical core mass''.
@@
 In this study we consider a low planetesimal accretion regime in which the luminosity evolution of the atmosphere is dominated by Kelvin-Helmholtz contraction.
-We use the atmosphere structure and cooling model developed in Piso \& Youdin (in prep.) to derive the profile and evolution of atmospheres composed of a gas described by a realistic equation of state.
+We use the atmosphere structure and cooling model developed in Piso \& Youdin (2013) to derive the profile and evolution of atmospheres composed of a gas described by a realistic equation of state.
 We find that the minimum core mass (which we denote as \textit{critical core mass}) to form a giant planet before the dissipation of the protoplanetary disk is substantially increased compared to an ideal gas polytrope when non-ideal effects such as hydrogen dissociation and ionization are taken into account.
@@
 
-One of the prevalent theories of giant planet formation is the core accretion model  (\citealt{mizuno78}, \citealt{stevenson82}, \citealt{boden86}, \citealt{wuchterl93}, \citealt{dangelo11}).
+One of the prevalent theories of giant planet formation is the core accretion model  (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{boden86}, \citealt{wuchterl93}, \citealt{dangelo11}).
 In this model, a solid core is first formed; the core grows, and once it becomes large enough it can accumulate a massive atmosphere.
-In the standard core accretion models, the atmosphere is heated due to accretion of planetesimals, and as a result it radiates away energy.
-The envelope is therefore in a steady state at all times, and the atmosphere mass is a function of the core mass.
-It is thus found that there exists a minimum core mass past which hydrostatic equilibrium breaks down and rapid, unstable gas accretion commences: this is called the ``critical core mass''.
+In standard core accretion models, in order to form a core large enough to attract an atmosphere, a high planetesimal accretion rate is needed, on average.
+The atmosphere is therefore heated due to accretion of planetesimals, and as a result it radiates away energy.
+The envelope is in a steady state at all times, and the atmosphere mass is a function of the core mass, which implies that every core mass maps uniquely to one atmosphere mass, for a given set of disk conditions and a planetesimal accretion rate.
+Moreover, larger cores hold fractionally larger envelopes.
+As a result, the atmosphere grows faster than the core, and eventually the atmosphere mass becomes comparable to the mass of the core.
+At this stage, hydrostatic equilibrium breaks down and a rapid phase of runaway gas accretion commences.
+This core mass $M\co$, for which the atmosphere mass $M_{\rm{atm}}$ satisfies $M_{\rm{atm}}\sim M\co$, is called the ``critical core mass'' in standard studies.
 
@@
 If the planetesimal accretion rate is very low, the atmosphere can no longer gain energy due to accretion of solids, and is instead dominated by gas accretion, contracting on a Kelvin-Helmholtz timescale.
-In Piso \& Youdin (in prep.) we studied the formation of giant planet atmospheres under the assumption that Kelvin-Helmoholtz gas contraction dominates the luminosity evolution of the atmosphere over planetesimal accretion.
+In Piso \& Youdin (2013; hereafter PY13) we studied the formation of giant planet atmospheres under the assumption that Kelvin-Helmoholtz gas contraction dominates the luminosity evolution of the atmosphere over planetesimal accretion.
 We built quasi-static two-layer atmosphere models with an inner convective region and an outer radiative region that matches smoothly onto the protoplanetary disk.
 We derived a cooling model to connect series of quasti-static atmospheres, and thus obtained an evolutionary history of the envelope.
-We defined the time at which unstable atmosphere collapse commences as the \texit{crossover time} $t_{\rm{co}}$, at which  $M_{\rm atm}(t_{\rm{co}})\sim M_{\rm c}$.
+We defined the time $t$ at which unstable atmosphere collapse commences as the \textit{crossover time} $t_{\rm{co}}$, at which  $M_{\rm atm}(t_{\rm{co}})\sim M_{\rm c}$.
 From this we defined as \textit{critical core mass} the minimum core mass for a protoplanet to initiate runaway gas accretion during the lifetime of the protoplanetary disk.
@@
 
- Piso \& Youdin (in prep.) assume that the nebular gas can be described by a polytropic equation of state (EOS) corresponding to an ideal diatomic gas: $\delad=2/7$.
+ PY13 assume that the nebular gas can be described by a polytropic equation of state (EOS) corresponding to an ideal diatomic gas: $\delad=2/7$.
 In reality, however, non-ideal effects, such as gas dissociation and ionization, have to be taken into account.
@@
 In this study, we use the \cite{saumon95} EOS tables to describe the nebular gas.
-We generate atmosphere profiles and estimate atmospheric cooling timescales for a variety of disk conditions.
+We generate atmosphere profiles and estimate the atmospheric crossover time for a variety of disk conditions.
 From this, we determine the minimum core mass required for runaway gas accretion to commence within the typical life timescale of the protoplanetary disk.
@@
 This paper is organized as follows.
-In section \ref{sec2} we review the quasi-static and evolution models derived in Piso \& Youdin (in prep.).
+In section \ref{sec2} we review the quasi-static and evolution models derived in PY13.
 We discuss the variations in the adiabatic gradient, and hence in the atmosphere structure, caused by a non-ideal equation of state in section \ref{deladtable}, and discuss the implications of this variability on the atmosphere evolution time in section \ref{EOSeffects}.
@@
 
-In this section we review the model developed in Piso \& Youdin (in prep.) for the structure and evolution of a planetary atmosphere embedded in a protoplanetary disk.
-We describe the assumptions of the model and the properties of our assumed protoplanetary disk in section \ref{model}, and we summarize the equations describing the structure and time evolution of a static atmosphere in section \ref{struct}.
+In this section we review the model developed in PY13 for the structure and evolution of a planetary atmosphere embedded in a protoplanetary disk.
+We describe the assumptions of the model and the properties of our assumed protoplanetary disk in section \ref{model}, and we summarize the equations describing the structure and time evolution of the atmosphere in section \ref{struct}.
 
@@
 The two regions are separated by the Schwarzschild criterion for convective instability.
-We denote the surface between the two regions as the radiative-convective boundary (RCB), which is henceforth defined by a radius $r=R_{\cb}$.
-The evolution of the atmosphere is dominated by Kelvin-Helmholtz contraction rather than planetesimal accretion, and the luminosity is assumed to be constant throughout the outer radiative region.
-The atmosphere is considered to be spherically symmetric and self-gravitating.
+We denote the surface between the two regions as the radiative-convective boundary (RCB), defined by a radius $r=R_{\cb}$.
+We assume a low planetesimal accretion regime in which the atmosphere evolution is dominated by Kelvin-Helmholtz contraction while planetesimal accretion is neglected.
+Moreover, we assume that the luminosity is constant throughout the outer radiative region.
+The atmosphere is spherically symmetric and self-gravitating.
 It consists of a hydrogen-helium mixture, with hydrogen and helium mass fractions of 0.7 and 0.3, respectively.
@@
 
-As a fiducial disk model, we use the minimum mass, passively irradiated model of  \citet{chiang10}. In this model, the surface density and mid-plane temperature are given by
+The temperature and pressure at the outer boundary of the atmosphere are given by the nebular temperature and pressure.
+As a disk model, we use the minimum mass, passively irradiated model of  \citet{chiang10}.
+The surface density, mid-plane temperature and mid-plane pressure are given by
 
@@
 \label{eq:diskparam}
-\Sigma\di&=&2200 F_{\Sigma} a^{-3/2}\,\, \text{g cm}^{-2} \slabel{eq:diska}\\
-T\di &=& 120 F_T a^{-3/7} \,\,K, \slabel{eq:diskb}
+\Sigma\di&=&2200\, (a/\text{AU})^{-3/2}\,\, \text{g cm}^{-2} \slabel{eq:diska}\\
+T\di &=& 120\, (a/\text{AU})^{-3/7} \,\,K, \slabel{eq:diskb} \\
+P\di&=&110\,  (a/\text{AU})^{-45/14} \,\, \text{dyn cm}^{-2} \slabel{eq:Pd}
 \end{subeqnarray}
 
-\noindent with $a$ the semi-major axis in AU, and $F_{\Sigma}$ and $F_T$ normalization factors that adjust the disk mass and temperature relative to the minimum mass solar nebula (MMSN).
-In this study, we assume $F_{\Sigma}=F_T=1$.
-The resulting mid-plane pressure is given by
-
-\begin{equation}
-\label{eq:Pd}
-P\di=110 F_{\Sigma} \sqrt{F_T m_*} a^{-45/14} \,\, \text{dyn cm}^{-2}
-\end{equation}
-
-\noindent for a mean molecular weight $\mu=2.35$. Here $m_* \equiv M_*/M_{\odot}$, where $M_*$ is the mass of the central star and $M_{\odot}$ is the mass of the Sun. We choose $m_*=1$.
-
-%We further define several characteristic length scales that are important for this problem. \\
-%
-%\textbf{The Core Radius} represents the physical radius of the protoplanetary core:
-%
-%\begin{equation}
-%\label{eq:rc}
-%R_c \equiv \Big(\frac{3 M_c}{4 \pi \rho_c}\Big)^{1/3},
-%\end{equation}
-%
-%\noindent with $M_c$ and $\rho_c$ the core mass and density, respectively. Since we assume that the core mass does not change with time, the core radius is also fixed for a given model. \\
-%
-%\textbf{The Bondi Radius} represents the distance from the planet at which the thermal energy of the nebular gas is of the order of the gravitational energy binding the gas to the planet.
-It is defined as
-%
-%\begin{equation}
-%\label{eq:RB}
-%R_B \equiv \frac{G M_p}{c_s^2}=\frac{G M_p}{\mathcal{R} T_d},
-%\end{equation}
-%
-%\noindent where $G$ is the gravitational constant, $M_p$ is the total mass of the planet, $c_s$ is the sound speed, and $\mathcal{R}$ is the reduced gas constant: $\mathcal{R}=k_b/(\mu m_p)$, with $k_b$ the Boltzmann constant and $m_p$ the proton mass.
-\\
-%
-%
-%\textbf{The Hill Radius} represents the distance at which the gravitational attraction of the planet and of the host star are comparable. It is given by
-%
-%\begin{equation}
-%\label{eq:RHill}
-%R_H \equiv a \Big(\frac{M_p}{M_{\odot}}\Big)^{1/3}
-%\end{equation}
-%
-%\noindent Outside the Hill radius, the gravity of the core is too weak to affect the nebular gas. Consequently, the planet will be able to accrete only gas that lies within its Hill sphere. \\
-%
-%\textbf{The Disk Scale Height} is defined by
-%
-%\begin{equation}
-%\label{eq:H}
-%H \equiv \frac{c_s^2}{\Omega},
-%\end{equation}
-%
-%\noindent where $\Omega$ is the Keplerian angular velocity.
-The disk scale height is a measure of the thickness of the protoplanetary disk.
-Using equation (\ref{eq:diska}), we find that the scale height for our assumed disk model can be expressed as
-%
-%\begin{equation}
-%\label{eq:H2}
-%H=0.022 \sqrt{F_T} a^{9/7} \,\, \mathrm{AU}
-%\end{equation}
-%
-%For small planet masses, $R_B<R_H<H$.
-As the atmosphere becomes more massive, tidal truncation becomes more important than thermal constraints, and $R_H<R_B<H$.
-Finally, for high mass atmospheres the disk scale height becomes smaller than the Hill radius of the planet, and spherical symmetry breaks, as the planet is now affected by the vertical profile of the disk.
-%
-%%Some previous studies of atmosphere accretion (e.g., \citealt{stevenson82}, \citealt{wuchterl93}, \citealt{rafikov06}) consider static envelopes, in which the luminosity is solely supplied by planetesimal accretion and fully radiated away by the atmosphere.
-In other studies, the time evolution is explicitly taken into account and full time dependent models are developed (e.g., \citealt{ikoma00}).
-We follow an intermediate approach and consider quasi static evolution.
-Our model for the atmosphere growth time is described in section \ref{...}.
-
-%For the atmosphere structure, we develop a two-layer atmosphere model with the following assumptions:
-%
-%\begin{enumerate}
-%\item The planet consists of a solid core of fixed mass and a two-layer atmosphere composed of an inner convective zone and an outer radiative zone that matches smoothly on to the disk, as mentioned above.
-The two regions are separated by the Schwarzschild criterion for convective instability (e.g., \citealt{thompson06})
-%\item The luminosity evolution of the atmosphere is dominated by Kelvin-Helmholtz contraction rather than planetesimal accretion. 
-%\item The luminosity is assumed to be constant throughout the outer radiative region.
-%\item The envelope evolves through stages of quasi-static equilibrium.
-%\end{enumerate}
+\noindent with $a$ the semi-major axis and for a mean molecular weight $\mu=2.35$.
 
@@
 
-An atmosphere in hydrostatic equilibrium is described by the following structure equations:
+The structure of a static atmosphere is described by the standard equations of hydrostatic balance and thermal equilibrium:
 
@@
 \label{eq:struct}
-\frac{dP}{dr}&=&-\frac{G m}{r^2}\rho \slabel{eq:structa} \\
-\frac{dm}{dr}&=&4 \pi r^2 \rho\slabel{eq:structb} \\
-\frac{dT}{dr}&=&\nabla \frac{T}{P}\frac{dP}{dr}\slabel{eq:structc} \\
-\frac{dL}{dr}&=&4 \pi r^2 \rho \epsilon_g\slabel{eq:structd}, 
+\frac{d P}{d r}&=&-\frac{G m}{r^2}\rho \slabel{eq:structa} \\
+\frac{d m}{d r}&=&4 \pi r^2 \rho\slabel{eq:structb} \\
+\frac{d T}{d r}&=&\nabla \frac{T}{P}\frac{d P}{d r}\slabel{eq:structc} \\
+\frac{d L}{d r}&=&4 \pi r^2 \rho (\epsilon + \epsilon_g)\slabel{eq:structd}, 
 \end{subeqnarray}
 
-\noindent where $r$ is the radial coordinate, $P$, $T$ and $\rho$ are the gas pressure, temperature and density, respectively, $m$ is the mass enclosed by the radius $r$, $L$ is the luminosity from the surface of radius $r$, and $\epsilon_g \equiv -T \frac{ds}{dt}$ represents the heating per unit mass due to gravitational contraction, with $s$ the specific gas entropy.
-The temperature gradient $\nabla \equiv \frac{d \ln T}{d \ln P}$ has different expressions depending on the primary means of energy transport throughout the atmosphere.
-We assume that energy can be transported either through radiation or convection.
-When the luminosity is carried by radiative diffusion, the temperature gradient is given by
+\noindent with $r$ the radial coordinate, $P$, $T$ and $\rho$ the gas pressure, temperature and density, respectively, $m$ the mass enclosed by the radius $r$, $L$ the luminosity from the surface of radius $r$.
+The $\epsilon$ term represents the rate at which internal heat is generated per unit mass, while $\epsilon_g \equiv -T \frac{ds}{dt}$ is the heating per unit mass due to gravitational contraction, with $s$ the specific gas entropy.
+We do not take into account any internal energy sources and we therefore set $\epsilon=0$.
+The expression for the temperature gradient $\nabla \equiv \frac{d \ln T}{d \ln P}$ depends on whether energy is transported throughout the atmosphere by radiation or convection.
+In the case of radiative diffusion for an optically thick gas, the temperature gradient is
 
@@
 
-\noindent where $\sigma$ is the Stefan-Boltzmann constant and $\kappa$ is the opacity.
-Alternatively, when the energy is transported outwards through convective motions, the temperature gradient becomes
+\noindent with $\sigma$ the Stefan-Boltzmann constant and $\kappa$ the dust opacity.
+In our models the atmosphere is optically thick throughout the outer boundary.
+On the other hand, when the energy is transported outwards through convection, the temperature gradient is given by
 
@@
 
-\noindent where $\delad$ is the adiabatic temperature gradient.
-The process that dominates energy transport throughout the atmosphere is determined by the Schwarzschild criterion (e.g., \citealt{thompson06}): the atmosphere is stable against convection when
-
-\begin{equation}
-\label{eq:schwarz}
-\nabla < \delad
-\end{equation}
-
-\noindent and convectively unstable when the reverse is true.
-In order for convection to be effective, $\nabla \approx \delad$.
-Therefore, we find that the temperature gradient is given by $\nabla=\mathrm{min}(\delad, \delrad)$.
-
-In order for equation set (\ref{eq:struct}) to be solvable, it has to be supplemented by an equation of state (EOS) that relates pressure, temperature and density $P=P(\rho, T)$, and an opacity law for $\kappa$.
-In this study we use the interpolated EOS tables of \citet{saumon95} for a helium fraction $Y=0.3$.
+\noindent with $\delad$ the adiabatic temperature gradient.
+The convective and radiative layers of the envelope are separated by the Schwarzschild criterion (e.g., \citealt{thompson06}): the atmosphere is stable against convection when $\nabla < \delad$ and convectively unstable when $\nabla < \delad$.
+For effective convection, $\nabla \approx \delad$.
+The temperature gradient is thus given by $\nabla=\mathrm{min}(\delad, \delrad)$.
+
+The equation set (\ref{eq:struct}) is supplemented by an equation of state relating pressure, temperature and density, as well as an opacity law.
+In this study we use the interpolated EOS tables of \citet{saumon95} for a helium mass fraction $Y=0.3$.
 More details on the EOS tables and the methodology of combining the separate tables for hydrogen and helium are presented in section \ref{deladtable} and \App{EOStables}.
 
-We assume an opacity power law of the form
+We assume a standard ISM opacity power law given by
 
@@
 To estimate $\alpha$, $\beta$ and $\kappa_0$ we use the \citet{bell94} opacity laws for ice grains: $\alpha =0 $, $\beta=2$ and $\kappa_0=2$.
-We note that these values are valid only for low disk temperatures: $T_d \lesssim 100 K$.
-As such, we only consider cool atmospheres that form in the outer part of the protoplanetary disk ($a \geq 5$ AU).
-
-We now discuss our choice of core parameters and boundary conditions.
-We assume a solid core of fixed mass $M_c$ with a radius $R_{\rm c}=(3 M_{\rm c}/4 \pi \rho_{\rm c})^{1/3}$, where $\rho_{\rm c}$ is the core density.
+These values are valid only for low disk temperatures: $T_d \lesssim 100 K$.
+We therefore consider protoplanets that form in the outer parts of the disk where temperatures are low, i.e.
+$a \geq 5$ AU.
+Moreover, dust settling or grain growth will result in different numerical values for the $\kappa_0$ and $\beta$ coefficients.
+
+We further discuss our choice of core parameters and boundary conditions.
+We assume a solid core of fixed mass $M_{\rm c}$ with a radius $R_{\rm c}=(3 M_{\rm c}/4 \pi \rho_{\rm c})^{1/3}$, where $\rho_{\rm c}$ is the core density.
 We choose $\rho_{\rm c}=3.2$ g cm$^{-3}$ (e.g., \citealt{pap99}).
-Furthermore, we assume the atmosphere matches on to the protoplanetary at a distance equal to its Hill radius, the distance at which the gravitational attraction of the planet and of the host star are comparable.
-Its radius is given by
-
-\begin{equation}
-\label{eq:RHill}
-R_{\rm H} \equiv a \Big(\frac{M_{\rm p}}{3 M_{\odot}}\Big)^{1/3}
-\end{equation}
-
-\noindent Outside the Hill radius, the gravity of the planet is overcome by the tidal gravity from its host star, and hence only gas that lies within the Hill sphere can be gravitationally bound to the planet.
-We choose the effective outer boundary of the atmosphere to be the surface defined by the Bondi radius.
-The Bondi radius represents the distance from the planet at which the thermal energy of the nebular gas is of the order of the gravitational energy binding the gas to the planet.
-It is defined as
-
-\begin{equation}
-\label{eq:RB}
-R_{\rm B} \equiv \frac{G M_{\rm p}}{c_s^2}=\frac{G M_{\rm p}}{\mathcal{R} T\di},
-\end{equation}
-
-\noindent where $G$ is the gravitational constant, $M_{\rm p}$ is the total mass of the planet, $c_s$ is the sound speed, and $\mathcal{R}$ is the reduced gas constant: $\mathcal{R}=k_b/(\mu m_p)$, with $k_b$ the Boltzmann constant and $m_p$ the proton mass.
-Outside the Bondi sphere, the core gravity is too weak to significantly affect the nebular gas, justifying the choice of Bondi radius as the relevant scale for the atmosphere.
-We note that the Hill radius is still the correct scale for matching on to the disk, as perturbations of the nebular gas still occur past the Bondi radius due to the gravitational influence of the core.
-The above choice for the atmosphere boundary is only applicable when the Bondi radius is smaller than the Hill radius.
-If, on the other hand, $R_{\rm B}>R_{\rm H}$, the atmosphere only extends out to the Hill radius, since material cannot be gravitationally bound to the protoplanet outside the Hill sphere.
+The atmosphere is assumed to match on to the protoplanetary at the Hill radius, $R_{\rm H} \equiv a \Big(\frac{M_{\rm p}}{3 M_{\odot}}\Big)^{1/3}$, the distance at which the gravitational attraction of the planet and the tidal gravity due to the host star are equal.
+Outside the Hill sphere, the gravity of the planet is overcome by the tidal gravity from its host star, and hence only gas that lies within the Hill sphere can be gravitationally bound to the planet.
+The effective outer boundary of the atmosphere is the surface defined by the Bondi radius, $R_{\rm B} \equiv \frac{G M_{\rm p}}{c_{\rm s}^2}=\frac{G M_{\rm p}}{\mathcal{R} T\di}$.
+This is the distance from the planet at which the thermal energy of the nebular gas is of the order of the gravitational energy of the planet.
+Here $G$ is the gravitational constant, $M_{\rm p}$ is the total mass of the planet, $c_{\rm s}$ is the isothermal sound speed, and $\mathcal{R}$ is the reduced gas constant: $\mathcal{R}=k_b/(\mu m_p)$, with $k_b$ the Boltzmann constant and $m_p$ the proton mass.
+Outside the Bondi sphere, the gravity of the planet is too weak to significantly affect the nebular gas, which justifies the choice of Bondi radius as the relevant radius of the atmosphere.
+However, the nebular gas is still perturbed outside the Bondi sphere, and therefore the Hill radius is the correct scale for matching on to the disk.
+This choice for the atmosphere boundary applies only when the Bondi radius is smaller than the Hill radius; when $R_{\rm B}>R_{\rm H}$, the atmosphere only extends out to the Hill radius, since material cannot be gravitationally bound to the protoplanet outside the Hill sphere.
 At the Hill radius, the temperature and pressure are given by the nebular temperature and pressure: $T(R_{\rm H})=T_{\rm d}$ and $P(R_{\rm H})=P_{\rm d}$.
@@
 
-Lastly, we review the cooling model developed in Piso \& Youdin (in prep.) used to determine the time evolution of the atmosphere between subsequent static models.
+Lastly, we review the cooling model developed in PY13 used to determine the time evolution of the atmosphere between subsequent static models.
 A protoplanetary atmosphere embedded in a gas disk satisfies the following cooling equation:
@@
 
-This cooling model applies at any radius $R$ where the mass enclosed is $M$, for example the Bondi radius, the Hill radius, or the radius of the radiative-convective boundary.
-$L$ is the total luminosity, $L_{\rm c}$ is the luminosity from the solid core, and may include planetesimal accretion and radioactive decay.
-$\Gamma$ is the heat generation rate, $\dot{E}$ is the rate at which total energy (internal and gravitational) is lost, and  $e_{\mathrm{acc}}$ is the specific total energy brought in by mass accreting at the rate $\dot{M}$: $e_{\mathrm{acc}}=u-G M/R$.
+Here, $L$ is the total luminosity, $L_{\rm c}$ is the luminosity from the solid core, and may include planetesimal accretion and radioactive decay, $\Gamma$ is the rate of internal heat generation, $\dot{E}$ is the rate at which total energy (internal and gravitational) is lost, and  $e_{\mathrm{acc}}$ is the specific total energy brought in by mass accreting at the rate $\dot{M}$: $e_{\mathrm{acc}}=u-G M/R$.
 The last term represents the work done on a surface mass element.
 
-As a consequence of the equations above, both the atmosphere structure and the gas accretion rate are uniquely determined by the current atmosphere mass.
+%As a consequence of the equations above, both the atmosphere structure and the gas accretion rate are uniquely determined by the current atmosphere mass.
 As this mass accretion rate is slow compared to the time it takes to relax to this solution, we can make a quasi-static model of the atmosphere growth.
-By connecting sets of subsequent static atmospheres through the cooling equation (\ref{eq:coolingglobal}) we can obtain an evolutionary atmosphere series.
-
-%\subsection{Quasi-Static Two-Layer Model}
-%\label{twolayer}
-%
-%%The model and equations presented in this section so far can be applied to any protoplanetary atmosphere that satisfies the assumptions listed in section \ref{model}. 
-%
-%As a consequence of the equations derived in section \ref{cooling}, both the atmosphere structure and the gas accretion rate are uniquely determined by the current atmosphere mass.
-As this mass accretion rate is slow compared to the time it takes to relax to this solution, we can make a quasi-static model of the atmosphere growth.
-In this section we therefore describe the procedure to obtain an evolutionary series from the cooling model introduced in \ref{cooling} between consecutive two-layer static atmospheres (cf.
-\ref{model}).
-%
-%We use the boundary conditions and opacity laws described in section \ref{struct}.
-The numerical integration is performed through the shooting method, which solves the boundary value problem by reducing it to an initial value problem: trial values are chosen for the parameters at one of the boundaries, then the equations are integrated and the resulting values at the other boundary are com- pared to the actual boundary conditions.
-The procedure is repeated until convergence is achieved.
-We start with a total atmosphere mass $M_i$, where the index $i$ labels each evolutionary stage.
-For this mass we guess a trial luminosity $L=L_{\mathrm{guess}}$.
-The assumption of constant luminosity throughout the radiative zone sets the right hand side of equation (\ref{eq:structb}) to zero, i.e.
-the time dependence is neglected.
-We integrate the structure equations (\ref{eq:struct}) inwards from $R_{\mathrm{out}}=R_H$, with the boundary conditions $T(R_{\mathrm{out}})=T_d$ and $P(R_{\mathrm{out}})=P_d$.
-The numerical integration gives a value for the core mass implied by the trial solution $M_c=M_{c, \mathrm{guess}}$.
-We adjust $L_{\mathrm{guess}}$ until $M_{c, \mathrm{guess}}$ converges to the actual core mass $M_c$.
-%
-%The time evolution of the atmosphere is obtained using the global cooling model described in section \ref{cooling}.
-We neglect the luminosity due to planetesimal accretion and direct heat generation; as such, the first two terms in equation (\ref{eq:coolingglobal}) are set to zero.
-Equation (\ref{eq:coolingglobal}) is then evaluated at the RCB, since we assume that luminosity is primarily generated in the convective zone.
-In what follows, $X_{RCB}$ denotes quantity $X$ evaluated at the RCB.
-The time step $\Delta t_i$ between two consecutive models of masses $M_i$ and $M_{i+1}$ is calculated as follows:
-%
-%\begin{eqnarray}\nonumber
-%\label{eq:dti}
-%<L>_i \Delta t_i&=&-\Delta E_{RCB,i} + <e_{\mathrm{acc}, RCB}>_i \Delta M_{RCB, i} \\
-%&&  - <P_{RCB}>_i \Delta V_{RCB, M, i} 
-%\end{eqnarray}  
-%
-%\noindent where we denote 
-%
-%\begin{subeqnarray}
-%\Delta X_i&=&X_{i+1}-X_i \\
-%<X>_i&=&\frac{X_i+X_{i+1}}{2}
-%\end{subeqnarray}
-%\\
-%\noindent The subscript $M$ in the volume expression above signifies evaluating the change in volume at constant mass. 
-%
-%By connecting sets of subsequent static atmospheres through the procedure described above we therefore obtain an evolutionary atmosphere series.
-
-%
-
-%\section{Quasi-Static Kelvin-Helmholtz Contraction}
-%\label{KH}
-%
-%\textbf{Plots of radial / P-T profiles, also M-L, M-t profiles; compare the evolutionary profiles with the polytropic case.
-The text below is copy-pasted from Paper I - things will be similar, but needs a lot of rephrasing.
-}
-%
-%In this section we describe our choices for the gas, disk and core parameters, and present some typical atmosphere numerical models.
-As mentioned in section \ref{model}, we assume that the central star is a Sun-like star with $M_*=M_{\odot}$.
-We are primarily interested in the outer regions of the protoplanetary disk, so we generate atmosphere models between 10 and 100 AU.
-As a fiducial case, we assume that the atmosphere is composed of a hydrogen-helium mixture with the helium mass fraction $Y=0.3$., resulting in a mean molecular weight $\mu=2.35$.
-We do not take into account the dependence of the adiabatic index on the helium fraction, and therefore assume the adiabatic index for a diatomic gas $\delad=2/7$.
-We further explore how the atmosphere evolution varies with adiabatic index and molecular weight.
-We generate model atmospheres for monatomic adiabatic index $\delad=2/5$ and for a purely molecular hydrogen composition ($\mu=2$).
-The comparison between different gas compositions is deferred to section \ref{critcore}.
-%
-%Figure 1 shows instantaneous atmosphere radial profiles at $a=10$ AU accreting around a fixed core of mass $M_{\rm{c}}=5 M_{\oplus}$ for several total planet masses.
-In what follows,  if the $R_B<R_H$ hierarchy holds, we define the planet mass as the total mass enclosed within the Bondi radius of the atmosphere, and regard the gas between the Bondi and Hill radii as perturbations due to the gravitational influence of the planet.
-For illustrative purposes we mark the position of the Bondi radius and of the radiative convective boundary.
-We notice that the pressure and temperature in the convective regions of the atmospheres scale as $1/r$, which is the expected behavior of an ideal gas adiabat with a polytopic equation of state (e.g., \citealt{rafikov06}).
-In the outer radiative region of the atmosphere, the pressure has a nearly exponential profile, which is in agreement with the predictions of the analytic model (see section \ref{analytic} and \App{sec:analytic}).
-In addition, the atmosphere temperature at the radiative-convective boundary $T_{\rm{RCB}}$ only differs from the disk temperature $T_{\rm{d}}$ by an order unity factor, and so the atmosphere is nearly isothermal throughout the radiative zone.
-This is also in agreement with the analytic predictions.
-%, and justifies
-%%
-%%\begin{figure}[h]
-%%\centering
-%%\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/PTm_profiles_v2.pdf}
-%%%\vspace{-0.5in}
-%%\caption{Example pressure, temperature and mass profiles as a function of radius.}
-%%\end{figure}
-%
-%The location of the Bondi radius moves further out as the mass of the atmosphere increases, as the Bondi radius scales with mass.
-The location of the radiative-convective boundary has a more interesting behavior.
-As stated in section \ref{intro}, the nebular gas initially settles into equilibrium around the protoplanetary core on a short timescale relative to the thermal timescale, and is isentropic throughout with an entropy equal to that of the disk.
-In the context of our model, this atmosphere represents the minimum atmosphere mass for which a hydrostatic solution exists.
-This atmosphere is purely convective.
-As gas accumulates and the atmosphere starts cooling, an outer radiative region forms.
-This layer is thin for low mass atmospheres, as can be seen in Fig.
-1: for the $5.21 M_{\oplus}$ atmosphere, the ratio between the thickness of the radiative and convective regions is low.
-An increase in atmosphere mass causes the radiative zone to expand and the convective region to contract.
-As more and more gas is accumulated, the convective zone eventually starts expanding again; however, the ratio between the radii of the radiative and convective regions continually increases with atmosphere mass, i.e.
-the radiative regions of these types of atmospheres are getting deeper and deeper.
-\textbf{[This needs to be rephrased, somewhat vague and repetitive.]}
-%
-%We further use the global cooling model developed in section \ref{cooling} to obtain an evolutionary series from static atmosphere profiles.
-Figure 2 shows the luminosity and cooling time evolution with atmosphere mass.
-Gas accretion is initially rapid, but slows down significantly after a relatively small mass increase.
-For comparison, we also plot the analytic results derived in section \ref{analytic}.
-We see that there is only a very brief agreement between the analytic and full numerical results.
-At early times, the radiative zone is too shallow and so the approximation that $P_{\rm RCB} \gg P_{\rm d}$ used in section \ref{analytic} breaks down.
-At late times, the self-gravity of the atmosphere becomes important.
-%We see that self-gravity starts to matter
-%
-%%\begin{figure}[h]
-%%\centering
-%%\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Lt_profiles_v2.pdf}
-%%%\vspace{-0.5in}
-%%\caption{Luminosity and time evolution with mass. The analytic result is plotted for comparison.}
-%%\end{figure}
-%
-%%\subsection{The End of Quasi-Static Evolution}
-%%\label{endoftime}
-%%
-%%
-%%The right-hand side of equation (\ref{eq:dti}) decreases with increasing atmosphere mass, and eventually becomes negative when the mass of the envelope becomes comparable to the core mass $M_{\rm{atm}} \sim M_{\rm{c}}$.
-This implies a negative luminosity from the RCB.
-Physically, the luminosity $L$ represents the total heat flux through the surface defined by the RCB.
-The heat flux due to convective motions is always directed outwards, and hence it is positive.
-Moreover, the temperature gradient $dT/dr$ is negative, as the temperature decreases towards the outside.
-As a result (see equation (\ref{eq:structd})), the heat flux due to radiative diffusion is also positive.
-This implies that, in the absence of planetesimal accretion and internal heat sources, there is no physical mechanism that can justify a negative luminosity.
-The quasi-static, constant luminosity approximation breaks down at this point and a more detailed, time-dependent model is necessary.
-%
-%
-%%As can be inferred from Figure 2, the time steps between subsequent evolutionary models become increasingly smaller as the atmosphere grows in mass.
-Around the time when the atmosphere mass is comparable to the core mass $M_{\mathrm{atm}} \sim M_{\mathrm{c}}$, we find that the time steps become negative.
+
+As such, we can obtain an evolutionary series for the atmosphere by connecting sets of subsequent static atmospheres through the cooling equation (\ref{eq:coolingglobal}).
+Details of our numerical procedure are described in PY13.
 
@@
 
-In this section we explain the differences in the behavior of thermodynamic variables between an ideal gas polytrope and a realistic equation of state by studying the dependence of the adiabatic gradient (defined in equation \ref{eq:delad}) on temperature and pressure.
+In this section we explain the differences in behavior of thermodynamic variables between an ideal gas polytrope and a gas described by a realistic equation of state by studying the dependence of the adiabatic gradient (defined in equation \ref{eq:delad}) on temperature and pressure.
 Figure \ref{fig:deladmap} shows a contour plot of the adiabatic gradient as a function of gas temperature and pressure, which was obtained by interpolating and extending the \citet{saumon95} equation of state tables as referenced in section \ref{sec2} and described in \App{EOStables}.
@@
 \begin{enumerate}
-\item Intermediate temperature regime (300 K $\lesssim T \lesssim 3000$), where the hydrogen-helium mixture behaves as an ideal gas characterized by a polytropic equation of state.
+\item Intermediate temperature regime (300 K $\lesssim T \lesssim 3000$ K), where the hydrogen-helium mixture behaves like an ideal gas characterized by a polytropic equation of state.
 \item High temperature regime ($T \gtrsim$ 3000 K), where dissociation of molecular hydrogen occurs, followed by ionization of atomic hydrogen.
@@
 This energy depends on the ionization fraction and can be determined from the Saha equation (see e.g., \citealt{kippenhahn90}).
-The ionization fraction only depends on gas temperature and density (see \App{deladioniz}), and hence only the equation of state.
+The ionization fraction only depends on gas temperature and density (see \App{deladioniz}), and hence only on the equation of state.
 An expression of the adiabatic gradient as a function of the ionization fraction can be derived.
@@
 
-The dissociation of molecular hydrogen is dictated by an equation similar to the Saha equation, with the ionization energy replaced by dissociation energy (equal to 4.27 eV for molecular hydrogen, \citealt{mandl89}).
+The dissociation of molecular hydrogen is dictated by an equation similar to the Saha equation, with the ionization energy replaced by dissociation energy (equal to 4.27 eV for molecular hydrogen, e.g.
+\citealt{mandl89}).
 The adiabatic gradient therefore has an analogous behavior, consistent with Fig.
@@
 
-The internal energy and specific heat per unit mass associated with rotation for the individual isomers and for the equilibrium mixture can be derived from equations (\ref{eq:Zpara}), (\ref{eq:Zortho}), (\ref{eq:Zrspin}), (\ref{eq:u}) and (\ref{eq:cv}) and plotted in Figure \ref{fig:ucvr}.
+The internal energy and specific heat per unit mass associated with rotation for the individual isomers and for the equilibrium mixture can be derived from equations (\ref{eq:Zpara}), (\ref{eq:Zortho}), (\ref{eq:Zrspin}), (\ref{eq:u}) and (\ref{eq:cv}) and plotted in Figure \ref{fig:ucvr} (see also \citealt{farkas35}, Figure 1).
 The para state has no rotational energy for low temperatures, since all the molecules occupy the rotational level with $j=0$.
@@
 %%\vspace{-0.5in}
-\caption{Internal energy and heat capacity per unit mass for the hydrogen spins isomers and the equilibrium mixture as a function of temperature.}
+\caption{Internal energy and heat capacity per unit mass for the hydrogen spins isomers and the equilibrium mixture as a function of temperature. See \citet{farkas35}, Figure 1, for a similar plot.}
 \label{fig:ucvr}
@@
 
-Variations in the adiabatic gradient have two competing effects on the atmosphere evolution: they affect the luminosity of the envelope and the amount of energy needed to accrete more mass, i.e.
-$dE/dM$.
-Together, they result in changes in the growth time of the atmosphere, and therefore in the crossover time and critical core mass.
+Variations in the adiabatic gradient $\delad$ have two competing effects on the atmosphere evolution: they yield a lower envelope luminosity when compared to an ideal gas, and a larger amount of energy that needs to be radiated per unit mass, which slows down accretion.
+Together, they result in changes in the growth time of the atmosphere, and therefore in the crossover time and critical core mass (both defined in section \ref{intro}).
 In this section we discuss how the variable adiabatic gradient discussed in section \ref{deladtable} affects the atmosphere evolution when compared to an ideal gas of constant $\delad$.
 We explore the effect of partial dissociation and hydrogen spin isomers (see section \ref{deladtable}) on the atmosphere luminosity and cooling time evolution.
-As the adiabatic gradient is variable throughout the atmosphere profiles described in section (\textbf{some previous section}), we first investigate the differences in atmosphere profiles and evolution for ideal gas polytropes with different adiabatic gradients in section \ref{deladpoly}.
+Due to the equation of state described in section \ref{deladtable}, a realistic atmosphere will not have a constant adiabatic gradient $\delad$, as was assumed in PY13.
+We thus first investigate the differences in atmosphere profiles and evolution for ideal gas polytropes with different adiabatic gradients in section \ref{deladpoly}.
 We then show how the variable adiabatic gradient affects the time evolution of the atmosphere in section \ref{deladeffect}.
@@
 The polytrope with a larger adiabatic gradient has a more shallow convective zone, and hence a deeper radiative region, since a larger temperature gradient delays the onset of convection.
-As the luminosity in the radiative region is inversely proportional to the depth of the radiative zone (see equation (\ref{eq:structd})), a deeper radiative region results in a lower luminosity, which explains the results in the top panel of Figure \ref{fig:Ltplotpoly}.
+As the luminosity in the radiative region is inversely proportional to the depth of the radiative zone, a deeper radiative region results in a lower luminosity, which explains the results in the top panel of Figure \ref{fig:Ltplotpoly}.
 
@@
 
-We study the energy behavior for the two polytropes in order to explain the $dE/dM$ effect.
-We have shown in Piso \& Youdin (in prep.) that polytropes with $\delad=2/7$ have most of the energy concentrated at the bottom of the atmosphere, while polytropes with $\delad=2/5$ have the bulk of the energy towards the outer boundary.
+In spite of the low luminosity of the atmosphere with $\delad=2/5$, the envelope still grows faster than in the case of a diatomic gas, as shown in the bottom panel of Figure \ref{fig:Ltplotpoly}.
+This is due to fact that the amount of energy per unit mass that needs to be radiated away, i.e.
+$|dE/dM|$, is lower for the $\delad=2/5$ polytrope.
+We have shown in PY13 that polytropes with $\delad=2/7$ have most of the energy concentrated at the bottom of the atmosphere, while polytropes with $\delad=2/5$ have the bulk of the energy towards the outer boundary.
 The bottom panel of Figure \ref{fig:ETrplotpoly} shows an instantaneous energy profile for the same total mass $M_{\rm{tot}} = 11.8 M_{\oplus}$, confirming this behavior.
-It takes more energy to bring in gas deep in the atmosphere for an envelope that has the bulk of its energy concentrated towards the bottom,  resulting in a larger $dE/dM$.
+It takes more energy to bring in gas deep in the atmosphere for an envelope that has the bulk of its energy concentrated towards the bottom,  resulting in a larger $|dE/dM|$.
 The energy effect prevails over the luminosity effect, resulting in a longer cooling time for the envelope, as shown in the bottom panel of Figure \ref{fig:Ltplotpoly}.
@@
 
-We now discuss the differences in luminosity and $dE/dM$ between an ideal gas with constant adiabatic gradient and atmospheres with variations in $\delad$ as prescribed by the equation of state discussed in section \ref{EOSeffects}.
+In this section we use the results obtained in section \ref{deladpoly} to explain the effects on the atmosphere evolution of a realistic equation of state.
+We explore the effect of hydrogen dissociation at the high temperatures in the inner part of the atmosphere, and the effect of hydrogen spin isomers at the low temperatures at the top of the atmosphere separately.
+In order to do this, we generate atmosphere profiles in which the nebular gas is assumed to be described by a combination of ideal and realistic equations of state, depending on the temperature.
+To explain the effects of the hydrogen spin isomers in the outer parts of the atmosphere, we assume that the realistic EOS effects only matter at low temperatures; conversely, we assume that the realistic EOS effects are important only at high temperatures in order to study the effects of hydrogen dissociation deep in the atmosphere.
+As such, we generate three sets of atmosphere profiles, choosing $a=10$ AU and $M\co=5 M_{\oplus}$.
+The first one is described by a realistic EOS for temperatures larger than 500 K and by an ideal gas with $\delad=0.3$ for $T<500$ K.
+The second profile consists of an ideal gas polytrope with $\delad=0.3$ for $T>500$ K and a realistic gas in the low temperature regime.
+Finally, the last profile corresponds to an ideal gas with $\delad=0.3$, the adiabatic gradient of an ideal hydrogen-helium mixture.
+We choose $T=500$ K as the reference temperature since the adiabatic gradient of the realistic gas is roughly constant around this temperature (see Figure \ref{fig:deladmap}).
+The differences in atmosphere structure and evolution between the ideal gas and the realistic gas at low (high) temperatures highlight the effects of hydrogen spin isomers (hydrogen dissociation).
+Figure \ref{fig:tplotall} illustrates the differences in time evolution between these profiles.
+We see that both dissociation and spin isomers have a comparable effect on the atmosphere growth, and result in slower cooling, and therefore a longer crossover time, when compared to the polytropic ideal gas equation of state.
+The cooling time is dependent on both the total energy released due to the contraction of the envelope and the luminosity of the atmosphere.
+In what follows we explore the relative influence of these two factors separately.
+
+%The resulting time evolution is shown in Figure \ref{fig:tplotall}.
+We see that both dissociation and spin isomers have a comparable effect on the atmosphere growth, and result in slower cooling, and therefore a longer crossover time, when compared to the polytropic ideal gas equation of state.
+The cooling time is dependent on both the total energy released due to the contraction of the envelope and the luminosity of the atmosphere.
+In what follows we explore the relative influence of these two factors separately.
+
+%We now discuss the differences in luminosity and $dE/dM$ between an ideal gas with constant adiabatic gradient and atmospheres with variations in $\delad$ as prescribed by the equation of state discussed in section \ref{EOSeffects}.
 In what follows we describe our choices of equation of state combinations.
@@
 The first one corresponds to an ideal gas of constant adiabatic index $\delad=0.3$.
-The second one is described by the real EOS for temperatures larger than 500 K and by an ideal gas with $\delad=0.3$ for $T<500$ K.
-Finally, the third profile consists of an ideal gas polytrope with $\delad=0.3$ for $T>500$ K and a real gas in the low temperature regime.
+The second one is described by the realistic EOS for temperatures larger than 500 K and by an ideal gas with $\delad=0.3$ for $T<500$ K.
+Finally, the third profile consists of an ideal gas polytrope with $\delad=0.3$ for $T>500$ K and a realistic gas in the low temperature regime.
 We compare the first two profiles to show the dissociation effects, and the first and third profile to show the effects of ortho- and parahydrogen.
@@
 \caption{Cooling time evolution as a function of total mass (core + atmosphere) for a variety of EOS combinations, for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=10 M_{\oplus}$.
-The cooling time is larger both due to hydrogen dissociation and spin effects when compared to an ideal gas polytrope.}
+The cooling time is larger for the realistic EOS both due to hydrogen dissociation and spin effects when compared to an ideal gas polytrope.}
 \label{fig:tplotall}
@@
 
-The right panel of Figure \ref{fig:TLrplot} shows the luminosity evolution with mass for the three combinations of equations of state described above, as well as for the complete real gas EOS.
-We use again instantaneous atmosphere profiles to explain the differences.
-The left panel of Figure \ref{fig:TLrplot} shows the instantaneous temperature profile and the location of the radiative-convective boundary for a total fixed mass $M_{\rm{tot}}=11.8 M_{\oplus}$.
-The real equation of state for low temperatures is characterized by a lower adiabatic index in the outer regions, due to the spin effects, and is therefore dominant in the radiative zone.
+The right panel of Figure \ref{fig:TLrplot} shows the luminosity evolution with mass for the three combinations of equations of state described above, as well as for the complete realistic gas EOS.
+Similarly to section \ref{deladpoly}, we use instantaneous atmosphere profiles to explain the differences.
+The left panel of Figure \ref{fig:TLrplot} shows the instantaneous temperature profile and the location of the radiative-convective boundary for a total fixed mass (core + atmosphere) $M_{\rm{tot}}=11.8 M_{\oplus}$.
+The realistic equation of state for low temperatures is characterized by a lower adiabatic index in the outer regions, due to the spin effects, and is therefore dominant in the radiative zone.
 As a result, it generates a deeper radiative zone with a lower luminosity, which explains the results in the left panel of Figure \ref{fig:TLrplot}.
-Moreover, since the cooling time is inversely proportional to the luminosity, the spin effect will result in a longer cooling (and crossover) time.
+Moreover, since the cooling time is inversely proportional to the luminosity, the spin effect will result in a longer cooling time.
 
 The energy behavior is shown in Figure \ref{fig:Erplot}.
-The real EOS for high temperatures has a low adiabatic index deep in the atmosphere, due to hydrogen dissociation, and thus the bulk of its energy concentrated at the bottom of the atmosphere, for the reasons described in section \ref{deladpoly}.
-It takes more energy to add mass more mass deep in the atmosphere, and $dE/dM$ is larger as a result.
+The realistic EOS for high temperatures has a low adiabatic index deep in the atmosphere, due to hydrogen dissociation, and thus the bulk of its energy concentrated at the bottom of the atmosphere, for the reasons described in section \ref{deladpoly}.
+It takes more energy to add mass more mass deep in the atmosphere, and $|dE/dM|$ is larger as a result.
 %
@@
 %%\vspace{-0.5in}
-\caption{Left panel: Instantaneous emperature profile as a function of radius for a variety of EOS combinations, for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=10 M_{\oplus}$.
+\caption{Left panel: Instantaneous temperature profile as a function of radius for a variety of EOS combinations, for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=10 M_{\oplus}$.
 The total mass (core + atmosphere) is $11.8 M_{\oplus}$.
 The location of the radiative-convective boundary is marked.
-The effect of hydrogen spin isomers in the outer region of the atmosphere sets the location of the radiative-convective boundary.
+The effect of hydrogen spin isomers at low temperatures in the outer region of the atmosphere sets the location of the radiative-convective boundary.
 Right panel: Luminosity evolution as a function of total mass (core + atmosphere) for a variety of EOS combinations, for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=5 M_{\oplus}$.
-Dissociation deep in the atmosphere increases the luminosity when compared to an ideal gas polytrope, while the existence of the hydrogen spin isomers in the outer part of the atmosphere results in a lower luminosity.
+The existence of the hydrogen spin isomers at low temperatures near the top of the atmosphere results in a lower luminosity.
 The two effects combined yield a lower luminosity for the realistic equation of state when compared to the polytrope.
@@
 The total mass (core + atmosphere) is $11.8 M_{\oplus}$.
-Hydrogen dissociation deep in the atmospheres causes the bulk of the energy to be concentrated at the bottom of the atmosphere.}
+Hydrogen dissociation deep in the atmospheres causes the bulk of the energy to be concentrated at the bottom of the atmosphere.
+This increases the amount of energy per unit mass that needs to be radiated away, i.e.
+$|dE/dM|$, resulting in a longer crossover time for the realistic EOS when compared to the polytrope.}
 \label{fig:Erplot}
@@
 In this section we put together the results obtained in section \ref{EOSeffects} and determine the minimum core mass to initiate runaway gas accretion during the lifetime of the protoplanetary disk assuming that the nebular gas is described by a realistic equation of state as prescribed by the \citet{saumon95} EOS tables (see section \ref{EOSeffects}).
-As in Piso \& Youdin (in prep.), we define this minimum core mass as the \textit{critical core mass}.
+As in PY13, we define this minimum core mass as the \textit{critical core mass}.
 Moreover, we define the time elapsed until runaway gas accretion is initiated when $M_{\rm{atm}} \sim M\co$ as the \textit{crossover time}.
 In this section we first explore the dependence of the crossover time on the core mass for a fixed semi-major axis.
-We then determine the critical core mass to form a giant planet before the dissipation of the gas in the protoplanetary disk, assuming that the nebular gas is described by a realistic hydrogen-helium mixture, and we compare this with the results from Piso \& Youdin (in prep.) for an ideal diatomic gas.
+We then determine the critical core mass to form a giant planet before the dissipation of the gas in the protoplanetary disk, assuming that the nebular gas is described by a realistic hydrogen-helium mixture, and we compare this with the results from PY13 for an ideal diatomic gas.
 
@@
 Figure \ref{fig:tvsMplot} displays the time evolution and the crossover time for core masses between 7 and 16 $M_{\oplus}$ at $a=10$ AU in our fiducial disk.
-The crossover time is shorter for lower mass cores, consistent with the results of Piso \& Youdin (in prep.).
+The crossover time is shorter for higher mass cores, consistent with the results of PY13.
 
@@
 Figure \ref{fig:Mvsaplot} shows the critical core mass for a massive atmosphere to form during a typical lifetime of a protoplanetary disk $t=3$ Myrs, for a gas described by a realistic equation of state.
-For comparison, we also plot the results of Piso \& Youdin (in prep.) for an ideal diatomic gas.
+For comparison, we also plot the results of PY13 for an ideal diatomic gas.
 The use of a realistic equation of state increases the critical core mass by more than a factor of 2.
@@
 \caption{The minimum core mass for an atmosphere to initiate runaway gas accretion within the lifetime of a typical protoplanetary disk $t \sim 3$ Myrs as a function of semi-major axis, for a realistic hydrogen-helium mixture.
-The results of Piso \& Youdin (in prep.) for an ideal diatomic gas are plotted for comparison.
+The results of PY13 for an ideal diatomic gas are plotted for comparison.
 The realistic equation of state yields core masses larger by more than a factor of 2 when compared to the polytrope.}
@@
 At the limit, $L_{\rm{acc}}=L_{\rm{KH}}$.
-For a given atmosphere model we can therefore estimate the maximum planetesimal accretion rate during the gas contraction phase in order for the atmosphere to be dominated by gas contraction.
+For a given atmosphere model we can therefore estimate the maximum planetesimal accretion rate during the contraction of the envelope in order for the atmosphere to be dominated by gas contraction.
 We choose as a fiducial case an atmosphere forming at 40 AU and with a core mass of $10 M_{\oplus}$, described by a realistic equation of state.
-For this choice of parameters, the atmosphere crossover time is $t \sim$ 2.7 Myrs, which is within the typical life time of a protoplanetary disk.
+For this choice of parameters, the atmosphere crossover time is $t _{\rm{co}}\sim$ 2.7 Myrs, which is within the typical life time of a protoplanetary disk.
 The results are presented in Figure \ref{fig:accrates}.
@@
 %\vspace{-0.5in}
-\caption{Various relevant accretion rates.
-$\dot{M}_{\rm{atm}}$ is the growth rate of the atmosphere as estimated by our model, and $\dot{M}_{\rm{c,KH}}$ is the maximum planetesimal accretion rate during the gas contraction phase in order for our regime to be valid.
-For comparison, we plot the core accretion rate $\dot{M}_{\rm{c,acc}}$ necessary to grow the core on the same time scale as the atmosphere $\tau \sim 2.7$ Myrs, and a typical planetesimal accretion rate where the random velocity of the planetesimals is given by the Hill velocity due to the core.}
+\caption{Various relevant accretion rates in the case of a planet forming at 10 AU and with a core mass $M\co=10 M_{\oplus}$.
+The $\dot{M}_{\rm{atm}}$ curve represents the growth rate of the atmosphere as estimated by our model, and $\dot{M}_{\rm{c,KH}}$ is the maximum planetesimal accretion rate during the gas contraction phase in order for our regime to be valid.
+For comparison, we plot the core accretion rate $\dot{M}_{\rm{c,acc}}$ necessary to grow the core on the same time scale as the atmosphere $\tau \sim 2.7$ Myrs, and a typically planetesimal accretion rate $\dot{M}_{\rm{c,Hill}}$ where the random velocity of the planetesimals is given by the Hill velocity due to the core.}
 \label{fig:accrates}
@@
 
-We label the resulting minimum core accretion rate as $\dot{M}_{\rm{c,KH}}$.
+We label the resulting minimum core accretion rate as $\dot{M}_{\rm{c,KH}}$, with  $\dot{M}_{\rm{c,KH}}=\frac{L_{\rm{KH}} R\co}{G M\co}$ from equation (\ref{eq:Lacc}).
 The atmosphere growth rate $\dot{M}_{\rm{atm}}$ is also plotted for comparison.
-We therefore see that the core accretion rate has to be $\sim2-3$ orders of magnitude lower than the atmosphere accretion rate for our assumptions to be valid.
-If the core had accreted planetesimals at this constant rate since it started forming, then the formation of a core massive enough to attract an atmosphere would not have been possible within typical disk life timescales, which indicates a slowing down of the planetesimal accretion regime.
+We see that the core accretion rate has to be $\sim2-3$ orders of magnitude lower than the atmosphere accretion rate for our assumptions to be valid.
+If the core had accreted planetesimals at this constant rate since it started forming, then the formation of a core massive enough to attract an atmosphere would not have been possible within a typical disk lifetime, implying that planetesimal accretion must have been larger initially, then slowed down.
 Possible explanations for that include the core having formed in the inner part of the disk and later migrated outwards, or the core having been depleted of planetesimals due to a giant neighbor.
-We also estimate the core accretion rate
+We further estimate the core accretion rate needed for the core to form on the same timescale as our model atmosphere, $\tau=2.7$ Myrs:
 
@@
 \end{equation}
-
-\noindent needed for the core to form on the same timescale as our model atmosphere, $\tau=2.7$ Myrs, as well as a typical planetesimal accretion rate, for which the random velocities of the planetesimals are of the order of the Hill velocity around the protoplanetary core \citep{goldreich04}.
-We denote this latter rate as $\dot{M}_{\rm{c,typical}}$.
+For reference, we also plot a typically assumed planetesimal accretion rate, for which the random velocities of the planetesimals are of the order of the Hill velocity around the protoplanetary core (e.g., \citealt{goldreich04}).
+We denote this latter rate as $\dot{M}_{\rm{c,Hill}}$.
 This is the accretion rate at the boundary between the dispersion dominated and shear dominated regimes.
-It is easy to see that  $\dot{M}_{\rm{c,typical}}$ is more than one order of magnitude lower than the gas accretion rate of our model atmosphere $\dot{M}_{\rm{atm}}$, and lower than the core accretion rate $\dot{M}_{\rm{c,acc}}$ needed to grow the core and the atmosphere at the same time within the disk life time.
+We estimate $\dot{M}_{\rm{c,Hill}}$ following \citet{rafikov06} (equation A1):
+
+\begin{equation}
+\label{eq:MdotHill}
+\dot{M}_{\rm{c,Hill}}=\Omega \Sigma_{\rm p} R\co R_{\rm H},
+\end{equation}
+where $\Sigma_{\rm p}$ is the surface density of solids, assumed to satisfy $\Sigma\di \approx 100 \Sigma_{\rm p}$ for a dust-to-gas ratio of 0.01.
+
+ %It is easy to see that  $\dot{M}_{\rm{c,typical}}$ is more than one order of magnitude lower than the gas accretion rate of our model atmosphere $\dot{M}_{\rm{atm}}$, and lower than the core accretion rate $\dot{M}_{\rm{c,acc}}$ needed to grow the core and the atmosphere at the same time within the disk life time.
 As such, the formation of a giant planet by growing the core first, then letting the atmosphere cool is faster than growing the core and the atmosphere at the same time at a steady planetesimal accretion rate.
@@
 \citet{rafikov06} assumes an ideal gas polytropic equation of state and a lower opacity than the standard ISM opacity that we use in our calculations (see equation \ref{eq:opacitylaw}), due to grain growth.
-In Piso \& Youdin (in prep.) we showed that a reduction in opacity results in a shorter crossover time and therefore a lower critical core mass.
+In PY13 we showed that a reduction in opacity results in a shorter crossover time and therefore a lower critical core mass.
 In this section, we calculate the critical core mass for an ideal gas polytrope with the standard ISM opacity reduced by a factor of 100, which is comparable to the opacity law used by \citet{rafikov06}.
@@
 \label{eq:critraf}
-M_{\rm{crit, acc}} \sim \Big[\frac{\dot{M}(M_{\rm{c}})}{64 \pi^2 C} \frac{\kappa_0}{\sigma G^3} \frac{1}{R\co M\co^{1/3}} \Big(\frac{k}{\mu}\Big)^4\Big]^{3/5},
+M_{\rm{crit, acc}} \sim \Big[\frac{\dot{M}(M_{\rm{c}})}{64 \pi^2 C} \frac{\kappa_0}{\sigma G^3} \frac{1}{R\co M\co^{1/3}} \Big(\frac{k_b}{\mu}\Big)^4\Big]^{3/5},
 \end{equation}
@@
 \label{eq:cminc}
-\Delta M_{\rm{c}} = \int_0^{t_{\rm{crit}}} \dot{M_{\rm{c}}} dt \approx \sum_i \dot{M_{\rm{c}}}_i \Delta t_i,
-\end{equation}
-
- \noindent where $t_{\rm{crit}}$ is the time elapsed until runaway gas accretion commences, and the accretion rate $ \dot{M_{\rm{c}}}_i $ is given by
+\Delta M_{\rm{c}} = \int_0^{t_{\rm{co}}} \dot{M_{\rm{c}}} dt \approx \sum_i \dot{M_{\rm{c}}}_i \Delta t_i,
+\end{equation}
+
+ \noindent where the accretion rate $ \dot{M_{\rm{c}}}_i $ is given by
 
@@
  In this paper we have studied giant planet formation under the assumption that the planetesimal accretion rate is negligible and the atmosphere evolution is dominated by gas contraction.
-We have used the model developed in Piso \& Youdin (in prep.) to build atmosphere profiles assuming that the nebular gas obeys a realistic equation of state that takes into account non-ideal effects.
+We have used the model developed in PY13 to build atmosphere profiles assuming that the nebular gas obeys a realistic equation of state that takes into account non-ideal effects.
 We found that the variations in the adiabatic index due to the realistic equation of state result in a significantly larger crossover time, and therefore critical core mass, when compared to an ideal gas polytrope.

### 2013-06-20 tex/draft/draft2.tex
updated analytic section

@@
 
-We consider a two-layer atmosphere structure model as presented in section \ref{model}.
-We approximate the outer radiative zone of the atmosphere as nearly isothermal.
+We consider a two-layer atmosphere structure model as presented in section \ref{model}, and we make two key assumptions: we neglect the self-gravity of the atmosphere, and we assume that the outer radiative region is very thick and hence the pressure at the radiative-convective boundary is much larger than the disk pressure $P\di$.
+We will show in section \ref{KH} that these assumptions limit the applicability of this simplified model during the late and early stages of the atmosphere evolution, respectively.
+We also approximate the structure of the radiative zone as nearly isothermal, and note that this assumption is consistent with the numerical solutions (see section \ref{profiles}).
+
+As self-gravity is quantitatively important, we expect our analytic model to only provide scalings for the full numerical model rather than absolute values.
+
+%We approximate the outer radiative zone of the atmosphere as nearly isothermal.
 Moreover, we do not take into account the self-gravity of the atmosphere in either region.
@@
 
-In the next subsections we derive the structure and cooling behavior of a two-layer non self-gravitating atmosphere, deferring some of the calculations to \App{sec:analytic}.
+We use the quasi-static model of section \ref{sec2} to derive expressions for the mass, energy and luminosity of a static non self-gravitating atmosphere.
+We use this to derive the time evolution of the atmosphere and determine the minimum core mass for runaway gas accretion in section \ref{coolingan}.
+We defer the details of our calculations to \App{sec:analytic}.
+
+%In the next subsections we derive the structure and cooling behavior of a two-layer non self-gravitating atmosphere, deferring some of the calculations to \App{sec:analytic}.
 We estimate the temperature and pressure corrections at the radiative-convective boundary due to the fact that the outer radiative region of the atmosphere is not purely isothermal in section \ref{RCBcorr}.
@@
 
-\subsection{Isothermal Atmosphere}
-\label{iso}
-
-We now consider the structure of a low mass (non self-gravitating) isothermal atmosphere.
-We assume the atmosphere matches onto a constant background density, $\rho_{\rm d}$, at a distance $r_{\rm fit} = n_{\rm fit} \RB$, where $R_{\rm B}$ is the Bondi radius defined in equation (\ref{eq:RB}).
-From equation (\ref{eq:structa}) and without taking into account the self-gravity of the atmosphere, the resulting density profile is
-\begin{equation} \label{eq:rhoiso}
-\rho = \rho_{\rm d} \exp \left({R_B \over r} - {1 \over n_{\rm{fit}}} \right) \approx   \rho_{\rm{d}} \exp \left(R_B \over r  \right),
-\end{equation} 
-where the approximate inequality holds deep inside the atmosphere ($r \ll \RB$) for any $n_{\rm fit} \gtrsim 1$.
-However, the choice of boundary condition does have an order unity effect on the density near the Bondi radius.
-In our study, $n_{\rm{fit}}=R_{\rm H}/R_{\rm B}$.
-
-The mass of the atmosphere is determined by integrating equation (\ref{eq:structb}) from the core to the Bondi radius using the density profile (\ref{eq:rhoiso}).
-Planets can attract massive atmospheres if $\theta\co \equiv R_B/R\co \gg 1$.
-In this case
-\begin{equation} \label{eq:MatmISO}
-M_{\rm iso} \approx 4 \pi \rho\di {R\co^4 \over R_B} e^{R_B/R\co} = 4 \pi \rho\di {R\co^3 \over \theta\co} e^{\theta\co}\, .
-\end{equation} 
-This result is the leading order term in a series expansion.
-Furthermore, because the atmospheric scale-height at $R\co$ is $H_\rho = |dr /d\ln\rho| = R\co^2/R_B$, the result is intuitively the correct order of magnitude.
-
-\subsection{Temperature and Pressure Corrections at the Radiative-Convective Boundary}
-\label{RCBcorr}
-
-%\subsection{Temperature Contrast at Convective Boundary}
-In this section we estimate the temperature and pressure corrections at the radiative-convective boundary due to the fact that the radiative region is not purely isothermal.
-We show that the temperature contrast between the temperature at the radiative-convective boundary $T_{\rm RCB}$ and the ambient disk $T_{\rm d}$, is modest.
-From equation (\ref{eq:delrad}), we express the radiative lapse rate
-\begin{equation}\label{eq:delradan}
-\delrad = {3 \kappa P \over 64 \pi  G M \sigma T^4} L = \nabla\di {(P/P_{\rm d})^{1+\alpha} \over (T/T_{\rm d})^{4-\beta}},
-\end{equation}
-
-\noindent where the second equality follows from the opacity law (\ref{eq:opacitylaw}), and $\nabla_{\rm d}$ is the radiative temperature gradient at the disk:
-
-\begin{equation}
-\label{eq:delo}
-\nabla \di \equiv \frac{3 \kappa(P{\di}, T{\di}) P{\di}}{64 \pi G M \sigma T_{\rm d}^4} L.
-\end{equation}
-
-\noindent $M$ is the sum of the core mass and the atmosphere mass below the pressure level $P$.
-If the mass in the radiative zone is small, then we can hold $M$ fixed at the sum of the core and convective zone masses: $M = M\co + M_{\cb}$.
-With this assumption, we get a constant value for $\nabla_{\rm d}$.
-As $\delrad=d \ln T/d \ln P$, the temperature profile integrates to
-\begin{equation}\label{eq:radTP}
-\left(T \over T_{\rm d}\right)^{4-\beta} - 1 = {\nabla\di \over \nabla_\infty} \left[\left({P \over P_{\rm d}}\right)^{1+\alpha} - 1 \right] \, ,
-\end{equation} 
-where $\nabla_\infty = (1+\alpha)/(4-\beta)$ is the radiative temperature gradient for $T ,P \rightarrow \infty$.
-We  apply \Eqs{eq:delradan}{eq:radTP} at the radiative-convective boundary (where $\delrad = \delad$) under the assumption that the pressure there $P\cb$ satisfies the relation $P\cb \gg P_{\rm d}$.
-This is motivated by the fact that the density, and hence pressure profile, falls exponentially (see equation \ref{eq:rhoiso}).
-The resulting temperature contrast at the radiative-convective boundary is
-\begin{equation}\label{eq:Tcb}
-\chi \equiv {T\cb\over T_{\rm d}} \simeq \left(1 - {\nabla_{\rm ad} \over \nabla_\infty}\right)^{-\frac{1}{4-\beta}} \, .
-\end{equation} 
-For  $\alpha=0$ and $\beta =2$,  and assuming $\delad = 2/7$, we find $\chi \approx 1.5$. Values of $\chi$ and $\nabla_\infty$ for other choices of $\beta$ are summarized in \App{sec:analytic}.
-
-%\footnote{Our definitions of $\nabla\di$ and $\nabla_\infty$ are precisely opposite to \citet{rafikov06}, but consistent with other works and the general convention of labeling a quantity $f$ in the disk as $f\di$.}
-
-The pressure at the convective boundary  follows from \Eqs{eq:radTP}{eq:Tcb} as
-\begin{equation}
-\label{eq:Pcbapprox}
-{P\cb\over P_{\rm d}} \simeq \left({\delad/\nabla\di \over 1 - \delad/\nabla_\infty}\right)^{1 \over 1 + \alpha}
-\end{equation} 
-This pressure contrast can be quite large since $\nabla_{\rm d} \ll 1$.
-
- We now eliminate $\nabla\di$ from equation (\ref{eq:Pcbapprox}) to obtain a relation between temperature and pressure as a function of the pressure at the radiative-convective boundary $P\cb$.
-From equation (\ref{eq:radTP}), it follows that
- \begin{equation}\label{eq:TP}
-{T \over T_{\rm d}} = \left\{1 + {1 \over {\nabla_\infty \over \delad} - 1} \left[ \left({P \over P\cb}\right)^{1+\alpha} -  \left({P_{\rm d} \over P\cb}\right)^{1+\alpha}  \right] \right\}^{1 \over 4-\beta}\, .
-\end{equation} 
- We can then determine the radius of the convective boundary $R\cb$ from the hydrostatic balance equation (\ref{eq:struct a}) as 
-\begin{equation}\label{eq:RCBint}
-{R_B \over R\cb} = \int_{P\di}^{P\cb} {T \over T_{\rm d}} {dP \over P}\, .
-\end{equation} 
-An isothermal atmosphere gives a simple logarithmic dependence on $P\cb$.  However using \Eq{eq:TP} in the integral gives
-\begin{equation}\label{eq:Rcb}
-{R_B \over r\cb} = \ln \left(P\cb \over P\di \right) - \ln \theta \, ,
-\end{equation} 
-with an extra correction term, $\theta < 1$.  The form we chose for the correction term allows us to relate the disk and radiative-convective boundary pressures as :
+%\subsection{Simplified cooling model}
+%\label{coolingan}
+
+%In this section we
+
+%We now use a simplified version of the cooling model introduced in section \ref{cooling} in order to obtain an analytic cooling model.
+As before, we assume a polytropic equation of state and neglect self-gravity.
+
+\subsection{Mass, Energy and Luminosity}
+\label{MELan}
+
+In this section, we derive simplified analytic expressions for the mass, energy and luminosity of the atmosphere.
+
+Our assumption of a nearly isothermal radiative region yields simple expressions for the temperature and pressure at the radiative-convective boundary as a function of the nebular temperature and pressure.
+These expressions will helps us estimate the mass and energy of the atmosphere.
+We account for the departure from a perfectly isothermal layer (i.e., constant temperature and exponential pressure profile, see also equation \ref{eq:rhoiso}) by introducing two correction terms $\chi$ and $\theta$ such that
+
+\begin{equation}
+\label{eq:Tcb}
+T\cb = \chi T\di
+\end{equation} 
+and
  \begin{equation}\label{eq:PcbRcb}
-P\cb = \theta P_{\rm d} e^{R_B/R\cb} \, .
-\end{equation}   
-In the $P\cb \gg P_{\rm d}$ limit, the correction term is an order unity constant that depends on $\alpha$, $\beta$ and $\delad$.
-Similarly to the temperature correction factor $\chi$,  $\theta$ accounts for the fact that the radiative region is not perfectly isothermal.
-For $\delad=2/7$, $\alpha=0$ and $\beta=2$, we find $\theta \approx 0.556$.
-Values of $\theta$ for other choices of $\beta$ are summarized in \App{sec:analytic}.
+P\cb = \theta P_{\rm d} e^{R_{\rm B}/R\cb} \, ,
+\end{equation}
+where $\chi$ can be approximated as
+\begin{equation}
+\label{eq:chi}
+\chi \simeq \Big(1-\frac{\delad}{\nabla_{\infty}} \Big)^{-\frac{1}{4-\beta}} ,
+\end{equation}
+with $\nabla_{\infty}=(1+\alpha)/(4-\beta)$ the radiative temperature gradient as $T, P \rightarrow \infty$.
 A simple analytic expression for $\theta$ is not possible.
-
-\subsection{Simplified cooling model}
-\label{coolingan}
-We now use a simplified version of the cooling model introduced in section \ref{cooling} in order to obtain an analytic cooling model.
-As before, we assume a polytropic equation of state and neglect self-gravity.
-
-\subsubsection{Luminosity,  Energy \& Mass}
-From equation (\ref{eq:delradan}), the luminosity that emerges at the radiative-convective boundary is
-\begin{equation} \label{eq:Lcb}
-L\cb = {64 \pi G M\cb \sigma T\cb^4 \over 3 \kappa P\cb } \nabla_{\rm ad} \approx L\di {P_{\rm d} \over P\cb}, % \left({P\di \over P\cb}\right)%^{1+\alpha}
-\end{equation} 
-where we drop the pressure dependence from the opacity law (\ref{eq:opacitylaw}) and define 
-\begin{equation} 
-L\di \equiv {64 \pi G M\cb \sigma T_{\rm d}^4 \over 3 \kappa(T_{\rm d}) P_{\rm d}} \nabla_{\rm ad}\chi^{4-\beta}, 
-\end{equation} 
-with $T = \chi T_{\rm d}$ at the radiative-convective boundary.
-
- For $\beta = 2$ and $F_\kappa = 1$ in the opacity law (\ref{eq:opacitylaw}), this gives the \citet{bell94} opacity for icy grains.
-By varying $F_\kappa$, dust depletion or enhancement can be considered.
-Grain properties affect both $F_\kappa$ and $\beta$, which generally satisfies  $1/2 \lesssim \beta \lesssim 2$ (aside from discontinuities across sublimation regions, see \citealt{semenov03}).
-
-To make further analytic progress, we ignore the self-gravity in the convective zone, holding the mass fixed at the core mass $M\co$.
-The density profile of an adiabatic atmosphere follows from hydrostatic balance as
+Both $\chi$ and $\theta$ depend on $\alpha$, $\beta$ and the adiabatic gradient $\delad$.
+For $\alpha=0$, $\beta=2$ and $\delad=2/7$, we find $\chi \approx 1.5$ and $\theta \approx 0.556$.
+Values for $\theta$ and $\chi$ for other choices of $\beta$ are summarized in \App{sec:analytic}, Table I.
+
+We then determine the density profile in the convective region of the atmosphere to estimate the mass of the atmosphere.
+Integrating the equation of hydrostatic balance (\ref{eq:structa}) and holding the mass fixed at $M\co$, we find
+
 \begin{equation}\label{eq:rhoconv} 
@@
 
-  Deep in the atmosphere, where $r \ll R\cb \lesssim R_B'$ the density profile is $\rho \propto r^{-1/(\gamma -1)}$.
-Since the energy scales as $\rho r^2 \propto r^{(2\gamma -3)/(\gamma - 1)}$, only polytropes with $\gamma < 3/2$ (i.e.
-$\gamma = 7/5$, but not $\gamma = 5/3$) have the bulk of energy at the bottom of the atmospheres.
-We will thus focus on the $\gamma = 7/5$ case, even though dissociation occurs deep in real protoplanetary atmospheres.
-
-The total (thermal and gravitational) energy in an adiabatic atmosphere could be evaluated from the virial theorem (see section \ref{cooling}).
-More simply, we use the result for temperature profiles in deep (but non-self-gravitating) convective regions:
+From this we can calculate the mass of the convective region as
+
+\begin{eqnarray} 
+\label{eq:Matman}
+M_{\rm atm} &=& 4 \pi \int_{R\co}^{R\cb} \rho r^2 dr \\
+&\approx& {5 \pi^2 \over 4} \rho\cb {R_B'}^{5/2} \sqrt{R\cb}, % \approx 0.54 \rho\cb {R_B'}^{5/2} \sqrt{r\cb} \\
+%&=& 0.54{R_B'^{3} \sqrt{\chi} \over  \sqrt{ \ln[P\cb/(\theta P\di)]}} \, .
+\end{eqnarray}
+in the limit $R\co \ll R\cb \ll R_B'$ and where $\gamma=7/5$ in the final expression.
+The mass in the isothermal region can be estimated as $M_{\rm iso} \sim 4 \pi \rho\cb r\cb^4/R_{\rm B}$ (see \App{sec:analytic} for a derivation).
+We see that $M_{\rm{iso}} \ll M_{\rm{atm}}$ and therefore it can be neglected.
+
+The total (thermal and gravitational) energy in an adiabatic atmosphere can be evaluated from the virial theorem (see section \ref{cooling}).
+More simply, we use the result for temperature profiles in deep, but non-self-gravitating, convective regions:
 \begin{equation}
@@
 \end{equation} 
-Thus the total energy for $\gamma < 3/2$ is thus
+The total energy follows as
 \begin{eqnarray} 
 E &=& - 4 \pi \nabla_{\rm ad} G M\co \int_{R\co}^{r\cb} \rho r dr \\
-&\approx& - 4 \pi P\cb {R_B'}^{1 \over \nabla_{\rm ad}} \left(\gamma-1 \over 3 - 2 \gamma\right)  R\co^{2\gamma-3\over \gamma-1}  \\
+&\approx& - 4 \pi P\cb {R_B'}^{1 \over \nabla_{\rm ad}} \left(\gamma-1 \over 3 - 2 \gamma\right)  R\co^{2\gamma-3\over \gamma-1}  \slabel{eq:Ean} \\ 
 &\approx& - 8 \pi P\cb {R_B'^{7/2} \over \sqrt{R\co}}
@@
 \end{eqnarray} 
-where the final expression takes $\gamma = 7/5$. % and then revert to the standard Bondi radius $R_B = G M\co/(\Rg T\di)$ in terms of the disk temperature.
-
-The mass of the adiabatic atmosphere is given by
-\begin{eqnarray} 
-\label{eq:Matman}
-M_{\rm atm} &=& 4 \pi \int_{R\co}^{R\cb} \rho r^2 dr \\
-&=& {5 \pi^2 \over 4} \rho\cb {R_B'}^{5/2} \sqrt{R\cb}, % \approx 0.54 \rho\cb {R_B'}^{5/2} \sqrt{r\cb} \\
-%&=& 0.54{R_B'^{3} \sqrt{\chi} \over  \sqrt{ \ln[P\cb/(\theta P\di)]}} \, .
-\end{eqnarray}
-in the limit $R\co \ll R\cb \ll R_B'$.
-From equation (\ref{eq:MatmISO}) and replacing $R_{\rm c}$ with $R\cb$, the mass in the isothermal region is given by $M_{\rm iso} \sim 4 \pi \rho\cb r\cb^4/R_B \ll M_{\rm atm}$, and can be neglected.
-
-We can eliminate $R\cb$ using \Eq{eq:PcbRcb}.  The ratio of atmosphere to core mass is then  
+when $\gamma = 7/5$. % and then revert to the standard Bondi radius $R_B = G M\co/(\Rg T\di)$ in terms of the disk temperature.
+
+We note that the approximation in the energy expression of equation (\ref{eq:Ean}) is valid only deep in the adiabatic atmosphere, where  $\rho \propto r^{-1/(\gamma -1)}$ from equation (\ref{eq:rhoconv}).
+As the energy scales as $\rho r^2 \propto r^{(2\gamma -3)/(\gamma - 1)}$, only polytropes with $\gamma < 3/2$ (i.e.
+$\gamma = 7/5$, but not $\gamma = 5/3$) have the bulk of energy at the bottom of the atmospheres.
+Equation (\ref{eq:Ean}) is therefore only valid when $\gamma<3/2$.
+% We will thus focus on the $\gamma = 7/5$ case, even though dissociation occurs deep in real protoplanetary atmospheres.
+
+Finally, we can estimate the luminosity that emerges at the radiative-convective boundary from equation (\ref{eq:structc}) assuming $\delrad=\delad$ at the radiative-convective boundary, as
+
+\begin{equation} \label{eq:Lcb}
+L\cb = {64 \pi G M\cb \sigma T\cb^4 \over 3 \kappa P\cb } \nabla_{\rm ad} \approx L\di {P_{\rm d} \over P\cb}, % \left({P\di \over P\cb}\right)%^{1+\alpha}
+\end{equation} 
+where we drop the pressure dependence from the opacity law (\ref{eq:opacitylaw}) and define 
+\begin{equation} 
+L\di \equiv {64 \pi G M\cb \sigma T_{\rm d}^4 \over 3 \kappa(T_{\rm d}) P_{\rm d}} \nabla_{\rm ad}\chi^{4-\beta}, 
+\end{equation}
+
+ For $\beta = 2$ and $F_\kappa = 1$ in the opacity law (\ref{eq:opacitylaw}), this gives the \citet{bell94} opacity for icy grains.
+By varying $F_\kappa$, dust depletion or enhancement can be considered.
+Grain properties affect both $F_\kappa$ and $\beta$, which generally satisfies  $1/2 \lesssim \beta \lesssim 2$ (aside from discontinuities across sublimation regions, see \citealt{semenov03}).
+
+%  The physical origin of this logarithm is the exponential increase of pressure with depth in the radiative zone.
+
+\subsection{Simplified Cooling Model}
+\label{coolingan}
+
+We use the equations obtained in section \ref{MELan} to estimate the atmosphere mass $M_{\rm{atm}}$ for which $M_{\rm{atm}} \sim M\co$ and runaway accretion commences.
+We then use a simplified version of the cooling model developed in section \ref{cooling} to estimate the time elapsed until collapse.
+
+We can eliminate $R\cb$ from equation (\ref{eq:Matman}) using \Eq{eq:PcbRcb}.  The ratio of atmosphere to core mass is then  
 \begin{equation} \label{eq:crit}
@@
 \end{equation} 
-For the atmosphere to initiate unstable gas accretion, $M_{\rm atm} = M\co$. We thus require
+For $M_{\rm atm} = M\co$ in equation (\ref{eq:crit}) we require
 \begin{equation} \label{eq:Pcbc}
-P\cb = \xi P_M
+P\cb = \xi P_M ,
 \end{equation} 
@@
 Numerical solutions exist for $\xi \ll 1$, but are unphysical as they imply $P\cb < P_{\rm d}$.
-%  The physical origin of this logarithm is the exponential increase of pressure with depth in the radiative zone.
-
-\subsubsection{Cooling Time}
-Neglecting the surface terms from equation (\ref{eq:coolingglobal}), the total time to cool the atmosphere from an initially fully adiabatic state is \footnote{The relevance of the neglected surface terms is discussed in \App{sec:analytic}}
+
+If we neglect the surface terms from equation (\ref{eq:coolingglobal}) \footnote{The relevance of the neglected surface terms is discussed in \App{sec:analytic}}, the total time elapsed until $M_{\rm{atm}}= M\co$ is given by
+
 \begin{eqnarray} 
 t_{\rm  cool} &=& -\int {d E \over L} = - \int_{P\di}^{P\cb } {d E/d P\cb \over L} dP\cb \\
-&\approx& 4 \pi {P\cb^{2} \over P\di} {R_B'^{7/2} \over L\di \sqrt{R\co}} \label{eq:tcool}\, ,
+&\approx& 4 \pi {P\cb^{2} \over P\di} {R_B'^{7/2} \over L\di \sqrt{R\co}} \label{eq:tcool}\, .
 %&\approx& {0.16 }{P\cb^{2} \over P\di} {R_B'^{7/2} \over L\di \sqrt{R\co}}
 \end{eqnarray} 
-where we neglect self-gravity and take $M\cb = M\co$ in $L\di$.
+%where we neglect self-gravity and take $M\cb = M\co$ in $L\di$.
 
@@
 for $\beta = 2$ and $\mc \equiv M_{\rm c} / M_{\oplus}$.  
-This timescale is too long compared to the average lifetime of a protoplanetary disk --- this is likely due to our neglect of self-gravity.
-Nevertheless, the trends are informative.
-A reduction in opacity gives faster cooling, as long as the optically thick assumption holds.
-Lower disk temperatures also decrease the crossover time;  even though higher temperatures result in higher luminosities, they also give higher dust opacities.
-For different $\beta$ values, the temperature dependence is $t_{\rm co} \propto F_T^{ \beta + 1/2}$.
-The cooling timescale only weakly depends on disk mass or pressure, via the  logarithmic factor $\xi$.
-This weak dependence is a consequence of cooling coming from the radiative-convective boundary.
+This timescale is too long compared to the average lifetime of a protoplanetary disk.
+We will show in section \ref{KH} that this is due to our neglect of self-gravity in the analytic model.
+Nevertheless, the trends in the atmosphere evolution predicted by our analytic model are informative.
+A reduction in opacity accelerates cooling as long as the optically thick assumption holds.
+Lower disk temperatures also decrease the crossover time --- even though higher temperatures result in higher luminosities, they also result in higher dust opacities.
+For different $\beta$ values, the crossover time depends on the disk temperature as $t_{\rm co} \propto F_T^{ \beta + 1/2}$.
+The cooling timescale depends on the disk pressure only weakly via the  logarithmic factor $\xi$.
+We show in section \ref{TPeffects} that the numerical model results in a qualitatively similar dependence on disk temperature and pressure.
+%This weak dependence is a consequence of cooling coming from the radiative-convective boundary.
 Note that the scaling laws do not reflect the changes to $\xi$ which remains order unity.
@@
 
-\subsubsection{Critical Core Mass}\label{sec:critmass}
+\subsection{Critical Core Mass}\label{sec:critmass}
 %We use the results of the previous section to derive the minimum core mass that can form a giant planet by gas accretion.
@@
 
-We can define a \textit{critical more mass} as the core mass for which $M_{\rm{atm}}=M_{\rm c}$ within a typical gas disk lifetime:
+We use the results of section \ref{coolingan} to estimate the core mass for which  runaway accretion commences, i.e. $M_{\atm}=M\co$, within a typical lifetime of a protoplanetary disk: 
 \begin{equation}
@@
 \end{equation} 
-From equation (\ref{eq:tcoolan}), the resulting critical core mass is
+We denote this core mass as the \textit{critical core mass} $M_{\rm crit}$. From equation (\ref{eq:tcoolan}), $M_{\rm{crit}}$ is given by
 \begin{equation}\label{eq:Mcrit}
@@
 \end{equation}
-Similarly to the crossover time, the critical mass is too large due to the neglect of self-gravity.  The scaling with disk properties is similar: all quantities are raised to the $3/5$ power.
+
+Similarly to the crossover time, the critical mass is too large due to our neglect of self-gravity.  The scaling with disk properties is similar: all quantities are raised to the $3/5$ power.
 
@@
 
+\subsection{Isothermal Atmosphere}
+\label{iso}
+
+We consider the structure of a low mass non self-gravitating isothermal atmosphere.
+We assume the atmosphere matches onto a constant background density, $\rho_{\rm d}$, at a distance $r_{\rm fit} = n_{\rm fit} \RB$, where $R_{\rm B}$ is the Bondi radius defined in equation (\ref{eq:RB}).
+From equation (\ref{eq:structa}) the resulting density profile is
+\begin{equation} \label{eq:rhoiso}
+\rho = \rho_{\rm d} \exp \left({R_{\rm B} \over r} - {1 \over n_{\rm{fit}}} \right) \approx   \rho_{\rm{d}} \exp \left(R_{\rm B} \over r  \right),
+\end{equation} 
+where the approximate inequality is valid deep inside the atmosphere ($r \ll \RB$) for any $n_{\rm fit} \gtrsim 1$.
+However, the choice of boundary condition does have an order unity effect on the density near the Bondi radius.
+
+The mass of the atmosphere is determined by integrating equation (\ref{eq:structb}) from the core to the Bondi radius using the density profile (\ref{eq:rhoiso}).
+Planets can attract massive atmospheres if $\theta\co \equiv R_B/R\co \gg 1$.
+In this case
+\begin{equation} \label{eq:MatmISO}
+M_{\rm iso} \approx 4 \pi \rho\di {R\co^4 \over R_B} e^{R_B/R\co} = 4 \pi \rho\di {R\co^3 \over \theta\co} e^{\theta\co}\, .
+\end{equation} 
+This result is the leading order term in a series expansion.
+Furthermore, because the atmospheric scale-height at $R\co$ is $H_\rho = |dr /d\ln\rho| = R\co^2/R_B$, the result is intuitively the correct order of magnitude.
+
+\subsection{Temperature and Pressure Corrections at the Radiative-Convective Boundary}
+\label{RCBcorr}
+
+%\subsection{Temperature Contrast at Convective Boundary}
+In this section we estimate the temperature and pressure corrections at the radiative-convective boundary due to the fact that the radiative region is not purely isothermal.
+We show that the temperature contrast between the temperature at the radiative-convective boundary $T_{\rm RCB}$ and the ambient disk $T_{\rm d}$, is modest.
+From equation (\ref{eq:delrad}), we express the radiative lapse rate
+\begin{equation}\label{eq:delradan}
+\delrad = {3 \kappa P \over 64 \pi  G M \sigma T^4} L = \nabla\di {(P/P_{\rm d})^{1+\alpha} \over (T/T_{\rm d})^{4-\beta}},
+\end{equation}
+
+\noindent where the second equality follows from the opacity law (\ref{eq:opacitylaw}), and $\nabla_{\rm d}$ is the radiative temperature gradient at the disk:
+
+\begin{equation}
+\label{eq:delo}
+\nabla \di \equiv \frac{3 \kappa(P{\di}, T{\di}) P{\di}}{64 \pi G M \sigma T_{\rm d}^4} L.
+\end{equation}
+
+\noindent The mass $M$ is the sum of the core mass and the atmosphere mass below the pressure level $P$.
+If the mass in the radiative zone is small, then we can hold $M$ fixed at the sum of the core and convective zone masses: $M = M\co + M_{\cb}$.
+With this assumption, we get a constant value for $\nabla_{\rm d}$.
+As $\delrad=d \ln T/d \ln P$, the temperature profile integrates to
+\begin{equation}\label{eq:radTP}
+\left(T \over T_{\rm d}\right)^{4-\beta} - 1 = {\nabla\di \over \nabla_\infty} \left[\left({P \over P_{\rm d}}\right)^{1+\alpha} - 1 \right] \, ,
+\end{equation} 
+where $\nabla_\infty = (1+\alpha)/(4-\beta)$ is the radiative temperature gradient for $T ,P \rightarrow \infty$.
+By applying \Eqs{eq:delradan}{eq:radTP} at the radiative-convective boundary (where $\delrad = \delad$) under the assumption that the pressure there $P\cb$ satisfies the relation $P\cb \gg P_{\rm d}$, we end up with the expression in equation (\ref{eq:Tcb}), $T\cb=\chi T\di$, with $\chi$ defined in equation (\ref{eq:chi}).
+
+ . %\begin{equation}\label{eq:Tcb}
+%\chi \equiv {T\cb\over T_{\rm d}} \simeq \left(1 - {\nabla_{\rm ad} \over \nabla_\infty}\right)^{-\frac{1}{4-\beta}} \, .
+%\end{equation} 
+%For  $\alpha=0$ and $\beta =2$,  and assuming $\delad = 2/7$, we find $\chi \approx 1.5$. Values of $\chi$ and $\nabla_\infty$ for other choices of $\beta$ are summarized in \App{sec:analytic}.
+
+%\footnote{Our definitions of $\nabla\di$ and $\nabla_\infty$ are precisely opposite to \citet{rafikov06}, but consistent with other works and the general convention of labeling a quantity $f$ in the disk as $f\di$.}
+
+The pressure at the convective boundary  follows from \Eqs{eq:radTP}{eq:Tcb} as
+\begin{equation}
+\label{eq:Pcbapprox}
+{P\cb\over P_{\rm d}} \simeq \left({\delad/\nabla\di \over 1 - \delad/\nabla_\infty}\right)^{1 \over 1 + \alpha}
+\end{equation} 
+This pressure contrast can be quite large since $\nabla_{\rm d} \ll 1$.
+
+ We now eliminate $\nabla\di$ from equation (\ref{eq:Pcbapprox}) to obtain a relation between temperature and pressure as a function of the pressure at the radiative-convective boundary $P\cb$.
+From equation (\ref{eq:radTP}), it follows that
+ \begin{equation}\label{eq:TP}
+{T \over T_{\rm d}} = \left\{1 + {1 \over {\nabla_\infty \over \delad} - 1} \left[ \left({P \over P\cb}\right)^{1+\alpha} -  \left({P_{\rm d} \over P\cb}\right)^{1+\alpha}  \right] \right\}^{1 \over 4-\beta}\, .
+\end{equation} 
+ We can then determine the radius of the convective boundary $R\cb$ from the hydrostatic balance equation (\ref{eq:structa}) as 
+\begin{equation}\label{eq:RCBint}
+{R_B \over R\cb} = \int_{P\di}^{P\cb} {T \over T_{\rm d}} {dP \over P}\, .
+\end{equation} 
+An isothermal atmosphere gives a simple logarithmic dependence on $P\cb$.  However, using \Eq{eq:TP} in the integral gives
+\begin{equation}\label{eq:Rcb}
+{R_B \over r\cb} = \ln \left(P\cb \over P\di \right) - \ln \theta \, ,
+\end{equation} 
+with an extra correction term, $\theta < 1$.  From this we arrive at the relation between $P\cb$ and $P\di$ given by equation (\ref{eq:PcbRcb}).
+
+% The form we chose for the correction term allows us to relate the disk and radiative-convective boundary pressures as :
+% \begin{equation}\label{eq:PcbRcb}
+%P\cb = \theta P_{\rm d} e^{R_B/R\cb} \, .
+%\end{equation}   
+%In the $P\cb \gg P_{\rm d}$ limit, the correction term is an order unity constant that depends on $\alpha$, $\beta$ and $\delad$.
+Similarly to the temperature correction factor $\chi$,  $\theta$ accounts for the fact that the radiative region is not perfectly isothermal.
+For $\delad=2/7$, $\alpha=0$ and $\beta=2$, we find $\theta \approx 0.556$.
+Values of $\theta$ for other choices of $\beta$ are summarized in \App{sec:analytic}.
+A simple analytic expression for $\theta$ is not possible.
+
 \subsection{Estimate of Atmosphere Mass Outside the Bondi Radius}

### 2013-07-01 python/Analytical/analytic_comp.py
repo update in progress

@@ -152,7 +152,7 @@ class AnSols:
 
     #Lcb = Lo * Pd / Pcb
 
-    def __init__(self, mu=2.35, a=100, FT=FT, **kwargs):
+    def __init__(self, mu=2.35, a=10, FT=FT, **kwargs):
         #disk
         self.Td = kwargs.get('Td', Tdisk(a, FT))
         self.Pd = kwargs.get('Pd', Pdisk(a, mstar, FSigma, FT))

### 2013-07-01 python/RadSGPoly/plot_helper.py

@@ -55,7 +55,7 @@ def t_vs_Mc_fixed_a(delad, Y, a, returnt = 0, opacity = kdust):
                   Cv = Cvfn(Y, delad), Pd = Pdisk(a, mstar, FSigma, FT), \
                   Td = Tdisk(a, FT), kappa = opacity)
 
-    if opacity == kdust:
+    if opacity == kdust or opacity == kdust10:
 
         if a == 1.0 or a  == 3.0 or a == 2.0 or a == 1.5 or a == 2.5 or a == 1.75:
             t = 0 * numpy.ndarray(shape = (5), dtype = float)
@@ -127,12 +127,15 @@ def t_vs_Mc_fixed_a(delad, Y, a, returnt = 0, opacity = kdust):
                 if delad == 2./7 and Y == 0.0:
                     Mc = numpy.linspace(11, 15, 5)
                 elif delad == 2./7 and Y == 0.3:
-                    Mc = numpy.linspace(1, 14, 14)               
+                    Mc = numpy.linspace(1, 14, 14)    
                 elif delad == 2./5 and Y == 0.3:
                     Mc = numpy.linspace(6, 10, 5)
 
             elif a == 10.0 and delad == 2./7 and Y == 0.3:
+                if opacity == kdust:
                     Mc = numpy.linspace(5, 14, 10)
+                else:
+                    Mc = numpy.linspace(1, 10, 10)
 
             else:
                 if delad == 2./7 and Y == 0.0:
@@ -153,7 +156,10 @@ def t_vs_Mc_fixed_a(delad, Y, a, returnt = 0, opacity = kdust):
             Mtotcrit[i] = temp[0].MB[-1]
 
             MBonditemp = temp[0].MB
-            dt = temp[-1]
+            if opacity == kdust:
+                dt = temp[-1]
+            else:
+                dt = temp[-1] / 10
             
             t[i] = sum(dt)
 

### 2013-07-01 python/RadSGPoly/profiles_poly.py

@@ -149,16 +149,16 @@ def profiles_write(n, nMpoints, L1, L2, Mmin, Mmax, filename, prms = prms, \
 
     if savefile == 1:
         if disk == 1:
-            if prms.kappa == kdust:
+            if prms.kappa == kdust or prms.kappa == kdust10:
                 paramfilename = userpath + '/dat_ana/MODELS/RadSGPoly/' + 'delad' + \
                                 str(prms.delad)[:4] + '/Y' + str(prms.Y) + '/' + \
                                 str(prms.a) + 'AU/' + filename
                 numpy.savez_compressed(paramfilename, model = model, param = param, \
                                    prof = prof)
-            elif prms.kappa == kdust10:
-                paramfilename = userpath + '/dat_ana/MODELS/RadSGPoly/' + 'delad' + \
-                                str(prms.delad)[:4] + '/Y' + str(prms.Y) + '/kdust10/' + \
-                                str(prms.a) + 'AU/' + filename
+##            elif prms.kappa == kdust10:
+##                paramfilename = userpath + '/dat_ana/MODELS/RadSGPoly/' + 'delad' + \
+##                                str(prms.delad)[:4] + '/Y' + str(prms.Y) + '/kdust10/' + \
+##                                str(prms.a) + 'AU/' + filename
                 numpy.savez_compressed(paramfilename, model = model, param = param, \
                                    prof = prof)
             elif prms.kappa == kdust100:
@@ -186,14 +186,14 @@ def profiles_write(n, nMpoints, L1, L2, Mmin, Mmax, filename, prms = prms, \
 
 def atmload(filename, prms = prms, disk = 1):
     if disk == 1:
-        if prms.kappa == kdust:
+        if prms.kappa == kdust or prms.kappa == kdust10:
             npzdat = numpy.load(userpath + '/dat_ana/MODELS/RadSGPoly/' + 'delad' + \
                             str(prms.delad)[:4] + '/Y' + str(prms.Y) + '/' + \
                             str(prms.a) + 'AU/' + filename)
-        elif prms.kappa == kdust10:
-            npzdat = numpy.load(userpath + '/dat_ana/MODELS/RadSGPoly/' + 'delad' + \
-                                str(prms.delad)[:4] + '/Y' + str(prms.Y) + '/kdust10/' + \
-                                str(prms.a) + 'AU/' + filename)
+##        elif prms.kappa == kdust10:
[…]

### 2013-07-02 tex/draft_paper2/draft_paper2.tex

@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/EOS/delad_S_mixt.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/delad_S_mixt.pdf}
 %%\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/EOSeffects/ortho_para_energy.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/ortho_para_energy.pdf}
 %%\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/EOSeffects/Ltplot_poly.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/Ltplot_poly.pdf}
 %%\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/EOSeffects/TErplot_poly.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/TErplot_poly.pdf}
 %%\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/EOSeffects/tplot.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/tplot.pdf}
 %%\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/EOSeffects/TLr_plot.pdf}
+\includegraphics[width=\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/TLr_plot.pdf}
 %%\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/EOSeffects/Er_plot.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/Er_plot.pdf}
 %%\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/t_vs_M_10au.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/t_vs_M_10au.pdf}
 %%\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/Mc_vs_a_poly_real_paper.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/Mc_vs_a_poly_real_paper.pdf}
 %%\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/acc_rates_paper.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/acc_rates_paper.pdf}
 %\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/Mc_vs_a_poly_raf2_paper.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/Mc_vs_a_poly_raf2_paper.pdf}
 %\vspace{-0.5in}
@@
 \centering
-\includegraphics[scale=.8]{../../figs/EOS/delad_S_H.pdf}
+\includegraphics[scale=.8]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/delad_S_H.pdf}
 \caption{Contour plot of the adiabatic gradient $\delad$ for the hydrogen extended table. The black curves represent constant entropy curves.}
@@
 \centering
-\includegraphics[scale=.8]{../../figs/EOS/delad_S_He.pdf}
+\includegraphics[scale=.8]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/delad_S_He.pdf}
 \caption{Contour plot for the adiabatic gradient $\delad$ for the helium extended table. The black curves represent constant entropy curves.}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/EOSeffects/delad_ionization.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/delad_ionization.pdf}
 %%\vspace{-0.5in}

### 2013-07-02 tex/draft/draft3.tex
revised section 2 (in draft3)

% !TEX TS-program = pdflatexmk

%\documentstyle[aas2pp4,epsf]{article}
%%%\documentstyle[aaspp4,epsf]{article}
%\documentstyle[12pt,aasms]{article}    % this is for a preprint
%(single-spaced)
%\documentstyle[aaspp4,epsf]{article} % this is for small print
%\documentstyle[12pt, aaspp4]{article}

%\documentstyle[11pt,aaspp]{article}
%\documentclass[12pt, preprint]{aastex}

%\documentclass[manuscript]{aastex}
\documentclass[apj, numberedappendix]{emulateapj}

%\documentclass[12pt, preprint,numberedappendix]{emulateapj}
%\documentstyle[12pt,aasms]{article}    % this is for submittal
                                       % (double-spaced)

%\documentstyle[12pt,aasms]{article}   \usepackage{emulateapj5}

\usepackage{graphicx} 
\usepackage{graphics}                       
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{subeqnarray}
%\bibliographystyle{apj}

\newcommand{\emgr}[1]{\emph{ \color{gray} #1}}

\newcommand{\ie}{i.e.\ }
\newcommand{\eg}{e.g.\ }
\newcommand{\p}{\partial}
\newcommand{\xv}{\vc{x}}
\newcommand{\kv}{\vc{k}}
\newcommand{\brak}[1]{\langle #1\rangle}

\newcommand{\gcc}{\;\mathrm{g\; cm^{-3}}}
\newcommand{\gsc}{\;\mathrm{g\; cm^{-2}}}
\newcommand{\cm}{\; {\rm cm}}
\newcommand{\mm}{\; {\rm mm}}
%\newcommand{\ps}{\; {\rm s^{-1}}}
\newcommand{\km}{\; {\rm km}}
%\newcommand{\au}{\; \varpi_{\rm AU}}

\newcommand{\AU}{\; {\rm AU}}
\newcommand{\yr}{\; {\rm yr}}
\def\K{\; {\rm K}}

\newcommand{\vcs}[1]{\mbox{\boldmath{$\scriptstyle{#1}$}}}
\newcommand{\vc}[1]{\mbox{\boldmath{$#1$}}}
\newcommand{\nab}{\vc{\nabla}}
\DeclareMathSymbol{\varOmega}{\mathord}{letters}{"0A}
\DeclareMathSymbol{\varSigma}{\mathord}{letters}{"06}
\DeclareMathSymbol{\varPsi}{\mathord}{letters}{"09}

\newcommand{\eq}[1]{equation\,(\ref{#1})}
\newcommand{\Eq}[1]{Equation\,(\ref{#1})}
\newcommand{\Eqs}[2]{Equations (\ref{#1}) and~(\ref{#2})}
\newcommand{\Eqss}[2]{Equations (\ref{#1})--(\ref{#2})}
\newcommand{\Eqsss}[3]{Equations (\ref{#1}), (\ref{#2}) and~(\ref{#3})}
\newcommand{\App}[1]{Appendix~\ref{#1}}
\newcommand{\Sec}[1]{Sect.~\ref{#1}}
\newcommand{\Chap}[1]{Chapter~\ref{#1}}
\newcommand{\Fig}[1]{Fig.~\ref{#1}}
\newcommand{\Figs}[2]{Figs.~\ref{#1} and \ref{#2}}
\newcommand{\Figss}[2]{Figs.~\ref{#1}--\ref{#2}} 
\newcommand{\Tab}[1]{Table \ref{#1}}

\newenvironment{packed_item}{
\begin{itemize}
  \setlength{\itemsep}{1pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}
}{\end{itemize}}

\newcommand{\delad}{\nabla_{\rm ad}}
\newcommand{\delrad}{\nabla_{\rm rad}}
\newcommand{\Rg}{\mathcal{R}}
\newcommand{\RB}{R_{\rm B}}
\newcommand{\RH}{R_{\rm H}}
\newcommand{\co}{_{\rm c}}
\newcommand{\pla}{_{\rm pl}}
\newcommand{\di}{_{\rm d}}
\newcommand{\cb}{_{\rm RCB}}
\newcommand{\mc}{m_{\rm c \oplus}}
\newcommand{\mcn}[1] { m_{ \rm c #1 \oplus} }
\newcommand{\MC}{M_{\rm crit}}
\newcommand{\au}{a_\oplus}
\newcommand{\aun}[1]{ a_{#1\oplus} }

\begin{document}
\bibliographystyle{apj}

\shortauthors{Piso \& Youdin}

\title{On the Minimum Core Mass for Giant Planet Formation}

\author{Ana-Maria A. Piso}
\affil{Harvard-Smithsonian Center for Astrophysics}

\author{Andrew N. Youdin}
\affil{JILA, University of Colorado at Boulder}

\begin{abstract}
%The core accretion model proposes that giant planets form by the accretion of gas onto a solid protoplanetary core. Previous studies have found that there exists a ``critical core mass'' past which hydrostatic solutions can no longer be found and unstable atmosphere collapse occurs. In standard calculations of the critical core mass, planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse. In this study we consider the extreme case in which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the planet. We develop a two-layer atmosphere model with an inner convective region and an outer radiative zone that matches onto the protoplanetary disk, and we determine the minimum core mass for a giant planet to form within the typical disk life timescale for a variety of disk conditions, which we denote as  ``critical core mass''.  We find that the absolute minimum core mass required to nucleate atmosphere collapse within the disk lifetime is smaller for planets forming further away from their host stars. Moreover, the critical core mass is strongly dependent on disk temperature, opacity and mean molecular weight of the gas. Our results yield lower mass cores than corresponding studies for large planetesimal accretion rates. We therefore show that it is easier to form a planet by growing the core first, then accreting a massive gaseous envelope, rather than forming the core and atmosphere simultaneously.

\end{abstract}

\section{Introduction}
\label{intro}

%Current theories of giant planet formation postulate that these planets form either through core accretion (refs), in which solid planetesimals collide and grow into a massive solid core, which then accretes a gaseous envelope, or due to a gravitational instability in the protoplanetary disk that leads to fragmentation of the disk into self-gravitating clumps (refs, inc. \citealt{dangelo11}).  %quote murray-clay, kratter 10, rafikov 05, etc.
%
%Standard core accretion models (refs) assume that the core and the atmosphere grow at the same time, and that planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse, while the heat generated by the gravitational (Kelvin-Helmholtz) contraction of the atmosphere is neglected. These studies consider that the planet atmosphere is in steady state, in which all the luminosity due to planetesimal accretion is radiated away by the envelope, and  find that there exists a minimum (``critical'') core mass past which hydrostatic solutions can no longer be found and unstable atmosphere collapse occurs. 
%
%Forming giant planets at wide separations in the disk poses theoretical challenges. On the one hand, gravitational instability generates objects that are too massive to explain the current observed properties of exoplanets (refs, inc. \citealt{rafikov05}). On the other hand, planetesimal accretion is slow at large distances in the disk, and therefore large cores may not be able to form before the dissipation of the disk (refs). I would therefore be easier if giant planets could form from smaller cores which would need less time to grow. 
%
%In this study, we show that giant planets can grow faster from small protoplanetary cores that are fully formed before significant gas accretion occurs. In this scenario, the planetesimal accretion rate is significantly slowed down during the gas contraction phase of the atmosphere. This reduction can arise due to dynamical clearing, or due to the core having formed in the inner parts of the disk and migrated outwards, etc. In this situation, the atmosphere evolution is dominated by the Kelvin-Helmholtz contraction of the envelope. The atmosphere is no longer in a steady state, but rather it accretes gas as it loses energy through radiation. 
%
%In our model we therefore assume that the luminosity evolution of the atmosphere is dominated by gas contraction, while the planetesimal accretion rate is negligible. As a result, the protoplanetary core has a fixed mass. We consider that the atmosphere evolves in time through stages of quasi-static equilibrium. Once the mass of the gaseous envelope becomes comparable to the mass of the solid core, the self-gravity of the atmosphere can no longer balance the pressure gradient and unstable hydrodynamic collapse commences. The time required for the atmosphere to grow to this stage is the characteristic growth time of the atmosphere. For a set of fixed gas and disk conditions, there exists a minimum core mass for which the atmosphere can grow on the time scale described above within the life time of the protoplanetary disk, which we define as the ``critical core mass''. 
%
%We develop a two-layer atmosphere model, with a convective inner region and a radiative outer region that matches smoothly on to the protoplanetary disk, and develop a cooling model that evolves the atmosphere in time. We aim to find the critical core mass for a giant planet to form before the dissipation of the disk.

Kelvin-Helmholz (KH) contraction considered previously (Ikoma, PapNel05).  Say why useful.  This paper develops a simplified model of KH contraction to both elucidate the relevant physical processes and explore a wide range of disk parameter space...

This paper is organized as follows. In section \ref{sec:model} we describe the assumptions of our atmosphere model, and derive the basic equations that govern the structure and evolution of the atmosphere. In section \ref{sec:coolingan}, we present a simplified analytic model that predicts the qualitative behavior of the numerical model. Results for atmospheric structure and evolution are presented in section \ref{sec:KH}, and implications for the critical core mass are presented in section \ref{sec:critical}.  The discussion in section \ref{sec:neglected} addresses various approximations and neglected effects.  We summarize our findings in section \ref{sec:conclusions}.  The appendices contain...

%Some previous studies of atmosphere accretion (e.g., \citealt{stevenson82}, \citealt{wuchterl93}, \citealt{rafikov06}) consider static envelopes, in which the luminosity is solely supplied by planetesimal accretion and fully radiated away by the atmosphere. In other studies, the time evolution is explicitly taken into account and full time dependent models are developed (e.g., \citealt{ikoma00}). We follow an intermediate approach and consider quasi static evolution. Our model for the atmosphere growth time is described in section \ref{...}.

 %The simplified treatment allows us to explore and better understand  the effect of crucial parameters including core mass, disk properties and opacity.

\section{Atmosphere Models} \label{sec:model}

To model the growth of planetary atmospheres around a solid core, we develop a simplified two layer model for time-dependent atmospheric cooling, i.e. Kelvin-Helmholtz (KH) contraction.  With a convective interior and radiative exterior, this model is motivated by similar models of hot Jupiters \citep{ab06, ym10}.

Our model can accurately model the growth of the atmosphere up to the crossover mass, when the atmosphere mass equals the mass of the core.   Beyond the crossover mass,  our approximate treatment of the radiative zone breaks down.  Since subsequent growth is a rapid runaway process \emph{(cite a numerical core accretion paper)}, our model can investigate to good accuracy the timescale requirement of core accretion.  Our simplified treatment is also inappropriate for hot, short-period planets, where dust sublimation gives deeper radiative zones that require more detailed models.

Our main assumptions are summarized as follows:
\begin{enumerate}
\item The atmosphere is spherically symmetric and remains in hydrostatic balance during its thermal evolution.
\item The core mass and radius are fixed in evolutionary calculations, neglecting ongoing planetesimal or dust accretion.
\item At the planet's Hill radius, the atmospheric temperature and pressure match the conditions of the disk midplane.
\item The only source of planetary luminosity is the gravitational contraction of the atmosphere.  
\item  In the radiative zone, luminosity generation is neglected, i.e. the luminosity is held constant.
\item A global cooling model connects independent static solutions into a time-dependent sequence.
\item A polytropic EOS is assumed for simplicity.
\item Dust grains provide the opacity in the radiative zone, which remains cool enough to avoid dust sublimation.
\end{enumerate}

The remainder of this section develops our model in more detail.

\subsection{Disk and Opacity Model}\label{sec:disk}

We adopt a minimum mass solar nebula (MMSN) model for a passively irradiated disk \citep{chiang10}. With the semi-major axis $a$ normalized to the outer disk as $\aun{10} = a/(10 \text{ AU})$, the gas surface density and mid-plane temperature are  
\begin{subeqnarray} \label{eq:diskparam}
\varSigma\di  &=& 70 \,F_\varSigma \aun{10}^{-3/2} ~{\rm g~cm}^{-2} \\
T\di &=& 45  \,F_T\, \aun{10}^{-3/7} ~{\rm K} \, .
%\Sigma\di&=&2200 F_{\Sigma} a^{-3/2}\,\, \text{g cm}^{-2} \slabel{eq:diska}\\
%T\di &=& 120 F_T a^{-3/7} \, \text{K}, \slabel{eq:diskb}
\end{subeqnarray}
The normalization factors $F_{\Sigma}$ and $F_T$  adjust the model relative to the fiducial MMSN.  We fix $F_{\varSigma}=F_T=1$ unless noted otherwise.

For a vertically isothermal disk in hydrostatic balance (with no self-gravity), the mid-plane pressure of disk gas is 
\begin{equation}
\label{eq:Pd}
P\di = 6.9 \times 10^{-3} F_\varSigma \sqrt{F_T} \, \aun{10}^{-45/14}~{\rm dyne~cm^2}
%P\di=1.1 \times 10^{-4} F_{\Sigma} \sqrt{F_T} a^{-45/14} \,\, \text{dyne cm}^{-2} %\times \sqrt{m_\ast}
\end{equation}
for a molecular weight of $\mu=2.35$ proton masses and a Solar mass star.  %Changes to the stellar mass (not considered here) would affect both the dynamic mass and, by heating the disk, $T\di$.
%The fiducial pressure is a paltry 7 nanobars.

The (thermodynamically isothermal) sound speed in the disk is
\begin{equation}
c\di = \sqrt{\Rg T\di} = 0.4 \sqrt{F_T} \aun{10}^{3/14} ~\text{km s}^{-1}
\end{equation}  
in terms of the specific gas constant $\Rg$.  The disk scale-height is 
\begin{equation}
H\di = {c\di / \varOmega} = 0.42 \sqrt{F_T}  \, \aun{10}^{9/7} \AU\, .
\end{equation} 
in terms of the Keplerian frequency $\varOmega = \sqrt{G M_\ast/a^3}$ with $G$ the gravitational constant and $M_\ast$ the stellar (in this work Solar) mass.

We assume a dust opacity following \citet{bell94}:
\begin{equation}
\label{eq:opacitylaw}
\kappa= 2 F_\kappa  \left(\frac{T}{100\; \rm{K}}\right)^{\beta} \; \mathrm{cm^2 ~ g^{-1}},
\end{equation}
with a powerlaw index $\beta = 2$ and normalization $F_\kappa = 1$ unless noted otherwise.  Roughly speaking, grain growth will lower both $F_\kappa$ and $\beta$, while dust abundance scales with $F_\kappa$.   Section \ref{sec:opEOS} discusses dust sublimation and more realistic opacity laws.

\subsection{Length Scales}
\label{sec:scales}

The characteristic length scales for protoplanetary atmospheres are crucial for choosing boundary conditions and for understanding the validity of  spherical symmetry in a disk of scaleheight $H\di$.  The radius of the solid core:

\begin{equation}
\label{eq:rc}
R\co \equiv \left(\frac{3 M\co}{4 \pi \rho\co}\right)^{1/3} \approx 10^{-4} \mcn{10}^{1/3} ~\text{AU},
\end{equation}
where the core mass, $M\co$, is normalized to 10 Earth masses as $\mcn{10} \equiv M\co/(10~M_\oplus)$. The core density is held fixed at $\rho\co=3.2$ g cm$^{-3}$.  We thus neglect  the detailed equation of state of the solids core \citep{fortney07}.

A planet can bind a dense atmosphere if the escape speed exceeds the sound speed.  This criterion is satisfied inside the Bondi radius
\begin{equation}
\label{eq:RB}
\RB \equiv \frac{G M\pla}{c\di^2} \approx 0.17 \, {\mcn{10}  \, \aun{10}^{3/7} \over F_T} ~\AU
\end{equation}
where the planet mass, $M\pla = M\co + M_\mathrm{atm}$, includes the mass of the core and atmosphere (defined as lying within the Bondi radius).  The numerical estimates above and in \Eq{eq:RHill} assume the core mass dominates, $M\pla \sim M\co$.

Stellar tides dominate the planet's gravity beyond the Hill radius
\begin{equation}
\label{eq:RHill}
R_{\rm H} = \left(M\pla \over 3 M_\ast \right)^{1/3}a \approx 0.22 \, {\mcn{10}^{1/3} \, \aun{10} }~\AU
\end{equation}
where hydrostatic balance clearly breaks down.

The relevant length scales of the atmosphere and disk satisfy the relation $\RB H\di^2 = 3 R_{\rm H}^3$.  The lengthscales are roughly equal at the ``thermal mass''
\begin{equation}
M_{\rm th} > {c\di^{3} \over G \varOmega} \approx 25 \, {F_T^{3/2} \over \sqrt{m_\ast} } \, \aun{10}^{6/7}~ M_\oplus \, .
\end{equation}

In the low mass regime, $M\pla < M_{\rm th}/\sqrt{3}$, the lengthscales order as $\RB< \RH<H\di$ (R06).  In this regime, many studies assume the atmosphere matches the disk conditions at $\RB$.  We however use $\RH$ as the matching radius in both this low mass and other higher mass regimes.  This choice is justified by the fact that, for hydrostatic solutions, the  density at $\RB$ exceeds the disk's background density by an order unity factor (R06).  This modest density change has a similarly modest effect on our results.

For a finite range of intermediate masses, $M_{\rm th}/\sqrt{3} < M\pla < 3 M_{\rm th}$, the Hill radius is the smallest scale, satisfying both $\RH < \RB$ and $\RH < H\di$.  Spherical symmetry remains a good, if imperfect, approximation because the bound atmosphere within $\RH$ doesn't see the vertical stratification of the disk on scales $\gtrsim H\di$.

At higher planet masses where $M\pla > 3 M_{\rm th}$ and $H\di < \RH < \RB$, spherical symmetry is no longer a good approximation, due to both the vertical stratification of the disk and gap opening.   See \S\ref{sec:hydro} for discussion of neglected non-hydrostatic effects on all mass scales.

Note that while $\RH$ is the outer boundary of our structure calculations, we define planet masses to include only the mass inside the smaller of $\RB$ or $\RH$.    This conservative choice in quoting planet masses is usually a minor distinction because (when $\RB < \RH$) the gas between $\RB$ and $\RH$ is weakly compressed.

\subsection{Structure Equations and Boundary Conditions}
\label{sec:struct}

Our atmosphere calculations use the standard structure equations of mass conservation, hydrostatic balance, thermal gradients, and energy conservation:
\begin{subeqnarray}
\label{eq:struct}
\frac{dm}{dr}&=&4 \pi r^2 \rho\slabel{eq:structb} \\
\frac{dP}{dr}&=&-\frac{G m}{r^2}\rho \slabel{eq:structa} \\
\frac{dT}{dr}&=&\nabla \frac{T}{P}\frac{dP}{dr}\slabel{eq:structc} \\
\frac{dL}{dr}&=&4 \pi r^2 \rho \left(\epsilon - \left. T {\partial S \over \partial t} \right|_m \right)\slabel{eq:structd}, 
\end{subeqnarray}
\noindent where $r$ is the radial coordinate, $P$, $T$, $\rho$  and $L$ are the gas pressure, temperature, density and luminosity, respectively.  The enclosed mass  at radius $r$ is $m$. \Eq{eq:structc} simply defines the temperature gradient  $\nabla \equiv d \ln T/d \ln P$.  In radiative zones radiative diffusion gives a temperature gradient
\begin{equation}
\label{eq:delrad}
\delrad \equiv \frac{3 \kappa P}{64 \pi G m \sigma T^4} L,
\end{equation}
where $\sigma$ is the Stefan-Boltzmann constant.  In convectively unstable regions, efficient convection gives an isentropic temperature gradient with $\nabla = \delad$, the adiabatic gradient. According to the Schwarzschild criterion, convective instability occurs when $\delrad > \delad$.  Thus $\nabla = \min(\delrad, \delad)$ set the temperature gradient.

In the energy equation (\ref{eq:structd}), $\epsilon$ represents all local sources of heat input, which excludes the motion of the atmosphere itself.  In stars, nuclear burning contributes to $\epsilon$.  In a protoplanetary atmosphere, dissipative drag on planetesimals contributes to $\epsilon$.  Our simplified models set $\epsilon = 0$, consistent with our neglect of planetesimal accretion luminosity at the base of the atmosphere.  The $\epsilon_{\rm g} = -T \partial S / \partial t$ term gives the energy input from gravitational contraction.\footnote{In general, any motion in a non-stationary atmosphere is accounted for by this term.  The partial time derivative is performed on shells of fixed mass.}  The partial time derivative would normally require our radial derivatives should be partial derivatives.  However our subsequent developments will replace the local energy equation (\ref{eq:structd}) with global energy balance, reverting the structure equations to time-independent ordinary differential equations (ODEs).

To solve the equation set (\ref{eq:struct}) an equation of state (EOS) is required for closure. In our study, we adopt an ideal gas law with a polytropic EOS 
\begin{subeqnarray}
P &=& \rho \Rg T \, ,\slabel{eq:idealgas} \\
P &=&K \rho^{\gamma} \, , \slabel{eq:polyEOS}%\equiv K \rho^{1/(1-\delad)}, 
\end{subeqnarray}
where $K$ is the adiabatic constant. The adiabatic index  $\gamma = 1/(1 - \delad)$.   An ideal monatomic gas has $\delad = 2/5$.  This work uses $\delad=2/7$ for an ideal diatomic gas.  While our reference mean molecular weight ($\mu = 2.35$ proton masses) includes Helium, we ignore Helium's effect on the EOS, which is already greatly simplified. The second law or thermodynamics gives the relative entropy as
\begin{equation}
S = \Rg \ln \left(T^{1/\delad} \over P \right) 
\end{equation} 
eliminating the need for $K$.

Boundary conditions must be satisfied at both the base and the top of the atmosphere with $m(R\co) = M\co$, $T(\RH) = T\di$ and $P(\RH) = P\di$.  In principle our solutions describe atmospheres with $L(R\co) = 0$. In practice, since we do not directly integrate \Eq{eq:structd} we need not directly impose this boundary condition, as described in \S\ref{sec:twolayer}.

\subsection{Global Cooling of an Embedded Planet}\label{cooling}

This section describes the global energy balance of a planet embedded in a gas disk, or more generally any spherical, hydrostatic object in pressure equilibrium with a background medium.  The total atmospheric energy includes gravitational and internal energies, $E = E_G + U$:
\begin{subeqnarray}
E_G&=&-\int_{M\co}^M \frac{G m}{r} dm \, , \label{eq:Eg} \\
U&=&\int_{M\co}^M u dm \, .\slabel{eq:U}
\end{subeqnarray}
The specific internal energy $u = C_V T = \Rg (\delad^{-1} -1) T$ for a polytropic EOS.  For a star or coreless planet, $L\co = M\co = 0$.

We start with the global energy balance for an isolated planet with a free surface:
\begin{equation}
\label{eq:coolingstar}
L_M = L\co + \Gamma - \dot{E}.
\end{equation}
The surface luminosity, $L_M$, includes contributions from the core luminosity $L\co$ --  e.g.\ planetesimal accretion or radioactive decay -- from the total heat generation $\Gamma$ -- given by the integral of $\epsilon$ over the object -- and from the rate of change of atmospheric energy $\dot{E}$, a loss term.

For an object with no core luminosity (or no core) and no internal heat sources, the energy equation $L_M = -\dot{E}$ describes KH contraction in its simplest form.  For a main sequence star $L_M = \Gamma$ as the internal heat of nuclear burning supplies the total luminosity.

For a protoplanetary atmosphere embedded in a gas disk, the full energy equation, 
\begin{equation}
\label{eq:coolingglobal}
L_M=L\co+\Gamma-\dot{E}+e_M \dot{M} - P_M \left. \frac{\partial V_M}{\partial t}\right|_M \, ,
\end{equation}
acquires surface terms as derived in  \App{sec:globalderiv}.  The energy accreted across the surface is given by the specific energy $e_M = u_M-G M/R$ and the mass accretion rate $\dot{M}$.  The work done by the surface is $P_M \partial V_M/ \partial t$, with the partial derivative performed at fixed mass.  This generalized energy equation applies on any spherical shell where hydrostatic balance holds.\footnote{\Eq{eq:coolingglobal} also applies in the interior of objects with a free surface.  The work term, $P_M \p V_M/\p t$, vanishes at the free surface.  The accretion energy, $e_M \dot{M}$,  vanishes for a truly isolated object, but would be included to account for accretion onto (or mass loss from) an otherwise free surface.}   Thus $M$ no longer refers to a uniquely defined total mass, but to the chosen mass level, where the instantaneous radius is $R$.  Values on this shell are labelled by $M$ subscripts.

For static solutions, which are not the focus of this Paper, the surface terms (and also $\dot{E})$ vanish.  Static solutions are valid when imposed heat sources, i.e.\ $L\co$ and $\Gamma$ exceed the atmospheric losses.  Quantitatively, static solutions apply when the evolutionary timescale exceeds  the KH timescale,
\begin{equation}
\tau_{\rm KH} \sim {|E| \over L_M}, 
\end{equation} 
where surface terms are assumed to be subdominant.  Thus evolutionary calculations, including the quasistatic calculations of this Paper, are needed to consider the fastest possible evolution that occurs on $\tau_{\rm KH}$.

\subsection{The Two-Layer Model} \label{sec:twolayer}

To simplify our calculations of atmospheric contraction we use a two layer model with a bottom convective region and an upper radiative layer.   The existence of such a structure is well known from previous studies (R06) and can be readily understood.  Before the protoplanetary atmosphere can cool, it has the entropy of the disk.  As the atmosphere cools the deep interior remains convective.  Convective interiors are a common feature of low mass cool objects (brown dwarfs and planets) that results from the behavior of $\delrad$ for realistic opacity laws.  However the entropy of the deep interior decreases as the atmosphere cools.  A region of outwardly increasing entropy, i.e.\ a radiative layer, is required to connect the convective interior to the disk.  A more complicated structure, with radiative windows in the convection zone, is possible as discussed in \S\ref{sec:opEOS}.

In convective regions, the adiabatic structure is independent of luminosity and can be calculated without local energy balance, \Eq{eq:structd}.  Thus for fully convective objects, a cooling sequence can be established by connecting a series of adiabatic solutions using a global energy equation, $L_M = -\dot{E}$ or \Eq{eq:coolingstar}.  Such methods are commonly used for their computational efficiency and are sometime referred to as ``following the adiabats," since the steady state solutions evolve in order of decreasing entropy \citep{marleau13}.

In the radiative zone,  local energy balance, \Eq{eq:structd}, does affect the atmospheric structure.  We proceed by assuming that the majority of energy is lost from the convective interior, and thus the luminosity can be treated as constant in the outer radiative zone.  With this approximation we can construct solutions from equations (\ref{eq:struct}a -- c) that ``follow the mass," i.e.\ gradually increase the atmospheric mass.  We then use the global energy balance, \Eq{eq:coolingglobal} to connect these solution in a cooling sequence.  The validity of neglecting luminosity generation in the radiative zone can be checked \emph{a posteriori}.

To make a single atmosphere model (indexed by $i$) we choose a planet mass $M_i$.  At the outer boundary, at $\RH(M_i)$, the temperature and pressure are set to the disk values.  To integrate equations (\ref{eq:struct}a--c), the luminosity is required to compute $\delrad$.  The correct value of the luminosity is the eigenvalue of the problem, which we find by the shooting method.  Only for the eigenvalue of luminosity does the integrated value of mass at the core, $m(R\co)$, match the actual core mass, $M\co$.

To establish the time difference between neighboring solutions, we apply \Eq{eq:coolingglobal} at the radiative-convective boundary (RCB) of the solutions.  Using exact solutions, energy balance could be evaluated at any level.  The approximation of constant luminosity in the radiative zone makes the RCB  the self-consistent choice.   After setting $\Gamma = L\co = 0$, we solve for the elapsed time $\Delta t$  between states $i$ and $i +1$ by finite differencing to give
\begin{equation}
\label{eq:dti}
\Delta t = %\left(
{ -\Delta E + \brak{e} \Delta M - \brak{P}  \Delta V_{\brak{M}} \over \brak{L} }\, .
\end{equation} 
Brackets indicate an average of, and Delta ($\Delta$) indicates a difference between, neighboring solutions.  All values are evaluated at the RCB.  Due to the partial derivative in \Eq{eq:coolingglobal}, the volume difference $\Delta V_{<M>}$ is performed at fixed mass, here the average of the masses at the RCB.

\section{Analytic Cooling Model}
\label{sec:coolingan}

\section{Quasi-Static Kelvin-Helmholtz Contraction}
\label{sec:KH}

\section{Critical Core Mass}
\label{sec:critical}

\section{Neglected Effects}\label{sec:neglected}

\subsection{Hydrodynamic Effects}\label{sec:hydro}
The neglect of hydrodynamical effects in our model is best discussed in terms of the thermal mass, $M_{\rm th}$, and the lengthscales introduced in \S\ref{sec:scales}.  In the low mass regime, $M\pla < M_{\rm th}/ \sqrt{3}$, where $\RB < \RH$ we assume that hydrostatic balance holds out to the outer boundary at $\RH$.   In this low mass regime, \citet[]{Orm13} calculated the flow patterns driven by stellar tides and disk headwinds.     On scales $\gtrsim \RB$ the flows no longer circulate the planet: they belong to the disk.  Nevertheless, the density structure outside $\RB$ remains spherically symmetric and hydrostatic.  Even if these flows do not destroy hydrostatic balance, they could affect the planet's cooling.  We expect such effects to be weak, as heat losses at greater depths dominate planetary cooling, but more study is needed.

At higher masses, non-hydrostatic effects become more severe.  At $M\pla \gtrsim M_{\rm th}$ planets can open significant gaps \citep{zhu13}.  At yet higher masses accretion instabilities could occur \citep{AylBat12}.  However in this high mass regime we already know that the spherically symmetric approximation breaks down, as does our approximate cooling model (\emph{discuss elsewhere and cite}).

Thus by restricting our attention to low masses, neglected hydrodynamic effects should be minor.   Moreover since $M_{\rm th} \propto a^{6/7}$ increases with disk radius, spherical hydrostratic models like ours have a greater range of applicability in the outer regions of disks.

\subsection{Realistic Opacities and EOS}\label{sec:opEOS}
\emph{some comments on sublimation and radiative windows here (current text rough). upshot is the outer disk is good.}

Real dust opacities exhibit a more complicated behavior that depends on grain composition.  Opacities drop by order unity when ice grains sublimate for $T \gtrsim 150$ K and they drop by orders of magnitude when silicate grains evaporate above $T \gtrsim 1500$ K \citep{semenov03, FerAle05}.  While protoplanetary atmospheres get significantly hotter than 1500 K, our model depends on opacity only in the upper radiative zones, which are cool enough to remain dusty.  For our opacity approximations to be valid, we must restrict our model to the lower temperatures of the outer disk.

A second radiative zone inside the convection zone, a ``radiative window," is possible.  Dust sublimation at shallow depths -- giving a low $\kappa$ and $P$ in \Eq{eq:delrad} -- could cause such a window.   Our two layer model ignores such complications.

\section{Conclusions} \label{sec:conclusions}

\bibliographystyle{apj}
\bibliography{refs}

\appendix
\section{Derivation of the Global Energy Equation}\label{sec:globalderiv}

To derive  the global energy equation (\ref{eq:coolingglobal}) for an embedded protoplanets, we generalize the analogous calculations in stellar structure theory, e.g.\ in \S4.3 of \citet{kippenhahn90}.  For our problem we add the effects of finite core radius, surface pressure and mass accretion.We start with the local energy equation (\ref{eq:structd}) whose more natural form in Lagrangian (mass) coordinates is $\p L/ \p m = \epsilon - T \p S /\p t$.  Integrating from the core to a higher shell with enclosed mass $M$ gives:
\begin{subeqnarray}
L - L\co &=& \int_{M\co}^M {\p L \over \p m} dm \\
&=& \int_{M\co}^M \left(\epsilon - T {\p S \over \p t} \right)dm \\
&=& \Gamma  - \int_{M\co}^M{\p u \over \p t} dm +  \int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm\slabel{eq:DLc}\, .
\end{subeqnarray} 
with $\Gamma = \int \epsilon dm$ the integral of the direct heating rate and applying the law of thermodynamics in the final step.

The global energy equation is derived by eliminating the partial time derivatives in \Eq{eq:DLc}, which are performed at a fixed mass,
in favor of total time derivatives, denoted with overdots.  %The physical distinction is that total derivatives include mass  accreted through the outer boundary.  
For instance the surface radius, $R$, evolves as  
\begin{equation}\label{eq:Rdot}
 \dot{R} = {\p R \over \p t} + {\dot{M} \over 4 \pi R^2 \rho_M}
\end{equation} 
where $\p R/\p t$ gives the Lagrangian contraction of the ``original" shell, and mass accretion through the upper boundary at rate $\dot{M}$ also changes the shell location.  
%The subscript $M$  denotes quantities at the upper boundary of total mass $M$ (though it is omitted from $M$ and $R$).  
Similarly the volume, $V = (4 \pi/3)r^3$, and pressure at the outer shell evolve as
\begin{subeqnarray}\label{eq:dot}
\dot{V}_M &=&  {\p V_{\rm M} \over \p t} + {\dot{M} \over \rho_{\rm M}}  \\
 \dot{P}_M &=& {\p P_{\rm M} \over \p t} + {\p P_M \over \p m}\dot{M} =  {\p P_{\rm M} \over \p t} - {G M  \over 4 \pi R^4} \dot{M}\, .
\end{subeqnarray} 
This derivation holds the core mass and radius fixed, $\dot{M}\co = \dot{R}\co = 0$.  Therefore the core pressure satisfies
\begin{equation}\label{eq:Pcdot}
 \dot{P}\co = \p P\co / \p t \, .
\end{equation}

The internal energy integral follows simply from  Leibniz's rule as
\begin{equation}\label{eq:udot}
\int_{M\co}^{M(t)}{\p u \over \p t} dm = \dot{U}  -  \dot{M}u_M\, .
\end{equation}

To make further progress we require the virial theorem:
\begin{equation}
\label{eq:virial}
E_G=-3 \int_{M\co}^M \frac{P}{\rho} dm + 4 \pi (R^3 P_M-R\co^3 P\co)
\end{equation}
which follows from \Eqsss{eq:structb}{eq:structa}{eq:Eg} by integrating hydrostatic balance in Lagrangian coordinates.  As an aside, the integral in equation (\ref{eq:virial}) can be evaluated for a polytropic EOS to give simple expressions for the total energy:
\begin{subeqnarray}
E&=&(1-\zeta)U+4 \pi (R^3 P_M-R\co^3 P\co) \slabel{eq:vira} \\
&=&\frac{\zeta-1}{\zeta}E_G+\frac{4 \pi}{\zeta} (R^3 P_M-R\co^3 P\co) \slabel{eq:virb}
\end{subeqnarray}
where $\zeta \equiv 3(\gamma - 1)$.  We will not make this assumption and will keep the EOS general.

To express the work integral, the final term in \Eq{eq:DLc}, in terms of changes to gravitational energy we first take the
 time derivative of \Eq{eq:virial}:
\begin{eqnarray}\label{eq:EGdot}
\dot{E}_G = 3  \int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm -3 \int_{M\co}^M {\p P\over \p t}{dm \over \rho} 
 -  3{P_M \over \rho_M} \dot{M}+ 3 \dot{P}_M V_M -3 \dot{P}\co V\co  + 3  P_M {\dot{ V}_M} \, . 
\end{eqnarray} 
%where the volumes, $V_M = 4 \pi R^3/3$ and $V\co = 4 \pi R\co^3/3$. 
The first integral in \Eq{eq:EGdot} is the one we want, but the next one must be eliminated.  The time derivative of \Eq{eq:Eg} (times four) gives
\begin{subeqnarray}
 4 \dot{E}_G &=&  -4 {G M \dot{M} \over R} + 4 \int_{M\co}^M {G m \over r^2}{\p r \over \p t} dm\\ 
&=&   -4 {G M \dot{M} \over R} + 4 \pi \int_{M\co}^M r^3{\p \over \p m}{\p P \over \p t} dm \slabel{eq:4EGb} \\
&=&  -4 {G M \dot{M} \over R} -3  \int_{M\co}^M {\p P\over \p t}{dm \over \rho}  + 3 V_M {\p P_M \over \p t} -3 V\co {\p P\co \over \p t} \slabel{eq:4EGc}
\end{subeqnarray} 
where \Eqs{eq:4EGb}{eq:4EGc} use hydrostatic balance  and integration by parts.

%To eliminate the time derivates of pressure, we take the time derivative of the hydrostatic balance equation for $\p^2 P / \p m\p t$ and integrate over $4\pi r^3 dm$ (as in the virial equation derivation) to get
%\begin{equation}\label{eq:dHBdt}
%3 \dot{P}_M V_M -3 \dot{P}\co V\co -3 \int_{M\co}^M {\p P\over \p t}{dm \over \rho}  = 4 \dot{E}_G + 4{G M \over R} \dot{M}  \, .
%\end{equation} 
%Combining \Eqs{eq:EGdot}{eq:dHBdt} gives 
%\begin{eqnarray}\label{eq:rhodot}
%\int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm  &=& - \dot{E}_G - {4 \over 3}{G M\over R} \dot{M} + {P_M \over \rho_M} \dot{M} -  P_M \dot{V}_M  \, , \nonumber \\
%&=&- \dot{E}_G - {4 \over 3}{G M\over R} \dot{M}  -  P_M {\p V_M \over \p t}  \, ,
%\end{eqnarray} 
%where the final step uses \Eq{eq:Rdot}.

Subtracting \Eqs{eq:udot}{eq:4EGc} and rearranging terms with the help of \Eqsss{eq:Rdot}{eq:dot}{eq:Pcdot} gives
\begin{eqnarray}\label{eq:PdVint}
\int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm  &=&  - \dot{E}_G - {G M \dot{M} \over R} - P_M {\p V_M \over \p t} \,  .
\end{eqnarray} 
Combining \Eqsss{eq:DLc}{eq:udot}{eq:PdVint}, we reproduce \Eq{eq:coolingglobal} with the accreted specific energy $e_M \equiv u_M - GM/R$.

\end{document}

### 2013-07-04 tex/draft/draft3.tex
section 3 edited (except for section intro) & added to draft3

@@
 \item Dust grains provide the opacity in the radiative zone, which remains cool enough to avoid dust sublimation.
+\item Because subsequent growth accelerates, the time to reach the crossover mass is a good approximation of the total time to form a gas giant.
 \end{enumerate}
-
 The remainder of this section develops our model in more detail.
@@
 
+\emph{TODO: this intro still needs edits}
+
+In this section we derive a simplified analytic model for the atmosphere structure and evolution, based on the assumptions and equations described in section \ref{sec2}.
+We use this model to predict the qualitative behavior of our more detailed numerical model.
+%, and to explain expected scalings (\textbf{Maybe take out this sentence.})
+
+We consider a two-layer atmosphere structure model as presented in section \ref{model}, and we make two key assumptions: we neglect the self-gravity of the atmosphere, and we assume that the outer radiative region is very thick and hence the pressure at the radiative-convective boundary is much larger than the disk pressure $P\di$.
+We will show in section \ref{KH} that these assumptions limit the applicability of this simplified model during the late and early stages of the atmosphere evolution, respectively.
+We also approximate the structure of the radiative zone as nearly isothermal, and note that this assumption is consistent with the numerical solutions (see section \ref{profiles}).
+
+As self-gravity is quantitatively important, we expect our analytic model to only provide scalings for the full numerical model rather than absolute values.
+
+%We approximate the outer radiative zone of the atmosphere as nearly isothermal.
+Moreover, we do not take into account the self-gravity of the atmosphere in either region.
+We derive the structure of an isothermal non self-gravitating atmosphere is section \ref{iso}.
+
+We use the quasi-static model of section \ref{sec2} to derive expressions for the mass, energy and luminosity of a static non self-gravitating atmosphere.
+We use this to derive the time evolution of the atmosphere and determine the minimum core mass for runaway gas accretion in section \ref{coolingan}.
+We defer the details of our calculations to \App{sec:analytic}.
+
+%\emph{shallow radiative zone passed quickly...}
+
+\subsection{Two Layer Structure}
+To apply the two layer cooling model analytically, we require expressions for the atmospheric structure.
+Conditions at the RCB are crucial as they set the interior adiabat and the radiative losses.
+(Recall that luminosity generation in the radiative zone is neglected.)  We express the temperature and pressure of the RCB, at the radius $R\cb$, as
+\begin{subeqnarray}
+T\cb &=& \chi T\di \slabel{eq:Tcb} \\
+P\cb &=& \theta P_{\rm d} e^{R_{\rm B}/R\cb} \, .\slabel{eq:PcbRcb}
+\end{subeqnarray}
+
+If the radiative zone were replaced by an isothermal layer, then hydrostatic balance would give $\chi = \theta = 1$.
+Standard radiative structure calculations (see \App{sec:analytic} for details) give
+\begin{equation}
+\label{eq:chi}
+\chi \simeq \Big(1-\frac{\delad}{\nabla_{\infty}} \Big)^{-\frac{1}{4-\beta}} \simeq 1.53 \, ,
+\end{equation}
+for $P\cb \gg P\di$, our regime of interest, and with the radiative temperature gradient at infinite depth, $\nabla_\infty = 1/2$, for our dust opacity.
+The radiative zones are thus nearly isothermal, as also found by R06.
+While  a simple closed form expression is not possible, the factor $\theta \simeq 0.556$ for our parameters.
+
+Given the conditions at the RCB, hydrostatic balance gives the density and temperature profiles along the adiabat,
+\begin{subeqnarray}
+\rho &=& \rho\cb \left[ 1 + {\RB' \over r} - {\RB' \over R\cb}  \right]^{1/(\gamma -1)} \slabel{eq:rhoconv}  \\
+T	&=& T\cb \left[ 1 + {\RB' \over r} - {\RB' \over R\cb}  \right] \slabel{eq:Tconv}\, .
+\end{subeqnarray} 
+%Note (to self) that the "1+" is neglibible in Rcb << RB limit taken elsewhere. 
+The effective Bondi radius
+\begin{equation}
+\RB' \equiv {G M\co \over C_P T\cb} = {\delad \over \chi} \RB
+\end{equation} 
+simplifies expressions, with $C_P = \Rg / \delad$ the specific heat at constant pressure.
+
+Even simpler expressions hold deep in the adiabatic interior, where $r \ll R\cb \ll \RB'$ gives
+\begin{subeqnarray}
+\rho &\simeq& \rho\cb \left[ {\RB' \over r}\right]^{1/(\gamma -1)} \propto r^{-5/2} \slabel{eq:rhoconv}  \\
+T	&\simeq& T\cb {\RB' \over r} = {G M\co \over C_P r} \slabel{eq:Tconvdeep}\, .
+\end{subeqnarray} 
+While the radial density profile depends on the adiabatic index, the temperature scaling does not.  The total specific energy at depth
+\begin{equation}\label{eq:ean}
+e = e_g + u = -\delad {GM\co \over r} \, .
+\end{equation} 
+is thus generically proportional to the gravitational potential, $e_G = -GM\co/r$. 
+%AY: note $u$ introduced already in section 2.  Furthermore meaning obvious from context here.  e_G is good to restate in the non-self-grav limit which changed.
+
+\subsection{Mass, Energy and Luminosity}
+\label{MELan}
+The most relevant quantities for the global cooling model are the integrated energy, luminosity, and atmospheric mass.
+In our non-self-gravitating limit, the mass of our nearly isothermal radiative zones is less than the convective interior, as shown in \App{sec:analytic}.
+
+\Eq{eq:rhoconv} thus integrates to give the atmospheric mass
+\begin{eqnarray} 
+\label{eq:Matman}
+M_{\rm atm} %&=& 4 \pi \int_{R\co}^{R\cb} \rho r^2 dr \\  %AY: any interested reader knows this step, right?
+&=& {5 \pi^2 \over 4} \rho\cb {\RB'}^{5/2} \sqrt{R\cb}, 
+%AY: alternate forms
+% \approx 0.54 \rho\cb {\RB'}^{5/2} \sqrt{r\cb} \\
+%&=& 0.54{\RB'^{3} \sqrt{\chi} \over  \sqrt{ \ln[P\cb/(\theta P\di)]}} \, .
+\end{eqnarray}
+in the relevant limit $R\co \ll R\cb \ll \RB'$ and for $\gamma =  7/5$.  Mass is concentrated near the outer regions of the convective zone, a result that holds more generally for $\gamma > 4/3$.
+
+Using \Eq{eq:PcbRcb} to eliminate $R\cb$, the critical ratio of atmosphere to core mass becomes
+\begin{equation} \label{eq:crit}
+{M_{\rm atm} \over M\co} = {P\cb \over \xi P_M},
+\end{equation} 
+where we define a characteristic pressure and a logarithmic factor
+\begin{subeqnarray} 
+P_M &\equiv& {4 \delad^{3/2} \over 5 \pi^2 \sqrt{\chi} } {G M\co^2 \over {\RB'}^4}\, \\
+\xi &\equiv& \sqrt{\ln[ P\cb/(\theta P_{\rm d})]} \slabel{eq:xidef}\, .%P_M \approx {1.85 \over \sqrt{\chi}} {G M\co^2 \over {\RB'}^4}\, .
+\end{subeqnarray}
+
+The atmosphere mass increases as radiative losses lower the internal adiabat and increase $P\cb$.  The crossover mass, $M_{\rm atm} = M\co$, is reached when
+\begin{equation} \label{eq:Pcbc}
+P\cb = \xi P_M ,
+\end{equation} 
+i.e.\ near the characteristic pressure $P_M$.
+The critical value of the order unity factor $\xi$ is found by eliminating $P\cb$ from \Eqs{eq:xidef}{eq:Pcbc}.
+This logarithmic factor complicates our analytic description, but since it remains order unity we simply hold it fixed in our scalings.
+
+The total energy is concentrated towards the core if $u \rho  r^3 \propto \rho r^2$ drops with increasing $r$.
+This condition requires $\gamma < 3/2$, which our choice of $\gamma = 7/5$ satisfies, but $\gamma = 5/3$ does not.
+Integration of \Eq{eq:ean} over the mass of the atmosphere thus gives
+\begin{subeqnarray} 
+E &=& - 4 \pi \nabla_{\rm ad} G M\co \int_{R\co}^{R\cb} \rho r dr \\
+&\approx& - 4 \pi P\cb {\RB'}^{1 \over \nabla_{\rm ad}} \left(\gamma-1 \over 3 - 2 \gamma\right)  R\co^{2\gamma-3\over \gamma-1}  \slabel{eq:Ean} \\ 
+&\approx& - 8 \pi P\cb {\RB'^{7/2} \over \sqrt{R\co}} \slabel{eq:Eus}
+%&\approx& -0.31 P\cb {(\RB / \chi)^{7/2} \over \sqrt{R\co}}
+%&\approx& -0.31 P\cb {\RB'^{7/2} \over \sqrt{R\co}}
+\end{subeqnarray} 
+with $\gamma < 3/2$ and $\gamma = 7/5$ in \Eqs{eq:Ean}{eq:Eus}, respectively. % and then revert to the standard Bondi radius $\RB = G M\co/(\Rg T\di)$ in terms of the disk temperature.
+
+Applying marginal convective stability, $\delrad = \delad$, to \Eq{eq:delrad} gives the emergent luminosity from the radiative-convective boundary, 
+\begin{equation} \label{eq:Lcb}
+L\cb = {64 \pi G M\cb \sigma T\cb^4 \over 3 \kappa P\cb } \nabla_{\rm ad} \approx L\di {P_{\rm d} \over P\cb}\, , % \left({P\di \over P\cb}\right)%^{1+\alpha}
+\end{equation} 
+where we define 
+\begin{equation} 
+L\di \equiv {64 \pi G M\cb \sigma T_{\rm d}^4 \over 3 \kappa(T_{\rm d}) P_{\rm d}} \nabla_{\rm ad}\chi^{4-\beta}.
+\end{equation} 
+The inverse scaling $L\cb \propto 1/P\cb$ shows that luminosity drops as the atmosphere cools and $P\cb$ deepens.
+This result relies on the pressure independence of  dust opacities.
+For fully non-self-gravitating results, we replace $M\cb$, the mass up to the RCB, with the core mass, but the mass of the convective atmosphere could be included for a slightly higher order estimate.
+
+\subsection{Simplified Cooling Model}
+\label{coolingan}
+%We can eliminate $R\cb$ from equation (\ref{eq:Matman}) using \Eq{eq:PcbRcb}.  The ratio of atmosphere to core mass is then  
+%\begin{equation} \label{eq:crit}
+%{M_{\rm atm} \over M\co} \approx {P\cb /P_M \over  \sqrt{\ln[ P\cb/(\theta P_{\rm d})]}},
+%\end{equation} 
+%where we introduce a characteristic pressure
+%\begin{equation} 
+%P_M \equiv {4 \delad^{3/2} \over 5 \pi^2 \sqrt{\chi} } {G M\co^2 \over {\RB'}^4}\, .
+%%P_M \approx {1.85 \over \sqrt{\chi}} {G M\co^2 \over {\RB'}^4}\, .
+%\end{equation} 
+%For $M_{\rm atm} = M\co$ in equation (\ref{eq:crit}) we require
+%\begin{equation} \label{eq:Pcbc}
+%P\cb = \xi P_M ,
+%\end{equation} 
+%where  the logarithmic factor
+%\begin{equation}\label{eq:xi}
+%\xi \equiv \sqrt{\ln[ P\cb/(\theta P_{\rm d})]} = \sqrt{\ln[ \xi P_M / (\theta P_{\rm d})]}
+%\end{equation} 
+%is found by numerically solving the above transcendental equation.
+The physical solution has $\xi >1$, but is typically of order unity.
+Numerical solutions exist for $\xi \ll 1$, but are unphysical as they imply $P\cb < P_{\rm d}$.
+
+Our analytic cooling model uses $L = -\dot{E}$,  neglecting the surface terms in \Eq{eq:coolingglobal}.\footnote{\App{sec:analytic} shows that these terms are negligible for a non-self-gravitating model, but they do become relevant when self-gravity is important and included.}  Applying \Eqs{eq:Eus}{eq:Lcb} the time it takes to cool the atmosphere until the RCB reaches a given pressure depth, $P\cb$, is
+\begin{subeqnarray} 
+t_{\rm  cool} &=&- \int_{P\di}^{P\cb } {d E/d P\cb \over L} dP\cb \\
+&\approx& 4 \pi {P\cb^{2} \over P\di} {\RB'^{7/2} \over L\di \sqrt{R\co}} \label{eq:tcool}\, .
+%&\approx& {0.16 }{P\cb^{2} \over P\di} {\RB'^{7/2} \over L\di \sqrt{R\co}}
+\end{subeqnarray} 
+The initial RCB depth, set to $P\di$ as a formality, is of little importance because the cooling slows it proceeds with $t_{\rm cool} \propto P\cb^2$.
+
+The most relevant cooling time is the crossover time, $t_{\rm co}$, required to reach the crossover mass, $M_{\rm{atm}} = M_{\rm c}$. \Eqs{eq:tcool}{eq:Pcbc} give this time as
+\begin{eqnarray} 
+\label{eq:tcoolan}
+t_{\rm co} &\approx & 2 \times 10^8 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2  \over \mcn{10}^{5/3} \aun{10}^{15 / 14}} \yr \, .
+\end{eqnarray} 
+The corresponding critical core mass, $\MC$, has a crossover time that equals the typical lifetime of a protoplanetary disk
+\begin{equation}
+t\di = 3 \times 10^6 ~{\rm yr} \,.
+\end{equation} 
+Setting $t_{\rm co} = t\di$ gives
+\begin{equation}\label{eq:Mcrit}
+\MC \approx 100 {F_T^{3/2} F_\kappa^{3/5}   \left(\xi \over 2.6 \right)^{6/5} \over \aun{10}^{9 / 14}} \; M_\oplus.
+\end{equation}
+
+Both $t_{\rm co}$ and $\MC$ are too large to be interesting, or to be correct based on previous results and the numerical results in this Paper.
+The main reason for this discrepancy is the neglect of self-gravity, as shown in section \ref{sec:KH}.
+The magnitude of the discrepancy may be surprising as one might expect self-gravity to be a modest factor for $M_{\rm atm} \leq M\co$.
+Clearly this seemingly reasonable expectation must be applied with caution, an interesting result in itself.
+
+The analytic model is also useful because, despite the amplitude error, it explains the basic scaling of our (and other) numerical models.
+As is well known, a lower opacity, here scaling with $F_\kappa$, accelerates growth and gives a lower critical core mass.
+Lower disk temperatures also decrease both $t_{\rm co}$ and $\MC$.
+Furthermore the decline in both quantities with semimajor axis is completely generated by the disk temperature law (ignoring $\xi$).
+The temperature dependence are a competition between two main effects.
+The higher luminosity, $\propto T^{4-\beta}$, at higher temperatures gives faster cooling, which opposes the overall effect.
+The larger Bondi radius, and the effectively stronger gravity of the core, is the dominant effect that gives faster cooling at lower temperatures.
+For different $\beta$ values, $t_{\rm co} \propto F_T^{ \beta + 1/2}$, so the slope of the dust opacity is an important moderating effect on the cooling.
+
+The cooling timescale and critical core mass depend only weakly on the disk pressure, via the  logarithmic factor $\xi$.
+Section \ref{sec:TPeffects} shows that these analytic predictions for disk temperature and pressure dependence compare favorably with numerical results.
+%This weak dependence is a consequence of cooling coming from the radiative-convective boundary.
+Note that the scaling laws do not reflect the changes to $\xi$ which remains order unity.
+% and is here given the appropriate value for the nominal parameters.
+
+\emph{TODO: consider adding "revised" analytic expressions for $t_{\rm co}$ and $\MC$ (e.g.
+lower Matm criterion).
+People like scalings, but the obviously wrong amplitudes will prevent people from using them unless we correct.}
+
 \section{Quasi-Static Kelvin-Helmholtz Contraction}
 \label{sec:KH}
+
+\subsubsection{Disk Temperature and Pressure}
+\label{sec:TPeffects}
 
@@
 
+\section{Analytic Cooling Model Details}\label{sec:analytic}
+TODO (academic): Figure out if theta < 1 is actually required? Should be but need to check
+
 \end{document}

### 2013-07-05 tex/draft/draft3.tex
section 3 intro edits (into draft 3)

@@
 
-\emph{TODO: this intro still needs edits}
+This section develops the analytic version of our two-layer model, which is useful for predicting and explaining the results of numerical evolution models.
+To obtain our analytic results, we neglect self-gravity.
+As shown in section \ref{sec:KH}, self-gravity becomes quantitatively important at surprisingly low atmospheric masses, $M_{\rm atm} \gtrsim 0.2 M\co$.
+Thus the analytic model cannot accurately model the the later stages of evolution.
+Nevertheless, the analytic model does successfully explain general trends with e.g.\ opacity and disk temperature and pressure.
 
-In this section we derive a simplified analytic model for the atmosphere structure and evolution, based on the assumptions and equations described in section \ref{sec2}.
-We use this model to predict the qualitative behavior of our more detailed numerical model.
-%, and to explain expected scalings (\textbf{Maybe take out this sentence.})
-
-We consider a two-layer atmosphere structure model as presented in section \ref{model}, and we make two key assumptions: we neglect the self-gravity of the atmosphere, and we assume that the outer radiative region is very thick and hence the pressure at the radiative-convective boundary is much larger than the disk pressure $P\di$.
-We will show in section \ref{KH} that these assumptions limit the applicability of this simplified model during the late and early stages of the atmosphere evolution, respectively.
-We also approximate the structure of the radiative zone as nearly isothermal, and note that this assumption is consistent with the numerical solutions (see section \ref{profiles}).
-
-As self-gravity is quantitatively important, we expect our analytic model to only provide scalings for the full numerical model rather than absolute values.
-
-%We approximate the outer radiative zone of the atmosphere as nearly isothermal.
-Moreover, we do not take into account the self-gravity of the atmosphere in either region.
-We derive the structure of an isothermal non self-gravitating atmosphere is section \ref{iso}.
-
-We use the quasi-static model of section \ref{sec2} to derive expressions for the mass, energy and luminosity of a static non self-gravitating atmosphere.
-We use this to derive the time evolution of the atmosphere and determine the minimum core mass for runaway gas accretion in section \ref{coolingan}.
-We defer the details of our calculations to \App{sec:analytic}.
-
-%\emph{shallow radiative zone passed quickly...}
+The analytic model also assumes that the upper radiative radiative layer is thick, with a pressure depth $P\cb \gg P\di$.
+This approximation limits the ability to study the initial cooling of very low mass atmospheres.
+However since early evolution is quite rapid, this approximation negligibly affects total cooling times and criticial core masses, our main foci.
 
@@
 While  a simple closed form expression is not possible, the factor $\theta \simeq 0.556$ for our parameters.
+According to \Eq{eq:PcbRcb} our pressure criterion also gives $R\cb \ll \RB$, but the inequality is logarithmically weaker.
 

### 2013-07-15 tex/draft/draft3.tex
section 2 edits w.r.t. Ana's comments

@@
 \newcommand{\co}{_{\rm c}}
-\newcommand{\pla}{_{\rm pl}}
+\newcommand{\pla}{_{\rm p}}
 \newcommand{\di}{_{\rm d}}
@@
 \newcommand{\mc}{m_{\rm c \oplus}}
-\newcommand{\mcn}[1] { m_{ \rm c #1 \oplus} }
+\newcommand{\mcn}[1] { m_{ \rm c #1} }
+\newcommand{\mpn}[1] { m_{ \rm p #1} }
 \newcommand{\MC}{M_{\rm crit}}
 \newcommand{\au}{a_\oplus}
-\newcommand{\aun}[1]{ a_{#1\oplus} }
+\newcommand{\aun}[1]{ a_{#1} }
+%\newcommand{\aun}[1]{ a_{#1\oplus} }
 
@@
 This paper is organized as follows.
-In section \ref{sec:model} we describe the assumptions of our atmosphere model, and derive the basic equations that govern the structure and evolution of the atmosphere.
-In section \ref{sec:coolingan}, we present a simplified analytic model that predicts the qualitative behavior of the numerical model.
-Results for atmospheric structure and evolution are presented in section \ref{sec:KH}, and implications for the critical core mass are presented in section \ref{sec:critical}.
-The discussion in section \ref{sec:neglected} addresses various approximations and neglected effects.
-We summarize our findings in section \ref{sec:conclusions}.
+In Section \ref{sec:model} we describe the assumptions of our atmosphere model, and derive the basic equations that govern the structure and evolution of the atmosphere.
+In Section \ref{sec:coolingan}, we present a simplified analytic model that predicts the qualitative behavior of the numerical model.
+Results for atmospheric structure and evolution are presented in Section \ref{sec:KH}, and implications for the critical core mass are presented in Section \ref{sec:critical}.
+The discussion in Section \ref{sec:neglected} addresses various approximations and neglected effects.
+We summarize our findings in Section \ref{sec:conclusions}.
 The appendices contain...
@@
 We follow an intermediate approach and consider quasi static evolution.
-Our model for the atmosphere growth time is described in section \ref{...}.
+Our model for the atmosphere growth time is described in Section \ref{...}.
 
@@
 Our model can accurately model the growth of the atmosphere up to the crossover mass, when the atmosphere mass equals the mass of the core.
-Beyond the crossover mass,  our approximate treatment of the radiative zone breaks down.
-Since subsequent growth is a rapid runaway process \emph{(cite a numerical core accretion paper)}, our model can investigate to good accuracy the timescale requirement of core accretion.
+Beyond the crossover mass,  our approximate treatment of the radiative zone (explained below) breaks down.
+Since subsequent growth is a rapid runaway process \citep{pollack96}, our model can investigate to good accuracy the timescale requirement of core accretion.
 Our simplified treatment is also inappropriate for hot, short-period planets, where dust sublimation gives deeper radiative zones that require more detailed models.
@@
 \item A global cooling model connects independent static solutions into a time-dependent sequence.
-\item A polytropic EOS is assumed for simplicity.
+\item A polytropic EOS is assumed for simplicity. %AP: define EOS here if not already
 \item Dust grains provide the opacity in the radiative zone, which remains cool enough to avoid dust sublimation.
-\item Because subsequent growth accelerates, the time to reach the crossover mass is a good approximation of the total time to form a gas giant.
+\item Because gas accretion accelerates after the crossover mass, the time to reach the crossover mass is a good approximation of the total time to form a gas giant.
 \end{enumerate}
@@
 with a powerlaw index $\beta = 2$ and normalization $F_\kappa = 1$ unless noted otherwise.
-Roughly speaking, grain growth will lower both $F_\kappa$ and $\beta$, while dust abundance scales with $F_\kappa$.
+Grain growth tends to lower both $F_\kappa$ and $\beta$, while dust abundance scales with $F_\kappa$.
 Section \ref{sec:opEOS} discusses dust sublimation and more realistic opacity laws.
@@
 The core density is held fixed at $\rho\co=3.2$ g cm$^{-3}$.
-We thus neglect  the detailed equation of state of the solids core \citep{fortney07}.
+We thus neglect  the detailed equation of state of the solid core \citep{fortney07}.
 
@@
 \label{eq:RB}
-\RB \equiv \frac{G M\pla}{c\di^2} \approx 0.17 \, {\mcn{10}  \, \aun{10}^{3/7} \over F_T} ~\AU
+\RB \equiv \frac{G M\pla}{c\di^2} \approx 0.17 \, {\mpn{10}  \, \aun{10}^{3/7} \over F_T} ~\AU
 \end{equation}
-where the planet mass, $M\pla = M\co + M_\mathrm{atm}$, includes the mass of the core and atmosphere (defined as lying within the Bondi radius).
-The numerical estimates above and in \Eq{eq:RHill} assume the core mass dominates, $M\pla \sim M\co$.
+where the enclosed planet mass, $M\pla = M\co + M_\mathrm{atm}$, includes the core and any atmosphere within the Bondi radius.  The scalings thus use $\mpn{10} \equiv M\pla /(10~M_\oplus)$.
 
@@
 \label{eq:RHill}
-R_{\rm H} = \left(M\pla \over 3 M_\ast \right)^{1/3}a \approx 0.22 \, {\mcn{10}^{1/3} \, \aun{10} }~\AU
+R_{\rm H} = \left(M\pla \over 3 M_\ast \right)^{1/3}a \approx 0.22 \, {\mpn{10}^{1/3} \, \aun{10} }~\AU
 \end{equation}
-where hydrostatic balance clearly breaks down.
+where hydrostatic balance breaks down.
+In \Eq{eq:RHill}, $M\pla$ includes mass enclosed within $\RH$.
+During the early stages of evolution, $M\pla \sim M\co$, and the core mass can be used to get estimates of both $\RB$ and $\RH$.
 
@@
 For a finite range of intermediate masses, $M_{\rm th}/\sqrt{3} < M\pla < 3 M_{\rm th}$, the Hill radius is the smallest scale, satisfying both $\RH < \RB$ and $\RH < H\di$.
-Spherical symmetry remains a good, if imperfect, approximation because the bound atmosphere within $\RH$ doesn't see the vertical stratification of the disk on scales $\gtrsim H\di$.
+Spherical symmetry remains a good, if imperfect, approximation because the disk is only weakly vertically stratified on  scales $\lesssim H\di$.
 
@@
 where $\sigma$ is the Stefan-Boltzmann constant.
-In convectively unstable regions, efficient convection gives an isentropic temperature gradient with $\nabla = \delad$, the adiabatic gradient.
+In convectively unstable regions, efficient convection gives an isentropic temperature gradient with $\nabla = \delad$, the adiabatic gradient (i.e.\ the temperature gradient at constant entropy).
 According to the Schwarzschild criterion, convective instability occurs when $\delrad > \delad$.
-Thus $\nabla = \min(\delrad, \delad)$ set the temperature gradient.
+Thus $\nabla = \min(\delrad, \delad)$ sets the temperature gradient.
 
@@
 The $\epsilon_{\rm g} = -T \partial S / \partial t$ term gives the energy input from gravitational contraction.\footnote{In general, any motion in a non-stationary atmosphere is accounted for by this term.
-The partial time derivative is performed on shells of fixed mass.}  The partial time derivative would normally require our radial derivatives should be partial derivatives.
+The partial time derivative is performed on shells of fixed mass.}  The partial time derivative would normally require our radial derivatives to be partial derivatives.
 However our subsequent developments will replace the local energy equation (\ref{eq:structd}) with global energy balance, reverting the structure equations to time-independent ordinary differential equations (ODEs).
@@
 
+%AP/AY: add virial theorem here?
+
 We start with the global energy balance for an isolated planet with a free surface:
@@
 acquires surface terms as derived in  \App{sec:globalderiv}.
-The energy accreted across the surface is given by the specific energy $e_M = u_M-G M/R$ and the mass accretion rate $\dot{M}$.
+The energy accreted across the surface is given by the specific energy, $e_M = u_M-G M/R$, and the mass accretion rate of gas, $\dot{M}$.
 The work done by the surface is $P_M \partial V_M/ \partial t$, with the partial derivative performed at fixed mass.
@@
 To obtain our analytic results, we neglect self-gravity.
-As shown in section \ref{sec:KH}, self-gravity becomes quantitatively important at surprisingly low atmospheric masses, $M_{\rm atm} \gtrsim 0.2 M\co$.
+As shown in Section \ref{sec:KH}, self-gravity becomes quantitatively important at surprisingly low atmospheric masses, $M_{\rm atm} \gtrsim 0.2 M\co$.
 Thus the analytic model cannot accurately model the the later stages of evolution.
@@
 \begin{subeqnarray}
-\rho &\simeq& \rho\cb \left[ {\RB' \over r}\right]^{1/(\gamma -1)} \propto r^{-5/2} \slabel{eq:rhoconv}  \\
+\rho &\simeq& \rho\cb \left[ {\RB' \over r}\right]^{1/(\gamma -1)} \propto r^{-5/2} \slabel{eq:rhoconvdeep}  \\
 T	&\simeq& T\cb {\RB' \over r} = {G M\co \over C_P r} \slabel{eq:Tconvdeep}\, .
@@
 Both $t_{\rm co}$ and $\MC$ are too large to be interesting, or to be correct based on previous results and the numerical results in this Paper.
-The main reason for this discrepancy is the neglect of self-gravity, as shown in section \ref{sec:KH}.
+The main reason for this discrepancy is the neglect of self-gravity, as shown in Section \ref{sec:KH}.
 The magnitude of the discrepancy may be surprising as one might expect self-gravity to be a modest factor for $M_{\rm atm} \leq M\co$.

### 2013-07-18 tex/draft/draft2.tex
updated abstract and intro in draft2

@@
 In standard calculations of the critical core mass, planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse.
-In this study we consider the extreme case in which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the planet.
-We develop a two-layer atmosphere model with an inner convective region and an outer radiative zone that matches onto the protoplanetary disk, and we determine the minimum core mass for a giant planet to form within the typical disk life timescale for a variety of disk conditions, which we denote as  \textit{critical core mass}.
-We find that the absolute minimum core mass required to nucleate atmosphere collapse within the disk lifetime is smaller for planets forming further away from their host stars.
-Moreover, the critical core mass is strongly dependent on disk temperature, opacity and mean molecular weight of the gas.
+In this study we consider the limiting case in which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the planet.
+We develop a two-layer atmosphere model with an inner convective region and an outer radiative zone that matches onto the protoplanetary disk, and we determine the minimum core mass for a giant planet to form within the typical disk lifetime for a variety of disk conditions.
+We denote this mass as  \textit{critical core mass}.
+The absolute minimum core mass required to nucleate atmosphere collapse is $\sim 8 M_{\oplus}$ at 5 AU and steadily decreases to $\sim 3.5 M_{\oplus}$ at 100 AU, for an ideal diatomic gas with a solar composition and a standard ISM opacity law.
+Lower opacity and disk temperature significantly reduce the critical core mass, while a decrease in the mean molecular weight of the nebular gas results in a larger critical core mass.
 Our results yield lower mass cores than corresponding studies for large planetesimal accretion rates.
@@
 
+%We find that the absolute minimum core mass required to nucleate atmosphere collapse within the disk lifetime is smaller for planets forming further away from their host stars.
+Moreover, the critical core mass is strongly dependent on disk temperature, opacity and mean molecular weight of the gas.
+
 \end{abstract}
@@
 
-Standard core accretion models (e.g., \citealt{stevenson82}, \citealt{rafikov06}) assume that the core and the atmosphere grow simultaneously, and that planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse.
-These studies consider that the planet atmosphere is in a steady state at all times, where all the incoming energy from planetesimal accretion is radiated away by the envelope.
-The core and the atmosphere grow simultaneously, and the mass of the envelope is a function of the core mass.
-As the atmosphere mass becomes comparable to the mass of the solid core, hydrostatic balance no longer holds and runaway gas accretion commences.
-For a set of disk conditions it is therefore found that there is one well-defined core mass past which hydrostatic solutions no longer exist and unstable atmosphere collapse occurs: this is defined as the ``critical core mass''.
-Standard calculations find the critical core mass to be about $10 M_{\oplus}$ (e.g., \citealt{stevenson82}, \citealt{rafikov06}).
-
-%Standard static calculations of giant planet formation that consider planetesimal accretion as the main source of luminosity (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{rafikov06}) assume that the atmosphere is at all times in a steady state at which the luminosity due to planetesimal accretion is fully radiated away by the atmosphere.
-Therefore, for a given set of disk conditions, there is a maximum mass the core can have before the onset of unstable accretion: this is defined as the critical core mass.
+%Detailed core accretion studies (\citealt{boden86}, \citealt{pollack96}, \citealt{alibert05}) have shown that this giant planet formation model can be separated in three phases.
+In phase 1 \citep{alibert05} a solid core forms by accreting planetesimals from its feeding zone at a high rate.
 
@@
 
-In this study, we show that giant planets can grow faster from small protoplanetary cores that are fully formed before significant gas accretion occurs.
-In this scenario, the planetesimal accretion rate is significantly slowed down during the gas contraction phase of the atmosphere.
+Analytic core accretion models (e.g., \citealt{stevenson82}, \citealt{rafikov06}) assume that the core and the atmosphere grow simultaneously, and that planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse.
+These studies consider that the planet atmosphere is in a steady state at all times, in which all the incoming energy from planetesimal accretion is radiated away by the envelope.
+The core and the atmosphere grow simultaneously, and the mass of the envelope is a function of the core mass.
+As the atmosphere mass becomes comparable to the mass of the solid core, hydrostatic balance no longer holds and runaway gas accretion commences.
+For a set of disk conditions and an assumed planetesimal accretion rate it is therefore found that there is one well-defined core mass past which hydrostatic solutions no longer exist and unstable atmosphere collapse occurs: this is defined as the ``critical core mass''.
+Typically, the critical core mass is found to be about $10 M_{\oplus}$ (e.g., \citealt{stevenson82}, \citealt{rafikov06}).
+
+%Standard static calculations of giant planet formation that consider planetesimal accretion as the main source of luminosity (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{rafikov06}) assume that the atmosphere is at all times in a steady state at which the luminosity due to planetesimal accretion is fully radiated away by the atmosphere.
+Therefore, for a given set of disk conditions, there is a maximum mass the core can have before the onset of unstable accretion: this is defined as the critical core mass.
+
+%    In this situation,
+
+%the atmosphere evolution is dominated by the contraction of the envelope rather than planetesimal accretion.
+The atmosphere is no longer in a steady state, but rather it accretes gas as it loses energy through radiation.
+%In this scenario, the planetesimal accretion rate is significantly slowed down during the gas contraction phase of the atmosphere.
 This reduction can arise, for example, due to dynamical clearing, or due to the core having formed in the inner parts of the disk and migrated outwards.
 In this situation, the atmosphere evolution is dominated by the contraction of the envelope rather than planetesimal accretion.
+
+However, full time dependent simulations for the evolution of giant planets do not assume a constant planetesimal accretion rate, like the studies above, and instead calculate the planetesimal accretion rate as part of an evolutionary model for the core and atmosphere.
+\citet{pollack96} find that planetesimal accretion significantly slows down before a massive gaseous envelope can form, due to depletion of the planet's feeding zone, and gas accretion starts to dominate the atmosphere evolution.
+In this situation, planetesimal accretion can no longer balance radiative losses, and instead the atmosphere gains energy from the Kelvin-Helmholtz contraction of the envelope.
+This phase of slow gas accretion is found to be much longer than the formation of the solid core (\citealt{pollack96}, \citealt{alibert05}), hence the evolutionary timescale of the atmosphere is set by Kelvin-Helmholtz contraction rather than planetesimal accretion.
+
+In this study, we show that giant planets can grow faster from small protoplanetary cores that are fully formed before significant gas accretion occurs.
 The atmosphere is no longer in a steady state, but rather it accretes gas as it loses energy through radiation.
-
 In our model we therefore assume that the luminosity evolution of the atmosphere is dominated by gas contraction, while the planetesimal accretion rate is negligible.
@@
 Once the mass of the gaseous envelope becomes comparable to the mass of the solid core, the self-gravity of the atmosphere can no longer balance the pressure gradient and unstable hydrodynamic collapse commences.
-The time required for the atmosphere to grow to this stage is the characteristic growth time of the atmosphere.
+We define the time required for the atmosphere to grow to this stage as the \textit{crossover time}.
 For a set of fixed gas and disk conditions, there exists a minimum core mass for which the atmosphere can grow on the time scale described above within the life time of the protoplanetary disk, which we define as the \textit{critical core mass}.

### 2013-07-20 tex/draft/draft3.tex
section 4 edits in draft 3

@@
 For an object with no core luminosity (or no core) and no internal heat sources, the energy equation $L_M = -\dot{E}$ describes KH contraction in its simplest form.
-For a main sequence star $L_M = \Gamma$ as the internal heat of nuclear burning supplies the total luminosity.
+For a main sequence star $L_M = \Gamma$ as the heat of nuclear burning supplies the total luminosity.
 
@@
 (Recall that luminosity generation in the radiative zone is neglected.)  We express the temperature and pressure of the RCB, at the radius $R\cb$, as
-\begin{subeqnarray}
+\begin{subeqnarray}\label{eq:cb2}
 T\cb &=& \chi T\di \slabel{eq:Tcb} \\
@@
 Even simpler expressions hold deep in the adiabatic interior, where $r \ll R\cb \ll \RB'$ gives
-\begin{subeqnarray}
+\begin{subeqnarray}\label{eq:deep}
 \rho &\simeq& \rho\cb \left[ {\RB' \over r}\right]^{1/(\gamma -1)} \propto r^{-5/2} \slabel{eq:rhoconvdeep}  \\
@@
 \end{subeqnarray} 
-While the radial density profile depends on the adiabatic index, the temperature scaling does not.  The total specific energy at depth
+While the radial density profile depends on the adiabatic index, the temperature scaling does not.  For reference, if self-gravity is included, the temperature profile satifies
+\begin{equation} \label{eq:TPsg}
+{dT \over dr} = - {G m(r) \over C_P r^2}\, .
+\end{equation} 
+Thus the enclosed mass in self-gravitating models gives a temperature profile flatter than $T \propto r^{-1}$.
+
+The total specific energy at depth
 \begin{equation}\label{eq:ean}
@@
 
-\subsubsection{Disk Temperature and Pressure}
-\label{sec:TPeffects}
+We now present some representative examples of the structure and evolution of our model atmospheres, calculated as described in \S\ref{sec:model}, with some comparisons made to the non-self-gravitating analytic model of \S\ref{sec:coolingan}.
+Radial structure is presented in \S\ref{sec:profiles}, time evolution is described in \S\ref{sec:timeev} and the regimes of validity of our two-layer cooling model are examined in \S\ref{sec:endoftime}.
+
+\begin{figure}[tb]
+\centering
+%AY: figure should probably use R_B for consistency with text.
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/PTm_profiles_v2.pdf}
+%\vspace{-0.5in}
+\caption{Radial profiles of atmospheric pressure, temperature and enclosed mass (including the core) for a $5 M_{\oplus}$ core at $a=60$ AU.
+Solid, dotted and dashed lines correspond to solutions with total mass (core and atmosphere) of $5.12 M_{\oplus}$, $5.85 M_{\oplus}$ and $9.22 M_{\oplus}$, respectively.
+Circles and triangles mark the locations of the Bondi radii and of the radiative-convective boundaries, respectively.
+The radial profiles extend from the core to the Hill radius.}
+\label{fig:profiles}
+\end{figure}
+
+\subsection{Atmospheric Stucture}
+\label{sec:profiles}
+Figure \ref{fig:profiles} shows radial profiles at different stages of atmospheric growth.
+In these examples, we fix the core mass at $M_{\rm{c}}=5 M_{\oplus}$ and the radial disk location to $a=60$ AU.
+The quoted mass values include the core plus atmosphere within the smaller of $\RB$ or $\RH$, which for these cases is $\RB$.
+The 9.22 $M_{\oplus}$ solution is the highest mass we can reach in our evolutionary sequence, as we explain further in \S\ref{sec:endoftime}.
+
+Deep in a non-self-gravitating atmosphere, \Eq{eq:deep} shows that $T \propto r^{-1}$ and $P\propto r^{-1/\delad}$.
+This behavior is seen in the low mass solutions in \Fig{fig:profiles}.
+As explained by \Eq{eq:TPsg}, the higher mass solutions show a flatter profile in $T$ and also in $P \propto T^{1/\delad}$.
+In agreement with \Eq{eq:cb2}, the radiative zone remains nearly isothermal -- even in the more massive solutions -- and thus shows a nearly exponential increase in pressure with inverse depth.
+
+The higher mass atmosphere solutions have lower entropy, i.e.\ cooling allows the atmosphere to accrete more gas.
+This is evident in \Fig{fig:profiles} from the higher internal pressures of the higher mass solutions.
+The somewhat higher temperatures in the higher mass solutions do not cause the entropy to increase.
+
+%AY: to say something about starting at disk entropy (not really needed) we should show it better in the figure.
+Also unclear if you want to say much more about evolution before figure 2 shows the evolution.
+%In the initial stages of its evolution, the atmosphere is purely convective and isentropic, with an entropy equal to that of the disk.
+In the context of our numerical model, this is the minimum atmosphere mass for which a hydrostatic solution exists.
+Further gas accumulation brings in energy that has to be radiated away, and an initially thin outer radiative region forms, which becomes thicker as the atmosphere becomes more massive.
+
+\subsection{Time Evolution}\label{sec:timeev}
+
+\begin{figure}[tb]
+\centering
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Lt_profiles_v2.pdf}
+%\vspace{-0.5in}
+\caption{Evolution of the luminosity and elapsed time during atmospheric growth around a $5 M_{\oplus}$ core at $60$ AU.
+The luminosity is initially high, then decreases as the atmosphere grows in mass and the radiative zone becomes optically thicker.
+Due to the neglect of self-gravity, the analytic model (\emph{dashed curve}) gives a further drop in luminosity and a longer evolution time.}
+\label{fig:Ltplot}
+\end{figure}
+
+The cooling model of \S\ref{sec:twolayer} is used to connect solutions of different atmospheric masses into an evolutionary sequence.
+\Fig{fig:Ltplot} shows the luminosity evolution and the elapsed time as a function of atmospheric mass for a $5 M_{\oplus}$ core at $60$ AU, i.e.\ the same parameters as in \Fig{fig:profiles}.
+During the early stages of atmospheric growth, the luminosity drops sharply.
+This behavior is seen in both the full numerical solutions (with solid lines) and the analytic model (with dashed lines).
+As seen in \Eq{eq:Lcb}, the increased pressure depth of the RCB increases the optical depth ($\propto \kappa P)$ and decreases the radiative luminosity.
+
+%AY: I don't want to say 5.12 is the lowest mass b/c it's not the adiabat, and I don't see what's gained by saying it's very near the lowest mass.
+
+At later stages of evolution, the numerical model gives a flat luminosity with increasing mass and also time (not shown).
+By contrast the analytic model gives a luminosity that continues to drop with increasing mass.
+The difference is due to the neglect of self-gravity in the analytic model.
+With $L\cb \propto M\cb T\cb^4/ (\kappa\cb  P\cb)$, accounting for the higher enclosed mass gives a somewhat higher luminosity in the self-gravitating model.
+However the main effect is the higher temperature in the self-gravitating models, which have the flatter temperature profiles as described above.
+
+%The flattening of the luminosity curve is a generic feature of numerical core accretion models
+
+%We find only a brief agreement between the analytic and numerical results.
+In the early stages, the radiative zone is too shallow and the assumption that $P_{\rm RCB} \gg P_{\rm d}$ used in the analytic model breaks down.
+
+\Fig{fig:LtvsMopacity} shows the effect of decreasing the atmospheric dust abundance on the atmospheric evolution.
+The opacity normalization in \Eq{eq:opacitylaw} is reduced by factors of 10 and 100.
+Lower opacities result in higher luminosities and faster evolution, a well-established result that we confirm for our model.
+Atmospheric dust opacities are difficult to robustly predict.
+Ablation of infalling solids is a dust source, while settling through the radiative zone is a sink.
+Grain growth both reduces dust opacities per unit mass and favors settling.
+Our scenario of negligible ongoing particle accretion tends to favor low dust opacities.
+To be conservative, however, our reference case considers full Solar abundances, but faster evolution is clearly possible.
+
+\begin{figure}[tb]
+\centering
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/opacity_effect.pdf}
+%\vspace{-0.5in}
+\caption{Similar to \Fig{fig:Ltplot}, but showing the effect of reducing the dust abundance by factors of 10 and 100 from standard Solar abundances.
+The reduced dust opacities give higher luminosities and faster atmospheric growth.}  %The actual magnitude of dust opacities is difficult to robustly predict.}
+\label{fig:LtvsMopacity}
+\end{figure}
+
+\Fig{fig:growthtime} plots the evolution of the atmospheric growth timescale, $M_{\rm atm}/\dot{M}$,  around a $5 M_{\oplus}$ core at several locations in the disk.
+These growth times are short both early -- when the radiative zone is relatively transparent -- and late -- when self-gravity accelerates growth.
+Thus the time required to accrete an atmosphere is dominated by intermediate masses, $M_{\rm atm} \sim 1-3 M_{\oplus}$ for the chosen parameters.
+The accelerated growth at higher masses is also evident in \Fig{fig:Ltplot} (bottom panel) as the flattening of the $t$ vs.\ $M_{\rm atm}$ curve.
+
+Generically for the models we considered, growth times peaked for $M_{\rm{atm}} < M_{\rm c}$, i.e.\ before the crossover mass.
+Thus cumulative accretion timescales can be well approximated by only considering the evolution prior to the crossover mass.
+This result is fortuitous, as we next show how our model assumptions can break down during the later stages of atmospheric growth.
+
+\begin{figure}[tb]
+\centering
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mt_profile.pdf}
+%\vspace{-0.5in}
+\caption{Evolution of the atmospheric growth timescale with mass around a $5 M_{\oplus}$ solid core  located at 10, 30 or 60 AU.
+Growth is slowest for $M_{\mathrm{atm}} \sim 1 - 3 M_{\oplus}$, i.e.\ before the crossover mass at $M_{\rm atm} = M\co$.}
+\label{fig:growthtime}
+\end{figure}
+
+\subsection{Validity of the Two-Layer Cooling Model}
+\label{sec:endoftime}
+
+To examine the regimes where our cooling model is valid, we compare the model luminosity to the additional neglected luminosity, $L_{\rm negl}$, that a more detailed model would have generated in the radiative zone.
+To compute $L_{\rm negl}$ we compute the entropy difference between successive radiative zone solutions.
+We then integrate the energy equation, $\p L / \p m = - T \p S/ \p t$, over the average depth of the radiative zone.\footnote{While useful as a diagnostic, the neglected luminosity cannot reliably correct the global cooling model because where $L_{\rm negl}$ is significant, the structure of the radiative zone would be inaccurate.}
+
+ \Fig{fig:coolingterms} shows that the neglected luminosity is indeed negligible during the early stages of evolution.
+However, $L_{\rm negl}$ exceeds the model luminosity, $L$, for $M_{\rm atm} > 3 M_\oplus$, meaning that our model is becoming inaccurate.
+Nevertheless, our cooling model gives both conservative and reasonably reliable estimates of core accretion timescales.
+Our results are conservative -- in the sense of overestimating growth times -- because the neglected luminosity would give more rapid evolution.
+And our results are reasonably accurate because the bulk of growth time is spent at low masses, when $L_{\rm negl}$ is small and safely neglected.
+
+The (non-vanishing) individual terms in the global cooling model of \Eq{eq:coolingglobal}, evaluated at the RCB, are also plotted in \Fig{fig:coolingterms}.
+At low masses, the change in energy, $- \dot{E}$, makes the dominant contribution to luminosity.
+As the mass increases the surface terms become more significant, led by the accretion energy.
+However the surface terms are everywhere smaller than $L_{\rm negl}$.
+Thus while the surface terms appear to reduce the total luminosity (compared to $-\dot{E}$), $L_{\rm negl}$ can more than compensate, showing again that our model conservatively underestimates cooling.
+
+At yet higher masses than shown in \Fig{fig:coolingterms}, our cooling model breaks down.
+Mathematically, the breakdown manifests as negative timesteps when advancing to a higher atmospheric mass.
+This unphysical result has no deep significance, as we already know that the neglected luminosity is large enough to give incorrect results.
+Sometimes this breakdown occurs even before the crossover mass is reached.
+The main practical effect of this behavior is that we halt the evolution where the model breaks down whenever we cannot reach the crossover mass.
+Fortunately the elapsed time to either the crossover mass or the breakdown is well-behaved and, most importantly, dominated by lower atmospheric masses where the model is accurate.
+
+\begin{figure}[tb]
+\centering
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cooling_a10_Mc5_rcb_paper.pdf}
+%\vspace{-0.5in}
+\caption{Luminosity evolution around a $5 M_\oplus$ core at 10 AU.
+In addition to the total model luminosity $L$, the contributing terms in the cooling model of \Eq{eq:coolingglobal}; $-\dot{E}$, $e_{\rm M} \dot{M}$ and $-P \p V/ \p t |_M$; are shown.
+The $L_{\rm negl}$ curve gives the neglected luminosity that would have been generated in the radiative zone, but is not included in our two-layer cooling model.}
+\label{fig:coolingterms}
+\end{figure}
 
@@
 \label{sec:critical}
+
+%AY make sure this AU range comment is incorporated in some form:  We are primarily interested in the outer regions of the disk, so we generate atmosphere models at distances between 5 and 100 AU in our fiducial disk.
+We emphasize that the location in the disk only affects the temperature and pressure of the nebular gas.
+As our model applies only when planetesimal accretion occurs too slowly to substantially change the mass of the core, we consider the evolution of an atmosphere around a core of fixed mass.
+
+\subsubsection{Disk Temperature and Pressure}
+\label{sec:TPeffects}
 

### 2013-07-30 tex/draft/draft3.tex
section 4 responses to Ana's comments

@@
 Solid, dotted and dashed lines correspond to solutions with total mass (core and atmosphere) of $5.12 M_{\oplus}$, $5.85 M_{\oplus}$ and $9.22 M_{\oplus}$, respectively.
-Circles and triangles mark the locations of the Bondi radii and of the radiative-convective boundaries, respectively.
+(See text for a description of evolution to yet higher masses.)   Circles and triangles mark the locations of the Bondi radii and of the radiative-convective boundaries, respectively.
 The radial profiles extend from the core to the Hill radius.}
@@
 The difference is due to the neglect of self-gravity in the analytic model.
-With $L\cb \propto M\cb T\cb^4/ (\kappa\cb  P\cb)$, accounting for the higher enclosed mass gives a somewhat higher luminosity in the self-gravitating model.
-However the main effect is the higher temperature in the self-gravitating models, which have the flatter temperature profiles as described above.
+From \Eq{eq:Lcb}, $L\cb \propto M\cb T\cb^4/ (\kappa\cb  P\cb)$.
+Thus accounting for the higher enclosed mass gives a somewhat higher luminosity in the self-gravitating model.
+However the main effect is the higher temperature in the self-gravitating models.
+The flatter temperature profiles with radius, as described above, explain why the self-gravitating models are hotter.
 
@@
 The opacity normalization in \Eq{eq:opacitylaw} is reduced by factors of 10 and 100.
-Lower opacities result in higher luminosities and faster evolution, a well-established result that we confirm for our model.
+Lower opacities result in higher luminosities and faster evolution, a well-established result \citep{HubBod05} that we confirm for our model.
 Atmospheric dust opacities are difficult to robustly predict.
@@
 \Fig{fig:growthtime} plots the evolution of the atmospheric growth timescale, $M_{\rm atm}/\dot{M}$,  around a $5 M_{\oplus}$ core at several locations in the disk.
-These growth times are short both early -- when the radiative zone is relatively transparent -- and late -- when self-gravity accelerates growth.
-Thus the time required to accrete an atmosphere is dominated by intermediate masses, $M_{\rm atm} \sim 1-3 M_{\oplus}$ for the chosen parameters.
-The accelerated growth at higher masses is also evident in \Fig{fig:Ltplot} (bottom panel) as the flattening of the $t$ vs.\ $M_{\rm atm}$ curve.
+This instantaneous growth time shows clearly that the atmosphere spends the bulk of its time growing though intermediate atmospheric masses, $\sim 1 -3$ $M_\oplus$ in this case.
+Growth times are short both early -- when the radiative zone is relatively transparent -- and late -- when self-gravity accelerates growth.
+The accelerated growth at higher masses is also evident in \Figs{fig:Ltplot}{fig:LtvsMopacity} (bottom panels) as the flattening of the $t$ vs.\ $M_{\rm atm}$ curve.
 
-Generically for the models we considered, growth times peaked for $M_{\rm{atm}} < M_{\rm c}$, i.e.\ before the crossover mass.
-Thus cumulative accretion timescales can be well approximated by only considering the evolution prior to the crossover mass.
-This result is fortuitous, as we next show how our model assumptions can break down during the later stages of atmospheric growth.
+The fact that growth times peak before the crossover mass, when $M_{\rm{atm}} < M_{\rm c}$, is very helpful for understanding atmospheric growth.
+The estimated timescale to runaway growth can be well approximated by only considering the time it takes to grow to the crossover mass, or even to a somewhat lower mass.
+This insensitivity to the upper mass threshold is characteristic of accelerating growth, and is fortuitious.
+We next show how our model assumptions can break down during the later stages of atmospheric growth.
 
@@
  \Fig{fig:coolingterms} shows that the neglected luminosity is indeed negligible during the early stages of evolution.
-However, $L_{\rm negl}$ exceeds the model luminosity, $L$, for $M_{\rm atm} > 3 M_\oplus$, meaning that our model is becoming inaccurate.
+However, $L_{\rm negl}$ exceeds the model luminosity, $L$, at high masses, $M_{\rm atm} > 3 M_\oplus$ in this case, meaning that our model is becoming inaccurate.
 Nevertheless, our cooling model gives both conservative and reasonably reliable estimates of core accretion timescales.
 Our results are conservative -- in the sense of overestimating growth times -- because the neglected luminosity would give more rapid evolution.
-And our results are reasonably accurate because the bulk of growth time is spent at low masses, when $L_{\rm negl}$ is small and safely neglected.
+Moreover our results are reasonably accurate because the bulk of growth time is spent at low masses, when $L_{\rm negl}$ is small and safely neglected.
 
@@
 %\vspace{-0.5in}
-\caption{Luminosity evolution around a $5 M_\oplus$ core at 10 AU.
+%\caption{Luminosity evolution around a $5 M_\oplus$ core at 10 AU.
 In addition to the total model luminosity $L$, the contributing terms in the cooling model of \Eq{eq:coolingglobal}; $-\dot{E}$, $e_{\rm M} \dot{M}$ and $-P \p V/ \p t |_M$; are shown.
@@
 
-\subsubsection{Disk Temperature and Pressure}
+\subsection{Disk Temperature and Pressure}
 \label{sec:TPeffects}

### 2013-08-04 tex/draft/draft4.tex
updated plots and secs. 2,3,4 in draft4

--- 2013-07-30 tex/draft/draft3.tex
@@
 Beyond the crossover mass,  our approximate treatment of the radiative zone (explained below) breaks down.
-Since subsequent growth is a rapid runaway process \citep{pollack96}, our model can investigate to good accuracy the timescale requirement of core accretion.
+Since subsequent growth is a rapid runaway process (e.g., \citealt{pollack96}), our model can investigate to good accuracy the timescale requirement of core accretion.
 Our simplified treatment is also inappropriate for hot, short-period planets, where dust sublimation gives deeper radiative zones that require more detailed models.
@@
 \item Dust grains provide the opacity in the radiative zone, which remains cool enough to avoid dust sublimation.
-\item Because gas accretion accelerates after the crossover mass, the time to reach the crossover mass is a good approximation of the total time to form a gas giant.
+\item Because gas accretion accelerates after the crossover mass, the time to reach the crossover mass, i.e. the crossover time, is a good approximation of the total time to form a gas giant.
 \end{enumerate}
@@
 
-A planet can bind a dense atmosphere if the escape speed exceeds the sound speed.  This criterion is satisfied inside the Bondi radius
+A planet can bind a dense atmosphere if its escape speed exceeds the sound speed.  This criterion is satisfied inside the Bondi radius
 \begin{equation}
@@
 
-The relevant length scales of the atmosphere and disk satisfy the relation $\RB H\di^2 = 3 R_{\rm H}^3$.  The lengthscales are roughly equal at the ``thermal mass''
+The relevant length scales of the atmosphere and disk satisfy the relation $\RB H\di^2 = 3 R_{\rm H}^3$.  The lengthscales are roughly equal at the ``thermal mass'' (e.g., \citealt{menou04})
 \begin{equation}
@@
 \end{equation}
-where $\sigma$ is the Stefan-Boltzmann constant.
-In convectively unstable regions, efficient convection gives an isentropic temperature gradient with $\nabla = \delad$, the adiabatic gradient (i.e.\ the temperature gradient at constant entropy).
-According to the Schwarzschild criterion, convective instability occurs when $\delrad > \delad$.
-Thus $\nabla = \min(\delrad, \delad)$ sets the temperature gradient.
+where $\sigma$ is the Stefan-Boltzmann constant.  In convectively unstable regions, efficient convection gives an isentropic temperature gradient with $\nabla = \delad$, the adiabatic gradient
+
+\begin{equation}
+\label{eq:delad}
+\delad \equiv \Big(\frac{d \ln T}{d \ln P}\Big)_{\rm{ad}}
+\end{equation}
+According to the Schwarzschild criterion, convective instability occurs when $\delrad > \delad$.  Thus $\nabla = \min(\delrad, \delad)$ sets the temperature gradient.
 
@@
 The partial time derivative is performed on shells of fixed mass.}  The partial time derivative would normally require our radial derivatives to be partial derivatives.
-However our subsequent developments will replace the local energy equation (\ref{eq:structd}) with global energy balance, reverting the structure equations to time-independent ordinary differential equations (ODEs).
+However, our subsequent developments will replace the local energy equation (\ref{eq:structd}) with global energy balance (see section \S\ref{cooling}), reverting the structure equations to time-independent ordinary differential equations (ODEs).
 
@@
 
-We start with the global energy balance for an isolated planet with a free surface:
+We start with the global energy balance for an isolated planet of mass $M$ with a free surface:
 \begin{equation}
@@
 To simplify our calculations of atmospheric contraction we use a two layer model with a bottom convective region and an upper radiative layer.
-The existence of such a structure is well known from previous studies (R06) and can be readily understood.
+The existence of such a structure is well known from previous studies (e.g., R06) and can be readily understood.
 Before the protoplanetary atmosphere can cool, it has the entropy of the disk.
@@
 To integrate equations (\ref{eq:struct}a--c), the luminosity is required to compute $\delrad$.
-The correct value of the luminosity is the eigenvalue of the problem, which we find by the shooting method.
+The correct value of the luminosity is the eigenvalue of the problem, which we find by the shooting method \citep{NR}.
 Only for the eigenvalue of luminosity does the integrated value of mass at the core, $m(R\co)$, match the actual core mass, $M\co$.
@@
 Thus the analytic model cannot accurately model the the later stages of evolution.
-Nevertheless, the analytic model does successfully explain general trends with e.g.\ opacity and disk temperature and pressure.
+Nevertheless, the analytic model does successfully explain general trends with opacity and disk temperature and pressure, for example.
 
 The analytic model also assumes that the upper radiative radiative layer is thick, with a pressure depth $P\cb \gg P\di$.
-This approximation limits the ability to study the initial cooling of very low mass atmospheres.
+This approximation limits the ability to study the initial cooling of very low mass atmospheres with thin outer radiative zones.
 However since early evolution is quite rapid, this approximation negligibly affects total cooling times and criticial core masses, our main foci.
@@
 
-Given the conditions at the RCB, hydrostatic balance gives the density and temperature profiles along the adiabat,
+Given the conditions at the RCB, hydrostatic balance gives the density and temperature profiles in the inner region along the adiabat,
 \begin{subeqnarray}
@@
 \end{equation} 
-is thus generically proportional to the gravitational potential, $e_G = -GM\co/r$. 
+is thus generically proportional to the gravitational potential, $e_g = -GM\co/r$. 
 %AY: note $u$ introduced already in section 2.  Furthermore meaning obvious from context here.  e_G is good to restate in the non-self-grav limit which changed.
@@
 \label{MELan}
-The most relevant quantities for the global cooling model are the integrated energy, luminosity, and atmospheric mass.
+The most relevant quantities for the global cooling model (equation \ref{eq:coolingglobal}) are the integrated energy, luminosity, and atmospheric mass.
 In our non-self-gravitating limit, the mass of our nearly isothermal radiative zones is less than the convective interior, as shown in \App{sec:analytic}.
@@
 
-Our analytic cooling model uses $L = -\dot{E}$,  neglecting the surface terms in \Eq{eq:coolingglobal}.\footnote{\App{sec:analytic} shows that these terms are negligible for a non-self-gravitating model, but they do become relevant when self-gravity is important and included.}  Applying \Eqs{eq:Eus}{eq:Lcb} the time it takes to cool the atmosphere until the RCB reaches a given pressure depth, $P\cb$, is
+Our analytic cooling model uses $L = -\dot{E}$,  neglecting the surface terms in \Eq{eq:coolingglobal}.\footnote{\App{sec:analytic} shows that these terms are negligible for a non-self-gravitating model, but they do become relevant when self-gravity is important and included.}  Applying \Eqs{eq:Eus}{eq:Lcb}, the time it takes to cool the atmosphere until the RCB reaches a given pressure depth, $P\cb$, is
 \begin{subeqnarray} 
-t_{\rm  cool} &=&- \int_{P\di}^{P\cb } {d E/d P\cb \over L} dP\cb \\
+t_{\rm  cool} &=&- \int_{P\di}^{P\cb } {d E/d P\cb \over L \cb} dP\cb \\
 &\approx& 4 \pi {P\cb^{2} \over P\di} {\RB'^{7/2} \over L\di \sqrt{R\co}} \label{eq:tcool}\, .
@@
 \end{subeqnarray} 
-The initial RCB depth, set to $P\di$ as a formality, is of little importance because the cooling slows it proceeds with $t_{\rm cool} \propto P\cb^2$.
+The initial RCB depth, set to $P\di$ as a formality, is of little importance because the cooling slows as it proceeds with $t_{\rm cool} \propto P\cb^2$.
 
@@
 %AY: figure should probably use R_B for consistency with text.
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/PTm_profiles_v2.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/PTm_profiles_v2_smallM.pdf}
 %\vspace{-0.5in}
 \caption{Radial profiles of atmospheric pressure, temperature and enclosed mass (including the core) for a $5 M_{\oplus}$ core at $a=60$ AU.
-Solid, dotted and dashed lines correspond to solutions with total mass (core and atmosphere) of $5.12 M_{\oplus}$, $5.85 M_{\oplus}$ and $9.22 M_{\oplus}$, respectively.
+Solid, dotted and dashed lines correspond to solutions with total mass (core and atmosphere) of $5.01 M_{\oplus}$, $5.10 M_{\oplus}$, $6.00 M_{\oplus}$ and $9.24 M_{\oplus}$, respectively.
 (See text for a description of evolution to yet higher masses.)   Circles and triangles mark the locations of the Bondi radii and of the radiative-convective boundaries, respectively.
@@
 The quoted mass values include the core plus atmosphere within the smaller of $\RB$ or $\RH$, which for these cases is $\RB$.
-The 9.22 $M_{\oplus}$ solution is the highest mass we can reach in our evolutionary sequence, as we explain further in \S\ref{sec:endoftime}.
+The 9.24 $M_{\oplus}$ solution is the highest mass we can reach in our evolutionary sequence, as we explain further in \S\ref{sec:endoftime}.
 
@@
 In agreement with \Eq{eq:cb2}, the radiative zone remains nearly isothermal -- even in the more massive solutions -- and thus shows a nearly exponential increase in pressure with inverse depth.
+
+The atmosphere starts out as purely convective and isentropic, as seen in \Fig{fig:profiles} for the 5.01 $M_{\oplus}$ solution which has no radiative zone.
+Further gas accumulation brings in energy that has to be radiated away, forming a radiative region that quickly deepens.
+The fast pressure deepening of the radiative region is evident in \Fig{fig:profiles} --- even a low mass atmosphere (i.e.
+the 5.10 $M_{\oplus}$ solution) has an extended radiative zone.
 
@@
 Thus accounting for the higher enclosed mass gives a somewhat higher luminosity in the self-gravitating model.
-However the main effect is the higher temperature in the self-gravitating models.
+However, the main effect is the higher temperature in the self-gravitating models.
 The flatter temperature profiles with radius, as described above, explain why the self-gravitating models are hotter.
@@
 
-\Fig{fig:LtvsMopacity} shows the effect of decreasing the atmospheric dust abundance on the atmospheric evolution.
-The opacity normalization in \Eq{eq:opacitylaw} is reduced by factors of 10 and 100.
+Before discussing time evolution (both the elapsed time from the bottom panel of \Fig{fig:Ltplot} and the characteristic growth time of the atmosphere), we briefly discuss the effect of decreasing the atmospheric dust abundance on the atmospheric evolution.
+In \Fig{fig:LtvsMopacity}, the opacity normalization in \Eq{eq:opacitylaw} is reduced by factors of 10 and 100.
 Lower opacities result in higher luminosities and faster evolution, a well-established result \citep{HubBod05} that we confirm for our model.
@@
 To be conservative, however, our reference case considers full Solar abundances, but faster evolution is clearly possible.
+The effect of opacity reduction on the critical core mass is deferred to \S\ref{sec:critical}.
 
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mt_profile.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mt_profile_temp.pdf}
 %\vspace{-0.5in}
@@
 Our results are conservative -- in the sense of overestimating growth times -- because the neglected luminosity would give more rapid evolution.
-Moreover our results are reasonably accurate because the bulk of growth time is spent at low masses, when $L_{\rm negl}$ is small and safely neglected.
+Moreover, our results are reasonably accurate because the bulk of growth time is spent at low masses, when $L_{\rm negl}$ is small and safely neglected.
 
@@
 The main practical effect of this behavior is that we halt the evolution where the model breaks down whenever we cannot reach the crossover mass.
-Fortunately the elapsed time to either the crossover mass or the breakdown is well-behaved and, most importantly, dominated by lower atmospheric masses where the model is accurate.
+Fortunately, the elapsed time to either the crossover mass or the breakdown is well-behaved and, most importantly, dominated by lower atmospheric masses where the model is accurate.
 

### 2013-09-11 tex/draft/draft4.tex
updated section 6 in draft4

@@
 \subsection{Hydrodynamic Effects}\label{sec:hydro}
-The neglect of hydrodynamical effects in our model is best discussed in terms of the thermal mass, $M_{\rm th}$, and the lengthscales introduced in \S\ref{sec:scales}.
-In the low mass regime, $M\pla < M_{\rm th}/ \sqrt{3}$, where $\RB < \RH$ we assume that hydrostatic balance holds out to the outer boundary at $\RH$.
+The neglect of hydrodynamical effects in our model is best discussed in terms of the thermal mass, $M_{\rm th}$, and the length scales introduced in \S\ref{sec:scales}.
+In the low mass regime, $M\pla < M_{\rm th}/ \sqrt{3}$, where $\RB < \RH$, we assume that hydrostatic balance holds out to the outer boundary at $\RH$.
 In this low mass regime, \citet[]{Orm13} calculated the flow patterns driven by stellar tides and disk headwinds.
@@
 At yet higher masses accretion instabilities could occur \citep{AylBat12}.
-However in this high mass regime we already know that the spherically symmetric approximation breaks down, as does our approximate cooling model (\emph{discuss elsewhere and cite}).
+However, in this high mass regime we already know that the spherically symmetric approximation breaks down (see \S\ref{sec:scales}).
+%, as does our approximate cooling model (\emph{discuss elsewhere and cite}).
 
@@
 
-\subsection{Realistic Opacities and EOS}\label{sec:opEOS}
-\emph{some comments on sublimation and radiative windows here (current text rough). upshot is the outer disk is good.}
+\subsection{Realistic Opacities}\label{sec:op}
+%\emph{some comments on sublimation and radiative windows here (current text rough). upshot is the outer disk is good.}
 
@@
 Opacities drop by order unity when ice grains sublimate for $T \gtrsim 150$ K and they drop by orders of magnitude when silicate grains evaporate above $T \gtrsim 1500$ K \citep{semenov03, FerAle05}.
-While protoplanetary atmospheres get significantly hotter than 1500 K, our model depends on opacity only in the upper radiative zones, which are cool enough to remain dusty.
+Our model assumes that radiative regions only exist in the outer part of the atmosphere.
+As such, even though protoplanetary atmospheres get significantly hotter than 1500 K, our model depends on opacity only in the upper radiative zones, which are cool enough to remain dusty.
 For our opacity approximations to be valid, we must restrict our model to the lower temperatures of the outer disk.
 
-A second radiative zone inside the convection zone, a ``radiative window," is possible.
-Dust sublimation at shallow depths -- giving a low $\kappa$ and $P$ in \Eq{eq:delrad} -- could cause such a window.
+A second radiative zone inside the convection zone, a ``radiative window," is possible, due to the opacity drops caused by the evaporation of ice and metal grains.
 Our two layer model ignores such complications.
+Grain growth also impacts the opacity in the outer regions as well as the existence or depth of radiative windows.
+We explore more realistic opacity laws in  In Piso et al.
+(2013, in prep.).
+
+% we show that the existence of inner radiative windows does not impact our simplified model significantly, i.e.
+most of the luminosity is generated in the innermost convective region of the atmosphere and therefore the assumption that luminosity is constant in the radiative layers still holds.
+
+ \subsection{Equation of State}
+ \label{sec:EOS}
+
+ In our model we assume an ideal gas law and a polytropic EOS, given by equations (\ref{eq:idealgas}) and (\ref{eq:polyEOS}), respectively.
+However, non-ideal effects such as partial dissociation and ionization have to be taken into account.
+One solution is using tabulated equation of state tables for hydrogen and helium mixtures.
+In Piso et al.
+(2013, in prep) we use the \citet{saumon95} EOS tables and extend them to lower pressures and temperatures as required by our disk assumptions.
 

### 2013-10-02 tex/draft/draft4.tex
modified paper I intro in draft4

@@
 
-\section{Introduction}
-\label{intro}
-
 %Current theories of giant planet formation postulate that these planets form either through core accretion (refs), in which solid planetesimals collide and grow into a massive solid core, which then accretes a gaseous envelope, or due to a gravitational instability in the protoplanetary disk that leads to fragmentation of the disk into self-gravitating clumps (refs, inc.
@@
 
-Kelvin-Helmholz (KH) contraction considered previously (Ikoma, PapNel05).
+%Kelvin-Helmholz (KH) contraction considered previously (Ikoma, PapNel05).
 Say why useful.
@@
 
-This paper is organized as follows.
+%This paper is organized as follows.
 In Section \ref{sec:model} we describe the assumptions of our atmosphere model, and derive the basic equations that govern the structure and evolution of the atmosphere.
@@
 The appendices contain...
+
+\section{Introduction}
+\label{intro}
+
+Current theories of giant planet formation postulate that these planets form either through core accretion or gravitational instability (GI).
+In the core accretion model (e.g., \citealt{harris78}, \citealt{mizuno78}, \citealt{stevenson82}, \citealt{boden86}, \citealt{wuchterl93}, \citealt{pollack96}, \citealt{dangelo11}), solid planetesimals collide and grow into a massive solid core which then accretes a gaseous envelope.
+GI theories postulate that an instability in the protoplanetary disk leads to the fragmentation of the disk into self-gravitating clumps (e.g., \citealt{harris78}, \citealt{boss97}, \citealt{rafikov05}, \citealt{kratter10}, \citealt{dangelo11}).
+%quote murray-clay, kratter 10, rafikov 05, etc.
+
+%Detailed core accretion studies (\citealt{boden86}, \citealt{pollack96}, \citealt{alibert05}) have shown that this giant planet formation model can be separated in three phases.
+In phase 1 \citep{alibert05} a solid core forms by accreting planetesimals from its feeding zone at a high rate.
+
+Forming giant planets at wide separations in the protoplanetary disk poses theoretical challenges.
+Even though GI is the preferred mechanism at large distances (e.g., \citealt{rafikov05}), fine-tuning is still required in terms of disk conditions and infall history for planetary mass objects to be the outcome of GI \citep{kratter10}.
+Moreover, core accretion times are thought to be too long in the outer disk, largely due to the time it takes to accrete the solid core.
+Some models, however, allow for rapid accretion of solids even at large separations (e.g., \citealt{dones93}, \citealt{kenyon09}, \citealt{lambrechts12}).
+We thus consider the timescale constraints imposed by the accretion of the gaseous atmosphere before the protoplanetary disk dissipates.
+
+%On the one hand, GI generates objects that are too massive to explain the current observed properties of exoplanets (e.g., \citealt{rafikov05}).
+On the other hand, planetesimal accretion is slow at large distances in the disk, and therefore large cores may not be able to form before the dissipation of the gas in the disk (e.g., \citealt{rafikov06}).
+It would therefore be easier if giant planets could form from smaller cores which need less time to grow.
+
+Several types of core accretion models have been studied previously.
+Analytic models (e.g., \citealt{stevenson82}, \citealt{rafikov06}) assume that the core and the atmosphere grow simultaneously, and that the luminosity of the gaseous envelope is entirely supplied by planetesimal accretion.
+These studies consider that the atmosphere is in a steady state at all times, in which all the incoming energy from planetesimal accretion is radiated away by the envelope.
+The core and the atmosphere grow simultaneously, and for a fixed planetesimal accretion rate every core mass maps uniquely to one atmosphere mass.
+As the atmosphere mass becomes comparable to the mass of the solid core, hydrostatic balance no longer holds and runaway gas accretion commences; this process is called the ``core accretion instability''.
+Thus, there is one well-defined core mass past which hydrostatic solutions no longer exist and unstable atmosphere collapse occurs: this is defined as the ``critical core mass''.
+Typically, the critical core mass is found to be about $10 M_{\oplus}$ (e.g., \citealt{stevenson82}, \citealt{rafikov06}).
+
+%Standard static calculations of giant planet formation that consider planetesimal accretion as the main source of luminosity (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{rafikov06}) assume that the atmosphere is at all times in a steady state at which the luminosity due to planetesimal accretion is fully radiated away by the atmosphere.
+Therefore, for a given set of disk conditions, there is a maximum mass the core can have before the onset of unstable accretion: this is defined as the critical core mass.
+
+%    In this situation,
+
+%the atmosphere evolution is dominated by the contraction of the envelope rather than planetesimal accretion.
+The atmosphere is no longer in a steady state, but rather it accretes gas as it loses energy through radiation.
+%In this scenario, the planetesimal accretion rate is significantly slowed down during the gas contraction phase of the atmosphere.
+This reduction can arise, for example, due to dynamical clearing, or due to the core having formed in the inner parts of the disk and migrated outwards.
+In this situation, the atmosphere evolution is dominated by the contraction of the envelope rather than planetesimal accretion.
+
+However, full time dependent simulations for the evolution of giant planets do not assume a constant planetesimal accretion rate, like the studies above, and instead calculate the planetesimal accretion rate as part of an evolutionary model for the core and the envelope.
+In these studies, the atmosphere luminosity is supplied not only by planetesimal accretion, but also by the Kelvin-Helmholtz (KH) contraction of the envelope itself.
+\citet{pollack96} find that planetesimal accretion significantly slows down before a massive gaseous envelope can form, due to depletion of the planet's feeding zone, and gas accretion starts to dominate the atmosphere evolution.
+In this situation, planetesimal accretion can no longer balance radiative losses, and the atmosphere cools and contracts .
+This phase of slow gas accretion is found to be much longer than the formation of the solid core (\citealt{pollack96}, \citealt{alibert05}), hence the evolutionary timescale of the atmosphere is set by KH contraction rather than planetesimal accretion.
+
+Since KH contraction dominates the envelope accretion timescale in many core accretion models, it is useful to consider atmospheric growth onto a core of fixed mass (\citealt{ikoma00}, \citealt{pn05}).
+This is the regime we assume in this work.
+Specifically, we study the atmosphere evolution of giant planets under the assumption that the solid core is fully grown and thus the planetesimal accretion rate is negligible.
+As a result, the luminosity of the gaseous envelope is solely supplied by atmospheric contraction.
+The envelope is not in a steady state, but rather it accretes gas as it cools radiatively.
+We consider that the atmosphere evolves in time through stages of quasi-static equilibrium.
+Once the mass of the gaseous envelope becomes comparable to the mass of the solid core, the self-gravity of the atmosphere can no longer balance the pressure gradient and a rapid phase of runaway gas accretion commences.
+We define the time required for the atmosphere to grow to this stage as the \textit{crossover time}.
+For a set of fixed gas and disk conditions, there exists a minimum core mass for which the atmosphere can grow on the time scale described above within the life time of the protoplanetary disk, which we define as the \textit{critical core mass}.
+
+%The atmosphere is no longer in a steady state, but rather it accretes gas as it loses energy through radiation.
+In our model we therefore assume that the luminosity evolution of the atmosphere is dominated by gas accretion, while the planetesimal accretion rate is negligible.
+As a result, the protoplanetary core has a fixed mass.
+
+%In this study, we show that giant planets can grow faster from small protoplanetary cores that are fully formed before significant gas accretion occurs.
+
+%We build a two-layer atmosphere model, with a convective inner region and a radiative outer region that matches smoothly on to the protoplanetary disk, and develop a cooling model that evolves the atmosphere in time.
+We aim to find the critical core mass for a giant planet to form before disk dissipation.
+
+This paper is organized as follows.
+In section \S\ref{sec:model} we describe the assumptions of our atmosphere model, and derive the basic equations that govern the structure and evolution of the atmosphere.
+In section \S\ref{sec:coolingan}, we present a simplified analytic model that predicts the qualitative behavior of the numerical model.
+We describe our numerical results in section \S\ref{sec:KH}, and determine the critical core mass for planet formation during the life time of the protoplanetary disk in section \S\ref{sec:critical}.
+We discuss some of the approximations of our model in section \S\ref{sec:neglected} and summarize our findings in section \S\ref{sec:conclusions}.
 

### 2013-10-03 tex/draft/draft4.tex
revised paper I intro

@@
 In the core accretion model (e.g., \citealt{harris78}, \citealt{mizuno78}, \citealt{stevenson82}, \citealt{boden86}, \citealt{wuchterl93}, \citealt{pollack96}, \citealt{dangelo11}), solid planetesimals collide and grow into a massive solid core which then accretes a gaseous envelope.
-GI theories postulate that an instability in the protoplanetary disk leads to the fragmentation of the disk into self-gravitating clumps (e.g., \citealt{harris78}, \citealt{boss97}, \citealt{rafikov05}, \citealt{kratter10}, \citealt{dangelo11}).
+GI theories postulate that giant planets can be formed due to the fragmentation of the protoplanetary disk into self-gravitating clumps (e.g., \citealt{harris78}, \citealt{boss97}, \citealt{rafikov05}, \citealt{kratter10}, \citealt{dangelo11}).
 %quote murray-clay, kratter 10, rafikov 05, etc.
@@
 Forming giant planets at wide separations in the protoplanetary disk poses theoretical challenges.
-Even though GI is the preferred mechanism at large distances (e.g., \citealt{rafikov05}), fine-tuning is still required in terms of disk conditions and infall history for planetary mass objects to be the outcome of GI \citep{kratter10}.
-Moreover, core accretion times are thought to be too long in the outer disk, largely due to the time it takes to accrete the solid core.
+Even though GI works better at large distances than in the inner disk (e.g., \citealt{rafikov05}), fine-tuning is still required in terms of disk conditions and infall history for planetary mass objects to be the outcome of GI \citep{kratter10}.
+On the other hand, core accretion times are thought to be too long in the outer disk, largely due to the time it takes to accrete the solid core.
 Some models, however, allow for rapid accretion of solids even at large separations (e.g., \citealt{dones93}, \citealt{kenyon09}, \citealt{lambrechts12}).
@@
 Several types of core accretion models have been studied previously.
-Analytic models (e.g., \citealt{stevenson82}, \citealt{rafikov06}) assume that the core and the atmosphere grow simultaneously, and that the luminosity of the gaseous envelope is entirely supplied by planetesimal accretion.
+In static models, which are mostly analytic, the accretion rate of planetesimals regulates the growth of the atmosphere  (e.g., \citealt{stevenson82}, \citealt{rafikov06}).
+The luminosity of the gaseous envelope is thus entirely supplied by planetesimal accretion.
 These studies consider that the atmosphere is in a steady state at all times, in which all the incoming energy from planetesimal accretion is radiated away by the envelope.
-The core and the atmosphere grow simultaneously, and for a fixed planetesimal accretion rate every core mass maps uniquely to one atmosphere mass.
-As the atmosphere mass becomes comparable to the mass of the solid core, hydrostatic balance no longer holds and runaway gas accretion commences; this process is called the ``core accretion instability''.
-Thus, there is one well-defined core mass past which hydrostatic solutions no longer exist and unstable atmosphere collapse occurs: this is defined as the ``critical core mass''.
+For a fixed planetesimal accretion rate, the atmosphere mass grows non-linearly as the core mass increases.
+As the atmosphere mass becomes comparable to the mass of the solid core, hydrostatic balance no longer holds and the static model breaks down.
+Unstable collapse occurs if the core mass further increases, a process known as the ``core accretion instability''.
+The maximum core mass for which the atmosphere is still in hydrostatic equilibrium is defined as the ``critical core mass''.
 Typically, the critical core mass is found to be about $10 M_{\oplus}$ (e.g., \citealt{stevenson82}, \citealt{rafikov06}).
@@
 
-However, full time dependent simulations for the evolution of giant planets do not assume a constant planetesimal accretion rate, like the studies above, and instead calculate the planetesimal accretion rate as part of an evolutionary model for the core and the envelope.
+Full time dependent simulations of giant planet evolution take into account the Kelvin-Helmholtz (KH) contraction of the gaseous envelope.
+As a result, the atmosphere is no longer in a steady state, but rather it accretes gas as it cools radiatively.
+These evolutionary models for core and atmosphere growth allow for the planetesimal accretion rate to be directly calculated as a function of time rather than prescribed, as it is the case in the static models.
+\citet{pollack96} find that the planet growth can be separated into three relatively well defined phases.
+Phase 1 is dominated by the growth of the core due to runaway planetesimal accretion, and ends when the planet's feeding zone has been depleted of solids.
+At this stage, planetesimal accretion can no longer balance radiative losses, and the atmosphere cools and contracts.
+This is phase 2, during which the envelope mass steadily increases, while core growth is no longer significant.
+Once the atmosphere mass becomes comparable to the core mass, growth is continually accelerated, leading to runaway gas accretion, i.e.
+phase 3.
+Phase 2 is found to be much longer than the other two phases (see also \citealt{alibert05}), hence the evolutionary timescale of the atmosphere is set by KH contraction.
+
+%However, full time dependent simulations for the evolution of giant planets do not assume a constant planetesimal accretion rate, like the studies above, and instead calculate the planetesimal accretion rate as part of an evolutionary model for the core and the envelope.
 In these studies, the atmosphere luminosity is supplied not only by planetesimal accretion, but also by the Kelvin-Helmholtz (KH) contraction of the envelope itself.
@@
 
-Since KH contraction dominates the envelope accretion timescale in many core accretion models, it is useful to consider atmospheric growth onto a core of fixed mass (\citealt{ikoma00}, \citealt{pn05}).
-This is the regime we assume in this work.
-Specifically, we study the atmosphere evolution of giant planets under the assumption that the solid core is fully grown and thus the planetesimal accretion rate is negligible.
-As a result, the luminosity of the gaseous envelope is solely supplied by atmospheric contraction.
+Since KH contraction dominates the envelope accretion timescale in many core accretion models, it is useful to study atmospheric growth onto a core of fixed mass (\citealt{ikoma00}, \citealt{pn05}).
+This regime is considered in this work.
+We explore the atmosphere evolution around fully grown cores of different masses, and determine the minimum core mass needed to form a giant planet before disk dissipation for a variety of disk conditions and properties of the nebular gas.
+We note that our models are a special case of the time dependent studies, with the planetesimal accretion rate set to zero.
+
+% Specifically, we study the atmosphere evolution of giant planets under the assumption that the solid core is fully grown and thus the planetesimal accretion rate is negligible.
+
+%As a result, the luminosity of the gaseous envelope is solely supplied by atmospheric contraction.
 The envelope is not in a steady state, but rather it accretes gas as it cools radiatively.

### 2013-10-05 tex/draft/draft4.tex
updated intro

@@
 
-Current theories of giant planet formation postulate that these planets form either through core accretion or gravitational instability (GI).
-In the core accretion model (e.g., \citealt{harris78}, \citealt{mizuno78}, \citealt{stevenson82}, \citealt{boden86}, \citealt{wuchterl93}, \citealt{pollack96}, \citealt{dangelo11}), solid planetesimals collide and grow into a massive solid core which then accretes a gaseous envelope.
-GI theories postulate that giant planets can be formed due to the fragmentation of the protoplanetary disk into self-gravitating clumps (e.g., \citealt{harris78}, \citealt{boss97}, \citealt{rafikov05}, \citealt{kratter10}, \citealt{dangelo11}).
+Current theories postulate that giant planets form either through gravitational instability (GI) or core accretion (e.g., \citealt{dangelo11}, \citealt{youdin13}).
+In the core accretion model (e.g., \citealt{harris78}, \citealt{mizuno78}, \citealt{stevenson82}, \citealt{boden86}, \citealt{pollack96}), solid planetesimals collide and grow into a massive solid core which then accretes a gaseous envelope.
+GI theories postulate that giant planets can be formed due to the fragmentation of the protoplanetary disk into self-gravitating clumps (e.g., \citealt{boss97}).
 %quote murray-clay, kratter 10, rafikov 05, etc.
@@
 Even though GI works better at large distances than in the inner disk (e.g., \citealt{rafikov05}), fine-tuning is still required in terms of disk conditions and infall history for planetary mass objects to be the outcome of GI \citep{kratter10}.
-On the other hand, core accretion times are thought to be too long in the outer disk, largely due to the time it takes to accrete the solid core.
+On the other hand, the time to accrete the solid core is too long in the outer disk.
 Some models, however, allow for rapid accretion of solids even at large separations (e.g., \citealt{dones93}, \citealt{kenyon09}, \citealt{lambrechts12}).
@@
 As a result, the atmosphere is no longer in a steady state, but rather it accretes gas as it cools radiatively.
-These evolutionary models for core and atmosphere growth allow for the planetesimal accretion rate to be directly calculated as a function of time rather than prescribed, as it is the case in the static models.
+In this scenario there are two main sources that contribute to the luminosity evolution of the envelope --- planetesimal accretion and gas contraction.
+These evolutionary models for core and atmosphere growth include time dependent planetesimal accretion rates that are determined by the distribution of solids in the disk, as well as the planet mass and capture cross-section.
+This is is contrast with the static models in which the planetesimal accretion rate is usually constant with time.
 \citet{pollack96} find that the planet growth can be separated into three relatively well defined phases.
@@
 
+%Aside from dominating the envelope accretion timescale in many core accretion models, KH contraction sets  To investigate these issues, we consider the growth of atmospheres in the limiting case of a fixed core mass with no ongoing accretion of solids.
+We thus obtain absolute lower limits on the critical core mass because additional heat sources limit the ability of the atmosphere to cool and undergo Kelvin-Helmholtz contraction.
+
 Since KH contraction dominates the envelope accretion timescale in many core accretion models, it is useful to study atmospheric growth onto a core of fixed mass (\citealt{ikoma00}, \citealt{pn05}).
-This regime is considered in this work.
-We explore the atmosphere evolution around fully grown cores of different masses, and determine the minimum core mass needed to form a giant planet before disk dissipation for a variety of disk conditions and properties of the nebular gas.
+This regime is the one considered in this work.
+We explore the atmosphere evolution around fully grown cores of different masses, and determine the minimum core mass to form a giant planet before disk dissipation for a variety of disk conditions and properties of the nebular gas.
 We note that our models are a special case of the time dependent studies, with the planetesimal accretion rate set to zero.
+Our neglect of planetesimal accretion allows us to obtain absolute lower limits on the core mass needed to form a gas giant, as additional heat sources limit the ability of the atmosphere to cool and undergo KH contraction.
 

### 2013-10-05 tex/draft/draft3.tex
minor section 2 edits (including eigenvalue discussion)

@@
 \begin{figure}[tb]
-\centering
+%\centering
 \includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cooling_a10_Mc5_rcb_paper.pdf}
@@
 
-\section{Critical Core Mass}
+\section{Results for Giant Planet Formation}
 \label{sec:critical}
+
+We now use our structure and evolution models to calculate the necessary conditions for gas giant formation by core accretion.
+Adopting a median disk lifetime of 3 Myr, we require a core to accrete its atmosphere within the lifetime of the gas disk.
+Our results for atmosphere growth times -- for a range of fixed core masses and disk conditions -- are presented in \S\ref{sec:tcross}.
+Section \ref{sec:critcore} gives our results for critical core masses, the minimum masses which can accrete a massive atmosphere within the disk lifetime.
+
+Our models will focus on giant planet formation between 5 and 100 AU.
+The outer disk is of particular interest for direct imaging searches.
+While the growth of atmospheres close to the star is also important, our cooling model is not appropriate for these hotter atmospheres.
+Their thick radiative zones where dust sublimates would require a different treatment.atmos
 
@@
 
-\subsection{Disk Temperature and Pressure}
+\subsection{Crossover Time}
+\label{sec:tcross}
+The time to grow a massive atmosphere is well characterized by the crossover time, $t_{\rm co}$.
+As in \S\ref{sec:coolingan}, the standard definition of $t_{\rm co}$ is the time for the atmosphere to grow equal to the core mass.
+The precise atmospheric mass threshold used to define $t_{\rm co}$ is not very significant.
+More important is that atmospheric growth starts to accelerate by the chosen mass, as was shown above.
+As explained in \S\ref{sec:endoftime}, we thus can (and do) safely use a lower mass threshold for $t_{\rm co}$ if the cooling model breaks down somewhat before the crossover mass is reached.
+
+%\subsubsection{Effects of Core Mass}
+%\label{Mct}
+
+\Fig{fig:tvsM} displays $t_{\rm{co}}$ for a range of core masses at $a=10$ AU in our fiducial disk.
+Atmospheres grow faster around more massive cores due to stronger gravity.
+Our analytical results in \Eq{eq:tcoolan} under-predict the steepness of the core mass dependence, which is somewhat expected due to the neglect of self-gravity in the analytic model.
+For the parameters of \Fig{fig:tvsM}, the critical core mass that gives $t_{\rm co} = 3$ Myr is between $7$ and $8 \, M_\oplus$.
+Before examining   $M_{\rm crit}$ in more detail, we consider how $t_{\rm co}$ depends on disk and atmosphere parameters.
+
+\begin{figure}[tb]
+%\centering
+\hspace{-.1in}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cumul_coolingtime_vs_Matm_10au_mu235.pdf}
+%\vspace{-0.5in}
+\caption{Time to grow an atmosphere of mass $M_{\rm{atm}}$ for cores with fixed masses between $5 M_{\oplus}$ and $14 M_{\oplus}$ (as labelled) at $a=10$ AU in our fiducial disk.
+The circles mark the crossover time where $M_{\rm{atm}} = M_{\rm c}$.
+Larger core masses give faster atmospheric growth, both to a fixed atmosphere mass and to the crossover mass.}
+%AY say this somewhere? (as in the bottom panel of \Fig{fig:Ltplot})
+\label{fig:tvsM}
+\end{figure}
+
+\subsubsection{Disk Temperature and Pressure}
 \label{sec:TPeffects}
+
+\Fig{fig:TPeffects} shows how the crossover time depends on disk temperature, $T\di$, and pressure, $P\di$.
+While a disk model, like \Eq{eq:diskparam}, gives values for both $T\di$ and $P\di$ vs.\ stellocentric distance, we here examine independent variations of  $T\di$ and $P\di$ to understand these effects separately.
+The left panel of
+
+\begin{figure*}[tb]
+\centering
+\includegraphics[width=1.\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/TdPd_effect.pdf}
+%\vspace{-0.5in}
+\caption{(Left:) Crossover time as a function of disk temperature for fixed disk pressure $P_{\rm d}=0.067$ dyn cm$^{-2}$ given by equation (\ref{eq:Pd}) at 10 AU and a core of fixed mass $M_{\rm c}=5 M_{\oplus}$.
+Right panel: crossover time as a function of disk pressure for fixed disk temperature $T_{\rm d}=44.73$ K given by equation (\ref{eq:diskb}) at 10 AU and a core of fixed mass  $M_{\rm c}=5 M_{\oplus}$.
+The analytic scalings are plotted for comparison, for an effective crossover mass $M_{\rm{co}}=f M_{\rm c}$ with $f=0.13$.
+Gas accretion is faster for lower temperatures and higher pressures.}
+\label{fig:TPeffects}
+\end{figure*}
+
+\subsection{Critical Core Mass}
+\label{sec:critcore}
 

### 2013-10-05 tex/draft/draft4.tex
minor section 2 edits (including eigenvalue discussion)

--- 2013-10-03 tex/draft/draft4.tex
@@
 \item The only source of planetary luminosity is the gravitational contraction of the atmosphere.  
-\item In the radiative zone, luminosity generation is neglected, i.e. the luminosity is held constant.
+\item Cooling of the atmosphere is dominated by the convective interior.
+Thus the luminosity in the radiative exterior is held spatially constant.
+%In the radiative zone, luminosity generation is neglected, i.e.
+the luminosity is held constant.
 \item A global cooling model connects independent static solutions into a time-dependent sequence.
@@
 This choice is justified by the fact that, for hydrostatic solutions, the  density at $\RB$ exceeds the disk's background density by an order unity factor (R06).
-This modest density change has a similarly modest effect on our results.
+Nevertheless, since the density change is modest, the choice of outer boundary condition has a similarly modest effect on our results.
 
@@
 Our simplified models set $\epsilon = 0$, consistent with our neglect of planetesimal accretion luminosity at the base of the atmosphere.
-The $\epsilon_{\rm g} = -T \partial S / \partial t$ term gives the energy input from gravitational contraction.\footnote{In general, any motion in a non-stationary atmosphere is accounted for by this term.
+The $\epsilon_{\rm g} = -T \partial S / \partial t$ term gives the energy input from gravitational contraction.\footnote{In general, any motion is accounted for by this term.
 The partial time derivative is performed on shells of fixed mass.}  The partial time derivative would normally require our radial derivatives to be partial derivatives.
@@
 \end{subeqnarray}
-The specific internal energy $u = C_V T = \Rg (\delad^{-1} -1) T$ for a polytropic EOS.  For a star or coreless planet, $L\co = M\co = 0$.
+The specific internal energy $u = C_V T = \Rg (\delad^{-1} -1) T$ for a polytropic EOS.  For a star or coreless planet, $M\co = 0$.
 
@@
 For an object with no core luminosity (or no core) and no internal heat sources, the energy equation $L_M = -\dot{E}$ describes KH contraction in its simplest form.
-For a main sequence star $L_M = \Gamma$ as the heat of nuclear burning supplies the total luminosity.
-
-For a protoplanetary atmosphere embedded in a gas disk, the full energy equation, 
+When internal heat sources dominate, $L_M = \Gamma$, e.g.\ for nuclear burning in an main sequence star.
+
+A protoplanetary atmosphere embedded in a gas disk lacks a free surface.  For objects without a free surface (or interior to a free surface), the full energy equation, 
 \begin{equation}
@@
 acquires surface terms as derived in  \App{sec:globalderiv}.
+The surface can be at any mass level $M$, where the instantaneous radius is $R$.
+Other surface quantities are labelled by $M$ subscripts.
 The energy accreted across the surface is given by the specific energy, $e_M = u_M-G M/R$, and the mass accretion rate of gas, $\dot{M}$.
 The work done by the surface is $P_M \partial V_M/ \partial t$, with the partial derivative performed at fixed mass.
-This generalized energy equation applies on any spherical shell where hydrostatic balance holds.\footnote{\Eq{eq:coolingglobal} also applies in the interior of objects with a free surface.
+%This generalized energy equation applies on any spherical shell where hydrostatic balance holds.
+The choice of the mass shell $M$\footnote{\Eq{eq:coolingglobal} also applies in the interior of objects with a free surface.
 The work term, $P_M \p V_M/\p t$, vanishes at the free surface.
@@
 For static solutions, which are not the focus of this Paper, the surface terms (and also $\dot{E})$ vanish.
-Static solutions are valid when imposed heat sources, i.e.\ $L\co$ and $\Gamma$ exceed the atmospheric losses.
-Quantitatively, static solutions apply when the evolutionary timescale exceeds  the KH timescale,
+Static solutions are valid when imposed heat sources, i.e.\ $L\co$ and $\Gamma$, exceed the atmospheric losses.
+Quantitatively, static solutions apply when the KH timescale,
 \begin{equation}
@@
 \end{equation} 
-where surface terms are assumed to be subdominant.
-Thus evolutionary calculations, including the quasistatic calculations of this Paper, are needed to consider the fastest possible evolution that occurs on $\tau_{\rm KH}$.
+is shorter than the actual evolutionary timescale.  Thus, $\tau_{\rm KH}$, which our models calculate, gives strong lower limits on the time to form giant planets by core accretion.
 
@@
 With this approximation we can construct solutions from equations (\ref{eq:struct}a -- c) that ``follow the mass," i.e.\ gradually increase the atmospheric mass.
-We then use the global energy balance, \Eq{eq:coolingglobal} to connect these solution in a cooling sequence.
+We then use the global energy balance, \Eq{eq:coolingglobal} to place these solutions in a cooling sequence.
 The validity of neglecting luminosity generation in the radiative zone can be checked \emph{a posteriori}.
@@
 To integrate equations (\ref{eq:struct}a--c), the luminosity is required to compute $\delrad$.
-The correct value of the luminosity is the eigenvalue of the problem, which we find by the shooting method \citep{NR}.
-Only for the eigenvalue of luminosity does the integrated value of mass at the core, $m(R\co)$, match the actual core mass, $M\co$.
+The correct value of the luminosity is not known in advance, and is the eigenvalue of the problem.
+The boundary conditions can only be satisfied for the correct value of the luminosity eigenvalue, which we find by the shooting method.
+Specifically, we alter the luminosity until the integrated value of mass at the core, $m(R\co)$, matches the actual core mass, $M\co$.
 
 To establish the time difference between neighboring solutions, we apply \Eq{eq:coolingglobal} at the radiative-convective boundary (RCB) of the solutions.
-Using exact solutions, energy balance could be evaluated at any level.
-The approximation of constant luminosity in the radiative zone makes the RCB  the self-consistent choice.
-After setting $\Gamma = L\co = 0$, we solve for the elapsed time $\Delta t$  between states $i$ and $i +1$ by finite differencing to give
+In principle, energy balance could be evaluated at any level.
+Our approximate treatment of the radiative zone makes the RCB  the preferred location.
+With $\Gamma = L\co = 0$,  the elapsed time $\Delta t$  between states $i$ and $i +1$ is given by the finite difference
 \begin{equation}
@@
 \end{equation} 
-Brackets indicate an average of, and Delta ($\Delta$) indicates a difference between, neighboring solutions.
+Brackets indicate an average of, and ``$\Delta$" indicates a difference between, the two states.
 All values are evaluated at the RCB.

### 2013-10-05 python/RadSGPoly/shoot2.py

@@ -26,17 +26,19 @@ Loosely there are (at least) 6 levels of founctions here:
 import sys
 import numpy as np
 from utils.constants import G, Me, Re, Msun, sigma, Pdisk, Tdisk, RHill, \
-                            kdust, AU, Myr
-from utils.parameters import R, FT, FSigma, mstar, rhoc, a
-from utils.mynumsci import nextguess, zbrac, mylogspace, ezroot
+                            kdust, AU, Myr, kb, mp
+from utils.parameters import FT, FSigma, mstar, rhoc, a
+from utils.mynumsci import nextguess, zbrac, mylogspace, ezroot, ngQuad
 from scipy.integrate import odeint
 from scipy.optimize import brentq, newton, root, fsolve
 from scipy.interpolate import interp1d
 from collections import namedtuple
 from utils.userpath import userpath, aydat
-import pickle
 #datfolder = userpath + "/dat/MODELS/RadSGPoly2/" 
 
+
+mu = 2.35
+R = kb / (mp*mu)
 p4 = 4*np.pi
 mcore  = 5*Me 
 rcore = (3*mcore/p4/rhoc)**(1./3)
@@ -53,6 +55,15 @@ Previ = namedtuple('Previ',
                    'Mmax, lL, T_m, P_m, r, L, Sold, Mconv, Mcint')
 Previunterp = namedtuple('Previunterp', 
                          'Mmax, lL, T, P, m, r, L, Sold, Mconv, Mcint')
+
+def prmsSet(mc, a=a, rhoc=rhoc):
+    """create prms with different core mass, radius or disk radius.  
+    Other params TODO"""
+
+    rc = (3*mc/p4/rhoc)**(1./3)
+    return Params(mc, rc, a, R, 2./7, Mstar=mstar*Msun, match='Hill',
+              Po=Pdisk(a,mstar, FSigma, FT), To=Tdisk(a,FT), kappa=kdust)
+
 
 def evostruct_r(y, r, dt, prev, prms=prms, verbose=0):
     """
@@ -480,7 +491,7 @@ def makeprev(M, lL=0, dt=0, prev=0, prms=prms, ngrid=1000):
     Rout = rout(M, prms)
     rgrid = mylogspace(Rout, prms.rco, ngrid)
     if lL == 0:
-        lL = lLtop(M, 24, 25)
+        lL = lLtop(M, 24, 25, prms=prms)
     if dt == 0:
[…]

### 2013-10-05 python/utils/mynumsci.py

@@ -31,11 +31,20 @@ def nextguess(xnew, ys, xs, maxdy = 0):
             dypred = maxdy*abs(ys[1] - ys[0])
     return ys[1] + dypred
 
+def ngQuad(xnew, ys, xs):
+    """
+    Quadratic extrapolation (or interpolation).  Requires 3 values of xs and 
+    3 (or 3, n array) of `ys`"""
+
+    poly = np.polyfit(xs, ys, 2)
+    return poly[0] * xnew**2 + poly[1] * xnew + poly[2]
+
 class nf(float):
     """
     Defines a class that forces representation of float to have only one digit
     after decimal.  Also removes trailing zero so '1.0' becomes '1'.
-    Found here: http://matplotlib.org/examples/pylab_examples/contour_label_demo.html
+    Found here: 
+        http://matplotlib.org/examples/pylab_examples/contour_label_demo.html
     """
     def __repr__(self):
         str = '%.1f' % (self.__float__(),)

### 2013-10-08 tex/draft/draft2.tex
merge commit

@@
 
-Current theories of giant planet formation postulate that these planets form either through core accretion (e.g., \citealt{harris78}, \citealt{mizuno78}, \citealt{stevenson82}, \citealt{boden86}, \citealt{wuchterl93}, \citealt{pollack96}, \citealt{dangelo11}), where solid planetesimals collide and grow into a massive solid core which then accretes a gaseous envelope, or due to a gravitational instability in the protoplanetary disk that leads to fragmentation of the disk into self-gravitating clumps (e.g., \citealt{harris78}, \citealt{boss97}, \citealt{rafikov05}, \citealt{kratter10}, \citealt{dangelo11}).
+Current theories of giant planet formation postulate that these planets form either through core accretion or gravitational instability (GI).
+In the core accretion model (e.g., \citealt{harris78}, \citealt{mizuno78}, \citealt{stevenson82}, \citealt{boden86}, \citealt{wuchterl93}, \citealt{pollack96}, \citealt{dangelo11}), solid planetesimals collide and grow into a massive solid core which then accretes a gaseous envelope.
+GI theories postulate that an instability in the protoplanetary disk leads to the fragmentation of the disk into self-gravitating clumps (e.g., \citealt{harris78}, \citealt{boss97}, \citealt{rafikov05}, \citealt{kratter10}, \citealt{dangelo11}).
 %quote murray-clay, kratter 10, rafikov 05, etc.
@@
 Forming giant planets at wide separations in the protoplanetary disk poses theoretical challenges.
-On the one hand, gravitational instability generates objects that are too massive to explain the current observed properties of exoplanets (e.g., \citealt{rafikov05}).
+Even though GI is the preferred mechanism at large distances (e.g., \citealt{rafikov05}), fine-tuning is still required in terms of disk conditions and infall history for planetary mass objects to be the outcome of GI \citep{kratter10}.
+Moreover, core accretion times are thought to be too long in the outer disk, largely due to the time it takes to accrete the solid core.
+Some models, however, allow for rapid accretion of solids even at large separations (e.g., \citealt{dones93}, \citealt{kenyon09}, \citealt{lambrechts12}).
+We thus consider the timescale constraints imposed by the accretion of the gaseous atmosphere before the protoplanetary disk dissipates.
+
+%On the one hand, GI generates objects that are too massive to explain the current observed properties of exoplanets (e.g., \citealt{rafikov05}).
 On the other hand, planetesimal accretion is slow at large distances in the disk, and therefore large cores may not be able to form before the dissipation of the gas in the disk (e.g., \citealt{rafikov06}).
@@
 
-Analytic core accretion models (e.g., \citealt{stevenson82}, \citealt{rafikov06}) assume that the core and the atmosphere grow simultaneously, and that planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse.
-These studies consider that the planet atmosphere is in a steady state at all times, in which all the incoming energy from planetesimal accretion is radiated away by the envelope.
-The core and the atmosphere grow simultaneously, and the mass of the envelope is a function of the core mass.
-As the atmosphere mass becomes comparable to the mass of the solid core, hydrostatic balance no longer holds and runaway gas accretion commences.
-For a set of disk conditions and an assumed planetesimal accretion rate it is therefore found that there is one well-defined core mass past which hydrostatic solutions no longer exist and unstable atmosphere collapse occurs: this is defined as the ``critical core mass''.
+Several types of core accretion models have been studied previously.
+Analytic models (e.g., \citealt{stevenson82}, \citealt{rafikov06}) assume that the core and the atmosphere grow simultaneously, and that the luminosity of the gaseous envelope is entirely supplied by planetesimal accretion.
+These studies consider that the atmosphere is in a steady state at all times, in which all the incoming energy from planetesimal accretion is radiated away by the envelope.
+The core and the atmosphere grow simultaneously, and for a fixed planetesimal accretion rate every core mass maps uniquely to one atmosphere mass.
+As the atmosphere mass becomes comparable to the mass of the solid core, hydrostatic balance no longer holds and runaway gas accretion commences; this process is called the ``core accretion instability''.
+Thus, there is one well-defined core mass past which hydrostatic solutions no longer exist and unstable atmosphere collapse occurs: this is defined as the ``critical core mass''.
 Typically, the critical core mass is found to be about $10 M_{\oplus}$ (e.g., \citealt{stevenson82}, \citealt{rafikov06}).
@@
 
-However, full time dependent simulations for the evolution of giant planets do not assume a constant planetesimal accretion rate, like the studies above, and instead calculate the planetesimal accretion rate as part of an evolutionary model for the core and atmosphere.
+However, full time dependent simulations for the evolution of giant planets do not assume a constant planetesimal accretion rate, like the studies above, and instead calculate the planetesimal accretion rate as part of an evolutionary model for the core and the envelope.
+In these studies, the atmosphere luminosity is supplied not only by planetesimal accretion, but also by the Kelvin-Helmholtz (KH) contraction of the envelope itself.
 \citet{pollack96} find that planetesimal accretion significantly slows down before a massive gaseous envelope can form, due to depletion of the planet's feeding zone, and gas accretion starts to dominate the atmosphere evolution.
-In this situation, planetesimal accretion can no longer balance radiative losses, and instead the atmosphere gains energy from the Kelvin-Helmholtz contraction of the envelope.
-This phase of slow gas accretion is found to be much longer than the formation of the solid core (\citealt{pollack96}, \citealt{alibert05}), hence the evolutionary timescale of the atmosphere is set by Kelvin-Helmholtz contraction rather than planetesimal accretion.
-
-In this study, we show that giant planets can grow faster from small protoplanetary cores that are fully formed before significant gas accretion occurs.
-The atmosphere is no longer in a steady state, but rather it accretes gas as it loses energy through radiation.
-In our model we therefore assume that the luminosity evolution of the atmosphere is dominated by gas contraction, while the planetesimal accretion rate is negligible.
-As a result, the protoplanetary core has a fixed mass.
+In this situation, planetesimal accretion can no longer balance radiative losses, and the atmosphere cools and contracts .
+This phase of slow gas accretion is found to be much longer than the formation of the solid core (\citealt{pollack96}, \citealt{alibert05}), hence the evolutionary timescale of the atmosphere is set by KH contraction rather than planetesimal accretion.
+
+Since KH contraction dominates the envelope accretion timescale in many core accretion models, it is useful to consider atmospheric growth onto a core of fixed mass (\citealt{ikoma00}, \citealt{pn05}).
+This is the regime we assume in this work.
+Specifically, we study the atmosphere evolution of giant planets under the assumption that the solid core is fully grown and thus the planetesimal accretion rate is negligible.
+As a result, the luminosity of the gaseous envelope is solely supplied by atmospheric contraction.
+The envelope is not in a steady state, but rather it accretes gas as it cools radiatively.
 We consider that the atmosphere evolves in time through stages of quasi-static equilibrium.
-Once the mass of the gaseous envelope becomes comparable to the mass of the solid core, the self-gravity of the atmosphere can no longer balance the pressure gradient and unstable hydrodynamic collapse commences.
+Once the mass of the gaseous envelope becomes comparable to the mass of the solid core, the self-gravity of the atmosphere can no longer balance the pressure gradient and a rapid phase of runaway gas accretion commences.
 We define the time required for the atmosphere to grow to this stage as the \textit{crossover time}.
@@
 
-We build a two-layer atmosphere model, with a convective inner region and a radiative outer region that matches smoothly on to the protoplanetary disk, and develop a cooling model that evolves the atmosphere in time.
+%The atmosphere is no longer in a steady state, but rather it accretes gas as it loses energy through radiation.
+In our model we therefore assume that the luminosity evolution of the atmosphere is dominated by gas accretion, while the planetesimal accretion rate is negligible.
+As a result, the protoplanetary core has a fixed mass.
+
+%In this study, we show that giant planets can grow faster from small protoplanetary cores that are fully formed before significant gas accretion occurs.
+
+%We build a two-layer atmosphere model, with a convective inner region and a radiative outer region that matches smoothly on to the protoplanetary disk, and develop a cooling model that evolves the atmosphere in time.
 We aim to find the critical core mass for a giant planet to form before disk dissipation.

### 2013-10-08 tex/draft_paper2/draft_paper2.tex
merge commit

@@
 The two regions are separated by the Schwarzschild criterion for convective instability.
-We denote the surface between the two regions as the radiative-convective boundary (RCB), defined by a radius $r=R_{\cb}$.
-We assume a low planetesimal accretion regime in which the atmosphere evolution is dominated by Kelvin-Helmholtz contraction while planetesimal accretion is neglected.
+We denote the surface between the two regions as the radiative-convective boundary (RCB), defined by a radius $r=R\cb$.
+We assume a low planetesimal accretion regime in which the atmosphere evolution is dominated by KH contraction while planetesimal accretion is neglected.
 Moreover, we assume that the luminosity is constant throughout the outer radiative region.
@@
 \noindent with $\delad$ the adiabatic temperature gradient.
-The convective and radiative layers of the envelope are separated by the Schwarzschild criterion (e.g., \citealt{thompson06}): the atmosphere is stable against convection when $\nabla < \delad$ and convectively unstable when $\nabla < \delad$.
+The convective and radiative layers of the envelope are separated by the Schwarzschild criterion (e.g., \citealt{thompson06}): the atmosphere is stable against convection when $\nabla < \delad$ and convectively unstable when $\nabla > \delad$.
 For effective convection, $\nabla \approx \delad$.
@@
 
-The equation set (\ref{eq:struct}) is supplemented by an equation of state relating pressure, temperature and density, as well as an opacity law.
+The equation set (\ref{eq:struct}) is supplemented by an equation of state (EOS) relating pressure, temperature and density, as well as an opacity law.
 In this study we use the interpolated EOS tables of \citet{saumon95} for a helium mass fraction $Y=0.3$.
@@
 $a \geq 5$ AU.
-Moreover, dust settling or grain growth will result in different numerical values for the $\kappa_0$ and $\beta$ coefficients.
+Dust settling or grain growth will result in different numerical values for the $\kappa_0$ and $\beta$ coefficients.
 
@@
 
-As such, we can obtain an evolutionary series for the atmosphere by connecting sets of subsequent static atmospheres through the cooling equation (\ref{eq:coolingglobal}).
+We obtain an evolutionary series for the atmosphere by connecting sets of subsequent static atmospheres through the cooling equation (\ref{eq:coolingglobal}).
 Details of our numerical procedure are described in PY13.
@@
 
-In this section we explain the differences in behavior of thermodynamic variables between an ideal gas polytrope and a gas described by a realistic equation of state by studying the dependence of the adiabatic gradient (defined in equation \ref{eq:delad}) on temperature and pressure.
-Figure \ref{fig:deladmap} shows a contour plot of the adiabatic gradient as a function of gas temperature and pressure, which was obtained by interpolating and extending the \citet{saumon95} equation of state tables as referenced in section \ref{sec2} and described in \App{EOStables}.
+This section explains how non-ideal EOS effects influence the behavior of the thermodynamic variables when compared to an ideal gas polytrope.
+Figure \ref{fig:deladmap} shows a contour plot of the adiabatic gradient (defined in equation \ref{eq:delad})  as a function of gas temperature and pressure, which was obtained by interpolating and extending the \citet{saumon95} EOS tables as described in \App{EOStables}.
 
@@
 \begin{enumerate}
-\item Intermediate temperature regime (300 K $\lesssim T \lesssim 3000$ K), where the hydrogen-helium mixture behaves like an ideal gas characterized by a polytropic equation of state.
+\item Intermediate temperature regime (300 K $\lesssim T \lesssim 3000$ K), where the hydrogen-helium mixture behaves like an ideal gas with a polytropic EOS.
 \item High temperature regime ($T \gtrsim$ 3000 K), where dissociation of molecular hydrogen occurs, followed by ionization of atomic hydrogen.
-\item Low temperature regime ($T \lesssim 300$ K), where the temperature becomes low enough for rotational motion to reduce.
+\item Low temperature regime ($T \lesssim 300$ K), where the rotational states of the hydrogen molecule are not fully excited. % temperature is low enough for rotational motion to reduce.
 \end{enumerate}
 
-We note that helium behaves like an ideal monatomic gas in our regime of interest ($\delad=2/5$).
-As such, its presence in the atmosphere only causes a small, constant upper shift in the adiabatic gradient of the mixture.
+We note that helium behaves like an ideal monatomic gas with \delad=2/5$ in our regime of interest.
+Its presence in the atmosphere thus only causes a small, constant upper shift in the adiabatic gradient of the mixture.
 
@@
 The black curves represent constant entropy adiabats, with the labels the natural logarithm of the absolute entropy per unit mass.
-At high temperatures, hydrogen dissociates and ionizes, while at low temperatures the rotational states of the hydrogen molecule are only partially excited and it therefore no longer behaves like an ideal diatomic gas.
-The discontinuity between the original EOS tables and our extension is due to the fact that we did not take into account dissociation and ionization.
-However, in our regime of interest (encompassed by the constant entropy curves), the extension matches smoothly with the original tables.}
+At high temperatures, hydrogen dissociates and ionizes, while at low temperatures the rotational states of the hydrogen molecule are only partially excited and it no longer behaves like an ideal diatomic gas.
+Our extension matches smoothly with the original tables in our regime of interest (encompassed by the constant entropy curves).}
 \label{fig:deladmap}
@@
 
-At low temperatures, hydrogen exists in molecular form, which has a stable configuration.
-As the temperature becomes higher than $T \sim 2000-3000$ K, the internal energy becomes large enough to break the covalent bond between the atoms, and hydrogen starts dissociating.
-At temperatures of the order of $10^4$ K, the internal energy becomes large enough to remove electrons from the atoms, and hydrogen ionizes.
+At low temperatures, hydrogen exists in molecular form and has a stable configuration.
+As the temperature becomes higher than $T \sim 2000-3000$ K, the internal energy of the hydrogen molecule becomes large enough to break the covalent bond between the atoms, and hydrogen starts dissociating.
+Further energy increase at temperatures of $\sim 10^4$ K results in electrons being removed from atoms, i.e.
+hydrogen ionizes.
 In stellar and giant planet interiors there is little overlap between the two processes: hydrogen is almost entirely dissociated into atoms by the time ionization becomes important.
 
-For a mixture of molecular and atomic hydrogen, we expect the adiabatic gradient to have an intermediate value between monatomic and diatomic gas, while for a mixture of protons and electrons the adiabatic index is just $2/5$ as for a monatomic ideal gas.
+We see in Figure \ref{fig:deladmap} that the adiabatic gradient decreases significantly in regions of partial dissociation and partial ionization.
+This behavior is different than that of a mixture of molecular and atomic hydrogen (a mixture of protons and electrons), where $2/7<\delad<2/5$ ($\delad=2/5$).
+We  explain this in what follows.
+% where hydrogen is either partially dissociated or partially ionized.
+
+%For a mixture of molecular and atomic hydrogen, we expect the adiabatic gradient to have an intermediate value between monatomic and diatomic gas, while for a mixture of protons and electrons the adiabatic index is just $2/5$ as for a monatomic ideal gas.
 However, we notice in Fig.
@@
 
-We first discuss ionization.
 For a mixture of ideal gases, the total internal energy is given by the sum of the internal energies of the individual gases.
-When a gas is ionized, however, the energy used to ionize the atoms has to also be taken into account.
-This energy depends on the ionization fraction and can be determined from the Saha equation (see e.g., \citealt{kippenhahn90}).
-The ionization fraction only depends on gas temperature and density (see \App{deladioniz}), and hence only on the equation of state.
-An expression of the adiabatic gradient as a function of the ionization fraction can be derived.
-We present this expression in \App{deladioniz}.
-The adiabatic gradient is $\delad=2/5$ when there is no ionization (i.e., only atomic hydrogen) or when the plasma is fully ionized, but decreases significantly during partial ionization, reaching a minimum when half of the gas is ionized.
+When a gas dissociates, however, the energy used to break down the molecules has to also be taken into account.
+This energy depends on the dissociation fraction and can be determined from the Saha equation (see e.g., \citealt{kippenhahn90}).
+The dissociation fraction only depends on gas temperature and density (see \App{deladioniz}), and hence only on the EOS.
+An expression for the adiabatic gradient as a function of the dissociation fraction is presented in \App{deladioniz}.
+As expected, the adiabatic gradient is $\delad=2/7$ for pure molecular hydrogen and $\delad=2/5$ when hydrogen is fully dissociated, but decreases significantly during partial dissociation and it is smallest when half of the gas is dissociated.
 
 This is consistent with the behavior we see in Figure \ref{fig:deladmap}.
-At constant entropy, we expect an increase in pressure to increase the internal energy of the system, thus causing the temperature to also rise significantly.
-In the case of partial ionization, however, part of the internal energy is used to remove the electrons from atoms, and therefore there is less energy available to increase the temperature of the system.
-This behavior of the constant entropy curves is seen in Figure \ref{fig:deladmap}.
-
-The dissociation of molecular hydrogen is dictated by an equation similar to the Saha equation, with the ionization energy replaced by dissociation energy (equal to 4.27 eV for molecular hydrogen, e.g.
-\citealt{mandl89}).
-The adiabatic gradient therefore has an analogous behavior, consistent with Fig.
+At constant entropy, an increase in pressure results in an increase in the internal energy of the system, which usually causes the temperature to rise significantly.
+During partial dissociation, however, part of the internal energy is used to remove the electrons from atoms, and therefore there is less energy available to increase the temperature of the system.
+This behavior of constant entropy curves is seen in Figure \ref{fig:deladmap}.
+
+The ionization of atomic hydrogen is also dictated by the Saha equation, with the dissociation energy replaced by ionization energy (equal to 13.6 eV for hydrogen, e.g.
+\citealt{mandl89}), hence the adiabatic gradient has an analogous behavior, consistent with Fig.
 \ref{fig:deladmap}.
@@
 As a diatomic molecule, hydrogen has five degrees of freedom, three associated with translational motion and two associated with rotation.
-At room temperature, the rotational states are fully excited and the full rotational effects are seen.
+At room temperature, the rotational states are fully excited.
 The excitation temperature for rotation is $\Theta_r \approx 85$ K for the hydrogen molecule (e.g., \citealt{kittel}); as the gas temperature becomes comparable to $\Theta_r$, fewer rotational states are excited and rotation entirely ceases as $T \rightarrow 0$.
@@
 
-In this section we discuss the quantum effects of the hydrogen isomeric forms and the way they affect the rotational energy and heat capacity of the hydrogen molecule at low temperatures, and hence the adiabatic gradient.
+%In this section we discuss the quantum effects of the hydrogen isomeric forms and the way they affect the rotational energy and heat capacity of the hydrogen molecule at low temperatures, and hence the adiabatic gradient.
 
@@
 Together, they result in changes in the growth time of the atmosphere, and therefore in the crossover time and critical core mass (both defined in section \ref{intro}).
-In this section we discuss how the variable adiabatic gradient discussed in section \ref{deladtable} affects the atmosphere evolution when compared to an ideal gas of constant $\delad$.
 We explore the effect of partial dissociation and hydrogen spin isomers (see section \ref{deladtable}) on the atmosphere luminosity and cooling time evolution.
@@
 
-We first discuss the effect of the variable adiabatic gradient on luminosity.
-Figure \ref{fig:ETrplotpoly}, top panel, shows the temperature profile as well as the location of the radiative-convective boundary for the two polytropes at a fixed total mass $M_{\rm{tot}}=11.8 M_{\oplus}$.
-The polytrope with a larger adiabatic gradient has a more shallow convective zone, and hence a deeper radiative region, since a larger temperature gradient delays the onset of convection.
-As the luminosity in the radiative region is inversely proportional to the depth of the radiative zone, a deeper radiative region results in a lower luminosity, which explains the results in the top panel of Figure \ref{fig:Ltplotpoly}.
+%We first discuss the effect of the variable adiabatic gradient on luminosity.
+
+Figure \ref{fig:ETrplotpoly}, top panel, shows the temperature profile for the two polytropes at a fixed total mass $M_{\rm{tot}}=15 M_{\oplus}$.
+The polytrope with a larger adiabatic gradient, $\delad=2/5$, has a more shallow convective zone, and hence a deeper radiative region, since a larger temperature gradient delays the onset of convection.
+A deeper radiative layer increases the number of steps that photons need to diffuse, resulting in a lower luminosity as seen in the top panel of Figure \ref{fig:Ltplotpoly}.
 
@@
 \caption{Instantaneous temperature and total energy profiles as a function of radius for polytropes with different adiabatic indices, for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=10 M_{\oplus}$.
-The total mass (core + atmosphere) is $11.8 M_{\oplus}$.
+The total mass (core + atmosphere) is $15 M_{\oplus}$.
 The location of the radiative-convective boundary is marked.
@@
 
-In spite of the low luminosity of the atmosphere with $\delad=2/5$, the envelope still grows faster than in the case of a diatomic gas, as shown in the bottom panel of Figure \ref{fig:Ltplotpoly}.
+%In spite of the low luminosity of the atmosphere with $\delad=2/5$, the envelope still grows faster than in the case of a diatomic gas, as shown in the bottom panel of Figure \ref{fig:Ltplotpoly}.
 This is due to fact that the amount of energy per unit mass that needs to be radiated away, i.e.
@@
 We have shown in PY13 that polytropes with $\delad=2/7$ have most of the energy concentrated at the bottom of the atmosphere, while polytropes with $\delad=2/5$ have the bulk of the energy towards the outer boundary.
-The bottom panel of Figure \ref{fig:ETrplotpoly} shows an instantaneous energy profile for the same total mass $M_{\rm{tot}} = 11.8 M_{\oplus}$, confirming this behavior.
-It takes more energy to bring in gas deep in the atmosphere for an envelope that has the bulk of its energy concentrated towards the bottom,  resulting in a larger $|dE/dM|$.
-The energy effect prevails over the luminosity effect, resulting in a longer cooling time for the envelope, as shown in the bottom panel of Figure \ref{fig:Ltplotpoly}.
+
+The bottom panel of Figure \ref{fig:ETrplotpoly} shows an instantaneous energy profile for the same total mass $M_{\rm{tot}} = 15 M_{\oplus}$.
+The bulk of the energy is concentrated deep in the atmosphere for the $\delad=2/7$ polytrope and towards the outer boundary for the $\delad=2/5$ polytrope.
+Qualitatively, this can be explained through a simple analytic argument: the density profile in an adiabatic, non-self gravitating atmosphere composed of an ideal gas scales as $\rho(r) \sim r^{1/(\gamma-1)}$ with $\gamma$ the adiabatic index (see PY13).
+Since the energy per unit mass scales as $e(r) \sim \rho(r) r^2$, we find that only polytropes with $\gamma<3/2$, i.e.
+$\delad<1/3$, have the total energy concentrated towards the core.
+This is satisfied by $\delad=2/7$ but not by $\delad=2/5$, in agreement with the results in the bottom panel of Figure \ref{fig:ETrplotpoly}.
+It takes more energy to bring in gas deep in the atmosphere for an envelope that has the bulk of its energy concentrated towards the bottom,  which increases $|dE/dM|$ for the $\delad=2/7$ polytrope.
+
+We find that the energy effect prevails over the luminosity effect, resulting in a longer cooling time for the polytrope with the lower adiabatic gradient (i.e., $\delad=2/7$), as shown in the bottom panel of Figure \ref{fig:Ltplotpoly}.
 
@@
 \centering
-\includegraphics[width=\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/TLr_plot.pdf}
+\includegraphics[width=\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/all_plot.pdf}
 %%\vspace{-0.5in}
@@
 We then determine the critical core mass to form a giant planet before the dissipation of the gas in the protoplanetary disk, assuming that the nebular gas is described by a realistic hydrogen-helium mixture, and we compare this with the results from PY13 for an ideal diatomic gas.
+Finally, we determine the critical core mass under more realistic opacity assumptions.
 
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/t_vs_M_10au.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/tco_vs_Matm_10au.pdf}
 %%\vspace{-0.5in}
-\caption{Time to grow an atmosphere of mass $M_{\rm{atm}}$ for cores with fixed masses between $7 M_{\oplus}$ and $16 M_{\oplus}$ at $a=10$ AU in our fiducial disk.
+\caption{Time to grow an atmosphere of mass $M_{\rm{atm}}$ for cores with fixed masses between $10 M_{\oplus}$ and $20 M_{\oplus}$ at $a=10$ AU in our fiducial disk.
 The circles mark the crossover time where $M_{\rm{atm}} \sim M_{\rm c}$.
@@
 
-\section{Model Relevance in Planet Formation Theory}
+The above results were obtained under the assumption that the dust opacity in the radiative region of the atmosphere is given by the standard ISM opacity (see also section \ref{struct}).
+However, our scenario of low planetesimal accretion is likely to favor lower dust opacities, due to grain growth and dust settling.
+Grain growth, in particular, lowers the absolute value of the opacity and changes the particle size distribution when compared to the standard ISM size distribution (e.g., \citealt{pollack85}).
+
+Although grain growth has been observed in protoplanetary disks (e.g., \citealt{beckwith90}, \citealt{beckwith91}, \citealt{perez12}), the size distribution of dust particles has not been observationally constrained.
+Typically, the grain size distribution is assumed to be a power law:
+
+\begin{equation}
+\label{eq:graindistr}
+n \sim a^{-p},
+\end{equation}
+where $a$ is the particle radius and $p=3.5$ (corresponding to a ``normal'' collisional cascade) or $p=2.5$ (an approximation for coagulation).
+In this work we use the  \citet{dalessio01} frequency-dependent opacity tables to obtain the temperature-dependent Rosseland mean opacity $\kappa$,  assuming a maximum particle size of 1 cm and a grain size distribution given by equation (\ref{eq:graindistr}) with $p=3.5$.
+
+The \citet{dalessio01} opacities are only relevant at temperatures that are sufficiently low for dust grains to remain solid ($T \lesssim 1000$ K).
+In the temperature regime where dust sublimates we use the \citet{bell94} analytic opacity laws, ensuring smooth transition from the grain growth opacities.
+
+The sharp drop in opacity ($\kappa \sim T^{-24}$) due to dust sublimation lowers the radiative temperature gradient significantly (see equation \ref{eq:delrad}), and may therefore generate radiative layers within the inner region of the atmosphere (see \App{radwindow}, Figure \ref{fig:delvsr}).
+This can potentially pose challenges for our model, since the additional luminosity generated in these radiative windows may be large enough to make our assumption of constant luminosity throughout the radiative layers invalid.
+However, the inner radiative windows are either very thin (Figure \ref{fig:delvsr}, middle panel), or the temperature gradient is flat enough to result in a roughly constant luminosity throughout the region (Figure \ref{fig:delvsr}, bottom panel; also see equation \ref{eq:structc}).
+This results in a negligible extra luminosity generated in the radiative windows.
+%moreover...
+(\textbf{add part about sigma T4/tau}).
+
+Despite the existence of one or more radiative windows in the planet atmosphere, the luminosity that emerges at the outer boundary is set by the top radiative window, as is the case for the standard inner convective-outer radiative envelopes.
+The equation of radiative diffusion (\ref{eq:structc}) can be approximated in terms of the radiative flux $\mathcal{F}$ as
+
+\begin{equation}
+\label{eq:sigmatau4}
+\mathcal{F} \sim \frac{\sigma T\cb^4}{\tau},
+\end{equation}
+where $\tau$ is the optical depth of the upper radiative layer. Furthermore, we find that $\tau$ is dominated by the optical depth of the first scale height, i.e.
+
+\begin{equation}
+\label{eq:taucb}
+\tau \sim \tau\cb=\kappa\cb \rho\cb H\cb,
+\end{equation}
+where 
+\begin{equation}
+\label{eq:Hcb}
+H\cb=\frac{k_b T\cb}{\mu m_H} \frac{R\cb^2}{G M\cb},
+\end{equation}
+with $k_b$ the Boltzmann constant, $\mu$ the mean molecular weight of the gas, $m_H$ the proton mass, and all other quantities evaluated at the bottom of the radiative layer.
+By substituting equations (\ref{eq:Hcb}) and (\ref{eq:taucb}) into equation (\ref{eq:sigmatau4}), and noting that $\mathcal{F}=L/(4 \pi (R\cb+H\cb)^2) \approx L / (4 \pi R\cb^2)$, we recover equation (\ref{eq:delrad}) applied at the boundary between the uppermost convective region of the atmosphere and the outer radiative layer (where $\delrad=\delad$).
+
+Due to the variable number and position within the planet atmosphere of radiative windows, and therefore of radiative-convective boundaries, we cannot calculate the time evolution of different atmospheres if we evaluate our cooling equation (\ref{eq:coolingglobal}) at the RCB, as we do in our standard model.
+We choose to evaluate the cooling time at the Bondi radius instead (since our cooling model applies at any radius $R$, see section \ref{struct}).
+We note that our choice of $R$ does not change the estimate of the atmosphere evolution time, to order of magnitude, since the additional luminosity generated in all radiative regions is negligible (see PY13).
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/Mcrit_vs_a_gg.pdf}
+%%\vspace{-0.5in}
+\caption{Critical core mass as a function of semi-major axis for grain growth opacity.}
+\label{fig:Mcritvsagg}
+\end{figure}
+
+Figure \ref{fig:Mcritvsagg} shows the resulting critical core mass as a function of semi-major axis.
+The critical core mass is lower than in the standard case, and less sensitive to the distance in the disk, i.e.
+the boundary conditions (temperature and pressure).
+We have shown in PY13 that the critical core mass is highly dependent on disk temperature rather than pressure.
+The temperature dependence, however, is mainly due to opacity: for the simplified analytic model developed in PY13 we found that the crossover time $t_{\rm co} \sim T^{\beta+1/2}$, with $\beta$ the power-law exponent in equation (\ref{eq:opacitylaw}).
+Opacity is less sensitive to temperature variations for larger grains and has an almost flat profile (see Figure \ref{fig:opacity}), which results in $\beta \ll 1$ and a much weaker temperature (and therefore semi-major axis) dependence of the critical core mass, as seen in Figure \ref{fig:Mcritvsagg}.
+Moreover, grain growth reduces the absolute value of the opacity, which results in an overall lower crossover time and critical core mass.
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/tco_vs_a_Mc4_comp.pdf}
+%%\vspace{-0.5in}
+\caption{Crossover time for different grain size distribution.}
+\label{fig:p25p35}
+\end{figure}
+
+In our calculation so far we assumed that the size distribution of dust grains is the power law (\ref{eq:graindistr}) with $p=3.5$, which corresponds to a normal collisional cascade.
+If, however, coagulation is taken into account, the exponent $p$ can be approximated as $p=2.5$ \citep{dalessio01}.
+This results in a flatter and significantly lower opacity, which may substantially reduce the critical core mass.
+However, we have found that our model breaks down for low core masses ($M_{\rm c} \lesssim 3 M_{\oplus}$) under this assumption, i.e.
+the outer radiative region becomes deep enough that the luminosity generated in this region can no longer be neglected.
+Figure \ref{fig:p25p35} shows the crossover time as a function of semi-major axis for the lowest core mass in which our model is valid, $M_{\rm c}=4 M_{\oplus}$.
+The crossover time is more than one order of magnitude lower when coagulation is accounted for, which implies that the critical core mass may be, in fact, significantly lower than presented in Figure \ref{fig:Mcritvsagg}.
+
+%\section{Model Relevance in Planet Formation Theory}
+\section{Effects of Planetesimal Accretion}
 \label{acc}
@@
 In this study we have considered atmospheres for which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the atmosphere.
-This is different from standard calculations, in which the atmosphere is heated by planetesimal accretion.
-In this section we compare our results for the critical core mass to analogous results from the standard calculations.
+This is different from standard analytic calculations, in which the atmosphere is heated by planetesimal accretion.
+In this section we compare our results for the critical core mass to analogous results from the analytic calculations.
 We discuss the core accretion rates that are necessary for our regime to be valid in section \ref{raf1}.
@@
 For a given atmosphere model we can therefore estimate the maximum planetesimal accretion rate during the contraction of the envelope in order for the atmosphere to be dominated by gas contraction.
-We choose as a fiducial case an atmosphere forming at 40 AU and with a core mass of $10 M_{\oplus}$, described by a realistic equation of state.
-For this choice of parameters, the atmosphere crossover time is $t _{\rm{co}}\sim$ 2.7 Myrs, which is within the typical life time of a protoplanetary disk.
+We choose as a fiducial case an atmosphere forming at 60 AU and with a core mass of $5 M_{\oplus}$.
+Since analytic studies assume an ideal gas EOS, our fiducial atmosphere is described by an ideal gas polytrope with constant adiabatic gradient $\delad=2/7$ and mean molecular weight $\mu=2.35$ (see also PY13).
+For this choice of parameters, the atmosphere crossover time is $t _{\rm{co}}\sim$ 2.1 Myrs, which is within the typical life time of a protoplanetary disk.
 The results are presented in Figure \ref{fig:accrates}.
@@
 %\vspace{-0.5in}
-\caption{Various relevant accretion rates in the case of a planet forming at 10 AU and with a core mass $M\co=10 M_{\oplus}$.
+\caption{Various relevant accretion rates in the case of a planet forming at 60 AU and with a core mass $M\co=5 M_{\oplus}$.
 The $\dot{M}_{\rm{atm}}$ curve represents the growth rate of the atmosphere as estimated by our model, and $\dot{M}_{\rm{c,KH}}$ is the maximum planetesimal accretion rate during the gas contraction phase in order for our regime to be valid.
-For comparison, we plot the core accretion rate $\dot{M}_{\rm{c,acc}}$ necessary to grow the core on the same time scale as the atmosphere $\tau \sim 2.7$ Myrs, and a typically planetesimal accretion rate $\dot{M}_{\rm{c,Hill}}$ where the random velocity of the planetesimals is given by the Hill velocity due to the core.}
+For comparison, we plot the core accretion rate $\dot{M}_{\rm{c,acc}}$ necessary to grow the core on the same time scale as the atmosphere $\tau \sim 2.1$ Myrs, and a typically assumed planetesimal accretion rate $\dot{M}_{\rm{c,Hill}}$ where the random velocity of the planetesimals is given by the Hill velocity due to the core.}
 \label{fig:accrates}
@@
 Possible explanations for that include the core having formed in the inner part of the disk and later migrated outwards, or the core having been depleted of planetesimals due to a giant neighbor.
-We further estimate the core accretion rate needed for the core to form on the same timescale as our model atmosphere, $\tau=2.7$ Myrs:
+We further estimate the core accretion rate needed for the core to form on the same timescale as our model atmosphere, $\tau=2.1$ Myrs:
 
@@
 Next, we are interested in whether hydrodynamic gas accumulation due to planetesimal accretion can already commence before the atmosphere becomes unstable due to Kelvin-Helmholtz contraction, as our regime is no longer the relevant one under such conditions.
-A core that forms on the same timescale as our model atmosphere accretes planetesimals at a rate given by equation (\ref{eq:Mcdot}).
-This accretion rate is dependent on the core mass, which is steadily increasing.
+From equation (\ref{eq:Lacc}), the planetesimal accretion rate during the atmosphere evolution has to satisfy
+\begin{equation}
+\label{eq:McdotKH}
+\dot{M\co} \lesssim \frac{L_{\rm KH} R\co}{G M\co}
+\end{equation}
+for our regime to be valid, i.e.
+$L_{\rm acc} \lesssim L_{L_{\rm KH}}$.
+The maximum planetesimal accretion rate $\dot{M}_{\rm c, KH}$ that satisfies this condition (when $L_{\rm acc}=L_{\rm KH}$) is denoted by the dotted line in Figure \ref{fig:accrates} for $a=60$ AU and $M\co=5M_{\oplus}$.
+If unstable atmosphere collapse does not occur due to planetesimal accretion for the lowest value of $M_{\rm c, KH}$, then runway gas accretion can only occur in the KH dominated regime.
+In what follows we calculate the critical core mass $M_{\rm c, KH}$ corresponding to $\dot{M}_{\rm c, KH}$ and show that it is higher than our calculated critical core mass.
+
+%A core that forms on the same timescale as our model atmosphere accretes planetesimals at a rate given by equation (\ref{eq:Mcdot}).
+
+%This accretion rate is dependent on the core mass, which is steadily increasing.
 We therefore compare the critical core mass due to planetesimal accretion at this rate $M_{\rm{crit,acc}}$ to the critical core mass as defined in our estimates $M_{\rm{c,crit}}$ (see section \ref{critical}).
@@
 
-In order to estimate the critical core mass due to planetesimal accretion $M_{\rm{crit,acc}}$, we use the results of \citet{rafikov06} for low luminosity atmospheres forming in the outer disk ($>2-5$ AU), consistent with our region of interest.
-\citet{rafikov06} assumes an ideal gas polytropic equation of state and a lower opacity than the standard ISM opacity that we use in our calculations (see equation \ref{eq:opacitylaw}), due to grain growth.
-In PY13 we showed that a reduction in opacity results in a shorter crossover time and therefore a lower critical core mass.
-In this section, we calculate the critical core mass for an ideal gas polytrope with the standard ISM opacity reduced by a factor of 100, which is comparable to the opacity law used by \citet{rafikov06}.
+In order to estimate $M_{\rm c, KH}$, we use the results of \citet{rafikov06} for low luminosity atmospheres forming in the outer disk ($>2-5$ AU), consistent with our region of interest.
+\citet{rafikov06} assumes an ideal gas polytropic equation of state and a lower opacity than the standard ISM opacity that we use in our calculations (see equation \ref{eq:opacitylaw}).
+We thus calculate the critical core mass for an ideal gas polytrope with the standard ISM opacity reduced by a factor of 100, which is comparable to the opacity law used by \citet{rafikov06} (see PY13).
 
@@
 \label{eq:critraf}
-M_{\rm{crit, acc}} \sim \Big[\frac{\dot{M}(M_{\rm{c}})}{64 \pi^2 C} \frac{\kappa_0}{\sigma G^3} \frac{1}{R\co M\co^{1/3}} \Big(\frac{k_b}{\mu}\Big)^4\Big]^{3/5},
-\end{equation}
-
-\noindent with all the constants as defined in previous sections, and $C$ a constant depending on the adiabatic gradient and disk properties (see \citealt{rafikov06}, equation B3).
-We calculate $M_{\rm{crit,acc}}$ for a range of core masses.
-The result is displayed in Figure \ref{fig:raf2}.
+M_{\rm{crit, KH}} \sim \Big[\frac{\dot{M}(M_{\rm{c}})}{64 \pi^2 C} \frac{\kappa_0}{\sigma G^3} \frac{1}{R\co M\co^{1/3}} \Big(\frac{k_b}{\mu}\Big)^4\Big]^{3/5},
+\end{equation}
+
+\noindent with all the constants as defined in previous sections, and $C$ an order unity constant depending on the adiabatic gradient and disk properties (see \citealt{rafikov06}, equation B3).
+The result is displayed in Figure \ref{fig:raf2}; the critical core mass corresponding to planetesimal accretion at the rates displayed in Figure \ref{fig:accrates} is displayed for comparison.
 
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/Mc_vs_a_poly_raf2_paper.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/Mc_vs_a_poly_comp.pdf}
 %\vspace{-0.5in}
-\caption{Comparison between the critical core mass $M_{\rm{crit, acc}}$ due to planetesimal accretion and the assumed fixed core mass when gas contraction dominates, for a growth time of $\tau=2.6$ Myrs.
+\caption{Comparison between the critical core mass $M_{\rm{crit, KG}}$ due to planetesimal accretion and the assumed fixed core mass when gas contraction dominates, for a growth time of $\tau=2.1$ Myrs.
 Our results yield lower core masses than in the standard case.}
@@
 
-We see that the critical core mass due to planetesimal accretion is smaller than in the case in which planetesimal accretion dominates the evolution of the atmosphere.
+The critical core mass due to planetesimal accretion is smaller than in the case in which planetesimal accretion dominates the evolution of the atmosphere.
 This brings us to two conclusions.
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/delad_ionization.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/delad_dissociation.pdf}
 %%\vspace{-0.5in}
@@
 
+\section{Radiative Windows} \label{radwindow}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/Opacity/del_vs_r.pdf}
+%%\vspace{-0.5in}
+\caption{Radiative and adiabatic gradient as a function of radius. The sharp drom in opacity due to dust sublimation can generate one or more radiative windows.}
+\label{fig:delvsr}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/kappa_grain_growth_SBL_paper.pdf}
+%%\vspace{-0.5in}
+\caption{Opacities...}
+\label{fig:opacity}
+\end{figure}
+
 \end{document}

### 2013-10-12 tex/draft/draft4.tex
latest draft4 version

@@
 \begin{abstract}
-%The core accretion model proposes that giant planets form by the accretion of gas onto a solid protoplanetary core.
-Previous studies have found that there exists a ``critical core mass'' past which hydrostatic solutions can no longer be found and unstable atmosphere collapse occurs.
-In standard calculations of the critical core mass, planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse.
-In this study we consider the extreme case in which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the planet.
-We develop a two-layer atmosphere model with an inner convective region and an outer radiative zone that matches onto the protoplanetary disk, and we determine the minimum core mass for a giant planet to form within the typical disk life timescale for a variety of disk conditions, which we denote as  ``critical core mass''.
-We find that the absolute minimum core mass required to nucleate atmosphere collapse within the disk lifetime is smaller for planets forming further away from their host stars.
-Moreover, the critical core mass is strongly dependent on disk temperature, opacity and mean molecular weight of the gas.
+
+The core accretion model proposes that giant planets form by the accretion of gas onto a solid protoplanetary core.
+Previous studies have found that there exists a ``critical core mass'' past which hydrostatic solutions no longer exist and unstable atmosphere collapse occurs.
+This core mass is typically quoted to be around $10 M_{\oplus}$.
+In static calculations of the critical core mass, planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse.
+In this study we consider the limiting case in which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the planet.
+We develop a two-layer atmosphere model with an inner convective region and an outer radiative zone that matches onto the protoplanetary disk, and we determine the minimum core mass for a giant planet to form within the typical disk lifetime for a variety of disk conditions.
+We denote this mass as  \textit{critical core mass}.
+The absolute minimum core mass required to nucleate atmosphere collapse is $\sim 8 M_{\oplus}$ at 5 AU and steadily decreases to $\sim 3.5 M_{\oplus}$ at 100 AU, for an ideal diatomic gas with a solar composition and a standard ISM dust opacity.
+Lower opacity and disk temperature significantly reduce the critical core mass, while a decrease in the mean molecular weight of the nebular gas results in a larger critical core mass.
 Our results yield lower mass cores than corresponding studies for large planetesimal accretion rates.
-We therefore show that it is easier to form a planet by growing the core first, then accreting a massive gaseous envelope, rather than forming the core and atmosphere simultaneously.
+As additional heating due to planetesimal accretion leads to slower cooling, our results represent an absolute minimum for the core mass needed to form a giant planet before disk dissipation.
 
@@
 Grain growth tends to lower both $F_\kappa$ and $\beta$, while dust abundance scales with $F_\kappa$.
-Section \ref{sec:opEOS} discusses dust sublimation and more realistic opacity laws.
+Section \S\ref{sec:op} discusses dust sublimation and more realistic opacity laws.
 
@@
 \Eq{eq:structc} simply defines the temperature gradient  $\nabla \equiv d \ln T/d \ln P$.
-In radiative zones radiative diffusion gives a temperature gradient
+In optically thick radiative zones, radiative diffusion gives a temperature gradient
 \begin{equation}
@@
 \end{equation}
-where $\sigma$ is the Stefan-Boltzmann constant.  In convectively unstable regions, efficient convection gives an isentropic temperature gradient with $\nabla = \delad$, the adiabatic gradient
+where $\sigma$ is the Stefan-Boltzmann constant.
+In our models the atmosphere is optically thick throughout the whole radiative region (see \App{opacityan} for an analytic estimate of the optical depth).
+In convectively unstable regions, efficient convection gives an isentropic temperature gradient with $\nabla = \delad$, the adiabatic gradient
 
@@
 
-If the radiative zone were replaced by an isothermal layer, then hydrostatic balance would give $\chi = \theta = 1$.
-Standard radiative structure calculations (see \App{sec:analytic} for details) give
+If the radiative zone were replaced by an isothermal layer, then hydrostatic balance would give $\chi = \theta = 1$.  Standard radiative structure calculations (see \App{RCBcorr} for details) give 
 \begin{equation}
@@
 The most relevant quantities for the global cooling model (equation \ref{eq:coolingglobal}) are the integrated energy, luminosity, and atmospheric mass.
-In our non-self-gravitating limit, the mass of our nearly isothermal radiative zones is less than the convective interior, as shown in \App{sec:analytic}.
+In our non-self-gravitating limit, the mass of our nearly isothermal radiative zones is less than the convective interior, as shown in \App{iso}.
 
@@
 
-Our analytic cooling model uses $L = -\dot{E}$,  neglecting the surface terms in \Eq{eq:coolingglobal}.\footnote{\App{sec:analytic} shows that these terms are negligible for a non-self-gravitating model, but they do become relevant when self-gravity is important and included.}  Applying \Eqs{eq:Eus}{eq:Lcb}, the time it takes to cool the atmosphere until the RCB reaches a given pressure depth, $P\cb$, is
+Our analytic cooling model uses $L = -\dot{E}$,  neglecting the surface terms in \Eq{eq:coolingglobal}.\footnote{\App{surfterms} shows that these terms are negligible for a non-self-gravitating model, but they do become relevant when self-gravity is important and included.}  Applying \Eqs{eq:Eus}{eq:Lcb}, the time it takes to cool the atmosphere until the RCB reaches a given pressure depth, $P\cb$, is
 \begin{subeqnarray} 
@@
 %\vspace{-0.5in}
-%\caption{Luminosity evolution around a $5 M_\oplus$ core at 10 AU.
-In addition to the total model luminosity $L$, the contributing terms in the cooling model of \Eq{eq:coolingglobal}; $-\dot{E}$, $e_{\rm M} \dot{M}$ and $-P \p V/ \p t |_M$; are shown.
-The $L_{\rm negl}$ curve gives the neglected luminosity that would have been generated in the radiative zone, but is not included in our two-layer cooling model.}
+\caption{The relative contribution of the various terms in the cooling \Eq{eq:coolingglobal} to the atmosphere luminosity, for a planet forming around a $5 M_\oplus$ core at 10 AU.
+The $L_{\rm negl}$ curve gives the extra luminosity that would have been generated in the radiative zone, but is not included in our two-layer cooling model.
+The neglected luminosity is lower than the model luminosity $L$ for low atmosphere mass, and becomes comparable at higher masses ($>3 M_{\oplus}$ in this particular case).
+As atmosphere growth is slowest at lower masses, neglecting the additional luminosity generated in the radiative layer yields accurate results, to order of magnitude.}
 \label{fig:coolingterms}
@@
 
-\section{Conclusions} \label{sec:conclusions}
+\section{Summary} \label{sec:conclusions}
+
+ In this paper we study the formation of giant planet atmospheres under the assumption of a low planetesimal accretion regime in which KH contraction dominates the luminosity evolution of the atmosphere.
+We consider atmosphere growth around a core of fixed mass, and we build quasi-static two-layer atmosphere models with an inner convective region and an outer radiative region that matches smoothly onto the protoplanetary disk at the Hill radius.
+In our static models, we use an ideal gas polytropic EOS, and assume a standard ISM power-law opacity for dust grains.
+We derive a cooling model to connect series of quasti-static atmospheres and obtain an evolutionary history of the envelope.
+We define the time $t$ at which unstable atmosphere collapse commences as the \textit{crossover time} $t_{\rm{co}}$, for which $M_{\rm atm}(t_{\rm{co}})\sim M_{\rm c}$.
+From this we define as \textit{critical core mass} the minimum core mass for a protoplanet to initiate runaway gas accretion during the lifetime of the protoplanetary disk.
+We calculate the critical core mass for a variety of disk conditions, nebular gas compositions and opacities.
+The critical core mass is $\sim 8 M_{\oplus}$ at 5 AU in our fiducial disk and decreases to $\sim 3.5 M_{\oplus}$ at 100 AU.
+A larger mean molecular weight of the nebular gas increases the critical core mass, while a factor of ten reduction in opacity results in a critical core mass twice as low.
+
+Our results are lower than the typically quoted critical core mass of 10 $M_{\oplus}$ in static studies.
+A giant planet can thus form faster if the core grows first, then accretion of solids slows down and a massive atmosphere accumulates, rather than if the core and envelope grow simultaneously in a high planetesimal accretion rate regime.
+Moreover, we are able to place robust absolute lower limits on the core mass needed to form a giant planet, since additional core growth and heating limit the ability of the atmosphere to radiate energy and cool.
+
+ %We also find that the critical core mass to form a giant planet within the life time of the disk (of about 3 Myrs) is smaller than the typical 10 $M_{\oplus}$ value yielded by studies that assume that the atmosphere evolution is dominated by the luminosity due to planetesimal accretion.
+As such, it is faster to form a planet by growing the core first in a fast planetesimal accretion regime (e.g., the core forms in the inner disk, then migrates outwards), then significantly reduce planetesimal accretion and allow a massive atmosphere to accumulate.
 
@@
 
-To derive  the global energy equation (\ref{eq:coolingglobal}) for an embedded protoplanets, we generalize the analogous calculations in stellar structure theory, e.g.\ in \S4.3 of \citet{kippenhahn90}.
-For our problem we add the effects of finite core radius, surface pressure and mass accretion.We start with the local energy equation (\ref{eq:structd}) whose more natural form in Lagrangian (mass) coordinates is $\p L/ \p m = \epsilon - T \p S /\p t$.
+To derive  the global energy equation (\ref{eq:coolingglobal}) for an embedded protoplanet, we generalize the analogous calculations in stellar structure theory, e.g.\ in \S4.3 of \citet{kippenhahn90}.
+For our problem, we add the effects of finite core radius, surface pressure and mass accretion.We start with the local energy equation (\ref{eq:structd}) whose more natural form in Lagrangian (mass) coordinates is $\p L/ \p m = \epsilon - T \p S /\p t$.
 Integrating from the core to a higher shell with enclosed mass $M$ gives:
@@
 &=& \int_{M\co}^M \left(\epsilon - T {\p S \over \p t} \right)dm \\
-&=& \Gamma  - \int_{M\co}^M{\p u \over \p t} dm +  \int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm\slabel{eq:DLc}\, .
-\end{subeqnarray} 
-with $\Gamma = \int \epsilon dm$ the integral of the direct heating rate and applying the law of thermodynamics in the final step.
+&=& \Gamma  - \int_{M\co}^M{\p u \over \p t} dm +  \int_{M\co}^M {P \over \rho^2} {\p \rho \over \p t} dm\slabel{eq:DLc}\, ,
+\end{subeqnarray} 
+with $\Gamma = \int \epsilon dm$ the integral of the direct heating rate and applying the first law of thermodynamics in the final step.
 
@@
 in favor of total time derivatives, denoted with overdots.  %The physical distinction is that total derivatives include mass  accreted through the outer boundary.  
-For instance the surface radius, $R$, evolves as  
+For instance, the surface radius $R$ of the shell with enclosed mass $M$ evolves as  
 \begin{equation}\label{eq:Rdot}
- \dot{R} = {\p R \over \p t} + {\dot{M} \over 4 \pi R^2 \rho_M}
+ \dot{R} = {\p R \over \p t} + {\dot{M} \over 4 \pi R^2 \rho_M},
 \end{equation} 
@@
 %The subscript $M$  denotes quantities at the upper boundary of total mass $M$ (though it is omitted from $M$ and $R$).  
-Similarly the volume, $V = (4 \pi/3)r^3$, and pressure at the outer shell evolve as
+Similarly, the volume $V = (4 \pi/3)R^3$ and pressure at the outer shell evolve as
 \begin{subeqnarray}\label{eq:dot}
@@
 \end{equation}
-
 The internal energy integral follows simply from  Leibniz's rule as
@@
 \end{equation} 
-
-To make further progress we require the virial theorem:
+To make further progress we use the virial theorem:
 \begin{equation}
 \label{eq:virial}
-E_G=-3 \int_{M\co}^M \frac{P}{\rho} dm + 4 \pi (R^3 P_M-R\co^3 P\co)
+E_G=-3 \int_{M\co}^M \frac{P}{\rho} dm + 4 \pi (R^3 P_M-R\co^3 P\co),
 \end{equation}
@@
 \section{Analytic Cooling Model Details}\label{sec:analytic}
-TODO (academic): Figure out if theta < 1 is actually required? Should be but need to check
+%TODO (academic): Figure out if theta < 1 is actually required? Should be but need to check
+
+\subsection{Isothermal Atmosphere}
+\label{iso}
+
+We consider the structure of a non self-gravitating isothermal atmosphere that extends outward from the radiative-convective boundary RCB and matches onto the disk density, $\rho_{\rm d}$, at a distance $r_{\rm fit} = n_{\rm fit} \RB$, where $R_{\rm B}$ is the Bondi radius defined in equation (\ref{eq:RB}).
+From equation (\ref{eq:structa}) the resulting density profile is
+\begin{equation} \label{eq:rhoiso}
+\rho = \rho_{\rm d} \exp \left({R_{\rm B} \over r} - {1 \over n_{\rm{fit}}} \right) \approx   \rho_{\rm{d}} \exp \left(R_{\rm B} \over r  \right),
+\end{equation} 
+where the approximate inequality is valid deep inside the atmosphere ($r \ll \RB$) for any $n_{\rm fit} \gtrsim 1$.
+However, the choice of boundary condition does have an order unity effect on the density near the Bondi radius.
+
+The mass of the atmosphere is determined by integrating equation (\ref{eq:structb}) from the RCB to the Bondi radius using the density profile (\ref{eq:rhoiso}) and can be approximated as
+\begin{equation} \label{eq:MatmISO}
+M_{\rm iso} \approx 4 \pi \rho\di {R\cb^4 \over R_{\rm B}} e^{R_B/R\cb} = 4\pi \rho\cb \frac{R\cb^4}{R_{\rm B}}
+\end{equation}
+with $\rho\cb$ the density at the RCB.
+This result is the leading order term in a series expansion.
+By comparing the expression above and \Eq{eq:Matman} under the assumption that $R\cb \ll R_{\rm B}$, we see that the mass of the outer radiative region (which is nearly isothermal) is negligible when compared with the atmosphere mass in the convective layer, as stated in section \S\ref{sec:coolingan}.
+% Furthermore, because the atmospheric scale-height at $R\co$ is $H_\rho = |dr /d\ln\rho| = R\co^2/R_B$, the result is intuitively the correct order of magnitude.
+%Planets can attract massive atmospheres if $\theta\co \equiv R_B/R\co \gg 1$.
+
+\subsection{Temperature and Pressure Corrections at the Radiative-Convective Boundary}
+\label{RCBcorr}
+
+%\subsection{Temperature Contrast at Convective Boundary}
+In this section we estimate the temperature and pressure corrections at the RCB due to the fact that the radiative region is not purely isothermal.
+From equation (\ref{eq:delrad}), we express the radiative lapse rate
+\begin{equation}\label{eq:delradan}
+\delrad = {3 \kappa P \over 64 \pi  G M \sigma T^4} L = \nabla\di {P/P_{\rm d} \over (T/T_{\rm d})^{4-\beta}},
+\end{equation}
+
+\noindent where the second equality follows from the opacity law (\ref{eq:opacitylaw}) and $\nabla_{\rm d}$ is the radiative temperature gradient at the disk:
+
+\begin{equation}
+\label{eq:delo}
+\nabla \di \equiv \frac{3 \kappa(T{\di}) P{\di}}{64 \pi G M \sigma T_{\rm d}^4} L.
+\end{equation}
+Here $M$ is the total planet mass.
+Since our analytic model neglects self-gravity, $M=M\co$ and therefore $\nabla\di$ is constant.
+From equation (\ref{eq:delradan}) and $\delrad=d \ln T / d\ln P$, the temperature profile in the radiative region integrates to
+\begin{equation}\label{eq:radTP}
+\left(T \over T_{\rm d}\right)^{4-\beta} - 1 = {\nabla\di \over \nabla_\infty} \left(\left{P \over P_{\rm d}}\right - 1 \right) \, ,
+\end{equation} 
+where $\nabla_\infty = 1/(4-\beta)$ is the radiative temperature gradient for $T ,P \rightarrow \infty$.
+Applying \Eqs{eq:delradan}{eq:radTP} at the RCB (where $\delrad = \delad$) under the assumption that $P\cb \gg P_{\rm d}$ results in  $T\cb=\chi T\di$ as in \Eq{eq:Tcb}, with $\chi$ defined in \Eq{eq:chi}.
+
+The pressure at the RCB follows from \Eqs{eq:radTP}{eq:Tcb} as
+\begin{equation}
+\label{eq:Pcbapprox}
+{P\cb\over P_{\rm d}} \simeq {\delad/\nabla\di \over 1 - \delad/\nabla_\infty}.
+\end{equation} 
+%This pressure contrast can be quite large since $\nabla_{\rm d} \ll 1$.
+ We can eliminate $\nabla\di$ from equation (\ref{eq:Pcbapprox}) to obtain a relation between temperature and pressure in the radiative zone as a function of the RCB pressure $P\cb$.
+From \Eq{eq:radTP}, it follows that
+ \begin{equation}\label{eq:TP}
+{T \over T_{\rm d}} = \left[1 + {1 \over {\nabla_\infty \over \delad} - 1} \left({P \over P\cb} -  {P_{\rm d} \over P\cb}\right) \right]^{1 \over 4-\beta}\, .
+\end{equation} 
+ We can then determine the radius of the convective boundary $R\cb$ from the \Eq{eq:structa} as 
+\begin{equation}\label{eq:RCBint}
+{R_B \over R\cb} = \int_{P\di}^{P\cb} {T \over T_{\rm d}} {dP \over P}\, .
+\end{equation}
+Evaluating the integral leads to 
+\begin{equation}\label{eq:Rcb}
+{R_B \over r\cb} = \ln \left(P\cb \over P\di \right) - \ln \theta \, ,
+\end{equation} 
+with an extra correction term $\theta < 1$, when compared to an isothermal atmosphere (see Equation \ref{eq:rhoiso}).
+From this we arrive at the relation between $P\cb$ and $P\di$ given by Equation (\ref{eq:PcbRcb}).
+As opposed from the temperature correction factor $\chi$, an analytic expression for  $\theta$ cannot be obtained.
+Estimates for $\chi$ and $\theta$ for different values of the exponent $\beta$ in the opacity law (\ref{eq:opacitylaw}) are presented in Table 1.
+
+%As shown in section \S\ref{iso, an isothermal atmosphere gives a simple logarithmic dependence on $P\cb$.  However, using \Eq{eq:TP} in the integral gives
+
+% The form we chose for the correction term allows us to relate the disk and radiative-convective boundary pressures as :
+% \begin{equation}\label{eq:PcbRcb}
+%P\cb = \theta P_{\rm d} e^{R_B/R\cb} \, .
+%\end{equation}   
+%In the $P\cb \gg P_{\rm d}$ limit, the correction term is an order unity constant that depends on $\alpha$, $\beta$ and $\delad$.
+Similarly to the temperature correction factor $\chi$,  $\theta$ accounts for the fact that the radiative region is not perfectly isothermal.
+For $\delad=2/7$, $\alpha=0$ and $\beta=2$, we find $\theta \approx 0.556$.
+Values of $\theta$ for other choices of $\beta$ are summarized in \App{sec:analytic}.
+A simple analytic expression for $\theta$ is not possible.
+
+\subsection{The Opacity Effect}
+\label{opacityan}
+A  lower opacity  decreases the critical core mass.
+Reducing the opacity by a factor of one hundred results in a critical core mass one order of magnitude lower than in the standard ISM case, specifically  $M_{\rm{crit}} \approx9 M_\oplus$ for the parameters in \Eq{eq:Mcrit}.
+The reduction is not  as strong as the nominal scaling would imply, $0.01^{3/5} \approx 0.06$, because $\xi$ increases.
+
+Even with significantly lower opacities, radiative diffusion remains a good approximation at the RCB. For $\beta = 2$, we estimate the optical depth as
+\begin{equation}
+\tau\cb \sim {\kappa\cb P\cb \over g} \sim 7 \times 10^4 {F_T^4 F_\kappa \over \left(\mc \over 10 \right) \left(\au \over 10\right)^{12 \over 7}}, 
+\end{equation} 
+where $P\cb \sim P_M$ for a self-gravitating atmosphere and $g \sim G M\co/R_B^2$, with both approximations good to within the order unity factor $\xi$.
+We see that $\tau\cb \gg 1$ even for $F_\kappa \lesssim 0.01$ out to very wide separations, hence the atmosphere remains optically thick at the RCB.
+
+%A hotter disk would increase core masses.
+Instead of our passive disk model, adopting the standard Hayashi temperature profile would increase core masses by $\sim 50\%$.
+A hotter accretion phase would further increase core masses, but such phases are presumably short lived.
+
+\subsection{Surface Terms}
+\label{surfterms}
+In this section we check the relevance of the neglected surface terms in \Eq{eq:coolingglobal}.
+We first show that accretion energy is only a small correction at the RCB, which is where we apply our cooling model.
+A rough comparison (ignoring terms of order $\xi$) of  accretion luminosity vs.
+$\dot{E}$ gives
+\begin{equation}
+{G M \dot{M} \over R \dot{E}} = {G M  \over R}{dM \over dE} \sim {G M\co^2 \over R_B E}{P\cb \over P_M} \sim \sqrt{R\co \over R_B} \ll 1,
+\end{equation} 
+where we assume $P\cb \sim P_M$  for a massive atmosphere.
+Accretion energy at the protoplanetary surface is thus very weak for marginally self-gravitating atmospheres, and even weaker for lower mass atmospheres.
+A similar scaling analysis shows that the work term $P\surf \p V\surf/\p t$ is similarly weak.
+Nevertheless, our numerical calculations include these surface terms in a more realistic and complete model of self-gravitating atmospheres.
+
+%\subsection{Hydrogen Dissociation}
+%The dissociation of molecular hydrogen deep in the atmospheres of accreting protoplanets plays a significant role in the energetics of core accretion.
+In the high density regions $r  \ll R\cb$ of a convective atmosphere, the thermal plus gravitational energy scales as
+%\begin{equation}
+%dE = -4 \pi \nabla_{\rm ad}^{1/\nabla_{\rm ad}} \rho\cb R_B'^{1/(\gamma-1)} r^{\frac{2\gamma - 3}{\gamma - 1}} {dr \over r}
+%\end{equation} 
+%If $\gamma < 3/2$ then the main contribution to the energy is at the bottom of the atmosphere, i.e.\ the core.
+This is the case for a diatomic ideal gas ($\gamma = 1.4$) or a solar mixture of hydrogen and helium ($\gamma \approx 1.43$).
+However a monatomic gas has $\gamma = 5/3 > 3/2$.
+In this case, the atmosphere's energy is no longer concentrated near the bottom, but will be concentrated near the top of the convective zone.
+%
+%A likely structure is an atmosphere that is dissociated near the base, but becomes molecular near the top of the convective zone.
+In this case the atmosphere's energy budget would be concentrated near the atomic to molecular transition.
+%
+%The energy required to dissociate a hydrogen molecule, $I = 4.467$ eV can be significant to the overall energy budget.
+%Since
+%\begin{equation}
+%{I \over \nabla_{\rm ad} G M\co (2 m_{\rm H})/r} \approx 3 \left(M\co \over 10 M_\oplus \right)^{-2/3} {r \over R_{\rm c}}
+%\end{equation} 
+%we see that this energy is always relevant.
+%
+%We can use the Saha equation to determine where dissociation is significant,
+%\begin{equation}
+%{n_{\rm H}^2 \over n_{\rm H_2}} = \left(\pi m_{\rm H} k T \over h \right)^{3/2} e^{-I/(kT)}
+%\end{equation} 
+%We introduce a reaction coordinate $\delta$ so that $n_{\rm H} = 2 \delta n_o$ and $n_{\rm H_2} = (1-\delta) n_o$ with $n_o = \rho X/(2 m_{\rm H})$ the number density when all hydrogen is molecular.
+We express equilibrium as
+%\begin{equation}
+%{\delta^2 \over 1-\delta} = f_\mu {P_{\rm diss}(T) \over P}
+%\end{equation} 
+%with the characteristic pressure below which dissociation occurs is
+%\begin{equation}
+%P_{\rm diss} = {\left(kT\right)^{5/2} \over 4} \left( \pi m_{\rm H} \over h^2\right)^{3/2}  e^{-I/(kT)}
+%\end{equation} 
+%and the order unity factor
+%\begin{equation}
+%f_\mu = 2\delta + (1-\delta) + Y/2 + Z/\mu_Z
+%\end{equation} 
+%accounts for variations in the mean molecular weight with dissociation.  (Take $\mu_Z = 31/2$, but not too significant.)
+%
+%Thus dissociation occurs where $P \lesssim P_{\rm diss}(T)$.
+At disk temperatures (say 150 K) the dissociation pressure is negligibly small ($\sim 10^{-141}$ dyne cm$^{-2}$) and no dissociation occurs.
+However at core temperatures the dissociation pressure is quite large especially for massive cores.
+Dissociation is guaranteed.
+
+\begin{deluxetable}{cccccc}  % <--- column justification (center/left/right)
+\gdef \numcols {6}
+\tablecolumns{\numcols}
+\tablecaption{Parameters Describing Structure of Radiative Zone.}
+\tablehead{   \multicolumn{\numcols}{c} {$\gamma = 7/5$ ($\delad = 2/7$)} }  
+\startdata
+ $\beta$   		 &1/2  	& 3/4 &1   		& 3/2  		& 2   \\
+ $\nabla_\infty$ & 2/7 \tablenotemark{a}  	&  4/13	& 1/3 	& 2/5 	 	& 1/2 \\
+ $\chi$ 		 & \nodata &  2.25245 &1.91293 	& 1.65054 	& 1.52753 \\
+ $\theta $  		 &\nodata   & 0.145032	&0.285824   &0.456333   & 0.556069   \\
+\enddata
+\tablenotetext{a}{Since $\delad = \nabla_\infty$ there is no convective transition at depth for this case.}
+\end{deluxetable}
+
+%\subsection{Estimate of Atmosphere Mass Outside the Bondi Radius}
+%
+%Here we consider the mass exterior to the Bondi radius.
+For a meaningful evaluation we only include the mass coming from the overdensity relative to the background density.
+The resulting external mass for an isothermal atmosphere is
+%\begin{subeqnarray}
+%M_{\rm ext} &=& 4 \pi \int_{\RB}^{r_{\rm fit}} (\rho - \rho\di) r^2 dr \\
+%&=& M\co \theta\co \int_1^{n_{\rm fit}} 3 \left[ \exp \left({1 \over x} - {1 \over n_{\rm fit}}\right) - 1 \right] x^2 dx  \nonumber \\
+%&\equiv& M\co \theta\co I(n_{\rm fit})
+%\end{subeqnarray} 
+%where $\theta_c=R_B/R_{\rm c}$ and the dimensionless integral $I(n_{\rm fit})$ obeys the limits $I(1) = 0$ and $I \rightarrow n_{\rm fit}^2/2$ as $n_{\rm fit} \rightarrow \infty$.
+Since this external mass does not converge, the choice of an outer boundary does matter in principle.
+In practice, however, the assumption that   $r_{\rm fit} = R_{\rm H}$ limits $n_{\rm fit}$ to modest values
+%\begin{equation}
+%n_{\rm fit} = {R_{\rm H} \over \RB} \approx 1.3 {\aun{10}^{4/7} \over \mcn{10}^{2/3}}{F_T \over  m_\ast^{1/3}}.
+%\end{equation} 
+%Since for instance $I(2) = 1.1$, these modest $n_{\rm fit}$ values will only produce a small external mass.
 

### 2013-10-16 tex/draft/draft4.tex
section 3 minor edits

@@
 As shown in Section \ref{sec:KH}, self-gravity becomes quantitatively important at surprisingly low atmospheric masses, $M_{\rm atm} \gtrsim 0.2 M\co$.
-Thus the analytic model cannot accurately model the the later stages of evolution.
-Nevertheless, the analytic model does successfully explain general trends with opacity and disk temperature and pressure, for example.
+Thus the analytic model cannot accurately model the later stages of evolution.
+Nevertheless, this model does explain general trends, including the dependence on opacity and disk temperature and pressure.
 
@@
 This approximation limits the ability to study the initial cooling of very low mass atmospheres with thin outer radiative zones.
-However since early evolution is quite rapid, this approximation negligibly affects total cooling times and criticial core masses, our main foci.
+However since early evolution is quite rapid, this approximation negligibly affects total cooling times and critical core masses, our main foci.
 
@@
 
-If the radiative zone were replaced by an isothermal layer, then hydrostatic balance would give $\chi = \theta = 1$.  Standard radiative structure calculations (see \App{RCBcorr} for details) give 
+The leading constants would be unity, $\chi = \theta = 1$, if the radiative zone were replaced by an isothermal layer.
+In practice deviations from unity are modest.
+Standard radiative structure calculations (see \App{RCBcorr} for details) give
 \begin{equation}
@@
 While  a simple closed form expression is not possible, the factor $\theta \simeq 0.556$ for our parameters.
-According to \Eq{eq:PcbRcb} our pressure criterion also gives $R\cb \ll \RB$, but the inequality is logarithmically weaker.
-
-Given the conditions at the RCB, hydrostatic balance gives the density and temperature profiles in the inner region along the adiabat,
+%According to \Eq{eq:PcbRcb} our pressure criterion also gives $R\cb \ll \RB$, but the inequality is logarithmically weaker.
+
+Given the conditions at the RCB,  the density and temperature profiles along the interior adiabat,
 \begin{subeqnarray}
 \rho &=& \rho\cb \left[ 1 + {\RB' \over r} - {\RB' \over R\cb}  \right]^{1/(\gamma -1)} \slabel{eq:rhoconv}  \\
-T	&=& T\cb \left[ 1 + {\RB' \over r} - {\RB' \over R\cb}  \right] \slabel{eq:Tconv}\, .
+T	&=& T\cb \left[ 1 + {\RB' \over r} - {\RB' \over R\cb}  \right] \slabel{eq:Tconv}\, ,
 \end{subeqnarray} 
 %Note (to self) that the "1+" is neglibible in Rcb << RB limit taken elsewhere. 
-The effective Bondi radius
+follow from hydrostatic balance.  We introduce an effective Bondi radius
 \begin{equation}
@@
 \end{equation} 
-simplifies expressions, with $C_P = \Rg / \delad$ the specific heat at constant pressure.
-
-Even simpler expressions hold deep in the adiabatic interior, where $r \ll R\cb \ll \RB'$ gives
+to simplify expressions, with $C_P = \Rg / \delad$ the specific heat at constant pressure.
+
+Deep in the adiabatic interior, where $r \ll R\cb \ll \RB'$, the profiles follow simple power laws,
 \begin{subeqnarray}\label{eq:deep}
@@
 \end{subeqnarray} 
-While the radial density profile depends on the adiabatic index, the temperature scaling does not.  For reference, if self-gravity is included, the temperature profile satifies
+While the radial density profile depends on the adiabatic index, the $r^{-1}$ temperature scaling is universal.  For reference, if self-gravity is included, the temperature profile satisfies
 \begin{equation} \label{eq:TPsg}
@@
 \end{equation} 
-Thus the enclosed mass in self-gravitating models gives a temperature profile flatter than $T \propto r^{-1}$.
-
-The total specific energy at depth
+Thus in self-gravitating models, the enclosed mass gives a temperature profile flatter than $T \propto r^{-1}$.
+
+Returning to our non-self-gravitating model, the total specific energy at depth
 \begin{equation}\label{eq:ean}
@@
 \end{equation} 
-is thus generically proportional to the gravitational potential, $e_g = -GM\co/r$. 
-%AY: note $u$ introduced already in section 2.  Furthermore meaning obvious from context here.  e_G is good to restate in the non-self-grav limit which changed.
+is simply proportional to the gravitational potential, $e_g = -GM\co/r$.
 
@@
 \label{MELan}
-The most relevant quantities for the global cooling model (equation \ref{eq:coolingglobal}) are the integrated energy, luminosity, and atmospheric mass.
+The most relevant quantities for global cooling are the integrated energy, luminosity, and atmospheric mass.
 In our non-self-gravitating limit, the mass of our nearly isothermal radiative zones is less than the convective interior, as shown in \App{iso}.
 
-\Eq{eq:rhoconv} thus integrates to give the atmospheric mass
+The atmospheric mass is thus given by the integration of \Eq{eq:rhoconv}, 
 \begin{eqnarray} 
@@
 \end{eqnarray}
-in the relevant limit $R\co \ll R\cb \ll \RB'$ and for $\gamma =  7/5$.  Mass is concentrated near the outer regions of the convective zone, a result that holds more generally for $\gamma > 4/3$.
+in the relevant limit $R\co \ll R\cb \ll \RB'$ and for $\gamma =  7/5$.  Mass is concentrated near the outer regions of the convective zone, a result that holds for $\gamma > 4/3$.
 
@@
 
-The total energy is concentrated towards the core if $u \rho  r^3 \propto \rho r^2$ drops with increasing $r$.
-This condition requires $\gamma < 3/2$, which our choice of $\gamma = 7/5$ satisfies, but $\gamma = 5/3$ does not.
+The total energy is concentrated towards the core if $|u| \rho  r^3 \propto \rho r^2$ drops with increasing $r$.
+This condition requires $\gamma < 3/2$, which our choice of $\gamma = 7/5$ satisfies, but $\gamma = 5/3$ would not.
 Integration of \Eq{eq:ean} over the mass of the atmosphere thus gives
@@
 
-Applying marginal convective stability, $\delrad = \delad$, to \Eq{eq:delrad} gives the emergent luminosity from the radiative-convective boundary, 
+The emergent luminosity from the radiative-convective boundary, 
 \begin{equation} \label{eq:Lcb}
@@
 \end{equation} 
-where we define 
+follows from \Eq{eq:delrad} and marginal convective stability, $\delrad = \delad$, where we define 
 \begin{equation} 
@@
 This result relies on the pressure independence of  dust opacities.
-For fully non-self-gravitating results, we replace $M\cb$, the mass up to the RCB, with the core mass, but the mass of the convective atmosphere could be included for a slightly higher order estimate.
-
-\subsection{Simplified Cooling Model}
+For fully non-self-gravitating results, we replace $M\cb$, the mass up to the RCB, with the core mass, but the mass of the convective atmosphere can be included for a slightly higher order estimate.
+
+\subsection{Cooling Times \& Core Masses}
 \label{coolingan}
@@
 The magnitude of the discrepancy may be surprising as one might expect self-gravity to be a modest factor for $M_{\rm atm} \leq M\co$.
-Clearly this seemingly reasonable expectation must be applied with caution, an interesting result in itself.
-
-The analytic model is also useful because, despite the amplitude error, it explains the basic scaling of our (and other) numerical models.
+Clearly this seemingly reasonable expectation can be misleading, an interesting result in itself.
+
+Aside from this insight, the analytic model is useful because, despite the amplitude error, it explains the basic scaling of numerical cooling models.
 As is well known, a lower opacity, here scaling with $F_\kappa$, accelerates growth and gives a lower critical core mass.
 Lower disk temperatures also decrease both $t_{\rm co}$ and $\MC$.
-Furthermore the decline in both quantities with semimajor axis is completely generated by the disk temperature law (ignoring $\xi$).
+Furthermore the decline in both quantities with semimajor axis is completely generated by the disk temperature law (ignoring the logarithmic $\xi$).
 The temperature dependence are a competition between two main effects.
@@
 \begin{equation}\label{eq:radTP}
-\left(T \over T_{\rm d}\right)^{4-\beta} - 1 = {\nabla\di \over \nabla_\infty} \left(\left{P \over P_{\rm d}}\right - 1 \right) \, ,
+\left(T \over T_{\rm d}\right)^{4-\beta} - 1 = {\nabla\di \over \nabla_\infty} \left( {P \over P_{\rm d}} - 1 \right) \, ,
 \end{equation} 
@@
 Accretion energy at the protoplanetary surface is thus very weak for marginally self-gravitating atmospheres, and even weaker for lower mass atmospheres.
-A similar scaling analysis shows that the work term $P\surf \p V\surf/\p t$ is similarly weak.
+A similar scaling analysis shows that the work term $P_M \p V_M/\p t$ is similarly weak.
 Nevertheless, our numerical calculations include these surface terms in a more realistic and complete model of self-gravitating atmospheres.

### 2013-10-17 tex/draft/draft4.tex
section 4 edits      Big problem was that luminosity explanation was still wrong/incomplete. Also claims that ignoring L_negl is conservative don't seem rigorous and were removed as unneeded anyway. Fig. 5 caption was both long and incomplete (no mention of RCB/negative values).

@@
 
-Deep in a non-self-gravitating atmosphere, \Eq{eq:deep} shows that $T \propto r^{-1}$ and $P\propto r^{-1/\delad}$.
+The lowest mass atmosphere -- which we take as our initial state -- is fully convective with the same entropy as the disk.
+In \Fig{fig:profiles} this state is the 5.01 $M_{\oplus}$ solution with no radiative zone.
+Atmospheric cooling and contraction allows the accretion of more gas.
+In the convective zones, higher mass solutions contain lower entropy and higher pressures.
+A radiative zone emerges to connect the lower entropy interior to the higher entropy disk.
+\Fig{fig:profiles} shows that this radiative zone is already fairly deep in the 5.10 $M_\oplus$ solution.
+
+The detailed structure of these profiles is well approximated by our non-self-gravitating, analytic solutions.
+Deep in convective zones, \Eq{eq:deep} gives that $T \propto r^{-1}$ and $P\propto r^{-1/\delad}$.
 This behavior is seen in the low mass solutions in \Fig{fig:profiles}.
-As explained by \Eq{eq:TPsg}, the higher mass solutions show a flatter profile in $T$ and also in $P \propto T^{1/\delad}$.
-In agreement with \Eq{eq:cb2}, the radiative zone remains nearly isothermal -- even in the more massive solutions -- and thus shows a nearly exponential increase in pressure with inverse depth.
-
-The atmosphere starts out as purely convective and isentropic, as seen in \Fig{fig:profiles} for the 5.01 $M_{\oplus}$ solution which has no radiative zone.
-Further gas accumulation brings in energy that has to be radiated away, forming a radiative region that quickly deepens.
-The fast pressure deepening of the radiative region is evident in \Fig{fig:profiles} --- even a low mass atmosphere (i.e.
-the 5.10 $M_{\oplus}$ solution) has an extended radiative zone.
-
-The higher mass atmosphere solutions have lower entropy, i.e.\ cooling allows the atmosphere to accrete more gas.
-This is evident in \Fig{fig:profiles} from the higher internal pressures of the higher mass solutions.
-The somewhat higher temperatures in the higher mass solutions do not cause the entropy to increase.
-
-%AY: to say something about starting at disk entropy (not really needed) we should show it better in the figure.
-Also unclear if you want to say much more about evolution before figure 2 shows the evolution.
-%In the initial stages of its evolution, the atmosphere is purely convective and isentropic, with an entropy equal to that of the disk.
-In the context of our numerical model, this is the minimum atmosphere mass for which a hydrostatic solution exists.
-Further gas accumulation brings in energy that has to be radiated away, and an initially thin outer radiative region forms, which becomes thicker as the atmosphere becomes more massive.
+As explained by \Eq{eq:TPsg}, the high mass solutions show a slightly flatter profile in $T$ and also in $P \propto T^{1/\delad}$.
+In agreement with \Eq{eq:cb2}, the radiative zones remain nearly isothermal, even for the higher masses.
+Consequently the pressure increases nearly exponentially with inverse depth.
 
@@
 This behavior is seen in both the full numerical solutions (with solid lines) and the analytic model (with dashed lines).
-As seen in \Eq{eq:Lcb}, the increased pressure depth of the RCB increases the optical depth ($\propto \kappa P)$ and decreases the radiative luminosity.
-
-%AY: I don't want to say 5.12 is the lowest mass b/c it's not the adiabat, and I don't see what's gained by saying it's very near the lowest mass.
-
-At later stages of evolution, the numerical model gives a flat luminosity with increasing mass and also time (not shown).
-By contrast the analytic model gives a luminosity that continues to drop with increasing mass.
-The difference is due to the neglect of self-gravity in the analytic model.
-From \Eq{eq:Lcb}, $L\cb \propto M\cb T\cb^4/ (\kappa\cb  P\cb)$.
-Thus accounting for the higher enclosed mass gives a somewhat higher luminosity in the self-gravitating model.
-However, the main effect is the higher temperature in the self-gravitating models.
-The flatter temperature profiles with radius, as described above, explain why the self-gravitating models are hotter.
-
-%The flattening of the luminosity curve is a generic feature of numerical core accretion models
-
-%We find only a brief agreement between the analytic and numerical results.
-In the early stages, the radiative zone is too shallow and the assumption that $P_{\rm RCB} \gg P_{\rm d}$ used in the analytic model breaks down.
-
-Before discussing time evolution (both the elapsed time from the bottom panel of \Fig{fig:Ltplot} and the characteristic growth time of the atmosphere), we briefly discuss the effect of decreasing the atmospheric dust abundance on the atmospheric evolution.
+The increasing pressure depth of the RCB with mass causes increases in the optical depth ($\propto \kappa P)$ and decreases in the radiative luminosity.
+This behavior is described in \Eqs{eq:crit}{eq:Lcb}.
+
+At later stages of evolution, the numerical model in \Fig{fig:Ltplot} shows a flat luminosity with increasing mass and also time (not shown).
+By contrast the non-self-gravitating analytic model gives a luminosity that continues to drop with increasing mass.
+To understand this difference note that the scaling of \Eq{eq:Lcb}, $L\cb \propto M\cb T\cb^4/ (\kappa\cb  P\cb)$, holds in both cases.
+Thus accounting for the higher enclosed mass in the self-gravitating model gives a somewhat higher luminosity, as desired.
+However, the main effect is that \Eq{eq:crit} -- which describes a nearly linear relation between atmospheric mass and RCB pressure --  breaks down for self-gravitating solutions.
+This behavior can be seen in the top panel of \Fig{fig:profiles} where the $P\cb$ increases significantly from 5.10 to 6.0 $M_\oplus$, but only increases relatively modestly with further growth to 9.24 $M_\oplus$.
+To allow these relatively lower $P\cb$ values (and thus higher luminosities) in the higher mass solutions, the radius $R\cb$ must shift outwards, as is shown.
+
+The accelerated growth in the numerical model, as shown in the bottom panel of \Fig{fig:Ltplot} is also a direct result of the enhanced cooling (higher luminsoties) in the self-gravitating models.
+Clearly self-gravity plays a very important role in atmospheric evolution, even well before the crossover mass.
+
+%Before discussing time evolution (both the elapsed time from the bottom panel of \Fig{fig:Ltplot} and the characteristic growth time of the atmosphere), we briefly discuss the effect of decreasing the atmospheric dust abundance on the atmospheric evolution.
+%AY abo e sentence is bad, we want to discuss science and not "direct traffic"
+
 In \Fig{fig:LtvsMopacity}, the opacity normalization in \Eq{eq:opacitylaw} is reduced by factors of 10 and 100.
-Lower opacities result in higher luminosities and faster evolution, a well-established result \citep{HubBod05} that we confirm for our model.
-Atmospheric dust opacities are difficult to robustly predict.
-Ablation of infalling solids is a dust source, while settling through the radiative zone is a sink.
+Lower opacities result in higher luminosities and faster evolution.
+Our model thus confirms this well-established result \citep{HubBod05}.
+While clearly an important effect, atmospheric dust opacities are difficult to robustly predict.
+Ablation of infalling solids is a dust source.
+Sinks include the sequestration of solids in the core and dust settling through the radiative zone.
 Grain growth both reduces dust opacities per unit mass and favors settling.
 Our scenario of negligible ongoing particle accretion tends to favor low dust opacities.
-To be conservative, however, our reference case considers full Solar abundances, but faster evolution is clearly possible.
-The effect of opacity reduction on the critical core mass is deferred to \S\ref{sec:critical}.
+To be conservative, however, our reference case considers full Solar abundances.
+The effect of opacity reduction on the critical core mass is described in \S\ref{sec:critical}.
 
@@
 
-\Fig{fig:growthtime} plots the evolution of the atmospheric growth timescale, $M_{\rm atm}/\dot{M}$,  around a $5 M_{\oplus}$ core at several locations in the disk.
+\Fig{fig:growthtime} plots the evolution of the atmospheric growth timescale, $M_{\rm atm}/\dot{M}$,  around a $5 M_{\oplus}$ core at several locations in our reference disk model.
 This instantaneous growth time shows clearly that the atmosphere spends the bulk of its time growing though intermediate atmospheric masses, $\sim 1 -3$ $M_\oplus$ in this case.
@@
 
-The fact that growth times peak before the crossover mass, when $M_{\rm{atm}} < M_{\rm c}$, is very helpful for understanding atmospheric growth.
+The fact that growth times peak before the crossover mass, when $M_{\rm{atm}} < M_{\rm c}$, is helpful for understanding atmospheric growth.
 The estimated timescale to runaway growth can be well approximated by only considering the time it takes to grow to the crossover mass, or even to a somewhat lower mass.
-This insensitivity to the upper mass threshold is characteristic of accelerating growth, and is fortuitious.
-We next show how our model assumptions can break down during the later stages of atmospheric growth.
+This insensitivity to the upper mass threshold is characteristic of accelerating growth.
+We next show how this behavior is fortuitous, as our model assumptions can break down during the later stages of atmospheric growth.
 
@@
 To compute $L_{\rm negl}$ we compute the entropy difference between successive radiative zone solutions.
-We then integrate the energy equation, $\p L / \p m = - T \p S/ \p t$, over the average depth of the radiative zone.\footnote{While useful as a diagnostic, the neglected luminosity cannot reliably correct the global cooling model because where $L_{\rm negl}$ is significant, the structure of the radiative zone would be inaccurate.}
+We then integrate the energy equation, $\p L / \p m = - T \p S/ \p t$, over the average depth of the radiative zone.\footnote{While useful as a diagnostic, the neglected luminosity cannot reliably correct the global cooling model because the effects of $L_{\rm negl}$ on the structure of the radiative zone are still ignored.}
 
  \Fig{fig:coolingterms} shows that the neglected luminosity is indeed negligible during the early stages of evolution.
-However, $L_{\rm negl}$ exceeds the model luminosity, $L$, at high masses, $M_{\rm atm} > 3 M_\oplus$ in this case, meaning that our model is becoming inaccurate.
-Nevertheless, our cooling model gives both conservative and reasonably reliable estimates of core accretion timescales.
-Our results are conservative -- in the sense of overestimating growth times -- because the neglected luminosity would give more rapid evolution.
-Moreover, our results are reasonably accurate because the bulk of growth time is spent at low masses, when $L_{\rm negl}$ is small and safely neglected.
+However, $L_{\rm negl}$ exceeds the model luminosity, $L$, at high masses, $M_{\rm atm} > 3 M_\oplus$ in this case, meaning that our model becomes inaccurate at late states.
+Nevertheless, our cooling model gives  reasonably reliable estimates of core accretion timescales.
+As shown above, the growth time is mostly spent at low masses, when $L_{\rm negl}$ is small and safely neglected.
+ %Our results are conservative -- in the sense of overestimating growth times -- because the neglected luminosity would give more rapid evolution.  
+ %AY: not sure we can back up the conservative claim because of the detailed effects on structure that haven't been calculated.
 
@@
 At low masses, the change in energy, $- \dot{E}$, makes the dominant contribution to luminosity.
-As the mass increases the surface terms become more significant, led by the accretion energy.
+As the mass increases, the surface terms become more significant, led by the accretion energy.
 However the surface terms are everywhere smaller than $L_{\rm negl}$.
-Thus while the surface terms appear to reduce the total luminosity (compared to $-\dot{E}$), $L_{\rm negl}$ can more than compensate, showing again that our model conservatively underestimates cooling.
+Thus wherever our model is accurate -- including the crucial early phases of growth -- surface terms are a minor correction.
+%while the surface terms appear to reduce the total luminosity (compared to $-\dot{E}$), $L_{\rm negl}$ can more than compensate, showing again that our model conservatively underestimates cooling.
+%AY:again don't think conservative claim can be so easily justified.
 
 At yet higher masses than shown in \Fig{fig:coolingterms}, our cooling model breaks down.
-Mathematically, the breakdown manifests as negative timesteps when advancing to a higher atmospheric mass.
-This unphysical result has no deep significance, as we already know that the neglected luminosity is large enough to give incorrect results.
+Mathematically, the breakdown manifests as negative time steps when advancing to a higher atmospheric mass.
+This unphysical result has no deep significance, as the neglected luminosity is already large enough to give inaccurate results.
 Sometimes this breakdown occurs even before the crossover mass is reached.
@@
 %\vspace{-0.5in}
-\caption{The relative contribution of the various terms in the cooling \Eq{eq:coolingglobal} to the atmosphere luminosity, for a planet forming around a $5 M_\oplus$ core at 10 AU.
-The $L_{\rm negl}$ curve gives the extra luminosity that would have been generated in the radiative zone, but is not included in our two-layer cooling model.
-The neglected luminosity is lower than the model luminosity $L$ for low atmosphere mass, and becomes comparable at higher masses ($>3 M_{\oplus}$ in this particular case).
-As atmosphere growth is slowest at lower masses, neglecting the additional luminosity generated in the radiative layer yields accurate results, to order of magnitude.}
+\caption{Individual terms in the atmospheric cooling model of \Eq{eq:coolingglobal}, for a $5 M_\oplus$ core at 10 AU.
+The dashed curve for accretion energy indicates a negative contribution.
+All quantities are evaluated at the RCB, except for $L_{\rm negl}$, the extra luminosity that would have been generated in the radiative zone, but is neglected in our model.
+The neglected luminosity is a small correction to the model luminosity $L$ for $M_{\rm atm} \lesssim 3 M_{\oplus}$.
+Since these low masses dominate growth times, our model is roughly accurate.}
 \label{fig:coolingterms}

### 2013-10-18 tex/draft/draft4.tex
minor corrections secs. 2-4, mostly typos

@@
 
-Full time dependent simulations of giant planet evolution take into account the Kelvin-Helmholtz (KH) contraction of the gaseous envelope.
+Full time-dependent simulations of giant planet evolution take into account the Kelvin-Helmholtz (KH) contraction of the gaseous envelope.
 As a result, the atmosphere is no longer in a steady state, but rather it accretes gas as it cools radiatively.
 In this scenario there are two main sources that contribute to the luminosity evolution of the envelope --- planetesimal accretion and gas contraction.
-These evolutionary models for core and atmosphere growth include time dependent planetesimal accretion rates that are determined by the distribution of solids in the disk, as well as the planet mass and capture cross-section.
+These evolutionary models for core and atmosphere growth include time-dependent planetesimal accretion rates that are determined by the distribution of solids in the disk, as well as the planet mass and capture cross-section.
 This is is contrast with the static models in which the planetesimal accretion rate is usually constant with time.
@@
 We explore the atmosphere evolution around fully grown cores of different masses, and determine the minimum core mass to form a giant planet before disk dissipation for a variety of disk conditions and properties of the nebular gas.
-We note that our models are a special case of the time dependent studies, with the planetesimal accretion rate set to zero.
+We note that our models are a special case of the time-dependent studies, with the planetesimal accretion rate set to zero.
 Our neglect of planetesimal accretion allows us to obtain absolute lower limits on the core mass needed to form a gas giant, as additional heat sources limit the ability of the atmosphere to cool and undergo KH contraction.
@@
 To model the growth of planetary atmospheres around a solid core, we develop a simplified two layer model for time-dependent atmospheric cooling, i.e.
-Kelvin-Helmholtz (KH) contraction.
+KH contraction.
 With a convective interior and radiative exterior, this model is motivated by similar models of hot Jupiters \citep{ab06, ym10}.
@@
 \item A global cooling model connects independent static solutions into a time-dependent sequence.
-\item A polytropic EOS is assumed for simplicity. %AP: define EOS here if not already
+\item A polytropic equation of state (EOS) is assumed for simplicity. %AP: define EOS here if not already
 \item Dust grains provide the opacity in the radiative zone, which remains cool enough to avoid dust sublimation.
@@
 \label{eq:Pd}
-P\di = 6.9 \times 10^{-3} F_\varSigma \sqrt{F_T} \, \aun{10}^{-45/14}~{\rm dyne~cm^2}
+P\di = 6.9 \times 10^{-3} F_\varSigma \sqrt{F_T} \, \aun{10}^{-45/14}~{\rm dyn~cm^{-2}}
 %P\di=1.1 \times 10^{-4} F_{\Sigma} \sqrt{F_T} a^{-45/14} \,\, \text{dyne cm}^{-2} %\times \sqrt{m_\ast}
@@
 \end{equation}  
-in terms of the specific gas constant $\Rg$.  The disk scale-height is 
+in terms of the specific gas constant $\Rg$.  The disk scale height is 
 \begin{equation}
@@
 \end{equation} 
-in terms of the Keplerian frequency $\varOmega = \sqrt{G M_\ast/a^3}$ with $G$ the gravitational constant and $M_\ast$ the stellar (in this work Solar) mass.
+in terms of the Keplerian frequency $\varOmega = \sqrt{G M_\ast/a^3}$, with $G$ the gravitational constant and $M_\ast$ the stellar (in this work Solar) mass.
 
@@
 
-The characteristic length scales for protoplanetary atmospheres are crucial for choosing boundary conditions and for understanding the validity of  spherical symmetry in a disk of scaleheight $H\di$.
-The radius of the solid core:
+The characteristic length scales for protoplanetary atmospheres are crucial for choosing boundary conditions and for understanding the validity of  spherical symmetry in a disk of scale height $H\di$.
+The radius of the solid core is
 
@@
 
-A planet can bind a dense atmosphere if its escape speed exceeds the sound speed.  This criterion is satisfied inside the Bondi radius
+A planet can bind a dense atmosphere if its escape velocity exceeds the sound speed.  This criterion is satisfied inside the Bondi radius
 \begin{equation}
@@
 \label{eq:RHill}
-R_{\rm H} = \left(M\pla \over 3 M_\ast \right)^{1/3}a \approx 0.22 \, {\mpn{10}^{1/3} \, \aun{10} }~\AU
+R_{\rm H} = \left(M\pla \over 3 M_\ast \right)^{1/3}a \approx 0.22 \, {\mpn{10}^{1/3} \, \aun{10} }~\AU ,
 \end{equation}
@@
 
-The relevant length scales of the atmosphere and disk satisfy the relation $\RB H\di^2 = 3 R_{\rm H}^3$.  The lengthscales are roughly equal at the ``thermal mass'' (e.g., \citealt{menou04})
+The relevant length scales of the atmosphere and disk satisfy the relation $\RB H\di^2 = 3 R_{\rm H}^3$.  The length scales are roughly equal at the ``thermal mass'' (e.g., \citealt{menou04})
 \begin{equation}
@@
 
-In the low mass regime, $M\pla < M_{\rm th}/\sqrt{3}$, the lengthscales order as $\RB< \RH<H\di$ (R06).
+In the low mass regime, $M\pla < M_{\rm th}/\sqrt{3}$, the length scales order as $\RB< \RH<H\di$ (\citealt{rafikov06}, hereafter R06).
 In this regime, many studies assume the atmosphere matches the disk conditions at $\RB$.
-We however use $\RH$ as the matching radius in both this low mass and other higher mass regimes.
+We, however, use $\RH$ as the matching radius in both this low mass and other higher mass regimes.
 This choice is justified by the fact that, for hydrostatic solutions, the  density at $\RB$ exceeds the disk's background density by an order unity factor (R06).
@@
 \end{subeqnarray}
-\noindent where $r$ is the radial coordinate, $P$, $T$, $\rho$  and $L$ are the gas pressure, temperature, density and luminosity, respectively.
+\noindent where $r$ is the radial coordinate, and $P$, $T$, $\rho$  and $L$ are the gas pressure, temperature, density and luminosity, respectively.
 The enclosed mass  at radius $r$ is $m$.
@@
 \label{eq:delad}
-\delad \equiv \Big(\frac{d \ln T}{d \ln P}\Big)_{\rm{ad}}
+\delad \equiv \Big(\frac{d \ln T}{d \ln P}\Big)_{\rm{ad}}.
 \end{equation}
@@
 
-To solve the equation set (\ref{eq:struct}) an equation of state (EOS) is required for closure. In our study, we adopt an ideal gas law with a polytropic EOS 
+To solve the equation set (\ref{eq:struct}), an EOS is required for closure. In our study, we adopt an ideal gas law with a polytropic EOS 
 \begin{subeqnarray}
@@
 where $K$ is the adiabatic constant.
-The adiabatic index  $\gamma = 1/(1 - \delad)$.
-An ideal monatomic gas has $\delad = 2/5$.
-This work uses $\delad=2/7$ for an ideal diatomic gas.
+The adiabatic index is  $\gamma = 1/(1 - \delad)$.
+This work uses $\delad=2/7$ for an ideal diatomic gas (by comparison, an ideal monatomic gas has $\delad = 2/5$).
 While our reference mean molecular weight ($\mu = 2.35$ proton masses) includes Helium, we ignore Helium's effect on the EOS, which is already greatly simplified.
@@
 \begin{equation}
-S = \Rg \ln \left(T^{1/\delad} \over P \right) 
+S = \Rg \ln \left(T^{1/\delad} \over P \right) ,
 \end{equation} 
@@
 Boundary conditions must be satisfied at both the base and the top of the atmosphere with $m(R\co) = M\co$, $T(\RH) = T\di$ and $P(\RH) = P\di$.
-In principle our solutions describe atmospheres with $L(R\co) = 0$.
+In principle, our solutions describe atmospheres with $L(R\co) = 0$.
 In practice, since we do not directly integrate \Eq{eq:structd} we need not directly impose this boundary condition, as described in \S\ref{sec:twolayer}.
@@
 This section describes the global energy balance of a planet embedded in a gas disk, or more generally any spherical, hydrostatic object in pressure equilibrium with a background medium.
-The total atmospheric energy includes gravitational and internal energies, $E = E_G + U$:
+The total atmospheric energy includes gravitational and internal energies, $E = E_G + U$, with
 \begin{subeqnarray}
@@
 \end{subeqnarray}
-The specific internal energy $u = C_V T = \Rg (\delad^{-1} -1) T$ for a polytropic EOS.  For a star or coreless planet, $M\co = 0$.
+The specific internal energy is $u = C_V T = \Rg (\delad^{-1} -1) T$ for a polytropic EOS.  For a star or coreless planet, $M\co = 0$.
 
@@
 The surface can be at any mass level $M$, where the instantaneous radius is $R$.
-Other surface quantities are labelled by $M$ subscripts.
+Other surface quantities are labeled by $M$ subscripts.
 The energy accreted across the surface is given by the specific energy, $e_M = u_M-G M/R$, and the mass accretion rate of gas, $\dot{M}$.
@@
 
-For static solutions, which are not the focus of this Paper, the surface terms (and also $\dot{E})$ vanish.
+For static solutions, which are not the focus of this paper, the surface terms (and also $\dot{E})$ vanish.
 Static solutions are valid when imposed heat sources, i.e.\ $L\co$ and $\Gamma$, exceed the atmospheric losses.
@@
 Before the protoplanetary atmosphere can cool, it has the entropy of the disk.
-As the atmosphere cools the deep interior remains convective.
+As the atmosphere cools, the deep interior remains convective.
 Convective interiors are a common feature of low mass cool objects (brown dwarfs and planets) that results from the behavior of $\delrad$ for realistic opacity laws.
-However the entropy of the deep interior decreases as the atmosphere cools.
+However, the entropy of the deep interior decreases as the atmosphere cools.
 A region of outwardly increasing entropy, i.e.\ a radiative layer, is required to connect the convective interior to the disk.
-A more complicated structure, with radiative windows in the convection zone, is possible as discussed in \S\ref{sec:opEOS}.
+A more complicated structure, with radiative windows in the convection zone, is possible as discussed in \S\ref{sec:op}.
 
 In convective regions, the adiabatic structure is independent of luminosity and can be calculated without local energy balance, \Eq{eq:structd}.
-Thus for fully convective objects, a cooling sequence can be established by connecting a series of adiabatic solutions using a global energy equation, $L_M = -\dot{E}$ or \Eq{eq:coolingstar}.
+Thus, for fully convective objects, a cooling sequence can be established by connecting a series of adiabatic solutions using a global energy equation, $L_M = -\dot{E}$ or \Eq{eq:coolingstar}.
 Such methods are commonly used for their computational efficiency and are sometime referred to as ``following the adiabats," since the steady state solutions evolve in order of decreasing entropy \citep{marleau13}.
@@
 We proceed by assuming that the majority of energy is lost from the convective interior, and thus the luminosity can be treated as constant in the outer radiative zone.
-With this approximation we can construct solutions from equations (\ref{eq:struct}a -- c) that ``follow the mass," i.e.\ gradually increase the atmospheric mass.
-We then use the global energy balance, \Eq{eq:coolingglobal} to place these solutions in a cooling sequence.
+With this approximation we can construct solutions from Equations (\ref{eq:struct}a -- c) that ``follow the mass," i.e.\ gradually increase the atmospheric mass.
+We then use global energy balance, \Eq{eq:coolingglobal}, to place these solutions in a cooling sequence.
 The validity of neglecting luminosity generation in the radiative zone can be checked \emph{a posteriori}.
 
-To make a single atmosphere model (indexed by $i$) we choose a planet mass $M_i$.
+To make a single static atmosphere model (indexed by $i$), we choose a planet mass $M_i$.
 At the outer boundary, at $\RH(M_i)$, the temperature and pressure are set to the disk values.
-To integrate equations (\ref{eq:struct}a--c), the luminosity is required to compute $\delrad$.
+To integrate Equations (\ref{eq:struct}a--c), the luminosity is required to compute $\delrad$.
 The correct value of the luminosity is not known in advance, and is the eigenvalue of the problem.
@@
 To obtain our analytic results, we neglect self-gravity.
-As shown in Section \ref{sec:KH}, self-gravity becomes quantitatively important at surprisingly low atmospheric masses, $M_{\rm atm} \gtrsim 0.2 M\co$.
+As shown in  \S\ref{sec:KH}, self-gravity becomes quantitatively important at surprisingly low atmospheric masses, $M_{\rm atm} \gtrsim 0.2 M\co$.
 Thus the analytic model cannot accurately model the later stages of evolution.
@@
 This approximation limits the ability to study the initial cooling of very low mass atmospheres with thin outer radiative zones.
-However since early evolution is quite rapid, this approximation negligibly affects total cooling times and critical core masses, our main foci.
+However, since early evolution is quite rapid, this approximation negligibly affects total cooling times and critical core masses, our main foci.
 
@@
 \end{subeqnarray}
-
 The leading constants would be unity, $\chi = \theta = 1$, if the radiative zone were replaced by an isothermal layer.
-In practice deviations from unity are modest.
+In practice, deviations from unity are modest.
 Standard radiative structure calculations (see \App{RCBcorr} for details) give
@@
 \end{equation} 
-Thus in self-gravitating models, the enclosed mass gives a temperature profile flatter than $T \propto r^{-1}$.
-
-Returning to our non-self-gravitating model, the total specific energy at depth
+Thus in self-gravitating models the enclosed mass gives a temperature profile flatter than $T \propto r^{-1}$.
+
+Returning to our non-self-gravitating model, the total specific energy at depth,
 \begin{equation}\label{eq:ean}
-e = e_g + u = -\delad {GM\co \over r} \, .
+e = e_g + u = -\delad {GM\co \over r} \, ,
 \end{equation} 
@@
 The total energy is concentrated towards the core if $|u| \rho  r^3 \propto \rho r^2$ drops with increasing $r$.
-This condition requires $\gamma < 3/2$, which our choice of $\gamma = 7/5$ satisfies, but $\gamma = 5/3$ would not.
+This condition requires $\gamma < 3/2$, which our choice of $\gamma = 7/5$ satisfies, but $\gamma = 5/3$ (monatomic gas) would not.
 Integration of \Eq{eq:ean} over the mass of the atmosphere thus gives
@@
 
-The emergent luminosity from the radiative-convective boundary, 
+The emergent luminosity from the RCB, 
 \begin{equation} \label{eq:Lcb}
@@
 
-Both $t_{\rm co}$ and $\MC$ are too large to be interesting, or to be correct based on previous results and the numerical results in this Paper.
+Both $t_{\rm co}$ and $\MC$ are too large to be interesting, or to be correct based on previous results and the numerical results in this paper.
 The main reason for this discrepancy is the neglect of self-gravity, as shown in Section \ref{sec:KH}.
@@
 Lower disk temperatures also decrease both $t_{\rm co}$ and $\MC$.
-Furthermore the decline in both quantities with semimajor axis is completely generated by the disk temperature law (ignoring the logarithmic $\xi$).
-The temperature dependence are a competition between two main effects.
+Furthermore, the decline in both quantities with semi-major axis is completely generated by the disk temperature law (ignoring the logarithmic $\xi$).
+The temperature dependence is a competition between two main effects.
 The higher luminosity, $\propto T^{4-\beta}$, at higher temperatures gives faster cooling, which opposes the overall effect.
@@
 We now present some representative examples of the structure and evolution of our model atmospheres, calculated as described in \S\ref{sec:model}, with some comparisons made to the non-self-gravitating analytic model of \S\ref{sec:coolingan}.
-Radial structure is presented in \S\ref{sec:profiles}, time evolution is described in \S\ref{sec:timeev} and the regimes of validity of our two-layer cooling model are examined in \S\ref{sec:endoftime}.
+Radial structure is presented in \S\ref{sec:profiles}, time evolution is described in \S\ref{sec:timeev}, and the regimes of validity of our two-layer cooling model are examined in \S\ref{sec:endoftime}.
 
@@
 \caption{Radial profiles of atmospheric pressure, temperature and enclosed mass (including the core) for a $5 M_{\oplus}$ core at $a=60$ AU.
-Solid, dotted and dashed lines correspond to solutions with total mass (core and atmosphere) of $5.01 M_{\oplus}$, $5.10 M_{\oplus}$, $6.00 M_{\oplus}$ and $9.24 M_{\oplus}$, respectively.
+Solid, dotted, dashed and dot-dashed lines correspond to solutions with total mass (core and atmosphere) of $5.01 M_{\oplus}$, $5.10 M_{\oplus}$, $6.00 M_{\oplus}$ and $9.24 M_{\oplus}$, respectively.
 (See text for a description of evolution to yet higher masses.)   Circles and triangles mark the locations of the Bondi radii and of the radiative-convective boundaries, respectively.
@@
 In agreement with \Eq{eq:cb2}, the radiative zones remain nearly isothermal, even for the higher masses.
-Consequently the pressure increases nearly exponentially with inverse depth.
+Consequently, the pressure increases nearly exponentially with inverse depth.
 
@@
 At later stages of evolution, the numerical model in \Fig{fig:Ltplot} shows a flat luminosity with increasing mass and also time (not shown).
-By contrast the non-self-gravitating analytic model gives a luminosity that continues to drop with increasing mass.
-To understand this difference note that the scaling of \Eq{eq:Lcb}, $L\cb \propto M\cb T\cb^4/ (\kappa\cb  P\cb)$, holds in both cases.
-Thus accounting for the higher enclosed mass in the self-gravitating model gives a somewhat higher luminosity, as desired.
+By contrast, the non-self-gravitating analytic model gives a luminosity that continues to drop with increasing mass.
+To understand this difference, we note that the scaling of \Eq{eq:Lcb}, $L\cb \propto M\cb T\cb^4/ (\kappa\cb  P\cb)$, holds in both cases; thus accounting for the higher enclosed mass in the self-gravitating model gives a somewhat higher luminosity, as desired.
 However, the main effect is that \Eq{eq:crit} -- which describes a nearly linear relation between atmospheric mass and RCB pressure --  breaks down for self-gravitating solutions.
 This behavior can be seen in the top panel of \Fig{fig:profiles} where the $P\cb$ increases significantly from 5.10 to 6.0 $M_\oplus$, but only increases relatively modestly with further growth to 9.24 $M_\oplus$.
-To allow these relatively lower $P\cb$ values (and thus higher luminosities) in the higher mass solutions, the radius $R\cb$ must shift outwards, as is shown.
-
-The accelerated growth in the numerical model, as shown in the bottom panel of \Fig{fig:Ltplot} is also a direct result of the enhanced cooling (higher luminsoties) in the self-gravitating models.
-Clearly self-gravity plays a very important role in atmospheric evolution, even well before the crossover mass.
+To allow these relatively lower $P\cb$ values (and thus higher luminosities) in the higher mass solutions, the radius $R\cb$ must shift outwards, as shown in \Fig{fig:profiles}.
+
+The accelerated growth in the numerical model, as shown in the bottom panel of \Fig{fig:Ltplot}, is also a direct result of the enhanced cooling (higher luminosities) in the self-gravitating models.
+We see that self-gravity plays a very important role in atmospheric evolution, even well before the crossover mass.
 
@@
 This unphysical result has no deep significance, as the neglected luminosity is already large enough to give inaccurate results.
-Sometimes this breakdown occurs even before the crossover mass is reached.
+Sometimes this breakdown occurs even before the crossover mass is reached, although still significantly after the phase of slowest growth.
 The main practical effect of this behavior is that we halt the evolution where the model breaks down whenever we cannot reach the crossover mass.

### 2013-10-20 tex/draft/draft4.tex
updated abstract, summary, and minor edits throughout

@@
 The core accretion model proposes that giant planets form by the accretion of gas onto a solid protoplanetary core.
-Previous studies have found that there exists a ``critical core mass'' past which hydrostatic solutions no longer exist and unstable atmosphere collapse occurs.
-This core mass is typically quoted to be around $10 M_{\oplus}$.
-In static calculations of the critical core mass, planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse.
-In this study we consider the limiting case in which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the planet.
-We develop a two-layer atmosphere model with an inner convective region and an outer radiative zone that matches onto the protoplanetary disk, and we determine the minimum core mass for a giant planet to form within the typical disk lifetime for a variety of disk conditions.
-We denote this mass as  \textit{critical core mass}.
+The minimum (or critical) core mass to form a gas giant is typically quoted as $10 M_{\oplus}$.
+This value, however, depends on several factors, such as the location in the disk, nebular gas composition, or the accretion rate of solids.
+In this study we consider the limiting case of atmosphere growth around a core of fixed mass, in which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the envelope.
+We develop a two-layer atmosphere model with an inner convective region and an outer radiative zone that matches onto the protoplanetary disk, and we determine the minimum core mass for a giant planet to form within typical disk life timescales for a variety of disk conditions.
+We are primarily interested in the outer regions of the protoplanetary disk, where our assumptions of spherical symmetry and dust grains opacity are valid, and where the increasing number of directly imaged giant planets helps constrain formation models.
 The absolute minimum core mass required to nucleate atmosphere collapse is $\sim 8 M_{\oplus}$ at 5 AU and steadily decreases to $\sim 3.5 M_{\oplus}$ at 100 AU, for an ideal diatomic gas with a solar composition and a standard ISM dust opacity.
@@
 As additional heating due to planetesimal accretion leads to slower cooling, our results represent an absolute minimum for the core mass needed to form a giant planet before disk dissipation.
+
+%Previous studies have found that there exists a ``critical core mass'' past which hydrostatic solutions no longer exist and unstable atmosphere collapse occurs.
+This core mass is typically quoted to be around $10 M_{\oplus}$.
+In static calculations of the critical core mass, planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse.
+In this study we consider the limiting case in which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the planet.
 
@@
 
-To model the growth of planetary atmospheres around a solid core, we develop a simplified two layer model for time-dependent atmospheric cooling, i.e.
+To model the growth of planetary atmospheres around a solid core, we develop a simplified two-layer model for time-dependent atmospheric cooling, i.e.
 KH contraction.
@@
 We adopt a minimum mass solar nebula (MMSN) model for a passively irradiated disk \citep{chiang10}.
-With the semi-major axis $a$ normalized to the outer disk as $\aun{10} = a/(10 \text{ AU})$, the gas surface density and mid-plane temperature are
+With the semimajor axis $a$ normalized to the outer disk as $\aun{10} = a/(10 \text{ AU})$, the gas surface density and mid-plane temperature are
 \begin{subeqnarray} \label{eq:diskparam}
@@
 
-To simplify our calculations of atmospheric contraction we use a two layer model with a bottom convective region and an upper radiative layer.
+To simplify our calculations of atmospheric contraction we use a two-layer model with a bottom convective region and an upper radiative layer.
 The existence of such a structure is well known from previous studies (e.g., R06) and can be readily understood.
@@
 
-To make a single static atmosphere model (indexed by $i$), we choose a planet mass $M_i$.
+To obtain a single atmosphere profile (indexed by $i$), we choose a planet mass $M_i$.
 At the outer boundary, at $\RH(M_i)$, the temperature and pressure are set to the disk values.
@@
 \subsection{Two Layer Structure}
-To apply the two layer cooling model analytically, we require expressions for the atmospheric structure.
+To apply the two-layer cooling model analytically, we require expressions for the atmospheric structure.
 Conditions at the RCB are crucial as they set the interior adiabat and the radiative losses.
@@
 \end{equation} 
-Thus in self-gravitating models the enclosed mass gives a temperature profile flatter than $T \propto r^{-1}$.
+Thus, the enclosed mass gives a temperature profile flatter than $T \propto r^{-1}$ in self-gravitating models.
 
@@
 Lower disk temperatures also decrease both $t_{\rm co}$ and $\MC$.
-Furthermore, the decline in both quantities with semi-major axis is completely generated by the disk temperature law (ignoring the logarithmic $\xi$).
+Furthermore, the decline in both quantities with semimajor axis is completely generated by the disk temperature law (ignoring the logarithmic $\xi$).
 The temperature dependence is a competition between two main effects.
@@
 The cooling timescale and critical core mass depend only weakly on the disk pressure, via the  logarithmic factor $\xi$.
-Section \ref{sec:TPeffects} shows that these analytic predictions for disk temperature and pressure dependence compare favorably with numerical results.
+Section \S\ref{sec:TPeffects} shows that these analytic predictions for disk temperature and pressure dependence compare favorably with numerical results.
+Moreover, the analytic estimates quantitatively reproduce the numerical results to reasonable accuracy, if the effective crossover mass for unstable atmosphere collapse is a fraction of the core mass, i.e.
+$M_{\rm atm}=f M\co$.
+Under this assumption, Equations (\ref{eq:tcoolan}) and (\ref{eq:Mcrit}) become
+\begin{eqnarray} 
+\label{eq:tcoolanscaled}
+t_{\rm co} &\approx & 2 \times 10^8 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2 f^2  \over \mcn{10}^{5/3} \aun{10}^{15 / 14}} \yr \\ \slabel{eq:tcoolanf}
+ & \approx & 3 \times 10^6 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2  \over \mcn{10}^{5/3} \aun{10}^{15 / 14}} \yr
+\end{eqnarray} 
+and
+\begin{eqnarray}\label{eq:Mcritscaled}
+\MC &\approx & 100 {F_T^{3/2} F_\kappa^{3/5}   \left(\xi \over 2.6 \right)^{6/5} f^{6/5} \over \aun{10}^{9 / 14}} \; M_\oplus \\ \slabel{eq:Mcritf}
+& \approx & 8\, {F_T^{3/2} F_\kappa^{3/5}   \left(\xi \over 2.6 \right)^{6/5} \over \aun{10}^{9 / 14}} \; M_\oplus\ ,
+\end{eqnarray}
+with $f=0.13$ in Equations (\ref{eq:tcoolanf}) and (\ref{eq:Mcritf}).
+We choose this particular fractional scaling to match our numerical results at 10 AU (see section \S\ref{sec:critical} for more details).
+We note that the scaling above is valid if the logarithmic factor $\xi$ is fixed; however, the analytic estimates presented in section \S\ref{sec:critical} explicitly take into account the dependence of $\xi$ on the core mass (see Equations \ref{eq:xidef} and \ref{eq:Pcbc}).
+
  %This weak dependence is a consequence of cooling coming from the radiative-convective boundary.
@@
 
-\emph{TODO: consider adding "revised" analytic expressions for $t_{\rm co}$ and $\MC$ (e.g.
+%\emph{TODO: consider adding "revised" analytic expressions for $t_{\rm co}$ and $\MC$ (e.g.
 lower Matm criterion).
@@
 %\vspace{-0.5in}
-\caption{Similar to \Fig{fig:Ltplot}, but showing the effect of reducing the dust abundance by factors of 10 and 100 from standard Solar abundances.
+\caption{Similar to \Fig{fig:Ltplot}, but for a $5 M_{\oplus}$ core forming at 10 AU; we show the effect of reducing the dust abundance by factors of 10 and 100 from standard Solar abundances.
 The reduced dust opacities give higher luminosities and faster atmospheric growth.}  %The actual magnitude of dust opacities is difficult to robustly predict.}
@@
 This unphysical result has no deep significance, as the neglected luminosity is already large enough to give inaccurate results.
-Sometimes this breakdown occurs even before the crossover mass is reached, although still significantly after the phase of slowest growth.
+Sometimes this breakdown occurs even before the crossover mass is reached, although still significantly after growth has started to accelerate.
+In our models the breakdown occurs at lower atmosphere mass as the semimajor axis increases, since accelerated and eventually runaway growth occurs sooner at larger distances, as seen in \Fig{fig:growthtime}.
 The main practical effect of this behavior is that we halt the evolution where the model breaks down whenever we cannot reach the crossover mass.
@@
 A second radiative zone inside the convection zone, a ``radiative window," is possible, due to the opacity drops caused by the evaporation of ice and metal grains.
-Our two layer model ignores such complications.
+Our two-layer model ignores such complications.
 Grain growth also impacts the opacity in the outer regions as well as the existence or depth of radiative windows.
@@
 
- In this paper we study the formation of giant planet atmospheres under the assumption of a low planetesimal accretion regime in which KH contraction dominates the luminosity evolution of the atmosphere.
+%In this paper we study the formation of giant planets through core accretion.
+
+In this paper we build evolutionary models of giant planets forming in a protoplanetary disk.
+We consider atmosphere growth around a core of fixed mass, and determine the minimum core mass for which unstable atmosphere collapse occurs before the dissipation of the protoplanetary disk, at various locations in the disk.
+For a standard ISM power-law opacity and a polytropic EOS for the nebular gas, the minimum core mass is $\sim 8 M_{\oplus}$ at 5 AU and decreases to $\sim 3.5 M_{\oplus}$ at 100 AU.
+A larger mean molecular weight of the nebular gas increases the critical core mass, while a factor of ten reduction in opacity results in a critical core mass twice as low.
+Our results are applicable in the outer regions of the disk ($\gtrsim 5$ AU), where planets can undergo runaway gas accretion before opening a gap in the disk.
+If follows that core accretion at wide separations may result in massive gas giants, which could explain the formation of some of the directly imaged giant planets (\citealt{marois08}; \citealt{kalas08}; \citealt{lagrange10}).
+
+Our results are lower than the typically quoted critical core mass of 10 $M_{\oplus}$ in static studies.
+A giant planet can thus form faster if the core grows first, then accretion of solids slows down and a massive atmosphere accumulates, rather than if the core and envelope grow simultaneously in a high planetesimal accretion rate regime.
+We note that our model is a limiting case of core accretion, in which planetesimal accretion is negligible and the evolution of the gaseous envelope is dominated by KH contraction.
+We are thus able to place robust absolute lower limits on the core mass needed to form a giant planet, since additional core growth and heating limit the ability of the atmosphere to radiate energy and cool.
+Our results are not inconsistent with the existence of ice giants at large distances (such as Uranus and Neptune in our own Solar System), as ongoing accretion of solids can stall atmospheric growth and prevent the accumulation of a massive envelope before disk dissipation.
+
+% In this paper we study the formation of giant planet atmospheres under the assumption of a low planetesimal accretion regime in which KH contraction dominates the luminosity evolution of the atmosphere.
 We consider atmosphere growth around a core of fixed mass, and we build quasi-static two-layer atmosphere models with an inner convective region and an outer radiative region that matches smoothly onto the protoplanetary disk at the Hill radius.
@@
 A larger mean molecular weight of the nebular gas increases the critical core mass, while a factor of ten reduction in opacity results in a critical core mass twice as low.
-
-Our results are lower than the typically quoted critical core mass of 10 $M_{\oplus}$ in static studies.
+% 
+%Our results are lower than the typically quoted critical core mass of 10 $M_{\oplus}$ in static studies.
 A giant planet can thus form faster if the core grows first, then accretion of solids slows down and a massive atmosphere accumulates, rather than if the core and envelope grow simultaneously in a high planetesimal accretion rate regime.

### 2013-10-23 tex/draft/draft4.tex
edits to absract: reorganize and mention analytic model

@@
 
-\title{On the Minimum Core Mass for Giant Planet Formation}
+\title{On the Minimum Core Mass for Giant Planet Formation at Wide Separations}
 
@@
 
-The core accretion model proposes that giant planets form by the accretion of gas onto a solid protoplanetary core.
+The core accretion hypothesis proposes that giant planets form by the accretion of gas onto a solid protoplanetary core.
 The minimum (or critical) core mass to form a gas giant is typically quoted as $10 M_{\oplus}$.
-This value, however, depends on several factors, such as the location in the disk, nebular gas composition, or the accretion rate of solids.
-In this study we consider the limiting case of atmosphere growth around a core of fixed mass, in which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the envelope.
-We develop a two-layer atmosphere model with an inner convective region and an outer radiative zone that matches onto the protoplanetary disk, and we determine the minimum core mass for a giant planet to form within typical disk life timescales for a variety of disk conditions.
-We are primarily interested in the outer regions of the protoplanetary disk, where our assumptions of spherical symmetry and dust grains opacity are valid, and where the increasing number of directly imaged giant planets helps constrain formation models.
-The absolute minimum core mass required to nucleate atmosphere collapse is $\sim 8 M_{\oplus}$ at 5 AU and steadily decreases to $\sim 3.5 M_{\oplus}$ at 100 AU, for an ideal diatomic gas with a solar composition and a standard ISM dust opacity.
-Lower opacity and disk temperature significantly reduce the critical core mass, while a decrease in the mean molecular weight of the nebular gas results in a larger critical core mass.
-Our results yield lower mass cores than corresponding studies for large planetesimal accretion rates.
-As additional heating due to planetesimal accretion leads to slower cooling, our results represent an absolute minimum for the core mass needed to form a giant planet before disk dissipation.
-
-%Previous studies have found that there exists a ``critical core mass'' past which hydrostatic solutions no longer exist and unstable atmosphere collapse occurs.
-This core mass is typically quoted to be around $10 M_{\oplus}$.
-In static calculations of the critical core mass, planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse.
-In this study we consider the limiting case in which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the planet.
+This value, however, depends on several factors, such as the location in the disk, atmospheric opacity, and the accretion rate of solids.
+Motivated by ongoing direct imaging searches for giant planets, this study focuses on core mass requirements in the outer disk.
+Our model investigates the growth of atmospheres around solid cores of fixed mass.
+Without planetesimal accretion to heat the atmosphere, our limiting case gives the fastest possible atmospheric growth by Kelvin-Helmholtz contraction.
+We develop a two-layer, spherical atmosphere model with an inner convective region and an outer radiative zone that matches onto the protoplanetary disk.
+For different locations in the disk, we determine the minimum core mass for a giant planet to form within a typical disk lifetime of 3 Myr.
+The minimum core mass required to nucleate atmosphere collapse is $\sim 8 M_{\oplus}$ at 5 AU and steadily decreases to $\sim 3.5 M_{\oplus}$ at 100 AU, for a solar composition atmosphere with a standard ISM dust opacity.
+This trend is mainly explained by the cooler temperatures in the outer disk.
+Lower pressure and densities in the outer disk have a negligible effect on core mass requirements.
+At all distances, decreasing the dust opacity and/or increasing the mean molecular weight of the atmosphere reduces the critical core mass.
+The ongoing accretion of solids would delay or prevent atmospheric runaway growth.
+Thus our low values for minimum core masses do not contradict the existence of ice giants, but emphasize the importance of the accretion history of solids.
+We also develop a non-self-gravitating, analytic cooling model to help explain our numerical trends.
+Additionally, this comparison reveals that self-gravity significantly affects atmospheric evolution even when the atmosphere is only $\sim 20\%$ as massive as the core.
 

### 2013-10-23 tex/draft/draft2.tex
draft4 tex + pdf, but several other files added automatically when trying to merge the pdf

@@
 
-Standard calculations that consider planetesimal accretion as the main source of luminosity (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{rafikov06}) assume a steady state at all times and find that, for a given set of disk conditions, there is a maximum mass the core can have before the onset of unstable accretion: this is defined as the critical core mass (see section \ref{intro}).
+%Standard calculations that consider planetesimal accretion as the main source of luminosity (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{rafikov06}) assume a steady state at all times and find that, for a given set of disk conditions, there is a maximum mass the core can have before the onset of unstable accretion: this is defined as the critical core mass (see section \ref{intro}).
 In our model the atmosphere is no longer in steady state because energy balance requires gas accretion, and the atmosphere contracts on a Kelvin-Helmholtz time scale.
@@
 
-In this section we explore the dependence of the crossover time on core, disk and gas parameters, and determine the minimum core mass needed to initiate runaway gas accretion within the lifetime of the protoplanetary disk.
+In section \S\ref{coolingan} we defined the crossover time $t_{\rm co}$ as the time elapsed for the atmosphere to reach the crossover mass at $M_{\rm atm}=M\co$.
+Here we explore the dependence of the crossover time on core, disk and gas parameters, and determine the critical core mass needed to initiate runaway gas accretion within the lifetime of the protoplanetary disk.
 In section \ref{Mct} we show how the crossover time varies with core mass for fixed disk parameters, demonstrating that higher mass cores achieve runaway accretion more quickly.
 We explore how the crossover time is affected by by the mean molecular weight of the nebular gas in section \ref{muopacity} and by the disk temperature and pressure in section \ref{TPeffects}.
-Finally, we determine the minimum core mass necessary for a giant planet to form before the nebular gas dissipates in section \ref{critcore}; we define this minimum mass as the \textit{critical core mass} for runaway accretion to commence.
+Finally, we determine the critical core mass in section \ref{critcore} and explore its dependence on the dust opacity.
 
@@
 
-We begin by generating atmosphere evolution profiles for cores of different masses under a fixed set of disk conditions.
+We generate atmosphere evolution profiles for cores of different masses under a fixed set of disk conditions.
 Figure \ref{fig:tvsM} displays $t_{\rm{co}}$ for $M_{\rm c}$ between 5 and 14 $M_{\oplus}$ at $a=10$ AU in our fiducial disk.
 The envelope grows faster if the core is more massive, as it can bind a gaseous atmosphere more quickly due to larger gravity.
-This can also be seen from the analytical results in equation (\ref{eq:tcoolan}).
+This is consistent with our analytic results in equation (\ref{eq:tcoolan}).
 As such, there is a minimum core mass required to form a planet within a fixed crossover time, e.g.
-the typical lifetime of the protoplanetary disk, which we defined above as the critical core mass.
+the typical lifetime of the protoplanetary disk, which we defined as the critical core mass.
 We defer numerical estimates of the critical core mass to section \ref{critcore}.
@@
 A realistic gas mixture, however, has variable adiabatic gradient and molecular weight, due to non-ideal effects and processes like dissociation or ionization.
-Future work will include realistic equation of state tables, and we expect quantitatively different results.
+Future work will include realistic EOS tables, and we expect quantitatively different results.
 In this study we analyze only the effect of the mean molecular weight of the nebular gas on the crossover time.
@@
 
-\subsubsection{Disk Temperature and Pressure}
+\subsubsection{Role of Disk Temperature and Pressure}
 \label{TPeffects}
@@
 
-We find that the crossover time is shorter for lower disk temperatures and higher pressures. We discuss the two effects separately.
-
-An increase in disk temperature results in a longer crossover time, as a higher temperature causes an increase in opacity.
+The crossover time is shorter for lower disk temperatures and higher pressures. We discuss the two effects separately.
+
+An increase in disk temperature results in a longer crossover time, as a higher temperature increases the dust opacity.
 As a result, photons diffuse through the radiative layer more slowly, producing a lower planetary luminosity.
@@
 Different opacity assumptions will result in a different scaling with temperature of $t_{\rm{co}}$.
-We note, however, that the power law dependence of the crossover time on disk temperature found in our analytic model breaks down for the numerical model for low $T\di$.
+We note that the power law dependence of the crossover time on disk temperature found in our analytic model breaks down in the numerical model for low $T\di$.
 Nevertheless, the trends predicted by the scaled analytic approximation in the left panel of Figure \ref{fig:TPeffects} are still informative.
 For the full numerical solution, a factor of $\sim 3$ reduction in the disk temperature results in an decrease of the growth time by a factor of $\sim 30$.
-The atmosphere evolution is therefore strongly dependent on the temperature in the protoplanetary disk.
-
-From equations (\ref{eq:tcoolan}) and (\ref{eq:xi}), the crossover time is inversely proportional to disk pressure.
-A higher disk pressure therefore results in a shorter crossover time.
+The atmosphere evolution is thus highly dependent on the temperature in the protoplanetary disk.
+
+From equations (\ref{eq:tcoolan}) and (\ref{eq:xi}), the crossover time is inversely proportional to disk pressure, and hence a higher disk pressure results in a shorter crossover time.
 Increasing the disk pressure by a factor of $\sim 100$ only reduces the crossover time by a factor of $\sim 2$.
-As such, the crossover time is only weakly dependent on the nebular pressure.
+The crossover time is therefore only weakly dependent on the nebular pressure.
 
@@
 
-We can now combine the results obtained in section \ref{Mct} to determine the minimum core mass necessary for a protoplanet to initiate runaway gas accretion within the disk lifetime, i.e.
-to have a crossover time less than the disk lifetime.
-We  denote this mass the \textit{critical core mass}.
-
-%In the previous subsections we explored the effect of disk and gas properties on the growth time of an atmosphere accreting around a protoplanetary core of fixed mass.
-The main question we address here is under what conditions the atmosphere can grow before the gas in the protoplanetary disk is dissipated.
-Specifically, we determine the minimum core mass that is necessary for a protoplanet to initiate runaway gas accretion within the disk lifetime, i.e.
-to have a crossover time equal to the disk life timescale.
-We  denoted this mass the \textit{critical core mass} in section \ref{Mct}.
-
+We can now combine the results obtained in section \ref{Mct} to determine the minimum core mass to form a giant planet before the dissipation of the disk,.
+i.e.
+the critical core mass.
 Observations of protoplanetary disks show that they dissipate over a time scale of a few million years (e.g., \citealt{lagrange00}, \citealt{haisch01}, \citealt{goldreich04}).
@@
 %\vspace{-0.5in}
-\caption{The minimum core mass for an atmosphere to initiate runaway gas accretion within the lifetime of a typical protoplanetary disk $t \sim 3$ Myrs as a function of semi-major axis, for different mean molecular weight of the nebular gas.
-The analytic scaled results are plotted for comparison.
+\caption{The critical core mass for a disk lifetime $\tau \sim 3$ Myrs as a function of semi-major axis, for different mean molecular weight of the nebular gas.
+The scaled analytic results are plotted for comparison.
 The critical core mass is lower at larger stellocentric distances.}
@@
 
- In Figure \ref{fig:Mcvsa}, we see that the critical core mass decreases for larger separations, primarily due to the fact the disk temperature is lower in the outer regions of the disk.
-As we showed in section \ref{Mct}, the atmosphere grows faster for lower temperatures, which offsets the need for a larger core mass.
+Figure \ref{fig:Mcvsa}  shows that the critical core mass decreases for larger separations, primarily because the disk temperature is lower in the outer disk.
+As shown in section \ref{Mct}, the atmosphere grows faster for lower temperatures, which offsets the need for a larger core mass.
 The critical core mass is larger for a lower mean molecular weight, consistent with the results derived in section \ref{Mct}.
-We also explore the opacity effect on the critical core mass in Figure \ref{fig:Mcritopacity}.
-For an opacity, and therefore crossover time, that is an order of magnitude larger, the critical core mass is increased by roughly a factor of 2.
+Figure \ref{fig:Mcritopacity} shows the effect of opacity reduction on the critical core mass.
+For an opacity one order of magnitude lower than the standard ISM opacity, the critical core mass decreases by roughly a factor of two.
 

### 2013-10-23 tex/draft/draft4.tex
draft4 tex + pdf, but several other files added automatically when trying to merge the pdf

@@
 
-The fact that growth times peak before the crossover mass, when $M_{\rm{atm}} < M_{\rm c}$, is helpful for understanding atmospheric growth.
-The estimated timescale to runaway growth can be well approximated by only considering the time it takes to grow to the crossover mass, or even to a somewhat lower mass.
+The fact that growth times peak when $M_{\rm{atm}} < M_{\rm c}$, before the crossover mass, is helpful for understanding atmospheric growth.
+In contrast with our analytic model, where the time to runaway accretion cannot be calculated self-consistently, the numerical model provides information about when unstable collapse occurs, i.e.
+when growth times decrease.
+It follows that the time to runaway accretion, $t_{\rm run}$, can be estimated as the time at which the atmosphere growth time $M_{\rm atm}/\dot{M}$ drops to a fraction $f$ of its maximum value.
+This allows for a more rigorous estimate of the runaway accretion timescale, and thus of the critical core mass, as opposed to choosing $M_{\rm atm} \sim M\co$ as the criterion for collapse.
+In practice, however, the continuous accelerated growth in the late stages of the atmosphere evolution renders the estimated timescale to envelope collapse, and therefore the critical core mass, insensitive to the exact choice of runaway accretion time or atmosphere mass.
+Since $M_{\rm atm} < M\co$ when growth times peak, we may proceed by analogy with the analytic model and set $t=t_{\rm co}$, when $M_{\rm atm}=M\co$, as the criterion for collapse.
+We note that in some cases our cooling model breaks down before the crossover mass, as explained in section \S\ref{sec:endoftime}, hence evolution is stopped before $M_{\rm atm}=M\co$.
+However, this breakdown occurs for $M_{\rm atm} \sim 0.8-0.9 M\co$, comparable to the crossover mass, so the additional time to reach crossover is negligible.
+It follows that $t_{\rm run} \approx t_{\rm co}$.
+By analogy with the analytic results we thus choose $t_{\rm co}$ as the timescale to runaway accretion, as it will give the same numerical results as $t_{\rm run}$ and it is easier to understand in the context of other core accretion models.
+
+%The estimated timescale to runaway growth can be well approximated by only considering the time it takes to grow to the crossover mass, or even to a somewhat lower mass.
 This insensitivity to the upper mass threshold is characteristic of accelerating growth.
-We next show how this behavior is fortuitous, as our model assumptions can break down during the later stages of atmospheric growth.
+
+%We next show how this behavior is fortuitous, as our model assumptions can break down during the later stages of atmospheric growth.
 
@@
 Sometimes this breakdown occurs even before the crossover mass is reached, although still significantly after growth has started to accelerate.
+As the breakdown always occurs after growth times peak, it does not affect our estimate of the timescale to runaway accretion $t_{\rm run}$.
 In our models the breakdown occurs at lower atmosphere mass as the semimajor axis increases, since accelerated and eventually runaway growth occurs sooner at larger distances, as seen in \Fig{fig:growthtime}.

### 2013-10-23 tex/draft_paper2/draft_paper2.tex
draft4 tex + pdf, but several other files added automatically when trying to merge the pdf

@@
 
-The core accretion model assumes that giant planets form through gas accretion on to a solid core.
-The core and the atmosphere initially grow simultaneously through stages of quasistatic equilibrium.
-Once the core becomes massive enough, the atmosphere is no longer in hydrostatic balance and a rapid phase of runaway gas accretion commences.
-The minimum core mass for which unstable atmosphere collapse occurs is typically called the ``critical core mass''.
-In standard calculations of the critical core mass, planetesimal accretion dominates the atmosphere evolution, and the energy deposited by incoming planetesimal on to the core is radiated away by the atmosphere.
-In this study we consider a low planetesimal accretion regime in which the luminosity evolution of the atmosphere is dominated by Kelvin-Helmholtz contraction.
-We use the atmosphere structure and cooling model developed in Piso \& Youdin (2013) to derive the profile and evolution of atmospheres composed of a gas described by a realistic equation of state.
-We find that the minimum core mass (which we denote as \textit{critical core mass}) to form a giant planet before the dissipation of the protoplanetary disk is substantially increased compared to an ideal gas polytrope when non-ideal effects such as hydrogen dissociation and ionization are taken into account.
-Moreover, our results yield lower mass cores than corresponding studies for large planetesimal accretion rates.
-We therefore show that it is easier to form a planet by growing the core first, then accreting a massive gaseous envelope, rather than forming the core and atmosphere simultaneously.
+The core accretion model postulates that giant planets form through accretion of gas onto a solid core.
+When a core embedded in a protoplanetary disk accretes an atmosphere with mass comparable to itself, the envelope is no longer in hydrostatic balance and unstable atmosphere collapse occurs.
+Formation of a core requires rapid planetesimal accretion, which deposits energy into the atmosphere.
+Heating increases the core mass required for collapse.
+However, the planetesimal accretion rate need not be constant --- in particular, it may decline over time.
+In this study we consider the limiting regime in which planetesimal accretion is negligible and the luminosity evolution of a core's atmosphere is dominated by Kelvin-Helmholtz contraction.
+We enhance the model of Piso \& Youdin (submitted), which derives the evolution of atmospheres embedded in a gas disk, to employ a more realistic equation of state and dust opacity.
+We find that the minimum core mass required to form a giant planet before the dissipation of the protoplanetary disk increases by more than a factor of two when non-ideal effects such as hydrogen dissociation and quantum rotational effects are included in the equation of state, when compared to an ideal gas polytrope.
+However, this minimum core mass decreases back to $M_{\rm{c, crit}}\sim5 M_{\oplus}$ once we assume realistic dust opacities that take into account grain growth.
+Our results yield lower critical core masses than corresponding studies for large planetesimal accretion rates.
+We therefore conclude that it is easier to form a planet by growing the core first, then accreting a massive gaseous envelope, rather than forming the core and atmosphere simultaneously.
 
@@
 
-One of the prevalent theories of giant planet formation is the core accretion model  (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{boden86}, \citealt{wuchterl93}, \citealt{dangelo11}).
-In this model, a solid core is first formed; the core grows, and once it becomes large enough it can accumulate a massive atmosphere.
+One of the prevalent theories of giant planet formation is the core accretion model  (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{boden86}, \citealt{dangelo11}).
+It stipulates that Jupiter-sized planets form due to accretion of planetesimals into a solid core that grows large enough to attract a massive atmosphere.
+
+In models that assume a high planetesimal accretion rate (e.g., \citealt{rafikov06}), the gaseous envelope is heated solely by planetesimals and is therefore in a steady state at all times, in which all the incoming energy due to planetesimals is radiated away.
+In this regime, the core and atmosphere grow simultaneously, with the atmosphere increasing in mass faster than the core.
+As the envelope and core become comparable in mass, hydrostatic balance no longer holds and a rapid phase of unstable atmosphere collapse commences.
+The maximum core mass for which the atmosphere is still in hydrostatic equilibrium is defined in these studies as the ``critical core mass''.
+For a fixed planetesimal accretion rate and a set of disk conditions, the critical core mass is uniquely determined.
+
+Time-dependent core accretion models (e.g., \citealt{pollack96}, \citealt{ikoma00}) include variable planetesimal rates which allow for the Kelvin-Helmholtz (KH) contraction of the gaseous envelope to become important.
+In this regime, the atmosphere luminosity is not only supplied by planetesimals, but also by gas contraction.
+The envelope is no longer in a steady state, but rather it accretes gas as it cools.
+\citet{pollack96} find that the evolution of a giant planet can be separated into three phases.
+The core forms during phase 1 due to runaway planetesimal accretion, while the envelope mass stays small.
+Once the planet's feeding zone has been depleted of solids, planetesimal accretion is significantly reduced and can no longer balance radiative losses.
+As a result, the atmosphere cools while undergoing KH contraction; this is phase 2.
+The envelope mass now steadily increases while core growth is stalled.
+Runaway gas accretion begins once the atmosphere mass is comparable to the core mass in phase 3.
+\citet{pollack96} find that phase 2 lasts much longer than phases 1 and 3, and so the evolutionary timescale of the planet is set by KH contraction.
+Studying atmosphere growth onto a solid core of fixed mass (e.g., \citealt{pn05}) can thus provide accurate estimates for the timescale required to form a giant planet.
+
+%It is thus useful to consider atmosphere growth onto a solid core of fixed mass (e.g., \citealt{pn05}).
+
+Piso \& Youdin (submitted; hereafter Paper I) study atmosphere evolution around a fully grown core with fixed mass, assuming the luminosity of the envelope is solely due to KH contraction while planetesimal accretion is neglected.
+They build quasi-static two-layer atmosphere models, and derive a cooling model to connect static profiles in time and obtain an evolutionary history.
+They define the time $t$ at which unstable atmosphere collapse commences as the \textit{crossover time} $t_{\rm{co}}$, at which  $M_{\rm atm}(t_{\rm{co}})\sim M_{\rm c}$.
+From this they define as \textit{critical core mass} the minimum core mass for a protoplanet to undergo runaway gas accretion during the lifetime of the disk.
+They calculate the critical core mass for a variety of disk conditions, nebular gas compositions and opacities, and find that the critical core mass is smaller for lower disk temperatures (i.e., larger stellocentric distances), lower opacities and larger mean molecular weight of the gas.
+
+This study builds on the results of Paper I by making two important additions: we include realistic equation of state (EOS) effects, as well as realistic dust opacities.
+
+ Paper I assumes that the nebular gas is an ideal diatomic gas described by a polytropic equation of state (EOS).
+However, non-ideal effects such as dissociation and ionization have to be taken into account in order to obtain better quantitative results.
+In this study, we consider atmosphere growth assuming that the nebular gas is described by a realistic EOS given by the \citet{saumon95} EOS tables, and determine the critical core mass for a giant planet to form before the dissipation of the protoplanetary disk.
+We find that the critical core mass is more than twice as large than in the ideal case if realistic thermodynamic effects are taken into account.
+
+Another simplification in Paper I is the assumption that the dust opacity in radiative regions is that of interstellar grains.
+In reality, opacities in protoplanetary disks, while not tightly constrained observationally, are unlikely to be interstellar.
+As we consider a regime of low planetesimal accretion in which solids have been segregated from the gas and are not being added back in, it is likely that the grain opacity will be lower than the standard interstellar medium (ISM) opacity.
+This effect is explored in Paper I, and it is found that a factor of ten opacity reduction results in a critical core mass twice as small.
+In this study, we use realistic opacity tables that are calculated based on observations of protoplanetary disks and take into account grain growth.
+We find that the resulting critical core mass may be significantly lower when ISM opacities are used, i.e.
+by more than a factor of three at 5 AU, and around 1.5 times lower at 100 AU.
+%We therefore see that it is less sensitive to disk temperature and hence semi-major axis.
+
+This paper is organized as follows.
+In section \S\ref{sec2} we review the quasi-static and cooling models derived in Paper I.
+We discuss the variations in the adiabatic gradient caused by non-ideal EOS effects in section \S\ref{deladtable}, and their implications for atmosphere evolution in section \S\ref{EOSeffects}.
+We determine the minimum core mass to form a giant planet during the disk lifetime in section \S\ref{critical}, with realistic EOS and dust opacity assumptions.
+In section \S\ref{acc} we compare our results to those obtained by studies that consider planetesimal accretion as the dominant source of energy.
+Finally, we summarize our findings in section \S\ref{conclusions}.
+
+%a solid core is first formed; the core grows, and once it becomes large enough it can accumulate a massive atmosphere.
 In standard core accretion models, in order to form a core large enough to attract an atmosphere, a high planetesimal accretion rate is needed, on average.
@@
 
-However, the planetesimal accretion rate is not necessarily constant at a given location in the protoplanetary disk throughout the disk lifetime (e.g., \citealt{ikoma00}).
+%However, the planetesimal accretion rate is not necessarily constant at a given location in the protoplanetary disk throughout the disk lifetime (e.g., \citealt{ikoma00}).
 If the planetesimal accretion rate is very low, the atmosphere can no longer gain energy due to accretion of solids, and is instead dominated by gas accretion, contracting on a Kelvin-Helmholtz timescale.
@@
 
- PY13 assume that the nebular gas can be described by a polytropic equation of state (EOS) corresponding to an ideal diatomic gas: $\delad=2/7$.
+ %PY13 assume that the nebular gas can be described by a polytropic equation of state (EOS) corresponding to an ideal diatomic gas: $\delad=2/7$.
 In reality, however, non-ideal effects, such as gas dissociation and ionization, have to be taken into account.
@@
 
-This paper is organized as follows.
-In section \ref{sec2} we review the quasi-static and evolution models derived in PY13.
-We discuss the variations in the adiabatic gradient, and hence in the atmosphere structure, caused by a non-ideal equation of state in section \ref{deladtable}, and discuss the implications of this variability on the atmosphere evolution time in section \ref{EOSeffects}.
-We determine the minimum core mass to form a giant planet during the disk lifetime when the nebular gas is described by a realistic equation of state in section \ref{critical}.
-In section \ref{acc} we compare our results to similar results obtained in studies that consider planetesimal accretion as the dominant source of energy.
-Finally, we summarize our findings in section \ref{conclusions}.
-
 % In section \ref{sec2} we describe the assumptions of our atmosphere model, and derive the basic equations that govern the structure and evolution of the atmosphere.
@@
 
-In this section we review the model developed in PY13 for the structure and evolution of a planetary atmosphere embedded in a protoplanetary disk.
-We describe the assumptions of the model and the properties of our assumed protoplanetary disk in section \ref{model}, and we summarize the equations describing the structure and time evolution of the atmosphere in section \ref{struct}.
+In this section we review the model developed in Paper I for the structure and evolution of a planetary atmosphere embedded in a protoplanetary disk.
+We describe the assumptions of the model and the properties of our assumed protoplanetary disk in section \S\ref{model}, and we summarize the equations describing the structure and time evolution of the atmosphere in section \S\ref{struct}.
 
@@
 
-We assume that the planet consists of a solid core of fixed mass and a two-layer atmosphere composed of an inner convective zone and an outer radiative zone that matches smoothly on to the disk.
-The two regions are separated by the Schwarzschild criterion for convective instability.
+We assume that the planet consists of a solid core of fixed mass and a two-layer atmosphere composed of an inner convective region and an outer radiative zone that matches smoothly onto the disk.
+The two regions are separated by the Schwarzschild criterion for convective instability (see section \S\ref{struct}) .
 We denote the surface between the two regions as the radiative-convective boundary (RCB), defined by a radius $r=R\cb$.
-We assume a low planetesimal accretion regime in which the atmosphere evolution is dominated by KH contraction while planetesimal accretion is neglected.
-Moreover, we assume that the luminosity is constant throughout the outer radiative region.
-The atmosphere is spherically symmetric and self-gravitating.
-It consists of a hydrogen-helium mixture, with hydrogen and helium mass fractions of 0.7 and 0.3, respectively.
-We further assume that the envelope evolves through stages of quasi-static equilibrium.
+We assume a low planetesimal accretion regime in which the atmosphere evolution is dominated by KH contraction.
+Moreover, we assume that the luminosity is constant throughout the outer radiative region (see section \S\ref{critical} for additional discussion).
+The atmosphere is spherically symmetric, self-gravitating and in hydrostatic balance.
+We note that spherical symmetry confines us to the outer regions of the disk ($a\gtrsim5$ AU), where the disk scale height is larger than the radius at which the planet matches onto the disk (see Paper I for further details).
+The nebular gas is composed of a hydrogen-helium mixture, with hydrogen and helium mass fractions of 0.7 and 0.3, respectively.
+We assume that the envelope evolves through stages of quasi-static equilibrium.
 
@@
 
-\noindent with $r$ the radial coordinate, $P$, $T$ and $\rho$ the gas pressure, temperature and density, respectively, $m$ the mass enclosed by the radius $r$, $L$ the luminosity from the surface of radius $r$.
+\noindent with $r$ the radial coordinate, $P$, $T$ and $\rho$ the gas pressure, temperature and density, respectively, $m$ the mass enclosed by the radius $r$, $L$ the luminosity from the surface of radius $r$, and $G$ the gravitational constant.
 The $\epsilon$ term represents the rate at which internal heat is generated per unit mass, while $\epsilon_g \equiv -T \frac{ds}{dt}$ is the heating per unit mass due to gravitational contraction, with $s$ the specific gas entropy.
-We do not take into account any internal energy sources and we therefore set $\epsilon=0$.
-The expression for the temperature gradient $\nabla \equiv \frac{d \ln T}{d \ln P}$ depends on whether energy is transported throughout the atmosphere by radiation or convection.
+We do not take into account any internal energy sources and set $\epsilon=0$.
+The temperature gradient $\nabla \equiv \frac{d \ln T}{d \ln P}$ depends on whether energy is transported throughout the atmosphere by radiation or convection.
 In the case of radiative diffusion for an optically thick gas, the temperature gradient is
@@
 In our models the atmosphere is optically thick throughout the outer boundary.
-On the other hand, when the energy is transported outwards through convection, the temperature gradient is given by
+Where energy is transported through convection the temperature gradient is given by
 
@@
 \end{equation}
-\\
 
@@
 The equation set (\ref{eq:struct}) is supplemented by an equation of state (EOS) relating pressure, temperature and density, as well as an opacity law.
-In this study we use the interpolated EOS tables of \citet{saumon95} for a helium mass fraction $Y=0.3$.
-More details on the EOS tables and the methodology of combining the separate tables for hydrogen and helium are presented in section \ref{deladtable} and \App{EOStables}.
-
-We assume a standard ISM opacity power law given by
+Paper I assumes an ideal gas polytropic EOS, $P=K\rho^{\gamma}$, with $K$ the adiabatic constant and $\gamma$ the adiabatic index.
+In this study we use realistic EOS tables, as explained in section \S\ref{deladtable}.
+Furthermore, Paper I assumes a power-law opacity given by
 
@@
 \label{eq:opacitylaw}
-\kappa=\kappa_0 \Big(\frac{P}{P_{\rm{ref}}}\Big)^{\alpha} \Big(\frac{T}{T_{\rm{ref}}}\Big)^{\beta},
-\end{equation}
-
-\noindent with $\alpha$, $\beta$, $\kappa_0$ constants, and $T_{\rm{ref}}$ and $P_{\rm{ref}}$ a normalizing temperature and pressure, respectively.
-To estimate $\alpha$, $\beta$ and $\kappa_0$ we use the \citet{bell94} opacity laws for ice grains: $\alpha =0 $, $\beta=2$ and $\kappa_0=2$.
-These values are valid only for low disk temperatures: $T_d \lesssim 100 K$.
-We therefore consider protoplanets that form in the outer parts of the disk where temperatures are low, i.e.
-$a \geq 5$ AU.
-Dust settling or grain growth will result in different numerical values for the $\kappa_0$ and $\beta$ coefficients.
-
-We further discuss our choice of core parameters and boundary conditions.
-We assume a solid core of fixed mass $M_{\rm c}$ with a radius $R_{\rm c}=(3 M_{\rm c}/4 \pi \rho_{\rm c})^{1/3}$, where $\rho_{\rm c}$ is the core density.
+\kappa=2 F_{\kappa} \Big(\frac{T}{T_{\rm ref}}\Big)^{\beta},
+\end{equation}  
+\noindent with $\beta$ and $F_{\kappa}$ constants, and $T_{\rm{ref}}$ a normalizing temperature.
+To estimate $\beta$ and $F_{\kappa}$ we use the \citet{bell94} opacity laws for ice grains: $\beta=2$, $F_{\kappa}=1$ and $T_{\rm ref}=100$ K.
+We note that these values are valid only for low disk temperatures, $T_d \lesssim 100 K$, where all dust components are present.
+Dust settling and grain growth lower both the normalization factor $F_{\kappa}$ and the power-law coefficient $\beta$.
+In this paper, we explore more realistic opacity laws and their effect on the critical core mass when compared to the ISM power-law (\ref{eq:opacitylaw}), as discussed in section \S\ref{critical}.
+
+%In addition to the \citet{bell94} opacity, we also explore more realistic opacity laws, as discussed in section \S\ref{critical}.
+
+%We further discuss our choice of core parameters and boundary conditions.
+
+We assume that the atmosphere forms around a solid core of fixed mass $M_{\rm c}$ with a radius $R_{\rm c}=(3 M_{\rm c}/4 \pi \rho_{\rm c})^{1/3}$, where $\rho_{\rm c}$ is the core density.
 We choose $\rho_{\rm c}=3.2$ g cm$^{-3}$ (e.g., \citealt{pap99}).
-The atmosphere is assumed to match on to the protoplanetary at the Hill radius, $R_{\rm H} \equiv a \Big(\frac{M_{\rm p}}{3 M_{\odot}}\Big)^{1/3}$, the distance at which the gravitational attraction of the planet and the tidal gravity due to the host star are equal.
-Outside the Hill sphere, the gravity of the planet is overcome by the tidal gravity from its host star, and hence only gas that lies within the Hill sphere can be gravitationally bound to the planet.
+The atmosphere is assumed to match on to the protoplanetary disk at the Hill radius $R_{\rm H} \equiv a \Big(\frac{M_{\rm p}}{3 M_{\odot}}\Big)^{1/3}$, the distance at which the gravitational attraction of the planet and the tidal gravity due to the host star are equal.
+Outside the Hill sphere, the planet gravity is overcome by stellar tidal gravity, and hence only gas that lies within the Hill sphere can be gravitationally bound to the planet.
 The effective outer boundary of the atmosphere is the surface defined by the Bondi radius, $R_{\rm B} \equiv \frac{G M_{\rm p}}{c_{\rm s}^2}=\frac{G M_{\rm p}}{\mathcal{R} T\di}$.
 This is the distance from the planet at which the thermal energy of the nebular gas is of the order of the gravitational energy of the planet.
-Here $G$ is the gravitational constant, $M_{\rm p}$ is the total mass of the planet, $c_{\rm s}$ is the isothermal sound speed, and $\mathcal{R}$ is the reduced gas constant: $\mathcal{R}=k_b/(\mu m_p)$, with $k_b$ the Boltzmann constant and $m_p$ the proton mass.
-Outside the Bondi sphere, the gravity of the planet is too weak to significantly affect the nebular gas, which justifies the choice of Bondi radius as the relevant radius of the atmosphere.
+Here $M_{\rm p}$ is the total mass of the planet, $c_{\rm s}$ is the isothermal sound speed, and $\mathcal{R}$ is the reduced gas constant: $\mathcal{R}=k_b/(\mu m_p)$, with $k_b$ the Boltzmann constant and $m_p$ the proton mass.
+For $R_{\rm B}<R_{\rm H}$, several studies assume that the atmosphere matches onto the disk at $R_{\rm B}$ (e.g., \citealt{ikoma00}, \citealt{pollack96}).
+One reason we always choose $R_{\rm H}$ as the matching radius is that, for hydrostatic solutions, the density at $R_B$ is larger than the disk density by an order unity factor \citep{rafikov06}.
+Additionally, even though gas flows no longer circulate the planet outside $R_{\rm B}$, the density structure between $R_{\rm B}$ and $R_{\rm H}$ remains spherical and is still well described by hydrostatic balance \citep{ormel13}, which justifies our outer boundary choice.
+At the Hill radius, the temperature and pressure are given by the nebular temperature and pressure: $T(R_{\rm H})=T_{\rm d}$ and $P(R_{\rm H})=P_{\rm d}$.
+We note, however, that we define the planet mass as the mass enclosed inside the smaller of $R_{\rm B}$ or $R_{\rm H}$.
+
+%Outside the Bondi sphere, the gravity of the planet is too weak to significantly affect the nebular gas, which justifies the choice of Bondi radius as the relevant radius of the atmosphere.
 However, the nebular gas is still perturbed outside the Bondi sphere, and therefore the Hill radius is the correct scale for matching on to the disk.
 This choice for the atmosphere boundary applies only when the Bondi radius is smaller than the Hill radius; when $R_{\rm B}>R_{\rm H}$, the atmosphere only extends out to the Hill radius, since material cannot be gravitationally bound to the protoplanet outside the Hill sphere.
-At the Hill radius, the temperature and pressure are given by the nebular temperature and pressure: $T(R_{\rm H})=T_{\rm d}$ and $P(R_{\rm H})=P_{\rm d}$.
-For a given core mass, the atmosphere profile and evolution are therefore uniquely determined by the outer boundary conditions.
 
@@
 
-Lastly, we review the cooling model developed in PY13 used to determine the time evolution of the atmosphere between subsequent static models.
+Lastly, we review the cooling model developed in Paper I used to determine the time evolution of the atmosphere between subsequent static models.
 A protoplanetary atmosphere embedded in a gas disk satisfies the following cooling equation:
@@
 \end{equation}
-
-Here, $L$ is the total luminosity, $L_{\rm c}$ is the luminosity from the solid core, and may include planetesimal accretion and radioactive decay, $\Gamma$ is the rate of internal heat generation, $\dot{E}$ is the rate at which total energy (internal and gravitational) is lost, and  $e_{\mathrm{acc}}$ is the specific total energy brought in by mass accreting at the rate $\dot{M}$: $e_{\mathrm{acc}}=u-G M/R$.
-The last term represents the work done on a surface mass element.
+Here, $L$ is the total luminosity, $L_{\rm c}$ is the luminosity from the solid core, which may include planetesimal accretion and radioactive decay, and $\Gamma$ is the rate of internal heat generation.
+We set $L\co=\Gamma=0$.
+The $\dot{E}$ term is the rate at which total energy (internal and gravitational) is lost, $e_{\mathrm{acc}}$ is the specific total energy brought in by mass accreting at the rate $\dot{M}$: $e_{\mathrm{acc}}=u-G M/R$, and the last term represents the work done on a surface mass element.
 
@@
 We obtain an evolutionary series for the atmosphere by connecting sets of subsequent static atmospheres through the cooling equation (\ref{eq:coolingglobal}).
-Details of our numerical procedure are described in PY13.
+Details of our numerical procedure are described in Paper I.
 
@@
 
-This section explains how non-ideal EOS effects influence the behavior of the thermodynamic variables when compared to an ideal gas polytrope.
-Figure \ref{fig:deladmap} shows a contour plot of the adiabatic gradient (defined in equation \ref{eq:delad})  as a function of gas temperature and pressure, which was obtained by interpolating and extending the \citet{saumon95} EOS tables as described in \App{EOStables}.
-
+In order to describe the nebular gas, we use the interpolated EOS tables of \citet{saumon95} for a helium mass fraction $Y=0.3$, and extend them to lower temperatures and pressures corresponding to the conditions in our fiducial disk.
+More details on our extension procedure and on the methodology of combining the separate tables for hydrogen and helium are presented in \App{EOStables}.
+
+For an ideal gas polytropic EOS, the adiabatic gradient is constant.
+In contrast, non-ideal effects such as dissociation or ionization produce temperature-dependent variations in $\delad$.
+Figure \ref{fig:deladmap} shows a contour plot of the adiabatic gradient (defined in equation \ref{eq:delad})  as a function of gas temperature and pressure.
 We distinguish three separate temperature regimes:
+
+%, which was obtained by interpolating and extending the \citet{saumon95} EOS tables as described in \App{EOStables}.
 
@@
 \begin{enumerate}
-\item Intermediate temperature regime (300 K $\lesssim T \lesssim 3000$ K), where the hydrogen-helium mixture behaves like an ideal gas with a polytropic EOS.
-\item High temperature regime ($T \gtrsim$ 3000 K), where dissociation of molecular hydrogen occurs, followed by ionization of atomic hydrogen.
+\item Intermediate temperature regime (300 K $\lesssim T \lesssim 2000$ K), where the hydrogen-helium mixture behaves like an ideal gas with a polytropic EOS.
+\item High temperature regime ($T \gtrsim$ 2000 K), where dissociation of molecular hydrogen occurs, followed by ionization of atomic hydrogen for $T \gtrsim 10,000$ K.
 \item Low temperature regime ($T \lesssim 300$ K), where the rotational states of the hydrogen molecule are not fully excited. % temperature is low enough for rotational motion to reduce.
@@
 \caption{Contour plot of the adiabatic gradient $\delad$ for a hydrogen-helium mixture as a function of gas temperature and pressure.
-The upper right rectangle encloses the region described by the original \citet{saumon95} EOS tables, while the rest of the plot is our extension to lower temperatures and pressures.
-The black curves represent constant entropy adiabats, with the labels the natural logarithm of the absolute entropy per unit mass.
-At high temperatures, hydrogen dissociates and ionizes, while at low temperatures the rotational states of the hydrogen molecule are only partially excited and it no longer behaves like an ideal diatomic gas.
-Our extension matches smoothly with the original tables in our regime of interest (encompassed by the constant entropy curves).}
+The upper-right rectangle encloses the region described by the original \citet{saumon95} EOS tables, while the rest of the plot is our extension.
+The black curves represent constant entropy adiabats, with the labels $\log_{10}(S)$ [erg K$^{-1}$ g$^{-1}$].
+The regions in which the EOS is either invalid or not computed are masked in white.
+Our extension is only valid for $T \lesssim 2000$ K, since it does not take into account hydrogen dissociation.
+We choose $T=1500$ K as a conservative temperature cutoff.
+\citet{saumon95} do not compute the EOS at very high pressures, since hydrogen is solid or may form a Coulomb lattice in this regime, and thus their EOS treatment is no longer valid.
+While the boundaries of the region in which the free-energy EOS treatment fails can be determined from fundamental thermodynamic constraints, such calculations are not the object of this work.
+Instead, we note that less pressure is needed for hydrogen to become solid at lower temperatures, and so we choose as boundary a constant entropy curve ($\log(S)=8.4$) above the region in which the \citet{saumon95} model fails.
+The expressions derived in \App{EOStables} are sufficient to give good results for the colored regions of the extended map, which fully cover the temperature and pressures ranges required by our models.}
 \label{fig:deladmap}
 \end{figure}
+%but not the white ones. We note, however, that the temperature and pressure ranges required by our models are fully covered by the colored regions, where our expressions are valid
+
+%If follows that the expressions derived in \App{EOStables} cannot be used to extend the original tables in this temperature and pressure regime.
+
+%At high temperatures, hydrogen dissociates and ionizes, while at low temperatures the rotational states of the hydrogen molecule are only partially excited and it no longer behaves like an ideal diatomic gas.
 
@@
 
-For temperatures less than $\sim 2000$ K but larger than $\sim 300$ K, the hydrogen molecule is not energetic enough to dissociate and hydrogen behaves as an ideal diatomic gas.
+For temperatures less than $\sim$$2000$ K but larger than $\sim$$300$ K, the hydrogen molecule is not energetic enough to dissociate and hydrogen behaves as an ideal diatomic gas.
 We see this in Fig.
 \ref{fig:deladmap} for 300 K $\lesssim T \lesssim 3000$ K, where the adiabatic gradient is approximately constant.
-The helium component of the gas causes a slight increase in the adiabatic index: $\delad \approx 0.3$ in this temperature range rather than 2/7 as is the case for a diatomic gas.
+The helium component of the gas causes a slight increase in the adiabatic index: $\delad \approx 0.3$ in this temperature range rather than 2/7 as is the case for a diatomic gas \footnote{Recall $\delad \equiv \frac{\gamma-1}{\gamma}$ for an ideal gas, with $\gamma$ the adiabatic index: $\gamma=7/5$ for a diatomic gas and 5/3 for a monatomic gas.}.
 
@@
 As the temperature becomes higher than $T \sim 2000-3000$ K, the internal energy of the hydrogen molecule becomes large enough to break the covalent bond between the atoms, and hydrogen starts dissociating.
-Further energy increase at temperatures of $\sim 10^4$ K results in electrons being removed from atoms, i.e.
+Further energy increase at temperatures of $\sim$$10^4$ K results in electrons being removed from atoms, i.e.
 hydrogen ionizes.
@@
 We see in Figure \ref{fig:deladmap} that the adiabatic gradient decreases significantly in regions of partial dissociation and partial ionization.
-This behavior is different than that of a mixture of molecular and atomic hydrogen (a mixture of protons and electrons), where $2/7<\delad<2/5$ ($\delad=2/5$).
+This behavior is different than that of a mixture of molecular and atomic hydrogen where $2/7<\delad<2/5$, or a mixture of protons and electrons where $\delad=2/5$.
 We  explain this in what follows.
@@
 For a mixture of ideal gases, the total internal energy is given by the sum of the internal energies of the individual gases.
-When a gas dissociates, however, the energy used to break down the molecules has to also be taken into account.
-This energy depends on the dissociation fraction and can be determined from the Saha equation (see e.g., \citealt{kippenhahn90}).
-The dissociation fraction only depends on gas temperature and density (see \App{deladioniz}), and hence only on the EOS.
-An expression for the adiabatic gradient as a function of the dissociation fraction is presented in \App{deladioniz}.
-As expected, the adiabatic gradient is $\delad=2/7$ for pure molecular hydrogen and $\delad=2/5$ when hydrogen is fully dissociated, but decreases significantly during partial dissociation and it is smallest when half of the gas is dissociated.
-
+When a gas dissociates, however, part of the internal energy of the system is used to break down molecules, which reduces the amount of energy available to increase the temperature of the system, thus lowering the adiabatic gradient.
+The energy required to dissociate depends on the dissociation fraction, which can be determined from the Saha equation (see e.g., \citealt{kippenhahn90}).
+The dissociation fraction only depends on gas temperature and density, and hence only on the EOS.
+
+An expression for the adiabatic gradient as a function of the dissociation fraction is presented in \App{deladdiss}.
+As expected, the adiabatic gradient is $\delad=2/7$ for pure molecular hydrogen and $\delad=2/5$ when hydrogen is fully dissociated, but decreases significantly during partial dissociation and is smallest when half of the gas is dissociated.
 This is consistent with the behavior we see in Figure \ref{fig:deladmap}.
-At constant entropy, an increase in pressure results in an increase in the internal energy of the system, which usually causes the temperature to rise significantly.
+
+%At constant entropy, an increase in pressure results in an increase in the internal energy of the system, which usually causes the temperature to rise significantly.
 During partial dissociation, however, part of the internal energy is used to remove the electrons from atoms, and therefore there is less energy available to increase the temperature of the system.
@@
 
-The ionization of atomic hydrogen is also dictated by the Saha equation, with the dissociation energy replaced by ionization energy (equal to 13.6 eV for hydrogen, e.g.
-\citealt{mandl89}), hence the adiabatic gradient has an analogous behavior, consistent with Fig.
+The ionization of atomic hydrogen is also dictated by the Saha equation, with the dissociation energy replaced by ionization energy, hence the adiabatic gradient has an analogous behavior, consistent with Fig.
 \ref{fig:deladmap}.
@@
 As a diatomic molecule, hydrogen has five degrees of freedom, three associated with translational motion and two associated with rotation.
-At room temperature, the rotational states are fully excited.
-The excitation temperature for rotation is $\Theta_r \approx 85$ K for the hydrogen molecule (e.g., \citealt{kittel}); as the gas temperature becomes comparable to $\Theta_r$, fewer rotational states are excited and rotation entirely ceases as $T \rightarrow 0$.
+The excitation temperature for rotation is $\Theta_r \approx 85$ K for the hydrogen molecule (e.g., \citealt{kittel}), hence the rotational states are fully excited at room temperature.
+As the gas temperature becomes comparable to $\Theta_r$, fewer rotational states are excited and rotation entirely ceases as $T \rightarrow 0$.
 %Here we explore the quantum mechanical effects associated with rotation.
@@
 
- Molecular hydrogen occurs in two isomeric forms: orthohydrogen, with parallel proton spins, and parahydrogen, with antiparallel proton spins.
-The Pauli exclusion principle requires the total wavefunction of two fermions, such as protons, to be antisymmetric.
+ Molecular hydrogen occurs in two isomeric forms, parahydrogen and orthohydrogen.
+Parahydrogen has antiparallel proton spins and thus an antisymmetric wave function.
+From the Pauli exclusion principle it follows that it can only occupy symmetric rotational states with even angular quantum number $j$ \citep{farkas35}.
+In contrast, orthohydrogen has parallel proton spins and a symmetric wave function, and can therefore only occupy states with odd $j$.
+At equilibrium, the relative abundance of the ortho- and para- states is given by the ratio of their partition functions, described in \App{EOStables}.
+At $T \rightarrow 0$ all hydrogen molecules are in the ground state with $j=0$, which corresponds to parahydrogen.
+As the temperature increases, parahydrogen starts converting into orthohydrogen, resulting in an ortho-para equilibrium ratio of 3:1 at room temperature.
+
+The adiabatic gradient scales as $\delad \sim 1/c_{\rm v}=1/(c_{\rm v,t}+c_{\rm v,r}) \sim 1/c_{\rm v, r}$, with $c_{\rm v}$ the specific heat capacity at constant volume, and $c_{\rm v, t}$ and $c_{\rm v,r}$ the translational and rotational components, respectively.
+The second equality is due to the fact that $c_{v, t}$ is independent of temperature and equal to $3\mathcal{R}/2$.
+We can therefore understand the $\delad$ behavior by studying the dependence on temperature of $c_{\rm v, r}$ of the ortho-para mixture.
+
+The internal energy per unit mass and specific heat capacity associated with rotation for the individual isomers and for the equilibrium mixture can be derived from their partition functions (see \App{EOStables} for details), and are plotted in Figure  \ref{fig:ucvr} (after \citealt{farkas35}, Figure 1).
+At low temperatures, parahydrogen is in the $j=0$ state and has no rotational energy, while orthohydrogen is in the $j=1$ state and has the energy of its first rotational level.
+Both para- and orthohydrogen, as well as their equilibrium mixture, behave like monatomic gases at low temperatures and thus have zero rotational heat capacity.
+This is consistent with $\delad=2/5$ at low temperatures as seen in Figure \ref{fig:deladmap}.
+As the temperature increases, the energetically higher-lying rotational states of para- and orthohydrogen are populated and the heat capacity of both spin isomers increases as a result.
+We note that the heat capacity of the equilibrium mixture is not a weighted average of the heat capacities of the individual components because it takes into account both the rotational energy uptake of para- and orthohydrogen, and also the shift in their equilibrium concentrations with temperature.
+This results in a peak in the heat capacity of the mixture around $\sim$$50$ K, as seen in the bottom plot of Figure \ref{fig:ucvr}.
+As $\delad \sim 1/c_{\rm{v,r}}$, it follows that the adiabatic gradient has to reduce, reach a minimum, then increase as the temperature rises, as shown in Figure \ref{fig:deladmap}.
+
+% As the temperature increases, the energetically higher-lying ($j=1$) ortho-hydrogen is formed, and the concomitant energy increase is seen as a peak in the heat capacity of the equilibrium mixture.
+
+%There are two significant maxima in the heat capacities of parahydrogen and of the mixture.
+At very low temperatures, the heat capacity of parahydrogen is zero because only the lowest accessible energy level $j=0$ is occupied and a temperature increase does not provide enough energy to populate the next higher level.
+When the temperature becomes sufficiently high to populate the second lowest level $j=2$, the heat capacity rapidly increases, passes through a maximum and starts to decrease when the second lowest level becomes saturated.
+
+%the adiabatic gradient is inversely proportional to the heat capacity, it
+
+%means that  the former has to first decrease from 2/5 as the temperature increases, reach a minimum around 50 K ($\delad \approx 0.25$ from Fig.
+\ref{fig:deladmap}), then gradually increase to 2/7 as for a diatomic gas.
+This behavior is illustrated in Figure \ref{fig:deladmap}.
+
+%The maxima in the ortho-para mixture appears when parahydrogen starts converting into orthohydrogen.
+The heat capacity of the equilibrium mixture is not a weighted average of the heat capacities of the individual components because it takes into account both the rotational energy uptake of para- and orthohydrogen, and also the shift in their equilibrium concentrations with temperature.
+At $T=0$, only parahydrogen is present in the equilibrium mixture; as the temperature is increased, the energetically higher-lying ($j=1$) ortho-hydrogen is formed, and the concomitant energy increase is seen as a peak in the heat capacity.
+As the adiabatic gradient is inversely proportional to the heat capacity, it means that  the former has to first decrease from 2/5 as the temperature increases, reach a minimum around 50 K ($\delad \approx 0.25$ from Fig.
+\ref{fig:deladmap}), then gradually increase to 2/7 as for a diatomic gas.
+This behavior is illustrated in Figure \ref{fig:deladmap}.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/ortho_para_energy.pdf}
+%%\vspace{-0.5in}
+\caption{Internal energy per unit mass and specific heat capacity associated with rotation for parahydrogen (dashed blue), orthohydrogen (dotted green), and the equilibrium mixture (solids red) as a function of temperature.
+After \citet{farkas35}, Figure 1.}
+\label{fig:ucvr}
+\end{figure}
+
+ %The partition functions for ortho- and parahydrogen are described in \App{EOStables}.
+ %At equilibrium, the relative abundance of the ortho and para states is given by the ratio of their partition functions.
+For very low temperatures there is only parahydrogen, as molecules are in the ground state with $j=0$, which corresponds to the para state.
+As the temperature is increased, parahydrogen starts converting into orthohydrogen, resulting in an ortho-para equilibrium ratio of 3:1 at room temperature.
+
+  %orthohydrogen, with parallel proton spins, and parahydrogen, with antiparallel proton spins.
+
+ %The Pauli exclusion principle requires the total wavefunction of two fermions, such as protons, to be antisymmetric.
 As such, a symmetric spin wavefunction requires an antisymmetric rotational wavefunction, and vice-versa (\citealt{farkas35}).
@@
 The partition functions for ortho- and parahydrogen are described in \App{EOStables}.
- At equilibrium, the relative abundance of the ortho and para states is given by the ratio of their partition functions.
+ %At equilibrium, the relative abundance of the ortho and para states is given by the ratio of their partition functions.
 For very low temperatures there is only parahydrogen, as molecules are in the ground state with $j=0$, which corresponds to the para state.
@@
 
-The internal energy and specific heat per unit mass associated with rotation for the individual isomers and for the equilibrium mixture can be derived from equations (\ref{eq:Zpara}), (\ref{eq:Zortho}), (\ref{eq:Zrspin}), (\ref{eq:u}) and (\ref{eq:cv}) and plotted in Figure \ref{fig:ucvr} (see also \citealt{farkas35}, Figure 1).
-The para state has no rotational energy for low temperatures, since all the molecules occupy the rotational level with $j=0$.
-Orthohydrogen, on the other hand, is in the $j=1$ state, and so has an energy given by the energy of its first rotational level.
-Since all the hydrogen mixtures behave like monatomic gases at low temperatures, their rotational heat capacity is zero in this region.
-This is consistent with $\delad=2/5$ at low temperatures as seen in Fig.
-\ref{fig:deladmap}.
-There are two significant maxima in the heat capacities of parahydrogen and of the mixture.
-At very low temperatures, the heat capacity of parahydrogen is zero because only the lowest accessible energy level $j=0$ is occupied and a temperature increase does not provide enough energy to populate the next higher level.
-When the temperature becomes sufficiently high to populate the second lowest level $j=2$, the heat capacity rapidly increases, passes through a maximum and starts to decrease when the second lowest level becomes saturated.
-The maxima in the ortho-para mixture appears around the time when parahydrogen starts converting into orthohydrogen.
-The heat capacity of the equilibrium mixture is not a weighted average of the heat capacities of the individual components because it takes into account both the rotational energy uptake of para- and ortho-hydrogen, and also the shift in their equilibrium concentrations with temperature.
-At $T=0$, only para-hydrogen is present in the equilibrium mixture; as the temperature is increased, the energetically higher-lying ($j=1$) ortho-hydrogen is formed, and the concomitant energy increase is seen as a peak in the heat capacity.
-As the adiabatic gradient is inversely proportional to the heat capacity, it means that  the former has to first decrease from 2/5 as the temperature increases, reach a minimum around 50 K ($\delad \approx 0.25$ from Fig.
-\ref{fig:deladmap}), then gradually increase to 2/7 as for a diatomic gas.
-This behavior is illustrated in Fig.
-\ref{fig:deladmap}.
+%\vspace{0.1in}
+
+\section{Equation of State Effects on Atmosphere Evolution}
+\label{EOSeffects}
+
+Variations in the adiabatic gradient $\delad$ due to partial dissociation and quantum rotational effects (see section \S\ref{deladtable}) have two competing effects on the evolution of a core's atmosphere: (1) they yield a lower envelope luminosity, and (2) a larger amount of energy that needs to be radiated per unit mass, when compared to an ideal gas.
+Both effects combined slow down accretion and increase the growth time of the atmosphere, and therefore the crossover time and critical core mass.
+% (both defined in section \ref{intro}).
+%In this section we explore the effect of partial dissociation and hydrogen spin isomers (see section \ref{deladtable}) on the atmosphere evolution.
+%Due to the equation of state described in section \ref{deladtable}, a realistic atmosphere will not have a constant adiabatic gradient $\delad$, as was assumed in PY13.
+We thus first investigate the differences in atmosphere profiles and evolution for ideal gas polytropes with different adiabatic gradients in section \ref{deladpoly}.
+We then show how the variable adiabatic gradient affects the time evolution of the atmosphere in section \ref{deladeffect}.
+
+%\subsection{Effect of Adiabatic Gradient on Atmosphere Cooling Time}
+%\label{deladeffect}
+
+%In this section we discuss the effect of the variable adiabatic index described in subsection \ref{deladinterp} on the atmosphere cooling time.
+
+%We now explore the separate effects of hydrogen dissociation at high temperatures and spin isomers at low temperatures.
+From the adiabatic gradient table shown in Figure \ref{fig:deladmap} we find that $\delad$ has the flattest behavior around $T=500$ K.
+In this region, the gas behaves like an ideal gas with constant polytropic index $\delad \approx 0.3$, with the shift from diatomic gas caused by the helium in the mixture.
+We generate three sets of atmosphere profiles.
+The first one corresponds to an ideal gas of constant adiabatic index $\delad=0.3$.
+The second one is described by the real EOS for temperatures larger than 500 K and by an ideal gas with $\delad=0.3$ for $T<500$ K.
+Finally, the third profile consists of an ideal gas polytrope with $\delad=0.3$ for $T>500$ K and a real gas in the low temperature regime.
+We compare the first two profiles to show the dissociation effects, and the first and third profile to show the effects of ortho- and parahydrogen.
+
+%We now explore the effects of
+
+%\subsection{Dissociation and Spin Isomers Effects}
+%\label{deladeffect}
+
+%In order to investigate the separate effects of hydrogen dissociation at high temperatures and spin isomers at low temperatures on the atmosphere evolution, we generate
+
+In order to explain the quantum rotational effects at low temperatures, we make the following thought experiment: we imagine that realistic EOS effects only affect the upper atmosphere where temperatures are low, and that the gas is ideal and polytropic deep in the envelope.
+Conversely, we study the effects of dissociation at high temperatures by assuming that realistic EOS effects only matter at the bottom of the atmosphere where temperatures are high, and that the gas is polytropic in the outer regions.
+Figure \ref{fig:tplotall} shows the time evolution of atmospheres forming at 10 AU around cores of mass $M\co=10 M_{\oplus}$ and described by various equations of state, as follows:
+
+\begin{itemize}
+\item Dashed curve (1) --- ideal gas polytrope with $\delad=0.3$.
+\item Dotted curve (2) --- ideal gas polytrope with $\delad=0.3$ for $T>500$ K and realistic EOS for $T<500$ K.
+\item Dash-dotted curve (3) --- ideal gas polytrope with $\delad=0.3$ for $T<500$ and realistic EOS for $T>500$ K.
+\item Solid curve (4) --- fully realistic EOS.
+\end{itemize}
+We choose $T=500$ K as the cutoff temperature because the hydrogen-helium mixture behaves like an ideal gas, with $\delad=0.3$, in this temperature regime (see Figure \ref{fig:deladmap}).
+As such, the difference in cooling times between curves (1) and (2) shows the quantum rotational effects on the atmosphere evolution, while the difference between curves (1) and (3) accounts for hydrogen dissociation.
+We see that both dissociation and rotational states have a comparable effect on the atmosphere growth and result in slower cooling.
+The cooling time is dependent on both the total energy released due to the contraction of the envelope and the luminosity of the atmosphere.
+We explore the relative influence of these two factors separately.
 
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/ortho_para_energy.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/tplot.pdf}
 %%\vspace{-0.5in}
-\caption{Internal energy and heat capacity per unit mass for the hydrogen spins isomers and the equilibrium mixture as a function of temperature. See \citet{farkas35}, Figure 1, for a similar plot.}
-\label{fig:ucvr}
+\caption{Elapsed time as a function of total mass (core + atmosphere) for a variety of EOS combinations (see text), for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=10 M_{\oplus}$.
+Both hydrogen dissociation at high temperatures deep in the atmosphere and quantum rotational effects at low temperatures in the outer envelope result in slower cooling when compared to an ideal gas polytrope.}
+\label{fig:tplotall}
 \end{figure}
 
-%\vspace{0.1in}
-
-\section{Equation of State Effects on Atmosphere Evolution}
-\label{EOSeffects}
-
-Variations in the adiabatic gradient $\delad$ have two competing effects on the atmosphere evolution: they yield a lower envelope luminosity when compared to an ideal gas, and a larger amount of energy that needs to be radiated per unit mass, which slows down accretion.
-Together, they result in changes in the growth time of the atmosphere, and therefore in the crossover time and critical core mass (both defined in section \ref{intro}).
-We explore the effect of partial dissociation and hydrogen spin isomers (see section \ref{deladtable}) on the atmosphere luminosity and cooling time evolution.
-Due to the equation of state described in section \ref{deladtable}, a realistic atmosphere will not have a constant adiabatic gradient $\delad$, as was assumed in PY13.
-We thus first investigate the differences in atmosphere profiles and evolution for ideal gas polytropes with different adiabatic gradients in section \ref{deladpoly}.
-We then show how the variable adiabatic gradient affects the time evolution of the atmosphere in section \ref{deladeffect}.
-
-\subsection{Ideal Gas Polytropes with Different Adiabatic Gradient}
-\label{deladpoly}
-
-In this section we investigate the differences in luminosity and $dE/dM$, and the resulting time evolution, between ideal gas polytropes with different adiabatic gradients: $\delad=2/7$ (diatomic gas) and $\delad=2/5$ (monatomic gas).
+\begin{figure*}[tb]
+\centering
+\includegraphics[width=\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/all_plot.pdf}
+%%\vspace{-0.5in}
+\caption{We explore atmosphere growth around a core of $M\co=10 M_{\oplus}$ forming at 10 AU in our fiducial disk, for the EOS choices in Figure \ref{fig:tplotall}.
+Upper-left panel:  Instantaneous temperature profile for a total mass (core + atmosphere) of $12 M_{\oplus}$ as a function of radius.
+The location of the RCB is marked.
+The quantum rotational effects at low temperatures in the outer region of the atmosphere dictate the location of the RCB.
+Upper-right panel:  Luminosity evolution as a function of total mass (core + atmosphere).
+The quantum rotational effects at low temperatures near the top of the atmosphere result in a lower luminosity for the realistic EOS (see text).
+Lower-left panel: Instantaneous energy profiles as a function on radius for a total mass (core + atmosphere) of $12 M_{\oplus}$.
+Hydrogen dissociation deep in the atmosphere causes the bulk of the energy to be concentrated at the bottom of the atmosphere.
+This increases the amount of energy per unit mass that needs to be radiated away, i.e.
+$|dE/dM|$, resulting in a longer crossover time for the realistic EOS when compared to the polytrope.
+Lower-right panel:  $|dE/dM|$ evolution as a function of total mass (core + atmosphere).
+More energy per unit mass has to be radiated away due to hydrogen dissociation when compared to an ideal gas polytrope, which results in slower growth.}
+\label{fig:all_plot}
+\end{figure*}
+
+The upper-right panel of Figure \ref{fig:all_plot} shows the luminosity evolution with mass for atmospheres (1)-(4), for an example core with $M\co=10 M_{\oplus}$ forming at 10 AU.
+We see that the lower luminosity of the fully realistic EOS (4) when compared to the polytrope (1) is due to the realistic EOS effects in the outer atmosphere (2).
+This behavior is directly correlated to the depth of the radiative region of the envelope, which is shown in the upper-left panel of Figure \ref{fig:all_plot} for a total planet mass of $12 M_{\oplus}$.
+We see that the realistic EOS for low temperatures (2) results in a deeper location of the RCB when compared to the ideal gas polytrope (1).
+This deeper RCB translates into a larger number of steps that photons need to diffuse to escape from the RCB, and therefore a lower luminosity (cf.
+equation \ref{eq:delrad}).
+In fact, a simple estimate of the luminosity emerging at the RCB can be found as follows.
+
+The equation of radiative diffusion (\ref{eq:structc}) can be approximated in terms of the radiative flux $\mathcal{F}$ as
+
+\begin{equation}
+\label{eq:sigmatau4}
+\mathcal{F} \sim \frac{\sigma T\cb^4}{\tau},
+\end{equation}
+where $\tau$ is the optical depth of the upper radiative layer. Furthermore, we find that $\tau$ is dominated by the optical depth of the first scale height, i.e.
+
+\begin{equation}
+\label{eq:taucb}
+\tau \sim \tau\cb=\kappa\cb \rho\cb H\cb,
+\end{equation}
+where 
+\begin{equation}
+\label{eq:Hcb}
+H\cb=\frac{k_b T\cb}{\mu m_H} \frac{R\cb^2}{G M\cb},
+\end{equation}
+with $\mu$ the mean molecular weight of the gas, $m_H$ the proton mass, and all other quantities evaluated at the bottom of the radiative layer.
+By substituting equations (\ref{eq:Hcb}) and (\ref{eq:taucb}) into equation (\ref{eq:sigmatau4}), and noting that $\mathcal{F}=L/(4 \pi (R\cb+H\cb)^2) \approx L / (4 \pi R\cb^2)$, we recover equation (\ref{eq:delrad}) applied at the boundary between the uppermost convective region of the atmosphere and the outer radiative layer (where $\delrad=\delad$).
+
+%We found that the RCB for the fully realistic EOS (4) is roughly at the same location as the RCB for the realistic EOS at low temperatures (2) for low atmosphere masses, and it moves further out relatively as the atmosphere mass increases.
+This explains our choice of $M_{\rm tot}=12 M_{\oplus}$ as a representative case to illustrate the effect of spin isomers.
+Envelope growth is slowest at this relative atmosphere-to-core mass (see PY13)...
+
+Dissociation deep in the planet's atmosphere increases the amount of energy per unit mass that needs to be radiated by the envelope, which further slows down growth.
+Figure \ref{fig:all_plot}, lower-left plot, shows the total energy (internal and gravitational) profile as a function of the radial coordinate, for the same example planet.
+The bulk of the energy is concentrated deep in the atmosphere for the realistic EOS at high temperatures (3) and the fully realistic EOS curve (4), which shows that hydrogen dissociation in the inner atmosphere dictates the energy behavior of the envelope.
+In contrast, energy is concentrated towards the outer boundary for the $\delad=0.3$ polytrope (1).
+Qualitatively, this can be explained through a simple analytic argument: the density profile in an adiabatic, non-self gravitating atmosphere composed of an ideal gas scales as $\rho(r) \sim r^{1/\delad-1}$ (see also Paper I).
+Since the energy per unit mass scales as $e(r) \sim \rho(r) r^2 \sim r^{1/\delad+1}$, it is easy to see that adiabats with lower adiabatic gradient have their energy more tightly packed towards the interior of the envelope, which is the case during dissociation when $\delad$ decreases significantly (see section \S\ref{deladtable} and Figure \ref{fig:deladmap}).
+It takes more energy to bring in gas deep in the atmosphere for an envelope that has the bulk of its energy concentrated towards the bottom,  which increases the amount of energy per unit mass that needs to be radiated, i.e.
+$|dE/dM|$, and therefore the cooling time, for the realistic EOS at high temperatures (3) when compared to the ideal gas polytrope (1).
+This increase in $|dE/dM|$ is shown in the bottom-right panel of Figure \ref{fig:all_plot}.
+
+We have seen that the quantum rotational effect at the outer boundary dictates the location of the radiative zone, and therefore the luminosity behavior, while dissociation deep in the atmosphere dictates the total energy behavior.
+Overall, both effects result in a longer time for the atmosphere to evolve.
+
+%we find that only polytropes with $\gamma<3/2$, i.e.
+$\delad<1/3$, have the total energy concentrated towards the core.
+This is satisfied by $\delad=2/7$ but not by $\delad=2/5$, in agreement with the results in the bottom panel of Figure \ref{fig:ETrplotpoly}.
+%
+%Similarly to section \ref{deladpoly}, we use instantaneous atmosphere profiles to explain the differences.
+The left panel of Figure \ref{fig:TLrplot} shows the instantaneous temperature profile and the location of the radiative-convective boundary for a total fixed mass (core + atmosphere) $M_{\rm{tot}}=11.8 M_{\oplus}$.
+The realistic equation of state for low temperatures is characterized by a lower adiabatic index in the outer regions, due to the spin effects, and is therefore dominant in the radiative zone.
+As a result, it generates a deeper radiative zone with a lower luminosity, which explains the results in the left panel of Figure \ref{fig:TLrplot}.
+Moreover, since the cooling time is inversely proportional to the luminosity, the spin effect will result in a longer cooling time.
+%
+%The energy behavior is shown in Figure \ref{fig:Erplot}.
+The realistic EOS for high temperatures has a low adiabatic index deep in the atmosphere, due to hydrogen dissociation, and thus the bulk of its energy concentrated at the bottom of the atmosphere, for the reasons described in section \ref{deladpoly}.
+It takes more energy to add mass more mass deep in the atmosphere, and $|dE/dM|$ is larger as a result.
+%%
+%
+%We have seen that the spin effect at the outer boundary dictates the location of the radiative zone, and therefore the luminosity behavior, while dissociation deep in the atmosphere dictates the energy behavior.
+Overall, both effects result in a longer time for the atmosphere to evolve.
+%
+%\subsection{Ideal Gas Polytropes with Different Adiabatic Gradient}
+%\label{deladpoly}
+%
+%In this section we investigate the differences in luminosity and $dE/dM$, and the resulting time evolution, between ideal gas polytropes with different adiabatic gradients: $\delad=2/7$ (diatomic gas) and $\delad=2/5$ (monatomic gas).
 We assume both gases have the same mean molecular weight.
 %We use these results to explain the separate effects of dissociation and spin on the time evolution of the atmosphere.
-
-We generate atmosphere profiles for the two different adiabatic indices at $a=10$ AU and for a core mass $M_{\rm c}=10 M_{\oplus}$, and estimate the luminosity and cooling time evolution as described in section \ref{sec2}.
+%
+%We generate atmosphere profiles for the two different adiabatic indices at $a=10$ AU and for a core mass $M_{\rm c}=10 M_{\oplus}$, and estimate the luminosity and cooling time evolution as described in section \ref{sec2}.
 The results are shown in Figure \ref{fig:Ltplotpoly}.
@@
 We use instantaneous atmosphere profiles to explain these effects.
-
-%We first discuss the effect of the variable adiabatic gradient on luminosity.
-
-Figure \ref{fig:ETrplotpoly}, top panel, shows the temperature profile for the two polytropes at a fixed total mass $M_{\rm{tot}}=15 M_{\oplus}$.
+%
+%%We first discuss the effect of the variable adiabatic gradient on luminosity. 
+%
+%Figure \ref{fig:ETrplotpoly}, top panel, shows the temperature profile for the two polytropes at a fixed total mass $M_{\rm{tot}}=15 M_{\oplus}$.
 The polytrope with a larger adiabatic gradient, $\delad=2/5$, has a more shallow convective zone, and hence a deeper radiative region, since a larger temperature gradient delays the onset of convection.
 A deeper radiative layer increases the number of steps that photons need to diffuse, resulting in a lower luminosity as seen in the top panel of Figure \ref{fig:Ltplotpoly}.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/Ltplot_poly.pdf}
-%%\vspace{-0.5in}
-\caption{Luminosity and time evolution as a function of total mass (core + atmosphere) for polytropes with different adiabatic indices, for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=10 M_{\oplus}$.
+%
+%\begin{figure}[h]
+%\centering
+%\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/Ltplot_poly.pdf}
+%%%\vspace{-0.5in}
+%\caption{Luminosity and time evolution as a function of total mass (core + atmosphere) for polytropes with different adiabatic indices, for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=10 M_{\oplus}$.
 A larger adiabatic index results in both a lower luminosity and a shorter cooling time.}
-\label{fig:Ltplotpoly}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/TErplot_poly.pdf}
-%%\vspace{-0.5in}
-\caption{Instantaneous temperature and total energy profiles as a function of radius for polytropes with different adiabatic indices, for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=10 M_{\oplus}$.
+%\label{fig:Ltplotpoly}
+%\end{figure}
+%
+%\begin{figure}[h]
+%\centering
+%\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/TErplot_poly.pdf}
+%%%\vspace{-0.5in}
+%\caption{Instantaneous temperature and total energy profiles as a function of radius for polytropes with different adiabatic indices, for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=10 M_{\oplus}$.
 The total mass (core + atmosphere) is $15 M_{\oplus}$.
@@
 A lower adiabatic gradient results in a more shallow radiative region (upper panel), and in the total energy being concentrated at the bottom of the atmosphere (lower panel).}
-\label{fig:ETrplotpoly}
-\end{figure}
-
-%In spite of the low luminosity of the atmosphere with $\delad=2/5$, the envelope still grows faster than in the case of a diatomic gas, as shown in the bottom panel of Figure \ref{fig:Ltplotpoly}.
+%\label{fig:ETrplotpoly}
+%\end{figure}
+%
+%%In spite of the low luminosity of the atmosphere with $\delad=2/5$, the envelope still grows faster than in the case of a diatomic gas, as shown in the bottom panel of Figure \ref{fig:Ltplotpoly}.
 This is due to fact that the amount of energy per unit mass that needs to be radiated away, i.e.
@@
 We have shown in PY13 that polytropes with $\delad=2/7$ have most of the energy concentrated at the bottom of the atmosphere, while polytropes with $\delad=2/5$ have the bulk of the energy towards the outer boundary.
-
-The bottom panel of Figure \ref{fig:ETrplotpoly} shows an instantaneous energy profile for the same total mass $M_{\rm{tot}} = 15 M_{\oplus}$.
+%
+%
+%The bottom panel of Figure \ref{fig:ETrplotpoly} shows an instantaneous energy profile for the same total mass $M_{\rm{tot}} = 15 M_{\oplus}$.
 The bulk of the energy is concentrated deep in the atmosphere for the $\delad=2/7$ polytrope and towards the outer boundary for the $\delad=2/5$ polytrope.
@@
 It takes more energy to bring in gas deep in the atmosphere for an envelope that has the bulk of its energy concentrated towards the bottom,  which increases $|dE/dM|$ for the $\delad=2/7$ polytrope.
-
-We find that the energy effect prevails over the luminosity effect, resulting in a longer cooling time for the polytrope with the lower adiabatic gradient (i.e., $\delad=2/7$), as shown in the bottom panel of Figure \ref{fig:Ltplotpoly}.
+%
+%We find that the energy effect prevails over the luminosity effect, resulting in a longer cooling time for the polytrope with the lower adiabatic gradient (i.e., $\delad=2/7$), as shown in the bottom panel of Figure \ref{fig:Ltplotpoly}.
 
@@
 
-%\subsection{Effect of Adiabatic Gradient on Atmosphere Cooling Time}
-%\label{deladeffect}
-
-%In this section we discuss the effect of the variable adiabatic index described in subsection \ref{deladinterp} on the atmosphere cooling time.
-
-%We now explore the separate effects of hydrogen dissociation at high temperatures and spin isomers at low temperatures.
-From the adiabatic gradient table shown in Figure \ref{fig:deladmap} we find that $\delad$ has the flattest behavior around $T=500$ K.
-In this region, the gas behaves like an ideal gas with constant polytropic index $\delad \approx 0.3$, with the shift from diatomic gas caused by the helium in the mixture.
-We generate three sets of atmosphere profiles.
-The first one corresponds to an ideal gas of constant adiabatic index $\delad=0.3$.
-The second one is described by the real EOS for temperatures larger than 500 K and by an ideal gas with $\delad=0.3$ for $T<500$ K.
-Finally, the third profile consists of an ideal gas polytrope with $\delad=0.3$ for $T>500$ K and a real gas in the low temperature regime.
-We compare the first two profiles to show the dissociation effects, and the first and third profile to show the effects of ortho- and parahydrogen.
-
-%We now explore the effects of
-
-\subsection{Dissociation and Spin Isomers Effects}
-\label{deladeffect}
-
-In this section we use the results obtained in section \ref{deladpoly} to explain the effects on the atmosphere evolution of a realistic equation of state.
+%As compared to the polytrope, the real EOS therefore generates a deeper radiative zone with a lower luminosity, due to the lower adiabatic index in the outer regions, as well as an atmosphere with the bulk of its energy concentrated at the bottom, due to the low $\delad$ caused by dissociation.
+As shown above for the ideal gas polytropes, and remembering that $\Delta t \sim -\Delta E/L$, we see that both effects result in a longer time for the atmosphere to evolve.
+
+% for an atmosphere described by an ideal gas EOS with $\delad=0.3$ (dashed curve), an atmosphere composed of an ideal gas with $\delad=0.3$ for $T>500$ K and a realistic gas
+
+%In this section we use the results obtained in section \ref{deladpoly} to explain the effects on the atmosphere evolution of a realistic equation of state.
 We explore the effect of hydrogen dissociation at the high temperatures in the inner part of the atmosphere, and the effect of hydrogen spin isomers at the low temperatures at the top of the atmosphere separately.
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/tplot.pdf}
-%%\vspace{-0.5in}
-\caption{Cooling time evolution as a function of total mass (core + atmosphere) for a variety of EOS combinations, for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=10 M_{\oplus}$.
-The cooling time is larger for the realistic EOS both due to hydrogen dissociation and spin effects when compared to an ideal gas polytrope.}
-\label{fig:tplotall}
-\end{figure}
-
 %We now investigate the effect of the low adiabatic index caused by hydrogen dissociation, on the one hand, and spin isomers on the other hand, on the luminosity and cooling time evolution of the atmosphere, in light of the discussion in section \ref{deladpoly}.
@@
 
-The right panel of Figure \ref{fig:TLrplot} shows the luminosity evolution with mass for the three combinations of equations of state described above, as well as for the complete realistic gas EOS.
-Similarly to section \ref{deladpoly}, we use instantaneous atmosphere profiles to explain the differences.
-The left panel of Figure \ref{fig:TLrplot} shows the instantaneous temperature profile and the location of the radiative-convective boundary for a total fixed mass (core + atmosphere) $M_{\rm{tot}}=11.8 M_{\oplus}$.
-The realistic equation of state for low temperatures is characterized by a lower adiabatic index in the outer regions, due to the spin effects, and is therefore dominant in the radiative zone.
-As a result, it generates a deeper radiative zone with a lower luminosity, which explains the results in the left panel of Figure \ref{fig:TLrplot}.
-Moreover, since the cooling time is inversely proportional to the luminosity, the spin effect will result in a longer cooling time.
-
-The energy behavior is shown in Figure \ref{fig:Erplot}.
-The realistic EOS for high temperatures has a low adiabatic index deep in the atmosphere, due to hydrogen dissociation, and thus the bulk of its energy concentrated at the bottom of the atmosphere, for the reasons described in section \ref{deladpoly}.
-It takes more energy to add mass more mass deep in the atmosphere, and $|dE/dM|$ is larger as a result.
-%
-
-We have seen that the spin effect at the outer boundary dictates the location of the radiative zone, and therefore the luminosity behavior, while dissociation deep in the atmosphere dictates the energy behavior.
-Overall, both effects result in a longer time for the atmosphere to evolve.
-
-%As compared to the polytrope, the real EOS therefore generates a deeper radiative zone with a lower luminosity, due to the lower adiabatic index in the outer regions, as well as an atmosphere with the bulk of its energy concentrated at the bottom, due to the low $\delad$ caused by dissociation.
-As shown above for the ideal gas polytropes, and remembering that $\Delta t \sim -\Delta E/L$, we see that both effects result in a longer time for the atmosphere to evolve.
-
-\begin{figure*}[tb]
-\centering
-\includegraphics[width=\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/all_plot.pdf}
-%%\vspace{-0.5in}
-\caption{Left panel: Instantaneous temperature profile as a function of radius for a variety of EOS combinations, for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=10 M_{\oplus}$.
-The total mass (core + atmosphere) is $11.8 M_{\oplus}$.
-The location of the radiative-convective boundary is marked.
-The effect of hydrogen spin isomers at low temperatures in the outer region of the atmosphere sets the location of the radiative-convective boundary.
-Right panel: Luminosity evolution as a function of total mass (core + atmosphere) for a variety of EOS combinations, for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=5 M_{\oplus}$.
-The existence of the hydrogen spin isomers at low temperatures near the top of the atmosphere results in a lower luminosity.
-The two effects combined yield a lower luminosity for the realistic equation of state when compared to the polytrope.
-}
-\label{fig:TLrplot}
-\end{figure*}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/Er_plot.pdf}
-%%\vspace{-0.5in}
-\caption{Instantaneous energy profiles as a function on radius for a variety of EOS combinations, for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=10 M_{\oplus}$.
+%
+%
+%\begin{figure}[h]
+%\centering
+%\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/Er_plot.pdf}
+%%%\vspace{-0.5in}
+%\caption{Instantaneous energy profiles as a function on radius for a variety of EOS combinations, for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=10 M_{\oplus}$.
 The total mass (core + atmosphere) is $11.8 M_{\oplus}$.
@@
 $|dE/dM|$, resulting in a longer crossover time for the realistic EOS when compared to the polytrope.}
-\label{fig:Erplot}
-\end{figure}
+%\label{fig:Erplot}
+%\end{figure}
 
@@
 
-In this section we put together the results obtained in section \ref{EOSeffects} and determine the minimum core mass to initiate runaway gas accretion during the lifetime of the protoplanetary disk assuming that the nebular gas is described by a realistic equation of state as prescribed by the \citet{saumon95} EOS tables (see section \ref{EOSeffects}).
-As in PY13, we define this minimum core mass as the \textit{critical core mass}.
-Moreover, we define the time elapsed until runaway gas accretion is initiated when $M_{\rm{atm}} \sim M\co$ as the \textit{crossover time}.
-In this section we first explore the dependence of the crossover time on the core mass for a fixed semi-major axis.
-We then determine the critical core mass to form a giant planet before the dissipation of the gas in the protoplanetary disk, assuming that the nebular gas is described by a realistic hydrogen-helium mixture, and we compare this with the results from PY13 for an ideal diatomic gas.
+In this section we put together the results obtained in section \S\ref{EOSeffects} and determine the minimum core mass to initiate runaway gas accretion during the lifetime of the protoplanetary disk.
+As in Paper I, we define this minimum core mass as the \textit{critical core mass}, and the time elapsed until runaway gas accretion is initiated when $M_{\rm{atm}} \sim M\co$ as the \textit{crossover time}.
+We first explore the dependence of the crossover time on the core mass for a fixed semi-major axis.
+We then determine the critical core mass to form a giant planet from a gas composed of a realistic hydrogen-helium mixture, and we compare this with the results from Paper I for an ideal diatomic gas.
 Finally, we determine the critical core mass under more realistic opacity assumptions.
@@
 
-Figure \ref{fig:tvsMplot} displays the time evolution and the crossover time for core masses between 7 and 16 $M_{\oplus}$ at $a=10$ AU in our fiducial disk.
-The crossover time is shorter for higher mass cores, consistent with the results of PY13.
+Figure \ref{fig:tvsMplot} displays the time evolution and the crossover time for atmospheres forming around cores with masses between 10 $M_{\oplus}$ and 20 $M_{\oplus}$ at $a=10$ AU in our fiducial disk.
+The crossover time is shorter for higher mass cores, consistent with the results of Paper I.
 
@@
 %%\vspace{-0.5in}
-\caption{Time to grow an atmosphere of mass $M_{\rm{atm}}$ for cores with fixed masses between $10 M_{\oplus}$ and $20 M_{\oplus}$ at $a=10$ AU in our fiducial disk.
+\caption{Elapsed time as a function of atmosphere mass, for cores with fixed masses between $10 M_{\oplus}$ and $20 M_{\oplus}$ at $a=10$ AU in our fiducial disk.
 The circles mark the crossover time where $M_{\rm{atm}} \sim M_{\rm c}$.
@@
 
-Figure \ref{fig:Mvsaplot} shows the critical core mass for a massive atmosphere to form during a typical lifetime of a protoplanetary disk $t=3$ Myrs, for a gas described by a realistic equation of state.
-For comparison, we also plot the results of PY13 for an ideal diatomic gas.
-The use of a realistic equation of state increases the critical core mass by more than a factor of 2.
+Figure \ref{fig:Mvsaplot} shows the critical core mass for a massive atmosphere to form during a typical lifetime of a protoplanetary disk $t=3$ Myrs (e.g., \citealt{jay99}), for a gas described by a realistic EOS and an ISM dust opacity given by equation \ref{eq:opacitylaw}.
+The results of Paper I for an ideal diatomic gas are plotted for comparison.
+The inclusion of realistic EOS effects results in a critical core mass more than twice as large than under the ideal gas assumption.
 As such, non-ideal effects substantially affect the core mass needed to form a giant planet  before the dissipation of the protoplanetary disk.
@@
 %%\vspace{-0.5in}
-\caption{The minimum core mass for an atmosphere to initiate runaway gas accretion within the lifetime of a typical protoplanetary disk $t \sim 3$ Myrs as a function of semi-major axis, for a realistic hydrogen-helium mixture.
-The results of PY13 for an ideal diatomic gas are plotted for comparison.
+\caption{The minimum core mass for an atmosphere to initiate runaway gas accretion within the lifetime of a typical protoplanetary disk $t \sim 3$ Myrs as a function of semi-major axis, for a realistic hydrogen-helium mixture and a standard ISM power-law opacity.
+The results of Paper I for an ideal diatomic gas are plotted for comparison.
 The realistic equation of state yields core masses larger by more than a factor of 2 when compared to the polytrope.}
@@
 
-The above results were obtained under the assumption that the dust opacity in the radiative region of the atmosphere is given by the standard ISM opacity (see also section \ref{struct}).
+The results above were obtained under the assumption that the dust opacity in the radiative region of the atmosphere is given by the standard ISM opacity (see also section \S\ref{struct}).
 However, our scenario of low planetesimal accretion is likely to favor lower dust opacities, due to grain growth and dust settling.
 Grain growth, in particular, lowers the absolute value of the opacity and changes the particle size distribution when compared to the standard ISM size distribution (e.g., \citealt{pollack85}).
-
-Although grain growth has been observed in protoplanetary disks (e.g., \citealt{beckwith90}, \citealt{beckwith91}, \citealt{perez12}), the size distribution of dust particles has not been observationally constrained.
+\textbf{Didn't understand comment...}
+
+%Observations of dust in protoplanetary disks (e.g., \citealt{beckwith90}, \citealt{beckwith91}, \citealt{perez12}) have shown evidence for grain growth and that the particle size distribution is not interstellar.
+
+Although grain growth and evidence for non-ISM size distribution have been observed in protoplanetary disks (e.g., \citealt{beckwith90}, \citealt{beckwith91}, \citealt{perez12}), the size distribution of dust particles has not been tightly constrained.
 Typically, the grain size distribution is assumed to be a power law:
@@
 where $a$ is the particle radius and $p=3.5$ (corresponding to a ``normal'' collisional cascade) or $p=2.5$ (an approximation for coagulation).
-In this work we use the  \citet{dalessio01} frequency-dependent opacity tables to obtain the temperature-dependent Rosseland mean opacity $\kappa$,  assuming a maximum particle size of 1 cm and a grain size distribution given by equation (\ref{eq:graindistr}) with $p=3.5$.
+In this work we use the  \citet{dalessio01} frequency-dependent opacity tables to obtain the temperature-dependent Rosseland mean opacity $\kappa$,  assuming a maximum particle size of 1 cm and a grain size distribution given by equation (\ref{eq:graindistr}) with $p=3.5$ as a fiducial case.
+Other choices for the power-law coefficient $p$ are discussed later in this section.
 
@@
 The sharp drop in opacity ($\kappa \sim T^{-24}$) due to dust sublimation lowers the radiative temperature gradient significantly (see equation \ref{eq:delrad}), and may therefore generate radiative layers within the inner region of the atmosphere (see \App{radwindow}, Figure \ref{fig:delvsr}).
-This can potentially pose challenges for our model, since the additional luminosity generated in these radiative windows may be large enough to make our assumption of constant luminosity throughout the radiative layers invalid.
-However, the inner radiative windows are either very thin (Figure \ref{fig:delvsr}, middle panel), or the temperature gradient is flat enough to result in a roughly constant luminosity throughout the region (Figure \ref{fig:delvsr}, bottom panel; also see equation \ref{eq:structc}).
+This could in principle pose challenges for our model, since the additional luminosity generated in these radiative windows may be large enough to make our assumption of constant luminosity throughout the radiative layers invalid.
+In practice, however, the inner radiative windows are either very thin (Figure \ref{fig:delvsr}, middle panel), or the temperature gradient is flat enough to result in a roughly constant luminosity throughout the region (Figure \ref{fig:delvsr}, bottom panel; also see equation \ref{eq:structc}).
 This results in a negligible extra luminosity generated in the radiative windows.
-%moreover...
-(\textbf{add part about sigma T4/tau}).
-
-Despite the existence of one or more radiative windows in the planet atmosphere, the luminosity that emerges at the outer boundary is set by the top radiative window, as is the case for the standard inner convective-outer radiative envelopes.
-The equation of radiative diffusion (\ref{eq:structc}) can be approximated in terms of the radiative flux $\mathcal{F}$ as
-
-\begin{equation}
-\label{eq:sigmatau4}
-\mathcal{F} \sim \frac{\sigma T\cb^4}{\tau},
-\end{equation}
-where $\tau$ is the optical depth of the upper radiative layer. Furthermore, we find that $\tau$ is dominated by the optical depth of the first scale height, i.e.
-
-\begin{equation}
-\label{eq:taucb}
-\tau \sim \tau\cb=\kappa\cb \rho\cb H\cb,
-\end{equation}
-where 
-\begin{equation}
-\label{eq:Hcb}
-H\cb=\frac{k_b T\cb}{\mu m_H} \frac{R\cb^2}{G M\cb},
-\end{equation}
-with $k_b$ the Boltzmann constant, $\mu$ the mean molecular weight of the gas, $m_H$ the proton mass, and all other quantities evaluated at the bottom of the radiative layer.
-By substituting equations (\ref{eq:Hcb}) and (\ref{eq:taucb}) into equation (\ref{eq:sigmatau4}), and noting that $\mathcal{F}=L/(4 \pi (R\cb+H\cb)^2) \approx L / (4 \pi R\cb^2)$, we recover equation (\ref{eq:delrad}) applied at the boundary between the uppermost convective region of the atmosphere and the outer radiative layer (where $\delrad=\delad$).
-
-Due to the variable number and position within the planet atmosphere of radiative windows, and therefore of radiative-convective boundaries, we cannot calculate the time evolution of different atmospheres if we evaluate our cooling equation (\ref{eq:coolingglobal}) at the RCB, as we do in our standard model.
-We choose to evaluate the cooling time at the Bondi radius instead (since our cooling model applies at any radius $R$, see section \ref{struct}).
-We note that our choice of $R$ does not change the estimate of the atmosphere evolution time, to order of magnitude, since the additional luminosity generated in all radiative regions is negligible (see PY13).
+While the above is true for our fiducial opacity law, the radiative windows may become non-negligible for other opacity choices and sufficiently low core masses, as we show later in this section.
+We note that despite the existence of one or more radiative windows in the planet atmosphere, the luminosity that emerges at the outer boundary is set by the top radiative window, as is the case for the standard inner convective-outer radiative envelopes.
+
+Due to the variable number and position of radiative windows, and therefore radiative-convective boundaries, within the planet atmosphere, we cannot consistently calculate the time evolution of different atmospheres if we evaluate our cooling equation (\ref{eq:coolingglobal}) at the RCB, as we do in our standard model.
+We choose to evaluate the cooling time at the Bondi radius instead (since our cooling model applies at any radius $R$, see section \S\ref{struct}).
+We note that our choice of $R$ does not change the estimate of the atmosphere evolution time, to order of magnitude, since the additional luminosity generated in all radiative regions is negligible for our opacity choice (also see Paper I).
 
@@
 %%\vspace{-0.5in}
-\caption{Critical core mass as a function of semi-major axis for grain growth opacity.}
+\caption{Critical core mass as a function of semi-major axis for radiative opacities that account for grain growth.
+The critical core mass is lower than in our fiducial case where dust grains have a standard ISM distribution.}
 \label{fig:Mcritvsagg}
@@
 the boundary conditions (temperature and pressure).
-We have shown in PY13 that the critical core mass is highly dependent on disk temperature rather than pressure.
-The temperature dependence, however, is mainly due to opacity: for the simplified analytic model developed in PY13 we found that the crossover time $t_{\rm co} \sim T^{\beta+1/2}$, with $\beta$ the power-law exponent in equation (\ref{eq:opacitylaw}).
+We have shown in Paper I that the critical core mass is highly dependent on disk temperature rather than pressure.
+The temperature dependence, however, is mainly due to opacity: for the simplified analytic model developed in Paper I we found that the crossover time $t_{\rm co} \sim T^{\beta+1/2}$, with $\beta$ the power-law exponent in equation (\ref{eq:opacitylaw}).
 Opacity is less sensitive to temperature variations for larger grains and has an almost flat profile (see Figure \ref{fig:opacity}), which results in $\beta \ll 1$ and a much weaker temperature (and therefore semi-major axis) dependence of the critical core mass, as seen in Figure \ref{fig:Mcritvsagg}.
@@
 %%\vspace{-0.5in}
-\caption{Crossover time for different grain size distribution.}
+\caption{Crossover time for different grain size distribution, for a planet forming around a core $M\co=4 M_{\oplus}$.
+The standard ISM result is plotted for comparison.
+The crossover time is more than one order of magnitude lower when coagulation is accounted for, i.e.
+$p=2.5$.}
 \label{fig:p25p35}
@@
 
-In this study we have considered atmospheres for which planetesimal accretion is negligible and Kelvin-Helmholtz contraction dominates the luminosity evolution of the atmosphere.
-This is different from standard analytic calculations, in which the atmosphere is heated by planetesimal accretion.
-In this section we compare our results for the critical core mass to analogous results from the analytic calculations.
-We discuss the core accretion rates that are necessary for our regime to be valid in section \ref{raf1}.
-We then compare our results with planetesimal accretion results under similar assumptions in section \ref{raf2}.
+In this study we have considered protoplanets with fully formed cores for which planetesimal accretion is negligible and KH contraction dominates the luminosity evolution of the atmosphere.
+This is different from calculations that assume high planetesimal accretion rates and find that the atmosphere is in steady state and solely heated due to accretion of solids.
+In this section we compare our results for the critical core mass to analogous results from steady-state analytic calculations.
+We discuss the core accretion rates that are necessary for our regime to be valid in section \S\ref{raf1}.
+In section \S\ref{raf3}, we estimate core growth during atmosphere evolution at the maximum rate for which the KH regime is valid, and show it is negligible.
+Finally, we compare our results with those assuming fast planetesimal accretion in section \S\ref{raf2}.
 
@@
 
-We estimate the planetesimal accretion rate consistent with our assumptions that $L_{\mathrm{acc}} \ll L_{\rm{KH}}$. Here $L_{\rm{acc}}$ is the accretion luminosity given by
+Kelvin-Helmholtz contraction dominates an atmosphere's luminosity if  $L_{\mathrm{acc}} < L_{\rm{KH}}$, where $L_{\rm{acc}}$ is the accretion luminosity,
 
@@
 \end{equation}
-
-\noindent where $\dot{M_{\rm{c}}}$ is the planetesimal accretion rate, and $L_{\rm{KH}}$ is the luminosity of the atmosphere due to gas contraction obtained from our static model described in section \ref{sec2}.
-At the limit, $L_{\rm{acc}}=L_{\rm{KH}}$.
-For a given atmosphere model we can therefore estimate the maximum planetesimal accretion rate during the contraction of the envelope in order for the atmosphere to be dominated by gas contraction.
-We choose as a fiducial case an atmosphere forming at 60 AU and with a core mass of $5 M_{\oplus}$.
-Since analytic studies assume an ideal gas EOS, our fiducial atmosphere is described by an ideal gas polytrope with constant adiabatic gradient $\delad=2/7$ and mean molecular weight $\mu=2.35$ (see also PY13).
-For this choice of parameters, the atmosphere crossover time is $t _{\rm{co}}\sim$ 2.1 Myrs, which is within the typical life time of a protoplanetary disk.
+and $K_{\rm KH}$ is given by equation (\ref{eq:coolingglobal}) with $L\co=\Gamma=0$ (see section \ref{struct}). This condition is satisfied as long as the planetesimal accretion rate
+
+\begin{equation}
+\label{eq:McdotKH}
+\dot{M\co}<\dot{M}_{\rm c, KH} \equiv \frac{L_{\rm KH} R\co}{G M\co}.
+\end{equation}
+
+To calculate $\dot{M}_{\rm c, KH}$, we choose as a fiducial case an atmosphere forming at 10 AU and with a core mass of $10 M_{\oplus}$.
+Since analytic studies assume an ideal gas EOS, for ease of comparison we choose an ideal gas polytrope with constant adiabatic gradient $\delad=2/7$ and mean molecular weight $\mu=2.35$ (see also Paper I).
+For this choice of parameters, the atmosphere crossover time is $t _{\rm{co}}\sim$ 1.2 Myrs, which is within the typical lifetime of a protoplanetary disk.
 The results are presented in Figure \ref{fig:accrates}.
+The atmosphere growth rate $\dot{M}_{\rm{atm}}$ is also plotted for comparison.
+We see that the core accretion rate has to be $\sim2-3$ orders of magnitude lower than the atmosphere accretion rate for our assumptions to be valid.
+If the core had accreted planetesimals at the $\dot{M}_{\rm c, KH}$ rate since it started forming, it couldn't have grown large enough to attract an atmosphere within typical disk lifetime.
+This is consistent with the requirement of our model that the planetesimal accretion rate is initially large during core growth, then significantly reduces as the gaseous envelope accumulates.
+This is a plausible scenario, as the core may have formed in the inner part of the disk and was later scattered outwards \citep{ida13}, or the planet's feeding zone could have been depleted of solids due to a giant neighbor.
+We further estimate two reference accretion rates.
+The first one is the core accretion rate $\dot{M}_{\rm c, acc}$ needed to grow the core to $M\co=10 M_{\oplus}$ on the same timescale as our model atmosphere, $\tau=1.2$ Myrs:
+
+\begin{equation}
+\label{eq:Mcdot}
+\dot{M}_{\rm{c,acc}}(M_{\rm{c}}) \equiv \frac{M_{\rm{c}}}{\tau}.
+\end{equation}
+The second reference planetesimal accretion rate is $\dot{M}_{\rm c, Hill}$, a typically assumed planetesimal accretion rate for which the random velocities of the planetesimals are of the order of the Hill velocity around the protoplanetary core (for a review, see \citealt{goldreich04}).
+This is the accretion rate at the boundary between the dispersion dominated and shear dominated regimes.
+Following \citet{rafikov06} (equation A1),
+
+\begin{equation}
+\label{eq:MdotHill}
+\dot{M}_{\rm{c,Hill}}=\Omega \Sigma_{\rm p} R\co R_{\rm H},
+\end{equation}
+where $\Sigma_{\rm p}$ is the surface density of solids, assumed to satisfy $\Sigma\di \approx 100 \Sigma_{\rm p}$ for a dust-to-gas ratio of 0.01.
 
@@
 %\vspace{-0.5in}
-\caption{Various relevant accretion rates in the case of a planet forming at 60 AU and with a core mass $M\co=5 M_{\oplus}$.
-The $\dot{M}_{\rm{atm}}$ curve represents the growth rate of the atmosphere as estimated by our model, and $\dot{M}_{\rm{c,KH}}$ is the maximum planetesimal accretion rate during the gas contraction phase in order for our regime to be valid.
-For comparison, we plot the core accretion rate $\dot{M}_{\rm{c,acc}}$ necessary to grow the core on the same time scale as the atmosphere $\tau \sim 2.1$ Myrs, and a typically assumed planetesimal accretion rate $\dot{M}_{\rm{c,Hill}}$ where the random velocity of the planetesimals is given by the Hill velocity due to the core.}
+\caption{Various accretion rates in the case of a planet forming at 10 AU and with a core mass $M\co=10 M_{\oplus}$, for a polyropic EOS and ISM power-law opacity.
+For this choice of parameters, the crossover time (defined in section \S\ref{critical}) is $t_{\rm co} \sim 1.2$ Myrs.
+The $\dot{M}_{\rm{atm}}$ (solid blue) curve represents the growth rate of the atmosphere as estimated by our model, and $\dot{M}_{\rm{c,KH}}$ (dotted red) is the maximum planetesimal accretion rate during the gas contraction phase in order for our regime to be valid, i.e.
+$L_{\mathrm{acc}} < L_{\rm{KH}}$ (see text).
+For comparison, we plot the core accretion rate $\dot{M}_{\rm{c,acc}}$ (dashed green) necessary to grow the core on a timescale $\tau=t_{\rm co}$, and a typically assumed planetesimal accretion rate $\dot{M}_{\rm{c,Hill}}$ (dash-dotted light blue) for which the random velocity of the planetesimals is given by the Hill velocity due to the core (see text).
+We note that $\dot{M}_{\rm c, KH}$ and $\dot{M}_{\rm c, Hill}$ are both lower than $\dot{M}_{\rm c, acc}$, consistent with our assumptions that planetesimal accretion must have slowed down after core growth.}
 \label{fig:accrates}
@@
 
-We label the resulting minimum core accretion rate as $\dot{M}_{\rm{c,KH}}$, with  $\dot{M}_{\rm{c,KH}}=\frac{L_{\rm{KH}} R\co}{G M\co}$ from equation (\ref{eq:Lacc}).
-The atmosphere growth rate $\dot{M}_{\rm{atm}}$ is also plotted for comparison.
-We see that the core accretion rate has to be $\sim2-3$ orders of magnitude lower than the atmosphere accretion rate for our assumptions to be valid.
-If the core had accreted planetesimals at this constant rate since it started forming, then the formation of a core massive enough to attract an atmosphere would not have been possible within a typical disk lifetime, implying that planetesimal accretion must have been larger initially, then slowed down.
-Possible explanations for that include the core having formed in the inner part of the disk and later migrated outwards, or the core having been depleted of planetesimals due to a giant neighbor.
-We further estimate the core accretion rate needed for the core to form on the same timescale as our model atmosphere, $\tau=2.1$ Myrs:
-
-\begin{equation}
-\label{eq:Mcdot}
-\dot{M}_{\rm{c,acc}}(M_{\rm{c}}) \equiv \frac{M_{\rm{c}}}{\tau}
-\end{equation}
-For reference, we also plot a typically assumed planetesimal accretion rate, for which the random velocities of the planetesimals are of the order of the Hill velocity around the protoplanetary core (e.g., \citealt{goldreich04}).
-We denote this latter rate as $\dot{M}_{\rm{c,Hill}}$.
-This is the accretion rate at the boundary between the dispersion dominated and shear dominated regimes.
-We estimate $\dot{M}_{\rm{c,Hill}}$ following \citet{rafikov06} (equation A1):
-
-\begin{equation}
-\label{eq:MdotHill}
-\dot{M}_{\rm{c,Hill}}=\Omega \Sigma_{\rm p} R\co R_{\rm H},
-\end{equation}
-where $\Sigma_{\rm p}$ is the surface density of solids, assumed to satisfy $\Sigma\di \approx 100 \Sigma_{\rm p}$ for a dust-to-gas ratio of 0.01.
+\subsection{Core Growth during KH Contraction}
+\label{raf3}
+
+In this section we investigate whether planetesimal accretion during the gas contraction phase at the rate $\dot{M}<\dot{M\co}_{\rm{KH}}$  can alter the core mass enough to affect the time evolution of the atmosphere.
+We can quantitatively estimate the increase in core mass as
+
+\begin{equation}
+\label{eq:cminc}
+\Delta M_{\rm{c}} = \int_0^{t_{\rm{co}}} \dot{M_{\rm{c}}} dt \approx \sum_i \dot{M_{\rm{c}}}_i \Delta t_i,
+\end{equation}
+
+ \noindent where the accretion rate $ \dot{M_{\rm{c}}}_i $ is given by
+
+ \begin{equation}
+ \label{eq:Mdotexp}
+ \dot{M_{\rm{c}}}_i =\frac{L_i R\co}{G M_{\rm{c}}} 
+ \end{equation}
+
+ \noindent from equation (\ref{eq:Lacc}), with $L_i$ the luminosity of the atmosphere at time $t_i$ in our model.
+For $M_{\rm{c}}=10 M_{\oplus}$, we find $\Delta M_{\rm{c}} \approx 0.2 M_{\oplus} \ll 10 M_{\oplus}$.
+It follows that a significant increase in core mass that could potentially alter the time evolution of the atmosphere would occur on a much longer timescale then the crossover time for the unperturbed atmosphere.
+The time evolution of the atmosphere is thus insensitive to core mass changes at a rate imposed by the assumption that $L_{\rm{acc}}<L_{\rm{KH}}$.
 
@@
 
-\subsection{Comparison with Standard Results}
+\subsection{Comparison with Steady-State Results}
 \label{raf2}
 
-Next, we are interested in whether hydrodynamic gas accumulation due to planetesimal accretion can already commence before the atmosphere becomes unstable due to Kelvin-Helmholtz contraction, as our regime is no longer the relevant one under such conditions.
-From equation (\ref{eq:Lacc}), the planetesimal accretion rate during the atmosphere evolution has to satisfy
-\begin{equation}
-\label{eq:McdotKH}
-\dot{M\co} \lesssim \frac{L_{\rm KH} R\co}{G M\co}
-\end{equation}
-for our regime to be valid, i.e.
-$L_{\rm acc} \lesssim L_{L_{\rm KH}}$.
-The maximum planetesimal accretion rate $\dot{M}_{\rm c, KH}$ that satisfies this condition (when $L_{\rm acc}=L_{\rm KH}$) is denoted by the dotted line in Figure \ref{fig:accrates} for $a=60$ AU and $M\co=5M_{\oplus}$.
+Next, we check that runaway gas accumulation due to planetesimal accretion cannot already commence before the atmosphere becomes unstable due to KH contraction, as our regime is no longer relevant under such conditions.
+We recall that in steady-state studies, the critical core mass is  the maximum core mass for which the atmosphere is still in hydrostatic balance; runaway gas accumulation occurs if the core mass grows further.
+A higher planetesimal accretion rate results in a larger critical core mass for the steady-state models.
+As such, if unstable atmosphere collapse does not occur due to planetesimal accretion for the lowest value of $\dot{M}_{\rm c, KH}$ over the course of the atmosphere's growth (see, e.g., Figure \ref{fig:accrates}), then it can only occur in the KH dominated regime.
+In what follows we calculate the critical core mass $M_{\rm crit, KH}$ corresponding to $\min(\dot{M}_{\rm c, KH})$ and show that it is higher that the critical core mass we determined in the KH dominated regime.
+
+%The maximum planetesimal accretion rate $\dot{M}_{\rm c, KH}$ that satisfies equation (\ref{eq:McdotKH})
+
+%is denoted by the dotted line in Figure \ref{fig:accrates} for $a=60$ AU and $M\co=5M_{\oplus}$.
 If unstable atmosphere collapse does not occur due to planetesimal accretion for the lowest value of $M_{\rm c, KH}$, then runway gas accretion can only occur in the KH dominated regime.
@@
 
-In order to estimate $M_{\rm c, KH}$, we use the results of \citet{rafikov06} for low luminosity atmospheres forming in the outer disk ($>2-5$ AU), consistent with our region of interest.
-\citet{rafikov06} assumes an ideal gas polytropic equation of state and a lower opacity than the standard ISM opacity that we use in our calculations (see equation \ref{eq:opacitylaw}).
-We thus calculate the critical core mass for an ideal gas polytrope with the standard ISM opacity reduced by a factor of 100, which is comparable to the opacity law used by \citet{rafikov06} (see PY13).
-
-By relating his expression for the critical core mass to a given core mass dependent planetesimal accretion rate $\dot{M}(M_{\rm{c}})$, we find the following expression for the critical core mass when accretion luminosity dominates the evolution of the atmosphere:
+In order to estimate $M_{\rm crit, KH}$, we use the results of \citet{rafikov06} for low luminosity atmospheres forming in the outer disk ($>2-5$ AU), consistent with our region of interest.
+\citet{rafikov06} assumes an ideal gas polytropic EOS and a lower opacity than the standard ISM opacity that we use in our calculations (see equation \ref{eq:opacitylaw}).
+We thus calculate the critical core mass for an ideal gas polytrope with the normalization factor $F_{\kappa}$ reduced by a hundred, which is comparable to the opacity law used by \citet{rafikov06} \footnote{The power-law opacity of \citet{rafikov06} is scaled to the (semi-major axis dependent) disk temperature, while our opacity is scaled to an absolute reference temperature.
+We thus cannot directly use the \citet{rafikov06} opacities for our comparison.}.
+%(see Paper I).
+
+Following \citet{rafikov06}, we find the following expression for the critical core mass when accretion luminosity dominates the evolution of the atmosphere:
 
@@
 \label{eq:critraf}
-M_{\rm{crit, KH}} \sim \Big[\frac{\dot{M}(M_{\rm{c}})}{64 \pi^2 C} \frac{\kappa_0}{\sigma G^3} \frac{1}{R\co M\co^{1/3}} \Big(\frac{k_b}{\mu}\Big)^4\Big]^{3/5},
-\end{equation}
-
-\noindent with all the constants as defined in previous sections, and $C$ an order unity constant depending on the adiabatic gradient and disk properties (see \citealt{rafikov06}, equation B3).
+M_{\rm{crit, KH}} \sim \Big[\frac{\min[\dot{M}_{\rm c, KH}(M_{\rm{c}})]}{64 \pi^2 C} \frac{\kappa_0}{\sigma G^3} \frac{1}{R\co M\co^{1/3}} \Big(\frac{k_b}{\mu}\Big)^4\Big]^{3/5},
+\end{equation}
+with $C$ an order unity constant depending on the adiabatic gradient and disk properties (see \citealt{rafikov06}, equation B3).
+From equation (\ref{eq:McdotKH}), the accretion rate $\dot{M}_{\rm c, KH}$ depends on the core mass $M\co$.
+We find $M_{\rm crit, KH}$ numerically by setting $M\co=M_{\rm crit, KH}$ on the right-hand side of equation (\ref{eq:critraf}).
 The result is displayed in Figure \ref{fig:raf2}; the critical core mass corresponding to planetesimal accretion at the rates displayed in Figure \ref{fig:accrates} is displayed for comparison.
@@
 %\vspace{-0.5in}
-\caption{Comparison between the critical core mass $M_{\rm{crit, KG}}$ due to planetesimal accretion and the assumed fixed core mass when gas contraction dominates, for a growth time of $\tau=2.1$ Myrs.
-Our results yield lower core masses than in the standard case.}
+\caption{Comparison between the critical core mass $M_{\rm{crit, KH}}$ due to planetesimal accretion and the assumed fixed core mass when gas contraction dominates.
+Our results yield lower core masses than in the fast planetesimal accretion case (e.g., \citealt{rafikov06}).
+The critical core mass corresponding to $\dot{M}_{\rm c, Hill}$ and $\dot{M}_{\rm c, acc}$ from Figure \ref{fig:accrates} is plotted for comparison.}
 \label{fig:raf2}
@@
 
-The critical core mass due to planetesimal accretion is smaller than in the case in which planetesimal accretion dominates the evolution of the atmosphere.
+The critical core mass due to KH contraction is smaller than in the case in which planetesimal accretion dominates the evolution of the atmosphere.
 This brings us to two conclusions.
 First, we confirm that planetesimal accretion can be safely ignored in our regime of interest.
-Secondly, this comparison tells us that it is easier and more efficient to form a planet by growing the core first, then accreting a massive envelope, rather than by growing the core and atmosphere in parallel.
-Moreover, our result represents a true, absolute minimum on the core mass that is needed to form a giant planet during the lifetime of the protoplanetary disk, as our core no longer grows.
+Secondly, this comparison tells us that the core mass needed to form a giant planet is smaller if the core forms first, then accretes a massive envelope, than in the case where the core and atmosphere grown simultaneously in a high planetesimal accretion regime.
+Moreover, our result represents a true, absolute minimum on the core mass needed to form a giant planet during the lifetime of the protoplanetary disk, as our core no longer grows.
 
@@
 
-As a final check, we investigate whether planetesimal accretion during the gas contraction phase at the rate $\dot{M}_{\rm{KH}}$ imposed by the condition that $L_{\rm{acc}}<L_{\rm{KH}}$ can alter the core mass enough to affect the time evolution of the atmosphere.
-We can quantitatively estimate the increase in core mass as
-
-\begin{equation}
-\label{eq:cminc}
-\Delta M_{\rm{c}} = \int_0^{t_{\rm{co}}} \dot{M_{\rm{c}}} dt \approx \sum_i \dot{M_{\rm{c}}}_i \Delta t_i,
-\end{equation}
-
- \noindent where the accretion rate $ \dot{M_{\rm{c}}}_i $ is given by
-
- \begin{equation}
- \label{eq:Mdotexp}
- \dot{M_{\rm{c}}}_i =\frac{L_i R\co}{G M_{\rm{c}}} 
- \end{equation}
-
- \noindent from equation (\ref{eq:Lacc}), with $L_i$ the luminosity of the atmosphere at time $t_i$ in our model.
-For $M_{\rm{c}}=10 M_{\oplus}$, we find $\Delta M_{\rm{c}} \approx 0.2 M_{\oplus}$.
-This mass increase is therefore negligible in comparison with the initial core mass.
-It follows that a significant increase in core mass that could potentially alter the time evolution of the atmosphere would occur on a longer time scale than the mass-doubling time for the unperturbed atmosphere.
-Therefore, the time evolution of the atmosphere is insensitive to core mass changes at a rate imposed by the assumption that $L_{\rm{acc}}<L_{\rm{KH}}$.
-
 %\section{Model Validity??? \textbf{(for now, in lack of a better title)}}
@@
 
- In this paper we have studied giant planet formation under the assumption that the planetesimal accretion rate is negligible and the atmosphere evolution is dominated by gas contraction.
-We have used the model developed in PY13 to build atmosphere profiles assuming that the nebular gas obeys a realistic equation of state that takes into account non-ideal effects.
-We found that the variations in the adiabatic index due to the realistic equation of state result in a significantly larger crossover time, and therefore critical core mass, when compared to an ideal gas polytrope.
-While for an ideal diatomic gas the minimum core mass to form a giant planet under the assumptions of our model is lower than the typically quoted value of $10 M_{\oplus}$ (see Piso \& Youdin in prep.), the inclusion of non-ideal effects brings this values back to around $10 M_{\oplus}$,
-
- We also compared our results to standard studies that assume that the evolution of the gaseous envelope is dominated by planetesimal accretion.
-We found that that our model yields lower core masses than the standard results.
-It is therefore easier to form a giant planet by growing the core first, then reducing the planetesimal accretion rate and let the atmosphere evolve on a Kelvin-Helmholtz time scale.
-Moreover, our results represent a true minimum on the core mass needed to form a giant planet during the typical lifetime of a protoplanetary disk.
+ In this paper we study the formation of giant planets embedded in a gas disk.
+We consider atmosphere evolution around fully grown cores and determine the minimum (critical) core mass required to form a gas giant during the typical lifetime of a protoplanetary disk.
+We improve the model developed in Paper I (\textbf{does this also need to be written out as Piso \& Youdin?}) by including realistic equation of state (EOS) tables and dust opacities.
+For a realistic EOS and interstellar grain opacity, the critical core mass is $\sim$$20 M_{\oplus}$ at 5 AU and drops to $\sim$$6 M_{\oplus}$ in our fiducial disk.
+These results are more than twice as large than those calculated in Paper I for a polytropic EOS, and bring the critical core mass to $\sim$$10 M_{\oplus}$, the value typically quoted in many core accretion studies.
+When realistic opacities that include grain growth are used, the critical core mass is significantly lower, i.e.
+$\sim$$6 M_{\oplus}$ at 5 AU and $\sim$$4 M_{\oplus}$ at 100 AU.
+Different assumptions about the grain size distribution (e.g., that take into account coagulation) may further reduce the critical core mass.
+
+Our results yield lower core masses than analogous results that consider high planetesimal accretion rates for which the core and atmosphere grow simultaneously.
+It is thus possible to form a giant planet from a smaller core if the core grows first, then the accretion rate of solids is reduced and a gaseous envelope is accumulated.
+Moreover, since additional heat sources such as planetesimal accretion limit the ability of the atmosphere to cool and undergo Kelvin-Helmholtz contraction, our results represent a true minimum on the core mass needed to form a giant planet during the typical lifetime of a protoplanetary disk.
+
+% In this paper we study giant planet formation assuming that planetesimal accretion is negligible and the atmosphere evolution is dominated by KH gas contraction.
+We use the model developed in Paper I to determine the critical core mass to form a giant planet before disk dissipation assuming that the nebular gas obeys a realistic EOS that takes into account non-ideal effects.
+We find that variations in the adiabatic gradient due to thermodynamic effects such as dissociation and hydrogen spin isomers result in a critical core mass more than twice as large than in the case of a gas described by an ideal gas polytrope.
+This brings the critical core mass to $M_{\rm{c, crit}} \sim 10 M_{\oplus}$, which is the value typically quoted in many core accretion studies (see PY 13).
+We further make realistic assumptions about the dust opacity in the protoplanetary disk by taking into account grain growth.
+The resulting critical core mass is much less sensitive to outer boundary conditions (disk temperature and pressure) and is found to be $M_{\rm{c, crit}} \sim 5 M_{\oplus}$.
+Different grain size distribution assumptions (i.e., that take into account coagulation) may further reduce the critical core mass.
+
+ %While for an ideal diatomic gas the minimum core mass to form a giant planet under the assumptions of our model is lower than the typically quoted value of $10 M_{\oplus}$ (see PY13), the inclusion of non-ideal effects brings this values back to around $10 M_{\oplus}$
+
+% We also compare our results to studies that assume high planetesimal accretion rates, due to which the atmosphere is in steady state and entirely heated by planetesimals, and find that that our model yields lower critical core masses.
+This shows that it is easier to form a giant planet by growing the core first, then reducing the planetesimal accretion rate and letting the atmosphere evolve on a KH time scale.
+Moreover, since additional heat sources such as planetesimal accretion limit the ability of the atmosphere to cool and undergo KH contraction, our results represent a true minimum on the core mass needed to form a giant planet during the typical lifetime of a protoplanetary disk.
 
@@
 
-In this section we explain the procedure for extending and interpolating the \cite{saumon95} equation of state tables.
-The equation of state takes into account non ideal interactions, and includes physical treatments of dissociation and ionization.
+In this study we consider atmosphere growth in the outer parts of protoplanetary disks ($a \lesssim 100$ AU), where temperature and pressure drop to $T \sim 20$ K and $P \sim10^{-4}$ dyn cm$^{-2}$ for our MMSN disk model (see equations \ref{eq:diskb} and \ref{eq:Pd}).
+We assume that the nebular gas is described by a realistic equation of state (EOS), as prescribed by the \citet{saumon95} EOS tables.
+However, these tables only cover a relatively high range of temperatures and pressures, i.e.
+$2.1 < \log_{10} T(\rm{K})<7.06$ and $4.0<\log_{10}P$(dyn cm$^{-2})<19.0$.
+We thus need to extend the tables to lower $T$ and $P$, as required by our disk model.
+We choose $\log_{10} T =1.0$ and $ \log_{10}P=-4.4$ as our lower boundaries for temperature and pressure, respectively, resulting in an extended temperature and pressure grid
+
+\begin{eqnarray}
+1.0 & < & \log_{10} T <2.1, \\ 
+-4.4& < & \log_{10} P<4.0. \\
+\end{eqnarray} 
+The other thermodynamic variables in the tables are calculated as follows.
+
+%In this section we explain the procedure for extending and interpolating the \cite{saumon95} EOS tables.
+The EOS takes into account non ideal interactions, and includes physical treatments of dissociation and ionization.
 However, the \cite{saumon95} EOS tables only cover a relatively high range of temperatures and pressures: $2.10 < \log_{10} T(\rm{K})<7.06$ and $4<\log_{10}P$(dyn cm$^{-2})<19$.
 We consider cold disks, where the temperature and pressure drop to $\sim 20$ K and $\sim 10^{-4}$ dyn cm$^{-2}$, respectively (see equations (\ref{eq:diskb}) and (\ref{eq:Pd})).
-As such, it is necessary to extend the \cite{saumon95} EOS tables to lower temperature and pressure values.
-
-We choose $\log_{10} T (\rm{K})=1$ and $ \log_{10}P$(dyn cm$^{-2})=-4.4$ as our lower boundaries for temperature and pressure, respectively.
+It is therefore necessary to extend the \cite{saumon95} EOS tables to lower temperature and pressure values.
+
+%We choose $\log_{10} T (\rm{K})=1$ and $ \log_{10}P$(dyn cm$^{-2})=-4.4$ as our lower boundaries for temperature and pressure, respectively.
 Our temperature and pressure grid becomes: $1 < \log_{10} T(\rm{K})<7.06$ and $-4.4<\log_{10}P$(dyn cm$^{-2})<19$.
@@
 
-\begin{equation}
-\label{eq:z}
-Z=Z_t Z_r Z_v Z_e Z_n,
-\end{equation}
-
-\noindent where $Z_t$, $Z_r$, $Z_v$, $Z_e$ and $Z_n$ are the partition functions associated with translation, rotation, vibration, electronic excitation and nuclear excitation, respectively.
-For hydrogen, electronic and nuclear excitation are only significant at temperatures higher than our region of interest ($\theta_e \approx 12000$ K and $\theta_n >> \theta_e$, where $\theta_e$ and $\theta_n$ are the characteristic temperatures for electronic and nuclear excitation, respectively).
-As such, we will only take into account the translation, rotation and vibration of the hydrogen molecule:
+%\begin{equation}
+%\label{eq:z}
+%Z=Z_t Z_r Z_v Z_e Z_n,
+%\end{equation}
 
@@
 
-The partition function associated with the motion of the center of mass of the molecule is given by (in the classical limit):
+\noindent where $Z_t$, $Z_r$, $Z_v$ are the partition functions associated with translation, rotation, and vibration, respectively \footnote{We ignore electronic and nuclear excitation as they are only important at temperatures much higher than our regime of interest}.
+Our derivations follow \citet{kittel} %For hydrogen, electronic and nuclear excitation are only significant at temperatures higher than our region of interest ($\theta_e \approx 12000$ K and $\theta_n >> \theta_e$, where $\theta_e$ and $\theta_n$ are the characteristic temperatures for electronic and nuclear excitation, respectively).
+As such, we will only take into account the translation, rotation and vibration of the hydrogen molecule:
+
+%In what follows we present and briefly derive expressions for thermodynamic variables based on quantum mechanics principles. More details on the derivations can be found in \citet{kittel}.
+
+In the classical limit, the partition function associated with the motion of the center of mass of a gas molecule of mass $m$ is given by:
 
@@
 \end{equation}
-
-\noindent where $\beta=1/(k T)$ and $V$ is the volume. The rotational partition function is generally written as:
+with  $T$ and $V$ the gas temperature and volume, respectively, $\hbar$ the reduced Planck constant, and  $\beta=1/(k_b T)$. The rotational partition function is generally written as:
 
@@
 However, molecular hydrogen occurs in two isomeric forms: orthohydrogen, with the proton spins aligned parallel to each other, and parahydrogen, with the proton spins aligned antiparallel.
-Parahydrogen can only a have symmetric (even) wave function associated with rotation, while orthohydrogen can only have an antisymmetric (odd) wave function associated with rotation (see section \ref{deladtable} for an explanation why).
+Parahydrogen can only a have symmetric (even) wave function associated with rotation, while orthohydrogen can only have an antisymmetric (odd) wave function associated with rotation (see section \S\ref{deladtable} for an explanation why).
 The rotational partition functions for ortho- and parahydrogen can thus be written as:
@@
 \end{equation}
-
 The factor of 3 above accounts for the three-fold degeneracy of the ortho state.
@@
 \end{equation}
-
 In our range of temperatures of interest, we found that $Z_r$ converges after about 25 terms in the series.
@@
 
-If the partition function of a system of $N$ particles is known in terms of $(V, T, N)$, the internal energy and entropy of the system can be determined as follows:
-
-\begin{equation}
-\label{eq:U}
-U_N=k T^2 \Big(\frac{\partial \log{Z}}{\partial T}\Big)_{V, N}
-\end{equation}
-
-\begin{equation}
-\label{eq:S}
-S_N=k \log{Z} + \frac{U_N}{T}
-\end{equation}
-
-The energy, and entropy per mass and specific heat capacity will subsequently be:
+If the partition function of a system particles is known in terms of $(V, T)$, the internal energy per unit mass, entropy per unit mass and specific heat capacity can be written as
+
+%\begin{equation}
+%\label{eq:U}
+%U_N=k T^2 \Big(\frac{\partial \log{Z}}{\partial T}\Big)_{V, N}
+%\end{equation}
+%
+%\begin{equation}
+%\label{eq:S}
+%S_N=k \log{Z} + \frac{U_N}{T}
+%\end{equation}
+%
+%The energy, and entropy per mass and specific heat capacity will subsequently be:
 
@@
 \label{eq:u}
-U=\mathcal{R} T^2 \Big(\frac{\partial \log{Z}}{\partial T}\Big)_{V, N}
-\end{equation}
-
+U=\mathcal{R} T^2 \Big(\frac{\partial \log{Z}}{\partial T}\Big)_{V}
+\end{equation}
 \begin{equation}
@@
 \end{equation}
-
 \begin{equation}
 \label{eq:cv}
-C_v=\Big(\frac{\partial U}{\partial T}\Big)_{V, N}
-\end{equation}
-
-Since $Z=Z_t Z_r Z_v$, it is easy to notice that $U=U_t+U_r+U_v$ and $S=S_t+S_r+S_v$, where $U_t$, $U_r$, $U_v$, $S_t$, $S_r$, $S_v$ are the quantities corresponding to the individual translation, rotation and partition functions, respectively.
-
-It can be shown that the entropy per mass due to translational motion can be expressed as:
+C_v=\Big(\frac{\partial U}{\partial T}\Big)_{V}.
+\end{equation}
+Since $Z=Z_t Z_r Z_v$, it is easy to notice that $U=U_t+U_r+U_v$ and $S=S_t+S_r+S_v$, where $U_t$, $U_r$, $U_v$, $S_t$, $S_r$, $S_v$ are the quantities corresponding to the individual translation, rotation and vibration partition functions, respectively.
+
+The entropy per mass due to translational motion can be expressed as:
 
@@
 Equation (\ref{eq:st}) is known as the Sackur-Tetrode formula, and it is only applicable to an ideal gas.
-It can also be easily shown that the internal energy per mass due to translational motion is given by:
+The internal energy per mass due to translational motion is given by:
 
@@
 \end{equation}
-
-\noindent and
-
+and
 \begin{equation}
 \label{eq:sP}
-S_T=\frac{\partial \log{S}}{\partial \log{P}} \Big |_T
-\end{equation}
-
-We calculate $S_T$ and $S_P$ using the table values for $S$, $T$ and $P$, and a linear central difference formula.
-
-\item{\textbf{Adiabatic gradient $\nabla_{ad}$}}. The adiabatic gradient is defined as:
+S_P=\frac{\partial \log{S}}{\partial \log{P}} \Big |_T
+\end{equation}
+We calculate $S_T$ and $S_P$ through finite differencing.
+
+\item{\textbf{Adiabatic gradient $\delad$}. The adiabatic gradient is defined as:
 
@@
 \label{eq:deladSP}
-\nabla_{ad}=\frac{\partial \log{T}}{\partial \log{P}} \Big |_S = -\frac{S_P}{S_T}
+\delad=\frac{\partial \log{T}}{\partial \log{P}} \Big |_S = -\frac{S_P}{S_T}
 \end{equation}
@@
 We evaluate it from the tabulated values for $S_T$ and $S_P$ determined above.
-Figure 1 shows a contour plot of the adiabatic index for the extended EOS table, while the black lines represent constant entropy curves.
-The upper right part of the plot ($\log T>2.1$ and $\log P>4$) is based on the \cite{saumon95} EOS table, while the rest of the plot is our extension.
-We see that the two tables join smoothly for entropy curves between $8.80<\log{S}$(K g$^{-1})<9.07$.
+Figure \ref{fig:deladH} shows a contour plot of $\delad$ as a function of temperature and pressure for the extended EOS table.
+%Our extension smoothly matches the original tables for $8.80<\log{S}$(K g$^{-1})<9.07$ (\textbf{numbers are wrong, change once you have the final figure version}).
 
@@
 \includegraphics[scale=.8]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/delad_S_H.pdf}
-\caption{Contour plot of the adiabatic gradient $\delad$ for the hydrogen extended table. The black curves represent constant entropy curves.}
+\caption{Contour plot of the hydrogen adiabatic gradient $\delad$ as a function of gas temperature and pressure.
+The upper right rectangle encloses the region described by the original \citet{saumon95} EOS tables, while the rest of the plot is our extension to lower temperatures and pressures.
+The black curves represent constant entropy adiabats, with the labels the natural logarithm of the absolute entropy per unit mass.
+At high temperatures, hydrogen dissociates and ionizes, while at low temperatures the rotational states of the hydrogen molecule are only partially excited and it no longer behaves like an ideal diatomic gas.
+Regions in which the EOS is invalid or has not been computed are masked in white (see Figure \ref{fig:deladmap} caption for an explanation).}
+\label{fig:deladH}
 \end{figure}
@@
 We extend the helium EOS tables based on a similar procedure.
-Since helium is primarily neutral and atomic at low temperatures and pressures, we treat it as an ideal monoatomic gas, and subsequently only take into account the translational components of the necessary thermodynamic quantities (see subsection \ref{hydrogen} above for details).
-The analogous $\nabla_{ad}$ contour plot for helium can be seen in Figure 2.
-We notice that, in the case of helium, the original and extended table join smoothly for entropy curves between $8.29<\log{S}$(K g$^{-1})<8.77$.
+Since helium is primarily neutral and atomic at low temperatures and pressures, we treat it as an ideal monoatomic gas, and thus only take into account the translational component of the partition function (\ref{eq:Zt}).
+Figure \ref{fig:deladHe} shows $\delad$ as a function of temperature and pressure for the extended EOS table.
+%The original and extended table join smoothly for entropy curves between $8.29<\log{S}$(K g$^{-1})<8.77$ in this case (\textbf{numbers are wrong, change once you have the final figure version}).
 
@@
 \includegraphics[scale=.8]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/delad_S_He.pdf}
-\caption{Contour plot for the adiabatic gradient $\delad$ for the helium extended table. The black curves represent constant entropy curves.}
+\caption{Contour plot of the helium adiabatic gradient $\delad$ as a function of gas temperature and pressure.
+The upper right rectangle encloses the region described by the original \citet{saumon95} EOS tables, while the rest of the plot is our extension to lower temperatures and pressures.
+The black curves represent constant entropy adiabats, with the labels the natural logarithm of the absolute entropy per unit mass.
+Helium ionizes at $T \gtrsim 10,000$ K, but behaves as an ideal monatomic gas otherwise.
+We choose $T=7,000$ K as a conservative temperature cutoff above which our extension is no longer valid (masked in white).
+The EOS has not been computed in the lower-right region of the plot --- see Figure \ref{fig:deladmap} for an explanation.}
+\label{fig:deladHe}
 \end{figure}
@@
 
-\section{Adiabatic Gradient during Partial Ionization}\label{deladioniz}
-
-For a partially ionized gas, the total internal energy includes contributions from the individual internal energies of neutral atoms, ions and electrons, as well as from the ionization energy.
-Specifically, if we denote the internal energies of neutral hydrogen, protons and electrons as $U_{H}$, $U_+$ and $U_e$, respectively, then the total internal energy of the gas is given by:
-
-\begin{equation}
-U=U_H+U_+ + U_e + x \chi,
-\end{equation}
-
-\noindent where $x$ is the ionization fraction and $\chi$ is the ionization energy (equal to -13.6 eV for hydrogen).
-The ionization fraction can be determined from the Saha equation (see e.g., \citealt{kippenhahn90}).
+\section{Adiabatic Gradient during Partial Dissociation}\label{deladdiss}
+
+%The adiabatic gradient during dissociation or ionization is a function of temperature and the dissociation or ionization fraction $x$.
+The Saha equation (e.g., \citealt{kippenhahn90}) relates $x$ to the gas
+
+The total internal energy of a partially dissociated gas includes contributions from the individual internal energies of the molecules and atoms, as well as from the dissociation energy.
+The dissociation energy depends on the dissociation fraction $x$ (i.e., the fraction of molecules that have dissociated), which can be found from the Saha equation (see e.g., \citealt{kippenhahn90}, Chapter 14) as a function of temperature and density,
 
@@
 \label{eq:saha}
-\frac{x^2}{1-x} \frac{\rho}{m_H}=\frac{(2 \pi m_e k_B T)^{3/2}}{h^3} e^{-\chi/k_B T},
-\end{equation}
-
-\noindent where $m_e$ is the mass of the electron and $h$ is Planck's constant.
-It can be seen from the Saha equation that the ionization fraction depends only on the gas temperature and density: $x=x(T, \rho)$.
-As such, all the thermodynamic quantities also depend only on the gas temperature and density, and hence on the equation of state.
-The adiabatic gradient is given by (see \citealt{kippenhahn90}, chapter 14 for a derivation):
-
-\begin{equation}
-\delad=\frac{2+x (1-x) \Phi_H}{5+x (1-x) \Phi_H^2},
-\end{equation} 
-with $\Phi_H=\frac{5}{2}+\frac{\chi}{k T}$.
-Figure \ref{fig:deladion} shows the behavior of $\delad$ for partially ionized hydrogen.
-We recover $\delad=2/5$ for $x=0$ (pure atomic hydrogen) and $x=1$ (fully ionized plasma).
-The adiabatic gradient decreases significantly for intermediate values of $x$, becoming smaller than 0.1 at its minimum (for $x=0.5$).
+\frac{x^2}{1-x} \propto \frac{T^{3/2}}{\rho} e^{-\chi/k_B T},
+\end{equation} 
+with $\chi$ the dissociation energy, equal to 4.48 eV for molecular hydrogen \citep{blanksby03}.
+
+The above also holds true for ionization, with the dissociation energy replaced by ionization energy $\chi=13.6$ eV for atomic hydrogen \citep{mandl89}.
+From the Saha equation one can find an expression for $\rho$ as a function of $T$ and $x$, then derive the adiabatic gradient directly from its definition (equation \ref{eq:delad}), taking into account the fact that the mean molecular weight in the ideal gas law varies with $x$, hence the pressure will not only be a function of $T$ and $\rho$ but also of $x$ (see \citealt{kippenhahn90}, Chapter 14.3 for a detailed derivation).
+The final expression for the adiabatic gradient during ionization is
+
+\begin{equation}
+\label{eq:deladioniz}
+\delad=\frac{2+x (1-x) \Phi_H}{5+x(1-x)\Phi_H^2},
+\end{equation}
+with $\Phi_H \equiv \frac{5}{2}+\frac{\chi}{k_B T}$.
+The derivation of $\delad$ during dissociation is more involved mathematically (see, e.g., \citealt{vardya60}) and leads to a slightly more complicated final expression,
+
+%One can derive a simple expression for the adiabatic gradient as a function of $x$ and $T$ from it's definition (equation \ref{eq:delad}), using the
+
+%The adiabatic gradient follows as (see \citealt{vardya60} for a detailed derivation)
+
+\begin{equation}
+\label{eq:deladdiss}
+\delad=\frac{1+x+ \frac{x(1-x^2)}{2} \frac{\chi}{k_B T}}{5 x + \frac{7(1-x)}{2} + \frac{x(1-x^2)}{2} \Big(\frac{\chi}{k_B T}\Big)^2}.
+\end{equation} 
+We recover $\delad=2/7$ for $x=0$ (no ongoing dissociation hence hydrogen is purely molecular and diatomic) and $\delad=2/5$ for $x=1$ (hydrogen is fully dissociated into atoms and hence monatomic).
+Figure \ref{fig:deladdiss} shows the dependence of $\delad$ on the dissociation fraction, for $T=3000$ K, the temperature at which dissociation typically occurs \citep{langmuir12}.
+The adiabatic gradient drops substantially during partial dissociation, since part of the internal energy is used in dissociation rather than in increasing the temperature of the system.
 
@@
 %%\vspace{-0.5in}
-\caption{Adiabatic gradient as a function of the hydrogen ionization fraction $x$.
-The adiabatic gradient is $\delad=2/5$ for pure atomic hydrogen ($x=0$) and fully ionized hydrogen ($x=1$), and drops to low values during partial ionization.}
-\label{fig:deladion}
+\caption{Adiabatic gradient as a function of the hydrogen dissociation fraction $x$.
+The adiabatic gradient is $\delad=2/7$ for pure molecular hydrogen ($x=0$) and $\delad=2/5$ for fully atomic hydrogen ($x=1$), and drops to low values during partial dissociation.}
+\label{fig:deladdiss}
 \end{figure}
 
-\section{Radiative Windows} \label{radwindow}
+%with $\chi$ and $T$ the dissociation energy and temperature, respectively, equal to $\chi=4.48$ eV\citep{blanksby03} and $T\sim3000$ K \citep{langmuir12} for molecular hydrogen.
+
+%Specifically, if we denote the internal energies of neutral hydrogen, protons and electrons as $U_{H}$, $U_+$ and $U_e$, respectively, then the total internal energy of the gas is given by:
+
+%\begin{equation}
+%U=U_H+U_+ + U_e + x \chi,
+%\end{equation}
+%
+%\noindent where $x$ is the ionization fraction and $\chi$ is the ionization energy (equal to -13.6 eV for hydrogen).
+The ionization fraction can be determined from the Saha equation (see e.g., \citealt{kippenhahn90}).
+%
+%\begin{equation}
+%\label{eq:saha}
+%\frac{x^2}{1-x} \frac{\rho}{m_H}=\frac{(2 \pi m_e k_B T)^{3/2}}{h^3} e^{-\chi/k_B T},
+%\end{equation}
+%
+%\noindent where $m_e$ is the mass of the electron and $h$ is Planck's constant.
+It can be seen from the Saha equation that the ionization fraction depends only on the gas temperature and density: $x=x(T, \rho)$.
+As such, all the thermodynamic quantities also depend only on the gas temperature and density, and hence on the equation of state.
+The adiabatic gradient is given by (see \citealt{kippenhahn90}, chapter 14 for a derivation):
+%
+%\begin{equation}
+%\delad=\frac{2+x (1-x) \Phi_H}{5+x (1-x) \Phi_H^2},
+%\end{equation} 
+%with $\Phi_H=\frac{5}{2}+\frac{\chi}{k T}$.
+Figure \ref{fig:deladion} shows the behavior of $\delad$ for partially ionized hydrogen.
+We recover $\delad=2/5$ for $x=0$ (pure atomic hydrogen) and $x=1$ (fully ionized plasma).
+The adiabatic gradient decreases significantly for intermediate values of $x$, becoming smaller than 0.1 at its minimum (for $x=0.5$).
+
+\section{Grain growth opacity and radiative windows} \label{radwindow}
+
+The opacity of the interstellar medium is reasonably well constrained and approximate analytic expressions for the Rosseland mean opacity as a function of temperature and density are derived in \citet{bell94}.
+For low temperatures ($T \lesssim 100$ K) at which ice grains are present, opacity scales with temperature as $\kappa \sim T^2$.
+Sublimation of ice grains at $\sim 150$ K and of metal grains at $\sim 1000$ K results in sharp opacity drops.
+This is shown in Figure \ref{fig:opacity} for a gas density $\rho=10^{-8}$ g cm$^{-3}$, which is typical for the outer regions of protoplanetary disks.
+\citet{semenov03} calculate Rosseland mean opacities in protoplanetary disks for grains of different sizes and structure.
+As shown in Figure \ref{fig:opacity}, their results are in good agreement with \citet{bell94}.
+However, \citet{semenov03} do not take grain growth into account, which is likely to occur in protoplanetary disks, particularly when most solids have been accreted.
+\citet{dalessio01} compute wavelength dependent opacities for a range of maximum particle sizes and different size distributions.
+Figure \ref{fig:opacity} shows the integrated Rosseland mean opacity for a maximum particle size of 1 cm and a power law size distribution $n \sim a^{-p}$, with $a$ the grain size and $p=3.5$ for a normal collisional cascade and $p=2.5$ when coagulation is taken into account.
+We see in Figure \ref{fig:opacity} that this yields a mean opacity that is both lower and less sensitive to temperature, when compared to \citet{bell94} or \citet{semenov03}.
+However, observations of dust in protoplanetary disks have only been made at low temperatures (i.e., before dust sublimates).
+As we see in Figure \ref{fig:opacity}, the opacity dramatically decreases during dust sublimation for ISM grains, which can result in regions in the inner parts of planetary atmospheres where energy is transported radiatively, i.e.
+radiative windows.
+We thus use the \citet{bell94} opacities for $T \gtrsim 1000$ K, ensuring they smoothly match the \citet{dalessio01} opacities for lower temperatures.
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/kappa_grain_growth_SBL_paper.pdf}
+%%\vspace{-0.5in}
+\caption{Rosseland mean opacity of dust grains as a function of temperature for different opacity assumptions.
+The dashed black curve shows the \citet{bell94} analytic ISM opacity for $\rho=10^{-8}$ g cm$^{-3}$.
+The solid black curve shows the tabulated opacity of \citet{semenov03} for a dust composition of 'normal' silicates.
+The dashed red curve shows the \citet{dalessio01} opacity, which takes grain growth into account, for a maximum particle size of 1 cm and a normal collisional cascade grain size distribution ($p=3.5$).
+The solid red curve is the same as the dashed red curve, but it accounts for coagulation ($p=2.5$).}
+\label{fig:opacity}
+\end{figure}
+
+The significant opacity drop due to the sublimation of ice and metal grains lowers the radiative temperature gradient $\delrad$, which may result in one or more inner radiative layers inside the atmosphere of a protoplanet.
+This is displayed in Figure \ref{fig:delvsr}: depending on the semi-major axis and core mass, the opacity drop will generate no radiative window (top panel), one radiative window (middle panel), or two radiative windows (bottom panel).
 
@@
 %%\vspace{-0.5in}
-\caption{Radiative and adiabatic gradient as a function of radius. The sharp drom in opacity due to dust sublimation can generate one or more radiative windows.}
+\caption{Snapshots of the radiative and adiabatic gradient as a function of the radial coordinate, for planets with different core masses forming at various locations in the disk.
+The sharp drop in opacity due to dust sublimation may generate one or more radiative windows.
+Top panel: no radiative window for $a=100$ AU and $M\co=3 M_{\oplus}$.
+Middle panel: the opacity drop produces one radiative window for $a=50$ AU and $M\co=5 M_{\oplus}$.
+Bottom panel: the sharp drop in opacity results in two radiative windows for $a=20$ AU and $M\co=3 M_{\oplus}$.}
 \label{fig:delvsr}
@@
 
-\begin{figure}[h!]
-\centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/kappa_grain_growth_SBL_paper.pdf}
-%%\vspace{-0.5in}
-\caption{Opacities...}
-\label{fig:opacity}
-\end{figure}
-
 \end{document}

### 2013-10-24 tex/draft/draft3.tex
section 5

--- 2013-07-30 tex/draft/draft3.tex
@@
 %\newcommand{\aun}[1]{ a_{#1\oplus} }
+\def\crit{_{\rm{c, crit}}}
 
@@
 The relevant length scales of the atmosphere and disk satisfy the relation $\RB H\di^2 = 3 R_{\rm H}^3$.  The lengthscales are roughly equal at the ``thermal mass''
-\begin{equation}
+\begin{equation}\label{eq:Mth}
 M_{\rm th} > {c\di^{3} \over G \varOmega} \approx 25 \, {F_T^{3/2} \over \sqrt{m_\ast} } \, \aun{10}^{6/7}~ M_\oplus \, .
@@
 
-\section{Critical Core Mass}
+\section{Results for Giant Planet Formation}
 \label{sec:critical}
 
-%AY make sure this AU range comment is incorporated in some form:  We are primarily interested in the outer regions of the disk, so we generate atmosphere models at distances between 5 and 100 AU in our fiducial disk.
-We emphasize that the location in the disk only affects the temperature and pressure of the nebular gas.
-As our model applies only when planetesimal accretion occurs too slowly to substantially change the mass of the core, we consider the evolution of an atmosphere around a core of fixed mass.
+We now use our structure and evolution models to calculate the necessary conditions for gas giant formation by core accretion.
+Adopting a median disk lifetime of 3 Myr, we examine the conditions for a fixed mass core to to undergo runaway gas accretion within the lifetime of the gas disk.
+Our results for atmosphere growth times -- for a range of core masses and disk conditions -- are presented in \S\ref{sec:tcross}.
+Section \ref{sec:critcore} gives our results for critical core masses, the minimum masses which can accrete a massive atmosphere within the disk lifetime.
 
-\subsection{Disk Temperature and Pressure}
+Our models will focus on giant planet formation between 5 and 100 AU.
+The outer disk is of particular interest for direct imaging searches.
+The growth of atmospheres close to the star is also important, but spherical accretion models (including ours) are less applicable here.
+In the inner disk, critical core masses increase, yet lower mass planets start to open gaps and outgrow the disk scale height, see \Eq{eq:Mth}.
+These concerns prevent us from applying our model to the inner disk.
+
+\subsection{Runaway Growth Time Scale}
+\label{sec:tcross}
+The time to undergo runaway growth of the atmosphere, $t_{\rm run}$, sets a minimum timescale for the formation of giant planets.
+Due to the accelerating nature of runaway growth, the precise threshold chosen for $t_{\rm run}$ (explained in \S ?) is of minor significance.
+
+\begin{figure}[tb]
+%\centering
+\hspace{-.1in}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cumul_coolingtime_vs_Matm_10au_mu235.pdf}
+%\vspace{-0.5in}
+\caption{Time to grow an atmosphere of mass $M_{\rm{atm}}$ for cores with fixed masses between $5 M_{\oplus}$ and $14 M_{\oplus}$ (as labelled) at $10$ AU in our fiducial disk.
+Circles mark the runaway growth time, $t_{\rm run}$, which occurs at roughly the crossover mass, $M_{\rm{atm}} = M_{\rm c}$.
+Larger core masses give faster atmospheric growth, to a fixed atmosphere mass and also to runaway.
+However, for larger $M\co$, runaway growth commences at higher $M_{\rm atm}/M\co$ values.}
+\label{fig:tvsM}
+\end{figure}
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.48\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/coolingtime_vs_Mc_10au.pdf}
+%\vspace{-0.5in}
+\caption{Runaway growth times, $t_{\rm{run}}$, vs.\ core mass at $10$ AU, for two values of mean molecular weight.
+Our numerical model (\emph{solid curves}) is compared to our non-self-gravitating analytic model (\emph{dashed curves}).
+A typical protoplanetary disk life time of $3$ Myr is plotted for comparison.
+The crossover time is larger for a lower mean molecular weight.}
+\label{fig:tvsMcomp}
+\end{figure}
+
+\subsubsection{Effects of Core Mass}
+\label{Mct}
+
+\Fig{fig:tvsM} demonstrates the growth of atmospheric mass with time for several core masses at $10$ AU in our fiducial disk.
+Atmospheres grow faster around more massive cores due to stronger gravitational binding.
+The endpoint of each curve marks $t_{\rm run}$.
+Runaway growth occurs near the crossover mass, when $M_{\rm atm} \sim M_{\rm co}$ in agreement with previous studies.
+At lower core masses, however, runaway occurs at lower $M_{\rm atm}/M\co$ ratios.
+Thus, when compared to the naive assumption that runaway occurs at a universal $M_{\rm atm}/M\co$ ratio,  effects that lower the critical core mass are amplified by the earlier onset of runaway growth.
+
+\Fig{fig:tvsMcomp} shows how $t_{\rm run}$ varies with core mass, also at 10 AU.
+The numerical results are plotted against our non-self-gravitating analytic model, described in \S\ref{sec:coolingan}.
+The analytic model reproduces the general decline in $t_{\rm run}$ with core mass.
+The numerical model, which includes self-gravity, has a somewhat steeper mass dependence.
+A modest correction due to self-gravity is unsurprising, and consistent with the above-mentioned trend in $M_{\rm atm}/M\co$ ratios.
+Moreover, in the analytic theory, crucial quantities like $M_{\rm atm}$ and $L$ (both roughly $\propto M\co^3$ near crossover) have non-linear dependence on core mass, offering plenty of opportunity for self-gravitational corrections.
+
+The effect of mean molecular weight is also shown in \Fig{fig:tvsMcomp}.
+Lower $\mu$ values give longer growth times, and vice-versa.
+For $\mu = 2.0$, which represents the idealized case of an H$_2$ atmosphere devoid of Helium, $t_{\rm run}$ increases by factors of $\sim 2 - 3$.
+Thus fairly drastic changes in atmospheric composition are required for $\mu$ to significantly affect core accretion timescales.
+
+Looking ahead to our discussion of the critical core mass in \S\ref{sec:critcore}, \Fig{fig:tvsMcomp} shows that $t_{\rm run}$ equals a characteristic disk lifetime of 3 Myr for a core mass of  $M_{\rm crit} \approx 7 \, M_\oplus$  (for the standard case of $\mu = 2.35$).
+
+\begin{figure*}[tb]
+\centering
+\includegraphics[width=1.\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/TdPd_effect.pdf}
+\vspace{-0.3in}
+\caption{Crossover time as a function of disk temperature (\emph{left}) and pressure (\emph{right}) around a $M_{\rm c} = 5 M_{\oplus}$ core.
+The disk pressure or temperature (\emph{left} or \emph{right}, respectively) are fixed at values for 10 AU in our disk model.
+The analytic scalings are plotted for comparison, as described in the text.
+Gas accretion slows down significantly at higher temperatures, but only speeds up modestly as the disk pressure or density increase.}
+\label{fig:TPeffects}
+\end{figure*}
+
+\subsubsection{Disk Temperature and Pressure}
 \label{sec:TPeffects}
+
+\Fig{fig:TPeffects} shows how the crossover time varies with disk temperature, $T\di$, or pressure, $P\di$, holding the other quantity fixed.
+The analytic model roughly reproduces the temperature and pressure scalings, again with some discrepancies due mainly to the neglect of self-gravity.
+Temperature variations are much more significant than pressure variations (note the difference in logarithmic and linear axes).
+Since midplane disk conditions depend only on temperature and pressure in our model\footnote{See \Eq{eq:diskparam}.
+When gap opening is considered in models of later growth stages, the orbital frequency and effective viscosity become relevant as well.}, the dominant effect of disk location is temperature.
+
+The decline in growth times at colder temperatures arises from a variety of competing effects.
+Briefly, the larger Bondi radius and lower dust opacity at lower temperatures accelerate growth, overpowering the inherently lower cooling luminosity at lower temperatures.
+Growth times depend only weakly on, but do fall slightly with, pressure.
+The nearly exponential increase in pressure and density with depth through the radiative zone means that cooling, which is regulated at the RCB, has only a logarithmic dependence on disk pressure.
+Section \ref{sec:coolingan} explains these effects in more detail.
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_new.pdf}
+%\vspace{-0.5in}
+\caption{The critical core mass as a function of semi-major axis, for a disk lifetime $3$ Myrs and two values of the mean molecular weight ($\mu = 2.35$ is standard).
+The decline in $M_{\rm c,crit}$ with distance is a robust result for standard disk models.
+The analytic model, which neglects self-gravity, over-predicts the steepness of the decline.}
+\label{fig:Mcvsa}
+\end{figure}
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_opacity.pdf}
+%\vspace{-0.5in}
+\caption{Critical core masses vs.
+distance for standard and reduced (by a factor of ten) dust opacities.
+Lower opacities give significantly lower $M\crit$ values.
+Atmospheric opacity remains a large uncertainty in core accretion models.}
+\label{fig:Mcritopacity}
+\end{figure}
+
+\subsection{Critical Core Mass}
+\label{sec:critcore}
+
+The critical core mass declines with distance from the star, as shown in \Fig{fig:Mcvsa} for our standard disk model.
+The main reason for the decline, as explained above, is that atmospheres grow faster at colder temperatures.
+The lower densities and pressures in the outer disk have a much smaller effect.
+Since the vast majority of disk models have a disk temperature that declines with $a$, our qualitative result is robust.
+We also show that higher $\mu$ values give lower values of $M\crit$.
+
+The average power-law decline from 1 to 100 AU (not plotted) is $M\crit \propto a^{-0.3}$, for both choices of $\mu$.
+While not a drastic decline, the ability of distant low mass cores to accrete gas efficiently is significant for the interpretation of direct imaging surveys.
+However our model offers no guarantee of copious giant planets at large distances.
+Many histories of solid accretion are possible, and solid cores may grow too slowly to allow the rapid gas accretion that we model.
+
+Our non-self-gravitating analytic model (dashed curves in \Fig{fig:Mcvsa}) over-predicts the steepness of the decline in $M\crit$ vs.\ a.
+This discrepancy is not surprising, as we have shown that self-gravity strongly affects evolution, even before runaway growth begins and the crossover mass is reached.
+
+\Fig{fig:Mcritopacity} shows that reducing the opacity by an order of magnitude significantly reduces $M\crit$.
+Furthermore, this opacity effect is stronger at larger distances.
+The opacity reduction by a factor of 10 lowers $M\crit$ at 5 AU by a factor of $\sim 2.5$ and at 100 AU by a factor of $\sim 3.5$.
+Atmospheric opacity is a dominant uncertainty in core accretion modelling.
+However, unless atmospheric opacity varies significantly with disk radius, the general decline in $M\crit$ vs.\ $a$ should hold.
+For our scenario of negligible ongoing planetesimal accretion, it is tempting to think that dust could settle out of the radiative zone, lowering the opacity \citep{podolak03}.
+However, since opacity near the RCB is a crucial factor, we speculate that convective overshoot may prevent very low opacities.
 

### 2013-10-24 tex/draft/draft4.tex
section 5

@@
 The relevant length scales of the atmosphere and disk satisfy the relation $\RB H\di^2 = 3 R_{\rm H}^3$.  The length scales are roughly equal at the ``thermal mass'' (e.g., \citealt{menou04})
-\begin{equation}
+\begin{equation}\label{eq:Mth}
 M_{\rm th} > {c\di^{3} \over G \varOmega} \approx 25 \, {F_T^{3/2} \over \sqrt{m_\ast} } \, \aun{10}^{6/7}~ M_\oplus \, .
@@
 
-\section{Critical Core Mass}
+\section{Results for Giant Planet Formation}
 \label{sec:critical}
 
-%AY make sure this AU range comment is incorporated in some form:  We are primarily interested in the outer regions of the disk, so we generate atmosphere models at distances between 5 and 100 AU in our fiducial disk.
-We emphasize that the location in the disk only affects the temperature and pressure of the nebular gas.
-As our model applies only when planetesimal accretion occurs too slowly to substantially change the mass of the core, we consider the evolution of an atmosphere around a core of fixed mass.
-
-\subsection{Disk Temperature and Pressure}
+We now use our structure and evolution models to calculate the necessary conditions for gas giant formation by core accretion.
+Adopting a median disk lifetime of 3 Myr, we examine the conditions for a fixed mass core to to undergo runaway gas accretion within the lifetime of the gas disk.
+Our results for atmosphere growth times -- for a range of core masses and disk conditions -- are presented in \S\ref{sec:tcross}.
+Section \ref{sec:critcore} gives our results for critical core masses, the minimum masses which can accrete a massive atmosphere within the disk lifetime.
+
+Our models will focus on giant planet formation between 5 and 100 AU.
+The outer disk is of particular interest for direct imaging searches.
+The growth of atmospheres close to the star is also important, but spherical accretion models (including ours) are less applicable here.
+In the inner disk, critical core masses increase, yet lower mass planets start to open gaps and outgrow the disk scale height, see \Eq{eq:Mth}.
+These concerns prevent us from applying our model to the inner disk.
+
+\subsection{Runaway Growth Time Scale}
+\label{sec:tcross}
+The time to undergo runaway growth of the atmosphere, $t_{\rm run}$, sets a minimum timescale for the formation of giant planets.
+Due to the accelerating nature of runaway growth, the precise threshold chosen for $t_{\rm run}$ (explained in \S ?) is of minor significance.
+
+\begin{figure}[tb]
+%\centering
+\hspace{-.1in}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cumul_coolingtime_vs_Matm_10au_mu235.pdf}
+%\vspace{-0.5in}
+\caption{Time to grow an atmosphere of mass $M_{\rm{atm}}$ for cores with fixed masses between $5 M_{\oplus}$ and $14 M_{\oplus}$ (as labelled) at $10$ AU in our fiducial disk.
+Circles mark the runaway growth time, $t_{\rm run}$, which occurs at roughly the crossover mass, $M_{\rm{atm}} = M_{\rm c}$.
+Larger core masses give faster atmospheric growth, to a fixed atmosphere mass and also to runaway.
+However, for larger $M\co$, runaway growth commences at higher $M_{\rm atm}/M\co$ values.}
+\label{fig:tvsM}
+\end{figure}
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.48\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/coolingtime_vs_Mc_10au.pdf}
+%\vspace{-0.5in}
+\caption{Runaway growth times, $t_{\rm{run}}$, vs.\ core mass at $10$ AU, for two values of mean molecular weight.
+Our numerical model (\emph{solid curves}) is compared to our non-self-gravitating analytic model (\emph{dashed curves}).
+A typical protoplanetary disk life time of $3$ Myr is plotted for comparison.
+The crossover time is larger for a lower mean molecular weight.}
+\label{fig:tvsMcomp}
+\end{figure}
+
+\subsubsection{Effects of Core Mass}
+\label{Mct}
+
+\Fig{fig:tvsM} demonstrates the growth of atmospheric mass with time for several core masses at $10$ AU in our fiducial disk.
+Atmospheres grow faster around more massive cores due to stronger gravitational binding.
+The endpoint of each curve marks $t_{\rm run}$.
+Runaway growth occurs near the crossover mass, when $M_{\rm atm} \sim M_{\rm co}$ in agreement with previous studies.
+At lower core masses, however, runaway occurs at lower $M_{\rm atm}/M\co$ ratios.
+Thus, when compared to the naive assumption that runaway occurs at a universal $M_{\rm atm}/M\co$ ratio,  effects that lower the critical core mass are amplified by the earlier onset of runaway growth.
+
+\Fig{fig:tvsMcomp} shows how $t_{\rm run}$ varies with core mass, also at 10 AU.
+The numerical results are plotted against our non-self-gravitating analytic model, described in \S\ref{sec:coolingan}.
+The analytic model reproduces the general decline in $t_{\rm run}$ with core mass.
+The numerical model, which includes self-gravity, has a somewhat steeper mass dependence.
+A modest correction due to self-gravity is unsurprising, and consistent with the above-mentioned trend in $M_{\rm atm}/M\co$ ratios.
+Moreover, in the analytic theory, crucial quantities like $M_{\rm atm}$ and $L$ (both roughly $\propto M\co^3$ near crossover) have non-linear dependence on core mass, offering plenty of opportunity for self-gravitational corrections.
+
+The effect of mean molecular weight is also shown in \Fig{fig:tvsMcomp}.
+Lower $\mu$ values give longer growth times, and vice-versa.
+For $\mu = 2.0$, which represents the idealized case of an H$_2$ atmosphere devoid of Helium, $t_{\rm run}$ increases by factors of $\sim 2 - 3$.
+Thus fairly drastic changes in atmospheric composition are required for $\mu$ to significantly affect core accretion timescales.
+
+Looking ahead to our discussion of the critical core mass in \S\ref{sec:critcore}, \Fig{fig:tvsMcomp} shows that $t_{\rm run}$ equals a characteristic disk lifetime of 3 Myr for a core mass of  $M_{\rm crit} \approx 7 \, M_\oplus$  (for the standard case of $\mu = 2.35$).
+
+\begin{figure*}[tb]
+\centering
+\includegraphics[width=1.\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/TdPd_effect.pdf}
+\vspace{-0.3in}
+\caption{Crossover time as a function of disk temperature (\emph{left}) and pressure (\emph{right}) around a $M_{\rm c} = 5 M_{\oplus}$ core.
+The disk pressure or temperature (\emph{left} or \emph{right}, respectively) are fixed at values for 10 AU in our disk model.
+The analytic scalings are plotted for comparison, as described in the text.
+Gas accretion slows down significantly at higher temperatures, but only speeds up modestly as the disk pressure or density increase.}
+\label{fig:TPeffects}
+\end{figure*}
+
+\subsubsection{Disk Temperature and Pressure}
 \label{sec:TPeffects}
+
+\Fig{fig:TPeffects} shows how the crossover time varies with disk temperature, $T\di$, or pressure, $P\di$, holding the other quantity fixed.
+The analytic model roughly reproduces the temperature and pressure scalings, again with some discrepancies due mainly to the neglect of self-gravity.
+Temperature variations are much more significant than pressure variations (note the difference in logarithmic and linear axes).
+Since midplane disk conditions depend only on temperature and pressure in our model\footnote{See \Eq{eq:diskparam}.
+When gap opening is considered in models of later growth stages, the orbital frequency and effective viscosity become relevant as well.}, the dominant effect of disk location is temperature.
+
+The decline in growth times at colder temperatures arises from a variety of competing effects.
+Briefly, the larger Bondi radius and lower dust opacity at lower temperatures accelerate growth, overpowering the inherently lower cooling luminosity at lower temperatures.
+Growth times depend only weakly on, but do fall slightly with, pressure.
+The nearly exponential increase in pressure and density with depth through the radiative zone means that cooling, which is regulated at the RCB, has only a logarithmic dependence on disk pressure.
+Section \ref{sec:coolingan} explains these effects in more detail.
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_new.pdf}
+%\vspace{-0.5in}
+\caption{The critical core mass as a function of semi-major axis, for a disk lifetime $3$ Myrs and two values of the mean molecular weight ($\mu = 2.35$ is standard).
+The decline in $M_{\rm c,crit}$ with distance is a robust result for standard disk models.
+The analytic model, which neglects self-gravity, over-predicts the steepness of the decline.}
+\label{fig:Mcvsa}
+\end{figure}
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_opacity.pdf}
+%\vspace{-0.5in}
+\caption{Critical core masses vs.
+distance for standard and reduced (by a factor of ten) dust opacities.
+Lower opacities give significantly lower $M\crit$ values.
+Atmospheric opacity remains a large uncertainty in core accretion models.}
+\label{fig:Mcritopacity}
+\end{figure}
+
+\subsection{Critical Core Mass}
+\label{sec:critcore}
+
+The critical core mass declines with distance from the star, as shown in \Fig{fig:Mcvsa} for our standard disk model.
+The main reason for the decline, as explained above, is that atmospheres grow faster at colder temperatures.
+The lower densities and pressures in the outer disk have a much smaller effect.
+Since the vast majority of disk models have a disk temperature that declines with $a$, our qualitative result is robust.
+We also show that higher $\mu$ values give lower values of $M\crit$.
+
+The average power-law decline from 1 to 100 AU (not plotted) is $M\crit \propto a^{-0.3}$, for both choices of $\mu$.
+While not a drastic decline, the ability of distant low mass cores to accrete gas efficiently is significant for the interpretation of direct imaging surveys.
+However our model offers no guarantee of copious giant planets at large distances.
+Many histories of solid accretion are possible, and solid cores may grow too slowly to allow the rapid gas accretion that we model.
+
+Our non-self-gravitating analytic model (dashed curves in \Fig{fig:Mcvsa}) over-predicts the steepness of the decline in $M\crit$ vs.\ a.
+This discrepancy is not surprising, as we have shown that self-gravity strongly affects evolution, even before runaway growth begins and the crossover mass is reached.
+
+\Fig{fig:Mcritopacity} shows that reducing the opacity by an order of magnitude significantly reduces $M\crit$.
+Furthermore, this opacity effect is stronger at larger distances.
+The opacity reduction by a factor of 10 lowers $M\crit$ at 5 AU by a factor of $\sim 2.5$ and at 100 AU by a factor of $\sim 3.5$.
+Atmospheric opacity is a dominant uncertainty in core accretion modelling.
+However, unless atmospheric opacity varies significantly with disk radius, the general decline in $M\crit$ vs.\ $a$ should hold.
+For our scenario of negligible ongoing planetesimal accretion, it is tempting to think that dust could settle out of the radiative zone, lowering the opacity \citep{podolak03}.
+However, since opacity near the RCB is a crucial factor, we speculate that convective overshoot may prevent very low opacities.
 
@@
 Thus by restricting our attention to low masses, neglected hydrodynamic effects should be minor.
-Moreover since $M_{\rm th} \propto a^{6/7}$ increases with disk radius, spherical hydrostratic models like ours have a greater range of applicability in the outer regions of disks.
+Moreover since $M_{\rm th} \propto a^{6/7}$ increases with disk radius, spherical hydrostatic models like ours have a greater range of applicability in the outer regions of disks.
 
@@
 Grain growth also impacts the opacity in the outer regions as well as the existence or depth of radiative windows.
-We explore more realistic opacity laws in  In Piso et al.
+We explore more realistic opacity laws in Piso et al.
 (2013, in prep.).

### 2013-10-24 tex/draft/draft4.tex
section 6 edits

@@
 To solve the equation set (\ref{eq:struct}), an EOS is required for closure. In our study, we adopt an ideal gas law with a polytropic EOS 
-\begin{subeqnarray}
+\begin{subeqnarray}\label{eq:idealEOS}
 P &=& \rho \Rg T \, ,\slabel{eq:idealgas} \\
@@
 At yet higher masses accretion instabilities could occur \citep{AylBat12}.
-However, in this high mass regime we already know that the spherically symmetric approximation breaks down (see \S\ref{sec:scales}).
+However, in this high mass regime, the spherically symmetric approximation has already broken down.
 %, as does our approximate cooling model (\emph{discuss elsewhere and cite}).
@@
 \subsection{Realistic Opacities}\label{sec:op}
-%\emph{some comments on sublimation and radiative windows here (current text rough). upshot is the outer disk is good.}
-
-Real dust opacities exhibit a more complicated behavior that depends on grain composition.
+
+Treating dust (and total) opacities as a power law in temperature is a simplification.
 Opacities drop by order unity when ice grains sublimate for $T \gtrsim 150$ K and they drop by orders of magnitude when silicate grains evaporate above $T \gtrsim 1500$ K \citep{semenov03, FerAle05}.
-Our model assumes that radiative regions only exist in the outer part of the atmosphere.
-As such, even though protoplanetary atmospheres get significantly hotter than 1500 K, our model depends on opacity only in the upper radiative zones, which are cool enough to remain dusty.
-For our opacity approximations to be valid, we must restrict our model to the lower temperatures of the outer disk.
-
-A second radiative zone inside the convection zone, a ``radiative window," is possible, due to the opacity drops caused by the evaporation of ice and metal grains.
-Our two-layer model ignores such complications.
-Grain growth also impacts the opacity in the outer regions as well as the existence or depth of radiative windows.
-We explore more realistic opacity laws in Piso et al.
-(2013, in prep.).
-
-% we show that the existence of inner radiative windows does not impact our simplified model significantly, i.e.
-most of the luminosity is generated in the innermost convective region of the atmosphere and therefore the assumption that luminosity is constant in the radiative layers still holds.
+Grain growth and composition also affect opacities.
+Piso et al.
+(2013) explores more realistic opacity laws.
+
+For this work, we justify a simplified dust opacity by the cool temperatures of both the outer disk and our nearly isothermal radiative zones (see \S\ref{sec:profiles}).
+While  convective interiors get significantly hotter than 1500 K, opacity does not affect the structure of adiabatic convecting regions.
+A possible caveat is the existence of  radiative zones sandwiched inside the convection interior.
+Such ``radiative windows" arise if the opacity drop from ice and metal grain sublimation occurs at sufficiently low pressures.
+Our two-layer model ignores this possibility, but radiative windows are known to exist in hot Jupiter models \citep{burrows97, ab06} and have been seen in some core accretion models (Lissauer, personal communication).
+The role of radiative windows in core accretion models remains to be explored in detail.
 
@@
 
- In our model we assume an ideal gas law and a polytropic EOS, given by equations (\ref{eq:idealgas}) and (\ref{eq:polyEOS}), respectively.
-However, non-ideal effects such as partial dissociation and ionization have to be taken into account.
-One solution is using tabulated equation of state tables for hydrogen and helium mixtures.
-In Piso et al.
-(2013, in prep) we use the \citet{saumon95} EOS tables and extend them to lower pressures and temperatures as required by our disk assumptions.
+Our model uses an ideal gas law and a polytropic EOS, given by \Eq{eq:idealEOS}.
+However, non-ideal effects can affect atmospheric structure and evolution.
+In the lower atmosphere, H$_2$ can partially dissociate at high temperatures.
+In the upper atmosphere, H$_2$ rotational levels can become depopulated at lower temeratures.
+Piso et al.
+(in prep) uses the \citet{saumon95} EOS, with extensions to lower pressures and temperatures as needed, to explore these effects.
 

### 2013-10-24 tex/draft/draft4.tex
minor section 5 edits

@@
 We now use our structure and evolution models to calculate the necessary conditions for gas giant formation by core accretion.
-Adopting a median disk lifetime of 3 Myr, we examine the conditions for a fixed mass core to to undergo runaway gas accretion within the lifetime of the gas disk.
+Adopting a median disk lifetime of 3 Myr, we examine the conditions for a fixed mass core to undergo runaway gas accretion within the lifetime of the gas disk.
 Our results for atmosphere growth times -- for a range of core masses and disk conditions -- are presented in \S\ref{sec:tcross}.
@@
 %\vspace{-0.5in}
-\caption{Time to grow an atmosphere of mass $M_{\rm{atm}}$ for cores with fixed masses between $5 M_{\oplus}$ and $14 M_{\oplus}$ (as labelled) at $10$ AU in our fiducial disk.
+\caption{Time to grow an atmosphere of mass $M_{\rm{atm}}$ for cores with fixed masses between $5 M_{\oplus}$ and $14 M_{\oplus}$ (as labeled) at $10$ AU in our fiducial disk.
 Circles mark the runaway growth time, $t_{\rm run}$, which occurs at roughly the crossover mass, $M_{\rm{atm}} = M_{\rm c}$.
@@
 %\vspace{-0.5in}
-\caption{Runaway growth times, $t_{\rm{run}}$, vs.\ core mass at $10$ AU, for two values of mean molecular weight.
+\caption{Runaway growth time, $t_{\rm{run}}$, vs.\ core mass at $10$ AU, for two values of the mean molecular weight.
 Our numerical model (\emph{solid curves}) is compared to our non-self-gravitating analytic model (\emph{dashed curves}).
 A typical protoplanetary disk life time of $3$ Myr is plotted for comparison.
-The crossover time is larger for a lower mean molecular weight.}
+The runaway growth time is larger for a lower mean molecular weight.}
 \label{fig:tvsMcomp}
@@
 The endpoint of each curve marks $t_{\rm run}$.
-Runaway growth occurs near the crossover mass, when $M_{\rm atm} \sim M_{\rm co}$ in agreement with previous studies.
+Runaway growth occurs near the crossover mass, when $M_{\rm atm} \sim M\co$, in agreement with previous studies.
 At lower core masses, however, runaway occurs at lower $M_{\rm atm}/M\co$ ratios.
@@
 \vspace{-0.3in}
-\caption{Crossover time as a function of disk temperature (\emph{left}) and pressure (\emph{right}) around a $M_{\rm c} = 5 M_{\oplus}$ core.
+\caption{Runaway growth time as a function of disk temperature (\emph{left}) and pressure (\emph{right}) around a $M_{\rm c} = 5 M_{\oplus}$ core.
 The disk pressure or temperature (\emph{left} or \emph{right}, respectively) are fixed at values for 10 AU in our disk model.
@@
 
-\Fig{fig:TPeffects} shows how the crossover time varies with disk temperature, $T\di$, or pressure, $P\di$, holding the other quantity fixed.
+\Fig{fig:TPeffects} shows how the runaway accretion time varies with disk temperature, $T\di$, or pressure, $P\di$, holding the other quantity fixed.
 The analytic model roughly reproduces the temperature and pressure scalings, again with some discrepancies due mainly to the neglect of self-gravity.
@@
 %\vspace{-0.5in}
-\caption{The critical core mass as a function of semi-major axis, for a disk lifetime $3$ Myrs and two values of the mean molecular weight ($\mu = 2.35$ is standard).
+\caption{The critical core mass as a function of semi-major axis, for a disk lifetime of $3$ Myrs and two values of the mean molecular weight ($\mu = 2.35$ is standard).
 The decline in $M_{\rm c,crit}$ with distance is a robust result for standard disk models.
@@
 While not a drastic decline, the ability of distant low mass cores to accrete gas efficiently is significant for the interpretation of direct imaging surveys.
-However our model offers no guarantee of copious giant planets at large distances.
+However, our model offers no guarantee of copious giant planets at large distances.
 Many histories of solid accretion are possible, and solid cores may grow too slowly to allow the rapid gas accretion that we model.

### 2013-10-25 tex/draft/draft4.tex
summary edits

--- 2013-10-24 tex/draft/draft4.tex
@@
 
-%In this paper we study the formation of giant planets through core accretion.
-
-In this paper we build evolutionary models of giant planets forming in a protoplanetary disk.
-We consider atmosphere growth around a core of fixed mass, and determine the minimum core mass for which unstable atmosphere collapse occurs before the dissipation of the protoplanetary disk, at various locations in the disk.
-For a standard ISM power-law opacity and a polytropic EOS for the nebular gas, the minimum core mass is $\sim 8 M_{\oplus}$ at 5 AU and decreases to $\sim 3.5 M_{\oplus}$ at 100 AU.
-A larger mean molecular weight of the nebular gas increases the critical core mass, while a factor of ten reduction in opacity results in a critical core mass twice as low.
-Our results are applicable in the outer regions of the disk ($\gtrsim 5$ AU), where planets can undergo runaway gas accretion before opening a gap in the disk.
-If follows that core accretion at wide separations may result in massive gas giants, which could explain the formation of some of the directly imaged giant planets (\citealt{marois08}; \citealt{kalas08}; \citealt{lagrange10}).
-
-Our results are lower than the typically quoted critical core mass of 10 $M_{\oplus}$ in static studies.
-A giant planet can thus form faster if the core grows first, then accretion of solids slows down and a massive atmosphere accumulates, rather than if the core and envelope grow simultaneously in a high planetesimal accretion rate regime.
-We note that our model is a limiting case of core accretion, in which planetesimal accretion is negligible and the evolution of the gaseous envelope is dominated by KH contraction.
-We are thus able to place robust absolute lower limits on the core mass needed to form a giant planet, since additional core growth and heating limit the ability of the atmosphere to radiate energy and cool.
-Our results are not inconsistent with the existence of ice giants at large distances (such as Uranus and Neptune in our own Solar System), as ongoing accretion of solids can stall atmospheric growth and prevent the accumulation of a massive envelope before disk dissipation.
-
-% In this paper we study the formation of giant planet atmospheres under the assumption of a low planetesimal accretion regime in which KH contraction dominates the luminosity evolution of the atmosphere.
-We consider atmosphere growth around a core of fixed mass, and we build quasi-static two-layer atmosphere models with an inner convective region and an outer radiative region that matches smoothly onto the protoplanetary disk at the Hill radius.
-In our static models, we use an ideal gas polytropic EOS, and assume a standard ISM power-law opacity for dust grains.
-We derive a cooling model to connect series of quasti-static atmospheres and obtain an evolutionary history of the envelope.
-We define the time $t$ at which unstable atmosphere collapse commences as the \textit{crossover time} $t_{\rm{co}}$, for which $M_{\rm atm}(t_{\rm{co}})\sim M_{\rm c}$.
-From this we define as \textit{critical core mass} the minimum core mass for a protoplanet to initiate runaway gas accretion during the lifetime of the protoplanetary disk.
-We calculate the critical core mass for a variety of disk conditions, nebular gas compositions and opacities.
-The critical core mass is $\sim 8 M_{\oplus}$ at 5 AU in our fiducial disk and decreases to $\sim 3.5 M_{\oplus}$ at 100 AU.
-A larger mean molecular weight of the nebular gas increases the critical core mass, while a factor of ten reduction in opacity results in a critical core mass twice as low.
-% 
-%Our results are lower than the typically quoted critical core mass of 10 $M_{\oplus}$ in static studies.
-A giant planet can thus form faster if the core grows first, then accretion of solids slows down and a massive atmosphere accumulates, rather than if the core and envelope grow simultaneously in a high planetesimal accretion rate regime.
-Moreover, we are able to place robust absolute lower limits on the core mass needed to form a giant planet, since additional core growth and heating limit the ability of the atmosphere to radiate energy and cool.
-
- %We also find that the critical core mass to form a giant planet within the life time of the disk (of about 3 Myrs) is smaller than the typical 10 $M_{\oplus}$ value yielded by studies that assume that the atmosphere evolution is dominated by the luminosity due to planetesimal accretion.
-As such, it is faster to form a planet by growing the core first in a fast planetesimal accretion regime (e.g., the core forms in the inner disk, then migrates outwards), then significantly reduce planetesimal accretion and allow a massive atmosphere to accumulate.
+We study the formation of giant planets by the core accretion mechanism.
+Our models start with a solid core that is embedded in a gas disk and no longer accreting solids.
+We determine -- as a function of disk location, core mass, and the atmosphere's mean molecular weight and opacity -- whether runaway atmospheric growth can occur within a typical disk lifetime of 3 Myr.
+By neglecting the accretion luminosity of planetesimals and smaller solids, we obtain the fastest allowed rate of gas accretion.
+Of relevance to direct maging surveys, we address core accretion in the outer disk.
+Our model approximations, including spherical accretion and low mass radiative exteriors with opacities dominated by dust, are tuned to conditions in the outer disk.
+Our main findings are as follows:
+
+\begin{enumerate}
+\item The minimum or critical core mass, $M\crit$, for giant planet formation declines with distance in standard protoplanetary disk models.
+For our reference case, the critical mass is $\sim 8.5 M_{\oplus}$ at $a = 5$ AU, decreasing to $\sim 3.5 M_{\oplus}$ at $a =100$ AU.
+This decline roughly follows $M\crit \propto a^{-0.3}$.
+
+\item The drop in disk temperature with radial distance explains the decrease in critical core masses.  The lower pressures and densities in the outer disk only weakly suppress atmospheric growth.
+
+\item Reducing dust opacities by a factor of 10 reduces critical core masses by a factor of $\sim 3$.  This reduction is somewhat stronger (weaker) at larger (smaller) separations from the star.
+
+\item A larger mean molecular weight reduces critical core masses, in agreement with \citet{HorIko11}.
+If enrichment in heavy elements correlates with increased dust opacity, then the stronger opacity effect will dominate, increasing $M\crit$.
+
+\item Runaway growth begins roughly at the crossover mass, when atmosphere and core masses are equal, in agreement with previous work \citep{pollack96}.
+For larger core masses, runway growth begins at larger ratios of atmosphere to core mass.
+
+\item Self-gravity affects atmospheric evolution before crossover.  Significant self-gravitational corrections appear when the atmosphere is only $\sim 20 \%$ as massive as the core.
+
+\end{enumerate}
+
+Rapid gas accretion onto low mass cores could explain the origin of distant directly imaged giant planets \citep{marois08, lagrange10}.
+However our model does address the details of how solid cores grow, and many possibilities exist.
+For a giant planet to form with a core near the minimum mass, the core's growth must  first be rapid and then slow significantly, as in phases I and II, respectively, of \cite{pollack96}.
+
+Initial core growth must be fast, compared to the disk lifetime, to get a sufficiently massive core.
+Subsequent slowing of core growth is required to allow the atmosphere to cool and contract.
+To be more quantitative, the minimum cooling luminosity in \Fig{fig:Ltplot}, $L\approx 3.5 \times 10^{24}\;\mathrm{erg s}^{-1}$, could be cancelled by low levels of heating from solid accretion.
+Specifically a core mass doubling timescale of $\sim 400$ Myr would provide enough heating to stall atmospheric cooling and growth.
+
+Thus for giant planets to form around around cores near $M\crit$, core growth must stall quite severely.
+The accretion and scattering of local planetesimals, combined with the inward migration of smaller solids past the core, offers a plausible means to sufficiently isolate the core.
+On the other hand, non-negligible accretion of solids onto cores could continue for the lifetime of the gas disk.
+In these cases, relatively high mass cores would remain as ice giants, akin to Uranus and Neptune.
+Ongoing direct imaging surveys and their successors \citep{hinz12, macintosh12} will help discriminate among the various planet formation pathways in the outskirts of protoplanetary disks.
 

### 2013-10-25 tex/draft/draft4.tex
section 5 edits + very minor abstract and section 6 edits

@@
 For different locations in the disk, we determine the minimum core mass for a giant planet to form within a typical disk lifetime of 3 Myr.
-The minimum core mass required to nucleate atmosphere collapse is $\sim 8 M_{\oplus}$ at 5 AU and steadily decreases to $\sim 3.5 M_{\oplus}$ at 100 AU, for a solar composition atmosphere with a standard ISM dust opacity.
+The minimum core mass required to nucleate atmosphere collapse is $\sim$$8.5 M_{\oplus}$ at 5 AU and steadily decreases to $\sim$$3.5 M_{\oplus}$ at 100 AU, for a Solar composition atmosphere with a standard ISM dust opacity.
 This trend is mainly explained by the cooler temperatures in the outer disk.
-Lower pressure and densities in the outer disk have a negligible effect on core mass requirements.
+Lower pressures and densities in the outer disk have a negligible effect on core mass requirements.
 At all distances, decreasing the dust opacity and/or increasing the mean molecular weight of the atmosphere reduces the critical core mass.
@@
 We also develop a non-self-gravitating, analytic cooling model to help explain our numerical trends.
-Additionally, this comparison reveals that self-gravity significantly affects atmospheric evolution even when the atmosphere is only $\sim 20\%$ as massive as the core.
+Additionally, this comparison reveals that self-gravity significantly affects atmospheric evolution even when the atmosphere is only $\sim$$20\%$ as massive as the core.
 
@@
 Our results for atmosphere growth times -- for a range of core masses and disk conditions -- are presented in \S\ref{sec:tcross}.
-Section \ref{sec:critcore} gives our results for critical core masses, the minimum masses which can accrete a massive atmosphere within the disk lifetime.
-
-Our models will focus on giant planet formation between 5 and 100 AU.
-The outer disk is of particular interest for direct imaging searches.
+Section \ref{sec:critcore} gives our results for critical core masses, the minimum masses needed to accrete a massive atmosphere within the disk lifetime.
+
+Our models focus on giant planet formation between 5 and 100 AU, as the outer disk is of particular interest for direct imaging searches.
 The growth of atmospheres close to the star is also important, but spherical accretion models (including ours) are less applicable here.
@@
 \label{sec:tcross}
-The time to undergo runaway growth of the atmosphere, $t_{\rm run}$, sets a minimum timescale for the formation of giant planets.
+The time to undergo atmospheric runaway growth, $t_{\rm run}$, sets a minimum timescale for the formation of giant planets.
 Due to the accelerating nature of runaway growth, the precise threshold chosen for $t_{\rm run}$ (explained in \S ?) is of minor significance.
@@
 Circles mark the runaway growth time, $t_{\rm run}$, which occurs at roughly the crossover mass, $M_{\rm{atm}} = M_{\rm c}$.
-Larger core masses give faster atmospheric growth, to a fixed atmosphere mass and also to runaway.
-However, for larger $M\co$, runaway growth commences at higher $M_{\rm atm}/M\co$ values.}
+Both the time to reach a fixed atmosphere mass and the runaway growth time are shorter for larger cores.
+For larger $M\co$, runaway growth commences at higher $M_{\rm atm}/M\co$ values.}
 \label{fig:tvsM}
@@
 \caption{Runaway growth time, $t_{\rm{run}}$, vs.\ core mass at $10$ AU, for two values of the mean molecular weight.
-Our numerical model (\emph{solid curves}) is compared to our non-self-gravitating analytic model (\emph{dashed curves}).
+Our numerical model (\emph{solid curves}) is compared to our non-self-gravitating analytic model (\emph{dashed curves}, with $t_{\rm run}$ as prescribed by Equation \ref{eq:tcoolanf}).
 A typical protoplanetary disk life time of $3$ Myr is plotted for comparison.
@@
 
-\Fig{fig:tvsM} demonstrates the growth of atmospheric mass with time for several core masses at $10$ AU in our fiducial disk.
+\Fig{fig:tvsM} shows the growth of atmospheric mass with time for several core masses at $10$ AU in our fiducial disk.
 Atmospheres grow faster around more massive cores due to stronger gravitational binding.
@@
 Runaway growth occurs near the crossover mass, when $M_{\rm atm} \sim M\co$, in agreement with previous studies.
-At lower core masses, however, runaway occurs at lower $M_{\rm atm}/M\co$ ratios.
-Thus, when compared to the naive assumption that runaway occurs at a universal $M_{\rm atm}/M\co$ ratio,  effects that lower the critical core mass are amplified by the earlier onset of runaway growth.
+Lower core masses undergo runaway accretion at fractionally smaller atmosphere masses.
+%Thus, when compared to the assumption that runaway occurs at a universal $M_{\rm atm}/M\co$ ratio,  effects that lower the critical core mass are amplified by the earlier onset of runaway growth.
 
@@
 The effect of mean molecular weight is also shown in \Fig{fig:tvsMcomp}.
-Lower $\mu$ values give longer growth times, and vice-versa.
+A lower $\mu$ gives longer growth times, as it results in a smaller atmosphere scale height and thus a weaker pressure gradient to oppose self-gravity, which accelerates growth.
 For $\mu = 2.0$, which represents the idealized case of an H$_2$ atmosphere devoid of Helium, $t_{\rm run}$ increases by factors of $\sim 2 - 3$.
@@
 
-Looking ahead to our discussion of the critical core mass in \S\ref{sec:critcore}, \Fig{fig:tvsMcomp} shows that $t_{\rm run}$ equals a characteristic disk lifetime of 3 Myr for a core mass of  $M_{\rm crit} \approx 7 \, M_\oplus$  (for the standard case of $\mu = 2.35$).
+%Looking ahead to our discussion of the critical core mass in \S\ref{sec:critcore}, \Fig{fig:tvsMcomp} shows that $t_{\rm run}$ equals a characteristic disk lifetime of 3 Myr for a core mass of  $M_{\rm crit} \approx 7 \, M_\oplus$  (for the standard case of $\mu = 2.35$).
 
@@
 The disk pressure or temperature (\emph{left} or \emph{right}, respectively) are fixed at values for 10 AU in our disk model.
-The analytic scalings are plotted for comparison, as described in the text.
+The analytic scalings given by \Eq{eq:tcoolanf} are plotted for comparison, as described in the text.
 Gas accretion slows down significantly at higher temperatures, but only speeds up modestly as the disk pressure or density increase.}
@@
 
-\subsubsection{Disk Temperature and Pressure}
+\subsubsection{Effects of Disk Temperature and Pressure}
 \label{sec:TPeffects}
@@
 
-The decline in growth times at colder temperatures arises from a variety of competing effects.
+The decline in growth times at lower temperatures arises from a variety of competing effects.
 Briefly, the larger Bondi radius and lower dust opacity at lower temperatures accelerate growth, overpowering the inherently lower cooling luminosity at lower temperatures.
@@
 %\vspace{-0.5in}
-\caption{The critical core mass as a function of semi-major axis, for a disk lifetime of $3$ Myrs and two values of the mean molecular weight ($\mu = 2.35$ is standard).
+\caption{The critical core mass as a function of semi-major axis, for a disk lifetime of $3$ Myrs and two values of the mean molecular weight ($\mu = 2.35$ is standard for a Hydrogen-Helium Solar mixture).
 The decline in $M_{\rm c,crit}$ with distance is a robust result for standard disk models.
@@
 The critical core mass declines with distance from the star, as shown in \Fig{fig:Mcvsa} for our standard disk model.
-The main reason for the decline, as explained above, is that atmospheres grow faster at colder temperatures.
+The main reason for the decline, as explained above, is that atmospheres grow faster at lower temperatures.
 The lower densities and pressures in the outer disk have a much smaller effect.
@@
 
-Our non-self-gravitating analytic model (dashed curves in \Fig{fig:Mcvsa}) over-predicts the steepness of the decline in $M\crit$ vs.\ a.
+Our non-self-gravitating analytic model (dashed curves in \Fig{fig:Mcvsa}) over-predicts the steepness of the decline in $M\crit$ with $a$.
 This discrepancy is not surprising, as we have shown that self-gravity strongly affects evolution, even before runaway growth begins and the crossover mass is reached.
@@
 Furthermore, this opacity effect is stronger at larger distances.
-The opacity reduction by a factor of 10 lowers $M\crit$ at 5 AU by a factor of $\sim 2.5$ and at 100 AU by a factor of $\sim 3.5$.
+The opacity reduction by a factor of 10 lowers $M\crit$ at 5 AU by a factor of $\sim$$2.5$ and at 100 AU by a factor of $\sim$$3.5$.
 Atmospheric opacity is a dominant uncertainty in core accretion modelling.
-However, unless atmospheric opacity varies significantly with disk radius, the general decline in $M\crit$ vs.\ $a$ should hold.
+However, unless atmospheric opacity varies significantly with disk radius, the general decline in $M\crit$ with increasing $a$ should hold.
 For our scenario of negligible ongoing planetesimal accretion, it is tempting to think that dust could settle out of the radiative zone, lowering the opacity \citep{podolak03}.
-However, since opacity near the RCB is a crucial factor, we speculate that convective overshoot may prevent very low opacities.
+Nonetheless, since opacity near the RCB is a crucial factor, we speculate that convective overshoot may prevent very low opacities.
 
@@
 Piso et al.
-(2013) explores more realistic opacity laws.
+(in prep.) explores more realistic opacity laws.
 

### 2013-10-25 tex/draft/draft4.tex
minor section 7 typo fixes

@@
 By neglecting the accretion luminosity of planetesimals and smaller solids, we obtain the fastest allowed rate of gas accretion.
-Of relevance to direct maging surveys, we address core accretion in the outer disk.
+Of relevance to direct imaging surveys, we address core accretion in the outer disk.
 Our model approximations, including spherical accretion and low mass radiative exteriors with opacities dominated by dust, are tuned to conditions in the outer disk.
@@
 \item The minimum or critical core mass, $M\crit$, for giant planet formation declines with distance in standard protoplanetary disk models.
-For our reference case, the critical mass is $\sim 8.5 M_{\oplus}$ at $a = 5$ AU, decreasing to $\sim 3.5 M_{\oplus}$ at $a =100$ AU.
+For our reference case, the critical mass is $\sim$$8.5 M_{\oplus}$ at $a = 5$ AU, decreasing to $\sim$$3.5 M_{\oplus}$ at $a =100$ AU.
 This decline roughly follows $M\crit \propto a^{-0.3}$.
@@
 
-\item Reducing dust opacities by a factor of 10 reduces critical core masses by a factor of $\sim 3$.  This reduction is somewhat stronger (weaker) at larger (smaller) separations from the star.
+\item Reducing dust opacities by a factor of 10 reduces critical core masses by a factor of $\sim$$3$.  This reduction is somewhat stronger (weaker) at larger (smaller) separations from the star.
 
@@
 \item Runaway growth begins roughly at the crossover mass, when atmosphere and core masses are equal, in agreement with previous work \citep{pollack96}.
-For larger core masses, runway growth begins at larger ratios of atmosphere to core mass.
-
-\item Self-gravity affects atmospheric evolution before crossover.  Significant self-gravitational corrections appear when the atmosphere is only $\sim 20 \%$ as massive as the core.
+For larger core masses, runaway growth begins at larger ratios of atmosphere to core mass.
+
+\item Self-gravity affects atmospheric evolution before crossover.  Significant self-gravitational corrections appear when the atmosphere is only $\sim$$20 \%$ as massive as the core.
 
@@
 Rapid gas accretion onto low mass cores could explain the origin of distant directly imaged giant planets \citep{marois08, lagrange10}.
-However our model does address the details of how solid cores grow, and many possibilities exist.
+However, our model does not address the details of how solid cores grow, and many possibilities exist.
 For a giant planet to form with a core near the minimum mass, the core's growth must  first be rapid and then slow significantly, as in phases I and II, respectively, of \cite{pollack96}.
@@
 To be more quantitative, the minimum cooling luminosity in \Fig{fig:Ltplot}, $L\approx 3.5 \times 10^{24}\;\mathrm{erg s}^{-1}$, could be cancelled by low levels of heating from solid accretion.
-Specifically a core mass doubling timescale of $\sim 400$ Myr would provide enough heating to stall atmospheric cooling and growth.
-
-Thus for giant planets to form around around cores near $M\crit$, core growth must stall quite severely.
+Specifically, a core mass doubling timescale of $\sim$$400$ Myr would provide enough heating to stall atmospheric cooling and growth.
+
+Thus for giant planets to form around cores near $M\crit$, core growth must stall quite severely.
 The accretion and scattering of local planetesimals, combined with the inward migration of smaller solids past the core, offers a plausible means to sufficiently isolate the core.

### 2013-10-25 tex/draft/draft4.tex
minor edits to Ana's edits

@@
 \caption{Time to grow an atmosphere of mass $M_{\rm{atm}}$ for cores with fixed masses between $5 M_{\oplus}$ and $14 M_{\oplus}$ (as labeled) at $10$ AU in our fiducial disk.
-Circles mark the runaway growth time, $t_{\rm run}$, which occurs at roughly the crossover mass, $M_{\rm{atm}} = M_{\rm c}$.
+Circles mark the runaway growth time, $t_{\rm run}$, which occurs at roughly the crossover mass, $M_{\rm{atm}} = M\co$.
 Both the time to reach a fixed atmosphere mass and the runaway growth time are shorter for larger cores.
@@
 The effect of mean molecular weight is also shown in \Fig{fig:tvsMcomp}.
-A lower $\mu$ gives longer growth times, as it results in a smaller atmosphere scale height and thus a weaker pressure gradient to oppose self-gravity, which accelerates growth.
-For $\mu = 2.0$, which represents the idealized case of an H$_2$ atmosphere devoid of Helium, $t_{\rm run}$ increases by factors of $\sim 2 - 3$.
+A lower $\mu$ gives longer growth times, because more cooling is required to compress the atmosphere.
+This effect is explained by the increase in the atmospheric scale height $\propto 1/\mu$ while the Bondi radius decreases, $\RB \propto \mu$.
+For $\mu = 2.0$, which represents the idealized case of an H$_2$ atmosphere completely devoid of Helium, $t_{\rm run}$ increases by factors of $\sim 2 - 3$.
 Thus fairly drastic changes in atmospheric composition are required for $\mu$ to significantly affect core accretion timescales.
@@
 %\vspace{-0.5in}
-\caption{The critical core mass as a function of semi-major axis, for a disk lifetime of $3$ Myrs and two values of the mean molecular weight ($\mu = 2.35$ is standard for a Hydrogen-Helium Solar mixture).
+\caption{The critical core mass as a function of semi-major axis, for a disk lifetime of $3$ Myrs and two values of the mean molecular weight ($\mu = 2.35$ is for Solar abundances).
 The decline in $M_{\rm c,crit}$ with distance is a robust result for standard disk models.
@@
 To be more quantitative, the minimum cooling luminosity in \Fig{fig:Ltplot}, $L\approx 3.5 \times 10^{24}\;\mathrm{erg s}^{-1}$, could be cancelled by low levels of heating from solid accretion.
-Specifically, a core mass doubling timescale of $\sim$$400$ Myr would provide enough heating to stall atmospheric cooling and growth.
+Specifically, a core mass doubling timescale of $\sim$$400$ Myr, or faster, would provide enough heating to stall atmospheric cooling and growth.
 
 Thus for giant planets to form around cores near $M\crit$, core growth must stall quite severely.
-The accretion and scattering of local planetesimals, combined with the inward migration of smaller solids past the core, offers a plausible means to sufficiently isolate the core.
+The accretion and scattering of local planetesimals, combined with the inward aerodynamic drift of smaller solids past the core, offers a plausible means to sufficiently isolate the core.
 On the other hand, non-negligible accretion of solids onto cores could continue for the lifetime of the gas disk.

### 2013-10-25 tex/draft/draft4.tex
minor edits

@@
 For different locations in the disk, we determine the minimum core mass for a giant planet to form within a typical disk lifetime of 3 Myr.
-The minimum core mass required to nucleate atmosphere collapse is $\sim$$8.5 M_{\oplus}$ at 5 AU and steadily decreases to $\sim$$3.5 M_{\oplus}$ at 100 AU, for a Solar composition atmosphere with a standard ISM dust opacity.
+The minimum core mass required to nucleate atmosphere collapse is $\sim$$8.5 M_{\oplus}$ at 5 AU and steadily decreases to $\sim$$3.5 M_{\oplus}$ at 100 AU, for a Solar composition atmosphere with a standard dust opacity.
 This trend is mainly explained by the cooler temperatures in the outer disk.
@@
 
-We assume a dust opacity following \citet{bell94}:
+We assume a dust opacity characteristic of interstellar grains, following \citet{bell94}:
 \begin{equation}
@@
 %\vspace{-0.5in}
-\caption{The critical core mass as a function of semi-major axis, for a disk lifetime of $3$ Myrs and two values of the mean molecular weight ($\mu = 2.35$ is for Solar abundances).
+\caption{The critical core mass as a function of semimajor axis, for a disk lifetime of $3$ Myrs and two values of the mean molecular weight ($\mu = 2.35$ is for Solar abundances).
 The decline in $M_{\rm c,crit}$ with distance is a robust result for standard disk models.
@@
 By neglecting the accretion luminosity of planetesimals and smaller solids, we obtain the fastest allowed rate of gas accretion.
-Of relevance to direct imaging surveys, we address core accretion in the outer disk.
+We address core accretion in the outer disk, as it is relevant to direct imaging surveys.
 Our model approximations, including spherical accretion and low mass radiative exteriors with opacities dominated by dust, are tuned to conditions in the outer disk.
@@
 \begin{enumerate}
-\item The minimum or critical core mass, $M\crit$, for giant planet formation declines with distance in standard protoplanetary disk models.
+\item The minimum or critical core mass, $M\crit$, for giant planet formation declines with stellocentric distance in standard protoplanetary disk models.
 For our reference case, the critical mass is $\sim$$8.5 M_{\oplus}$ at $a = 5$ AU, decreasing to $\sim$$3.5 M_{\oplus}$ at $a =100$ AU.
@@
 This result is the leading order term in a series expansion.
-By comparing the expression above and \Eq{eq:Matman} under the assumption that $R\cb \ll R_{\rm B}$, we see that the mass of the outer radiative region (which is nearly isothermal) is negligible when compared with the atmosphere mass in the convective layer, as stated in section \S\ref{sec:coolingan}.
+By comparing the expression above and \Eq{eq:Matman} under the assumption that $R\cb \ll R_{\rm B}$, we see that the mass of the outer radiative region (which is nearly isothermal) is negligible when compared with the atmosphere mass in the convective layer, as stated in Section \ref{sec:coolingan}.
 % Furthermore, because the atmospheric scale-height at $R\co$ is $H_\rho = |dr /d\ln\rho| = R\co^2/R_B$, the result is intuitively the correct order of magnitude.

### 2013-10-25 tex/draft/draft4.tex
section 4 edits

--- 2013-10-25 tex/draft/draft4.tex
@@
 
-We now present some representative examples of the structure and evolution of our model atmospheres, calculated as described in \S\ref{sec:model}, with some comparisons made to the non-self-gravitating analytic model of \S\ref{sec:coolingan}.
-Radial structure is presented in \S\ref{sec:profiles}, time evolution is described in \S\ref{sec:timeev}, and the regimes of validity of our two-layer cooling model are examined in \S\ref{sec:endoftime}.
+We present examples of the structure and evolution of our model atmospheres, calculated as described in \S\ref{sec:model}, with some comparisons to the non-self-gravitating analytic model of \S\ref{sec:coolingan}.
+Radial structure is presented in \S\ref{sec:profiles}, time evolution is described in \S\ref{sec:timeev}, and validity of our two-layer cooling model is examined in \S\ref{sec:endoftime}.
 
@@
 %\vspace{-0.5in}
-\caption{Radial profiles of atmospheric pressure, temperature and enclosed mass (including the core) for a $5 M_{\oplus}$ core at $a=60$ AU.
+\caption{Radial profiles of atmospheric pressure, temperature and enclosed mass (core included) for a $5 M_{\oplus}$ core at $a=60$ AU.
 Solid, dotted, dashed and dot-dashed lines correspond to solutions with total mass (core and atmosphere) of $5.01 M_{\oplus}$, $5.10 M_{\oplus}$, $6.00 M_{\oplus}$ and $9.24 M_{\oplus}$, respectively.
@@
 \label{sec:profiles}
-Figure \ref{fig:profiles} shows radial profiles at different stages of atmospheric growth.
-In these examples, we fix the core mass at $M_{\rm{c}}=5 M_{\oplus}$ and the radial disk location to $a=60$ AU.
-The quoted mass values include the core plus atmosphere within the smaller of $\RB$ or $\RH$, which for these cases is $\RB$.
+Figure \ref{fig:profiles} shows radial profiles at different stages of atmospheric growth around $5 M_{\oplus}$ core at $60$ AU.
+Quoted mass values include the core plus atmosphere within the smaller of $\RB$ or $\RH$, which for these cases is $\RB$.
 The 9.24 $M_{\oplus}$ solution is the highest mass we can reach in our evolutionary sequence, as we explain further in \S\ref{sec:endoftime}.
@@
 
-The detailed structure of these profiles is well approximated by our non-self-gravitating, analytic solutions.
-Deep in convective zones, \Eq{eq:deep} gives that $T \propto r^{-1}$ and $P\propto r^{-1/\delad}$.
+The atmospheric structure is well approximated by our non-self-gravitating, analytic solutions.
+Deep in convective zones, \Eq{eq:deep} gives $T \propto r^{-1}$ and $P\propto r^{-1/\delad}$.
 This behavior is seen in the low mass solutions in \Fig{fig:profiles}.
@@
 This behavior is seen in both the full numerical solutions (with solid lines) and the analytic model (with dashed lines).
-The increasing pressure depth of the RCB with mass causes increases in the optical depth ($\propto \kappa P)$ and decreases in the radiative luminosity.
+With increasing atmospheric mass, the pressure depth of the RCB increases, along with the optical depth ($\propto \kappa P\cb)$.
+Consequently, the radiative luminosity decreases.
 This behavior is described in \Eqs{eq:crit}{eq:Lcb}.
@@
 By contrast, the non-self-gravitating analytic model gives a luminosity that continues to drop with increasing mass.
-To understand this difference, we note that the scaling of \Eq{eq:Lcb}, $L\cb \propto M\cb T\cb^4/ (\kappa\cb  P\cb)$, holds in both cases; thus accounting for the higher enclosed mass in the self-gravitating model gives a somewhat higher luminosity, as desired.
+To understand this difference, consider the scaling of \Eq{eq:Lcb}, $L\cb \propto M\cb T\cb^4/ (\kappa\cb  P\cb)$, which holds in both cases.
+Accounting for the higher enclosed mass in the self-gravitating model gives a somewhat higher luminosity, as desired.
 However, the main effect is that \Eq{eq:crit} -- which describes a nearly linear relation between atmospheric mass and RCB pressure --  breaks down for self-gravitating solutions.
 This behavior can be seen in the top panel of \Fig{fig:profiles} where the $P\cb$ increases significantly from 5.10 to 6.0 $M_\oplus$, but only increases relatively modestly with further growth to 9.24 $M_\oplus$.
-To allow these relatively lower $P\cb$ values (and thus higher luminosities) in the higher mass solutions, the radius $R\cb$ must shift outwards, as shown in \Fig{fig:profiles}.
+In the higher mass solutions, the relatively low $P\cb$ values (and thus the relatively high luminosities) require an outward shift in $R\cb$, as shown in \Fig{fig:profiles}.
 
@@
 This instantaneous growth time shows clearly that the atmosphere spends the bulk of its time growing though intermediate atmospheric masses, $\sim 1 -3$ $M_\oplus$ in this case.
-Growth times are short both early -- when the radiative zone is relatively transparent -- and late -- when self-gravity accelerates growth.
-The accelerated growth at higher masses is also evident in \Figs{fig:Ltplot}{fig:LtvsMopacity} (bottom panels) as the flattening of the $t$ vs.\ $M_{\rm atm}$ curve.
-
-The fact that growth times peak when $M_{\rm{atm}} < M_{\rm c}$, before the crossover mass, is helpful for understanding atmospheric growth.
+Growth times are short both early -- when the radiative zone is transparent -- and late -- when self-gravity accelerates growth.
+
+The fact that growth times have a well defined maximum is a characteristic of runaway growth.
+Unlike our analytic model, which must assume that runaway growth begins near the crossover mass, our numerical model allows us to measure when runaway growth begins.
+Runaway growth clearly does not begin at a universal value of $M_{\rm atm}/M\co$.
+\Fig{fig:growthtime} shows that further from the star runaway growth begins at smaller $M_{\rm atm}/M\co$.
+\Fig{fig:tvsM} (described in the next section) shows how the onset of runaway growth depends on core mass.
+
+We quantify the runaway growth timescale, $t_{\rm run}$, as the time when $M_{\rm atm}/\dot{M}$ drops to 10\% of its maximum value.
+The choice of 10\% is arbitrary; the precise threshold chosen is relatively unimportant because growth continues to accelerate.
+
+%The onset of runaway growth at high masses is also evident in \Figs{fig:Ltplot}{fig:LtvsMopacity} (bottom panels) as the flattening of the $t$ vs.\ $M_{\rm atm}$ curve.
+
+%peak when $M_{\rm{atm}} < M_{\rm c}$, before the crossover mass, is helpful for understanding atmospheric growth.
 In contrast with our analytic model, where the time to runaway accretion cannot be calculated self-consistently, the numerical model provides information about when unstable collapse occurs, i.e.
@@
 
-%The estimated timescale to runaway growth can be well approximated by only considering the time it takes to grow to the crossover mass, or even to a somewhat lower mass.
-This insensitivity to the upper mass threshold is characteristic of accelerating growth.
-
-%We next show how this behavior is fortuitous, as our model assumptions can break down during the later stages of atmospheric growth.
-
 \begin{figure}[tb]
@@
 
-To examine the regimes where our cooling model is valid, we compare the model luminosity to the additional neglected luminosity, $L_{\rm negl}$, that a more detailed model would have generated in the radiative zone.
-To compute $L_{\rm negl}$ we compute the entropy difference between successive radiative zone solutions.
+We examine the  validity of our cooling model by compare our model luminosity to the neglected luminosity, $L_{\rm negl}$,  that a more detailed model would generate in the radiative zone.
+We compute $L_{\rm negl}$ from the entropy difference between successive radiative zone solutions.
 We then integrate the energy equation, $\p L / \p m = - T \p S/ \p t$, over the average depth of the radiative zone.\footnote{While useful as a diagnostic, the neglected luminosity cannot reliably correct the global cooling model because the effects of $L_{\rm negl}$ on the structure of the radiative zone are still ignored.}
@@
  \Fig{fig:coolingterms} shows that the neglected luminosity is indeed negligible during the early stages of evolution.
-However, $L_{\rm negl}$ exceeds the model luminosity, $L$, at high masses, $M_{\rm atm} > 3 M_\oplus$ in this case, meaning that our model becomes inaccurate at late states.
-Nevertheless, our cooling model gives  reasonably reliable estimates of core accretion timescales.
-As shown above, the growth time is mostly spent at low masses, when $L_{\rm negl}$ is small and safely neglected.
- %Our results are conservative -- in the sense of overestimating growth times -- because the neglected luminosity would give more rapid evolution.  
- %AY: not sure we can back up the conservative claim because of the detailed effects on structure that haven't been calculated.
-
-The (non-vanishing) individual terms in the global cooling model of \Eq{eq:coolingglobal}, evaluated at the RCB, are also plotted in \Fig{fig:coolingterms}.
+However, $L_{\rm negl}$ exceeds the model luminosity, $L$, at high masses, $M_{\rm atm} > 3 M_\oplus$ in this case.
+Our cooling model is thus inaccurate at higher masses.
+However the model remains reasonably accurate up to the beginning stages of runaway growth, which is sufficient for our purposes of widely exploring parameter space and exploring trends.
+
+The individual terms in the global cooling model of \Eq{eq:coolingglobal}, evaluated at the RCB, are also plotted in \Fig{fig:coolingterms}.
 At low masses, the change in energy, $- \dot{E}$, makes the dominant contribution to luminosity.
@@
 Thus wherever our model is accurate -- including the crucial early phases of growth -- surface terms are a minor correction.
-%while the surface terms appear to reduce the total luminosity (compared to $-\dot{E}$), $L_{\rm negl}$ can more than compensate, showing again that our model conservatively underestimates cooling.
-%AY:again don't think conservative claim can be so easily justified.
-
-At yet higher masses than shown in \Fig{fig:coolingterms}, our cooling model breaks down.
-Mathematically, the breakdown manifests as negative time steps when advancing to a higher atmospheric mass.
-This unphysical result has no deep significance, as the neglected luminosity is already large enough to give inaccurate results.
-Sometimes this breakdown occurs even before the crossover mass is reached, although still significantly after growth has started to accelerate.
-As the breakdown always occurs after growth times peak, it does not affect our estimate of the timescale to runaway accretion $t_{\rm run}$.
-In our models the breakdown occurs at lower atmosphere mass as the semimajor axis increases, since accelerated and eventually runaway growth occurs sooner at larger distances, as seen in \Fig{fig:growthtime}.
-The main practical effect of this behavior is that we halt the evolution where the model breaks down whenever we cannot reach the crossover mass.
-Fortunately, the elapsed time to either the crossover mass or the breakdown is well-behaved and, most importantly, dominated by lower atmospheric masses where the model is accurate.
+The neglect of surface terms in the analytic model is thus not a serious omission.
+
+At even higher masses than shown in \Fig{fig:coolingterms}, our cooling model breaks down.
+Mathematically, the breakdown manifests as negative time step (due to an increase in internal entropy) when advancing to a higher atmospheric mass.
+This unphysical result has no deep significance, as the neglected luminosity overwhelms the model luminosity by this point.
+Effectively the model is telling us when to ignore it.
+Our results are unaffected by the breakdown, which occurs comfortably after $t_{\rm run}$, our threshold for runaway growth.
 
@@
 
-\item Runaway growth begins roughly at the crossover mass, when atmosphere and core masses are equal, in agreement with previous work \citep{pollack96}.
-For larger core masses, runaway growth begins at larger ratios of atmosphere to core mass.
-
-\item Self-gravity affects atmospheric evolution before crossover.  Significant self-gravitational corrections appear when the atmosphere is only $\sim$$20 \%$ as massive as the core.
+\item Runaway growth begins roughly at the crossover mass, when atmosphere and core masses are equal, $M_{\rm atm} \sim M\co$, in agreement with previous work \citep{pollack96}.
+Further from the star, runaway growth begins at smaller $M_{\rm atm} \sim M\co$ ratios.
+For larger core masses, runaway growth begins at larger values of $M_{\rm atm} \sim M\co$.
+
+\item Self-gravity affects atmospheric evolution before crossover.  Significant self-gravitational corrections appear when the atmosphere is only $\sim$$10 \%$ as massive as the core.
 

### 2013-10-25 tex/draft/draft4.tex
added ackowledgements
very minor typos in section 4

@@
 \label{sec:profiles}
-Figure \ref{fig:profiles} shows radial profiles at different stages of atmospheric growth around $5 M_{\oplus}$ core at $60$ AU.
+Figure \ref{fig:profiles} shows radial profiles at different stages of atmospheric growth around a $5 M_{\oplus}$ core at $60$ AU.
 Quoted mass values include the core plus atmosphere within the smaller of $\RB$ or $\RH$, which for these cases is $\RB$.
@@
 
-We examine the  validity of our cooling model by compare our model luminosity to the neglected luminosity, $L_{\rm negl}$,  that a more detailed model would generate in the radiative zone.
+We examine the  validity of our cooling model by comparing our model luminosity to the neglected luminosity, $L_{\rm negl}$,  that a more detailed model would generate in the radiative zone.
 We compute $L_{\rm negl}$ from the entropy difference between successive radiative zone solutions.
@@
 Our cooling model is thus inaccurate at higher masses.
-However the model remains reasonably accurate up to the beginning stages of runaway growth, which is sufficient for our purposes of widely exploring parameter space and exploring trends.
+However, the model remains reasonably accurate up to the beginning stages of runaway growth, which is sufficient for our purposes of widely exploring parameter space and exploring trends.
 
@@
 At even higher masses than shown in \Fig{fig:coolingterms}, our cooling model breaks down.
-Mathematically, the breakdown manifests as negative time step (due to an increase in internal entropy) when advancing to a higher atmospheric mass.
+Mathematically, the breakdown manifests as a negative time step (due to an increase in internal entropy) when advancing to a higher atmospheric mass.
 This unphysical result has no deep significance, as the neglected luminosity overwhelms the model luminosity by this point.
@@
 Ongoing direct imaging surveys and their successors \citep{hinz12, macintosh12} will help discriminate among the various planet formation pathways in the outskirts of protoplanetary disks.
+
+\acknowledgements{We thank Ruth Murray-Clay for detailed feedback and skillful proof reading. We also thank Eugene Chiang for a valuable conversation.}
 

### 2013-10-25 tex/draft/draft4.tex
tex file updated with figure changes

@@
 %AY: figure should probably use R_B for consistency with text.
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/PTm_profiles_v2_smallM.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/PTm_profiles_v2_smallM.eps}
 %\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Lt_profiles_v2.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Lt_profiles_v2.eps}
 %\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/opacity_effect.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/opacity_effect.eps}
 %\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mt_profile_temp.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mt_profile_temp.eps}
 %\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cooling_a10_Mc5_rcb_paper.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cooling_a10_Mc5_rcb.eps}
 %\vspace{-0.5in}
@@
 \hspace{-.1in}
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cumul_coolingtime_vs_Matm_10au_mu235.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cumul_coolingtime_vs_Matm_10au_mu235.eps}
 %\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.48\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/coolingtime_vs_Mc_10au.pdf}
+\includegraphics[width=0.48\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/coolingtime_vs_Mc_10au.eps}
 %\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=1.\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/TdPd_effect.pdf}
+\includegraphics[width=1.\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/TdPd_effect.eps}
 \vspace{-0.3in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_new.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_new.eps}
 %\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_opacity.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_opacity.eps}
 %\vspace{-0.5in}

### 2013-10-25 tex/draft/draft4.tex
minor section 4 edits + removed model breakdown part

@@
 \newcommand{\Chap}[1]{Chapter~\ref{#1}}
-\newcommand{\Fig}[1]{Fig.~\ref{#1}}
+\newcommand{\Fig}[1]{Figure~\ref{#1}}
 \newcommand{\Figs}[2]{Figs.~\ref{#1} and \ref{#2}}
@@
 \caption{Radial profiles of atmospheric pressure, temperature and enclosed mass (core included) for a $5 M_{\oplus}$ core at $a=60$ AU.
-Solid, dotted, dashed and dot-dashed lines correspond to solutions with total mass (core and atmosphere) of $5.01 M_{\oplus}$, $5.10 M_{\oplus}$, $6.00 M_{\oplus}$ and $9.24 M_{\oplus}$, respectively.
-(See text for a description of evolution to yet higher masses.)   Circles and triangles mark the locations of the Bondi radii and of the radiative-convective boundaries, respectively.
-The radial profiles extend from the core to the Hill radius.}
+Solid, dotted, dashed and dot-dashed lines correspond to solutions with total mass (core and atmosphere) of $5.01 M_{\oplus}$, $5.10 M_{\oplus}$, $6.00 M_{\oplus}$ and $8.99 M_{\oplus}$, respectively.
+Circles and triangles mark the locations of the Bondi radii and of the radiative-convective boundaries, respectively.
+The radial profiles extend from the core to the Hill radius.} %  (See text for a description of evolution to yet higher masses.)
 \label{fig:profiles}
@@
 Quoted mass values include the core plus atmosphere within the smaller of $\RB$ or $\RH$, which for these cases is $\RB$.
-The 9.24 $M_{\oplus}$ solution is the highest mass we can reach in our evolutionary sequence, as we explain further in \S\ref{sec:endoftime}.
+The 8.99 $M_{\oplus}$ solution is the mass at which runaway growth begins according to the criteria described in \S\ref{sec:timeev}.
+% we can reach in our evolutionary sequence, as we explain further in \S\ref{sec:endoftime}.
 
@@
 Atmospheric cooling and contraction allows the accretion of more gas.
-In the convective zones, higher mass solutions contain lower entropy and higher pressures.
+In the convective zones, higher mass solutions have lower entropy and higher pressures.
 A radiative zone emerges to connect the lower entropy interior to the higher entropy disk.
@@
 
-The cooling model of \S\ref{sec:twolayer} is used to connect solutions of different atmospheric masses into an evolutionary sequence.
+The cooling model of \S\ref{sec:twolayer} is used to connect solutions with different atmospheric masses into an evolutionary sequence.
 \Fig{fig:Ltplot} shows the luminosity evolution and the elapsed time as a function of atmospheric mass for a $5 M_{\oplus}$ core at $60$ AU, i.e.\ the same parameters as in \Fig{fig:profiles}.
@@
 At later stages of evolution, the numerical model in \Fig{fig:Ltplot} shows a flat luminosity with increasing mass and also time (not shown).
-By contrast, the non-self-gravitating analytic model gives a luminosity that continues to drop with increasing mass.
+By contrast, the non-self-gravitating analytic model gives a luminosity that continues to drop as the atmosphere becomes more massive.
 To understand this difference, consider the scaling of \Eq{eq:Lcb}, $L\cb \propto M\cb T\cb^4/ (\kappa\cb  P\cb)$, which holds in both cases.
@@
 The accelerated growth in the numerical model, as shown in the bottom panel of \Fig{fig:Ltplot}, is also a direct result of the enhanced cooling (higher luminosities) in the self-gravitating models.
-We see that self-gravity plays a very important role in atmospheric evolution, even well before the crossover mass.
+We see that self-gravity plays a very important role in atmospheric evolution, even well before the crossover mass, when $M_{\rm atm}=M\co$.
 
@@
 
-The fact that growth times have a well defined maximum is a characteristic of runaway growth.
-Unlike our analytic model, which must assume that runaway growth begins near the crossover mass, our numerical model allows us to measure when runaway growth begins.
+The fact that growth times have a well defined maximum is a characteristic of accelerating growth.
+Unlike our analytic model, which must assume that runaway growth begins near the crossover mass, our numerical model allows us to measure when runaway accretion begins.
 Runaway growth clearly does not begin at a universal value of $M_{\rm atm}/M\co$.
@@
 
-At even higher masses than shown in \Fig{fig:coolingterms}, our cooling model breaks down.
+%At even higher masses than shown in \Fig{fig:coolingterms}, our cooling model breaks down.
 Mathematically, the breakdown manifests as a negative time step (due to an increase in internal entropy) when advancing to a higher atmospheric mass.

### 2013-10-25 tex/draft/draft4.tex
partial section 3 edits (lunchtime)

--- 2013-10-25 tex/draft/draft4.tex
@@
 
-This section develops the analytic version of our two-layer model, which is useful for predicting and explaining the results of numerical evolution models.
-To obtain our analytic results, we neglect self-gravity.
-As shown in  \S\ref{sec:KH}, self-gravity becomes quantitatively important at surprisingly low atmospheric masses, $M_{\rm atm} \gtrsim 0.2 M\co$.
-Thus the analytic model cannot accurately model the later stages of evolution.
-Nevertheless, this model does explain general trends, including the dependence on opacity and disk temperature and pressure.
-
-The analytic model also assumes that the upper radiative radiative layer is thick, with a pressure depth $P\cb \gg P\di$.
-This approximation limits the ability to study the initial cooling of very low mass atmospheres with thin outer radiative zones.
-However, since early evolution is quite rapid, this approximation negligibly affects total cooling times and critical core masses, our main foci.
+This section develops the analytic version of our two-layer model.
+This analytic model is less accurate, primarily because it neglects self-gravity which we show (in \S\ref{sec:KH}) to be important at rather low atmospheric masses, $M_{\rm atm} \gtrsim 0.1 M\co$.
+Nevertheless the analytic model is useful for understanding atmospheric evolution and interpreting the numerical results.
+
+The analytic model also assumes that the upper radiative radiative layer is thick enough that $P\cb \gg P\di$.
+This approximation ignores the earliest stages of cooling, which are rapid enough to be negligible.
+Finally the analytic model ignores the surface terms in \Eq{eq:coolingglobal}, which we show to be a modest correction in \S\ref{sec:endoftime}.
 
@@
 To apply the two-layer cooling model analytically, we require expressions for the atmospheric structure.
-Conditions at the RCB are crucial as they set the interior adiabat and the radiative losses.
+Conditions at the RCB are crucial as they set the interior adiabat and the radiative losses from the interior.
 (Recall that luminosity generation in the radiative zone is neglected.)  We express the temperature and pressure of the RCB, at the radius $R\cb$, as
@@
 \end{equation}
-for $P\cb \gg P\di$, our regime of interest, and with the radiative temperature gradient at infinite depth, $\nabla_\infty = 1/2$, for our dust opacity.
-The radiative zones are thus nearly isothermal, as also found by R06.
-While  a simple closed form expression is not possible, the factor $\theta \simeq 0.556$ for our parameters.
+for $P\cb \gg P\di$, our regime of interest, and with the radiative temperature gradient at depth, $\nabla_\infty = 1/2$, for our dust opacity.
+The radiative zones are thus nearly isothermal, as found by R06.
+A numerical integration gives $\theta \simeq 0.556$ for our parameters.
 %According to \Eq{eq:PcbRcb} our pressure criterion also gives $R\cb \ll \RB$, but the inequality is logarithmically weaker.
@@
 \end{subeqnarray} 
-While the radial density profile depends on the adiabatic index, the $r^{-1}$ temperature scaling is universal.  For reference, if self-gravity is included, the temperature profile satisfies
+While the radial density profile depends on the adiabatic index, the $r^{-1}$ temperature scaling is universal.  In self-gravitating models, the temperature gradient,
 \begin{equation} \label{eq:TPsg}
-{dT \over dr} = - {G m(r) \over C_P r^2}\, .
-\end{equation} 
-Thus, the enclosed mass gives a temperature profile flatter than $T \propto r^{-1}$ in self-gravitating models.
+{dT \over dr} = - {G m(r) \over C_P r^2}\, ,
+\end{equation} 
+gives a profile that is flatter than $T \propto r^{-1}$.
 
@@
 \begin{equation} \label{eq:crit}
-{M_{\rm atm} \over M\co} = {P\cb \over \xi P_M},
-\end{equation} 
-where we define a characteristic pressure and a logarithmic factor
+{M_{\rm atm} \over M\co} = {P\cb \over \xi P_M}\, ,
+\end{equation} 
+where we define a characteristic pressure and a logarithmic factor:
 \begin{subeqnarray} 
@@
 The critical value of the order unity factor $\xi$ is found by eliminating $P\cb$ from \Eqs{eq:xidef}{eq:Pcbc}.
-This logarithmic factor complicates our analytic description, but since it remains order unity we simply hold it fixed in our scalings.
+This logarithmic factor complicates our analytic description.
+Since it remains order unity we simply hold it fixed in our scalings, but not in plots.
 
@@
 \end{equation} 
-The inverse scaling $L\cb \propto 1/P\cb$ shows that luminosity drops as the atmosphere cools and $P\cb$ deepens.
+The scaling $L\cb \propto 1/P\cb$ shows that luminosity drops as the atmosphere cools and $P\cb$ deepens.
 This result relies on the pressure independence of  dust opacities.
@@
 \label{coolingan}
-%We can eliminate $R\cb$ from equation (\ref{eq:Matman}) using \Eq{eq:PcbRcb}.  The ratio of atmosphere to core mass is then  
-%\begin{equation} \label{eq:crit}
-%{M_{\rm atm} \over M\co} \approx {P\cb /P_M \over  \sqrt{\ln[ P\cb/(\theta P_{\rm d})]}},
-%\end{equation} 
-%where we introduce a characteristic pressure
-%\begin{equation} 
-%P_M \equiv {4 \delad^{3/2} \over 5 \pi^2 \sqrt{\chi} } {G M\co^2 \over {\RB'}^4}\, .
-%%P_M \approx {1.85 \over \sqrt{\chi}} {G M\co^2 \over {\RB'}^4}\, .
-%\end{equation} 
-%For $M_{\rm atm} = M\co$ in equation (\ref{eq:crit}) we require
-%\begin{equation} \label{eq:Pcbc}
-%P\cb = \xi P_M ,
-%\end{equation} 
-%where  the logarithmic factor
-%\begin{equation}\label{eq:xi}
-%\xi \equiv \sqrt{\ln[ P\cb/(\theta P_{\rm d})]} = \sqrt{\ln[ \xi P_M / (\theta P_{\rm d})]}
-%\end{equation} 
-%is found by numerically solving the above transcendental equation.
-The physical solution has $\xi >1$, but is typically of order unity.
-Numerical solutions exist for $\xi \ll 1$, but are unphysical as they imply $P\cb < P_{\rm d}$.
-
-Our analytic cooling model uses $L = -\dot{E}$,  neglecting the surface terms in \Eq{eq:coolingglobal}.\footnote{\App{surfterms} shows that these terms are negligible for a non-self-gravitating model, but they do become relevant when self-gravity is important and included.}  Applying \Eqs{eq:Eus}{eq:Lcb}, the time it takes to cool the atmosphere until the RCB reaches a given pressure depth, $P\cb$, is
+
+Our analytic cooling model uses $L = -\dot{E}$,  neglecting the surface terms in \Eq{eq:coolingglobal}.\footnote{\App{surfterms} shows that these terms are negligible for a non-self-gravitating model.}  %They do become relevant when self-gravity is important and included, as shown in \S\ref{sec:endoftime}.}
+Applying \Eqs{eq:Eus}{eq:Lcb}, the time it takes to cool the atmosphere until the RCB reaches a given pressure depth, $P\cb$, is 
 \begin{subeqnarray} 
@@
 
-The most relevant cooling time is the crossover time, $t_{\rm co}$, required to reach the crossover mass, $M_{\rm{atm}} = M_{\rm c}$. \Eqs{eq:tcool}{eq:Pcbc} give this time as
+We expect runaway growth to begin around the crossover mass, $M_{\rm{atm}} = M_{\rm c}$. \Eqs{eq:tcool}{eq:Pcbc} give the time to crossover as
 \begin{eqnarray} 
@@
 \end{eqnarray} 
-The corresponding critical core mass, $\MC$, has a crossover time that equals the typical lifetime of a protoplanetary disk
+We estimate critical core mass, $\MC$, has a crossover time that equals the typical lifetime of a protoplanetary disk
 \begin{equation}
@@
 \label{sec:profiles}
-Figure \ref{fig:profiles} shows radial profiles at different stages of atmospheric growth around a $5 M_{\oplus}$ core at $60$ AU.
+Figure \ref{fig:profiles} shows radial profiles at different stages of atmospheric growth around $5 M_{\oplus}$ core at $60$ AU.
 Quoted mass values include the core plus atmosphere within the smaller of $\RB$ or $\RH$, which for these cases is $\RB$.
@@
 
-We examine the  validity of our cooling model by comparing our model luminosity to the neglected luminosity, $L_{\rm negl}$,  that a more detailed model would generate in the radiative zone.
+We examine the  validity of our cooling model by compare our model luminosity to the neglected luminosity, $L_{\rm negl}$,  that a more detailed model would generate in the radiative zone.
 We compute $L_{\rm negl}$ from the entropy difference between successive radiative zone solutions.
@@
 Our cooling model is thus inaccurate at higher masses.
-However, the model remains reasonably accurate up to the beginning stages of runaway growth, which is sufficient for our purposes of widely exploring parameter space and exploring trends.
+However the model remains reasonably accurate up to the beginning stages of runaway growth, which is sufficient for our purposes of widely exploring parameter space and exploring trends.
 
@@
 At even higher masses than shown in \Fig{fig:coolingterms}, our cooling model breaks down.
-Mathematically, the breakdown manifests as a negative time step (due to an increase in internal entropy) when advancing to a higher atmospheric mass.
+Mathematically, the breakdown manifests as negative time step (due to an increase in internal entropy) when advancing to a higher atmospheric mass.
 This unphysical result has no deep significance, as the neglected luminosity overwhelms the model luminosity by this point.
@@
 Ongoing direct imaging surveys and their successors \citep{hinz12, macintosh12} will help discriminate among the various planet formation pathways in the outskirts of protoplanetary disks.
-
-\acknowledgements{We thank Ruth Murray-Clay for detailed feedback and skillful proof reading. We also thank Eugene Chiang for a valuable conversation.}
 

### 2013-10-25 tex/draft/draft4.tex
removed c subscript from Mcrit

@@
 \caption{The critical core mass as a function of semimajor axis, for a disk lifetime of $3$ Myrs and two values of the mean molecular weight ($\mu = 2.35$ is for Solar abundances).
-The decline in $M_{\rm c,crit}$ with distance is a robust result for standard disk models.
+The decline in $\MC$ with distance is a robust result for standard disk models.
 The analytic model, which neglects self-gravity, over-predicts the steepness of the decline.}
@@
 distance for standard and reduced (by a factor of ten) dust opacities.
-Lower opacities give significantly lower $M\crit$ values.
+Lower opacities give significantly lower $\MC$ values.
 Atmospheric opacity remains a large uncertainty in core accretion models.}
@@
 Since the vast majority of disk models have a disk temperature that declines with $a$, our qualitative result is robust.
-We also show that higher $\mu$ values give lower values of $M\crit$.
-
-The average power-law decline from 1 to 100 AU (not plotted) is $M\crit \propto a^{-0.3}$, for both choices of $\mu$.
+We also show that higher $\mu$ values give lower values of $\MC$.
+
+The average power-law decline from 1 to 100 AU (not plotted) is $\MC \propto a^{-0.3}$, for both choices of $\mu$.
 While not a drastic decline, the ability of distant low mass cores to accrete gas efficiently is significant for the interpretation of direct imaging surveys.
@@
 
-Our non-self-gravitating analytic model (dashed curves in \Fig{fig:Mcvsa}) over-predicts the steepness of the decline in $M\crit$ with $a$.
+Our non-self-gravitating analytic model (dashed curves in \Fig{fig:Mcvsa}) over-predicts the steepness of the decline in $\MC$ with $a$.
 This discrepancy is not surprising, as we have shown that self-gravity strongly affects evolution, even before runaway growth begins and the crossover mass is reached.
 
-\Fig{fig:Mcritopacity} shows that reducing the opacity by an order of magnitude significantly reduces $M\crit$.
+\Fig{fig:Mcritopacity} shows that reducing the opacity by an order of magnitude significantly reduces $\MC$.
 Furthermore, this opacity effect is stronger at larger distances.
-The opacity reduction by a factor of 10 lowers $M\crit$ at 5 AU by a factor of $\sim$$2.5$ and at 100 AU by a factor of $\sim$$3.5$.
+The opacity reduction by a factor of 10 lowers $\MC$ at 5 AU by a factor of $\sim$$2.5$ and at 100 AU by a factor of $\sim$$3.5$.
 Atmospheric opacity is a dominant uncertainty in core accretion modelling.
-However, unless atmospheric opacity varies significantly with disk radius, the general decline in $M\crit$ with increasing $a$ should hold.
+However, unless atmospheric opacity varies significantly with disk radius, the general decline in $\MC$ with increasing $a$ should hold.
 For our scenario of negligible ongoing planetesimal accretion, it is tempting to think that dust could settle out of the radiative zone, lowering the opacity \citep{podolak03}.
@@
 \begin{enumerate}
-\item The minimum or critical core mass, $M\crit$, for giant planet formation declines with stellocentric distance in standard protoplanetary disk models.
+\item The minimum or critical core mass, $\MC$, for giant planet formation declines with stellocentric distance in standard protoplanetary disk models.
 For our reference case, the critical mass is $\sim$$8.5 M_{\oplus}$ at $a = 5$ AU, decreasing to $\sim$$3.5 M_{\oplus}$ at $a =100$ AU.
-This decline roughly follows $M\crit \propto a^{-0.3}$.
+This decline roughly follows $\MC \propto a^{-0.3}$.
 
@@
 \item A larger mean molecular weight reduces critical core masses, in agreement with \citet{HorIko11}.
-If enrichment in heavy elements correlates with increased dust opacity, then the stronger opacity effect will dominate, increasing $M\crit$.
+If enrichment in heavy elements correlates with increased dust opacity, then the stronger opacity effect will dominate, increasing $\MC$.
 
@@
 
-Thus for giant planets to form around cores near $M\crit$, core growth must stall quite severely.
+Thus for giant planets to form around cores near $\MC$, core growth must stall quite severely.
 The accretion and scattering of local planetesimals, combined with the inward aerodynamic drift of smaller solids past the core, offers a plausible means to sufficiently isolate the core.
@@
 A  lower opacity  decreases the critical core mass.
-Reducing the opacity by a factor of one hundred results in a critical core mass one order of magnitude lower than in the standard ISM case, specifically  $M_{\rm{crit}} \approx9 M_\oplus$ for the parameters in \Eq{eq:Mcrit}.
+Reducing the opacity by a factor of one hundred results in a critical core mass one order of magnitude lower than in the standard ISM case, specifically  $\MC \approx9 M_\oplus$ for the parameters in \Eq{eq:Mcrit}.
 The reduction is not  as strong as the nominal scaling would imply, $0.01^{3/5} \approx 0.06$, because $\xi$ increases.

### 2013-10-25 tex/draft/draft4.tex
section 3 edits,  eps back to pdf for now

--- 2013-10-25 tex/draft/draft4.tex
@@
 \usepackage{graphicx} 
-\usepackage{graphics}                       
+%\usepackage{graphics}                       
 \usepackage{amsmath}
@@
 \usepackage{subeqnarray}
+\usepackage{epstopdf}
 %\bibliographystyle{apj}
@@
 \end{eqnarray} 
-We estimate critical core mass, $\MC$, has a crossover time that equals the typical lifetime of a protoplanetary disk
+We estimate critical core mass, $\MC$, by equating the crossover time with a typical protoplanetary disk lifetime,
 \begin{equation}
@@
 As is well known, a lower opacity, here scaling with $F_\kappa$, accelerates growth and gives a lower critical core mass.
-Lower disk temperatures also decrease both $t_{\rm co}$ and $\MC$.
-Furthermore, the decline in both quantities with semimajor axis is completely generated by the disk temperature law (ignoring the logarithmic $\xi$).
+The cooling timescale and critical core mass depend only weakly on the disk pressure, via the logarithmic factor $\xi$.
+
+Lower disk temperatures decrease both $t_{\rm co}$ and $\MC$.
+The decline in both quantities with semimajor axis is completely explained by the temperature  profile (ignoring the logarithmic factor $\xi$).
 The temperature dependence is a competition between two main effects.
-The higher luminosity, $\propto T^{4-\beta}$, at higher temperatures gives faster cooling, which opposes the overall effect.
-The larger Bondi radius, and the effectively stronger gravity of the core, is the dominant effect that gives faster cooling at lower temperatures.
-For different $\beta$ values, $t_{\rm co} \propto F_T^{ \beta + 1/2}$, so the slope of the dust opacity is an important moderating effect on the cooling.
-
-The cooling timescale and critical core mass depend only weakly on the disk pressure, via the  logarithmic factor $\xi$.
-Section \S\ref{sec:TPeffects} shows that these analytic predictions for disk temperature and pressure dependence compare favorably with numerical results.
-Moreover, the analytic estimates quantitatively reproduce the numerical results to reasonable accuracy, if the effective crossover mass for unstable atmosphere collapse is a fraction of the core mass, i.e.
-$M_{\rm atm}=f M\co$.
-Under this assumption, Equations (\ref{eq:tcoolan}) and (\ref{eq:Mcrit}) become
-\begin{eqnarray} 
-\label{eq:tcoolanscaled}
-t_{\rm co} &\approx & 2 \times 10^8 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2 f^2  \over \mcn{10}^{5/3} \aun{10}^{15 / 14}} \yr \\ \slabel{eq:tcoolanf}
- & \approx & 3 \times 10^6 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2  \over \mcn{10}^{5/3} \aun{10}^{15 / 14}} \yr
-\end{eqnarray} 
-and
-\begin{eqnarray}\label{eq:Mcritscaled}
-\MC &\approx & 100 {F_T^{3/2} F_\kappa^{3/5}   \left(\xi \over 2.6 \right)^{6/5} f^{6/5} \over \aun{10}^{9 / 14}} \; M_\oplus \\ \slabel{eq:Mcritf}
-& \approx & 8\, {F_T^{3/2} F_\kappa^{3/5}   \left(\xi \over 2.6 \right)^{6/5} \over \aun{10}^{9 / 14}} \; M_\oplus\ ,
-\end{eqnarray}
-with $f=0.13$ in Equations (\ref{eq:tcoolanf}) and (\ref{eq:Mcritf}).
-We choose this particular fractional scaling to match our numerical results at 10 AU (see section \S\ref{sec:critical} for more details).
-We note that the scaling above is valid if the logarithmic factor $\xi$ is fixed; however, the analytic estimates presented in section \S\ref{sec:critical} explicitly take into account the dependence of $\xi$ on the core mass (see Equations \ref{eq:xidef} and \ref{eq:Pcbc}).
-
- %This weak dependence is a consequence of cooling coming from the radiative-convective boundary.
-Note that the scaling laws do not reflect the changes to $\xi$ which remains order unity.
-% and is here given the appropriate value for the nominal parameters.
-
-%\emph{TODO: consider adding "revised" analytic expressions for $t_{\rm co}$ and $\MC$ (e.g.
-lower Matm criterion).
-People like scalings, but the obviously wrong amplitudes will prevent people from using them unless we correct.}
+The Bondi radius, which quantifies the strength of gravity, decreases for colder temperatures.
+On the other hand, the luminosity, $\propto T^{4-\beta}$, is lower at colder temperatures, which slows cooling and opposes the overall effect.
+With $t_{\rm co} \propto F_T^{ \beta + 1/2}$, the slope of the dust opacity,  $\beta$, is an important factor in regulating the resulting growth time.
+
+To facilitate comparison with our numerical results, we rescale the analytic results as a rough correction for the neglect of self gravity.
+The goal of the rescaling is not to get a perfect agreement, but to aid in the interpretation of general trends.
+To rescale, we simply assume that runaway occurs at a modified crossover mass, $M_{\rm atm} = f M\co$.
+From \Eqs{eq:crit}{eq:Pcbc}, this change amounts to the replacing $\xi \rightarrow f\xi$.
+We require $f < 1$ to give the faster growth seen in numerical models.
+We choose $f = 0.13$, which normalizes the analytic and numerical models at 10 AU, for standard paramters.
+The revised scalings are
+\begin{subeqnarray}
+t_{\rm run} & \approx & 3 \times 10^6 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2  \over \mcn{10}^{5/3} \aun{10}^{15 / 14}} \yr  \slabel{eq:tcoolanf} \\
+\MC & \approx & 8\, {F_T^{3/2} F_\kappa^{3/5}   \left(\xi \over 2.6 \right)^{6/5} \over \aun{10}^{9 / 14}} \; M_\oplus\ \, ,  \slabel{eq:Mcritf}
+\end{subeqnarray} 
+now using $t_{\rm run}$ instead of $t_{\rm co}$ for the runaway growth timescale, which is no longer tied to the crossover mass.
+The numerical models are more accurate and self-consistent because they measure runaway growth directly, instead of relying on any assumed $M_{\rm atm}/M\co$ threshold.
+When plotting the analytic scalings we include the variations in $\xi$ from \Eq{eq:xidef}.
 
@@
 %AY: figure should probably use R_B for consistency with text.
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/PTm_profiles_v2_smallM.eps}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/PTm_profiles_v2_smallM.pdf}
 %\vspace{-0.5in}
@@
 
-\subsection{Atmospheric Stucture}
+\subsection{Atmospheric Structure}
 \label{sec:profiles}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Lt_profiles_v2.eps}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Lt_profiles_v2.pdf}
 %\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/opacity_effect.eps}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/opacity_effect.pdf}
 %\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mt_profile_temp.eps}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mt_profile_temp.pdf}
 %\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cooling_a10_Mc5_rcb.eps}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cooling_a10_Mc5_rcb.pdf}
 %\vspace{-0.5in}
@@
 \hspace{-.1in}
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cumul_coolingtime_vs_Matm_10au_mu235.eps}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cumul_coolingtime_vs_Matm_10au_mu235.pdf}
 %\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.48\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/coolingtime_vs_Mc_10au.eps}
+\includegraphics[width=0.48\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/coolingtime_vs_Mc_10au.pdf}
 %\vspace{-0.5in}
 \caption{Runaway growth time, $t_{\rm{run}}$, vs.\ core mass at $10$ AU, for two values of the mean molecular weight.
-Our numerical model (\emph{solid curves}) is compared to our non-self-gravitating analytic model (\emph{dashed curves}, with $t_{\rm run}$ as prescribed by Equation \ref{eq:tcoolanf}).
+Our numerical model (\emph{solid curves}) is compared to our non-self-gravitating analytic model (\emph{dashed curves}, from Equation \ref{eq:tcoolanf}).
 A typical protoplanetary disk life time of $3$ Myr is plotted for comparison.
@@
 \centering
-\includegraphics[width=1.\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/TdPd_effect.eps}
+\includegraphics[width=1.\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/TdPd_effect.pdf}
 \vspace{-0.3in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_new.eps}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_new.pdf}
 %\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_opacity.eps}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_opacity.pdf}
 %\vspace{-0.5in}

### 2013-10-25 tex/draft/draft4.tex
minor section 3 edits / typo fixes

@@
 This section develops the analytic version of our two-layer model.
-This analytic model is less accurate, primarily because it neglects self-gravity which we show (in \S\ref{sec:KH}) to be important at rather low atmospheric masses, $M_{\rm atm} \gtrsim 0.1 M\co$.
-Nevertheless the analytic model is useful for understanding atmospheric evolution and interpreting the numerical results.
-
-The analytic model also assumes that the upper radiative radiative layer is thick enough that $P\cb \gg P\di$.
+This analytic model is less accurate, primarily because it neglects self-gravity, which we show (in \S\ref{sec:KH}) to be important at rather low atmospheric masses, $M_{\rm atm} \gtrsim 0.1 M\co$.
+Nevertheless, the analytic model is useful for understanding atmospheric evolution and interpreting the numerical results.
+
+The analytic model also assumes that the upper radiative layer is thick enough that $P\cb \gg P\di$.
 This approximation ignores the earliest stages of cooling, which are rapid enough to be negligible.
-Finally the analytic model ignores the surface terms in \Eq{eq:coolingglobal}, which we show to be a modest correction in \S\ref{sec:endoftime}.
+Finally, the analytic model ignores the surface terms in \Eq{eq:coolingglobal}, which we show to be a modest correction in \S\ref{sec:endoftime}.
 
 \subsection{Two Layer Structure}
-To apply the two-layer cooling model analytically, we require expressions for the atmospheric structure.
+In order to apply the two-layer cooling model analytically, we require expressions for the atmospheric structure.
 Conditions at the RCB are crucial as they set the interior adiabat and the radiative losses from the interior.
@@
 \begin{equation}
-\RB' \equiv {G M\co \over C_P T\cb} = {\delad \over \chi} \RB
+\RB' \equiv {G M\co \over C_PT\cb} = {\delad \over \chi} \RB
 \end{equation} 
@@
 This logarithmic factor complicates our analytic description.
-Since it remains order unity we simply hold it fixed in our scalings, but not in plots.
+Since it remains order unity, we simply hold it fixed in our scalings, but not in plots.
 
@@
 \end{eqnarray} 
-We estimate critical core mass, $\MC$, by equating the crossover time with a typical protoplanetary disk lifetime,
+We estimate the critical core mass, $\MC$, by equating the crossover time with a typical protoplanetary disk lifetime,
 \begin{equation}
@@
 The main reason for this discrepancy is the neglect of self-gravity, as shown in Section \ref{sec:KH}.
-The magnitude of the discrepancy may be surprising as one might expect self-gravity to be a modest factor for $M_{\rm atm} \leq M\co$.
+The magnitude of the discrepancy may be surprising, as one might expect self-gravity to be a modest factor for $M_{\rm atm} \leq M\co$.
 Clearly this seemingly reasonable expectation can be misleading, an interesting result in itself.
@@
 The temperature dependence is a competition between two main effects.
-The Bondi radius, which quantifies the strength of gravity, decreases for colder temperatures.
+The Bondi radius, which quantifies the strength of gravity, decreases for lower temperatures.
 On the other hand, the luminosity, $\propto T^{4-\beta}$, is lower at colder temperatures, which slows cooling and opposes the overall effect.
@@
 We require $f < 1$ to give the faster growth seen in numerical models.
-We choose $f = 0.13$, which normalizes the analytic and numerical models at 10 AU, for standard paramters.
+We choose $f = 0.13$, which normalizes the analytic and numerical models at 10 AU, for standard parameters.
 The revised scalings are
@@
 The numerical models are more accurate and self-consistent because they measure runaway growth directly, instead of relying on any assumed $M_{\rm atm}/M\co$ threshold.
-When plotting the analytic scalings we include the variations in $\xi$ from \Eq{eq:xidef}.
+When plotting the analytic scalings, we include the variations in $\xi$ from \Eq{eq:xidef}.
 

### 2013-10-25 tex/draft/draft4.tex
further section 3 edits

@@
 This section develops the analytic version of our two-layer model.
-This analytic model is less accurate, primarily because it neglects self-gravity, which we show (in \S\ref{sec:KH}) to be important at rather low atmospheric masses, $M_{\rm atm} \gtrsim 0.1 M\co$.
+This analytic model is less accurate than our numerical model, primarily because it neglects self-gravity, which we show (in \S\ref{sec:KH}) to be important at rather low atmospheric masses, $M_{\rm atm} \gtrsim 0.1 M\co$.
 Nevertheless, the analytic model is useful for understanding atmospheric evolution and interpreting the numerical results.
@@
 
-Using \Eq{eq:PcbRcb} to eliminate $R\cb$, the critical ratio of atmosphere to core mass becomes
+Using \Eq{eq:PcbRcb} to eliminate $R\cb$, the ratio of atmosphere to core mass becomes
 \begin{equation} \label{eq:crit}
@@
 This logarithmic factor complicates our analytic description.
-Since it remains order unity, we simply hold it fixed in our scalings, but not in plots.
+Since it remains order unity, we simply hold it fixed in our scalings, but not when plotting the results.
 
@@
 Aside from this insight, the analytic model is useful because, despite the amplitude error, it explains the basic scaling of numerical cooling models.
-As is well known, a lower opacity, here scaling with $F_\kappa$, accelerates growth and gives a lower critical core mass.
+As is well known \citep{HubBod05}, a lower opacity, here scaling with $F_\kappa$, accelerates growth and gives a lower critical core mass.
 The cooling timescale and critical core mass depend only weakly on the disk pressure, via the logarithmic factor $\xi$.

### 2013-10-27 tex/draft/draft4.tex
intro edits

--- 2013-10-25 tex/draft/draft4.tex
@@
 
-Current theories postulate that giant planets form either through gravitational instability (GI) or core accretion (e.g., \citealt{dangelo11}, \citealt{youdin13}).
-In the core accretion model (e.g., \citealt{harris78}, \citealt{mizuno78}, \citealt{stevenson82}, \citealt{boden86}, \citealt{pollack96}), solid planetesimals collide and grow into a massive solid core which then accretes a gaseous envelope.
-GI theories postulate that giant planets can be formed due to the fragmentation of the protoplanetary disk into self-gravitating clumps (e.g., \citealt{boss97}).
+Models of giant planet fall in two categories:  core accretion or gravitational instability \citep{dangelo11, youdin13}.
+In core accretion models, a solid core grows until it becomes massive enough to rapidly accrete gas \citep{PerCam74, mizuno78}.
+Gravitational instability (GI) theories investigate the fragmentation of the protoplanetary disk into bound clumps \citep{cameron78, boss97}.
 %quote murray-clay, kratter 10, rafikov 05, etc.
@@
 
-Forming giant planets at wide separations in the protoplanetary disk poses theoretical challenges.
-Even though GI works better at large distances than in the inner disk (e.g., \citealt{rafikov05}), fine-tuning is still required in terms of disk conditions and infall history for planetary mass objects to be the outcome of GI \citep{kratter10}.
-On the other hand, the time to accrete the solid core is too long in the outer disk.
-Some models, however, allow for rapid accretion of solids even at large separations (e.g., \citealt{dones93}, \citealt{kenyon09}, \citealt{lambrechts12}).
-We thus consider the timescale constraints imposed by the accretion of the gaseous atmosphere before the protoplanetary disk dissipates.
-
-%On the one hand, GI generates objects that are too massive to explain the current observed properties of exoplanets (e.g., \citealt{rafikov05}).
-On the other hand, planetesimal accretion is slow at large distances in the disk, and therefore large cores may not be able to form before the dissipation of the gas in the disk (e.g., \citealt{rafikov06}).
-It would therefore be easier if giant planets could form from smaller cores which need less time to grow.
-
-Several types of core accretion models have been studied previously.
-In static models, which are mostly analytic, the accretion rate of planetesimals regulates the growth of the atmosphere  (e.g., \citealt{stevenson82}, \citealt{rafikov06}).
-The luminosity of the gaseous envelope is thus entirely supplied by planetesimal accretion.
-These studies consider that the atmosphere is in a steady state at all times, in which all the incoming energy from planetesimal accretion is radiated away by the envelope.
-For a fixed planetesimal accretion rate, the atmosphere mass grows non-linearly as the core mass increases.
-As the atmosphere mass becomes comparable to the mass of the solid core, hydrostatic balance no longer holds and the static model breaks down.
-Unstable collapse occurs if the core mass further increases, a process known as the ``core accretion instability''.
-The maximum core mass for which the atmosphere is still in hydrostatic equilibrium is defined as the ``critical core mass''.
-Typically, the critical core mass is found to be about $10 M_{\oplus}$ (e.g., \citealt{stevenson82}, \citealt{rafikov06}).
-
-%Standard static calculations of giant planet formation that consider planetesimal accretion as the main source of luminosity (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{rafikov06}) assume that the atmosphere is at all times in a steady state at which the luminosity due to planetesimal accretion is fully radiated away by the atmosphere.
-Therefore, for a given set of disk conditions, there is a maximum mass the core can have before the onset of unstable accretion: this is defined as the critical core mass.
-
-%    In this situation,
-
-%the atmosphere evolution is dominated by the contraction of the envelope rather than planetesimal accretion.
-The atmosphere is no longer in a steady state, but rather it accretes gas as it loses energy through radiation.
-%In this scenario, the planetesimal accretion rate is significantly slowed down during the gas contraction phase of the atmosphere.
-This reduction can arise, for example, due to dynamical clearing, or due to the core having formed in the inner parts of the disk and migrated outwards.
-In this situation, the atmosphere evolution is dominated by the contraction of the envelope rather than planetesimal accretion.
-
-Full time-dependent simulations of giant planet evolution take into account the Kelvin-Helmholtz (KH) contraction of the gaseous envelope.
+Forming giant planets at wide separations in the protoplanetary disk poses theoretical challenges for both theories.
+While disks cannot fragment close to their host star, GI is more promising in the outer disk \citep{matzner05, rafikov05}.
+However,  for GI to form planetary mass objects versus brown dwarfs, both instantaneous disk conditions and the history of gas infall must be finely tuned \citep{kratter10}.
+The main concern for core accretion models, which are successful in the inner disk, is that they operate too slowly in the outer disk.
+The few Myr lifetime of gas disks sets the constraint \citep{}.
+However, mechanisms for the rapid accretion of solids, even at large separations, exist \citep{dones93, kenyon09, lambrechts12}.
+We thus consider the complementary problem of gas accretion timescales in the outer disk.
+
+Core accretion models fall in several categories: static, quasi-static evolutionary and hydrodynamic.
+In static models, the accretion of planetesimals provides a steady luminosity that determines the structure and mass of the atmosphere \citep{stevenson82}.
+For a given planetesimal accretion rate, static solutions only exist up to a maximum core mass, referred to as the ``critical core mass",  $\MC$.
+Above this mass, the atmosphere is assumed to collapse and rapidly accrete disk gas, a process known as the ``core accretion instability."  Many studies reproduce a canonical $\MC \sim 10 M_\oplus$.
+However, \citet[hereafter R06]{rafikov06} finds a wide range of values, $0.1 M_\oplus \lesssim \MC \lesssim 100 M_\oplus$, depending on the planetesimal accretion rate and disk parameters \citep[see also][]{Raf11}.
+A limitation of static models is the neglect of heat generated by atmospheric collapse, which can transform the core accretion instability into a slower process of Kelvin-Helmholtz (KH) contraction.
+
+Quasi-static models include time-dependent atmospheric evolution due to KH contraction, and typically include a time-varying planetesimal accretion rate \citep{boden86, alibert05}.
+Successful quasi-static models describe three phases of giant planet formation \citep{pollack96}.
+In phase 1, rapid planetesimal accretion gives significant core growth, but prevents the accretion of a massive atmosphere.
+When planetesimal accretion abates, as the core's feeding zone depletes, phase 2 begins.
+During phase 2, the atmosphere grows gradually, cooling by KH contraction.
+An ongoing trickle of planetesimal accretion can heat the atmosphere and slow this contraction.
+Eventually the atmosphere reaches the ``crossover mass" when it equals the mass of the now slowly growing core.
+Phase 3, the runaway growth of the atmosphere, begins around the crossover mass.
+This runaway occurs quasi-statically, i.e.\ in hydrostatic balance, but very rapidly compared to disk lifetimes.
+
+Dynamical models are needed to understand how runaway growth ends and final planet masses are determined.
+Relevant processes -- including gap opening in disks and the transition of accretion from spherical to planar -- can be simulated and then added to atmospheric evolution calculations \citep{LisHub09}.
+
+This work develops quasi-static models in which the core mass is fixed and no longer accreting solids.
+By isolating the process of KH contraction, we determine the minimum time scale for runaway atmospheric growth, and thus for giant planet formation by core accretion.
+Previous studies have considered this type of growth \citep{ikoma00, pn05}, which is an end-member case of phase 2 and 3 growth in more detailed quasi-static models.
+Our study develops simplified cooling models for the purposes of broadly exploring parameter space -- especially in the outer disk -- and developing a more detailed understanding of atmospheric accretion.
+
+%Full time-dependent simulations of giant planet evolution take into account the Kelvin-Helmholtz (KH) contraction of the gaseous envelope.
 As a result, the atmosphere is no longer in a steady state, but rather it accretes gas as it cools radiatively.
@@
 
-%However, full time dependent simulations for the evolution of giant planets do not assume a constant planetesimal accretion rate, like the studies above, and instead calculate the planetesimal accretion rate as part of an evolutionary model for the core and the envelope.
-In these studies, the atmosphere luminosity is supplied not only by planetesimal accretion, but also by the Kelvin-Helmholtz (KH) contraction of the envelope itself.
-\citet{pollack96} find that planetesimal accretion significantly slows down before a massive gaseous envelope can form, due to depletion of the planet's feeding zone, and gas accretion starts to dominate the atmosphere evolution.
-In this situation, planetesimal accretion can no longer balance radiative losses, and the atmosphere cools and contracts .
-This phase of slow gas accretion is found to be much longer than the formation of the solid core (\citealt{pollack96}, \citealt{alibert05}), hence the evolutionary timescale of the atmosphere is set by KH contraction rather than planetesimal accretion.
-
-%Aside from dominating the envelope accretion timescale in many core accretion models, KH contraction sets  To investigate these issues, we consider the growth of atmospheres in the limiting case of a fixed core mass with no ongoing accretion of solids.
-We thus obtain absolute lower limits on the critical core mass because additional heat sources limit the ability of the atmosphere to cool and undergo Kelvin-Helmholtz contraction.
-
-Since KH contraction dominates the envelope accretion timescale in many core accretion models, it is useful to study atmospheric growth onto a core of fixed mass (\citealt{ikoma00}, \citealt{pn05}).
+%Since KH contraction dominates the envelope accretion timescale in many core accretion models, it is useful to study atmospheric growth onto a core of fixed mass (\citealt{ikoma00}, \citealt{pn05}).
 This regime is the one considered in this work.
@@
 
-% Specifically, we study the atmosphere evolution of giant planets under the assumption that the solid core is fully grown and thus the planetesimal accretion rate is negligible.
-
-%As a result, the luminosity of the gaseous envelope is solely supplied by atmospheric contraction.
-The envelope is not in a steady state, but rather it accretes gas as it cools radiatively.
-We consider that the atmosphere evolves in time through stages of quasi-static equilibrium.
-Once the mass of the gaseous envelope becomes comparable to the mass of the solid core, the self-gravity of the atmosphere can no longer balance the pressure gradient and a rapid phase of runaway gas accretion commences.
-We define the time required for the atmosphere to grow to this stage as the \textit{crossover time}.
-For a set of fixed gas and disk conditions, there exists a minimum core mass for which the atmosphere can grow on the time scale described above within the life time of the protoplanetary disk, which we define as the \textit{critical core mass}.
-
-%The atmosphere is no longer in a steady state, but rather it accretes gas as it loses energy through radiation.
-In our model we therefore assume that the luminosity evolution of the atmosphere is dominated by gas accretion, while the planetesimal accretion rate is negligible.
-As a result, the protoplanetary core has a fixed mass.
-
-%In this study, we show that giant planets can grow faster from small protoplanetary cores that are fully formed before significant gas accretion occurs.
-
-%We build a two-layer atmosphere model, with a convective inner region and a radiative outer region that matches smoothly on to the protoplanetary disk, and develop a cooling model that evolves the atmosphere in time.
-We aim to find the critical core mass for a giant planet to form before disk dissipation.
-
 This paper is organized as follows.
-In section \S\ref{sec:model} we describe the assumptions of our atmosphere model, and derive the basic equations that govern the structure and evolution of the atmosphere.
-In section \S\ref{sec:coolingan}, we present a simplified analytic model that predicts the qualitative behavior of the numerical model.
-We describe our numerical results in section \S\ref{sec:KH}, and determine the critical core mass for planet formation during the life time of the protoplanetary disk in section \S\ref{sec:critical}.
-We discuss some of the approximations of our model in section \S\ref{sec:neglected} and summarize our findings in section \S\ref{sec:conclusions}.
-
-%Some previous studies of atmosphere accretion (e.g., \citealt{stevenson82}, \citealt{wuchterl93}, \citealt{rafikov06}) consider static envelopes, in which the luminosity is solely supplied by planetesimal accretion and fully radiated away by the atmosphere.
-In other studies, the time evolution is explicitly taken into account and full time dependent models are developed (e.g., \citealt{ikoma00}).
-We follow an intermediate approach and consider quasi static evolution.
-Our model for the atmosphere growth time is described in Section \ref{...}.
-
- %The simplified treatment allows us to explore and better understand  the effect of crucial parameters including core mass, disk properties and opacity.
-
-\section{Atmosphere Models} \label{sec:model}
+Section \ref{sec:model} describes our model of atmospheric growth.
+In Section \S\ref{sec:coolingan}, we develop a simplified analytic version of our atmospheric model.
+We describe the structure and evolution of our atmosphere solutions in Section \S\ref{sec:KH}.
+Section \S\ref{sec:critical} gives our final results for growth timescales and critical core masses.
+We discuss some neglected effects in section \S\ref{sec:neglected} and summarize our findings in Section \S\ref{sec:conclusions}.
+Some detailed derivations are presented in the appendices.
+
+\section{Atmospheric Accretion Model} \label{sec:model}
 
@@
 Beyond the crossover mass,  our approximate treatment of the radiative zone (explained below) breaks down.
-Since subsequent growth is a rapid runaway process (e.g., \citealt{pollack96}), our model can investigate to good accuracy the timescale requirement of core accretion.
+Since subsequent growth is a rapid runaway process, our model can investigate to good accuracy the timescale requirement of core accretion.
 Our simplified treatment is also inappropriate for hot, short-period planets, where dust sublimation gives deeper radiative zones that require more detailed models.

### 2013-10-27 tex/draft/draft4.tex

@@
 This section develops the analytic version of our two-layer model.
-This analytic model is less accurate, primarily because it neglects self-gravity which we show (in \S\ref{sec:KH}) to be important at rather low atmospheric masses, $M_{\rm atm} \gtrsim 0.1 M\co$.
-Nevertheless the analytic model is useful for understanding atmospheric evolution and interpreting the numerical results.
-
-The analytic model also assumes that the upper radiative radiative layer is thick enough that $P\cb \gg P\di$.
+This analytic model is less accurate than our numerical model, primarily because it neglects self-gravity, which we show (in \S\ref{sec:KH}) to be important at rather low atmospheric masses, $M_{\rm atm} \gtrsim 0.1 M\co$.
+Nevertheless, the analytic model is useful for understanding atmospheric evolution and interpreting the numerical results.
+
+The analytic model also assumes that the upper radiative layer is thick enough that $P\cb \gg P\di$.
 This approximation ignores the earliest stages of cooling, which are rapid enough to be negligible.
-Finally the analytic model ignores the surface terms in \Eq{eq:coolingglobal}, which we show to be a modest correction in \S\ref{sec:endoftime}.
+Finally, the analytic model ignores the surface terms in \Eq{eq:coolingglobal}, which we show to be a modest correction in \S\ref{sec:endoftime}.
 
 \subsection{Two Layer Structure}
-To apply the two-layer cooling model analytically, we require expressions for the atmospheric structure.
+In order to apply the two-layer cooling model analytically, we require expressions for the atmospheric structure.
 Conditions at the RCB are crucial as they set the interior adiabat and the radiative losses from the interior.
@@
 This logarithmic factor complicates our analytic description.
-Since it remains order unity we simply hold it fixed in our scalings, but not in plots.
+Since it remains order unity, we simply hold it fixed in our scalings, but not when plotting the results.
 

### 2013-10-27 tex/draft/draft4.tex
minor intro edits

@@
 
-Models of giant planet fall in two categories:  core accretion or gravitational instability \citep{dangelo11, youdin13}.
+Models of giant planet formation fall in two categories:  core accretion or gravitational instability \citep{dangelo11, youdin13}.
 In core accretion models, a solid core grows until it becomes massive enough to rapidly accrete gas \citep{PerCam74, mizuno78}.
@@
 
-Forming giant planets at wide separations in the protoplanetary disk poses theoretical challenges for both theories.
+Forming giant planets at wide separations in the protoplanetary disk poses theoretical challenges for both models.
 While disks cannot fragment close to their host star, GI is more promising in the outer disk \citep{matzner05, rafikov05}.
@@
 The main concern for core accretion models, which are successful in the inner disk, is that they operate too slowly in the outer disk.
-The few Myr lifetime of gas disks sets the constraint \citep{}.
+The few Myr lifetime of gas disks sets the constraint \citep{hillenbrand05}.
 However, mechanisms for the rapid accretion of solids, even at large separations, exist \citep{dones93, kenyon09, lambrechts12}.
@@
 In phase 1, rapid planetesimal accretion gives significant core growth, but prevents the accretion of a massive atmosphere.
-When planetesimal accretion abates, as the core's feeding zone depletes, phase 2 begins.
+When planetesimal accretion abates, as the core's feeding zone is depleted of solids, phase 2 begins.
 During phase 2, the atmosphere grows gradually, cooling by KH contraction.
 An ongoing trickle of planetesimal accretion can heat the atmosphere and slow this contraction.
-Eventually the atmosphere reaches the ``crossover mass" when it equals the mass of the now slowly growing core.
+Eventually, the atmosphere reaches the ``crossover mass", when it equals the mass of the now slowly growing core.
 Phase 3, the runaway growth of the atmosphere, begins around the crossover mass.
@@
 This work develops quasi-static models in which the core mass is fixed and no longer accreting solids.
-By isolating the process of KH contraction, we determine the minimum time scale for runaway atmospheric growth, and thus for giant planet formation by core accretion.
+By isolating the process of KH contraction, we determine the minimum timescale for runaway atmospheric growth, and thus for giant planet formation by core accretion.
 Previous studies have considered this type of growth \citep{ikoma00, pn05}, which is an end-member case of phase 2 and 3 growth in more detailed quasi-static models.

### 2013-10-28 tex/draft/draft4.tex
minor overall edits

@@
 The minimum (or critical) core mass to form a gas giant is typically quoted as $10 M_{\oplus}$.
-This value, however, depends on several factors, such as the location in the disk, atmospheric opacity, and the accretion rate of solids.
+This value, however, depends on several factors, such as the location in the protoplanetary disk, atmospheric opacity, and the accretion rate of solids.
 Motivated by ongoing direct imaging searches for giant planets, this study focuses on core mass requirements in the outer disk.
@@
 
-This work develops quasi-static models in which the core mass is fixed and no longer accreting solids.
+This work develops quasi-static models in which the core has a fixed mass and is no longer accreting solids.
 By isolating the process of KH contraction, we determine the minimum timescale for runaway atmospheric growth, and thus for giant planet formation by core accretion.
@@
 This paper is organized as follows.
-Section \ref{sec:model} describes our model of atmospheric growth.
+Section \S\ref{sec:model} describes our model of atmospheric growth.
 In Section \S\ref{sec:coolingan}, we develop a simplified analytic version of our atmospheric model.
@@
 Section \S\ref{sec:critical} gives our final results for growth timescales and critical core masses.
-We discuss some neglected effects in section \S\ref{sec:neglected} and summarize our findings in Section \S\ref{sec:conclusions}.
+We discuss some neglected effects in Section \S\ref{sec:neglected} and summarize our findings in Section \S\ref{sec:conclusions}.
 Some detailed derivations are presented in the appendices.
@@
 \begin{equation}
-c\di = \sqrt{\Rg T\di} = 0.4 \sqrt{F_T} \aun{10}^{3/14} ~\text{km s}^{-1}
+c\di = \sqrt{\Rg T\di} = 0.4 \sqrt{F_T} \aun{10}^{3/14} ~\text{km s}^{-1} \,,
 \end{equation}  
@@
 \begin{equation}
-H\di = {c\di / \varOmega} = 0.42 \sqrt{F_T}  \, \aun{10}^{9/7} \AU\, .
+H\di = {c\di / \varOmega} = 0.42 \sqrt{F_T}  \, \aun{10}^{9/7} \AU\, ,
 \end{equation} 
@@
 \end{equation}
-with a powerlaw index $\beta = 2$ and normalization $F_\kappa = 1$ unless noted otherwise.
+with a power-law index $\beta = 2$ and normalization $F_\kappa = 1$ unless noted otherwise.
 Grain growth tends to lower both $F_\kappa$ and $\beta$, while dust abundance scales with $F_\kappa$.
@@
 \label{eq:RB}
-\RB \equiv \frac{G M\pla}{c\di^2} \approx 0.17 \, {\mpn{10}  \, \aun{10}^{3/7} \over F_T} ~\AU
+\RB \equiv \frac{G M\pla}{c\di^2} \approx 0.17 \, {\mpn{10}  \, \aun{10}^{3/7} \over F_T} ~\AU,
 \end{equation}
@@
 where hydrostatic balance breaks down.
-In \Eq{eq:RHill}, $M\pla$ includes mass enclosed within $\RH$.
+In \Eq{eq:RHill}, $M\pla$ includes the mass enclosed within $\RH$.
 During the early stages of evolution, $M\pla \sim M\co$, and the core mass can be used to get estimates of both $\RB$ and $\RH$.
@@
 
-In the low mass regime, $M\pla < M_{\rm th}/\sqrt{3}$, the length scales order as $\RB< \RH<H\di$ (\citealt{rafikov06}, hereafter R06).
+In the low mass regime, $M\pla < M_{\rm th}/\sqrt{3}$, the length scales order as $\RB< \RH<H\di$ (R06).
 In this regime, many studies assume the atmosphere matches the disk conditions at $\RB$.
@@
 
-At higher planet masses where $M\pla > 3 M_{\rm th}$ and $H\di < \RH < \RB$, spherical symmetry is no longer a good approximation, due to both the vertical stratification of the disk and gap opening.
-See \S\ref{sec:hydro} for discussion of neglected non-hydrostatic effects on all mass scales.
-
-Note that while $\RH$ is the outer boundary of our structure calculations, we define planet masses to include only the mass inside the smaller of $\RB$ or $\RH$.
+At higher planet masses where $M\pla > 3 M_{\rm th}$ and $H\di < \RH < \RB$, spherical symmetry is no longer a good approximation, due to both the vertical stratification of the disk and gap opening (see \S\ref{sec:hydro} for a discussion of neglected non-hydrostatic effects on all mass scales).
+
+Note that while $\RH$ is the outer boundary in our structure calculations, we define planet masses to include only the mass inside the smaller of $\RB$ or $\RH$.
 This conservative choice in quoting planet masses is usually a minor distinction because (when $\RB < \RH$) the gas between $\RB$ and $\RH$ is weakly compressed.
@@
 
-Boundary conditions must be satisfied at both the base and the top of the atmosphere with $m(R\co) = M\co$, $T(\RH) = T\di$ and $P(\RH) = P\di$.
+Boundary conditions must be satisfied at both the base and the top of the atmosphere, with $m(R\co) = M\co$, $T(\RH) = T\di$ and $P(\RH) = P\di$.
 In principle, our solutions describe atmospheres with $L(R\co) = 0$.
@@
 
-To simplify our calculations of atmospheric contraction we use a two-layer model with a bottom convective region and an upper radiative layer.
+To simplify our calculations of atmospheric contraction, we use a two-layer model with a bottom convective region and an upper radiative layer.
 The existence of such a structure is well known from previous studies (e.g., R06) and can be readily understood.
@@
 \end{equation} 
-to simplify expressions, with $C_P = \Rg / \delad$ the specific heat at constant pressure.
+to simplify expressions, with $C_P = \Rg / \delad$ the specific heat capacity at constant pressure.
 
@@
 This logarithmic factor complicates our analytic description.
-Since it remains order unity, we simply hold it fixed in our scalings, but not when plotting the results.
-
-The total energy is concentrated towards the core if $|u| \rho  r^3 \propto \rho r^2$ drops with increasing $r$.
+Since it remains order unity, we simply hold it fixed in our scalings.
+%, but not when plotting the results.
+
+The total energy is concentrated towards the core if $|e| \rho  r^3 \propto \rho r^2$ drops with increasing $r$.
 This condition requires $\gamma < 3/2$, which our choice of $\gamma = 7/5$ satisfies, but $\gamma = 5/3$ (monatomic gas) would not.
@@
 &\approx& - 4 \pi P\cb {\RB'}^{1 \over \nabla_{\rm ad}} \left(\gamma-1 \over 3 - 2 \gamma\right)  R\co^{2\gamma-3\over \gamma-1}  \slabel{eq:Ean} \\ 
-&\approx& - 8 \pi P\cb {\RB'^{7/2} \over \sqrt{R\co}} \slabel{eq:Eus}
+&\approx& - 8 \pi P\cb {\RB'^{7/2} \over \sqrt{R\co}} \slabel{eq:Eus} \, ,
 %&\approx& -0.31 P\cb {(\RB / \chi)^{7/2} \over \sqrt{R\co}}
@@
 We present examples of the structure and evolution of our model atmospheres, calculated as described in \S\ref{sec:model}, with some comparisons to the non-self-gravitating analytic model of \S\ref{sec:coolingan}.
-Radial structure is presented in \S\ref{sec:profiles}, time evolution is described in \S\ref{sec:timeev}, and validity of our two-layer cooling model is examined in \S\ref{sec:endoftime}.
+Radial structure is presented in \S\ref{sec:profiles}, time evolution is described in \S\ref{sec:timeev}, and the validity of our two-layer cooling model is examined in \S\ref{sec:endoftime}.
 
@@
 However, the main effect is that \Eq{eq:crit} -- which describes a nearly linear relation between atmospheric mass and RCB pressure --  breaks down for self-gravitating solutions.
-This behavior can be seen in the top panel of \Fig{fig:profiles} where the $P\cb$ increases significantly from 5.10 to 6.0 $M_\oplus$, but only increases relatively modestly with further growth to 9.24 $M_\oplus$.
+This behavior can be seen in the top panel of \Fig{fig:profiles} where the $P\cb$ increases significantly from 5.10 to 6.0 $M_\oplus$, but only increases relatively modestly with further growth to 8.99 $M_\oplus$.
 In the higher mass solutions, the relatively low $P\cb$ values (and thus the relatively high luminosities) require an outward shift in $R\cb$, as shown in \Fig{fig:profiles}.
@@
 The fact that growth times have a well defined maximum is a characteristic of accelerating growth.
-Unlike our analytic model, which must assume that runaway growth begins near the crossover mass, our numerical model allows us to measure when runaway accretion begins.
+Unlike our analytic model, which must assume that runaway growth begins near the crossover mass, our numerical model allows us to measure when runaway accretion starts.
 Runaway growth clearly does not begin at a universal value of $M_{\rm atm}/M\co$.
@@
 As the mass increases, the surface terms become more significant, led by the accretion energy.
-However the surface terms are everywhere smaller than $L_{\rm negl}$.
+However, the surface terms are everywhere smaller than $L_{\rm negl}$.
 Thus wherever our model is accurate -- including the crucial early phases of growth -- surface terms are a minor correction.
@@
 
-\subsection{Runaway Growth Time Scale}
+\subsection{Runaway Growth Timescale}
 \label{sec:tcross}
 The time to undergo atmospheric runaway growth, $t_{\rm run}$, sets a minimum timescale for the formation of giant planets.
-Due to the accelerating nature of runaway growth, the precise threshold chosen for $t_{\rm run}$ (explained in \S ?) is of minor significance.
+Due to the accelerating nature of runaway growth, the precise threshold chosen for $t_{\rm run}$ (explained in \S\ref{sec:timeev}) is of minor significance.
 
@@
 Thus by restricting our attention to low masses, neglected hydrodynamic effects should be minor.
-Moreover since $M_{\rm th} \propto a^{6/7}$ increases with disk radius, spherical hydrostatic models like ours have a greater range of applicability in the outer regions of disks.
+Moreover, since $M_{\rm th} \propto a^{6/7}$ increases with disk radius, spherical hydrostatic models like ours have a greater range of applicability in the outer regions of disks.
 
@@
 To derive  the global energy equation (\ref{eq:coolingglobal}) for an embedded protoplanet, we generalize the analogous calculations in stellar structure theory, e.g.\ in \S4.3 of \citet{kippenhahn90}.
-For our problem, we add the effects of finite core radius, surface pressure and mass accretion.We start with the local energy equation (\ref{eq:structd}) whose more natural form in Lagrangian (mass) coordinates is $\p L/ \p m = \epsilon - T \p S /\p t$.
+For our problem, we add the effects of finite core radius, surface pressure and mass accretion.
+We start with the local energy equation (\ref{eq:structd}), whose more natural form in Lagrangian (mass) coordinates is $\p L/ \p m = \epsilon - T \p S /\p t$.
 Integrating from the core to a higher shell with enclosed mass $M$ gives:
@@
 \end{subeqnarray} 
-with $\Gamma = \int \epsilon dm$ the integral of the direct heating rate and applying the first law of thermodynamics in the final step.
+with $\Gamma = \int \epsilon dm$ the integral of the direct heating rate, and applying the first law of thermodynamics in the final step.
 
@@
 \end{subeqnarray} 
-This derivation holds the core mass and radius fixed, $\dot{M}\co = \dot{R}\co = 0$.  Therefore the core pressure satisfies
+This derivation holds the core mass and radius fixed, $\dot{M}\co = \dot{R}\co = 0$.  Therefore, the core pressure satisfies
 \begin{equation}\label{eq:Pcdot}
@@
 \end{equation} 
-To make further progress we use the virial theorem:
+To make further progress, we use the virial theorem:
 \begin{equation}
@@
 E&=&(1-\zeta)U+4 \pi (R^3 P_M-R\co^3 P\co) \slabel{eq:vira} \\
-&=&\frac{\zeta-1}{\zeta}E_G+\frac{4 \pi}{\zeta} (R^3 P_M-R\co^3 P\co) \slabel{eq:virb}
+&=&\frac{\zeta-1}{\zeta}E_G+\frac{4 \pi}{\zeta} (R^3 P_M-R\co^3 P\co) \slabel{eq:virb} \, ,
 \end{subeqnarray}
@@
 
-To express the work integral, the final term in \Eq{eq:DLc}, in terms of changes to gravitational energy we first take the
+To express the work integral, i.e. the final term in \Eq{eq:DLc}, in terms of changes to gravitational energy, we first take the
  time derivative of \Eq{eq:virial}:
@@
 %where the volumes, $V_M = 4 \pi R^3/3$ and $V\co = 4 \pi R\co^3/3$. 
-The first integral in \Eq{eq:EGdot} is the one we want, but the next one must be eliminated.  The time derivative of \Eq{eq:Eg} (times four) gives
+The first integral in \Eq{eq:EGdot} is the one we want, but the second one must be eliminated.  The time derivative of \Eq{eq:Eg} (times four) gives
 \begin{subeqnarray}
@@
 &=&   -4 {G M \dot{M} \over R} + 4 \pi \int_{M\co}^M r^3{\p \over \p m}{\p P \over \p t} dm \slabel{eq:4EGb} \\
-&=&  -4 {G M \dot{M} \over R} -3  \int_{M\co}^M {\p P\over \p t}{dm \over \rho}  + 3 V_M {\p P_M \over \p t} -3 V\co {\p P\co \over \p t} \slabel{eq:4EGc}
+&=&  -4 {G M \dot{M} \over R} -3  \int_{M\co}^M {\p P\over \p t}{dm \over \rho}  + 3 V_M {\p P_M \over \p t} -3 V\co {\p P\co \over \p t} \slabel{eq:4EGc} \, ,
 \end{subeqnarray} 
@@
 
-We consider the structure of a non self-gravitating isothermal atmosphere that extends outward from the radiative-convective boundary RCB and matches onto the disk density, $\rho_{\rm d}$, at a distance $r_{\rm fit} = n_{\rm fit} \RB$, where $R_{\rm B}$ is the Bondi radius defined in equation (\ref{eq:RB}).
+We consider the structure of a non-self-gravitating, isothermal atmosphere that extends outward from the radiative-convective boundary (RCB) and matches onto the disk density, $\rho_{\rm d}$, at a distance $r_{\rm fit} = n_{\rm fit} \RB$, where $R_{\rm B}$ is the Bondi radius defined in equation (\ref{eq:RB}).
 From equation (\ref{eq:structa}) the resulting density profile is
@@
 \begin{equation} \label{eq:MatmISO}
-M_{\rm iso} \approx 4 \pi \rho\di {R\cb^4 \over R_{\rm B}} e^{R_B/R\cb} = 4\pi \rho\cb \frac{R\cb^4}{R_{\rm B}}
+M_{\rm iso} \approx 4 \pi \rho\di {R\cb^4 \over R_{\rm B}} e^{R_B/R\cb} = 4\pi \rho\cb \frac{R\cb^4}{R_{\rm B}} \, ,
 \end{equation}
@@
 %\subsection{Temperature Contrast at Convective Boundary}
-In this section we estimate the temperature and pressure corrections at the RCB due to the fact that the radiative region is not purely isothermal.
-From equation (\ref{eq:delrad}), we express the radiative lapse rate
+We estimate the temperature and pressure corrections at the RCB due to the fact that the radiative region is not purely isothermal. From equation (\ref{eq:delrad}), we express the radiative lapse rate
 \begin{equation}\label{eq:delradan}
@@
 \end{equation} 
- We can then determine the radius of the convective boundary $R\cb$ from the \Eq{eq:structa} as 
+ We can then determine the RCB radius $R\cb$ from \Eq{eq:structa} as 
 \begin{equation}\label{eq:RCBint}
@@
 A  lower opacity  decreases the critical core mass.
-Reducing the opacity by a factor of one hundred results in a critical core mass one order of magnitude lower than in the standard ISM case, specifically  $\MC \approx9 M_\oplus$ for the parameters in \Eq{eq:Mcrit}.
+Reducing the opacity by a factor of one hundred results in a critical core mass one order of magnitude lower than in the standard ISM case, for our analytic model.
 The reduction is not  as strong as the nominal scaling would imply, $0.01^{3/5} \approx 0.06$, because $\xi$ increases.

### 2013-10-28 tex/draft/draft4.tex
edits to section 2, mostly

@@
 We also develop a non-self-gravitating, analytic cooling model to help explain our numerical trends.
-Additionally, this comparison reveals that self-gravity significantly affects atmospheric evolution even when the atmosphere is only $\sim$$20\%$ as massive as the core.
-
+Additionally, this comparison reveals that self-gravity significantly affects atmospheric evolution even when the atmosphere is only $\sim$$10\%$ as massive as the core.
 \end{abstract}
-
-%Current theories of giant planet formation postulate that these planets form either through core accretion (refs), in which solid planetesimals collide and grow into a massive solid core, which then accretes a gaseous envelope, or due to a gravitational instability in the protoplanetary disk that leads to fragmentation of the disk into self-gravitating clumps (refs, inc.
-\citealt{dangelo11}).
-%quote murray-clay, kratter 10, rafikov 05, etc.
-%
-%Standard core accretion models (refs) assume that the core and the atmosphere grow at the same time, and that planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse, while the heat generated by the gravitational (Kelvin-Helmholtz) contraction of the atmosphere is neglected.
-These studies consider that the planet atmosphere is in steady state, in which all the luminosity due to planetesimal accretion is radiated away by the envelope, and  find that there exists a minimum (``critical'') core mass past which hydrostatic solutions can no longer be found and unstable atmosphere collapse occurs.
-%
-%Forming giant planets at wide separations in the disk poses theoretical challenges.
-On the one hand, gravitational instability generates objects that are too massive to explain the current observed properties of exoplanets (refs, inc.
-\citealt{rafikov05}).
-On the other hand, planetesimal accretion is slow at large distances in the disk, and therefore large cores may not be able to form before the dissipation of the disk (refs).
-I would therefore be easier if giant planets could form from smaller cores which would need less time to grow.
-%
-%In this study, we show that giant planets can grow faster from small protoplanetary cores that are fully formed before significant gas accretion occurs.
-In this scenario, the planetesimal accretion rate is significantly slowed down during the gas contraction phase of the atmosphere.
-This reduction can arise due to dynamical clearing, or due to the core having formed in the inner parts of the disk and migrated outwards, etc.
-In this situation, the atmosphere evolution is dominated by the Kelvin-Helmholtz contraction of the envelope.
-The atmosphere is no longer in a steady state, but rather it accretes gas as it loses energy through radiation.
-%
-%In our model we therefore assume that the luminosity evolution of the atmosphere is dominated by gas contraction, while the planetesimal accretion rate is negligible.
-As a result, the protoplanetary core has a fixed mass.
-We consider that the atmosphere evolves in time through stages of quasi-static equilibrium.
-Once the mass of the gaseous envelope becomes comparable to the mass of the solid core, the self-gravity of the atmosphere can no longer balance the pressure gradient and unstable hydrodynamic collapse commences.
-The time required for the atmosphere to grow to this stage is the characteristic growth time of the atmosphere.
-For a set of fixed gas and disk conditions, there exists a minimum core mass for which the atmosphere can grow on the time scale described above within the life time of the protoplanetary disk, which we define as the ``critical core mass''.
-%
-%We develop a two-layer atmosphere model, with a convective inner region and a radiative outer region that matches smoothly on to the protoplanetary disk, and develop a cooling model that evolves the atmosphere in time.
-We aim to find the critical core mass for a giant planet to form before the dissipation of the disk.
-
-%Kelvin-Helmholz (KH) contraction considered previously (Ikoma, PapNel05).
-Say why useful.
-This paper develops a simplified model of KH contraction to both elucidate the relevant physical processes and explore a wide range of disk parameter space...
-
-%This paper is organized as follows.
-In Section \ref{sec:model} we describe the assumptions of our atmosphere model, and derive the basic equations that govern the structure and evolution of the atmosphere.
-In Section \ref{sec:coolingan}, we present a simplified analytic model that predicts the qualitative behavior of the numerical model.
-Results for atmospheric structure and evolution are presented in Section \ref{sec:KH}, and implications for the critical core mass are presented in Section \ref{sec:critical}.
-The discussion in Section \ref{sec:neglected} addresses various approximations and neglected effects.
-We summarize our findings in Section \ref{sec:conclusions}.
-The appendices contain...
 
@@
 Gravitational instability (GI) theories investigate the fragmentation of the protoplanetary disk into bound clumps \citep{cameron78, boss97}.
-%quote murray-clay, kratter 10, rafikov 05, etc.
-
-%Detailed core accretion studies (\citealt{boden86}, \citealt{pollack96}, \citealt{alibert05}) have shown that this giant planet formation model can be separated in three phases.
-In phase 1 \citep{alibert05} a solid core forms by accreting planetesimals from its feeding zone at a high rate.
 
@@
 By isolating the process of KH contraction, we determine the minimum timescale for runaway atmospheric growth, and thus for giant planet formation by core accretion.
-Previous studies have considered this type of growth \citep{ikoma00, pn05}, which is an end-member case of phase 2 and 3 growth in more detailed quasi-static models.
+Previous studies have considered this type of growth \citep{ikoma00, pn05}, which is an end-member case of phases 2 and 3 in more detailed quasi-static models.
+Especially since phase 2 atmospheric growth is often the slowest stage of core accretion, understanding the fastest allowed rates of gas accretion is crucial importance.
 Our study develops simplified cooling models for the purposes of broadly exploring parameter space -- especially in the outer disk -- and developing a more detailed understanding of atmospheric accretion.
-
-%Full time-dependent simulations of giant planet evolution take into account the Kelvin-Helmholtz (KH) contraction of the gaseous envelope.
-As a result, the atmosphere is no longer in a steady state, but rather it accretes gas as it cools radiatively.
-In this scenario there are two main sources that contribute to the luminosity evolution of the envelope --- planetesimal accretion and gas contraction.
-These evolutionary models for core and atmosphere growth include time-dependent planetesimal accretion rates that are determined by the distribution of solids in the disk, as well as the planet mass and capture cross-section.
-This is is contrast with the static models in which the planetesimal accretion rate is usually constant with time.
-\citet{pollack96} find that the planet growth can be separated into three relatively well defined phases.
-Phase 1 is dominated by the growth of the core due to runaway planetesimal accretion, and ends when the planet's feeding zone has been depleted of solids.
-At this stage, planetesimal accretion can no longer balance radiative losses, and the atmosphere cools and contracts.
-This is phase 2, during which the envelope mass steadily increases, while core growth is no longer significant.
-Once the atmosphere mass becomes comparable to the core mass, growth is continually accelerated, leading to runaway gas accretion, i.e.
-phase 3.
-Phase 2 is found to be much longer than the other two phases (see also \citealt{alibert05}), hence the evolutionary timescale of the atmosphere is set by KH contraction.
-
-%Since KH contraction dominates the envelope accretion timescale in many core accretion models, it is useful to study atmospheric growth onto a core of fixed mass (\citealt{ikoma00}, \citealt{pn05}).
-This regime is the one considered in this work.
-We explore the atmosphere evolution around fully grown cores of different masses, and determine the minimum core mass to form a giant planet before disk dissipation for a variety of disk conditions and properties of the nebular gas.
-We note that our models are a special case of the time-dependent studies, with the planetesimal accretion rate set to zero.
-Our neglect of planetesimal accretion allows us to obtain absolute lower limits on the core mass needed to form a gas giant, as additional heat sources limit the ability of the atmosphere to cool and undergo KH contraction.
 
@@
 
-Our model can accurately model the growth of the atmosphere up to the crossover mass, when the atmosphere mass equals the mass of the core.
-Beyond the crossover mass,  our approximate treatment of the radiative zone (explained below) breaks down.
-Since subsequent growth is a rapid runaway process, our model can investigate to good accuracy the timescale requirement of core accretion.
-Our simplified treatment is also inappropriate for hot, short-period planets, where dust sublimation gives deeper radiative zones that require more detailed models.
+Our model can  treat atmospheric growth up to the early stages of runaway growth, around the crossover mass.
+At higher masses, our approximate treatment of the radiative zone (explained below) breaks down.
+However, evolution after the onset of runaway growth contributes minimally to the total planet formation timescale, which we can model to good accuracy.
+%Our treatment is also inappropriate for hot, short-period planets, where dust sublimation gives deeper radiative zones that require more detailed models.
 
@@
 \item The core mass and radius are fixed in evolutionary calculations, neglecting ongoing planetesimal or dust accretion.
-\item At the planet's Hill radius, the atmospheric temperature and pressure match the conditions of the disk midplane.
+\item At the planet's Hill radius, the atmospheric temperature and pressure match the conditions in the disk midplane.
 \item The only source of planetary luminosity is the gravitational contraction of the atmosphere.  
-\item Cooling of the atmosphere is dominated by the convective interior.
-Thus the luminosity in the radiative exterior is held spatially constant.
-%In the radiative zone, luminosity generation is neglected, i.e.
-the luminosity is held constant.
+\item Cooling of the atmosphere is dominated by the convective interior.  Thus the luminosity in the radiative exterior is held spatially constant. 
 \item A global cooling model connects independent static solutions into a time-dependent sequence.
-\item A polytropic equation of state (EOS) is assumed for simplicity. %AP: define EOS here if not already
+\item A polytropic equation of state (EOS) is assumed for simplicity.
 \item Dust grains provide the opacity in the radiative zone, which remains cool enough to avoid dust sublimation.
-\item Because gas accretion accelerates after the crossover mass, the time to reach the crossover mass, i.e. the crossover time, is a good approximation of the total time to form a gas giant.
+%\item Because gas accretion accelerates after the crossover mass, the time to reach the crossover mass, i.e. the crossover time, is a good approximation of the total time to form a gas giant.
 \end{enumerate}
-The remainder of this section develops our model in more detail.
+%The remainder of this section develops our model in more detail.
 
@@
 
-The characteristic length scales for protoplanetary atmospheres are crucial for choosing boundary conditions and for understanding the validity of  spherical symmetry in a disk of scale height $H\di$.
-The radius of the solid core is
-
+The characteristic length scales for protoplanetary atmospheres are crucial for choosing boundary conditions and for understanding the validity of  spherical symmetry in a gas disk of scale height $H\di$.
+
+The solid core has a radius
 \begin{equation}
@@
 \end{equation}
-where the enclosed planet mass, $M\pla = M\co + M_\mathrm{atm}$, includes the core and any atmosphere within the Bondi radius.  The scalings thus use $\mpn{10} \equiv M\pla /(10~M_\oplus)$.
+where the enclosed planet mass, $M\pla = M\co + M_\mathrm{atm}$, includes the core and any atmosphere within the Bondi radius and $\mpn{10} \equiv M\pla /(10~M_\oplus)$.
+The contribution of the atmospheric mass is small in early evolutionary stages.
 
@@
 \end{equation}
-where hydrostatic balance breaks down.
-In \Eq{eq:RHill}, $M\pla$ includes the mass enclosed within $\RH$.
-During the early stages of evolution, $M\pla \sim M\co$, and the core mass can be used to get estimates of both $\RB$ and $\RH$.
+where hydrostatic balance breaks down.  (Here we use the same symbol, $M\pla$, to now mean the mass within $\RH$.)
 
@@
 
-In the low mass regime, $M\pla < M_{\rm th}/\sqrt{3}$, the length scales order as $\RB< \RH<H\di$ (R06).
+In the low mass regime, $M\pla < M_{\rm th}/\sqrt{3}$, the length scales order as $\RB< \RH<H\di$.
 In this regime, many studies assume the atmosphere matches the disk conditions at $\RB$.
-We, however, use $\RH$ as the matching radius in both this low mass and other higher mass regimes.
-This choice is justified by the fact that, for hydrostatic solutions, the  density at $\RB$ exceeds the disk's background density by an order unity factor (R06).
-Nevertheless, since the density change is modest, the choice of outer boundary condition has a similarly modest effect on our results.
+We use $\RH$ as the matching radius in all regimes.
+We are accounting for the fact that, in hydrostatic models, the density at
+ $\RB$ is higher than the background density (see R06). Nevertheless, since the density change is modest, these two boundary conditions give similar results.
 
@@
 
-At higher planet masses where $M\pla > 3 M_{\rm th}$ and $H\di < \RH < \RB$, spherical symmetry is no longer a good approximation, due to both the vertical stratification of the disk and gap opening (see \S\ref{sec:hydro} for a discussion of neglected non-hydrostatic effects on all mass scales).
-
-Note that while $\RH$ is the outer boundary in our structure calculations, we define planet masses to include only the mass inside the smaller of $\RB$ or $\RH$.
-This conservative choice in quoting planet masses is usually a minor distinction because (when $\RB < \RH$) the gas between $\RB$ and $\RH$ is weakly compressed.
+At higher planet masses where $M\pla > 3 M_{\rm th}$ and $H\di < \RH < \RB$, spherical symmetry is no longer a good approximation, due to both the vertical stratification of the disk and gap opening.
+See \S\ref{sec:hydro} for further discussion of neglected hydrodynamic effects.
+
+While $\RH$ is always the outer boundary in our structure calculations, we quote planet masses as the enclosed mass within the smaller of $\RB$ and $\RH$.
+%  This conservative choice in quoting planet masses is usually a minor distinction because (when $\RB < \RH$) the gas between $\RB$ and $\RH$ is weakly compressed.
 
@@
 \end{subeqnarray}
-\noindent where $r$ is the radial coordinate, and $P$, $T$, $\rho$  and $L$ are the gas pressure, temperature, density and luminosity, respectively.
+\noindent where $r$ is the radial coordinate, $L$ is the luminosity, and $P$, $T$, $\rho$  and $S$ are the gas pressure, temperature, density and entropy, respectively.
 The enclosed mass  at radius $r$ is $m$.
 \Eq{eq:structc} simply defines the temperature gradient  $\nabla \equiv d \ln T/d \ln P$.
-In optically thick radiative zones, radiative diffusion gives a temperature gradient
+
+Radiation diffusion gives a temperature gradient
 \begin{equation}
 \label{eq:delrad}
-\delrad \equiv \frac{3 \kappa P}{64 \pi G m \sigma T^4} L,
-\end{equation}
-where $\sigma$ is the Stefan-Boltzmann constant.
-In our models the atmosphere is optically thick throughout the whole radiative region (see \App{opacityan} for an analytic estimate of the optical depth).
+\delrad \equiv \frac{3 \kappa P}{64 \pi G m \sigma T^4} L\, ,
+\end{equation}
+with $\sigma$ the Stefan-Boltzmann constant.
+In our models, optically thick diffusion is a good approximation throughout the radiative zones.
 In convectively unstable regions, efficient convection gives an isentropic temperature gradient with $\nabla = \delad$, the adiabatic gradient
@@
 
-In the energy equation (\ref{eq:structd}), $\epsilon$ represents all local sources of heat input, which excludes the motion of the atmosphere itself.
-In stars, nuclear burning contributes to $\epsilon$.
+In the energy equation (\ref{eq:structd}), $\epsilon$ represents all local sources of heat input, except for the motion of the atmosphere itself.
+From stellar structure, $\epsilon$ may be familiar as a nuclear burning term.
 In a protoplanetary atmosphere, dissipative drag on planetesimals contributes to $\epsilon$.
-Our simplified models set $\epsilon = 0$, consistent with our neglect of planetesimal accretion luminosity at the base of the atmosphere.
-The $\epsilon_{\rm g} = -T \partial S / \partial t$ term gives the energy input from gravitational contraction.\footnote{In general, any motion is accounted for by this term.
-The partial time derivative is performed on shells of fixed mass.}  The partial time derivative would normally require our radial derivatives to be partial derivatives.
+Our models set $\epsilon = 0$, consistent with our neglect of planetesimal accretion.
+
+The energy input from gravitational contraction, $\epsilon_{\rm g} = -T \partial S / \partial t$, is crucial for a cooling model.\footnote{In general, any motion is accounted for by this term.
+The partial time derivative is performed on shells of fixed mass.}  The partial time derivative would normally require our radial derivative to be partial as well.
 However, our subsequent developments will replace the local energy equation (\ref{eq:structd}) with global energy balance (see section \S\ref{cooling}), reverting the structure equations to time-independent ordinary differential equations (ODEs).
@@
 P &=& \rho \Rg T \, ,\slabel{eq:idealgas} \\
-P &=&K \rho^{\gamma} \, , \slabel{eq:polyEOS}%\equiv K \rho^{1/(1-\delad)}, 
-\end{subeqnarray}
-where $K$ is the adiabatic constant.
-The adiabatic index is  $\gamma = 1/(1 - \delad)$.
-This work uses $\delad=2/7$ for an ideal diatomic gas (by comparison, an ideal monatomic gas has $\delad = 2/5$).
-While our reference mean molecular weight ($\mu = 2.35$ proton masses) includes Helium, we ignore Helium's effect on the EOS, which is already greatly simplified.
-The second law or thermodynamics gives the relative entropy as
-\begin{equation}
-S = \Rg \ln \left(T^{1/\delad} \over P \right) ,
-\end{equation} 
-eliminating the need for $K$.
+S &=& \Rg \ln \left(T^{1/\delad} \over P \right) \, ,\slabel{eq:polyEOS}  % P &=&K \rho^{\gamma} \, , 
+\end{subeqnarray}
+with $S$ a relative entropy.
+This work uses $\delad=2/7$, i.e.\  a polytropic index $\gamma \equiv 1/(1 - \delad) = 7/5$, for an ideal diatomic gas.\footnote{\Eq{eq:polyEOS} is equivalent to the standard polytropic relation,  $P =K \rho^{\gamma}$, if the polytropic index $K$ replaces $S$.}    We thus neglect the presence of monatomic Helium in our choice of polytropes, a common practice in idealized studies.
+We do however account for Helium in our reference value of $\mu = 2.35$ proton masses.
 
 Boundary conditions must be satisfied at both the base and the top of the atmosphere, with $m(R\co) = M\co$, $T(\RH) = T\di$ and $P(\RH) = P\di$.
-In principle, our solutions describe atmospheres with $L(R\co) = 0$.
-In practice, since we do not directly integrate \Eq{eq:structd} we need not directly impose this boundary condition, as described in \S\ref{sec:twolayer}.
+Our model atmospheres also obey $L(R\co) = 0$.
+Along with \Eq{eq:structd}, this boundary condition is incorporated in the global cooling model described below.
 
@@
 
-This section describes the global energy balance of a planet embedded in a gas disk, or more generally any spherical, hydrostatic object in pressure equilibrium with a background medium.
+We now consider the global energy balance of a planet embedded in a gas disk.
+More generally our derivation applies to any spherical, hydrostatic object in pressure equilibrium with a background medium.
 The total atmospheric energy includes gravitational and internal energies, $E = E_G + U$, with
@@
 
-%AP/AY: add virial theorem here?
-
-We start with the global energy balance for an isolated planet of mass $M$ with a free surface:
+We start with a standard result, global energy balance for an isolated, i.e.\ not embedded, planet:
 \begin{equation}
@@
 \end{equation}
-The surface luminosity, $L_M$, includes contributions from the core luminosity $L\co$ --  e.g.\ planetesimal accretion or radioactive decay -- from the total heat generation $\Gamma$ -- given by the integral of $\epsilon$ over the object -- and from the rate of change of atmospheric energy $\dot{E}$, a loss term.
+The surface luminosity, $L_M$, includes the core luminosity $L\co$ from e.g.\ planetesimal accretion or radioactive decay.
+The total heat generation $\Gamma$  is the integral of $\epsilon$ over the atmosphere.
+The rate of change of atmospheric energy, $\dot{E}$, is a loss term.
 
@@
 The surface can be at any mass level $M$, where the instantaneous radius is $R$.
-Other surface quantities are labeled by $M$ subscripts.
+Surface quantities are labeled by $M$ subscripts (except for $R$).
 The energy accreted across the surface is given by the specific energy, $e_M = u_M-G M/R$, and the mass accretion rate of gas, $\dot{M}$.
 The work done by the surface is $P_M \partial V_M/ \partial t$, with the partial derivative performed at fixed mass.
-%This generalized energy equation applies on any spherical shell where hydrostatic balance holds.
-The choice of the mass shell $M$\footnote{\Eq{eq:coolingglobal} also applies in the interior of objects with a free surface.
-The work term, $P_M \p V_M/\p t$, vanishes at the free surface.
-The accretion energy, $e_M \dot{M}$,  vanishes for a truly isolated object, but would be included to account for accretion onto (or mass loss from) an otherwise free surface.}   Thus $M$ no longer refers to a uniquely defined total mass, but to the chosen mass level, where the instantaneous radius is $R$.
-Values on this shell are labelled by $M$ subscripts.
 
@@
 
-To simplify our calculations of atmospheric contraction, we use a two-layer model with a bottom convective region and an upper radiative layer.
-The existence of such a structure is well known from previous studies (e.g., R06) and can be readily understood.
+To simplify our evolutionary calculations, we develop a two-layer atmospheric model with a convective interior and a radiative exterior.
+The existence of this layered structure is well known from previous studies (e.g., R06) and can be readily understood.
 Before the protoplanetary atmosphere can cool, it has the entropy of the disk.
@@
 We proceed by assuming that the majority of energy is lost from the convective interior, and thus the luminosity can be treated as constant in the outer radiative zone.
-With this approximation we can construct solutions from Equations (\ref{eq:struct}a -- c) that ``follow the mass," i.e.\ gradually increase the atmospheric mass.
-We then use global energy balance, \Eq{eq:coolingglobal}, to place these solutions in a cooling sequence.
-The validity of neglecting luminosity generation in the radiative zone can be checked \emph{a posteriori}.
-
-To obtain a single atmosphere profile (indexed by $i$), we choose a planet mass $M_i$.
+
+With this approximation for the radiative zone, we can construct an array of solutions to Equations (\ref{eq:struct}a -- c) in order of increasing atmospheric mass.
+We then use global energy balance, \Eq{eq:coolingglobal}, to ``follow the mass" and place these solutions in a cooling sequence.
+The validity of neglecting luminosity generation in the radiative zone can be checked \emph{a posteriori}, see \S\ref{sec:endoftime}.
+
+To obtain a single atmosphere solution (indexed by $i$), we choose a planet mass $M_i$.
 At the outer boundary, at $\RH(M_i)$, the temperature and pressure are set to the disk values.
@@
 Our approximate treatment of the radiative zone makes the RCB  the preferred location.
-With $\Gamma = L\co = 0$,  the elapsed time $\Delta t$  between states $i$ and $i +1$ is given by the finite difference
+The elapsed time $\Delta t$  between states $i$ and $i +1$ is given by the finite difference
 \begin{equation}
@@
 \Delta t = %\left(
-{ -\Delta E + \brak{e} \Delta M - \brak{P}  \Delta V_{\brak{M}} \over \brak{L} }\, .
-\end{equation} 
-Brackets indicate an average of, and ``$\Delta$" indicates a difference between, the two states.
+{ -\Delta E + \brak{e} \Delta M - \brak{P}  \Delta V_{\brak{M}} \over \brak{L} }\, ,
+\end{equation} 
+using \Eq{eq:coolinggobal} with $\Gamma = L\co = 0$.
+Brackets indicate an average of, and $\Delta$ indicates a difference between, the two states.
 All values are evaluated at the RCB.
@@
 \centering
-%AY: figure should probably use R_B for consistency with text.
 \includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/PTm_profiles_v2_smallM.pdf}
@@
 We see that self-gravity plays a very important role in atmospheric evolution, even well before the crossover mass, when $M_{\rm atm}=M\co$.
-
-%Before discussing time evolution (both the elapsed time from the bottom panel of \Fig{fig:Ltplot} and the characteristic growth time of the atmosphere), we briefly discuss the effect of decreasing the atmospheric dust abundance on the atmospheric evolution.
-%AY abo e sentence is bad, we want to discuss science and not "direct traffic"
 
@@
 %The onset of runaway growth at high masses is also evident in \Figs{fig:Ltplot}{fig:LtvsMopacity} (bottom panels) as the flattening of the $t$ vs.\ $M_{\rm atm}$ curve.
-
-%peak when $M_{\rm{atm}} < M_{\rm c}$, before the crossover mass, is helpful for understanding atmospheric growth.
-In contrast with our analytic model, where the time to runaway accretion cannot be calculated self-consistently, the numerical model provides information about when unstable collapse occurs, i.e.
-when growth times decrease.
-It follows that the time to runaway accretion, $t_{\rm run}$, can be estimated as the time at which the atmosphere growth time $M_{\rm atm}/\dot{M}$ drops to a fraction $f$ of its maximum value.
-This allows for a more rigorous estimate of the runaway accretion timescale, and thus of the critical core mass, as opposed to choosing $M_{\rm atm} \sim M\co$ as the criterion for collapse.
-In practice, however, the continuous accelerated growth in the late stages of the atmosphere evolution renders the estimated timescale to envelope collapse, and therefore the critical core mass, insensitive to the exact choice of runaway accretion time or atmosphere mass.
-Since $M_{\rm atm} < M\co$ when growth times peak, we may proceed by analogy with the analytic model and set $t=t_{\rm co}$, when $M_{\rm atm}=M\co$, as the criterion for collapse.
-We note that in some cases our cooling model breaks down before the crossover mass, as explained in section \S\ref{sec:endoftime}, hence evolution is stopped before $M_{\rm atm}=M\co$.
-However, this breakdown occurs for $M_{\rm atm} \sim 0.8-0.9 M\co$, comparable to the crossover mass, so the additional time to reach crossover is negligible.
-It follows that $t_{\rm run} \approx t_{\rm co}$.
-By analogy with the analytic results we thus choose $t_{\rm co}$ as the timescale to runaway accretion, as it will give the same numerical results as $t_{\rm run}$ and it is easier to understand in the context of other core accretion models.
 
@@
 
-%Looking ahead to our discussion of the critical core mass in \S\ref{sec:critcore}, \Fig{fig:tvsMcomp} shows that $t_{\rm run}$ equals a characteristic disk lifetime of 3 Myr for a core mass of  $M_{\rm crit} \approx 7 \, M_\oplus$  (for the standard case of $\mu = 2.35$).
-
 \begin{figure*}[tb]
@@
 However, in this high mass regime, the spherically symmetric approximation has already broken down.
-%, as does our approximate cooling model (\emph{discuss elsewhere and cite}).
 
@@
 
-\acknowledgements{We thank Ruth Murray-Clay for detailed feedback and skillful proof reading. We also thank Eugene Chiang for a valuable conversation.}
+\acknowledgements{We thank Ruth Murray-Clay for detailed feedback and skillful proof reading.
+We also thank Eugene Chiang forvaluable comments.
+ANY thanks Phil Armitage and Scott Kenyon for stimulating conversations.}
 

### 2013-10-28 tex/draft/draft4.tex
editing main assumptions

@@
 
-To model the growth of planetary atmospheres around a solid core, we develop a simplified two-layer model for time-dependent atmospheric cooling, i.e.
+To model the accretion of gas by a solid protoplanet, we develop a simplified two-layer model for time-dependent atmospheric growth via cooling, i.e.
 KH contraction.
@@
 At higher masses, our approximate treatment of the radiative zone (explained below) breaks down.
-However, evolution after the onset of runaway growth contributes minimally to the total planet formation timescale, which we can model to good accuracy.
+Since evolution after the onset of runaway growth contributes minimally to the total planet formation timescale, we can model growth times to good accuracy.
 %Our treatment is also inappropriate for hot, short-period planets, where dust sublimation gives deeper radiative zones that require more detailed models.
@@
 \begin{enumerate}
-\item The atmosphere is spherically symmetric and remains in hydrostatic balance during its thermal evolution.
+\item The atmosphere is spherically symmetric, remains in hydrostatic balance, and matches onto the disk's midplane temperature and pressure at the planet's Hill radius.
 \item The core mass and radius are fixed in evolutionary calculations, neglecting ongoing planetesimal or dust accretion.
-\item At the planet's Hill radius, the atmospheric temperature and pressure match the conditions in the disk midplane.
-\item The only source of planetary luminosity is the gravitational contraction of the atmosphere.  
+%\item At the planet's Hill radius, the atmospheric temperature and pressure match the conditions in the disk midplane.
+\item Gravitational contraction of the atmosphere is the only source of planetary luminosity.  
 \item Cooling of the atmosphere is dominated by the convective interior.  Thus the luminosity in the radiative exterior is held spatially constant. 
-\item A global cooling model connects independent static solutions into a time-dependent sequence.
+%\item A global cooling model connects independent static solutions into a time-dependent sequence.
 \item A polytropic equation of state (EOS) is assumed for simplicity.

### 2013-10-28 tex/draft/draft4.tex
minor edits + updated Fig. 1

@@
 Previous studies have considered this type of growth \citep{ikoma00, pn05}, which is an end-member case of phases 2 and 3 in more detailed quasi-static models.
-Especially since phase 2 atmospheric growth is often the slowest stage of core accretion, understanding the fastest allowed rates of gas accretion is crucial importance.
+Especially since phase 2 atmospheric growth is often the slowest stage of core accretion, understanding the fastest allowed rates of gas accretion is of crucial importance.
 Our study develops simplified cooling models for the purposes of broadly exploring parameter space -- especially in the outer disk -- and developing a more detailed understanding of atmospheric accretion.
@@
 This work uses $\delad=2/7$, i.e.\  a polytropic index $\gamma \equiv 1/(1 - \delad) = 7/5$, for an ideal diatomic gas.\footnote{\Eq{eq:polyEOS} is equivalent to the standard polytropic relation,  $P =K \rho^{\gamma}$, if the polytropic index $K$ replaces $S$.}    We thus neglect the presence of monatomic Helium in our choice of polytropes, a common practice in idealized studies.
-We do however account for Helium in our reference value of $\mu = 2.35$ proton masses.
+We do, however, account for Helium in our reference value of $\mu = 2.35$ proton masses.
 
@@
 We now consider the global energy balance of a planet embedded in a gas disk.
-More generally our derivation applies to any spherical, hydrostatic object in pressure equilibrium with a background medium.
+More generally, our derivation applies to any spherical, hydrostatic object in pressure equilibrium with a background medium.
 The total atmospheric energy includes gravitational and internal energies, $E = E_G + U$, with
@@
 For an object with no core luminosity (or no core) and no internal heat sources, the energy equation $L_M = -\dot{E}$ describes KH contraction in its simplest form.
-When internal heat sources dominate, $L_M = \Gamma$, e.g.\ for nuclear burning in an main sequence star.
+When internal heat sources dominate, $L_M = \Gamma$, e.g.\ for nuclear burning in a main sequence star.
 
@@
 \end{equation} 
-using \Eq{eq:coolinggobal} with $\Gamma = L\co = 0$.
+using \Eq{eq:coolingglobal} with $\Gamma = L\co = 0$.
 Brackets indicate an average of, and $\Delta$ indicates a difference between, the two states.
@@
 \caption{Radial profiles of atmospheric pressure, temperature and enclosed mass (core included) for a $5 M_{\oplus}$ core at $a=60$ AU.
-Solid, dotted, dashed and dot-dashed lines correspond to solutions with total mass (core and atmosphere) of $5.01 M_{\oplus}$, $5.10 M_{\oplus}$, $6.00 M_{\oplus}$ and $8.99 M_{\oplus}$, respectively.
+Solid, dotted, dashed and dot-dashed lines correspond to solutions with total mass (core and atmosphere) of $5.01 M_{\oplus}$, $5.10 M_{\oplus}$, $6.00 M_{\oplus}$ and $8.99 M_{\oplus}$, respectively (see text for an explanation of mass choices).
 Circles and triangles mark the locations of the Bondi radii and of the radiative-convective boundaries, respectively.

### 2013-10-28 tex/draft/draft4.tex
missing space in acknowledgements
minor, mostly section 3 and grant acknowledgement

@@
 This section develops the analytic version of our two-layer model.
-This analytic model is less accurate than our numerical model, primarily because it neglects self-gravity, which we show (in \S\ref{sec:KH}) to be important at rather low atmospheric masses, $M_{\rm atm} \gtrsim 0.1 M\co$.
+This analytic model is less accurate than our numerical model, primarily because it neglects self-gravity.
+We we show (in \S\ref{sec:KH}) that self-gravity becomes important at rather low atmospheric masses, $M_{\rm atm} \gtrsim 0.1 M\co$.
 Nevertheless, the analytic model is useful for understanding atmospheric evolution and interpreting the numerical results.
@@
 \end{subeqnarray} 
-The initial RCB depth, set to $P\di$ as a formality, is of little importance because the cooling slows as it proceeds with $t_{\rm cool} \propto P\cb^2$.
+The initial RCB depth is set to $P\di$ as a formality.  The cooling slows as it proceeds with $t_{\rm cool} \propto P\cb^2$.
 
@@
 The main reason for this discrepancy is the neglect of self-gravity, as shown in Section \ref{sec:KH}.
-The magnitude of the discrepancy may be surprising, as one might expect self-gravity to be a modest factor for $M_{\rm atm} \leq M\co$.
-Clearly this seemingly reasonable expectation can be misleading, an interesting result in itself.
+One might expect self-gravity to be only a modest correction for $M_{\rm atm} \leq M\co$.
+This seemingly reasonable expectation can be misleading, an interesting result in itself.
 
@@
 The cooling timescale and critical core mass depend only weakly on the disk pressure, via the logarithmic factor $\xi$.
+%AY: add density explanation?
 
@@
 \caption{Radial profiles of atmospheric pressure, temperature and enclosed mass (core included) for a $5 M_{\oplus}$ core at $a=60$ AU.
-Solid, dotted, dashed and dot-dashed lines correspond to solutions with total mass (core and atmosphere) of $5.01 M_{\oplus}$, $5.10 M_{\oplus}$, $6.00 M_{\oplus}$ and $8.99 M_{\oplus}$, respectively (see text for an explanation of mass choices).
+Solid, dotted, dashed and dot-dashed lines correspond to solutions with total mass (core and atmosphere) of $5.01 M_{\oplus}$, $5.10 M_{\oplus}$, $6.00 M_{\oplus}$ and $8.99 M_{\oplus}$, respectively (see text for significance of these masses).
 Circles and triangles mark the locations of the Bondi radii and of the radiative-convective boundaries, respectively.
@@
 Temperature variations are much more significant than pressure variations (note the difference in logarithmic and linear axes).
-Since midplane disk conditions depend only on temperature and pressure in our model\footnote{See \Eq{eq:diskparam}.
-When gap opening is considered in models of later growth stages, the orbital frequency and effective viscosity become relevant as well.}, the dominant effect of disk location is temperature.
-
-The decline in growth times at lower temperatures arises from a variety of competing effects.
-Briefly, the larger Bondi radius and lower dust opacity at lower temperatures accelerate growth, overpowering the inherently lower cooling luminosity at lower temperatures.
+Since midplane disk conditions depend only on temperature and pressure in our model,\footnote{See \Eq{eq:diskparam}.
+When gap opening is considered in models of later growth stages, the orbital frequency and effective viscosity become relevant as well.} the dominant effect of disk location is temperature.
+
+The decline in growth times at lower temperatures arises from a balance of competing effects.
+The larger Bondi radius and lower dust opacity at lower temperatures accelerate growth, overpowering the inherently lower cooling luminosity at lower temperatures.
 Growth times depend only weakly on, but do fall slightly with, pressure.
 The nearly exponential increase in pressure and density with depth through the radiative zone means that cooling, which is regulated at the RCB, has only a logarithmic dependence on disk pressure.
-Section \ref{sec:coolingan} explains these effects in more detail.
+Section \ref{sec:coolingan} shows how these effects arise in our analytic model.
 
@@
 \acknowledgements{We thank Ruth Murray-Clay for detailed feedback and skillful proof reading.
-We also thank Eugene Chiang forvaluable comments.
-ANY thanks Phil Armitage and Scott Kenyon for stimulating conversations.}
+We also thank Eugene Chiang for valuable comments.
+ANY thanks Phil Armitage and Scott Kenyon for stimulating conversations.
+ANY acknowledges partial grant support from NSF AST 1313021 and NASA NNX13AI58G.}
 

### 2013-10-28 tex/draft/draft4.tex
minor, mostly T,P explanation

@@
 The main concern for core accretion models, which are successful in the inner disk, is that they operate too slowly in the outer disk.
-The few Myr lifetime of gas disks sets the constraint \citep{hillenbrand05}.
+The few Myr lifetime of gas disks sets the constraint \citep{williams11}.
 However, mechanisms for the rapid accretion of solids, even at large separations, exist \citep{dones93, kenyon09, lambrechts12}.
@@
 Aside from this insight, the analytic model is useful because, despite the amplitude error, it explains the basic scaling of numerical cooling models.
-As is well known \citep{HubBod05}, a lower opacity, here scaling with $F_\kappa$, accelerates growth and gives a lower critical core mass.
+As is well known \citep{HubBod05}, a lower opacity, here scaling with $F_\kappa$, allows faster cooling and gives a lower critical core mass.
 The cooling timescale and critical core mass depend only weakly on the disk pressure, via the logarithmic factor $\xi$.
+The almost exponential rise in radiative zone pressure with depth, see \Eq{eq:PcbRcb}, explains why disk pressure has only a weak effect on cooling at the RCB.
 %AY: add density explanation?
@@
 
-We present examples of the structure and evolution of our model atmospheres, calculated as described in \S\ref{sec:model}, with some comparisons to the non-self-gravitating analytic model of \S\ref{sec:coolingan}.
+We examine the structure and evolution of our model atmospheres, calculated as described in \S\ref{sec:model}.
+For comparison, we also show results of the non-self-gravitating analytic model of \S\ref{sec:coolingan}.
 Radial structure is presented in \S\ref{sec:profiles}, time evolution is described in \S\ref{sec:timeev}, and the validity of our two-layer cooling model is examined in \S\ref{sec:endoftime}.
@@
 In agreement with \Eq{eq:cb2}, the radiative zones remain nearly isothermal, even for the higher masses.
-Consequently, the pressure increases nearly exponentially with inverse depth.
+Consequently, the pressure increases nearly exponentially with depth.
 
@@
 
-The decline in growth times at lower temperatures arises from a balance of competing effects.
-The larger Bondi radius and lower dust opacity at lower temperatures accelerate growth, overpowering the inherently lower cooling luminosity at lower temperatures.
+The decline in growth times with lower temperatures arises from a balance of competing effects.
+The cooling luminosity is inherently smaller at lower temperatures.
+Overpowering this effect, the larger Bondi radius and lower dust opacity act to accelerate growth at lower temperatures.
+
 Growth times depend only weakly on, but do fall slightly with, pressure.
-The nearly exponential increase in pressure and density with depth through the radiative zone means that cooling, which is regulated at the RCB, has only a logarithmic dependence on disk pressure.
-Section \ref{sec:coolingan} shows how these effects arise in our analytic model.
+This result may be surprising, given that the disk is the source of atmospheric mass and the atmosphere must match onto the disk's density and pressure.
+The nearly exponential increase in pressure with depth through the radiative zone explains this effect.
+Cooling is largely regulated at the RCB, and a modest change in RCB depth compensates for large variations in disk pressure.
+
+Section \ref{sec:coolingan} shows how these temperature and pressure effects arise in our analytic model.
 

### 2013-10-28 tex/draft/draft4.tex
very minor edit
minor, mostly sec 4 and fig captions

@@
 %\vspace{-0.5in}
-\caption{Radial profiles of atmospheric pressure, temperature and enclosed mass (core included) for a $5 M_{\oplus}$ core at $a=60$ AU.
+\caption{Radial profiles of atmospheric pressure, temperature and enclosed mass (core included) for a $5 M_{\oplus}$ core at $60$ AU.
 Solid, dotted, dashed and dot-dashed lines correspond to solutions with total mass (core and atmosphere) of $5.01 M_{\oplus}$, $5.10 M_{\oplus}$, $6.00 M_{\oplus}$ and $8.99 M_{\oplus}$, respectively (see text for significance of these masses).
@@
 Quoted mass values include the core plus atmosphere within the smaller of $\RB$ or $\RH$, which for these cases is $\RB$.
-The 8.99 $M_{\oplus}$ solution is the mass at which runaway growth begins according to the criteria described in \S\ref{sec:timeev}.
-% we can reach in our evolutionary sequence, as we explain further in \S\ref{sec:endoftime}.
-
-The lowest mass atmosphere -- which we take as our initial state -- is fully convective with the same entropy as the disk.
+The 8.99 $M_{\oplus}$ solution is the mass
+that satisfies our runaway growth criteria (described in \S\ref{sec:timeev}). % we can reach in our evolutionary sequence, as we explain further in \S\ref{sec:endoftime}.
+
+The lowest mass atmosphere -- which we take as our initial state -- is fully convective and shares the disk's entropy.
 In \Fig{fig:profiles} this state is the 5.01 $M_{\oplus}$ solution with no radiative zone.
-Atmospheric cooling and contraction allows the accretion of more gas.
+
+Cooling and contraction allow the atmosphere to accrete more gas.
 In the convective zones, higher mass solutions have lower entropy and higher pressures.
@@
 The luminosity is initially high, then decreases as the atmosphere grows in mass and the radiative zone becomes optically thicker.
-Due to the neglect of self-gravity, the analytic model (\emph{dashed curve}) gives a further drop in luminosity and a longer evolution time.}
+Due to the neglect of self-gravity, the analytic model (\emph{dashed curve}) gives luminosities that are too low and evolution times that are too long.}
 \label{fig:Ltplot}
@@
 The cooling model of \S\ref{sec:twolayer} is used to connect solutions with different atmospheric masses into an evolutionary sequence.
-\Fig{fig:Ltplot} shows the luminosity evolution and the elapsed time as a function of atmospheric mass for a $5 M_{\oplus}$ core at $60$ AU, i.e.\ the same parameters as in \Fig{fig:profiles}.
+\Fig{fig:Ltplot} shows the luminosity evolution and the elapsed time as a function of atmospheric mass for the same parameters as in \Fig{fig:profiles}.
+
 During the early stages of atmospheric growth, the luminosity drops sharply.
-This behavior is seen in both the full numerical solutions (with solid lines) and the analytic model (with dashed lines).
+This behavior is seen in both the full numerical solutions and the analytic model.
 With increasing atmospheric mass, the pressure depth of the RCB increases, along with the optical depth ($\propto \kappa P\cb)$.
@@
 
-The accelerated growth in the numerical model, as shown in the bottom panel of \Fig{fig:Ltplot}, is also a direct result of the enhanced cooling (higher luminosities) in the self-gravitating models.
-We see that self-gravity plays a very important role in atmospheric evolution, even well before the crossover mass, when $M_{\rm atm}=M\co$.
-
-In \Fig{fig:LtvsMopacity}, the opacity normalization in \Eq{eq:opacitylaw} is reduced by factors of 10 and 100.
+The accelerated growth in the numerical model, as shown in the bottom panel of \Fig{fig:Ltplot}, is also a direct result of the higher cooling luminosities in the self-gravitating models.
+We see that self-gravity strongly affects atmospheric evolution, even well before the crossover mass, when $M_{\rm atm}=M\co$.
+
+In \Fig{fig:LtvsMopacity}, our standard dust opacity, \Eq{eq:opacitylaw}, is reduced by factors of 10 and 100.
 Lower opacities result in higher luminosities and faster evolution.
-Our model thus confirms this well-established result \citep{HubBod05}.
+Our model thus confirms a well-established result \citep{HubBod05}.
 While clearly an important effect, atmospheric dust opacities are difficult to robustly predict.
@@
 %\vspace{-0.5in}
-\caption{Similar to \Fig{fig:Ltplot}, but for a $5 M_{\oplus}$ core forming at 10 AU; we show the effect of reducing the dust abundance by factors of 10 and 100 from standard Solar abundances.
-The reduced dust opacities give higher luminosities and faster atmospheric growth.}  %The actual magnitude of dust opacities is difficult to robustly predict.}
+\caption{The effect of dust abundance on atmospheric evolution.
+Reducing dust opacities by factors of 10 and 100 from standard Solar abundances gives higher luminosities and faster atmospheric growth.
+Plotted quantities are similar to \Fig{fig:Ltplot}, but for a $5 M_{\oplus}$ core at 10 AU.
+}  %The actual magnitude of dust opacities is difficult to robustly predict.}
 \label{fig:LtvsMopacity}
@@
 Unlike our analytic model, which must assume that runaway growth begins near the crossover mass, our numerical model allows us to measure when runaway accretion starts.
-Runaway growth clearly does not begin at a universal value of $M_{\rm atm}/M\co$.
-\Fig{fig:growthtime} shows that further from the star runaway growth begins at smaller $M_{\rm atm}/M\co$.
+Runaway growth does not begin at a universal value of $M_{\rm atm}/M\co$.
+Further from the star, runaway growth begins at smaller $M_{\rm atm}/M\co$, as \Fig{fig:growthtime} shows.
 \Fig{fig:tvsM} (described in the next section) shows how the onset of runaway growth depends on core mass.

### 2013-10-28 tex/draft/draft4.tex
Scott's comments

@@
 
+The first assumption of spherical accretion breaks down early in the inner disk, as explained below.  However our focus is the outer disk, where this standard approximation is better justified.
+
 \subsection{Disk and Opacity Model}\label{sec:disk}
@@
 \end{subeqnarray}
-The normalization factors $F_{\Sigma}$ and $F_T$  adjust the model relative to the fiducial MMSN.  We fix $F_{\varSigma}=F_T=1$ unless noted otherwise.
+The normalization factors $F_{\Sigma}$ and $F_T$  adjust the model relative to the fiducial MMSN.
+We fix $F_{\varSigma}=F_T=1$ unless noted otherwise.
+Some observations support a flatter surface density profile, $\varSigma \propto a^{-1}$ \citep{andrews10}.
+We find that gas density and pressure are weak corrections to atmospheric growth.
+However, a greater surface density of solids in the outer disk could favor the rapid growth of cores \citep{bromley11}.
 
@@
 where the core mass, $M\co$, is normalized to 10 Earth masses as $\mcn{10} \equiv M\co/(10~M_\oplus)$.
-The core density is held fixed at $\rho\co=3.2$ g cm$^{-3}$.
+The core density is held fixed at $\rho\co=3.2$ g cm$^{-3}$, representing a mixture of ice and rocky material \citep{pap99}.
 We thus neglect  the detailed equation of state of the solid core \citep{fortney07}.
@@
 This analytic model is less accurate than our numerical model, primarily because it neglects self-gravity.
-We we show (in \S\ref{sec:KH}) that self-gravity becomes important at rather low atmospheric masses, $M_{\rm atm} \gtrsim 0.1 M\co$.
+We show (in \S\ref{sec:KH}) that self-gravity becomes important at rather low atmospheric masses, $M_{\rm atm} \gtrsim 0.1 M\co$.
 Nevertheless, the analytic model is useful for understanding atmospheric evolution and interpreting the numerical results.
@@
 Both $t_{\rm co}$ and $\MC$ are too large to be interesting, or to be correct based on previous results and the numerical results in this paper.
-The main reason for this discrepancy is the neglect of self-gravity, as shown in Section \ref{sec:KH}.
+The main reason for this discrepancy is the neglect of self-gravity.
+Section \ref{sec:KH} explores in detail the effects of self-gravity on atmospheric structure, luminosity and evolution.
 One might expect self-gravity to be only a modest correction for $M_{\rm atm} \leq M\co$.
@@
 To rescale, we simply assume that runaway occurs at a modified crossover mass, $M_{\rm atm} = f M\co$.
-From \Eqs{eq:crit}{eq:Pcbc}, this change amounts to the replacing $\xi \rightarrow f\xi$.
+From \Eqs{eq:crit}{eq:Pcbc}, this change amounts to replacing $\xi \rightarrow f\xi$.
 We require $f < 1$ to give the faster growth seen in numerical models.
@@
 %\vspace{-0.5in}
-\caption{Evolution of the atmospheric growth timescale with mass around a $5 M_{\oplus}$ solid core  located at 10, 30 or 60 AU.
+\caption{Evolution of the atmospheric growth timescale with mass around a $5 M_{\oplus}$ solid core  located at 10, 30 or 60 AU for standard Solar opacities.
 Growth is slowest for $M_{\mathrm{atm}} \sim 1 - 3 M_{\oplus}$, i.e.\ before the crossover mass at $M_{\rm atm} = M\co$.}
@@
 \acknowledgements{We thank Ruth Murray-Clay for detailed feedback and skillful proof reading.
-We also thank Eugene Chiang for valuable comments.
-ANY thanks Phil Armitage and Scott Kenyon for stimulating conversations.
-ANY acknowledges partial grant support from NSF AST 1313021 and NASA NNX13AI58G.}
+We also thank Eugene Chiang and Scott Kenyon for valuable comments.
+ANY thanks Phil Armitage for stimulating conversations.
+
+ANY acknowledges support from the {\it NASA } {\it ATP} and {\it OSS} grant NNX10AF35G and the  {\it NASA} {\it OPR} grant NNX11AM37G while at the CfA, and from the {\it NSF AST}  grant 1313021 and the {\it NASA OSS}  grant NNX13AI58G while at JILA.}
 

### 2013-10-29 tex/draft/draft4.tex
minor edits

@@
 
-The first assumption of spherical accretion breaks down early in the inner disk, as explained below.  However our focus is the outer disk, where this standard approximation is better justified.
+The first assumption of spherical accretion breaks down early in the inner disk, as explained below.  However, our focus is the outer disk, where this standard approximation is better justified.
 
@@
 We fix $F_{\varSigma}=F_T=1$ unless noted otherwise.
-Some observations support a flatter surface density profile, $\varSigma \propto a^{-1}$ \citep{andrews10}.
-We find that gas density and pressure are weak corrections to atmospheric growth.
+Some observations support a flatter surface density profile, $\varSigma\di \propto a^{-1}$ \citep{andrews10}.
+We find that gas density and pressure are weak corrections to atmospheric growth (see Section \S\ref{sec:critical}).
 However, a greater surface density of solids in the outer disk could favor the rapid growth of cores \citep{bromley11}.
@@
 The core density is held fixed at $\rho\co=3.2$ g cm$^{-3}$, representing a mixture of ice and rocky material \citep{pap99}.
-We thus neglect  the detailed equation of state of the solid core \citep{fortney07}.
+We thus neglect  the detailed EOS of the solid core \citep{fortney07}.
 
@@
 %\vspace{-0.5in}
-\caption{Evolution of the atmospheric growth timescale with mass around a $5 M_{\oplus}$ solid core  located at 10, 30 or 60 AU for standard Solar opacities.
+\caption{Evolution of the atmospheric growth timescale with mass around a $5 M_{\oplus}$ solid core  located at 10, 30 or 60 AU, for standard Solar opacities.
 Growth is slowest for $M_{\mathrm{atm}} \sim 1 - 3 M_{\oplus}$, i.e.\ before the crossover mass at $M_{\rm atm} = M\co$.}

### 2013-10-29 tex/draft/draft4.tex
added Rafikov discussion

@@
 (in prep) uses the \citet{saumon95} EOS, with extensions to lower pressures and temperatures as needed, to explore these effects.
+
+\subsection{Planetesimal Accretion}
+\label{sec:placc}
+
+Our regime assumes that the protoplanetary core is fully grown and planetesimal accretion has ceased.
+However, it is possible, in theory, for the atmosphere to reach crossover and initiate runaway gas accretion during core growth, i.e.
+before planetesimal accretion becomes low enough for KH contraction to start dominating.
+This concern is explored in detail in Piso et al.
+(in prep.), where we compare our results to the static model of R06.
+We find that our model yields lower critical core masses than the R06 corresponding study for large planetesimal accretion rates.
+It follows that the accretion of solids can be safely ignored in our regime, and that giant planet formation is faster with variable planetesimal accretion rates.
+
+%an additional concern for our regime is whether the atmosphere becomes critical before KH contraction starts dominating, i.e.
+the atmosphere may reach crossover before planetesimal accretion slows down to the maximum rate imposed by our model, which satisfies L_KH = L_acc
 

### 2013-10-29 tex/draft/draft4.tex
R06 comparison edits

--- 2013-10-29 tex/draft/draft4.tex
@@
 For a given planetesimal accretion rate, static solutions only exist up to a maximum core mass, referred to as the ``critical core mass",  $\MC$.
-Above this mass, the atmosphere is assumed to collapse and rapidly accrete disk gas, a process known as the ``core accretion instability."  Many studies reproduce a canonical $\MC \sim 10 M_\oplus$.
-However, \citet[hereafter R06]{rafikov06} finds a wide range of values, $0.1 M_\oplus \lesssim \MC \lesssim 100 M_\oplus$, depending on the planetesimal accretion rate and disk parameters \citep[see also][]{Raf11}.
-A limitation of static models is the neglect of heat generated by atmospheric collapse, which can transform the core accretion instability into a slower process of Kelvin-Helmholtz (KH) contraction.
+Above this mass, the atmosphere is assumed to collapse and rapidly accrete disk gas, a process known as the ``core accretion instability."
+
+Many static studies reproduce a canonical $\MC \sim 10 M_\oplus$.
+However, \citet[hereafter R06]{Raf11, rafikov06} finds a wide range of values, $0.1 M_\oplus \lesssim \MC \lesssim 100 M_\oplus$, depending on the planetesimal accretion rate and disk parameters.
+% \citep[see also][]{Raf11}.
+Section \ref{sec:placc} compares our results to R06.
+A general limitation of static models is the neglect of heat generated by atmospheric collapse, which can transform the core accretion instability into a slower process of Kelvin-Helmholtz (KH) contraction.
 
@@
 
+\subsection{Core growth}
+\label{sec:placc}
+
+Our model purposefully neglects the accretion of solids to study the fastest rates of gas accretion.
+Our $\MC$ values thus differ from the $\MC$ values in static models, such as R06.
+For similar parameters, we generally obtain lower $\MC$ values than R06 because our atmospheres are not heated by ongoing planetesimal accretion.
+Piso et al.\ (in prep) presents a quantitative comparison.
+
+At low planetesimal accretion rates, $\dot{M}\co$, a static model could formally give lower $\MC$ values than our evolutionary calculations.
+R06 shows that $\MC \propto \dot{M}\co^{3/5}$ in static models.
+The underlying assumption of static models, a negligibly short KH contraction time, fails whenever static models give a lower $\MC$.
+Thus our results give a firm lower limit on $\MC$ which complement the results of static models with planetesimal accretion.
+
 \section{Summary} \label{sec:conclusions}
@@
 By neglecting the accretion luminosity of planetesimals and smaller solids, we obtain the fastest allowed rate of gas accretion.
+
 We address core accretion in the outer disk, as it is relevant to direct imaging surveys.
@@
 The accretion and scattering of local planetesimals, combined with the inward aerodynamic drift of smaller solids past the core, offers a plausible means to sufficiently isolate the core.
+
 On the other hand, non-negligible accretion of solids onto cores could continue for the lifetime of the gas disk.
@@
 \acknowledgements{We thank Ruth Murray-Clay for detailed feedback and skillful proof reading.
-We also thank Eugene Chiang and Scott Kenyon for valuable comments.
+We appreciate valuable comments from Eugene Chiang and Scott Kenyon.
 ANY thanks Phil Armitage for stimulating conversations.

### 2013-10-30 tex/PaperI/draft4.tex
shortened abstract

--- 2013-10-29 tex/draft/draft4.tex
@@
 \author{Ana-Maria A. Piso}
-\affil{Harvard-Smithsonian Center for Astrophysics}
+\affil{Harvard-Smithsonian Center for Astrophysics, 60 Garden Street, Cambridge, MA 02138}
 
 \author{Andrew N. Youdin}
-\affil{JILA, University of Colorado at Boulder}
+\affil{JILA, University of Colorado and NIST, 440 UCB, Boulder, CO 80309-0440}
 
@@
 
-The core accretion hypothesis proposes that giant planets form by the accretion of gas onto a solid protoplanetary core.
+The core accretion hypothesis proposes that giant planets form by gas accretion onto a solid protoplanetary core.
+The minimum (or critical) core mass to form a gas giant is typically quoted as $10 M_{\oplus}$.
+This value, however, depends on several factors, including the location in the protoplanetary disk, atmospheric opacity, and the accretion rate of solids.
+Motivated by ongoing direct imaging searches for giant planets, this study focuses on core mass requirements in the outer disk.
+Our model investigates atmospheric growth around solid cores of fixed mass.
+Without planetesimal accretion heating the envelope, our limiting case gives the fastest possible atmospheric growth by Kelvin-Helmholtz contraction.
+We develop a two-layer, spherical atmosphere model with an inner convective region and an outer radiative zone that matches onto the disk.
+For different disk locations, we determine the minimum core mass for a giant planet to form within a typical disk lifetime of 3 Myr.
+The minimum core mass is $\sim$$8.5 M_{\oplus}$ at 5 AU and decreases to $\sim$$3.5 M_{\oplus}$ at 100 AU, for a Solar composition atmosphere with a standard dust opacity.
+This trend is mainly explained by the cooler temperatures in the outer disk, while disk pressures have a negligible effect.
+At all distances, decreasing the dust opacity and/or increasing the mean molecular weight of the gas reduces the critical core mass.
+We also develop a non-self-gravitating, analytic cooling model, which reveals that self-gravity significantly affects atmospheric evolution even when the atmosphere is only $\sim$$10\%$ as massive as the core.
+
+%The core accretion hypothesis proposes that giant planets form by the accretion of gas onto a solid protoplanetary core.
 The minimum (or critical) core mass to form a gas giant is typically quoted as $10 M_{\oplus}$.

### 2013-10-30 tex/PaperI/draft4.tex
more abstract pruning

@@
-% !TEX TS-program = pdflatexmk
-
-%\documentstyle[aas2pp4,epsf]{article}
-%%%\documentstyle[aaspp4,epsf]{article}
-%\documentstyle[12pt,aasms]{article}    % this is for a preprint
-%(single-spaced)
-%\documentstyle[aaspp4,epsf]{article} % this is for small print
-%\documentstyle[12pt, aaspp4]{article}
-
-%\documentstyle[11pt,aaspp]{article}
-%\documentclass[12pt, preprint]{aastex}
-
-%\documentclass[manuscript]{aastex}
 \documentclass[apj, numberedappendix]{emulateapj}
 
-%\documentclass[12pt, preprint,numberedappendix]{emulateapj}
-%\documentstyle[12pt,aasms]{article}    % this is for submittal
-                                       % (double-spaced)
-
-%\documentstyle[12pt,aasms]{article}   \usepackage{emulateapj5}
-
 \usepackage{graphicx} 
-%\usepackage{graphics}                       
 \usepackage{amsmath}
@@
 \newcommand{\km}{\; {\rm km}}
-%\newcommand{\au}{\; \varpi_{\rm AU}}
 
@@
 \newcommand{\aun}[1]{ a_{#1} }
-%\newcommand{\aun}[1]{ a_{#1\oplus} }
 \def\crit{_{\rm{c, crit}}}
@@
 
-The core accretion hypothesis proposes that giant planets form by gas accretion onto a solid protoplanetary core.
-The minimum (or critical) core mass to form a gas giant is typically quoted as $10 M_{\oplus}$.
-This value, however, depends on several factors, including the location in the protoplanetary disk, atmospheric opacity, and the accretion rate of solids.
-Motivated by ongoing direct imaging searches for giant planets, this study focuses on core mass requirements in the outer disk.
-Our model investigates atmospheric growth around solid cores of fixed mass.
-Without planetesimal accretion heating the envelope, our limiting case gives the fastest possible atmospheric growth by Kelvin-Helmholtz contraction.
-We develop a two-layer, spherical atmosphere model with an inner convective region and an outer radiative zone that matches onto the disk.
-For different disk locations, we determine the minimum core mass for a giant planet to form within a typical disk lifetime of 3 Myr.
-The minimum core mass is $\sim$$8.5 M_{\oplus}$ at 5 AU and decreases to $\sim$$3.5 M_{\oplus}$ at 100 AU, for a Solar composition atmosphere with a standard dust opacity.
-This trend is mainly explained by the cooler temperatures in the outer disk, while disk pressures have a negligible effect.
-At all distances, decreasing the dust opacity and/or increasing the mean molecular weight of the gas reduces the critical core mass.
-We also develop a non-self-gravitating, analytic cooling model, which reveals that self-gravity significantly affects atmospheric evolution even when the atmosphere is only $\sim$$10\%$ as massive as the core.
+In the core accretion hypothesis, giant planets form by gas accretion onto solid protoplanetary cores.
+While typically quoted as  $10 M_\oplus$, the minimum (or critical) core mass to form a gas giant varies, depending on several factors: location in the protoplanetary disk, atmospheric opacity, and the accretion rate of solids.
+Motivated by ongoing direct imaging searches for giant planets, this study investigates core mass requirements in the outer disk.
+To determine the fastest allowed rates of gas accretion, we consider solid cores that no longer accrete planetesimals, as this would heat the gas envelope.
+Our spherical, two-layer atmospheric cooling model includes an inner convective region and an outer radiative zone, which matches onto the disk.
+We determine the minimum core mass for a giant planet to form within a typical disk lifetime of 3 Myr.
+The minimum core mass declines with disk radius, from $\sim$$8.5 M_{\oplus}$ at 5 AU to $\sim$$3.5 M_{\oplus}$ at 100 AU, with standard dust opacities.
+Cooler temperatures in the outer disk explain this trend, while variations in disk density are less influential.
+At all distances, a lower dust opacity or higher mean molecular weight reduces the critical core mass.
+Our non-self-gravitating, analytic cooling model reveals that self-gravity significantly affects early atmospheric evolution, starting when the atmosphere is only $\sim$$10\%$ as massive as the core.
 

### 2013-10-30 tex/PaperI/ms.tex
edited abstract

--- 2013-10-30 tex/PaperI/draft4.tex
@@
 In the core accretion hypothesis, giant planets form by gas accretion onto solid protoplanetary cores.
-While typically quoted as  $10 M_\oplus$, the minimum (or critical) core mass to form a gas giant varies, depending on several factors: location in the protoplanetary disk, atmospheric opacity, and the accretion rate of solids.
+The minimum (or critical) core mass to form a gas giant is typically quoted as $10 M_{\oplus}$.
+This value, however, depends on several factors, such as the location in the protoplanetary disk, atmospheric opacity, and the accretion rate of solids.
 Motivated by ongoing direct imaging searches for giant planets, this study investigates core mass requirements in the outer disk.
-To determine the fastest allowed rates of gas accretion, we consider solid cores that no longer accrete planetesimals, as this would heat the gas envelope.
-Our spherical, two-layer atmospheric cooling model includes an inner convective region and an outer radiative zone, which matches onto the disk.
+To determine the fastest allowed rates of gas accretion, we consider solid cores that no longer accrete planetesimals, as this would heat the gaseous envelope.
+Our spherical, two-layer atmospheric cooling model includes an inner convective region and an outer radiative zone that matches onto the disk.
 We determine the minimum core mass for a giant planet to form within a typical disk lifetime of 3 Myr.
 The minimum core mass declines with disk radius, from $\sim$$8.5 M_{\oplus}$ at 5 AU to $\sim$$3.5 M_{\oplus}$ at 100 AU, with standard dust opacities.
-Cooler temperatures in the outer disk explain this trend, while variations in disk density are less influential.
+Lower temperatures in the outer disk explain this trend, while variations in disk density are less influential.
 At all distances, a lower dust opacity or higher mean molecular weight reduces the critical core mass.

### 2013-10-30 tex/PaperI/ms.tex
very minor abs tweak
arXiv posting version, updated refs + style files

@@
 \usepackage{epstopdf}
+%\usepackage{natbib}
 %\bibliographystyle{apj}
@@
 \begin{document}
-\bibliographystyle{apj}
 
@@
 The minimum (or critical) core mass to form a gas giant is typically quoted as $10 M_{\oplus}$.
-This value, however, depends on several factors, such as the location in the protoplanetary disk, atmospheric opacity, and the accretion rate of solids.
+The actual value depends on several factors: the location in the protoplanetary disk, atmospheric opacity, and the accretion rate of solids.
 Motivated by ongoing direct imaging searches for giant planets, this study investigates core mass requirements in the outer disk.
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/PTm_profiles_v2_smallM.pdf}
+\includegraphics[width=0.5\textwidth]{PTm_profiles_v2_smallM.pdf}
 %\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Lt_profiles_v2.pdf}
+\includegraphics[width=0.5\textwidth]{Lt_profiles_v2.pdf}
 %\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/opacity_effect.pdf}
+\includegraphics[width=0.5\textwidth]{opacity_effect.pdf}
 %\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mt_profile_temp.pdf}
+\includegraphics[width=0.5\textwidth]{Mt_profile_temp.pdf}
 %\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cooling_a10_Mc5_rcb.pdf}
+\includegraphics[width=0.5\textwidth]{cooling_a10_Mc5_rcb.pdf}
 %\vspace{-0.5in}
@@
 \hspace{-.1in}
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cumul_coolingtime_vs_Matm_10au_mu235.pdf}
+\includegraphics[width=0.5\textwidth]{cumul_coolingtime_vs_Matm_10au_mu235.pdf}
 %\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.48\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/coolingtime_vs_Mc_10au.pdf}
+\includegraphics[width=0.48\textwidth]{coolingtime_vs_Mc_10au.pdf}
 %\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=1.\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/TdPd_effect.pdf}
+\includegraphics[width=1.\textwidth]{TdPd_effect.pdf}
 \vspace{-0.3in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_new.pdf}
+\includegraphics[width=0.5\textwidth]{Mcrit_vs_a_3Myrs_new.pdf}
 %\vspace{-0.5in}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_opacity.pdf}
+\includegraphics[width=0.5\textwidth]{Mcrit_vs_a_3Myrs_opacity.pdf}
 %\vspace{-0.5in}

### 2013-11-13 tex/draft_paper2/draft_paper2.tex
paper II tex file + figures

@@
 
-The core accretion model postulates that giant planets form through accretion of gas onto a solid core.
-When a core embedded in a protoplanetary disk accretes an atmosphere with mass comparable to itself, the envelope is no longer in hydrostatic balance and unstable atmosphere collapse occurs.
-Formation of a core requires rapid planetesimal accretion, which deposits energy into the atmosphere.
-Heating increases the core mass required for collapse.
-However, the planetesimal accretion rate need not be constant --- in particular, it may decline over time.
-In this study we consider the limiting regime in which planetesimal accretion is negligible and the luminosity evolution of a core's atmosphere is dominated by Kelvin-Helmholtz contraction.
-We enhance the model of Piso \& Youdin (submitted), which derives the evolution of atmospheres embedded in a gas disk, to employ a more realistic equation of state and dust opacity.
-We find that the minimum core mass required to form a giant planet before the dissipation of the protoplanetary disk increases by more than a factor of two when non-ideal effects such as hydrogen dissociation and quantum rotational effects are included in the equation of state, when compared to an ideal gas polytrope.
-However, this minimum core mass decreases back to $M_{\rm{c, crit}}\sim5 M_{\oplus}$ once we assume realistic dust opacities that take into account grain growth.
+The core accretion model postulates that giant planets form through gas accretion onto a solid core.
+To accumulate a massive atmosphere, this core's mass must exceed a minimum value $M_{\rm crit}$ often quoted as $10 M_{\oplus}$.
+The value of $M_{\rm crit}$, however, depends on the rate at which planetesimals accrete onto the core and heat its atmosphere.
+In this study we calculate the minimum core mass for atmospheric collapse in the limiting regime in which core growth is halted and the luminosity evolution of a core's atmosphere is dominated by Kelvin-Helmholtz contraction.
+We build on the idealized model of Piso \& Youdin (submitted), which derives the evolution of atmospheres accreting around protoplanetary cores and calculates the minimum (or critical) core mass to form a gas giant before disk dissipation.
+We enhance this model by employing a more realistic equation of state and dust opacity.
+We find that $M_{\rm crit}$ increases by more than a factor of two when non-ideal effects such as hydrogen dissociation and quantum rotation are included in the equation of state, when compared to an ideal gas polytrope.
+Lower opacities due to grain growth reduce $M_{\rm crit}$.
+Taken together, we calculate $M_{\rm crit} \sim 7 M_{\oplus}$ at 5 AU, decreasing to $\sim$$4.5 M_{\oplus}$ at 100 AU, for a realistic EOS and a grain size distribution $n \propto a^{-3.5}$ with a maximum particle size of 1 cm.
+If $n \propto a^{-2.5}$, appropriate for coagulation, $M_{\rm crit}$ is up to one order of magnitude smaller.
+We show that giant planets may grow from smaller cores if core growth is halted, which is particularly relevant for understanding the origin of gas giants directly imaged at wide separations.
+
+%Thus halting a core's growth may facilitate giant planet formation in the outer disk, which helps understand the origin of giant planets directly imaged at wide separations.
+
+%It is thus easier to form gas giants from fully grown cores, which may facilitate giant planet formation through core accretion in the outer disk and may be helpful for understanding the origin of directly imaged giant planets at wide separations.
+
+ %However, this minimum core mass decreases back to $M_{\rm{c, crit}}\sim5 M_{\oplus}$ once we assume realistic dust opacities that take into account grain growth.
 Our results yield lower critical core masses than corresponding studies for large planetesimal accretion rates.
 We therefore conclude that it is easier to form a planet by growing the core first, then accreting a massive gaseous envelope, rather than forming the core and atmosphere simultaneously.
+
+%requires rapid planetesimal accretion, which deposits energy into the atmosphere.
+
+%Heating increases the core mass required for collapse.
+However, the planetesimal accretion rate need not be constant --- in particular, it may decline over time.
+In this study we consider the limiting regime
+
+ %When a core embedded in a protoplanetary disk accretes an atmosphere with mass comparable to itself, the envelope is no longer in hydrostatic balance and unstable atmosphere collapse occurs.
 
@@
 
-%\textbf{Explain the context of the physical problem.
-Briefly describe the results of paper I, and say how that gave us a qualitative understanding of the numbers (critical core mass, cooling time etc.).
-Say that, however, in reality gas gas is non-ideal, non-polytropic etc., and that these effects need to be taken into account.
-Then describe what we do in the paper with the EOS.
-(Maybe also mention what we actually find in the intro?
-Although maybe too early to say it in the intro...)}
-
-%Current theories of giant planet formation postulate that these planets form either through core accretion (refs), in which solid planetesimals collide and grow into a massive solid core, which then accretes a gaseous envelope, or due to a gravitational instability in the protoplanetary disk that leads to fragmentation of the disk into self-gravitating clumps (refs, inc.
-\citealt{dangelo11}).
-%quote murray-clay, kratter 10, rafikov 05, etc.
-%
-%Standard core accretion models (refs) assume that the core and the atmosphere grow at the same time, and that planetesimal accretion deposits enough heat to alter the luminosity of the atmosphere, increasing the core mass required for the atmosphere to collapse, while the heat generated by the gravitational (Kelvin-Helmholtz) contraction of the atmosphere is neglected.
-These studies consider that the planet atmosphere is in steady state, in which all the luminosity due to planetesimal accretion is radiated away by the envelope, and  find that there exists a minimum (``critical'') core mass past which hydrostatic solutions can no longer be found and unstable atmosphere collapse occurs.
-%
-%Forming giant planets at wide separations in the disk poses theoretical challenges.
-On the one hand, gravitational instability generates objects that are too massive to explain the current observed properties of exoplanets (refs, inc.
-\citealt{rafikov05}).
-On the other hand, planetesimal accretion is slow at large distances in the disk, and therefore large cores may not be able to form before the dissipation of the disk (refs).
-It would therefore be easier if giant planets could form from smaller cores which would need less time to grow.
-
-One of the prevalent theories of giant planet formation is the core accretion model  (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{boden86}, \citealt{dangelo11}).
-It stipulates that Jupiter-sized planets form due to accretion of planetesimals into a solid core that grows large enough to attract a massive atmosphere.
-
-In models that assume a high planetesimal accretion rate (e.g., \citealt{rafikov06}), the gaseous envelope is heated solely by planetesimals and is therefore in a steady state at all times, in which all the incoming energy due to planetesimals is radiated away.
+%Fast accretion implies large critical masses.
+Because protoplanetary disks dissipate on timescales of a few Myrs, such cores must grow quickly to have access to the nebular gas.
+If cores and their atmospheres grow simultaneously, fast core growth requires large planetesimal accretion rates and hence large critical core masses.
+Fast core growth to large critical masses poses particular challenges in the outer disk, where dynamical times are long.
+Core accretion at wide separations thus requires either growth of cores at rates faster than typically assumed or, perhaps counterintuitively, moderate core growth rates followed by a halt to planetesimal accretion.
+In this study we explore the latter possibility.
+
+Core accretion --- a prominent theory of giant planet formation --- stipulates that Jupiter-sized planets form when planetesimal accretion produces a solid core large enough to attract a massive atmosphere   (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{boden86}, \citealt{dangelo11}).
+Because protoplanetary disks dissipate on timescales of a few Myrs, cores must grow quickly to accrete disk gas.
+Forming large cores is particularly challenging in the outer disk, where long dynamical times slow core growth.
+%, making the production of massive cores within the disk lifetime less likely.
+
+An \textit{in situ} solution to this problem requires accelerating core growth to rates larger than those typically assumed, but fast growth produces an additional difficulty.
+Though the minimum, or critical, core mass to form a giant planet is typically quoted to be $M_{\rm crit}\sim 10M_{\oplus}$, the actual value depends on how quickly planetesimals accrete (e.g., \citealt{ikoma00}, \citealt{pollack96}, \citealt{rafikov06}).
+Fast accretion heats a core's  atmosphere, increasing pressure support and hence $M_{\rm crit}$.
+However, planetesimal accretion rates need not be constant over time.
+If  accretion slows before the nebular gas dissipates, the target $M_{\rm crit}$ for successful giant planet formation is substantially reduced.
+%At distances $\gtrsim40-50$ AU, simultaneous core and atmosphere growth to critical masses may not be possible within a disk lifetime \citep{rafikov11}.
+This problem may be circumvented if planetesimal accretion slows before the nebular gas dissipates, $M_{\rm crit}$ may be substantially reduced if .
+%  An alternative solution is core growth at moderate rates and thus to smaller masses, followed by a halt to planetesimal accretion.
+This study explore the latter scenario.
+
+%Fast accretion generates an additional difficulty.
+
+%An \textit{in situ} solution to this problem requires accelerating core growth to rates faster than those typically assumed.
+However, the minimum, or critical, core mass to form a giant planet, though typically quoted to be $M_{\rm crit}\sim 10M_{\oplus}$, depends on how quickly planetesimals accrete (e.g., \citealt{ikoma00}, \citealt{pollack96}, \citealt{rafikov06}).
+Fast accretion increases $M_{\rm crit}$.
+However, $M_{\rm crit}$ may be substantially reduced if planetesimal accretion slows before the nebular gas dissipates.
+
+%or fast  to grow the core and atmosphere simultaneously, much faster rates would be required ---  fast planetesimal accretion increases the core mass necessary for atmospheric collapse.
+during the era of fast accretion requires larger critical core masses.
+
+%Core accretion at wide separations thus requires either growth of cores at rates faster than typically assumed in the first regime of high planetesimal accretion or, perhaps counterintuitively, moderate core growth rates followed by a halt to planetesimal accretion in the second regime --- i.e., the three-phase regime assumed by \citet{pollack96}.
+
+ %If cores and their atmospheres grow simultaneously, fast core growth requires large planetesimal accretion rates and hence large critical core masses.
+Core growth to these critical masses is particularly challenging in the outer disk, where long dynamical times slow core growth, making production of a massive core within the disk lifetime less likely.
+
+%Fast accretion implies large critical core masses. 
+In order to calculate the minimum value of $M_{\rm crit}$, we first need to understand how $M_{\rm crit}$ depends on the planetesimal accretion rate.
+%the evolution of a core's atmosphere depends on the rate at which planetesimals are accreted. 
+For high planetesimal accretion rates, atmospheres are primarily heated by planetesimals (e.g., \citealt{rafikov06}).
+As a result, the core's envelope is in a steady state at all times, in which all the incoming energy due to planetesimals is radiated away.
+As the envelope and core become comparable in mass, hydrostatic balance no longer holds and runaway gas accretion commences.
+In this case, $M_{\rm crit}$ is the maximum core mass for which the atmosphere is still in hydrostatic equilibrium.
+For a fixed planetesimal accretion rate and a set of disk conditions, $M_{\rm crit}$ is uniquely determined.
+
+%In models that assume a high planetesimal accretion rate (e.g., \citealt{rafikov06}), the gaseous envelope is heated solely by planetesimals and is therefore in a steady state at all times, in which all the incoming energy due to planetesimals is radiated away.
 In this regime, the core and atmosphere grow simultaneously, with the atmosphere increasing in mass faster than the core.
-As the envelope and core become comparable in mass, hydrostatic balance no longer holds and a rapid phase of unstable atmosphere collapse commences.
-The maximum core mass for which the atmosphere is still in hydrostatic equilibrium is defined in these studies as the ``critical core mass''.
-For a fixed planetesimal accretion rate and a set of disk conditions, the critical core mass is uniquely determined.
-
-Time-dependent core accretion models (e.g., \citealt{pollack96}, \citealt{ikoma00}) include variable planetesimal rates which allow for the Kelvin-Helmholtz (KH) contraction of the gaseous envelope to become important.
-In this regime, the atmosphere luminosity is not only supplied by planetesimals, but also by gas contraction.
-The envelope is no longer in a steady state, but rather it accretes gas as it cools.
+
+However, planetesimal accretion rates need not be constant at a given location in the protoplanetary disk throughout the disk's lifetime --- in particular, planetesimal accretion may significantly reduce as the planet's feeding zone is depleted of solids.
+Time-dependent core accretion models (e.g., \citealt{pollack96}, \citealt{ikoma00}) take this into account and include variable planetesimal accretion rates which allow for the Kelvin-Helmholtz (KH) contraction of the gaseous envelope to become important.
+In this regime, the envelope is no longer in a steady state, but rather it accretes gas as it cools.
 \citet{pollack96} find that the evolution of a giant planet can be separated into three phases.
@@
 \citet{pollack96} find that phase 2 lasts much longer than phases 1 and 3, and so the evolutionary timescale of the planet is set by KH contraction.
-Studying atmosphere growth onto a solid core of fixed mass (e.g., \citealt{pn05}) can thus provide accurate estimates for the timescale required to form a giant planet.
-
-%It is thus useful to consider atmosphere growth onto a solid core of fixed mass (e.g., \citealt{pn05}).
-
-Piso \& Youdin (submitted; hereafter Paper I) study atmosphere evolution around a fully grown core with fixed mass, assuming the luminosity of the envelope is solely due to KH contraction while planetesimal accretion is neglected.
-They build quasi-static two-layer atmosphere models, and derive a cooling model to connect static profiles in time and obtain an evolutionary history.
-They define the time $t$ at which unstable atmosphere collapse commences as the \textit{crossover time} $t_{\rm{co}}$, at which  $M_{\rm atm}(t_{\rm{co}})\sim M_{\rm c}$.
-From this they define as \textit{critical core mass} the minimum core mass for a protoplanet to undergo runaway gas accretion during the lifetime of the disk.
-They calculate the critical core mass for a variety of disk conditions, nebular gas compositions and opacities, and find that the critical core mass is smaller for lower disk temperatures (i.e., larger stellocentric distances), lower opacities and larger mean molecular weight of the gas.
-
-This study builds on the results of Paper I by making two important additions: we include realistic equation of state (EOS) effects, as well as realistic dust opacities.
-
- Paper I assumes that the nebular gas is an ideal diatomic gas described by a polytropic equation of state (EOS).
+%Studying atmosphere growth onto a solid core of fixed mass (e.g., \citealt{pn05}) can thus provide accurate estimates for the timescale required to form a giant planet.
+
+%Fast core growth to critical masses is particularly challenging in the outer disk.
+Long dynamical times at these distances slow core growth, making production of a massive core within the disk lifetime less likely.
+We thus investigate the possibility of forming gas giant at wide separations from smaller cores.
+%Core accretion at wide separations thus requires either growth of cores at rates faster than typically assumed or, perhaps counterintuitively, moderate core growth rates followed by a halt to planetesimal accretion.
+
+%We note, however, that long dynamical timescales in the outer disk slow core growth.
+Core accretion at wide separations thus requires either growth of cores at rates faster than typically assumed in the first regime of high planetesimal accretion or, perhaps counterintuitively, moderate core growth rates followed by a halt to planetesimal accretion in the second regime --- i.e., the three-phase regime assumed by \citet{pollack96}.
+
+It is not clear \textit{a priori} which of the two regimes --- atmosphere evolution dominated by planetesimal accretion versus KH contraction --- is the most relevant, as the conclusion of \citet{pollack96} depends on their assumptions about planetesimal accretion history.
+In particular, a core's feeding zone may be replenished by migration of solids through the protoplanetary disk.
+In this study, we accept the three-stage growth, in which the core is fully formed before it accumulates a significant atmosphere.
+During the slow phase of atmospheric growth we thus assume that the luminosity of the envelope is only due to KH contraction and neglect ongoing planetesimal accretion.
+Our results yield lower core masses than studies that assume large planetesimal accretion rates, confirming that it may indeed be easier to grow a giant planet from a fully formed core.
+Moreover, as additional heating due to accretion of solids limits the envelope's ability to cool and thus increases the core mass required for collapse, our results give a robust minimum for the core mass needed to form a giant planet before disk dissipation.
+
+%Both the steady state regime, where planetesimal accretion dominates the atmosphere evolution, and the KH regime, Therefore, it may be easier to grow a giant planet from a fully formed core, which is particularly relevant in the outer disk, where low planetesimal accretion rates prevent the growth of large cores.
+
+We follow the model developed in Piso \& Youdin (submitted; hereafter Paper I) and consider quasi-static two-layer atmospheres, embedded in the protoplanetary disk.
+Paper I calculates the minimum (critical) core mass to form a gas giant at several locations in the protoplanetary disk.
+We build on the results of Paper I by making two important additions: we include realistic equation of state (EOS) effects, as well as realistic dust opacities.
+
+ Paper I assumes that the nebular gas is ideal, diatomic and can be described by a polytropic equation of state (EOS).
 However, non-ideal effects such as dissociation and ionization have to be taken into account in order to obtain better quantitative results.
-In this study, we consider atmosphere growth assuming that the nebular gas is described by a realistic EOS given by the \citet{saumon95} EOS tables, and determine the critical core mass for a giant planet to form before the dissipation of the protoplanetary disk.
-We find that the critical core mass is more than twice as large than in the ideal case if realistic thermodynamic effects are taken into account.
-
-Another simplification in Paper I is the assumption that the dust opacity in radiative regions is that of interstellar grains.
+In this study, we consider atmosphere growth assuming a realistic gas with an EOS prescribed by the \citet{saumon95} tables.
+%We find that the critical core mass is more than twice as large than in the ideal gas case.
+
+ Another simplified assumption of Paper I is that the dust opacity in radiative regions is that of interstellar grains.
 In reality, opacities in protoplanetary disks, while not tightly constrained observationally, are unlikely to be interstellar.
-As we consider a regime of low planetesimal accretion in which solids have been segregated from the gas and are not being added back in, it is likely that the grain opacity will be lower than the standard interstellar medium (ISM) opacity.
-This effect is explored in Paper I, and it is found that a factor of ten opacity reduction results in a critical core mass twice as small.
+In particular, grain growth and dust settling --- which is particularly relevant to our regime of low planetesimal accretion in which solids have been segregated from the gas and are not being added back in --- lower the opacity.
 In this study, we use realistic opacity tables that are calculated based on observations of protoplanetary disks and take into account grain growth.
-We find that the resulting critical core mass may be significantly lower when ISM opacities are used, i.e.
+%We find that the resulting critical core mass may be significantly lower when ISM opacities are used, i.e.
 by more than a factor of three at 5 AU, and around 1.5 times lower at 100 AU.
-%We therefore see that it is less sensitive to disk temperature and hence semi-major axis.
+%We therefore see that it is less sensitive to disk temperature and hence semimajor axis.
+
+% making two important additions: we include realistic equation of state (EOS) effects, as well as realistic dust opacities.
 
 This paper is organized as follows.
-In section \S\ref{sec2} we review the quasi-static and cooling models derived in Paper I.
-We discuss the variations in the adiabatic gradient caused by non-ideal EOS effects in section \S\ref{deladtable}, and their implications for atmosphere evolution in section \S\ref{EOSeffects}.
-We determine the minimum core mass to form a giant planet during the disk lifetime in section \S\ref{critical}, with realistic EOS and dust opacity assumptions.
-In section \S\ref{acc} we compare our results to those obtained by studies that consider planetesimal accretion as the dominant source of energy.
-Finally, we summarize our findings in section \S\ref{conclusions}.
-
-%a solid core is first formed; the core grows, and once it becomes large enough it can accumulate a massive atmosphere.
-In standard core accretion models, in order to form a core large enough to attract an atmosphere, a high planetesimal accretion rate is needed, on average.
-The atmosphere is therefore heated due to accretion of planetesimals, and as a result it radiates away energy.
-The envelope is in a steady state at all times, and the atmosphere mass is a function of the core mass, which implies that every core mass maps uniquely to one atmosphere mass, for a given set of disk conditions and a planetesimal accretion rate.
-Moreover, larger cores hold fractionally larger envelopes.
-As a result, the atmosphere grows faster than the core, and eventually the atmosphere mass becomes comparable to the mass of the core.
-At this stage, hydrostatic equilibrium breaks down and a rapid phase of runaway gas accretion commences.
-This core mass $M\co$, for which the atmosphere mass $M_{\rm{atm}}$ satisfies $M_{\rm{atm}}\sim M\co$, is called the ``critical core mass'' in standard studies.
-
-%However, the planetesimal accretion rate is not necessarily constant at a given location in the protoplanetary disk throughout the disk lifetime (e.g., \citealt{ikoma00}).
-If the planetesimal accretion rate is very low, the atmosphere can no longer gain energy due to accretion of solids, and is instead dominated by gas accretion, contracting on a Kelvin-Helmholtz timescale.
-In Piso \& Youdin (2013; hereafter PY13) we studied the formation of giant planet atmospheres under the assumption that Kelvin-Helmoholtz gas contraction dominates the luminosity evolution of the atmosphere over planetesimal accretion.
-We built quasi-static two-layer atmosphere models with an inner convective region and an outer radiative region that matches smoothly onto the protoplanetary disk.
-We derived a cooling model to connect series of quasti-static atmospheres, and thus obtained an evolutionary history of the envelope.
-We defined the time $t$ at which unstable atmosphere collapse commences as the \textit{crossover time} $t_{\rm{co}}$, at which  $M_{\rm atm}(t_{\rm{co}})\sim M_{\rm c}$.
-From this we defined as \textit{critical core mass} the minimum core mass for a protoplanet to initiate runaway gas accretion during the lifetime of the protoplanetary disk.
-We studied this minimum mass for a variety of disk conditions, nebular gas compositions and opacities.
-We found that the critical core mass decreases for larger stellocentric distances, and is smaller for lower disk temperatures and opacities and for a higher mean molecular weight of the gas.
-
- %We found that the critical core mass to form a giant planet within the life time of the disk is smaller than the results yielded by studies that assume that the atmosphere evolution is dominated by the luminosity due to planetesimal accretion.
-We have showed that the planetesimal accretion rate needed to grow the core on a typical disk time scale is larger than the expected planetesimal accretion rates at large separations.
-As such, it is faster to form a planet by growing the core first in a fast planetesimal accretion regime (e.g., the core forms in the inner disk, then migrates outwards), then significantly reduce planetesimal accretion and allow a massive atmosphere to accumulate.
-
- %PY13 assume that the nebular gas can be described by a polytropic equation of state (EOS) corresponding to an ideal diatomic gas: $\delad=2/7$.
-In reality, however, non-ideal effects, such as gas dissociation and ionization, have to be taken into account.
-A realistic hydrogen-helium mixture can be described using tabulated equation of state tables.
-In this study, we use the \cite{saumon95} EOS tables to describe the nebular gas.
-We generate atmosphere profiles and estimate the atmospheric crossover time for a variety of disk conditions.
-From this, we determine the minimum core mass required for runaway gas accretion to commence within the typical life timescale of the protoplanetary disk.
-We find that the realistic equation of state yields larger critical core masses compared to the ideal gas polytrope.
-%We show, however, that our results are still smaller than corresponding studies for large planetesimal accretion rates.
-
-%Youdin \& Piso (2013) showed that giant planets can grow faster from small protoplanetary cores that are fully formed before significant gas accretion occurs.
-In this scenario, the planetesimal accretion rate is significantly slowed down during the gas contraction phase of the atmosphere.
-This reduction can arise due to dynamical clearing, or due to the core having formed in the inner parts of the disk and migrated outwards, etc.
-In this situation, the atmosphere evolution is dominated by the Kelvin-Helmholtz contraction of the envelope.
-The atmosphere is no longer in a steady state, but rather it accretes gas as it loses energy through radiation.
-
-%In our model we therefore assume that the luminosity evolution of the atmosphere is dominated by gas contraction, while the planetesimal accretion rate is negligible.
-As a result, the protoplanetary core has a fixed mass.
-We consider that the atmosphere evolves in time through stages of quasi-static equilibrium.
-Once the mass of the gaseous envelope becomes comparable to the mass of the solid core, the self-gravity of the atmosphere can no longer balance the pressure gradient and unstable hydrodynamic collapse commences.
-The time required for the atmosphere to grow to this stage is the characteristic growth time of the atmosphere.
-For a set of fixed gas and disk conditions, there exists a minimum core mass for which the atmosphere can grow on the time scale described above within the life time of the protoplanetary disk, which we define as the ``critical core mass''.
-
-%We develop a two-layer atmosphere model, with a convective inner region and a radiative outer region that matches smoothly on to the protoplanetary disk, and develop a cooling model that evolves the atmosphere in time.
-We aim to find the critical core mass for a giant planet to form before the dissipation of the disk.
-
-% In section \ref{sec2} we describe the assumptions of our atmosphere model, and derive the basic equations that govern the structure and evolution of the atmosphere.
-In section \ref{analytic}, we present a simplified analytic model that predicts the qualitative behavior of the numerical model.
-We describe our results in section \ref{KH}, and determined the critical core mass for planet formation during the life time of the protoplanetary disk in section \ref{critical}.
-We discuss our results in section \ref{discussion} and summarize our findings in section \ref{conclusions}.
-
-   %We assume that the luminosity is primarily generated in the convective zone of the atmosphere,
-
-%Giant planets play a fundamental role in shaping the orbital structure of planetary systems, and in affecting the delivery of volatiles to terrestrial planets in the
+In Section \S\ref{sec2} we review the quasi-static and cooling models derived in Paper I.
+We discuss the variations in the adiabatic gradient caused by the non-ideal EOS in Section \S\ref{deladtable}, and their effect on atmosphere evolution in Section \S\ref{EOSeffects}.
+We determine the minimum core mass to form a giant planet during the disk lifetime in Section \S\ref{critical}, with realistic EOS and dust opacity assumptions.
+In Section \S\ref{acc} we compare our results to those obtained by studies that consider planetesimal accretion as the dominant source of energy.
+Finally, we summarize our findings in Section \S\ref{conclusions}.
 
@@
 
-In this section we review the model developed in Paper I for the structure and evolution of a planetary atmosphere embedded in a protoplanetary disk.
-We describe the assumptions of the model and the properties of our assumed protoplanetary disk in section \S\ref{model}, and we summarize the equations describing the structure and time evolution of the atmosphere in section \S\ref{struct}.
+We begin by reviewing the model developed in Paper I for the structure and evolution of a planetary atmosphere embedded in a protoplanetary disk.
+We describe the assumptions of the model and the properties of our assumed protoplanetary disk in \S\ref{model}, and we summarize the equations describing the structure and time evolution of the atmosphere in \S\ref{struct}.
 
@@
 We assume that the planet consists of a solid core of fixed mass and a two-layer atmosphere composed of an inner convective region and an outer radiative zone that matches smoothly onto the disk.
-The two regions are separated by the Schwarzschild criterion for convective instability (see section \S\ref{struct}) .
+The two regions are separated by the Schwarzschild criterion for convective instability (see \S\ref{struct}) .
 We denote the surface between the two regions as the radiative-convective boundary (RCB), defined by a radius $r=R\cb$.
-We assume a low planetesimal accretion regime in which the atmosphere evolution is dominated by KH contraction.
-Moreover, we assume that the luminosity is constant throughout the outer radiative region (see section \S\ref{critical} for additional discussion).
-The atmosphere is spherically symmetric, self-gravitating and in hydrostatic balance.
+Moreover, we assume that the luminosity is constant throughout the outer radiative region (see Section \S\ref{critical} for additional discussion).
+We assume a low planetesimal accretion regime in which the atmosphere evolution is dominated by KH contraction.The atmosphere is spherically symmetric, self-gravitating and in hydrostatic balance.
 We note that spherical symmetry confines us to the outer regions of the disk ($a\gtrsim5$ AU), where the disk scale height is larger than the radius at which the planet matches onto the disk (see Paper I for further details).
@@
 \Sigma\di&=&2200\, (a/\text{AU})^{-3/2}\,\, \text{g cm}^{-2} \slabel{eq:diska}\\
-T\di &=& 120\, (a/\text{AU})^{-3/7} \,\,K, \slabel{eq:diskb} \\
-P\di&=&110\,  (a/\text{AU})^{-45/14} \,\, \text{dyn cm}^{-2} \slabel{eq:Pd}
+T\di &=& 120\, (a/\text{AU})^{-3/7} \,\,K \slabel{eq:diskb} \\
+P\di&=&110\,  (a/\text{AU})^{-45/14} \,\, \text{dyn cm}^{-2} \slabel{eq:Pd} \, ,
 \end{subeqnarray}
 
-\noindent with $a$ the semi-major axis and for a mean molecular weight $\mu=2.35$.
+\noindent with $a$ the semimajor axis and for a mean molecular weight $\mu=2.35$.
 
@@
 
-\noindent with $r$ the radial coordinate, $P$, $T$ and $\rho$ the gas pressure, temperature and density, respectively, $m$ the mass enclosed by the radius $r$, $L$ the luminosity from the surface of radius $r$, and $G$ the gravitational constant.
-The $\epsilon$ term represents the rate at which internal heat is generated per unit mass, while $\epsilon_g \equiv -T \frac{ds}{dt}$ is the heating per unit mass due to gravitational contraction, with $s$ the specific gas entropy.
+\noindent where $r$ is the radial coordinate, $P$, $T$ and $\rho$ are the gas pressure, temperature and density, respectively, $m$ is the mass enclosed by the radius $r$, $L$  is the luminosity from the surface of radius $r$, and $G$ is the gravitational constant.
+The gas is heated at a rate of $\epsilon_g \equiv -T \frac{ds}{dt}$ per unit mass due to gravitational contraction, where $s$ is the specific gas entropy, while  $\epsilon$ represents the rate at which internal heat is generated per unit mass.
 We do not take into account any internal energy sources and set $\epsilon=0$.
@@
 In our models the atmosphere is optically thick throughout the outer boundary.
-Where energy is transported through convection the temperature gradient is given by
+Where energy is transported by convection, the temperature gradient is
 
@@
 The convective and radiative layers of the envelope are separated by the Schwarzschild criterion (e.g., \citealt{thompson06}): the atmosphere is stable against convection when $\nabla < \delad$ and convectively unstable when $\nabla > \delad$.
-For effective convection, $\nabla \approx \delad$.
+Since convective energy transport is highly efficient,  $\nabla \approx \delad$ in convecting regions.
 The temperature gradient is thus given by $\nabla=\mathrm{min}(\delad, \delrad)$.
 
-The equation set (\ref{eq:struct}) is supplemented by an equation of state (EOS) relating pressure, temperature and density, as well as an opacity law.
+The Equation set (\ref{eq:struct}) is supplemented by an equation of state (EOS) relating pressure, temperature and density, as well as an opacity law.
 Paper I assumes an ideal gas polytropic EOS, $P=K\rho^{\gamma}$, with $K$ the adiabatic constant and $\gamma$ the adiabatic index.
-In this study we use realistic EOS tables, as explained in section \S\ref{deladtable}.
+In this study we use realistic EOS tables, as explained in Section \S\ref{deladtable}.
 Furthermore, Paper I assumes a power-law opacity given by
@@
 \noindent with $\beta$ and $F_{\kappa}$ constants, and $T_{\rm{ref}}$ a normalizing temperature.
-To estimate $\beta$ and $F_{\kappa}$ we use the \citet{bell94} opacity laws for ice grains: $\beta=2$, $F_{\kappa}=1$ and $T_{\rm ref}=100$ K.
-We note that these values are valid only for low disk temperatures, $T_d \lesssim 100 K$, where all dust components are present.
-Dust settling and grain growth lower both the normalization factor $F_{\kappa}$ and the power-law coefficient $\beta$.
-In this paper, we explore more realistic opacity laws and their effect on the critical core mass when compared to the ISM power-law (\ref{eq:opacitylaw}), as discussed in section \S\ref{critical}.
+Paper I sets $\beta=2$, $F_{\kappa}=1$ and $T_{\rm ref}=100$ K, appropriate for ice grains in the interstellar medium (ISM) \citep{bell94}.
+However, these values are valid only for low disk temperatures, $T\di \lesssim 100 K$, where all dust components are present.
+Furthermore, dust settling and grain growth lower both the normalization factor $F_{\kappa}$ and the power-law coefficient $\beta$.
+In this paper, we explore more realistic opacity laws and their effect on the critical core mass when compared to the ISM power-law, as discussed in Section \S\ref{sec:opacity}.
 
@@
 We choose $\rho_{\rm c}=3.2$ g cm$^{-3}$ (e.g., \citealt{pap99}).
-The atmosphere is assumed to match on to the protoplanetary disk at the Hill radius $R_{\rm H} \equiv a \Big(\frac{M_{\rm p}}{3 M_{\odot}}\Big)^{1/3}$, the distance at which the gravitational attraction of the planet and the tidal gravity due to the host star are equal.
-Outside the Hill sphere, the planet gravity is overcome by stellar tidal gravity, and hence only gas that lies within the Hill sphere can be gravitationally bound to the planet.
+The atmosphere matches onto the protoplanetary disk at the Hill radius $R_{\rm H} \equiv a \Big(\frac{M_{\rm p}}{3 M_{\odot}}\Big)^{1/3}$, where $M_{\rm p}$ is the total planet mass, $M_{\rm p}=M\co+M_{\rm atm}$.
+The Hill radius is the distance from the planet at which the gravitational attraction of the planet and the tidal gravity due to the host star are equal.
+Outside the Hill sphere, stellar tides unbind gas from the planet.
 The effective outer boundary of the atmosphere is the surface defined by the Bondi radius, $R_{\rm B} \equiv \frac{G M_{\rm p}}{c_{\rm s}^2}=\frac{G M_{\rm p}}{\mathcal{R} T\di}$.
 This is the distance from the planet at which the thermal energy of the nebular gas is of the order of the gravitational energy of the planet.
-Here $M_{\rm p}$ is the total mass of the planet, $c_{\rm s}$ is the isothermal sound speed, and $\mathcal{R}$ is the reduced gas constant: $\mathcal{R}=k_b/(\mu m_p)$, with $k_b$ the Boltzmann constant and $m_p$ the proton mass.
+Here $c_{\rm s}$ is the isothermal sound speed and $\mathcal{R}$ is the reduced gas constant: $\mathcal{R}=k_B/(\mu m_p)$, with $k_B$ the Boltzmann constant and $m_p$ the proton mass.
 For $R_{\rm B}<R_{\rm H}$, several studies assume that the atmosphere matches onto the disk at $R_{\rm B}$ (e.g., \citealt{ikoma00}, \citealt{pollack96}).
-One reason we always choose $R_{\rm H}$ as the matching radius is that, for hydrostatic solutions, the density at $R_B$ is larger than the disk density by an order unity factor \citep{rafikov06}.
+We choose the Hill radius as the outer boundary because the temperature and pressure at $R_{\rm H}$ are those of the disk,  $T(R_{\rm H})=T_{\rm d}$ and $P(R_{\rm H})=P_{\rm d}$.
+Outside $R_{\rm B}$, gas flows no longer circulate the planet, but rather belong to the disk.
+However, the density structure between $R_{\rm B}$ and $R_{\rm H}$ remains spherical and is still well described by hydrostatic balance \citep{ormel13}.
+The flows outside $R_{\rm B}$ could affect the atmosphere's cooling; however, we expect this effect to be weak, as heat losses predominantly occur deep in the planetary atmosphere.
+This justifies our outer boundary choice.
+We note, however, that we define the planet mass as the mass enclosed inside the smaller of $R_{\rm B}$ or $R_{\rm H}$.
+
+%, which justifies our outer boundary choice.
+
+%One reason we always choose $R_{\rm H}$ as the matching radius is that, for hydrostatic solutions, the density at $R_B$ is larger than the disk density by an order unity factor \citep{rafikov06}.
 Additionally, even though gas flows no longer circulate the planet outside $R_{\rm B}$, the density structure between $R_{\rm B}$ and $R_{\rm H}$ remains spherical and is still well described by hydrostatic balance \citep{ormel13}, which justifies our outer boundary choice.
@@
 
-Lastly, we review the cooling model developed in Paper I used to determine the time evolution of the atmosphere between subsequent static models.
+Lastly, we review the cooling model developed in Paper I, used to determine the time evolution of the atmosphere between subsequent static models.
 A protoplanetary atmosphere embedded in a gas disk satisfies the following cooling equation:
@@
 \label{eq:coolingglobal}
-L=L_c+\Gamma-\dot{E}+e_{\mathrm{acc}}\dot{M}-P_M \frac{\partial V_M}{\partial t}
+L=L_c+\Gamma-\dot{E}+e_{\mathrm{acc}}\dot{M}-P_M \frac{\partial V_M}{\partial t}.
 \end{equation}
@@
 We set $L\co=\Gamma=0$.
-The $\dot{E}$ term is the rate at which total energy (internal and gravitational) is lost, $e_{\mathrm{acc}}$ is the specific total energy brought in by mass accreting at the rate $\dot{M}$: $e_{\mathrm{acc}}=u-G M/R$, and the last term represents the work done on a surface mass element.
+The $\dot{E}$ term is the rate at which total energy (internal and gravitational) is lost, and $e_{\mathrm{acc}}$ is the specific total energy brought in by gas accreting at the rate $\dot{M}$: $e_{\mathrm{acc}}=u-G M/R$, where $u$ is the internal energy per unit mass.
+Finally, the last term represents the work done on a surface mass element.
 
@@
 
-We obtain an evolutionary series for the atmosphere by connecting sets of subsequent static atmospheres through the cooling equation (\ref{eq:coolingglobal}).
+We obtain an evolutionary series for the atmosphere by connecting sets of subsequent static atmospheres through the cooling Equation (\ref{eq:coolingglobal}).
 Details of our numerical procedure are described in Paper I.
+
+%Paper I finds that atmosphere growth continuously accelerates after self-gravity becomes important, which happens at relatively low atmosphere masses, $M_{\rm atm} \lessim 0.1 M\co$.
 
@@
 In order to describe the nebular gas, we use the interpolated EOS tables of \citet{saumon95} for a helium mass fraction $Y=0.3$, and extend them to lower temperatures and pressures corresponding to the conditions in our fiducial disk.
-More details on our extension procedure and on the methodology of combining the separate tables for hydrogen and helium are presented in \App{EOStables}.
+\App{EOStables} describes our extension procedure.
+
+%More details on our extension procedure and on the methodology of combining the separate tables for hydrogen and helium are presented in \App{EOStables}.
 
@@
 In contrast, non-ideal effects such as dissociation or ionization produce temperature-dependent variations in $\delad$.
-Figure \ref{fig:deladmap} shows a contour plot of the adiabatic gradient (defined in equation \ref{eq:delad})  as a function of gas temperature and pressure.
+Figure \ref{fig:deladmap} shows a contour plot of the adiabatic gradient (defined in Equation \ref{eq:delad})  as a function of gas temperature and pressure.
 We distinguish three separate temperature regimes:
@@
 
-We note that helium behaves like an ideal monatomic gas with \delad=2/5$ in our regime of interest.
-Its presence in the atmosphere thus only causes a small, constant upper shift in the adiabatic gradient of the mixture.
+We note that helium behaves like an ideal monatomic gas with $\delad=2/5$ in our regime of interest.
+Its presence in the atmosphere thus only causes a small, constant upward shift in the adiabatic gradient of the mixture.
 
@@
 While the boundaries of the region in which the free-energy EOS treatment fails can be determined from fundamental thermodynamic constraints, such calculations are not the object of this work.
-Instead, we note that less pressure is needed for hydrogen to become solid at lower temperatures, and so we choose as boundary a constant entropy curve ($\log(S)=8.4$) above the region in which the \citet{saumon95} model fails.
+Instead, we choose as boundary a constant entropy curve ($\log(S)=8.4$) above the region in which the \citet{saumon95} model fails.
 The expressions derived in \App{EOStables} are sufficient to give good results for the colored regions of the extended map, which fully cover the temperature and pressures ranges required by our models.}
@@
 
-In what follows we explain the behavior of the adiabatic gradient in the three temperature regimes separately.
+%In what follows we explain the behavior of the adiabatic gradient in the three temperature regimes separately.
 
@@
 
-For temperatures less than $\sim$$2000$ K but larger than $\sim$$300$ K, the hydrogen molecule is not energetic enough to dissociate and hydrogen behaves as an ideal diatomic gas.
-We see this in Fig.
-\ref{fig:deladmap} for 300 K $\lesssim T \lesssim 3000$ K, where the adiabatic gradient is approximately constant.
+For $300$ K $\lesssim T \lesssim 2000$ K, the hydrogen molecule is not energetic enough to dissociate and hydrogen behaves as an ideal diatomic gas with constant $\delad$ (Figure \ref{fig:deladmap}).
 The helium component of the gas causes a slight increase in the adiabatic index: $\delad \approx 0.3$ in this temperature range rather than 2/7 as is the case for a diatomic gas \footnote{Recall $\delad \equiv \frac{\gamma-1}{\gamma}$ for an ideal gas, with $\gamma$ the adiabatic index: $\gamma=7/5$ for a diatomic gas and 5/3 for a monatomic gas.}.
@@
 As the temperature becomes higher than $T \sim 2000-3000$ K, the internal energy of the hydrogen molecule becomes large enough to break the covalent bond between the atoms, and hydrogen starts dissociating.
-Further energy increase at temperatures of $\sim$$10^4$ K results in electrons being removed from atoms, i.e.
-hydrogen ionizes.
+At $T \gtrsim 10,000$ K, hydrogen ionizes.
 In stellar and giant planet interiors there is little overlap between the two processes: hydrogen is almost entirely dissociated into atoms by the time ionization becomes important.
 
-We see in Figure \ref{fig:deladmap} that the adiabatic gradient decreases significantly in regions of partial dissociation and partial ionization.
-This behavior is different than that of a mixture of molecular and atomic hydrogen where $2/7<\delad<2/5$, or a mixture of protons and electrons where $\delad=2/5$.
-We  explain this in what follows.
-% where hydrogen is either partially dissociated or partially ionized.
-
-%For a mixture of molecular and atomic hydrogen, we expect the adiabatic gradient to have an intermediate value between monatomic and diatomic gas, while for a mixture of protons and electrons the adiabatic index is just $2/5$ as for a monatomic ideal gas.
-However, we notice in Fig.
-\ref{fig:deladmap} that the adiabatic gradient decreases significantly in the regions where hydrogen is either partially dissociated or partially ionized.
-We further explain this behavior.
-
+As displayed in Figure \ref{fig:deladmap}, the adiabatic gradient decreases significantly in regions of partial dissociation and partial ionization.
+This behavior is different than that of a mixture of molecular and atomic hydrogen, for which $2/7<\delad<2/5$, or a mixture of protons and electrons, for which $\delad=2/5$.
 For a mixture of ideal gases, the total internal energy is given by the sum of the internal energies of the individual gases.
@@
 
-%At constant entropy, an increase in pressure results in an increase in the internal energy of the system, which usually causes the temperature to rise significantly.
-During partial dissociation, however, part of the internal energy is used to remove the electrons from atoms, and therefore there is less energy available to increase the temperature of the system.
-This behavior of constant entropy curves is seen in Figure \ref{fig:deladmap}.
-
-The ionization of atomic hydrogen is also dictated by the Saha equation, with the dissociation energy replaced by ionization energy, hence the adiabatic gradient has an analogous behavior, consistent with Fig.
+The ionization of atomic hydrogen is also dictated by the Saha equation, with the dissociation energy replaced by ionization energy, hence the adiabatic gradient behaves analogously, consistent with Fig.
 \ref{fig:deladmap}.
 
-%We now investigate the effect of the low adiabatic index caused by hydrogen dissociation on the luminosity and cooling time evolution of the atmosphere. The left panel of Figure (x) shows the
-
-%The resulting luminosity and time evolution are shown in Figure \ref{fig:Ltplotall}.
-We use again instantaneous atmosphere profiles to explain the differences.
-Figure \ref{fig:ETrprofall} shows the instantaneous temperature and energy profiles, as well as the location of the Bondi radius for a total fixed mass $M_{\rm{tot}}=11.8 M_{\oplus}$.
-We see that the spin effect at the outer boundary dictates the location of the radiative zone, and therefore the luminosity behavior, while dissociation deep in the atmosphere dictates the energy behavior.
-As compared to the polytrope, the real EOS therefore generates a deeper radiative zone with a lower luminosity, due to the lower adiabatic index in the outer regions, as well as an atmosphere with the bulk of its energy concentrated at the bottom, due to the low $\delad$ caused by dissociation.
-As shown above for the ideal gas polytropes, and remembering that $\Delta t \sim -\Delta E/L$, we see that both effects result in a longer time for the atmosphere to evolve.
-
 \vspace{0.2in}
@@
 As a diatomic molecule, hydrogen has five degrees of freedom, three associated with translational motion and two associated with rotation.
-The excitation temperature for rotation is $\Theta_r \approx 85$ K for the hydrogen molecule (e.g., \citealt{kittel}), hence the rotational states are fully excited at room temperature.
-As the gas temperature becomes comparable to $\Theta_r$, fewer rotational states are excited and rotation entirely ceases as $T \rightarrow 0$.
-%Here we explore the quantum mechanical effects associated with rotation.
-%We present expressions for the partition function associated with rotation and derive relevant thermodynamic variables in \App{EOStables}.
-For $T \ll \Theta_r$, the partition function is just equal to one, and no rotational states are excited for very low temperatures and hydrogen behaves like a monatomic gas.
-On the other hand, if $T \gg \Theta_r$, the molecule is energetic enough for all the rotational states to be fully excited.
-
-%In this section we discuss the quantum effects of the hydrogen isomeric forms and the way they affect the rotational energy and heat capacity of the hydrogen molecule at low temperatures, and hence the adiabatic gradient.
+The excitation temperature for rotation is $\Theta_r \approx 85$ K \citep{kittel}, hence the rotational states are fully excited at room temperature.
+As the gas temperature becomes comparable to $\Theta_r$, fewer rotational states are excited, and rotation entirely ceases as $T \rightarrow 0$.
+We note that at temperatures larger than $\gtrsim 6000$ K, vibrational motions also become important.
+While at $T\lesssim2000$ K, where our extension of the \citet{saumon95} EOS tables is valid, vibrational motion is negligible, we include these effects in our extension of the EOS for completion (see \App{EOStables} for details).
 
@@
 
-The adiabatic gradient scales as $\delad \sim 1/c_{\rm v}=1/(c_{\rm v,t}+c_{\rm v,r}) \sim 1/c_{\rm v, r}$, with $c_{\rm v}$ the specific heat capacity at constant volume, and $c_{\rm v, t}$ and $c_{\rm v,r}$ the translational and rotational components, respectively.
-The second equality is due to the fact that $c_{v, t}$ is independent of temperature and equal to $3\mathcal{R}/2$.
-We can therefore understand the $\delad$ behavior by studying the dependence on temperature of $c_{\rm v, r}$ of the ortho-para mixture.
-
-The internal energy per unit mass and specific heat capacity associated with rotation for the individual isomers and for the equilibrium mixture can be derived from their partition functions (see \App{EOStables} for details), and are plotted in Figure  \ref{fig:ucvr} (after \citealt{farkas35}, Figure 1).
-At low temperatures, parahydrogen is in the $j=0$ state and has no rotational energy, while orthohydrogen is in the $j=1$ state and has the energy of its first rotational level.
-Both para- and orthohydrogen, as well as their equilibrium mixture, behave like monatomic gases at low temperatures and thus have zero rotational heat capacity.
-This is consistent with $\delad=2/5$ at low temperatures as seen in Figure \ref{fig:deladmap}.
-As the temperature increases, the energetically higher-lying rotational states of para- and orthohydrogen are populated and the heat capacity of both spin isomers increases as a result.
-We note that the heat capacity of the equilibrium mixture is not a weighted average of the heat capacities of the individual components because it takes into account both the rotational energy uptake of para- and orthohydrogen, and also the shift in their equilibrium concentrations with temperature.
-This results in a peak in the heat capacity of the mixture around $\sim$$50$ K, as seen in the bottom plot of Figure \ref{fig:ucvr}.
-As $\delad \sim 1/c_{\rm{v,r}}$, it follows that the adiabatic gradient has to reduce, reach a minimum, then increase as the temperature rises, as shown in Figure \ref{fig:deladmap}.
-
-% As the temperature increases, the energetically higher-lying ($j=1$) ortho-hydrogen is formed, and the concomitant energy increase is seen as a peak in the heat capacity of the equilibrium mixture.
-
-%There are two significant maxima in the heat capacities of parahydrogen and of the mixture.
-At very low temperatures, the heat capacity of parahydrogen is zero because only the lowest accessible energy level $j=0$ is occupied and a temperature increase does not provide enough energy to populate the next higher level.
-When the temperature becomes sufficiently high to populate the second lowest level $j=2$, the heat capacity rapidly increases, passes through a maximum and starts to decrease when the second lowest level becomes saturated.
-
-%the adiabatic gradient is inversely proportional to the heat capacity, it
-
-%means that  the former has to first decrease from 2/5 as the temperature increases, reach a minimum around 50 K ($\delad \approx 0.25$ from Fig.
-\ref{fig:deladmap}), then gradually increase to 2/7 as for a diatomic gas.
-This behavior is illustrated in Figure \ref{fig:deladmap}.
-
-%The maxima in the ortho-para mixture appears when parahydrogen starts converting into orthohydrogen.
-The heat capacity of the equilibrium mixture is not a weighted average of the heat capacities of the individual components because it takes into account both the rotational energy uptake of para- and orthohydrogen, and also the shift in their equilibrium concentrations with temperature.
-At $T=0$, only parahydrogen is present in the equilibrium mixture; as the temperature is increased, the energetically higher-lying ($j=1$) ortho-hydrogen is formed, and the concomitant energy increase is seen as a peak in the heat capacity.
-As the adiabatic gradient is inversely proportional to the heat capacity, it means that  the former has to first decrease from 2/5 as the temperature increases, reach a minimum around 50 K ($\delad \approx 0.25$ from Fig.
-\ref{fig:deladmap}), then gradually increase to 2/7 as for a diatomic gas.
-This behavior is illustrated in Figure \ref{fig:deladmap}.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/ortho_para_energy.pdf}
-%%\vspace{-0.5in}
-\caption{Internal energy per unit mass and specific heat capacity associated with rotation for parahydrogen (dashed blue), orthohydrogen (dotted green), and the equilibrium mixture (solids red) as a function of temperature.
-After \citet{farkas35}, Figure 1.}
-\label{fig:ucvr}
-\end{figure}
-
- %The partition functions for ortho- and parahydrogen are described in \App{EOStables}.
- %At equilibrium, the relative abundance of the ortho and para states is given by the ratio of their partition functions.
-For very low temperatures there is only parahydrogen, as molecules are in the ground state with $j=0$, which corresponds to the para state.
-As the temperature is increased, parahydrogen starts converting into orthohydrogen, resulting in an ortho-para equilibrium ratio of 3:1 at room temperature.
-
-  %orthohydrogen, with parallel proton spins, and parahydrogen, with antiparallel proton spins.
-
- %The Pauli exclusion principle requires the total wavefunction of two fermions, such as protons, to be antisymmetric.
-As such, a symmetric spin wavefunction requires an antisymmetric rotational wavefunction, and vice-versa (\citealt{farkas35}).
-Parahydrogen has an antisymmetric spin wavefunction, which means that it can only occupy symmetric rotational states and hence the angular quantum number $j$ has to be even.
-By analogy, orthohydrogen must have an antisymmetric rotational wavefunction and can only occupy states with odd $j$.
-The partition functions for ortho- and parahydrogen are described in \App{EOStables}.
- %At equilibrium, the relative abundance of the ortho and para states is given by the ratio of their partition functions.
-For very low temperatures there is only parahydrogen, as molecules are in the ground state with $j=0$, which corresponds to the para state.
-As the temperature is increased, parahydrogen starts converting into orthohydrogen, resulting in an ortho-para equilibrium ratio of 3:1 at room temperature.
-
-%\begin{figure}[h]
-%\centering
-%\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/EOSeffects/ortho_para_ratio.pdf}
-%%%\vspace{-0.5in}
-%\caption{Relative abundance of ortho- and parahydrogen as a function of temperature.}
-%\label{fig:Zrel}
-%\end{figure}
-
-%\vspace{0.1in}
+%During the para-to-ortho conversion, part of the internal energy of the hydrogen molecule is used
+
+Similarly to the case of dissociation, as $T$ increases, part of the rotational energy $U_{\rm r}$ of the hydrogen molecule is used to convert parahydrogen to orthohydrogen rather than to only increase the temperature, so $U_{\rm r} \propto T$ no longer holds.
+The adiabatic gradient $\delad$ thus varies with temperature.
+\App{deladspin} explores the temperature dependence in detail.
+These $\delad$ variations, which occur for $T \lesssim 300$ K, are displayed in Figure \ref{fig:deladmap}.
+At higher temperatures, the ortho-para 3:1 equilibrium ratio is reached; no further isomer conversion occurs, and thus $\delad$ remains relatively constant until dissociation temperatures are reached.
 
@@
 
-Variations in the adiabatic gradient $\delad$ due to partial dissociation and quantum rotational effects (see section \S\ref{deladtable}) have two competing effects on the evolution of a core's atmosphere: (1) they yield a lower envelope luminosity, and (2) a larger amount of energy that needs to be radiated per unit mass, when compared to an ideal gas.
-Both effects combined slow down accretion and increase the growth time of the atmosphere, and therefore the crossover time and critical core mass.
-% (both defined in section \ref{intro}).
-%In this section we explore the effect of partial dissociation and hydrogen spin isomers (see section \ref{deladtable}) on the atmosphere evolution.
-%Due to the equation of state described in section \ref{deladtable}, a realistic atmosphere will not have a constant adiabatic gradient $\delad$, as was assumed in PY13.
-We thus first investigate the differences in atmosphere profiles and evolution for ideal gas polytropes with different adiabatic gradients in section \ref{deladpoly}.
-We then show how the variable adiabatic gradient affects the time evolution of the atmosphere in section \ref{deladeffect}.
-
-%\subsection{Effect of Adiabatic Gradient on Atmosphere Cooling Time}
-%\label{deladeffect}
-
-%In this section we discuss the effect of the variable adiabatic index described in subsection \ref{deladinterp} on the atmosphere cooling time.
-
-%We now explore the separate effects of hydrogen dissociation at high temperatures and spin isomers at low temperatures.
-From the adiabatic gradient table shown in Figure \ref{fig:deladmap} we find that $\delad$ has the flattest behavior around $T=500$ K.
-In this region, the gas behaves like an ideal gas with constant polytropic index $\delad \approx 0.3$, with the shift from diatomic gas caused by the helium in the mixture.
-We generate three sets of atmosphere profiles.
-The first one corresponds to an ideal gas of constant adiabatic index $\delad=0.3$.
-The second one is described by the real EOS for temperatures larger than 500 K and by an ideal gas with $\delad=0.3$ for $T<500$ K.
-Finally, the third profile consists of an ideal gas polytrope with $\delad=0.3$ for $T>500$ K and a real gas in the low temperature regime.
-We compare the first two profiles to show the dissociation effects, and the first and third profile to show the effects of ortho- and parahydrogen.
-
-%We now explore the effects of
-
-%\subsection{Dissociation and Spin Isomers Effects}
-%\label{deladeffect}
-
-%In order to investigate the separate effects of hydrogen dissociation at high temperatures and spin isomers at low temperatures on the atmosphere evolution, we generate
-
-In order to explain the quantum rotational effects at low temperatures, we make the following thought experiment: we imagine that realistic EOS effects only affect the upper atmosphere where temperatures are low, and that the gas is ideal and polytropic deep in the envelope.
-Conversely, we study the effects of dissociation at high temperatures by assuming that realistic EOS effects only matter at the bottom of the atmosphere where temperatures are high, and that the gas is polytropic in the outer regions.
+Variations in the adiabatic gradient $\delad$ due to partial dissociation and quantum rotational effects (see \S\ref{deladtable}) have two competing effects on the evolution of a core's atmosphere.
+They yield: (1) a lower envelope luminosity, and (2) a larger amount of radiated energy per unit of accreted mass, when compared to an ideal gas.
+Both effects combined slow down accretion and increase the growth time of the atmosphere and thus the critical core mass.
+
+In order to explain how quantum rotational states at low temperatures affect atmospheric growth, we perform the following thought experiment: we imagine that the EOS deviates from a polytrope only in the upper atmosphere where temperatures are low, and that the gas is ideal and polytropic deep in the envelope.
+Conversely, we study the effects of dissociation at high temperatures by assuming that the EOS is realistic at the bottom of the atmosphere where temperatures are high, but that the gas is polytropic in the outer regions.
 Figure \ref{fig:tplotall} shows the time evolution of atmospheres forming at 10 AU around cores of mass $M\co=10 M_{\oplus}$ and described by various equations of state, as follows:
@@
 We choose $T=500$ K as the cutoff temperature because the hydrogen-helium mixture behaves like an ideal gas, with $\delad=0.3$, in this temperature regime (see Figure \ref{fig:deladmap}).
-As such, the difference in cooling times between curves (1) and (2) shows the quantum rotational effects on the atmosphere evolution, while the difference between curves (1) and (3) accounts for hydrogen dissociation.
+As such, the time evolution difference between curves (1) and (2) shows the quantum rotational effects on the atmosphere evolution, while the difference between curves (1) and (3) accounts for hydrogen dissociation.
 We see that both dissociation and rotational states have a comparable effect on the atmosphere growth and result in slower cooling.
-The cooling time is dependent on both the total energy released due to the contraction of the envelope and the luminosity of the atmosphere.
+The growth time is dependent on both the total energy that must be released (i.e., the rate at which energy is released) and on the luminosity of the atmosphere.
 We explore the relative influence of these two factors separately.
@@
 %%\vspace{-0.5in}
-\caption{Elapsed time as a function of total mass (core + atmosphere) for a variety of EOS combinations (see text), for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=10 M_{\oplus}$.
+\caption{Elapsed time to grow a planet of total mass (core + atmosphere) for a variety of EOS combinations (see text), for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=10 M_{\oplus}$.
 Both hydrogen dissociation at high temperatures deep in the atmosphere and quantum rotational effects at low temperatures in the outer envelope result in slower cooling when compared to an ideal gas polytrope.}
@@
 \caption{We explore atmosphere growth around a core of $M\co=10 M_{\oplus}$ forming at 10 AU in our fiducial disk, for the EOS choices in Figure \ref{fig:tplotall}.
-Upper-left panel:  Instantaneous temperature profile for a total mass (core + atmosphere) of $12 M_{\oplus}$ as a function of radius.
+Upper-left panel:  Instantaneous temperature profile as a function of the radial coordinate when the total mass (core + atmosphere) is $12 M_{\oplus}$.
 The location of the RCB is marked.
-The quantum rotational effects at low temperatures in the outer region of the atmosphere dictate the location of the RCB.
-Upper-right panel:  Luminosity evolution as a function of total mass (core + atmosphere).
-The quantum rotational effects at low temperatures near the top of the atmosphere result in a lower luminosity for the realistic EOS (see text).
-Lower-left panel: Instantaneous energy profiles as a function on radius for a total mass (core + atmosphere) of $12 M_{\oplus}$.
-Hydrogen dissociation deep in the atmosphere causes the bulk of the energy to be concentrated at the bottom of the atmosphere.
-This increases the amount of energy per unit mass that needs to be radiated away, i.e.
-$|dE/dM|$, resulting in a longer crossover time for the realistic EOS when compared to the polytrope.
-Lower-right panel:  $|dE/dM|$ evolution as a function of total mass (core + atmosphere).
-More energy per unit mass has to be radiated away due to hydrogen dissociation when compared to an ideal gas polytrope, which results in slower growth.}
+Quantum rotational states at low temperatures in the outer region of the atmosphere result in a deeper RCB.
+Upper-right panel:  Luminosity evolution as a function of total mass.
+Quantum rotational effects at low temperatures near the top of the atmosphere result in a lower luminosity for the realistic EOS (see text).
+Lower-left panel: Instantaneous energy profiles as a function on radius for a total mass of $12 M_{\oplus}$.
+Hydrogen dissociation deep in the atmosphere causes the bulk of the energy to be concentrated at the bottom of the envelope.
+This increases the amount of energy per unit mass that needs to be radiated away ($|dE/dM|$) resulting in a longer crossover time for the realistic EOS when compared to the polytrope.
+Lower-right panel:  $|dE/dM|$ evolution as a function of total mass.
+Hydrogen dissociation deep in the atmosphere results in the bulk of the energy being concentrated at the bottom of the envelope, which increases the energy per unit mass that has to be radiated away when compared to an ideal gas polytrope.
+This results in slower growth.}
 \label{fig:all_plot}
@@
 The upper-right panel of Figure \ref{fig:all_plot} shows the luminosity evolution with mass for atmospheres (1)-(4), for an example core with $M\co=10 M_{\oplus}$ forming at 10 AU.
-We see that the lower luminosity of the fully realistic EOS (4) when compared to the polytrope (1) is due to the realistic EOS effects in the outer atmosphere (2).
+We see that the lower luminosity of the fully realistic EOS (4) when compared to the polytrope (1) is due to the realistic EOS in the outer atmosphere (2).
 This behavior is directly correlated to the depth of the radiative region of the envelope, which is shown in the upper-left panel of Figure \ref{fig:all_plot} for a total planet mass of $12 M_{\oplus}$.
@@
 This deeper RCB translates into a larger number of steps that photons need to diffuse to escape from the RCB, and therefore a lower luminosity (cf.
-equation \ref{eq:delrad}).
-In fact, a simple estimate of the luminosity emerging at the RCB can be found as follows.
-
-The equation of radiative diffusion (\ref{eq:structc}) can be approximated in terms of the radiative flux $\mathcal{F}$ as
-
-\begin{equation}
-\label{eq:sigmatau4}
-\mathcal{F} \sim \frac{\sigma T\cb^4}{\tau},
-\end{equation}
-where $\tau$ is the optical depth of the upper radiative layer. Furthermore, we find that $\tau$ is dominated by the optical depth of the first scale height, i.e.
-
-\begin{equation}
-\label{eq:taucb}
-\tau \sim \tau\cb=\kappa\cb \rho\cb H\cb,
-\end{equation}
-where 
-\begin{equation}
-\label{eq:Hcb}
-H\cb=\frac{k_b T\cb}{\mu m_H} \frac{R\cb^2}{G M\cb},
-\end{equation}
-with $\mu$ the mean molecular weight of the gas, $m_H$ the proton mass, and all other quantities evaluated at the bottom of the radiative layer.
-By substituting equations (\ref{eq:Hcb}) and (\ref{eq:taucb}) into equation (\ref{eq:sigmatau4}), and noting that $\mathcal{F}=L/(4 \pi (R\cb+H\cb)^2) \approx L / (4 \pi R\cb^2)$, we recover equation (\ref{eq:delrad}) applied at the boundary between the uppermost convective region of the atmosphere and the outer radiative layer (where $\delrad=\delad$).
+Equation \ref{eq:delrad}).
 
@@
 In contrast, energy is concentrated towards the outer boundary for the $\delad=0.3$ polytrope (1).
-Qualitatively, this can be explained through a simple analytic argument: the density profile in an adiabatic, non-self gravitating atmosphere composed of an ideal gas scales as $\rho(r) \sim r^{1/\delad-1}$ (see also Paper I).
-Since the energy per unit mass scales as $e(r) \sim \rho(r) r^2 \sim r^{1/\delad+1}$, it is easy to see that adiabats with lower adiabatic gradient have their energy more tightly packed towards the interior of the envelope, which is the case during dissociation when $\delad$ decreases significantly (see section \S\ref{deladtable} and Figure \ref{fig:deladmap}).
-It takes more energy to bring in gas deep in the atmosphere for an envelope that has the bulk of its energy concentrated towards the bottom,  which increases the amount of energy per unit mass that needs to be radiated, i.e.
-$|dE/dM|$, and therefore the cooling time, for the realistic EOS at high temperatures (3) when compared to the ideal gas polytrope (1).
+Qualitatively, this can be explained through a simple analytic argument: the density profile in an adiabatic, non-self-gravitating atmosphere composed of an ideal gas scales as $\rho(r) \sim r^{1/\delad-1}$ (see Paper I).
+Since the total energy scales as $E(r) \sim \rho(r) r^2 \sim r^{1/\delad+1}$, adiabats with lower $\delad$ have their energy more tightly packed towards the interior of the envelope, which is the case during dissociation when $\delad$ decreases significantly (see Section \S\ref{deladtable} and Figure \ref{fig:deladmap}).
+It takes more energy to bring gas deep into the atmosphere for an envelope that has the bulk of its energy concentrated towards the bottom,  which increases the amount of energy per unit mass that needs to be radiated, i.e.
+$|dE/dM|$.
+The cooling time is therefore larger for the realistic EOS at high temperatures (3) when compared to the ideal gas polytrope (1).
 This increase in $|dE/dM|$ is shown in the bottom-right panel of Figure \ref{fig:all_plot}.
 
-We have seen that the quantum rotational effect at the outer boundary dictates the location of the radiative zone, and therefore the luminosity behavior, while dissociation deep in the atmosphere dictates the total energy behavior.
+We have seen that quantum rotational states transitions due to the low temperatures at the top of the atmosphere increase the thickness of the radiative zone, and decrease the luminosity of the envelope, while dissociation at high temperatures deep in the atmosphere increases the amount of energy needed to increase the atmosphere mass by a fixed amount.
 Overall, both effects result in a longer time for the atmosphere to evolve.
@@
 
+\section{Opacity Effects on Atmosphere Evolution}
+\label{sec:opacity}
+
+So far in our model we have assumed that the dust opacity in the radiative region of the atmosphere is given by the standard ISM opacity.
+However, our scenario of low planetesimal accretion is likely to favor lower dust opacities, due to grain growth and dust settling.
+Grain growth, in particular, changes the particle size distribution when compared to the standard ISM size distribution (e.g., \citealt{pollack85}) and lowers the absolute value of the opacity.
+Enhanced metallicity due to planetesimal accretion, by construction not present during our atmosphere's growth, cannot make up for this reduction.
+
+%Observations of dust in protoplanetary disks (e.g., \citealt{beckwith90}, \citealt{beckwith91}, \citealt{perez12}) have shown evidence for grain growth and that the particle size distribution is not interstellar.
+
+Although grain growth and evidence for a non-ISM size distribution have been observed in protoplanetary disks (e.g., \citealt{beckwith90}, \citealt{beckwith91}, \citealt{perez12}), the size distribution of dust particles has not been tightly constrained.
+Typically, the grain size distribution is assumed to be a power-law:
+
+\begin{equation}
+\label{eq:graindistr}
+n(a) \sim a^{-p},
+\end{equation}
+where $n(a)$ is the number density of particles of size $a$ and $p=3.5$ (corresponding to a ``normal'' collisional cascade) or $p=2.5$ (an approximation for coagulation).
+In this work we use the  \citet{dalessio01} frequency-dependent opacity tables to obtain the temperature-dependent Rosseland mean opacity $\kappa$.
+We take as a fiducial case a maximum particle size of 1 cm and a grain size distribution given by Equation (\ref{eq:graindistr}) with $p=3.5$.
+Other choices for the power-law coefficient $p$ are discussed later in this section.
+
+The \citet{dalessio01} opacities are only relevant at temperatures that are sufficiently low for dust grains to remain solid ($T \lesssim 1000$ K).
+In the temperature regime where dust sublimates we use the \citet{bell94} analytic opacity laws, ensuring smooth transition from the grain growth opacities.
+
+The sharp drop in opacity ($\kappa \sim T^{-24}$, see Figure \ref{fig:delvsr}) due to dust sublimation lowers the radiative temperature gradient significantly (see Equation \ref{eq:delrad}), and may therefore generate radiative layers within the inner region of the atmosphere (see \App{radwindow}, Figure \ref{fig:delvsr}).
+This could in principle pose challenges for our model, since the additional luminosity generated in these radiative windows may be large enough to make our assumption of constant luminosity throughout the radiative layers invalid.
+In practice, however, the inner radiative windows are either very thin (Figure \ref{fig:delvsr}, middle panel), or the temperature gradient is flat enough to result in a roughly constant luminosity throughout the region (Figure \ref{fig:delvsr}, bottom panel; also see Equation \ref{eq:structc}).
+This results in a negligible extra luminosity generated in the radiative windows.
+However, the radiative windows may become non-negligible for other opacity choices and sufficiently low core masses, as we show later in this section.
+We note that despite the existence of one or more radiative windows in the planet's atmosphere, the luminosity that emerges at the outer boundary is set by the outer radiative layer.
+In fact, a simple estimate of the luminosity emerging at the RCB can be found as follows.
+
+The equation of radiative diffusion (\ref{eq:structc}) can be approximated in terms of the radiative flux $\mathcal{F}$ as
+
+\begin{equation}
+\label{eq:sigmatau4}
+\mathcal{F} \sim \frac{\sigma T\cb^4}{\tau},
+\end{equation}
+where $\tau$ is the optical depth of the upper radiative layer. Furthermore, we find that $\tau$ is dominated by the optical depth of the first scale height, i.e.
+
+\begin{equation}
+\label{eq:taucb}
+\tau \sim \tau\cb=\kappa\cb \rho\cb H\cb,
+\end{equation}
+with
+\begin{equation}
+\label{eq:Hcb}
+H\cb=\frac{k_B T\cb}{\mu m_p} \frac{R\cb^2}{G M\cb},
+\end{equation}
+where $\mu$ is the mean molecular weight of the gas, $m_p$ the proton mass, and all other quantities are evaluated at the RCB.
+By substituting Equations (\ref{eq:Hcb}) and (\ref{eq:taucb}) into Equation (\ref{eq:sigmatau4}), and noting that $\mathcal{F}=L/(4 \pi (R\cb+H\cb)^2) \approx L / (4 \pi R\cb^2)$, we recover Equation (\ref{eq:delrad}) applied at the boundary between the uppermost convective region of the atmosphere and the outer radiative layer (where $\delrad=\delad$) as
+
+\begin{equation}
+\label{eq:Lapprox}
+L \sim \frac{4 \pi \sigma T\cb^3 G M\cb \mu m_p}{k_B \kappa\cb \rho\cb}.
+\end{equation}
+Equation (\ref{eq:Lapprox}) can be used to estimate the luminosity emerging at the top of the convective layer of the atmosphere.
+We have verified that the luminosity emerging at the bottom of the outer radiative layer in our models can indeed be approximated by Equation (\ref{eq:Lapprox}).
+%As $T\cb$ only differs from the disk temperature by an order unity factor, $L\sim 1/\rho\cb$.
+The change in EOS thus results in a deeper RCB and a lower luminosity.
+
+Due to the variable number and position of radiative windows, and therefore radiative-convective boundaries, within the planet atmosphere, we cannot consistently calculate the time evolution of different atmospheres if we evaluate our cooling Equation (\ref{eq:coolingglobal}) at the RCB, as we do in our standard model.
+We choose to evaluate the cooling time at the Bondi radius instead (since our cooling model applies at any radius $R$, see section \S\ref{struct}).
+We note that our choice of $R$ does not change the estimate of the atmosphere evolution time, to order of magnitude, since the additional luminosity generated in all radiative regions is negligible for our opacity choice (also see Paper I).
+
 %\subsection{Outer Boundary Effects}
@@
 
-In this section we put together the results obtained in section \S\ref{EOSeffects} and determine the minimum core mass to initiate runaway gas accretion during the lifetime of the protoplanetary disk.
-As in Paper I, we define this minimum core mass as the \textit{critical core mass}, and the time elapsed until runaway gas accretion is initiated when $M_{\rm{atm}} \sim M\co$ as the \textit{crossover time}.
-We first explore the dependence of the crossover time on the core mass for a fixed semi-major axis.
+In this section we put together the results obtained in Sections \S\ref{EOSeffects}  and \S\ref{sec:opacity}, and determine the minimum core mass to initiate runaway gas accretion during the lifetime of the protoplanetary disk, i.e.
+the critical core mass.
+As in Paper I, we quantify the runaway growth timescale $t_{\rm run}$ as the time at which the atmosphere growth time $M_{\rm atm}/\dot{M}$ drops to 10\% of its maximum value (see Paper I for details).
+We first explore the dependence of the runaway growth time on the core mass for a fixed semimajor axis.
 We then determine the critical core mass to form a giant planet from a gas composed of a realistic hydrogen-helium mixture, and we compare this with the results from Paper I for an ideal diatomic gas.
@@
 
-Figure \ref{fig:tvsMplot} displays the time evolution and the crossover time for atmospheres forming around cores with masses between 10 $M_{\oplus}$ and 20 $M_{\oplus}$ at $a=10$ AU in our fiducial disk.
-The crossover time is shorter for higher mass cores, consistent with the results of Paper I.
+Figure \ref{fig:tvsMplot} displays the time evolution and the runaway growth time for atmospheres forming around cores with masses between 10 $M_{\oplus}$ and 20 $M_{\oplus}$ at $a=10$ AU in our fiducial disk.
+Higher mass cores have both shorter $t_{\rm run}$ and fractionally larger atmosphere masses at runaway, consistent with the results of Paper I.
 
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/tco_vs_Matm_10au.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/t_vs_M_10au.pdf}
 %%\vspace{-0.5in}
 \caption{Elapsed time as a function of atmosphere mass, for cores with fixed masses between $10 M_{\oplus}$ and $20 M_{\oplus}$ at $a=10$ AU in our fiducial disk.
-The circles mark the crossover time where $M_{\rm{atm}} \sim M_{\rm c}$.
+The circles mark the runaway growth time.
 The numbers are labeling the core mass in Earth masses.
-A larger core mass results in a shorter crossover time.}
+A larger core mass results in a lower $t_{\rm run}$ and a higher $M_{\rm atm}/M\co$ at runaway.}
 \label{fig:tvsMplot}
@@
 
-Figure \ref{fig:Mvsaplot} shows the critical core mass for a massive atmosphere to form during a typical lifetime of a protoplanetary disk $t=3$ Myrs (e.g., \citealt{jay99}), for a gas described by a realistic EOS and an ISM dust opacity given by equation \ref{eq:opacitylaw}.
+Figure \ref{fig:Mvsaplot} shows the critical core mass for a massive atmosphere to form during a typical lifetime of a protoplanetary disk $t=3$ Myrs (e.g., \citealt{jay99}), for a gas described by a realistic EOS and an ISM dust opacity given by Equation (\ref{eq:opacitylaw}) with $F_{\kappa}=1$ and $\beta=2$.
 The results of Paper I for an ideal diatomic gas are plotted for comparison.
-The inclusion of realistic EOS effects results in a critical core mass more than twice as large than under the ideal gas assumption.
-As such, non-ideal effects substantially affect the core mass needed to form a giant planet  before the dissipation of the protoplanetary disk.
+The use of a realistic EOS results in a critical core mass more than twice as large as that produced by a polytrope.
+As such, non-ideal effects substantially increase the core mass needed to form a giant planet  before the dissipation of the protoplanetary disk.
 
@@
 %%\vspace{-0.5in}
-\caption{The minimum core mass for an atmosphere to initiate runaway gas accretion within the lifetime of a typical protoplanetary disk $t \sim 3$ Myrs as a function of semi-major axis, for a realistic hydrogen-helium mixture and a standard ISM power-law opacity.
+\caption{The minimum core mass for an atmosphere to initiate runaway gas accretion within the lifetime of a typical protoplanetary disk $t \sim 3$ Myrs as a function of semimajor axis, for a realistic hydrogen-helium mixture and a standard ISM power-law opacity.
 The results of Paper I for an ideal diatomic gas are plotted for comparison.
-The realistic equation of state yields core masses larger by more than a factor of 2 when compared to the polytrope.}
+The realistic EOS yields core masses larger by more than a factor of 2 when compared to the polytrope.}
 \label{fig:Mvsaplot}
 \end{figure}
-
-The results above were obtained under the assumption that the dust opacity in the radiative region of the atmosphere is given by the standard ISM opacity (see also section \S\ref{struct}).
-However, our scenario of low planetesimal accretion is likely to favor lower dust opacities, due to grain growth and dust settling.
-Grain growth, in particular, lowers the absolute value of the opacity and changes the particle size distribution when compared to the standard ISM size distribution (e.g., \citealt{pollack85}).
-\textbf{Didn't understand comment...}
-
-%Observations of dust in protoplanetary disks (e.g., \citealt{beckwith90}, \citealt{beckwith91}, \citealt{perez12}) have shown evidence for grain growth and that the particle size distribution is not interstellar.
-
-Although grain growth and evidence for non-ISM size distribution have been observed in protoplanetary disks (e.g., \citealt{beckwith90}, \citealt{beckwith91}, \citealt{perez12}), the size distribution of dust particles has not been tightly constrained.
-Typically, the grain size distribution is assumed to be a power law:
-
-\begin{equation}
-\label{eq:graindistr}
-n \sim a^{-p},
-\end{equation}
-where $a$ is the particle radius and $p=3.5$ (corresponding to a ``normal'' collisional cascade) or $p=2.5$ (an approximation for coagulation).
-In this work we use the  \citet{dalessio01} frequency-dependent opacity tables to obtain the temperature-dependent Rosseland mean opacity $\kappa$,  assuming a maximum particle size of 1 cm and a grain size distribution given by equation (\ref{eq:graindistr}) with $p=3.5$ as a fiducial case.
-Other choices for the power-law coefficient $p$ are discussed later in this section.
-
-The \citet{dalessio01} opacities are only relevant at temperatures that are sufficiently low for dust grains to remain solid ($T \lesssim 1000$ K).
-In the temperature regime where dust sublimates we use the \citet{bell94} analytic opacity laws, ensuring smooth transition from the grain growth opacities.
-
-The sharp drop in opacity ($\kappa \sim T^{-24}$) due to dust sublimation lowers the radiative temperature gradient significantly (see equation \ref{eq:delrad}), and may therefore generate radiative layers within the inner region of the atmosphere (see \App{radwindow}, Figure \ref{fig:delvsr}).
-This could in principle pose challenges for our model, since the additional luminosity generated in these radiative windows may be large enough to make our assumption of constant luminosity throughout the radiative layers invalid.
-In practice, however, the inner radiative windows are either very thin (Figure \ref{fig:delvsr}, middle panel), or the temperature gradient is flat enough to result in a roughly constant luminosity throughout the region (Figure \ref{fig:delvsr}, bottom panel; also see equation \ref{eq:structc}).
-This results in a negligible extra luminosity generated in the radiative windows.
-While the above is true for our fiducial opacity law, the radiative windows may become non-negligible for other opacity choices and sufficiently low core masses, as we show later in this section.
-We note that despite the existence of one or more radiative windows in the planet atmosphere, the luminosity that emerges at the outer boundary is set by the top radiative window, as is the case for the standard inner convective-outer radiative envelopes.
-
-Due to the variable number and position of radiative windows, and therefore radiative-convective boundaries, within the planet atmosphere, we cannot consistently calculate the time evolution of different atmospheres if we evaluate our cooling equation (\ref{eq:coolingglobal}) at the RCB, as we do in our standard model.
-We choose to evaluate the cooling time at the Bondi radius instead (since our cooling model applies at any radius $R$, see section \S\ref{struct}).
-We note that our choice of $R$ does not change the estimate of the atmosphere evolution time, to order of magnitude, since the additional luminosity generated in all radiative regions is negligible for our opacity choice (also see Paper I).
 
@@
 %%\vspace{-0.5in}
-\caption{Critical core mass as a function of semi-major axis for radiative opacities that account for grain growth.
-The critical core mass is lower than in our fiducial case where dust grains have a standard ISM distribution.}
+\caption{Critical core mass as a function of semimajor axis for radiative opacities that account for grain growth (green triangles, with $p=3.5$ and $a_{\rm max}=1$ cm; see text for details).
+The critical core mass is lower than it would be if dust grains had an ISM-like size distribution (blue circles).}
 \label{fig:Mcritvsagg}
@@
 
-Figure \ref{fig:Mcritvsagg} shows the resulting critical core mass as a function of semi-major axis.
-The critical core mass is lower than in the standard case, and less sensitive to the distance in the disk, i.e.
+Figure \ref{fig:Mcritvsagg} shows the critical core mass as a function of semimajor axis, for a grain growth opacity with a size distribution given by the power-law (\ref{eq:graindistr}) with $p=3.5$ and maximum particle size of 1 cm.
+The critical core mass is lower than in the standard case, and less sensitive to location in the disk, i.e.
 the boundary conditions (temperature and pressure).
 We have shown in Paper I that the critical core mass is highly dependent on disk temperature rather than pressure.
-The temperature dependence, however, is mainly due to opacity: for the simplified analytic model developed in Paper I we found that the crossover time $t_{\rm co} \sim T^{\beta+1/2}$, with $\beta$ the power-law exponent in equation (\ref{eq:opacitylaw}).
-Opacity is less sensitive to temperature variations for larger grains and has an almost flat profile (see Figure \ref{fig:opacity}), which results in $\beta \ll 1$ and a much weaker temperature (and therefore semi-major axis) dependence of the critical core mass, as seen in Figure \ref{fig:Mcritvsagg}.
-Moreover, grain growth reduces the absolute value of the opacity, which results in an overall lower crossover time and critical core mass.
+The temperature dependence, however, is mainly due to opacity.
+For the simplified analytic model developed in Paper I we found that the crossover time $t_{\rm co} \sim T_{\rm d}^{\beta+1/2}$, with $\beta$ the power-law exponent in Equation (\ref{eq:opacitylaw}) and $t_{\rm co}$ a proxy for $t_{\rm run}$, defined as the time when $M_{\rm atm}=M\co$, as $t_{\rm run}$ cannot be determined self-consistently for the analytic model.
+Opacity is less sensitive to temperature variations for larger grains and has an almost flat profile (see Figure \ref{fig:opacity}), which results in $\beta \ll 1$ and a much weaker temperature (and therefore semimajor axis) dependence of the critical core mass, as seen in Figure \ref{fig:Mcritvsagg}.
+Moreover, grain growth reduces the absolute value of the opacity, which results in an overall lower runaway accretion time and critical core mass.
+% As noted in Section \S\ref{sec:opacity}, the critical core mass may be significantly lower if coagulation is taken into account, i.e.
+$p=2.5$.
 
@@
 %%\vspace{-0.5in}
-\caption{Crossover time for different grain size distribution, for a planet forming around a core $M\co=4 M_{\oplus}$.
-The standard ISM result is plotted for comparison.
-The crossover time is more than one order of magnitude lower when coagulation is accounted for, i.e.
+\caption{Runaway accretion times for different grain size distributions, for an atmosphere forming around a core with $M\co=4 M_{\oplus}$.
+The lines marked by green and red triangles have grain growth opacities with $a_{\rm max}=1$ cm, and $p=3.5$ and $p=2.5$, respectively (see text for details).
+The blue circle line has an ISM power-law opacity.
+The runaway accretion time is more than one order of magnitude lower when coagulation is accounted for, i.e.
 $p=2.5$.}
@@
 
-In our calculation so far we assumed that the size distribution of dust grains is the power law (\ref{eq:graindistr}) with $p=3.5$, which corresponds to a normal collisional cascade.
+So far we have assumed that the size distribution of dust grains is a power-law (\ref{eq:graindistr}) with $p=3.5$ which corresponds to a normal collisional cascade.
+If, however, coagulation is taken into account, the exponent $p$ can be approximated as $p=2.5$ \citep{dalessio01}.
+This results in a flatter and significantly lower opacity, which may substantially reduce the critical core mass.
+However, we have found that our model breaks down for low core masses ($M\co \lesssim 3 M_{\oplus}$) under our assumption of constant luminosity in the outer radiative layer.
+Figure \ref{fig:p25p35} shows the crossover time $t_{\rm co}$ as a function of semimajor axis for the lowest core mass for which our model is valid, $M\co=4 M_{\oplus}$.
+The crossover time is more than one order of magnitude lower for $p=2.5$, which implies that the critical core mass may be, in fact, significantly lower than presented in Figure \ref{fig:Mcritvsagg}.
+
+%Figure \ref{fig:p25p35} shows the crossover time $t_{\rm co}$ as a function of semimajor axis for a core of mass $M\co=4 M_{\oplus}$ and for two choices of the exponent $p$ in the power-law (\ref{eq:graindistr}): $p=3.5$, which corresponds to a normal collisional cascade, and $p=2.5$, which accounts for coagulation \citep{dalessio01}.
+We choose $M\co=4 M_{\oplus}$ as our model breaks down for lower core masses ($M\co \lesssim 3 M_{\oplus}$) under our assumption of constant luminosity in the outer radiative layer.
+The crossover time is lower for both choices of $p$ when compared to $t_{\rm co}$ when an ISM opacity is assumed.
+Moreover, $t_{\rm co}$ is more than one order of magnitude lower for $p=2.5$ than for $p=3.5$.
+As $t_{\rm co}$ is lower than the typical disk lifetime of a few Myrs for $M\co=4 M_{\oplus}$ and $p=2.5$, the critical core mass under these assumptions is lower than the core mass for which our model is valid.
+We thus cannot accurately calculate $M_{\rm crit}$ for $p=2.5$ and choose $p=3.5$ as our fiducial case.
+We note, however, that the critical core mass may be, in fact, significantly lower that the results we present in \S\ref{critical}.
+
+%So far we have assumed that the size distribution of dust grains is a power-law (\ref{eq:graindistr}) with $p=3.5$, which corresponds to a normal collisional cascade.
 If, however, coagulation is taken into account, the exponent $p$ can be approximated as $p=2.5$ \citep{dalessio01}.
@@
 However, we have found that our model breaks down for low core masses ($M_{\rm c} \lesssim 3 M_{\oplus}$) under this assumption, i.e.
-the outer radiative region becomes deep enough that the luminosity generated in this region can no longer be neglected.
-Figure \ref{fig:p25p35} shows the crossover time as a function of semi-major axis for the lowest core mass in which our model is valid, $M_{\rm c}=4 M_{\oplus}$.
-The crossover time is more than one order of magnitude lower when coagulation is accounted for, which implies that the critical core mass may be, in fact, significantly lower than presented in Figure \ref{fig:Mcritvsagg}.
+the outer radiative zone becomes deep enough that the luminosity generated in this region can no longer be neglected.
+The crossover time is more than one order of magnitude lower for $p=2.5$, which implies that the critical core mass may be, in fact, significantly lower than presented in Figure \ref{fig:Mcritvsagg}.
 
@@
 This is different from calculations that assume high planetesimal accretion rates and find that the atmosphere is in steady state and solely heated due to accretion of solids.
-In this section we compare our results for the critical core mass to analogous results from steady-state analytic calculations.
-We discuss the core accretion rates that are necessary for our regime to be valid in section \S\ref{raf1}.
-In section \S\ref{raf3}, we estimate core growth during atmosphere evolution at the maximum rate for which the KH regime is valid, and show it is negligible.
-Finally, we compare our results with those assuming fast planetesimal accretion in section \S\ref{raf2}.
+In this section we compare our results for the critical core mass to analogous results from steady-state fast planetesimal accretion calculations.
+We discuss the core accretion rates that are necessary for our regime to be valid in \S\ref{raf1}.
+In \S\ref{raf3}, we estimate core growth during atmosphere evolution at the maximum rate for which the KH regime is valid, and show it is negligible.
+Finally, we compare our results with those assuming fast planetesimal accretion in \S\ref{raf2}.
 
@@
 \end{equation}
-and $K_{\rm KH}$ is given by equation (\ref{eq:coolingglobal}) with $L\co=\Gamma=0$ (see section \ref{struct}). This condition is satisfied as long as the planetesimal accretion rate
+and $L_{\rm KH}$ is given by Equation (\ref{eq:coolingglobal}) with $L\co=\Gamma=0$. This condition is satisfied as long as the planetesimal accretion rate
 
@@
 
-To calculate $\dot{M}_{\rm c, KH}$, we choose as a fiducial case an atmosphere forming at 10 AU and with a core mass of $10 M_{\oplus}$.
-Since analytic studies assume an ideal gas EOS, for ease of comparison we choose an ideal gas polytrope with constant adiabatic gradient $\delad=2/7$ and mean molecular weight $\mu=2.35$ (see also Paper I).
-For this choice of parameters, the atmosphere crossover time is $t _{\rm{co}}\sim$ 1.2 Myrs, which is within the typical lifetime of a protoplanetary disk.
-The results are presented in Figure \ref{fig:accrates}.
-The atmosphere growth rate $\dot{M}_{\rm{atm}}$ is also plotted for comparison.
-We see that the core accretion rate has to be $\sim2-3$ orders of magnitude lower than the atmosphere accretion rate for our assumptions to be valid.
-If the core had accreted planetesimals at the $\dot{M}_{\rm c, KH}$ rate since it started forming, it couldn't have grown large enough to attract an atmosphere within typical disk lifetime.
-This is consistent with the requirement of our model that the planetesimal accretion rate is initially large during core growth, then significantly reduces as the gaseous envelope accumulates.
-This is a plausible scenario, as the core may have formed in the inner part of the disk and was later scattered outwards \citep{ida13}, or the planet's feeding zone could have been depleted of solids due to a giant neighbor.
-We further estimate two reference accretion rates.
-The first one is the core accretion rate $\dot{M}_{\rm c, acc}$ needed to grow the core to $M\co=10 M_{\oplus}$ on the same timescale as our model atmosphere, $\tau=1.2$ Myrs:
+To illustrate the magnitude of $\dot{M}_{\rm c, KH}$, we choose as a fiducial case an atmosphere forming at 10 AU and with a core mass of $10 M_{\oplus}$.
+Since analytic studies of critical core masses at high planetesimal accretion rates assume an ideal gas EOS, for ease of comparison we choose an ideal gas polytrope with constant adiabatic gradient $\delad=2/7$ and mean molecular weight $\mu=2.35$ (see also Paper I).
+For this choice of parameters, the runaway accretion time is $t _{\rm run}\sim$ 1.3 Myrs, which is within the typical lifetime of a protoplanetary disk.
+We also estimate two reference accretion rates.
+The first one is the core accretion rate $\dot{M}_{\rm c, acc}$ needed to grow the core to $M\co=10 M_{\oplus}$ on the same timescale as our model atmosphere, $\tau=1.3$ Myrs:
 
@@
 The second reference planetesimal accretion rate is $\dot{M}_{\rm c, Hill}$, a typically assumed planetesimal accretion rate for which the random velocities of the planetesimals are of the order of the Hill velocity around the protoplanetary core (for a review, see \citealt{goldreich04}).
-This is the accretion rate at the boundary between the dispersion dominated and shear dominated regimes.
-Following \citet{rafikov06} (equation A1),
+Following R06 (equation A1),
+
+%This is the accretion rate at the boundary between the dispersion dominated and shear dominated regimes.
 
@@
 where $\Sigma_{\rm p}$ is the surface density of solids, assumed to satisfy $\Sigma\di \approx 100 \Sigma_{\rm p}$ for a dust-to-gas ratio of 0.01.
+
+Figure \ref{fig:accrates} shows that $\dot{M}_{\rm c,KH}$ is $\sim2-3$ orders of magnitude lower than $\dot{M}_{\rm c, acc}$.
+Had the core accreted planetesimals at the $\dot{M}_{\rm c, KH}$ rate since it started forming, it could not have grown large enough to attract an atmosphere within typical disk lifetime.
+This is consistent with the results of \citet{pollack96}, which find that the planetesimal accretion rate decreases with time as the core's feeding zone is depleted, and with the requirement of our model that the planetesimal accretion rate is initially large during core growth, then significantly reduces as the gaseous envelope accumulates.
+This is a plausible scenario, as the core may have formed in the inner part of the disk and was later scattered outwards \citep{ida13}, or the planet's feeding zone could have been depleted of solids due to a giant neighbor.
 
@@
 %\vspace{-0.5in}
-\caption{Various accretion rates in the case of a planet forming at 10 AU and with a core mass $M\co=10 M_{\oplus}$, for a polyropic EOS and ISM power-law opacity.
-For this choice of parameters, the crossover time (defined in section \S\ref{critical}) is $t_{\rm co} \sim 1.2$ Myrs.
-The $\dot{M}_{\rm{atm}}$ (solid blue) curve represents the growth rate of the atmosphere as estimated by our model, and $\dot{M}_{\rm{c,KH}}$ (dotted red) is the maximum planetesimal accretion rate during the gas contraction phase in order for our regime to be valid, i.e.
+\caption{Various accretion rates for a planet forming at 10 AU and with a core mass $M\co=10 M_{\oplus}$, using a polyropic EOS and ISM power-law opacity.
+For this choice of parameters, the runaway accretion time is $t_{\rm run} \sim 1.3$ Myrs.
+The $\dot{M}_{\rm{atm}}$ (solid blue) curve represents the growth rate of the atmosphere as estimated by our model.
+The core accretion rate $\dot{M}_{\rm c, acc}$ (dashed green) necessary to grow the core on the timescale $t_{\rm run}$ is larger than $\dot{M}_{\rm c, KH}$ (dotted red), the maximum planetesimal accretion rate during KH contraction for which our regime is valid (see text).
+The $\dot{M}_{\rm c, acc}$ rate is also larger than the frequently used planetesimal accretion rate $\dot{M}_{\rm c, Hill}$ (dashed-dotted light blue) for which the random velocity of the planetesimals is given by the Hill velocity due to the core (see text).
+This motivates our requirement  that planetesimal accretion must have slowed down after core growth for our model to be valid.}
+
+%and $\dot{M}_{\rm{c,KH}}$ (dotted red) is the maximum planetesimal accretion rate during the gas contraction phase in order for our regime to be valid, i.e.
 $L_{\mathrm{acc}} < L_{\rm{KH}}$ (see text).
-For comparison, we plot the core accretion rate $\dot{M}_{\rm{c,acc}}$ (dashed green) necessary to grow the core on a timescale $\tau=t_{\rm co}$, and a typically assumed planetesimal accretion rate $\dot{M}_{\rm{c,Hill}}$ (dash-dotted light blue) for which the random velocity of the planetesimals is given by the Hill velocity due to the core (see text).
-We note that $\dot{M}_{\rm c, KH}$ and $\dot{M}_{\rm c, Hill}$ are both lower than $\dot{M}_{\rm c, acc}$, consistent with our assumptions that planetesimal accretion must have slowed down after core growth.}
+For comparison, we plot the core accretion rate $\dot{M}_{\rm{c,acc}}$ (dashed green) necessary to grow the core on a timescale $\tau=t_{\rm run}$, and a frequently used planetesimal accretion rate $\dot{M}_{\rm{c,Hill}}$ (dash-dotted light blue) for which the random velocity of the planetesimals is given by the Hill velocity due to the core (see text).
+We note that $\dot{M}_{\rm c, KH}$ and $\dot{M}_{\rm c, Hill}$ are both lower than $\dot{M}_{\rm c, acc}$, motivating our requirement  that planetesimal accretion must have slowed down after core growth for our model to be valid.}
 \label{fig:accrates}
@@
 
-In this section we investigate whether planetesimal accretion during the gas contraction phase at the rate $\dot{M}<\dot{M\co}_{\rm{KH}}$  can alter the core mass enough to affect the time evolution of the atmosphere.
+Planetesimal accretion during the KH contraction phase of atmosphere growth at a rate $\dot{M}<\dot{M\co}_{\rm{KH}}$  cannot alter the core mass enough to affect the time evolution of the atmosphere.
 We can quantitatively estimate the increase in core mass as
@@
 \label{eq:cminc}
-\Delta M_{\rm{c}} = \int_0^{t_{\rm{co}}} \dot{M_{\rm{c}}} dt \approx \sum_i \dot{M_{\rm{c}}}_i \Delta t_i,
+\Delta M_{\rm{c}} = \int_0^{t_{\rm{run}}} \dot{M_{\rm{c}}} dt \approx \sum_i \dot{M_{\rm{c}}}_i \Delta t_i,
 \end{equation}
@@
 
- \noindent from equation (\ref{eq:Lacc}), with $L_i$ the luminosity of the atmosphere at time $t_i$ in our model.
-For $M_{\rm{c}}=10 M_{\oplus}$, we find $\Delta M_{\rm{c}} \approx 0.2 M_{\oplus} \ll 10 M_{\oplus}$.
+ \noindent from Equation (\ref{eq:Lacc}), with $L_i$ the luminosity of the atmosphere at time $t_i$ in our model.
+For $M_{\rm{c}}=10 M_{\oplus}$, we find $\Delta M_{\rm{c}} \approx 0.05 M_{\oplus} \ll 10 M_{\oplus}$.
 It follows that a significant increase in core mass that could potentially alter the time evolution of the atmosphere would occur on a much longer timescale then the crossover time for the unperturbed atmosphere.
@@
 
-Next, we check that runaway gas accumulation due to planetesimal accretion cannot already commence before the atmosphere becomes unstable due to KH contraction, as our regime is no longer relevant under such conditions.
-We recall that in steady-state studies, the critical core mass is  the maximum core mass for which the atmosphere is still in hydrostatic balance; runaway gas accumulation occurs if the core mass grows further.
-A higher planetesimal accretion rate results in a larger critical core mass for the steady-state models.
-As such, if unstable atmosphere collapse does not occur due to planetesimal accretion for the lowest value of $\dot{M}_{\rm c, KH}$ over the course of the atmosphere's growth (see, e.g., Figure \ref{fig:accrates}), then it can only occur in the KH dominated regime.
-In what follows we calculate the critical core mass $M_{\rm crit, KH}$ corresponding to $\min(\dot{M}_{\rm c, KH})$ and show that it is higher that the critical core mass we determined in the KH dominated regime.
+We compare our results for the critical core mass $M_{\rm crit}$ with those of studies that assume large planetesimal accretion rates.
+These studies find that the atmosphere is in steady state at all times and that static solutions only exist up to a maximum core mass, which they define as the critical core mass.
+If the $M_{\rm crit}$ found by static studies were lower than the $M_{\rm crit}$ in our regime of negligible solids accretion, the atmosphere would undergo runaway gas accretion before KH contraction became dominant, and our regime would not be relevant.
+We show, however, that our model yields lower core masses than those found when fast planetesimal accretion is considered.
+
+In static studies, the critical core mass is larger for higher planetesimal accretion, as additional heating increases the core mass required for collapse.
+As such, if atmosphere collapse does not occur due to planetesimal accretion for the lowest value of $\dot{M}_{\rm c, KH}$ over the course of the atmosphere's growth (see, e.g., Figure \ref{fig:accrates}), then it can only occur in the KH dominated regime.
+%In what follows we calculate the critical core mass $M_{\rm crit, KH}$ corresponding to $\min(\dot{M}_{\rm c, KH})$ and show that it is higher that the critical core mass we determined in the KH dominated regime.
 
@@
 
-In order to estimate $M_{\rm crit, KH}$, we use the results of \citet{rafikov06} for low luminosity atmospheres forming in the outer disk ($>2-5$ AU), consistent with our region of interest.
-\citet{rafikov06} assumes an ideal gas polytropic EOS and a lower opacity than the standard ISM opacity that we use in our calculations (see equation \ref{eq:opacitylaw}).
-We thus calculate the critical core mass for an ideal gas polytrope with the normalization factor $F_{\kappa}$ reduced by a hundred, which is comparable to the opacity law used by \citet{rafikov06} \footnote{The power-law opacity of \citet{rafikov06} is scaled to the (semi-major axis dependent) disk temperature, while our opacity is scaled to an absolute reference temperature.
-We thus cannot directly use the \citet{rafikov06} opacities for our comparison.}.
+In order to estimate the critical core mass $M_{\rm crit, KH}$ corresponding to planetesimal accretion at the rate $\dot{M}_{\rm c, KH}$, we use the results of R06 for low luminosity atmospheres forming in the outer disk ($>2-5$ AU), consistent with our region of interest.
+R06 assumes an ideal gas polytropic EOS and a lower opacity than the standard ISM (see Equation \ref{eq:opacitylaw}).
+We thus calculate the critical core mass for an ideal gas polytrope with the normalization factor $F_{\kappa}$ reduced by a hundred, which is comparable to the opacity law used by R06 \footnote{The power-law opacity of R06 is scaled to the (semimajor axis dependent) disk temperature, while our opacity is scaled to an absolute reference temperature.
+We thus cannot directly use the R06 opacities for our comparison.}.
 %(see Paper I).
 
-Following \citet{rafikov06}, we find the following expression for the critical core mass when accretion luminosity dominates the evolution of the atmosphere:
+Following R06, we find that the critical core mass when accretion luminosity dominates the evolution of the atmosphere can be expressed as
 
@@
 \end{equation}
-with $C$ an order unity constant depending on the adiabatic gradient and disk properties (see \citealt{rafikov06}, equation B3).
-From equation (\ref{eq:McdotKH}), the accretion rate $\dot{M}_{\rm c, KH}$ depends on the core mass $M\co$.
-We find $M_{\rm crit, KH}$ numerically by setting $M\co=M_{\rm crit, KH}$ on the right-hand side of equation (\ref{eq:critraf}).
+with $C$ an order unity constant depending on the adiabatic gradient and disk properties (see R06, Equation B3).
+From Equation (\ref{eq:McdotKH}), the accretion rate $\dot{M}_{\rm c, KH}$ depends on the core mass $M\co$.
+We find $M_{\rm crit, KH}$ numerically by setting $M\co=M_{\rm crit, KH}$ on the right-hand side of Equation (\ref{eq:critraf}).
 The result is displayed in Figure \ref{fig:raf2}; the critical core mass corresponding to planetesimal accretion at the rates displayed in Figure \ref{fig:accrates} is displayed for comparison.
@@
 %\vspace{-0.5in}
-\caption{Comparison between the critical core mass $M_{\rm{crit, KH}}$ due to planetesimal accretion and the assumed fixed core mass when gas contraction dominates.
+\caption{Comparison between the critical core mass $M_{\rm{crit, KH}}$ given significant planetesimal accretion and the critical core mass when gas contraction dominates.
 Our results yield lower core masses than in the fast planetesimal accretion case (e.g., \citealt{rafikov06}).
@@
 
-The critical core mass due to KH contraction is smaller than in the case in which planetesimal accretion dominates the evolution of the atmosphere.
-This brings us to two conclusions.
-First, we confirm that planetesimal accretion can be safely ignored in our regime of interest.
+The critical core mass in the regime where KH contraction dominates is smaller than in the case in which planetesimal accretion dominates the evolution of the atmosphere.
+This leads to two conclusions:
+
+\begin{enumerate}
+\item Planetesimal accretion can be safely ignored in our regime.
+\item Giant planets can form from smaller cores if planetesimal accretion significantly reduces during atmosphere growth. 
+\end{enumerate}
+
+As additional heating due to planetesimals limits the atmosphere's ability to cool, our result represents a true minimum for the core mass needed to form a gas giant during the lifetime of the protoplanetary disk.
+
+% First, we confirm that planetesimal accretion can be safely ignored in our regime of interest.
 Secondly, this comparison tells us that the core mass needed to form a giant planet is smaller if the core forms first, then accretes a massive envelope, than in the case where the core and atmosphere grown simultaneously in a high planetesimal accretion regime.
@@
 
-% \textbf{This works for this particular choice of core, disk etc.
-conditions, and I am expecting it to work for most of our parameter space, but it would probably be good to explore this numerically, at least to order of magnitude, with the estimates we have.}
-
-%This means that for large core masses, planetesimal accretion will lead to runaway gas accretion before gas contraction starts dominating, and hence our model is not applicable in that regime.
-However, we have found in section \ref{critcore} that unstable atmospheres collapse occurs within the life time of the disk for protoplanetary cores smaller than this.
-
-%\section{Model Validity??? \textbf{(for now, in lack of a better title)}}
-%
-%\textbf{Discuss geometric effects (spherical symmetry in the inner disk), opacity effects (also in the 	inner disk) and planetesimal accretion; from this build the parameter space where the model is 	valid and confirm from the plots in the previous section that this is where we're looking.
-Part of the text regarding the planetesimal accretion discussion is below, as initially written for Paper I}
-%
-%\subsection{Geometric Effects}
-%\subsection{Opacity Effects}
-%
-%
-%\section{Discussion}
-%\label{discussion}
-
-%In this section we discuss the parameter space of validity of our model.
-We calculate the conditions under which planetesimal accretion can be ignored, and review the effects of some of the approximations that come into our model.
-%
-%  
- \section{Conclusions}
+ \section{Summary}
  \label{conclusions}
@@
 We consider atmosphere evolution around fully grown cores and determine the minimum (critical) core mass required to form a gas giant during the typical lifetime of a protoplanetary disk.
-We improve the model developed in Paper I (\textbf{does this also need to be written out as Piso \& Youdin?}) by including realistic equation of state (EOS) tables and dust opacities.
-For a realistic EOS and interstellar grain opacity, the critical core mass is $\sim$$20 M_{\oplus}$ at 5 AU and drops to $\sim$$6 M_{\oplus}$ in our fiducial disk.
+We improve the model developed in Piso \& Youdin (submitted, hereafter Paper I) by including realistic equation of state (EOS) tables and dust opacities.
+
+ For a realistic EOS and grain growth opacity with maximum particle size $a_{\rm max}=1$ cm and a power-law size distribution (\ref{eq:graindistr}) with $p=3.5$, the critical core mass is $\sim$$7 M_{\oplus}$ at 5 AU in our fiducial disk and drops to $\sim$$4.5 M_{\oplus}$ at 100 AU.
+This result for the critical core mass $M_{\rm crit}$ is similar to the one obtained in Paper I for an ideal gas polytropic EOS and interstellar opacity.
+The realistic EOS and grain growth opacity have two competing effects on $M_{\rm crit}$:
+
+ \begin{itemize}
+ \item Realistic EOS effects increase $M_{\rm crit}$ by a factor of $\sim$$2$ when compared to the ideal gas polytrope.
+ \item Grain growth opacities decrease $M_{\rm crit}$ by a factor of $\sim$$3$ at 5 AU and by a factor of $\sim$$1.5$ at 100 AU, when compared to ISM opacities, for a particle distribution given by the power-law (\ref{eq:graindistr}) with $p=3.5$ and a maximum particle size of 1 cm.
+The critical core mass is less sensitive to the location in the disk when realistic opacities are used.
+If $p=2.5$, an approximation for coagulation,  $M_{\rm crit}$ further reduces.
+ \end{itemize}
+
+ %interstellar grain opacity, the critical core mass is $\sim$$20 M_{\oplus}$ at 5 AU and drops to $\sim$$6 M_{\oplus}$ in our fiducial disk.
 These results are more than twice as large than those calculated in Paper I for a polytropic EOS, and bring the critical core mass to $\sim$$10 M_{\oplus}$, the value typically quoted in many core accretion studies.
@@
 %---------------------------------------------------------------------------
+
+\acknowledgements{We thank Sean Andrews for providing us with the \citet{dalessio01} opacity tables.
+We also thank Eugene Chiang and Steven Cranmer for useful conversations.
+ANY thanks Phil Arras for his help with entropy units conventions in the equation of state tables.}
+
 \bibliographystyle{apj}
@@
 
-In this study we consider atmosphere growth in the outer parts of protoplanetary disks ($a \lesssim 100$ AU), where temperature and pressure drop to $T \sim 20$ K and $P \sim10^{-4}$ dyn cm$^{-2}$ for our MMSN disk model (see equations \ref{eq:diskb} and \ref{eq:Pd}).
+In this study we consider atmosphere growth in the outer parts of protoplanetary disks ($5<a<100$ AU), where temperature and pressure drop to as little as $T \sim 20$ K and $P \sim10^{-4}$ dyn cm$^{-2}$ for our MMSN disk model (see equations \ref{eq:diskb} and \ref{eq:Pd}).
 We assume that the nebular gas is described by a realistic equation of state (EOS), as prescribed by the \citet{saumon95} EOS tables.
@@
 We thus need to extend the tables to lower $T$ and $P$, as required by our disk model.
-We choose $\log_{10} T =1.0$ and $ \log_{10}P=-4.4$ as our lower boundaries for temperature and pressure, respectively, resulting in an extended temperature and pressure grid
+We calculate $\delad$ for
 
 \begin{eqnarray}
-1.0 & < & \log_{10} T <2.1, \\ 
--4.4& < & \log_{10} P<4.0. \\
+1.0 & < & \log_{10} T <2.1 \\ 
+-4.4& < & \log_{10} P<4.0 \\
 \end{eqnarray} 
-The other thermodynamic variables in the tables are calculated as follows.
+using the following method.
 
@@
 
-For a system of particles, the partition function can be written as the product of all partition functions associated with each type of energy that the system can have:
+For a system of particles, the partition function can be written as the product of all partition functions associated with each type of internal energy:
 
@@
 
-\noindent where $Z_t$, $Z_r$, $Z_v$ are the partition functions associated with translation, rotation, and vibration, respectively \footnote{We ignore electronic and nuclear excitation as they are only important at temperatures much higher than our regime of interest}.
-Our derivations follow \citet{kittel} %For hydrogen, electronic and nuclear excitation are only significant at temperatures higher than our region of interest ($\theta_e \approx 12000$ K and $\theta_n >> \theta_e$, where $\theta_e$ and $\theta_n$ are the characteristic temperatures for electronic and nuclear excitation, respectively).
+\noindent where $Z_t$, $Z_r$, $Z_v$ are the partition functions associated with translation, rotation, and vibration, respectively.
+\footnote{We ignore electronic and nuclear excitation as they are only important at temperatures much higher than our regime of interest.} Our derivations follow \citet{kittel} %For hydrogen, electronic and nuclear excitation are only significant at temperatures higher than our region of interest ($\theta_e \approx 12000$ K and $\theta_n >> \theta_e$, where $\theta_e$ and $\theta_n$ are the characteristic temperatures for electronic and nuclear excitation, respectively).
 As such, we will only take into account the translation, rotation and vibration of the hydrogen molecule:
@@
 However, molecular hydrogen occurs in two isomeric forms: orthohydrogen, with the proton spins aligned parallel to each other, and parahydrogen, with the proton spins aligned antiparallel.
-Parahydrogen can only a have symmetric (even) wave function associated with rotation, while orthohydrogen can only have an antisymmetric (odd) wave function associated with rotation (see section \S\ref{deladtable} for an explanation why).
+Parahydrogen can only a have symmetric (even) wave function associated with rotation, while orthohydrogen can only have an antisymmetric (odd) wave function associated with rotation.
 The rotational partition functions for ortho- and parahydrogen can thus be written as:
@@
 \end{equation}
-The factor of 3 above accounts for the three-fold degeneracy of the ortho state.
+The factor of 3 in Equation (\ref{eq:Zortho}) accounts for the three-fold degeneracy of the ortho state.
 
@@
 \end{equation}
-Since $Z=Z_t Z_r Z_v$, it is easy to notice that $U=U_t+U_r+U_v$ and $S=S_t+S_r+S_v$, where $U_t$, $U_r$, $U_v$, $S_t$, $S_r$, $S_v$ are the quantities corresponding to the individual translation, rotation and vibration partition functions, respectively.
+Since $Z=Z_t Z_r Z_v$, we may write $U=U_t+U_r+U_v$ and $S=S_t+S_r+S_v$, where $U_t$, $U_r$, $U_v$, $S_t$, $S_r$, $S_v$ are the quantities corresponding to the individual translation, rotation and vibration partition functions, respectively.
 
@@
 As such, the density in this region follows the ideal gas law $P=\rho \mathcal{R} T$.
-\item{\textbf{Internal energy per mass.}} $U=U_t+U_r+U_v$, where $U_t$ is given by equation (\ref{eq:ut}), and $U_r$, $U_v$ are determined using equations (\ref{eq:u}), (\ref{eq:Zrspin}) and (\ref{eq:Zv}) above.
+\item{\textbf{Internal energy per mass.}} $U=U_t+U_r+U_v$, where $U_t$ is given by Equation(\ref{eq:ut}), and $U_r$ and $U_v$ are determined using equations (\ref{eq:u}), (\ref{eq:Zrspin}) and (\ref{eq:Zv}) above.
 \item{\textbf{Entropy per unit mass}}.
-Similarly, $S=S_t+S_r+S_v$, where $S_t$ is given by equation (\ref{eq:st}), and $S_r$, $S_v$ can be determined from equation (\ref{eq:s}) and the calculated expressions for $U_r$ and $U_v$, respectively.
+Similarly, $S=S_t+S_r+S_v$, where $S_t$ is given by Equation (\ref{eq:st}), and $S_r$ and $S_v$ can be determined from Equation (\ref{eq:s}) and the calculated expressions for $U_r$ and $U_v$, respectively.
 \item{\textbf{Entropy logarithmic derivatives}}. The logarithmic derivatives $S_T$ and $S_P$ are given by:
@@
 
-\item{\textbf{Adiabatic gradient $\delad$}. The adiabatic gradient is defined as:
+\item{\textbf{Adiabatic gradient $\delad$}}. The adiabatic gradient is defined as:
 
@@
 
-Lastly, we obtain the equation of state tables for the hydrogen-helium mixture thorough the procedure described in \citet{saumon95}, for a helium mass fraction $Y=0.3$.
+Lastly, we combine Figures \ref{fig:deladH} and \ref{fig:deladHe} to obtain the EOS tables for the hydrogen-helium mixture thorough the procedure described in \citet{saumon95}, for a helium mass fraction $Y=0.3$ (Figure \ref{fig:deladmap}).
 
@@
 
-\section{Adiabatic Gradient during Partial Dissociation}\label{deladdiss}
+\section{Adiabatic Gradient Variations}
+\label{alldelad}
+
+\subsection{Adiabatic Gradient during Partial Dissociation}\label{deladdiss}
 
@@
 \end{equation} 
-with $\chi$ the dissociation energy, equal to 4.48 eV for molecular hydrogen \citep{blanksby03}.
+where $\chi$=4.48 eV is the dissociation energy for molecular hydrogen \citep{blanksby03}.
 
 The above also holds true for ionization, with the dissociation energy replaced by ionization energy $\chi=13.6$ eV for atomic hydrogen \citep{mandl89}.
-From the Saha equation one can find an expression for $\rho$ as a function of $T$ and $x$, then derive the adiabatic gradient directly from its definition (equation \ref{eq:delad}), taking into account the fact that the mean molecular weight in the ideal gas law varies with $x$, hence the pressure will not only be a function of $T$ and $\rho$ but also of $x$ (see \citealt{kippenhahn90}, Chapter 14.3 for a detailed derivation).
+From the Saha equation one can find an expression for $\rho$ as a function of $T$ and $x$, then derive the adiabatic gradient directly from its definition (Equation \ref{eq:delad}), taking into account the fact that the mean molecular weight in the ideal gas law varies with $x$, hence the pressure will not only be a function of $T$ and $\rho$ but also of $x$ (see \citealt{kippenhahn90}, Chapter 14.3 for a detailed derivation).
 The final expression for the adiabatic gradient during ionization is
@@
 \label{eq:deladdiss}
-\delad=\frac{1+x+ \frac{x(1-x^2)}{2} \frac{\chi}{k_B T}}{5 x + \frac{7(1-x)}{2} + \frac{x(1-x^2)}{2} \Big(\frac{\chi}{k_B T}\Big)^2}.
-\end{equation} 
-We recover $\delad=2/7$ for $x=0$ (no ongoing dissociation hence hydrogen is purely molecular and diatomic) and $\delad=2/5$ for $x=1$ (hydrogen is fully dissociated into atoms and hence monatomic).
+\delad=\frac{1+x+ \frac{x(1-x^2)}{2} \frac{\chi}{k_B T}}{5 x + \frac{7(1-x)}{2} + \frac{x(1-x^2)}{2} \Big(\frac{\chi}{k_B T}\Big)^2} \, .
+\end{equation} 
+Using Equation (\ref{eq:deladdiss}), we recover $\delad=2/7$ for $x=0$ (no ongoing dissociation hence hydrogen is purely molecular and diatomic) and $\delad=2/5$ for $x=1$ (hydrogen is fully dissociated into atoms and hence monatomic).
 Figure \ref{fig:deladdiss} shows the dependence of $\delad$ on the dissociation fraction, for $T=3000$ K, the temperature at which dissociation typically occurs \citep{langmuir12}.
@@
 
+\subsection{Adiabatic Gradient during Conversion of Spin Isomers}
+\label{deladspin}
+
+The adiabatic gradient scales as $\delad \sim 1/c_{\rm V}=1/(c_{\rm V,t}+c_{\rm V,r}) \sim 1/c_{\rm V, r}$, where $c_{\rm V}$ is the specific heat capacity at constant volume, and $c_{\rm V, t}$ and $c_{\rm V,r}$ are the translational and rotational components, respectively.
+The second equality is due to the fact that $c_{\rm V, t}=3\mathcal{R}/2$ is independent of temperature .
+We can therefore understand how conversion between spin isomers affects $\delad$ by studying the dependence on temperature of $c_{\rm V, r}$ of the ortho-para mixture.
+
+The internal energy per unit mass and specific heat capacity associated with rotation for the individual isomers and for the equilibrium mixture can be derived from their partition functions (see \App{EOStables} for details), and are plotted in Figure  \ref{fig:ucvr} (after \citealt{farkas35}, Figure 1).
+At low temperatures, parahydrogen is in the $j=0$ state and has no rotational energy, while orthohydrogen is in the $j=1$ state and has the energy of its first rotational level.
+Both para- and orthohydrogen, as well as their equilibrium mixture, behave like monatomic gases at low temperatures and thus have zero rotational heat capacity.
+This is consistent with $\delad=2/5$ at low temperatures as seen in Figure \ref{fig:deladmap}.
+As the temperature increases, the energetically higher-lying rotational states of para- and orthohydrogen are populated and the heat capacity of both spin isomers increases as a result.
+We note that the heat capacity of the equilibrium mixture is not a weighted average of the heat capacities of the individual components because it takes into account both the rotational energy uptake of para- and orthohydrogen, and also the shift in their equilibrium concentrations with temperature.
+This results in a peak in the heat capacity of the mixture around $\sim$$50$ K, as seen in the bottom plot of Figure \ref{fig:ucvr}.
+As $\delad \sim 1/c_{\rm{V,r}}$, it follows that the adiabatic gradient has to reduce, reach a minimum, then increase as the temperature rises, as shown in Figure \ref{fig:deladmap}.
+
+% As the temperature increases, the energetically higher-lying ($j=1$) ortho-hydrogen is formed, and the concomitant energy increase is seen as a peak in the heat capacity of the equilibrium mixture.
+
+%There are two significant maxima in the heat capacities of parahydrogen and of the mixture.
+At very low temperatures, the heat capacity of parahydrogen is zero because only the lowest accessible energy level $j=0$ is occupied and a temperature increase does not provide enough energy to populate the next higher level.
+When the temperature becomes sufficiently high to populate the second lowest level $j=2$, the heat capacity rapidly increases, passes through a maximum and starts to decrease when the second lowest level becomes saturated.
+
+%the adiabatic gradient is inversely proportional to the heat capacity, it
+
+%means that  the former has to first decrease from 2/5 as the temperature increases, reach a minimum around 50 K ($\delad \approx 0.25$ from Fig.
+\ref{fig:deladmap}), then gradually increase to 2/7 as for a diatomic gas.
+This behavior is illustrated in Figure \ref{fig:deladmap}.
+
+%The maxima in the ortho-para mixture appears when parahydrogen starts converting into orthohydrogen.
+The heat capacity of the equilibrium mixture is not a weighted average of the heat capacities of the individual components because it takes into account both the rotational energy uptake of para- and orthohydrogen, and also the shift in their equilibrium concentrations with temperature.
+At $T=0$, only parahydrogen is present in the equilibrium mixture; as the temperature is increased, the energetically higher-lying ($j=1$) ortho-hydrogen is formed, and the concomitant energy increase is seen as a peak in the heat capacity.
+As the adiabatic gradient is inversely proportional to the heat capacity, it means that  the former has to first decrease from 2/5 as the temperature increases, reach a minimum around 50 K ($\delad \approx 0.25$ from Fig.
+\ref{fig:deladmap}), then gradually increase to 2/7 as for a diatomic gas.
+This behavior is illustrated in Figure \ref{fig:deladmap}.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/ortho_para_energy.pdf}
+%%\vspace{-0.5in}
+\caption{Internal energy per unit mass and specific heat capacity associated with rotation for parahydrogen (dashed blue), orthohydrogen (dotted green), and the equilibrium mixture (solids red) as a function of temperature.
+After \citet{farkas35}, Figure 1.}
+\label{fig:ucvr}
+\end{figure}
+
 %with $\chi$ and $T$ the dissociation energy and temperature, respectively, equal to $\chi=4.48$ eV\citep{blanksby03} and $T\sim3000$ K \citep{langmuir12} for molecular hydrogen.
@@
 For low temperatures ($T \lesssim 100$ K) at which ice grains are present, opacity scales with temperature as $\kappa \sim T^2$.
-Sublimation of ice grains at $\sim 150$ K and of metal grains at $\sim 1000$ K results in sharp opacity drops.
+Sublimation of ice grains at $\sim$$150$ K and of metal grains at $\sim$$1000$ K results in sharp opacity drops.
 This is shown in Figure \ref{fig:opacity} for a gas density $\rho=10^{-8}$ g cm$^{-3}$, which is typical for the outer regions of protoplanetary disks.
@@
 As shown in Figure \ref{fig:opacity}, their results are in good agreement with \citet{bell94}.
-However, \citet{semenov03} do not take grain growth into account, which is likely to occur in protoplanetary disks, particularly when most solids have been accreted.
+However, \citet{semenov03} do not take grain growth into account, which is likely to occur in protoplanetary disks, particularly at the late times when cores form.
 \citet{dalessio01} compute wavelength dependent opacities for a range of maximum particle sizes and different size distributions.
@@
 radiative windows.
-We thus use the \citet{bell94} opacities for $T \gtrsim 1000$ K, ensuring they smoothly match the \citet{dalessio01} opacities for lower temperatures.
+We thus use the \citet{bell94} opacities for $T \gtrsim 1000$ K, ensuring that they smoothly match the \citet{dalessio01} opacities for lower temperatures.
 
@@
 The solid black curve shows the tabulated opacity of \citet{semenov03} for a dust composition of 'normal' silicates.
-The dashed red curve shows the \citet{dalessio01} opacity, which takes grain growth into account, for a maximum particle size of 1 cm and a normal collisional cascade grain size distribution ($p=3.5$).
+The dashed red curve shows the \citet{dalessio01} opacity, which takes grain growth into account, for a maximum particle size of 1 cm and a standard collisional cascade grain size distribution ($p=3.5$).
 The solid red curve is the same as the dashed red curve, but it accounts for coagulation ($p=2.5$).}
@@
 The significant opacity drop due to the sublimation of ice and metal grains lowers the radiative temperature gradient $\delrad$, which may result in one or more inner radiative layers inside the atmosphere of a protoplanet.
-This is displayed in Figure \ref{fig:delvsr}: depending on the semi-major axis and core mass, the opacity drop will generate no radiative window (top panel), one radiative window (middle panel), or two radiative windows (bottom panel).
+This is displayed in Figure \ref{fig:delvsr}: depending on the semimajor axis and core mass, the opacity drop will generate no radiative window (top panel), one radiative window (middle panel), or two radiative windows (bottom panel).
 
@@
 \centering
-\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/Opacity/del_vs_r.pdf}
+\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/del_vs_r.pdf}
 %%\vspace{-0.5in}
 \caption{Snapshots of the radiative and adiabatic gradient as a function of the radial coordinate, for planets with different core masses forming at various locations in the disk.
+The nebular gas is described by a realistic EOS.
 The sharp drop in opacity due to dust sublimation may generate one or more radiative windows.
 Top panel: no radiative window for $a=100$ AU and $M\co=3 M_{\oplus}$.
-Middle panel: the opacity drop produces one radiative window for $a=50$ AU and $M\co=5 M_{\oplus}$.
-Bottom panel: the sharp drop in opacity results in two radiative windows for $a=20$ AU and $M\co=3 M_{\oplus}$.}
+Middle panel: the sharp opacity decrease produces one radiative window for $a=50$ AU and $M\co=5 M_{\oplus}$.
+Bottom panel: the decrease in opacity results in two radiative windows for $a=20$ AU and $M\co=3 M_{\oplus}$.}
 \label{fig:delvsr}

### 2013-11-20 tex/draft_paper2/draft_paper2.tex
section 4 comments

@@
 \usepackage{graphicx} 
-\usepackage{graphics}                       
 \usepackage{amsmath}
@@
 \usepackage{subeqnarray}
+\usepackage{ulem}
 %\bibliographystyle{apj}
@@
 
-\title{Quantitative Estimates of Minimum Core Masses for Giant Planet Formation}
+\title{Quantitative Estimates of Minimum Core Masses for Giant Planet Formation
+{\bf Paper I was quantitative.}}
 \author{Ana-Maria A. Piso}
@@
 Instead, we choose as boundary a constant entropy curve ($\log(S)=8.4$) above the region in which the \citet{saumon95} model fails.
-The expressions derived in \App{EOStables} are sufficient to give good results for the colored regions of the extended map, which fully cover the temperature and pressures ranges required by our models.}
+The expressions derived in \App{EOStables} are sufficient to give good results for the colored regions of the extended map, which fully cover the temperature and pressures ranges required by our models.
+{\bf This caption is extremely long.
+I would find a way to defer much of this to text.}}
 \label{fig:deladmap}
@@
 
-\section{Equation of State Effects on Atmosphere Evolution}
+\section{Equation of State Effects on Atmosphere Evolution \\
+{\bf long section title.  Just "Role of EOS"?}}
 \label{EOSeffects}
 
-Variations in the adiabatic gradient $\delad$ due to partial dissociation and quantum rotational effects (see \S\ref{deladtable}) have two competing effects on the evolution of a core's atmosphere.
+Variations in the adiabatic gradient $\delad$ due to partial dissociation and quantum {\bf (`Quantum' is not a useful descriptor (IMO).
+All EOS effects are quantum, noting molecular or H$_2$ rotation would be more descriptive.)} rotational effects (see \S\ref{deladtable}) have two competing {\bf (Not `competing,' both contribute and can/do go in same direction)} effects on \sout{the} {\bf atmospheric} evolution \sout{of a core's atmosphere}.
+{\bf Putting the equation $\dot{M} = -L/(dE/dM)$ early in the explanation would help.}
 They yield: (1) a lower envelope luminosity, and (2) a larger amount of radiated energy per unit of accreted mass, when compared to an ideal gas.
-Both effects combined slow down accretion and increase the growth time of the atmosphere and thus the critical core mass.
-
-In order to explain how quantum rotational states at low temperatures affect atmospheric growth, we perform the following thought experiment: we imagine that the EOS deviates from a polytrope only in the upper atmosphere where temperatures are low, and that the gas is ideal and polytropic deep in the envelope.
+Both effects {\bf A bit confusing if you mean EOS effects or resulting evolution effects.
+Maybe avoid reusing `effects`.)} combined slow down accretion and increase the growth time of the atmosphere and thus the critical core mass.
+
+{\bf In what follows it may be clearer to start with the motivation: isolating the two EOS effects, i.e.
+dissociation and rotation.
+Then introduce the two patched EOSs as how we do this.
+Currently it's less clear and more drawn out where this is going.}
+
+In order to explain how quantum rotational states at low temperatures affect atmospheric growth, we perform the following thought experiment {\bf (I don't see how this is a thought experiment, or why we have to imagine.
+I would just describe the EOSs)}: we imagine that the EOS deviates from a polytrope only in the upper atmosphere where temperatures are low, and that the gas is ideal and polytropic deep in the envelope.
 Conversely, we study the effects of dissociation at high temperatures by assuming that the EOS is realistic at the bottom of the atmosphere where temperatures are high, but that the gas is polytropic in the outer regions.
-Figure \ref{fig:tplotall} shows the time evolution of atmospheres forming at 10 AU around cores of mass $M\co=10 M_{\oplus}$ and described by various equations of state, as follows:
-
+{{\bf Finish the description of the EOSs first.
+Then describe and interpret Fig.
+2 starting in a new paragraph.
+Starting the fig description at the end of a methods discussion lost me.)}  Figure \ref{fig:tplotall} shows the time evolution of atmospheres forming at 10 AU around cores of mass $M\co=10 M_{\oplus}$ and described by various equations of state, as follows:
+
+{\bf I would make this list numbered and omit the reference to the curve styles, which is clear in figures.}
 \begin{itemize}
@@
 \item Dash-dotted curve (3) --- ideal gas polytrope with $\delad=0.3$ for $T<500$ and realistic EOS for $T>500$ K.
-\item Solid curve (4) --- fully realistic EOS.
+\item Solid curve (4) --- fully realistic EOS. ({\bf Here and elsewhere, I don't like ``fully" realistic, as it implies a perfect EOS.  I would just say realistic at all $T$.})
 \end{itemize}
 We choose $T=500$ K as the cutoff temperature because the hydrogen-helium mixture behaves like an ideal gas, with $\delad=0.3$, in this temperature regime (see Figure \ref{fig:deladmap}).
-As such, the time evolution difference between curves (1) and (2) shows the quantum rotational effects on the atmosphere evolution, while the difference between curves (1) and (3) accounts for hydrogen dissociation.
-We see that both dissociation and rotational states have a comparable effect on the atmosphere growth and result in slower cooling.
+{\bf (I find the references to curve numbers distracting in general.
+I would try to do less of this and focus on describing the physical effects.
+The reader still needs to look at the figure, and there the curves are clearly labelled.)} \sout{As such}, the time evolution difference between curves (1) and (2) shows the quantum rotational effects on the atmosphere evolution, while the difference between curves (1) and (3) accounts for hydrogen dissociation.
+We see that both dissociation and rotational states have a comparable effect {\bf (due to log scale, I would not call the effects comparable, just b/c they go in same direction)} on the atmosphere growth and result in slower cooling.
 The growth time is dependent on both the total energy that must be released (i.e., the rate at which energy is released) and on the luminosity of the atmosphere.
-We explore the relative influence of these two factors separately.
+{\bf `Rate of energy release' and luminosity are the same thing.
+You may be calling $dE/dM$ a `rate' which is confusing.
+Again, I think if you introduce and physically explain $\dot{M} = -L/(dE/dM)$ at the outset, then the resulting explanations can more compactly and clearly refer to the terms that matter.}  We explore the \sout{relative} influence of these two factors separately.
+{\bf Haven't you already started exploring these factors separately?
+Why say this again?
+Also perhaps worth a note (if not an explanation) that the combined effect is significantly greater than either individually (again noting log scale).}
 
@@
 %%\vspace{-0.5in}
-\caption{We explore atmosphere growth around a core of $M\co=10 M_{\oplus}$ forming at 10 AU in our fiducial disk, for the EOS choices in Figure \ref{fig:tplotall}.
+\caption{{\bf (Left and right panels could be different figs.
+The left panel could benefit from showing $P$ or $\rho$.
+Also $E_{\rm tot}$ is not yet well defined; the local quantity $e \rho r^3$ might be more instructive.
+A greater range of $E$ would show what the polytrope is doing better.
+The right panel could benefit from adding $\dot{M}$.
+We call $E_{\rm tot} \rightarrow E$ elsewhere, so should be consistent.
+Why don't x-axes line up in right panel, are masses different?
+What is the discontinuity in $dE/dM$ for the polytrope around 14 $M_\oplus$?
+(A kink would be course resolution, but a discontinuity is harder to understand.)  I don't think you need an abs.\ val.
+around $-dE/dM$.)} We explore atmosphere growth around a core of $M\co=10 M_{\oplus}$ forming at 10 AU in our fiducial disk, for the EOS choices in Figure \ref{fig:tplotall}.
 Upper-left panel:  Instantaneous temperature profile as a function of the radial coordinate when the total mass (core + atmosphere) is $12 M_{\oplus}$.
@@
 
-The upper-right panel of Figure \ref{fig:all_plot} shows the luminosity evolution with mass for atmospheres (1)-(4), for an example core with $M\co=10 M_{\oplus}$ forming at 10 AU.
+The upper-right panel of Figure \ref{fig:all_plot} shows the luminosity evolution with mass for \sout{atmospheres (1)-(4)} {\bf our model EOSs}, for a\sout{n example core with} $M\co=10 M_{\oplus}$ {\bf core} \sout{forming} {\bf (as core isn't forming, this phrase is misleading in addition to being too lengthy)} at 10 AU.
 We see that the lower luminosity of the fully realistic EOS (4) when compared to the polytrope (1) is due to the realistic EOS in the outer atmosphere (2).
@@
 We see that the realistic EOS for low temperatures (2) results in a deeper location of the RCB when compared to the ideal gas polytrope (1).
-This deeper RCB translates into a larger number of steps that photons need to diffuse to escape from the RCB, and therefore a lower luminosity (cf.
+{\bf This is an accurate observation.
+The deeper question is why does the EOS change result in a deeper RCB?
+This should be possible to understand, right?}  This deeper RCB translates into a larger number of steps that photons need to diffuse to escape from the RCB, and therefore a lower luminosity (cf.
 Equation \ref{eq:delrad}).
+{\bf Better to just say it's an optical depth effect than reexplaining how radiative diffusion works.
+Also it's the pressure depth (not the radius) that correlates more closely with optical depth.
+A plot of $P(r)$ might thus help as noted in fig caption.}
 
@@
 Dissociation deep in the planet's atmosphere increases the amount of energy per unit mass that needs to be radiated by the envelope, which further slows down growth. 
-Figure \ref{fig:all_plot}, lower-left plot, shows the total energy (internal and gravitational) profile as a function of the radial coordinate, for the same example planet.
-The bulk of the energy is concentrated deep in the atmosphere for the realistic EOS at high temperatures (3) and the fully realistic EOS curve (4), which shows that hydrogen dissociation in the inner atmosphere dictates the energy behavior of the envelope.
+{\bf Why is this true?
+Dissociation absorbs energy which by itself decreases the amount of energy that needs to be radiated.}  Figure \ref{fig:all_plot}, lower-left plot, shows the total energy (internal and gravitational) profile as a function of the radial coordinate, \sout{for the same example planet}.
+{\bf Two choices: (1) define this cumulative energy more clearly and give it a different symbol than $E_{\rm tot}$ that's already used, maybe $E(r)$, or (2) plot the local (noncumulative) quantity $e \rho r^3$, which should peak in the interesting places.} The bulk of the energy is concentrated deep in the atmosphere for the realistic EOS at high temperatures (3) and the fully realistic EOS curve (4), which shows that hydrogen dissociation in the inner atmosphere dictates the energy behavior of the envelope.
 In contrast, energy is concentrated towards the outer boundary for the $\delad=0.3$ polytrope (1).
-Qualitatively, this can be explained through a simple analytic argument: the density profile in an adiabatic, non-self-gravitating atmosphere composed of an ideal gas scales as $\rho(r) \sim r^{1/\delad-1}$ (see Paper I).
-Since the total energy scales as $E(r) \sim \rho(r) r^2 \sim r^{1/\delad+1}$, adiabats with lower $\delad$ have their energy more tightly packed towards the interior of the envelope, which is the case during dissociation when $\delad$ decreases significantly (see Section \S\ref{deladtable} and Figure \ref{fig:deladmap}).
-It takes more energy to bring gas deep into the atmosphere for an envelope that has the bulk of its energy concentrated towards the bottom,  which increases the amount of energy per unit mass that needs to be radiated, i.e.
+{\bf The key question here is how does the slope of the energy affect the quantity that matters, $|dE/dM|$?
+This is not clear yet.
+I think the shift to lower temperatures, seen in the $T(r)$ plot is helpful, but perhaps the EOS plots should also include an internal energy map.}
+Qualitatively, this can be explained through a simple analytic argument: the density profile in an adiabatic, non-self-gravitating atmosphere composed of an ideal gas scales as $\rho(r) \sim r^{1/\delad-1}$ (see Paper I) {\bf (The sign is wrong here and the error propagates to the energy.
+Should be $\propto$ not $\sim$.)}.
+Since the total energy scales as $E(r) \sim \rho(r) r^2 \sim r^{1/\delad+1}$, {\bf (This is a new/contradictory use of $E$, what is given here is $dE/d\ln(r)$.
+Scaling is wrong as noted above.
+Again $\propto$ not $\sim$.)} adiabats with lower $\delad$ have their energy more tightly packed towards the interior of the envelope, which is the case during dissociation when $\delad$ decreases significantly (see Section \S\ref{deladtable} and Figure \ref{fig:deladmap}).
+It takes more energy to bring gas deep into the atmosphere for an envelope that has the bulk of its energy concentrated towards the bottom,  {\bf (This argument isn't clear, new gas isn't brought deep into the atmosphere, it is added at the surface.)} which increases the amount of energy per unit mass that needs to be radiated, i.e.
 $|dE/dM|$.
-The cooling time is therefore larger for the realistic EOS at high temperatures (3) when compared to the ideal gas polytrope (1).
+The cooling time is therefore larger {\bf with dissociation} \sout{for the realistic EOS at high temperatures} (3) when compared to the ideal gas polytrope (1).
 This increase in $|dE/dM|$ is shown in the bottom-right panel of Figure \ref{fig:all_plot}.
 
-We have seen that quantum rotational states transitions due to the low temperatures at the top of the atmosphere increase the thickness of the radiative zone, and decrease the luminosity of the envelope, while dissociation at high temperatures deep in the atmosphere increases the amount of energy needed to increase the atmosphere mass by a fixed amount.
-Overall, both effects result in a longer time for the atmosphere to evolve.
+We have seen that quantum rotational states transitions due to the low temperatures at the top of the atmosphere {\bf (already too long and convoluted)} increase the thickness of the radiative zone, and decrease the luminosity of the envelope, while dissociation at high temperatures deep in the atmosphere increases the amount of energy needed to increase the atmosphere mass by a fixed amount.
+{\bf Way too long a sentence!} Overall, both effects result in a longer time for the atmosphere to evolve.
 
@@
 We also thank Eugene Chiang and Steven Cranmer for useful conversations.
-ANY thanks Phil Arras for his help with entropy units conventions in the equation of state tables.}
+ANY thanks Phil Arras for a methodical explanation of entropy.}
 

### 2013-11-22 tex/draft_paper2/draft_paper2.tex
updated section 4 tex file

--- 2013-11-13 tex/draft_paper2/draft_paper2.tex
@@
 
-\section{Equation of State Effects on Atmosphere Evolution}
+\section{Role of Equation of State}
 \label{EOSeffects}
 
-Variations in the adiabatic gradient $\delad$ due to partial dissociation and quantum rotational effects (see \S\ref{deladtable}) have two competing effects on the evolution of a core's atmosphere.
-They yield: (1) a lower envelope luminosity, and (2) a larger amount of radiated energy per unit of accreted mass, when compared to an ideal gas.
-Both effects combined slow down accretion and increase the growth time of the atmosphere and thus the critical core mass.
-
-In order to explain how quantum rotational states at low temperatures affect atmospheric growth, we perform the following thought experiment: we imagine that the EOS deviates from a polytrope only in the upper atmosphere where temperatures are low, and that the gas is ideal and polytropic deep in the envelope.
+Variations in the adiabatic gradient $\delad$ due to partial dissociation and molecular rotational effects (see \S\ref{deladtable}) affect atmospheric evolution by yielding: (1) a lower envelope luminosity $L$, and (2) a larger amount of radiated energy per unit of accreted mass $-dE/dM$, when compared to an ideal gas.
+As a result, the rate of change in atmospheric mass,
+
+\begin{equation}
+\label{eq:dMdt}
+\dot{M} = -\frac{L}{dE/dM},
+\end{equation}
+(cf.
+Equation \ref{eq:coolingglobal} and ignoring surface terms) is lower than in the ideal case.
+The $\delad$ variations thus slow down gas accretion and increase the growth time of the atmosphere, and hence the critical core mass.
+
+To understand the separate effects of dissociation and rotation on atmospheric growth, we consider the resulting EOS deviations from an ideal gas independently.  
+%{\bf In what follows it may be clearer to start with the motivation: isolating the two EOS effects, i.e.
+dissociation and rotation.
+Then introduce the two patched EOSs as how we do this.
+Currently it's less clear and more drawn out where this is going.}
+In order to explain how quantum rotational states at low temperatures affect atmospheric growth, we 
+% {\bf (I don't see how this is a thought experiment, or why we have to imagine.  I would just describe the EOSs)}: 
+imagine \textbf{(I used 'imagine' to make it clear that this is just an exercise to isolate the EOS effects, rather than a physically motivated assumption)} that the EOS deviates from a polytrope only in the upper atmosphere where temperatures are low, and that the gas is ideal and polytropic deep in the envelope.
 Conversely, we study the effects of dissociation at high temperatures by assuming that the EOS is realistic at the bottom of the atmosphere where temperatures are high, but that the gas is polytropic in the outer regions.
+We compare the behavior of each EOS described above with that of an ideal gas polytrope.
+Lastly, we use the realistic EOS at all temperatures to study the combined effects of dissociation and rotation.
+
+%{{\bf Finish the description of the EOSs first.  Then describe and interpret Fig. 2 starting in a new paragraph.  Starting the fig description at the end of a methods discussion lost me.)}
+
 Figure \ref{fig:tplotall} shows the time evolution of atmospheres forming at 10 AU around cores of mass $M\co=10 M_{\oplus}$ and described by various equations of state, as follows:
-
-\begin{itemize}
-\item Dashed curve (1) --- ideal gas polytrope with $\delad=0.3$.
-\item Dotted curve (2) --- ideal gas polytrope with $\delad=0.3$ for $T>500$ K and realistic EOS for $T<500$ K.
-\item Dash-dotted curve (3) --- ideal gas polytrope with $\delad=0.3$ for $T<500$ and realistic EOS for $T>500$ K.
-\item Solid curve (4) --- fully realistic EOS.
-\end{itemize}
+%{\bf I would make this list numbered and omit the reference to the curve styles, which is clear in figures.}
+\begin{enumerate}
+\item Ideal gas polytrope with $\delad=0.3$.
+\item Ideal gas polytrope with $\delad=0.3$ for $T>500$ K and realistic EOS for $T<500$ K (hereafter low-$T$ realistic EOS).
+\item Ideal gas polytrope with $\delad=0.3$ for $T<500$ and realistic EOS for $T>500$ K  (hereafter high-$T$ realistic EOS).
+\item Realistic EOS at all $T$. 
+%({\bf Here and elsewhere, I don't like ``fully" realistic, as it implies a perfect EOS.  I would just say realistic at all $T$.})
+\end{enumerate}
 We choose $T=500$ K as the cutoff temperature because the hydrogen-helium mixture behaves like an ideal gas, with $\delad=0.3$, in this temperature regime (see Figure \ref{fig:deladmap}). 
-As such, the time evolution difference between curves (1) and (2) shows the quantum rotational effects on the atmosphere evolution, while the difference between curves (1) and (3) accounts for hydrogen dissociation.
-We see that both dissociation and rotational states have a comparable effect on the atmosphere growth and result in slower cooling.
-The growth time is dependent on both the total energy that must be released (i.e., the rate at which energy is released) and on the luminosity of the atmosphere.
-We explore the relative influence of these two factors separately.
+%{\bf (I find the references to curve numbers distracting in general.
+I would try to do less of this and focus on describing the physical effects.
+The reader still needs to look at the figure, and there the curves are clearly labelled.)} \sout{As such},
+%By separately comparing curves (1) and (2), and (1) and (3)  while the difference between curves (1) and (3) accounts for hydrogen dissociation. 
+Both dissociation and rotation result in slower cooling for both the high-$T$ and low-$T$ realistic EOSs. 
+%decrease the atmospheric accretion rate $\dot{M}$ and thus
+%This follows from Equation (\ref{eq:dMdt}), with rotation decreasing $L$ and dissociation increasing $-dE/dM$, as we show further. 
+Noting the logarithmic scale in Figure \ref{fig:tplotall}, we see that the combined effect of rotation and dissociation is significantly greater than either individually.
+%consistent with Equation \ref{eq:dMdt}, since dissociation increases $-dE/dM$ while molecular rotation decreases $L$.
+
+ %From Equation \ref{eq:dMdt}, the atmospheric growth time is dependent on
+
+%The growth time is dependent on both the total energy that must be released (i.e., the rate at which energy is released) and on the luminosity of the atmosphere.
+{\bf `Rate of energy release' and luminosity are the same thing.
+You may be calling $dE/dM$ a `rate' which is confusing.
+Again, I think if you introduce and physically explain $\dot{M} = -L/(dE/dM)$ at the outset, then the resulting explanations can more compactly and clearly refer to the terms that matter.}  We explore the \sout{relative} influence of these two factors separately.
+{\bf Haven't you already started exploring these factors separately?
+Why say this again?
+Also perhaps worth a note (if not an explanation) that the combined effect is significantly greater than either individually (again noting log scale).}
 
@@
 \caption{Elapsed time to grow a planet of total mass (core + atmosphere) for a variety of EOS combinations (see text), for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=10 M_{\oplus}$.
-Both hydrogen dissociation at high temperatures deep in the atmosphere and quantum rotational effects at low temperatures in the outer envelope result in slower cooling when compared to an ideal gas polytrope.}
+Both hydrogen dissociation at high temperatures deep in the atmosphere and molecular rotational effects at low temperatures in the outer envelope result in slower cooling when compared to an ideal gas polytrope.}
 \label{fig:tplotall}
@@
 \centering
-\includegraphics[width=\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/all_plot.pdf}
+\includegraphics[width=\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/all_plot_test.pdf}
 %%\vspace{-0.5in}
-\caption{We explore atmosphere growth around a core of $M\co=10 M_{\oplus}$ forming at 10 AU in our fiducial disk, for the EOS choices in Figure \ref{fig:tplotall}.
-Upper-left panel:  Instantaneous temperature profile as a function of the radial coordinate when the total mass (core + atmosphere) is $12 M_{\oplus}$.
+\caption{
+%{\bf (Left and right panels could be different figs.
+The left panel could benefit from showing $P$ or $\rho$.
+Also $E_{\rm tot}$ is not yet well defined; the local quantity $e \rho r^3$ might be more instructive.
+A greater range of $E$ would show what the polytrope is doing better.
+The right panel could benefit from adding $\dot{M}$.
+We call $E_{\rm tot} \rightarrow E$ elsewhere, so should be consistent.
+Why don't x-axes line up in right panel, are masses different?
+What is the discontinuity in $dE/dM$ for the polytrope around 14 $M_\oplus$?
+(A kink would be course resolution, but a discontinuity is harder to understand.)  I don't think you need an abs.\ val.
+around $-dE/dM$.)}
+We explore atmosphere growth around a core of $M\co=10 M_{\oplus}$ forming at 10 AU in our fiducial disk, for the EOS choices in Figure \ref{fig:tplotall}.
+Upper-left panel: Radial pressure profile for a total mass (core + atmosphere) of $12 M_{\oplus}$.
 The location of the RCB is marked.
@@
 Upper-right panel:  Luminosity evolution as a function of total mass.
-Quantum rotational effects at low temperatures near the top of the atmosphere result in a lower luminosity for the realistic EOS (see text).
-Lower-left panel: Instantaneous energy profiles as a function on radius for a total mass of $12 M_{\oplus}$.
-Hydrogen dissociation deep in the atmosphere causes the bulk of the energy to be concentrated at the bottom of the envelope.
-This increases the amount of energy per unit mass that needs to be radiated away ($|dE/dM|$) resulting in a longer crossover time for the realistic EOS when compared to the polytrope.
-Lower-right panel:  $|dE/dM|$ evolution as a function of total mass.
-Hydrogen dissociation deep in the atmosphere results in the bulk of the energy being concentrated at the bottom of the envelope, which increases the energy per unit mass that has to be radiated away when compared to an ideal gas polytrope.
-This results in slower growth.}
+Rotational effects at low temperatures near the top of the atmosphere result in a lower luminosity for the realistic EOS (see text).
+Middle-left panel: Total local energy $e \rho r^3$ radial profile for a total mass of $12 M_{\oplus}$.
+Hydrogen dissociation deep in the atmosphere causes more of the energy to be concentrated at the bottom of the envelope when compared to an ideal gas polytrope.
+This increases the amount of energy per unit mass that needs to be radiated away, $-dE/dM$.
+Middle-right panel:  $-dE/dM$ evolution as a function of total mass.
+Hydrogen dissociation deep in the atmosphere results in a more negative total energy $E$.
+Consequently, $-dE/dM$ is larger when compared to the ideal gas polytrope.
+Lower-left panel:  Radial temperature profile for a total mass of $12 M_{\oplus}$.
+Hydrogen dissociation deep in the atmosphere decreases the temperature at the core.
+This lowers the internal energy and thus results in a more negative total energy $E$.
+Lower-right panel: the atmospheric accretion rate $\dot{M}$ as a function of total mass.
+Both dissociation and rotation reduce $\dot{M}$ and thus slow down atmospheric growth.}
 \label{fig:all_plot}
@@
 
-The upper-right panel of Figure \ref{fig:all_plot} shows the luminosity evolution with mass for atmospheres (1)-(4), for an example core with $M\co=10 M_{\oplus}$ forming at 10 AU.
-We see that the lower luminosity of the fully realistic EOS (4) when compared to the polytrope (1) is due to the realistic EOS in the outer atmosphere (2).
-This behavior is directly correlated to the depth of the radiative region of the envelope, which is shown in the upper-left panel of Figure \ref{fig:all_plot} for a total planet mass of $12 M_{\oplus}$.
+The upper-right panel of Figure \ref{fig:all_plot} shows the luminosity evolution with mass for our model EOSs, for a $M\co=10 M_{\oplus}$ core at 10 AU.
+The realistic EOS effects in the outer envelope decrease the atmospheric luminosity $L$ when compared to an ideal gas polytrope.
+We can explain this analytically.
+By substituting Equation (\ref{eq:structa}) into Equation (\ref{eq:structc}) in the convective region of the atmosphere, and assuming the nebular gas obeys the ideal gas law, we can write hydrostatic balance along an adiabat as
+
+\begin{equation}
+\label{eq:dTdr}
+\frac{d T}{d r}=-\frac{\delad G m(r)}{\mathcal{R} r^2},
+\end{equation}
+where $\mathcal{R}$ is the specific gas constant.
+This result is valid for the $\delad=0.3$ polytrope.
+As the low$-T$ realistic EOS only slightly (see Figure \ref{fig:deladmap}) deviates from an ideal gas in the outermost convective layers, we can assume that the gas described by the low-$T$ EOS can be approximated as ideal throughout the convective interior.
+The low-$T$ EOS thus also obeys Equation (\ref{eq:dTdr}) with $\delad \approx const.$.
+We further neglect self-gravity in Equation (\ref{eq:dTdr}) and set $m(r)=M\co$.
+Integrating Equation (\ref{eq:dTdr}) from the core to the RCB under these assumptions yields
+\begin{equation}
+\label{eq:dTdrint}
+T\cb=T\co - \frac{\delad G M\co}{\mathcal{R}} \Big(\frac{1}{R\co} - \frac{1}{R\cb}\Big).
+\end{equation}  
+Deep in the convective interior, $T \gg T\cb$ and $r \ll R\cb$. From virial equilibrium, the temperature in the inner atmosphere can thus be expressed as
+\begin{equation}
+\label{eq:virialT}
+T(r) \simeq \frac{\nabla_{\rm ad} G M\co}{\mathcal{R} r}.
+\end{equation}
+%The temperature scale height deep in the convective interior is $H_T \sim r$, with $dT/dr \sim -T/H_T \sim -T/r$.
+From Equation (\ref{eq:structc}) it immediately follows that the pressure scale height, for which $dP/dr \sim -P/H_P$, is $H_P \sim r \delad$.
+Deep in the interior and for $m(r)=M\co$, hydrostatic balance (\ref{eq:structa}) thus becomes
+
+%\begin{equation}
+%\label{eq:structint}
+%\frac{P}{r \delad} \sim -\frac{G M\co}{r^2}. 
+%\end{equation}
+%From this and the definition of the isothermal sound speed
+%\begin{equation}
+%\label{eq:cs}
+%c_s^2 = \frac{P}{\rho} =\mathcal{R} T, 
+%\end{equation} 
+%with both expressions applied at the core where $r=R\co$, $T=T\co$ and $\delad=\nabla_{\rm ad, c}$, we find
+Applying equation (\ref{eq:virialT}) at the core, where $T=T\co$, $r=R\co$ and $\delad=\nabla_{\rm ad, c}$, gives
+\begin{equation}
+\label{eq:Tc}
+T\co \simeq \frac{\nabla_{\rm ad, c} G M\co}{\mathcal{R} R\co}.
+\end{equation}
+As the adiabatic gradient at the core is the same for both the ideal gas polytrope and the low$-T$ realistic EOS, $T\co$ has the same value for both EOSs.
+This is shown in Figure \ref{fig:all_plot}, lower-left panel.
+Substituting Equation (\ref{eq:Tc}) into Equation (\ref{eq:dTdrint}) applied at the RCB gives
+\begin{equation}
+\label{eq:Tcb}
+T\cb \simeq \frac{\nabla_{\rm ad, RCB} G M\co}{\mathcal{R} R\cb},
+\end{equation}
+%We note that the expression above and the definition of the pressure scale height $H_P \equiv c_s^2 r^2/(G M\co)$ imply that the pressure scale height at the RCB is the same as in the deep interior, $H_P \sim \delad r$.
+and thus
+\begin{equation}
+\label{eq:rcb}
+R\cb \propto \frac{\nabla_{\rm ad, RCB}}{T\cb}.
+\end{equation}
+An approximate expression for $T\cb$ as a function of the adiabatic gradient and disk temperature can be derived using Equation (\ref{eq:delrad}) applied at the RCB (where $\delad=\delrad$) and $\delrad=d \ln T/ d \ln P$ (see Paper I for details).
+For $\beta=2$ in the opacity law (\ref{eq:opacitylaw}), $T\cb \simeq T\di (1- 2 \nabla_{\rm ad, RCB})^{-1/2}$, and thus
+\begin{equation}
+\label{eq:rcbdel}
+R\cb \propto \frac{\nabla_{\rm ad, RCB}}{(1- 2 \nabla_{\rm ad, RCB})^{-1/2}}.
+\end{equation}
+For the ideal gas polytrope $\nabla_{\rm ad, RCB}=0.3$, and for the low-$T$ realistic EOS $\nabla_{\rm ad, RCB}\approx 0.25$.
+For these particular choices, Equation (\ref{eq:rcbdel}) yields a smaller $R\cb$ for the low$-T$ realistic EOS.
+As such, molecular rotation increases the physical extent, and thus the optical depth of the radiative zone.
+%Qualitatively, this can be explained through a simple analytic argument: for a non-self-gravitating atmosphere composed of an ideal gas, the temperature profile in convective regions scales as $T(r) \propto \delad/r$ (see Paper I for details; it can also be derived from Equations \ref{eq:structa} and \ref{eq:structc} applied in convective regions).
+Applying the expression above at the RCB yields $T\cb \propto \delad\cb/ R\cb$.
+As the outer radiative layer of the atmosphere is nearly isothermal, $T\cb \approx T\di=const.$, to order unity corrections.
+It easily follows that $R\cb \propto \deladcb$   Figure \ref{fig:all_plot_temp}, middle-left panel shows
+This is directly correlated to a larger pressure depth of the RCB for the low-$T$ realistic EOS, as shown in the upper-left panel of Figure \ref{fig:all_plot} for a total planet mass of $12 M_{\oplus}$.
+As $L \propto 1/P\cb$ (cf.
+Equation \ref{eq:delrad}), the deeper RCB results in a lower $L$ and thus $\dot{M}$ (Figure \ref{fig:all_plot}, lower-right panel).
+
+%directly correlated to the depth of the radiative region of the envelope, which is shown in the upper-left panel of Figure \ref{fig:all_plot} for a total planet mass of $12 M_{\oplus}$.
 We see that the realistic EOS for low temperatures (2) results in a deeper location of the RCB when compared to the ideal gas polytrope (1).
-This deeper RCB translates into a larger number of steps that photons need to diffuse to escape from the RCB, and therefore a lower luminosity (cf.
+{\bf This is an accurate observation.
+%The deeper question is why does the EOS change result in a deeper RCB?
+This should be possible to understand, right?}  This deeper RCB translates into a larger number of steps that photons need to diffuse to escape from the RCB, and therefore a lower luminosity (cf.
 Equation \ref{eq:delrad}).
+{\bf Better to just say it's an optical depth effect than reexplaining how radiative diffusion works.
+Also it's the pressure depth (not the radius) that correlates more closely with optical depth.
+A plot of $P(r)$ might thus help as noted in fig caption.}
+
+%We see that the lower luminosity of the fully realistic EOS (4) when compared to the polytrope (1) is due to the realistic EOS in the outer atmosphere (2). T
 
@@
 
-Dissociation deep in the planet's atmosphere increases the amount of energy per unit mass that needs to be radiated by the envelope, which further slows down growth.
-Figure \ref{fig:all_plot}, lower-left plot, shows the total energy (internal and gravitational) profile as a function of the radial coordinate, for the same example planet.
-The bulk of the energy is concentrated deep in the atmosphere for the realistic EOS at high temperatures (3) and the fully realistic EOS curve (4), which shows that hydrogen dissociation in the inner atmosphere dictates the energy behavior of the envelope.
-In contrast, energy is concentrated towards the outer boundary for the $\delad=0.3$ polytrope (1).
-Qualitatively, this can be explained through a simple analytic argument: the density profile in an adiabatic, non-self-gravitating atmosphere composed of an ideal gas scales as $\rho(r) \sim r^{1/\delad-1}$ (see Paper I).
-Since the total energy scales as $E(r) \sim \rho(r) r^2 \sim r^{1/\delad+1}$, adiabats with lower $\delad$ have their energy more tightly packed towards the interior of the envelope, which is the case during dissociation when $\delad$ decreases significantly (see Section \S\ref{deladtable} and Figure \ref{fig:deladmap}).
-It takes more energy to bring gas deep into the atmosphere for an envelope that has the bulk of its energy concentrated towards the bottom,  which increases the amount of energy per unit mass that needs to be radiated, i.e.
+%Dissociation deep in the planet's atmosphere increases the amount of energy per unit mass that needs to be radiated by the envelope, which further slows down growth.
+
+%{\bf Why is this true?  Dissociation absorbs energy which by itself decreases the amount of energy that needs to be radiated.}
+
+Figure \ref{fig:all_plot}, middle-left panel, shows a radial profile of the total local energy $-e \rho r^3$, with $e$ the total specific energy, for the same example planet.
+\textbf{(the core + total mass choices need to be specified; "the same example planet" is shorter than "for a planet with Mc = ...", and clear enough I think.)}
+%the cumulative total energy (internal and gravitational) profile as a function of the radial coordinate, \sout{for the same example planet}.
+{\bf Two choices: (1) define this cumulative energy more clearly and give it a different symbol than $E_{\rm tot}$ that's already used, maybe $E(r)$, or (2) plot the local (noncumulative) quantity $e \rho r^3$, which should peak in the interesting places.}
+The bulk of the energy is concentrated in the atmospheric interior for all EOSs.
+However, the energy of the high-$T$ realistic EOS and of the realistic EOS at all $T$ is much larger in magnitude (more negative) near the core when compared to an ideal gas polytrope.
+This is due to the fact that $\delad$ decreases in dissociation regions (Figure \ref{fig:deladmap}), which results in a shallower temperature gradient and thus lower temperatures near the core.
+Figure \ref{fig:all_plot}, bottom-left panel shows the radial temperature profile (for $M\co=10 M_{\oplus}$ and $M_{\rm tot}=12 M_{\oplus}$), with temperatures lower near the core for the high-$T$ realistic EOS.
+As the internal energy $U \propto T$, dissociation decreases $U$ and thus the total energy deep in the interior.
+This explains the local energy behavior shown in Figure \ref{fig:all_plot}, middle-left panel.
+Note that, in some astrophysical contexts, this loss of thermal energy due to dissociation can be so large as to trigger dynamical instabilities and eventual collapse in higher mass objects, such as in protostars \citep{larson69} or during the runaway growth of giant planets \citep{bodenheimer80}.
+
+The energy behavior can also be explained qualitatively through a simple analytic argument: the density profile in an adiabatic, non-self-gravitating atmosphere composed of an ideal gas scales as $\rho(r) \propto r^{-1/\delad+1}$, and the total specific energy is $e(r) \propto 1/r$ (see Paper I).
+Thus $e \rho r^3 \sim r^{-1/\delad+3}$.
+It follows that adiabats with lower $\delad$ have their energy more tightly packed towards the interior of the envelope, which is the case during dissociation.
+This is consistent with the energy behavior in Figure \ref{fig:all_plot}, middle-left panel.
+
+We note that the amount of energy that needs to be released in order to accrete gas at the outer boundary is going to be roughly equal for all EOSs, as it has to match the outer boundary conditions of the nebula.
+However, a more negative total energy in the interior, as is the case for the high-$T$ EOS, requires a larger amount of energy to be radiated away to bind the next batch of gas.
+This implies a larger $-dE/dM$ during atmospheric evolution (middle-right panel of Figure \ref{fig:all_plot}), and thus a lower $\dot{M}$ for the high-T EOS (lower-right panel of Figure \ref{fig:all_plot}).
+
+%The high-$T$ realistic EOS has a significantly higher binding energy than the ideal gas polytrope.
+
+%which increases the amount of energy that needs to be radiated away in order to accrete more gas, i.e. $-dE/dM$.
+
+ %Dissociation deep in the atmosphere reduces $\delad$, and thus the temperature gradient.
+Figure \ref{fig:all_plot}, bottom-left panel shows the radial temperature profile (for $M\co=10 M_{\oplus}$ and $M_{\rm tot}=12 M_{\oplus}$).
+A lower adiabatic gradient results in lower temperatures near the core for the high-$T$ EOS when compared to the ideal gas polytrope.
+As the internal energy $U \propto T$, dissociation decreases $U$.
+The total energy $E$ thus becomes more negative
+
+%The bulk of the energy for the high-$T$ realistic EOS and the realistic EOS at all $T$ is , which shows that hydrogen dissociation in the inner atmosphere dictates the energy behavior of the envelope.
+In contrast, energy is concentrated towards the outer boundary for the $\delad=0.3$ polytrope.
+%{\bf The key question here is how does the slope of the energy affect the quantity that matters, $|dE/dM|$?
+This is not clear yet.
+I think the shift to lower temperatures, seen in the $T(r)$ plot is helpful, but perhaps the EOS plots should also include an internal energy map.}
+%Qualitatively, this can be explained through a simple analytic argument: the density profile in an adiabatic, non-self-gravitating atmosphere composed of an ideal gas scales as $\rho(r) \propto r^{-1/\delad+1}$, and the total specific energy is $e(r) \propto 1/r$ (see Paper I).
+Thus the total energy scales as $E(r) \propto e(r) \rho(r) r^3 \sim r^{-1/\delad+3}$.
+It follows that adiabats with lower $\delad$ have their energy more tightly packed towards the interior of the envelope, which is the case during dissociation when $\delad$ decreases significantly (see Section \S\ref{deladtable} and Figure \ref{fig:deladmap}).
+Dissociation thus causes the atmospheric interior to dominate the total binding energy, which increases the amount of energy that needs to be radiated away in order to accrete more gas, i.e.
+$-dE/dM$.
+
+%The larger $-dE/dM$ due to dissociation can also be understood directly from the $\delad$ behavior.
+Dissociation deep in the atmosphere reduces $\delad$, and thus the temperature gradient.
+Figure \ref{fig:all_plot}, bottom-left panel shows the radial temperature profile (for $M\co=10 M_{\oplus}$ and $M_{\rm tot}=12 M_{\oplus}$).
+A lower adiabatic gradient results in lower temperatures near the core for the high-$T$ EOS when compared to the ideal gas polytrope.
+As the internal energy $U \propto T$, dissociation decreases $U$.
+The total energy $E$ thus becomes more negative, which translates into a larger $-dE/dM$ during atmospheric evolution (middle-right panel of Figure \ref{fig:all_plot}), and thus a lower $\dot{M}$ for the high-T EOS (lower-right panel of Figure \ref{fig:all_plot}).
+Note that, in some astrophysical contexts, this loss of thermal energy due to dissociation can be so large as to trigger dynamical instabilities and eventual collapse in higher mass objects, such as in protostars \citep{larson69} or during the runaway growth of giant planets \citep{bodenheimer80}.
+
+% It takes more energy to bring gas deep into the atmosphere for an envelope that has the bulk of its energy concentrated towards the bottom,  {\bf (This argument isn't clear, new gas isn't brought deep into the atmosphere, it is added at the surface.)} which increases the amount of energy per unit mass that needs to be radiated, i.e.
 $|dE/dM|$.
-The cooling time is therefore larger for the realistic EOS at high temperatures (3) when compared to the ideal gas polytrope (1).
+The cooling time is therefore larger {\bf with dissociation} \sout{for the realistic EOS at high temperatures} (3) when compared to the ideal gas polytrope (1).
 This increase in $|dE/dM|$ is shown in the bottom-right panel of Figure \ref{fig:all_plot}.
 
-We have seen that quantum rotational states transitions due to the low temperatures at the top of the atmosphere increase the thickness of the radiative zone, and decrease the luminosity of the envelope, while dissociation at high temperatures deep in the atmosphere increases the amount of energy needed to increase the atmosphere mass by a fixed amount.
-Overall, both effects result in a longer time for the atmosphere to evolve.
+From Equation (\ref{eq:dMdt}), realistic EOS effects reduce the atmosphere accretion rate $\dot{M}$ by decreasing $L$ and increasing $-dE/dM$.
+This results in slower atmospheric growth and therefore a larger critical core mass, as we show in Section \S\ref{critical}.
+
+%Molecular rotational effects at the top of the atmosphere increase the luminosity $L$, 
+%
+%We have seen that quantum rotational states transitions due to the low temperatures at the top of the atmosphere {\bf (already too long and convoluted)} increase the thickness of the radiative zone, and decrease the luminosity of the envelope, while dissociation at high temperatures deep in the atmosphere increases the amount of energy needed to increase the atmosphere mass by a fixed amount.
+{\bf Way too long a sentence!} Overall, both effects result in a longer time for the atmosphere to evolve.
 

### 2014-01-16 tex/PaperI/ms.tex
revised draft + referee response

@@
 We determine the minimum core mass for a giant planet to form within a typical disk lifetime of 3 Myr.
-The minimum core mass declines with disk radius, from $\sim$$8.5 M_{\oplus}$ at 5 AU to $\sim$$3.5 M_{\oplus}$ at 100 AU, with standard dust opacities.
+The minimum core mass declines with disk radius, from $\sim$$8.5 M_{\oplus}$ at 5 AU to $\sim$$3.5 M_{\oplus}$ at 100 AU, with standard \textbf{interstellar grain} opacities.
 Lower temperatures in the outer disk explain this trend, while variations in disk density are less influential.
@@
 \item Gravitational contraction of the atmosphere is the only source of planetary luminosity.  
-\item Cooling of the atmosphere is dominated by the convective interior.  Thus the luminosity in the radiative exterior is held spatially constant. 
+\item Cooling of the atmosphere is dominated by the convective interior.
+Thus the luminosity in the radiative exterior is held spatially constant.
+\textbf{We justify this assumption in \S\ref{sec:twolayer} and confirm its validity in \S\ref{sec:endoftime}.}
 %\item A global cooling model connects independent static solutions into a time-dependent sequence.
 \item A polytropic equation of state (EOS) is assumed for simplicity.
-\item Dust grains provide the opacity in the radiative zone, which remains cool enough to avoid dust sublimation.
+\item Dust grains provide the opacity in the radiative zone.
+\textbf{For the cool temperatures in the outer disk, our assumption of an inner convective --- outer radiative structure self-consistently ensures that the radiative layer remains cool enough to avoid dust sublimation.
+\footnote{Other opacity choices are discussed in \S\ref{sec:op}}.} %which remains cool enough to avoid dust sublimation.
+\textbf{This follows self-consistently }
 %\item Because gas accretion accelerates after the crossover mass, the time to reach the crossover mass, i.e. the crossover time, is a good approximation of the total time to form a gas giant.
@@
 We use $\RH$ as the matching radius in all regimes.
-We are accounting for the fact that, in hydrostatic models, the density at
- $\RB$ is higher than the background density (see R06). Nevertheless, since the density change is modest, these two boundary conditions give similar results.
+\textbf{We believe $\RH$ to be the best outer boundary choice, as it allows us to include the density perturbations outside the Bondi radius in the structure calculations.
+This is also consistent with the fact that, in hydrostatic models, the density at $\RB$ is higher than the background density (see R06).
+Nevertheless, since the density change is modest, these two boundary conditions give similar results.}
+
+ %We are accounting for the fact that, in hydrostatic models, the density at 
+ %$\RB$ is higher than the background density (see R06). Nevertheless, since the density change is modest, these two boundary conditions give similar results.
 
@@
 While $\RH$ is always the outer boundary in our structure calculations, we quote planet masses as the enclosed mass within the smaller of $\RB$ and $\RH$.
+\textbf{In contrast with the choice of outer boundary, which has physical consequences, we note that the choice of atmospheric mass is primarily a convention that does not affect atmospheric structure or evolution.
+Despite the possible confusion between two outer boundaries, we believe this to be the preferred location for several reasons.
+Our choice of envelope mass is conservative, as it gives the smaller mass when $\RB < \RH$, and is consistent with previous works that use $\RB$ as the outer boundary (e.g., \citealt{ikoma00}, \citealt{pollack96}).
+The extra mass between $\RB$ and $\RH$, however large, is nebular gas weakly compressed at the disk midplane density, unlikely to remain bound to the planet if the disk were to dissipate quickly.
+}
+
  %  This conservative choice in quoting planet masses is usually a minor distinction because (when $\RB < \RH$) the gas between $\RB$ and $\RH$ is weakly compressed.
+
+%\textbf{Finally, the mass at which core growth rapidly decreases due to depletion of planetesimals from the core's feeding zone is called the isolation mass.
+For our fiducial disk model and for an assumed dust-to-gas ratio of 0.01, the isolation mass is given by}
+%
+%\begin{equation}
+%\label{eq:Misol}
+%M_{\rm isol} \sim 2 \pi a \times 2 \RH \times \Sigma_{\rm p} \approx 0.2 a_{10}^{3/4} ~ M_{\oplus},
+%\end{equation}
+%\textbf{where $\Sigma_p$ is the surface density of solids.
+Since in our regime the core is no longer accreting solids, we require $M_{\rm isol}$ to always be smaller than the core mass at which runaway gas accretion commences, i.e.
+the critical core mass.}
 
@@
 However, our subsequent developments will replace the local energy equation (\ref{eq:structd}) with global energy balance (see section \S\ref{cooling}), reverting the structure equations to time-independent ordinary differential equations (ODEs).
+%\textbf{}
 
@@
 In the radiative zone,  local energy balance, \Eq{eq:structd}, does affect the atmospheric structure.
-We proceed by assuming that the majority of energy is lost from the convective interior, and thus the luminosity can be treated as constant in the outer radiative zone.
+We proceed by assuming that the majority of energy is lost from the convective interior, and thus the luminosity can be treated as constant in the outer radiative zone, \textbf{i.e.
+the RHS of \Eq{eq:structd} is set to zero.
+This assumption simplifies the numerical problem, as static solutions can now be found by simply integrating Equations (\ref{eq:struct}a -- c) with our choice of boundary conditions.
+We show in \S\ref{sec:endoftime} that the approximation of constant luminosity in the outer layer holds for our regime of interest.
+Under our assumptions, the  convective interior fully sets the atmospheric luminosity.
+We note, however, that the location and extent of the convective region is set by the matching of the upper radiative layer to the nebula.} %is set by the convective interior, and the radiative upper layer   luminosity  emerging from the convective interior   Since most of the luminosity emerges from the inner convective region of the atmosphere,
 
@@
 Conditions at the RCB are crucial as they set the interior adiabat and the radiative losses from the interior.
-(Recall that luminosity generation in the radiative zone is neglected.)  We express the temperature and pressure of the RCB, at the radius $R\cb$, as
+(Recall that luminosity generation in the radiative zone is neglected.)  \textbf{Rather than setting the outer boundary at $\RH$ as in the numerical model, we assume that our non-self-gravitating atmosphere matches onto the disk at a distance $r_{\rm fit}=n_{\rm fit} \RB$, where $n_{\rm fit}>1$ (see \App{iso} and R06).
+We note that the exact choice of $n_{\rm fit}$ is not important for the non-self-gravitating atmosphere structure, as it only has an order unity effect on the density at $\RB$.
+For $n_{\rm fit} \rightarrow \infty$, }we express the temperature and pressure of the RCB, at the radius $R\cb$, as
 \begin{subeqnarray}\label{eq:cb2}
@@
 This condition requires $\gamma < 3/2$, which our choice of $\gamma = 7/5$ satisfies, but $\gamma = 5/3$ (monatomic gas) would not.
+\textbf{While a realistic gas may become monatomic at the high temperatures near the core due to hydrogen dissociation, in this work we assume that the gas is diatomic throughout, with $\gamma=7/5$ and thus the bulk of its total energy concentrated near the core.
+The effect of a realistic EOS on the atmospheric energy distribution is discussed in Piso, Youdin, \& Murray-Clay (in prep, hereafter Paper II).}
+
 Integration of \Eq{eq:ean} over the mass of the atmosphere thus gives
@@
 
-To facilitate comparison with our numerical results, we rescale the analytic results as a rough correction for the neglect of self gravity.
+To facilitate comparison with our numerical results, we rescale the analytic results as a rough correction for the neglect of self-gravity.
 The goal of the rescaling is not to get a perfect agreement, but to aid in the interpretation of general trends.
 To rescale, we simply assume that runaway occurs at a modified crossover mass, $M_{\rm atm} = f M\co$.
-From \Eqs{eq:crit}{eq:Pcbc}, this change amounts to replacing $\xi \rightarrow f\xi$.
-We require $f < 1$ to give the faster growth seen in numerical models.
+\textbf{Here $f<1$, as growth is slower in our analytic model due to the neglect of self-gravity.
+For the purpose of this scaling, we are only interested in atmospheric growth up to the modified crossover mass defined above.
+We are thus not concerned about evolution to yet higher masses where the analytic approximation becomes inaccurate.} Mathematically, this choice of crossover mass amounts to replacing $\xi \rightarrow f\xi$ (cf.
+Equations \ref{eq:crit} and \ref{eq:Pcbc}).
 We choose $f = 0.13$, which normalizes the analytic and numerical models at 10 AU, for standard parameters.
@@
 now using $t_{\rm run}$ instead of $t_{\rm co}$ for the runaway growth timescale, which is no longer tied to the crossover mass.
-The numerical models are more accurate and self-consistent because they measure runaway growth directly, instead of relying on any assumed $M_{\rm atm}/M\co$ threshold.
+\textbf{Here $t_{\rm run}$ is the time at which $M_{\rm atm}=f M\co$.} The numerical models are more accurate and self-consistent because they measure runaway growth directly, instead of relying on any assumed $M_{\rm atm}/M\co$ threshold.
 When plotting the analytic scalings, we include the variations in $\xi$ from \Eq{eq:xidef}.
@@
 
-The atmospheric structure is well approximated by our non-self-gravitating, analytic solutions.
-Deep in convective zones, \Eq{eq:deep} gives $T \propto r^{-1}$ and $P\propto r^{-1/\delad}$.
-This behavior is seen in the low mass solutions in \Fig{fig:profiles}.
-As explained by \Eq{eq:TPsg}, the high mass solutions show a slightly flatter profile in $T$ and also in $P \propto T^{1/\delad}$.
-In agreement with \Eq{eq:cb2}, the radiative zones remain nearly isothermal, even for the higher masses.
-Consequently, the pressure increases nearly exponentially with depth.
+\textbf{The atmospheric structure is well approximated by our non-self-gravitating, analytic solutions.
+Deep in convective zones, \Eq{eq:Tconvdeep} gives $T \propto r^{-1}$ and $P\propto r^{-1/\delad}$.
+This behavior is seen in all solutions in \Fig{fig:profiles} for $r\ll R\cb$.
+Near the core, where $m(r) \approx M\co$, \Eq{eq:deep} gives an expression for the core temperature, $T\co=\frac{G M\co}{C_{\rm P} R\co}$, which holds regardless of the atmosphere mass, as shown in Figure \ref{fig:profiles}.
+As the RCB is approached, \Eq{eq:Tconvdeep} is no longer valid, particularly for large envelope masses.
+Instead, these high mass solutions show a slightly flatter profile in $T$ and also in $P \propto T^{1/\delad}$, as explained by \Eq{eq:TPsg}.}
+
+In agreement with \Eq{eq:cb2}, the radiative zones remain nearly isothermal, even for the higher masses.  Consequently, the pressure increases nearly exponentially with depth.
 
@@
 In the higher mass solutions, the relatively low $P\cb$ values (and thus the relatively high luminosities) require an outward shift in $R\cb$, as shown in \Fig{fig:profiles}.
+\textbf{This shift does not occur in the analytic solution, where $R\cb$ continually decreases with mass (cf.
+Equations \ref{eq:crit} and \ref{eq:PcbRcb}).}
 
@@
 We see that self-gravity strongly affects atmospheric evolution, even well before the crossover mass, when $M_{\rm atm}=M\co$.
+\textbf{The early divergence between the analytic and numerical curves in Figure \ref{fig:Ltplot} shows that self-gravity becomes important even when the atmosphere is only a few percent as massive as the core.
+Our analytic model starts deviating from the numerical model at slightly larger $M_{\rm atm}/M\co$ closer to the star (not shown).
+However, we have found that self-gravity is always significant at $M_{\rm atm} \leq 10\% M\co$ in our regime of interest.}
 
@@
 
-We now use our structure and evolution models to calculate the necessary conditions for gas giant formation by core accretion.
+%We now use our structure and evolution models to calculate the necessary conditions for gas giant formation by core accretion.   
+\textbf{We now use our structure and evolution models to estimate the timescales and core masses necessary for giant planet formation, for our assumed disk model, nebular gas properties and dust grain opacity.}
 Adopting a median disk lifetime of 3 Myr, we examine the conditions for a fixed mass core to undergo runaway gas accretion within the lifetime of the gas disk.
@@
 This effect is explained by the increase in the atmospheric scale height $\propto 1/\mu$ while the Bondi radius decreases, $\RB \propto \mu$.
-For $\mu = 2.0$, which represents the idealized case of an H$_2$ atmosphere completely devoid of Helium, $t_{\rm run}$ increases by factors of $\sim 2 - 3$.
+\textbf{The larger atmospheric scale height due to a lower $\mu$ results in a larger pressure depth at the RCB, $P\cb$, and thus a lower $R\cb$, cf.
+\Eq{eq:PcbRcb}.} For $\mu = 2.0$, which represents the idealized case of an H$_2$ atmosphere completely devoid of Helium, $t_{\rm run}$ increases by factors of $\sim 2 - 3$.
 Thus fairly drastic changes in atmospheric composition are required for $\mu$ to significantly affect core accretion timescales.
+\textbf{While we note that changes in the mean molecular weight of the gas also affect the adiabatic gradient $\delad$, we defer the discussion of the influence of $\delad$ on atmospheric evolution to Paper II.}
 
@@
 The opacity reduction by a factor of 10 lowers $\MC$ at 5 AU by a factor of $\sim$$2.5$ and at 100 AU by a factor of $\sim$$3.5$.
-Atmospheric opacity is a dominant uncertainty in core accretion modelling.
+Atmospheric opacity is a dominant uncertainty in core accretion modeling.
 However, unless atmospheric opacity varies significantly with disk radius, the general decline in $\MC$ with increasing $a$ should hold.
@@
 Nonetheless, since opacity near the RCB is a crucial factor, we speculate that convective overshoot may prevent very low opacities.
+
+\textbf{A larger disk lifetime further reduces $M_{\rm crit}$.
+Recent studies have shown that gas disks may live up to $\sim$$10-12$ Myr \citep{bell13}.
+As $M_{\rm crit} \sim t\di^{-3/5}$ (cf.
+Equation \ref{eq:tcoolan}), a disk lifetime of 10 Myrs decreases the critical core mass by a factor of two.}
+
+%\textbf{As the core is assumed to form through accretion of solids, we require that $M_{\rm crit}>M_{\rm isol}$.
+Since $M_{\rm crit}$ decreases with semi major axis, while $M_{\rm isol}$ increases with $a$ (cf.
+Equation \ref{eq:Misol}), it is possible that the isolation mass may become larger than the critical core mass at large orbital distances.
+From \Eq{eq:Misol}, $M_{\rm isol} \approx 1 M_{\oplus}$ at 100 AU, which is smaller than $M_{\rm crit}$ under standard opacity assumptions, but comparable for a reduced opacity (see Figure \ref{fig:Mcritopacity}).
+A flatter surface density profile (see \S\ref{sec:disk}) may further increase $M_{\rm isol}$.
+We note, however, that the estimate of \Eq{eq:Misol} assumes an initial uniform dust-to-gas ratio throughout the disk, although the dust-to-gas ratio may be distance dependent.
+A lower dust-to-gas ratio in the outer disk reduces $M_{\rm isol}$.
+As stated in \S\ref{sec:scales}, we assume that the isolation mass is always lower than our calculated $M_{\rm crit}$.}
+
+%\textbf{Finally, the mass at which core growth rapidly decreases due to depletion of planetesimals from the core's feeding zone is called the isolation mass.
+For our fiducial disk model and for an assumed dust-to-gas ratio of 0.01, the isolation mass is given by}
+%
+%\begin{equation}
+%\label{eq:Misol}
+%M_{\rm isol} \sim 2 \pi a \times 2 \RH \times \Sigma_{\rm p} \approx 0.2 a_{10}^{3/4} ~ M_{\oplus},
+%\end{equation}
+%\textbf{where $\Sigma_p$ is the surface density of solids.
+Since in our regime the core is no longer accreting solids, we require $M_{\rm isol}$ to always be smaller than the core mass at which runaway gas accretion commences, i.e.
+the critical core mass.}
+
+\subsection{\textbf{Comparison with Previous Studies}}
+\label{sec:comp}
+
+\textbf{Our results are consistent with those of previous core accretion studies.
+The solution of \citet{stevenson82}, which considers fully radiative atmospheres with constant opacity, predicts that the critical core mass is lower for lower envelope opacity and higher mean molecular weight of the nebular gas.
+Additionally, \citet{stevenson82} finds that the critical core mass decreases weakly with increasing envelope radius (via a logarithmic factor), i.e.
+with increasing orbital distance.
+Both of the above qualitative predictions are reproduced by our result.}
+
+\textbf{In the regime of no ongoing planetesimal accretion, \citet{ikoma00} find an approximate analytic expression for the envelope formation time (the equivalent of the runaway accretion time $t_{\rm run}$ in our models), }
+
+\begin{equation}
+\label{eq:tauenv}
+\tau_{\rm env} \sim 3 \times 10^8 \Big(\frac{M\co}{M_{\oplus}}\Big)^{-2.5} \Big(\frac{\kappa}{1 \text{cm$^2$ s$^{-1}$}}\Big) \,\,\, \text{yr},
+\end{equation}
+\textbf{at $a=5.2$ AU, Jupiter's location.
+The result above reproduces the linear opacity dependence of the atmospheric evolutionary times in our model (see \S\ref{sec:timeev}).
+Moreover, we can approximate $t_{\rm run}$ in our model at 5 AU as}
+\begin{equation}
+\label{eq:trun5au}
+t_{\rm run} \sim 5 \times 10^8 \Big(\frac{M\co}{M_{\oplus}}\Big)^{-2.4} \,\,\, \text{yr},
+\end{equation} 
+\textbf{for the opacity-law (\ref{eq:opacitylaw}). Our results are thus consistent with those of \citet{ikoma00} at 5 AU, both in terms of scaling and magnitude.}
+
+\textbf{Time-dependent studies that incorporate planetesimal accretion generally find larger formation timescales and critical core masses than those of this work.
+\citet{pollack96} find an evolutionary time and crossover mass of $\sim$$7.5$ Myrs and $\sim$$16 M_{\oplus}$, respectively, for a MMSN disk model at 5.2 AU and interstellar grain opacity.
+These values, however, decrease to $\sim$$3$ Myrs and $\sim$$12 M_{\oplus}$ if planetesimal accretion is entirely shut off during the gas accretion phase.
+The core accretion model of \citet{HubBod05} for Jupiter's formation predicts an evolutionary time of 3.3 Myrs for a $10 M_{\oplus}$ core, for an interstellar opacity, which is also consistent with our result at 5 AU.}
+
+%\textbf{Lastly, the decline in $M_{\rm crit}$ with semimajor axis is also found by R06 for planets forming at $\gtrsim 5$ AU.
+A quantitative comparison with our results requires detailed calculations of the }
 
@@
 
+\textbf{The importance of envelope opacity is an established factor in core accretion calculations (e.g., \citealt{stevenson82}, \citealt{ikoma00}, R06), and several studies explore the influence of opacity on the timescales for giant planet formation in detail (e.g., \citealt{HubBod05})}.
 Treating dust (and total) opacities as a power law in temperature is a simplification.
@@
 Grain growth and composition also affect opacities.
-Piso et al.
-(in prep.) explores more realistic opacity laws.
+\textbf{Our scenario of no ongoing accretion of solids may result in a grain-free atmosphere, which could substantially reduce the critical core mass ($\sim$$1M_{\oplus}$ in the case of Jupiter, \citealt{hori10}).} Paper II explores more realistic opacity laws.
 
@@
 In the upper atmosphere, H$_2$ rotational levels can become depopulated at lower temeratures.
-Piso et al.
-(in prep) uses the \citet{saumon95} EOS, with extensions to lower pressures and temperatures as needed, to explore these effects.
+Paper II uses the \citet{saumon95} EOS, with extensions to lower pressures and temperatures as needed, to explore these effects.
 
@@
 For similar parameters, we generally obtain lower $\MC$ values than R06 because our atmospheres are not heated by ongoing planetesimal accretion.
-Piso et al.\ (in prep) presents a quantitative comparison.
+Paper II presents a quantitative comparison.
+
+%For giant planet formation to occur with a core mass close to our $M_{\rm crit}$, \textit{in-situ} formation of the core requires $M_{\rm crit}$ to be smaller than the isolation mass, $M_{\rm iso}$.
+}
+
+%\textbf{As the core is assumed to form through accretion of solids, we require that $M_{\rm crit}>M_{\rm isol}$.
+Since $M_{\rm crit}$ decreases with semi major axis, while $M_{\rm isol}$ increases with $a$ (cf.
+Equation \ref{eq:Misol}), it is possible that the isolation mass may become larger than the critical core mass at large orbital distances.
+From \Eq{eq:Misol}, $M_{\rm isol} \approx 1 M_{\oplus}$ at 100 AU, which is smaller than $M_{\rm crit}$ under standard opacity assumptions, but comparable for a reduced opacity (see Figure \ref{fig:Mcritopacity}).
+A flatter surface density profile (see \S\ref{sec:disk}) may further increase $M_{\rm isol}$.
+We note, however, that the estimate of \Eq{eq:Misol} assumes an initial uniform dust-to-gas ratio throughout the disk, although the dust-to-gas ratio may be distance dependent.
+A lower dust-to-gas ratio in the outer disk reduces $M_{\rm isol}$.
+As stated in \S\ref{sec:scales}, we assume that the isolation mass is always lower than our calculated $M_{\rm crit}$.}
 
@@
 \item Runaway growth begins roughly at the crossover mass, when atmosphere and core masses are equal, $M_{\rm atm} \sim M\co$, in agreement with previous work \citep{pollack96}.
-Further from the star, runaway growth begins at smaller $M_{\rm atm} \sim M\co$ ratios.
-For larger core masses, runaway growth begins at larger values of $M_{\rm atm} \sim M\co$.
+Further from the star, runaway growth begins at smaller $M_{\rm atm} / M\co$ ratios.
+For larger core masses, runaway growth begins at larger values of $M_{\rm atm} / M\co$.
 
@@
 However, our model does not address the details of how solid cores grow, and many possibilities exist.
+\textbf{Our brief treatment of core formation is purposeful, as there are large uncertainties in both planetesimal formation and the subsequent processes of core growth, particularly at large orbital distances.
+Core formation in the outer disk is challenging due to large dynamical times, and cores are unlikely to reach critical masses at $\gtrsim 40-50$ AU under classic core accretion scenarios \citep{Raf11}.
+However, modern models of aerodynamic pebble accretion (\citealt{ormel10}; \citealt{lambrechts12}) can give faster growth.
+Moreover, cores can form closer to the star, then migrate or be scattered outwards by other giants in the system \citep{ida13}.}
+
  For a giant planet to form with a core near the minimum mass, the core's growth must  first be rapid and then slow significantly, as in phases I and II, respectively, of \cite{pollack96}.
-
 Initial core growth must be fast, compared to the disk lifetime, to get a sufficiently massive core.
@@
 On the other hand, non-negligible accretion of solids onto cores could continue for the lifetime of the gas disk.
-In these cases, relatively high mass cores would remain as ice giants, akin to Uranus and Neptune.
-Ongoing direct imaging surveys and their successors \citep{hinz12, macintosh12} will help discriminate among the various planet formation pathways in the outskirts of protoplanetary disks.
+\textbf{For our fiducial disk model, the isolation mass, $M_{\rm iso} \propto a^{3/4}$, increases with orbital distance, and may thus become larger than $M_{\rm crit}$ at large separations.
+In this case, the ongoing accretion of solids increases the critical core mass by inhibiting atmospheric growth, and may prevent accumulation of a significant envelope before the disk dissipates.
+It follows that} relatively high mass cores would remain as ice giants, akin to Uranus and Neptune.
+\textbf{Non-traditional accretion models such as pebble accretion or scattering are thus needed in the outer disk in order to avoid slow growth to the isolation mass, which would stall atmospheric evolution.}  Ongoing direct imaging surveys and their successors \citep{hinz12, macintosh12} will help discriminate among the various planet formation pathways in the outskirts of protoplanetary disks.
 

### 2014-02-06 tex/PaperI/ms.tex
minor changes in progress to referees comments

@@
 \item Dust grains provide the opacity in the radiative zone.
-\textbf{For the cool temperatures in the outer disk, our assumption of an inner convective --- outer radiative structure self-consistently ensures that the radiative layer remains cool enough to avoid dust sublimation.
-\footnote{Other opacity choices are discussed in \S\ref{sec:op}}.} %which remains cool enough to avoid dust sublimation.
-\textbf{This follows self-consistently }
+\textbf{For the cool temperatures in the outer disk, the radiative layer remains cool enough to avoid dust sublimation.} Other opacity choices are discussed in \S\ref{sec:op}.
 %\item Because gas accretion accelerates after the crossover mass, the time to reach the crossover mass, i.e. the crossover time, is a good approximation of the total time to form a gas giant.
@@
 
-The first assumption of spherical accretion breaks down early in the inner disk, as explained below.  However, our focus is the outer disk, where this standard approximation is better justified.
+The first assumption of spherical accretion breaks down early in the inner disk, as explained below.  However, our focus is on the outer disk, where this standard approximation is better justified.
 
@@
 \end{equation}
-where hydrostatic balance breaks down.  (Here we use the same symbol, $M\pla$, to now mean the mass within $\RH$.)
+where spherical symmetry and hydrostatic balance break down.  (Here we use the same symbol, $M\pla$, to now mean the mass within $\RH$.)
 
@@
 In this regime, many studies assume the atmosphere matches the disk conditions at $\RB$.
-We use $\RH$ as the matching radius in all regimes.
-\textbf{We believe $\RH$ to be the best outer boundary choice, as it allows us to include the density perturbations outside the Bondi radius in the structure calculations.
-This is also consistent with the fact that, in hydrostatic models, the density at $\RB$ is higher than the background density (see R06).
-Nevertheless, since the density change is modest, these two boundary conditions give similar results.}
+We use $\RH$ as the matching radius, i.e.\ outer boundary, in all regimes.
+\textbf{This choice reflects the fact that, even outside the Bondi radius, hydrostatic densities exceed the background density (see R06).
+Nevertheless, since the density contrast is modest outside $\RB$, these two boundary conditions give similar results.
+We do not extend our models past $\RH$ where tidal forces dominate.
+}
 
@@
 While $\RH$ is always the outer boundary in our structure calculations, we quote planet masses as the enclosed mass within the smaller of $\RB$ and $\RH$.
-\textbf{In contrast with the choice of outer boundary, which has physical consequences, we note that the choice of atmospheric mass is primarily a convention that does not affect atmospheric structure or evolution.
-Despite the possible confusion between two outer boundaries, we believe this to be the preferred location for several reasons.
-Our choice of envelope mass is conservative, as it gives the smaller mass when $\RB < \RH$, and is consistent with previous works that use $\RB$ as the outer boundary (e.g., \citealt{ikoma00}, \citealt{pollack96}).
-The extra mass between $\RB$ and $\RH$, however large, is nebular gas weakly compressed at the disk midplane density, unlikely to remain bound to the planet if the disk were to dissipate quickly.
-}
+\textbf{When $\RB < \RH$, this means that our enclosed mass extends to a smaller radius than the outer boundary.
+This choice is motivated as follows.
+First, we want to conservatively define planet mass in a way that does not exaggerate atmospheric growth and is more consistent with previous works that use $\RB$ as the outer boundary in this regime (e.g., \citealt{ikoma00}, \citealt{pollack96}).
+Second, most of the gas outside $\RB$ is weakly bound, and unlikely to remain attached to planet if the disk suddenly dissipates.
+Our convention for enclosed mass does not affect our structure and evolution calculations, which use the physical mass at every radius.}
 
@@
 We show in \S\ref{sec:endoftime} that the approximation of constant luminosity in the outer layer holds for our regime of interest.
-Under our assumptions, the  convective interior fully sets the atmospheric luminosity.
+radiative luminosity is thus set by the location of, and physical conditions at, the ra ...}
+
+{Under our assumptions, the  convective interior fully sets the atmospheric luminosity.
 We note, however, that the location and extent of the convective region is set by the matching of the upper radiative layer to the nebula.} %is set by the convective interior, and the radiative upper layer   luminosity  emerging from the convective interior   Since most of the luminosity emerges from the inner convective region of the atmosphere,

### 2014-02-07 tex/PaperI/ms.tex
edits of revisions up to section 3.1

@@
 %\item A global cooling model connects independent static solutions into a time-dependent sequence.
-\item A polytropic equation of state (EOS) is assumed for simplicity.
+\item The atmosphere obeys a polytropic equation of state (EOS), \textbf{with adiabatic index $\gamma = 7/5$ for an ideal diatomic gas.
+Corrections from a realistic EOS are discussed in \S\ref{sec:EOS} and deferred to future work.}
 \item Dust grains provide the opacity in the radiative zone.
@@
 the RHS of \Eq{eq:structd} is set to zero.
-This assumption simplifies the numerical problem, as static solutions can now be found by simply integrating Equations (\ref{eq:struct}a -- c) with our choice of boundary conditions.
+This assumption greatly simplifies the numerical problem.
+Instead of solving time dependent partial differential equations, we can solve for  static solutions to a set of ODEs, Equations (\ref{eq:struct}a -- c), which we connect in a time series as described below.
 We show in \S\ref{sec:endoftime} that the approximation of constant luminosity in the outer layer holds for our regime of interest.
-radiative luminosity is thus set by the location of, and physical conditions at, the ra ...}
-
-{Under our assumptions, the  convective interior fully sets the atmospheric luminosity.
+}
+
+%{Under our assumptions, the  convective interior fully sets the atmospheric luminosity.
 We note, however, that the location and extent of the convective region is set by the matching of the upper radiative layer to the nebula.} %is set by the convective interior, and the radiative upper layer   luminosity  emerging from the convective interior   Since most of the luminosity emerges from the inner convective region of the atmosphere,
-
-With this approximation for the radiative zone, we can construct an array of solutions to Equations (\ref{eq:struct}a -- c) in order of increasing atmospheric mass.
-We then use global energy balance, \Eq{eq:coolingglobal}, to ``follow the mass" and place these solutions in a cooling sequence.
-The validity of neglecting luminosity generation in the radiative zone can be checked \emph{a posteriori}, see \S\ref{sec:endoftime}.
 
@@
 At the outer boundary, at $\RH(M_i)$, the temperature and pressure are set to the disk values.
-To integrate Equations (\ref{eq:struct}a--c), the luminosity is required to compute $\delrad$.
+A luminosity value is required to compute $\delrad$ and integrate Equations (\ref{eq:struct}a--c).
 The correct value of the luminosity is not known in advance, and is the eigenvalue of the problem.
@@
 Specifically, we alter the luminosity until the integrated value of mass at the core, $m(R\co)$, matches the actual core mass, $M\co$.
-
-To establish the time difference between neighboring solutions, we apply \Eq{eq:coolingglobal} at the radiative-convective boundary (RCB) of the solutions.
+\textbf{Physically the luminosity determines, consistent with the structure equations and the Schwarzschild criterion, the location of the radiative-convective boundary (RCB).}
+
+\textbf{To understand time evolution, we construct} an array of solutions to Equations (\ref{eq:struct}a -- c) in order of increasing atmospheric mass.
+We then use global energy balance, \Eq{eq:coolingglobal}, to ``follow the mass" and place these solutions in a cooling sequence.
+To establish the time difference between neighboring solutions, we apply \Eq{eq:coolingglobal} at the RCB.
 In principle, energy balance could be evaluated at any level.
@@
 The analytic model also assumes that the upper radiative layer is thick enough that $P\cb \gg P\di$.
-This approximation ignores the earliest stages of cooling, which are rapid enough to be negligible.
-Finally, the analytic model ignores the surface terms in \Eq{eq:coolingglobal}, which we show to be a modest correction in \S\ref{sec:endoftime}.
+This approximation ignores the earliest stages of cooling, which are rapid enough to be of minor importance.
+The analytic model also ignores the surface terms in \Eq{eq:coolingglobal}, which we show to be a modest correction in \S\ref{sec:endoftime}.
+\textbf{Finally, while the numerical models set the outer boundary at $\RH$, we simplify the analytic calculations by setting the outer boundary to infinity, effectively neglecting the finite scale-height of the disk.
+As shown in \App{iso}, the effect of this approximation is minor for $\RB \lesssim \RH$, the primary case of interest (see also R06).}
+
+\textbf{Of all the approximations, the neglect of self-gravity is by far the most significant, as we have verified by comparison to numerical integrations that only neglect self-gravity.}
 
@@
 Conditions at the RCB are crucial as they set the interior adiabat and the radiative losses from the interior.
-(Recall that luminosity generation in the radiative zone is neglected.)  \textbf{Rather than setting the outer boundary at $\RH$ as in the numerical model, we assume that our non-self-gravitating atmosphere matches onto the disk at a distance $r_{\rm fit}=n_{\rm fit} \RB$, where $n_{\rm fit}>1$ (see \App{iso} and R06).
-We note that the exact choice of $n_{\rm fit}$ is not important for the non-self-gravitating atmosphere structure, as it only has an order unity effect on the density at $\RB$.
-For $n_{\rm fit} \rightarrow \infty$, }we express the temperature and pressure of the RCB, at the radius $R\cb$, as
+(Recall that luminosity generation in the radiative zone is neglected.)  We express the temperature and pressure of the RCB, at the radius $R\cb$, as
 \begin{subeqnarray}\label{eq:cb2}
@@
 This condition requires $\gamma < 3/2$, which our choice of $\gamma = 7/5$ satisfies, but $\gamma = 5/3$ (monatomic gas) would not.
-\textbf{While a realistic gas may become monatomic at the high temperatures near the core due to hydrogen dissociation, in this work we assume that the gas is diatomic throughout, with $\gamma=7/5$ and thus the bulk of its total energy concentrated near the core.
-The effect of a realistic EOS on the atmospheric energy distribution is discussed in Piso, Youdin, \& Murray-Clay (in prep, hereafter Paper II).}
 
@@
 To rescale, we simply assume that runaway occurs at a modified crossover mass, $M_{\rm atm} = f M\co$.
+%AY: improve this discussion to make more clear that we are not actually implying that runaway growth begins early and make vernally less rambling.
 \textbf{Here $f<1$, as growth is slower in our analytic model due to the neglect of self-gravity.
@@
 Thus fairly drastic changes in atmospheric composition are required for $\mu$ to significantly affect core accretion timescales.
-\textbf{While we note that changes in the mean molecular weight of the gas also affect the adiabatic gradient $\delad$, we defer the discussion of the influence of $\delad$ on atmospheric evolution to Paper II.}
+\textbf{While we note that changes in the mean molecular weight of the gas also affect the adiabatic gradient $\delad$, we defer the discussion of the influence of $\delad$ on atmospheric evolution to Piso, Youdin \& Murray-Clay (in prep., hereafter Paper II).}
 

### 2014-02-10 tex/PaperI/ms.tex
minor edits

@@
 Nevertheless, since the density contrast is modest outside $\RB$, these two boundary conditions give similar results.
-We do not extend our models past $\RH$ where tidal forces dominate.
+We do not extend our models past $\RH$, where tidal forces dominate.
 }
@@
 First, we want to conservatively define planet mass in a way that does not exaggerate atmospheric growth and is more consistent with previous works that use $\RB$ as the outer boundary in this regime (e.g., \citealt{ikoma00}, \citealt{pollack96}).
-Second, most of the gas outside $\RB$ is weakly bound, and unlikely to remain attached to planet if the disk suddenly dissipates.
+Second, most of the gas outside $\RB$ is weakly bound, and unlikely to remain attached to the planet if the disk suddenly dissipates.
 Our convention for enclosed mass does not affect our structure and evolution calculations, which use the physical mass at every radius.}
@@
 Specifically, we alter the luminosity until the integrated value of mass at the core, $m(R\co)$, matches the actual core mass, $M\co$.
-\textbf{Physically the luminosity determines, consistent with the structure equations and the Schwarzschild criterion, the location of the radiative-convective boundary (RCB).}
+\textbf{Physically, the luminosity determines the location of the radiative-convective boundary (RCB), consistent with the structure equations and the Schwarzschild criterion.}
 

### 2014-02-12 tex/PaperI/ms.tex
text edits up to 5.2 some (not all) changes to response to match

@@
 
-To facilitate comparison with our numerical results, we rescale the analytic results as a rough correction for the neglect of self-gravity.
+To facilitate comparison with our numerical results, \textbf{we rescale the analytic results as follows.
+Since growth times are too slow in the non-self-gravitating analytic model, we modify the runaway grown criterion to occur at an effective crossover mass $M_{\rm atm} = f M\co$ with $f < 1$.
+Specifically we choose $f = 0.13$, because with this value the analytic model gives the same critical core mass at 10 AU as the numerical model, for our standard choices of other parameters.
+This prescription does not mean that runaway growth physically occurs at low atmosphere masses, nor does it cause perfect agreement between the models.
+Rather by artificially accelerating growth in the analytic model, we can more easily compare the parameter scalings of the two models.
+Mathematically, the modified crossover mass amounts to replacing $\xi \rightarrow f\xi$ in our analysis.}  With $f = 0.13$, the revised scalings are
+\begin{subeqnarray}
+t_{\rm run} & \approx & 3 \times 10^6 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2  \over \mcn{10}^{5/3} \aun{10}^{15 / 14}} \yr  \slabel{eq:tcoolanf} \\
+\MC & \approx & 8\, {F_T^{3/2} F_\kappa^{3/5}   \left(\xi \over 2.6 \right)^{6/5} \over \aun{10}^{9 / 14}} \; M_\oplus\ \, ,  \slabel{eq:Mcritf}
+\end{subeqnarray} 
+now using $t_{\rm run}$, defined as the time when $M_{\rm atm} = f M\co$,  instead of $t_{\rm co}$ for the runaway growth timescale.
+When plotting these rescaled analytic results, we include the variations in $\xi$ from \Eq{eq:xidef}.
+
+%To facilitate comparison with our numerical results, we rescale the analytic results as a rough correction for the neglect of self-gravity.
 The goal of the rescaling is not to get a perfect agreement, but to aid in the interpretation of general trends.
 To rescale, we simply assume that runaway occurs at a modified crossover mass, $M_{\rm atm} = f M\co$.
-%AY: improve this discussion to make more clear that we are not actually implying that runaway growth begins early and make vernally less rambling.
-\textbf{Here $f<1$, as growth is slower in our analytic model due to the neglect of self-gravity.
+%\textbf{Here $f<1$, as growth is slower in our analytic model due to the neglect of self-gravity.
 For the purpose of this scaling, we are only interested in atmospheric growth up to the modified crossover mass defined above.
@@
 The revised scalings are
-\begin{subeqnarray}
-t_{\rm run} & \approx & 3 \times 10^6 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2  \over \mcn{10}^{5/3} \aun{10}^{15 / 14}} \yr  \slabel{eq:tcoolanf} \\
-\MC & \approx & 8\, {F_T^{3/2} F_\kappa^{3/5}   \left(\xi \over 2.6 \right)^{6/5} \over \aun{10}^{9 / 14}} \; M_\oplus\ \, ,  \slabel{eq:Mcritf}
-\end{subeqnarray} 
-now using $t_{\rm run}$ instead of $t_{\rm co}$ for the runaway growth timescale, which is no longer tied to the crossover mass.
+%\begin{subeqnarray}
+%t_{\rm run} & \approx & 3 \times 10^6 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2  \over \mcn{10}^{5/3} \aun{10}^{15 / 14}} \yr  \slabel{eq:tcoolanf} \\
+%\MC & \approx & 8\, {F_T^{3/2} F_\kappa^{3/5}   \left(\xi \over 2.6 \right)^{6/5} \over \aun{10}^{9 / 14}} \; M_\oplus\ \, ,  \slabel{eq:Mcritf}
+%\end{subeqnarray} 
+%now using $t_{\rm run}$ instead of $t_{\rm co}$ for the runaway growth timescale, which is no longer tied to the crossover mass.
 \textbf{Here $t_{\rm run}$ is the time at which $M_{\rm atm}=f M\co$.} The numerical models are more accurate and self-consistent because they measure runaway growth directly, instead of relying on any assumed $M_{\rm atm}/M\co$ threshold.
@@
 \textbf{The atmospheric structure is well approximated by our non-self-gravitating, analytic solutions.
-Deep in convective zones, \Eq{eq:Tconvdeep} gives $T \propto r^{-1}$ and $P\propto r^{-1/\delad}$.
-This behavior is seen in all solutions in \Fig{fig:profiles} for $r\ll R\cb$.
-Near the core, where $m(r) \approx M\co$, \Eq{eq:deep} gives an expression for the core temperature, $T\co=\frac{G M\co}{C_{\rm P} R\co}$, which holds regardless of the atmosphere mass, as shown in Figure \ref{fig:profiles}.
-As the RCB is approached, \Eq{eq:Tconvdeep} is no longer valid, particularly for large envelope masses.
+Deep in convective zones, thermal energy is a fixed fraction of the gravitational potential energy, giving $T \propto r^{-1}$ and $P\propto r^{-1/\delad}$  as in \Eq{eq:deep}.
+This behavior is seen in \Fig{fig:profiles} for $r\ll R\cb$.
+Near the core, \Eq{eq:Tconvdeep} gives the core temperature, $T\co= G M\co / (C_{\rm P} R\co)$, unaffected by  the overlying atmosphere mass which does not contribute to the gravitational potential.
+Closer to the RCB, self-gravity is no longer negligible, particularly for large envelope masses.
 Instead, these high mass solutions show a slightly flatter profile in $T$ and also in $P \propto T^{1/\delad}$, as explained by \Eq{eq:TPsg}.}
@@
 In the higher mass solutions, the relatively low $P\cb$ values (and thus the relatively high luminosities) require an outward shift in $R\cb$, as shown in \Fig{fig:profiles}.
-\textbf{This shift does not occur in the analytic solution, where $R\cb$ continually decreases with mass (cf.
-Equations \ref{eq:crit} and \ref{eq:PcbRcb}).}
-
-The accelerated growth in the numerical model, as shown in the bottom panel of \Fig{fig:Ltplot}, is also a direct result of the higher cooling luminosities in the self-gravitating models.
-We see that self-gravity strongly affects atmospheric evolution, even well before the crossover mass, when $M_{\rm atm}=M\co$.
-\textbf{The early divergence between the analytic and numerical curves in Figure \ref{fig:Ltplot} shows that self-gravity becomes important even when the atmosphere is only a few percent as massive as the core.
-Our analytic model starts deviating from the numerical model at slightly larger $M_{\rm atm}/M\co$ closer to the star (not shown).
-However, we have found that self-gravity is always significant at $M_{\rm atm} \leq 10\% M\co$ in our regime of interest.}
+\textbf{This shift does not occur in the analytic solution, where $R\cb$ continually decreases with atmospheric mass (cf.
+Equations \ref{eq:PcbRcb} and \ref{eq:crit}).}
+
+The accelerated growth in the numerical model, as shown in the bottom panel of \Fig{fig:Ltplot}, is also a direct result of the higher cooling luminosities with self-gravity included.
+\textbf{Even when $M_{\rm atm} / M\co$ is only few percent, i.e.\ well before the crossover mass, the effect of self-gravity is  quite evident.
+Closer to the star, the effects of self-gravity are not as strong for low atmosphere masses.
+Nevertheless our models all show that self-gravity noticeably accelerates growth for $M_{\rm atm} > 0.1 M\co$.}
 
@@
 %We now use our structure and evolution models to calculate the necessary conditions for gas giant formation by core accretion.   
-\textbf{We now use our structure and evolution models to estimate the timescales and core masses necessary for giant planet formation, for our assumed disk model, nebular gas properties and dust grain opacity.}
-Adopting a median disk lifetime of 3 Myr, we examine the conditions for a fixed mass core to undergo runaway gas accretion within the lifetime of the gas disk.
-Our results for atmosphere growth times -- for a range of core masses and disk conditions -- are presented in \S\ref{sec:tcross}.
-Section \ref{sec:critcore} gives our results for critical core masses, the minimum masses needed to accrete a massive atmosphere within the disk lifetime.
+\textbf{We now use our structure and evolution models to estimate the timescales and minimum core masses for giant planet formation for a range of disk conditions and other model parameters.
+Our results for atmosphere growth times -- the time for a core of fixed mass to undergo runaway gas accretion -- are presented in \S\ref{sec:tcross}.
+Section \ref{sec:critcore} gives our results for critical core masses, the minimum values that trigger runaway atmospheric growth within a plausible disk lifetime, here 3 Myr.}
 
@@
 A lower $\mu$ gives longer growth times, because more cooling is required to compress the atmosphere.
-This effect is explained by the increase in the atmospheric scale height $\propto 1/\mu$ while the Bondi radius decreases, $\RB \propto \mu$.
-\textbf{The larger atmospheric scale height due to a lower $\mu$ results in a larger pressure depth at the RCB, $P\cb$, and thus a lower $R\cb$, cf.
-\Eq{eq:PcbRcb}.} For $\mu = 2.0$, which represents the idealized case of an H$_2$ atmosphere completely devoid of Helium, $t_{\rm run}$ increases by factors of $\sim 2 - 3$.
+\textbf{This effect is intuitive since the atmospheric scale height $\propto 1/\mu$ is more extended for lower $\mu$.
+The Bondi radius decreases as $\RB \propto \mu$ giving a smaller gravitational sphere of influence (and a weaker compression at $\RH$ when that is the more relevant scale).} For $\mu = 2.0$, which represents the idealized case of an H$_2$ atmosphere completely devoid of Helium, $t_{\rm run}$ increases by factors of $\sim 2 - 3$.
 Thus fairly drastic changes in atmospheric composition are required for $\mu$ to significantly affect core accretion timescales.
-\textbf{While we note that changes in the mean molecular weight of the gas also affect the adiabatic gradient $\delad$, we defer the discussion of the influence of $\delad$ on atmospheric evolution to Piso, Youdin \& Murray-Clay (in prep., hereafter Paper II).}
+\textbf{In principle, changes in the mean molecular weight of the gas also affect the EOS, including $\delad$, but such effects are not considered here, see \S\ref{sec:EOS}.}
 
@@
 \textbf{A larger disk lifetime further reduces $M_{\rm crit}$.
-Recent studies have shown that gas disks may live up to $\sim$$10-12$ Myr \citep{bell13}.
+Recent studies have shown that gas disks may live up to $\propto$$10-12$ Myr \citep{bell13}.
 As $M_{\rm crit} \sim t\di^{-3/5}$ (cf.
-Equation \ref{eq:tcoolan}), a disk lifetime of 10 Myrs decreases the critical core mass by a factor of two.}
+Equation \ref{eq:tcoolan}), a disk lifetime of 10 Myrs decreases the critical core mass by a factor of two.
+Of course we also require some fraction of the disk lifetime for the core to grow and possibly migrate.}
 
@@
 \section{Neglected Effects}\label{sec:neglected}
+\textbf{Since one goal of this paper was to obtain a detailed understanding of atmospheric evolution with a simple model, we have necessarily ignored many  effects of potential significance.
+We briefly address the most important of these and note that a followup work, Piso, Youdin \& Murray-Clay (in prep., hereafter Paper II), will extend these models to address some of them.}
 

### 2014-02-13 tex/PaperI/ms.tex
more edits, final significant ones hopefully

@@
 By isolating the process of KH contraction, we determine the minimum timescale for runaway atmospheric growth, and thus for giant planet formation by core accretion.
-Previous studies have considered this type of growth \citep{ikoma00, pn05}, which is an end-member case of phases 2 and 3 in more detailed quasi-static models.
+Previous studies have considered this type of growth \citep[hereafter I00; PN05, respectively]{ikoma00, pn05}, which is an end-member case of phases 2 and 3 in more detailed quasi-static models.
 Especially since phase 2 atmospheric growth is often the slowest stage of core accretion, understanding the fastest allowed rates of gas accretion is of crucial importance.
@@
 Equation \ref{eq:tcoolan}), a disk lifetime of 10 Myrs decreases the critical core mass by a factor of two.
-Of course we also require some fraction of the disk lifetime for the core to grow and possibly migrate.}
+Of course, some fraction of the disk lifetime must be allocated to the growth and possibly migration of the core.}
 
@@
 
-\textbf{Our results are consistent with those of previous core accretion studies.
-The solution of \citet{stevenson82}, which considers fully radiative atmospheres with constant opacity, predicts that the critical core mass is lower for lower envelope opacity and higher mean molecular weight of the nebular gas.
-Additionally, \citet{stevenson82} finds that the critical core mass decreases weakly with increasing envelope radius (via a logarithmic factor), i.e.
-with increasing orbital distance.
-Both of the above qualitative predictions are reproduced by our result.}
-
-\textbf{In the regime of no ongoing planetesimal accretion, \citet{ikoma00} find an approximate analytic expression for the envelope formation time (the equivalent of the runaway accretion time $t_{\rm run}$ in our models), }
-
+We can directly compare our results with other studies of protoplanetary atmospheric growth in the absence of planetesimal accretion, notably those in I00 and PN05.
+A major distinguishing feature of our study is that we study a range of disk conditions, while I00 and PN05 focused on growth at the location of Jupiter, 5.2 AU, in their disk models.
+Also both I00 and PN05 solved the full time dependent energy equation and considered more detailed EOSs, in addition to other detailed differences in model parameters.
+Despite these differences, the qualitative and quantitative agreement with our study is good.
+
+\textbf{I00 give an approximate analytic fit to their models' envelope formation time (the equivalent of the runaway accretion time $t_{\rm run}$ in our models), at $a=5.2$ AU, }
 \begin{equation}
 \label{eq:tauenv}
-\tau_{\rm env} \sim 3 \times 10^8 \Big(\frac{M\co}{M_{\oplus}}\Big)^{-2.5} \Big(\frac{\kappa}{1 \text{cm$^2$ s$^{-1}$}}\Big) \,\,\, \text{yr},
-\end{equation}
-\textbf{at $a=5.2$ AU, Jupiter's location.
-The result above reproduces the linear opacity dependence of the atmospheric evolutionary times in our model (see \S\ref{sec:timeev}).
-Moreover, we can approximate $t_{\rm run}$ in our model at 5 AU as}
+\tau_{\rm env} \sim 3 \times 10^8 \Big(\frac{M\co}{M_{\oplus}}\Big)^{-2.5} \Big(\frac{\kappa}{1 \text{cm$^2$ s$^{-1}$}}\Big) \,\,\, \text{yr}\, .
+\end{equation}
+\textbf{Our model reproduces the linear opacity dependence (see \S\ref{sec:timeev}), even though we use a temperature dependent opacity, \Eq{eq:opacitylaw}, instead of a constant $\kappa$.
+Our growth times at 5 AU,}
 \begin{equation}
@@
 \end{equation} 
-\textbf{for the opacity-law (\ref{eq:opacitylaw}). Our results are thus consistent with those of \citet{ikoma00} at 5 AU, both in terms of scaling and magnitude.}
+\textbf{agree well with the I00 results, both in magnitude and scaling with core mass.}
+
+One goal of our study was to explain the extended luminosity minimum in evolutionary models that characterizes phase 2 in core accretion models as described in the introduction.
+Previous work (see Figure 2 in I00 and Figures 3 and 4 in PN05) show that this luminosity minimum is an intrinsic feature of atmospheric cooling even in the absence of planetesimal accretion.
+We not only reproduce this effect in a simple model (see \Figs{fig:Ltplot}{fig:LtvsMopacity}) but show that self-gravity is the essential ingredient to produce a broad luminosity minimum well before the crossover mass.
 
 \textbf{Time-dependent studies that incorporate planetesimal accretion generally find larger formation timescales and critical core masses than those of this work.
+This result is expected since additional energy from planetesimal accretion limits the ability of the atmosphere to cool.
 \citet{pollack96} find an evolutionary time and crossover mass of $\sim$$7.5$ Myrs and $\sim$$16 M_{\oplus}$, respectively, for a MMSN disk model at 5.2 AU and interstellar grain opacity.
@@
 
-%\textbf{Lastly, the decline in $M_{\rm crit}$ with semimajor axis is also found by R06 for planets forming at $\gtrsim 5$ AU.
-A quantitative comparison with our results requires detailed calculations of the }
+\textbf{A direct comparison with studies that only include planetesimal accretion, and neglect atmospheric cooling, is difficult.
+Note however that I00 establishes the correspondence between the minimum luminosity in evolutionary models and the minimum planetesimal accretion rate needed in static models, which gives rise to the classical critical core mass of static models \citep{mizuno78, stevenson82}.}
+
+\textbf{In terms of modern static studies, our results complement (and borrow some tools from) R06 and \citet{Raf11}.
+These studies considered the disk radius dependence of core accretion for protoplanets that continuously accrete planetesimals.
+We show that the limits on core accretion at large radial distance which \cite{Raf11} (and references therein) claim can be overcome if planetesimal accretion shuts off and the atmosphere is allowed to cool.
+The plausibility of rapid core growth followed by negligible subsequent solid accretion admittedly remains highly uncertain, as discussed in more detail in the Summary.}
 
@@
 Rapid gas accretion onto low mass cores could explain the origin of distant directly imaged giant planets \citep{marois08, lagrange10}.
-However, our model does not address the details of how solid cores grow, and many possibilities exist.
-\textbf{Our brief treatment of core formation is purposeful, as there are large uncertainties in both planetesimal formation and the subsequent processes of core growth, particularly at large orbital distances.
+However, our model does not address the details of how solid cores grow, as many possibilities exist and many uncertainties remain.
+%\textbf{Our brief treatment of core formation is purposeful, as there are large uncertainties in both planetesimal formation and the subsequent processes of core growth, particularly at large orbital distances.
 Core formation in the outer disk is challenging due to large dynamical times, and cores are unlikely to reach critical masses at $\gtrsim 40-50$ AU under classic core accretion scenarios \citep{Raf11}.
@@
 Moreover, cores can form closer to the star, then migrate or be scattered outwards by other giants in the system \citep{ida13}.}
-
- For a giant planet to form with a core near the minimum mass, the core's growth must  first be rapid and then slow significantly, as in phases I and II, respectively, of \cite{pollack96}.
+For a giant planet to form with a core near the minimum masses we derive, core growth must  first be rapid and then slow significantly, as in phases 1 and 2, respectively, of \cite{pollack96}.
+
 Initial core growth must be fast, compared to the disk lifetime, to get a sufficiently massive core.
-Subsequent slowing of core growth is required to allow the atmosphere to cool and contract.
+\textbf{Such rapid core growth is possible in a variety of scenarios, including the fastest gas-free planetesimal accretion rates \citep{dones93} and -- probably more relevantly for gas rich disks --  the aerodynamic accretion of mm-m sized ``pebbles" and ``boulders" in gas disks \citep{ormel10, lambrechts12}.
+Moreover, cores could form rapidly closer to the star, then migrate or be scattered outwards by already formed giants \citep{ida13}.}
+
+\textbf{A stronger constraint is that} core growth subsequently slow severely, to allow the atmosphere to cool and contract.
 To be more quantitative, the minimum cooling luminosity in \Fig{fig:Ltplot}, $L\approx 3.5 \times 10^{24}\;\mathrm{erg s}^{-1}$, could be cancelled by low levels of heating from solid accretion.
-Specifically, a core mass doubling timescale of $\sim$$400$ Myr, or faster, would provide enough heating to stall atmospheric cooling and growth.
-
-Thus for giant planets to form around cores near $\MC$, core growth must stall quite severely.
-The accretion and scattering of local planetesimals, combined with the inward aerodynamic drift of smaller solids past the core, offers a plausible means to sufficiently isolate the core.
-
-On the other hand, non-negligible accretion of solids onto cores could continue for the lifetime of the gas disk.
-\textbf{For our fiducial disk model, the isolation mass, $M_{\rm iso} \propto a^{3/4}$, increases with orbital distance, and may thus become larger than $M_{\rm crit}$ at large separations.
-In this case, the ongoing accretion of solids increases the critical core mass by inhibiting atmospheric growth, and may prevent accumulation of a significant envelope before the disk dissipates.
-It follows that} relatively high mass cores would remain as ice giants, akin to Uranus and Neptune.
-\textbf{Non-traditional accretion models such as pebble accretion or scattering are thus needed in the outer disk in order to avoid slow growth to the isolation mass, which would stall atmospheric evolution.}  Ongoing direct imaging surveys and their successors \citep{hinz12, macintosh12} will help discriminate among the various planet formation pathways in the outskirts of protoplanetary disks.
-
-\acknowledgements{We thank Ruth Murray-Clay for detailed feedback and skillful proof reading.
+A core mass doubling timescale of $\sim$$400$ Myr, or faster, would thus provide enough heating to stall atmospheric cooling and growth.
+\textbf{An additional concern is that isolation masses tend to grow with disk radius, as $M_{\rm iso} \propto \varSigma_{\rm p}^{3/2} a^2 \propto a^{3/4}$ under the approximation that the surface density of accreted planetesimals ($\varSigma_{\rm p}$) scales with the gas \citep{youdin13}.
+While this behavior is nominally inconsistent with final core masses that decline with distance, the predictive power of the isolation mass is imperfect.
+For starters, the efficiency of planetesimal formation remains uncertain.
+Moreover the locality of core growth, which underlies the isolation mass, disappears when accreted solids drift and/or cores migrate significantly.
+}
+
+Thus while our calculations show that low mass cores can grow into gas giants in the outer disk, ongoing solid accretion could prevent significant atmospheric growth.
+In the Solar System, the ice giants Uranus and Neptune,  with core (here ice and rock) masses of $\sim 13 -15$ M$_\oplus$, argue for the latter possibility.
+Ongoing exoplanet imaging surveys and their successors \citep{hinz12, macintosh12, close14} will help discriminate among the various planet formation pathways in the outskirts of protoplanetary disks.
+
+\vspace*{5 mm}
+
+\acknowledgements{We thank the referee for a thorough report and Ruth Murray-Clay for detailed feedback and skillful proof reading.
 We appreciate valuable comments from Eugene Chiang and Scott Kenyon.

### 2014-02-13 tex/PaperI/ms.tex
extra stevenson 82 reference for mu

@@
 A lower $\mu$ gives longer growth times, because more cooling is required to compress the atmosphere.
-\textbf{This effect is intuitive since the atmospheric scale height $\propto 1/\mu$ is more extended for lower $\mu$.
+\textbf{This effect is both well established in core accretion studies \citep{stevenson82} and intuitive since the atmospheric scale height $\propto 1/\mu$ is more extended for lower $\mu$.
 The Bondi radius decreases as $\RB \propto \mu$ giving a smaller gravitational sphere of influence (and a weaker compression at $\RH$ when that is the more relevant scale).} For $\mu = 2.0$, which represents the idealized case of an H$_2$ atmosphere completely devoid of Helium, $t_{\rm run}$ increases by factors of $\sim 2 - 3$.

### 2014-02-14 tex/PaperI/ms.tex
updated tex and response, mostly minor edits

@@
 Previous studies have considered this type of growth \citep[hereafter I00; PN05, respectively]{ikoma00, pn05}, which is an end-member case of phases 2 and 3 in more detailed quasi-static models.
-Especially since phase 2 atmospheric growth is often the slowest stage of core accretion, understanding the fastest allowed rates of gas accretion is of crucial importance.
+Especially since phase 2 of atmospheric growth is often the slowest stage of core accretion, understanding the fastest allowed rates of gas accretion is of crucial importance.
 Our study develops simplified cooling models for the purposes of broadly exploring parameter space -- especially in the outer disk -- and developing a more detailed understanding of atmospheric accretion.
@@
 To facilitate comparison with our numerical results, \textbf{we rescale the analytic results as follows.
-Since growth times are too slow in the non-self-gravitating analytic model, we modify the runaway grown criterion to occur at an effective crossover mass $M_{\rm atm} = f M\co$ with $f < 1$.
-Specifically we choose $f = 0.13$, because with this value the analytic model gives the same critical core mass at 10 AU as the numerical model, for our standard choices of other parameters.
+Since growth times are too slow in the non-self-gravitating analytic model, we modify the runaway growth criterion to occur at an effective crossover mass $M_{\rm atm} = f M\co$, with $f < 1$.
+Specifically, we choose $f = 0.13$, because with this value the analytic model gives the same critical core mass at 10 AU as the numerical model, for our standard choices of other parameters.
 This prescription does not mean that runaway growth physically occurs at low atmosphere masses, nor does it cause perfect agreement between the models.
@@
 Closer to the star, the effects of self-gravity are not as strong for low atmosphere masses.
-Nevertheless our models all show that self-gravity noticeably accelerates growth for $M_{\rm atm} > 0.1 M\co$.}
+Nevertheless, all our models show that self-gravity noticeably accelerates growth for $M_{\rm atm} > 0.1 M\co$.}
 
@@
 \textbf{This effect is both well established in core accretion studies \citep{stevenson82} and intuitive since the atmospheric scale height $\propto 1/\mu$ is more extended for lower $\mu$.
-The Bondi radius decreases as $\RB \propto \mu$ giving a smaller gravitational sphere of influence (and a weaker compression at $\RH$ when that is the more relevant scale).} For $\mu = 2.0$, which represents the idealized case of an H$_2$ atmosphere completely devoid of Helium, $t_{\rm run}$ increases by factors of $\sim 2 - 3$.
+The Bondi radius decreases as $\RB \propto \mu$, giving a smaller gravitational sphere of influence (and a weaker compression at $\RH$ when that is the more relevant scale).
+The larger atmospheric scale height due to a lower $\mu$ also results in a larger pressure depth at the RCB, $P\cb$, and thus a lower $R\cb$, cf.
+\Eq{eq:PcbRcb}.} For $\mu = 2.0$, which represents the idealized case of an H$_2$ atmosphere completely devoid of Helium, $t_{\rm run}$ increases by factors of $\sim 2 - 3$.
 Thus fairly drastic changes in atmospheric composition are required for $\mu$ to significantly affect core accretion timescales.
-\textbf{In principle, changes in the mean molecular weight of the gas also affect the EOS, including $\delad$, but such effects are not considered here, see \S\ref{sec:EOS}.}
+\textbf{In principle, changes in the mean molecular weight of the gas also affect the EOS, including $\delad$, but such effects are not considered here (see \S\ref{sec:EOS}).}
 
@@
 
-We can directly compare our results with other studies of protoplanetary atmospheric growth in the absence of planetesimal accretion, notably those in I00 and PN05.
-A major distinguishing feature of our study is that we study a range of disk conditions, while I00 and PN05 focused on growth at the location of Jupiter, 5.2 AU, in their disk models.
-Also both I00 and PN05 solved the full time dependent energy equation and considered more detailed EOSs, in addition to other detailed differences in model parameters.
-Despite these differences, the qualitative and quantitative agreement with our study is good.
+\textbf{We can directly compare our results with other studies of protoplanetary atmospheric growth in the absence of planetesimal accretion, notably those in I00 and PN05.
+A major distinguishing feature of our study is that we explore a range of disk conditions, while I00 and PN05 focus on growth at the location of Jupiter, 5.2 AU, in their disk models.
+Moreover, both I00 and PN05 solve the full time dependent energy equation and consider more detailed EOSs, in addition to other detailed differences in model parameters.
+Despite these differences, the qualitative and quantitative agreement with our study is good.}
 
@@
 
-One goal of our study was to explain the extended luminosity minimum in evolutionary models that characterizes phase 2 in core accretion models as described in the introduction.
-Previous work (see Figure 2 in I00 and Figures 3 and 4 in PN05) show that this luminosity minimum is an intrinsic feature of atmospheric cooling even in the absence of planetesimal accretion.
-We not only reproduce this effect in a simple model (see \Figs{fig:Ltplot}{fig:LtvsMopacity}) but show that self-gravity is the essential ingredient to produce a broad luminosity minimum well before the crossover mass.
+\textbf{One goal of our study was to explain the extended luminosity minimum in evolutionary models that characterizes phase 2 in core accretion models, as described in the introduction.
+Previous work (see Figure 2 in I00 and Figures 3 and 4 in PN05) shows that this luminosity minimum is an intrinsic feature of atmospheric cooling even in the absence of planetesimal accretion.
+We not only reproduce this effect in our simplified model (see \Figs{fig:Ltplot}{fig:LtvsMopacity}), but we also show that self-gravity is the essential ingredient to produce a broad luminosity minimum well before the crossover mass.}
 
@@
 \textbf{A direct comparison with studies that only include planetesimal accretion, and neglect atmospheric cooling, is difficult.
-Note however that I00 establishes the correspondence between the minimum luminosity in evolutionary models and the minimum planetesimal accretion rate needed in static models, which gives rise to the classical critical core mass of static models \citep{mizuno78, stevenson82}.}
+Note, however, that I00 establishes the correspondence between the minimum luminosity in evolutionary models and the minimum planetesimal accretion rate needed in static models, which gives rise to the classical critical core mass of static models \citep{mizuno78, stevenson82}.}
 
 \textbf{In terms of modern static studies, our results complement (and borrow some tools from) R06 and \citet{Raf11}.
-These studies considered the disk radius dependence of core accretion for protoplanets that continuously accrete planetesimals.
-We show that the limits on core accretion at large radial distance which \cite{Raf11} (and references therein) claim can be overcome if planetesimal accretion shuts off and the atmosphere is allowed to cool.
-The plausibility of rapid core growth followed by negligible subsequent solid accretion admittedly remains highly uncertain, as discussed in more detail in the Summary.}
+These studies consider the disk radius dependence of core accretion for protoplanets that continuously accrete planetesimals.
+We show that the limits on core accretion at large radial distance claimed by \cite{Raf11} (and references therein) can be overcome if planetesimal accretion shuts off and the atmosphere is allowed to cool.
+Nevertheless, the plausibility of rapid core growth followed by negligible subsequent solid accretion admittedly remains highly uncertain, as discussed in more detail in \S\ref{sec:conclusions}.}
 
@@
 \textbf{Since one goal of this paper was to obtain a detailed understanding of atmospheric evolution with a simple model, we have necessarily ignored many  effects of potential significance.
-We briefly address the most important of these and note that a followup work, Piso, Youdin \& Murray-Clay (in prep., hereafter Paper II), will extend these models to address some of them.}
+We briefly address the most important of these and note that a followup work, Piso, Youdin \& Murray-Clay (in prep., hereafter Paper II), will extend out current models to address some of these effects.}
 
@@
 For starters, the efficiency of planetesimal formation remains uncertain.
-Moreover the locality of core growth, which underlies the isolation mass, disappears when accreted solids drift and/or cores migrate significantly.
+Moreover, the locality of core growth, which underlies the isolation mass, disappears when accreted solids drift and/or cores migrate significantly.
 }

### 2014-02-14 tex/PaperI/ms.tex
'highly uncertain' -> 'uncertain', I didn't need to be overly negative there

@@
 We show that the limits on core accretion at large radial distance claimed by \cite{Raf11} (and references therein) can be overcome if planetesimal accretion shuts off and the atmosphere is allowed to cool.
-Nevertheless, the plausibility of rapid core growth followed by negligible subsequent solid accretion admittedly remains highly uncertain, as discussed in more detail in \S\ref{sec:conclusions}.}
+Nevertheless, the plausibility of rapid core growth followed by negligible subsequent solid accretion admittedly remains uncertain, as discussed in more detail in \S\ref{sec:conclusions}.}
 

### 2014-02-14 tex/PaperI/ms.tex
edit mu description.  Done?

@@
 \begin{subeqnarray} 
-P_M &\equiv& {4 \delad^{3/2} \over 5 \pi^2 \sqrt{\chi} } {G M\co^2 \over {\RB'}^4}\,  \\
+P_M &\equiv& {4 \delad^{3/2} \over 5 \pi^2 \sqrt{\chi} } {G M\co^2 \over {\RB'}^4} \,  \label{eq:PM} \\
 \xi &\equiv& \sqrt{\ln[ P\cb/(\theta P_{\rm d})]} \slabel{eq:xidef}\, .%P_M \approx {1.85 \over \sqrt{\chi}} {G M\co^2 \over {\RB'}^4}\, .
@@
 \textbf{This effect is both well established in core accretion studies \citep{stevenson82} and intuitive since the atmospheric scale height $\propto 1/\mu$ is more extended for lower $\mu$.
-The Bondi radius decreases as $\RB \propto \mu$, giving a smaller gravitational sphere of influence (and a weaker compression at $\RH$ when that is the more relevant scale).
-The larger atmospheric scale height due to a lower $\mu$ also results in a larger pressure depth at the RCB, $P\cb$, and thus a lower $R\cb$, cf.
-\Eq{eq:PcbRcb}.} For $\mu = 2.0$, which represents the idealized case of an H$_2$ atmosphere completely devoid of Helium, $t_{\rm run}$ increases by factors of $\sim 2 - 3$.
+Moreover, the Bondi radius decreases as $\RB \propto \mu$, giving a smaller gravitational sphere of influence (and a weaker compression at $\RH$ when that is the more relevant scale).
+To see how this affects cooling times, note that the characteristic RCB depth near runaway scales as $P\cb \sim P_{\rm M} \propto \mu^{-4}$ from \Eq{eq:PM}.
+Thus $L \propto 1/P\cb \propto \mu^4$ explains the trend of slower cooling for lower $\mu$.}
+%he larger atmospheric scale height due to a lower $\mu$ also results in a larger pressure depth at the RCB, $P\cb$, and thus a lower $R\cb$, cf. \Eq{eq:PcbRcb}.}
+
+For $\mu = 2.0$, which represents the idealized case of an H$_2$ atmosphere completely devoid of Helium, $t_{\rm run}$ increases by factors of $\sim 2 - 3$.
 Thus fairly drastic changes in atmospheric composition are required for $\mu$ to significantly affect core accretion timescales.
@@
 \section{Neglected Effects}\label{sec:neglected}
-\textbf{Since one goal of this paper was to obtain a detailed understanding of atmospheric evolution with a simple model, we have necessarily ignored many  effects of potential significance.
-We briefly address the most important of these and note that a followup work, Piso, Youdin \& Murray-Clay (in prep., hereafter Paper II), will extend out current models to address some of these effects.}
+\textbf{Since one goal of this paper was to obtain a detailed understanding of atmospheric evolution with a simple model, we have necessarily ignored   effects of potential significance.
+We briefly address the most important of these and note that a followup work, Piso, Youdin \& Murray-Clay (in prep., hereafter Paper II), will extend our current models to address some of these effects.}
 

### 2014-02-18 tex/draft_paper2/draft_paper2.tex
Paper II draft

@@
 
-Variations in the adiabatic gradient $\delad$ due to partial dissociation and molecular rotational effects (see \S\ref{deladtable}) affect atmospheric evolution by yielding: (1) a lower envelope luminosity $L$, and (2) a larger amount of radiated energy per unit of accreted mass $-dE/dM$, when compared to an ideal gas.
+Variations in the adiabatic gradient $\delad$ due to partial dissociation and fractional occupation of $\rm{H}_2$ rotational states (see \S\ref{deladtable}) affect atmospheric evolution by yielding: (1) a lower envelope luminosity $L$, and (2) a larger amount of radiated energy per unit of accreted mass $-dE/dM$, when compared to an ideal gas.
 As a result, the rate of change in atmospheric mass,
@@
 (cf.
-Equation \ref{eq:coolingglobal} and ignoring surface terms) is lower than in the ideal case.
+Equation \ref{eq:coolingglobal} and ignoring surface terms, which only become significant in the late stages of atmospheric evolution) is lower than in the ideal case.
 The $\delad$ variations thus slow down gas accretion and increase the growth time of the atmosphere, and hence the critical core mass.
 
-To understand the separate effects of dissociation and rotation on atmospheric growth, we consider the resulting EOS deviations from an ideal gas independently.  
-%{\bf In what follows it may be clearer to start with the motivation: isolating the two EOS effects, i.e.
+Modifications in the EOS affect $L$ and $-dE/dM$ because both of these quantities depend on the global structure of the envelope.
+From \Eq{eq:delrad} applied at the RCB where $\delad=\delrad$, and for a fixed atmospheric mass, the luminosity emerging from the convective interior scales as $L \propto T\cb^4/P\cb$.
+We have shown in Paper I that the outer radiative layer of the atmosphere is nearly isothermal, and that $T\cb$ is only a factor unity correction from the disk temperature, while the pressure in the radiative region increases exponentially with depth.
+It follows that $L$ primarily scales as $1/P\cb$.
+This pressure depth at the RCB depends on both the interior structure of the atmosphere, which is affected by dissociation (at high temperatures), and on the matching to the nebula through the radiative region, where the gas is non-ideal due to the variable occupation of $\rm{H}_2$ rotational states (at low temperatures).
+Similarly, the EOS modifications due to dissociation and spin isomers affect the way in which the total energy $E(r)$ is distributed throughout the envelope, and thus $-dE/dM$.
+
+Since both dissociation at high temperatures and variable occupation of rotation states at low temperatures affect atmospheric structure and evolution, as discussed above, it is helpful to separate these effects and study them independently.
+
+%
+%the amount of energy radiated per unit of accreted mass, $-dE/dM$, depends on how the total energy $E(r)$ is distributed throughout the entire envelope, and thus on the 
+%
+%
+% is thus affected by both dissociation in the deep interior and the $\rm{H}_2$ partially excited rotational states at the top of the envelope. 
+%
+%In order to understand how $L$ and 
+%
+%It follows that both $L$ and $-dE/dM$ depend on the global structure of the atmosphere, and therefore on the EOS.  
+%
+% Understanding how variations in the adiabatic gradient due to a realistic EOS affect the energy behavior thus requires
+
+%To understand the separate effects of dissociation and rotation on atmospheric growth, we consider the resulting EOS deviations from an ideal gas independently.  
+%%{\bf In what follows it may be clearer to start with the motivation: isolating the two EOS effects, i.e.
 dissociation and rotation.
@@
 Currently it's less clear and more drawn out where this is going.}
-In order to explain how quantum rotational states at low temperatures affect atmospheric growth, we 
-% {\bf (I don't see how this is a thought experiment, or why we have to imagine.  I would just describe the EOSs)}: 
-imagine \textbf{(I used 'imagine' to make it clear that this is just an exercise to isolate the EOS effects, rather than a physically motivated assumption)} that the EOS deviates from a polytrope only in the upper atmosphere where temperatures are low, and that the gas is ideal and polytropic deep in the envelope.
+%In order to explain how quantum rotational states at low temperatures affect atmospheric growth, we 
+%% {\bf (I don't see how this is a thought experiment, or why we have to imagine.  I would just describe the EOSs)}: 
+%imagine \textbf{(I used 'imagine' to make it clear that this is just an exercise to isolate the EOS effects, rather than a physically motivated assumption)} that the EOS deviates from a polytrope only in the upper atmosphere where temperatures are low, and that the gas is ideal and polytropic deep in the envelope.
 Conversely, we study the effects of dissociation at high temperatures by assuming that the EOS is realistic at the bottom of the atmosphere where temperatures are high, but that the gas is polytropic in the outer regions.
@@
 %By separately comparing curves (1) and (2), and (1) and (3)  while the difference between curves (1) and (3) accounts for hydrogen dissociation. 
-Both dissociation and rotation result in slower cooling for both the high-$T$ and low-$T$ realistic EOSs. 
+Both the high-$T$ and low-$T$ EOSs yield slower cooling when compared to the ideal gas polytrope --- due to dissociation for the high-$T$ EOS, and due to fractional occupation of $\rm H_2$ rotational states for the low-$T$ EOS.
+%Both dissociation and fractional occupation of $\rm H_2$ rotational states result in slower cooling for both the high-$T$ and low-$T$ realistic EOSs. 
 %decrease the atmospheric accretion rate $\dot{M}$ and thus
 %This follows from Equation (\ref{eq:dMdt}), with rotation decreasing $L$ and dissociation increasing $-dE/dM$, as we show further. 
-Noting the logarithmic scale in Figure \ref{fig:tplotall}, we see that the combined effect of rotation and dissociation is significantly greater than either individually.
+Noting the logarithmic scale in Figure \ref{fig:tplotall}, we see that the combined effect of fractional occupation of rotational states and dissociation is significantly greater than either individually.
 %consistent with Equation \ref{eq:dMdt}, since dissociation increases $-dE/dM$ while molecular rotation decreases $L$.
@@
 \caption{Elapsed time to grow a planet of total mass (core + atmosphere) for a variety of EOS combinations (see text), for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=10 M_{\oplus}$.
-Both hydrogen dissociation at high temperatures deep in the atmosphere and molecular rotational effects at low temperatures in the outer envelope result in slower cooling when compared to an ideal gas polytrope.}
+Both hydrogen dissociation at high temperatures deep in the atmosphere and fractional occupation of $\rm H_2$ rotational states at low temperatures in the outer envelope result in slower cooling when compared to an ideal gas polytrope.}
 \label{fig:tplotall}
@@
 The location of the RCB is marked.
-Quantum rotational states at low temperatures in the outer region of the atmosphere result in a deeper RCB.
+The variable occupation of rotational states at low temperatures in the outer region of the atmosphere results in a deeper RCB.
 Upper-right panel:  Luminosity evolution as a function of total mass.
-Rotational effects at low temperatures near the top of the atmosphere result in a lower luminosity for the realistic EOS (see text).
-Middle-left panel: Total local energy $e \rho r^3$ radial profile for a total mass of $12 M_{\oplus}$.
+Fractional occupation of $\rm H_2$ rotational states at low temperatures near the top of the atmosphere results in a lower luminosity for the realistic EOS (see text).
+Middle-left panel: Total energy at scale $r$, $-e \rho r^3$, profile for a total mass of $12 M_{\oplus}$.
 Hydrogen dissociation deep in the atmosphere causes more of the energy to be concentrated at the bottom of the envelope when compared to an ideal gas polytrope.
@@
 Lower-right panel: the atmospheric accretion rate $\dot{M}$ as a function of total mass.
-Both dissociation and rotation reduce $\dot{M}$ and thus slow down atmospheric growth.}
+Both dissociation and fractional occupation of rotational states reduce $\dot{M}$ and thus slow down atmospheric growth.}
 \label{fig:all_plot}
@@
 
-The upper-right panel of Figure \ref{fig:all_plot} shows the luminosity evolution with mass for our model EOSs, for a $M\co=10 M_{\oplus}$ core at 10 AU.
-The realistic EOS effects in the outer envelope decrease the atmospheric luminosity $L$ when compared to an ideal gas polytrope.
-We can explain this analytically.
-By substituting Equation (\ref{eq:structa}) into Equation (\ref{eq:structc}) in the convective region of the atmosphere, and assuming the nebular gas obeys the ideal gas law, we can write hydrostatic balance along an adiabat as
-
-\begin{equation}
-\label{eq:dTdr}
-\frac{d T}{d r}=-\frac{\delad G m(r)}{\mathcal{R} r^2},
-\end{equation}
-where $\mathcal{R}$ is the specific gas constant.
-This result is valid for the $\delad=0.3$ polytrope.
-As the low$-T$ realistic EOS only slightly (see Figure \ref{fig:deladmap}) deviates from an ideal gas in the outermost convective layers, we can assume that the gas described by the low-$T$ EOS can be approximated as ideal throughout the convective interior.
-The low-$T$ EOS thus also obeys Equation (\ref{eq:dTdr}) with $\delad \approx const.$.
-We further neglect self-gravity in Equation (\ref{eq:dTdr}) and set $m(r)=M\co$.
-Integrating Equation (\ref{eq:dTdr}) from the core to the RCB under these assumptions yields
-\begin{equation}
-\label{eq:dTdrint}
-T\cb=T\co - \frac{\delad G M\co}{\mathcal{R}} \Big(\frac{1}{R\co} - \frac{1}{R\cb}\Big).
-\end{equation}  
-Deep in the convective interior, $T \gg T\cb$ and $r \ll R\cb$. From virial equilibrium, the temperature in the inner atmosphere can thus be expressed as
-\begin{equation}
-\label{eq:virialT}
-T(r) \simeq \frac{\nabla_{\rm ad} G M\co}{\mathcal{R} r}.
-\end{equation}
-%The temperature scale height deep in the convective interior is $H_T \sim r$, with $dT/dr \sim -T/H_T \sim -T/r$.
-From Equation (\ref{eq:structc}) it immediately follows that the pressure scale height, for which $dP/dr \sim -P/H_P$, is $H_P \sim r \delad$.
-Deep in the interior and for $m(r)=M\co$, hydrostatic balance (\ref{eq:structa}) thus becomes
-
-%\begin{equation}
-%\label{eq:structint}
-%\frac{P}{r \delad} \sim -\frac{G M\co}{r^2}. 
-%\end{equation}
-%From this and the definition of the isothermal sound speed
-%\begin{equation}
-%\label{eq:cs}
-%c_s^2 = \frac{P}{\rho} =\mathcal{R} T, 
-%\end{equation} 
-%with both expressions applied at the core where $r=R\co$, $T=T\co$ and $\delad=\nabla_{\rm ad, c}$, we find
-Applying equation (\ref{eq:virialT}) at the core, where $T=T\co$, $r=R\co$ and $\delad=\nabla_{\rm ad, c}$, gives
-\begin{equation}
-\label{eq:Tc}
-T\co \simeq \frac{\nabla_{\rm ad, c} G M\co}{\mathcal{R} R\co}.
-\end{equation}
-As the adiabatic gradient at the core is the same for both the ideal gas polytrope and the low$-T$ realistic EOS, $T\co$ has the same value for both EOSs.
-This is shown in Figure \ref{fig:all_plot}, lower-left panel.
-Substituting Equation (\ref{eq:Tc}) into Equation (\ref{eq:dTdrint}) applied at the RCB gives
-\begin{equation}
-\label{eq:Tcb}
-T\cb \simeq \frac{\nabla_{\rm ad, RCB} G M\co}{\mathcal{R} R\cb},
-\end{equation}
-%We note that the expression above and the definition of the pressure scale height $H_P \equiv c_s^2 r^2/(G M\co)$ imply that the pressure scale height at the RCB is the same as in the deep interior, $H_P \sim \delad r$.
-and thus
-\begin{equation}
-\label{eq:rcb}
-R\cb \propto \frac{\nabla_{\rm ad, RCB}}{T\cb}.
-\end{equation}
-An approximate expression for $T\cb$ as a function of the adiabatic gradient and disk temperature can be derived using Equation (\ref{eq:delrad}) applied at the RCB (where $\delad=\delrad$) and $\delrad=d \ln T/ d \ln P$ (see Paper I for details).
-For $\beta=2$ in the opacity law (\ref{eq:opacitylaw}), $T\cb \simeq T\di (1- 2 \nabla_{\rm ad, RCB})^{-1/2}$, and thus
-\begin{equation}
-\label{eq:rcbdel}
-R\cb \propto \frac{\nabla_{\rm ad, RCB}}{(1- 2 \nabla_{\rm ad, RCB})^{-1/2}}.
-\end{equation}
-For the ideal gas polytrope $\nabla_{\rm ad, RCB}=0.3$, and for the low-$T$ realistic EOS $\nabla_{\rm ad, RCB}\approx 0.25$.
-For these particular choices, Equation (\ref{eq:rcbdel}) yields a smaller $R\cb$ for the low$-T$ realistic EOS.
-As such, molecular rotation increases the physical extent, and thus the optical depth of the radiative zone.
-%Qualitatively, this can be explained through a simple analytic argument: for a non-self-gravitating atmosphere composed of an ideal gas, the temperature profile in convective regions scales as $T(r) \propto \delad/r$ (see Paper I for details; it can also be derived from Equations \ref{eq:structa} and \ref{eq:structc} applied in convective regions).
-Applying the expression above at the RCB yields $T\cb \propto \delad\cb/ R\cb$.
-As the outer radiative layer of the atmosphere is nearly isothermal, $T\cb \approx T\di=const.$, to order unity corrections.
-It easily follows that $R\cb \propto \deladcb$   Figure \ref{fig:all_plot_temp}, middle-left panel shows
-This is directly correlated to a larger pressure depth of the RCB for the low-$T$ realistic EOS, as shown in the upper-left panel of Figure \ref{fig:all_plot} for a total planet mass of $12 M_{\oplus}$.
-As $L \propto 1/P\cb$ (cf.
-Equation \ref{eq:delrad}), the deeper RCB results in a lower $L$ and thus $\dot{M}$ (Figure \ref{fig:all_plot}, lower-right panel).
-
-%directly correlated to the depth of the radiative region of the envelope, which is shown in the upper-left panel of Figure \ref{fig:all_plot} for a total planet mass of $12 M_{\oplus}$.
-We see that the realistic EOS for low temperatures (2) results in a deeper location of the RCB when compared to the ideal gas polytrope (1).
-{\bf This is an accurate observation.
-%The deeper question is why does the EOS change result in a deeper RCB?
-This should be possible to understand, right?}  This deeper RCB translates into a larger number of steps that photons need to diffuse to escape from the RCB, and therefore a lower luminosity (cf.
-Equation \ref{eq:delrad}).
-{\bf Better to just say it's an optical depth effect than reexplaining how radiative diffusion works.
-Also it's the pressure depth (not the radius) that correlates more closely with optical depth.
-A plot of $P(r)$ might thus help as noted in fig caption.}
-
-%We see that the lower luminosity of the fully realistic EOS (4) when compared to the polytrope (1) is due to the realistic EOS in the outer atmosphere (2). T
-
-%We found that the RCB for the fully realistic EOS (4) is roughly at the same location as the RCB for the realistic EOS at low temperatures (2) for low atmosphere masses, and it moves further out relatively as the atmosphere mass increases.
-This explains our choice of $M_{\rm tot}=12 M_{\oplus}$ as a representative case to illustrate the effect of spin isomers.
-Envelope growth is slowest at this relative atmosphere-to-core mass (see PY13)...
-
-%Dissociation deep in the planet's atmosphere increases the amount of energy per unit mass that needs to be radiated by the envelope, which further slows down growth.
-
-%{\bf Why is this true?  Dissociation absorbs energy which by itself decreases the amount of energy that needs to be radiated.}
-
-Figure \ref{fig:all_plot}, middle-left panel, shows a radial profile of the total local energy $-e \rho r^3$, with $e$ the total specific energy, for the same example planet.
-\textbf{(the core + total mass choices need to be specified; "the same example planet" is shorter than "for a planet with Mc = ...", and clear enough I think.)}
+In Figure \ref{fig:all_plot}, middle-right and upper-right panels, we show the evolution with planet mass of $-dE/dM$ and $L$, respectively, for the EOSs described above.
+The EOS variations (i.e., the realistic EOS at all $T$) reduce $L$ by a factor of $\sim$$2$ and $-dE/dM$ by a factor of $\sim$$5$, when compared to the ideal gas polytrope.
+Atmospheric growth is therefore primarily affected by the changes in the energy profile $E(r)$ due to the realistic EOS, with the luminosity $L$ having a secondary effect.
+In what follows we explore each contribution separately.
+
+\subsection{Role of $-dE/dM$}
+
+Figure \ref{fig:all_plot}, middle-left panel, shows a radial profile of the total (internal + gravitational) energy at scale $r$, $-e \rho r^3$, with $e$ the total specific energy.
 %the cumulative total energy (internal and gravitational) profile as a function of the radial coordinate, \sout{for the same example planet}.
@@
 However, the energy of the high-$T$ realistic EOS and of the realistic EOS at all $T$ is much larger in magnitude (more negative) near the core when compared to an ideal gas polytrope.
-This is due to the fact that $\delad$ decreases in dissociation regions (Figure \ref{fig:deladmap}), which results in a shallower temperature gradient and thus lower temperatures near the core.
-Figure \ref{fig:all_plot}, bottom-left panel shows the radial temperature profile (for $M\co=10 M_{\oplus}$ and $M_{\rm tot}=12 M_{\oplus}$), with temperatures lower near the core for the high-$T$ realistic EOS.
-As the internal energy $U \propto T$, dissociation decreases $U$ and thus the total energy deep in the interior.
+This is due to the fact that $\delad$ decreases in dissociation regions (Figure \ref{fig:deladmap}).
+Dissociation increases the number of particles, which tends to increase entropy.
+In order for entropy to stay constant, the temperature must therefore decrease, which also lowers the adiabatic gradient.
+A lower $\delad$ results in a shallower temperature gradient in dissociation regions, and hence in lower temperatures near the core (cf.
+Figure \ref{fig:all_plot}, bottom-left panel).
+As the specific internal energy $u(r) \propto T(r)$, dissociation decreases $u(r)$ and thus the total energy deep in the interior.
 This explains the local energy behavior shown in Figure \ref{fig:all_plot}, middle-left panel.
@@
 The energy behavior can also be explained qualitatively through a simple analytic argument: the density profile in an adiabatic, non-self-gravitating atmosphere composed of an ideal gas scales as $\rho(r) \propto r^{-1/\delad+1}$, and the total specific energy is $e(r) \propto 1/r$ (see Paper I).
-Thus $e \rho r^3 \sim r^{-1/\delad+3}$.
-It follows that adiabats with lower $\delad$ have their energy more tightly packed towards the interior of the envelope, which is the case during dissociation.
+Thus $- e \rho r^3 \sim r^{-1/\delad+3}$.
+It is easy to see that the total energy is concentrated near the core if $\delad<1/3$, and at at the outer boundary otherwise.
+In our regime of interest, $\delad<1/3$, on average, hence the energy always peaks near the core (cf.
+Figure \ref{fig:all_plot}, middle-left panel).
+More importantly, the $e$ dependence on $\delad$ also implies that adiabats with lower $\delad$ have their energy more tightly packed towards the interior of the envelope, which is the case during dissociation.
 This is consistent with the energy behavior in Figure \ref{fig:all_plot}, middle-left panel.
@@
 This implies a larger $-dE/dM$ during atmospheric evolution (middle-right panel of Figure \ref{fig:all_plot}), and thus a lower $\dot{M}$ for the high-T EOS (lower-right panel of Figure \ref{fig:all_plot}).
+
+While dissociation at high temperatures is primarily responsible for the larger $-dE/dM$ for a realistic EOS, the variable occupation of $\rm H_2$ rotational states at low temperatures also increases $-e \rho r^3$ (Figure \ref{fig:all_plot}, middle-left panel), and hence $-dE/dM$ (Figure \ref{fig:all_plot}, middle-right panel), albeit less significantly.
+As both dissociation and spin isomers modify the energy profile throughout the envelope, the rate of change in atmospheric mass $\dot{M}$, and thus the atmospheric cooling time, is primarily affected by changes in $-dE/dM$ rather than $L$ (see Equation \ref{eq:dMdt}).
+
+%However, the fact that $-dE/dM$ is affected by both dissociation and spin isomers results in it
 
@@
 
+\begin{figure*}[tb]
+\centering
+\includegraphics[width=\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/varying_delad_4panel_3.pdf}
+%%\vspace{-0.5in}
+\caption{We explore atmosphere growth around a core with $M\co=10 M_{\oplus}$ forming at 10 AU, and with total mass (core + atmosphere) 12 $M_{\oplus}$, for the EOS choice of \Eq{eq:deladvar} and various $\nabla_{\rm ad,1}$ and $\nabla_{\rm ad,2}$.
+In each panel, the circles mark the RCB location.
+The horizontal dash-dotted lines show the temperature $T_{\rm switch}=500$ K at which the adiabatic gradient changes.
+Upper-left panel: radial temperature profile for the same $\nabla_{\rm ad,1}$ in the interior and various $\nabla_{\rm ad,2}$ in the exterior.
+The analytic profile of Equation (\ref{eq:Tdeep}) is over-plotted for comparison (\textit{dotted line}) in regions of agreement.
+The analytic temperature correction at the RCB (\textit{dashed}) is in good agreement with the numerical results.
+A larger $\nabla_{\rm ad,2}$ yields a deeper radiative region.
+Lower-left panel: radial pressure profile for the same $\nabla_{\rm ad,1}$ in the interior and various $\nabla_{\rm ad,2}$ in the exterior.
+The dashed black line is the analytic prediction for the pressure profile in the radiative zone, which agrees with the numerical results.
+Upper-right panel: radial temperature profile for various $\nabla_{\rm ad,1}$ in the interior and the same $\nabla_{\rm ad,2}$ in the exterior.
+The analytic profile of Equation (\ref{eq:Tdeep}) is over-plotted for comparison (\textit{dotted line}) in regions of agreement.
+The analytic temperature correction at the RCB (\textit{dashed}) is in good agreement with the numerical results.
+A lower $\nabla_{\rm ad,1}$ yields a more shallow radiative region.
+Lower-right panel: radial pressure profile for various $\nabla_{\rm ad,1}$ in the interior and the same $\nabla_{\rm ad,2}$ in the exterior.
+The dashed black line is the analytic prediction for the pressure profile in the radiative zone, which agrees with the numerical results.}
+\label{fig:varying_delad}
+\end{figure*}
+
+\subsection{Role of $L$}
+
+The EOS variations in the cold outer envelope affect $\dot{M}$ by decreasing the atmosphere's luminosity compared to an ideal gas polytrope (Figure \ref{fig:all_plot}, upper-right panel, for the same parameters as Figure \ref{fig:tplotall}).
+From Figures \ref{fig:deladmap} and \ref{fig:all_plot}, bottom-left panel, we see that the adiabatic gradient of the low-$T$ realistic EOS increases as the temperature decreases throughout the convective zone, then becomes lower near the RCB.
+This results in a larger adiabatic gradient, on average, for the low-$T$ realistic EOS in the outer parts of the convective region of the atmosphere.
+
+To understand how a variable, but overall larger $\delad$ in the outer atmosphere affects evolution, we study the simplified problem of atmospheres composed of an ideal gas with adiabatic gradient
+
+\begin{equation}
+\label{eq:deladvar}
+\delad = \left\{
+\begin{array}{l l}
+\nabla_{\rm ad, 1}, & \quad T > 500 \text{ K} \\
+\nabla_{\rm ad, 2}, & \quad T < 500 \text{ K}
+\end{array} \right ,
+\end{equation}   
+where $\nabla_{\rm ad, 1}$ and $\nabla_{\rm ad,2}$ are constant.
+Figure \ref{fig:varying_delad}, lower-left panel, shows the radial pressure profile for $\nabla_{\rm ad,1}=0.3$ and various $\nabla_{\rm ad,2}$, for the same parameters as Figures \ref{fig:tplotall} and \ref{fig:all_plot}.
+The numerical solutions reproduce the analytic expectation of a nearly exponential pressure profile in the radiative region, $P(r) \propto \exp(\RB/r)$.
+Thus $L \propto 1/P\cb \propto 1/\exp(\RB/R\cb)$ is directly correlated to the physical depth of the RCB.
+
+The location of the RCB is set by the matching between the convective interior and the outer radiative zone. Deep in the atmosphere, where $r \ll R\cb$, the temperature profile is given by
+
+\begin{equation}
+\label{eq:Tdeep}
+T(r) \approx \frac{\nabla_{\rm ad,1} G M\co}{\mathcal{R} r}
+\end{equation}   
+(see Paper I for a derivation), consistent with the upper-left panel of Figure \ref{fig:varying_delad}.
+Even though $\nabla_{\rm ad,1}$ is an order unity coefficient, the temperature scaling with $\nabla_{\rm ad,1}$ in \Eq{eq:Tdeep} is exact, as shown in the upper-right panel of Figure \ref{fig:varying_delad} for various $\nabla_{\rm ad,1}$.
+It follows that the radial temperature profile in the atmospheric interior is primarily set by the inner adiabatic gradient.
+Thus the radius $R_{\rm switch}$ at which the adiabatic gradient changes, i.e.
+where $T(R_{\rm switch})=500$ K, shows little variation with $\nabla_{\rm ad,2}$.
+In contrast with the deep interior, we cannot rigorously apply \Eq{eq:Tdeep} at $R_{\rm switch}$, although the approximation is still valid to order unity.
+
+The RCB temperature $T\cb$ is an order unity correction from the disk temperature that depends on the adiabatic gradient in the outer envelope, $\nabla_{\rm ad,2}$.
+We have showed in Paper I that $T\cb \simeq T\di (1- 2 \nabla_{\rm ad, 2})^{-1/2}$, for $\beta=2$ in the opacity-law (\ref{eq:opacitylaw}), modestly increases with $\nabla_{\rm ad,2}$, consistent with the upper-left panel of Figure \ref{fig:varying_delad}.
+Moreover, a constant $\nabla_{\rm ad,2}$ in the exterior yields the same value for $T\cb$, independent of $\nabla_{\rm ad,1}$ (Figure \ref{fig:varying_delad}, upper-right panel), again consistent with the analytic prediction.
+
+Figure \ref{fig:varying_delad} shows that a larger adiabatic gradient in the outer atmosphere results in a deeper radiative region.
+We can understand this effect by assuming that $R_{\rm switch}$ and $T\cb$ are fixed, i.e.
+ignoring the order unity variations in these quantities due to changes in the adiabatic gradient, as discussed above.
+A steeper adiabat (i.e., with a larger adiabatic gradient) that starts at a given depth (here $R_{\rm switch}$) will match the fixed RCB temperature at a smaller radius.
+A quantitative scaling of $R\cb$ is not possible, as the physical quantities that dictate the RCB location are only order unity corrections to the (also order unity) adiabatic gradient in the outer atmosphere.
+However, relatively small variations in the physical depth of the RCB can impact the RCB pressure significantly, as $P\cb$ increases exponentially with depth, as explained above.
+It follows that $L \propto 1/P\cb$ decreases with increasing outer $\nabla_{\rm ad}$; thus the overall larger adiabatic gradient due to the variable occupation of rotation states at low temperatures decreases $L$ and slows down growth.
+
+%ignoring the order unity variations in    discussed above due to changes in the adiabatic gradient. For
+
+%coefficients in $R_{\rm switch}$ and $T\cb$ discussed above that arise due to $\delad$ variations.
+For a fixed $R_{\rm switch}$, a steeper adiabat (i.e., larger $\nabla_{\rm ad,2}$) will match the (assumed fixed) $T\cb$ at a smaller $R\cb$.
+A quantitative scaling of $R\cb$ with $\nabla_{\rm ad,2}$ is not possible, as all the relevant quantities ($R_{\rm switch}$, $T\cb$, and $R\cb$) are only order unity corrections to the (also order unity) variable $\nabla_{\rm ad,2}$.
+However, relatively small changes in $R\cb$ can impact the RCB pressure significantly since $P\cb$ is an exponential function of $R\cb$, as explained above.
+As $L \propto 1/P\cb$, the atmospheric luminosity decreases with increasing $\nabla_{\rm ad,2}$; thus the overall larger adiabatic gradient due to the variable occupation of rotation states at low temperatures decreases $L$ and slows down growth.
+
 From Equation (\ref{eq:dMdt}), realistic EOS effects reduce the atmosphere accretion rate $\dot{M}$ by decreasing $L$ and increasing $-dE/dM$.
 This results in slower atmospheric growth and therefore a larger critical core mass, as we show in Section \S\ref{critical}.
+
+%Figure \ref{fig:varying_delad}, top panel, shows the resulting radial temperature profile, for various values of $\nabla_{\rm ad, 2}$.
+The core and atmosphere masses are the same as in Figure \ref{fig:all_plot}.
+The temperature profiles overlap in the inner atmosphere, where all polytropes have the same adiabatic gradient $\nabla_{\rm ad,1}=0.3$.
+A larger $\nabla_{\rm ad,2}$ in the outer envelope results in a larger RCB temperature and a deeper radiative region.
+The radius $R_{\rm sw}$ at which $T(R_{\rm sw})=T_{\rm sw}=500$ K modestly increases with $\nabla_{\rm ad,2}$.
+
+%\begin{figure}[h]
+%\centering
+%\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravRealEOS/PaperFigs/varying_delad.pdf}
+%%%\vspace{-0.5in}
+%\caption{Top panel: Radial temperature profiles for polytropes with the same adiabatic gradient in the interior and varying $\delad$ at low temperatures.
+The circles mark the RCB location.
+The dashed black line is the temperature $T_{\rm sw}=500$ K at which $\delad$ changes.
+A larger $\delad$ in the outer envelope increases the RCB temperature and the depth of the radiative region of the atmosphere.
+Bottom panel: Radial pressure profiles for polytropes with the same adiabatic gradient in the interior and varying $\delad$ at low temperatures.
+The circles mark the RCB location.
+The pressure profiles in the radiative region are nearly exponential and overlap for the different polytropes.}
+%\label{fig:varying_delad}
+%\end{figure}
+
+% We can explain these effects analytically.
+By substituting Equation (\ref{eq:structa}) into Equation (\ref{eq:structc}) in the convective region of the atmosphere, and assuming the nebular gas obeys the ideal gas law, we can write hydrostatic balance along an adiabat as
+%
+%\begin{equation}
+%\label{eq:dTdr}
+%\frac{d T}{d r}=-\frac{\delad G m(r)}{\mathcal{R} r^2},
+%\end{equation}
+%where $\mathcal{R}$ is the specific gas constant.
+For a non-self-gravitating atmosphere with an EOS described by Equation (\ref{eq:deladvar}), an analytic expression for the temperature profile can be obtained by integrating Equation (\ref{eq:dTdr}) piece-wise from the core to $R_{\rm sw}$ and from $R_{\rm sw}$ to the RCB, respectively.
+We have showed in Paper I that the temperature dependence deep in a non-self-gravitating atmosphere (where $r \ll R\cb$) with constant $\delad$ throughout can be expressed as
+%
+%
+%Figure \ref{fig:varying_delad} shows that the $R_{\rm sw} \ll R\cb$ approximation, and thus the
+%\begin{equation}
+%\label{eq:Tswapprox}
+%T_{\rm sw} \approx \frac{\nabla_{\rm ad,1} G M\co}{\mathcal{R} R_{\rm sw}}
+%\end{equation}  
+%approximation, become less accurate as $\nabla_{\rm ad, 2}$ increases.
+However, the variations in $R_{\rm sw}$ are small enough that we can assume that $R_{\rm sw}$ is in principle the same for all polytropes.
+Integrating Equation (\ref{eq:dTdr}) from $R_{\rm sw}$ to $R\cb$ thus gives
+%
+%\begin{equation}
+%\label{eq:rcbteor}
+%\frac{1}{R\cb} = \frac{1}{R_{\rm sw}} - \frac{\mathcal{R}}{\nabla_{\rm ad,2} G M\co} (T_{\rm sw}-T\cb).
+%\end{equation}
+%An approximate expression for $T\cb$ as a function of the adiabatic gradient $\nabla_{\rm ad,2}$ and disk temperature can be derived using Equation (\ref{eq:delrad}) applied at the RCB (where $\delad=\delrad$) and $\delrad=d \ln T/ d \ln P$ (see Paper I for details).
+For $\beta=2$ in the opacity law (\ref{eq:opacitylaw}), $T\cb \simeq T\di (1- 2 \nabla_{\rm ad, 2})^{-1/2}$, which increases with $\nabla_{\rm ad,2}$ as shown in Figure \ref{fig:varying_delad}.
+Both $T\cb$ and $\nabla_{\rm ad, 2}$ thus reduce the second term on the right-hand side of Equation (\ref{eq:rcbteor}) when $\nabla_{\rm ad,2}$ is larger, and hence $R\cb$.
+The slightly larger $R_{\rm sw}$ cannot make up for this effect, and therefore a larger adiabatic gradient in the outer atmosphere increases the thickness of the radiative region, consistent with Figures \ref{fig:varying_delad}, top panel, and \ref{fig:all_plot}, middle-left panel.
+It follows that the RCB location is set by the interior structure of the atmosphere, rather than the EOS in the outer envelope.
+This is also shown Figure \ref{fig:varying_delad_in}, where polytropes with the same $\delad$ at the top of the atmosphere have different RCB locations as the interior adiabatic gradient varies.
+%
+%
+%
+%An overall larger $\delad$ in the outer atmosphere due to molecular rotation thus increases the physical extent of the radiative zone.
+As the radiative region is nearly isothermal, the density and pressure increase exponentially with depth, as shown in Figure \ref{fig:varying_delad}, bottom panel.
+The lower $R\cb$ for the low-$T$ realistic EOS thus results in a larger pressure depth of the RCB, as shown in the upper-left panel of Figure \ref{fig:all_plot}.
+As $L \propto 1/P\cb$ (cf.
+Equation \ref{eq:delrad}), the deeper RCB results in a lower $L$ and thus $\dot{M}$ (Figure \ref{fig:all_plot}, lower-right panel).
+\footnote{The $\delad$ and $T\cb$ terms in Equation (\ref{eq:delrad}) applied at the RCB only modestly affect $L$ as $\delad$ varies.}
+
+%Qualitatively, this can be explained through a simple analytic argument: for a non-self-gravitating atmosphere composed of an ideal gas, the temperature profile in convective regions scales as $T(r) \propto \delad/r$ (see Paper I for details; it can also be derived from Equations \ref{eq:structa} and \ref{eq:structc} applied in convective regions).
+Applying the expression above at the RCB yields $T\cb \propto \delad\cb/ R\cb$.
+As the outer radiative layer of the atmosphere is nearly isothermal, $T\cb \approx T\di=const.$, to order unity corrections.
+It easily follows that $R\cb \propto \deladcb$   Figure \ref{fig:all_plot_temp}, middle-left panel shows
+%This is directly correlated to
+
+%directly correlated to the depth of the radiative region of the envelope, which is shown in the upper-left panel of Figure \ref{fig:all_plot} for a total planet mass of $12 M_{\oplus}$.
+We see that the realistic EOS for low temperatures (2) results in a deeper location of the RCB when compared to the ideal gas polytrope (1).
+{\bf This is an accurate observation.
+%The deeper question is why does the EOS change result in a deeper RCB?
+This should be possible to understand, right?}  This deeper RCB translates into a larger number of steps that photons need to diffuse to escape from the RCB, and therefore a lower luminosity (cf.
+Equation \ref{eq:delrad}).
+{\bf Better to just say it's an optical depth effect than reexplaining how radiative diffusion works.
+Also it's the pressure depth (not the radius) that correlates more closely with optical depth.
+A plot of $P(r)$ might thus help as noted in fig caption.}
+
+%We see that the lower luminosity of the fully realistic EOS (4) when compared to the polytrope (1) is due to the realistic EOS in the outer atmosphere (2). T
+
+%We found that the RCB for the fully realistic EOS (4) is roughly at the same location as the RCB for the realistic EOS at low temperatures (2) for low atmosphere masses, and it moves further out relatively as the atmosphere mass increases.
+This explains our choice of $M_{\rm tot}=12 M_{\oplus}$ as a representative case to illustrate the effect of spin isomers.
+Envelope growth is slowest at this relative atmosphere-to-core mass (see PY13)...
+
+%Dissociation deep in the planet's atmosphere increases the amount of energy per unit mass that needs to be radiated by the envelope, which further slows down growth.
+
+%{\bf Why is this true?  Dissociation absorbs energy which by itself decreases the amount of energy that needs to be radiated.}
+
+%hence  The $\delad$ variations given by Equation (\ref{eq:deladvar})
+
+%This result is valid for the $\delad=0.3$ polytrope. 
+%
+%As the low$-T$ realistic EOS only slightly (see Figure \ref{fig:deladmap}) deviates from an ideal gas in the outermost convective layers, we can assume that the gas described by the low-$T$ EOS can be approximated as ideal throughout the convective interior.
+The low-$T$ EOS thus also obeys Equation (\ref{eq:dTdr}) with $\delad \approx const.$.
+We further neglect self-gravity in Equation (\ref{eq:dTdr}) and set $m(r)=M\co$.
+Integrating Equation (\ref{eq:dTdr}) from the core to the RCB under these assumptions yields
+%\begin{equation}
+%\label{eq:dTdrint}
+%T\cb=T\co - \frac{\delad G M\co}{\mathcal{R}} \Big(\frac{1}{R\co} - \frac{1}{R\cb}\Big).
+%\end{equation}  
+%Deep in the convective interior, $T \gg T\cb$ and $r \ll R\cb$. From virial equilibrium, the temperature in the inner atmosphere can thus be expressed as
+%\begin{equation}
+%\label{eq:virialT}
+%T(r) \simeq \frac{\nabla_{\rm ad} G M\co}{\mathcal{R} r}.
+%\end{equation}
+%%The temperature scale height deep in the convective interior is $H_T \sim r$, with $dT/dr \sim -T/H_T \sim -T/r$.
+From Equation (\ref{eq:structc}) it immediately follows that the pressure scale height, for which $dP/dr \sim -P/H_P$, is $H_P \sim r \delad$.
+Deep in the interior and for $m(r)=M\co$, hydrostatic balance (\ref{eq:structa}) thus becomes
+%
+%%\begin{equation}
+%%\label{eq:structint}
+%%\frac{P}{r \delad} \sim -\frac{G M\co}{r^2}. 
+%%\end{equation}
+%%From this and the definition of the isothermal sound speed
+%%\begin{equation}
+%%\label{eq:cs}
+%%c_s^2 = \frac{P}{\rho} =\mathcal{R} T, 
+%%\end{equation} 
+%%with both expressions applied at the core where $r=R\co$, $T=T\co$ and $\delad=\nabla_{\rm ad, c}$, we find
+%Applying equation (\ref{eq:virialT}) at the core, where $T=T\co$, $r=R\co$ and $\delad=\nabla_{\rm ad, c}$, gives
+%\begin{equation}
+%\label{eq:Tc}
+%T\co \simeq \frac{\nabla_{\rm ad, c} G M\co}{\mathcal{R} R\co}.
+%\end{equation}
+%As the adiabatic gradient at the core is the same for both the ideal gas polytrope and the low$-T$ realistic EOS, $T\co$ has the same value for both EOSs.
+This is shown in Figure \ref{fig:all_plot}, lower-left panel.
+Substituting Equation (\ref{eq:Tc}) into Equation (\ref{eq:dTdrint}) applied at the RCB gives
+%\begin{equation}
+%\label{eq:Tcb}
+%T\cb \simeq \frac{\nabla_{\rm ad, RCB} G M\co}{\mathcal{R} R\cb},
+%\end{equation}
+%%We note that the expression above and the definition of the pressure scale height $H_P \equiv c_s^2 r^2/(G M\co)$ imply that the pressure scale height at the RCB is the same as in the deep interior, $H_P \sim \delad r$.
+%and thus
+%\begin{equation}
+%\label{eq:rcb}
+%R\cb \propto \frac{\nabla_{\rm ad, RCB}}{T\cb}.
+%\end{equation}
+%An approximate expression for $T\cb$ as a function of the adiabatic gradient and disk temperature can be derived using Equation (\ref{eq:delrad}) applied at the RCB (where $\delad=\delrad$) and $\delrad=d \ln T/ d \ln P$ (see Paper I for details).
+For $\beta=2$ in the opacity law (\ref{eq:opacitylaw}), $T\cb \simeq T\di (1- 2 \nabla_{\rm ad, RCB})^{-1/2}$, and thus
+%\begin{equation}
+%\label{eq:rcbdel}
+%R\cb \propto \frac{\nabla_{\rm ad, RCB}}{(1- 2 \nabla_{\rm ad, RCB})^{-1/2}}.
+%\end{equation}
+%For the ideal gas polytrope $\nabla_{\rm ad, RCB}=0.3$, and for the low-$T$ realistic EOS $\nabla_{\rm ad, RCB}\approx 0.25$.
+For these particular choices, Equation (\ref{eq:rcbdel}) yields a smaller $R\cb$ for the low$-T$ realistic EOS.
 
@@
 
+%\section{Location of radiative-convective boundary for polytropes with varying $\delad$} \label{vardelad}
+
 \section{Grain growth opacity and radiative windows} \label{radwindow}

### 2014-03-03 tex/draft_paper2/draft_paper2.tex
minor Paper II edits, but more coming today

@@
 %Fast accretion implies large critical core masses. 
-In order to calculate the minimum value of $M_{\rm crit}$, we first need to understand how $M_{\rm crit}$ depends on the planetesimal accretion rate.
+In order to calculate the minimum value of $M_{\rm crit}$, we need to understand how $M_{\rm crit}$ depends on the planetesimal accretion rate.
 %the evolution of a core's atmosphere depends on the rate at which planetesimals are accreted. 
@@
 
-\section{Model Review}
+\section{Atmospheric Model Review}
 \label{sec2}
@@
 
-We begin by reviewing the model developed in Paper I for the structure and evolution of a planetary atmosphere embedded in a protoplanetary disk.
+In this section we review the model developed in Paper I for the structure and evolution of a planetary atmosphere embedded in a protoplanetary disk.
 We describe the assumptions of the model and the properties of our assumed protoplanetary disk in \S\ref{model}, and we summarize the equations describing the structure and time evolution of the atmosphere in \S\ref{struct}.
@@
 We assume that the planet consists of a solid core of fixed mass and a two-layer atmosphere composed of an inner convective region and an outer radiative zone that matches smoothly onto the disk.
-The two regions are separated by the Schwarzschild criterion for convective instability (see \S\ref{struct}) .
-We denote the surface between the two regions as the radiative-convective boundary (RCB), defined by a radius $r=R\cb$.
-Moreover, we assume that the luminosity is constant throughout the outer radiative region (see Section \S\ref{critical} for additional discussion).
+The two regions are separated by the Schwarzschild criterion for convective instability (see \S\ref{struct}).
+The surface between the two regions is the radiative-convective boundary (RCB), defined by a radius $r=R\cb$.
+Moreover, we assume that the luminosity is constant throughout the radiative region (see Section \S\ref{critical} for additional discussion).
 We assume a low planetesimal accretion regime in which the atmosphere evolution is dominated by KH contraction.The atmosphere is spherically symmetric, self-gravitating and in hydrostatic balance.
@@
 The temperature and pressure at the outer boundary of the atmosphere are given by the nebular temperature and pressure.
-As a disk model, we use the minimum mass, passively irradiated model of  \citet{chiang10}.
-The surface density, mid-plane temperature and mid-plane pressure are given by
+We use the minimum mass, passively irradiated disk model of  \citet{chiang10}.
+The surface density, mid-plane temperature and mid-plane pressure are
 
@@
 \end{subeqnarray}
-
-\noindent with $a$ the semimajor axis and for a mean molecular weight $\mu=2.35$.
+where $a$ is the semimajor axis and for a mean molecular weight $\mu=2.35$.
 
@@
 
-\noindent with $\sigma$ the Stefan-Boltzmann constant and $\kappa$ the dust opacity.
+\noindent where $\sigma$ is the Stefan-Boltzmann constant and $\kappa$ is the dust opacity.
 In our models the atmosphere is optically thick throughout the outer boundary.
@@
 \end{equation}
-
-\noindent with $\delad$ the adiabatic temperature gradient.
+with $\delad$ the adiabatic temperature gradient.
 The convective and radiative layers of the envelope are separated by the Schwarzschild criterion (e.g., \citealt{thompson06}): the atmosphere is stable against convection when $\nabla < \delad$ and convectively unstable when $\nabla > \delad$.
@@
 
-At low temperatures, hydrogen exists in molecular form and has a stable configuration.
-As the temperature becomes higher than $T \sim 2000-3000$ K, the internal energy of the hydrogen molecule becomes large enough to break the covalent bond between the atoms, and hydrogen starts dissociating.
+At low temperatures, hydrogen is molecular and has a stable configuration.
+At temperatures higher than $T \sim 2000-3000$ K, the internal energy of the hydrogen molecule becomes large enough to break the covalent bond between the atoms, and hydrogen starts dissociating.
 At $T \gtrsim 10,000$ K, hydrogen ionizes.
@@
 As expected, the adiabatic gradient is $\delad=2/7$ for pure molecular hydrogen and $\delad=2/5$ when hydrogen is fully dissociated, but decreases significantly during partial dissociation and is smallest when half of the gas is dissociated.
-This is consistent with the behavior we see in Figure \ref{fig:deladmap}.
+This is consistent with the behavior shown in Figure \ref{fig:deladmap}.
 
@@
 
-Similarly to the case of dissociation, as $T$ increases, part of the rotational energy $U_{\rm r}$ of the hydrogen molecule is used to convert parahydrogen to orthohydrogen rather than to only increase the temperature, so $U_{\rm r} \propto T$ no longer holds.
-The adiabatic gradient $\delad$ thus varies with temperature.
-\App{deladspin} explores the temperature dependence in detail.
-These $\delad$ variations, which occur for $T \lesssim 300$ K, are displayed in Figure \ref{fig:deladmap}.
+Similarly to the case of dissociation, as $T$ increases, part of the rotational energy of the hydrogen molecule is used to convert parahydrogen to orthohydrogen rather than to only increase the temperature.
+Thus $u_{\rm r} \propto T$, where $u_{\rm r}$ is the rotational energy per unit mass, no longer holds, and the adiabatic gradient $\delad$ is temperature dependent.
+These $\delad$ variations, which occur for $T \lesssim 300$ K, are explored in \App{deladspin} and displayed in Figure \ref{fig:deladmap}.
 At higher temperatures, the ortho-para 3:1 equilibrium ratio is reached; no further isomer conversion occurs, and thus $\delad$ remains relatively constant until dissociation temperatures are reached.
@@
 
-So far in our model we have assumed that the dust opacity in the radiative region of the atmosphere is given by the standard ISM opacity.
+Our calculations so far have assumed that the dust opacity in the radiative region of the atmosphere is given by the standard ISM opacity.
 However, our scenario of low planetesimal accretion is likely to favor lower dust opacities, due to grain growth and dust settling.
@@
 In this work we use the  \citet{dalessio01} frequency-dependent opacity tables to obtain the temperature-dependent Rosseland mean opacity $\kappa$.
-We take as a fiducial case a maximum particle size of 1 cm and a grain size distribution given by Equation (\ref{eq:graindistr}) with $p=3.5$.
+We take as a fiducial case a maximum particle size $a_{\rm max}=1$ cm and a grain size distribution given by Equation (\ref{eq:graindistr}) with $p=3.5$.
 Other choices for the power-law coefficient $p$ are discussed later in this section.
@@
 The \citet{dalessio01} opacities are only relevant at temperatures that are sufficiently low for dust grains to remain solid ($T \lesssim 1000$ K).
-In the temperature regime where dust sublimates we use the \citet{bell94} analytic opacity laws, ensuring smooth transition from the grain growth opacities.
-
-The sharp drop in opacity ($\kappa \sim T^{-24}$, see Figure \ref{fig:delvsr}) due to dust sublimation lowers the radiative temperature gradient significantly (see Equation \ref{eq:delrad}), and may therefore generate radiative layers within the inner region of the atmosphere (see \App{radwindow}, Figure \ref{fig:delvsr}).
+At higher temperatures, we use the \citet{bell94} analytic opacity laws, ensuring smooth transition from the grain growth opacities.
+
+The sharp drop in opacity ($\kappa \sim T^{-24}$, see Figure \ref{fig:delvsr}) due to dust sublimation lowers the radiative temperature gradient significantly (see Equation \ref{eq:delrad}), and may thus generate radiative layers within the inner region of the atmosphere (see \App{radwindow}, Figure \ref{fig:delvsr}).
 This could in principle pose challenges for our model, since the additional luminosity generated in these radiative windows may be large enough to make our assumption of constant luminosity throughout the radiative layers invalid.
@@
 However, the radiative windows may become non-negligible for other opacity choices and sufficiently low core masses, as we show later in this section.
-We note that despite the existence of one or more radiative windows in the planet's atmosphere, the luminosity that emerges at the outer boundary is set by the outer radiative layer.
-In fact, a simple estimate of the luminosity emerging at the RCB can be found as follows.
+Despite the existence of one or more radiative windows in the planet's atmosphere, the luminosity that emerges at the outer boundary is set by the outer radiative layer.
+A simple estimate of the luminosity emerging at the RCB can be found as follows.
 
@@
 We first explore the dependence of the runaway growth time on the core mass for a fixed semimajor axis.
-We then determine the critical core mass to form a giant planet from a gas composed of a realistic hydrogen-helium mixture, and we compare this with the results from Paper I for an ideal diatomic gas.
-Finally, we determine the critical core mass under more realistic opacity assumptions.
+We then determine the critical core mass $M_{\rm crit}$ to form a giant planet from a gas composed of a realistic hydrogen-helium mixture, and we compare this with the results from Paper I for an ideal diatomic gas.
+Finally, we determine $M_{\rm crit}$ under more realistic opacity assumptions.
 
@@
 The results of Paper I for an ideal diatomic gas are plotted for comparison.
-The use of a realistic EOS results in a critical core mass more than twice as large as that produced by a polytrope.
+The inclusion of realistic EOS effects increases $M_{\rm crit}$ by more than a factor of two when compared to an ideal gas polytrope.
 As such, non-ideal effects substantially increase the core mass needed to form a giant planet  before the dissipation of the protoplanetary disk.
@@
 
-Figure \ref{fig:Mcritvsagg} shows the critical core mass as a function of semimajor axis, for a grain growth opacity with a size distribution given by the power-law (\ref{eq:graindistr}) with $p=3.5$ and maximum particle size of 1 cm.
-The critical core mass is lower than in the standard case, and less sensitive to location in the disk, i.e.
+Figure \ref{fig:Mcritvsagg} shows $M_{\rm crit}$ as a function of semimajor axis, for a grain growth opacity with a size distribution given by the power-law (\ref{eq:graindistr}) with $p=3.5$ and maximum particle size $a_{\rm max}=1$ cm.
+The critical core mass is lower than in the standard interstellar opacity case, and less sensitive to location in the disk, i.e.
 the boundary conditions (temperature and pressure).
@@
 
-So far we have assumed that the size distribution of dust grains is a power-law (\ref{eq:graindistr}) with $p=3.5$ which corresponds to a normal collisional cascade.
-If, however, coagulation is taken into account, the exponent $p$ can be approximated as $p=2.5$ \citep{dalessio01}.
-This results in a flatter and significantly lower opacity, which may substantially reduce the critical core mass.
+In the grain size distribution (\ref{eq:graindistr}), the coefficient $p=3.5$ corresponds to a normal collisional cascade.
+If coagulation is taken into account, the exponent $p$ can be approximated as $p=2.5$ \citep{dalessio01}.
+This results in a flatter and significantly lower opacity, which may substantially reduce $M_{\rm crit}$.
 However, we have found that our model breaks down for low core masses ($M\co \lesssim 3 M_{\oplus}$) under our assumption of constant luminosity in the outer radiative layer.
 Figure \ref{fig:p25p35} shows the crossover time $t_{\rm co}$ as a function of semimajor axis for the lowest core mass for which our model is valid, $M\co=4 M_{\oplus}$.
-The crossover time is more than one order of magnitude lower for $p=2.5$, which implies that the critical core mass may be, in fact, significantly lower than presented in Figure \ref{fig:Mcritvsagg}.
+The crossover time is more than one order of magnitude lower for $p=2.5$, which implies that $M_{\rm crit}$ may be, in fact, significantly lower than presented in Figure \ref{fig:Mcritvsagg}.
 
@@
 
-In this study we have considered protoplanets with fully formed cores for which planetesimal accretion is negligible and KH contraction dominates the luminosity evolution of the atmosphere.
+This study considers protoplanets with fully formed cores for which planetesimal accretion is negligible and KH contraction dominates the luminosity evolution of the atmosphere.
 This is different from calculations that assume high planetesimal accretion rates and find that the atmosphere is in steady state and solely heated due to accretion of solids.
@@
 We discuss the core accretion rates that are necessary for our regime to be valid in \S\ref{raf1}.
-In \S\ref{raf3}, we estimate core growth during atmosphere evolution at the maximum rate for which the KH regime is valid, and show it is negligible.
+In \S\ref{raf3}, we estimate core growth during atmosphere evolution at the maximum rate for which the KH regime is valid, and show it is indeed negligible.
 Finally, we compare our results with those assuming fast planetesimal accretion in \S\ref{raf2}.
@@
 \end{equation} 
-
 To illustrate the magnitude of $\dot{M}_{\rm c, KH}$, we choose as a fiducial case an atmosphere forming at 10 AU and with a core mass of $10 M_{\oplus}$.
-Since analytic studies of critical core masses at high planetesimal accretion rates assume an ideal gas EOS, for ease of comparison we choose an ideal gas polytrope with constant adiabatic gradient $\delad=2/7$ and mean molecular weight $\mu=2.35$ (see also Paper I).
+Since analytic studies of critical core masses at high planetesimal accretion rates assume an ideal gas EOS, for ease of comparison we choose an ideal gas polytropic EOS with constant adiabatic gradient $\delad=2/7$ and mean molecular weight $\mu=2.35$ (see also Paper I).
 For this choice of parameters, the runaway accretion time is $t _{\rm run}\sim$ 1.3 Myrs, which is within the typical lifetime of a protoplanetary disk.
@@
 \label{eq:Mcdot}
-\dot{M}_{\rm{c,acc}}(M_{\rm{c}}) \equiv \frac{M_{\rm{c}}}{\tau}.
+\dot{M}_{\rm{c,acc}}(M_{\rm{c}}) \sim \frac{M_{\rm{c}}}{\tau}.
 \end{equation}
@@
 
-Planetesimal accretion during the KH contraction phase of atmosphere growth at a rate $\dot{M}<\dot{M\co}_{\rm{KH}}$  cannot alter the core mass enough to affect the time evolution of the atmosphere.
+Planetesimal accretion during the KH contraction phase of atmosphere growth at a rate $\dot{M}_{\rm c}<\dot{M}_{\rm{c,KH}}$  cannot alter the core mass enough to affect the time evolution of the atmosphere.
 We can quantitatively estimate the increase in core mass as
@@
 For $M_{\rm{c}}=10 M_{\oplus}$, we find $\Delta M_{\rm{c}} \approx 0.05 M_{\oplus} \ll 10 M_{\oplus}$.
-It follows that a significant increase in core mass that could potentially alter the time evolution of the atmosphere would occur on a much longer timescale then the crossover time for the unperturbed atmosphere.
-The time evolution of the atmosphere is thus insensitive to core mass changes at a rate imposed by the assumption that $L_{\rm{acc}}<L_{\rm{KH}}$.
+Core growth is negligible in our regime, and the time evolution of the atmosphere is thus insensitive to core mass changes at a rate imposed by the assumption that $L_{\rm{acc}}<L_{\rm{KH}}$.
 
@@
 
-We compare our results for the critical core mass $M_{\rm crit}$ with those of studies that assume large planetesimal accretion rates.
-These studies find that the atmosphere is in steady state at all times and that static solutions only exist up to a maximum core mass, which they define as the critical core mass.
+We compare our results for $M_{\rm crit}$ with those of studies that assume large planetesimal accretion rates.
+These studies find that the atmosphere is in steady state at all times and that static solutions only exist up to a maximum (critical) core mass.
 If the $M_{\rm crit}$ found by static studies were lower than the $M_{\rm crit}$ in our regime of negligible solids accretion, the atmosphere would undergo runaway gas accretion before KH contraction became dominant, and our regime would not be relevant.
@@
 R06 assumes an ideal gas polytropic EOS and a lower opacity than the standard ISM (see Equation \ref{eq:opacitylaw}).
-We thus calculate the critical core mass for an ideal gas polytrope with the normalization factor $F_{\kappa}$ reduced by a hundred, which is comparable to the opacity law used by R06 \footnote{The power-law opacity of R06 is scaled to the (semimajor axis dependent) disk temperature, while our opacity is scaled to an absolute reference temperature.
+We thus calculate $M_{\rm crit}$ for an ideal gas polytrope with the normalization factor $F_{\kappa}$ reduced by a hundred, which is comparable to the opacity law used by R06 \footnote{The power-law opacity of R06 is scaled to the (semimajor axis dependent) disk temperature, while our opacity is scaled to an absolute reference temperature.
 We thus cannot directly use the R06 opacities for our comparison.}.
@@
 \label{eq:u}
-U=\mathcal{R} T^2 \Big(\frac{\partial \log{Z}}{\partial T}\Big)_{V}
+u=\mathcal{R} T^2 \Big(\frac{\partial \log{Z}}{\partial T}\Big)_{V}
 \end{equation}
@@
 \end{equation}
-Since $Z=Z_t Z_r Z_v$, we may write $U=U_t+U_r+U_v$ and $S=S_t+S_r+S_v$, where $U_t$, $U_r$, $U_v$, $S_t$, $S_r$, $S_v$ are the quantities corresponding to the individual translation, rotation and vibration partition functions, respectively.
+Since $Z=Z_t Z_r Z_v$, we may write $u=u_t+u_r+u_v$ and $S=S_t+S_r+S_v$, where $u_t$, $u_r$, $u_v$, $S_t$, $S_r$, $S_v$ are the quantities corresponding to the individual translation, rotation and vibration partition functions, respectively.
 
@@
 \label{eq:ut}
-U_t=\frac{3}{2} \mathcal{R} T
+u_t=\frac{3}{2} \mathcal{R} T
 \end{equation}
@@
 As such, the density in this region follows the ideal gas law $P=\rho \mathcal{R} T$.
-\item{\textbf{Internal energy per mass.}} $U=U_t+U_r+U_v$, where $U_t$ is given by Equation(\ref{eq:ut}), and $U_r$ and $U_v$ are determined using equations (\ref{eq:u}), (\ref{eq:Zrspin}) and (\ref{eq:Zv}) above.
+\item{\textbf{Internal energy per mass.}} $u=u_t+u_r+u_v$, where $u_t$ is given by Equation(\ref{eq:ut}), and $u_r$ and $u_v$ are determined using equations (\ref{eq:u}), (\ref{eq:Zrspin}) and (\ref{eq:Zv}) above.
 \item{\textbf{Entropy per unit mass}}.
-Similarly, $S=S_t+S_r+S_v$, where $S_t$ is given by Equation (\ref{eq:st}), and $S_r$ and $S_v$ can be determined from Equation (\ref{eq:s}) and the calculated expressions for $U_r$ and $U_v$, respectively.
+Similarly, $S=S_t+S_r+S_v$, where $S_t$ is given by Equation (\ref{eq:st}), and $S_r$ and $S_v$ can be determined from Equation (\ref{eq:s}) and the calculated expressions for $u_r$ and $u_v$, respectively.
 \item{\textbf{Entropy logarithmic derivatives}}. The logarithmic derivatives $S_T$ and $S_P$ are given by:

### 2014-03-04 tex/draft_paper2/draft_paper2.tex
a few more Paper II edits

@@
 
-\title{Quantitative Estimates of Minimum Core Masses for Giant Planet Formation
-{\bf Paper I was quantitative.}}
+\title{Minimum Core Masses for Giant Planet Formation Using Realistic Equation of State Tables and Opacities}
 \author{Ana-Maria A. Piso}
@@
 The regions in which the EOS is either invalid or not computed are masked in white.
-Our extension is only valid for $T \lesssim 2000$ K, since it does not take into account hydrogen dissociation.
-We choose $T=1500$ K as a conservative temperature cutoff.
-\citet{saumon95} do not compute the EOS at very high pressures, since hydrogen is solid or may form a Coulomb lattice in this regime, and thus their EOS treatment is no longer valid.
-While the boundaries of the region in which the free-energy EOS treatment fails can be determined from fundamental thermodynamic constraints, such calculations are not the object of this work.
-Instead, we choose as boundary a constant entropy curve ($\log(S)=8.4$) above the region in which the \citet{saumon95} model fails.
-The expressions derived in \App{EOStables} are sufficient to give good results for the colored regions of the extended map, which fully cover the temperature and pressures ranges required by our models.
-{\bf This caption is extremely long.
-I would find a way to defer much of this to text.}}
+Further details about the masked regions can be found in \App{hydrogen}.}
 \label{fig:deladmap}
@@
 This behavior is different than that of a mixture of molecular and atomic hydrogen, for which $2/7<\delad<2/5$, or a mixture of protons and electrons, for which $\delad=2/5$.
-For a mixture of ideal gases, the total internal energy is given by the sum of the internal energies of the individual gases.
-When a gas dissociates, however, part of the internal energy of the system is used to break down molecules, which reduces the amount of energy available to increase the temperature of the system, thus lowering the adiabatic gradient.
+When a gas dissociates, part of the internal energy of the system is used to break down molecules, which reduces the amount of energy available to increase the temperature of the system, thus lowering the adiabatic gradient.
 The energy required to dissociate depends on the dissociation fraction, which can be determined from the Saha equation (see e.g., \citealt{kippenhahn90}).
@@
 \ref{fig:deladmap}.
+
+%For a mixture of ideal gases, the total internal energy is given by the sum of the internal energies of the individual gases.
 
@@
 Equation \ref{eq:coolingglobal} and ignoring surface terms, which only become significant in the late stages of atmospheric evolution) is lower than in the ideal case.
-The $\delad$ variations thus slow down gas accretion and increase the growth time of the atmosphere, and hence the critical core mass.
+The $\delad$ variations thus slow down gas accretion and increase the growth time of the atmosphere.
+Since envelope growth is faster for larger cores, the minimum core mass required to accumulate a massive envelope during the (assumed fixed) disk lifetime, i.e.
+$M_{\rm crit}$, increases when realistic EOS effects are included.
 
@@
 \item Ideal gas polytrope with $\delad=0.3$.
-\item Ideal gas polytrope with $\delad=0.3$ for $T>500$ K and realistic EOS for $T<500$ K (hereafter low-$T$ realistic EOS).
-\item Ideal gas polytrope with $\delad=0.3$ for $T<500$ and realistic EOS for $T>500$ K  (hereafter high-$T$ realistic EOS).
+\item Ideal gas polytrope with $\delad=0.3$ for $T>500$ K and realistic EOS for $T<500$ K (hereafter low-$T$ realistic EOS), which includes the effects of fractional occupation of $\rm H_2$ rotational states.
+\item Ideal gas polytrope with $\delad=0.3$ for $T<500$ and realistic EOS for $T>500$ K  (hereafter high-$T$ realistic EOS), which includes the effects of hydrogen dissociation.
 \item Realistic EOS at all $T$. 
@@
 %By separately comparing curves (1) and (2), and (1) and (3)  while the difference between curves (1) and (3) accounts for hydrogen dissociation. 
-Both the high-$T$ and low-$T$ EOSs yield slower cooling when compared to the ideal gas polytrope --- due to dissociation for the high-$T$ EOS, and due to fractional occupation of $\rm H_2$ rotational states for the low-$T$ EOS.
+Both the high-$T$ and low-$T$ EOSs yield slower gas accretion when compared to the ideal gas polytrope.
+
+% --- due to dissociation for the high-$T$ EOS, and due to fractional occupation of $\rm H_2$ rotational states for the low-$T$ EOS.
 %Both dissociation and fractional occupation of $\rm H_2$ rotational states result in slower cooling for both the high-$T$ and low-$T$ realistic EOSs. 
@@
 \caption{Elapsed time to grow a planet of total mass (core + atmosphere) for a variety of EOS combinations (see text), for a planet forming at 10 AU and with a fixed core mass $M_{\rm c}=10 M_{\oplus}$.
-Both hydrogen dissociation at high temperatures deep in the atmosphere and fractional occupation of $\rm H_2$ rotational states at low temperatures in the outer envelope result in slower cooling when compared to an ideal gas polytrope.}
+Both hydrogen dissociation at high temperatures deep in the atmosphere and fractional occupation of $\rm H_2$ rotational states at low temperatures in the outer envelope result in slower gas accretion when compared to an ideal gas polytrope.}
 \label{fig:tplotall}
@@
 The location of the RCB is marked.
-The variable occupation of rotational states at low temperatures in the outer region of the atmosphere results in a deeper RCB.
+The variable occupation of $\rm H_2$ rotational states in the outer region of the atmosphere results in a deeper RCB.
 Upper-right panel:  Luminosity evolution as a function of total mass.
-Fractional occupation of $\rm H_2$ rotational states at low temperatures near the top of the atmosphere results in a lower luminosity for the realistic EOS (see text).
-Middle-left panel: Total energy at scale $r$, $-e \rho r^3$, profile for a total mass of $12 M_{\oplus}$.
+The variable occupation of $\rm H_2$ rotational states in the outer region of the atmosphere results in a lower luminosity for the realistic EOS (see text).
+Middle-left panel: Profile of total energy at scale $r$, $-e \rho r^3$, for a total mass of $12 M_{\oplus}$.
 Hydrogen dissociation deep in the atmosphere causes more of the energy to be concentrated at the bottom of the envelope when compared to an ideal gas polytrope.
@@
 Dissociation increases the number of particles, which tends to increase entropy.
-In order for entropy to stay constant, the temperature must therefore decrease, which also lowers the adiabatic gradient.
+In order for entropy to stay constant with radius, the temperature must therefore decrease, which also lowers the adiabatic gradient.
 A lower $\delad$ results in a shallower temperature gradient in dissociation regions, and hence in lower temperatures near the core (cf.
@@
 As the specific internal energy $u(r) \propto T(r)$, dissociation decreases $u(r)$ and thus the total energy deep in the interior.
-This explains the local energy behavior shown in Figure \ref{fig:all_plot}, middle-left panel.
+This explains the rise in $-e \rho r^3$ in Figure \ref{fig:all_plot}, middle-left panel.
 Note that, in some astrophysical contexts, this loss of thermal energy due to dissociation can be so large as to trigger dynamical instabilities and eventual collapse in higher mass objects, such as in protostars \citep{larson69} or during the runaway growth of giant planets \citep{bodenheimer80}.
@@
 The EOS variations in the cold outer envelope affect $\dot{M}$ by decreasing the atmosphere's luminosity compared to an ideal gas polytrope (Figure \ref{fig:all_plot}, upper-right panel, for the same parameters as Figure \ref{fig:tplotall}).
-From Figures \ref{fig:deladmap} and \ref{fig:all_plot}, bottom-left panel, we see that the adiabatic gradient of the low-$T$ realistic EOS increases as the temperature decreases throughout the convective zone, then becomes lower near the RCB.
+From Figures \ref{fig:deladmap} and \ref{fig:all_plot}, bottom-left panel, we see that the adiabatic gradient of the low-$T$ realistic EOS increases with radius (i.e., with decreasing temperature) throughout the convective zone when $T<500$ K, then becomes lower near the RCB.
 This results in a larger adiabatic gradient, on average, for the low-$T$ realistic EOS in the outer parts of the convective region of the atmosphere.
@@
 the critical core mass.
-As in Paper I, we quantify the runaway growth timescale $t_{\rm run}$ as the time at which the atmosphere growth time $M_{\rm atm}/\dot{M}$ drops to 10\% of its maximum value (see Paper I for details).
-We first explore the dependence of the runaway growth time on the core mass for a fixed semimajor axis.
+As in Paper I, we quantify the runaway accretion time $t_{\rm run}$ as the time at which the atmosphere growth time $M_{\rm atm}/\dot{M}$ drops to 10\% of its maximum value (see Paper I for details).
+We first explore the dependence of $t_{\rm run}$ on the core mass for a fixed semimajor axis.
 We then determine the critical core mass $M_{\rm crit}$ to form a giant planet from a gas composed of a realistic hydrogen-helium mixture, and we compare this with the results from Paper I for an ideal diatomic gas.
@@
 The crossover time is more than one order of magnitude lower for $p=2.5$, which implies that $M_{\rm crit}$ may be, in fact, significantly lower than presented in Figure \ref{fig:Mcritvsagg}.
+In other words, grain growth can yield critical core masses up to an order of magnitude lower than in the case where interstellar opacities are used.
 
@@
 Figure \ref{fig:deladH} shows a contour plot of $\delad$ as a function of temperature and pressure for the extended EOS table.
+Our extension is only valid for $T \lesssim 2000$ K, since it does not take into account hydrogen dissociation.
+We choose $T=1500$ K as a conservative temperature cutoff.
+\citet{saumon95} do not compute the EOS at very high pressures, since hydrogen is solid or may form a Coulomb lattice in this regime, and thus their EOS treatment is no longer valid.
+While the boundaries of the region in which the free-energy EOS treatment fails can be determined from fundamental thermodynamic constraints, such calculations are not the object of this work.
+Instead, we choose as boundary a constant entropy curve ($\log(S)=8.4$) above the region in which the \citet{saumon95} model fails.
+The expressions derived above are sufficient to give good results for the colored regions of the extended map, which fully cover the temperature and pressures ranges required by our models.
+
  %Our extension smoothly matches the original tables for $8.80<\log{S}$(K g$^{-1})<9.07$ (\textbf{numbers are wrong, change once you have the final figure version}).
@@
 At high temperatures, hydrogen dissociates and ionizes, while at low temperatures the rotational states of the hydrogen molecule are only partially excited and it no longer behaves like an ideal diatomic gas.
-Regions in which the EOS is invalid or has not been computed are masked in white (see Figure \ref{fig:deladmap} caption for an explanation).}
+Regions in which the EOS is invalid or has not been computed are masked in white (see text).}
 \label{fig:deladH}
@@
 We choose $T=7,000$ K as a conservative temperature cutoff above which our extension is no longer valid (masked in white).
-The EOS has not been computed in the lower-right region of the plot --- see Figure \ref{fig:deladmap} for an explanation.}
+The EOS has not been computed in the lower-right region of the plot (see text).}
 \label{fig:deladHe}

### 2014-03-05 tex/draft_paper2/draft_paper2.tex
minor (compilation problem, needed to close bracket with \.)
couple minor edits

@@
 \noindent where $r$ is the radial coordinate, $P$, $T$ and $\rho$ are the gas pressure, temperature and density, respectively, $m$ is the mass enclosed by the radius $r$, $L$  is the luminosity from the surface of radius $r$, and $G$ is the gravitational constant.
-The gas is heated at a rate of $\epsilon_g \equiv -T \frac{ds}{dt}$ per unit mass due to gravitational contraction, where $s$ is the specific gas entropy, while  $\epsilon$ represents the rate at which internal heat is generated per unit mass.
+The gas is heated at a rate of $\epsilon_g \equiv -T ds/dt$ per unit mass due to gravitational contraction, where $s$ is the specific gas entropy, while  $\epsilon$ represents the rate at which internal heat is generated per unit mass.
 We do not take into account any internal energy sources and set $\epsilon=0$.
-The temperature gradient $\nabla \equiv \frac{d \ln T}{d \ln P}$ depends on whether energy is transported throughout the atmosphere by radiation or convection.
+The temperature gradient $\nabla \equiv d \ln T/d \ln P$ depends on whether energy is transported throughout the atmosphere by radiation or convection.
 In the case of radiative diffusion for an optically thick gas, the temperature gradient is
@@
 We choose $\rho_{\rm c}=3.2$ g cm$^{-3}$ (e.g., \citealt{pap99}).
-The atmosphere matches onto the protoplanetary disk at the Hill radius $R_{\rm H} \equiv a \Big(\frac{M_{\rm p}}{3 M_{\odot}}\Big)^{1/3}$, where $M_{\rm p}$ is the total planet mass, $M_{\rm p}=M\co+M_{\rm atm}$.
+The atmosphere matches onto the protoplanetary disk at the Hill radius $R_{\rm H} \equiv a [M_{\rm p} / (3 M_{\odot})]^{1/3}$, where $M_{\rm p}$ is the total planet mass, $M_{\rm p}=M\co+M_{\rm atm}$.
 The Hill radius is the distance from the planet at which the gravitational attraction of the planet and the tidal gravity due to the host star are equal.
 Outside the Hill sphere, stellar tides unbind gas from the planet.
-The effective outer boundary of the atmosphere is the surface defined by the Bondi radius, $R_{\rm B} \equiv \frac{G M_{\rm p}}{c_{\rm s}^2}=\frac{G M_{\rm p}}{\mathcal{R} T\di}$.
+The effective outer boundary of the atmosphere is the surface defined by the Bondi radius, $R_{\rm B} \equiv G M_{\rm p}/c_{\rm s}^2=G M_{\rm p} / (\mathcal{R} T\di)$.
 This is the distance from the planet at which the thermal energy of the nebular gas is of the order of the gravitational energy of the planet.
@@
 \nabla_{\rm ad, 2}, & \quad T < 500 \text{ K}
-\end{array} \right ,
+\end{array} 
+\right.
 \end{equation}   
@@
 This study considers protoplanets with fully formed cores for which planetesimal accretion is negligible and KH contraction dominates the luminosity evolution of the atmosphere.
-This is different from calculations that assume high planetesimal accretion rates and find that the atmosphere is in steady state and solely heated due to accretion of solids.
+This approach contrasts that of static models, which assume high planetesimal accretion rates and find that the atmosphere is in steady state and solely heated due to accretion of solids.
 In this section we compare our results for the critical core mass to analogous results from steady-state fast planetesimal accretion calculations.

### 2014-03-07 tex/PaperI/ms.tex
small Paper I edits

@@
 We use $\RH$ as the matching radius, i.e.\ outer boundary, in all regimes.
-\textbf{This choice reflects the fact that, even outside the Bondi radius, hydrostatic densities exceed the background density (see R06).
+%\textbf{This choice reflects the fact that, even outside the Bondi radius, hydrostatic densities exceed the background density (see R06).
 Nevertheless, since the density contrast is modest outside $\RB$, these two boundary conditions give similar results.
@@
 On scales $\gtrsim \RB$ the flows no longer circulate the planet: they belong to the disk.
-Nevertheless, the density structure outside $\RB$ remains spherically symmetric and hydrostatic.
 Even if these flows do not destroy hydrostatic balance, they could affect the planet's cooling.
 We expect such effects to be weak, as heat losses at greater depths dominate planetary cooling, but more study is needed.
+
+ %Nevertheless, the density structure outside $\RB$ remains spherically symmetric and hydrostatic.
 

### 2014-03-07 tex/PaperI/ms.tex
minor edits

@@
 We use $\RH$ as the matching radius, i.e.\ outer boundary, in all regimes.
+We discuss this choice further below.
 %\textbf{This choice reflects the fact that, even outside the Bondi radius, hydrostatic densities exceed the background density (see R06).
@@
 
-While $\RH$ is always the outer boundary in our structure calculations, we quote planet masses as the enclosed mass within the smaller of $\RB$ and $\RH$.
-\textbf{When $\RB < \RH$, this means that our enclosed mass extends to a smaller radius than the outer boundary.
-This choice is motivated as follows.
+We quote planet masses as the enclosed mass within the smaller of $\RB$ and $\RH$.
+\textbf{Thus when $\RB < \RH$, our computational domain -- which always ends at $\RH$ -- extends further than the location where we define atmosphere mass, within $\RB$.
+We emphasize that this mass definition does not affect the evolutionary calculation, which consistently includes the enclosed mass at all radii.
+We justify our mass convention as follows.
 First, we want to conservatively define planet mass in a way that does not exaggerate atmospheric growth and is more consistent with previous works that use $\RB$ as the outer boundary in this regime (e.g., \citealt{ikoma00}, \citealt{pollack96}).
-Second, most of the gas outside $\RB$ is weakly bound, and unlikely to remain attached to the planet if the disk suddenly dissipates.
-Our convention for enclosed mass does not affect our structure and evolution calculations, which use the physical mass at every radius.}
+Second, most of the gas outside $\RB$ is weakly bound, and unlikely to remain attached to the planet if the disk suddenly dissipates.}
+
+\textbf{The choice of $\RH$ as the outer boundary is based on this being the largest radius where spherical hydrostatic balance is approximately (but not exactly, see \S\ref{sec:hydro}) valid.
+When $\RB < \RH$, we thus include the fact that the density at $\RB$ slightly exceeds the background disk density (see R06).
+In practice, this effect is not very significant.
+We would get similar results by choosing the outer boundary at $\RB$ in this regime, as previous studies have done.}
 
@@
 In the low mass regime, $M\pla < M_{\rm th}/ \sqrt{3}$, where $\RB < \RH$, we assume that hydrostatic balance holds out to the outer boundary at $\RH$.
-In this low mass regime, \citet[]{Orm13} calculated the flow patterns driven by stellar tides and disk headwinds.
+In this low mass regime, \citet{Orm13} calculated the 2D (radial and azimuthal) flow patterns driven by stellar tides and disk headwinds.
 On scales $\gtrsim \RB$ the flows no longer circulate the planet: they belong to the disk.
-Even if these flows do not destroy hydrostatic balance, they could affect the planet's cooling.
-We expect such effects to be weak, as heat losses at greater depths dominate planetary cooling, but more study is needed.
+While the density structure still appears roughly spherical and hydrostatic, these flows could affect the planet's cooling.
+We expect such effects to be weak, as heat losses at greater depths dominate planetary cooling, but more study is needed, especially in 3D.
 

### 2014-03-13 tex/PaperI/ms.tex
some Paper II updates + accepted version of Paper I

@@
 We determine the minimum core mass for a giant planet to form within a typical disk lifetime of 3 Myr.
-The minimum core mass declines with disk radius, from $\sim$$8.5 M_{\oplus}$ at 5 AU to $\sim$$3.5 M_{\oplus}$ at 100 AU, with standard \textbf{interstellar grain} opacities.
+The minimum core mass declines with disk radius, from $\sim$$8.5 M_{\oplus}$ at 5 AU to $\sim$$3.5 M_{\oplus}$ at 100 AU, with standard interstellar grain opacities.
 Lower temperatures in the outer disk explain this trend, while variations in disk density are less influential.
@@
 Thus the luminosity in the radiative exterior is held spatially constant.
-\textbf{We justify this assumption in \S\ref{sec:twolayer} and confirm its validity in \S\ref{sec:endoftime}.}
+We justify this assumption in \S\ref{sec:twolayer} and confirm its validity in \S\ref{sec:endoftime}.
 %\item A global cooling model connects independent static solutions into a time-dependent sequence.
-\item The atmosphere obeys a polytropic equation of state (EOS), \textbf{with adiabatic index $\gamma = 7/5$ for an ideal diatomic gas.
-Corrections from a realistic EOS are discussed in \S\ref{sec:EOS} and deferred to future work.}
+\item The atmosphere obeys a polytropic equation of state (EOS), with adiabatic index $\gamma = 7/5$ for an ideal diatomic gas.
+Corrections from a realistic EOS are discussed in \S\ref{sec:EOS} and deferred to future work.
 \item Dust grains provide the opacity in the radiative zone.
-\textbf{For the cool temperatures in the outer disk, the radiative layer remains cool enough to avoid dust sublimation.} Other opacity choices are discussed in \S\ref{sec:op}.
+For the cool temperatures in the outer disk, the radiative layer remains cool enough to avoid dust sublimation.
+Other opacity choices are discussed in \S\ref{sec:op}.
 %\item Because gas accretion accelerates after the crossover mass, the time to reach the crossover mass, i.e. the crossover time, is a good approximation of the total time to form a gas giant.
@@
 We quote planet masses as the enclosed mass within the smaller of $\RB$ and $\RH$.
-\textbf{Thus when $\RB < \RH$, our computational domain -- which always ends at $\RH$ -- extends further than the location where we define atmosphere mass, within $\RB$.
+Thus when $\RB < \RH$, our computational domain -- which always ends at $\RH$ -- extends further than the location where we define atmosphere mass, within $\RB$.
 We emphasize that this mass definition does not affect the evolutionary calculation, which consistently includes the enclosed mass at all radii.
@@
 First, we want to conservatively define planet mass in a way that does not exaggerate atmospheric growth and is more consistent with previous works that use $\RB$ as the outer boundary in this regime (e.g., \citealt{ikoma00}, \citealt{pollack96}).
-Second, most of the gas outside $\RB$ is weakly bound, and unlikely to remain attached to the planet if the disk suddenly dissipates.}
-
-\textbf{The choice of $\RH$ as the outer boundary is based on this being the largest radius where spherical hydrostatic balance is approximately (but not exactly, see \S\ref{sec:hydro}) valid.
+Second, most of the gas outside $\RB$ is weakly bound, and unlikely to remain attached to the planet if the disk suddenly dissipates.
+
+The choice of $\RH$ as the outer boundary is based on this being the largest radius where spherical hydrostatic balance is approximately (but not exactly, see \S\ref{sec:hydro}) valid.
 When $\RB < \RH$, we thus include the fact that the density at $\RB$ slightly exceeds the background disk density (see R06).
 In practice, this effect is not very significant.
-We would get similar results by choosing the outer boundary at $\RB$ in this regime, as previous studies have done.}
+We would get similar results by choosing the outer boundary at $\RB$ in this regime, as previous studies have done.
 
@@
 In the radiative zone,  local energy balance, \Eq{eq:structd}, does affect the atmospheric structure.
-We proceed by assuming that the majority of energy is lost from the convective interior, and thus the luminosity can be treated as constant in the outer radiative zone, \textbf{i.e.
+We proceed by assuming that the majority of energy is lost from the convective interior, and thus the luminosity can be treated as constant in the outer radiative zone, i.e.
 the RHS of \Eq{eq:structd} is set to zero.
@@
 We show in \S\ref{sec:endoftime} that the approximation of constant luminosity in the outer layer holds for our regime of interest.
-}
 
@@
 Specifically, we alter the luminosity until the integrated value of mass at the core, $m(R\co)$, matches the actual core mass, $M\co$.
-\textbf{Physically, the luminosity determines the location of the radiative-convective boundary (RCB), consistent with the structure equations and the Schwarzschild criterion.}
-
-\textbf{To understand time evolution, we construct} an array of solutions to Equations (\ref{eq:struct}a -- c) in order of increasing atmospheric mass.
+Physically, the luminosity determines the location of the radiative-convective boundary (RCB), consistent with the structure equations and the Schwarzschild criterion.
+
+To understand time evolution, we construct an array of solutions to Equations (\ref{eq:struct}a -- c) in order of increasing atmospheric mass.
 We then use global energy balance, \Eq{eq:coolingglobal}, to ``follow the mass" and place these solutions in a cooling sequence.
@@
 The analytic model also ignores the surface terms in \Eq{eq:coolingglobal}, which we show to be a modest correction in \S\ref{sec:endoftime}.
-\textbf{Finally, while the numerical models set the outer boundary at $\RH$, we simplify the analytic calculations by setting the outer boundary to infinity, effectively neglecting the finite scale-height of the disk.
-As shown in \App{iso}, the effect of this approximation is minor for $\RB \lesssim \RH$, the primary case of interest (see also R06).}
-
-\textbf{Of all the approximations, the neglect of self-gravity is by far the most significant, as we have verified by comparison to numerical integrations that only neglect self-gravity.}
+Finally, while the numerical models set the outer boundary at $\RH$, we simplify the analytic calculations by setting the outer boundary to infinity, effectively neglecting the finite scale-height of the disk.
+As shown in \App{iso}, the effect of this approximation is minor for $\RB \lesssim \RH$, the primary case of interest (see also R06).
+
+Of all the approximations, the neglect of self-gravity is by far the most significant, as we have verified by comparison to numerical integrations that only neglect self-gravity.
 
@@
 
-To facilitate comparison with our numerical results, \textbf{we rescale the analytic results as follows.
+To facilitate comparison with our numerical results, we rescale the analytic results as follows.
 Since growth times are too slow in the non-self-gravitating analytic model, we modify the runaway growth criterion to occur at an effective crossover mass $M_{\rm atm} = f M\co$, with $f < 1$.
@@
 Rather by artificially accelerating growth in the analytic model, we can more easily compare the parameter scalings of the two models.
-Mathematically, the modified crossover mass amounts to replacing $\xi \rightarrow f\xi$ in our analysis.}  With $f = 0.13$, the revised scalings are
+Mathematically, the modified crossover mass amounts to replacing $\xi \rightarrow f\xi$ in our analysis.
+With $f = 0.13$, the revised scalings are
 \begin{subeqnarray}
@@
 
-\textbf{The atmospheric structure is well approximated by our non-self-gravitating, analytic solutions.
+The atmospheric structure is well approximated by our non-self-gravitating, analytic solutions.
 Deep in convective zones, thermal energy is a fixed fraction of the gravitational potential energy, giving $T \propto r^{-1}$ and $P\propto r^{-1/\delad}$  as in \Eq{eq:deep}.
@@
 Closer to the RCB, self-gravity is no longer negligible, particularly for large envelope masses.
-Instead, these high mass solutions show a slightly flatter profile in $T$ and also in $P \propto T^{1/\delad}$, as explained by \Eq{eq:TPsg}.}
+Instead, these high mass solutions show a slightly flatter profile in $T$ and also in $P \propto T^{1/\delad}$, as explained by \Eq{eq:TPsg}.
 
@@
 In the higher mass solutions, the relatively low $P\cb$ values (and thus the relatively high luminosities) require an outward shift in $R\cb$, as shown in \Fig{fig:profiles}.
-\textbf{This shift does not occur in the analytic solution, where $R\cb$ continually decreases with atmospheric mass (cf.
-Equations \ref{eq:PcbRcb} and \ref{eq:crit}).}
+This shift does not occur in the analytic solution, where $R\cb$ continually decreases with atmospheric mass (cf.
+Equations \ref{eq:PcbRcb} and \ref{eq:crit}).
 
 The accelerated growth in the numerical model, as shown in the bottom panel of \Fig{fig:Ltplot}, is also a direct result of the higher cooling luminosities with self-gravity included.
-\textbf{Even when $M_{\rm atm} / M\co$ is only few percent, i.e.\ well before the crossover mass, the effect of self-gravity is  quite evident.
+Even when $M_{\rm atm} / M\co$ is only few percent, i.e.\ well before the crossover mass, the effect of self-gravity is  quite evident.
 Closer to the star, the effects of self-gravity are not as strong for low atmosphere masses.
-Nevertheless, all our models show that self-gravity noticeably accelerates growth for $M_{\rm atm} > 0.1 M\co$.}
+Nevertheless, all our models show that self-gravity noticeably accelerates growth for $M_{\rm atm} > 0.1 M\co$.
 
@@
 %We now use our structure and evolution models to calculate the necessary conditions for gas giant formation by core accretion.   
-\textbf{We now use our structure and evolution models to estimate the timescales and minimum core masses for giant planet formation for a range of disk conditions and other model parameters.
+We now use our structure and evolution models to estimate the timescales and minimum core masses for giant planet formation for a range of disk conditions and other model parameters.
 Our results for atmosphere growth times -- the time for a core of fixed mass to undergo runaway gas accretion -- are presented in \S\ref{sec:tcross}.
-Section \ref{sec:critcore} gives our results for critical core masses, the minimum values that trigger runaway atmospheric growth within a plausible disk lifetime, here 3 Myr.}
+Section \ref{sec:critcore} gives our results for critical core masses, the minimum values that trigger runaway atmospheric growth within a plausible disk lifetime, here 3 Myr.
 
@@
 A lower $\mu$ gives longer growth times, because more cooling is required to compress the atmosphere.
-\textbf{This effect is both well established in core accretion studies \citep{stevenson82} and intuitive since the atmospheric scale height $\propto 1/\mu$ is more extended for lower $\mu$.
+This effect is both well established in core accretion studies \citep{stevenson82} and intuitive since the atmospheric scale height $\propto 1/\mu$ is more extended for lower $\mu$.
 Moreover, the Bondi radius decreases as $\RB \propto \mu$, giving a smaller gravitational sphere of influence (and a weaker compression at $\RH$ when that is the more relevant scale).
 To see how this affects cooling times, note that the characteristic RCB depth near runaway scales as $P\cb \sim P_{\rm M} \propto \mu^{-4}$ from \Eq{eq:PM}.
-Thus $L \propto 1/P\cb \propto \mu^4$ explains the trend of slower cooling for lower $\mu$.}
+Thus $L \propto 1/P\cb \propto \mu^4$ explains the trend of slower cooling for lower $\mu$.
 %he larger atmospheric scale height due to a lower $\mu$ also results in a larger pressure depth at the RCB, $P\cb$, and thus a lower $R\cb$, cf. \Eq{eq:PcbRcb}.}
@@
 Thus fairly drastic changes in atmospheric composition are required for $\mu$ to significantly affect core accretion timescales.
-\textbf{In principle, changes in the mean molecular weight of the gas also affect the EOS, including $\delad$, but such effects are not considered here (see \S\ref{sec:EOS}).}
+In principle, changes in the mean molecular weight of the gas also affect the EOS, including $\delad$, but such effects are not considered here (see \S\ref{sec:EOS}).
 
@@
 
-\textbf{A larger disk lifetime further reduces $M_{\rm crit}$.
+A larger disk lifetime further reduces $M_{\rm crit}$.
 Recent studies have shown that gas disks may live up to $\propto$$10-12$ Myr \citep{bell13}.
@@
 Equation \ref{eq:tcoolan}), a disk lifetime of 10 Myrs decreases the critical core mass by a factor of two.
-Of course, some fraction of the disk lifetime must be allocated to the growth and possibly migration of the core.}
+Of course, some fraction of the disk lifetime must be allocated to the growth and possibly migration of the core.
 
@@
 
-\subsection{\textbf{Comparison with Previous Studies}}
+\subsection{Comparison with Previous Studies}
 \label{sec:comp}
 
-\textbf{We can directly compare our results with other studies of protoplanetary atmospheric growth in the absence of planetesimal accretion, notably those in I00 and PN05.
+We can directly compare our results with other studies of protoplanetary atmospheric growth in the absence of planetesimal accretion, notably those in I00 and PN05.
 A major distinguishing feature of our study is that we explore a range of disk conditions, while I00 and PN05 focus on growth at the location of Jupiter, 5.2 AU, in their disk models.
 Moreover, both I00 and PN05 solve the full time dependent energy equation and consider more detailed EOSs, in addition to other detailed differences in model parameters.
-Despite these differences, the qualitative and quantitative agreement with our study is good.}
-
-\textbf{I00 give an approximate analytic fit to their models' envelope formation time (the equivalent of the runaway accretion time $t_{\rm run}$ in our models), at $a=5.2$ AU, }
+Despite these differences, the qualitative and quantitative agreement with our study is good.
+
+I00 give an approximate analytic fit to their models' envelope formation time (the equivalent of the runaway accretion time $t_{\rm run}$ in our models), at $a=5.2$ AU, 
 \begin{equation}
@@
 \end{equation}
-\textbf{Our model reproduces the linear opacity dependence (see \S\ref{sec:timeev}), even though we use a temperature dependent opacity, \Eq{eq:opacitylaw}, instead of a constant $\kappa$.
-Our growth times at 5 AU,}
+Our model reproduces the linear opacity dependence (see \S\ref{sec:timeev}), even though we use a temperature dependent opacity, \Eq{eq:opacitylaw}, instead of a constant $\kappa$.
+Our growth times at 5 AU,
 \begin{equation}
@@
 \end{equation} 
-\textbf{agree well with the I00 results, both in magnitude and scaling with core mass.}
-
-\textbf{One goal of our study was to explain the extended luminosity minimum in evolutionary models that characterizes phase 2 in core accretion models, as described in the introduction.
+agree well with the I00 results, both in magnitude and scaling with core mass.
+
+One goal of our study was to explain the extended luminosity minimum in evolutionary models that characterizes phase 2 in core accretion models, as described in the introduction.
 Previous work (see Figure 2 in I00 and Figures 3 and 4 in PN05) shows that this luminosity minimum is an intrinsic feature of atmospheric cooling even in the absence of planetesimal accretion.
-We not only reproduce this effect in our simplified model (see \Figs{fig:Ltplot}{fig:LtvsMopacity}), but we also show that self-gravity is the essential ingredient to produce a broad luminosity minimum well before the crossover mass.}
-
-\textbf{Time-dependent studies that incorporate planetesimal accretion generally find larger formation timescales and critical core masses than those of this work.
+We not only reproduce this effect in our simplified model (see \Figs{fig:Ltplot}{fig:LtvsMopacity}), but we also show that self-gravity is the essential ingredient to produce a broad luminosity minimum well before the crossover mass.
+
+Time-dependent studies that incorporate planetesimal accretion generally find larger formation timescales and critical core masses than those of this work.
 This result is expected since additional energy from planetesimal accretion limits the ability of the atmosphere to cool.
@@
 These values, however, decrease to $\sim$$3$ Myrs and $\sim$$12 M_{\oplus}$ if planetesimal accretion is entirely shut off during the gas accretion phase.
-The core accretion model of \citet{HubBod05} for Jupiter's formation predicts an evolutionary time of 3.3 Myrs for a $10 M_{\oplus}$ core, for an interstellar opacity, which is also consistent with our result at 5 AU.}
-
-\textbf{A direct comparison with studies that only include planetesimal accretion, and neglect atmospheric cooling, is difficult.
-Note, however, that I00 establishes the correspondence between the minimum luminosity in evolutionary models and the minimum planetesimal accretion rate needed in static models, which gives rise to the classical critical core mass of static models \citep{mizuno78, stevenson82}.}
-
-\textbf{In terms of modern static studies, our results complement (and borrow some tools from) R06 and \citet{Raf11}.
+The core accretion model of \citet{HubBod05} for Jupiter's formation predicts an evolutionary time of 3.3 Myrs for a $10 M_{\oplus}$ core, for an interstellar opacity, which is also consistent with our result at 5 AU.
+
+A direct comparison with studies that only include planetesimal accretion, and neglect atmospheric cooling, is difficult.
+Note, however, that I00 establishes the correspondence between the minimum luminosity in evolutionary models and the minimum planetesimal accretion rate needed in static models, which gives rise to the classical critical core mass of static models \citep{mizuno78, stevenson82}.
+
+In terms of modern static studies, our results complement (and borrow some tools from) R06 and \citet{Raf11}.
 These studies consider the disk radius dependence of core accretion for protoplanets that continuously accrete planetesimals.
 We show that the limits on core accretion at large radial distance claimed by \cite{Raf11} (and references therein) can be overcome if planetesimal accretion shuts off and the atmosphere is allowed to cool.
-Nevertheless, the plausibility of rapid core growth followed by negligible subsequent solid accretion admittedly remains uncertain, as discussed in more detail in \S\ref{sec:conclusions}.}
+Nevertheless, the plausibility of rapid core growth followed by negligible subsequent solid accretion admittedly remains uncertain, as discussed in more detail in \S\ref{sec:conclusions}.
 
 \section{Neglected Effects}\label{sec:neglected}
-\textbf{Since one goal of this paper was to obtain a detailed understanding of atmospheric evolution with a simple model, we have necessarily ignored   effects of potential significance.
-We briefly address the most important of these and note that a followup work, Piso, Youdin \& Murray-Clay (in prep., hereafter Paper II), will extend our current models to address some of these effects.}
+Since one goal of this paper was to obtain a detailed understanding of atmospheric evolution with a simple model, we have necessarily ignored   effects of potential significance.
+We briefly address the most important of these and note that a followup work, Piso, Youdin \& Murray-Clay (in prep., hereafter Paper II), will extend our current models to address some of these effects.
 
@@
 
-\textbf{The importance of envelope opacity is an established factor in core accretion calculations (e.g., \citealt{stevenson82}, \citealt{ikoma00}, R06), and several studies explore the influence of opacity on the timescales for giant planet formation in detail (e.g., \citealt{HubBod05})}.
+The importance of envelope opacity is an established factor in core accretion calculations (e.g., \citealt{stevenson82}, \citealt{ikoma00}, R06), and several studies explore the influence of opacity on the timescales for giant planet formation in detail (e.g., \citealt{HubBod05}).
 Treating dust (and total) opacities as a power law in temperature is a simplification.
@@
 Grain growth and composition also affect opacities.
-\textbf{Our scenario of no ongoing accretion of solids may result in a grain-free atmosphere, which could substantially reduce the critical core mass ($\sim$$1M_{\oplus}$ in the case of Jupiter, \citealt{hori10}).} Paper II explores more realistic opacity laws.
+Our scenario of no ongoing accretion of solids may result in a grain-free atmosphere, which could substantially reduce the critical core mass ($\sim$$1M_{\oplus}$ in the case of Jupiter, \citealt{hori10}).
+Paper II explores more realistic opacity laws.
 
@@
 Initial core growth must be fast, compared to the disk lifetime, to get a sufficiently massive core.
-\textbf{Such rapid core growth is possible in a variety of scenarios, including the fastest gas-free planetesimal accretion rates \citep{dones93} and -- probably more relevantly for gas rich disks --  the aerodynamic accretion of mm-m sized ``pebbles" and ``boulders" in gas disks \citep{ormel10, lambrechts12}.
-Moreover, cores could form rapidly closer to the star, then migrate or be scattered outwards by already formed giants \citep{ida13}.}
-
-\textbf{A stronger constraint is that} core growth subsequently slow severely, to allow the atmosphere to cool and contract.
+Such rapid core growth is possible in a variety of scenarios, including the fastest gas-free planetesimal accretion rates \citep{dones93} and -- probably more relevantly for gas rich disks --  the aerodynamic accretion of mm-m sized ``pebbles" and ``boulders" in gas disks \citep{ormel10, lambrechts12}.
+Moreover, cores could form rapidly closer to the star, then migrate or be scattered outwards by already formed giants \citep{ida13}.
+
+A stronger constraint is that core growth subsequently slow severely, to allow the atmosphere to cool and contract.
 To be more quantitative, the minimum cooling luminosity in \Fig{fig:Ltplot}, $L\approx 3.5 \times 10^{24}\;\mathrm{erg s}^{-1}$, could be cancelled by low levels of heating from solid accretion.
 A core mass doubling timescale of $\sim$$400$ Myr, or faster, would thus provide enough heating to stall atmospheric cooling and growth.
-\textbf{An additional concern is that isolation masses tend to grow with disk radius, as $M_{\rm iso} \propto \varSigma_{\rm p}^{3/2} a^2 \propto a^{3/4}$ under the approximation that the surface density of accreted planetesimals ($\varSigma_{\rm p}$) scales with the gas \citep{youdin13}.
+An additional concern is that isolation masses tend to grow with disk radius, as $M_{\rm iso} \propto \varSigma_{\rm p}^{3/2} a^2 \propto a^{3/4}$ under the approximation that the surface density of accreted planetesimals ($\varSigma_{\rm p}$) scales with the gas \citep{youdin13}.
 While this behavior is nominally inconsistent with final core masses that decline with distance, the predictive power of the isolation mass is imperfect.
@@
 Moreover, the locality of core growth, which underlies the isolation mass, disappears when accreted solids drift and/or cores migrate significantly.
-}
 

### 2014-03-13 tex/draft_paper2/draft_paper2.tex
some Paper II updates + accepted version of Paper I

@@
 
-\title{Minimum Core Masses for Giant Planet Formation Using Realistic Equation of State Tables and Opacities}
+\title{Minimum Core Masses for Giant Planet Formation: Using Realistic Equations of State and Opacities}
 \author{Ana-Maria A. Piso}
@@
 
-The core accretion model postulates that giant planets form through gas accretion onto a solid core.
-To accumulate a massive atmosphere, this core's mass must exceed a minimum value $M_{\rm crit}$ often quoted as $10 M_{\oplus}$.
-The value of $M_{\rm crit}$, however, depends on the rate at which planetesimals accrete onto the core and heat its atmosphere.
-In this study we calculate the minimum core mass for atmospheric collapse in the limiting regime in which core growth is halted and the luminosity evolution of a core's atmosphere is dominated by Kelvin-Helmholtz contraction.
-We build on the idealized model of Piso \& Youdin (submitted), which derives the evolution of atmospheres accreting around protoplanetary cores and calculates the minimum (or critical) core mass to form a gas giant before disk dissipation.
+We calculate the minimum core mass $M_{\rm crit}$ needed to form a giant planet through the core accretion mechanism before the dissipation of the protoplanetary disk.
+We build on the idealized model of Piso \& Youdin (ApJ, accepted), which derives the evolution of atmospheres accreting around protoplanetary cores and calculates $M_{\rm crit}$ in the limiting regime in which core growth is halted and the luminosity evolution of a core's atmosphere is dominated by Kelvin-Helmholtz contraction.
 We enhance this model by employing a more realistic equation of state and dust opacity.
@@
 If $n \propto a^{-2.5}$, appropriate for coagulation, $M_{\rm crit}$ is up to one order of magnitude smaller.
+The drop in $M_{\rm crit}$ with semimajor axis is less sharp when grain growth opacities are used than in the case of interstellar opacities.
 We show that giant planets may grow from smaller cores if core growth is halted, which is particularly relevant for understanding the origin of gas giants directly imaged at wide separations.
+
+%The core accretion model postulates that giant planets form through gas accretion onto a solid core.
+To accumulate a massive atmosphere, this core's mass must exceed a minimum value $M_{\rm crit}$ often quoted as $10 M_{\oplus}$.
+The value of $M_{\rm crit}$, however, depends on the rate at which planetesimals accrete onto the core and heat its atmosphere.
+
+ %In this study we calculate the minimum core mass for atmospheric collapse in the limiting regime in which core growth is halted and the luminosity evolution of a core's atmosphere is dominated by Kelvin-Helmholtz contraction.
 
@@
 An \textit{in situ} solution to this problem requires accelerating core growth to rates larger than those typically assumed, but fast growth produces an additional difficulty.
-Though the minimum, or critical, core mass to form a giant planet is typically quoted to be $M_{\rm crit}\sim 10M_{\oplus}$, the actual value depends on how quickly planetesimals accrete (e.g., \citealt{ikoma00}, \citealt{pollack96}, \citealt{rafikov06}).
+Though the minimum, or critical, core mass to form a giant planet is often quoted as $M_{\rm crit}\sim 10M_{\oplus}$, the actual value depends on how quickly planetesimals accrete (e.g., \citealt{ikoma00}, \citealt{pollack96}, \citealt{rafikov06}).
 Fast accretion heats a core's  atmosphere, increasing pressure support and hence $M_{\rm crit}$.
@@
 As a result, the core's envelope is in a steady state at all times, in which all the incoming energy due to planetesimals is radiated away.
-As the envelope and core become comparable in mass, hydrostatic balance no longer holds and runaway gas accretion commences.
+%As the envelope and core become comparable in mass, hydrostatic balance no longer holds and runaway gas accretion commences.
 In this case, $M_{\rm crit}$ is the maximum core mass for which the atmosphere is still in hydrostatic equilibrium.
@@
 
-We follow the model developed in Piso \& Youdin (submitted; hereafter Paper I) and consider quasi-static two-layer atmospheres, embedded in the protoplanetary disk.
+We follow the model developed in Piso \& Youdin (accepted to ApJ; hereafter Paper I) and consider quasi-static two-layer atmospheres, embedded in the protoplanetary disk.
 Paper I calculates the minimum (critical) core mass to form a gas giant at several locations in the protoplanetary disk.
@@
 
-\vspace{0.2in}
-
-\textbf{1. Intermediate T: Ideal Gas}
+%\vspace{0.2in}
+
+%\textbf{1. Intermediate T: Ideal Gas}
+
+\subsection{Intermediate $T$: Ideal Gas}
 
@@
 
-\vspace{0.2in}
-
-\textbf{2. High T: Dissociation and Ionization of Hydrogen.}
-
-At low temperatures, hydrogen is molecular and has a stable configuration.
-At temperatures higher than $T \sim 2000-3000$ K, the internal energy of the hydrogen molecule becomes large enough to break the covalent bond between the atoms, and hydrogen starts dissociating.
+%\vspace{0.2in}
+
+%\textbf{2. High T: Dissociation and Ionization of Hydrogen.}
+
+\subsection{High $T$: Dissociation and Ionization of Hydrogen}
+
+At low temperatures, hydrogen is molecular, with a stable configuration.
+At temperatures higher than $T \sim 2000-3000$ K, the internal energy of the hydrogen molecule is large enough for hydrogen to dissociate.
 At $T \gtrsim 10,000$ K, hydrogen ionizes.
@@
 
-As displayed in Figure \ref{fig:deladmap}, the adiabatic gradient decreases significantly in regions of partial dissociation and partial ionization.
-This behavior is different than that of a mixture of molecular and atomic hydrogen, for which $2/7<\delad<2/5$, or a mixture of protons and electrons, for which $\delad=2/5$.
-When a gas dissociates, part of the internal energy of the system is used to break down molecules, which reduces the amount of energy available to increase the temperature of the system, thus lowering the adiabatic gradient.
+As displayed in Figure \ref{fig:deladmap}, the adiabatic gradient decreases significantly in regions of partial dissociation and partial ionization.When a gas dissociates, part of the internal energy of the system is used to break down molecules, which reduces the amount of energy available to increase the temperature of the system, thus lowering the adiabatic gradient.
 The energy required to dissociate depends on the dissociation fraction, which can be determined from the Saha equation (see e.g., \citealt{kippenhahn90}).
 The dissociation fraction only depends on gas temperature and density, and hence only on the EOS.
+
+ %This behavior is different than that of a mixture of molecular and atomic hydrogen, for which $2/7<\delad<2/5$, or a mixture of protons and electrons, for which $\delad=2/5$.
 
@@
 
-\vspace{0.2in}
-
-\textbf{3. Low T: Hydrogen Rotation and Spin Isomers}
-
-As a diatomic molecule, hydrogen has five degrees of freedom, three associated with translational motion and two associated with rotation.
+%\vspace{0.2in}
+
+%\textbf{3. Low T: Hydrogen Rotation and Spin Isomers}
+
+\subsection{Low $T$: Hydrogen Rotation and Spin Isomers}
+
+The (diatomic) hydrogen molecule has five degrees of freedom, three associated with translational motion and two associated with rotation.
 The excitation temperature for rotation is $\Theta_r \approx 85$ K \citep{kittel}, hence the rotational states are fully excited at room temperature.
@@
 From the Pauli exclusion principle it follows that it can only occupy symmetric rotational states with even angular quantum number $j$ \citep{farkas35}.
-In contrast, orthohydrogen has parallel proton spins and a symmetric wave function, and can therefore only occupy states with odd $j$.
+In contrast, orthohydrogen has parallel proton spins and a symmetric wave function, and can thus only occupy states with odd $j$.
 At equilibrium, the relative abundance of the ortho- and para- states is given by the ratio of their partition functions, described in \App{EOStables}.
@@
 This is consistent with the energy behavior in Figure \ref{fig:all_plot}, middle-left panel.
-
-We note that the amount of energy that needs to be released in order to accrete gas at the outer boundary is going to be roughly equal for all EOSs, as it has to match the outer boundary conditions of the nebula.
-However, a more negative total energy in the interior, as is the case for the high-$T$ EOS, requires a larger amount of energy to be radiated away to bind the next batch of gas.
+It follows that a more negative total energy in the interior, as is the case for the high-$T$ EOS, requires a larger amount of energy to be radiated away to bind the next batch of gas.
 This implies a larger $-dE/dM$ during atmospheric evolution (middle-right panel of Figure \ref{fig:all_plot}), and thus a lower $\dot{M}$ for the high-T EOS (lower-right panel of Figure \ref{fig:all_plot}).
+
+%We note that the amount of energy that needs to be released in order to accrete gas at the outer boundary is going to be roughly equal for all EOSs, as it has to match the outer boundary conditions of the nebula.
+However,
 
@@
 The analytic profile of Equation (\ref{eq:Tdeep}) is over-plotted for comparison (\textit{dotted line}) in regions of agreement.
-The analytic temperature correction at the RCB (\textit{dashed}) is in good agreement with the numerical results.
 A larger $\nabla_{\rm ad,2}$ yields a deeper radiative region.
@@
 The analytic profile of Equation (\ref{eq:Tdeep}) is over-plotted for comparison (\textit{dotted line}) in regions of agreement.
-The analytic temperature correction at the RCB (\textit{dashed}) is in good agreement with the numerical results.
 A lower $\nabla_{\rm ad,1}$ yields a more shallow radiative region.
@@
 
-The EOS variations in the cold outer envelope affect $\dot{M}$ by decreasing the atmosphere's luminosity compared to an ideal gas polytrope (Figure \ref{fig:all_plot}, upper-right panel, for the same parameters as Figure \ref{fig:tplotall}).
+%The EOS variations in the cold outer envelope affect $\dot{M}$ by decreasing the atmosphere's luminosity compared to an ideal gas polytrope (Figure \ref{fig:all_plot}, upper-right panel, for the same parameters as Figure \ref{fig:tplotall}).
 From Figures \ref{fig:deladmap} and \ref{fig:all_plot}, bottom-left panel, we see that the adiabatic gradient of the low-$T$ realistic EOS increases with radius (i.e., with decreasing temperature) throughout the convective zone when $T<500$ K, then becomes lower near the RCB.
 This results in a larger adiabatic gradient, on average, for the low-$T$ realistic EOS in the outer parts of the convective region of the atmosphere.
+
+The variations in adiabatic gradient at low temperatures decrease the atmosphere's luminosity compared to an ideal gas polytrope.
+The adiabatic gradient for the low-$T$ realistic EOS increases with radius (i.e., with decreasing temperature) throughout the convective zone, then becomes lower near the RCB.
+This results in a lower $\delad$, on average, for the low-$T$ EOS in the outer parts of the envelope.
 
@@
 Moreover, a constant $\nabla_{\rm ad,2}$ in the exterior yields the same value for $T\cb$, independent of $\nabla_{\rm ad,1}$ (Figure \ref{fig:varying_delad}, upper-right panel), again consistent with the analytic prediction.
+Quantitatively, we have found that the analytic approximation for $T\cb$ agrees with the numerical result to within 5\%.
 
@@
 Despite the existence of one or more radiative windows in the planet's atmosphere, the luminosity that emerges at the outer boundary is set by the outer radiative layer.
-A simple estimate of the luminosity emerging at the RCB can be found as follows.
-
-The equation of radiative diffusion (\ref{eq:structc}) can be approximated in terms of the radiative flux $\mathcal{F}$ as
-
+To order of magnitude, Equation (\ref{eq:delrad}) can be expressed as
 \begin{equation}
@@
 \end{equation}
-where $\tau$ is the optical depth of the upper radiative layer. Furthermore, we find that $\tau$ is dominated by the optical depth of the first scale height, i.e.
-
-\begin{equation}
-\label{eq:taucb}
-\tau \sim \tau\cb=\kappa\cb \rho\cb H\cb,
-\end{equation}
-with
-\begin{equation}
-\label{eq:Hcb}
-H\cb=\frac{k_B T\cb}{\mu m_p} \frac{R\cb^2}{G M\cb},
-\end{equation}
-where $\mu$ is the mean molecular weight of the gas, $m_p$ the proton mass, and all other quantities are evaluated at the RCB.
+where $\mathcal{F}$ is the radiative flux through the RCB and $\tau \sim \kappa\cb \rho\cb H\cb$ is the optical depth of the first scale height of the radiative region, $H\cb \equiv k_{\rm b} T R^2/({\mu m_{\rm p}} G M)$, with all quantities evaluated at the RCB.
+While Equation (\ref{eq:delrad}) is always true as part of the structure Equations set (\ref{eq:struct}), Equation (\ref{eq:sigmatau4}) shows that the atmospheric luminosity is set by the diffusive flux through the outermost RCB.
+We have verified that Equation (\ref{eq:sigmatau4}) is valid for our model atmospheres.
+
+%A simple estimate of the luminosity emerging at the RCB can be found as follows. 
+%
+%The equation of radiative diffusion (\ref{eq:structc}) can be approximated in terms of the radiative flux $\mathcal{F}$ as 
+%
+%\begin{equation}
+%\label{eq:sigmatau4}
+%\mathcal{F} \sim \frac{\sigma T\cb^4}{\tau},
+%\end{equation}
+%where $\tau$ is the optical depth of the upper radiative layer. Furthermore, we find that $\tau$ is dominated by the optical depth of the first scale height, i.e. 
+%
+%\begin{equation}
+%\label{eq:taucb}
+%\tau \sim \tau\cb=\kappa\cb \rho\cb H\cb,
+%\end{equation}
+%with
+%\begin{equation}
+%\label{eq:Hcb}
+%H\cb=\frac{k_B T\cb}{\mu m_p} \frac{R\cb^2}{G M\cb},
+%\end{equation}
+%where $\mu$ is the mean molecular weight of the gas, $m_p$ the proton mass, and all other quantities are evaluated at the RCB.
 By substituting Equations (\ref{eq:Hcb}) and (\ref{eq:taucb}) into Equation (\ref{eq:sigmatau4}), and noting that $\mathcal{F}=L/(4 \pi (R\cb+H\cb)^2) \approx L / (4 \pi R\cb^2)$, we recover Equation (\ref{eq:delrad}) applied at the boundary between the uppermost convective region of the atmosphere and the outer radiative layer (where $\delrad=\delad$) as
-
-\begin{equation}
-\label{eq:Lapprox}
-L \sim \frac{4 \pi \sigma T\cb^3 G M\cb \mu m_p}{k_B \kappa\cb \rho\cb}.
-\end{equation}
-Equation (\ref{eq:Lapprox}) can be used to estimate the luminosity emerging at the top of the convective layer of the atmosphere.
+%
+%\begin{equation}
+%\label{eq:Lapprox}
+%L \sim \frac{4 \pi \sigma T\cb^3 G M\cb \mu m_p}{k_B \kappa\cb \rho\cb}.
+%\end{equation}
+%Equation (\ref{eq:Lapprox}) can be used to estimate the luminosity emerging at the top of the convective layer of the atmosphere.
 We have verified that the luminosity emerging at the bottom of the outer radiative layer in our models can indeed be approximated by Equation (\ref{eq:Lapprox}).
@@
 This approach contrasts that of static models, which assume high planetesimal accretion rates and find that the atmosphere is in steady state and solely heated due to accretion of solids.
-In this section we compare our results for the critical core mass to analogous results from steady-state fast planetesimal accretion calculations.
+As the envelope and core become comparable in mass, hydrostatic balance no longer holds and runaway gas accretion commences.
+In these models, $M_{\rm crit}$ is the maximum core mass for which the atmosphere is still in hydrostatic equilibrium.
+For a fixed planetesimal accretion rate and a set of disk conditions, $M_{\rm crit}$ is uniquely determined.
+In this section we compare our results for $M_{\rm crit}$ to analogous results from steady-state fast planetesimal accretion calculations.
 We discuss the core accretion rates that are necessary for our regime to be valid in \S\ref{raf1}.
@@
 This is consistent with the results of \citet{pollack96}, which find that the planetesimal accretion rate decreases with time as the core's feeding zone is depleted, and with the requirement of our model that the planetesimal accretion rate is initially large during core growth, then significantly reduces as the gaseous envelope accumulates.
-This is a plausible scenario, as the core may have formed in the inner part of the disk and was later scattered outwards \citep{ida13}, or the planet's feeding zone could have been depleted of solids due to a giant neighbor.
+This is a plausible scenario: the core could have formed through pebble accretion \citep{lambrechts12}, or in the inner part of the disk and was later scattered outwards by other giants in the system \citep{ida13}.
+
+%or the planet's feeding zone could have been depleted of solids due to a giant neighbor.
 
@@
 We consider atmosphere evolution around fully grown cores and determine the minimum (critical) core mass required to form a gas giant during the typical lifetime of a protoplanetary disk.
-We improve the model developed in Piso \& Youdin (submitted, hereafter Paper I) by including realistic equation of state (EOS) tables and dust opacities.
+We improve the model developed in Piso \& Youdin (accepted to ApJ, hereafter Paper I) by including realistic equation of state (EOS) tables and dust opacities.
 

### 2014-03-13 tex/draft_paper2/draft_paper2.tex
minor intro changes + ms pdf removed due to merge conflict

@@
 Fast accretion heats a core's  atmosphere, increasing pressure support and hence $M_{\rm crit}$.
-However, planetesimal accretion rates need not be constant over time.
+For high planetesimal accretion rates, atmospheres are primarily heated by planetesimals (e.g., \citealt{rafikov06}).
+In this case,  $M_{\rm crit}$ is uniquely determined for a fixed accretion rate of solids and a set of disk conditions.
+
+ %However, planetesimal accretion rates need not be constant over time.
 If  accretion slows before the nebular gas dissipates, the target $M_{\rm crit}$ for successful giant planet formation is substantially reduced.
@@
 %Fast accretion implies large critical core masses.
-In order to calculate the minimum value of $M_{\rm crit}$, we need to understand how $M_{\rm crit}$ depends on the planetesimal accretion rate.
-%the evolution of a core's atmosphere depends on the rate at which planetesimals are accreted. 
+
+%In order to calculate the minimum value of $M_{\rm crit}$, we need to understand how $M_{\rm crit}$ depends on the planetesimal accretion rate.
 For high planetesimal accretion rates, atmospheres are primarily heated by planetesimals (e.g., \citealt{rafikov06}).
 As a result, the core's envelope is in a steady state at all times, in which all the incoming energy due to planetesimals is radiated away.
+
+%the evolution of a core's atmosphere depends on the rate at which planetesimals are accreted.
+
 %As the envelope and core become comparable in mass, hydrostatic balance no longer holds and runaway gas accretion commences.
@@
 
-However, planetesimal accretion rates need not be constant at a given location in the protoplanetary disk throughout the disk's lifetime --- in particular, planetesimal accretion may significantly reduce as the planet's feeding zone is depleted of solids.
-Time-dependent core accretion models (e.g., \citealt{pollack96}, \citealt{ikoma00}) take this into account and include variable planetesimal accretion rates which allow for the Kelvin-Helmholtz (KH) contraction of the gaseous envelope to become important.
+However, planetesimal accretion rates need not be constant over time ---  in particular, planetesimal accretion may significantly reduce as the planet's feeding zone is depleted of solids.
+If  accretion slows before the nebular gas dissipates, the target $M_{\rm crit}$ for successful giant planet formation is substantially reduced.
+Moreover, $M_{\rm crit}$ is no longer uniquely determined ---  any core has the potential to initiate runaway gas accretion, provided it can accumulate a massive atmosphere before disk dissipation.
+
+In order to calculate the minimum value of $M_{\rm crit}$, we need to understand how $M_{\rm crit}$ depends on the planetesimal accretion rate.
+For high planetesimal accretion rates, the core's envelope is in steady state at all times, in which all the incoming energy due to planetesimals is radiated away.
+On the other hand, time-dependent core accretion models (e.g., \citealt{pollack96}, \citealt{ikoma00}) include variable planetesimal accretion rates which allow for the Kelvin-Helmholtz (KH) contraction of the gaseous envelope to become important.
 In this regime, the envelope is no longer in a steady state, but rather it accretes gas as it cools.
@@
 %Studying atmosphere growth onto a solid core of fixed mass (e.g., \citealt{pn05}) can thus provide accurate estimates for the timescale required to form a giant planet.
+
+%However, planetesimal accretion rates need not be constant at a given location in the protoplanetary disk throughout the disk's lifetime --- in particular, planetesimal accretion may significantly reduce as the planet's feeding zone is depleted of solids.
 

### 2014-04-03 tex/draft_paper2/draft_paper2.tex
almost final Paper II draft

@@
 
-We calculate the minimum core mass $M_{\rm crit}$ needed to form a giant planet through the core accretion mechanism before the dissipation of the protoplanetary disk.
+We calculate the minimum core mass, $M_{\rm crit}$, required for a giant planet to form by core accretion before the dissipation of its host protoplanetary disk.
 We build on the idealized model of Piso \& Youdin (ApJ, accepted), which derives the evolution of atmospheres accreting around protoplanetary cores and calculates $M_{\rm crit}$ in the limiting regime in which core growth is halted and the luminosity evolution of a core's atmosphere is dominated by Kelvin-Helmholtz contraction.
@@
 Lower opacities due to grain growth reduce $M_{\rm crit}$.
-Taken together, we calculate $M_{\rm crit} \sim 7 M_{\oplus}$ at 5 AU, decreasing to $\sim$$4.5 M_{\oplus}$ at 100 AU, for a realistic EOS and a grain size distribution $n \propto a^{-3.5}$ with a maximum particle size of 1 cm.
-If $n \propto a^{-2.5}$, appropriate for coagulation, $M_{\rm crit}$ is up to one order of magnitude smaller.
-The drop in $M_{\rm crit}$ with semimajor axis is less sharp when grain growth opacities are used than in the case of interstellar opacities.
+Taken together, we calculate $M_{\rm crit} \sim 7 M_{\oplus}$ at 5 AU, decreasing to $\sim$$4.5 M_{\oplus}$ at 100 AU, for a realistic EOS and a grain size distribution $n \propto s^{-3.5}$ with a maximum particle size $s=1$ cm.
+If $n \propto s^{-2.5}$, appropriate for coagulation, $M_{\rm crit}$ is up to one order of magnitude smaller.
+Because the opacities produced by large grains are not a strong function of temperature, $M_{\rm crit}$ varies only weakly with stellocentric distance.
+These results provide the minimum critical core mass for production of giant planets by core accretion, an important parameter for understanding the origin of gas giants directly imaged at wide separations.
+
+%The drop in $M_{\rm crit}$ with semimajor axis is less sharp when grain growth opacities are used than in the case of interstellar opacities.
 We show that giant planets may grow from smaller cores if core growth is halted, which is particularly relevant for understanding the origin of gas giants directly imaged at wide separations.
@@
 
-%Fast accretion implies large critical masses.
-Because protoplanetary disks dissipate on timescales of a few Myrs, such cores must grow quickly to have access to the nebular gas.
-If cores and their atmospheres grow simultaneously, fast core growth requires large planetesimal accretion rates and hence large critical core masses.
-Fast core growth to large critical masses poses particular challenges in the outer disk, where dynamical times are long.
-Core accretion at wide separations thus requires either growth of cores at rates faster than typically assumed or, perhaps counterintuitively, moderate core growth rates followed by a halt to planetesimal accretion.
-In this study we explore the latter possibility.
-
 Core accretion --- a prominent theory of giant planet formation --- stipulates that Jupiter-sized planets form when planetesimal accretion produces a solid core large enough to attract a massive atmosphere   (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{boden86}, \citealt{dangelo11}).
@@
 An \textit{in situ} solution to this problem requires accelerating core growth to rates larger than those typically assumed, but fast growth produces an additional difficulty.
-Though the minimum, or critical, core mass to form a giant planet is often quoted as $M_{\rm crit}\sim 10M_{\oplus}$, the actual value depends on how quickly planetesimals accrete (e.g., \citealt{ikoma00}, \citealt{pollack96}, \citealt{rafikov06}).
+Though the minimum, or critical, core mass to form a giant planet is often quoted as $M_{\rm crit}\sim 10M_{\oplus}$, the actual value depends on how quickly the core accretes planetesimals  (e.g.,  \citealt{pollack96}, \citealt{ikoma00}, \citealt{rafikov06}).
 Fast accretion heats a core's atmosphere, increasing pressure support and hence $M_{\rm crit}$.
-For high planetesimal accretion rates, atmospheres are primarily heated by planetesimals (e.g., \citealt{rafikov06}).
+\citet{rafikov11} finds that beyond 40-50 AU, a core cannot  reach $M_{\rm crit}$ while accreting at the rate it requires to grow during the disk lifetime.
+
+%Given standard core accretion rates, the $M_{\rm crit}$ you need to grow an atmosphere during the disk lifetime while the core is growing is too large, leading people to believe that core accretion doesn't work in the outer disk.
+
+However, time-dependent core accretion models, such as \citet{pollack96} and \citet{ikoma00}, suggest that a planets's feeding zone may be depleted of planetesimals before the gas disk dissipates, causing core growth to stall.
+As core growth slows, energy deposition due to planetesimal accretion no longer balances the planet's luminosity.
+The envelope cannot maintain a steady state, and it accretes gas while  undergoing Kelvin-Helmholtz (KH) contraction.
+A core  produces a giant planet if it has enough time to accumulate an atmosphere of its own mass before disk dissipation.
+Considering atmospheric growth in the absence of planetesimal accretion thus gives a minimum value for $M_{\rm crit}$.
+%, as additional heating due to accretion of solids would limit the envelope's ability to cool and increase the core mass required for collapse.
+%if core growth is stalled during the atmospheric accretion phase, the target $M_{\rm crit}$ decreases, conceivably making runaway atmospheric growth possible in the outer disk.
+
+%the envelope mass remains small during this initial phase of core growth, and runaway gas accretion cannot commence.
+Once the planet's feeding zone has been depleted of solids, planetesimal accretion slows enough that the resulting energy deposition can no longer balance radiative losses.
+
+%However, the planetesimal accretion rate, $\dot{M_{\rm c}}$, need not be constant over time and may significantly reduce as the planet's feeding zone is depleted of solids.
+
+%While a core is accreting the majority of its mass, $\dot{M_{\rm c}}$ is large, and the energy its atmosphere receives from infalling planetesimals balances its luminosity.
+The core's envelope is in steady state, and $M_{\rm crit}$ is uniquely determined for a fixed $\dot{M_{\rm c}}$ and a set of disk conditions.
+Time-dependent core accretion models, such as \citet{pollack96} and \citet{ikoma00}, suggest that the envelope mass remains small during this initial phase of core growth, and runaway gas accretion cannot commence.
+Once the planet's feeding zone has been depleted of solids, planetesimal accretion slows enough that the resulting energy deposition can no longer balance radiative losses.
+Core growth stalls, and the atmosphere cools, undergoes Kelvin-Helmholtz (KH) contraction, and steadily accretes gas.
+In this accretion regime,  $M_{\rm crit}$ is no longer unique ---  any core may produce a giant planet if it has enough time to accumulate an atmosphere of comparable mass to itself before disk dissipation.
+Given the accretion history employed by \cite{pollack96}, which neglects drift of planetesimals through the gas disk, the atmospheric cooling phase lasts much longer than the phase of core growth, and so the the time requires to form a giant planet is set by KH contraction.
+
+%It is not clear \textit{a priori} which regime --- atmosphere evolution dominated by planetesimal accretion versus KH contraction --- is the most relevant, as the conclusion of \citet{pollack96} depends on their assumptions about planetesimal accretion history.
+In particular, a core's feeding zone may be replenished by migration of solids through the protoplanetary disk.
+
+%This study does not model an accretion history, but instead calculates a minimum value for $M_{\rm crit}$.
+We accept the two-stage growth suggested by time-dependent models, in which the core is fully formed before it accumulates a significant atmosphere.
+During the slow phase of
+
+%In this paper, we calculate the $M_{\rm crit}$ required to nucleate runaway atmospheric growth for a fully formed core undergoing minimal planetesimal accretion.
+We assume that the luminosity of the envelope is solely due to KH contraction.
+As additional heating due to accretion of solids would limit the envelope's ability to cool, and thus increase the core mass required for collapse, our results give a robust minimum for the core mass needed to form a giant planet before disk dissipation.
+
+%Our results yield lower core masses than studies that assume large planetesimal accretion rates, confirming that if cores stop growing, they can accrete massive atmospheres more effectively.
+Moreover,
+%However, planetesimal accretion rates need not be constant over time.
+This is supported by time-dependent core accretion models, such as \citet{pollack96}, \citet{ikoma00}, which suggest that the evolution of a giant planet can be separated into three phases.
+In phase 1, the core forms via runaway planetesimal accretion, while the envelope mass stays small.
+Once the planet's feeding zone has been depleted of solids, planetesimal accretion is significantly reduced and can no longer balance radiative losses.
+As a result, core growth is stalled and the atmosphere cools while undergoing Kelvin-Helmholtz (KH) contraction; this is phase 2.
+The envelope mass steadily increases.
+When it becomes comparable to the core mass, runaway gas accretion begins in phase 3.
+\citet{pollack96} find that phase 2 lasts much longer than phases 1 and 3, and so the evolutionary timescale of the planet is set by KH contraction.
+
+%It is not clear \textit{a priori} which regime --- atmosphere evolution dominated by planetesimal accretion versus KH contraction --- is the most relevant, as the conclusion of \citet{pollack96} depends on their assumptions about planetesimal accretion history.
+In particular, a core's feeding zone may be replenished by migration of solids through the protoplanetary disk.
+This study does not model an accretion history, but instead calculates a minimum value for $M_{\rm crit}$.
+To this purpose, we must first understand how $M_{\rm crit}$ depends on the planetesimal accretion rate, $\dot{M_{\rm c}}$.
+For large $\dot{M_{\rm c}}$, the energy due to planetesimals balances the planet's luminosity and the core's envelope is in steady state.
+In this case, $M_{\rm crit}$ is uniquely determined for a fixed $\dot{M_{\rm c}}$ and a set of disk conditions.
+For small $\dot{M_{\rm c}}$, this energy balance is no longer maintained and the gaseous envelope loses energy via KH contraction.
+Moreover, $M_{\rm crit}$ is no longer unique ---  any core may initiate runaway gas accretion if it can accumulate a massive atmosphere before disk dissipation.
+
+%In this study, we accept the three-stage growth, in which the core is fully formed before it accumulates a significant atmosphere.
+During the slow phase of atmospheric growth we thus assume that the luminosity of the envelope is only due to KH contraction and neglect ongoing planetesimal accretion.
+Our results yield lower core masses than studies that assume large planetesimal accretion rates, confirming that if cores stop growing, they can accrete massive atmospheres more effectively.
+Moreover, as additional heating due to accretion of solids limits the envelope's ability to cool and thus increases the core mass required for collapse, our results give a robust minimum for the core mass needed to form a giant planet before disk dissipation.
+
+%Studying atmosphere growth onto a solid core of fixed mass (e.g., \citealt{pn05}) can thus provide accurate estimates for the timescale required to form a giant planet.
+
+%However, the planetesimal accretion rates, $\dot{M_{\rm c}}$, need not be constant over time, and may significantly reduce as the planet's feeding zone is depleted of solids.
+If  accretion slows before the nebular gas dissipates, the target $M_{\rm crit}$ for successful giant planet formation is substantially reduced.
+%
+%To calculate the minimum value of $M_{\rm crit}$, we must first understand how $M_{\rm crit}$ depends on $\dot{M_{\rm c}}$.
+For large $\dot{M_{\rm c}}$, planetesimals deposit sufficient energy to balance the planet's luminosity and the core's envelope is in steady state.
 In this case,  $M_{\rm crit}$ is uniquely determined for a fixed accretion rate of solids and a set of disk conditions.
-
- %However, planetesimal accretion rates need not be constant over time.
-If  accretion slows before the nebular gas dissipates, the target $M_{\rm crit}$ for successful giant planet formation is substantially reduced.
-%At distances $\gtrsim40-50$ AU, simultaneous core and atmosphere growth to critical masses may not be possible within a disk lifetime \citep{rafikov11}.
-This problem may be circumvented if planetesimal accretion slows before the nebular gas dissipates, $M_{\rm crit}$ may be substantially reduced if .
-%  An alternative solution is core growth at moderate rates and thus to smaller masses, followed by a halt to planetesimal accretion.
-This study explore the latter scenario.
-
-%Fast accretion generates an additional difficulty.
-
-%An \textit{in situ} solution to this problem requires accelerating core growth to rates faster than those typically assumed.
-However, the minimum, or critical, core mass to form a giant planet, though typically quoted to be $M_{\rm crit}\sim 10M_{\oplus}$, depends on how quickly planetesimals accrete (e.g., \citealt{ikoma00}, \citealt{pollack96}, \citealt{rafikov06}).
-Fast accretion increases $M_{\rm crit}$.
-However, $M_{\rm crit}$ may be substantially reduced if planetesimal accretion slows before the nebular gas dissipates.
-
-%or fast  to grow the core and atmosphere simultaneously, much faster rates would be required ---  fast planetesimal accretion increases the core mass necessary for atmospheric collapse.
-during the era of fast accretion requires larger critical core masses.
-
-%Core accretion at wide separations thus requires either growth of cores at rates faster than typically assumed in the first regime of high planetesimal accretion or, perhaps counterintuitively, moderate core growth rates followed by a halt to planetesimal accretion in the second regime --- i.e., the three-phase regime assumed by \citet{pollack96}.
-
- %If cores and their atmospheres grow simultaneously, fast core growth requires large planetesimal accretion rates and hence large critical core masses.
-Core growth to these critical masses is particularly challenging in the outer disk, where long dynamical times slow core growth, making production of a massive core within the disk lifetime less likely.
-
-%Fast accretion implies large critical core masses.
-
-%In order to calculate the minimum value of $M_{\rm crit}$, we need to understand how $M_{\rm crit}$ depends on the planetesimal accretion rate.
-For high planetesimal accretion rates, atmospheres are primarily heated by planetesimals (e.g., \citealt{rafikov06}).
-As a result, the core's envelope is in a steady state at all times, in which all the incoming energy due to planetesimals is radiated away.
-
-%the evolution of a core's atmosphere depends on the rate at which planetesimals are accreted.
-
-%As the envelope and core become comparable in mass, hydrostatic balance no longer holds and runaway gas accretion commences.
-In this case, $M_{\rm crit}$ is the maximum core mass for which the atmosphere is still in hydrostatic equilibrium.
-For a fixed planetesimal accretion rate and a set of disk conditions, $M_{\rm crit}$ is uniquely determined.
-
-%In models that assume a high planetesimal accretion rate (e.g., \citealt{rafikov06}), the gaseous envelope is heated solely by planetesimals and is therefore in a steady state at all times, in which all the incoming energy due to planetesimals is radiated away.
-In this regime, the core and atmosphere grow simultaneously, with the atmosphere increasing in mass faster than the core.
-
-However, planetesimal accretion rates need not be constant over time ---  in particular, planetesimal accretion may significantly reduce as the planet's feeding zone is depleted of solids.
-If  accretion slows before the nebular gas dissipates, the target $M_{\rm crit}$ for successful giant planet formation is substantially reduced.
+For small $\dot{M_{\rm c}}$, this energy balance cannot be maintained and the gaseous envelope loses energy via Kelvin-Helmholtz contraction, resulting in gas accretion.
 Moreover, $M_{\rm crit}$ is no longer uniquely determined ---  any core has the potential to initiate runaway gas accretion, provided it can accumulate a massive atmosphere before disk dissipation.
-
-In order to calculate the minimum value of $M_{\rm crit}$, we need to understand how $M_{\rm crit}$ depends on the planetesimal accretion rate.
-For high planetesimal accretion rates, the core's envelope is in steady state at all times, in which all the incoming energy due to planetesimals is radiated away.
-On the other hand, time-dependent core accretion models (e.g., \citealt{pollack96}, \citealt{ikoma00}) include variable planetesimal accretion rates which allow for the Kelvin-Helmholtz (KH) contraction of the gaseous envelope to become important.
-In this regime, the envelope is no longer in a steady state, but rather it accretes gas as it cools.
-\citet{pollack96} find that the evolution of a giant planet can be separated into three phases.
+%
+%Time-dependent core accretion models, such as \citet{pollack96}, \citet{ikoma00}, suggest that the evolution of a giant planet can be separated into three phases.
 The core forms during phase 1 due to runaway planetesimal accretion, while the envelope mass stays small.
@@
 %Studying atmosphere growth onto a solid core of fixed mass (e.g., \citealt{pn05}) can thus provide accurate estimates for the timescale required to form a giant planet.
+%
+%It is not clear \textit{a priori} which of the two regimes --- atmosphere evolution dominated by planetesimal accretion versus KH contraction --- is the most relevant, as the conclusion of \citet{pollack96} depends on their assumptions about planetesimal accretion history.
+In particular, a core's feeding zone may be replenished by migration of solids through the protoplanetary disk.
+In this study, we accept the three-stage growth, in which the core is fully formed before it accumulates a significant atmosphere.
+During the slow phase of atmospheric growth we thus assume that the luminosity of the envelope is only due to KH contraction and neglect ongoing planetesimal accretion.
+Our results yield lower core masses than studies that assume large planetesimal accretion rates, confirming that if cores stop growing, they can accrete massive atmospheres more effectively.
+Moreover, as additional heating due to accretion of solids limits the envelope's ability to cool and thus increases the core mass required for collapse, our results give a robust minimum for the core mass needed to form a giant planet before disk dissipation.
+%
+
+In this paper, we calculate the $M_{\rm crit}$ required to nucleate runaway atmospheric growth for a fully formed core.
+Following Piso \& Youdin (accepted to ApJ; hereafter Paper I), we model the planet's evolution using a series of quasi-static two-layer atmospheres embedded in a protoplanetary disk.
+We build on the results of Paper I by making two important additions: (1) a realistic equation of state (EOS), and (2) realistic dust opacities.
+
+We use the EOS calculated by \citet{saumon95}, which captures non-ideal effects such as dissociation, ionization, and selective occupation of quantum states at low temperatures.
+Opacities in protoplanetary disks are unlikely to be interstellar, and are lowered by grain growth and dust settling.
+We use the \citet{dalessio01} opacity tables, which are used to model observations of protoplanetary disks.
+
+This paper is organized as follows.
+In Section \ref{sec2} we review the quasi-static and cooling models derived in Paper I.
+We discuss variations in the adiabatic gradient caused by the non-ideal EOS in Section \ref{deladtable}, and their effect on atmosphere evolution in Section \ref{EOSeffects}.
+We introduce realistic opacities in Section \ref{sec:opacity}, and determine $M_{\rm crit}$ in Section \ref{critical}.
+In Section \ref{acc} we compare our results to those obtained by studies employing high planetesimal accretion rates.
+We summarize our findings in Section \ref{conclusions}.
+%-- which is particularly relevant to our regime of low planetesimal accretion in which solids have been segregated from the gas and are not being added back in ---
+
+%Fast accretion heats a core's  atmosphere, increasing pressure support and hence $M_{\rm crit}$.
+For high planetesimal accretion rates, atmospheres are primarily heated by planetesimals (e.g., \citealt{rafikov06}).
+In this case,  $M_{\rm crit}$ is uniquely determined for a fixed accretion rate of solids and a set of disk conditions.
+
+ %However, planetesimal accretion rates need not be constant over time.
+If  accretion slows before the nebular gas dissipates, the target $M_{\rm crit}$ for successful giant planet formation is substantially reduced.
+%At distances $\gtrsim40-50$ AU, simultaneous core and atmosphere growth to critical masses may not be possible within a disk lifetime \citep{rafikov11}.
+This problem may be circumvented if planetesimal accretion slows before the nebular gas dissipates, $M_{\rm crit}$ may be substantially reduced if .
+%  An alternative solution is core growth at moderate rates and thus to smaller masses, followed by a halt to planetesimal accretion.
+This study explore the latter scenario.
+
+%Fast accretion generates an additional difficulty.
+
+%An \textit{in situ} solution to this problem requires accelerating core growth to rates faster than those typically assumed.
+However, the minimum, or critical, core mass to form a giant planet, though typically quoted to be $M_{\rm crit}\sim 10M_{\oplus}$, depends on how quickly planetesimals accrete (e.g., \citealt{ikoma00}, \citealt{pollack96}, \citealt{rafikov06}).
+Fast accretion increases $M_{\rm crit}$.
+However, $M_{\rm crit}$ may be substantially reduced if planetesimal accretion slows before the nebular gas dissipates.
+
+%or fast  to grow the core and atmosphere simultaneously, much faster rates would be required ---  fast planetesimal accretion increases the core mass necessary for atmospheric collapse.
+during the era of fast accretion requires larger critical core masses.
+
+%Core accretion at wide separations thus requires either growth of cores at rates faster than typically assumed in the first regime of high planetesimal accretion or, perhaps counterintuitively, moderate core growth rates followed by a halt to planetesimal accretion in the second regime --- i.e., the three-phase regime assumed by \citet{pollack96}.
+
+ %If cores and their atmospheres grow simultaneously, fast core growth requires large planetesimal accretion rates and hence large critical core masses.
+Core growth to these critical masses is particularly challenging in the outer disk, where long dynamical times slow core growth, making production of a massive core within the disk lifetime less likely.
+
+%Fast accretion implies large critical core masses.
+
+%In order to calculate the minimum value of $M_{\rm crit}$, we need to understand how $M_{\rm crit}$ depends on the planetesimal accretion rate.
+For high planetesimal accretion rates, atmospheres are primarily heated by planetesimals (e.g., \citealt{rafikov06}).
+As a result, the core's envelope is in a steady state at all times, in which all the incoming energy due to planetesimals is radiated away.
+
+%the evolution of a core's atmosphere depends on the rate at which planetesimals are accreted.
+
+%As the envelope and core become comparable in mass, hydrostatic balance no longer holds and runaway gas accretion commences.
+In this case, $M_{\rm crit}$ is the maximum core mass for which the atmosphere is still in hydrostatic equilibrium.
+For a fixed planetesimal accretion rate and a set of disk conditions, $M_{\rm crit}$ is uniquely determined.
+
+%In models that assume a high planetesimal accretion rate (e.g., \citealt{rafikov06}), the gaseous envelope is heated solely by planetesimals and is therefore in a steady state at all times, in which all the incoming energy due to planetesimals is radiated away.
+In this regime, the core and atmosphere grow simultaneously, with the atmosphere increasing in mass faster than the core.
+
+%Moreover, $M_{\rm crit}$ is no longer uniquely determined ---  any core has the potential to initiate runaway gas accretion, provided it can accumulate a massive atmosphere before disk dissipation.
 
 %However, planetesimal accretion rates need not be constant at a given location in the protoplanetary disk throughout the disk's lifetime --- in particular, planetesimal accretion may significantly reduce as the planet's feeding zone is depleted of solids.
+
+%For high planetesimal accretion rates, the core's envelope is in steady state at all times, in which all the incoming energy due to planetesimals is radiated away.
+On the other hand, time-dependent core accretion models (e.g., \citealt{pollack96}, \citealt{ikoma00}) include variable planetesimal accretion rates which allow for the Kelvin-Helmholtz (KH) contraction of the gaseous envelope to become important.
+In this regime, the envelope is no longer in a steady state, but rather it accretes gas as it cools.
 
@@
 
-It is not clear \textit{a priori} which of the two regimes --- atmosphere evolution dominated by planetesimal accretion versus KH contraction --- is the most relevant, as the conclusion of \citet{pollack96} depends on their assumptions about planetesimal accretion history.
-In particular, a core's feeding zone may be replenished by migration of solids through the protoplanetary disk.
-In this study, we accept the three-stage growth, in which the core is fully formed before it accumulates a significant atmosphere.
-During the slow phase of atmospheric growth we thus assume that the luminosity of the envelope is only due to KH contraction and neglect ongoing planetesimal accretion.
-Our results yield lower core masses than studies that assume large planetesimal accretion rates, confirming that it may indeed be easier to grow a giant planet from a fully formed core.
-Moreover, as additional heating due to accretion of solids limits the envelope's ability to cool and thus increases the core mass required for collapse, our results give a robust minimum for the core mass needed to form a giant planet before disk dissipation.
-
 %Both the steady state regime, where planetesimal accretion dominates the atmosphere evolution, and the KH regime, Therefore, it may be easier to grow a giant planet from a fully formed core, which is particularly relevant in the outer disk, where low planetesimal accretion rates prevent the growth of large cores.
 
-We follow the model developed in Piso \& Youdin (accepted to ApJ; hereafter Paper I) and consider quasi-static two-layer atmospheres, embedded in the protoplanetary disk.
-Paper I calculates the minimum (critical) core mass to form a gas giant at several locations in the protoplanetary disk.
-We build on the results of Paper I by making two important additions: we include realistic equation of state (EOS) effects, as well as realistic dust opacities.
-
- Paper I assumes that the nebular gas is ideal, diatomic and can be described by a polytropic equation of state (EOS).
+ %Paper I assumes that the nebular gas is ideal, diatomic and can be described by a polytropic equation of state (EOS).
 However, non-ideal effects such as dissociation and ionization have to be taken into account in order to obtain better quantitative results.
@@
 %We find that the critical core mass is more than twice as large than in the ideal gas case.
-
- Another simplified assumption of Paper I is that the dust opacity in radiative regions is that of interstellar grains.
-In reality, opacities in protoplanetary disks, while not tightly constrained observationally, are unlikely to be interstellar.
-In particular, grain growth and dust settling --- which is particularly relevant to our regime of low planetesimal accretion in which solids have been segregated from the gas and are not being added back in --- lower the opacity.
-In this study, we use realistic opacity tables that are calculated based on observations of protoplanetary disks and take into account grain growth.
  %We find that the resulting critical core mass may be significantly lower when ISM opacities are used, i.e.
@@
 
-This paper is organized as follows.
-In Section \S\ref{sec2} we review the quasi-static and cooling models derived in Paper I.
-We discuss the variations in the adiabatic gradient caused by the non-ideal EOS in Section \S\ref{deladtable}, and their effect on atmosphere evolution in Section \S\ref{EOSeffects}.
-We determine the minimum core mass to form a giant planet during the disk lifetime in Section \S\ref{critical}, with realistic EOS and dust opacity assumptions.
-In Section \S\ref{acc} we compare our results to those obtained by studies that consider planetesimal accretion as the dominant source of energy.
-Finally, we summarize our findings in Section \S\ref{conclusions}.
+%Fast accretion implies large critical masses.
+Because protoplanetary disks dissipate on timescales of a few Myrs, such cores must grow quickly to have access to the nebular gas.
+If cores and their atmospheres grow simultaneously, fast core growth requires large planetesimal accretion rates and hence large critical core masses.
+Fast core growth to large critical masses poses particular challenges in the outer disk, where dynamical times are long.
+Core accretion at wide separations thus requires either growth of cores at rates faster than typically assumed or, perhaps counterintuitively, moderate core growth rates followed by a halt to planetesimal accretion.
+In this study we explore the latter possibility.
 
@@
 We assume that the planet consists of a solid core of fixed mass and a two-layer atmosphere composed of an inner convective region and an outer radiative zone that matches smoothly onto the disk.
-The two regions are separated by the Schwarzschild criterion for convective instability (see \S\ref{struct}).
-The surface between the two regions is the radiative-convective boundary (RCB), defined by a radius $r=R\cb$.
-Moreover, we assume that the luminosity is constant throughout the radiative region (see Section \S\ref{critical} for additional discussion).
-We assume a low planetesimal accretion regime in which the atmosphere evolution is dominated by KH contraction.The atmosphere is spherically symmetric, self-gravitating and in hydrostatic balance.
-We note that spherical symmetry confines us to the outer regions of the disk ($a\gtrsim5$ AU), where the disk scale height is larger than the radius at which the planet matches onto the disk (see Paper I for further details).
+The two regions are separated by the Schwarzschild criterion for convective instability (see \S\ref{struct}) at radius $r=R\cb$, known as the radiative-convective boundary (RCB).
+We assume that the luminosity is constant throughout the radiative region (see Section \S\ref{critical} for additional discussion).
+Our model applies when planetesimal accretion is minimal, so that the envelope's evolution is dominated by KH contraction (see also \S\ref{acc}).
+The atmosphere is spherically symmetric, self-gravitating and in hydrostatic balance.
+We apply our model in at semimajor axes $a\gtrsim5$ AU.
+Here the disk scale height is larger than the radius at which the planet matches onto the disk (see Paper I for further details), and hence spherical symmetry holds.
 The nebular gas is composed of a hydrogen-helium mixture, with hydrogen and helium mass fractions of 0.7 and 0.3, respectively.
-We assume that the envelope evolves through stages of quasi-static equilibrium.
+Because the envelope grows slowly, we calculate its evolution using a series of linked quasi-static equilibrium models.
 
@@
 \end{subeqnarray}
-where $a$ is the semimajor axis and for a mean molecular weight $\mu=2.35$.
+for a mean molecular weight $\mu=2.35$.
 
@@
 \noindent where $r$ is the radial coordinate, $P$, $T$ and $\rho$ are the gas pressure, temperature and density, respectively, $m$ is the mass enclosed by the radius $r$, $L$  is the luminosity from the surface of radius $r$, and $G$ is the gravitational constant.
-The gas is heated at a rate of $\epsilon_g \equiv -T ds/dt$ per unit mass due to gravitational contraction, where $s$ is the specific gas entropy, while  $\epsilon$ represents the rate at which internal heat is generated per unit mass.
+The gas is heated at a rate of $\epsilon_g \equiv -T dS/dt$ per unit mass due to gravitational contraction, where $S$ is the specific gas entropy, while  $\epsilon$ represents the rate at which internal heat is generated per unit mass.
 We do not take into account any internal energy sources and set $\epsilon=0$.
@@
 Paper I assumes an ideal gas polytropic EOS, $P=K\rho^{\gamma}$, with $K$ the adiabatic constant and $\gamma$ the adiabatic index.
-In this study we use realistic EOS tables, as explained in Section \S\ref{deladtable}.
+In this study we use the EOS tables of \citet{saumon95}, as explained in Section \S\ref{deladtable}.
 Furthermore, Paper I assumes a power-law opacity given by
@@
 Furthermore, dust settling and grain growth lower both the normalization factor $F_{\kappa}$ and the power-law coefficient $\beta$.
-In this paper, we explore more realistic opacity laws and their effect on the critical core mass when compared to the ISM power-law, as discussed in Section \S\ref{sec:opacity}.
+In this paper, we explore more realistic opacity laws derived from the power-law size distributions typically used to interpret observations of protoplanetary disks (Section \S\ref{sec:opacity}).
+
+%and their effect on the critical core mass when compared to the ISM power-law, as discussed in Section \S\ref{sec:opacity}.
 
@@
 We choose $\rho_{\rm c}=3.2$ g cm$^{-3}$ (e.g., \citealt{pap99}).
-The atmosphere matches onto the protoplanetary disk at the Hill radius $R_{\rm H} \equiv a [M_{\rm p} / (3 M_{\odot})]^{1/3}$, where $M_{\rm p}$ is the total planet mass, $M_{\rm p}=M\co+M_{\rm atm}$.
-The Hill radius is the distance from the planet at which the gravitational attraction of the planet and the tidal gravity due to the host star are equal.
-Outside the Hill sphere, stellar tides unbind gas from the planet.
-The effective outer boundary of the atmosphere is the surface defined by the Bondi radius, $R_{\rm B} \equiv G M_{\rm p}/c_{\rm s}^2=G M_{\rm p} / (\mathcal{R} T\di)$.
-This is the distance from the planet at which the thermal energy of the nebular gas is of the order of the gravitational energy of the planet.
-Here $c_{\rm s}$ is the isothermal sound speed and $\mathcal{R}$ is the reduced gas constant: $\mathcal{R}=k_B/(\mu m_p)$, with $k_B$ the Boltzmann constant and $m_p$ the proton mass.
+Two radial scales determine the extent of the atmosphere: the Hill radius, $R_{\rm H} \equiv a [M_{\rm p} / (3 M_{\odot})]^{1/3}$, where the gravitational attraction of the planet and the tidal gravity due to the host star are equal, and the Bondi radius, $R_{\rm B} \equiv G M_{\rm p}/c_{\rm s}^2=G M_{\rm p} / (\mathcal{R} T\di)$, where the thermal energy of the nebular gas is approximately the gravitational energy of the planet.
+Here, $M_{\rm p}$ is the total planet mass, $c_{\rm s}$ is the isothermal sound speed, $\mathcal{R}=k_B/(\mu m_p)$ is the reduced gas constant, $k_B$ is the Boltzmann constant, and $m_p$ the proton mass.
+We define the planet mass as the mass enclosed inside the smaller of $R_{\rm B}$ or $R_{\rm H}$.
 For $R_{\rm B}<R_{\rm H}$, several studies assume that the atmosphere matches onto the disk at $R_{\rm B}$ (e.g., \citealt{ikoma00}, \citealt{pollack96}).
-We choose the Hill radius as the outer boundary because the temperature and pressure at $R_{\rm H}$ are those of the disk,  $T(R_{\rm H})=T_{\rm d}$ and $P(R_{\rm H})=P_{\rm d}$.
+In all cases, we choose the Hill radius as our outer boundary because the temperature and pressure at $R_{\rm H}$ are those of the disk,  $T(R_{\rm H})=T_{\rm d}$ and $P(R_{\rm H})=P_{\rm d}$ (see Paper I for more details).
 Outside $R_{\rm B}$, gas flows no longer circulate the planet, but rather belong to the disk.
-However, the density structure between $R_{\rm B}$ and $R_{\rm H}$ remains spherical and is still well described by hydrostatic balance \citep{ormel13}.
-The flows outside $R_{\rm B}$ could affect the atmosphere's cooling; however, we expect this effect to be weak, as heat losses predominantly occur deep in the planetary atmosphere.
+However, when $R_{\rm B}<R_{\rm H}$, the density structure between $R_{\rm B}$ and $R_{\rm H}$ remains spherical and is still well described by hydrostatic balance \citep{ormel13}.
+
+%The flows outside $R_{\rm B}$ could affect the atmosphere's cooling; however, we expect this effect to be weak, as heat losses predominantly occur deep in the planetary atmosphere.
 This justifies our outer boundary choice.
-We note, however, that we define the planet mass as the mass enclosed inside the smaller of $R_{\rm B}$ or $R_{\rm H}$.
-
 %, which justifies our outer boundary choice.
@@
 
-Lastly, we review the cooling model developed in Paper I, used to determine the time evolution of the atmosphere between subsequent static models.
-A protoplanetary atmosphere embedded in a gas disk satisfies the following cooling equation:
+Finally, we employ a cooling model developed in Paper I to determine the time evolution of the atmosphere between subsequent static models.
+A protoplanetary atmosphere embedded in a gas disk emits a total luminosity
 
@@
 \end{equation}
-Here, $L$ is the total luminosity, $L_{\rm c}$ is the luminosity from the solid core, which may include planetesimal accretion and radioactive decay, and $\Gamma$ is the rate of internal heat generation.
+Here, $L_{\rm c}$ is the luminosity from the solid core, which may include planetesimal accretion and radioactive decay, and $\Gamma$ is the rate of internal heat generation.
 We set $L\co=\Gamma=0$.
-The $\dot{E}$ term is the rate at which total energy (internal and gravitational) is lost, and $e_{\mathrm{acc}}$ is the specific total energy brought in by gas accreting at the rate $\dot{M}$: $e_{\mathrm{acc}}=u-G M/R$, where $u$ is the internal energy per unit mass.
+The $\dot{E}$ term is the rate at which total energy (internal and gravitational) is lost, and $e_{\mathrm{acc}}$ is the specific total energy brought in by gas accreting at rate $\dot{M}$: $e_{\mathrm{acc}}=u-G M/R$, where $u$ is the internal energy per unit mass.
 Finally, the last term represents the work done on a surface mass element.
@@
 
-In order to describe the nebular gas, we use the interpolated EOS tables of \citet{saumon95} for a helium mass fraction $Y=0.3$, and extend them to lower temperatures and pressures corresponding to the conditions in our fiducial disk.
+For our gas equation of state, we use the interpolated EOS tables of \citet{saumon95} for a helium mass fraction $Y=0.3$, and extend them to lower temperatures and pressures corresponding to the conditions in our fiducial disk.
 \App{EOStables} describes our extension procedure.
+%useful parameter for understanding atmospheric structure, we choose to represent the EOS through $\delad$.
+
+%devote this section to the impact on $\delad$ of the realistic EOS.
 
@@
 
+ We choose to represent the EOS through the adiabatic gradient $\delad$ (defined in Equation \ref{eq:delad}).
+In contrast with an ideal gas polytrope, the EOS of a realistic gas cannot be fully specified by $\delad$.
+The \citet{saumon95} EOS tables are obtained using the free energy minimization method (see, e.g., \citealt{graboske69}), which provides additional information to relate the thermodynamic variables.
+Nevertheless, $\delad$ is a useful parameter for understanding atmospheric structure.
 For an ideal gas polytropic EOS, the adiabatic gradient is constant.
 In contrast, non-ideal effects such as dissociation or ionization produce temperature-dependent variations in $\delad$.
-Figure \ref{fig:deladmap} shows a contour plot of the adiabatic gradient (defined in Equation \ref{eq:delad})  as a function of gas temperature and pressure.
+Figure \ref{fig:deladmap} shows a contour plot of the adiabatic gradient  as a function of gas temperature and pressure.
 We distinguish three separate temperature regimes:
@@
 
-At low temperatures, hydrogen is molecular, with a stable configuration.
-At temperatures higher than $T \sim 2000-3000$ K, the internal energy of the hydrogen molecule is large enough for hydrogen to dissociate.
-At $T \gtrsim 10,000$ K, hydrogen ionizes.
+ Hydrogen is molecular at low temperatures, dissociates at $T \gtrsim 2000-3000$ K, and ionizes at  $T \gtrsim 10,000$ K.
 In stellar and giant planet interiors there is little overlap between the two processes: hydrogen is almost entirely dissociated into atoms by the time ionization becomes important.
@@
 As displayed in Figure \ref{fig:deladmap}, the adiabatic gradient decreases significantly in regions of partial dissociation and partial ionization.When a gas dissociates, part of the internal energy of the system is used to break down molecules, which reduces the amount of energy available to increase the temperature of the system, thus lowering the adiabatic gradient.
-The energy required to dissociate depends on the dissociation fraction, which can be determined from the Saha equation (see e.g., \citealt{kippenhahn90}).
+%The energy required to dissociate depends on the dissociation fraction, which can be determined from the Saha equation (see e.g., \citealt{kippenhahn90}).
 The dissociation fraction only depends on gas temperature and density, and hence only on the EOS.
@@
 As expected, the adiabatic gradient is $\delad=2/7$ for pure molecular hydrogen and $\delad=2/5$ when hydrogen is fully dissociated, but decreases significantly during partial dissociation and is smallest when half of the gas is dissociated.
-This is consistent with the behavior shown in Figure \ref{fig:deladmap}.
-
-The ionization of atomic hydrogen is also dictated by the Saha equation, with the dissociation energy replaced by ionization energy, hence the adiabatic gradient behaves analogously, consistent with Fig.
+Note that this behavior differs from a mixture of molecular and atomic hydrogen, for which $2/7<\delad<2/5$.
+%This behavior is shown in Figure \ref{fig:deladmap}.
+Ionization generates an analogous dip in $\delad$ at higher temperatures.
+
+%The ionization of atomic hydrogen is also dictated by the Saha equation, with the dissociation energy replaced by ionization energy, hence the adiabatic gradient behaves analogously, consistent with Fig.
 \ref{fig:deladmap}.
@@
 As the gas temperature becomes comparable to $\Theta_r$, fewer rotational states are excited, and rotation entirely ceases as $T \rightarrow 0$.
-We note that at temperatures larger than $\gtrsim 6000$ K, vibrational motions also become important.
+%We note that at temperatures larger than $\gtrsim 6000$ K, vibrational motions also become important.
 While at $T\lesssim2000$ K, where our extension of the \citet{saumon95} EOS tables is valid, vibrational motion is negligible, we include these effects in our extension of the EOS for completion (see \App{EOStables} for details).
@@
 Parahydrogen has antiparallel proton spins and thus an antisymmetric wave function.
-From the Pauli exclusion principle it follows that it can only occupy symmetric rotational states with even angular quantum number $j$ \citep{farkas35}.
-In contrast, orthohydrogen has parallel proton spins and a symmetric wave function, and can thus only occupy states with odd $j$.
-At equilibrium, the relative abundance of the ortho- and para- states is given by the ratio of their partition functions, described in \App{EOStables}.
+From the Pauli exclusion principle, it can only occupy symmetric rotational states with even angular quantum number $j$ \citep{farkas35}.
+In contrast, orthohydrogen has parallel proton spins, and can only occupy states with odd $j$.
 At $T \rightarrow 0$ all hydrogen molecules are in the ground state with $j=0$, which corresponds to parahydrogen.
-As the temperature increases, parahydrogen starts converting into orthohydrogen, resulting in an ortho-para equilibrium ratio of 3:1 at room temperature.
-
+As the temperature increases, parahydrogen starts converting into orthohydrogen, plateauing at an ortho-para equilibrium ratio of 3:1 for $T \gtrsim 300$ K.
+
+ %At equilibrium, the relative abundance of the ortho- and para- states is given by the ratio of their partition functions, described in \App{EOStables}.
 %During the para-to-ortho conversion, part of the internal energy of the hydrogen molecule is used
 
-Similarly to the case of dissociation, as $T$ increases, part of the rotational energy of the hydrogen molecule is used to convert parahydrogen to orthohydrogen rather than to only increase the temperature.
-Thus $u_{\rm r} \propto T$, where $u_{\rm r}$ is the rotational energy per unit mass, no longer holds, and the adiabatic gradient $\delad$ is temperature dependent.
-These $\delad$ variations, which occur for $T \lesssim 300$ K, are explored in \App{deladspin} and displayed in Figure \ref{fig:deladmap}.
-At higher temperatures, the ortho-para 3:1 equilibrium ratio is reached; no further isomer conversion occurs, and thus $\delad$ remains relatively constant until dissociation temperatures are reached.
+As $T$ increases from 0 to 300 K, part of the rotational energy of the hydrogen molecule is used to convert parahydrogen to orthohydrogen rather than to only increase $T$.
+As a result, $\delad$ varies with temperature, as  displayed in Figure \ref{fig:deladmap} (see \App{deladspin} for further discussion) .
+%At higher temperatures, the ortho-para 3:1 equilibrium ratio is reached; no further isomer conversion occurs, and thus $\delad$ remains relatively constant until dissociation temperatures are reached.
 
@@
 
-Variations in the adiabatic gradient $\delad$ due to partial dissociation and fractional occupation of $\rm{H}_2$ rotational states (see \S\ref{deladtable}) affect atmospheric evolution by yielding: (1) a lower envelope luminosity $L$, and (2) a larger amount of radiated energy per unit of accreted mass $-dE/dM$, when compared to an ideal gas.
+Variations in the EOS, and hence $\delad$, due to partial dissociation and fractional occupation of $\rm{H}_2$ rotational states affect atmospheric evolution by yielding: (1) a lower envelope luminosity $L$, and (2) a larger amount of radiated energy per unit of accreted mass $-dE/dM$, when compared to the polytropic EOS considered in Paper I.
 As a result, the rate of change in atmospheric mass,
@@
 (cf.
-Equation \ref{eq:coolingglobal} and ignoring surface terms, which only become significant in the late stages of atmospheric evolution) is lower than in the ideal case.
-The $\delad$ variations thus slow down gas accretion and increase the growth time of the atmosphere.
-Since envelope growth is faster for larger cores, the minimum core mass required to accumulate a massive envelope during the (assumed fixed) disk lifetime, i.e.
+Equation \ref{eq:coolingglobal} and ignoring surface terms, which only become significant in the late stages of atmospheric evolution) is lower than in the polytropic case.
+Slower gas accretion increases the growth time of the atmosphere.
+Since envelope growth is faster for larger cores, we calculate a larger $M_{\rm crit}$.
+% --- the minimum core mass required to accumulate a massive envelope during a fixed disk lifetime.%, i.e.
 $M_{\rm crit}$, increases when realistic EOS effects are included.
@@
 It follows that $L$ primarily scales as $1/P\cb$.
-This pressure depth at the RCB depends on both the interior structure of the atmosphere, which is affected by dissociation (at high temperatures), and on the matching to the nebula through the radiative region, where the gas is non-ideal due to the variable occupation of $\rm{H}_2$ rotational states (at low temperatures).
-Similarly, the EOS modifications due to dissociation and spin isomers affect the way in which the total energy $E(r)$ is distributed throughout the envelope, and thus $-dE/dM$.
-
-Since both dissociation at high temperatures and variable occupation of rotation states at low temperatures affect atmospheric structure and evolution, as discussed above, it is helpful to separate these effects and study them independently.
+This pressure depth at the RCB depends on both the interior structure of the atmosphere at high temperatures, and on the matching to the nebula through the radiative region at low temperatures.
+Similarly, the EOS affects the distribution of total energy $E(r)$ throughout the envelope, and thus $-dE/dM$.
+
+Since both dissociation deep in the convective interior and variable occupation of rotation states in the outer envelope affect atmospheric structure and evolution, it is helpful to separate these effects and study them independently.
 
@@
 \begin{enumerate}
-\item Ideal gas polytrope with $\delad=0.3$.
-\item Ideal gas polytrope with $\delad=0.3$ for $T>500$ K and realistic EOS for $T<500$ K (hereafter low-$T$ realistic EOS), which includes the effects of fractional occupation of $\rm H_2$ rotational states.
-\item Ideal gas polytrope with $\delad=0.3$ for $T<500$ and realistic EOS for $T>500$ K  (hereafter high-$T$ realistic EOS), which includes the effects of hydrogen dissociation.
+\item Ideal gas polytrope with $\delad=0.3$. % (hereafter EOS 1).
+\item Ideal gas polytrope with $\delad=0.3$ for $T>500$ K and realistic EOS for $T<500$ K, which includes the effects of fractional occupation of $\rm H_2$ rotational states.
+\item Ideal gas polytrope with $\delad=0.3$ for $T<500$ and realistic EOS for $T>500$ K , which includes the effects of hydrogen dissociation.
 \item Realistic EOS at all $T$. 
@@
 %By separately comparing curves (1) and (2), and (1) and (3)  while the difference between curves (1) and (3) accounts for hydrogen dissociation. 
-Both the high-$T$ and low-$T$ EOSs yield slower gas accretion when compared to the ideal gas polytrope.
+Both EOS 2 and EOS 3 yield slower gas accretion when compared to EOS 1.
+Noting the logarithmic scale in Figure \ref{fig:tplotall}, we see that the combined effect of fractional occupation of rotational states and dissociation is significantly greater than either individually.
 
@@
 %This follows from Equation (\ref{eq:dMdt}), with rotation decreasing $L$ and dissociation increasing $-dE/dM$, as we show further. 
-Noting the logarithmic scale in Figure \ref{fig:tplotall}, we see that the combined effect of fractional occupation of rotational states and dissociation is significantly greater than either individually.
 %consistent with Equation \ref{eq:dMdt}, since dissociation increases $-dE/dM$ while molecular rotation decreases $L$.
@@
 Middle-left panel: Profile of total energy at scale $r$, $-e \rho r^3$, for a total mass of $12 M_{\oplus}$.
-Hydrogen dissociation deep in the atmosphere causes more of the energy to be concentrated at the bottom of the envelope when compared to an ideal gas polytrope.
-This increases the amount of energy per unit mass that needs to be radiated away, $-dE/dM$.
+Hydrogen dissociation deep in the atmosphere causes more of the (negative) energy to be concentrated at the bottom of the envelope when compared to an ideal gas polytrope.
+This increases the amount of energy per unit mass, $-dE/dM$,  that needs to be radiated away to accrete the next parcel of gas.
 Middle-right panel:  $-dE/dM$ evolution as a function of total mass.
@@
 In Figure \ref{fig:all_plot}, middle-right and upper-right panels, we show the evolution with planet mass of $-dE/dM$ and $L$, respectively, for the EOSs described above.
-The EOS variations (i.e., the realistic EOS at all $T$) reduce $L$ by a factor of $\sim$$2$ and $-dE/dM$ by a factor of $\sim$$5$, when compared to the ideal gas polytrope.
-Atmospheric growth is therefore primarily affected by the changes in the energy profile $E(r)$ due to the realistic EOS, with the luminosity $L$ having a secondary effect.
+The realistic EOS reduces $L$ by a factor of $\sim$$2$ and $-dE/dM$ by a factor of $\sim$$5$, when compared to the ideal gas polytrope.
+Atmospheric growth is therefore primarily slowed by changes in the energy profile $E(r)$, with the luminosity $L$ having a secondary effect.
 In what follows we explore each contribution separately.
@@
 
-Figure \ref{fig:all_plot}, middle-left panel, shows a radial profile of the total (internal + gravitational) energy at scale $r$, $-e \rho r^3$, with $e$ the total specific energy.
+Variations in $\delad$ due to dissociation increase $-dE/dM$.
+Figure \ref{fig:all_plot}, middle-left panel, shows a radial profile of the total (negative) energy at scale $r$, $-e \rho r^3$.
+Here, $e$ is the total specific energy.
 %the cumulative total energy (internal and gravitational) profile as a function of the radial coordinate, \sout{for the same example planet}.
@@
 The bulk of the energy is concentrated in the atmospheric interior for all EOSs.
-However, the energy of the high-$T$ realistic EOS and of the realistic EOS at all $T$ is much larger in magnitude (more negative) near the core when compared to an ideal gas polytrope.
+However, $-e \rho r^3$ for EOS 3 and EOS 4 is much larger in magnitude (more negative) near the core when compared to the others.
 This is due to the fact that $\delad$ decreases in dissociation regions (Figure \ref{fig:deladmap}).
-Dissociation increases the number of particles, which tends to increase entropy.
-In order for entropy to stay constant with radius, the temperature must therefore decrease, which also lowers the adiabatic gradient.
-A lower $\delad$ results in a shallower temperature gradient in dissociation regions, and hence in lower temperatures near the core (cf.
+Dissociation increases the number of particles, which increases entropy.
+In order for entropy to stay constant with radius in the convective interior, the temperature must therefore decrease, which lowers $\delad$ and $|dT/dr|$, resulting in lower temperatures near the core (cf.
 Figure \ref{fig:all_plot}, bottom-left panel).
-As the specific internal energy $u(r) \propto T(r)$, dissociation decreases $u(r)$ and thus the total energy deep in the interior.
-This explains the rise in $-e \rho r^3$ in Figure \ref{fig:all_plot}, middle-left panel.
-Note that, in some astrophysical contexts, this loss of thermal energy due to dissociation can be so large as to trigger dynamical instabilities and eventual collapse in higher mass objects, such as in protostars \citep{larson69} or during the runaway growth of giant planets \citep{bodenheimer80}.
-
-The energy behavior can also be explained qualitatively through a simple analytic argument: the density profile in an adiabatic, non-self-gravitating atmosphere composed of an ideal gas scales as $\rho(r) \propto r^{-1/\delad+1}$, and the total specific energy is $e(r) \propto 1/r$ (see Paper I).
+Because  the specific internal energy $u \propto T$, dissociation decreases $u$ and thus the total energy deep in the interior.
+Note that this loss of thermal energy due to dissociation can be large enough to trigger dynamical instabilities and eventual collapse in higher mass objects, such as in protostars \citep{larson69} or during the runaway growth of giant planets \citep{bodenheimer80}.
+
+This energy behavior can also be explained qualitatively using $\delad$: the density profile in an adiabatic, non-self-gravitating atmosphere composed of an ideal gas scales as $\rho(r) \propto r^{-1/\delad+1}$, and the total specific energy is $e(r) \propto 1/r$ (see Paper I).
 Thus $- e \rho r^3 \sim r^{-1/\delad+3}$.
-It is easy to see that the total energy is concentrated near the core if $\delad<1/3$, and at at the outer boundary otherwise.
+The total energy is concentrated near the core if $\delad<1/3$, and at the outer boundary otherwise.
 In our regime of interest, $\delad<1/3$, on average, hence the energy always peaks near the core (cf.
 Figure \ref{fig:all_plot}, middle-left panel).
-More importantly, the $e$ dependence on $\delad$ also implies that adiabats with lower $\delad$ have their energy more tightly packed towards the interior of the envelope, which is the case during dissociation.
-This is consistent with the energy behavior in Figure \ref{fig:all_plot}, middle-left panel.
-It follows that a more negative total energy in the interior, as is the case for the high-$T$ EOS, requires a larger amount of energy to be radiated away to bind the next batch of gas.
-This implies a larger $-dE/dM$ during atmospheric evolution (middle-right panel of Figure \ref{fig:all_plot}), and thus a lower $\dot{M}$ for the high-T EOS (lower-right panel of Figure \ref{fig:all_plot}).
+More importantly, the dependence of $-e \rho r^3$ on $\delad$ also implies that adiabats with lower $\delad$ have their energy more tightly packed towards the interior of the envelope, which is the case during dissociation.
+
+A more negative total energy in the interior of a planet's atmosphere, as is the case for EOS 3, requires a larger amount of energy to be radiated away to bind the next batch of gas.
+This implies a larger $-dE/dM$ during atmospheric evolution (middle-right panel of Figure \ref{fig:all_plot}), and thus a lower $\dot{M}$ (lower-right panel of Figure \ref{fig:all_plot}).
 
@@
 
-While dissociation at high temperatures is primarily responsible for the larger $-dE/dM$ for a realistic EOS, the variable occupation of $\rm H_2$ rotational states at low temperatures also increases $-e \rho r^3$ (Figure \ref{fig:all_plot}, middle-left panel), and hence $-dE/dM$ (Figure \ref{fig:all_plot}, middle-right panel), albeit less significantly.
-As both dissociation and spin isomers modify the energy profile throughout the envelope, the rate of change in atmospheric mass $\dot{M}$, and thus the atmospheric cooling time, is primarily affected by changes in $-dE/dM$ rather than $L$ (see Equation \ref{eq:dMdt}).
+While dissociation at high temperatures is primarily responsible for the larger $-dE/dM$ for the realistic EOS, the variable occupation of $\rm H_2$ rotational states at low temperatures also increases $-e \rho r^3$ (Figure \ref{fig:all_plot}, middle-left panel), and hence $-dE/dM$ (Figure \ref{fig:all_plot}, middle-right panel), albeit less significantly.
+%As both dissociation and spin isomers modify the energy profile throughout the envelope, the rate of change in atmospheric mass $\dot{M}$, and thus the atmospheric cooling time, is primarily affected by changes in $-dE/dM$ rather than $L$ (see Equation \ref{eq:dMdt}).
 
@@
 %%\vspace{-0.5in}
-\caption{We explore atmosphere growth around a core with $M\co=10 M_{\oplus}$ forming at 10 AU, and with total mass (core + atmosphere) 12 $M_{\oplus}$, for the EOS choice of \Eq{eq:deladvar} and various $\nabla_{\rm ad,1}$ and $\nabla_{\rm ad,2}$.
+\caption{As an illustrative numerical experiment, we explore atmosphere growth around a core with $M\co=10 M_{\oplus}$ forming at 10 AU, and with total mass (core + atmosphere) 12 $M_{\oplus}$, for the EOS choice of \Eq{eq:deladvar} and various $\nabla_{\rm ad,1}$ and $\nabla_{\rm ad,2}$.
 In each panel, the circles mark the RCB location.
@@
 Upper-left panel: radial temperature profile for the same $\nabla_{\rm ad,1}$ in the interior and various $\nabla_{\rm ad,2}$ in the exterior.
-The analytic profile of Equation (\ref{eq:Tdeep}) is over-plotted for comparison (\textit{dotted line}) in regions of agreement.
 A larger $\nabla_{\rm ad,2}$ yields a deeper radiative region.
+Upper-right panel: radial temperature profile for various $\nabla_{\rm ad,1}$ in the interior and the same $\nabla_{\rm ad,2}$ in the exterior.
+In both upper panels, the analytic profile of Equation (\ref{eq:Tdeep}) is over-plotted for comparison (\textit{dotted line}) in regions of agreement.
+A lower $\nabla_{\rm ad,1}$ yields a more shallow radiative region.
 Lower-left panel: radial pressure profile for the same $\nabla_{\rm ad,1}$ in the interior and various $\nabla_{\rm ad,2}$ in the exterior.
-The dashed black line is the analytic prediction for the pressure profile in the radiative zone, which agrees with the numerical results.
-Upper-right panel: radial temperature profile for various $\nabla_{\rm ad,1}$ in the interior and the same $\nabla_{\rm ad,2}$ in the exterior.
-The analytic profile of Equation (\ref{eq:Tdeep}) is over-plotted for comparison (\textit{dotted line}) in regions of agreement.
-A lower $\nabla_{\rm ad,1}$ yields a more shallow radiative region.
 Lower-right panel: radial pressure profile for various $\nabla_{\rm ad,1}$ in the interior and the same $\nabla_{\rm ad,2}$ in the exterior.
-The dashed black line is the analytic prediction for the pressure profile in the radiative zone, which agrees with the numerical results.}
+In both lower panels, the dashed black line is the analytic prediction for the pressure profile in the radiative zone, which agrees with the numerical results.}
 \label{fig:varying_delad}
@@
 
-The variations in adiabatic gradient at low temperatures decrease the atmosphere's luminosity compared to an ideal gas polytrope.
-The adiabatic gradient for the low-$T$ realistic EOS increases with radius (i.e., with decreasing temperature) throughout the convective zone, then becomes lower near the RCB.
-This results in a lower $\delad$, on average, for the low-$T$ EOS in the outer parts of the envelope.
-
+%The variations in $\delad$ at low temperatures
+
+The variations in $\delad$ at low temperatures decrease the atmosphere's luminosity compared to an ideal gas polytrope.
+The adiabatic gradient of EOS 2 is larger, on average, than for EOS 1 in the outer part of the convective region \footnote {$\delad$ for EOS 2 increases with radius (i.e., with decreasing temperature) throughout the convective zone, then becomes lower near the RCB.}.
 To understand how a variable, but overall larger $\delad$ in the outer atmosphere affects evolution, we study the simplified problem of atmospheres composed of an ideal gas with adiabatic gradient
@@
 where $\nabla_{\rm ad, 1}$ and $\nabla_{\rm ad,2}$ are constant.
-Figure \ref{fig:varying_delad}, lower-left panel, shows the radial pressure profile for $\nabla_{\rm ad,1}=0.3$ and various $\nabla_{\rm ad,2}$, for the same parameters as Figures \ref{fig:tplotall} and \ref{fig:all_plot}.
-The numerical solutions reproduce the analytic expectation of a nearly exponential pressure profile in the radiative region, $P(r) \propto \exp(\RB/r)$.
+We explore the independent effects of varying $\nabla_{\rm ad,1}$ and $\nabla_{\rm ad,2}$ by alternately fixing one and varying the other   (Figure \ref{fig:varying_delad}).
+The adiabatic gradient $\nabla_{\rm ad,1}$ dictates the temperature profile deep in the convective interior, while $\nabla_{\rm ad,2}$ sets the RCB temperature $T\cb$.
+Thus we can find the RCB location, and hence the RCB pressure $P\cb \propto \exp(\RB/R\cb)$ (see Figure \ref{fig:varying_delad}, lower panels), by matching the convective interior and the outer radiative zone for a fixed atmosphere mass.
+Higher $P\cb$ implies lower luminosity.
+
+%the location of the RCB, is set by the matching between the convective interior and the outer radiative zone. As $P(r) \propto \exp(\RB/r)$ in the radiative region,
+
+%As $P(r) \propto \exp(\RB/r)$ in the outer radiative region,
+
+%Figure \ref{fig:varying_delad}, lower-left panel, shows the radial pressure profile for $\nabla_{\rm ad,1}=0.3$ and various $\nabla_{\rm ad,2}$, for the same parameters as Figures \ref{fig:tplotall} and \ref{fig:all_plot}, which are in agreement with the analytic expectation of an exponential profile.
+
+%The numerical solutions reproduce the analytic expectation of a nearly exponential pressure profile in the radiative region, $P(r) \propto \exp(\RB/r)$.
 Thus $L \propto 1/P\cb \propto 1/\exp(\RB/R\cb)$ is directly correlated to the physical depth of the RCB.
 
-The location of the RCB is set by the matching between the convective interior and the outer radiative zone. Deep in the atmosphere, where $r \ll R\cb$, the temperature profile is given by
-
+%The location of the RCB is set by the matching between the convective interior and the outer radiative zone.
+
+Deep in the atmosphere, where $r \ll R\cb$, virial equilibrium yields a temperature profile 
 \begin{equation}
 \label{eq:Tdeep}
-T(r) \approx \frac{\nabla_{\rm ad,1} G M\co}{\mathcal{R} r}
-\end{equation}   
-(see Paper I for a derivation), consistent with the upper-left panel of Figure \ref{fig:varying_delad}.
-Even though $\nabla_{\rm ad,1}$ is an order unity coefficient, the temperature scaling with $\nabla_{\rm ad,1}$ in \Eq{eq:Tdeep} is exact, as shown in the upper-right panel of Figure \ref{fig:varying_delad} for various $\nabla_{\rm ad,1}$.
-It follows that the radial temperature profile in the atmospheric interior is primarily set by the inner adiabatic gradient.
-Thus the radius $R_{\rm switch}$ at which the adiabatic gradient changes, i.e.
-where $T(R_{\rm switch})=500$ K, shows little variation with $\nabla_{\rm ad,2}$.
-In contrast with the deep interior, we cannot rigorously apply \Eq{eq:Tdeep} at $R_{\rm switch}$, although the approximation is still valid to order unity.
-
+T(r) \approx \frac{\nabla_{\rm ad,1} G M\co}{\mathcal{R} r}.
+\end{equation}   
+Even though $\nabla_{\rm ad,1}$ is an order unity coefficient, the temperature scaling with $\nabla_{\rm ad,1}$ in \Eq{eq:Tdeep} is exact (Figure \ref{fig:varying_delad}, upper-right panel).
+Because the radial temperature profile in the atmospheric interior is set by $\nabla_{\rm ad,1}$, the radius $R_{\rm switch}$ at which the adiabatic gradient changes shows little variation with $\nabla_{\rm ad,2}$.
+%We can therefore match the fixed interior temperature profile to the exterior in order to determine $R\cb$ as a function of $\nabla_{\rm ad,2}$.
+Note, however, that \Eq{eq:Tdeep} is only valid to order of magnitude at $R_{\rm switch}$.
+
+%n contrast with the deep interior, however, we cannot rigorously apply \Eq{eq:Tdeep} at $R_{\rm switch}$, although the approximation is still valid to order unity.
+
+%We can use 
 The RCB temperature $T\cb$ is an order unity correction from the disk temperature that depends on the adiabatic gradient in the outer envelope, $\nabla_{\rm ad,2}$.
-We have showed in Paper I that $T\cb \simeq T\di (1- 2 \nabla_{\rm ad, 2})^{-1/2}$, for $\beta=2$ in the opacity-law (\ref{eq:opacitylaw}), modestly increases with $\nabla_{\rm ad,2}$, consistent with the upper-left panel of Figure \ref{fig:varying_delad}.
+We have shown in Paper I that $T\cb \simeq T\di (1- 2 \nabla_{\rm ad, 2})^{-1/2}$ (for $\beta=2$ in Equation \ref{eq:opacitylaw}).
+This approximation agrees with the numerically calculated $T\cb$ shown in Figure \ref{fig:varying_delad} to within 5\%.
+Note that $T\cb$ modestly increases with $\nabla_{\rm ad,2}$ (upper-left panel), and is constant for fixed $\nabla_{\rm ad,2}$  regardless of $\nabla_{\rm ad,1}$ (upper-right panel).
+
+% modestly increases with $\nabla_{\rm ad,2}$, consistent with the upper-left panel of Figure \ref{fig:varying_delad}.
 Moreover, a constant $\nabla_{\rm ad,2}$ in the exterior yields the same value for $T\cb$, independent of $\nabla_{\rm ad,1}$ (Figure \ref{fig:varying_delad}, upper-right panel), again consistent with the analytic prediction.
 Quantitatively, we have found that the analytic approximation for $T\cb$ agrees with the numerical result to within 5\%.
-
-Figure \ref{fig:varying_delad} shows that a larger adiabatic gradient in the outer atmosphere results in a deeper radiative region.
-We can understand this effect by assuming that $R_{\rm switch}$ and $T\cb$ are fixed, i.e.
-ignoring the order unity variations in these quantities due to changes in the adiabatic gradient, as discussed above.
+When these temperature conditions are matched for a fixed atmosphere mass, Figure \ref{fig:varying_delad} shows that a larger adiabatic gradient in the outer atmosphere results in a deeper radiative region.
+We can understand this effect by assuming that $R_{\rm switch}$ is fixed.
 A steeper adiabat (i.e., with a larger adiabatic gradient) that starts at a given depth (here $R_{\rm switch}$) will match the fixed RCB temperature at a smaller radius.
-A quantitative scaling of $R\cb$ is not possible, as the physical quantities that dictate the RCB location are only order unity corrections to the (also order unity) adiabatic gradient in the outer atmosphere.
-However, relatively small variations in the physical depth of the RCB can impact the RCB pressure significantly, as $P\cb$ increases exponentially with depth, as explained above.
+A quantitative scaling of $R\cb$ is not possible, because Equation (\ref{eq:Tdeep})  is only valid to order of magnitude at $R_{\rm switch}$ and relatively small variations in the physical depth of the RCB can impact the RCB pressure significantly.
 It follows that $L \propto 1/P\cb$ decreases with increasing outer $\nabla_{\rm ad}$; thus the overall larger adiabatic gradient due to the variable occupation of rotation states at low temperatures decreases $L$ and slows down growth.
 
+ %which dictates the RCB location, is only an order unity correction to the (also order unity) adiabatic gradient in the outer atmosphere. 
 %ignoring the order unity variations in    discussed above due to changes in the adiabatic gradient. For
@@
 
-From Equation (\ref{eq:dMdt}), realistic EOS effects reduce the atmosphere accretion rate $\dot{M}$ by decreasing $L$ and increasing $-dE/dM$.
-This results in slower atmospheric growth and therefore a larger critical core mass, as we show in Section \S\ref{critical}.
+%From Equation (\ref{eq:dMdt}), the lower $L$ and larger $-dE/dM$ caused by non-ideal EOS effects reduce the atmospheric accretion rate $\dot{M}$, resulting in slower growth and thus larger $M_{\rm crit}$ (see \S\ref{critical}).
+
+%realistic EOS effects reduce the atmosphere accretion rate $\dot{M}$ This results in slower atmospheric growth and therefore a larger critical core mass, as we show in Section \S\ref{critical}.
 
@@
 However, our scenario of low planetesimal accretion is likely to favor lower dust opacities, due to grain growth and dust settling.
-Grain growth, in particular, changes the particle size distribution when compared to the standard ISM size distribution (e.g., \citealt{pollack85}) and lowers the absolute value of the opacity.
+Grain growth, in particular, lowers the absolute value of the opacity, and may change the particle size distribution when compared to the standard ISM size distribution (e.g., \citealt{pollack85}).
 Enhanced metallicity due to planetesimal accretion, by construction not present during our atmosphere's growth, cannot make up for this reduction.
@@
 \label{eq:graindistr}
-n(a) \sim a^{-p},
-\end{equation}
-where $n(a)$ is the number density of particles of size $a$ and $p=3.5$ (corresponding to a ``normal'' collisional cascade) or $p=2.5$ (an approximation for coagulation).
+n(s) \sim s^{-p},
+\end{equation}
+where $n(s)$ is the number density of particles of size $s$ and $p=3.5$ (corresponding to a ``normal'' collisional cascade; this is also the ISM size distribution, e.g.
+\citealt{pollack85}) or $p=2.5$ (an approximation for coagulation).
 In this work we use the  \citet{dalessio01} frequency-dependent opacity tables to obtain the temperature-dependent Rosseland mean opacity $\kappa$.
-We take as a fiducial case a maximum particle size $a_{\rm max}=1$ cm and a grain size distribution given by Equation (\ref{eq:graindistr}) with $p=3.5$.
+We take as a fiducial case a maximum particle size $s_{\rm max}=1$ cm and a grain size distribution given by Equation (\ref{eq:graindistr}) with $p=3.5$.
 Other choices for the power-law coefficient $p$ are discussed later in this section.
@@
 The sharp drop in opacity ($\kappa \sim T^{-24}$, see Figure \ref{fig:delvsr}) due to dust sublimation lowers the radiative temperature gradient significantly (see Equation \ref{eq:delrad}), and may thus generate radiative layers within the inner region of the atmosphere (see \App{radwindow}, Figure \ref{fig:delvsr}).
-This could in principle pose challenges for our model, since the additional luminosity generated in these radiative windows may be large enough to make our assumption of constant luminosity throughout the radiative layers invalid.
-In practice, however, the inner radiative windows are either very thin (Figure \ref{fig:delvsr}, middle panel), or the temperature gradient is flat enough to result in a roughly constant luminosity throughout the region (Figure \ref{fig:delvsr}, bottom panel; also see Equation \ref{eq:structc}).
+Additionally, the weak temperature dependence of grain growth opacities may cause larger outer radiative zones than when ISM opacities are used.
+This could pose two challenges for our model:
+
+\begin{enumerate}
+
+ \item The additional luminosity, $\Delta L$, generated in the outer radiative layer of the envelope may not satisfy $\Delta L \ll L$, where $L$ is the assumed fixed atmospheric luminosity.
+For $p=3.5$ in Equation (\ref{eq:graindistr}), we have checked that $\Delta L \ll L$ in all the cases presented in this study.
+For $p=2.5$, however, this approximation breaks down at low core masses, as we show in \S\ref{critical}.
+
+ %In practice, however, the inner radiative windows are either very narrow compared to the height of the convective regions (Figure \ref{fig:delvsr}, middle panel), or  have $\delad \approx \delrad$ throughout, which makes the distinction between convective and radiative layers less  pronounced (Figure \ref{fig:delvsr}, bottom panel).
+
+%This results in a negligible extra luminosity generated in the radiative windows.
+However, the radiative windows may become non-negligible for other opacity choices and sufficiently low core masses, as we show later in this section.
+
+\item Only about half of the atmosphere's luminosity is generated in the innermost convective layer when radiative windows exist.
+We must therefore check that our assumption of constant $L$ does not substantially change the structure of the atmosphere in the region of the radiative windows.
+Fortunately, these radiative windows are either very narrow compared to the height of the convective regions (Figure \ref{fig:delvsr}, middle panel), or  have $\delad \approx \delrad$ throughout, which makes the distinction between convective and radiative layers less  pronounced (Figure \ref{fig:delvsr}, bottom panel).
+We have verified using an extreme luminosity profile that our results reasonably approximate the atmosphere's structure interior to the outermost radiative layer (see \App{radwindow} for additional details).
+
+%By design, the luminosity in the radiative windows, $L_{\rm radw}$, must satisfy $L_{\rm radw}=L$, where $L$ is the assumed fixed luminosity at the top of the atmosphere.
+Non-negligible extra luminosity generated in the outer convective layers would, however, yield $L_{\rm radw}<L$, in conflict with our assumptions, and could change atmospheric structure.
+A simple way to check this \textit{a posteriori} is by rewriting the local energy equation (\ref{eq:structd}) as $\partial L/\partial m=-T \partial S/\partial t$, and integrating it throughout the atmosphere assuming that $\partial S/\partial t$ is fixed (e.g., \citealt{arras06}).
+If the value of this integral, $I$\footnote{This integral does not have units of luminosity, as we drop the constant $\partial S/\partial t$ term.}, throughout the innermost convective region is significantly larger than its value throughout the rest of the envelope, $\Delta I$, then our assumptions hold.
+For all the models for which $\Delta L \ll L$ holds (see paragraph above), we have found $\Delta I / I \lesssim 30\%$, and typically $\lesssim 10\%$ (see \App{radwindow} for additional details).
+
+\end{enumerate}
+
+% If radiative windows exist, our model requires most of the atmospheric luminosity to be generated in the innermost convective region.
+By design, the luminosity in the radiative windows, $L_{\rm radw}$, must satisfy $L_{\rm radw}=L$, where $L$ is the assumed fixed luminosity at the top of the atmosphere.
+Non-negligible extra luminosity generated in the outer layers would, however, yield $L_{\rm radw}<L$, in conflict with our assumptions, and could change atmospheric structure.
+A simple way to check this \textit{a posteriori} is by rewriting the local energy equation (\ref{eq:structd}) as $\partial L/\partial m=-T \partial S/\partial t$, and integrating it throughout the atmosphere assuming that $\partial S/\partial t$ is constant throughout the convective zone (e.g., \citealt{arras06}).
+If the value of this integral, $I$\footnote{This integral does not have units of luminosity, as we drop the constant $\partial S/\partial t$ term.}, throughout the innermost convective region is significantly larger than its value throughout the rest of the envelope, $\Delta I$, then our assumptions hold.
+For $p=3.5$ in the power-law (\ref{eq:graindistr}), we have found $\Delta I / I \lesssim 30\%$, and typically $\lesssim 10\%$ (see \App{radwindow} for additional details).
+For $p=2.5$, however, this approximation breaks down at low core masses, as we show in \S\ref{critical} --- specifically, the additional luminosity generated in either the radiative windows or the outer radiative layer is no longer non-negligible compared to the fixed atmospheric luminosity $L$.
+
+%For all our models of interest, we have found $\Delta I / I \lesssim 30\%$, and typically $\lesssim 10\%$ (see \App{radwindow} for additional details).
+
+%Additionally,
+
+%two challenges for our model, as follows:
+
+%\begin{enumerate}
+%
+% \item The additional luminosity, $\Delta L$, generated in these radiative windows may not satisfy $\Delta L \ll L$, where $L$ is the assumed fixed atmospheric luminosity.
+This would make our assumption of constant luminosity throughout radiative regions invalid.
+In practice, however, the inner radiative windows are either very narrow compared to the height of the convective regions (Figure \ref{fig:delvsr}, middle panel), or  have $\delad \approx \delrad$ throughout, which makes the distinction between convective and radiative layers less  pronounced (Figure \ref{fig:delvsr}, bottom panel).
+For $p=3.5$ in the power-law (\ref{eq:graindistr}), we have checked that $\Delta L \ll L$ in all the cases presented in this study.
+For $p=2.5$, however, this approximation breaks down at low core masses, as we show in \S\ref{critical}.
+%
+%%This results in a negligible extra luminosity generated in the radiative windows.
+However, the radiative windows may become non-negligible for other opacity choices and sufficiently low core masses, as we show later in this section.
+%
+%\item If radiative windows exist, our model requires most of the atmospheric luminosity to be generated in the innermost convective region.
+By design, the luminosity in the radiative windows, $L_{\rm radw}$, must satisfy $L_{\rm radw}=L$, where $L$ is the assumed fixed luminosity at the top of the atmosphere.
+Non-negligible extra luminosity generated in the outer convective layers would, however, yield $L_{\rm radw}<L$, in conflict with our assumptions, and could change atmospheric structure.
+A simple way to check this \textit{a posteriori} is by rewriting the local energy equation (\ref{eq:structd}) as $\partial L/\partial m=-T \partial S/\partial t$, and integrating it throughout the atmosphere assuming that $\partial S/\partial t$ is fixed (e.g., \citealt{arras06}).
+If the value of this integral, $I$\footnote{This integral does not have units of luminosity, as we drop the constant $\partial S/\partial t$ term.}, throughout the innermost convective region is significantly larger than its value throughout the rest of the envelope, $\Delta I$, then our assumptions hold.
+For all the models for which $\Delta L \ll L$ holds (see paragraph above), we have found $\Delta I / I \lesssim 30\%$, and typically $\lesssim 10\%$ (see \App{radwindow} for additional details).
+%
+%\end{enumerate}
+
+ %throughout the innermost convective layer and ensuring it is much larger than its value throughout the rest of the atmosphere. We have performed this check and found that
+
+%most of the atmospheric luminosity be generated in the innermost convective region of the envelope, since we
+
+%In practice, however, the inner radiative windows are either very thin (Figure \ref{fig:delvsr}, middle panel), or the temperature gradient is flat enough to result in a roughly constant luminosity throughout the region (Figure \ref{fig:delvsr}, bottom panel; also see Equation \ref{eq:structc}).
 This results in a negligible extra luminosity generated in the radiative windows.
@@
 To order of magnitude, Equation (\ref{eq:delrad}) can be expressed as
-\begin{equation}
-\label{eq:sigmatau4}
-\mathcal{F} \sim \frac{\sigma T\cb^4}{\tau},
-\end{equation}
-where $\mathcal{F}$ is the radiative flux through the RCB and $\tau \sim \kappa\cb \rho\cb H\cb$ is the optical depth of the first scale height of the radiative region, $H\cb \equiv k_{\rm b} T R^2/({\mu m_{\rm p}} G M)$, with all quantities evaluated at the RCB.
+%\begin{equation}
+%\label{eq:sigmatau4}
+%\mathcal{F} \sim \frac{\sigma T\cb^4}{\tau},
+%\end{equation}
+%where $\mathcal{F}$ is the radiative flux through the RCB and $\tau \sim \kappa\cb \rho\cb H\cb$ is the optical depth of the first scale height of the radiative region, $H\cb \equiv k_{\rm b} T R^2/({\mu m_{\rm p}} G M)$, with all quantities evaluated at the RCB.
 While Equation (\ref{eq:delrad}) is always true as part of the structure Equations set (\ref{eq:struct}), Equation (\ref{eq:sigmatau4}) shows that the atmospheric luminosity is set by the diffusive flux through the outermost RCB.
@@
 
+We thus find that our atmospheric and cooling model is valid in our regions of interest, with some exceptions discussed in \S\ref{critical}.
 Due to the variable number and position of radiative windows, and therefore radiative-convective boundaries, within the planet atmosphere, we cannot consistently calculate the time evolution of different atmospheres if we evaluate our cooling Equation (\ref{eq:coolingglobal}) at the RCB, as we do in our standard model.
 We choose to evaluate the cooling time at the Bondi radius instead (since our cooling model applies at any radius $R$, see section \S\ref{struct}).
-We note that our choice of $R$ does not change the estimate of the atmosphere evolution time, to order of magnitude, since the additional luminosity generated in all radiative regions is negligible for our opacity choice (also see Paper I).
+%We note that our choice of $R$ does not change the estimate of the atmosphere evolution time, to order of magnitude, since the additional luminosity generated in all radiative regions is negligible for our opacity choice (also see Paper I).
 
@@
 
-In this section we put together the results obtained in Sections \S\ref{EOSeffects}  and \S\ref{sec:opacity}, and determine the minimum core mass to initiate runaway gas accretion during the lifetime of the protoplanetary disk, i.e.
-the critical core mass.
-As in Paper I, we quantify the runaway accretion time $t_{\rm run}$ as the time at which the atmosphere growth time $M_{\rm atm}/\dot{M}$ drops to 10\% of its maximum value (see Paper I for details).
-We first explore the dependence of $t_{\rm run}$ on the core mass for a fixed semimajor axis.
+In this section we put together the results obtained in Sections \S\ref{EOSeffects}  and \S\ref{sec:opacity}, and determine the minimum core mass to initiate runaway gas accretion during the lifetime of the protoplanetary disk ($M_{\rm crit}$).
+As in Paper I, we quantify the runaway accretion time $t_{\rm run}$ as the time at which the atmosphere growth timescale $M_{\rm atm}/\dot{M}$ drops to 10\% of its maximum value (see Paper I for details).
+
+%We first explore the dependence of $t_{\rm run}$ on the core mass for a fixed semimajor axis.
 We then determine the critical core mass $M_{\rm crit}$ to form a giant planet from a gas composed of a realistic hydrogen-helium mixture, and we compare this with the results from Paper I for an ideal diatomic gas.
@@
 
-Figure \ref{fig:tvsMplot} displays the time evolution and the runaway growth time for atmospheres forming around cores with masses between 10 $M_{\oplus}$ and 20 $M_{\oplus}$ at $a=10$ AU in our fiducial disk.
+Figure \ref{fig:tvsMplot} displays the time evolution and the runaway growth time for atmospheres forming around cores with masses between 10 $M_{\oplus}$ and 20 $M_{\oplus}$ at $a=10$ AU in our fiducial disk, for a realistic EOS and standard ISM opacity.
 Higher mass cores have both shorter $t_{\rm run}$ and fractionally larger atmosphere masses at runaway, consistent with the results of Paper I.
@@
 %%\vspace{-0.5in}
-\caption{Elapsed time as a function of atmosphere mass, for cores with fixed masses between $10 M_{\oplus}$ and $20 M_{\oplus}$ at $a=10$ AU in our fiducial disk.
+\caption{Elapsed time as a function of atmosphere mass, for cores with fixed masses between $10 M_{\oplus}$ and $20 M_{\oplus}$ at $a=10$ AU in our fiducial disk, for a realistic EOS and standard ISM opacity.
 The circles mark the runaway growth time.
@@
 
-Figure \ref{fig:Mvsaplot} shows the critical core mass for a massive atmosphere to form during a typical lifetime of a protoplanetary disk $t=3$ Myrs (e.g., \citealt{jay99}), for a gas described by a realistic EOS and an ISM dust opacity given by Equation (\ref{eq:opacitylaw}) with $F_{\kappa}=1$ and $\beta=2$.
+Figure \ref{fig:Mvsaplot} shows the critical core mass for a massive atmosphere to form during a typical lifetime of a protoplanetary disk $t=3$ Myrs (e.g., \citealt{jay99}), for a gas described by a realistic EOS and an ISM dust opacity.
 The results of Paper I for an ideal diatomic gas are plotted for comparison.
 The inclusion of realistic EOS effects increases $M_{\rm crit}$ by more than a factor of two when compared to an ideal gas polytrope.
-As such, non-ideal effects substantially increase the core mass needed to form a giant planet  before the dissipation of the protoplanetary disk.
+%As such, non-ideal effects substantially increase the core mass needed to form a giant planet  before the dissipation of the protoplanetary disk.
 
@@
 %%\vspace{-0.5in}
-\caption{The minimum core mass for an atmosphere to initiate runaway gas accretion within the lifetime of a typical protoplanetary disk $t \sim 3$ Myrs as a function of semimajor axis, for a realistic hydrogen-helium mixture and a standard ISM power-law opacity.
+\caption{The minimum core mass for an atmosphere to initiate runaway gas accretion within the lifetime of a typical protoplanetary disk $t \sim 3$ Myrs as a function of semimajor axis, for a realistic hydrogen-helium mixture and a standard ISM opacity.
 The results of Paper I for an ideal diatomic gas are plotted for comparison.
@@
 %%\vspace{-0.5in}
-\caption{Critical core mass as a function of semimajor axis for radiative opacities that account for grain growth (green triangles, with $p=3.5$ and $a_{\rm max}=1$ cm; see text for details).
+\caption{Critical core mass as a function of semimajor axis for a realistic EOS and radiative opacities that account for grain growth (green triangles, with $p=3.5$ and $a_{\rm max}=1$ cm; see text for details).
 The critical core mass is lower than it would be if dust grains had an ISM-like size distribution (blue circles).}
@@
 
-Figure \ref{fig:Mcritvsagg} shows $M_{\rm crit}$ as a function of semimajor axis, for a grain growth opacity with a size distribution given by the power-law (\ref{eq:graindistr}) with $p=3.5$ and maximum particle size $a_{\rm max}=1$ cm.
-The critical core mass is lower than in the standard interstellar opacity case, and less sensitive to location in the disk, i.e.
-the boundary conditions (temperature and pressure).
-We have shown in Paper I that the critical core mass is highly dependent on disk temperature rather than pressure.
-The temperature dependence, however, is mainly due to opacity.
-For the simplified analytic model developed in Paper I we found that the crossover time $t_{\rm co} \sim T_{\rm d}^{\beta+1/2}$, with $\beta$ the power-law exponent in Equation (\ref{eq:opacitylaw}) and $t_{\rm co}$ a proxy for $t_{\rm run}$, defined as the time when $M_{\rm atm}=M\co$, as $t_{\rm run}$ cannot be determined self-consistently for the analytic model.
-Opacity is less sensitive to temperature variations for larger grains and has an almost flat profile (see Figure \ref{fig:opacity}), which results in $\beta \ll 1$ and a much weaker temperature (and therefore semimajor axis) dependence of the critical core mass, as seen in Figure \ref{fig:Mcritvsagg}.
-Moreover, grain growth reduces the absolute value of the opacity, which results in an overall lower runaway accretion time and critical core mass.
+Figure \ref{fig:Mcritvsagg} shows $M_{\rm crit}$ as a function of semimajor axis, for a realistic EOS and grain growth opacity with a size distribution given by Equation (\ref{eq:graindistr}) with $p=3.5$ and maximum particle size $s_{\rm max}=1$ cm.
+The critical core mass is lower than in the standard interstellar opacity case, and less sensitive to location in the disk.
+Location primarily affects the atmosphere through the opacity in the outer envelope, which depends on disk temperature.
+For the simplified analytic model developed in Paper I, we approximated $t_{\rm run}$ by  $t_{\rm co}$, the time when $M_{\rm atm}=M\co$, and found that $t_{\rm co} \sim T_{\rm d}^{\beta+1/2}$, with $\beta$ the power-law exponent in Equation (\ref{eq:opacitylaw}).
+Opacity is less sensitive to temperature variations for larger grains and has an almost flat profile (see Figure \ref{fig:opacity}), which results in $\beta \ll 1$ and a much weaker temperature (and therefore semimajor axis) dependence of $M_{\rm crit}$, as seen in Figure \ref{fig:Mcritvsagg}.
+Moreover, grain growth reduces the absolute value of the opacity, which also lowers $M_{\rm crit}$.
 % As noted in Section \S\ref{sec:opacity}, the critical core mass may be significantly lower if coagulation is taken into account, i.e.
@@
 %%\vspace{-0.5in}
-\caption{Runaway accretion times for different grain size distributions, for an atmosphere forming around a core with $M\co=4 M_{\oplus}$.
+\caption{Runaway accretion times for a realistic EOS and different grain size distributions, for an atmosphere forming around a core with $M\co=4 M_{\oplus}$.
 The lines marked by green and red triangles have grain growth opacities with $a_{\rm max}=1$ cm, and $p=3.5$ and $p=2.5$, respectively (see text for details).
@@
 
-In the grain size distribution (\ref{eq:graindistr}), the coefficient $p=3.5$ corresponds to a normal collisional cascade.
+In Equation (\ref{eq:graindistr}), the coefficient $p=3.5$ corresponds to a standard collisional cascade.
 If coagulation is taken into account, the exponent $p$ can be approximated as $p=2.5$ \citep{dalessio01}.
-This results in a flatter and significantly lower opacity, which may substantially reduce $M_{\rm crit}$.
+This results in a flatter and significantly lower opacity (see \App{radwindow}), which may substantially reduce $M_{\rm crit}$.
 However, we have found that our model breaks down for low core masses ($M\co \lesssim 3 M_{\oplus}$) under our assumption of constant luminosity in the outer radiative layer.
-Figure \ref{fig:p25p35} shows the crossover time $t_{\rm co}$ as a function of semimajor axis for the lowest core mass for which our model is valid, $M\co=4 M_{\oplus}$.
-The crossover time is more than one order of magnitude lower for $p=2.5$, which implies that $M_{\rm crit}$ may be, in fact, significantly lower than presented in Figure \ref{fig:Mcritvsagg}.
+Figure \ref{fig:p25p35} shows the runaway accretion time $t_{\rm run}$ as a function of semimajor axis for the lowest core mass for which our model is valid, $M\co=4 M_{\oplus}$.
+The runaway accretion time is more than one order of magnitude lower for $p=2.5$, which implies that $M_{\rm crit}$ may be, in fact, significantly lower than presented in Figure \ref{fig:Mcritvsagg}.
 In other words, grain growth can yield critical core masses up to an order of magnitude lower than in the case where interstellar opacities are used.
+
+In summary, we have found $M_{\rm crit}\sim 24 M_{\oplus}$ at 5 AU, steadily decreasing to $\sim$$6.5 M_{\oplus}$ at 100 AU, for a realistic EOS and interstellar opacity.
+For a grain growth opacity with a size distribution given by Equation (\ref{eq:graindistr}) with $p=3.5$, $M_{\rm crit}$ significantly drops to $\sim$$7 M_{\oplus}$ at 5 AU and $\sim$$4.5 M_{\oplus}$ at 100 AU.
+Accounting for coagulation (i.e., $p=2.5$) further reduces $M_{\rm crit}$ by up to one order of magnitude.
 
@@
 
+% In this scenario, the energy the envelope receives from planetesimals balances its luminosity.
+The core's atmosphere is in steady state, and  $M_{\rm crit}$ is uniquely determined for a fixed planetesimal accretion rate, $\dot{M_{\rm c}}$, and a set of disk conditions.
+For high $\dot{M_{\rm c}}$, the resulting $M_{\rm crit}$ needed to grow an atmosphere during the disk lifetime is too large, which has led to the belief that core accretion does not work at large separations.
+
 This study considers protoplanets with fully formed cores for which planetesimal accretion is negligible and KH contraction dominates the luminosity evolution of the atmosphere.
-This approach contrasts that of static models, which assume high planetesimal accretion rates and find that the atmosphere is in steady state and solely heated due to accretion of solids.
-As the envelope and core become comparable in mass, hydrostatic balance no longer holds and runaway gas accretion commences.
-In these models, $M_{\rm crit}$ is the maximum core mass for which the atmosphere is still in hydrostatic equilibrium.
-For a fixed planetesimal accretion rate and a set of disk conditions, $M_{\rm crit}$ is uniquely determined.
+This approach contrasts with that of models which assume high planetesimal accretion rates and find that the atmosphere is in steady state and solely heated due to accretion of solids.
+In both cases, as the envelope and core become comparable in mass, hydrostatic balance no longer holds and runaway gas accretion commences.
+For fast accretion, $M_{\rm crit}$ is uniquely defined as the maximum core mass for which the atmosphere is still in hydrostatic equilibrium for a fixed planetesimal accretion rate and a set of disk conditions.
 In this section we compare our results for $M_{\rm crit}$ to analogous results from steady-state fast planetesimal accretion calculations.
 We discuss the core accretion rates that are necessary for our regime to be valid in \S\ref{raf1}.
-In \S\ref{raf3}, we estimate core growth during atmosphere evolution at the maximum rate for which the KH regime is valid, and show it is indeed negligible.
+In \S\ref{raf3}, we estimate core growth at the maximum rate for which the KH regime is valid, and show it is negligible over the timescale on which the atmosphere evolves.
 Finally, we compare our results with those assuming fast planetesimal accretion in \S\ref{raf2}.
@@
 Figure \ref{fig:accrates} shows that $\dot{M}_{\rm c,KH}$ is $\sim2-3$ orders of magnitude lower than $\dot{M}_{\rm c, acc}$.
-Had the core accreted planetesimals at the $\dot{M}_{\rm c, KH}$ rate since it started forming, it could not have grown large enough to attract an atmosphere within typical disk lifetime.
-This is consistent with the results of \citet{pollack96}, which find that the planetesimal accretion rate decreases with time as the core's feeding zone is depleted, and with the requirement of our model that the planetesimal accretion rate is initially large during core growth, then significantly reduces as the gaseous envelope accumulates.
-This is a plausible scenario: the core could have formed through pebble accretion \citep{lambrechts12}, or in the inner part of the disk and was later scattered outwards by other giants in the system \citep{ida13}.
+Had the core accreted planetesimals at the $\dot{M}_{\rm c, KH}$ rate since it started forming, it could not have grown large enough to attract an atmosphere within a typical disk lifetime.
+Our model requires that the planetesimal accretion rate is initially large during core growth, then significantly reduces as the gaseous envelope accumulates, as suggested by, e.g., \citet{pollack96}.
+This is a plausible scenario: the core's feeding zone may be depleted of solids if it is not refilled by radial drift of planetesimals through the nebula, or the core may form in the inner part of the disk and later be scattered outwards by other giants in the system \citep{ida13}.
 
@@
 The core accretion rate $\dot{M}_{\rm c, acc}$ (dashed green) necessary to grow the core on the timescale $t_{\rm run}$ is larger than $\dot{M}_{\rm c, KH}$ (dotted red), the maximum planetesimal accretion rate during KH contraction for which our regime is valid (see text).
-The $\dot{M}_{\rm c, acc}$ rate is also larger than the frequently used planetesimal accretion rate $\dot{M}_{\rm c, Hill}$ (dashed-dotted light blue) for which the random velocity of the planetesimals is given by the Hill velocity due to the core (see text).
-This motivates our requirement  that planetesimal accretion must have slowed down after core growth for our model to be valid.}
+The frequently used planetesimal accretion rate $\dot{M}_{\rm c, Hill}$ (dashed-dotted light blue) for which the random velocity of the planetesimals is given by the Hill velocity due to the core (see text), also exceeds $\dot{M}_{\rm c, KH}$.This motivates our requirement  that planetesimal accretion must have slowed down after core growth for our model to be valid.}
 
@@
 Planetesimal accretion during the KH contraction phase of atmosphere growth at a rate $\dot{M}_{\rm c}<\dot{M}_{\rm{c,KH}}$  cannot alter the core mass enough to affect the time evolution of the atmosphere.
-We can quantitatively estimate the increase in core mass as
+We can quantitatively estimate the maximum increase in core mass as
 
@@
 We compare our results for $M_{\rm crit}$ with those of studies that assume large planetesimal accretion rates.
-These studies find that the atmosphere is in steady state at all times and that static solutions only exist up to a maximum (critical) core mass.
-If the $M_{\rm crit}$ found by static studies were lower than the $M_{\rm crit}$ in our regime of negligible solids accretion, the atmosphere would undergo runaway gas accretion before KH contraction became dominant, and our regime would not be relevant.
+If the $M_{\rm crit}$ found by these studies were lower than the $M_{\rm crit}$ in our regime of negligible solids accretion, the atmosphere would undergo runaway gas accretion before KH contraction became dominant, and our regime would not be relevant.
 We show, however, that our model yields lower core masses than those found when fast planetesimal accretion is considered.
 
-In static studies, the critical core mass is larger for higher planetesimal accretion, as additional heating increases the core mass required for collapse.
-As such, if atmosphere collapse does not occur due to planetesimal accretion for the lowest value of $\dot{M}_{\rm c, KH}$ over the course of the atmosphere's growth (see, e.g., Figure \ref{fig:accrates}), then it can only occur in the KH dominated regime.
+The critical core mass is larger for higher planetesimal accretion, as additional heating increases the core mass required for collapse.
+As such, if atmosphere collapse does not occur for the lowest value of $\dot{M}_{\rm c, KH}$ over the course of the atmosphere's growth, then it can only occur in the KH dominated regime.
 %In what follows we calculate the critical core mass $M_{\rm crit, KH}$ corresponding to $\min(\dot{M}_{\rm c, KH})$ and show that it is higher that the critical core mass we determined in the KH dominated regime.
@@
 %\vspace{-0.5in}
-\caption{Comparison between the critical core mass $M_{\rm{crit, KH}}$ given significant planetesimal accretion and the critical core mass when gas contraction dominates.
+\caption{Comparison between the critical core mass $M_{\rm{crit, KH}}$ given significant planetesimal accretion and the critical core mass when gas contraction dominates, for a polytropic EOS and interstellar opacity.
 Our results yield lower core masses than in the fast planetesimal accretion case (e.g., \citealt{rafikov06}).
@@
 
-As additional heating due to planetesimals limits the atmosphere's ability to cool, our result represents a true minimum for the core mass needed to form a gas giant during the lifetime of the protoplanetary disk.
+%As additional heating due to planetesimals limits the atmosphere's ability to cool, our result represents a true minimum for the core mass needed to form a gas giant during the lifetime of the protoplanetary disk.
 
@@
 
- For a realistic EOS and grain growth opacity with maximum particle size $a_{\rm max}=1$ cm and a power-law size distribution (\ref{eq:graindistr}) with $p=3.5$, the critical core mass is $\sim$$7 M_{\oplus}$ at 5 AU in our fiducial disk and drops to $\sim$$4.5 M_{\oplus}$ at 100 AU.
-This result for the critical core mass $M_{\rm crit}$ is similar to the one obtained in Paper I for an ideal gas polytropic EOS and interstellar opacity.
+ For a realistic EOS and grain growth opacity with maximum particle size $s_{\rm max}=1$ cm and a power-law size distribution (\ref{eq:graindistr}) with $p=3.5$, the critical core mass is $\sim$$7 M_{\oplus}$ at 5 AU in our fiducial disk and drops to $\sim$$4.5 M_{\oplus}$ at 100 AU.
 The realistic EOS and grain growth opacity have two competing effects on $M_{\rm crit}$:
@@
 The critical core mass is less sensitive to the location in the disk when realistic opacities are used.
-If $p=2.5$, an approximation for coagulation,  $M_{\rm crit}$ further reduces.
+If $p=2.5$, an approximation for coagulation,  $M_{\rm crit}$ further reduces by up to an order of magnitude.
  \end{itemize}
@@
 Different assumptions about the grain size distribution (e.g., that take into account coagulation) may further reduce the critical core mass.
+ \vspace{0.08in}
 
@@
 
-\noindent where $Z_t$, $Z_r$, $Z_v$ are the partition functions associated with translation, rotation, and vibration, respectively.
-\footnote{We ignore electronic and nuclear excitation as they are only important at temperatures much higher than our regime of interest.} Our derivations follow \citet{kittel} %For hydrogen, electronic and nuclear excitation are only significant at temperatures higher than our region of interest ($\theta_e \approx 12000$ K and $\theta_n >> \theta_e$, where $\theta_e$ and $\theta_n$ are the characteristic temperatures for electronic and nuclear excitation, respectively).
+\noindent where $Z_t$, $Z_r$, $Z_v$ are the partition functions associated with translation, rotation, and vibration, respectively.\footnote{We ignore electronic and nuclear excitation as they are only important at temperatures much higher than our regime of interest.} Our derivations follow \citet{kittel} %For hydrogen, electronic and nuclear excitation are only significant at temperatures higher than our region of interest ($\theta_e \approx 12000$ K and $\theta_n >> \theta_e$, where $\theta_e$ and $\theta_n$ are the characteristic temperatures for electronic and nuclear excitation, respectively).
 As such, we will only take into account the translation, rotation and vibration of the hydrogen molecule:
@@
 \label{eq:cv}
-C_v=\Big(\frac{\partial U}{\partial T}\Big)_{V}.
+c_{\rm V}=\Big(\frac{\partial U}{\partial T}\Big)_{V}.
 \end{equation}
@@
 
-We evaluate it from the tabulated values for $S_T$ and $S_P$ determined above.
+We evaluate $\delad$ from the tabulated values for $S_T$ and $S_P$ determined above.
 Figure \ref{fig:deladH} shows a contour plot of $\delad$ as a function of temperature and pressure for the extended EOS table.
 Our extension is only valid for $T \lesssim 2000$ K, since it does not take into account hydrogen dissociation.
+While we account for vibrational motion for completeness, its contribution is negligible in the temperature regime we are interested in.
 We choose $T=1500$ K as a conservative temperature cutoff.
@@
 %%\vspace{-0.5in}
-\caption{Internal energy per unit mass and specific heat capacity associated with rotation for parahydrogen (dashed blue), orthohydrogen (dotted green), and the equilibrium mixture (solids red) as a function of temperature.
+\caption{Internal energy per unit mass and specific heat capacity associated with rotation for parahydrogen (dashed blue), orthohydrogen (dotted green), and the equilibrium mixture (solid red) as a function of temperature.
 After \citet{farkas35}, Figure 1.}
@@
 \citet{dalessio01} compute wavelength dependent opacities for a range of maximum particle sizes and different size distributions.
-Figure \ref{fig:opacity} shows the integrated Rosseland mean opacity for a maximum particle size of 1 cm and a power law size distribution $n \sim a^{-p}$, with $a$ the grain size and $p=3.5$ for a normal collisional cascade and $p=2.5$ when coagulation is taken into account.
+Figure \ref{fig:opacity} shows the integrated Rosseland mean opacity for a maximum particle size of 1 cm and a power law size distribution $n \sim s^{-p}$, with $s$ the grain size and $p=3.5$ for a standard collisional cascade and $p=2.5$ when coagulation is taken into account.
 We see in Figure \ref{fig:opacity} that this yields a mean opacity that is both lower and less sensitive to temperature, when compared to \citet{bell94} or \citet{semenov03}.
-However, observations of dust in protoplanetary disks have only been made at low temperatures (i.e., before dust sublimates).
-As we see in Figure \ref{fig:opacity}, the opacity dramatically decreases during dust sublimation for ISM grains, which can result in regions in the inner parts of planetary atmospheres where energy is transported radiatively, i.e.
-radiative windows.
+However, \citet{dalessio01} only computes opacities for temperatures less than the dust sublimation temperature, appropriate for current observations of dust in protoplanetary disks.
+As we see in Figure \ref{fig:opacity}, the opacity dramatically decreases during dust sublimation.
 We thus use the \citet{bell94} opacities for $T \gtrsim 1000$ K, ensuring that they smoothly match the \citet{dalessio01} opacities for lower temperatures.
@@
 \caption{Snapshots of the radiative and adiabatic gradient as a function of the radial coordinate, for planets with different core masses forming at various locations in the disk.
-The nebular gas is described by a realistic EOS.
+The nebular gas is described by a realistic EOS, with a standard collisional cascade size distribution ($p=3.5$).
 The sharp drop in opacity due to dust sublimation may generate one or more radiative windows.
@@
 
+%This is not a problem, however, if most atmospheric luminosity is generated in the innermost convective region of the envelope (see \S\ref{sec:opacity}).
+We can check this \textit{a posteriori} by using the local energy equation,
+%\begin{equation}
+%\label{eq:localen}
+%\frac{\partial L}{\partial m}=-T \frac{\partial S}{\partial t},
+%\end{equation}
+%assuming that $\partial S/\partial t$ is constant for a given atmospheric profile. Our model is then valid if $\Delta I \ll I$, where
+%\begin{equation}
+%\label{eq:I}
+%I = \int_{M_{\rm c}}^{M_{\rm{RCB_1}}} T d m
+%\end{equation}
+%and
+%\begin{equation}
+%\label{eq:I}
+%\Delta I = \int_{M_{\rm{RCB_1}}}^{M_{\rm p}}T d m,
+%\end{equation}
+%with $\rm{RCB_1}$ the innermost RCB and $M_{\rm p}$ the total planet mass.
+We find $\Delta I /I \lesssim 30\%$ for all our models of interest (see, however, \S\ref{sec:opacity} and \S\ref{critical} for exceptions).
+
+Radiative windows could, in principle, change atmospheric structure and thus affect evolution if most luminosity is not generated in the innermost convective region of the envelope.
+We can check whether this is true \textit{a posteriori} by using the local energy equation,
+\begin{equation}
+\label{eq:localen}
+\frac{\partial L}{\partial m}=-T \frac{\partial S}{\partial t},
+\end{equation}
+and integrating it between $M\co$ and $M_{\rm{RCB_1}}$, where $\rm{RCB_1}$ the innermost RCB. Typically, we have found this luminosity to be about half of the assumed fixed atmospheric luminosity $L$.
+
+Self-consistently calculating $L(r)$ is not feasible for our code.
+Instead, we note that since  $\partial S/\partial t$ is fixed in convective regions \citep{arras06}, the luminosity profile is more centrally concentrated than $L \propto m$, which can be implemented simply.
+We calculated example profiles using $L \propto m$ and found that though this luminosity scaling may move the outermost RCB closer to the top of the envelope, it does not substantially change the luminosity emerging at the top of the atmosphere or the time evolution.
+%We therefore expect our results to approximate those for the true $L(r)$.
+
+%assuming that $\partial S/\partial t$ is constant for a given atmospheric profile. Our model is then valid if $\Delta I \ll I$, where
+%\begin{equation}
+%\label{eq:I}
+%I = \int_{M_{\rm c}}^{M_{\rm{RCB_1}}} T d m
+%\end{equation}
+%and
+%\begin{equation}
+%\label{eq:I}
+%\Delta I = \int_{M_{\rm{RCB_1}}}^{M_{\rm p}}T d m,
+%\end{equation}
+%with $\rm{RCB_1}$ the innermost RCB and $M_{\rm p}$ the total planet mass.
+We find $\Delta I /I \lesssim 30\%$ for all our models of interest (see, however, \S\ref{sec:opacity} and \S\ref{critical} for exceptions).
+
+%If radiative windows exist, our model requires most of the atmospheric luminosity to be generated in the innermost convective region.
+By design, the luminosity in the radiative windows, $L_{\rm radw}$, must satisfy $L_{\rm radw}=L$, where $L$ is the assumed fixed luminosity at the top of the atmosphere.
+Non-negligible extra luminosity generated in the outer convective layers would, however, yield $L_{\rm radw}<L$, in conflict with our assumptions, and could change atmospheric structure.
+A simple way to check this \textit{a posteriori} is by rewriting the local energy equation (\ref{eq:structd}) as $\partial L/\partial m=-T \partial S/\partial t$, and integrating it throughout the atmosphere assuming that $\partial S/\partial t$ is fixed (e.g., \citealt{arras06}).
+If the value of this integral, $I$\footnote{This integral does not have units of luminosity, as we drop the constant $\partial S/\partial t$ term.}, throughout the innermost convective region is significantly larger than its value throughout the rest of the envelope, $\Delta I$, then our assumptions hold.
+For all the models for which $\Delta L \ll L$ holds (see paragraph above), we have found $\Delta I / I \lesssim 30\%$, and typically $\lesssim 10\%$ (see \App{radwindow} for additional details).
+
+%A more rigorous check would be comparing our models with constant $L$ throughout with example profiles for which $L \propto \int T dm$.
+In practice, this calculation is not feasible for our code.
+However, we can make a simple estimate of how a non-constant $L$ affects atmospheric structure by assuming $L \propto m$ throughout the atmosphere.
+Note that $L \propto \int T dm$ makes the luminosity more centrally concentrated than if $L \propto m$, so our simple estimate is conservative.
+We have found that $L \propto m$ may change atmospheric structure by moving the outermost RCB closer to the top of the envelope, but this does not change the atmospheric luminosity or time evolution.
+
+%As an additional check that the atmospheric structure and evolution are not changed due to the radiative windows, we
+
 \end{document}
